### 2020-09-25 Questions.Rmd
Updated, need to work on questions

---
title: "Questions regarding modeling"
author: "Document Author"
header-includes:
- \usepackage{amsmath}
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github

---

After reading the book from Krushke and going back and forth between the "time-varying-anova-example" file and the Stan model "time-varying" there are certain things that I still don't understand:

1. Krushke uses diagrams (like one below) to illustrate how models are built
```{r Fig 1., echo=FALSE, fig.cap="Krushke's diagram for two groups, metric response analysis", out.width='60%'}
knitr::include_graphics("Diagram Krushke.png")
```

From what I have understood,in the diagram the response is expected to come from a $t$ distribution where the three parameters $\nu,\mu$ and $\sigma$.
The examples I have seen on Krushke using Stan following something like this:

\begin{align}
y_n=\alpha+\beta*x_n+\epsilon_n \tag{1}\\
\end{align}

where $\epsilon_n \sim normal(0,\sigma)$ 
or \[y_n \sim normal(\alpha+\beta x_n,\sigma)$ $(2)\]

The parameters in Stan are then defined:

* alpha
* beta 
* sigma

And the model appears in the model block as:
$y\sim normal(alpha+beta*x,sigma)$

I know you have mentioned this before, but I am a little confused by how the argument is passed in the model in "time-varying.stan". The parameter $\beta$ is a matrix of 6 rows and two columns, and then is transformed into $\mu$, another matrix with row number equal to the total number of observations and 2 columns. And each $\mu$ is a combination of the basis functions of the spline and the $\beta$ parameter.

When the final model is specified,I don't fully understand how the model would look if specified as in $1$.

### 2020-09-25 time-varying-anova-example.R
Updated

@@ -56,7 +56,7 @@ dat_fit <- list(
     N = nrow(dat_sim), #length (total number of observations) is No. of subjects*timepoints*treatments
     n_time = n_time,
     n_treatment = n_treatment,
-    q = ncol(X_bs), #wiggliness of the response
+    q = ncol(X_bs), #number of basis functions for the spline
     time_idx = as.numeric(as.factor(dat_sim$Day)), #all the timepoints per group per treatment
     treatment_idx = as.numeric(as.factor(dat_sim$Group)),
     y = dat_sim$observation, #all the generated responses

### 2020-09-27 Questions.Rmd
Updated "Questions" file

@@
 
-After reading the book from Krushke and going back and forth between the "time-varying-anova-example" file and the Stan model "time-varying" there are certain things that I still don't understand:
+After reading the book from Krushke and going back and forth between the "time-varying-anova-example" file and the Stan model "time-varying" there are certain things that I have understood:
 
-1. Krushke uses diagrams (like one below) to illustrate how models are built
-```{r Fig 1., echo=FALSE, fig.cap="Krushke's diagram for two groups, metric response analysis", out.width='60%'}
-knitr::include_graphics("Diagram Krushke.png")
-```
+* The specification of prior distributions: I did not understand at the beginning why you chose a Cauchy distribution for the error, but upon reading and talking with you I understand it is just a possible distribution, it has large tails and therefore can accomodate outliers well.
 
-From what I have understood,in the diagram the response is expected to come from a $t$ distribution where the three parameters $\nu,\mu$ and $\sigma$.
-The examples I have seen on Krushke using Stan following something like this:
+* I also understand most of the operations that take place and specially for the simulate_data function.
+
+Things that I still don't understand:
+
+1. Krushke's book uses Stan model specification that goes something like this:
 
@@
 
-where $\epsilon_n \sim normal(0,\sigma)$ 
-or \[y_n \sim normal(\alpha+\beta x_n,\sigma)$ $(2)\]
+where $\epsilon_n \sim normal(0,\sigma)$. 
+Alternatively, the model can be specified as:
+\begin{align}
+y_n \sim normal(\alpha+\beta x_n,\sigma) \tag{2}\\
+\end{align}
 
@@
 
+I understand this is a less complex model than the one that we are implementing, but the correspondence between the model notation and the specification in Stan makes it easy to understand for people that don't know complex statistical notation (such as myself).
+
 I know you have mentioned this before, but I am a little confused by how the argument is passed in the model in "time-varying.stan".
-The parameter $\beta$ is a matrix of 6 rows and two columns, and then is transformed into $\mu$, another matrix with row number equal to the total number of observations and 2 columns.
-And each $\mu$ is a combination of the basis functions of the spline and the $\beta$ parameter.
+The parameter $\beta$ is a matrix of 2 rows and 6 columns, and then is transformed into $\mu$, by multiplying $\beta$ (2 rows,6 columns) by  matrix $X$ (6 rows, 5 columns) .In this way, $\mu$ is a combination of the basis functions of the spline and the $\beta$ parameter, with size ( 2 rows, 5 columns).
 
-When the final model is specified,I don't fully understand how the model would look if specified as in $1$.
+But I looked upon the basis spline functions that the script creates (X_bs) and while I think I understand what X_bs is(a matrix that contains a basis function for a spline model on each column), I don't understand the role of this matrix in the Bayesian model.
+Does this mean that we are using X_bs to fit the data?
+And if that is the case, setting a different seed then would cause to have a better or worse fit?
+
+When the final model is specified,I don't fully understand how the model would look if specified as in $(1)$.It is running within a loop, and if taken literally, the model is:
+
+\begin{align}
+y_i \sim normal(\mu(time_i,treatment_i)) \tag{3}\\
+\end{align}
+Does this mean the model specifies an interaction between the $time$ and $treatment$ factors?
+I think this is the most important thing that I don't understand, because when the model is run and the "mu_post" variable is called in the R script, I just seen 100 matrices each size (2,5) but don't understand how the model created the numbers in each cell.
+
+Thank you for your help, if you can provide some guidance on the questions I wrote here, that would be very helpful!

### 2020-09-30 Questions.Rmd
updated Questions

@@
 * The specification of prior distributions: I did not understand at the beginning why you chose a Cauchy distribution for the error, but upon reading and talking with you I understand it is just a possible distribution, it has large tails and therefore can accomodate outliers well.
+
+    * Right, there are many different prior choices.
+In general, the choice of prior is not that important.
+You can also check the posterior distribution and compare with the prior to explore how sensitive the model is to the prior choice.
 
@@
 
+* The model we are fitting is very similar.
+The primary difference is that our model can be expressed as $y_n \sim normal(\alpha+\beta f(x_n),\sigma)$ where $f(x_n)$ is a function of the $x_n$.
+In the model above, $f(x_n) = x_n$ is the identity function whereas in the `time-varying.stan` model we use the b-spline function $f(\cdot)$.
+
 I know you have mentioned this before, but I am a little confused by how the argument is passed in the model in "time-varying.stan".
@@
 And if that is the case, setting a different seed then would cause to have a better or worse fit?
+
+* The $X_{bs}$ is the transformation of $X$ by the function $f(X)$.
+There are lots of functions $f(\cdot)$ that are widely used (log transformations, square roots, polynomials etc.).
+For example, $y_n \sim normal(\alpha+\beta_1 x_n + \beta_2 x_n^2, \sigma)$ is a function $f(x_n) = x_n + x_n^2$.
+The function $f(\cdot)$ is deterministic which means that the random seed should not matter.
 
@@
 
+* The model defines $\mu$ as a matrix where each row represents a treatment and each column represents a time step.
+Thus, $\mu[2, 4]$ is the mean response for treatment 2 at during the 4th time period.
+The basis functions in $X_{bs}$ allow for the time-varying process to be different for each treatment (each row of $\beta$ are the parameters for a treatment).
+The $X_{bs} \beta$ model the correlation in the $y$ variable over time based on how close in time the observations are.
+
 Thank you for your help, if you can provide some guidance on the questions I wrote here, that would be very helpful!

### 2020-10-01 time-varying-anova-example.R
updated data folder

@@ -2,6 +2,7 @@ library(tidyverse)
 library(splines)
 library(rstan)
 library(rstanarm)
+library(patchwork)
 
 
 ## number of observations -- currently using 3
@@ -17,7 +18,9 @@ simulate_data <- function(dat, n = 10, sd = 5) {
     dat_sim <- dat %>%
         slice(rep(1:n(), each = n)) %>%
         group_by(Group, Day) %>%
-        mutate(observation = rnorm(n, StO2, sd)) %>%
+        mutate(observation = rnorm(n, StO2, sd),
+               observation_log = rlnorm(n, log(StO2), log(sd)),
+               observation_tobit = pmax(rnorm(n, StO2, sd), 0.0001)) %>%
         ungroup()
 
     return(dat_sim)
@@ -41,14 +44,28 @@ X_bs <- bs(unique(dat_sim$Day), df = df, intercept = TRUE) #generates matrix for
 matplot(unique(dat_sim$Day), X_bs, type = 'l')
 
 
-ggplot(dat_sim, aes(x = Day, y = observation, color = Group)) +
+p1 <- ggplot(dat_sim, aes(x = Day, y = observation, color = Group)) +
     geom_point() +
     ggtitle("Simulated time-varying responses") +
     scale_color_viridis_d(end = 0.75) +
     geom_line(aes(y = StO2), alpha = 0.75) +
     theme_bw()
 
+p2 <- ggplot(dat_sim, aes(x = Day, y = observation_log, color = Group)) +
+    geom_point() +
+    ggtitle("Simulated log-scale time-varying responses") +
+    scale_color_viridis_d(end = 0.75) +
+    geom_line(aes(y = StO2), alpha = 0.75) +
+    theme_bw()
 
+p3 <- ggplot(dat_sim, aes(x = Day, y = observation_tobit, color = Group)) +
+    geom_point() +
+    ggtitle("Simulated tobit-scale time-varying responses") +
+    scale_color_viridis_d(end = 0.75) +
+    geom_line(aes(y = StO2), alpha = 0.75) +
+    theme_bw()
+
+p1 / p2 / p3
 
 
 library(rstan)
@@ -102,6 +119,24 @@ dat_plot %>%
     geom_ribbon(aes(ymin = lower_50, ymax = upper_50, fill = treatment), alpha = 0.5, color = NA) +
[…]

### 2020-10-01 Manuscripts/Manuscript_AM_v1.Rmd
Created `Manuscripts` folder, started references and files

---
title: "Bayesian statistics for repeated measures"
author: "Ariel Mundo, John Tipton, Timothy Muldoon"
subtitle: Their application and use in biomedical research
output: 
    html_document: default
    word_document:
     reference_docx: docx_template.docx
    #pdf_document: default
bibliography: refs.bib
---
Background
==============================================================

Biomedical research often uses longitudinal studies to analyze tumor response [@vishwanath2009].

# References

### 2020-10-01 Manuscripts/Manuscript_AM_v1.tex
Created `Manuscripts` folder, started references and files

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Bayesian statistics for repeated measures},
  pdfauthor={Ariel Mundo, John Tipton, Timothy Muldoon},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\title{Bayesian statistics for repeated measures}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Their application and use in biomedical research}
\author{Ariel Mundo, John Tipton, Timothy Muldoon}
\date{}

\begin{document}
\maketitle

\hypertarget{background}{%
\section{Background}\label{background}}

Biomedical research often uses longitudinal studies to analyze tumor
response {[}@Vishwanath2009{]}

\end{document}

### 2020-10-01 Manuscripts/Manuscript_AM_v1.Rmd
Updated Manuscript V1

@@
 ---
+Paper outline
+
+Background/Introduction
+
+* Paragraph 1: Definition of a longitudinal study. .Statistical advantages of repeated measures studies over cross sectional:power and number of subjects (references).
+
+* Paragraph 2: Challenges presented by longitudinal studies.
+Missing observations, and correlation between measurements.
+Limitations that these items raise for the traditional ANOVA methods of analysis.
+
+* Paragraph 3: Bayesian statistics as an alternative approach.
+Advantages over ANOVA and what inference can be made from it.
+Argue that while it is not commonly used in the biomedical arena, this paper aims at showing the implementation in an amenable manner to analyze non-linear trends in data.
+
+***
+
+* Recap on repeated measures ANOVA and the requisites that it needs to work properly
+-Sphericity
+-Variance-covariance matrix
+-Explain the *true* meaning of a _p-value_ and why it is not what researchers commonly think it is
+
+* Gentle introduction to Bayesian statistics
+-How it works
+-What advantages it has over ANOVA
+-What a confidence interval means probabilistically
+
 Background
 ==============================================================
+
+1.
 

### 2020-10-15 Manuscripts/Manuscript_AM_v1.Rmd
Started writing intro

@@
     html_document: default
+    css: style.css
     word_document:
      reference_docx: docx_template.docx
+mainfont: Times New Roman
     #pdf_document: default
 bibliography: refs.bib
+biblio-style: apalike
+link-citations: yes
 ---
+
 Paper outline
@@
 
-1.
+The study of the temporal changes in a variable of interest is a question that has been analyzed extensively in biomedical research.
+Typically, measurements are taken across multiple timepoints on the same subject(s) within a group.
+Examples of this type of approach include clinical studies on cancer breast and neck cancer[@sio2016;@kamstra2015], tumor response[@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010].
+Whereas this type of study presents advantages over a cross-sectional study in the number of subjects required to achieve a certain statistical power, and in its ability
+[@chavalarias2016]
 
-Biomedical research often uses longitudinal studies to analyze tumor response [@vishwanath2009].
+[@vishwanath2009].
+
+### Longitudinal vs
 

### 2020-10-19 Manuscripts/Manuscript_AM_v1.Rmd
Updated manuscript

@@
 author: "Ariel Mundo, John Tipton, Timothy Muldoon"
-subtitle: Their application and use in biomedical research
+bibliography: refs.bib
 output:
+  pdf_document: default
   html_document: default
@@
 mainfont: Times New Roman
-    #pdf_document: default
-bibliography: refs.bib
+link-citations: yes
+subtitle: Their application and use in biomedical research
 biblio-style: apalike
-link-citations: yes
 ---
 
-Paper outline
+## Paper outline
+ <span style="color: red;">I have been thinking that this type of paper probably does not need a relatively long introduction as the one I did for $JBO$.
+Below are some of my ideas on how to tackle this paper, your comments would be appreciated.
+Please keep in mind that I have started to write the Background/Intro part, and just have ideas on how to go from there</span>
 
 Background/Introduction
+Background
+==============================================================
 
-* Paragraph 1: Definition of a longitudinal study. .Statistical advantages of repeated measures studies over cross sectional:power and number of subjects (references).
+The study of the temporal changes in a variable of interest (a longitudinal analysis) is a question that has been analyzed extensively in biomedical research.
+Typically, measurements are taken across multiple timepoints on the same subject(s) within a group.
+Examples of this type of approach include clinical studies on cancer breast and neck cancer[@sio2016;@kamstra2015], tumor response[@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010].
+Whereas this type of study presents advantages over a cross-sectional study in the number of subjects required to achieve a certain statistical power, its ability to provide a time-correlated view of the phenomenon of interest, and, the potential to explore multiple relationships between different variables, the statistical analysis of such data is more challenging and requires careful consideration.
+
+Biomedical researchers typically employ a $frequentist$ approach to analyze longitudinal data.
+Such type of analysis is a hypothesis test using the $analysis of variance over repeated measures$ (repeated measures ANOVA or rm-ANOVA).
+Although the usual approach, rm-ANOVA is subject to multiple conditions to be valid, some of which are not easily verifiable.
+Moreover, this approach restricts the inferences it can extract from a longitudinal study, particularly when the data does not follow a linear trend (Figure 1.
+with line and "wiggly plot"), due to the inherent nature of the model and its limitations .Examples of such type of data are found in studies of tumor response to radio/chemotherapy in multiple models and clinical settings [[@vishwanath2009]].
+Therefore, this study has three goals: a) Present in an amenable and practical manner the requisities  of a $frequentist$ approach over longitudinal data and how these limit the analysis b) present a different area of statistical analysis for longitudinal biomedical data $Bayesian Statistics$, which does not use *p-values*, and is less restrictive thereby allowing  a statical analysis that is based on the data itself.
+c)Iimplement b) over a set of simulated data that matches previously reported trends in longitudinal biomedical studies.
+With an emphasis on reproducibility by providing the code and dataset used, this will provide biomedical researchers a clear view of the advantages of Bayesian statistics for the analysis of longitudinal data.
 
@@
 
-Background
-==============================================================
-
-The study of the temporal changes in a variable of interest is a question that has been analyzed extensively in biomedical research.
-Typically, measurements are taken across multiple timepoints on the same subject(s) within a group.
-Examples of this type of approach include clinical studies on cancer breast and neck cancer[@sio2016;@kamstra2015], tumor response[@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010].
-Whereas this type of study presents advantages over a cross-sectional study in the number of subjects required to achieve a certain statistical power, and in its ability
 [@chavalarias2016]
 
-[@vishwanath2009].
-
-### Longitudinal vs
-
 # References

### 2020-10-19 Manuscripts/Manuscript_AM_v1.tex
Updated manuscript

@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
 \PassOptionsToPackage{hyphens}{url}
@@
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
+  \usepackage{textcomp} % provides euro and other symbols
+\else % if luatex or xelatex
   \usepackage{unicode-math}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+  \setmainfont[]{Times New Roman}
 \fi
-% Use upquote if available, for straight quotes in verbatim environments
+% use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
   pdfauthor={Ariel Mundo, John Tipton, Timothy Muldoon},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
+  pdfborder={0 0 0},
+  breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
 \usepackage[margin=1in]{geometry}
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{-2}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+% set default figure placement to htbp
 \makeatletter
@@
 \makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 
@@
 
-\hypertarget{background}{%
-\section{Background}\label{background}}
+\hypertarget{paper-outline}{%
+\subsection{Paper outline}\label{paper-outline}}
 
-Biomedical research often uses longitudinal studies to analyze tumor
-response {[}@Vishwanath2009{]}
+{I have been thinking that this type of paper probably does not need a
+relatively long introduction as the one I did for \(JBO\). Below are
+some of my ideas on how to tackle this paper, your comments would be
+appreciated. Please keep in mind that I have started to write the
+Background/Intro part, and just have ideas on how to go from there}
+
+Background/Introduction Background
+==============================================================
+
+The study of the temporal changes in a variable of interest (a
+longitudinal analysis) is a question that has been analyzed extensively
+in biomedical research. Typically, measurements are taken across
+multiple timepoints on the same subject(s) within a group. Examples of
+this type of approach include clinical studies on cancer breast and neck
+cancer(Sio et al. \protect\hyperlink{ref-sio2016}{2016}; Kamstra et al.
+\protect\hyperlink{ref-kamstra2015}{2015}), tumor response(Roblyer et
+al. \protect\hyperlink{ref-roblyer2011}{2011}; Tank et al.
+\protect\hyperlink{ref-tank2020}{2020}; Pavlov et al.
+\protect\hyperlink{ref-pavlov2018}{2018}; Demidov et al.
+\protect\hyperlink{ref-demidov2018}{2018}), antibody expression(Ritter
+et al. \protect\hyperlink{ref-ritter2001}{2001}; Roth et al.
+\protect\hyperlink{ref-roth2017}{2017}), and cell metabolism(Jones et
+al. \protect\hyperlink{ref-jones2018}{2018}; Skala et al.
+\protect\hyperlink{ref-skala2010}{2010}). Whereas this type of study
+presents advantages over a cross-sectional study in the number of
+subjects required to achieve a certain statistical power, its ability to
+provide a time-correlated view of the phenomenon of interest, and, the
+potential to explore multiple relationships between different variables,
+the statistical analysis of such data is more challenging and requires
+careful consideration.
+
+Biomedical researchers typically employ a \(frequentist\) approach to
+analyze longitudinal data. Such type of analysis is a hypothesis test
+using the \(analysis of variance over repeated measures\) (repeated
+measures ANOVA or rm-ANOVA). Although the usual approach, rm-ANOVA is
+subject to multiple conditions to be valid, some of which are not easily
+verifiable. Moreover, this approach restricts the inferences it can
+extract from a longitudinal study, particularly when the data does not
+follow a linear trend (Figure 1. with line and ``wiggly plot''), due to
+the inherent nature of the model and its limitations .Examples of such
+type of data are found in studies of tumor response to
+radio/chemotherapy in multiple models and clinical settings
+{[}(Vishwanath et al. \protect\hyperlink{ref-vishwanath2009}{2009}){]}.
+Therefore, this study has three goals: a) Present in an amenable and
+practical manner the requisities of a \(frequentist\) approach over
+longitudinal data and how these limit the analysis b) present a
+different area of statistical analysis for longitudinal biomedical data
+\(Bayesian Statistics\), which does not use \emph{p-values}, and is less
+restrictive thereby allowing a statical analysis that is based on the
+data itself. c)Iimplement b) over a set of simulated data that matches
+previously reported trends in longitudinal biomedical studies. With an
+emphasis on reproducibility by providing the code and dataset used, this
+will provide biomedical researchers a clear view of the advantages of
+Bayesian statistics for the analysis of longitudinal data.
+
+\begin{itemize}
+\item
+  Paragraph 2: Challenges presented by longitudinal studies. Missing
+  observations, and correlation between measurements. Limitations that
+  these items raise for the traditional ANOVA methods of analysis.
+\item
+  Paragraph 3: Bayesian statistics as an alternative approach.
+  Advantages over ANOVA and what inference can be made from it. Argue
+  that while it is not commonly used in the biomedical arena, this paper
+  aims at showing the implementation in an amenable manner to analyze
+  non-linear trends in data.
+\end{itemize}
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+\begin{itemize}
+\item
+  Recap on repeated measures ANOVA and the requisites that it needs to
+  work properly -Sphericity -Variance-covariance matrix -Explain the
+  \emph{true} meaning of a \emph{p-value} and why it is not what
+  researchers commonly think it is
+\item
+  Gentle introduction to Bayesian statistics -How it works -What
+  advantages it has over ANOVA -What a confidence interval means
+  probabilistically
+\end{itemize}
+
+(Chavalarias et al. \protect\hyperlink{ref-chavalarias2016}{2016})
+
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-chavalarias2016}{}%
+Chavalarias, David, Joshua David Wallach, Alvin Ho Ting Li, and John PA
+Ioannidis. 2016. ``Evolution of Reporting P Values in the Biomedical
+Literature, 1990-2015.'' \emph{Jama} 315 (11): 1141--8.
+
+\leavevmode\hypertarget{ref-demidov2018}{}%
+Demidov, Valentin, Azusa Maeda, Mitsuro Sugita, Victoria Madge,
+Siddharth Sadanand, Costel Flueraru, and I Alex Vitkin. 2018.
+``Preclinical Longitudinal Imaging of Tumor Microvascular
+Radiobiological Response with Functional Optical Coherence Tomography.''
+\emph{Scientific Reports} 8 (1): 1--12.
+
+\leavevmode\hypertarget{ref-jones2018}{}%
+Jones, Jake D, Hallie E Ramser, Alan E Woessner, and Kyle P Quinn. 2018.
+``In Vivo Multiphoton Microscopy Detects Longitudinal Metabolic Changes
+Associated with Delayed Skin Wound Healing.'' \emph{Communications
+Biology} 1 (1): 1--8.
+
+\leavevmode\hypertarget{ref-kamstra2015}{}%
+Kamstra, JI, PU Dijkstra, M Van Leeuwen, JLN Roodenburg, and JA
+Langendijk. 2015. ``Mouth Opening in Patients Irradiated for Head and
+Neck Cancer: A Prospective Repeated Measures Study.'' \emph{Oral
+Oncology} 51 (5): 548--55.
+
+\leavevmode\hypertarget{ref-pavlov2018}{}%
+Pavlov, Mikhail V, Tatiana I Kalganova, Yekaterina S Lyubimtseva,
+Vladimir I Plekhanov, German Yurievich Golubyatnikov, Olga Y Ilyinskaya,
+Anna G Orlova, et al. 2018. ``Multimodal Approach in Assessment of the
+Response of Breast Cancer to Neoadjuvant Chemotherapy.'' \emph{Journal
+of Biomedical Optics} 23 (9): 091410.
+
+\leavevmode\hypertarget{ref-ritter2001}{}%
+Ritter, Gerd, Leonard S Cohen, Clarence Williams, Elizabeth C Richards,
+Lloyd J Old, and Sydney Welt. 2001. ``Serological Analysis of Human
+Anti-Human Antibody Responses in Colon Cancer Patients Treated with
+Repeated Doses of Humanized Monoclonal Antibody A33.'' \emph{Cancer
+Research} 61 (18): 6851--9.
+
+\leavevmode\hypertarget{ref-roblyer2011}{}%
+Roblyer, Darren, Shigeto Ueda, Albert Cerussi, Wendy Tanamai, Amanda
+Durkin, Rita Mehta, David Hsiang, et al. 2011. ``Optical Imaging of
+Breast Cancer Oxyhemoglobin Flare Correlates with Neoadjuvant
+Chemotherapy Response One Day After Starting Treatment.''
+\emph{Proceedings of the National Academy of Sciences} 108 (35):
+14626--31.
+
+\leavevmode\hypertarget{ref-roth2017}{}%
+Roth, Eli M, Anne C Goldberg, Alberico L Catapano, Albert Torri, George
+D Yancopoulos, Neil Stahl, Aurélie Brunet, Guillaume Lecorps, and Helen
+M Colhoun. 2017. ``Antidrug Antibodies in Patients Treated with
+Alirocumab.''
+
+\leavevmode\hypertarget{ref-sio2016}{}%
+Sio, Terence T, Pamela J Atherton, Brandon J Birckhead, David J
+Schwartz, Jeff A Sloan, Drew K Seisler, James A Martenson, et al. 2016.
+``Repeated Measures Analyses of Dermatitis Symptom Evolution in Breast
+Cancer Patients Receiving Radiotherapy in a Phase 3 Randomized Trial of
+Mometasone Furoate Vs Placebo (N06c4 {[}Alliance{]}).'' \emph{Supportive
+Care in Cancer} 24 (9): 3847--55.
+
+\leavevmode\hypertarget{ref-skala2010}{}%
+Skala, Melissa C, Andrew Nicholas Fontanella, Lan Lan, Joseph A Izatt,
+and Mark W Dewhirst. 2010. ``Longitudinal Optical Imaging of Tumor
+Metabolism and Hemodynamics.'' \emph{Journal of Biomedical Optics} 15
+(1): 011112.
+
+\leavevmode\hypertarget{ref-tank2020}{}%
+Tank, Anup, Hannah M Peterson, Vivian Pera, Syeda Tabassum, Anais
+Leproux, Thomas O'Sullivan, Eric Jones, et al. 2020. ``Diffuse Optical
+Spectroscopic Imaging Reveals Distinct Early Breast Tumor Hemodynamic
+Responses to Metronomic and Maximum Tolerated Dose Regimens.''
+\emph{Breast Cancer Research} 22 (1): 1--10.
+
+\leavevmode\hypertarget{ref-vishwanath2009}{}%
+Vishwanath, Karthik, Hong Yuan, William T Barry, Mark W Dewhirst, and
+Nimmi Ramanujam. 2009. ``Using Optical Spectroscopy to Longitudinally
+Monitor Physiological Changes Within Solid Tumors.'' \emph{Neoplasia} 11
+(9): 889--900.
 

### 2020-10-19 Manuscripts/Manuscript_AM_v1.Rmd
Generated first pdf!

@@
   pdf_document: default
-  html_document: default
-  css: style.css
   word_document:
     reference_docx: docx_template.docx
+  css: style.css
+  html_document: default
 mainfont: Times New Roman
@@
 Biomedical researchers typically employ a $frequentist$ approach to analyze longitudinal data.
-Such type of analysis is a hypothesis test using the $analysis of variance over repeated measures$ (repeated measures ANOVA or rm-ANOVA).
+Such type of analysis is a hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
 Although the usual approach, rm-ANOVA is subject to multiple conditions to be valid, some of which are not easily verifiable.

### 2020-10-20 Manuscripts/Manuscript_AM_v1.Rmd
Working on plots

@@
 output:
-  pdf_document: default
+  html_document: default
   word_document:
@@
   css: style.css
-  html_document: default
+  pdf_document: default
 mainfont: Times New Roman
@@
 
-Background/Introduction
-Background
-==============================================================
+## Background
 
-The study of the temporal changes in a variable of interest (a longitudinal analysis) is a question that has been analyzed extensively in biomedical research.
-Typically, measurements are taken across multiple timepoints on the same subject(s) within a group.
-Examples of this type of approach include clinical studies on cancer breast and neck cancer[@sio2016;@kamstra2015], tumor response[@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010].
-Whereas this type of study presents advantages over a cross-sectional study in the number of subjects required to achieve a certain statistical power, its ability to provide a time-correlated view of the phenomenon of interest, and, the potential to explore multiple relationships between different variables, the statistical analysis of such data is more challenging and requires careful consideration.
+The study of the temporal changes in a variable of interest (a longitudinal analysis) presents advantages over a cross-sectional study as it reduces the number of subjects required to achieve a certain statistical power[@guo2013], it enables a time-correlated view of the phenomenon of interest, and allows to explore multiple relationships between different variables.
+Typically, measurements are taken across multiple timepoints on the same subjects within  different group(s).
+Examples of this type of approach in biomedical research include clinical studies on breast and neck cancer[@sio2016;@kamstra2015], tumor response[@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010].
 
-Biomedical researchers typically employ a $frequentist$ approach to analyze longitudinal data.
+Researchers typically employ a $frequentist$ approach to analyze longitudinal data.
 Such type of analysis is a hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-Although the usual approach, rm-ANOVA is subject to multiple conditions to be valid, some of which are not easily verifiable.
+Although the usual approach, rm-ANOVA assumes that the data fulfills certain requisites: Constant variance across measurements (which is frequently unjustified) and  complete observations from each subject [@schober2018;@gueorguieva2004].
 Moreover, this approach restricts the inferences it can extract from a longitudinal study, particularly when the data does not follow a linear trend (Figure 1.
-with line and "wiggly plot"), due to the inherent nature of the model and its limitations .Examples of such type of data are found in studies of tumor response to radio/chemotherapy in multiple models and clinical settings [[@vishwanath2009]].
-Therefore, this study has three goals: a) Present in an amenable and practical manner the requisities  of a $frequentist$ approach over longitudinal data and how these limit the analysis b) present a different area of statistical analysis for longitudinal biomedical data $Bayesian Statistics$, which does not use *p-values*, and is less restrictive thereby allowing  a statical analysis that is based on the data itself.
+with line and "wiggly plot"), due to the inherent nature of the model and its limitations .Examples of such type of data are found in studies of tumor response to radio/chemotherapy in multiple models and clinical settings [@vishwanath2009].
+Therefore, this study has three goals: a) Present in an amenable and practical manner the requisites  of a $frequentist$ approach over longitudinal data and how these limit the analysis b) present a different area of statistical analysis for longitudinal biomedical data _Bayesian Statistics_, which does not use *p-values*, and is less restrictive thereby allowing  a statical analysis that is based on the data itself.
 c)Iimplement b) over a set of simulated data that matches previously reported trends in longitudinal biomedical studies.
 With an emphasis on reproducibility by providing the code and dataset used, this will provide biomedical researchers a clear view of the advantages of Bayesian statistics for the analysis of longitudinal data.
+
+```{r,FIGURE 1}
+rm(list=ls())
+library(ggplot2)
+library(splines)
+dt<-seq(1,6,0.1)
+y<-2*dt+1
+z<-rnorm(y,mean=y,sd=3)
+q<-0.01*dt^6-0.2*dt^5+0.2*dt^4+1
+zq<-rnorm(q,mean=q,sd=10)
+ab<-as.data.frame(cbind(dt,z,zq))
+
+x<-seq(0.6,500)
+knots <- c(0.25, 0.5, 0.75)
+theta <- c(0.6, 0.1, 0.5, 0.2, 0.8, 0.3)
+basis<-bs(x,knots=3,degree=1)
+y.spline<-basis %*% theta
+
+ggplot(data=ab,aes(x=dt,y=z))+geom_point()
+ggplot(data=ab,aes(x=dt,y=zq))+geom_point()
+
+```
 

### 2020-10-20 Manuscripts/Manuscript_AM_v1.Rmd
Udpated

@@
 
-x<-seq(0.6,500)
-knots <- c(0.25, 0.5, 0.75)
-theta <- c(0.6, 0.1, 0.5, 0.2, 0.8, 0.3)
-basis<-bs(x,knots=3,degree=1)
-y.spline<-basis %*% theta
-
-ggplot(data=ab,aes(x=dt,y=z))+geom_point()
-ggplot(data=ab,aes(x=dt,y=zq))+geom_point()
+# x<-seq(0.6,500)
+# knots <- c(0.25, 0.5, 0.75)
+# theta <- c(0.6, 0.1, 0.5, 0.2, 0.8, 0.3)
+# basis<-bs(x,knots=3,degree=1)
+# y.spline<-basis %*% theta
+# 
+# 
+# ggplot(data=ab,aes(x=dt,y=z))+geom_point()
+# ggplot(data=ab,aes(x=dt,y=zq))+geom_point()
 

### 2020-10-21 Manuscripts/Manuscript_AM_v1.Rmd
Updated

@@
 
-The study of the temporal changes in a variable of interest (a longitudinal analysis) presents advantages over a cross-sectional study as it reduces the number of subjects required to achieve a certain statistical power[@guo2013], it enables a time-correlated view of the phenomenon of interest, and allows to explore multiple relationships between different variables.
-Typically, measurements are taken across multiple timepoints on the same subjects within  different group(s).
-Examples of this type of approach in biomedical research include clinical studies on breast and neck cancer[@sio2016;@kamstra2015], tumor response[@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010].
+A longitudinal study is defined as that were a variable of interest is measured repeatedly in a group (or groups) of subjects.
+In biomedical research, this type of study is preferred when the intention is to observe the evolution of the effect of treatment (or treatments) across time, rather than analyzing the information at a single timepoint (a cross-sectional study).
+Clinical examples of this  approach in biomedical research include studies on breast and neck cancer[@sio2016;@kamstra2015], where in the first case  weekly measurements of skin toxicities in breast cancer patients with radiation-induced dermatitis were taken for up to 8 weeks, and mouth opening in head and neck cancer patients was measured at 6,12, 18, 24 and 36 months after radiotherapy (RT) in the latter.
+Similar examples are found in studies of tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010].
+From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach:it requires a lower number of subjects to reach a certain statistical power, and besides the previously mentioned time-effect evolution, it allows to determine the intra-variability of the response between different subjects [@guo2013; @fitzmaurice2012].
 
-Researchers typically employ a $frequentist$ approach to analyze longitudinal data.
-Such type of analysis is a hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-Although the usual approach, rm-ANOVA assumes that the data fulfills certain requisites: Constant variance across measurements (which is frequently unjustified) and  complete observations from each subject [@schober2018;@gueorguieva2004].
-Moreover, this approach restricts the inferences it can extract from a longitudinal study, particularly when the data does not follow a linear trend (Figure 1.
-with line and "wiggly plot"), due to the inherent nature of the model and its limitations .Examples of such type of data are found in studies of tumor response to radio/chemotherapy in multiple models and clinical settings [@vishwanath2009].
-Therefore, this study has three goals: a) Present in an amenable and practical manner the requisites  of a $frequentist$ approach over longitudinal data and how these limit the analysis b) present a different area of statistical analysis for longitudinal biomedical data _Bayesian Statistics_, which does not use *p-values*, and is less restrictive thereby allowing  a statical analysis that is based on the data itself.
+Researchers have typically employed a _frequentist_ approach to analyze longitudinal data.
+Such type of analysis is based on a hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
+However,a rm-ANOVA analysis not only assumes that the data fulfills certain requisites such as constant variance across measurements (which is frequently unjustified) and  complete observations from each subject [@schober2018;@gueorguieva2004], but it also requires the use of _post hoc_ analyses which inflate the false positivity rate[@liu2010].
+Recently, _linear mixed effects models_ (LMEMs) have started to be used by certain groups to analyze longitudinal data [@vishwanath2009;@skala2010].
+Generally speaking, LMEMs incorporate both _fixed_ and _random_ effects making them more flexible that rm-ANOVA as they can work with missing observations, and they allow to model the covariance of the parameters in different manners[@west2014].
+
+However, both LMEMs and rm-ANOVA restricts the inferences it can extract from a longitudinal study, particularly when the data does not follow a linear trend (Figure 1.
+with line and "wiggly plot"), due to the inherent nature of the model and its limitations .Examples of such type of data are found in studies of tumor response to radio/chemotherapy in multiple models and clinical settings [@vishwanath2009; @roblyer2011,@].
+Therefore, this study has three goals: a) Present in an amenable and practical manner the requisites  of a _frequentist_ approach over longitudinal data and how these limit the analysis b) present a different area of statistical analysis for longitudinal biomedical data _Bayesian Statistics_, which does not use *p-values*, and is less restrictive thereby allowing  a statical analysis that is based on the data itself.
 c)Iimplement b) over a set of simulated data that matches previously reported trends in longitudinal biomedical studies.

### 2020-10-22 Manuscripts/Manuscript_AM_v1.Rmd
Updated files for manuscript

@@
 ---
-title: "Bayesian statistics for repeated measures"
+title: "*Bayesian statistics for longitudinal studies in biomedical research*"
 author: "Ariel Mundo, John Tipton, Timothy Muldoon"
@@
 output:
-  html_document: default
+  pdf_document: default
   word_document:
@@
   css: style.css
-  pdf_document: default
+  html_document: default
 mainfont: Times New Roman
@@
 ## Paper outline
- <span style="color: red;">I have been thinking that this type of paper probably does not need a relatively long introduction as the one I did for $JBO$.
+ <span style="color: red;">I have been thinking that this type of paper probably does not need a relatively long introduction as the one I did for _JBO_.
 Below are some of my ideas on how to tackle this paper, your comments would be appreciated.
@@
 Such type of analysis is based on a hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-However,a rm-ANOVA analysis not only assumes that the data fulfills certain requisites such as constant variance across measurements (which is frequently unjustified) and  complete observations from each subject [@schober2018;@gueorguieva2004], but it also requires the use of _post hoc_ analyses which inflate the false positivity rate[@liu2010].
+However, a rm-ANOVA analysis not only assumes that the data fulfills certain requisites such as constant variance across measurements (which is frequently unjustified) and  complete observations from each subject [@schober2018;@gueorguieva2004], but it also requires the use of _post hoc_ analyses which inflate the false positivity rate[@liu2010].
 Recently, _linear mixed effects models_ (LMEMs) have started to be used by certain groups to analyze longitudinal data [@vishwanath2009;@skala2010].
@@
 
-However, both LMEMs and rm-ANOVA restricts the inferences it can extract from a longitudinal study, particularly when the data does not follow a linear trend (Figure 1.
-with line and "wiggly plot"), due to the inherent nature of the model and its limitations .Examples of such type of data are found in studies of tumor response to radio/chemotherapy in multiple models and clinical settings [@vishwanath2009; @roblyer2011,@].
-Therefore, this study has three goals: a) Present in an amenable and practical manner the requisites  of a _frequentist_ approach over longitudinal data and how these limit the analysis b) present a different area of statistical analysis for longitudinal biomedical data _Bayesian Statistics_, which does not use *p-values*, and is less restrictive thereby allowing  a statical analysis that is based on the data itself.
-c)Iimplement b) over a set of simulated data that matches previously reported trends in longitudinal biomedical studies.
+However, the nature of LMEMs and rm-ANOVA restrict the inferences they can extract from a longitudinal study when the data does not follow a linear trend (Figure 1.
+with line and "wiggly plot"), because the model in both cases does not allow for a consistent fit with the trend of the data.
+This particular non-linear behavior in longitudinal data has been reported in particular in studies that measure tumor response to radio/chemotherapy in preclinical and clinical settings  [@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994].
+In such circumstances, even if a _p-value_ with "significance" ( _p_<0.05) is obtained, the model lacks predictive power and this compromises the extent of the inferences that can be derived from the analysis.
+
+In contrast to the _frequentist_ rm-ANOVA and LMEM analysis, _Bayesian statistics_ represent a relatively new field that does not rely on _p-values_ and hypothesis test to analyze information.
+Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the comparisons and inferences derived form the analysis.On the other hand, the shift that Bayesian theory represents from the traditional statistical view in research and the set of computational tools required for the implementation of this type of models have limited their use in the biomedical research community.
+Based on this, the goals of this study are: a) to present the limitations of a _frequentist_ approach over longitudinal data, and demonstrate how these limitations in turn affect the results of the analysis b) introduce in a practical and amenable manner the theory of Bayesian statistics highlighting its applicability to biomedical research and c)Implement b) over a set of simulated data that matches previously reported trends in longitudinal biomedical studies.
 With an emphasis on reproducibility by providing the code and dataset used, this will provide biomedical researchers a clear view of the advantages of Bayesian statistics for the analysis of longitudinal data.
@@
 rm(list=ls())
+set.seed(123)
 library(ggplot2)

### 2020-10-22 Manuscripts/Manuscript_AM_v1.Rmd
Updated intro, refined title and author info

@@
 ---
-title: "*Bayesian statistics for longitudinal studies in biomedical research*"
-author: "Ariel Mundo, John Tipton, Timothy Muldoon"
+title: '**Bayesian statistics for longitudinal studies in biomedical research**'
+author: 
+- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
+- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
+- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
+
 bibliography: refs.bib
 output:
-  pdf_document: default
   word_document:
     reference_docx: docx_template.docx
+  pdf_document: default
   css: style.css
@@
 link-citations: yes
-subtitle: Their application and use in biomedical research
+subtitle: '_Their application and use in biomedical research_'
 biblio-style: apalike
@@
 ## Paper outline
- <span style="color: red;">I have been thinking that this type of paper probably does not need a relatively long introduction as the one I did for _JBO_.
-Below are some of my ideas on how to tackle this paper, your comments would be appreciated.
-Please keep in mind that I have started to write the Background/Intro part, and just have ideas on how to go from there</span>
+<span style="color: red;">\textcolor{red}{The paper Introduction has been updated, proposed sections appear at the end of the document as well as an initial graph.}</span>
 
@@
 Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the comparisons and inferences derived form the analysis.On the other hand, the shift that Bayesian theory represents from the traditional statistical view in research and the set of computational tools required for the implementation of this type of models have limited their use in the biomedical research community.
-Based on this, the goals of this study are: a) to present the limitations of a _frequentist_ approach over longitudinal data, and demonstrate how these limitations in turn affect the results of the analysis b) introduce in a practical and amenable manner the theory of Bayesian statistics highlighting its applicability to biomedical research and c)Implement b) over a set of simulated data that matches previously reported trends in longitudinal biomedical studies.
+Based on this, the goals of this study are: a) to present the limitations of a _frequentist_ approach (rm-ANOVA) over longitudinal data, and demonstrate how these limitations in turn affect the results of the analysis b) introduce in a practical and amenable manner the theory of Bayesian statistics highlighting its applicability to biomedical research and c)Implement b) over a set of simulated data that matches previously reported trends in longitudinal biomedical studies.
 With an emphasis on reproducibility by providing the code and dataset used, this will provide biomedical researchers a clear view of the advantages of Bayesian statistics for the analysis of longitudinal data.
 
-```{r,FIGURE 1}
+```{r,FIGURE 1,echo=FALSE, message=FALSE, warning=FALSE}
 rm(list=ls())
@@
 library(splines)
-dt<-seq(1,6,0.1)
-y<-2*dt+1
-z<-rnorm(y,mean=y,sd=3)
-q<-0.01*dt^6-0.2*dt^5+0.2*dt^4+1
-zq<-rnorm(q,mean=q,sd=10)
-ab<-as.data.frame(cbind(dt,z,zq))
+library(tidyverse)
+library(here)
 
-# x<-seq(0.6,500)
-# knots <- c(0.25, 0.5, 0.75)
-# theta <- c(0.6, 0.1, 0.5, 0.2, 0.8, 0.3)
-# basis<-bs(x,knots=3,degree=1)
-# y.spline<-basis %*% theta
-# 
-# 
-# ggplot(data=ab,aes(x=dt,y=z))+geom_point()
-# ggplot(data=ab,aes(x=dt,y=zq))+geom_point()
+linear<-read.csv(here("data","Simulated_linear.csv"))
+
+sim_linear <- function(dat, n = 10, sd = 5) {
+    linear_vals <- dat %>%
+        slice(rep(1:n(), each = n)) %>%
+        group_by(Group,Day) %>%
+        mutate(observation = rnorm(n, y, sd)) %>%
+        ungroup()
+
+    return(linear_vals)
+}
+n<-10
+sd<-2
+linear_vals <- sim_linear(linear, n, sd)
+linear_vals[] <- lapply(linear_vals, function(observation) ifelse(observation<=0, NA, observation)) #taking out values below 0
+
+ ggplot(linear_vals, aes(x = Day, y = observation, color = Group)) +
+    geom_point() +
+    ggtitle("Simulated linear time-varying responses") +
+    scale_color_viridis_d(end = 0.75) +
+    geom_line(aes(y = y), alpha = 0.75) +
+    theme_bw()
 
@@
 
-* Paragraph 2: Challenges presented by longitudinal studies.
-Missing observations, and correlation between measurements.
-Limitations that these items raise for the traditional ANOVA methods of analysis.
+* Section 1: Challenges presented by longitudinal studies: Missing observations, and correlation between measurements.
+How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)
 
-* Paragraph 3: Bayesian statistics as an alternative approach.
-Advantages over ANOVA and what inference can be made from it.
-Argue that while it is not commonly used in the biomedical arena, this paper aims at showing the implementation in an amenable manner to analyze non-linear trends in data.
+* Section 2: Bayesian statistics as an alternative approach.
+Gentle presentation of Bayes theorem _with a biomedical-related example(!)_ Advantages over ANOVA and how inference works.
+Argue that while it is not commonly used in the biomedical arena, it is a more accurate and flexible approach.
+
+* Section 3: Present the implementation of a spline-fitted model in R, using data simulated from [@vishwanath2009]
 
@@
 
-* Recap on repeated measures ANOVA and the requisites that it needs to work properly
--Sphericity
--Variance-covariance matrix
--Explain the *true* meaning of a _p-value_ and why it is not what researchers commonly think it is
-
-* Gentle introduction to Bayesian statistics
--How it works
--What advantages it has over ANOVA
--What a confidence interval means probabilistically
-
-[@chavalarias2016]
-
 # References

### 2020-10-22 Manuscripts/Manuscript_AM_v1.Rmd
Updated figure

@@
 output:
+  pdf_document: 
+    fig_caption: yes
+  html_document: default
+  css: style.css
   word_document:
     reference_docx: docx_template.docx
-  pdf_document: default
-  css: style.css
-  html_document: default
 mainfont: Times New Roman
@@
 
-```{r,FIGURE 1,echo=FALSE, message=FALSE, warning=FALSE}
+```{r,FIGURE 1,echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Simulated longitudinal data with a linear trend.'}
 rm(list=ls())

### 2020-10-27 Manuscripts/Manuscript_AM_v1.Rmd
Updated manuscript

@@
 
-A longitudinal study is defined as that were a variable of interest is measured repeatedly in a group (or groups) of subjects.
-In biomedical research, this type of study is preferred when the intention is to observe the evolution of the effect of treatment (or treatments) across time, rather than analyzing the information at a single timepoint (a cross-sectional study).
-Clinical examples of this  approach in biomedical research include studies on breast and neck cancer[@sio2016;@kamstra2015], where in the first case  weekly measurements of skin toxicities in breast cancer patients with radiation-induced dermatitis were taken for up to 8 weeks, and mouth opening in head and neck cancer patients was measured at 6,12, 18, 24 and 36 months after radiotherapy (RT) in the latter.
-Similar examples are found in studies of tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010].
-From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach:it requires a lower number of subjects to reach a certain statistical power, and besides the previously mentioned time-effect evolution, it allows to determine the intra-variability of the response between different subjects [@guo2013; @fitzmaurice2012].
+A longitudinal study is defined as one where a variable of interest is measured repeatedly in a group (or groups) of subjects.
+In biomedical research, this type of study is preferred when the intention is to observe the evolution of the effect of treatment across time, rather than analyzing the information at a single timepoint (a cross-sectional study).
+Clinical examples of this  approach in biomedical research include studies on breast and neck cancer[@sio2016;@kamstra2015]; in the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening measured was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT).
+Longitudinal studies have used also to measure tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010].
+From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach:it requires a lower number of subjects to reach a certain statistical power, and besides it being able to track the previously mentioned time-effect evolution on a group-by-group basis, it allows to determine the variability of the response within subjects [@guo2013; @fitzmaurice2012].
+In other words,a longitudinal study permits to quantify how the variable changes within each subject across time.
 
 Researchers have typically employed a _frequentist_ approach to analyze longitudinal data.
-Such type of analysis is based on a hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-However, a rm-ANOVA analysis not only assumes that the data fulfills certain requisites such as constant variance across measurements (which is frequently unjustified) and  complete observations from each subject [@schober2018;@gueorguieva2004], but it also requires the use of _post hoc_ analyses which inflate the false positivity rate[@liu2010].
+Such statistical view derives its name from the fact that it views probability as a limiting frequency [wagenmakers2008].
+In practice, this analysis for longitudinal data is  based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
+However, rm-ANOVA makes two key assumptions regarding the data: constant correlation across same-subject measurements and  complete observations from each subject [@schober2018;@gueorguieva2004].
+The first condition is frequently unjustified in many cases as  the correlation between measures diminishes when the time gap between them increases [@ugrinowitsch2004], and in that case violating that assumption increases the type I error rate[@lane2016].
+The reasons behind the second assumption being not met are varied: subjects can withdraw during the course of the study, attrition in animals due to injury or weight loss, or complications can arise that prevent the researcher from collecting measurements at a required timepoint.
+When incomplete observations  exist, rm-ANOVA requires to exclude the subject with missing observations from the analysis, which can result in increased costs for the study and possibly bias the results.
+
+Moreover, rm-ANOVA uses a _post hoc_ analysis to distinguish the differences between groups.
+Because a _post hoc_ analysis is based in multiple repeated comparisons, it can inflate the false positivity rate[@liu2010].
 Recently, _linear mixed effects models_ (LMEMs) have started to be used by certain groups to analyze longitudinal data [@vishwanath2009;@skala2010].
@@
 
-* Section 1: Challenges presented by longitudinal studies: Missing observations, and correlation between measurements.
+* Section 1:
+
+###Challenges presented by longitudinal studies:
+
+Missing observations, and correlation between measurements.
 How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)

### 2020-10-28 Manuscripts/Manuscript_AM_v1.Rmd
Updated intro check code

@@
 output:
+  html_document: default
   pdf_document: 
     fig_caption: yes
-  html_document: default
   css: style.css
@@
 
-Researchers have typically employed a _frequentist_ approach to analyze longitudinal data.
-Such statistical view derives its name from the fact that it views probability as a limiting frequency [wagenmakers2008].
-In practice, this analysis for longitudinal data is  based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-However, rm-ANOVA makes two key assumptions regarding the data: constant correlation across same-subject measurements and  complete observations from each subject [@schober2018;@gueorguieva2004].
-The first condition is frequently unjustified in many cases as  the correlation between measures diminishes when the time gap between them increases [@ugrinowitsch2004], and in that case violating that assumption increases the type I error rate[@lane2016].
-The reasons behind the second assumption being not met are varied: subjects can withdraw during the course of the study, attrition in animals due to injury or weight loss, or complications can arise that prevent the researcher from collecting measurements at a required timepoint.
-When incomplete observations  exist, rm-ANOVA requires to exclude the subject with missing observations from the analysis, which can result in increased costs for the study and possibly bias the results.
+Researchers have typically employed a _frequentist_ approach to analyze longitudinal data, a statistical view that derives its name from the fact that it regards probability as a limiting frequency [wagenmakers2008].
+when applied to longitudinal, a frequentist analysis is  based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
+This methodology makes two key assumptions regarding longitudinal data: a constant correlation across same-subject measurements and  complete observations from each subject [@schober2018;@gueorguieva2004].
+The first condition is frequently unjustified  as  the correlation between measures diminishes when the time gap between them increases [@ugrinowitsch2004], and in that case violating that assumption increases the type I error rate[@lane2016].
+The reasons behind the second assumption being not met in biomedical research are varied: subjects can withdraw during the course of the study, attrition in animals due to injury or weight loss, or complications can arise that prevent the researcher from collecting measurements at a certain timepoint.
+When these issues arise, rm-ANOVA requires to exclude the subject with missing observations from the analysis, which can lead to increased costs for the study if the desired statistical power is not met with the remaining subjects, and can lead to the exclusion of valuable information.Additionaly, rm-ANOVA uses a _post hoc_ analysis to assess significance in the differences between groups.
+Because a _post hoc_ analysis is based in multiple repeated comparisons, it can inflate the false positivity rate[@liu2010].
 
-Moreover, rm-ANOVA uses a _post hoc_ analysis to distinguish the differences between groups.
-Because a _post hoc_ analysis is based in multiple repeated comparisons, it can inflate the false positivity rate[@liu2010].
-Recently, _linear mixed effects models_ (LMEMs) have started to be used by certain groups to analyze longitudinal data [@vishwanath2009;@skala2010].
-Generally speaking, LMEMs incorporate both _fixed_ and _random_ effects making them more flexible that rm-ANOVA as they can work with missing observations, and they allow to model the covariance of the parameters in different manners[@west2014].
+Recently, _linear mixed effects models_ (LMEMs) have been used by certain groups to analyze biomedical longitudinal data [@vishwanath2009;@skala2010].
+Briefly, these models incorporate _fixed_ effects, which correspond to the levels of experimental factors in the study (e.g.
+the different drug regimens in a clinical trial), and _random_ effects, which account for random variation within the population [@pinheiro2006].
+These models are more flexible than rm-ANOVA as they can accomodate missing observations in the data, and  allow to model the covariance of the parameters in different manners[@west2014].
+On the other hand, they impose restrictions in the distribution of the errors and of the random effects, and assume linearity between the response and the parameters[@schielzeth2020;@pinheiro2006].
 
-However, the nature of LMEMs and rm-ANOVA restrict the inferences they can extract from a longitudinal study when the data does not follow a linear trend (Figure 1.
-with line and "wiggly plot"), because the model in both cases does not allow for a consistent fit with the trend of the data.
-This particular non-linear behavior in longitudinal data has been reported in particular in studies that measure tumor response to radio/chemotherapy in preclinical and clinical settings  [@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994].
-In such circumstances, even if a _p-value_ with "significance" ( _p_<0.05) is obtained, the model lacks predictive power and this compromises the extent of the inferences that can be derived from the analysis.
+Additionaly, the inherent nature of both LMEMs and rm-ANOVA restrict the inferences they can extract from a longitudinal study when the data does not follow a linear trend, because the model in both cases does not provide consistent fit with the trend of the data.
+This particular non-linear behavior is exemplified in longitudinal studies that measure tumor response to radio/chemotherapy in preclinical and clinical settings  [@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994].
+In such circumstances, it is even possible to obtain a "significant" _p-value_( _p_<0.05) due to the well-known variability of this metric [@nuzzo2014;@halsey2019], but the model will lack predictive power and this in turn will compromise the extent of the inferences that can be derived.
 
-In contrast to the _frequentist_ rm-ANOVA and LMEM analysis, _Bayesian statistics_ represent a relatively new field that does not rely on _p-values_ and hypothesis test to analyze information.
-Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the comparisons and inferences derived form the analysis.On the other hand, the shift that Bayesian theory represents from the traditional statistical view in research and the set of computational tools required for the implementation of this type of models have limited their use in the biomedical research community.
+In contrast to rm-ANOVA and LMEMs, _Bayesian statistics_ represent a relatively new field that does not rely on _p-values_ and hypothesis tests to analyze information.
+Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the comparisons and inferences derived form the analysis.
+On the other hand, the shift that Bayesian theory represents from the traditional statistical view in research and the set of computational tools required for the implementation of this type of models have limited their use in the biomedical research community.
 Based on this, the goals of this study are: a) to present the limitations of a _frequentist_ approach (rm-ANOVA) over longitudinal data, and demonstrate how these limitations in turn affect the results of the analysis b) introduce in a practical and amenable manner the theory of Bayesian statistics highlighting its applicability to biomedical research and c)Implement b) over a set of simulated data that matches previously reported trends in longitudinal biomedical studies.

### 2020-10-28 Manuscripts/Manuscript_AM_v1.Rmd
Updated

@@
 output:
-  html_document: default
+  word_document:
+    reference_docx: docx_template.docx
   pdf_document: 
@@
   css: style.css
-  word_document:
-    reference_docx: docx_template.docx
+  html_document: default
 mainfont: Times New Roman
@@
 
-Additionaly, the inherent nature of both LMEMs and rm-ANOVA restrict the inferences they can extract from a longitudinal study when the data does not follow a linear trend, because the model in both cases does not provide consistent fit with the trend of the data.
-This particular non-linear behavior is exemplified in longitudinal studies that measure tumor response to radio/chemotherapy in preclinical and clinical settings  [@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994].
+This assumption of linearity in LMEMs is also present in rm-ANOVA, and therefore both models restrict the inferences they can extract from a longitudinal study when the data does not follow a linear trend.In both cases the model does not provide a consistent fit with the trend of the data.
 In such circumstances, it is even possible to obtain a "significant" _p-value_( _p_<0.05) due to the well-known variability of this metric [@nuzzo2014;@halsey2019], but the model will lack predictive power and this in turn will compromise the extent of the inferences that can be derived.
+The non-linear behavior in longitudinal data is exemplified in different biomedical studies, including response to radio/chemotherapy in preclinical and clinical settings  [@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994].
+Since both the _frequentist_ (rm-ANOVA) and LMEM approaches are limited in the analysis of non-linear longitudinal information, there is a need to use statistical tools that allow both a consistent fit with the data and that allow adequate inferences.
 
-In contrast to rm-ANOVA and LMEMs, _Bayesian statistics_ represent a relatively new field that does not rely on _p-values_ and hypothesis tests to analyze information.
+In this regard, _Bayesian statistics_ represent a relatively new field in statistics that does not rely on _p-values_ and hypothesis tests to analyze information.
 Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the comparisons and inferences derived form the analysis.
-On the other hand, the shift that Bayesian theory represents from the traditional statistical view in research and the set of computational tools required for the implementation of this type of models have limited their use in the biomedical research community.
-Based on this, the goals of this study are: a) to present the limitations of a _frequentist_ approach (rm-ANOVA) over longitudinal data, and demonstrate how these limitations in turn affect the results of the analysis b) introduce in a practical and amenable manner the theory of Bayesian statistics highlighting its applicability to biomedical research and c)Implement b) over a set of simulated data that matches previously reported trends in longitudinal biomedical studies.
+On the other hand, the shift that Bayesian theory represents from the traditional _frequentist_ statistical view in research and the computational tools required for the implementation of this type of models have limited their use in the biomedical research community.
+However, the increasing recognition of the advantages of Bayesian statistics and the current development in computational tools are enabling researchers to transition to the field more rapidly.
+
+To this end, the goals of this study are: a) to present the limitations of (rm-ANOVA) and LMEMs over longitudinal data, and demonstrate how these limitations in turn affect the results of the analysis b) introduce in a practical and amenable manner the theory of Bayesian statistics highlighting its applicability to biomedical research and c)Implement a Bayesian analysis over a set of simulated data that matches previously reported trends in longitudinal biomedical studies.
 With an emphasis on reproducibility by providing the code and dataset used, this will provide biomedical researchers a clear view of the advantages of Bayesian statistics for the analysis of longitudinal data.
@@
 
-* Section 1:
+## Section 1:
 
-###Challenges presented by longitudinal studies:
+### Challenges presented by longitudinal studies:
 
+#### the _frequentist_ case for constant correlation
+
+One of the key assumptions for rm-ANOVA is the constant correlation among measurements.
+This frequently not the case as the correlation reduces as the time interval between two measurements increases [@ugrinowitsch2004].
+From a practical
+
+The first 
 Missing observations, and correlation between measurements.

### 2020-10-28 Manuscripts/Manuscript_AM_v1.Rmd
updated paper sections

@@
 
+- Why LMEMs are better than ANOVA
+
+- How LMEMs (using splines) and a Bayesian analysis can be used to analyze longitudinal data
+
 ```{r,FIGURE 1,echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Simulated longitudinal data with a linear trend.'}
@@
 
-#### the _frequentist_ case for constant correlation
+#### 1.1 the _frequentist_ case for constant correlation (ANOVA limitations)
+
+The first 
+Missing observations, and correlation between measurements.
+How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)
 
@@
 
-The first 
-Missing observations, and correlation between measurements.
-How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)
+#### 1.2 the case of LMEMs using splines and how they work and how they are better than rm-ANOVA
 
-* Section 2: Bayesian statistics as an alternative approach.
-Gentle presentation of Bayes theorem _with a biomedical-related example(!)_ Advantages over ANOVA and how inference works.
-Argue that while it is not commonly used in the biomedical arena, it is a more accurate and flexible approach.
+#### 1.3 Bayesian brief introduction, and compare the results of 1.2 to the results of Bayesian
 
-* Section 3: Present the implementation of a spline-fitted model in R, using data simulated from [@vishwanath2009]
+* Section 2: Implementation of both LMEMs and Bayesian and their results
+
+Present the implementation of a spline-fitted model in R, using data simulated from [@vishwanath2009]
 

### 2020-11-03 Manuscripts/Manuscript_AM_v1.Rmd
updated
Updated manuscript, need to work on first section

@@
 output:
-  word_document:
-    reference_docx: docx_template.docx
+  html_document: default
   pdf_document: 
@@
   css: style.css
-  html_document: default
+  word_document:
+    reference_docx: docx_template.docx
 mainfont: Times New Roman
@@
 the different drug regimens in a clinical trial), and _random_ effects, which account for random variation within the population [@pinheiro2006].
-These models are more flexible than rm-ANOVA as they can accomodate missing observations in the data, and  allow to model the covariance of the parameters in different manners[@west2014].
+These models are more flexible than rm-ANOVA as they can accommodate missing observations in the data, and  allow different modeling strategies for the covariance[@west2014].
 On the other hand, they impose restrictions in the distribution of the errors and of the random effects, and assume linearity between the response and the parameters[@schielzeth2020;@pinheiro2006].
 
-This assumption of linearity in LMEMs is also present in rm-ANOVA, and therefore both models restrict the inferences they can extract from a longitudinal study when the data does not follow a linear trend.In both cases the model does not provide a consistent fit with the trend of the data.
-In such circumstances, it is even possible to obtain a "significant" _p-value_( _p_<0.05) due to the well-known variability of this metric [@nuzzo2014;@halsey2019], but the model will lack predictive power and this in turn will compromise the extent of the inferences that can be derived.
-The non-linear behavior in longitudinal data is exemplified in different biomedical studies, including response to radio/chemotherapy in preclinical and clinical settings  [@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994].
-Since both the _frequentist_ (rm-ANOVA) and LMEM approaches are limited in the analysis of non-linear longitudinal information, there is a need to use statistical tools that allow both a consistent fit with the data and that allow adequate inferences.
+The assumption of linearity is a key component of LMEMs and rm-ANOVA, and therefore both models restrict the inferences they can extract from longitudinal data when it does not follow a linear trend.
+In biomedical research, such behavior in longitudinal data does arise and has been exemplified in different studies, including response to radio/chemotherapy in preclinical and clinical settings  [@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994].
+In such situations, an estimated model via rm-ANOVA or LMEMs may not provide a consistent fit with the observed variations in the variables of interest.
+Moreover, although it is possible to obtain a "significant" _p-value_( _p_<0.05) due to the well-known variability of this metric [@nuzzo2014;@halsey2019], the model will lack predictive power and this in turn will compromise the extent of the inferences that can be derived from it.
 
-In this regard, _Bayesian statistics_ represent a relatively new field in statistics that does not rely on _p-values_ and hypothesis tests to analyze information.
-Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the comparisons and inferences derived form the analysis.
-On the other hand, the shift that Bayesian theory represents from the traditional _frequentist_ statistical view in research and the computational tools required for the implementation of this type of models have limited their use in the biomedical research community.
-However, the increasing recognition of the advantages of Bayesian statistics and the current development in computational tools are enabling researchers to transition to the field more rapidly.
+Since both the _frequentist_ (rm-ANOVA) and LMEM approaches are limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to use statistical tools that allow the data to determine the fit of the model while enabling  inferences that are  both adequate and consistent from a statistical standpoint.
+
+Generalized additive models (GAMs) are a subset of generalized linear models that use _smooth functions_ (hencefort _smooths_) to estimate the parameters of a model.
+They have been used in palaeolimnology, ecology and clinical studies to model longitudinal data[@woolway2016;@hefley2017;@ko2007].
+Briefly, GAMs use a combination of multiple functions (basis functions)to construct the smooths of the model [@wood2017].
+Their main advantage over LMEMs and rm-ANOVA is that the model specification is directed by the _smooths_ rather than by a parametric relationship.
+This allows a consistent fit of the model with the data, and estimations of significance using the terms of the model.
+However, certain assumptions about the data are necessary: a normal distribution and constant variance of the residuals with the mean response.
+Therefore, GAMs provide a more suitable statistical method to analyze biomedical longitudinal data, when these assumptions are met.
+
+If the assumptions of GAMs do not hold,  the field of _Bayesian statistics_ represents a relatively new area of Statistics that does not rely on _p-values_ and hypothesis tests to analyze information.
+Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the number of comparisons and inferences derived form the analysis.
+On the other hand, the shift that Bayesian statistics represent from the traditional _frequentist_ statistical view in research, the computational tools required for its implementation, and the underlying mathematical theory have limited the use of this approach in the biomedical research community.
+However,Bayesian theory is intuitive and shares some principles with _frequentist_ statistics, and there is an increasing use and recognition of the advantages of their use across different areas of biomedical research such as clinical trial design and imaging[@biswas2009;@kelter2020;@kwon2020;@zhou2017].
+Additionaly,the current development in computational tools, specifically the programming language $\textsf{R}$, enable a rapid implementation of Bayesian models.
 
@@
 - How LMEMs (using splines) and a Bayesian analysis can be used to analyze longitudinal data
+
+https://www.frontiersin.org/articles/10.3389/fevo.2018.00149/full
 

### 2020-11-04 Manuscripts/Manuscript_AM_v1.Rmd
Updated abstract, section 1

@@
 However, certain assumptions about the data are necessary: a normal distribution and constant variance of the residuals with the mean response.
-Therefore, GAMs provide a more suitable statistical method to analyze biomedical longitudinal data, when these assumptions are met.
+Therefore, GAMs provide a more suitable statistical method to analyze biomedical longitudinal data, when these assumptions of the model are met by the data.
 
-If the assumptions of GAMs do not hold,  the field of _Bayesian statistics_ represents a relatively new area of Statistics that does not rely on _p-values_ and hypothesis tests to analyze information.
+However, it is possible that the assumptions of GAMs for longitudinal data do not hold under  certain circumstances.
+In that case, the field of _Bayesian statistics_ represents a relatively new area of Statistics that does not rely on _p-values_ and hypothesis tests to analyze information.
 Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the number of comparisons and inferences derived form the analysis.
@@
 However,Bayesian theory is intuitive and shares some principles with _frequentist_ statistics, and there is an increasing use and recognition of the advantages of their use across different areas of biomedical research such as clinical trial design and imaging[@biswas2009;@kelter2020;@kwon2020;@zhou2017].
-Additionaly,the current development in computational tools, specifically the programming language $\textsf{R}$, enable a rapid implementation of Bayesian models.
+Additionally,the current development in computational tools, specifically the programming language $\textsf{R}$, enable a rapid implementation of Bayesian models for longitudinal data.
 
-To this end, the goals of this study are: a) to present the limitations of (rm-ANOVA) and LMEMs over longitudinal data, and demonstrate how these limitations in turn affect the results of the analysis b) introduce in a practical and amenable manner the theory of Bayesian statistics highlighting its applicability to biomedical research and c)Implement a Bayesian analysis over a set of simulated data that matches previously reported trends in longitudinal biomedical studies.
-With an emphasis on reproducibility by providing the code and dataset used, this will provide biomedical researchers a clear view of the advantages of Bayesian statistics for the analysis of longitudinal data.
+Therefore, this study focuses in three areas in the analysis of longitudinal data from a biomedical perspective.
+First, it presents the limitations of (rm-ANOVA) and LMEMs over longitudinal data, and explains how these limitations in turn affect the results of the analysis.
+Secondly, it uses simulation to generate non-linear longitudinal data and presents the implementation of GAMs as a statistical tool that enables the analysis of data that shows this behavior.
+And finally, it introduces Bayesian statistics and presents their implementation with GAMs over simulated data to demonstrate the differences and benefits of this approach .
+With an emphasis on reproducibility by providing the code and dataset used, this study provides biomedical researchers with a clear view of the advantages of different approaches in the statistical analysis of longitudinal data.
 
@@
 
-#### 1.1 the _frequentist_ case for constant correlation (ANOVA limitations)
+### 1 The _frequentist_ case for longitudinal data
 
-The first 
+While the _repeated measures analysis of variance_ (rm-ANOVA) is the standard for the statistical analysis of longitudinal data, there are key assumptions that are made in order to make the model valid.
+From a practical view, they can be divided in three areas: linear relationship between covariates and response, constant correlation between measurements, and complete observations for all subjects.
+Each one of these assumptions is discussed below.
+
+#### 1.1 Linear relationship
+
+When a continuous and normally distributed response $y$ is measured from subject $n$, the basic model for rm-ANOVA becomes[@davis2002]:
+
+$$ y_{ij}=\mu_{ij}+\pi_{ij}+e_{ij}$$
+Missing observations, and correlation between measurements.
+How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)
+#### 1.1 Constant correlation
+
+While the _repeated measures analysis of variance_ (rm-ANOVA) is the norm in the biomedical community to analyze longitudinal data, two frequent situations that arise in longitudinal studies limit its applicability.
 Missing observations, and correlation between measurements.

### 2020-11-04 Manuscripts/Manuscript_AM_v1.Rmd
Updated section 1

@@
 
-When a continuous and normally distributed response $y$ is measured from subject $n$, the basic model for rm-ANOVA becomes[@davis2002]:
+In a biomedical longitudinal study, two or more groups of subjects (humans, mice, samples) are subject to a different treatments (e.g.
+group of mice receiving a novel drug vs.
+a group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
+Moreover, it is assumed that the collected response has two components: a _fixed_ and a _random_ component.
+The _fixed_ component can be understood as a constant value in the response which the researcher intends to measure, i.e, the effect of a novel drug in a subject.The _random_ component can be defined as "noise" caused by some some factors that are not of interest to the researcher, i.e., if the concentration of a drug is measured in some subjects within the same group in the early hours of the morning while others are measured in the afternoon, the researcher might consider this variability in the collection time of the measurement to introduce some "noise" in the signal.
+As their name suggests, this "random" variabliity needs to be modeled as a variable rather than as a constant value.
 
-$$ y_{ij}=\mu_{ij}+\pi_{ij}+e_{ij}$$
+Mathematically speaking, if a normally distributed response $y$ is measured repeatedly at $t$ time points from subjects in $p$ groups, where each group has a certain $n_p$ number of subjects, the the model for the response $y_{hij}$ becomes:
+
+$$ y_{hij}=\mu+\gamma_h+\tau_j+(\gamma \tau)_{hj}+\pi_{i(h)}+e_{hij}$$
+Where
+
+$i=1,...,n_h$ $j=1,...,t$, $h=1,...,p$; with $\pi_i{h} \sim N(0,\sigma^{2}_\epsilon)$ (independently normally distributed) and $e_{hij} \sim N(0,\sigma^{2}_\epsilon)$
+
+In this model, $\mu$ represents the group mean, $\gamma_h$ is the _fixed effect_ of group $h$, $\tau_j$ is the fixed effect of time $j$, and $(\gamma \tau)_{hj}$ represents the interaction of time and group effects.
+The term $\pi_{ij}$ represents the _random effects_ for each subject within each group.Finally, $e_{hij}$ represents the independent random error terms, which need to be normally distributed with mean 0 [@davis2002].
+
 Missing observations, and correlation between measurements.

### 2020-11-09 Manuscripts/Manuscript_AM_v1.Rmd
Updated
Updated background, references

@@
 
-A longitudinal study is defined as one where a variable of interest is measured repeatedly in a group (or groups) of subjects.
-In biomedical research, this type of study is preferred when the intention is to observe the evolution of the effect of treatment across time, rather than analyzing the information at a single timepoint (a cross-sectional study).
-Clinical examples of this  approach in biomedical research include studies on breast and neck cancer[@sio2016;@kamstra2015]; in the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening measured was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT).
+A longitudinal study is defined as that which is designed to repeatedly measure a variable of interest in a group (or groups) of subjects.
+In biomedical research, this type of study arises when the investigator intends to observe the evolution of the effect of a certain treatment across time, rather than analyzing it at a single time point (a cross-sectional study).
+Clinical examples of this  approach in biomedical research include studies on breast and neck cancer[@sio2016;@kamstra2015]; in the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT).
 Longitudinal studies have used also to measure tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010].
@@
 
-Researchers have typically employed a _frequentist_ approach to analyze longitudinal data, a statistical view that derives its name from the fact that it regards probability as a limiting frequency [wagenmakers2008].
-when applied to longitudinal, a frequentist analysis is  based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-This methodology makes two key assumptions regarding longitudinal data: a constant correlation across same-subject measurements and  complete observations from each subject [@schober2018;@gueorguieva2004].
-The first condition is frequently unjustified  as  the correlation between measures diminishes when the time gap between them increases [@ugrinowitsch2004], and in that case violating that assumption increases the type I error rate[@lane2016].
-The reasons behind the second assumption being not met in biomedical research are varied: subjects can withdraw during the course of the study, attrition in animals due to injury or weight loss, or complications can arise that prevent the researcher from collecting measurements at a certain timepoint.
-When these issues arise, rm-ANOVA requires to exclude the subject with missing observations from the analysis, which can lead to increased costs for the study if the desired statistical power is not met with the remaining subjects, and can lead to the exclusion of valuable information.Additionaly, rm-ANOVA uses a _post hoc_ analysis to assess significance in the differences between groups.
-Because a _post hoc_ analysis is based in multiple repeated comparisons, it can inflate the false positivity rate[@liu2010].
+Traditionally, a “frequentist” approach is used in biomedical research to derive inferences from the results of a longitudinal study.
+Such statistical view derives its name from the fact that it regards probability as a limiting frequency [wagenmakers2008] and its application is based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
+This methodology makes two key assumptions regarding longitudinal data: a constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study  [@schober2018;@gueorguieva2004].
 
-Recently, _linear mixed effects models_ (LMEMs) have been used by certain groups to analyze biomedical longitudinal data [@vishwanath2009;@skala2010].
-Briefly, these models incorporate _fixed_ effects, which correspond to the levels of experimental factors in the study (e.g.
-the different drug regimens in a clinical trial), and _random_ effects, which account for random variation within the population [@pinheiro2006].
-These models are more flexible than rm-ANOVA as they can accommodate missing observations in the data, and  allow different modeling strategies for the covariance[@west2014].
-On the other hand, they impose restrictions in the distribution of the errors and of the random effects, and assume linearity between the response and the parameters[@schielzeth2020;@pinheiro2006].
+However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases[@ugrinowitsch2004], and the violation of this assumption increases the false positivity rate [@lane2016].
+Moreover, it is unlikely that complete observations in all subjects are obtained in a biomedical study due to reasons that can be specific to different situations.
+In a clinical trial, voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and in both cases it is possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point.
 
-The assumption of linearity is a key component of LMEMs and rm-ANOVA, and therefore both models restrict the inferences they can extract from longitudinal data when it does not follow a linear trend.
-In biomedical research, such behavior in longitudinal data does arise and has been exemplified in different studies, including response to radio/chemotherapy in preclinical and clinical settings  [@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994].
-In such situations, an estimated model via rm-ANOVA or LMEMs may not provide a consistent fit with the observed variations in the variables of interest.
-Moreover, although it is possible to obtain a "significant" _p-value_( _p_<0.05) due to the well-known variability of this metric [@nuzzo2014;@halsey2019], the model will lack predictive power and this in turn will compromise the extent of the inferences that can be derived from it.
+When the aforementioned issues arise, rm-ANOVA requires to exclude all subjects with missing observations from the analysis, thereby increasing costs for the study if the desired statistical power is not met with the remaining observations as it makes necessary to enroll more subjects; and  raising the possibility that the rejection of those partial observations limits the demonstration of significant differences between groups.
+Additionally, rm-ANOVA uses a _post hoc_ analysis to assess the relevance between the measured response in different groups.
+This analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance.
+Because the p-value is highly variable [@halsey2015;@nuzzo2014], multiple comparisons can inflate the false positivity rate [@liu2010],consequently biasing the conclusions of the study.
 
-Since both the _frequentist_ (rm-ANOVA) and LMEM approaches are limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to use statistical tools that allow the data to determine the fit of the model while enabling  inferences that are  both adequate and consistent from a statistical standpoint.
+During the last decade, the biomedical community has started to recognize the limitations of a rm-ANOVA approach in the analysis of longitudinal information.
+This is exemplified by the pioneering use of  linear mixed effects models (LMEMs)  in certain groups to analyze tumor longitudinal data [@skala2010;@vishwanath2009].
+Briefly, these models incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g.
+the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population [@pinheiro2006].These models are more flexible than rm-ANOVA as they can accommodate missing observations for multiple subjects, and allow different modeling strategies for the variability within each measure in every subject [@west2014;@pinheiro2006].On the other hand, they impose restrictions in the distribution of the errors of the model and random effects [@gueorguieva2004;@schielzeth2020].
+
+Additionally, another assumption is made in both rm-ANOVA and LMEMs models, where a linear relationship is expected between the observed response and the covariates across the study [@pinheiro2006].This common assumption to both rm-ANOVA and LMEMs causes the models to be restrictive in their inferences when used in longitudinal data that does not follow a linear trend.
+In biomedical research, this particular behavior in longitudinal has been reported in studies of tumor response to radio/chemotherapy in preclinical and clinical settings[@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994].
+This studies have shown that that the collected signal does not follow a linear trend over time, and presents high variability at different time points, making the estimations of a LMEM or rm-ANOVA model inconsistent with the pattern of the observed variations.
+Additionally,  although it is possible that a post hoc analysis is able to find “significance”( _p-value_<0.05) by using multiple comparisons between the model terms, this estimator would be inherently biased because of the lack of fit between the information and the model.
+
+As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed distribution) to determine the fit of the model while enabling inferences that are both adequate and consistent from a statistical perspective.
 
@@
 Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the number of comparisons and inferences derived form the analysis.
-On the other hand, the shift that Bayesian statistics represent from the traditional _frequentist_ statistical view in research, the computational tools required for its implementation, and the underlying mathematical theory have limited the use of this approach in the biomedical research community.
-However,Bayesian theory is intuitive and shares some principles with _frequentist_ statistics, and there is an increasing use and recognition of the advantages of their use across different areas of biomedical research such as clinical trial design and imaging[@biswas2009;@kelter2020;@kwon2020;@zhou2017].
-Additionally,the current development in computational tools, specifically the programming language $\textsf{R}$, enable a rapid implementation of Bayesian models for longitudinal data.
+On the other hand, the shift that Bayesian statistics represent from the traditional "frequentist" statistical view in research, the computational tools required for its implementation, and the underlying mathematical theory have limited the use of this approach in the biomedical research community.
+However,Bayesian theory is intuitive and shares some principles with "frequentist" statistics, and there is an increasing use and recognition of the advantages of their use across different areas of biomedical research such as clinical trial design and imaging[@biswas2009;@kelter2020;@kwon2020;@zhou2017].
+
+Additionally,the current development in computational tools, specifically the programming language $\textsf{R}$, enable a rapid implementation of GAMs and Bayesian models for longitudinal data.In particular, $\textsf{R}$ has an extensive collection of documentation, functions and libraries that speed up the initial stages of the analsys and that enable the use of complex statistical methods without requiring a specialized set of programming skills from the user.
 
@@
 First, it presents the limitations of (rm-ANOVA) and LMEMs over longitudinal data, and explains how these limitations in turn affect the results of the analysis.
-Secondly, it uses simulation to generate non-linear longitudinal data and presents the implementation of GAMs as a statistical tool that enables the analysis of data that shows this behavior.
-And finally, it introduces Bayesian statistics and presents their implementation with GAMs over simulated data to demonstrate the differences and benefits of this approach .
-With an emphasis on reproducibility by providing the code and dataset used, this study provides biomedical researchers with a clear view of the advantages of different approaches in the statistical analysis of longitudinal data.
+Secondly, it uses $\textsf{R}$ to simulate non-linear longitudinal data following previously reported values in the literature, and presents the implementation of GAMs as a statistical tool for longitudinal data.
+And finally, it introduces Bayesian statistics and presents their implementation with GAMs to demonstrate the differences and benefits of this approach.
+with an emphasis in reproducibility  by providing the code, simulated dataset and a step-by-step guide of the computational implementation of different models, this study aims to encourage the exploration of modern statistical methods for biomedical longitudinal data, and to improve the  statistical standards in biomedical research.
 
@@
 
-### 1 The _frequentist_ case for longitudinal data
+### 1 The "frequentist" case for longitudinal data
 
-While the _repeated measures analysis of variance_ (rm-ANOVA) is the standard for the statistical analysis of longitudinal data, there are key assumptions that are made in order to make the model valid.
+The _repeated measures analysis of variance_ (rm-ANOVA) is the standard for the statistical analysis of longitudinal data, and there are key assumptions that are made in order to make the model valid.
 From a practical view, they can be divided in three areas: linear relationship between covariates and response, constant correlation between measurements, and complete observations for all subjects.
@@
 The term $\pi_{ij}$ represents the _random effects_ for each subject within each group.Finally, $e_{hij}$ represents the independent random error terms, which need to be normally distributed with mean 0 [@davis2002].
+The model then, is a linear combination of terms, and if plotted, it would a straight line.
+
+#### 1.2 Constant correlation
 

### 2020-11-09 Manuscripts/Manuscript_AM_v1.Rmd
Working on section 1

@@
 
-#### 1.2 Constant correlation
+_Question: How can one make a plot that tests the "limits" of the model, i.e how much "wiggliness" can this model accomodate? Make a plot to show the behavior of the model_
 
+#### 1.2 Covariance and correlation
+
+In a longitudinal study, multiple measures on the same subject creates a _correlation_ and _covariance_ issue that needs to be incorporated into the model.
+In this case, _correlation_ can be defined as the relationship between different variables, e.g.
+how much one variable changes when other variable changes.
+It can be further decomposed in _between-subject_ correlation and _within-subject correlation_.
+In the first case, _between-subject correlation_ measures the correlation between subject means, whereas _within-subject correlation_ describes how much a measurement in a subject changes in regard to another to another measurement _in the same subject_ [@roy2006].
+between repeated measures on the same subject can be defined as the relationship between the variability .
+The _correlation_ of the measures can be practically defined as is defined as the relationship between
 Missing observations, and correlation between measurements.
@@
 
-#### 1.2 the case of LMEMs using splines and how they work and how they are better than rm-ANOVA
+#### 1.2 the case of GMEMs using splines and how they work and how they are better than rm-ANOVA
 

### 2020-11-10 Manuscripts/Manuscript_AM_v1.Rmd
updated

--- 2020-11-09 Manuscripts/Manuscript_AM_v1.Rmd
@@
 output:
+  pdf_document: 
+    fig_caption: yes
+  html_document: default
+  css: style.css
   word_document:
     reference_docx: docx_template.docx
-  pdf_document: 
-    fig_caption: yes
-  css: style.css
-  html_document: default
 mainfont: Times New Roman
@@
 
-$$ y_{hij}=\mu+\gamma_h+\tau_j+(\gamma \tau)_{hj}+\pi_{i(h)}+e_{hij}$$
+\begin{equation}
+(\#eq:labelANOVA)
+y_{hij}=\mu+\gamma_h+\tau_j+(\gamma \tau)_{hj}+\pi_{i(h)}+e_{hij}
+\end{equation}
+
 Where
@@
 
-#### 1.2 Constant correlation
+_Question: How can one make a plot that tests the "limits" of the model, i.e how much "wiggliness" can this model accomodate? Make a plot to show the behavior of the model_
 
-Missing observations, and correlation between measurements.
-How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)
-#### 1.1 Constant correlation
+#### 1.2 Covariance and correlation in rm-ANOVA and LMEMs
 
-While the _repeated measures analysis of variance_ (rm-ANOVA) is the norm in the biomedical community to analyze longitudinal data, two frequent situations that arise in longitudinal studies limit its applicability.
-Missing observations, and correlation between measurements.
-How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)
+In a longitudinal study, the fact that multiple measures are taken on the same subject creates a a _covariance_ issue that needs to be incorporated into the model.
+Specifically, a parametric model (rm-ANOVA or LMEMs) assumes that there is no relationship between the repeated measures of different subjects (no correlation), but that there is a relationship for the repeated measures of the same subject.
+The latter is then taken in to account by a _covariance_ structure [@wolfinger1996].
+In this case, _covariance_ can be defined as the dependency  between two different values, e.g.
+if higher values of a variable correspond with higher values of another value, the covariance is positive.
+Although not immediately apparent, the reason for this specification arises from the fact that the model from \@ref(eq:labelANOVA) is re-written in matrix form for computational purposes (For an in-depth analysis see [@west2014;@weiss2005]).
+
+For an rm-ANOVA analysis, the _variance-covariance matrix_ or the matrix that specifies the relationships between the different repeated measures, needs to have a specific construction known as _compound symmetry_ (which is commonly known also as "sphericity" or "circularity").
+This assumes that variance of observations and the correlation between observations are constant,and when this is not the case, adjustments need to be made  [@weiss2005;@geisser1958;@huynh1976].However, it has been shown when the data violates the sphericity assumption, the false positivity rate is inflated [@haverkamp2017].This assumption is frequently unjustified as the correlation between measurements tends to change over time; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
+
+In the case of LMEMs, they have the advantage of allowing different structures of the variance-covariance matrix [@pinheiro2006].
+
+*make plot that explains between and within correlation*
+
+#### 1.3 Missing observations
+
+In a longitudinal study, missing observations are an important issue that arises frequently.
+In biomedical research, this situation is not unusual and can be caused by different reasons in the context of each study, but they are typically beyond the control of the investigator [molenberghs2004].statistical treatment of missing observations is particularly important for longitudinal studies.
+*talk about missing at random, mcar etc*
 
 One of the key assumptions for rm-ANOVA is the constant correlation among measurements.
-This frequently not the case as the correlation reduces as the time interval between two measurements increases [@ugrinowitsch2004].
+This frequently not the case as the correlation reduces as the time interval between two measurements increases [].
 From a practical
 
-#### 1.2 the case of LMEMs using splines and how they work and how they are better than rm-ANOVA
+#### 1.2 the case of GMEMs using splines and how they work and how they are better than rm-ANOVA
 

### 2020-11-11 Manuscripts/Manuscript_AM_v1.Rmd
Updated manuscript

@@
 In biomedical research, this type of study arises when the investigator intends to observe the evolution of the effect of a certain treatment across time, rather than analyzing it at a single time point (a cross-sectional study).
-Clinical examples of this  approach in biomedical research include studies on breast and neck cancer[@sio2016;@kamstra2015]; in the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT).
-Longitudinal studies have used also to measure tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010].
-From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach:it requires a lower number of subjects to reach a certain statistical power, and besides it being able to track the previously mentioned time-effect evolution on a group-by-group basis, it allows to determine the variability of the response within subjects [@guo2013; @fitzmaurice2012].
+Clinical examples of this  approach in biomedical research include studies on breast and neck cancer [@sio2016;@kamstra2015]; in the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT).
+Longitudinal studies have used also to measure tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
+From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach: it requires a lower number of subjects to reach a certain statistical power, and besides it being able to track the previously mentioned time-effect evolution on a group-by-group basis, it allows to determine the variability of the response within subjects [@guo2013; @fitzmaurice2012].
 In other words,a longitudinal study permits to quantify how the variable changes within each subject across time.
@@
 
-However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases[@ugrinowitsch2004], and the violation of this assumption increases the false positivity rate [@lane2016].
+However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases [@ugrinowitsch2004], and the violation of this assumption increases the false positivity rate [@lane2016].
 Moreover, it is unlikely that complete observations in all subjects are obtained in a biomedical study due to reasons that can be specific to different situations.
@@
 
-When the aforementioned issues arise, rm-ANOVA requires to exclude all subjects with missing observations from the analysis, thereby increasing costs for the study if the desired statistical power is not met with the remaining observations as it makes necessary to enroll more subjects; and  raising the possibility that the rejection of those partial observations limits the demonstration of significant differences between groups.
+When the aforementioned issues arise, rm-ANOVA requires exclusion of all subjects with missing observations from the analysis, thereby increasing costs for the study if the desired statistical power is not met with the remaining observations as it makes necessary to enroll more subjects; and  raising the possibility that the rejection of those partial observations limits the demonstration of significant differences between groups.
 Additionally, rm-ANOVA uses a _post hoc_ analysis to assess the relevance between the measured response in different groups.
@@
 
-Additionally, another assumption is made in both rm-ANOVA and LMEMs models, where a linear relationship is expected between the observed response and the covariates across the study [@pinheiro2006].This common assumption to both rm-ANOVA and LMEMs causes the models to be restrictive in their inferences when used in longitudinal data that does not follow a linear trend.
+Additionally, another assumption made in both rm-ANOVA and LMEMs models is that a linear relationship is expected between the observed response and the covariates across the study [@pinheiro2006].This common assumption to both rm-ANOVA and LMEMs causes the models to be restrictive in their inferences when used in longitudinal data that does not follow a linear trend.
 In biomedical research, this particular behavior in longitudinal has been reported in studies of tumor response to radio/chemotherapy in preclinical and clinical settings[@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994].
-This studies have shown that that the collected signal does not follow a linear trend over time, and presents high variability at different time points, making the estimations of a LMEM or rm-ANOVA model inconsistent with the pattern of the observed variations.
+These studies have shown that that the collected signal does not follow a linear trend over time, and presents high variability at different time points, making the estimations of a LMEM or rm-ANOVA model inconsistent with the pattern of the observed variations.
 Additionally,  although it is possible that a post hoc analysis is able to find “significance”( _p-value_<0.05) by using multiple comparisons between the model terms, this estimator would be inherently biased because of the lack of fit between the information and the model.
 
-As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed distribution) to determine the fit of the model while enabling inferences that are both adequate and consistent from a statistical perspective.
+As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumed distribution) to determine the fit of the model while enabling inferences that are both adequate and consistent from a statistical perspective.
 
 Generalized additive models (GAMs) are a subset of generalized linear models that use _smooth functions_ (hencefort _smooths_) to estimate the parameters of a model.
-They have been used in palaeolimnology, ecology and clinical studies to model longitudinal data[@woolway2016;@hefley2017;@ko2007].
-Briefly, GAMs use a combination of multiple functions (basis functions)to construct the smooths of the model [@wood2017].
+They have been used in palaeolimnology, ecology and clinical studies to model longitudinal data [@woolway2016;@hefley2017;@ko2007].
+Briefly, GAMs use a combination of multiple functions (basis functions) to construct the smooths of the model [@wood2017].
 Their main advantage over LMEMs and rm-ANOVA is that the model specification is directed by the _smooths_ rather than by a parametric relationship.
@@
 In that case, the field of _Bayesian statistics_ represents a relatively new area of Statistics that does not rely on _p-values_ and hypothesis tests to analyze information.
-Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the number of comparisons and inferences derived form the analysis.
+Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine significant differences and are able to expand the number of comparisons and inferences derived form the analysis.
 On the other hand, the shift that Bayesian statistics represent from the traditional "frequentist" statistical view in research, the computational tools required for its implementation, and the underlying mathematical theory have limited the use of this approach in the biomedical research community.
@@
 (\#eq:labelANOVA)
-y_{hij}=\mu+\gamma_h+\tau_j+(\gamma \tau)_{hj}+\pi_{i(h)}+e_{hij}
+y_{kit}=\mu+\gamma_k+\tau_t+(\gamma \tau)_{kt}+\pi_{i(k)}+e_{kit}
 \end{equation}
@@
 
-$i=1,...,n_h$ $j=1,...,t$, $h=1,...,p$; with $\pi_i{h} \sim N(0,\sigma^{2}_\epsilon)$ (independently normally distributed) and $e_{hij} \sim N(0,\sigma^{2}_\epsilon)$
+$i=1,...,n_K$ $t=1,...,T$, $k=1,...,K$; with $\pi_i{h} \sim N(0,\sigma^{2}_\epsilon)$ (independently normally distributed) and $e_{hij} \sim N(0,\sigma^{2}_\epsilon)$
 
@@
 
-In a longitudinal study, the fact that multiple measures are taken on the same subject creates a a _covariance_ issue that needs to be incorporated into the model.
+The fact that multiple measures are taken on the same subject creates a a _covariance_ issue that needs to be incorporated into the model.
 Specifically, a parametric model (rm-ANOVA or LMEMs) assumes that there is no relationship between the repeated measures of different subjects (no correlation), but that there is a relationship for the repeated measures of the same subject.
@@
 
-In a longitudinal study, missing observations are an important issue that arises frequently.
-In biomedical research, this situation is not unusual and can be caused by different reasons in the context of each study, but they are typically beyond the control of the investigator [molenberghs2004].statistical treatment of missing observations is particularly important for longitudinal studies.
+In a longitudinal study, missing observations are an issue that arises frequently.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator [molenberghs2004].Dropout from patients, or attrition or injury in animals are among the reasons for missing observations.
+Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
+In a MAR scenario, the pattern of the missing information is related to some variable in the dataset, but it is not related to the variable of interest [@scheffer2002].
+If the data are MCAR, this means that the missigness is completely unrelated to the collected information[@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
+In the case of LMEMs, one key advantage over rm-ANOVA is that the model can work with missing observations that are MAR, whereas rm-ANOVA needs to
+
 *talk about missing at random, mcar etc*
-
-One of the key assumptions for rm-ANOVA is the constant correlation among measurements.
-This frequently not the case as the correlation reduces as the time interval between two measurements increases [].
-From a practical
 

### 2020-11-16 Manuscripts/Manuscript_AM_v1.Rmd
Updated references and manuscript

@@
 
-#### 1.2 Covariance and correlation in rm-ANOVA and LMEMs
+#### 1.2 Covariance in rm-ANOVA and LMEMs
 
-The fact that multiple measures are taken on the same subject creates a a _covariance_ issue that needs to be incorporated into the model.
-Specifically, a parametric model (rm-ANOVA or LMEMs) assumes that there is no relationship between the repeated measures of different subjects (no correlation), but that there is a relationship for the repeated measures of the same subject.
-The latter is then taken in to account by a _covariance_ structure [@wolfinger1996].
-In this case, _covariance_ can be defined as the dependency  between two different values, e.g.
-if higher values of a variable correspond with higher values of another value, the covariance is positive.
-Although not immediately apparent, the reason for this specification arises from the fact that the model from \@ref(eq:labelANOVA) is re-written in matrix form for computational purposes (For an in-depth analysis see [@west2014;@weiss2005]).
+In a longitudinal study, there is a relationship between repeated measures on the same subject.
+In other words, there is an expected _variance_ between those measurements, and since repeated measures are also taken on the rest of the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
+The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
 
-For an rm-ANOVA analysis, the _variance-covariance matrix_ or the matrix that specifies the relationships between the different repeated measures, needs to have a specific construction known as _compound symmetry_ (which is commonly known also as "sphericity" or "circularity").
-This assumes that variance of observations and the correlation between observations are constant,and when this is not the case, adjustments need to be made  [@weiss2005;@geisser1958;@huynh1976].However, it has been shown when the data violates the sphericity assumption, the false positivity rate is inflated [@haverkamp2017].This assumption is frequently unjustified as the correlation between measurements tends to change over time; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
+For practical purposes, an rm-ANOVA analysis assumes that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
+This construction implies that the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
+However, it has been shown that this condition is is frequently unjustified [@maxwell2017] because the correlation between measurements tends to change over time; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].When the data violates the sphericity assumption, the false positivity rate is inflated [@haverkamp2017].
 

### 2020-11-17 Manuscripts/Manuscript_AM_v1.Rmd
Updated

@@
 
-When the aforementioned issues arise, rm-ANOVA requires exclusion of all subjects with missing observations from the analysis, thereby increasing costs for the study if the desired statistical power is not met with the remaining observations as it makes necessary to enroll more subjects; and  raising the possibility that the rejection of those partial observations limits the demonstration of significant differences between groups.
+When the aforementioned issues arise, rm-ANOVA requires exclusion of all subjects with missing observations from the analysis, thereby increasing costs for the study if the desired statistical power is not met with the remaining observations as it makes necessary to enroll more subjects [@ma2012]; and  raising the possibility that the rejection of those partial observations limits the demonstration of significant differences between groups.
 Additionally, rm-ANOVA uses a _post hoc_ analysis to assess the relevance between the measured response in different groups.
@@
 During the last decade, the biomedical community has started to recognize the limitations of a rm-ANOVA approach in the analysis of longitudinal information.
-This is exemplified by the pioneering use of  linear mixed effects models (LMEMs)  in certain groups to analyze tumor longitudinal data [@skala2010;@vishwanath2009].
+This is exemplified by the use of  linear mixed effects models (LMEMs)  in certain groups to analyze tumor longitudinal data [@skala2010;@vishwanath2009].
 Briefly, these models incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g.
@@
 
-Additionally, another assumption made in both rm-ANOVA and LMEMs models is that a linear relationship is expected between the observed response and the covariates across the study [@pinheiro2006].This common assumption to both rm-ANOVA and LMEMs causes the models to be restrictive in their inferences when used in longitudinal data that does not follow a linear trend.
+Additionally, another assumption made in both rm-ANOVA and LMEMs models is that a linear relationship exists between the observed response and the explanatory variables (covariates) across the study [@pinheiro2006].This common assumption to both rm-ANOVA and LMEMs causes the models to be restrictive in their inferences when used in longitudinal data that does not follow a linear trend.
 In biomedical research, this particular behavior in longitudinal has been reported in studies of tumor response to radio/chemotherapy in preclinical and clinical settings[@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994].
 These studies have shown that that the collected signal does not follow a linear trend over time, and presents high variability at different time points, making the estimations of a LMEM or rm-ANOVA model inconsistent with the pattern of the observed variations.
-Additionally,  although it is possible that a post hoc analysis is able to find “significance”( _p-value_<0.05) by using multiple comparisons between the model terms, this estimator would be inherently biased because of the lack of fit between the information and the model.
+Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” results ( _p_<0.05) by using multiple comparisons between the model terms, this metric is inherently biased due to the discrepancy between the data and the model.
 
@@
 \begin{equation}
-(\#eq:labelANOVA)
-y_{kit}=\mu+\gamma_k+\tau_t+(\gamma \tau)_{kt}+\pi_{i(k)}+e_{kit}
+\label{eq:ANOVA}
+$$y_{kit}=\mu+\gamma_k+\tau_t+(\gamma \tau)_{kt}+\pi_{i(k)}+e_{kit}$$
 \end{equation}
@@
 The term $\pi_{ij}$ represents the _random effects_ for each subject within each group.Finally, $e_{hij}$ represents the independent random error terms, which need to be normally distributed with mean 0 [@davis2002].
-The model then, is a linear combination of terms, and if plotted, it would a straight line.
+The model then, is a linear combination of terms, and if plotted, represents a line.
 
@@
 
-In a longitudinal study, there is a relationship between repeated measures on the same subject.
-In other words, there is an expected _variance_ between those measurements, and since repeated measures are also taken on the rest of the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
+In a longitudinal study there is an expected _variance_ between repeated measurements on the same subject, and since repeated measures are also taken on the rest of the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
 The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
@@
 For practical purposes, an rm-ANOVA analysis assumes that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
-This construction implies that the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
-However, it has been shown that this condition is is frequently unjustified [@maxwell2017] because the correlation between measurements tends to change over time; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].When the data violates the sphericity assumption, the false positivity rate is inflated [@haverkamp2017].
+Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
+However, it has been shown that this condition is frequently unjustified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].When the data violates the sphericity assumption, the false positivity rate is inflated [@haverkamp2017].
 
-In the case of LMEMs, they have the advantage of allowing different structures of the variance-covariance matrix [@pinheiro2006].
+In the case of LMEMs, one of their main advantages is that they allow multiple structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others[@pinheiro2006].
+The choice for the covariance matrix needs to be carefully considered by the investigator based on the analyzing the collected data.
 
@@
 
-In a longitudinal study, missing observations are an issue that arises frequently.
+Missing observations are an issue that arises frequently in longitudinal studies.
 In biomedical research, this situation can be caused by reasons beyond the control of the investigator [molenberghs2004].Dropout from patients, or attrition or injury in animals are among the reasons for missing observations.
@@
 In a MAR scenario, the pattern of the missing information is related to some variable in the dataset, but it is not related to the variable of interest [@scheffer2002].
-If the data are MCAR, this means that the missigness is completely unrelated to the collected information[@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
-In the case of LMEMs, one key advantage over rm-ANOVA is that the model can work with missing observations that are MAR, whereas rm-ANOVA needs to
+If the data are MCAR, this means that the missigness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
+An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
 
-*talk about missing at random, mcar etc*
+In the case of LMEMs, one key advantage over rm-ANOVA is that inferences from the model are valid when missing observations in the dataset exist that are MAR [@west2014].
+The pattern of missing observations can be considered MAR if the missing observations are not related any of the other variables measured in the study [@maxwell2017].
+For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study.
+Because in this case the missigness is unrelated to other variables of interest, the missing data can be considered MAR.
 
-#### 1.2 the case of GMEMs using splines and how they work and how they are better than rm-ANOVA
+This section has presented the limitations of rm-ANOVA to analyze longitudinal information and the differences of LMEMs with regards to missing data and the modeling of the covariance matrix.
+Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
+However, when the data presents high variability, LMEMs fail to capture the non-linear trend of the data.
+To analyze such type of data, we propose the use of generalized additive models (GAMs) in the following section.
+
+#### 2 GAMs as a special case of Generalized Linear Models:
+
+A GAM is a special case of the Generalized Linear Model (GLM), a framework that allows for response distributions that do not follow a normal distribution [@wood2017;@hastie1987].
+[@simpson2018].
+To allow more flexibility, GAMs use _smooth functions_  to model the relationship between the covariates and the response.
+The smooth functions are represented using splines over  evenly spaced ranges of the covariates known as _knots_.
+A GAM model can be represented as:
+
+\begin{equation}
+(\#eq:labelGAM)
+y_{t}=\beta_0+f(x_t)+\epsilon_t  
+\end{equation}
+
+Where $y_t$ is the response at time $t$, $\beta_0$ is the expected value at time 0, the change of $y$ over time is represented by the function $f(x_t)$ and $\epsilon_t$ represents the residuals.
+
+*This equation above does not have an "interaction term" similar to that of anova.*
+
+This model is similar in its construction to that of \ref{eq:ANOVA}, but the key difference relies on the fact that instead of a linear relationship for the covariates, a function is used.
 

### 2020-12-20 Manuscripts/Manuscript_AM_v1.Rmd
Updated
Update the GAM section

@@
 
+```{r,quadatric behavior}
+#equation y=-2*(x-3)^2
+
+```
+
 ## Section 1:
@@
 
-The _repeated measures analysis of variance_ (rm-ANOVA) is the standard for the statistical analysis of longitudinal data, and there are key assumptions that are made in order to make the model valid.
+The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data, and certain assumptions are necessary to make the model valid.
 From a practical view, they can be divided in three areas: linear relationship between covariates and response, constant correlation between measurements, and complete observations for all subjects.
-Each one of these assumptions is discussed below.
+Each one of these assumptions is discussed below from a biomedical perspective.
 
@@
 
-In a biomedical longitudinal study, two or more groups of subjects (humans, mice, samples) are subject to a different treatments (e.g.
-group of mice receiving a novel drug vs.
-a group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-Moreover, it is assumed that the collected response has two components: a _fixed_ and a _random_ component.
-The _fixed_ component can be understood as a constant value in the response which the researcher intends to measure, i.e, the effect of a novel drug in a subject.The _random_ component can be defined as "noise" caused by some some factors that are not of interest to the researcher, i.e., if the concentration of a drug is measured in some subjects within the same group in the early hours of the morning while others are measured in the afternoon, the researcher might consider this variability in the collection time of the measurement to introduce some "noise" in the signal.
-As their name suggests, this "random" variabliity needs to be modeled as a variable rather than as a constant value.
+In a biomedical longitudinal study, two or more groups of subjects (patients, mice, samples) are subject to a different treatments (e.g.
+a "treatment" group receives a novel drug vs.
+a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
+The collected response has both  _fixed_ and  _random_ components.
+The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the effect of the novel drug in the "treatment" group.The _random_ component can be defined as "noise" caused by some inherent variability within the study.
+For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while others are measured in the afternoon, it is possible that this this variability in the collection time of the measurement introduces some "noise" in the signal.
+As their name suggests, this "random" variabiliity needs to be modeled as a variable rather than as a constant value.
 
-Mathematically speaking, if a normally distributed response $y$ is measured repeatedly at $t$ time points from subjects in $p$ groups, where each group has a certain $n_p$ number of subjects, the the model for the response $y_{hij}$ becomes:
+Mathematically speaking, if a normally distributed response $y$ is measured repeatedly at $t$ time points from subjects in $p$ groups, where each group has a certain $n_p$ number of subjects, the rm-ANOVA the model for the response $y_{hij}$ becomes:
 
@@
 In this model, $\mu$ represents the group mean, $\gamma_h$ is the _fixed effect_ of group $h$, $\tau_j$ is the fixed effect of time $j$, and $(\gamma \tau)_{hj}$ represents the interaction of time and group effects.
-The term $\pi_{ij}$ represents the _random effects_ for each subject within each group.Finally, $e_{hij}$ represents the independent random error terms, which need to be normally distributed with mean 0 [@davis2002].
+The term $\pi_{ij}$ represents the _random effects_ for each subject within each group.
+Finally, $e_{hij}$ represents the independent random error terms, which need to be normally distributed with mean 0 [@davis2002].
 The model then, is a linear combination of terms, and if plotted, represents a line.
@@
 
-In a longitudinal study there is an expected _variance_ between repeated measurements on the same subject, and since repeated measures are also taken on the rest of the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
+In a longitudinal study there is an expected _variance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
 The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
 
-For practical purposes, an rm-ANOVA analysis assumes that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
+In the case of an rm-ANOVA analysis, it is assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
 Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
-However, it has been shown that this condition is frequently unjustified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].When the data violates the sphericity assumption, the false positivity rate is inflated [@haverkamp2017].
+However, it has been shown that this condition is frequently unjustified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser) but the effectiveness of each correction is dependent on the size of the sample and the number of repeated measurement[@haverkamp2017].
 
-In the case of LMEMs, one of their main advantages is that they allow multiple structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others[@pinheiro2006].
-The choice for the covariance matrix needs to be carefully considered by the investigator based on the analyzing the collected data.
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others [@pinheiro2006].
+Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself.
+Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
 
@@
 Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
-In a MAR scenario, the pattern of the missing information is related to some variable in the dataset, but it is not related to the variable of interest [@scheffer2002].
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002].
 If the data are MCAR, this means that the missigness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
@@
 
-In the case of LMEMs, one key advantage over rm-ANOVA is that inferences from the model are valid when missing observations in the dataset exist that are MAR [@west2014].
+In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR [@west2014].
 The pattern of missing observations can be considered MAR if the missing observations are not related any of the other variables measured in the study [@maxwell2017].
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study.
-Because in this case the missigness is unrelated to other variables of interest, the missing data can be considered MAR.
+For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.
 
-This section has presented the limitations of rm-ANOVA to analyze longitudinal information and the differences of LMEMs with regards to missing data and the modeling of the covariance matrix.
+_What limitations LMEMs have regarding missing observations?_
+
+This section has presented the assumptions of rm-ANOVA to analyze longitudinal information and  its differences when compared to LMEMs  regarding to missing data and the modeling of the covariance matrix.
 Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
 However, when the data presents high variability, LMEMs fail to capture the non-linear trend of the data.
-To analyze such type of data, we propose the use of generalized additive models (GAMs) in the following section.
+To analyze such type of data, we present generalized additive models (GAMs) as an alternative in the following section.
 
-#### 2 GAMs as a special case of Generalized Linear Models:
+#### 2 GAMs as a special case of Generalized Linear Models
 
@@
 [@simpson2018].
-To allow more flexibility, GAMs use _smooth functions_  to model the relationship between the covariates and the response.
+In contrast to rm-ANOVA or LMEMs, GAMs use _smooth functions_  to model the relationship between the covariates and the response.
 The smooth functions are represented using splines over  evenly spaced ranges of the covariates known as _knots_.

### 2020-12-23 splines-missing-data.R
JRT add equations and missing data

@@ -56,8 +56,6 @@ dat_missing %>%
     summarize(count = n())
 
 
-
-
 dat_missing$Group <- as.factor(dat_missing$Group)
 dat_sim$Group <- as.factor(dat_sim$Group)
 mod1 <- gam(observation ~ s(Day, by = Group, k = 5), data  = dat_missing)

### 2020-12-24 splines-missing-data.R
Trying to plot smooths over data

@@ -69,6 +69,46 @@ layout(matrix(1:4, 2, 2))
 plot(mod1, main = "Missing")
 plot(mod2, main = "Full")
 
+par(mfrow=c(1,1))
 
 gam.check(mod1)
 gam.check(mod2)
+
+#plotting smooths and simulated data
+#full data
+full_predict<-with(dat_sim,expand.grid(observation=seq(min(observation),
+                                                       max(observation),length=400),
+                                                    Group=Group,Day=Day))
+
+full_predict<-cbind(full_predict,
+                         predict(mod2,full_predict,
+                                 se.fit = TRUE,type='response'))
+
+
+ggplot(data=full_predict,aes(x=Day,y=fit))+geom_point()+facet_wrap(~Group)
+
+ggplot(data=dat_sim, aes(x=Day, y=observation, group=Group)) +
+    facet_wrap(~Group) +
+    geom_point(colour='black',size=1,alpha=0.5)+
+    geom_ribbon(aes(ymin=(fit - 2*se.fit), ymax=(fit + 2*se.fit), x=Day),
+                data=full_predict, alpha=0.3,
+                inherit.aes=FALSE) +
+    geom_line(aes(x=Day,y=(fit),color=factor(Group)), size=1,data=full_predict,show.legend = FALSE)
+
+#missing data
+missing_predict<-with(dat_missing,expand.grid(observation=seq(min(observation),
+                                                       max(observation),length=400),
+                                       Group=levels(Group),Day=Day))
+
+missing_predict<-cbind(missing_predict,
+                    predict(mod1,missing_predict,
+                            se.fit = TRUE,type='response'))
+
+
+ggplot(data=dat_missing, aes(x=Day, y=observation, group=Group)) +
+    facet_wrap(~Group) +
+    geom_point(colour='black',size=1,alpha=0.5)+
+    geom_ribbon(aes(ymin=(fit - 2*se.fit), ymax=(fit + 2*se.fit), x=Day),
+                data=missing_predict, alpha=0.3,
+                inherit.aes=FALSE) +
+    geom_line(aes(y=(fit),color=factor(Group)), size=1,data=missing_predict,show.legend = FALSE)

### 2020-12-25 splines-missing-data.R
Testing prediction plot

@@ -84,7 +84,6 @@ full_predict<-cbind(full_predict,
                          predict(mod2,full_predict,
                                  se.fit = TRUE,type='response'))
 
-
 ggplot(data=full_predict,aes(x=Day,y=fit))+geom_point()+facet_wrap(~Group)
 
 ggplot(data=dat_sim, aes(x=Day, y=observation, group=Group)) +

### 2020-12-25 Manuscripts/Manuscript_AM_v1.Rmd
Changes in manuscript
Updated manuscript background section

@@
 A longitudinal study is defined as that which is designed to repeatedly measure a variable of interest in a group (or groups) of subjects.
-In biomedical research, this type of study arises when the investigator intends to observe the evolution of the effect of a certain treatment across time, rather than analyzing it at a single time point (a cross-sectional study).
-Clinical examples of this  approach in biomedical research include studies on breast and neck cancer [@sio2016;@kamstra2015]; in the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT).
-Longitudinal studies have used also to measure tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
-From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach: it requires a lower number of subjects to reach a certain statistical power, and besides it being able to track the previously mentioned time-effect evolution on a group-by-group basis, it allows to determine the variability of the response within subjects [@guo2013; @fitzmaurice2012].
-In other words,a longitudinal study permits to quantify how the variable changes within each subject across time.
+In biomedical research, this type of study is employed when the intention is to observe the evolution of the effect of a certain treatment across time, rather than analyzing it at a single time point (a cross-sectional study).
+Clinical studies of breast and neck cancer have used a longitudinal approach [@sio2016;@kamstra2015].
+In the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT).
+Longitudinal studies have been also used to analyze tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
 
-Traditionally, a “frequentist” approach is used in biomedical research to derive inferences from the results of a longitudinal study.
-Such statistical view derives its name from the fact that it regards probability as a limiting frequency [wagenmakers2008] and its application is based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-This methodology makes two key assumptions regarding longitudinal data: a constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study  [@schober2018;@gueorguieva2004].
+Traditionally, a “frequentist” approach is used in biomedical research to derive inferences from a longitudinal study.
+Such statistical view derives its name from the fact that it regards probability as a limiting frequency[@wagenmakers2008] and its application is based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
+This methodology makes two key assumptions regarding longitudinal data: constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study (a condition also known as complete observations) [@gueorguieva2004;@schober2018].
 
-However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases [@ugrinowitsch2004], and the violation of this assumption increases the false positivity rate [@lane2016].
-Moreover, it is unlikely that complete observations in all subjects are obtained in a biomedical study due to reasons that can be specific to different situations.
-In a clinical trial, voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and in both cases it is possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point.
+However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases [@ugrinowitsch2004]. 
+In the case of the number of observations during the study, different situations can arise that prevent the collection of complete measurements for all subjects:  in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and it is even possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point and therefore violating the _complete observations_ assumption of rm-ANOVA.
 
-When the aforementioned issues arise, rm-ANOVA requires exclusion of all subjects with missing observations from the analysis, thereby increasing costs for the study if the desired statistical power is not met with the remaining observations as it makes necessary to enroll more subjects [@ma2012]; and  raising the possibility that the rejection of those partial observations limits the demonstration of significant differences between groups.
-Additionally, rm-ANOVA uses a _post hoc_ analysis to assess the relevance between the measured response in different groups.
-This analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance.
-Because the p-value is highly variable [@halsey2015;@nuzzo2014], multiple comparisons can inflate the false positivity rate [@liu2010],consequently biasing the conclusions of the study.
+When incomplete observations occur, an rm-ANOVA requires the exclusion of all subjects with missing observations from the analysis.
+This can result in increased costs if the desired statistical power is not met with the remaining observations, as it would be necessary to enroll more subjects.
+At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
+Additionally, rm-ANOVA uses a _post hoc_ analysis to assess differences between the measured response in different groups.
+A _post hoc_ analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance.
+Because the _p-value_ is highly variable, multiple comparisons can inflate the false positivity rate [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
 
-During the last decade, the biomedical community has started to recognize the limitations of a rm-ANOVA approach in the analysis of longitudinal information.
-This is exemplified by the use of  linear mixed effects models (LMEMs)  in certain groups to analyze tumor longitudinal data [@skala2010;@vishwanath2009].
+During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
+This is exemplified by the  use of  linear mixed effects models (LMEMs)  in certain groups to analyze longitudinal data [@skala2010;@vishwanath2009].
 Briefly, these models incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g.
-the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population [@pinheiro2006].These models are more flexible than rm-ANOVA as they can accommodate missing observations for multiple subjects, and allow different modeling strategies for the variability within each measure in every subject [@west2014;@pinheiro2006].On the other hand, they impose restrictions in the distribution of the errors of the model and random effects [@gueorguieva2004;@schielzeth2020].
+the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population.
+These models are more flexible than the traditional rm-ANOVA as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
+On the other hand, they impose restrictions in the distribution of the errors  of the random effects [@gueorguieva2004].
 
-Additionally, another assumption made in both rm-ANOVA and LMEMs models is that a linear relationship exists between the observed response and the explanatory variables (covariates) across the study [@pinheiro2006].This common assumption to both rm-ANOVA and LMEMs causes the models to be restrictive in their inferences when used in longitudinal data that does not follow a linear trend.
-In biomedical research, this particular behavior in longitudinal has been reported in studies of tumor response to radio/chemotherapy in preclinical and clinical settings[@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994].
-These studies have shown that that the collected signal does not follow a linear trend over time, and presents high variability at different time points, making the estimations of a LMEM or rm-ANOVA model inconsistent with the pattern of the observed variations.
-Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” results ( _p_<0.05) by using multiple comparisons between the model terms, this metric is inherently biased due to the discrepancy between the data and the model.
+One final assumption that is not initially evident for  both rm-ANOVA and LMEMs models pertains the relationship between the observed effect and the progression of the timeline of the study, which is expected to be linear [@pinheiro2006].
+This common assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend.
+In biomedical research, a special case of this behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
+These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent with the observed variation.
+In other words, the in both cases the model tries to explain highly-variable data using a linear trend, consequently biasing the estimates.
+Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) by using multiple comparisons between the model terms, the use of this estimator is limited in such scenario because it is intended to work in models that have a reasonable agreement with the data.
 
-As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumed distribution) to determine the fit of the model while enabling inferences that are both adequate and consistent from a statistical perspective.
+As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, while enabling inferences that are both adequate and consistent from a statistical perspective.
+In this regard, Generalized Additive Models (GAMs) present an alternative approach to analyze longitudinal data.
+While not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018].
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: They have an extensively developed theoretical background, enable the data to dictate the trend of the model, can accommodate missing observations and do not require constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more suitable statistical method to analyze non-linear biomedical longitudinal data.
 
-Generalized additive models (GAMs) are a subset of generalized linear models that use _smooth functions_ (hencefort _smooths_) to estimate the parameters of a model.
-They have been used in palaeolimnology, ecology and clinical studies to model longitudinal data [@woolway2016;@hefley2017;@ko2007].
-Briefly, GAMs use a combination of multiple functions (basis functions) to construct the smooths of the model [@wood2017].
-Their main advantage over LMEMs and rm-ANOVA is that the model specification is directed by the _smooths_ rather than by a parametric relationship.
-This allows a consistent fit of the model with the data, and estimations of significance using the terms of the model.
-However, certain assumptions about the data are necessary: a normal distribution and constant variance of the residuals with the mean response.
-Therefore, GAMs provide a more suitable statistical method to analyze biomedical longitudinal data, when these assumptions of the model are met by the data.
+ The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs.
+In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2011;@wood2016;@wood2017] that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g.
+hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+This programming language is also able to simulate data, an emerging strategy used to test statistical models [@haverkamp2017].
+This allows to create and  explore different alternatives for analysis without collecting information in the field, and reduces the time window between experiment design and its implementation.
 
-However, it is possible that the assumptions of GAMs for longitudinal data do not hold under  certain circumstances.
-In that case, the field of _Bayesian statistics_ represents a relatively new area of Statistics that does not rely on _p-values_ and hypothesis tests to analyze information.
-Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine significant differences and are able to expand the number of comparisons and inferences derived form the analysis.
-On the other hand, the shift that Bayesian statistics represent from the traditional "frequentist" statistical view in research, the computational tools required for its implementation, and the underlying mathematical theory have limited the use of this approach in the biomedical research community.
-However,Bayesian theory is intuitive and shares some principles with "frequentist" statistics, and there is an increasing use and recognition of the advantages of their use across different areas of biomedical research such as clinical trial design and imaging[@biswas2009;@kelter2020;@kwon2020;@zhou2017].
-
-Additionally,the current development in computational tools, specifically the programming language $\textsf{R}$, enable a rapid implementation of GAMs and Bayesian models for longitudinal data.In particular, $\textsf{R}$ has an extensive collection of documentation, functions and libraries that speed up the initial stages of the analsys and that enable the use of complex statistical methods without requiring a specialized set of programming skills from the user.
-
-Therefore, this study focuses in three areas in the analysis of longitudinal data from a biomedical perspective.
-First, it presents the limitations of (rm-ANOVA) and LMEMs over longitudinal data, and explains how these limitations in turn affect the results of the analysis.
-Secondly, it uses $\textsf{R}$ to simulate non-linear longitudinal data following previously reported values in the literature, and presents the implementation of GAMs as a statistical tool for longitudinal data.
-And finally, it introduces Bayesian statistics and presents their implementation with GAMs to demonstrate the differences and benefits of this approach.
-with an emphasis in reproducibility  by providing the code, simulated dataset and a step-by-step guide of the computational implementation of different models, this study aims to encourage the exploration of modern statistical methods for biomedical longitudinal data, and to improve the  statistical standards in biomedical research.
+Therefore, the purpose of this study is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas.
+First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity and correlation structures.
+Secondly, the key theoretical elements of GAMS are presented without using extensive derivations or complex mathematical notation that can obscure their understanding.
+Thirdly, simulated data that follows the trend of previously reported values [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research and the implementation of GAMs in such scenario.
+Finally, reproducibility is emphasized by providing the code to generate the data and the implementation of GAMs in $\textsf{R}$, in conjunction with a step-by-step guide to fit models of increasing complexity.
+In summary, the exploration of modern statistical techniques to analyze longitudinal data may allow biomedical researchers to consider the use of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a linear trend, and will also help to improve the standards for reproducibility  in biomedical research.
 
@@
 
-#### 1.3 Bayesian brief introduction, and compare the results of 1.2 to the results of Bayesian
+#### 1.3 Bayesian brief introduction, and compare the results of 1.2 to the results of Bayesianjo
 

### 2020-12-29 Manuscripts/Manuscript_AM_v1.Rmd
updated manuscript on GAM section and updated references

@@
 
-- Why LMEMs are better than ANOVA
-
-- How LMEMs (using splines) and a Bayesian analysis can be used to analyze longitudinal data
-
-https://www.frontiersin.org/articles/10.3389/fevo.2018.00149/full
-
 ```{r,FIGURE 1,echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Simulated longitudinal data with a linear trend.'}
@@
 
-The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data, and certain assumptions are necessary to make the model valid.
+The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research, but certain assumptions are necessary to make the model valid.
 From a practical view, they can be divided in three areas: linear relationship between covariates and response, constant correlation between measurements, and complete observations for all subjects.
-Each one of these assumptions is discussed below from a biomedical perspective.
+Each one of these assumptions is discussed below.
 
@@
 The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the effect of the novel drug in the "treatment" group.The _random_ component can be defined as "noise" caused by some inherent variability within the study.
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while others are measured in the afternoon, it is possible that this this variability in the collection time of the measurement introduces some "noise" in the signal.
+For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while others are measured in the afternoon, it is possible that the difference in the collection time of the measurement introduces some "noise" in the signal.
 As their name suggests, this "random" variabiliity needs to be modeled as a variable rather than as a constant value.
 
-Mathematically speaking, if a normally distributed response $y$ is measured repeatedly at $t$ time points from subjects in $p$ groups, where each group has a certain $n_p$ number of subjects, the rm-ANOVA the model for the response $y_{hij}$ becomes:
+Mathematically speaking, if a normally distributed response $y$ is measured repeatedly at $t$ time points from  $i$ subjects in $k$ groups, the rm-ANOVA the model for the response $y_{kit}$ becomes:
 
@@
 
-$i=1,...,n_K$ $t=1,...,T$, $k=1,...,K$; with $\pi_i{h} \sim N(0,\sigma^{2}_\epsilon)$ (independently normally distributed) and $e_{hij} \sim N(0,\sigma^{2}_\epsilon)$
+$i=1,...,n_K$ $t=1,...,T$, $k=1,...,K$; with $\pi_{ik} \sim N(0,\sigma^{2}_\epsilon)$ (independently normally distributed) and $e_{hik} \sim N(0,\sigma^{2}_\epsilon)$
 
-In this model, $\mu$ represents the group mean, $\gamma_h$ is the _fixed effect_ of group $h$, $\tau_j$ is the fixed effect of time $j$, and $(\gamma \tau)_{hj}$ represents the interaction of time and group effects.
-The term $\pi_{ij}$ represents the _random effects_ for each subject within each group.
-Finally, $e_{hij}$ represents the independent random error terms, which need to be normally distributed with mean 0 [@davis2002].
-The model then, is a linear combination of terms, and if plotted, represents a line.
+In this model, $\mu$ represents the group mean, $\gamma_k$ is the _fixed effect_ of each group $k$, $\tau_t$ is the fixed effect of time, and $(\gamma \tau)_{kt}$ represents the interaction of time and group effects.
+The term $\pi_{ik}$ represents the _random effects_ for each subject within each group.
+Finally, $e_{hik}$ represents the independent random error terms, which need to be normally distributed with mean 0 [@davis2002].
+The model then, assumes a linear relationship between the response and the covariate.
+This is better represented in the follo
 
@@
 However, it has been shown that this condition is frequently unjustified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser) but the effectiveness of each correction is dependent on the size of the sample and the number of repeated measurement[@haverkamp2017].
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser) the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
+In other words, if the data does not present constant correlation between repeated measurements, the assumptions required for an rm-ANOVA model are not met and the use of corrections may still not provide a reasonable adjustment that makes the model valid.
 
@@
 A GAM is a special case of the Generalized Linear Model (GLM), a framework that allows for response distributions that do not follow a normal distribution [@wood2017;@hastie1987].
-[@simpson2018].
-In contrast to rm-ANOVA or LMEMs, GAMs use _smooth functions_  to model the relationship between the covariates and the response.
-The smooth functions are represented using splines over  evenly spaced ranges of the covariates known as _knots_.
-A GAM model can be represented as:
+Following the notation by Simpson [@simpson2018] A GAM model can be represented as:
 
@@
 
-*This equation above does not have an "interaction term" similar to that of anova.*
+In contrast to rm-ANOVA or LMEMs, GAMs use _smooth functions_  to model the relationship between the covariates and the response.
+This approach is more advantageous than that of a LMEM or rm-ANOVA models as it allows a non-linear behavior of the data.
+One possibility is to use a polynomial function, but polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
 
-This model is similar in its construction to that of \ref{eq:ANOVA}, but the key difference relies on the fact that instead of a linear relationship for the covariates, a function is used.
+In GAMs the smooth functions are represented using _basis functions_ function over  evenly spaced ranges of the covariates known as _knots_.
+In particular, the _basis function_ used is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together[@wood2017;@simpson2018].
+
+This model is similar in its construction to that of \ref{eq:ANOVA}, but the key difference relies on the fact that instead of a linear combination of terms, a function is used to model the response.
+This allows
 

### 2020-12-31 Manuscripts/Manuscript_AM_v1.Rmd
updated manuscript and references

@@
 
-In GAMs the smooth functions are represented using _basis functions_ function over  evenly spaced ranges of the covariates known as _knots_.
-In particular, the _basis function_ used is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together[@wood2017;@simpson2018].
+In GAMs the smooth functions are represented using _basis functions_  over  evenly spaced ranges of the covariates known as _knots_.
+The _basis function_ used is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together[@wood2017;@simpson2018].
+Cubic splines have a long history in their use to solve non-parametric statistical problems and are the default choice to fit GAMs as they are the simplest option to obtain visual smoothness [@wegman1983].
 

### 2021-01-03 Manuscripts/Manuscript_AM_v1.Rmd
Updated manuscript and references

@@
 In this regard, Generalized Additive Models (GAMs) present an alternative approach to analyze longitudinal data.
-While not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018].
+Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and political science [@beck1998] .
 There are several advantages of GAMs over LMEMs and rm-ANOVA models: They have an extensively developed theoretical background, enable the data to dictate the trend of the model, can accommodate missing observations and do not require constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more suitable statistical method to analyze non-linear biomedical longitudinal data.
@@
 
-```{r,FIGURE 1,echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Simulated longitudinal data with a linear trend.'}
-rm(list=ls())
-set.seed(123)
-library(ggplot2)
-library(splines)
-library(tidyverse)
-library(here)
-
-linear<-read.csv(here("data","Simulated_linear.csv"))
-
-sim_linear <- function(dat, n = 10, sd = 5) {
-    linear_vals <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group,Day) %>%
-        mutate(observation = rnorm(n, y, sd)) %>%
-        ungroup()
-
-    return(linear_vals)
-}
-n<-10
-sd<-2
-linear_vals <- sim_linear(linear, n, sd)
-linear_vals[] <- lapply(linear_vals, function(observation) ifelse(observation<=0, NA, observation)) #taking out values below 0
-
- ggplot(linear_vals, aes(x = Day, y = observation, color = Group)) +
-    geom_point() +
-    ggtitle("Simulated linear time-varying responses") +
-    scale_color_viridis_d(end = 0.75) +
-    geom_line(aes(y = y), alpha = 0.75) +
-    theme_bw()
-
-```
-
-```{r,quadatric behavior}
-#equation y=-2*(x-3)^2
-
-```
-
 ## Section 1:
 
-### Challenges presented by longitudinal studies:
+### Challenges presented by longitudinal studies
 
@@
 
-*make plot that explains between and within correlation*
-
 #### 1.3 Missing observations
@@
 In contrast to rm-ANOVA or LMEMs, GAMs use _smooth functions_  to model the relationship between the covariates and the response.
-This approach is more advantageous than that of a LMEM or rm-ANOVA models as it allows a non-linear behavior of the data.
-One possibility is to use a polynomial function, but polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
+This approach is more advantageous than that of a LMEMs or rm-ANOVA as it does not restrict the model to a linear relationship.
+One possible function for $f(x_t)$  is a polynomial, but a major limitation is that they create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
 
-In GAMs the smooth functions are represented using _basis functions_  over  evenly spaced ranges of the covariates known as _knots_.
+To overcome this limitation, the smooth functions in GAMs  are represented using _basis functions_  over  evenly spaced ranges of the covariates known as _knots_.
 The _basis function_ used is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together[@wood2017;@simpson2018].
 Cubic splines have a long history in their use to solve non-parametric statistical problems and are the default choice to fit GAMs as they are the simplest option to obtain visual smoothness [@wegman1983].
-
-This model is similar in its construction to that of \ref{eq:ANOVA}, but the key difference relies on the fact that instead of a linear combination of terms, a function is used to model the response.
-This allows
 

### 2021-01-04 Manuscripts/Manuscript_AM_v1.Rmd
Updated manuscript

@@
 
-Mathematically speaking, if a normally distributed response $y$ is measured repeatedly at $t$ time points from  $i$ subjects in $k$ groups, the rm-ANOVA the model for the response $y_{kit}$ becomes:
+Mathematically speaking, a linear model can be written as:
+
+\begin{equation}
+\label{eq:linear_model}
+$$
+y = \beta_0 + time * \beta_1 + treatment * \beta_2 + time * treatment * \beta_3 \\
+y = \begin{cases}
+\beta_0 + time * \beta_1 & \mbox{if treatment 1}\\
+\beta_0 + time * \beta_1 + treatment * \beta_2 + time * treatment * \beta_3 & \mbox{if treatment 2}\\
+\end{cases}
+\\
+$$
+\end{equation}
 
@@
 The model then, assumes a linear relationship between the response and the covariate.
-This is better represented in the follo
 
-_Question: How can one make a plot that tests the "limits" of the model, i.e how much "wiggliness" can this model accomodate? Make a plot to show the behavior of the model_
+The rm-ANOVA model  can be re-written under the generalized linear model  (GLM) framework [@nelder1972].
+The main difference in this case is that a GLM model  uses a _linking function_ to connect the observed value to the linear model.
+Thus,
+
+\begin{equation}
+\label{eq:GLM}
+
+$$
+y = \beta_0 + f(time) * \boldsymbol{\beta}_1 + treatment * \beta_2 + f(time) * treatment * \boldsymbol{\beta}_3 \\
+y = \begin{cases}
+\beta_0 + f(time) * \boldsymbol{\beta}_1 & \mbox{if treatment 1} \\
+\beta_0 + f(time) * \boldsymbol{\beta}_1 + treatment * \beta_2 + f(time) * treatment * \boldsymbol{\beta}_3 & \mbox{if treatment 2}\\
+\end{cases}
+\\
+$$
+\end{equation}
+
+\begin{equation}
+\label{eq:ANOVA_GLM}
+$$y_{kit}=\mu+f(\gamma_k)+f(\tau_t)+(\gamma \tau)_{kt}+\pi_{i(k)}+e_{kit}$$
+\end{equation}
+
+The Linear model as a GLM: set $f(time) = time$ with the identity function $f(x) = x$
 

### 2021-01-04 Manuscripts/Manuscript_AM_v1.Rmd
Updated manuscript

--- 2021-01-03 Manuscripts/Manuscript_AM_v1.Rmd
@@
 ---
-title: '**Bayesian statistics for longitudinal studies in biomedical research**'
+title: '**The statistical analysis of non-linear longitudinal data in biomedical research**'
 author: 
@@
 output:
-  html_document: default
   pdf_document: 
     fig_caption: yes
+  html_document: default
   css: style.css
@@
 link-citations: yes
-subtitle: '_Their application and use in biomedical research_'
-biblio-style: apalike
+subtitle: '_Beyond repeated measures ANOVA and Linear Mixed Models_'
+csl: ieee.csl
 ---
@@
 For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while others are measured in the afternoon, it is possible that the difference in the collection time of the measurement introduces some "noise" in the signal.
-As their name suggests, this "random" variabiliity needs to be modeled as a variable rather than as a constant value.
+As their name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
 
-Mathematically speaking, if a normally distributed response $y$ is measured repeatedly at $t$ time points from  $i$ subjects in $k$ groups, the rm-ANOVA the model for the response $y_{kit}$ becomes:
+Mathematically speaking, a rm-ANOVA model can be written as:
 
 \begin{equation}
-\label{eq:ANOVA}
-$$y_{kit}=\mu+\gamma_k+\tau_t+(\gamma \tau)_{kt}+\pi_{i(k)}+e_{kit}$$
+  y_{i} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t} * treatment_{j}+\pi_{ij} +\epsilon_{tij}\\ (\#eq:linear-model)
 \end{equation}
 
-Where
+In this model $y_i$ is the response by subject $i$, which can be decomposed in a mean value $\mu$, _fixed effects_ of time ($time_t$), treatment ($time_j$) and their interaction; a _random effect_ $\pi_{ij}$ of each subject within each group, and errors per subject per time per group $\epsilon_{tij}$.
+Suppose two treatments groups are used in the study.
+Models per group can be generated if 0 is used for the first treatment group (Group A) and 1 for the second treatment group (Group B).
+The linear models then become:
 
-$i=1,...,n_K$ $t=1,...,T$, $k=1,...,K$; with $\pi_{ik} \sim N(0,\sigma^{2}_\epsilon)$ (independently normally distributed) and $e_{hik} \sim N(0,\sigma^{2}_\epsilon)$
+\begin{equation}
+  y = \begin{cases}
+  \mu + \beta_1*time_{t}+\epsilon_{ti}   & \mbox{if Group A}\\
+  \mu + \beta_1 * time_{t} + \beta_2 * treatment_{j} +\beta_3* time_{t} * treatment_{j}        +\pi_{i}+ \epsilon_{tij}  & \mbox{if Group B}\\
+  \end{cases}
+  (\#eq:ANOVA-by-group)
+\end{equation}
 
-In this model, $\mu$ represents the group mean, $\gamma_k$ is the _fixed effect_ of each group $k$, $\tau_t$ is the fixed effect of time, and $(\gamma \tau)_{kt}$ represents the interaction of time and group effects.
-The term $\pi_{ik}$ represents the _random effects_ for each subject within each group.
-Finally, $e_{hik}$ represents the independent random error terms, which need to be normally distributed with mean 0 [@davis2002].
-The model then, assumes a linear relationship between the response and the covariate.
-This is better represented in the follo
+Furthermore, \@ref(eq:ANOVA-by-group) can be re-written under the generalized linear model  (GLM) framework [@nelder1972].
+The main difference in this case is that a GLM model  uses a _linking function_ to connect the observed value to the linear model.
 
-_Question: How can one make a plot that tests the "limits" of the model, i.e how much "wiggliness" can this model accomodate? Make a plot to show the behavior of the model_
+Thus,
+
+\begin{equation}
+  y =\mu +\boldsymbol{\beta}_1* f(time) + \beta_2* treatment+\boldsymbol{\beta}_3* f(time)*treatment \\
+
+  y = \begin{cases}
+  \mu +\boldsymbol{\beta}_1* f(time)  & \mbox{if Group A} \\
+  \mu +\boldsymbol{\beta}_1* f(time)  + \beta_2* treatment + \boldsymbol{\beta}_3* f(time) *     treatment   & \mbox{if Group B}\\
+  \end{cases}
+  \\
+  (\#eq:GLM
+\end{equation}
+
+The linear model in \@ref(eq:linear-model) is a GLM where $f(time) = time$ with the identity function $f(x) = x$.
+
+Regardless of its presentation, \@ref(eq:linear-model), \@ref(eq:ANOVA-by-group) and \@ref(eq:GLM) show that the model expects a linear relationship between the covariates and the response.
+When the data does not follow a linear trend, the fit that the model produces does not accurately represent the changes in the data.
+To exemplify this consider simulated longitudinal data where the a normally distributed response of two groups of 10 subjects each is obtained, and an rm-ANOVA model such as \@ref(eq:linear-model) is fitted to the data (code for the simulated data is available in the Appendix).
+
+```{r, message = FALSE,include=FALSE}
+set.seed(11)
+library(patchwork)
+library(tidyverse)
+```
+
+```{r,include=FALSE}
+## Example with linear response
+example <- function(n_time = 6, 
+                    fun_type = "linear", 
+                    error_type = "correlated") {
+
+  if (!(fun_type %in% c("linear", "quadratic")))
+    stop('fun_type must be either "linear", or "quadratic"')
+  if (!(error_type %in% c("correlated", "independent")))
+    stop('fun_type must be either "correlated", or "independent"')
+
+  library(tidyverse)
+  library(mvnfast)
+  library(nlme)
+
+  x <- seq(0, 1, length.out = n_time)
+  mu <- matrix(0, length(x), 2)
+  # linear response
+  if (fun_type == "linear") {
+    mu[, 1] <- - (x - .5) + 0.25
+    mu[, 2] <- (x - .3)
+  } else {
+    # nonlinear response
+    mu[, 1] <- - 4 * (x - .5)^2 + 0.25
+    mu[, 2] <- 4 * (x - .5)^2
+  }
+  # matplot(mu, type = 'l')
+
+  y <- array(0, dim = c(length(x), 2, 10))
+  errors <- array(0, dim = c(length(x), 2, 10))
+
+  if (error_type == "independent") {
+    ## independent errors
+    for (i in 1:2) {
+      for (j in 1:10) {
+        errors[, i, j] <- rnorm(6, 0, 0.25)
+        y[, i, j] <- mu[, i] + errors[, i, j]
+      }
+    }
+  } else {
+    for (i in 1:2) {     # number of treatments
+      for (j in 1:10) {  # number of subjects
+        # compound symmetry errors
+        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
+        y[, i, j] <- mu[, i] + errors[, i, j]
+      }
+    }
+  }
+
+  ## subject random effects
+
+  ## visualizing the difference between independent errors and compound symmetry
+  ## why do we need to account for this -- overly confident inference
+
+  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
+  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
+  dimnames(mu) <- list(time = x, treatment = 1:2)
+  dat <- as.data.frame.table(y, responseName = "y")
+  dat_errors <- as.data.frame.table(errors, responseName = "errors")
+  dat_mu <- as.data.frame.table(mu, responseName = "mu")
+  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
+  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
+  dat$time <- as.numeric(as.character(dat$time))
+  dat <- dat %>%
+    mutate(subject = factor(paste(subject, treatment, sep = "-")))
+
+  ## repeated measures ANOVA in R
+  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
+  dat$preds_lm <- predict(fit_lm)
+
+  fit_lme <- lme(y ~ treatment + time + treatment:time,
+                 data = dat,
+                 random = ~ 1 | subject,
+                 correlation = corCompSymm(form = ~ 1 | subject)
+  )
+
+  pred_dat <- expand.grid(
+    treatment = factor(1:2), 
+    time = unique(dat$time)
+  )
+
+  dat$y_pred <- predict(fit_lme)
+
+  return(list(
+    dat = dat,
+    pred_dat = pred_dat,
+    fit_lm = fit_lm,
+    fit_lme = fit_lme
+
+  ))
+}
+
+plot_example <- function(sim_dat) {
+  library(patchwork)
+  ## Plot the simulated data
+  p1 <- sim_dat$dat %>%
+    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
+    geom_point() +
+    geom_line(aes(x = time, y = mu, color = treatment)) +
+    theme_bw(base_size = 12) +
+    ggtitle("Simulated data with true response function")
+
+  p2 <- sim_dat$dat %>%
+    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
+    geom_line(aes(size = "Subjects")) +
+    # facet_wrap(~ treatment) +
+    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1) +
+    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
+    ggtitle("Simulated data\nIndividual responses with population mean") +
+    theme_bw(base_size = 12)
+
+   p3 <- sim_dat$dat %>%
+    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
+    geom_line() +
+    # facet_wrap(~ treatment) +
+    ggtitle("Simulated errors") +
+    theme_bw(base_size = 12)
+
+  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
+    geom_point() +
+    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
+    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
+    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
+    theme_bw(base_size = 12) +
+    ggtitle("Fitted Model")
+  return((p1 + p3) / (p2 + p4))
+  # return((p1 + p3 + plot_layout(guides = "collect")) / (p2 + p4 + plot_layout(guides = "collect")))
+
+}
+```
+
+```{r, fig.width=16, fig.height=9, out.width="100%", echo=FALSE}
+# linear response, correlated errors (subject effect)
+plot_example(example(fun_type = "linear", error_type = "correlated")) +
+  plot_annotation(
+  title = 'Simulated with linear function and correlated errors',
+    theme = theme(plot.title = element_text(size = 20))
+) #& 
+  # theme(plot.title = element_text(size = 20))
+```
 
@@
 \begin{equation}
-(\#eq:labelGAM)
   y_{t}=\beta_0+f(x_t)+\epsilon_t  
+  (\#eq:GAM)
 \end{equation}
@@
 
-#### 1.3 Bayesian brief introduction, and compare the results of 1.2 to the results of Bayesianjo
-
-* Section 2: Implementation of both LMEMs and Bayesian and their results
-
-Present the implementation of a spline-fitted model in R, using data simulated from [@vishwanath2009]
-
 ***

### 2021-01-04 Manuscripts/Manuscript_AM_v1.tex
Updated manuscript

--- 2020-10-01 Manuscripts/Manuscript_AM_v1.tex
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+  \setmainfont[]{Times New Roman}
 \fi
@@
 \hypersetup{
-  pdftitle={Bayesian statistics for repeated measures},
-  pdfauthor={Ariel Mundo, John Tipton, Timothy Muldoon},
+  pdftitle={The statistical analysis of non-linear longitudinal data in biomedical research},
+  pdfauthor={Ariel Mundo; Timothy J. Muldoon; John R. Tipton},
   hidelinks,
@@
 
-\title{Bayesian statistics for repeated measures}
+\title{\textbf{The statistical analysis of non-linear longitudinal data in
+biomedical research}}
 \usepackage{etoolbox}
@@
 \makeatother
-\subtitle{Their application and use in biomedical research}
-\author{Ariel Mundo, John Tipton, Timothy Muldoon}
+\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
+\author{Ariel Mundo\footnote{Department of Biomedical Engineering, University of
+  Arkansas, Fayetteville} \and Timothy J. Muldoon\footnote{Department of Biomedical Engineering,
+  University of Arkansas, Fayetteville} \and John R. Tipton\footnote{Department of Mathematical Sciences, University
+  of Arkansas, Fayetteville}}
 \date{}
@@
 
+\hypertarget{paper-outline}{%
+\subsection{Paper outline}\label{paper-outline}}
+
+{\textcolor{red}{The paper Introduction has been updated, proposed sections appear at the end of the document as well as an initial graph.}}
+
 \hypertarget{background}{%
-\section{Background}\label{background}}
+\subsection{Background}\label{background}}
 
-Biomedical research often uses longitudinal studies to analyze tumor
-response {[}@Vishwanath2009{]}
+A longitudinal study is defined as that which is designed to repeatedly
+measure a variable of interest in a group (or groups) of subjects. In
+biomedical research, this type of study is employed when the intention
+is to observe the evolution of the effect of a certain treatment across
+time, rather than analyzing it at a single time point (a cross-sectional
+study). Clinical studies of breast and neck cancer have used a
+longitudinal approach {[}\protect\hyperlink{ref-sio2016}{1}{]},
+{[}\protect\hyperlink{ref-kamstra2015}{2}{]}. In the first case, weekly
+measurements of skin toxicities in patients with radiation-induced
+dermatitis were taken for up to 8 weeks; whereas in the latter mouth
+opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy
+(RT). Longitudinal studies have been also used to analyze tumor response
+{[}\protect\hyperlink{ref-roblyer2011}{3}{]}--{[}\protect\hyperlink{ref-demidov2018}{6}{]},
+antibody expression {[}\protect\hyperlink{ref-ritter2001}{7}{]},
+{[}\protect\hyperlink{ref-roth2017}{8}{]}, and cell metabolism
+{[}\protect\hyperlink{ref-jones2018}{9}{]},
+{[}\protect\hyperlink{ref-skala2010}{10}{]}.
+
+Traditionally, a ``frequentist'' approach is used in biomedical research
+to derive inferences from a longitudinal study. Such statistical view
+derives its name from the fact that it regards probability as a limiting
+frequency{[}\protect\hyperlink{ref-wagenmakers2008}{11}{]} and its
+application is based on a null hypothesis test using the \emph{analysis
+of variance over repeated measures} (repeated measures ANOVA or
+rm-ANOVA). This methodology makes two key assumptions regarding
+longitudinal data: constant correlation exists across same-subject
+measurements, and observations from each subject are obtained at all
+time points through the study (a condition also known as complete
+observations) {[}\protect\hyperlink{ref-gueorguieva2004}{12}{]},
+{[}\protect\hyperlink{ref-schober2018}{13}{]}.
+
+However, constant correlation is frequently unjustified as its value
+tends to diminish between measures when the time interval between them
+increases {[}\protect\hyperlink{ref-ugrinowitsch2004}{14}{]}. In the
+case of the number of observations during the study, different
+situations can arise that prevent the collection of complete
+measurements for all subjects: in a clinical trial voluntary withdrawal
+from one or multiple patients can occur, whereas attrition in animals
+due to injury or weight loss can occur in preclinical experiments, and
+it is even possible that unexpected complications with equipment or
+supplies arise, preventing the researcher from collecting measurements
+at a certain time point and therefore violating the \emph{complete
+observations} assumption of rm-ANOVA.
+
+When incomplete observations occur, an rm-ANOVA requires the exclusion
+of all subjects with missing observations from the analysis. This can
+result in increased costs if the desired statistical power is not met
+with the remaining observations, as it would be necessary to enroll more
+subjects. At the same time, if the excluded observations contain
+insightful information that is not used, their elimination from the
+analysis may limit the demonstration of significant differences between
+groups. Additionally, rm-ANOVA uses a \emph{post hoc} analysis to assess
+differences between the measured response in different groups. A
+\emph{post hoc} analysis is based on multiple repeated comparisons to
+estimate a \emph{p-value}, a metric that is widely used as a measure of
+significance. Because the \emph{p-value} is highly variable, multiple
+comparisons can inflate the false positivity rate
+{[}\protect\hyperlink{ref-liu2010}{15}{]},
+{[}\protect\hyperlink{ref-halsey2015}{16}{]}, consequently biasing the
+conclusions of the study.
+
+During the last decade, the biomedical community has started to
+recognize the limitations of rm-ANOVA in the analysis of longitudinal
+information. This is exemplified by the use of linear mixed effects
+models (LMEMs) in certain groups to analyze longitudinal data
+{[}\protect\hyperlink{ref-skala2010}{10}{]},
+{[}\protect\hyperlink{ref-vishwanath2009}{17}{]}. Briefly, these models
+incorporate \emph{fixed effects}, which correspond to the levels of
+experimental factors in the study (e.g.~the different drug regimens in a
+clinical trial), and \emph{random effects}, which account for random
+variation within the population. These models are more flexible than the
+traditional rm-ANOVA as they can accommodate missing observations for
+multiple subjects and allow different modeling strategies for the
+variability within each measure in every subject
+{[}\protect\hyperlink{ref-pinheiro2006}{18}{]}. On the other hand, they
+impose restrictions in the distribution of the errors of the random
+effects {[}\protect\hyperlink{ref-gueorguieva2004}{12}{]}.
+
+One final assumption that is not initially evident for both rm-ANOVA and
+LMEMs models pertains the relationship between the observed effect and
+the progression of the timeline of the study, which is expected to be
+linear {[}\protect\hyperlink{ref-pinheiro2006}{18}{]}. This common
+assumption in both rm-ANOVA ane LMEMs consequently restricts the
+inferences from the model when the data does not follow a linear trend.
+In biomedical research, a special case of this behavior in longitudinal
+data arises in measurements of tumor response in preclinical and
+clinical settings {[}\protect\hyperlink{ref-roblyer2011}{3}{]},
+{[}\protect\hyperlink{ref-skala2010}{10}{]},
+{[}\protect\hyperlink{ref-vishwanath2009}{17}{]}. These studies have
+shown that the collected signal does not follow a linear trend over
+time, and presents extreme variability at different time points, making
+the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent
+with the observed variation. In other words, the in both cases the model
+tries to explain highly-variable data using a linear trend, consequently
+biasing the estimates. Additionally, although it is possible that a
+\emph{post hoc} analysis is able to find ``significant''
+\emph{p-values}( \emph{p}\textless0.05) by using multiple comparisons
+between the model terms, the use of this estimator is limited in such
+scenario because it is intended to work in models that have a reasonable
+agreement with the data.
+
+As the ``frequentist'' rm-ANOVA and the more advanced LMEM approach are
+both limited in the analysis of non-linear longitudinal information,
+there is a need for biomedical researchers to explore the use of
+additional statistical tools that allow the information (and not an
+assumed trend) to determine the fit of the model, while enabling
+inferences that are both adequate and consistent from a statistical
+perspective. In this regard, Generalized Additive Models (GAMs) present
+an alternative approach to analyze longitudinal data. Although not
+commonly used in the biomedical community, these non-parametric models
+have been used to analyze temporal variations in geochemical and
+palaeoecological data
+{[}\protect\hyperlink{ref-rose2012}{19}{]}--{[}\protect\hyperlink{ref-simpson2018}{21}{]},
+health-environment interactions
+{[}\protect\hyperlink{ref-yang2012}{22}{]} and political science
+{[}\protect\hyperlink{ref-beck1998}{23}{]} . There are several
+advantages of GAMs over LMEMs and rm-ANOVA models: They have an
+extensively developed theoretical background, enable the data to dictate
+the trend of the model, can accommodate missing observations and do not
+require constant correlation between repeated measurements
+{[}\protect\hyperlink{ref-wood2017}{24}{]}.Therefore, GAMs can provide a
+more suitable statistical method to analyze non-linear biomedical
+longitudinal data.
+
+The current advances in programming languages designed for statistical
+analysis (specifically \(\textsf{R}\)), have eased the computational
+implementation of more complex models beyond LMEMs. In particular,
+\(\textsf{R}\) has an extensive collection of documentation and
+functions to fit GAMs in the package \emph{mgcv}
+{[}\protect\hyperlink{ref-wood2017}{24}{]}--{[}\protect\hyperlink{ref-wood2016}{26}{]}
+that not only speed up the initial stages of the analysis but enable the
+use of advanced modeling structures (e.g.~hierarchical models,
+confidence interval comparisons) without requiring advanced programming
+skills from the user. This programming language is also able to simulate
+data, an emerging strategy used to test statistical models
+{[}\protect\hyperlink{ref-haverkamp2017}{27}{]}. This allows to create
+and explore different alternatives for analysis without collecting
+information in the field, and reduces the time window between experiment
+design and its implementation.
+
+Therefore, the purpose of this study is to provide biomedical
+researchers with a clear understanding of the theory and the practical
+implementation of GAMs to analyze longitudinal data using by focusing on
+four areas. First, the limitations of rm-ANOVA and LMEMs to analyze
+longitudinal data are contrasted in the cases of missing observations,
+assumption of linearity and correlation structures. Secondly, the key
+theoretical elements of GAMS are presented without using extensive
+derivations or complex mathematical notation that can obscure their
+understanding. Thirdly, simulated data that follows the trend of
+previously reported values
+{[}\protect\hyperlink{ref-vishwanath2009}{17}{]} is used to illustrate
+the type of non-linear longitudinal data that occurs in biomedical
+research and the implementation of GAMs in such scenario. Finally,
+reproducibility is emphasized by providing the code to generate the data
+and the implementation of GAMs in \(\textsf{R}\), in conjunction with a
+step-by-step guide to fit models of increasing complexity.\\
+In summary, the exploration of modern statistical techniques to analyze
+longitudinal data may allow biomedical researchers to consider the use
+of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a
+linear trend, and will also help to improve the standards for
+reproducibility in biomedical research.
+
+\hypertarget{section-1}{%
+\subsection{Section 1:}\label{section-1}}
+
+\hypertarget{challenges-presented-by-longitudinal-studies}{%
+\subsubsection{Challenges presented by longitudinal
+studies}\label{challenges-presented-by-longitudinal-studies}}
+
+\hypertarget{the-frequentist-case-for-longitudinal-data}{%
+\subsubsection{1 The ``frequentist'' case for longitudinal
+data}\label{the-frequentist-case-for-longitudinal-data}}
+
+The \emph{repeated measures analysis of variance} (rm-ANOVA) is the
+standard statistical analysis for longitudinal data in biomedical
+research, but certain assumptions are necessary to make the model valid.
+From a practical view, they can be divided in three areas: linear
+relationship between covariates and response, constant correlation
+between measurements, and complete observations for all subjects. Each
+one of these assumptions is discussed below.
+
+\hypertarget{linear-relationship}{%
+\paragraph{1.1 Linear relationship}\label{linear-relationship}}
+
+In a biomedical longitudinal study, two or more groups of subjects
+(patients, mice, samples) are subject to a different treatments (e.g.~a
+``treatment'' group receives a novel drug vs.~a ``control'' group that
+receives a placebo), and measurements from each subject within each
+group are collected at specific time points. The collected response has
+both \emph{fixed} and \emph{random} components. The \emph{fixed}
+component can be understood as a constant value in the response which
+the researcher is interested in measuring, i.e, the effect of the novel
+drug in the ``treatment'' group.The \emph{random} component can be
+defined as ``noise'' caused by some inherent variability within the
+study. For example, if the blood concentration of the drug is measured
+in certain subjects in the early hours of the morning while others are
+measured in the afternoon, it is possible that the difference in the
+collection time of the measurement introduces some ``noise'' in the
+signal. As their name suggests, this ``random'' variability needs to be
+modeled as a variable rather than as a constant value.
+
+Mathematically speaking, a rm-ANOVA model can be written as:
+
+\begin{equation}
+  y_{i} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t} * treatment_{j}+\pi_{ij} +\epsilon_{tij}\\ (\#eq:linear-model)
+\end{equation}
+
+In this model \(y_i\) is the response by subject \(i\), which can be
+decomposed in a mean value \(\mu\), \emph{fixed effects} of time
+(\(time_t\)), treatment (\(time_j\)) and their interaction; a
+\emph{random effect} \(\pi_{ij}\) of each subject within each group, and
+errors per subject per time per group \(\epsilon_{tij}\). Suppose two
+treatments groups are used in the study. Models per group can be
+generated if 0 is used for the first treatment group (Group A) and 1 for
+the second treatment group (Group B). The linear models then become:
+
+\begin{equation}
+  y = \begin{cases}
+  \mu + \beta_1*time_{t}+\epsilon_{ti}   & \mbox{if Group A}\\
+  \mu + \beta_1 * time_{t} + \beta_2 * treatment_{j} +\beta_3* time_{t} * treatment_{j}        +\pi_{i}+ \epsilon_{tij}  & \mbox{if Group B}\\
+  \end{cases}
+  (\#eq:ANOVA-by-group)
+\end{equation}
+
+Furthermore, @ref(eq:ANOVA-by-group) can be re-written under the
+generalized linear model (GLM) framework
+{[}\protect\hyperlink{ref-nelder1972}{28}{]}. The main difference in
+this case is that a GLM model uses a \emph{linking function} to connect
+the observed value to the linear model.
+
+Thus,
+
+\begin{equation}
+  y =\mu +\boldsymbol{\beta}_1* f(time) + \beta_2* treatment+\boldsymbol{\beta}_3* f(time)*treatment \\
+
+  y = \begin{cases}
+  \mu +\boldsymbol{\beta}_1* f(time)  & \mbox{if Group A} \\
+  \mu +\boldsymbol{\beta}_1* f(time)  + \beta_2* treatment + \boldsymbol{\beta}_3* f(time) *     treatment   & \mbox{if Group B}\\
+  \end{cases}
+  \\
+  (\#eq:GLM
+\end{equation}
+
+The linear model in @ref(eq:linear-model) is a GLM where
+\(f(time) = time\) with the identity function \(f(x) = x\).
+
+Regardless of its presentation, @ref(eq:linear-model),
+@ref(eq:ANOVA-by-group) and @ref(eq:GLM) show that the model expects a
+linear relationship between the covariates and the response. When the
+data does not follow a linear trend, the fit that the model produces
+does not accurately represent the changes in the data. To exemplify this
+consider simulated longitudinal data where the a normally distributed
+response of two groups of 10 subjects each is obtained, and an rm-ANOVA
+model such as @ref(eq:linear-model) is fitted to the data (code for the
+simulated data is available in the Appendix).
+
+\begin{verbatim}
+## 
+## Attaching package: 'nlme'
+\end{verbatim}
+
+\begin{verbatim}
+## The following object is masked from 'package:dplyr':
+## 
+##     collapse
+\end{verbatim}
+
+\includegraphics[width=1\linewidth]{Manuscript_AM_v1_files/figure-latex/unnamed-chunk-3-1}
+
+\hypertarget{covariance-in-rm-anova-and-lmems}{%
+\paragraph{1.2 Covariance in rm-ANOVA and
+LMEMs}\label{covariance-in-rm-anova-and-lmems}}
+
+In a longitudinal study there is an expected \emph{variance} between
+repeated measurements on the same subject, and because repeated measures
+occur in the subjects within each group, there is a \emph{covariance}
+between measurements at each time point within each group. The
+\emph{covariance matrix} (also known as the variance-covariance matrix)
+is a matrix that captures the variation between and within subjects in a
+longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{29}{]} (For
+an in-depth analysis of the covariance matrix see
+{[}\protect\hyperlink{ref-west2014}{30}{]},
+{[}\protect\hyperlink{ref-weiss2005}{31}{]}).
+
+In the case of an rm-ANOVA analysis, it is assumed that the covariance
+matrix has a specific construction known as \emph{compound symmetry}
+(also known as ``sphericity'' or ``circularity''). Under this
+assumption, the between-subject variance and within-subject correlation
+are constant across time
+{[}\protect\hyperlink{ref-weiss2005}{31}{]}--{[}\protect\hyperlink{ref-huynh1976}{33}{]}.
+However, it has been shown that this condition is frequently unjustified
+because the correlation between measurements tends to change over time
+{[}\protect\hyperlink{ref-maxwell2017}{34}{]}; and it is higher between
+consecutive measurements
+{[}\protect\hyperlink{ref-gueorguieva2004}{12}{]},
+{[}\protect\hyperlink{ref-ugrinowitsch2004}{14}{]}. Although corrections
+can be made (such as Huyhn-Feldt or Greenhouse-Geisser) the
+effectiveness of each correction is limited because it depends on the
+size of the sample,the number of repeated
+measurements{[}\protect\hyperlink{ref-haverkamp2017}{27}{]}, and they
+are not robust if the group sizes are unbalanced
+{[}\protect\hyperlink{ref-keselman2001}{35}{]}. In other words, if the
+data does not present constant correlation between repeated
+measurements, the assumptions required for an rm-ANOVA model are not met
+and the use of corrections may still not provide a reasonable adjustment
+that makes the model valid.
+
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow
+different structures for the variance-covariance matrix including
+exponential, autoregressive of order 1, rational quadratic and others
+{[}\protect\hyperlink{ref-pinheiro2006}{18}{]}. Nevertheless, the
+analysis required to determine an appropriate variance-covariance
+structure for the data can be a long process by itself. Overall, the
+spherical assumption for rm-ANOVA may not capture the natural variations
+of the correlation in the data, and can bias the inferences from the
+analysis.
+
+\hypertarget{missing-observations}{%
+\paragraph{1.3 Missing observations}\label{missing-observations}}
+
+Missing observations are an issue that arises frequently in longitudinal
+studies. In biomedical research, this situation can be caused by reasons
+beyond the control of the investigator {[}molenberghs2004{]}.Dropout
+from patients, or attrition or injury in animals are among the reasons
+for missing observations. Statistically, missing information can be
+classified as \emph{missing at random} (MAR), \emph{missing completely
+at random} (MCAR), and \emph{missing not at random} (MNAR)
+{[}\protect\hyperlink{ref-weiss2005}{31}{]}. In a MAR scenario, the
+pattern of the missing information is related to some variable in the
+data, but it is not related to the variable of interest
+{[}\protect\hyperlink{ref-scheffer2002}{36}{]}. If the data are MCAR,
+this means that the missigness is completely unrelated to the collected
+information {[}\protect\hyperlink{ref-potthoff2006}{37}{]}, and in the
+case of MNAR the missing values are dependent on their value. An
+rm-ANOVA model assumes complete observations for all subjects, and
+therefore subjects with one or more missing observations are excluded
+from the analysis. This is inconvenient because the remaining subjects
+might not accurately represent the population, and statistical power is
+affected by this reduction in sample size
+{[}\protect\hyperlink{ref-ma2012}{38}{]}.
+
+In the case of LMEMs, inferences from the model are valid when missing
+observations in the data exist that are MAR
+{[}\protect\hyperlink{ref-west2014}{30}{]}. The pattern of missing
+observations can be considered MAR if the missing observations are not
+related any of the other variables measured in the study
+{[}\protect\hyperlink{ref-maxwell2017}{34}{]}. For example, if attrition
+occurs in all mice that had lower weights at the beginning of a
+chemotherapy response study, the missing data can be considered MAR
+because the missigness is unrelated to other variables of interest.
+
+\emph{What limitations LMEMs have regarding missing observations?}
+
+This section has presented the assumptions of rm-ANOVA to analyze
+longitudinal information and its differences when compared to LMEMs
+regarding to missing data and the modeling of the covariance matrix. Of
+notice, LMEMs offer a more robust and flexible approach than rm-ANOVA
+and if the data follows a linear trend, they provide an excellent choice
+to derive inferences from a repeated measures study. However, when the
+data presents high variability, LMEMs fail to capture the non-linear
+trend of the data. To analyze such type of data, we present generalized
+additive models (GAMs) as an alternative in the following section.
+
+\hypertarget{gams-as-a-special-case-of-generalized-linear-models}{%
+\paragraph{2 GAMs as a special case of Generalized Linear
+Models}\label{gams-as-a-special-case-of-generalized-linear-models}}
+
+A GAM is a special case of the Generalized Linear Model (GLM), a
+framework that allows for response distributions that do not follow a
+normal distribution {[}\protect\hyperlink{ref-wood2017}{24}{]},
+{[}\protect\hyperlink{ref-hastie1987}{39}{]}. Following the notation by
+Simpson {[}\protect\hyperlink{ref-simpson2018}{21}{]} A GAM model can be
+represented as:
+
+\begin{equation}
+  y_{t}=\beta_0+f(x_t)+\epsilon_t  
+  (\#eq:GAM)
+\end{equation}
+
+Where \(y_t\) is the response at time \(t\), \(\beta_0\) is the expected
+value at time 0, the change of \(y\) over time is represented by the
+function \(f(x_t)\) and \(\epsilon_t\) represents the residuals.
+
+In contrast to rm-ANOVA or LMEMs, GAMs use \emph{smooth functions} to
+model the relationship between the covariates and the response. This
+approach is more advantageous than that of a LMEMs or rm-ANOVA as it
+does not restrict the model to a linear relationship. One possible
+function for \(f(x_t)\) is a polynomial, but a major limitation is that
+they create a ``global'' fit as they assume that the same relationship
+exists everywhere, which can cause problems with the fit
+{[}\protect\hyperlink{ref-beck1998}{23}{]}.
+
+To overcome this limitation, the smooth functions in GAMs are
+represented using \emph{basis functions} over evenly spaced ranges of
+the covariates known as \emph{knots}. The \emph{basis function} used is
+a cubic spline, which is a smooth curve constructed from cubic
+polynomials joined
+together{[}\protect\hyperlink{ref-simpson2018}{21}{]},
+{[}\protect\hyperlink{ref-wood2017}{24}{]}. Cubic splines have a long
+history in their use to solve non-parametric statistical problems and
+are the default choice to fit GAMs as they are the simplest option to
+obtain visual smoothness {[}\protect\hyperlink{ref-wegman1983}{40}{]}.
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-sio2016}{}%
+{[}1{]} T. T. Sio \emph{et al.}, ``Repeated measures analyses of
+dermatitis symptom evolution in breast cancer patients receiving
+radiotherapy in a phase 3 randomized trial of mometasone furoate vs
+placebo (n06c4 {[}alliance{]}),'' \emph{Supportive Care in Cancer}, vol.
+24, no. 9, pp. 3847--3855, 2016.
+
+\leavevmode\hypertarget{ref-kamstra2015}{}%
+{[}2{]} J. Kamstra, P. Dijkstra, M. Van Leeuwen, J. Roodenburg, and J.
+Langendijk, ``Mouth opening in patients irradiated for head and neck
+cancer: A prospective repeated measures study,'' \emph{Oral Oncology},
+vol. 51, no. 5, pp. 548--555, 2015.
+
+\leavevmode\hypertarget{ref-roblyer2011}{}%
+{[}3{]} D. Roblyer \emph{et al.}, ``Optical imaging of breast cancer
+oxyhemoglobin flare correlates with neoadjuvant chemotherapy response
+one day after starting treatment,'' \emph{Proceedings of the National
+Academy of Sciences}, vol. 108, no. 35, pp. 14626--14631, 2011.
+
+\leavevmode\hypertarget{ref-tank2020}{}%
+{[}4{]} A. Tank \emph{et al.}, ``Diffuse optical spectroscopic imaging
+reveals distinct early breast tumor hemodynamic responses to metronomic
+and maximum tolerated dose regimens,'' \emph{Breast cancer research},
+vol. 22, no. 1, pp. 1--10, 2020.
+
+\leavevmode\hypertarget{ref-pavlov2018}{}%
+{[}5{]} M. V. Pavlov \emph{et al.}, ``Multimodal approach in assessment
+of the response of breast cancer to neoadjuvant chemotherapy,''
+\emph{Journal of biomedical optics}, vol. 23, no. 9, p. 091410, 2018.
+
+\leavevmode\hypertarget{ref-demidov2018}{}%
+{[}6{]} V. Demidov \emph{et al.}, ``Preclinical longitudinal imaging of
+tumor microvascular radiobiological response with functional optical
+coherence tomography,'' \emph{Scientific reports}, vol. 8, no. 1, pp.
+1--12, 2018.
+
+\leavevmode\hypertarget{ref-ritter2001}{}%
+{[}7{]} G. Ritter, L. S. Cohen, C. Williams, E. C. Richards, L. J. Old,
+and S. Welt, ``Serological analysis of human anti-human antibody
+responses in colon cancer patients treated with repeated doses of
+humanized monoclonal antibody a33,'' \emph{Cancer Research}, vol. 61,
+no. 18, pp. 6851--6859, 2001.
+
+\leavevmode\hypertarget{ref-roth2017}{}%
+{[}8{]} E. M. Roth \emph{et al.}, ``Antidrug antibodies in patients
+treated with alirocumab,'' 2017.
+
+\leavevmode\hypertarget{ref-jones2018}{}%
+{[}9{]} J. D. Jones, H. E. Ramser, A. E. Woessner, and K. P. Quinn, ``In
+vivo multiphoton microscopy detects longitudinal metabolic changes
+associated with delayed skin wound healing,'' \emph{Communications
+biology}, vol. 1, no. 1, pp. 1--8, 2018.
+
+\leavevmode\hypertarget{ref-skala2010}{}%
+{[}10{]} M. C. Skala, A. N. Fontanella, L. Lan, J. A. Izatt, and M. W.
+Dewhirst, ``Longitudinal optical imaging of tumor metabolism and
+hemodynamics,'' \emph{Journal of biomedical optics}, vol. 15, no. 1, p.
+011112, 2010.
+
+\leavevmode\hypertarget{ref-wagenmakers2008}{}%
+{[}11{]} E.-J. Wagenmakers, M. Lee, T. Lodewyckx, and G. J. Iverson,
+``Bayesian versus frequentist inference,'' in \emph{Bayesian evaluation
+of informative hypotheses}, Springer, 2008, pp. 181--207.
+
+\leavevmode\hypertarget{ref-gueorguieva2004}{}%
+{[}12{]} R. Gueorguieva and J. H. Krystal, ``Move over anova: Progress
+in analyzing repeated-measures data andits reflection in papers
+published in the archives of general psychiatry,'' \emph{Archives of
+general psychiatry}, vol. 61, no. 3, pp. 310--317, 2004.
+
+\leavevmode\hypertarget{ref-schober2018}{}%
+{[}13{]} P. Schober and T. R. Vetter, ``Repeated measures designs and
+analysis of longitudinal data: If at first you do not succeed---try, try
+again,'' \emph{Anesthesia and analgesia}, vol. 127, no. 2, p. 569, 2018.
+
+\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
+{[}14{]} C. Ugrinowitsch, G. W. Fellingham, and M. D. Ricard,
+``Limitations of ordinary least squares models in analyzing repeated
+measures data,'' \emph{Medicine and science in sports and exercise},
+vol. 36, pp. 2144--2148, 2004.
+
+\leavevmode\hypertarget{ref-liu2010}{}%
+{[}15{]} C. Liu, T. P. Cripe, and M.-O. Kim, ``Statistical issues in
+longitudinal data analysis for treatment efficacy studies in the
+biomedical sciences,'' \emph{Molecular Therapy}, vol. 18, no. 9, pp.
+1724--1730, 2010.
+
+\leavevmode\hypertarget{ref-halsey2015}{}%
+{[}16{]} L. G. Halsey, D. Curran-Everett, S. L. Vowler, and G. B.
+Drummond, ``The fickle p value generates irreproducible results,''
+\emph{Nature methods}, vol. 12, no. 3, pp. 179--185, 2015.
+
+\leavevmode\hypertarget{ref-vishwanath2009}{}%
+{[}17{]} K. Vishwanath, H. Yuan, W. T. Barry, M. W. Dewhirst, and N.
+Ramanujam, ``Using optical spectroscopy to longitudinally monitor
+physiological changes within solid tumors,'' \emph{Neoplasia}, vol. 11,
+no. 9, pp. 889--900, 2009.
+
+\leavevmode\hypertarget{ref-pinheiro2006}{}%
+{[}18{]} J. Pinheiro and D. Bates, \emph{Mixed-effects models in s and
+s-plus}. Springer Science \& Business Media, 2006.
+
+\leavevmode\hypertarget{ref-rose2012}{}%
+{[}19{]} N. L. Rose, H. Yang, S. D. Turner, and G. L. Simpson, ``An
+assessment of the mechanisms for the transfer of lead and mercury from
+atmospherically contaminated organic soils to lake sediments with
+particular reference to scotland, uk,'' \emph{Geochimica et Cosmochimica
+Acta}, vol. 82, pp. 113--135, 2012.
+
+\leavevmode\hypertarget{ref-pedersen2019}{}%
+{[}20{]} E. J. Pedersen, D. L. Miller, G. L. Simpson, and N. Ross,
+``Hierarchical generalized additive models in ecology: An introduction
+with mgcv,'' \emph{PeerJ}, vol. 7, p. e6876, 2019.
+
+\leavevmode\hypertarget{ref-simpson2018}{}%
+{[}21{]} G. L. Simpson, ``Modelling palaeoecological time series using
+generalised additive models,'' \emph{Frontiers in Ecology and
+Evolution}, vol. 6, p. 149, 2018.
+
+\leavevmode\hypertarget{ref-yang2012}{}%
+{[}22{]} L. Yang, G. Qin, N. Zhao, C. Wang, and G. Song, ``Using a
+generalized additive model with autoregressive terms to study the
+effects of daily temperature on mortality,'' \emph{BMC Medical Research
+Methodology}, vol. 12, no. 1, p. 165, 2012.
+
+\leavevmode\hypertarget{ref-beck1998}{}%
+{[}23{]} N. Beck and S. Jackman, ``Beyond linearity by default:
+Generalized additive models,'' \emph{American Journal of Political
+Science}, pp. 596--627, 1998.
+
+\leavevmode\hypertarget{ref-wood2017}{}%
+{[}24{]} S. N. Wood, \emph{Generalized additive models: An introduction
+with r}. CRC press, 2017.
+
+\leavevmode\hypertarget{ref-wood2011}{}%
+{[}25{]} S. N. Wood, N., Pya, and B. S"afken, ``Smoothing parameter and
+model selection for general smooth models (with discussion),''
+\emph{Journal of the American Statistical Association}, vol. 111, pp.
+1548--1575, 2016.
+
+\leavevmode\hypertarget{ref-wood2016}{}%
+{[}26{]} S. N. Wood, N., Pya, and B. S"afken, ``Smoothing parameter and
+model selection for general smooth models (with discussion),''
+\emph{Journal of the American Statistical Association}, vol. 111, pp.
+1548--1575, 2016.
+
+\leavevmode\hypertarget{ref-haverkamp2017}{}%
+{[}27{]} N. Haverkamp and A. Beauducel, ``Violation of the sphericity
+assumption and its effect on type-i error rates in repeated measures
+anova and multi-level linear models (mlm),'' \emph{Frontiers in
+psychology}, vol. 8, p. 1841, 2017.
+
+\leavevmode\hypertarget{ref-nelder1972}{}%
+{[}28{]} J. A. Nelder and R. W. Wedderburn, ``Generalized linear
+models,'' \emph{Journal of the Royal Statistical Society: Series A
+(General)}, vol. 135, no. 3, pp. 370--384, 1972.
+
+\leavevmode\hypertarget{ref-wolfinger1996}{}%
+{[}29{]} R. D. Wolfinger, ``Heterogeneous variance: Covariance
+structures for repeated measures,'' \emph{Journal of agricultural,
+biological, and environmental statistics}, pp. 205--230, 1996.
+
+\leavevmode\hypertarget{ref-west2014}{}%
+{[}30{]} B. T. West, K. B. Welch, and A. T. Galecki, \emph{Linear mixed
+models: A practical guide using statistical software}. CRC Press, 2014.
+
+\leavevmode\hypertarget{ref-weiss2005}{}%
+{[}31{]} R. E. Weiss, \emph{Modeling longitudinal data}. Springer
+Science \& Business Media, 2005.
+
+\leavevmode\hypertarget{ref-geisser1958}{}%
+{[}32{]} S. Geisser, S. W. Greenhouse, and others, ``An extension of
+box's results on the use of the \(F\) distribution in multivariate
+analysis,'' \emph{The Annals of Mathematical Statistics}, vol. 29, no.
+3, pp. 885--891, 1958.
+
+\leavevmode\hypertarget{ref-huynh1976}{}%
+{[}33{]} H. Huynh and L. S. Feldt, ``Estimation of the box correction
+for degrees of freedom from sample data in randomized block and
+split-plot designs,'' \emph{Journal of educational statistics}, vol. 1,
+no. 1, pp. 69--82, 1976.
+
+\leavevmode\hypertarget{ref-maxwell2017}{}%
+{[}34{]} S. E. Maxwell, H. D. Delaney, and K. Kelley, \emph{Designing
+experiments and analyzing data: A model comparison perspective}.
+Routledge, 2017.
+
+\leavevmode\hypertarget{ref-keselman2001}{}%
+{[}35{]} H. Keselman, J. Algina, and R. K. Kowalchuk, ``The analysis of
+repeated measures designs: A review,'' \emph{British Journal of
+Mathematical and Statistical Psychology}, vol. 54, no. 1, pp. 1--20,
+2001.
+
+\leavevmode\hypertarget{ref-scheffer2002}{}%
+{[}36{]} J. Scheffer, ``Dealing with missing data,'' 2002.
+
+\leavevmode\hypertarget{ref-potthoff2006}{}%
+{[}37{]} R. F. Potthoff, G. E. Tudor, K. S. Pieper, and V. Hasselblad,
+``Can one assess whether missing data are missing at random in medical
+studies?'' \emph{Statistical methods in medical research}, vol. 15, no.
+3, pp. 213--234, 2006.
+
+\leavevmode\hypertarget{ref-ma2012}{}%
+{[}38{]} Y. Ma, M. Mazumdar, and S. G. Memtsoudis, ``Beyond
+repeated-measures analysis of variance: Advanced statistical methods for
+the analysis of longitudinal data in anesthesia research,''
+\emph{Regional Anesthesia \& Pain Medicine}, vol. 37, no. 1, pp.
+99--105, 2012.
+
+\leavevmode\hypertarget{ref-hastie1987}{}%
+{[}39{]} T. Hastie and R. Tibshirani, ``Generalized additive models:
+Some applications,'' \emph{Journal of the American Statistical
+Association}, vol. 82, no. 398, pp. 371--386, 1987.
+
+\leavevmode\hypertarget{ref-wegman1983}{}%
+{[}40{]} E. J. Wegman and I. W. Wright, ``Splines in statistics,''
+\emph{Journal of the American Statistical Association}, vol. 78, no.
+382, pp. 351--365, 1983.
 

### 2021-01-04 rm-ANOVA equation.Rmd
Updated manuscript

---
title: "rm-ANOVA"
author: "Ariel Mundo"
date: "1/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\begin{equation}
\label{eq:ANOVA}
$$ y_{kit}=\mu+\gamma_k+\tau_t+(\gamma \tau)_{kt}+\pi_{i(k)}+e_{kit}$$
\end{equation}

Where

$i=1,...,n_K$ $t=1,...,T$, $k=1,...,K$; with $\pi_{ik} \sim N(0,\sigma^{2}_\epsilon)$ (independently normally distributed) and $e_{hik} \sim N(0,\sigma^{2}_\epsilon)$

In this model, $\mu$ represents the group mean, $\gamma_k$ is the _fixed effect_ of each group $k$, $\tau_t$ is the fixed effect of time, and $(\gamma \tau)_{kt}$ represents the interaction of time and group effects. The term $\pi_{ik}$ represents the _random effects_ for each subject within each group. Finally, $e_{hik}$ represents the independent random error terms, which need to be normally distributed with mean 0 [@davis2002]. The model then, assumes a linear relationship between the response and the covariate.

### 2021-01-05 Manuscripts/Manuscript_AM_v1.Rmd
Updated manuscript, html_2 document works now, need to add non-linear anova graph and text and finish summary of GAMs

@@
 - Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
-- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
+- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
+  Fayetteville]
 - John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
+subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
+bibliography: refs.bib
+csl: ieee.csl
+css: style.css
+output:
 
-bibliography: refs.bib
-output:
-  pdf_document: 
+  bookdown::html_document2: default
+  bookdown::pdf_document2:
+    keep_tex: yes
     fig_caption: yes
-  html_document: default
-  css: style.css
-  word_document:
-    reference_docx: docx_template.docx
-mainfont: Times New Roman
-link-citations: yes
-subtitle: '_Beyond repeated measures ANOVA and Linear Mixed Models_'
-csl: ieee.csl
+  bookdown::word_document2: default
+  link-citations: true
 ---
 
-## Paper outline
-<span style="color: red;">\textcolor{red}{The paper Introduction has been updated, proposed sections appear at the end of the document as well as an initial graph.}</span>
-
-## Background
+# Background
 
@@
 
-## Section 1:
+# Challenges presented by longitudinal studies
 
-### Challenges presented by longitudinal studies
-
-### 1 The "frequentist" case for longitudinal data
+## The "frequentist" case for longitudinal data
 
@@
 
-#### 1.1 Linear relationship
+## Linear relationship
 
@@
 \begin{equation}
-  y_{i} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t} * treatment_{j}+\pi_{ij} +\epsilon_{tij}\\ (\#eq:linear-model)
+  y_{i} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t} *                                    treatment_{j}+\pi_{ij} +\epsilon_{tij}\\ 
+  (\#eq:linear-model)
 \end{equation}
@@
   \mu + \beta_1*time_{t}+\epsilon_{ti}   & \mbox{if Group A}\\
-  \mu + \beta_1 * time_{t} + \beta_2 * treatment_{j} +\beta_3* time_{t} * treatment_{j}        +\pi_{i}+ \epsilon_{tij}  & \mbox{if Group B}\\
+  \mu + \beta_1 * time_{t} + \beta_2 * treatment_{j} +\beta_3* time_{t} * treatment_{j}+\pi_{i}+\epsilon_{tij}  & \mbox{if Group B}\\
   \end{cases}
@@
   \end{cases}
-  \\
-  (\#eq:GLM
+  (\#eq:GLM)
 \end{equation}
@@
 
-```{r,include=FALSE}
+```{r,include=FALSE,message=FALSE,echo=FALSE}
 ## Example with linear response
@@
 
-```{r, fig.width=16, fig.height=9, out.width="100%", echo=FALSE}
+```{r, fig.width=16, fig.height=9, out.width="100%", echo=FALSE,message=FALSE}
 # linear response, correlated errors (subject effect)
@@
 ```
+It is clear from this graph that
 
-#### 1.2 Covariance in rm-ANOVA and LMEMs
+```{r, fig.width=16, fig.height=9, out.width="100%"}
+# quadratic response, correlated errors (subject effect)
+plot_example(example(fun_type = "quadratic", error_type = "correlated")) +
+  plot_annotation(
+  title = 'Simulated with quadratic function and correlated errors',
+    theme = theme(plot.title = element_text(size = 20))
+) #& 
+  # theme(plot.title = element_text(size = 20))
+```
+
+## Covariance in rm-ANOVA and LMEMs
 
@@
 
-#### 1.3 Missing observations
+## Missing observations
 
@@
 
-#### 2 GAMs as a special case of Generalized Linear Models
+# GAMs as a special case of Generalized Linear Models
 

### 2021-01-05 Manuscripts/Manuscript_AM_v1.tex
Updated manuscript, html_2 document works now, need to add non-linear anova graph and text and finish summary of GAMs

@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-  \setmainfont[]{Times New Roman}
 \fi
@@
 \usepackage[margin=1in]{geometry}
+\usepackage{longtable,booktabs}
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx,grffile}
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 
-\title{\textbf{The statistical analysis of non-linear longitudinal data in
-biomedical research}}
-\usepackage{etoolbox}
-\makeatletter
-\providecommand{\subtitle}[1]{% add subtitle to \maketitle
-  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
-}
-\makeatother
-\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
-\author{Ariel Mundo\footnote{Department of Biomedical Engineering, University of
-  Arkansas, Fayetteville} \and Timothy J. Muldoon\footnote{Department of Biomedical Engineering,
-  University of Arkansas, Fayetteville} \and John R. Tipton\footnote{Department of Mathematical Sciences, University
-  of Arkansas, Fayetteville}}
+\title{\textbf{The statistical analysis of non-linear longitudinal data in biomedical research}}
+\author{Ariel Mundo\footnote{Department of Biomedical Engineering, University of Arkansas, Fayetteville} \and Timothy J.
+Muldoon\footnote{Department of Biomedical Engineering, University of Arkansas, Fayetteville} \and John R.
+Tipton\footnote{Department of Mathematical Sciences, University of Arkansas, Fayetteville}}
 \date{}
@@
 
-\hypertarget{paper-outline}{%
-\subsection{Paper outline}\label{paper-outline}}
+{
+\setcounter{tocdepth}{2}
+\tableofcontents
+}
+\hypertarget{background}{%
+\section{Background}\label{background}}
 
-{\textcolor{red}{The paper Introduction has been updated, proposed sections appear at the end of the document as well as an initial graph.}}
+A longitudinal study is defined as that which is designed to repeatedly measure a variable of interest in a group (or groups) of subjects.
+In biomedical research, this type of study is employed when the intention is to observe the evolution of the effect of a certain treatment across time, rather than analyzing it at a single time point (a cross-sectional study).
+Clinical studies of breast and neck cancer have used a longitudinal approach {[}1{]}, {[}2{]}.
+In the first case, weekly measurements of skin toxicities in patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT).
+Longitudinal studies have been also used to analyze tumor response {[}3{]}--{[}6{]}, antibody expression {[}7{]}, {[}8{]}, and cell metabolism {[}9{]}, {[}10{]}.
 
-\hypertarget{background}{%
-\subsection{Background}\label{background}}
+Traditionally, a ``frequentist'' approach is used in biomedical research to derive inferences from a longitudinal study.
+Such statistical view derives its name from the fact that it regards probability as a limiting frequency{[}11{]} and its application is based on a null hypothesis test using the \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
+This methodology makes two key assumptions regarding longitudinal data: constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study (a condition also known as complete observations) {[}12{]}, {[}13{]}.
 
-A longitudinal study is defined as that which is designed to repeatedly
-measure a variable of interest in a group (or groups) of subjects. In
-biomedical research, this type of study is employed when the intention
-is to observe the evolution of the effect of a certain treatment across
-time, rather than analyzing it at a single time point (a cross-sectional
-study). Clinical studies of breast and neck cancer have used a
-longitudinal approach {[}\protect\hyperlink{ref-sio2016}{1}{]},
-{[}\protect\hyperlink{ref-kamstra2015}{2}{]}. In the first case, weekly
-measurements of skin toxicities in patients with radiation-induced
-dermatitis were taken for up to 8 weeks; whereas in the latter mouth
-opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy
-(RT). Longitudinal studies have been also used to analyze tumor response
-{[}\protect\hyperlink{ref-roblyer2011}{3}{]}--{[}\protect\hyperlink{ref-demidov2018}{6}{]},
-antibody expression {[}\protect\hyperlink{ref-ritter2001}{7}{]},
-{[}\protect\hyperlink{ref-roth2017}{8}{]}, and cell metabolism
-{[}\protect\hyperlink{ref-jones2018}{9}{]},
-{[}\protect\hyperlink{ref-skala2010}{10}{]}.
+However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases {[}14{]}.
+In the case of the number of observations during the study, different situations can arise that prevent the collection of complete measurements for all subjects: in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and it is even possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point and therefore violating the \emph{complete observations} assumption of rm-ANOVA.
 
-Traditionally, a ``frequentist'' approach is used in biomedical research
-to derive inferences from a longitudinal study. Such statistical view
-derives its name from the fact that it regards probability as a limiting
-frequency{[}\protect\hyperlink{ref-wagenmakers2008}{11}{]} and its
-application is based on a null hypothesis test using the \emph{analysis
-of variance over repeated measures} (repeated measures ANOVA or
-rm-ANOVA). This methodology makes two key assumptions regarding
-longitudinal data: constant correlation exists across same-subject
-measurements, and observations from each subject are obtained at all
-time points through the study (a condition also known as complete
-observations) {[}\protect\hyperlink{ref-gueorguieva2004}{12}{]},
-{[}\protect\hyperlink{ref-schober2018}{13}{]}.
+When incomplete observations occur, an rm-ANOVA requires the exclusion of all subjects with missing observations from the analysis.
+This can result in increased costs if the desired statistical power is not met with the remaining observations, as it would be necessary to enroll more subjects.
+At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
+Additionally, rm-ANOVA uses a \emph{post hoc} analysis to assess differences between the measured response in different groups.
+A \emph{post hoc} analysis is based on multiple repeated comparisons to estimate a \emph{p-value}, a metric that is widely used as a measure of significance.
+Because the \emph{p-value} is highly variable, multiple comparisons can inflate the false positivity rate {[}15{]}, {[}16{]}, consequently biasing the conclusions of the study.
 
-However, constant correlation is frequently unjustified as its value
-tends to diminish between measures when the time interval between them
-increases {[}\protect\hyperlink{ref-ugrinowitsch2004}{14}{]}. In the
-case of the number of observations during the study, different
-situations can arise that prevent the collection of complete
-measurements for all subjects: in a clinical trial voluntary withdrawal
-from one or multiple patients can occur, whereas attrition in animals
-due to injury or weight loss can occur in preclinical experiments, and
-it is even possible that unexpected complications with equipment or
-supplies arise, preventing the researcher from collecting measurements
-at a certain time point and therefore violating the \emph{complete
-observations} assumption of rm-ANOVA.
+During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
+This is exemplified by the use of linear mixed effects models (LMEMs) in certain groups to analyze longitudinal data {[}10{]}, {[}17{]}.
+Briefly, these models incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g.~the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population.
+These models are more flexible than the traditional rm-ANOVA as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}18{]}.
+On the other hand, they impose restrictions in the distribution of the errors of the random effects {[}12{]}.
 
-When incomplete observations occur, an rm-ANOVA requires the exclusion
-of all subjects with missing observations from the analysis. This can
-result in increased costs if the desired statistical power is not met
-with the remaining observations, as it would be necessary to enroll more
-subjects. At the same time, if the excluded observations contain
-insightful information that is not used, their elimination from the
-analysis may limit the demonstration of significant differences between
-groups. Additionally, rm-ANOVA uses a \emph{post hoc} analysis to assess
-differences between the measured response in different groups. A
-\emph{post hoc} analysis is based on multiple repeated comparisons to
-estimate a \emph{p-value}, a metric that is widely used as a measure of
-significance. Because the \emph{p-value} is highly variable, multiple
-comparisons can inflate the false positivity rate
-{[}\protect\hyperlink{ref-liu2010}{15}{]},
-{[}\protect\hyperlink{ref-halsey2015}{16}{]}, consequently biasing the
-conclusions of the study.
+One final assumption that is not initially evident for both rm-ANOVA and LMEMs models pertains the relationship between the observed effect and the progression of the timeline of the study, which is expected to be linear {[}18{]}.
+This common assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend.
+In biomedical research, a special case of this behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings {[}3{]}, {[}10{]}, {[}17{]}.
+These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent with the observed variation.
+In other words, the in both cases the model tries to explain highly-variable data using a linear trend, consequently biasing the estimates.
+Additionally, although it is possible that a \emph{post hoc} analysis is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) by using multiple comparisons between the model terms, the use of this estimator is limited in such scenario because it is intended to work in models that have a reasonable agreement with the data.
 
-During the last decade, the biomedical community has started to
-recognize the limitations of rm-ANOVA in the analysis of longitudinal
-information. This is exemplified by the use of linear mixed effects
-models (LMEMs) in certain groups to analyze longitudinal data
-{[}\protect\hyperlink{ref-skala2010}{10}{]},
-{[}\protect\hyperlink{ref-vishwanath2009}{17}{]}. Briefly, these models
-incorporate \emph{fixed effects}, which correspond to the levels of
-experimental factors in the study (e.g.~the different drug regimens in a
-clinical trial), and \emph{random effects}, which account for random
-variation within the population. These models are more flexible than the
-traditional rm-ANOVA as they can accommodate missing observations for
-multiple subjects and allow different modeling strategies for the
-variability within each measure in every subject
-{[}\protect\hyperlink{ref-pinheiro2006}{18}{]}. On the other hand, they
-impose restrictions in the distribution of the errors of the random
-effects {[}\protect\hyperlink{ref-gueorguieva2004}{12}{]}.
+As the ``frequentist'' rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, while enabling inferences that are both adequate and consistent from a statistical perspective.
+In this regard, Generalized Additive Models (GAMs) present an alternative approach to analyze longitudinal data.
+Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data {[}19{]}--{[}21{]}, health-environment interactions {[}22{]} and political science {[}23{]} .
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: They have an extensively developed theoretical background, enable the data to dictate the trend of the model, can accommodate missing observations and do not require constant correlation between repeated measurements {[}24{]}.Therefore, GAMs can provide a more suitable statistical method to analyze non-linear biomedical longitudinal data.
 
-One final assumption that is not initially evident for both rm-ANOVA and
-LMEMs models pertains the relationship between the observed effect and
-the progression of the timeline of the study, which is expected to be
-linear {[}\protect\hyperlink{ref-pinheiro2006}{18}{]}. This common
-assumption in both rm-ANOVA ane LMEMs consequently restricts the
-inferences from the model when the data does not follow a linear trend.
-In biomedical research, a special case of this behavior in longitudinal
-data arises in measurements of tumor response in preclinical and
-clinical settings {[}\protect\hyperlink{ref-roblyer2011}{3}{]},
-{[}\protect\hyperlink{ref-skala2010}{10}{]},
-{[}\protect\hyperlink{ref-vishwanath2009}{17}{]}. These studies have
-shown that the collected signal does not follow a linear trend over
-time, and presents extreme variability at different time points, making
-the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent
-with the observed variation. In other words, the in both cases the model
-tries to explain highly-variable data using a linear trend, consequently
-biasing the estimates. Additionally, although it is possible that a
-\emph{post hoc} analysis is able to find ``significant''
-\emph{p-values}( \emph{p}\textless0.05) by using multiple comparisons
-between the model terms, the use of this estimator is limited in such
-scenario because it is intended to work in models that have a reasonable
-agreement with the data.
+The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of more complex models beyond LMEMs.
+In particular, \(\textsf{R}\) has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}24{]}--{[}26{]} that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+This programming language is also able to simulate data, an emerging strategy used to test statistical models {[}27{]}.
+This allows to create and explore different alternatives for analysis without collecting information in the field, and reduces the time window between experiment design and its implementation.
 
-As the ``frequentist'' rm-ANOVA and the more advanced LMEM approach are
-both limited in the analysis of non-linear longitudinal information,
-there is a need for biomedical researchers to explore the use of
-additional statistical tools that allow the information (and not an
-assumed trend) to determine the fit of the model, while enabling
-inferences that are both adequate and consistent from a statistical
-perspective. In this regard, Generalized Additive Models (GAMs) present
-an alternative approach to analyze longitudinal data. Although not
-commonly used in the biomedical community, these non-parametric models
-have been used to analyze temporal variations in geochemical and
-palaeoecological data
-{[}\protect\hyperlink{ref-rose2012}{19}{]}--{[}\protect\hyperlink{ref-simpson2018}{21}{]},
-health-environment interactions
-{[}\protect\hyperlink{ref-yang2012}{22}{]} and political science
-{[}\protect\hyperlink{ref-beck1998}{23}{]} . There are several
-advantages of GAMs over LMEMs and rm-ANOVA models: They have an
-extensively developed theoretical background, enable the data to dictate
-the trend of the model, can accommodate missing observations and do not
-require constant correlation between repeated measurements
-{[}\protect\hyperlink{ref-wood2017}{24}{]}.Therefore, GAMs can provide a
-more suitable statistical method to analyze non-linear biomedical
-longitudinal data.
-
-The current advances in programming languages designed for statistical
-analysis (specifically \(\textsf{R}\)), have eased the computational
-implementation of more complex models beyond LMEMs. In particular,
-\(\textsf{R}\) has an extensive collection of documentation and
-functions to fit GAMs in the package \emph{mgcv}
-{[}\protect\hyperlink{ref-wood2017}{24}{]}--{[}\protect\hyperlink{ref-wood2016}{26}{]}
-that not only speed up the initial stages of the analysis but enable the
-use of advanced modeling structures (e.g.~hierarchical models,
-confidence interval comparisons) without requiring advanced programming
-skills from the user. This programming language is also able to simulate
-data, an emerging strategy used to test statistical models
-{[}\protect\hyperlink{ref-haverkamp2017}{27}{]}. This allows to create
-and explore different alternatives for analysis without collecting
-information in the field, and reduces the time window between experiment
-design and its implementation.
-
-Therefore, the purpose of this study is to provide biomedical
-researchers with a clear understanding of the theory and the practical
-implementation of GAMs to analyze longitudinal data using by focusing on
-four areas. First, the limitations of rm-ANOVA and LMEMs to analyze
-longitudinal data are contrasted in the cases of missing observations,
-assumption of linearity and correlation structures. Secondly, the key
-theoretical elements of GAMS are presented without using extensive
-derivations or complex mathematical notation that can obscure their
-understanding. Thirdly, simulated data that follows the trend of
-previously reported values
-{[}\protect\hyperlink{ref-vishwanath2009}{17}{]} is used to illustrate
-the type of non-linear longitudinal data that occurs in biomedical
-research and the implementation of GAMs in such scenario. Finally,
-reproducibility is emphasized by providing the code to generate the data
-and the implementation of GAMs in \(\textsf{R}\), in conjunction with a
-step-by-step guide to fit models of increasing complexity.\\
-In summary, the exploration of modern statistical techniques to analyze
-longitudinal data may allow biomedical researchers to consider the use
-of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a
-linear trend, and will also help to improve the standards for
-reproducibility in biomedical research.
-
-\hypertarget{section-1}{%
-\subsection{Section 1:}\label{section-1}}
+Therefore, the purpose of this study is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas.
+First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity and correlation structures.
+Secondly, the key theoretical elements of GAMS are presented without using extensive derivations or complex mathematical notation that can obscure their understanding.
+Thirdly, simulated data that follows the trend of previously reported values {[}17{]} is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research and the implementation of GAMs in such scenario.
+Finally, reproducibility is emphasized by providing the code to generate the data and the implementation of GAMs in \(\textsf{R}\), in conjunction with a step-by-step guide to fit models of increasing complexity.\\
+In summary, the exploration of modern statistical techniques to analyze longitudinal data may allow biomedical researchers to consider the use of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a linear trend, and will also help to improve the standards for reproducibility in biomedical research.
 
 \hypertarget{challenges-presented-by-longitudinal-studies}{%
-\subsubsection{Challenges presented by longitudinal
-studies}\label{challenges-presented-by-longitudinal-studies}}
+\section{Challenges presented by longitudinal studies}\label{challenges-presented-by-longitudinal-studies}}
 
 \hypertarget{the-frequentist-case-for-longitudinal-data}{%
-\subsubsection{1 The ``frequentist'' case for longitudinal
-data}\label{the-frequentist-case-for-longitudinal-data}}
+\subsection{The ``frequentist'' case for longitudinal data}\label{the-frequentist-case-for-longitudinal-data}}
 
-The \emph{repeated measures analysis of variance} (rm-ANOVA) is the
-standard statistical analysis for longitudinal data in biomedical
-research, but certain assumptions are necessary to make the model valid.
-From a practical view, they can be divided in three areas: linear
-relationship between covariates and response, constant correlation
-between measurements, and complete observations for all subjects. Each
-one of these assumptions is discussed below.
+The \emph{repeated measures analysis of variance} (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research, but certain assumptions are necessary to make the model valid.
+From a practical view, they can be divided in three areas: linear relationship between covariates and response, constant correlation between measurements, and complete observations for all subjects.
+Each one of these assumptions is discussed below.
 
 \hypertarget{linear-relationship}{%
-\paragraph{1.1 Linear relationship}\label{linear-relationship}}
+\subsection{Linear relationship}\label{linear-relationship}}
 
-In a biomedical longitudinal study, two or more groups of subjects
-(patients, mice, samples) are subject to a different treatments (e.g.~a
-``treatment'' group receives a novel drug vs.~a ``control'' group that
-receives a placebo), and measurements from each subject within each
-group are collected at specific time points. The collected response has
-both \emph{fixed} and \emph{random} components. The \emph{fixed}
-component can be understood as a constant value in the response which
-the researcher is interested in measuring, i.e, the effect of the novel
-drug in the ``treatment'' group.The \emph{random} component can be
-defined as ``noise'' caused by some inherent variability within the
-study. For example, if the blood concentration of the drug is measured
-in certain subjects in the early hours of the morning while others are
-measured in the afternoon, it is possible that the difference in the
-collection time of the measurement introduces some ``noise'' in the
-signal. As their name suggests, this ``random'' variability needs to be
-modeled as a variable rather than as a constant value.
+In a biomedical longitudinal study, two or more groups of subjects (patients, mice, samples) are subject to a different treatments (e.g.~a ``treatment'' group receives a novel drug vs.~a ``control'' group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
+The collected response has both \emph{fixed} and \emph{random} components.
+The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the effect of the novel drug in the ``treatment'' group.The \emph{random} component can be defined as ``noise'' caused by some inherent variability within the study.
+For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while others are measured in the afternoon, it is possible that the difference in the collection time of the measurement introduces some ``noise'' in the signal.
+As their name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
 
@@
 \begin{equation}
-  y_{i} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t} * treatment_{j}+\pi_{ij} +\epsilon_{tij}\\ (\#eq:linear-model)
+  y_{i} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t} *                                    treatment_{j}+\pi_{ij} +\epsilon_{tij}\\ 
+  \label{eq:linear-model}
 \end{equation}
 
-In this model \(y_i\) is the response by subject \(i\), which can be
-decomposed in a mean value \(\mu\), \emph{fixed effects} of time
-(\(time_t\)), treatment (\(time_j\)) and their interaction; a
-\emph{random effect} \(\pi_{ij}\) of each subject within each group, and
-errors per subject per time per group \(\epsilon_{tij}\). Suppose two
-treatments groups are used in the study. Models per group can be
-generated if 0 is used for the first treatment group (Group A) and 1 for
-the second treatment group (Group B). The linear models then become:
+In this model \(y_i\) is the response by subject \(i\), which can be decomposed in a mean value \(\mu\), \emph{fixed effects} of time (\(time_t\)), treatment (\(time_j\)) and their interaction; a \emph{random effect} \(\pi_{ij}\) of each subject within each group, and errors per subject per time per group \(\epsilon_{tij}\).
+Suppose two treatments groups are used in the study.
+Models per group can be generated if 0 is used for the first treatment group (Group A) and 1 for the second treatment group (Group B).
+The linear models then become:
 
@@
   \mu + \beta_1*time_{t}+\epsilon_{ti}   & \mbox{if Group A}\\
-  \mu + \beta_1 * time_{t} + \beta_2 * treatment_{j} +\beta_3* time_{t} * treatment_{j}        +\pi_{i}+ \epsilon_{tij}  & \mbox{if Group B}\\
+  \mu + \beta_1 * time_{t} + \beta_2 * treatment_{j} +\beta_3* time_{t} * treatment_{j}+\pi_{i}+\epsilon_{tij}  & \mbox{if Group B}\\
   \end{cases}
-  (\#eq:ANOVA-by-group)
+  \label{eq:ANOVA-by-group}
 \end{equation}
 
-Furthermore, @ref(eq:ANOVA-by-group) can be re-written under the
-generalized linear model (GLM) framework
-{[}\protect\hyperlink{ref-nelder1972}{28}{]}. The main difference in
-this case is that a GLM model uses a \emph{linking function} to connect
-the observed value to the linear model.
+Furthermore, \eqref{eq:ANOVA-by-group} can be re-written under the generalized linear model (GLM) framework {[}28{]}.
+The main difference in this case is that a GLM model uses a \emph{linking function} to connect the observed value to the linear model.
 
@@
   \end{cases}
-  \\
-  (\#eq:GLM
+  \label{eq:GLM}
 \end{equation}
 
-The linear model in @ref(eq:linear-model) is a GLM where
-\(f(time) = time\) with the identity function \(f(x) = x\).
+The linear model in \eqref{eq:linear-model} is a GLM where \(f(time) = time\) with the identity function \(f(x) = x\).
 
-Regardless of its presentation, @ref(eq:linear-model),
-@ref(eq:ANOVA-by-group) and @ref(eq:GLM) show that the model expects a
-linear relationship between the covariates and the response. When the
-data does not follow a linear trend, the fit that the model produces
-does not accurately represent the changes in the data. To exemplify this
-consider simulated longitudinal data where the a normally distributed
-response of two groups of 10 subjects each is obtained, and an rm-ANOVA
-model such as @ref(eq:linear-model) is fitted to the data (code for the
-simulated data is available in the Appendix).
-
-\begin{verbatim}
-## 
-## Attaching package: 'nlme'
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:dplyr':
-## 
-##     collapse
-\end{verbatim}
+Regardless of its presentation, \eqref{eq:linear-model}, \eqref{eq:ANOVA-by-group} and \eqref{eq:GLM} show that the model expects a linear relationship between the covariates and the response.
+When the data does not follow a linear trend, the fit that the model produces does not accurately represent the changes in the data.
+To exemplify this consider simulated longitudinal data where the a normally distributed response of two groups of 10 subjects each is obtained, and an rm-ANOVA model such as \eqref{eq:linear-model} is fitted to the data (code for the simulated data is available in the Appendix).
 
@@
 \hypertarget{covariance-in-rm-anova-and-lmems}{%
-\paragraph{1.2 Covariance in rm-ANOVA and
-LMEMs}\label{covariance-in-rm-anova-and-lmems}}
+\subsection{Covariance in rm-ANOVA and LMEMs}\label{covariance-in-rm-anova-and-lmems}}
 
-In a longitudinal study there is an expected \emph{variance} between
-repeated measurements on the same subject, and because repeated measures
-occur in the subjects within each group, there is a \emph{covariance}
-between measurements at each time point within each group. The
-\emph{covariance matrix} (also known as the variance-covariance matrix)
-is a matrix that captures the variation between and within subjects in a
-longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{29}{]} (For
-an in-depth analysis of the covariance matrix see
-{[}\protect\hyperlink{ref-west2014}{30}{]},
-{[}\protect\hyperlink{ref-weiss2005}{31}{]}).
+In a longitudinal study there is an expected \emph{variance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
+The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}29{]} (For an in-depth analysis of the covariance matrix see {[}30{]}, {[}31{]}).
 
-In the case of an rm-ANOVA analysis, it is assumed that the covariance
-matrix has a specific construction known as \emph{compound symmetry}
-(also known as ``sphericity'' or ``circularity''). Under this
-assumption, the between-subject variance and within-subject correlation
-are constant across time
-{[}\protect\hyperlink{ref-weiss2005}{31}{]}--{[}\protect\hyperlink{ref-huynh1976}{33}{]}.
-However, it has been shown that this condition is frequently unjustified
-because the correlation between measurements tends to change over time
-{[}\protect\hyperlink{ref-maxwell2017}{34}{]}; and it is higher between
-consecutive measurements
-{[}\protect\hyperlink{ref-gueorguieva2004}{12}{]},
-{[}\protect\hyperlink{ref-ugrinowitsch2004}{14}{]}. Although corrections
-can be made (such as Huyhn-Feldt or Greenhouse-Geisser) the
-effectiveness of each correction is limited because it depends on the
-size of the sample,the number of repeated
-measurements{[}\protect\hyperlink{ref-haverkamp2017}{27}{]}, and they
-are not robust if the group sizes are unbalanced
-{[}\protect\hyperlink{ref-keselman2001}{35}{]}. In other words, if the
-data does not present constant correlation between repeated
-measurements, the assumptions required for an rm-ANOVA model are not met
-and the use of corrections may still not provide a reasonable adjustment
-that makes the model valid.
+In the case of an rm-ANOVA analysis, it is assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
+Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}31{]}--{[}33{]}.
+However, it has been shown that this condition is frequently unjustified because the correlation between measurements tends to change over time {[}34{]}; and it is higher between consecutive measurements {[}12{]}, {[}14{]}.
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser) the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}27{]}, and they are not robust if the group sizes are unbalanced {[}35{]}.
+In other words, if the data does not present constant correlation between repeated measurements, the assumptions required for an rm-ANOVA model are not met and the use of corrections may still not provide a reasonable adjustment that makes the model valid.
 
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow
-different structures for the variance-covariance matrix including
-exponential, autoregressive of order 1, rational quadratic and others
-{[}\protect\hyperlink{ref-pinheiro2006}{18}{]}. Nevertheless, the
-analysis required to determine an appropriate variance-covariance
-structure for the data can be a long process by itself. Overall, the
-spherical assumption for rm-ANOVA may not capture the natural variations
-of the correlation in the data, and can bias the inferences from the
-analysis.
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}18{]}.
+Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself.
+Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
 
 \hypertarget{missing-observations}{%
-\paragraph{1.3 Missing observations}\label{missing-observations}}
+\subsection{Missing observations}\label{missing-observations}}
 
-Missing observations are an issue that arises frequently in longitudinal
-studies. In biomedical research, this situation can be caused by reasons
-beyond the control of the investigator {[}molenberghs2004{]}.Dropout
-from patients, or attrition or injury in animals are among the reasons
-for missing observations. Statistically, missing information can be
-classified as \emph{missing at random} (MAR), \emph{missing completely
-at random} (MCAR), and \emph{missing not at random} (MNAR)
-{[}\protect\hyperlink{ref-weiss2005}{31}{]}. In a MAR scenario, the
-pattern of the missing information is related to some variable in the
-data, but it is not related to the variable of interest
-{[}\protect\hyperlink{ref-scheffer2002}{36}{]}. If the data are MCAR,
-this means that the missigness is completely unrelated to the collected
-information {[}\protect\hyperlink{ref-potthoff2006}{37}{]}, and in the
-case of MNAR the missing values are dependent on their value. An
-rm-ANOVA model assumes complete observations for all subjects, and
-therefore subjects with one or more missing observations are excluded
-from the analysis. This is inconvenient because the remaining subjects
-might not accurately represent the population, and statistical power is
-affected by this reduction in sample size
-{[}\protect\hyperlink{ref-ma2012}{38}{]}.
+Missing observations are an issue that arises frequently in longitudinal studies.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}molenberghs2004{]}.Dropout from patients, or attrition or injury in animals are among the reasons for missing observations.
+Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}31{]}.
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}36{]}.
+If the data are MCAR, this means that the missigness is completely unrelated to the collected information {[}37{]}, and in the case of MNAR the missing values are dependent on their value.
+An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}38{]}.
 
-In the case of LMEMs, inferences from the model are valid when missing
-observations in the data exist that are MAR
-{[}\protect\hyperlink{ref-west2014}{30}{]}. The pattern of missing
-observations can be considered MAR if the missing observations are not
-related any of the other variables measured in the study
-{[}\protect\hyperlink{ref-maxwell2017}{34}{]}. For example, if attrition
-occurs in all mice that had lower weights at the beginning of a
-chemotherapy response study, the missing data can be considered MAR
-because the missigness is unrelated to other variables of interest.
+In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR {[}30{]}.
+The pattern of missing observations can be considered MAR if the missing observations are not related any of the other variables measured in the study {[}34{]}.
+For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
 
@@
 
-This section has presented the assumptions of rm-ANOVA to analyze
-longitudinal information and its differences when compared to LMEMs
-regarding to missing data and the modeling of the covariance matrix. Of
-notice, LMEMs offer a more robust and flexible approach than rm-ANOVA
-and if the data follows a linear trend, they provide an excellent choice
-to derive inferences from a repeated measures study. However, when the
-data presents high variability, LMEMs fail to capture the non-linear
-trend of the data. To analyze such type of data, we present generalized
-additive models (GAMs) as an alternative in the following section.
+This section has presented the assumptions of rm-ANOVA to analyze longitudinal information and its differences when compared to LMEMs regarding to missing data and the modeling of the covariance matrix.
+Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
+However, when the data presents high variability, LMEMs fail to capture the non-linear trend of the data.
+To analyze such type of data, we present generalized additive models (GAMs) as an alternative in the following section.
 
 \hypertarget{gams-as-a-special-case-of-generalized-linear-models}{%
-\paragraph{2 GAMs as a special case of Generalized Linear
-Models}\label{gams-as-a-special-case-of-generalized-linear-models}}
+\section{GAMs as a special case of Generalized Linear Models}\label{gams-as-a-special-case-of-generalized-linear-models}}
 
-A GAM is a special case of the Generalized Linear Model (GLM), a
-framework that allows for response distributions that do not follow a
-normal distribution {[}\protect\hyperlink{ref-wood2017}{24}{]},
-{[}\protect\hyperlink{ref-hastie1987}{39}{]}. Following the notation by
-Simpson {[}\protect\hyperlink{ref-simpson2018}{21}{]} A GAM model can be
-represented as:
+A GAM is a special case of the Generalized Linear Model (GLM), a framework that allows for response distributions that do not follow a normal distribution {[}24{]}, {[}39{]}.
+Following the notation by Simpson {[}21{]} A GAM model can be represented as:
 
@@
   y_{t}=\beta_0+f(x_t)+\epsilon_t  
-  (\#eq:GAM)
+  \label{eq:GAM}
 \end{equation}
 
-Where \(y_t\) is the response at time \(t\), \(\beta_0\) is the expected
-value at time 0, the change of \(y\) over time is represented by the
-function \(f(x_t)\) and \(\epsilon_t\) represents the residuals.
+Where \(y_t\) is the response at time \(t\), \(\beta_0\) is the expected value at time 0, the change of \(y\) over time is represented by the function \(f(x_t)\) and \(\epsilon_t\) represents the residuals.
 
-In contrast to rm-ANOVA or LMEMs, GAMs use \emph{smooth functions} to
-model the relationship between the covariates and the response. This
-approach is more advantageous than that of a LMEMs or rm-ANOVA as it
-does not restrict the model to a linear relationship. One possible
-function for \(f(x_t)\) is a polynomial, but a major limitation is that
-they create a ``global'' fit as they assume that the same relationship
-exists everywhere, which can cause problems with the fit
-{[}\protect\hyperlink{ref-beck1998}{23}{]}.
+In contrast to rm-ANOVA or LMEMs, GAMs use \emph{smooth functions} to model the relationship between the covariates and the response.
+This approach is more advantageous than that of a LMEMs or rm-ANOVA as it does not restrict the model to a linear relationship.
+One possible function for \(f(x_t)\) is a polynomial, but a major limitation is that they create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}23{]}.
 
-To overcome this limitation, the smooth functions in GAMs are
-represented using \emph{basis functions} over evenly spaced ranges of
-the covariates known as \emph{knots}. The \emph{basis function} used is
-a cubic spline, which is a smooth curve constructed from cubic
-polynomials joined
-together{[}\protect\hyperlink{ref-simpson2018}{21}{]},
-{[}\protect\hyperlink{ref-wood2017}{24}{]}. Cubic splines have a long
-history in their use to solve non-parametric statistical problems and
-are the default choice to fit GAMs as they are the simplest option to
-obtain visual smoothness {[}\protect\hyperlink{ref-wegman1983}{40}{]}.
+To overcome this limitation, the smooth functions in GAMs are represented using \emph{basis functions} over evenly spaced ranges of the covariates known as \emph{knots}.
+The \emph{basis function} used is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together{[}21{]}, {[}24{]}.
+Cubic splines have a long history in their use to solve non-parametric statistical problems and are the default choice to fit GAMs as they are the simplest option to obtain visual smoothness {[}40{]}.
 
@@
 \leavevmode\hypertarget{ref-sio2016}{}%
-{[}1{]} T. T. Sio \emph{et al.}, ``Repeated measures analyses of
-dermatitis symptom evolution in breast cancer patients receiving
-radiotherapy in a phase 3 randomized trial of mometasone furoate vs
-placebo (n06c4 {[}alliance{]}),'' \emph{Supportive Care in Cancer}, vol.
-24, no. 9, pp. 3847--3855, 2016.
+{[}1{]} T.
+T.
+Sio \emph{et al.}, ``Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (n06c4 {[}alliance{]}),'' \emph{Supportive Care in Cancer}, vol.
+24, no.
+9, pp.
+3847--3855, 2016.
 
 \leavevmode\hypertarget{ref-kamstra2015}{}%
-{[}2{]} J. Kamstra, P. Dijkstra, M. Van Leeuwen, J. Roodenburg, and J.
-Langendijk, ``Mouth opening in patients irradiated for head and neck
-cancer: A prospective repeated measures study,'' \emph{Oral Oncology},
-vol. 51, no. 5, pp. 548--555, 2015.
+{[}2{]} J.
+Kamstra, P.
+Dijkstra, M.
+Van Leeuwen, J.
+Roodenburg, and J.
+Langendijk, ``Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study,'' \emph{Oral Oncology}, vol.
+51, no.
+5, pp.
+548--555, 2015.
 
 \leavevmode\hypertarget{ref-roblyer2011}{}%
-{[}3{]} D. Roblyer \emph{et al.}, ``Optical imaging of breast cancer
-oxyhemoglobin flare correlates with neoadjuvant chemotherapy response
-one day after starting treatment,'' \emph{Proceedings of the National
-Academy of Sciences}, vol. 108, no. 35, pp. 14626--14631, 2011.
+{[}3{]} D.
+Roblyer \emph{et al.}, ``Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment,'' \emph{Proceedings of the National Academy of Sciences}, vol.
+108, no.
+35, pp.
+14626--14631, 2011.
 
 \leavevmode\hypertarget{ref-tank2020}{}%
-{[}4{]} A. Tank \emph{et al.}, ``Diffuse optical spectroscopic imaging
-reveals distinct early breast tumor hemodynamic responses to metronomic
-and maximum tolerated dose regimens,'' \emph{Breast cancer research},
-vol. 22, no. 1, pp. 1--10, 2020.
+{[}4{]} A.
+Tank \emph{et al.}, ``Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens,'' \emph{Breast cancer research}, vol.
+22, no.
+1, pp.
+1--10, 2020.
 
 \leavevmode\hypertarget{ref-pavlov2018}{}%
-{[}5{]} M. V. Pavlov \emph{et al.}, ``Multimodal approach in assessment
-of the response of breast cancer to neoadjuvant chemotherapy,''
-\emph{Journal of biomedical optics}, vol. 23, no. 9, p. 091410, 2018.
+{[}5{]} M.
+V.
+Pavlov \emph{et al.}, ``Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy,'' \emph{Journal of biomedical optics}, vol.
+23, no.
+9, p.
+091410, 2018.
 
 \leavevmode\hypertarget{ref-demidov2018}{}%
-{[}6{]} V. Demidov \emph{et al.}, ``Preclinical longitudinal imaging of
-tumor microvascular radiobiological response with functional optical
-coherence tomography,'' \emph{Scientific reports}, vol. 8, no. 1, pp.
+{[}6{]} V.
+Demidov \emph{et al.}, ``Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography,'' \emph{Scientific reports}, vol.
+8, no.
+1, pp.
 1--12, 2018.
@@
 \leavevmode\hypertarget{ref-ritter2001}{}%
-{[}7{]} G. Ritter, L. S. Cohen, C. Williams, E. C. Richards, L. J. Old,
-and S. Welt, ``Serological analysis of human anti-human antibody
-responses in colon cancer patients treated with repeated doses of
-humanized monoclonal antibody a33,'' \emph{Cancer Research}, vol. 61,
-no. 18, pp. 6851--6859, 2001.
+{[}7{]} G.
+Ritter, L.
+S.
+Cohen, C.
+Williams, E.
+C.
+Richards, L.
+J.
+Old, and S.
+Welt, ``Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody a33,'' \emph{Cancer Research}, vol.
+61, no.
+18, pp.
+6851--6859, 2001.
 
 \leavevmode\hypertarget{ref-roth2017}{}%
-{[}8{]} E. M. Roth \emph{et al.}, ``Antidrug antibodies in patients
-treated with alirocumab,'' 2017.
+{[}8{]} E. M. Roth \emph{et al.}, ``Antidrug antibodies in patients treated with alirocumab,'' 2017.
 
 \leavevmode\hypertarget{ref-jones2018}{}%
-{[}9{]} J. D. Jones, H. E. Ramser, A. E. Woessner, and K. P. Quinn, ``In
-vivo multiphoton microscopy detects longitudinal metabolic changes
-associated with delayed skin wound healing,'' \emph{Communications
-biology}, vol. 1, no. 1, pp. 1--8, 2018.
+{[}9{]} J.
+D.
+Jones, H.
+E.
+Ramser, A.
+E.
+Woessner, and K.
+P.
+Quinn, ``In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing,'' \emph{Communications biology}, vol.
+1, no.
+1, pp.
+1--8, 2018.
 
 \leavevmode\hypertarget{ref-skala2010}{}%
-{[}10{]} M. C. Skala, A. N. Fontanella, L. Lan, J. A. Izatt, and M. W.
-Dewhirst, ``Longitudinal optical imaging of tumor metabolism and
-hemodynamics,'' \emph{Journal of biomedical optics}, vol. 15, no. 1, p.
+{[}10{]} M.
+C.
+Skala, A.
+N.
+Fontanella, L.
+Lan, J.
+A.
+Izatt, and M.
+W.
+Dewhirst, ``Longitudinal optical imaging of tumor metabolism and hemodynamics,'' \emph{Journal of biomedical optics}, vol.
+15, no.
+1, p.
 011112, 2010.
@@
 \leavevmode\hypertarget{ref-wagenmakers2008}{}%
-{[}11{]} E.-J. Wagenmakers, M. Lee, T. Lodewyckx, and G. J. Iverson,
-``Bayesian versus frequentist inference,'' in \emph{Bayesian evaluation
-of informative hypotheses}, Springer, 2008, pp. 181--207.
+{[}11{]} E.-J. Wagenmakers, M. Lee, T. Lodewyckx, and G. J. Iverson, ``Bayesian versus frequentist inference,'' in \emph{Bayesian evaluation of informative hypotheses}, Springer, 2008, pp. 181--207.
 
 \leavevmode\hypertarget{ref-gueorguieva2004}{}%
-{[}12{]} R. Gueorguieva and J. H. Krystal, ``Move over anova: Progress
-in analyzing repeated-measures data andits reflection in papers
-published in the archives of general psychiatry,'' \emph{Archives of
-general psychiatry}, vol. 61, no. 3, pp. 310--317, 2004.
+{[}12{]} R.
+Gueorguieva and J.
+H.
+Krystal, ``Move over anova: Progress in analyzing repeated-measures data andits reflection in papers published in the archives of general psychiatry,'' \emph{Archives of general psychiatry}, vol.
+61, no.
+3, pp.
+310--317, 2004.
 
 \leavevmode\hypertarget{ref-schober2018}{}%
-{[}13{]} P. Schober and T. R. Vetter, ``Repeated measures designs and
-analysis of longitudinal data: If at first you do not succeed---try, try
-again,'' \emph{Anesthesia and analgesia}, vol. 127, no. 2, p. 569, 2018.
+{[}13{]} P.
+Schober and T.
+R.
+Vetter, ``Repeated measures designs and analysis of longitudinal data: If at first you do not succeed---try, try again,'' \emph{Anesthesia and analgesia}, vol.
+127, no.
+2, p.
+569, 2018.
 
 \leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
-{[}14{]} C. Ugrinowitsch, G. W. Fellingham, and M. D. Ricard,
-``Limitations of ordinary least squares models in analyzing repeated
-measures data,'' \emph{Medicine and science in sports and exercise},
-vol. 36, pp. 2144--2148, 2004.
+{[}14{]} C.
+Ugrinowitsch, G.
+W.
+Fellingham, and M.
+D.
+Ricard, ``Limitations of ordinary least squares models in analyzing repeated measures data,'' \emph{Medicine and science in sports and exercise}, vol.
+36, pp.
+2144--2148, 2004.
 
 \leavevmode\hypertarget{ref-liu2010}{}%
-{[}15{]} C. Liu, T. P. Cripe, and M.-O. Kim, ``Statistical issues in
-longitudinal data analysis for treatment efficacy studies in the
-biomedical sciences,'' \emph{Molecular Therapy}, vol. 18, no. 9, pp.
+{[}15{]} C.
+Liu, T.
+P.
+Cripe, and M.-O.
+Kim, ``Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences,'' \emph{Molecular Therapy}, vol.
+18, no.
+9, pp.
 1724--1730, 2010.
@@
 \leavevmode\hypertarget{ref-halsey2015}{}%
-{[}16{]} L. G. Halsey, D. Curran-Everett, S. L. Vowler, and G. B.
-Drummond, ``The fickle p value generates irreproducible results,''
-\emph{Nature methods}, vol. 12, no. 3, pp. 179--185, 2015.
+{[}16{]} L. G. Halsey, D. Curran-Everett, S. L. Vowler, and G. B. Drummond, ``The fickle p value generates irreproducible results,'' \emph{Nature methods}, vol. 12, no. 3, pp. 179--185, 2015.
 
 \leavevmode\hypertarget{ref-vishwanath2009}{}%
-{[}17{]} K. Vishwanath, H. Yuan, W. T. Barry, M. W. Dewhirst, and N.
-Ramanujam, ``Using optical spectroscopy to longitudinally monitor
-physiological changes within solid tumors,'' \emph{Neoplasia}, vol. 11,
-no. 9, pp. 889--900, 2009.
+{[}17{]} K.
+Vishwanath, H.
+Yuan, W.
+T.
+Barry, M.
+W.
+Dewhirst, and N.
+Ramanujam, ``Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors,'' \emph{Neoplasia}, vol.
+11, no.
+9, pp.
+889--900, 2009.
 
 \leavevmode\hypertarget{ref-pinheiro2006}{}%
-{[}18{]} J. Pinheiro and D. Bates, \emph{Mixed-effects models in s and
-s-plus}. Springer Science \& Business Media, 2006.
+{[}18{]} J. Pinheiro and D. Bates, \emph{Mixed-effects models in s and s-plus}. Springer Science \& Business Media, 2006.
 
 \leavevmode\hypertarget{ref-rose2012}{}%
-{[}19{]} N. L. Rose, H. Yang, S. D. Turner, and G. L. Simpson, ``An
-assessment of the mechanisms for the transfer of lead and mercury from
-atmospherically contaminated organic soils to lake sediments with
-particular reference to scotland, uk,'' \emph{Geochimica et Cosmochimica
-Acta}, vol. 82, pp. 113--135, 2012.
+{[}19{]} N.
+L.
+Rose, H.
+Yang, S.
+D.
+Turner, and G.
+L.
+Simpson, ``An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, uk,'' \emph{Geochimica et Cosmochimica Acta}, vol.
+82, pp.
+113--135, 2012.
 
 \leavevmode\hypertarget{ref-pedersen2019}{}%
-{[}20{]} E. J. Pedersen, D. L. Miller, G. L. Simpson, and N. Ross,
-``Hierarchical generalized additive models in ecology: An introduction
-with mgcv,'' \emph{PeerJ}, vol. 7, p. e6876, 2019.
+{[}20{]} E. J. Pedersen, D. L. Miller, G. L. Simpson, and N. Ross, ``Hierarchical generalized additive models in ecology: An introduction with mgcv,'' \emph{PeerJ}, vol. 7, p. e6876, 2019.
 
 \leavevmode\hypertarget{ref-simpson2018}{}%
-{[}21{]} G. L. Simpson, ``Modelling palaeoecological time series using
-generalised additive models,'' \emph{Frontiers in Ecology and
-Evolution}, vol. 6, p. 149, 2018.
+{[}21{]} G. L. Simpson, ``Modelling palaeoecological time series using generalised additive models,'' \emph{Frontiers in Ecology and Evolution}, vol. 6, p. 149, 2018.
 
 \leavevmode\hypertarget{ref-yang2012}{}%
-{[}22{]} L. Yang, G. Qin, N. Zhao, C. Wang, and G. Song, ``Using a
-generalized additive model with autoregressive terms to study the
-effects of daily temperature on mortality,'' \emph{BMC Medical Research
-Methodology}, vol. 12, no. 1, p. 165, 2012.
+{[}22{]} L.
+Yang, G.
+Qin, N.
+Zhao, C.
+Wang, and G.
+Song, ``Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality,'' \emph{BMC Medical Research Methodology}, vol.
+12, no.
+1, p.
+165, 2012.
 
 \leavevmode\hypertarget{ref-beck1998}{}%
-{[}23{]} N. Beck and S. Jackman, ``Beyond linearity by default:
-Generalized additive models,'' \emph{American Journal of Political
-Science}, pp. 596--627, 1998.
+{[}23{]} N. Beck and S. Jackman, ``Beyond linearity by default: Generalized additive models,'' \emph{American Journal of Political Science}, pp. 596--627, 1998.
 
 \leavevmode\hypertarget{ref-wood2017}{}%
-{[}24{]} S. N. Wood, \emph{Generalized additive models: An introduction
-with r}. CRC press, 2017.
+{[}24{]} S. N. Wood, \emph{Generalized additive models: An introduction with r}. CRC press, 2017.
 
 \leavevmode\hypertarget{ref-wood2011}{}%
-{[}25{]} S. N. Wood, N., Pya, and B. S"afken, ``Smoothing parameter and
-model selection for general smooth models (with discussion),''
-\emph{Journal of the American Statistical Association}, vol. 111, pp.
+{[}25{]} S.
+N.
+Wood, N., Pya, and B.
+S"afken, ``Smoothing parameter and model selection for general smooth models (with discussion),'' \emph{Journal of the American Statistical Association}, vol.
+111, pp.
 1548--1575, 2016.
@@
 \leavevmode\hypertarget{ref-wood2016}{}%
-{[}26{]} S. N. Wood, N., Pya, and B. S"afken, ``Smoothing parameter and
-model selection for general smooth models (with discussion),''
-\emph{Journal of the American Statistical Association}, vol. 111, pp.
+{[}26{]} S.
+N.
+Wood, N., Pya, and B.
+S"afken, ``Smoothing parameter and model selection for general smooth models (with discussion),'' \emph{Journal of the American Statistical Association}, vol.
+111, pp.
 1548--1575, 2016.
@@
 \leavevmode\hypertarget{ref-haverkamp2017}{}%
-{[}27{]} N. Haverkamp and A. Beauducel, ``Violation of the sphericity
-assumption and its effect on type-i error rates in repeated measures
-anova and multi-level linear models (mlm),'' \emph{Frontiers in
-psychology}, vol. 8, p. 1841, 2017.
+{[}27{]} N.
+Haverkamp and A.
+Beauducel, ``Violation of the sphericity assumption and its effect on type-i error rates in repeated measures anova and multi-level linear models (mlm),'' \emph{Frontiers in psychology}, vol.
+8, p.
+1841, 2017.
 
 \leavevmode\hypertarget{ref-nelder1972}{}%
-{[}28{]} J. A. Nelder and R. W. Wedderburn, ``Generalized linear
-models,'' \emph{Journal of the Royal Statistical Society: Series A
-(General)}, vol. 135, no. 3, pp. 370--384, 1972.
+{[}28{]} J. A. Nelder and R. W. Wedderburn, ``Generalized linear models,'' \emph{Journal of the Royal Statistical Society: Series A (General)}, vol. 135, no. 3, pp. 370--384, 1972.
 
 \leavevmode\hypertarget{ref-wolfinger1996}{}%
-{[}29{]} R. D. Wolfinger, ``Heterogeneous variance: Covariance
-structures for repeated measures,'' \emph{Journal of agricultural,
-biological, and environmental statistics}, pp. 205--230, 1996.
+{[}29{]} R. D. Wolfinger, ``Heterogeneous variance: Covariance structures for repeated measures,'' \emph{Journal of agricultural, biological, and environmental statistics}, pp. 205--230, 1996.
 
 \leavevmode\hypertarget{ref-west2014}{}%
-{[}30{]} B. T. West, K. B. Welch, and A. T. Galecki, \emph{Linear mixed
-models: A practical guide using statistical software}. CRC Press, 2014.
+{[}30{]} B. T. West, K. B. Welch, and A. T. Galecki, \emph{Linear mixed models: A practical guide using statistical software}. CRC Press, 2014.
 
 \leavevmode\hypertarget{ref-weiss2005}{}%
-{[}31{]} R. E. Weiss, \emph{Modeling longitudinal data}. Springer
-Science \& Business Media, 2005.
+{[}31{]} R. E. Weiss, \emph{Modeling longitudinal data}. Springer Science \& Business Media, 2005.
 
 \leavevmode\hypertarget{ref-geisser1958}{}%
-{[}32{]} S. Geisser, S. W. Greenhouse, and others, ``An extension of
-box's results on the use of the \(F\) distribution in multivariate
-analysis,'' \emph{The Annals of Mathematical Statistics}, vol. 29, no.
-3, pp. 885--891, 1958.
+{[}32{]} S.
+Geisser, S.
+W.
+Greenhouse, and others, ``An extension of box's results on the use of the \(F\) distribution in multivariate analysis,'' \emph{The Annals of Mathematical Statistics}, vol.
+29, no.
+3, pp.
+885--891, 1958.
 
 \leavevmode\hypertarget{ref-huynh1976}{}%
-{[}33{]} H. Huynh and L. S. Feldt, ``Estimation of the box correction
-for degrees of freedom from sample data in randomized block and
-split-plot designs,'' \emph{Journal of educational statistics}, vol. 1,
-no. 1, pp. 69--82, 1976.
+{[}33{]} H.
+Huynh and L.
+S.
+Feldt, ``Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs,'' \emph{Journal of educational statistics}, vol.
+1, no.
+1, pp.
+69--82, 1976.
 
 \leavevmode\hypertarget{ref-maxwell2017}{}%
-{[}34{]} S. E. Maxwell, H. D. Delaney, and K. Kelley, \emph{Designing
-experiments and analyzing data: A model comparison perspective}.
-Routledge, 2017.
+{[}34{]} S. E. Maxwell, H. D. Delaney, and K. Kelley, \emph{Designing experiments and analyzing data: A model comparison perspective}. Routledge, 2017.
 
 \leavevmode\hypertarget{ref-keselman2001}{}%
-{[}35{]} H. Keselman, J. Algina, and R. K. Kowalchuk, ``The analysis of
-repeated measures designs: A review,'' \emph{British Journal of
-Mathematical and Statistical Psychology}, vol. 54, no. 1, pp. 1--20,
-2001.
+{[}35{]} H.
+Keselman, J.
+Algina, and R.
+K.
+Kowalchuk, ``The analysis of repeated measures designs: A review,'' \emph{British Journal of Mathematical and Statistical Psychology}, vol.
+54, no.
+1, pp.
+1--20, 2001.
 
@@
 \leavevmode\hypertarget{ref-potthoff2006}{}%
-{[}37{]} R. F. Potthoff, G. E. Tudor, K. S. Pieper, and V. Hasselblad,
-``Can one assess whether missing data are missing at random in medical
-studies?'' \emph{Statistical methods in medical research}, vol. 15, no.
-3, pp. 213--234, 2006.
+{[}37{]} R.
+F.
+Potthoff, G.
+E.
+Tudor, K.
+S.
+Pieper, and V.
+Hasselblad, ``Can one assess whether missing data are missing at random in medical studies?'' \emph{Statistical methods in medical research}, vol.
+15, no.
+3, pp.
+213--234, 2006.
 
 \leavevmode\hypertarget{ref-ma2012}{}%
-{[}38{]} Y. Ma, M. Mazumdar, and S. G. Memtsoudis, ``Beyond
-repeated-measures analysis of variance: Advanced statistical methods for
-the analysis of longitudinal data in anesthesia research,''
-\emph{Regional Anesthesia \& Pain Medicine}, vol. 37, no. 1, pp.
+{[}38{]} Y.
+Ma, M.
+Mazumdar, and S.
+G.
+Memtsoudis, ``Beyond repeated-measures analysis of variance: Advanced statistical methods for the analysis of longitudinal data in anesthesia research,'' \emph{Regional Anesthesia \& Pain Medicine}, vol.
+37, no.
+1, pp.
 99--105, 2012.
@@
 \leavevmode\hypertarget{ref-hastie1987}{}%
-{[}39{]} T. Hastie and R. Tibshirani, ``Generalized additive models:
-Some applications,'' \emph{Journal of the American Statistical
-Association}, vol. 82, no. 398, pp. 371--386, 1987.
+{[}39{]} T. Hastie and R. Tibshirani, ``Generalized additive models: Some applications,'' \emph{Journal of the American Statistical Association}, vol. 82, no. 398, pp. 371--386, 1987.
 
 \leavevmode\hypertarget{ref-wegman1983}{}%
-{[}40{]} E. J. Wegman and I. W. Wright, ``Splines in statistics,''
-\emph{Journal of the American Statistical Association}, vol. 78, no.
-382, pp. 351--365, 1983.
+{[}40{]} E. J. Wegman and I. W. Wright, ``Splines in statistics,'' \emph{Journal of the American Statistical Association}, vol. 78, no. 382, pp. 351--365, 1983.
 

### 2021-01-05 Manuscripts/Manuscript_AM_v1.Rmd
Minor update, knit does not work on laptop

@@
 css: style.css
+fontfamily: Arial
   output:
@@
 In the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT).
-Longitudinal studies have been also used to analyze tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
+Longitudinal studies have beenn also used to analyze tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
 

### 2021-01-05 Manuscripts/Manuscript_AM_v1.Rmd
Does not work on laptop

@@
   output:
-
   bookdown::html_document2: default

### 2021-01-05 Code and miscellaneous/Questions.Rmd
Re-organized folder

--- 2020-09-30 Questions.Rmd
@@
 
-* The specification of prior distributions: I did not understand at the beginning why you chose a Cauchy distribution for the error, but upon reading and talking with you I understand it is just a possible distribution, it has large tails and therefore can accomodate outliers well.
+* The specification of prior distributions: I did not understand at the beginning why you chose a Cauchy distribution for the error, but upon reading and talking with you I understand it is just a possible distribution, it has large tails and therefore can accommodate outliers well.
 

### 2021-01-05 Code and miscellaneous/linear-vs-glm-equations.Rmd
Re-organized folder

---
title: "Equations"
author: "John Tipton"
date: "12/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Linear model

$$
y = \beta_0 + time * \beta_1 + treatment * \beta_2 + time * treatment * \beta_3 \\
y = \begin{cases}
\beta_0 + time * \beta_1 & \mbox{if treatment 1}\\
\beta_0 + time * \beta_1 + treatment * \beta_2 + time * treatment * \beta_3 & \mbox{if treatment 2}\\
\end{cases}
\\
$$

plugging in the value treatment = 0 for treatment 1 and the value treatment = 1 for treatment 2

$$
y = \begin{cases}
\beta_0 + time * \beta_1 & \mbox{if treatment 1}\\
( \beta_0 + \beta_2)  + time * (\beta_1 + \beta_3) & \mbox{if treatment 2}\\
\end{cases}
$$

# GLM framework

$$
y = \beta_0 + f(time) * \boldsymbol{\beta}_1 + treatment * \beta_2 + f(time) * treatment * \boldsymbol{\beta}_3 \\
y = \begin{cases}
\beta_0 + f(time) * \boldsymbol{\beta}_1 & \mbox{if treatment 1} \\
\beta_0 + f(time) * \boldsymbol{\beta}_1 + treatment * \beta_2 + f(time) * treatment * \boldsymbol{\beta}_3 & \mbox{if treatment 2}\\
\end{cases}
\\
$$

$$
y = \begin{cases}
\beta_0 + f(time) * \boldsymbol{\beta}_1 & \mbox{if treatment 1}\\
(\beta_0 + \beta_2) + f(time) * (\boldsymbol{\beta}_1 + \boldsymbol{\beta}_3) & \mbox{if treatment 2}\\
\end{cases} \\
= \begin{cases}
\beta_0 + f(time) * \boldsymbol{\beta}_1 & \mbox{if treatment 1}\\
(\tilde{\beta}_2) + f(time) * (\boldsymbol{\tilde{\beta}}_3) & \mbox{if treatment 2}\\
\end{cases}
$$

The Linear model as a GLM: set $f(time) = time$ with the identity function $f(x) = x$

### 2021-01-05 Code and miscellaneous/non-linear-ANOVA.Rmd
Re-organized folder

--- 2021-01-05 Manuscripts/Manuscript_AM_v1.Rmd
@@
 ---
-title: '**The statistical analysis of non-linear longitudinal data in biomedical research**'
-author:
-- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
-- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
-  Fayetteville]
-- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
-subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
-bibliography: refs.bib
-csl: ieee.csl
-css: style.css
-output:
-
-  bookdown::html_document2: default
-  bookdown::pdf_document2:
-    keep_tex: yes
-    fig_caption: yes
-  bookdown::word_document2: default
-  link-citations: true
+title: "Non-linear ANOVA example"
+output: html_document
 ---
 
-# Background
-
-A longitudinal study is defined as that which is designed to repeatedly measure a variable of interest in a group (or groups) of subjects.
-In biomedical research, this type of study is employed when the intention is to observe the evolution of the effect of a certain treatment across time, rather than analyzing it at a single time point (a cross-sectional study).
-Clinical studies of breast and neck cancer have used a longitudinal approach [@sio2016;@kamstra2015].
-In the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT).
-Longitudinal studies have been also used to analyze tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
-
-Traditionally, a “frequentist” approach is used in biomedical research to derive inferences from a longitudinal study.
-Such statistical view derives its name from the fact that it regards probability as a limiting frequency[@wagenmakers2008] and its application is based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-This methodology makes two key assumptions regarding longitudinal data: constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study (a condition also known as complete observations) [@gueorguieva2004;@schober2018].
-
-However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases [@ugrinowitsch2004]. 
-In the case of the number of observations during the study, different situations can arise that prevent the collection of complete measurements for all subjects:  in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and it is even possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point and therefore violating the _complete observations_ assumption of rm-ANOVA.
-
-When incomplete observations occur, an rm-ANOVA requires the exclusion of all subjects with missing observations from the analysis.
-This can result in increased costs if the desired statistical power is not met with the remaining observations, as it would be necessary to enroll more subjects.
-At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
-Additionally, rm-ANOVA uses a _post hoc_ analysis to assess differences between the measured response in different groups.
-A _post hoc_ analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance.
-Because the _p-value_ is highly variable, multiple comparisons can inflate the false positivity rate [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
-
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-This is exemplified by the  use of  linear mixed effects models (LMEMs)  in certain groups to analyze longitudinal data [@skala2010;@vishwanath2009].
-Briefly, these models incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g.
-the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population.
-These models are more flexible than the traditional rm-ANOVA as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-On the other hand, they impose restrictions in the distribution of the errors  of the random effects [@gueorguieva2004].
-
-One final assumption that is not initially evident for  both rm-ANOVA and LMEMs models pertains the relationship between the observed effect and the progression of the timeline of the study, which is expected to be linear [@pinheiro2006].
-This common assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend.
-In biomedical research, a special case of this behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent with the observed variation.
-In other words, the in both cases the model tries to explain highly-variable data using a linear trend, consequently biasing the estimates.
-Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) by using multiple comparisons between the model terms, the use of this estimator is limited in such scenario because it is intended to work in models that have a reasonable agreement with the data.
-
-As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, while enabling inferences that are both adequate and consistent from a statistical perspective.
-In this regard, Generalized Additive Models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and political science [@beck1998] .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: They have an extensively developed theoretical background, enable the data to dictate the trend of the model, can accommodate missing observations and do not require constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more suitable statistical method to analyze non-linear biomedical longitudinal data.
-
- The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs.
-In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2011;@wood2016;@wood2017] that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g.
-hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-This programming language is also able to simulate data, an emerging strategy used to test statistical models [@haverkamp2017].
-This allows to create and  explore different alternatives for analysis without collecting information in the field, and reduces the time window between experiment design and its implementation.
-
-Therefore, the purpose of this study is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity and correlation structures.
-Secondly, the key theoretical elements of GAMS are presented without using extensive derivations or complex mathematical notation that can obscure their understanding.
-Thirdly, simulated data that follows the trend of previously reported values [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research and the implementation of GAMs in such scenario.
-Finally, reproducibility is emphasized by providing the code to generate the data and the implementation of GAMs in $\textsf{R}$, in conjunction with a step-by-step guide to fit models of increasing complexity.
-In summary, the exploration of modern statistical techniques to analyze longitudinal data may allow biomedical researchers to consider the use of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a linear trend, and will also help to improve the standards for reproducibility  in biomedical research.
-
-# Challenges presented by longitudinal studies
-
-## The "frequentist" case for longitudinal data
-
-The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research, but certain assumptions are necessary to make the model valid.
-From a practical view, they can be divided in three areas: linear relationship between covariates and response, constant correlation between measurements, and complete observations for all subjects.
-Each one of these assumptions is discussed below.
-
-## Linear relationship
-
-In a biomedical longitudinal study, two or more groups of subjects (patients, mice, samples) are subject to a different treatments (e.g.
-a "treatment" group receives a novel drug vs.
-a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response has both  _fixed_ and  _random_ components.
-The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the effect of the novel drug in the "treatment" group.The _random_ component can be defined as "noise" caused by some inherent variability within the study.
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while others are measured in the afternoon, it is possible that the difference in the collection time of the measurement introduces some "noise" in the signal.
-As their name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
-
-Mathematically speaking, a rm-ANOVA model can be written as:
-
-\begin{equation}
-  y_{i} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t} *                                    treatment_{j}+\pi_{ij} +\epsilon_{tij}\\ 
-  (\#eq:linear-model)
-\end{equation}
-
-In this model $y_i$ is the response by subject $i$, which can be decomposed in a mean value $\mu$, _fixed effects_ of time ($time_t$), treatment ($time_j$) and their interaction; a _random effect_ $\pi_{ij}$ of each subject within each group, and errors per subject per time per group $\epsilon_{tij}$.
-Suppose two treatments groups are used in the study.
-Models per group can be generated if 0 is used for the first treatment group (Group A) and 1 for the second treatment group (Group B).
-The linear models then become:
-
-\begin{equation}
-  y = \begin{cases}
-  \mu + \beta_1*time_{t}+\epsilon_{ti}   & \mbox{if Group A}\\
-  \mu + \beta_1 * time_{t} + \beta_2 * treatment_{j} +\beta_3* time_{t} * treatment_{j}+\pi_{i}+\epsilon_{tij}  & \mbox{if Group B}\\
-  \end{cases}
-  (\#eq:ANOVA-by-group)
-\end{equation}
-
-Furthermore, \@ref(eq:ANOVA-by-group) can be re-written under the generalized linear model  (GLM) framework [@nelder1972].
-The main difference in this case is that a GLM model  uses a _linking function_ to connect the observed value to the linear model.
-
-Thus,
-
-\begin{equation}
-  y =\mu +\boldsymbol{\beta}_1* f(time) + \beta_2* treatment+\boldsymbol{\beta}_3*             f(time)*treatment \\
-
-  y = \begin{cases}
-  \mu +\boldsymbol{\beta}_1* f(time)  & \mbox{if Group A} \\
-  \mu +\boldsymbol{\beta}_1* f(time)  + \beta_2* treatment + \boldsymbol{\beta}_3* f(time) *     treatment   & \mbox{if Group B}\\
-  \end{cases}
-  (\#eq:GLM)
-\end{equation}
-
-The linear model in \@ref(eq:linear-model) is a GLM where $f(time) = time$ with the identity function $f(x) = x$.
-
-Regardless of its presentation, \@ref(eq:linear-model), \@ref(eq:ANOVA-by-group) and \@ref(eq:GLM) show that the model expects a linear relationship between the covariates and the response.
-When the data does not follow a linear trend, the fit that the model produces does not accurately represent the changes in the data.
-To exemplify this consider simulated longitudinal data where the a normally distributed response of two groups of 10 subjects each is obtained, and an rm-ANOVA model such as \@ref(eq:linear-model) is fitted to the data (code for the simulated data is available in the Appendix).
-
-```{r, message = FALSE,include=FALSE}
+```{r, message = FALSE}
 set.seed(11)
@@
 
-```{r,include=FALSE,message=FALSE,echo=FALSE}
+```{r}
 ## Example with linear response
@@
 
-```{r, fig.width=16, fig.height=9, out.width="100%", echo=FALSE,message=FALSE}
+```{r, fig.width=16, fig.height=9, out.width="100%"}
+set.seed(111)
+# linear response, independent errors (no subject effect)
+plot_example(example(fun_type = "linear", error_type = "independent")) +
+  plot_annotation(
+  title = 'Simulated with linear function and independent errors', 
+  theme = theme(plot.title = element_text(size = 20))
+) #& 
+  # theme(plot.title = element_text(size = 20))
+```
+
+```{r, fig.width=16, fig.height=9, out.width="100%"}
 # linear response, correlated errors (subject effect)
@@
 ```
-It is clear from this graph that
+
+```{r, fig.width=16, fig.height=9, out.width="100%"}
+# quadratic response, independent errors (no subject effect)
+plot_example(example(fun_type = "quadratic", error_type = "independent")) +
+  plot_annotation(
+  title = 'Simulated with quadratic function and independent errors',
+    theme = theme(plot.title = element_text(size = 20))
+) #& 
+  # theme(plot.title = element_text(size = 20))
+```
 
@@
 ```
-
-## Covariance in rm-ANOVA and LMEMs
-
-In a longitudinal study there is an expected _variance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
-The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
-
-In the case of an rm-ANOVA analysis, it is assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
-Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
-However, it has been shown that this condition is frequently unjustified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser) the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
-In other words, if the data does not present constant correlation between repeated measurements, the assumptions required for an rm-ANOVA model are not met and the use of corrections may still not provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others [@pinheiro2006].
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself.
-Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
-
-## Missing observations
-
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator [molenberghs2004].Dropout from patients, or attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002].
-If the data are MCAR, this means that the missigness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
-
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR [@west2014].
-The pattern of missing observations can be considered MAR if the missing observations are not related any of the other variables measured in the study [@maxwell2017].
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.
-
-_What limitations LMEMs have regarding missing observations?_
-
-This section has presented the assumptions of rm-ANOVA to analyze longitudinal information and  its differences when compared to LMEMs  regarding to missing data and the modeling of the covariance matrix.
-Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high variability, LMEMs fail to capture the non-linear trend of the data.
-To analyze such type of data, we present generalized additive models (GAMs) as an alternative in the following section.
-
-# GAMs as a special case of Generalized Linear Models
-
-A GAM is a special case of the Generalized Linear Model (GLM), a framework that allows for response distributions that do not follow a normal distribution [@wood2017;@hastie1987].
-Following the notation by Simpson [@simpson2018] A GAM model can be represented as:
-
-\begin{equation}
-  y_{t}=\beta_0+f(x_t)+\epsilon_t  
-  (\#eq:GAM)
-\end{equation}
-
-Where $y_t$ is the response at time $t$, $\beta_0$ is the expected value at time 0, the change of $y$ over time is represented by the function $f(x_t)$ and $\epsilon_t$ represents the residuals.
-
-In contrast to rm-ANOVA or LMEMs, GAMs use _smooth functions_  to model the relationship between the covariates and the response.
-This approach is more advantageous than that of a LMEMs or rm-ANOVA as it does not restrict the model to a linear relationship.
-One possible function for $f(x_t)$  is a polynomial, but a major limitation is that they create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
-
-To overcome this limitation, the smooth functions in GAMs  are represented using _basis functions_  over  evenly spaced ranges of the covariates known as _knots_.
-The _basis function_ used is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together[@wood2017;@simpson2018].
-Cubic splines have a long history in their use to solve non-parametric statistical problems and are the default choice to fit GAMs as they are the simplest option to obtain visual smoothness [@wegman1983].
-
-***
-
-# References

### 2021-01-05 Code and miscellaneous/splines-missing-data.R

@@ -1,0 +1,113 @@
+library(mgcv)
+library(tidyverse)
+library(splines)
+library(rstan)
+library(rstanarm)
+library(patchwork)
+
+
+## number of observations -- currently using 3
+## use measurement error of sd as 5 (rough estimate)
+
+dat <- read_csv(here::here("data", "Vishwanath StO2.csv"))
+
+## plot the data
+ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
+    geom_line()
+
+simulate_data <- function(dat, n = 10, sd = 5) {
+    dat_sim <- dat %>%
+        slice(rep(1:n(), each = n)) %>%
+        group_by(Group, Day) %>%
+        mutate(observation = rnorm(n, StO2, sd),
+               observation_log = rlnorm(n, log(StO2), log(sd)),
+               observation_tobit = pmax(rnorm(n, StO2, sd), 0.0001)) %>%
+        ungroup()
+
+    return(dat_sim)
+}
+
+
+
+n <- 10
+sd <- 5
+set.seed(11)
+dat_sim <- simulate_data(dat, n, sd)
+df <- 6
+
+n_time <- length(unique(dat_sim$Day))
+n_treatment <- length(unique(dat_sim$Group))
+
+
+
+missing <- sample(1:100, 40)
+dat_missing <- dat_sim[-missing, ]
+
+
+ggplot(dat_missing, aes(x = Day, y = observation, color = Group)) +
+    geom_point() +
+    ggtitle("Simulated time-varying responses") +
+    scale_color_viridis_d(end = 0.75) +
+    geom_line(aes(y = StO2), alpha = 0.75) +
+    theme_bw()
+
+dat_missing %>%
+    group_by(Day, Group) %>%
+    summarize(count = n())
+
+
+dat_missing$Group <- as.factor(dat_missing$Group)
+dat_sim$Group <- as.factor(dat_sim$Group)
+mod1 <- gam(observation ~ s(Day, by = Group, k = 5), data  = dat_missing)
+mod2 <- gam(observation ~ s(Day, by = Group, k = 5), data  = dat_sim)
+
+summary(mod1)
+summary(mod2)
+
+par(ask=F)
+layout(matrix(1:4, 2, 2))
+plot(mod1, main = "Missing")
+plot(mod2, main = "Full")
+
+par(mfrow=c(1,1))
+
+gam.check(mod1)
+gam.check(mod2)
+
+#plotting smooths and simulated data
+#full data
[…]

### 2021-01-05 Code and miscellaneous/time-varying-anova-example.R

@@ -1,0 +1,280 @@
+library(tidyverse)
+library(splines)
+library(rstan)
+library(rstanarm)
+library(patchwork)
+
+
+## number of observations -- currently using 3
+## use measurement error of sd as 5 (rough estimate)
+
+dat <- read_csv(here::here("data", "Vishwanath StO2.csv"))
+
+## plot the data
+ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
+    geom_line()
+
+simulate_data <- function(dat, n = 10, sd = 5) {
+    dat_sim <- dat %>%
+        slice(rep(1:n(), each = n)) %>%
+        group_by(Group, Day) %>%
+        mutate(observation = rnorm(n, StO2, sd),
+               observation_log = rlnorm(n, log(StO2), log(sd)),
+               observation_tobit = pmax(rnorm(n, StO2, sd), 0.0001)) %>%
+        ungroup()
+
+    return(dat_sim)
+}
+
+
+
+n <- 10
+sd <- 5
+set.seed(11)
+dat_sim <- simulate_data(dat, n, sd)
+df <- 6
+
+n_time <- length(unique(dat_sim$Day))
+n_treatment <- length(unique(dat_sim$Group))
+
+## basis expansion over time that allows for smooth functional response
+X_bs <- bs(unique(dat_sim$Day), df = df, intercept = TRUE) #generates matrix for representing the family of piecewise polynomials with specificed knots
+# 2 knots as it is defined by df(6)-degree(3)-1 because of the intercept
+#each column of the matrix is a basis function
+matplot(unique(dat_sim$Day), X_bs, type = 'l')
+
+
+p1 <- ggplot(dat_sim, aes(x = Day, y = observation, color = Group)) +
+    geom_point() +
+    ggtitle("Simulated time-varying responses") +
+    scale_color_viridis_d(end = 0.75) +
+    geom_line(aes(y = StO2), alpha = 0.75) +
+    theme_bw()
+
+p2 <- ggplot(dat_sim, aes(x = Day, y = observation_log, color = Group)) +
+    geom_point() +
+    ggtitle("Simulated log-scale time-varying responses") +
+    scale_color_viridis_d(end = 0.75) +
+    geom_line(aes(y = StO2), alpha = 0.75) +
+    theme_bw()
+
+p3 <- ggplot(dat_sim, aes(x = Day, y = observation_tobit, color = Group)) +
+    geom_point() +
[…]

### 2021-01-05 Code and miscellaneous/time-varying-anova.R

@@ -1,0 +1,281 @@
+library(tidyverse)
+library(splines)
+library(rstan)
+library(rstanarm)
+
+set.seed(11)
+n_time      <- 10 #repeated observations
+n_treatment <- 3 #treatment groups
+n_reps      <- 5 #subjects per treatment
+## spline degrees of freedom
+## controls "wigglieness" of the response
+df          <- 4
+
+time      <- 1:n_time
+treatment <- 1:n_treatment
+
+## basis expansion over time that allows for smooth functional response
+X_bs <- bs(time, df = df) #generates matrix for representing the family of piecewise polynomials with specificed knots
+#and df
+## beta parameters
+## each treatment gets a different set of parameters
+beta <- matrix(rnorm(df * n_treatment), df, n_treatment) #generates samples from the normal distribution and
+#places them in a matrix with size df and n_treatment
+
+beta <- matrix(c(
+    4, 3, 2, 1,
+    -1, -2, -3, -4,
+    -4, -1, -1, -4), df, n_treatment)
+
+plot(X_bs %*% beta[, 1], ylim = c(0, 6), type = 'l')
+matplot(X_bs, type = 'l', add = TRUE)
+plot(X_bs %*% beta[, 2], ylim = c(-4, 4), type = 'l')
+matplot(t(t(X_bs) * beta[, 2]), type = 'l', add = TRUE)
+
+
+
+## mean time-varying response within each group
+mu <- X_bs %*% beta #matrix multiplication to generate the mean time-varying response
+
+dat_mu <- data.frame(
+    mu        = c(mu),
+    time      = rep(1:n_time, times = n_treatment),
+    treatment = factor(rep(1:n_treatment, each = n_time))
+)
+
+#this data frame has mean response for 10 observations for 3 treatment groups
+dat_mu %>%
+    ggplot(aes(x = time, y = mu, group = treatment, color = treatment)) +
+    geom_line() +
+    scale_color_viridis_d(end = 0.75) +
+    ggtitle("Simulated time-varying mean responses")
+
+## measurement/individual response processes
+## can make this more specific to the model as needed
+sigma <- 0.25
+
+##
+## generate the data
+##
+#generates empty arrays. n_time rows, n_treatment columns, n_reps (number of subjects per treatment group) number of arrays and then puts names on the rows
+#and columns of each array
[…]

### 2021-01-05 splines-missing-data.R
Moved files to 'Code and miscellaneous'

@@ -1,113 +1,0 @@
-library(mgcv)
-library(tidyverse)
-library(splines)
-library(rstan)
-library(rstanarm)
-library(patchwork)
-
-
-## number of observations -- currently using 3
-## use measurement error of sd as 5 (rough estimate)
-
-dat <- read_csv(here::here("data", "Vishwanath StO2.csv"))
-
-## plot the data
-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
-    geom_line()
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(observation = rnorm(n, StO2, sd),
-               observation_log = rlnorm(n, log(StO2), log(sd)),
-               observation_tobit = pmax(rnorm(n, StO2, sd), 0.0001)) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-
-
-n <- 10
-sd <- 5
-set.seed(11)
-dat_sim <- simulate_data(dat, n, sd)
-df <- 6
-
-n_time <- length(unique(dat_sim$Day))
-n_treatment <- length(unique(dat_sim$Group))
-
-
-
-missing <- sample(1:100, 40)
-dat_missing <- dat_sim[-missing, ]
-
-
-ggplot(dat_missing, aes(x = Day, y = observation, color = Group)) +
-    geom_point() +
-    ggtitle("Simulated time-varying responses") +
-    scale_color_viridis_d(end = 0.75) +
-    geom_line(aes(y = StO2), alpha = 0.75) +
-    theme_bw()
-
-dat_missing %>%
-    group_by(Day, Group) %>%
-    summarize(count = n())
-
-
-dat_missing$Group <- as.factor(dat_missing$Group)
-dat_sim$Group <- as.factor(dat_sim$Group)
-mod1 <- gam(observation ~ s(Day, by = Group, k = 5), data  = dat_missing)
-mod2 <- gam(observation ~ s(Day, by = Group, k = 5), data  = dat_sim)
-
-summary(mod1)
-summary(mod2)
-
-par(ask=F)
-layout(matrix(1:4, 2, 2))
-plot(mod1, main = "Missing")
-plot(mod2, main = "Full")
-
-par(mfrow=c(1,1))
-
-gam.check(mod1)
-gam.check(mod2)
-
-#plotting smooths and simulated data
-#full data
[…]

### 2021-01-05 time-varying-anova-example.R

@@ -1,280 +1,0 @@
-library(tidyverse)
-library(splines)
-library(rstan)
-library(rstanarm)
-library(patchwork)
-
-
-## number of observations -- currently using 3
-## use measurement error of sd as 5 (rough estimate)
-
-dat <- read_csv(here::here("data", "Vishwanath StO2.csv"))
-
-## plot the data
-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
-    geom_line()
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(observation = rnorm(n, StO2, sd),
-               observation_log = rlnorm(n, log(StO2), log(sd)),
-               observation_tobit = pmax(rnorm(n, StO2, sd), 0.0001)) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-
-
-n <- 10
-sd <- 5
-set.seed(11)
-dat_sim <- simulate_data(dat, n, sd)
-df <- 6
-
-n_time <- length(unique(dat_sim$Day))
-n_treatment <- length(unique(dat_sim$Group))
-
-## basis expansion over time that allows for smooth functional response
-X_bs <- bs(unique(dat_sim$Day), df = df, intercept = TRUE) #generates matrix for representing the family of piecewise polynomials with specificed knots
-# 2 knots as it is defined by df(6)-degree(3)-1 because of the intercept
-#each column of the matrix is a basis function
-matplot(unique(dat_sim$Day), X_bs, type = 'l')
-
-
-p1 <- ggplot(dat_sim, aes(x = Day, y = observation, color = Group)) +
-    geom_point() +
-    ggtitle("Simulated time-varying responses") +
-    scale_color_viridis_d(end = 0.75) +
-    geom_line(aes(y = StO2), alpha = 0.75) +
-    theme_bw()
-
-p2 <- ggplot(dat_sim, aes(x = Day, y = observation_log, color = Group)) +
-    geom_point() +
-    ggtitle("Simulated log-scale time-varying responses") +
-    scale_color_viridis_d(end = 0.75) +
-    geom_line(aes(y = StO2), alpha = 0.75) +
-    theme_bw()
-
-p3 <- ggplot(dat_sim, aes(x = Day, y = observation_tobit, color = Group)) +
-    geom_point() +
[…]

### 2021-01-05 time-varying-anova.R

@@ -1,281 +1,0 @@
-library(tidyverse)
-library(splines)
-library(rstan)
-library(rstanarm)
-
-set.seed(11)
-n_time      <- 10 #repeated observations
-n_treatment <- 3 #treatment groups
-n_reps      <- 5 #subjects per treatment
-## spline degrees of freedom
-## controls "wigglieness" of the response
-df          <- 4
-
-time      <- 1:n_time
-treatment <- 1:n_treatment
-
-## basis expansion over time that allows for smooth functional response
-X_bs <- bs(time, df = df) #generates matrix for representing the family of piecewise polynomials with specificed knots
-#and df
-## beta parameters
-## each treatment gets a different set of parameters
-beta <- matrix(rnorm(df * n_treatment), df, n_treatment) #generates samples from the normal distribution and
-#places them in a matrix with size df and n_treatment
-
-beta <- matrix(c(
-    4, 3, 2, 1,
-    -1, -2, -3, -4,
-    -4, -1, -1, -4), df, n_treatment)
-
-plot(X_bs %*% beta[, 1], ylim = c(0, 6), type = 'l')
-matplot(X_bs, type = 'l', add = TRUE)
-plot(X_bs %*% beta[, 2], ylim = c(-4, 4), type = 'l')
-matplot(t(t(X_bs) * beta[, 2]), type = 'l', add = TRUE)
-
-
-
-## mean time-varying response within each group
-mu <- X_bs %*% beta #matrix multiplication to generate the mean time-varying response
-
-dat_mu <- data.frame(
-    mu        = c(mu),
-    time      = rep(1:n_time, times = n_treatment),
-    treatment = factor(rep(1:n_treatment, each = n_time))
-)
-
-#this data frame has mean response for 10 observations for 3 treatment groups
-dat_mu %>%
-    ggplot(aes(x = time, y = mu, group = treatment, color = treatment)) +
-    geom_line() +
-    scale_color_viridis_d(end = 0.75) +
-    ggtitle("Simulated time-varying mean responses")
-
-## measurement/individual response processes
-## can make this more specific to the model as needed
-sigma <- 0.25
-
-##
-## generate the data
-##
-#generates empty arrays. n_time rows, n_treatment columns, n_reps (number of subjects per treatment group) number of arrays and then puts names on the rows
-#and columns of each array
[…]

### 2021-01-05 Manuscripts/Manuscript_AM_v1.Rmd
Updated manuscript, pending Dr. Tipton's comments. PDF works, Word works, HTML works

--- 2021-01-05 Manuscripts/Manuscript_AM_v1.Rmd
@@
 ---
-title: '**The statistical analysis of non-linear longitudinal data in biomedical research**'
+title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
 author:
@@
 subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
-bibliography: refs.bib
-csl: ieee.csl
-css: style.css
-fontfamily: Arial
 output:
-
-  bookdown::html_document2: default
   bookdown::pdf_document2:
@@
     fig_caption: yes
+    extra_dependencies: ["subfig","breqn"]
   bookdown::word_document2: default
-  link-citations: true
+  bookdown::html_document2: default
+  link-citations: yes
+  css: style.css
+  '': default
+csl: ieee.csl
+bibliography: refs.bib
 ---
@@
 \begin{equation}
-  y_{i} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t} *                                    treatment_{j}+\pi_{ij} +\epsilon_{tij}\\ 
+y_{i} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t}*treatment_{j}+\pi_{ij} +\epsilon_{tij}\\ 
 (\#eq:linear-model)
@@
 \begin{equation}
-  y =\mu +\boldsymbol{\beta}_1* f(time) + \beta_2* treatment+\boldsymbol{\beta}_3*             f(time)*treatment \\
-
-  y = \begin{cases}
-  \mu +\boldsymbol{\beta}_1* f(time)  & \mbox{if Group A} \\
-  \mu +\boldsymbol{\beta}_1* f(time)  + \beta_2* treatment + \boldsymbol{\beta}_3* f(time) *     treatment   & \mbox{if Group B}\\
-  \end{cases}
+  y =\mu +\boldsymbol{\beta}_1* f(time) + \beta_2* treatment+ \\
+  \boldsymbol{\beta}_3* f(time)*treatment \\
   (\#eq:GLM)
@@
 The linear model in \@ref(eq:linear-model) is a GLM where $f(time) = time$ with the identity function $f(x) = x$.
+If the GLM is then written in terms of two treatment groups as in \@ref(eq:ANOVA-by-group), the model becomes:
 
-Regardless of its presentation, \@ref(eq:linear-model), \@ref(eq:ANOVA-by-group) and \@ref(eq:GLM) show that the model expects a linear relationship between the covariates and the response.
+  \begin{equation}
+  y = \begin{cases}
+  \mu +\boldsymbol{\beta}_1* f(time)  & \mbox{if Group A} \\
+  \mu +\boldsymbol{\beta}_1* f(time)  + \beta_2* treatment + \boldsymbol{\beta}_3* f(time) *treatment   
+  & \mbox{if Group B}\\
+  \end{cases}
+  (\#eq:GLM-by-group)
+\end{equation}
+
+Regardless of its presentation, \@ref(eq:linear-model), \@ref(eq:ANOVA-by-group) and \@ref(eq:GLM-by-group) show that the model expects a linear relationship between the covariates and the response.
 When the data does not follow a linear trend, the fit that the model produces does not accurately represent the changes in the data.
-To exemplify this consider simulated longitudinal data where the a normally distributed response of two groups of 10 subjects each is obtained, and an rm-ANOVA model such as \@ref(eq:linear-model) is fitted to the data (code for the simulated data is available in the Appendix).
+To exemplify this consider simulated longitudinal data where the a normally distributed response of two groups of 10 subjects each is obtained, and an rm-ANOVA model identical to \@ref(eq:linear-model) is fitted to the data using the package _nlme_ (the code for the simulated data and models is available in the Appendix).
 
@@
     # nonlinear response
+
     mu[, 1] <- - 4 * (x - .5)^2 + 0.25
@@
 
-```{r, fig.width=16, fig.height=9, out.width="100%", echo=FALSE,message=FALSE}
+```{r linear-model, fig.width=16, fig.height=9, out.width="100%", echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with a rm-ANOVA model fitted.
+Top row, simulated data, lines represent mean response.
+Bottom row, fitted model, thick lines represent predicted mean response "}
 # linear response, correlated errors (subject effect)
@@
 ```
-It is clear from this graph that
 
-```{r, fig.width=16, fig.height=9, out.width="100%"}
+It is clear from \@ref(fig:linear-model) that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are identical to the "truth" of the simulated data.
+
+However, consider the case when the data follows a non-linear trend, such as the one simulated in \@ref(fig:quadratic-response).
+Here, the simulated data follows a quadratic behavior.
+It is clear from the figure that changes in each group occur through the timeline, although the final mean value is the same as the initial value.
+Fitting an rm-ANOVA model like \@ref(eq:linear-model) again produces the fit that appears at the bottom of \@ref(fig:quadratic-response).
+
+```{r quadratic-response, fig.width=16, fig.height=9, out.width="100%",echo=FALSE,fig.cap='Simulated quadratic responses from two groups with an rm-ANOVA model fitted.
+Top row, simulated data, lines represent mean response.
+Bottom row, fitted model, thick lines represent predicted mean response', message=FALSE}
 # quadratic response, correlated errors (subject effect)
@@
 ```
+
+In this case, when the predictions are compared to the simulated data it is clear that the model is not capturing the changes within each group throughout the timeline.
+This highlights the limitation of rm-ANOVA and LMEMs with longitudinal non-linear data, where the variations are not captured and modeled properly.
 
@@
 
-_What limitations LMEMs have regarding missing observations?_
-
 This section has presented the assumptions of rm-ANOVA to analyze longitudinal information and  its differences when compared to LMEMs  regarding to missing data and the modeling of the covariance matrix.
@@
 In contrast to rm-ANOVA or LMEMs, GAMs use _smooth functions_  to model the relationship between the covariates and the response.
-This approach is more advantageous than that of a LMEMs or rm-ANOVA as it does not restrict the model to a linear relationship.
+This approach is more advantageous as it does not restrict the model to a linear relationship.
 One possible function for $f(x_t)$  is a polynomial, but a major limitation is that they create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
@@
 Cubic splines have a long history in their use to solve non-parametric statistical problems and are the default choice to fit GAMs as they are the simplest option to obtain visual smoothness [@wegman1983].
+Therefore, GAMs overcome the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear, such as \@ref(fig:quadratic-response).
+Regarding longitudinal data, Pedersen et al [@pedersen2019] demonstrated the capabilities of GAMs in this area using ecological data.
+
+The use of GAMs to analyze biomedical longitudinal data, and the impact of missing observations in the fit of the model will be examined in detail in the following section using simulation.
 

### 2021-01-05 Manuscripts/Manuscript_AM_v1.tex
Updated manuscript, pending Dr. Tipton's comments. PDF works, Word works, HTML works

@@
 \hypersetup{
-  pdftitle={The statistical analysis of non-linear longitudinal data in biomedical research},
+  pdftitle={The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models},
   pdfauthor={Ariel Mundo; Timothy J. Muldoon; John R. Tipton},
@@
 \setcounter{secnumdepth}{5}
+\usepackage{subfig}
+\usepackage{breqn}
 
-\title{\textbf{The statistical analysis of non-linear longitudinal data in biomedical research}}
+\title{\textbf{The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models}}
+\usepackage{etoolbox}
+\makeatletter
+\providecommand{\subtitle}[1]{% add subtitle to \maketitle
+  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
+}
+\makeatother
+\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
 \author{Ariel Mundo\footnote{Department of Biomedical Engineering, University of Arkansas, Fayetteville} \and Timothy J.
@@
 In the first case, weekly measurements of skin toxicities in patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT).
-Longitudinal studies have been also used to analyze tumor response {[}3{]}--{[}6{]}, antibody expression {[}7{]}, {[}8{]}, and cell metabolism {[}9{]}, {[}10{]}.
+Longitudinal studies have beenn also used to analyze tumor response {[}3{]}--{[}6{]}, antibody expression {[}7{]}, {[}8{]}, and cell metabolism {[}9{]}, {[}10{]}.
 
@@
 \begin{equation}
-  y_{i} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t} *                                    treatment_{j}+\pi_{ij} +\epsilon_{tij}\\ 
+y_{i} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t}*treatment_{j}+\pi_{ij} +\epsilon_{tij}\\ 
 \label{eq:linear-model}
@@
 \begin{equation}
-  y =\mu +\boldsymbol{\beta}_1* f(time) + \beta_2* treatment+\boldsymbol{\beta}_3*             f(time)*treatment \\
-
-  y = \begin{cases}
-  \mu +\boldsymbol{\beta}_1* f(time)  & \mbox{if Group A} \\
-  \mu +\boldsymbol{\beta}_1* f(time)  + \beta_2* treatment + \boldsymbol{\beta}_3* f(time) *     treatment   & \mbox{if Group B}\\
-  \end{cases}
+  y =\mu +\boldsymbol{\beta}_1* f(time) + \beta_2* treatment+ \\
+  \boldsymbol{\beta}_3* f(time)*treatment \\
   \label{eq:GLM}
@@
 The linear model in \eqref{eq:linear-model} is a GLM where \(f(time) = time\) with the identity function \(f(x) = x\).
+If the GLM is then written in terms of two treatment groups as in \eqref{eq:ANOVA-by-group}, the model becomes:
 
-Regardless of its presentation, \eqref{eq:linear-model}, \eqref{eq:ANOVA-by-group} and \eqref{eq:GLM} show that the model expects a linear relationship between the covariates and the response.
+\begin{equation}
+  y = \begin{cases}
+  \mu +\boldsymbol{\beta}_1* f(time)  & \mbox{if Group A} \\
+  \mu +\boldsymbol{\beta}_1* f(time)  + \beta_2* treatment + \boldsymbol{\beta}_3* f(time) *treatment   
+  & \mbox{if Group B}\\
+  \end{cases}
+  \label{eq:GLM-by-group}
+\end{equation}
+
+Regardless of its presentation, \eqref{eq:linear-model}, \eqref{eq:ANOVA-by-group} and \eqref{eq:GLM-by-group} show that the model expects a linear relationship between the covariates and the response.
 When the data does not follow a linear trend, the fit that the model produces does not accurately represent the changes in the data.
-To exemplify this consider simulated longitudinal data where the a normally distributed response of two groups of 10 subjects each is obtained, and an rm-ANOVA model such as \eqref{eq:linear-model} is fitted to the data (code for the simulated data is available in the Appendix).
+To exemplify this consider simulated longitudinal data where the a normally distributed response of two groups of 10 subjects each is obtained, and an rm-ANOVA model identical to \eqref{eq:linear-model} is fitted to the data using the package \emph{nlme} (the code for the simulated data and models is available in the Appendix).
 
-\includegraphics[width=1\linewidth]{Manuscript_AM_v1_files/figure-latex/unnamed-chunk-3-1}
+\begin{figure}
+\includegraphics[width=1\linewidth]{Manuscript_AM_v1_files/figure-latex/linear-model-1} \caption{Simulated linear responses from two groups with a rm-ANOVA model fitted.
+Top row, simulated data, lines represent mean response.
+Bottom row, fitted model, thick lines represent predicted mean response }\label{fig:linear-model}
+\end{figure}
+
+It is clear from \ref{fig:linear-model} that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are identical to the ``truth'' of the simulated data.
+
+However, consider the case when the data follows a non-linear trend, such as the one simulated in \ref{fig:quadratic-response}.
+Here, the simulated data follows a quadratic behavior.
+It is clear from the figure that changes in each group occur through the timeline, although the final mean value is the same as the initial value.
+Fitting an rm-ANOVA model like \eqref{eq:linear-model} again produces the fit that appears at the bottom of \ref{fig:quadratic-response}.
+
+\begin{figure}
+\includegraphics[width=1\linewidth]{Manuscript_AM_v1_files/figure-latex/quadratic-response-1} \caption{Simulated quadratic responses from two groups with an rm-ANOVA model fitted.
+Top row, simulated data, lines represent mean response.
+Bottom row, fitted model, thick lines represent predicted mean response}\label{fig:quadratic-response}
+\end{figure}
+
+In this case, when the predictions are compared to the simulated data it is clear that the model is not capturing the changes within each group throughout the timeline.
+This highlights the limitation of rm-ANOVA and LMEMs with longitudinal non-linear data, where the variations are not captured and modeled properly.
 
@@
 
-\emph{What limitations LMEMs have regarding missing observations?}
-
 This section has presented the assumptions of rm-ANOVA to analyze longitudinal information and its differences when compared to LMEMs regarding to missing data and the modeling of the covariance matrix.
@@
 In contrast to rm-ANOVA or LMEMs, GAMs use \emph{smooth functions} to model the relationship between the covariates and the response.
-This approach is more advantageous than that of a LMEMs or rm-ANOVA as it does not restrict the model to a linear relationship.
+This approach is more advantageous as it does not restrict the model to a linear relationship.
 One possible function for \(f(x_t)\) is a polynomial, but a major limitation is that they create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}23{]}.
@@
 Cubic splines have a long history in their use to solve non-parametric statistical problems and are the default choice to fit GAMs as they are the simplest option to obtain visual smoothness {[}40{]}.
+Therefore, GAMs overcome the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear, such as \ref{fig:quadratic-response}.
+Regarding longitudinal data, Pedersen et al {[}20{]} demonstrated the capabilities of GAMs in this area using ecological data.
+
+The use of GAMs to analyze biomedical longitudinal data, and the impact of missing observations in the fit of the model will be examined in detail in the following section using simulation.
 

### 2021-01-15 Manuscripts/Manuscript_AM_v1.tex
Updated refernces, stared v2 of manuscript

@@
 L.
-Simpson, ``An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, uk,'' \emph{Geochimica et Cosmochimica Acta}, vol.
+Simpson, ``An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to Scotland, UK,'' \emph{Geochimica et Cosmochimica Acta}, vol.
 82, pp.
@@
 Wood, N., Pya, and B.
-S"afken, ``Smoothing parameter and model selection for general smooth models (with discussion),'' \emph{Journal of the American Statistical Association}, vol.
+Säfken, ``Smoothing parameter and model selection for general smooth models (with discussion),'' \emph{Journal of the American Statistical Association}, vol.
 111, pp.

### 2021-01-15 Manuscripts/Manuscript_AM_v2.Rmd
Updated refernces, stared v2 of manuscript

--- 2021-01-05 Manuscripts/Manuscript_AM_v1.Rmd
@@
 
-A longitudinal study is defined as that which is designed to repeatedly measure a variable of interest in a group (or groups) of subjects.
-In biomedical research, this type of study is employed when the intention is to observe the evolution of the effect of a certain treatment across time, rather than analyzing it at a single time point (a cross-sectional study).
-Clinical studies of breast and neck cancer have used a longitudinal approach [@sio2016;@kamstra2015].
-In the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT).
-Longitudinal studies have beenn also used to analyze tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
+A longitudinal study is designed to repeatedly measure a variable of interest in a group (or groups) of subjects, when the intention is to observe the evolution of the effect of a certain treatment across time rather than analyzing it at a single time point (e.g., a cross-sectional study).
+Medical research commonly uses longitudinal studies to analyze the evolution of patients before, during and after treatment.
+For example, Sio et al analyzed the evolution of dermatitis symptoms in breast cancer patients after radiotherapy (RT), by taking weekly measurements over the course of two moths [@sio2016], and Kamstra et al measured mouth opening at regular monthly intervals after RT in neck cancer patients [@kamstra2015].
+Biomedical research has used longitudinal studies as well to analyze tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
 
-Traditionally, a “frequentist” approach is used in biomedical research to derive inferences from a longitudinal study.
-Such statistical view derives its name from the fact that it regards probability as a limiting frequency[@wagenmakers2008] and its application is based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
+Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
+The frequentist paradigm regards probability as a limiting frequency[@wagenmakers2008] by assuming a null hypothesis using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
 This methodology makes two key assumptions regarding longitudinal data: constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study (a condition also known as complete observations) [@gueorguieva2004;@schober2018].
 
-However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases [@ugrinowitsch2004]. 
-In the case of the number of observations during the study, different situations can arise that prevent the collection of complete measurements for all subjects:  in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and it is even possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point and therefore violating the _complete observations_ assumption of rm-ANOVA.
+The assumption of constant correlation is frequently unjustified as correlation in time tends to diminish between measurements when the time interval between them increases [@ugrinowitsch2004]. 
+Due to a variety of causes, the number of observations during a study can vary between all subjects.
+For example, in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and it is even possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point and therefore violating the _complete observations_ assumption of  classical rm-ANOVA.
 
-When incomplete observations occur, an rm-ANOVA requires the exclusion of all subjects with missing observations from the analysis.
+When incomplete observations occur, an rm-ANOVA is fit by excluding all subjects with missing observations from the analysis[@gueorguieva2004].
 This can result in increased costs if the desired statistical power is not met with the remaining observations, as it would be necessary to enroll more subjects.
@@
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-This is exemplified by the  use of  linear mixed effects models (LMEMs)  in certain groups to analyze longitudinal data [@skala2010;@vishwanath2009].
-Briefly, these models incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g.
+The recognition on the limitations of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal data [@skala2010;@vishwanath2009].
+Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g.
 the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population.
-These models are more flexible than the traditional rm-ANOVA as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-On the other hand, they impose restrictions in the distribution of the errors  of the random effects [@gueorguieva2004].
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
+On the other hand, LMEMs impose restrictions in the distribution of the errors  of the random effects [@gueorguieva2004].
 
-One final assumption that is not initially evident for  both rm-ANOVA and LMEMs models pertains the relationship between the observed effect and the progression of the timeline of the study, which is expected to be linear [@pinheiro2006].
-This common assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend.
-In biomedical research, a special case of this behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
+One final assumption that is not initially evident for both rm-ANOVA and LMEMs models is that the mean response is expected to change linearly through time[@pinheiro2006].
+The linearity assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend.
+In biomedical research, a particular case of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent with the observed variation.
-In other words, the in both cases the model tries to explain highly-variable data using a linear trend, consequently biasing the estimates.
-Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) by using multiple comparisons between the model terms, the use of this estimator is limited in such scenario because it is intended to work in models that have a reasonable agreement with the data.
+In other words, both LMEMs and rm-ANOVA explain highly-variable data using a linear trend, consequently biasing the estimates when the true response is non-linear in time.
+Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) it has to be reminded that the validity of such metric relies on how adequate the model is to the data.
+In other words, the _p-value_ requires that the model and the data have good agreement and if that is not the case, a "Type 3" error (known as "model misspecification") occurs[@dennis2019].That would be the case if for example, a model that is only able to explain linear responses is fitted to data that has a quadratic behavior.
+In such scenario, the _p-value_ becomes invalid[@wang2019].
 
-As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, while enabling inferences that are both adequate and consistent from a statistical perspective.
-In this regard, Generalized Additive Models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and political science [@beck1998] .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: They have an extensively developed theoretical background, enable the data to dictate the trend of the model, can accommodate missing observations and do not require constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more suitable statistical method to analyze non-linear biomedical longitudinal data.
+As  the “frequentist” rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, enabling inference that is appropiate to the data.
+In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
+Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend of the model, can easily accommodate missing observations and can model non-constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data.
 
  The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs.
-In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2011;@wood2016;@wood2017] that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g.
+In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g.
 hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-This programming language is also able to simulate data, an emerging strategy used to test statistical models [@haverkamp2017].
+At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017].
 This allows to create and  explore different alternatives for analysis without collecting information in the field, and reduces the time window between experiment design and its implementation.
 
-Therefore, the purpose of this study is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity and correlation structures.
-Secondly, the key theoretical elements of GAMS are presented without using extensive derivations or complex mathematical notation that can obscure their understanding.
-Thirdly, simulated data that follows the trend of previously reported values [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research and the implementation of GAMs in such scenario.
+The purpose of this work is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas.
+First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity of response and constant correlation structures.
+Second, the key theoretical elements of GAMs are presented without using extensive derivations or complex mathematical notation that can obscure their understanding.
+Third, simulated data that follows the trend of previously reported values [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research to highlight the differences in inference between rm-ANOVA, LMEMs and GAMs and the implementation of the latter in such scenario.
 Finally, reproducibility is emphasized by providing the code to generate the data and the implementation of GAMs in $\textsf{R}$, in conjunction with a step-by-step guide to fit models of increasing complexity.
@@
 
-The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research, but certain assumptions are necessary to make the model valid.
-From a practical view, they can be divided in three areas: linear relationship between covariates and response, constant correlation between measurements, and complete observations for all subjects.
+The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research, but certain assumptions are necessary for the model to be valid.
+From a practical view, the assumptions can be divided in three areas: a linear relationship between covariates and response, a constant correlation between measurements, and complete observations for all subjects.
 Each one of these assumptions is discussed below.
@@
 
-In a biomedical longitudinal study, two or more groups of subjects (patients, mice, samples) are subject to a different treatments (e.g.
-a "treatment" group receives a novel drug vs.
+In a biomedical longitudinal study, two or more groups of subjects (patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs.
 a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
 The collected response has both  _fixed_ and  _random_ components.
-The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the effect of the novel drug in the "treatment" group.The _random_ component can be defined as "noise" caused by some inherent variability within the study.
+The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the effect of the novel drug in the "treatment" group.
+The _random_ component can be defined as "noise" caused by some inherent variability within the study.
 For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while others are measured in the afternoon, it is possible that the difference in the collection time of the measurement introduces some "noise" in the signal.
@@
 
-Mathematically speaking, a rm-ANOVA model can be written as:
+Mathematically speaking, a rm-ANOVA model with an interaction can be written as:
 
 \begin{equation}
-y_{i} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t}*treatment_{j}+\pi_{ij} +\epsilon_{tij}\\ 
+y_{ijt} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t}*treatment_{j}+\pi_{ij} +\varepsilon_{tij}\\ 
 (\#eq:linear-model)
@@
 
-In this model $y_i$ is the response by subject $i$, which can be decomposed in a mean value $\mu$, _fixed effects_ of time ($time_t$), treatment ($time_j$) and their interaction; a _random effect_ $\pi_{ij}$ of each subject within each group, and errors per subject per time per group $\epsilon_{tij}$.
+In this model $y_{ijt}$ is the response by subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\mu$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$.
+A _random effect_ $\pi_{ij}$ of each subject within each group, and independent errors per subject per time per group $\varepsilon_{tij}$, that represent random variation not explained by the _fixed_ or _random_ effects.
 Suppose two treatments groups are used in the study.
-Models per group can be generated if 0 is used for the first treatment group (Group A) and 1 for the second treatment group (Group B).
+Then the models per group can be written as below with $j=0$ representing the first treatment group (Group A) and $j=1$ representing the second treatment group (Group B).
 The linear models then become:
@@
 \begin{equation}
-y = \begin{cases}
-\mu + \beta_1*time_{t}+\epsilon_{ti}   & \mbox{if Group A}\\
-\mu + \beta_1 * time_{t} + \beta_2 * treatment_{j} +\beta_3* time_{t} * treatment_{j}+\pi_{i}+\epsilon_{tij}  & \mbox{if Group B}\\
+y_{ijt} = \begin{cases}
+\mu + \beta_1*time_{t}+\varepsilon_{ijt}+\pi_{ijt}   & \mbox{if Group A}\\
+\mu + \beta_1 * time_{t} + \beta_2 * treatment_{j} +\beta_3* time_{t} * treatment_{j}+\pi_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
 \end{cases}

### 2021-01-18 Manuscripts/Manuscript_AM_v2.Rmd
Most changes performed, need to update the basis function section

@@
 output:
+  bookdown::html_document2: default
+  bookdown::word_document2: default
   bookdown::pdf_document2:
@@
     extra_dependencies: ["subfig","breqn"]
-  bookdown::word_document2: default
-  bookdown::html_document2: default
   link-citations: yes
@@
 
-Furthermore, \@ref(eq:ANOVA-by-group) can be re-written under the generalized linear model  (GLM) framework [@nelder1972].
-The main difference in this case is that a GLM model  uses a _linking function_ to connect the observed value to the linear model.
-
-Thus,
+To further simplify the expression, let us substitute $\widetilde{\mu}=\mu+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B. The equation then becomes:
 
 \begin{equation}
-  y =\mu +\boldsymbol{\beta}_1* f(time) + \beta_2* treatment+ \\
-  \boldsymbol{\beta}_3* f(time)*treatment \\
-  (\#eq:GLM)
+y_{ijt} = \begin{cases}
+\mu + \beta_1*time_{t}+\varepsilon_{ijt}+\pi_{ijt}   & \mbox{if Group A}\\
+\widetilde{mu} + \widetilde{\beta_1} * time_{t} +\pi_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\end{cases}
+(\#eq:ANOVA-lines)
 \end{equation}
 
-The linear model in \@ref(eq:linear-model) is a GLM where $f(time) = time$ with the identity function $f(x) = x$.
-If the GLM is then written in terms of two treatment groups as in \@ref(eq:ANOVA-by-group), the model becomes:
-
-  \begin{equation}
-  y = \begin{cases}
-  \mu +\boldsymbol{\beta}_1* f(time)  & \mbox{if Group A} \\
-  \mu +\boldsymbol{\beta}_1* f(time)  + \beta_2* treatment + \boldsymbol{\beta}_3* f(time) *treatment   
-  & \mbox{if Group B}\\
-  \end{cases}
-  (\#eq:GLM-by-group)
-\end{equation}
-
-Regardless of its presentation, \@ref(eq:linear-model), \@ref(eq:ANOVA-by-group) and \@ref(eq:GLM-by-group) show that the model expects a linear relationship between the covariates and the response.
-When the data does not follow a linear trend, the fit that the model produces does not accurately represent the changes in the data.
-To exemplify this consider simulated longitudinal data where the a normally distributed response of two groups of 10 subjects each is obtained, and an rm-ANOVA model identical to \@ref(eq:linear-model) is fitted to the data using the package _nlme_ (the code for the simulated data and models is available in the Appendix).
+Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different slopes per group).
+In other words, When it is stated that an rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as \@ref(eq:linear-model), \@ref(eq:ANOVA-by-group) or \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data, the maximum flexibility being that it can assign different slopes to each group.
+Because the construction of the model is essentially the same in LMEMs, the expected linearity holds in such models, although it is possible to fit models with different slopes at the subject-level[@pinheiro2006].
+When the data does not follow a linear trend, the fit that an rm-ANOVA or LMEM model produces does not accurately represent the trends in the data.
+To demonstrate this, longitudinal data where a normally distributed response of two groups of 10 subjects each is simulated, and a rm-ANOVA model  (\@ref(eq:linear-model)) is fitted to the data using the package _nlme_ (the code for the simulated data and models is available in the Appendix).
 
@@
   library(nlme)
+  library(ggsci)
 
@@
     theme_bw(base_size = 12) +
-    ggtitle("Simulated data with true response function")
+    ggtitle("Simulated data with true response function")+
+    scale_color_aaas()
 
@@
     ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_bw(base_size = 12)
+    theme_bw(base_size = 12)+
+    scale_color_aaas()
 
@@
     ggtitle("Simulated errors") +
-    theme_bw(base_size = 12)
+    theme_bw(base_size = 12)+
+    scale_color_aaas()
 
@@
     theme_bw(base_size = 12) +
-    ggtitle("Fitted Model")
+    ggtitle("Fitted Model")+
+    scale_color_aaas()
+
   return((p1 + p3) / (p2 + p4))
@@
 
-```{r linear-model, fig.width=16, fig.height=9, out.width="100%", echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with a rm-ANOVA model fitted.
+```{r, linear-model, fig.width=16, fig.height=9, out.width="100%", echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with a rm-ANOVA model fitted.
 Top row, simulated data, lines represent mean response.
-Bottom row, fitted model, thick lines represent predicted mean response "}
+Bottom row, fitted model, thick lines represent predicted mean response"}
 # linear response, correlated errors (subject effect)
+txt<-20
 plot_example(example(fun_type = "linear", error_type = "correlated")) +
@@
   title = 'Simulated with linear function and correlated errors',
-    theme = theme(plot.title = element_text(size = 20))
-) #& 
-  # theme(plot.title = element_text(size = 20))
+    theme = theme(plot.title = element_text(size = txt),text=element_text(size=txt))
+)
 ```
 
-It is clear from \@ref(fig:linear-model) that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are identical to the "truth" of the simulated data.
+It is clear from Figure \@ref(fig:linear-model) that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the "truth" of the simulated data.
 
-However, consider the case when the data follows a non-linear trend, such as the one simulated in \@ref(fig:quadratic-response).
+However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:quadratic-response).
 Here, the simulated data follows a quadratic behavior.
-It is clear from the figure that changes in each group occur through the timeline, although the final mean value is the same as the initial value.
-Fitting an rm-ANOVA model like \@ref(eq:linear-model) again produces the fit that appears at the bottom of \@ref(fig:quadratic-response).
+The figure shows that changes in each group occur through the timeline, although the final mean value is the same as the initial value.
+Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears at the bottom right of Figure \@ref(fig:quadratic-response).
 
@@
 # quadratic response, correlated errors (subject effect)
-plot_example(example(fun_type = "quadratic", error_type = "correlated")) +
+library(ggsci)
+plot_example(example(fun_type = "linear", error_type = "correlated")) +
   plot_annotation(
-  title = 'Simulated with quadratic function and correlated errors',
-    theme = theme(plot.title = element_text(size = 20))
-) #& 
+  title = 'Simulated with linear function and correlated errors',
+    theme = theme(plot.title = element_text(size = txt),text=element_text(size=txt))
+)
   # theme(plot.title = element_text(size = 20))
@@
 
-In this case, when the predictions are compared to the simulated data it is clear that the model is not capturing the changes within each group throughout the timeline.
-This highlights the limitation of rm-ANOVA and LMEMs with longitudinal non-linear data, where the variations are not captured and modeled properly.
+In this case, compare the predictions to the simulated data and it is noticeable that the model is not capturing the changes within each group throughout the timeline.
+This highlights the limitation of rm-ANOVA and LMEMs with longitudinal non-linear data, where the model is only flexible enough to allow different slopes per group, but is unable to follow any trend that is highly variable.
 
@@
 
-In the case of an rm-ANOVA analysis, it is assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
+In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
 Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
 However, it has been shown that this condition is frequently unjustified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
+A visual representation of the
 Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser) the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
@@
 Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator [molenberghs2004].Dropout from patients, or attrition or injury in animals are among the reasons for missing observations.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004].
+Dropout from patients, attrition or injury in animals are among the reasons for missing observations.
 Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
@@
 
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR [@west2014].
+In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014].
 The pattern of missing observations can be considered MAR if the missing observations are not related any of the other variables measured in the study [@maxwell2017].
@@
 
+## The Basis function approach
+
 A GAM is a special case of the Generalized Linear Model (GLM), a framework that allows for response distributions that do not follow a normal distribution [@wood2017;@hastie1987].
@@
 \begin{equation}
-  y_{t}=\beta_0+f(x_t)+\epsilon_t  
+  y_{ijt}=\beta_0+f(x_t\mid \beta)+\varepsilon_{ijt}
   (\#eq:GAM)
@@
 
-Where $y_t$ is the response at time $t$, $\beta_0$ is the expected value at time 0, the change of $y$ over time is represented by the function $f(x_t)$ and $\epsilon_t$ represents the residuals.
+Where $y_{ijt}$ is the response at time $t$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta)$ and $\epsilon_{ijt}$ represents the residuals.
 

### 2021-01-20 Manuscripts/Manuscript_AM_v2.Rmd
Updated

@@
 
-## The Basis function approach
+## GAMs and Basis Functions
 
@@
 
-Where $y_{ijt}$ is the response at time $t$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta)$ and $\epsilon_{ijt}$ represents the residuals.
+Where $y_{ijt}$ is the response at time $t$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta)$ and $\varepsilon_{ijt}$ represents the residuals.
 
@@
 This approach is more advantageous as it does not restrict the model to a linear relationship.
-One possible function for $f(x_t)$  is a polynomial, but a major limitation is that they create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
+One possible function for $f(x_t\mid \beta)$ that allows for non-linear responses is a polynomial, but a major limitation is that they create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].In particular, as $t$ goes to $\pm \infty$, $f(x_t \mid \beta)$ goes to $\pm \infty$ which is almost always unrealistic.
 
+The model specification for GAMs requires that the _smooth functions_ are represented in a parametric way that fit within the GLM framework, and this step is achieved by using a _basis function_ to represent such functions.
+A _basis_ is a set of functions that capture the space where the smooths that approximate $f(x_t \mid \beta)$ exist [@simpson2018].
+In other words,
 To overcome this limitation, the smooth functions in GAMs  are represented using _basis functions_  over  evenly spaced ranges of the covariates known as _knots_.

### 2021-01-20 Manuscripts/Manuscript_AM_v2.Rmd
Updated

@@
 title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
+subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
 author:
@@
 - John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
-subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
 output:
+  rticles::jss_article
   bookdown::html_document2: default
@@
   '': default
-csl: ieee.csl
+csl: Elsevier.csl
 bibliography: refs.bib
@@
 \begin{equation}
-y_{ijt} = \mu+ \beta_1* time_{t} + \beta_2* treatment_{j} + \beta_3* time_{t}*treatment_{j}+\pi_{ij} +\varepsilon_{tij}\\ 
+y_{ijt} = \mu+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\pi_{ij} +\varepsilon_{tij}\\ 
 (\#eq:linear-model)
@@
 y_{ijt} = \begin{cases}
-\mu + \beta_1*time_{t}+\varepsilon_{ijt}+\pi_{ijt}   & \mbox{if Group A}\\
-\mu + \beta_1 * time_{t} + \beta_2 * treatment_{j} +\beta_3* time_{t} * treatment_{j}+\pi_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\mu + \beta_1\times time_{t}+\varepsilon_{ijt}+\pi_{ijt}   & \mbox{if Group A}\\
+\mu + \beta_1 \times time_{t} + \beta_2 \times treatment_{j} +\beta_3 \times time_{t} \times treatment_{j}+\pi_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
 \end{cases}
@@
 y_{ijt} = \begin{cases}
-\mu + \beta_1*time_{t}+\varepsilon_{ijt}+\pi_{ijt}   & \mbox{if Group A}\\
-\widetilde{mu} + \widetilde{\beta_1} * time_{t} +\pi_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\mu + \beta_1\times time_{t}+\varepsilon_{ijt}+\pi_{ijt}   & \mbox{if Group A}\\
+\widetilde{\mu} + \widetilde{\beta_1} \times time_{t} +\pi_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
 \end{cases}
@@
 
-The model specification for GAMs requires that the _smooth functions_ are represented in a parametric way that fit within the GLM framework, and this step is achieved by using a _basis function_ to represent such functions.
+The model specification for GAMs requires that the _smooth functions_ are represented in a parametric way that fit within the GLM framework, and this step is achieved by using _basis functions_ to represent them.
 A _basis_ is a set of functions that capture the space where the smooths that approximate $f(x_t \mid \beta)$ exist [@simpson2018].
-In other words,
-To overcome this limitation, the smooth functions in GAMs  are represented using _basis functions_  over  evenly spaced ranges of the covariates known as _knots_.
-The _basis function_ used is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together[@wood2017;@simpson2018].
-Cubic splines have a long history in their use to solve non-parametric statistical problems and are the default choice to fit GAMs as they are the simplest option to obtain visual smoothness [@wegman1983].
-Therefore, GAMs overcome the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear, such as \@ref(fig:quadratic-response).
+For the linear model in \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of \@ref(eq:linear-model) is restricted to a linear behavior.
+In the case of \@ref(eq:GAM), the basis function is $f(x_t\mid \beta)$, which means that the model allows relationships beyond linear for the covariates.
+
+In GAMs the _basis functions_  are represented over  evenly spaced ranges of the covariates known as _knots_.
+A commonly used _basis function_ is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations[@wood2017;@simpson2018].
+Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness [@wegman1983].
+Therefore, GAMs overcome the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear, such as Figure \@ref(fig:quadratic-response).
 Regarding longitudinal data, Pedersen et al [@pedersen2019] demonstrated the capabilities of GAMs in this area using ecological data.

### 2021-01-20 Manuscripts/Manuscript_AM_v2.tex
Updated

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models},
  pdfauthor={Ariel Mundo; Timothy J. Muldoon; John R. Tipton},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{subfig}
\usepackage{breqn}

\title{\textbf{The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
\author{true \and true \and Ariel Mundo\footnote{Department of Biomedical Engineering, University of Arkansas, Fayetteville} \and Timothy J. Muldoon\footnote{Department of Biomedical Engineering, University of Arkansas, Fayetteville} \and John R. Tipton\footnote{Department of Mathematical Sciences, University of Arkansas, Fayetteville}}
\date{}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{background}{%
\section{Background}\label{background}}

A longitudinal study is designed to repeatedly measure a variable of interest in a group (or groups) of subjects, when the intention is to observe the evolution of the effect of a certain treatment across time rather than analyzing it at a single time point (e.g., a cross-sectional study). Medical research commonly uses longitudinal studies to analyze the evolution of patients before, during and after treatment. For example, Sio et al analyzed the evolution of dermatitis symptoms in breast cancer patients after radiotherapy (RT), by taking weekly measurements over the course of two moths {[}1{]}, and Kamstra et al measured mouth opening at regular monthly intervals after RT in neck cancer patients {[}2{]}. Biomedical research has used longitudinal studies as well to analyze tumor response {[}3--6{]}, antibody expression {[}7,8{]}, and cell metabolism {[}9,10{]}.

Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study. The frequentist paradigm regards probability as a limiting frequency{[}11{]} by assuming a null hypothesis using the \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA). This methodology makes two key assumptions regarding longitudinal data: constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study (a condition also known as complete observations) {[}12,13{]}.

The assumption of constant correlation is frequently unjustified as correlation in time tends to diminish between measurements when the time interval between them increases {[}14{]}.
Due to a variety of causes, the number of observations during a study can vary between all subjects. For example, in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and it is even possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point and therefore violating the \emph{complete observations} assumption of classical rm-ANOVA.

When incomplete observations occur, an rm-ANOVA is fit by excluding all subjects with missing observations from the analysis{[}12{]}. This can result in increased costs if the desired statistical power is not met with the remaining observations, as it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups. Additionally, rm-ANOVA uses a \emph{post hoc} analysis to assess differences between the measured response in different groups. A \emph{post hoc} analysis is based on multiple repeated comparisons to estimate a \emph{p-value}, a metric that is widely used as a measure of significance. Because the \emph{p-value} is highly variable, multiple comparisons can inflate the false positivity rate {[}15,16{]}, consequently biasing the conclusions of the study.

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information. The recognition on the limitations of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal data {[}10,17{]}. Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g.~the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population. When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}18{]}. On the other hand, LMEMs impose restrictions in the distribution of the errors of the random effects {[}12{]}.

One final assumption that is not initially evident for both rm-ANOVA and LMEMs models is that the mean response is expected to change linearly through time{[}18{]}. The linearity assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend. In biomedical research, a particular case of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings {[}3,10,17{]}. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent with the observed variation. In other words, both LMEMs and rm-ANOVA explain highly-variable data using a linear trend, consequently biasing the estimates when the true response is non-linear in time.
Additionally, although it is possible that a \emph{post hoc} analysis is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) it has to be reminded that the validity of such metric relies on how adequate the model is to the data. In other words, the \emph{p-value} requires that the model and the data have good agreement and if that is not the case, a ``Type 3'' error (known as ``model misspecification'') occurs{[}19{]}.That would be the case if for example, a model that is only able to explain linear responses is fitted to data that has a quadratic behavior. In such scenario, the \emph{p-value} becomes invalid{[}20{]}.

As the ``frequentist'' rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, enabling inference that is appropiate to the data.
In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data. Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data {[}21--23{]}, health-environment interactions {[}24{]} and the dynamics of government in political science {[}25{]} . There are several advantages of GAMs over LMEMs and rm-ANOVA models: GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend of the model, can easily accommodate missing observations and can model non-constant correlation between repeated measurements {[}26{]}.Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data.

The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of more complex models beyond LMEMs. In particular, \(\textsf{R}\) has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}26,27{]} that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}28{]}. This allows to create and explore different alternatives for analysis without collecting information in the field, and reduces the time window between experiment design and its implementation.

The purpose of this work is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity of response and constant correlation structures. Second, the key theoretical elements of GAMs are presented without using extensive derivations or complex mathematical notation that can obscure their understanding. Third, simulated data that follows the trend of previously reported values {[}17{]} is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research to highlight the differences in inference between rm-ANOVA, LMEMs and GAMs and the implementation of the latter in such scenario. Finally, reproducibility is emphasized by providing the code to generate the data and the implementation of GAMs in \(\textsf{R}\), in conjunction with a step-by-step guide to fit models of increasing complexity. In summary, the exploration of modern statistical techniques to analyze longitudinal data may allow biomedical researchers to consider the use of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a linear trend, and will also help to improve the standards for reproducibility in biomedical research.

\hypertarget{challenges-presented-by-longitudinal-studies}{%
\section{Challenges presented by longitudinal studies}\label{challenges-presented-by-longitudinal-studies}}

\hypertarget{the-frequentist-case-for-longitudinal-data}{%
\subsection{The ``frequentist'' case for longitudinal data}\label{the-frequentist-case-for-longitudinal-data}}

The \emph{repeated measures analysis of variance} (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research, but certain assumptions are necessary for the model to be valid. From a practical view, the assumptions can be divided in three areas: a linear relationship between covariates and response, a constant correlation between measurements, and complete observations for all subjects. Each one of these assumptions is discussed below.

\hypertarget{linear-relationship}{%
\subsection{Linear relationship}\label{linear-relationship}}

In a biomedical longitudinal study, two or more groups of subjects (patients, mice, samples) are subject to different treatments (e.g., a ``treatment'' group receives a novel drug vs.~a ``control'' group that receives a placebo), and measurements from each subject within each group are collected at specific time points. The collected response has both \emph{fixed} and \emph{random} components. The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the effect of the novel drug in the ``treatment'' group. The \emph{random} component can be defined as ``noise'' caused by some inherent variability within the study. For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while others are measured in the afternoon, it is possible that the difference in the collection time of the measurement introduces some ``noise'' in the signal. As their name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.

Mathematically speaking, a rm-ANOVA model with an interaction can be written as:

\begin{equation}
y_{ijt} = \mu+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\pi_{ij} +\varepsilon_{tij}\\ 
\label{eq:linear-model}
\end{equation}

In this model \(y_{ijt}\) is the response by subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\mu\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\). A \emph{random effect} \(\pi_{ij}\) of each subject within each group, and independent errors per subject per time per group \(\varepsilon_{tij}\), that represent random variation not explained by the \emph{fixed} or \emph{random} effects.
Suppose two treatments groups are used in the study. Then the models per group can be written as below with \(j=0\) representing the first treatment group (Group A) and \(j=1\) representing the second treatment group (Group B). The linear models then become:

\begin{equation}
y_{ijt} = \begin{cases}
\mu + \beta_1\times time_{t}+\varepsilon_{ijt}+\pi_{ijt}   & \mbox{if Group A}\\
\mu + \beta_1 \times time_{t} + \beta_2 \times treatment_{j} +\beta_3 \times time_{t} \times treatment_{j}+\pi_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
\label{eq:ANOVA-by-group}
\end{equation}

To further simplify the expression, let us substitute \(\widetilde{\mu}=\mu+\beta_{2}\) and \(\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}\) in the equation for Group B. The equation then becomes:

\begin{equation}
y_{ijt} = \begin{cases}
\mu + \beta_1\times time_{t}+\varepsilon_{ijt}+\pi_{ijt}   & \mbox{if Group A}\\
\widetilde{\mu} + \widetilde{\beta_1} \times time_{t} +\pi_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
\label{eq:ANOVA-lines}
\end{equation}

Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different slopes per group). In other words, When it is stated that an rm-ANOVA model ``expects'' a linear relationship between the covariates and the response, this means that either presented as \eqref{eq:linear-model}, \eqref{eq:ANOVA-by-group} or \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data, the maximum flexibility being that it can assign different slopes to each group. Because the construction of the model is essentially the same in LMEMs, the expected linearity holds in such models, although it is possible to fit models with different slopes at the subject-level{[}18{]}.\\
When the data does not follow a linear trend, the fit that an rm-ANOVA or LMEM model produces does not accurately represent the trends in the data. To demonstrate this, longitudinal data where a normally distributed response of two groups of 10 subjects each is simulated, and a rm-ANOVA model (\eqref{eq:linear-model}) is fitted to the data using the package \emph{nlme} (the code for the simulated data and models is available in the Appendix).

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript_AM_v2_files/figure-latex/linear-model-1} \caption{Simulated linear responses from two groups with a rm-ANOVA model fitted. Top row, simulated data, lines represent mean response. Bottom row, fitted model, thick lines represent predicted mean response}\label{fig:linear-model}
\end{figure}

It is clear from Figure \ref{fig:linear-model} that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data.

However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:quadratic-response}. Here, the simulated data follows a quadratic behavior. The figure shows that changes in each group occur through the timeline, although the final mean value is the same as the initial value. Fitting an rm-ANOVA model \eqref{eq:linear-model} to this data produces the fit that appears at the bottom right of Figure \ref{fig:quadratic-response}.

\begin{figure}
\includegraphics[width=1\linewidth]{Manuscript_AM_v2_files/figure-latex/quadratic-response-1} \caption{Simulated quadratic responses from two groups with an rm-ANOVA model fitted. Top row, simulated data, lines represent mean response. Bottom row, fitted model, thick lines represent predicted mean response}\label{fig:quadratic-response}
\end{figure}

In this case, compare the predictions to the simulated data and it is noticeable that the model is not capturing the changes within each group throughout the timeline. This highlights the limitation of rm-ANOVA and LMEMs with longitudinal non-linear data, where the model is only flexible enough to allow different slopes per group, but is unable to follow any trend that is highly variable.

\hypertarget{covariance-in-rm-anova-and-lmems}{%
\subsection{Covariance in rm-ANOVA and LMEMs}\label{covariance-in-rm-anova-and-lmems}}

In a longitudinal study there is an expected \emph{variance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group. The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}29{]} (For an in-depth analysis of the covariance matrix see {[}30,31{]}).

In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity''). Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}31--33{]}. However, it has been shown that this condition is frequently unjustified because the correlation between measurements tends to change over time {[}34{]}; and it is higher between consecutive measurements {[}12,14{]}. A visual representation of the
Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser) the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}28{]}, and they are not robust if the group sizes are unbalanced {[}35{]}. In other words, if the data does not present constant correlation between repeated measurements, the assumptions required for an rm-ANOVA model are not met and the use of corrections may still not provide a reasonable adjustment that makes the model valid.

In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}18{]}. Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself. Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.

\hypertarget{missing-observations}{%
\subsection{Missing observations}\label{missing-observations}}

Missing observations are an issue that arises frequently in longitudinal studies. In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}36{]}. Dropout from patients, attrition or injury in animals are among the reasons for missing observations. Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}31{]}. In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}37{]}. If the data are MCAR, this means that the missigness is completely unrelated to the collected information {[}38{]}, and in the case of MNAR the missing values are dependent on their value. An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis. This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}39{]}.

In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}30{]}. The pattern of missing observations can be considered MAR if the missing observations are not related any of the other variables measured in the study {[}34{]}. For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.

This section has presented the assumptions of rm-ANOVA to analyze longitudinal information and its differences when compared to LMEMs regarding to missing data and the modeling of the covariance matrix. Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study. However, when the data presents high variability, LMEMs fail to capture the non-linear trend of the data. To analyze such type of data, we present generalized additive models (GAMs) as an alternative in the following section.

\hypertarget{gams-as-a-special-case-of-generalized-linear-models}{%
\section{GAMs as a special case of Generalized Linear Models}\label{gams-as-a-special-case-of-generalized-linear-models}}

\hypertarget{gams-and-basis-functions}{%
\subsection{GAMs and Basis Functions}\label{gams-and-basis-functions}}

A GAM is a special case of the Generalized Linear Model (GLM), a framework that allows for response distributions that do not follow a normal distribution {[}26,40{]}. Following the notation by Simpson {[}23{]} A GAM model can be represented as:

\begin{equation}
  y_{ijt}=\beta_0+f(x_t\mid \beta)+\varepsilon_{ijt}
  \label{eq:GAM}
\end{equation}

Where \(y_{ijt}\) is the response at time \(t\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the function \(f(x_t\mid \beta)\) and \(\varepsilon_{ijt}\) represents the residuals.

In contrast to rm-ANOVA or LMEMs, GAMs use \emph{smooth functions} to model the relationship between the covariates and the response. This approach is more advantageous as it does not restrict the model to a linear relationship. One possible function for \(f(x_t\mid \beta)\) that allows for non-linear responses is a polynomial, but a major limitation is that they create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}25{]}.In particular, as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta)\) goes to \(\pm \infty\) which is almost always unrealistic.

The model specification for GAMs requires that the \emph{smooth functions} are represented in a parametric way that fit within the GLM framework, and this step is achieved by using \emph{basis functions} to represent them. A \emph{basis} is a set of functions that capture the space where the smooths that approximate \(f(x_t \mid \beta)\) exist {[}23{]}. For the linear model in \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\). The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}41{]}, which in the case of \eqref{eq:linear-model} is restricted to a linear behavior. In the case of \eqref{eq:GAM}, the basis function is \(f(x_t\mid \beta)\), which means that the model allows relationships beyond linear for the covariates.

In GAMs the \emph{basis functions} are represented over evenly spaced ranges of the covariates known as \emph{knots}. A commonly used \emph{basis function} is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations{[}23,26{]}. Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness {[}42{]}. Therefore, GAMs overcome the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear, such as Figure \ref{fig:quadratic-response}. Regarding longitudinal data, Pedersen et al {[}22{]} demonstrated the capabilities of GAMs in this area using ecological data.

The use of GAMs to analyze biomedical longitudinal data, and the impact of missing observations in the fit of the model will be examined in detail in the following section using simulation.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-sio2016}{}%
{[}1{]} T.T. Sio, P.J. Atherton, B.J. Birckhead, D.J. Schwartz, J.A. Sloan, D.K. Seisler, J.A. Martenson, C.L. Loprinzi, P.C. Griffin, R.F. Morton, others, Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}), Supportive Care in Cancer. 24 (2016) 3847--3855.

\leavevmode\hypertarget{ref-kamstra2015}{}%
{[}2{]} J. Kamstra, P. Dijkstra, M. Van Leeuwen, J. Roodenburg, J. Langendijk, Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study, Oral Oncology. 51 (2015) 548--555.

\leavevmode\hypertarget{ref-roblyer2011}{}%
{[}3{]} D. Roblyer, S. Ueda, A. Cerussi, W. Tanamai, A. Durkin, R. Mehta, D. Hsiang, J.A. Butler, C. McLaren, W.-P. Chen, others, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences. 108 (2011) 14626--14631.

\leavevmode\hypertarget{ref-tank2020}{}%
{[}4{]} A. Tank, H.M. Peterson, V. Pera, S. Tabassum, A. Leproux, T. O'Sullivan, E. Jones, H. Cabral, N. Ko, R.S. Mehta, others, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research. 22 (2020) 1--10.

\leavevmode\hypertarget{ref-pavlov2018}{}%
{[}5{]} M.V. Pavlov, T.I. Kalganova, Y.S. Lyubimtseva, V.I. Plekhanov, G.Y. Golubyatnikov, O.Y. Ilyinskaya, A.G. Orlova, P.V. Subochev, D.V. Safonov, N.M. Shakhova, others, Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy, Journal of Biomedical Optics. 23 (2018) 091410.

\leavevmode\hypertarget{ref-demidov2018}{}%
{[}6{]} V. Demidov, A. Maeda, M. Sugita, V. Madge, S. Sadanand, C. Flueraru, I.A. Vitkin, Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography, Scientific Reports. 8 (2018) 1--12.

\leavevmode\hypertarget{ref-ritter2001}{}%
{[}7{]} G. Ritter, L.S. Cohen, C. Williams, E.C. Richards, L.J. Old, S. Welt, Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody a33, Cancer Research. 61 (2001) 6851--6859.

\leavevmode\hypertarget{ref-roth2017}{}%
{[}8{]} E.M. Roth, A.C. Goldberg, A.L. Catapano, A. Torri, G.D. Yancopoulos, N. Stahl, A. Brunet, G. Lecorps, H.M. Colhoun, Antidrug antibodies in patients treated with alirocumab, (2017).

\leavevmode\hypertarget{ref-jones2018}{}%
{[}9{]} J.D. Jones, H.E. Ramser, A.E. Woessner, K.P. Quinn, In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing, Communications Biology. 1 (2018) 1--8.

\leavevmode\hypertarget{ref-skala2010}{}%
{[}10{]} M.C. Skala, A.N. Fontanella, L. Lan, J.A. Izatt, M.W. Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics. 15 (2010) 011112.

\leavevmode\hypertarget{ref-wagenmakers2008}{}%
{[}11{]} E.-J. Wagenmakers, M. Lee, T. Lodewyckx, G.J. Iverson, Bayesian versus frequentist inference, in: Bayesian Evaluation of Informative Hypotheses, Springer, 2008: pp. 181--207.

\leavevmode\hypertarget{ref-gueorguieva2004}{}%
{[}12{]} R. Gueorguieva, J.H. Krystal, Move over anova: Progress in analyzing repeated-measures data andits reflection in papers published in the archives of general psychiatry, Archives of General Psychiatry. 61 (2004) 310--317.

\leavevmode\hypertarget{ref-schober2018}{}%
{[}13{]} P. Schober, T.R. Vetter, Repeated measures designs and analysis of longitudinal data: If at first you do not succeed---try, try again, Anesthesia and Analgesia. 127 (2018) 569.

\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
{[}14{]} C. Ugrinowitsch, G.W. Fellingham, M.D. Ricard, Limitations of ordinary least squares models in analyzing repeated measures data, Medicine and Science in Sports and Exercise. 36 (2004) 2144--2148.

\leavevmode\hypertarget{ref-liu2010}{}%
{[}15{]} C. Liu, T.P. Cripe, M.-O. Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy. 18 (2010) 1724--1730.

\leavevmode\hypertarget{ref-halsey2015}{}%
{[}16{]} L.G. Halsey, D. Curran-Everett, S.L. Vowler, G.B. Drummond, The fickle p value generates irreproducible results, Nature Methods. 12 (2015) 179--185.

\leavevmode\hypertarget{ref-vishwanath2009}{}%
{[}17{]} K. Vishwanath, H. Yuan, W.T. Barry, M.W. Dewhirst, N. Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia. 11 (2009) 889--900.

\leavevmode\hypertarget{ref-pinheiro2006}{}%
{[}18{]} J. Pinheiro, D. Bates, Mixed-effects models in s and s-plus, Springer Science \& Business Media, 2006.

\leavevmode\hypertarget{ref-dennis2019}{}%
{[}19{]} B. Dennis, J.M. Ponciano, M.L. Taper, S.R. Lele, Errors in statistical inference under model misspecification: Evidence, hypothesis testing, and aic, Frontiers in Ecology and Evolution. 7 (2019) 372.

\leavevmode\hypertarget{ref-wang2019}{}%
{[}20{]} B. Wang, Z. Zhou, H. Wang, X.M. Tu, C. Feng, The p-value and model specification in statistics, General Psychiatry. 32 (2019).

\leavevmode\hypertarget{ref-rose2012}{}%
{[}21{]} N.L. Rose, H. Yang, S.D. Turner, G.L. Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to Scotland, UK, Geochimica et Cosmochimica Acta. 82 (2012) 113--135.

\leavevmode\hypertarget{ref-pedersen2019}{}%
{[}22{]} E.J. Pedersen, D.L. Miller, G.L. Simpson, N. Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, PeerJ. 7 (2019) e6876.

\leavevmode\hypertarget{ref-simpson2018}{}%
{[}23{]} G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018) 149.

\leavevmode\hypertarget{ref-yang2012}{}%
{[}24{]} L. Yang, G. Qin, N. Zhao, C. Wang, G. Song, Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality, BMC Medical Research Methodology. 12 (2012) 165.

\leavevmode\hypertarget{ref-beck1998}{}%
{[}25{]} N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. (1998) 596--627.

\leavevmode\hypertarget{ref-wood2017}{}%
{[}26{]} S.N. Wood, Generalized additive models: An introduction with r, CRC press, 2017.

\leavevmode\hypertarget{ref-wood2016}{}%
{[}27{]} S.N. Wood, N., Pya, B. Säfken, Smoothing parameter and model selection for general smooth models (with discussion), Journal of the American Statistical Association. 111 (2016) 1548--1575.

\leavevmode\hypertarget{ref-haverkamp2017}{}%
{[}28{]} N. Haverkamp, A. Beauducel, Violation of the sphericity assumption and its effect on type-i error rates in repeated measures anova and multi-level linear models (mlm), Frontiers in Psychology. 8 (2017) 1841.

\leavevmode\hypertarget{ref-wolfinger1996}{}%
{[}29{]} R.D. Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics. (1996) 205--230.

\leavevmode\hypertarget{ref-west2014}{}%
{[}30{]} B.T. West, K.B. Welch, A.T. Galecki, Linear mixed models: A practical guide using statistical software, CRC Press, 2014.

\leavevmode\hypertarget{ref-weiss2005}{}%
{[}31{]} R.E. Weiss, Modeling longitudinal data, Springer Science \& Business Media, 2005.

\leavevmode\hypertarget{ref-geisser1958}{}%
{[}32{]} S. Geisser, S.W. Greenhouse, others, An extension of box's results on the use of the \(F\) distribution in multivariate analysis, The Annals of Mathematical Statistics. 29 (1958) 885--891.

\leavevmode\hypertarget{ref-huynh1976}{}%
{[}33{]} H. Huynh, L.S. Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics. 1 (1976) 69--82.

\leavevmode\hypertarget{ref-maxwell2017}{}%
{[}34{]} S.E. Maxwell, H.D. Delaney, K. Kelley, Designing experiments and analyzing data: A model comparison perspective, Routledge, 2017.

\leavevmode\hypertarget{ref-keselman2001}{}%
{[}35{]} H. Keselman, J. Algina, R.K. Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematical and Statistical Psychology. 54 (2001) 1--20.

\leavevmode\hypertarget{ref-molenberghs2004}{}%
{[}36{]} G. Molenberghs, H. Thijs, I. Jansen, C. Beunckens, M.G. Kenward, C. Mallinckrodt, R.J. Carroll, Analyzing incomplete longitudinal clinical trial data, Biostatistics. 5 (2004) 445--464.

\leavevmode\hypertarget{ref-scheffer2002}{}%
{[}37{]} J. Scheffer, Dealing with missing data, (2002).

\leavevmode\hypertarget{ref-potthoff2006}{}%
{[}38{]} R.F. Potthoff, G.E. Tudor, K.S. Pieper, V. Hasselblad, Can one assess whether missing data are missing at random in medical studies?, Statistical Methods in Medical Research. 15 (2006) 213--234.

\leavevmode\hypertarget{ref-ma2012}{}%
{[}39{]} Y. Ma, M. Mazumdar, S.G. Memtsoudis, Beyond repeated-measures analysis of variance: Advanced statistical methods for the analysis of longitudinal data in anesthesia research, Regional Anesthesia \& Pain Medicine. 37 (2012) 99--105.

\leavevmode\hypertarget{ref-hastie1987}{}%
{[}40{]} T. Hastie, R. Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association. 82 (1987) 371--386.

\leavevmode\hypertarget{ref-hefley2017}{}%
{[}41{]} T.J. Hefley, K.M. Broms, B.M. Brost, F.E. Buderman, S.L. Kay, H.R. Scharf, J.R. Tipton, P.J. Williams, M.B. Hooten, The basis function approach for modeling autocorrelation in ecological data, Ecology. 98 (2017) 632--646.

\leavevmode\hypertarget{ref-wegman1983}{}%
{[}42{]} E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365.

\end{document}

### 2021-01-20 Manuscripts/Manuscript_AM_v2.Rmd
Manuscript and figures updated

@@
 output:
-  rticles::jss_article
-  bookdown::html_document2: default
-  bookdown::word_document2: default
   bookdown::pdf_document2:
@@
     extra_dependencies: ["subfig","breqn"]
+  bookdown::word_document2: default
+  bookdown::html_document2: default
   link-citations: yes
@@
     geom_line(aes(x = time, y = mu, color = treatment)) +
-    theme_bw(base_size = 12) +
+    theme_classic() +
     ggtitle("Simulated data with true response function")+
+    theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
     scale_color_aaas()
@@
     ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_bw(base_size = 12)+
+    theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
+    theme_classic()+
     scale_color_aaas()
@@
     ggtitle("Simulated errors") +
-    theme_bw(base_size = 12)+
+     theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
+    theme_classic()+
     scale_color_aaas()
@@
     scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_bw(base_size = 12) +
+    theme_classic() +
     ggtitle("Fitted Model")+
+    theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
     scale_color_aaas()
@@
 
-```{r, linear-model, fig.width=16, fig.height=9, out.width="100%", echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with a rm-ANOVA model fitted.
+```{r, linear-model, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with a rm-ANOVA model fitted.
 Top row, simulated data, lines represent mean response.
@@
 # linear response, correlated errors (subject effect)
-txt<-20
+txt<-18
 plot_example(example(fun_type = "linear", error_type = "correlated")) +
-  plot_annotation(
-  title = 'Simulated with linear function and correlated errors',
-    theme = theme(plot.title = element_text(size = txt),text=element_text(size=txt))
+  plot_annotation(tag_levels='A',
+                  title = 'Simulated with linear function and correlated errors') &
+   theme(plot.title = element_text(size = 20),text=element_text(size=txt)
 )
@@
 
-```{r quadratic-response, fig.width=16, fig.height=9, out.width="100%",echo=FALSE,fig.cap='Simulated quadratic responses from two groups with an rm-ANOVA model fitted.
+```{r quadratic-response, fig.width=16, fig.height=16,echo=FALSE,fig.cap='Simulated quadratic responses from two groups with an rm-ANOVA model fitted.
 Top row, simulated data, lines represent mean response.
@@
 # quadratic response, correlated errors (subject effect)
-library(ggsci)
+txt<-18
 plot_example(example(fun_type = "linear", error_type = "correlated")) +
   plot_annotation(
-  title = 'Simulated with linear function and correlated errors',
-    theme = theme(plot.title = element_text(size = txt),text=element_text(size=txt))
+    tag_levels='A',
+  title = 'Simulated with linear function and correlated errors') &
+   theme(plot.title = element_text(size = 20),text=element_text(size=txt)
 )

### 2021-01-20 Manuscripts/Manuscript_AM_v2.tex
Manuscript and figures updated

@@
 \subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
-\author{true \and true \and Ariel Mundo\footnote{Department of Biomedical Engineering, University of Arkansas, Fayetteville} \and Timothy J.
+\author{Ariel Mundo\footnote{Department of Biomedical Engineering, University of Arkansas, Fayetteville} \and Timothy J.
 Muldoon\footnote{Department of Biomedical Engineering, University of Arkansas, Fayetteville} \and John R.
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript_AM_v2_files/figure-latex/linear-model-1} \caption{Simulated linear responses from two groups with a rm-ANOVA model fitted.
+\centering
+\includegraphics{Manuscript_AM_v2_files/figure-latex/linear-model-1.pdf}
+\caption{\label{fig:linear-model}Simulated linear responses from two groups with a rm-ANOVA model fitted.
 Top row, simulated data, lines represent mean response.
-Bottom row, fitted model, thick lines represent predicted mean response}\label{fig:linear-model}
+Bottom row, fitted model, thick lines represent predicted mean response}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{Manuscript_AM_v2_files/figure-latex/quadratic-response-1} \caption{Simulated quadratic responses from two groups with an rm-ANOVA model fitted.
+\centering
+\includegraphics{Manuscript_AM_v2_files/figure-latex/quadratic-response-1.pdf}
+\caption{\label{fig:quadratic-response}Simulated quadratic responses from two groups with an rm-ANOVA model fitted.
 Top row, simulated data, lines represent mean response.
-Bottom row, fitted model, thick lines represent predicted mean response}\label{fig:quadratic-response}
+Bottom row, fitted model, thick lines represent predicted mean response}
 \end{figure}

### 2021-01-28 Manuscripts/Manuscript_AM_v1.Rmd
File needs updates from master branch

@@
 ---
-title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
+title: '**The statistical analysis of non-linear longitudinal data in biomedical research
+  using generalized additive models**'
 author:
@@
 - John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
+csl: ieee.csl
+output:
+- default
+- default
+- yes
+- style.css
+- default
+- extra_dependencies:
+  - subfig
+  - breqn
+  fig_caption: yes
+  keep_tex: yes
 subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
-output:
-  bookdown::pdf_document2:
-    keep_tex: yes
-    fig_caption: yes
-    extra_dependencies: ["subfig","breqn"]
-  bookdown::word_document2: default
-  bookdown::html_document2: default
-  link-citations: yes
-  css: style.css
-  '': default
-csl: ieee.csl
 bibliography: refs.bib
@@
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-This is exemplified by the  use of  linear mixed effects models (LMEMs)  in certain groups to analyze longitudinal data [@skala2010;@vishwanath2009].
-Briefly, these models incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g.
+The recognition on the limitations of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal data [@skala2010;@vishwanath2009].
+Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g.
 the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population.
-These models are more flexible than the traditional rm-ANOVA as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-On the other hand, they impose restrictions in the distribution of the errors  of the random effects [@gueorguieva2004].
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
+On the other hand, LMEMs impose restrictions in the distribution of the errors  of the random effects [@gueorguieva2004].
 
-One final assumption that is not initially evident for  both rm-ANOVA and LMEMs models pertains the relationship between the observed effect and the progression of the timeline of the study, which is expected to be linear [@pinheiro2006].
-This common assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend.
-In biomedical research, a special case of this behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
+One final assumption that is not initially evident for both rm-ANOVA and LMEMs models is that the mean response is expected to change linearly through time[@pinheiro2006].
+The linearity assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend.
+In biomedical research, a particular case of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent with the observed variation.
-In other words, the in both cases the model tries to explain highly-variable data using a linear trend, consequently biasing the estimates.
+In other words, both LMEMs and rm-ANOVA explain highly-variable data using a linear trend, consequently biasing the estimates when the true response is non-linear in time.
+
 Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) by using multiple comparisons between the model terms, the use of this estimator is limited in such scenario because it is intended to work in models that have a reasonable agreement with the data.
 
-As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, while enabling inferences that are both adequate and consistent from a statistical perspective.
-In this regard, Generalized Additive Models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and political science [@beck1998] .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: They have an extensively developed theoretical background, enable the data to dictate the trend of the model, can accommodate missing observations and do not require constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more suitable statistical method to analyze non-linear biomedical longitudinal data.
+***
+Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) it has to be reminded that the validity of such metric relies on how adequate the model is to the data.
+In other words, the _p-value_ requires that the model and the data have good agreement and if that is not the case, a "Type 3" error (known as "model misspecification") occurs[@dennis2019].That would be the case if for example, a model that is only able to explain linear responses is fitted to data that has a quadratic behavior.
+In such scenario, the _p-value_ becomes invalid[@wang2019].
+
+***
+
+As  the “frequentist” rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, enabling inference that is appropiate to the data.
+In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
+Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend of the model, can easily accommodate missing observations and can model non-constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data.
 
  The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs.
-In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2011;@wood2016;@wood2017] that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g.
+In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g.
 hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-This programming language is also able to simulate data, an emerging strategy used to test statistical models [@haverkamp2017].
+At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017].
 This allows to create and  explore different alternatives for analysis without collecting information in the field, and reduces the time window between experiment design and its implementation.
 
-Therefore, the purpose of this study is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity and correlation structures.
-Secondly, the key theoretical elements of GAMS are presented without using extensive derivations or complex mathematical notation that can obscure their understanding.
-Thirdly, simulated data that follows the trend of previously reported values [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research and the implementation of GAMs in such scenario.
+The purpose of this work is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas.
+First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity of response and constant correlation structures.
+Second, the key theoretical elements of GAMs are presented without using extensive derivations or complex mathematical notation that can obscure their understanding.
+Third, simulated data that follows the trend of previously reported values [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research to highlight the differences in inference between rm-ANOVA, LMEMs and GAMs and the implementation of the latter in such scenario.
 Finally, reproducibility is emphasized by providing the code to generate the data and the implementation of GAMs in $\textsf{R}$, in conjunction with a step-by-step guide to fit models of increasing complexity.

### 2021-02-02 Manuscripts/Manuscript_AM_v2.Rmd
Working on comments

@@
 
-A longitudinal study is designed to repeatedly measure a variable of interest in a group (or groups) of subjects, when the intention is to observe the evolution of the effect of a certain treatment across time rather than analyzing it at a single time point (e.g., a cross-sectional study).
+A longitudinal study is designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of the treatment effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
 Medical research commonly uses longitudinal studies to analyze the evolution of patients before, during and after treatment.
-For example, Sio et al analyzed the evolution of dermatitis symptoms in breast cancer patients after radiotherapy (RT), by taking weekly measurements over the course of two moths [@sio2016], and Kamstra et al measured mouth opening at regular monthly intervals after RT in neck cancer patients [@kamstra2015].
-Biomedical research has used longitudinal studies as well to analyze tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
+For example, Sio et.
+al.
+analyzed the evolution of dermatitis symptoms in breast cancer patients after radiotherapy (RT), by taking weekly measurements over the course of two months [@sio2016], and Kamstra et al measured mouth opening at regular monthly intervals after RT in neck cancer patients [@kamstra2015].
+Other studies using longitudinal designs in biomedical research have tumor  analyzed tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
 
 Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as a limiting frequency[@wagenmakers2008] by assuming a null hypothesis using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-This methodology makes two key assumptions regarding longitudinal data: constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study (a condition also known as complete observations) [@gueorguieva2004;@schober2018].
+The frequentist paradigm regards probability as a limiting frequency[@wagenmakers2008] by assuming a null hypothesis under a statistical model that is often assumed to be an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
+The rm-ANOVA model typically makes two key assumptions regarding longitudinal data: constant correlation across same-subject measurements, and observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].
 
-The assumption of constant correlation is frequently unjustified as correlation in time tends to diminish between measurements when the time interval between them increases [@ugrinowitsch2004]. 
+The assumption of constant correlation (often known as the _compoun symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].
 Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and it is even possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point and therefore violating the _complete observations_ assumption of  classical rm-ANOVA.
+For example, in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in preclinical animal studies due to injury or weight loss is possible.
+It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
+In each of these missing data scenarios,the _complete observations_ assumption of  classical rm-ANOVA is violated.
 
-When incomplete observations occur, an rm-ANOVA is fit by excluding all subjects with missing observations from the analysis[@gueorguieva2004].
-This can result in increased costs if the desired statistical power is not met with the remaining observations, as it would be necessary to enroll more subjects.
+When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis[@gueorguieva2004].
+This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
 At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
@@
 Because the _p-value_ is highly variable, multiple comparisons can inflate the false positivity rate [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
+Another less known side effect of multiple comparisons is that they reduce statistical power ($\alpha$)[@nakagawa2004], which in turn can lead to fail to reject the null hypothesis although it is false (a type II error)[@albers2019].
 

### 2021-02-03 Manuscripts/Manuscript_AM_v2.Rmd
updated manuscript and references

@@
 A _post hoc_ analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance.
-Because the _p-value_ is highly variable, multiple comparisons can inflate the false positivity rate [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
-Another less known side effect of multiple comparisons is that they reduce statistical power ($\alpha$)[@nakagawa2004], which in turn can lead to fail to reject the null hypothesis although it is false (a type II error)[@albers2019].
+Because the _p-value_ is highly variable, multiple comparisons can inflate the false positivity rate (Type I error) [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
+Although corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), they in turn reduce statistical power ($\alpha$)[@nakagawa2004], and lead to increased Type II error (fail to reject the null hypothesis although it is false [@gelman2012;@albers2019].
+Therefore, the _post hoc_ comparisons commonly used in rm-ANOVA impose a tradeoff between Type I and II errors.
 
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-The recognition on the limitations of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal data [@skala2010;@vishwanath2009].
-Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g.
-the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population.
+The recognition on the limitations of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor data [@skala2010;@vishwanath2009].
+Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
 When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-On the other hand, LMEMs impose restrictions in the distribution of the errors  of the random effects [@gueorguieva2004].
+On the other hand, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
 
 One final assumption that is not initially evident for both rm-ANOVA and LMEMs models is that the mean response is expected to change linearly through time[@pinheiro2006].
-The linearity assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend.
+The linearity assumption in both rm-ANOVA and LMEMs  implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
 In biomedical research, a particular case of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent with the observed variation.
-In other words, both LMEMs and rm-ANOVA explain highly-variable data using a linear trend, consequently biasing the estimates when the true response is non-linear in time.
-Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) it has to be reminded that the validity of such metric relies on how adequate the model is to the data.
-In other words, the _p-value_ requires that the model and the data have good agreement and if that is not the case, a "Type 3" error (known as "model misspecification") occurs[@dennis2019].That would be the case if for example, a model that is only able to explain linear responses is fitted to data that has a quadratic behavior.
-In such scenario, the _p-value_ becomes invalid[@wang2019].
+These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit LMEMs or rm-ANOVA model inconsistent with the observed variation.
+Therefore, when LMEMs and rm-ANOVA are used to draw inference of such highly-variable data the estimates are inevitably biased, because non-linear data is provided to a model which can only accomodate linear trends.
 
-As  the “frequentist” rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, enabling inference that is appropiate to the data.
+Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) the validity of such metric relies on how adequate the model fits the data.
+In other words, the _p-value_ requires that the model and the data have good agreement and if that is not the case, a "Type 3" error (known as "model misspecification") occurs[@dennis2019].
+For example, a model that is only able to explain linear responses but is fitted to data that has a quadratic behavior results in _p-values_ and parameter estimates that are invalid[@wang2019].
+
+As  the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed behavior) to determine the trend in the data, which enables inference that is appropriate.
+
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
 Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend of the model, can easily accommodate missing observations and can model non-constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data.
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend of the model, can easily accommodate missing observations, and can model non-constant correlation between repeated measurements [@wood2017].
+Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data.
 
  The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs.
-In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g.
+In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
 hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
 At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017].
-This allows to create and  explore different alternatives for analysis without collecting information in the field, and reduces the time window between experiment design and its implementation.
+Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and can be also used for power calculations and study design questions.
 
-The purpose of this work is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity of response and constant correlation structures.
-Second, the key theoretical elements of GAMs are presented without using extensive derivations or complex mathematical notation that can obscure their understanding.
-Third, simulated data that follows the trend of previously reported values [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research to highlight the differences in inference between rm-ANOVA, LMEMs and GAMs and the implementation of the latter in such scenario.
-Finally, reproducibility is emphasized by providing the code to generate the data and the implementation of GAMs in $\textsf{R}$, in conjunction with a step-by-step guide to fit models of increasing complexity.
+This work provides biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas.
+First, the limitations of LMEMs and rm-ANOVA regarding missing observations, assumption of linearity of response and constant correlation structures is explained in detail.
+Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
+Third, simulated data that reproduces patterns in previously reported studies [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
+The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
+Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in $\textsf{R}$, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
+
 In summary, the exploration of modern statistical techniques to analyze longitudinal data may allow biomedical researchers to consider the use of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a linear trend, and will also help to improve the standards for reproducibility  in biomedical research.

### 2021-02-03 Manuscripts/Manuscript_AM_v2.Rmd
Updated manuscript and references

@@
 The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs.
-In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
+In particular, $\textsf{R}$[@r]has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
 hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
@@
 
-## The "frequentist" case for longitudinal data
+## The repeated measures ANOVA
 
@@
 a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response has both  _fixed_ and  _random_ components.
-The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the effect of the novel drug in the "treatment" group.
+The collected respnse is modeled with both  _fixed_ and  _random_ components.
+The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the average effect of the novel drug in the "treatment" group.
 The _random_ component can be defined as "noise" caused by some inherent variability within the study.
@@
 \begin{equation}
-y_{ijt} = \mu+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\pi_{ij} +\varepsilon_{tij}\\ 
+y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{tij}\\ 
 (\#eq:linear-model)
@@
 
-In this model $y_{ijt}$ is the response by subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\mu$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$.
-A _random effect_ $\pi_{ij}$ of each subject within each group, and independent errors per subject per time per group $\varepsilon_{tij}$, that represent random variation not explained by the _fixed_ or _random_ effects.
-Suppose two treatments groups are used in the study.
-Then the models per group can be written as below with $j=0$ representing the first treatment group (Group A) and $j=1$ representing the second treatment group (Group B).
-The linear models then become:
+In this model $y_{ijt}$ is the response by subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
+The _random effect_ $\mu_{ij}$, accounts for variability in each subject within each group by allowing different values for each subject.
+Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ or _random_ effects.
+In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
+"novel drug" or "saline" vs.
+"chemotherapy").
+Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $j=0$ representing the first treatment group (Group A) and $j=1$ representing the second treatment group (Group B).
+The linear models then can be expressed as
 
@@
 y_{ijt} = \begin{cases}
-\mu + \beta_1\times time_{t}+\varepsilon_{ijt}+\pi_{ijt}   & \mbox{if Group A}\\
-\mu + \beta_1 \times time_{t} + \beta_2 \times treatment_{j} +\beta_3 \times time_{t} \times treatment_{j}+\pi_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
+\beta_0 + \beta_1 \times time_{t} + \beta_2 \times treatment_{j} +\beta_3 \times time_{t} \times treatment_{j}+\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
 \end{cases}
@@
 
-To further simplify the expression, let us substitute $\widetilde{\mu}=\mu+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B. The equation then becomes:
+To further simplify the expression, substitute $\widetilde{\mu}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B.
+This substitution allows for a different intercept and slope for Group B.
+The model is written as
 
@@
 y_{ijt} = \begin{cases}
-\mu + \beta_1\times time_{t}+\varepsilon_{ijt}+\pi_{ijt}   & \mbox{if Group A}\\
-\widetilde{\mu} + \widetilde{\beta_1} \times time_{t} +\pi_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
+\widetilde{\mu} + \widetilde{\beta_1} \times time_{t} +\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
 \end{cases}
@@
 
-Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different slopes per group).
-In other words, When it is stated that an rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as \@ref(eq:linear-model), \@ref(eq:ANOVA-by-group) or \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data, the maximum flexibility being that it can assign different slopes to each group.
-Because the construction of the model is essentially the same in LMEMs, the expected linearity holds in such models, although it is possible to fit models with different slopes at the subject-level[@pinheiro2006].
-When the data does not follow a linear trend, the fit that an rm-ANOVA or LMEM model produces does not accurately represent the trends in the data.
-To demonstrate this, longitudinal data where a normally distributed response of two groups of 10 subjects each is simulated, and a rm-ANOVA model  (\@ref(eq:linear-model)) is fitted to the data using the package _nlme_ (the code for the simulated data and models is available in the Appendix).
+Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
+In other words, When it is stated that an rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data.
+If the data show non-linear behavior, the rm-ANOVa model will approximate this behavior with non-parallel lines.
+Because the construction of the model is essentially the same in LMEMs, the linearity assumption holds for these models, although it is possible to fit models with different slopes at the subject-level[@pinheiro2006].
+
+When the data does not follow a linear trend, the fit that an rm-ANOVA or LMEM produces may not accurately represent the trends in the data.
+To demonstrate this inaccurate representation, longitudinal data where a normally distributed response of two groups of 10 subjects each is simulated, and a rm-ANOVA model  (Equation \@ref(eq:linear-model)) is fitted to the data using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix).
+Briefly, the mean responses for each group are linear functions with different intercepts.
 

### 2021-02-04 Manuscripts/Manuscript_AM_v2.Rmd
Need to be able to put the two graphs as a single one

@@
 To demonstrate this inaccurate representation, longitudinal data where a normally distributed response of two groups of 10 subjects each is simulated, and a rm-ANOVA model  (Equation \@ref(eq:linear-model)) is fitted to the data using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix).
-Briefly, the mean responses for each group are linear functions with different intercepts.
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes (negative slope for Group 1 and positive slope for Group 2 ).
+In the second case, a quadratic function is used for the mean response per group (concave down for Group 1 and concave up for Group 2 ).
+From the mean response per group in the linear or quadratic case, individual responses for each subject in each group are simulated using covariance matrices for with compound symmetry (constant variance across time) or independent errors (errors that are generated from a a normal distribution but are not constant over time).
+Thus, the response per subject corresponds to the mean response per group plus the error (compound symmetry or independent).
+To this data, an rm-ANOVA model identical as Equation \@ref:eqn(linear-model) is fitted.
+The resulting simulated data, simulated errors, individual response and fitted parameters from the model are plotted.
 
@@
 
-  return((p1 + p3) / (p2 + p4))
+  return((p1+p3+p2+p4)+plot_layout(nrow=1))
   # return((p1 + p3 + plot_layout(guides = "collect")) / (p2 + p4 + plot_layout(guides = "collect")))
@@
 
-```{r, linear-model, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with a rm-ANOVA model fitted.
+```{r, linear-model-correlated, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with a rm-ANOVA model fitted.
 Top row, simulated data, lines represent mean response.
@@
 
+```{r linear-model-independent, fig.width=16, fig.height=16,echo=FALSE,fig.cap='Simulated quadratic responses from two groups with an rm-ANOVA model fitted.
+Top row, simulated data, lines represent mean response.
+Bottom row, fitted model, thick lines represent predicted mean response', message=FALSE}
+# quadratic response, correlated errors (subject effect)
+txt<-18
+plot_example(example(fun_type = "linear", error_type = "independent")) +
+  plot_annotation(
+    tag_levels='A',
+  title = 'Simulated with linear function and independent errors') &
+   theme(plot.title = element_text(size = 20),text=element_text(size=txt)
+)
+  # theme(plot.title = element_text(size = 20))
+```
+
 It is clear from Figure \@ref(fig:linear-model) that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the "truth" of the simulated data.
@@
 Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears at the bottom right of Figure \@ref(fig:quadratic-response).
-
-```{r quadratic-response, fig.width=16, fig.height=16,echo=FALSE,fig.cap='Simulated quadratic responses from two groups with an rm-ANOVA model fitted.
-Top row, simulated data, lines represent mean response.
-Bottom row, fitted model, thick lines represent predicted mean response', message=FALSE}
-# quadratic response, correlated errors (subject effect)
-txt<-18
-plot_example(example(fun_type = "linear", error_type = "correlated")) +
-  plot_annotation(
-    tag_levels='A',
-  title = 'Simulated with linear function and correlated errors') &
-   theme(plot.title = element_text(size = 20),text=element_text(size=txt)
-)
-  # theme(plot.title = element_text(size = 20))
-```
 

### 2021-02-08 Manuscripts/Manuscript_AM_v2.Rmd
Manuscript 2, testing graphs

@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
  ---
@@
 }
-```
 
-```{r, linear-model-correlated, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with a rm-ANOVA model fitted.
-Top row, simulated data, lines represent mean response.
-Bottom row, fitted model, thick lines represent predicted mean response"}
-# linear response, correlated errors (subject effect)
 txt<-18
-plot_example(example(fun_type = "linear", error_type = "correlated")) +
+A<-plot_example(example(fun_type = "linear", error_type = "correlated")) +
   plot_annotation(tag_levels='A',
@@
 )
-```
 
-```{r linear-model-independent, fig.width=16, fig.height=16,echo=FALSE,fig.cap='Simulated quadratic responses from two groups with an rm-ANOVA model fitted.
-Top row, simulated data, lines represent mean response.
-Bottom row, fitted model, thick lines represent predicted mean response', message=FALSE}
-# quadratic response, correlated errors (subject effect)
-txt<-18
-plot_example(example(fun_type = "linear", error_type = "independent")) +
+B<-plot_example(example(fun_type = "linear", error_type = "independent")) +
   plot_annotation(
@@
 )
-  # theme(plot.title = element_text(size = 20))
+
+C<-plot_example(example(fun_type = "quadratic", error_type = "correlated")) +
+  plot_annotation(tag_levels='A',
+  title = 'Simulated with quadratic function and independent errors')&
+    theme(plot.title = element_text(size = 20),text=element_text(size=txt)
+)
+
+D<-plot_example(example(fun_type = "quadratic", error_type = "independent")) +
+  plot_annotation(tag_levels='A',
+  title = 'Simulated with quadratic function and correlated errors')&
+  theme(plot.title = element_text(size = 20),text=element_text(size=txt)
+)
+
 ```
 
-It is clear from Figure \@ref(fig:linear-model) that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the "truth" of the simulated data.
+```{r, linear-models, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with a rm-ANOVA model fitted.
+Top row, simulated data, lines represent mean response.
+Bottom row, fitted model, thick lines represent predicted mean response"}
+# linear response, correlated errors (subject effect)
+A/B+plot_annotation(tag_levels='A')
+```
+
+It is clear from Figure \@ref(fig:linear-models) that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the "truth" of the simulated data.
 
@@
 Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears at the bottom right of Figure \@ref(fig:quadratic-response).
+
+```{r, quadratic-response, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
+Top row, simulated data, lines represent mean response.
+Bottom row, fitted model, thick lines represent predicted mean response"}
+# linear response, correlated errors (subject effect)
+C/D+plot_annotation(tag_levels='A')
+```
 

### 2021-02-08 Manuscripts/Manuscript_AM_v2.Rmd
Working on figures, issue with pdf_document2

@@
----
-output:
-  pdf_document: default
-  html_document: default
----
 ---

### 2021-02-09 Manuscripts/Manuscript_AM_v2.Rmd
Working on manuscript

--- 2021-02-04 Manuscripts/Manuscript_AM_v2.Rmd
@@
 In the second case, a quadratic function is used for the mean response per group (concave down for Group 1 and concave up for Group 2 ).
-From the mean response per group in the linear or quadratic case, individual responses for each subject in each group are simulated using covariance matrices for with compound symmetry (constant variance across time) or independent errors (errors that are generated from a a normal distribution but are not constant over time).
+From the mean response per group (linear or quadratic), individual responses for each subject within each group are simulated using covariance matrices for with compound symmetry (constant variance across time) or independent errors (errors that are generated from a a normal distribution but are not constant over time).
 Thus, the response per subject corresponds to the mean response per group plus the error (compound symmetry or independent).
-To this data, an rm-ANOVA model identical as Equation \@ref:eqn(linear-model) is fitted.
+To this data, an rm-ANOVA model identical as Equation \@ref:eq(linear-model) is fitted.
 The resulting simulated data, simulated errors, individual response and fitted parameters from the model are plotted.
@@
 
-  x <- seq(0, 1, length.out = n_time)
+  x <- seq(1,6, length.out = n_time)
   mu <- matrix(0, length(x), 2)
@@
   if (fun_type == "linear") {
-    mu[, 1] <- - (x - .5) + 0.25
-    mu[, 2] <- (x - .3)
+    mu[, 1] <- - (0.25*x)+2  
+    mu[, 2] <- 0.25*x+2
   } else {
@@
 
-    mu[, 1] <- - 4 * (x - .5)^2 + 0.25
-    mu[, 2] <- 4 * (x - .5)^2
+    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
+    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
   }
@@
     ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point() +
-    geom_line(aes(x = time, y = mu, color = treatment)) +
+    geom_point(show.legend=FALSE) +labs(y='response')+
+    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
     theme_classic() +
-    ggtitle("Simulated data with true response function")+
+    #ggtitle("Simulated data with true response function")+
     theme(plot.title = element_text(size = 30, 
@@
     ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects")) +
+    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
     # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1) +
+    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +
     scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
+    #ggtitle("Simulated data\nIndividual responses with population mean") +
     theme_classic()+
-    scale_color_aaas()
-
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line() +
-    # facet_wrap(~ treatment) +
-    ggtitle("Simulated errors") +
-     theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    theme_classic()+
-    scale_color_aaas()
-
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point() +
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    ggtitle("Fitted Model")+
      theme(plot.title = element_text(size = 30, 
@@
 
-  return((p1+p3+p2+p4)+plot_layout(nrow=1))
+   p3 <- sim_dat$dat %>%
+    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
+    geom_line(show.legend=FALSE) +labs(y='response')+
+     theme_classic()+
+    # facet_wrap(~ treatment) +
+    #ggtitle("Simulated errors") +
+     theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
+    scale_color_aaas()
+
+  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
+    geom_point()+labs(y='response')+
+    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
+    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
+    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
+    theme_classic() +
+    #ggtitle("Fitted Model")+
+    theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
+    scale_color_aaas()
+
+  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
   # return((p1 + p3 + plot_layout(guides = "collect")) / (p2 + p4 + plot_layout(guides = "collect")))
@@
 }
+
+txt<-18
+A<-plot_example(example(fun_type = "linear", error_type = "correlated"))
+
+B<-plot_example(example(fun_type = "linear", error_type = "independent"))
+
+C<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
+
+D<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
+
 ```
 
-```{r, linear-model-correlated, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with a rm-ANOVA model fitted.
-Top row, simulated data, lines represent mean response.
-Bottom row, fitted model, thick lines represent predicted mean response"}
+```{r, linear-models, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
+B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
+C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
+D,H: Estimations from the rm-ANOVA model for the mean group response.
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data."}
 # linear response, correlated errors (subject effect)
-txt<-18
-plot_example(example(fun_type = "linear", error_type = "correlated")) +
-  plot_annotation(tag_levels='A',
-                  title = 'Simulated with linear function and correlated errors') &
-   theme(plot.title = element_text(size = 20),text=element_text(size=txt)
-)
+A/B+plot_annotation(tag_levels = 'A')
 ```
 
-```{r linear-model-independent, fig.width=16, fig.height=16,echo=FALSE,fig.cap='Simulated quadratic responses from two groups with an rm-ANOVA model fitted.
-Top row, simulated data, lines represent mean response.
-Bottom row, fitted model, thick lines represent predicted mean response', message=FALSE}
-# quadratic response, correlated errors (subject effect)
-txt<-18
-plot_example(example(fun_type = "linear", error_type = "independent")) +
-  plot_annotation(
-    tag_levels='A',
-  title = 'Simulated with linear function and independent errors') &
-   theme(plot.title = element_text(size = 20),text=element_text(size=txt)
-)
-  # theme(plot.title = element_text(size = 20))
+In Figure \@ref(fig:linear-models)  panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
+In panel G,because the errors are independent, there is no expectation that responses are likely to follow a trend.
+It is clear from \@ref(fig:linear-models) fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the "truth" of the simulated data.In particular,
+
+However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:quadratic-response).
+Here, the mean response per group was simulated using a quadratic function and errors, individual responses and the rm-ANOVA model were produced in the same manner as \@ref(fig:linear-models) .
+The figure shows that changes in each group occur through the timeline, and that the mean value is the same as the initial value by the fifth time point.
+Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panels D and H in Figure \@ref(fig:quadratic-response).
+
+```{r, quadratic-response, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
+A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
+B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
+C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
+D,H: Estimations from the rm-ANOVA model for the mean group response.
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data."}
+# linear response, correlated errors (subject effect)
+C/D+plot_annotation(tag_levels='A')
 ```
 
-It is clear from Figure \@ref(fig:linear-model) that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the "truth" of the simulated data.
-
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:quadratic-response).
-Here, the simulated data follows a quadratic behavior.
-The figure shows that changes in each group occur through the timeline, although the final mean value is the same as the initial value.
-Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears at the bottom right of Figure \@ref(fig:quadratic-response).
-
-In this case, compare the predictions to the simulated data and it is noticeable that the model is not capturing the changes within each group throughout the timeline.
-This highlights the limitation of rm-ANOVA and LMEMs with longitudinal non-linear data, where the model is only flexible enough to allow different slopes per group, but is unable to follow any trend that is highly variable.
+In this case, a comparison of the predictions of the rm-ANOVA model to the the simulated data and indicates that the model is not capturing the changes within each group in a good way.
+This is expected as the rm-ANOVA model is only able to accommodate linear changes, which in this case fail to capture the changes over the timeline.
+Specifically, note that in Figure \@ref(fig:quadratic-response) panels D and H the model is not capturing the increase in the response through time points 2 and 4.
+If such incrase may indicate a physiologically relevant change, the model will not be able to capture it, potentially limiting inferences from the data.
 

### 2021-02-09 Manuscripts/Manuscript_AM_v2.tex
Working on manuscript

@@
 
-A longitudinal study is designed to repeatedly measure a variable of interest in a group (or groups) of subjects, when the intention is to observe the evolution of the effect of a certain treatment across time rather than analyzing it at a single time point (e.g., a cross-sectional study).
+A longitudinal study is designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of the treatment effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
 Medical research commonly uses longitudinal studies to analyze the evolution of patients before, during and after treatment.
-For example, Sio et al analyzed the evolution of dermatitis symptoms in breast cancer patients after radiotherapy (RT), by taking weekly measurements over the course of two moths {[}1{]}, and Kamstra et al measured mouth opening at regular monthly intervals after RT in neck cancer patients {[}2{]}.
-Biomedical research has used longitudinal studies as well to analyze tumor response {[}3--6{]}, antibody expression {[}7,8{]}, and cell metabolism {[}9,10{]}.
+For example, Sio et.
+al.~analyzed the evolution of dermatitis symptoms in breast cancer patients after radiotherapy (RT), by taking weekly measurements over the course of two months {[}1{]}, and Kamstra et al measured mouth opening at regular monthly intervals after RT in neck cancer patients {[}2{]}.
+Other studies using longitudinal designs in biomedical research have tumor analyzed tumor response {[}3--6{]}, antibody expression {[}7,8{]}, and cell metabolism {[}9,10{]}.
 
 Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as a limiting frequency{[}11{]} by assuming a null hypothesis using the \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-This methodology makes two key assumptions regarding longitudinal data: constant correlation exists across same-subject measurements, and observations from each subject are obtained at all time points through the study (a condition also known as complete observations) {[}12,13{]}.
+The frequentist paradigm regards probability as a limiting frequency{[}11{]} by assuming a null hypothesis under a statistical model that is often assumed to be an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
+The rm-ANOVA model typically makes two key assumptions regarding longitudinal data: constant correlation across same-subject measurements, and observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}12,13{]}.
 
-The assumption of constant correlation is frequently unjustified as correlation in time tends to diminish between measurements when the time interval between them increases {[}14{]}.
+The assumption of constant correlation (often known as the \emph{compoun symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}14{]}.
 Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and it is even possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point and therefore violating the \emph{complete observations} assumption of classical rm-ANOVA.
+For example, in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in preclinical animal studies due to injury or weight loss is possible.
+It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
+In each of these missing data scenarios,the \emph{complete observations} assumption of classical rm-ANOVA is violated.
 
-When incomplete observations occur, an rm-ANOVA is fit by excluding all subjects with missing observations from the analysis{[}12{]}.
-This can result in increased costs if the desired statistical power is not met with the remaining observations, as it would be necessary to enroll more subjects.
+When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis{[}12{]}.
+This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
 At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
@@
 A \emph{post hoc} analysis is based on multiple repeated comparisons to estimate a \emph{p-value}, a metric that is widely used as a measure of significance.
-Because the \emph{p-value} is highly variable, multiple comparisons can inflate the false positivity rate {[}15,16{]}, consequently biasing the conclusions of the study.
+Because the \emph{p-value} is highly variable, multiple comparisons can inflate the false positivity rate (Type I error) {[}15,16{]}, consequently biasing the conclusions of the study.
+Although corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}17{]}), they in turn reduce statistical power (\(\alpha\)){[}18{]}, and lead to increased Type II error (fail to reject the null hypothesis although it is false {[}19,20{]}.
+Therefore, the \emph{post hoc} comparisons commonly used in rm-ANOVA impose a tradeoff between Type I and II errors.
 
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-The recognition on the limitations of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal data {[}10,17{]}.
-Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g.~the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population.
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}18{]}.
-On the other hand, LMEMs impose restrictions in the distribution of the errors of the random effects {[}12{]}.
+The recognition on the limitations of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor data {[}10,21{]}.
+Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}22{]}.
+On the other hand, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}12,23{]}.
 
-One final assumption that is not initially evident for both rm-ANOVA and LMEMs models is that the mean response is expected to change linearly through time{[}18{]}.
-The linearity assumption in both rm-ANOVA ane LMEMs consequently restricts the inferences from the model when the data does not follow a linear trend.
-In biomedical research, a particular case of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings {[}3,10,17{]}.
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of and the estimations of an LMEM or rm-ANOVA model inconsistent with the observed variation.
-In other words, both LMEMs and rm-ANOVA explain highly-variable data using a linear trend, consequently biasing the estimates when the true response is non-linear in time.
-Additionally, although it is possible that a \emph{post hoc} analysis is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) it has to be reminded that the validity of such metric relies on how adequate the model is to the data.
-In other words, the \emph{p-value} requires that the model and the data have good agreement and if that is not the case, a ``Type 3'' error (known as ``model misspecification'') occurs{[}19{]}.That would be the case if for example, a model that is only able to explain linear responses is fitted to data that has a quadratic behavior.
-In such scenario, the \emph{p-value} becomes invalid{[}20{]}.
+One final assumption that is not initially evident for both rm-ANOVA and LMEMs models is that the mean response is expected to change linearly through time{[}22{]}.
+The linearity assumption in both rm-ANOVA and LMEMs implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
+In biomedical research, a particular case of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings {[}3,10,21{]}.
+These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit LMEMs or rm-ANOVA model inconsistent with the observed variation.
+Therefore, when LMEMs and rm-ANOVA are used to draw inference of such highly-variable data the estimates are inevitably biased, because non-linear data is provided to a model which can only accomodate linear trends.
 
-As the ``frequentist'' rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, enabling inference that is appropiate to the data.
+Additionally, although it is possible that a \emph{post hoc} analysis is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) the validity of such metric relies on how adequate the model fits the data.
+In other words, the \emph{p-value} requires that the model and the data have good agreement and if that is not the case, a ``Type 3'' error (known as ``model misspecification'') occurs{[}24{]}.
+For example, a model that is only able to explain linear responses but is fitted to data that has a quadratic behavior results in \emph{p-values} and parameter estimates that are invalid{[}25{]}.
+
+As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed behavior) to determine the trend in the data, which enables inference that is appropriate.
+
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data {[}21--23{]}, health-environment interactions {[}24{]} and the dynamics of government in political science {[}25{]} .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend of the model, can easily accommodate missing observations and can model non-constant correlation between repeated measurements {[}26{]}.Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data.
+Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data {[}26--28{]}, health-environment interactions {[}29{]} and the dynamics of government in political science {[}30{]} .
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend of the model, can easily accommodate missing observations, and can model non-constant correlation between repeated measurements {[}31{]}.
+Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data.
 
 The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of more complex models beyond LMEMs.
-In particular, \(\textsf{R}\) has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}26,27{]} that not only speed up the initial stages of the analysis but enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}28{]}.
-This allows to create and explore different alternatives for analysis without collecting information in the field, and reduces the time window between experiment design and its implementation.
+In particular, \(\textsf{R}\){[}{\textbf{???}}{]}has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}31,32{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}33{]}.
+Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and can be also used for power calculations and study design questions.
 
-The purpose of this work is to provide biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of rm-ANOVA and LMEMs to analyze longitudinal data are contrasted in the cases of missing observations, assumption of linearity of response and constant correlation structures.
-Second, the key theoretical elements of GAMs are presented without using extensive derivations or complex mathematical notation that can obscure their understanding.
-Third, simulated data that follows the trend of previously reported values {[}17{]} is used to illustrate the type of non-linear longitudinal data that occurs in biomedical research to highlight the differences in inference between rm-ANOVA, LMEMs and GAMs and the implementation of the latter in such scenario.
-Finally, reproducibility is emphasized by providing the code to generate the data and the implementation of GAMs in \(\textsf{R}\), in conjunction with a step-by-step guide to fit models of increasing complexity.
+This work provides biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas.
+First, the limitations of LMEMs and rm-ANOVA regarding missing observations, assumption of linearity of response and constant correlation structures is explained in detail.
+Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
+Third, simulated data that reproduces patterns in previously reported studies {[}21{]} is used to illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
+The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
+Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in \(\textsf{R}\), in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
+
 In summary, the exploration of modern statistical techniques to analyze longitudinal data may allow biomedical researchers to consider the use of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a linear trend, and will also help to improve the standards for reproducibility in biomedical research.
@@
 
-\hypertarget{the-frequentist-case-for-longitudinal-data}{%
-\subsection{The ``frequentist'' case for longitudinal data}\label{the-frequentist-case-for-longitudinal-data}}
+\hypertarget{the-repeated-measures-anova}{%
+\subsection{The repeated measures ANOVA}\label{the-repeated-measures-anova}}
 
@@
 In a biomedical longitudinal study, two or more groups of subjects (patients, mice, samples) are subject to different treatments (e.g., a ``treatment'' group receives a novel drug vs.~a ``control'' group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response has both \emph{fixed} and \emph{random} components.
-The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the effect of the novel drug in the ``treatment'' group.
+The collected respnse is modeled with both \emph{fixed} and \emph{random} components.
+The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the average effect of the novel drug in the ``treatment'' group.
 The \emph{random} component can be defined as ``noise'' caused by some inherent variability within the study.
@@
 \begin{equation}
-y_{ijt} = \mu+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\pi_{ij} +\varepsilon_{tij}\\ 
+y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{tij}\\ 
 \label{eq:linear-model}
@@
 
-In this model \(y_{ijt}\) is the response by subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\mu\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\).
-A \emph{random effect} \(\pi_{ij}\) of each subject within each group, and independent errors per subject per time per group \(\varepsilon_{tij}\), that represent random variation not explained by the \emph{fixed} or \emph{random} effects.
-Suppose two treatments groups are used in the study.
-Then the models per group can be written as below with \(j=0\) representing the first treatment group (Group A) and \(j=1\) representing the second treatment group (Group B).
-The linear models then become:
+In this model \(y_{ijt}\) is the response by subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
+The \emph{random effect} \(\mu_{ij}\), accounts for variability in each subject within each group by allowing different values for each subject.
+Independent errors \(\varepsilon_{tij}\) represent random variation not explained by the \emph{fixed} or \emph{random} effects.
+In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy'').
+Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(j=0\) representing the first treatment group (Group A) and \(j=1\) representing the second treatment group (Group B).
+The linear models then can be expressed as
 
@@
 
-To further simplify the expression, let us substitute \(\widetilde{\mu}=\mu+\beta_{2}\) and \(\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}\) in the equation for Group B. The equation then becomes:
+To further simplify the expression, substitute \(\widetilde{\mu}=\beta_0+\beta_{2}\) and \(\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}\) in the equation for Group B.
+This substitution allows for a different intercept and slope for Group B.
+The model is written as
 
@@
 
-Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different slopes per group).
-In other words, When it is stated that an rm-ANOVA model ``expects'' a linear relationship between the covariates and the response, this means that either presented as \eqref{eq:linear-model}, \eqref{eq:ANOVA-by-group} or \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data, the maximum flexibility being that it can assign different slopes to each group.
-Because the construction of the model is essentially the same in LMEMs, the expected linearity holds in such models, although it is possible to fit models with different slopes at the subject-level{[}18{]}.\\
-When the data does not follow a linear trend, the fit that an rm-ANOVA or LMEM model produces does not accurately represent the trends in the data.
-To demonstrate this, longitudinal data where a normally distributed response of two groups of 10 subjects each is simulated, and a rm-ANOVA model (\eqref{eq:linear-model}) is fitted to the data using the package \emph{nlme} (the code for the simulated data and models is available in the Appendix).
+Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
+In other words, When it is stated that an rm-ANOVA model ``expects'' a linear relationship between the covariates and the response, this means that either presented as Equation \eqref{eq:linear-model}, Equation \eqref{eq:ANOVA-by-group} or Equation \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data.
+If the data show non-linear behavior, the rm-ANOVa model will approximate this behavior with non-parallel lines.
+Because the construction of the model is essentially the same in LMEMs, the linearity assumption holds for these models, although it is possible to fit models with different slopes at the subject-level{[}22{]}.
+
+When the data does not follow a linear trend, the fit that an rm-ANOVA or LMEM produces may not accurately represent the trends in the data.
+To demonstrate this inaccurate representation, longitudinal data where a normally distributed response of two groups of 10 subjects each is simulated, and a rm-ANOVA model (Equation \eqref{eq:linear-model}) is fitted to the data using \(\textsf{R}\){[}{\textbf{???}}{]} and the package \emph{nlme}{[}{\textbf{???}}{]} (code in the Appendix).
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes (negative slope for Group 1 and positive slope for Group 2 ).
+In the second case, a quadratic function is used for the mean response per group (concave down for Group 1 and concave up for Group 2 ).
+From the mean response per group in the linear or quadratic case, individual responses for each subject in each group are simulated using covariance matrices for with compound symmetry (constant variance across time) or independent errors (errors that are generated from a a normal distribution but are not constant over time).
+Thus, the response per subject corresponds to the mean response per group plus the error (compound symmetry or independent).
+To this data, an rm-ANOVA model identical as Equation @ref:eqn(linear-model) is fitted.
+The resulting simulated data, simulated errors, individual response and fitted parameters from the model are plotted.
 
@@
 \centering
-\includegraphics{Manuscript_AM_v2_files/figure-latex/linear-model-1.pdf}
-\caption{\label{fig:linear-model}Simulated linear responses from two groups with a rm-ANOVA model fitted.
+\includegraphics{Manuscript_AM_v2_files/figure-latex/linear-models-1.pdf}
+\caption{\label{fig:linear-models}Simulated linear responses from two groups with a rm-ANOVA model fitted.
 Top row, simulated data, lines represent mean response.
@@
 
-It is clear from Figure \ref{fig:linear-model} that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data.
+It is clear from Figure \ref{fig:linear-models} that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data.
 
@@
 \includegraphics{Manuscript_AM_v2_files/figure-latex/quadratic-response-1.pdf}
-\caption{\label{fig:quadratic-response}Simulated quadratic responses from two groups with an rm-ANOVA model fitted.
+\caption{\label{fig:quadratic-response}Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
 Top row, simulated data, lines represent mean response.
@@
 In a longitudinal study there is an expected \emph{variance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}29{]} (For an in-depth analysis of the covariance matrix see {[}30,31{]}).
+The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}34{]} (For an in-depth analysis of the covariance matrix see {[}35,36{]}).
 
 In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}31--33{]}.
-However, it has been shown that this condition is frequently unjustified because the correlation between measurements tends to change over time {[}34{]}; and it is higher between consecutive measurements {[}12,14{]}.
+Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}36--38{]}.
+However, it has been shown that this condition is frequently unjustified because the correlation between measurements tends to change over time {[}39{]}; and it is higher between consecutive measurements {[}12,14{]}.
 A visual representation of the
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser) the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}28{]}, and they are not robust if the group sizes are unbalanced {[}35{]}.
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser) the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}33{]}, and they are not robust if the group sizes are unbalanced {[}40{]}.
 In other words, if the data does not present constant correlation between repeated measurements, the assumptions required for an rm-ANOVA model are not met and the use of corrections may still not provide a reasonable adjustment that makes the model valid.
 
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}18{]}.
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}22{]}.
 Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself.
@@
 Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}36{]}.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}41{]}.
 Dropout from patients, attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}31{]}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}37{]}.
-If the data are MCAR, this means that the missigness is completely unrelated to the collected information {[}38{]}, and in the case of MNAR the missing values are dependent on their value.
+Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}36{]}.
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}42{]}.
+If the data are MCAR, this means that the missigness is completely unrelated to the collected information {[}43{]}, and in the case of MNAR the missing values are dependent on their value.
 An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}39{]}.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}44{]}.
 
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}30{]}.
-The pattern of missing observations can be considered MAR if the missing observations are not related any of the other variables measured in the study {[}34{]}.
+In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}35{]}.
+The pattern of missing observations can be considered MAR if the missing observations are not related any of the other variables measured in the study {[}39{]}.
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
@@
 
-A GAM is a special case of the Generalized Linear Model (GLM), a framework that allows for response distributions that do not follow a normal distribution {[}26,40{]}.
-Following the notation by Simpson {[}23{]} A GAM model can be represented as:
+A GAM is a special case of the Generalized Linear Model (GLM), a framework that allows for response distributions that do not follow a normal distribution {[}31,45{]}.
+Following the notation by Simpson {[}28{]} A GAM model can be represented as:
 
@@
 This approach is more advantageous as it does not restrict the model to a linear relationship.
-One possible function for \(f(x_t\mid \beta)\) that allows for non-linear responses is a polynomial, but a major limitation is that they create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}25{]}.In particular, as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta)\) goes to \(\pm \infty\) which is almost always unrealistic.
+One possible function for \(f(x_t\mid \beta)\) that allows for non-linear responses is a polynomial, but a major limitation is that they create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}30{]}.In particular, as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta)\) goes to \(\pm \infty\) which is almost always unrealistic.
 
 The model specification for GAMs requires that the \emph{smooth functions} are represented in a parametric way that fit within the GLM framework, and this step is achieved by using \emph{basis functions} to represent them.
-A \emph{basis} is a set of functions that capture the space where the smooths that approximate \(f(x_t \mid \beta)\) exist {[}23{]}.
+A \emph{basis} is a set of functions that capture the space where the smooths that approximate \(f(x_t \mid \beta)\) exist {[}28{]}.
 For the linear model in \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}41{]}, which in the case of \eqref{eq:linear-model} is restricted to a linear behavior.
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}46{]}, which in the case of \eqref{eq:linear-model} is restricted to a linear behavior.
 In the case of \eqref{eq:GAM}, the basis function is \(f(x_t\mid \beta)\), which means that the model allows relationships beyond linear for the covariates.
@@
 In GAMs the \emph{basis functions} are represented over evenly spaced ranges of the covariates known as \emph{knots}.
-A commonly used \emph{basis function} is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations{[}23,26{]}.
-Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness {[}42{]}.
+A commonly used \emph{basis function} is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations{[}28,31{]}.
+Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness {[}47{]}.
 Therefore, GAMs overcome the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear, such as Figure \ref{fig:quadratic-response}.
-Regarding longitudinal data, Pedersen et al {[}22{]} demonstrated the capabilities of GAMs in this area using ecological data.
+Regarding longitudinal data, Pedersen et al {[}27{]} demonstrated the capabilities of GAMs in this area using ecological data.
 
@@
 
+\leavevmode\hypertarget{ref-abdi2010}{}%
+{[}17{]} H. Abdi, Holm's sequential bonferroni procedure, Encyclopedia of Research Design. 1 (2010) 1--8.
+
+\leavevmode\hypertarget{ref-nakagawa2004}{}%
+{[}18{]} S. Nakagawa, A farewell to bonferroni: The problems of low statistical power and publication bias, Behavioral Ecology. 15 (2004) 1044--1045.
+
+\leavevmode\hypertarget{ref-gelman2012}{}%
+{[}19{]} A. Gelman, J. Hill, M. Yajima, Why we (usually) don't have to worry about multiple comparisons, Journal of Research on Educational Effectiveness. 5 (2012) 189--211.
+
+\leavevmode\hypertarget{ref-albers2019}{}%
+{[}20{]} C. Albers, The problem with unadjusted multiple and sequential statistical testing, Nature Communications. 10 (2019) 1--4.
+
 \leavevmode\hypertarget{ref-vishwanath2009}{}%
-{[}17{]} K. Vishwanath, H. Yuan, W.T. Barry, M.W. Dewhirst, N. Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia. 11 (2009) 889--900.
+{[}21{]} K. Vishwanath, H. Yuan, W.T. Barry, M.W. Dewhirst, N. Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia. 11 (2009) 889--900.
 
 \leavevmode\hypertarget{ref-pinheiro2006}{}%
-{[}18{]} J. Pinheiro, D. Bates, Mixed-effects models in s and s-plus, Springer Science \& Business Media, 2006.
+{[}22{]} J. Pinheiro, D. Bates, Mixed-effects models in s and s-plus, Springer Science \& Business Media, 2006.
+
+\leavevmode\hypertarget{ref-barr2013}{}%
+{[}23{]} D.J. Barr, R. Levy, C. Scheepers, H.J. Tily, Random effects structure for confirmatory hypothesis testing: Keep it maximal, Journal of Memory and Language. 68 (2013) 255--278.
 
 \leavevmode\hypertarget{ref-dennis2019}{}%
-{[}19{]} B.
+{[}24{]} B.
 Dennis, J.M.
@@
 \leavevmode\hypertarget{ref-wang2019}{}%
-{[}20{]} B. Wang, Z. Zhou, H. Wang, X.M. Tu, C. Feng, The p-value and model specification in statistics, General Psychiatry. 32 (2019).
+{[}25{]} B. Wang, Z. Zhou, H. Wang, X.M. Tu, C. Feng, The p-value and model specification in statistics, General Psychiatry. 32 (2019).
 
 \leavevmode\hypertarget{ref-rose2012}{}%
-{[}21{]} N.L.
+{[}26{]} N.L.
 Rose, H.
@@
 \leavevmode\hypertarget{ref-pedersen2019}{}%
-{[}22{]} E.J. Pedersen, D.L. Miller, G.L. Simpson, N. Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, PeerJ. 7 (2019) e6876.
+{[}27{]} E.J. Pedersen, D.L. Miller, G.L. Simpson, N. Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, PeerJ. 7 (2019) e6876.
 
 \leavevmode\hypertarget{ref-simpson2018}{}%
-{[}23{]} G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018) 149.
+{[}28{]} G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018) 149.
 
 \leavevmode\hypertarget{ref-yang2012}{}%
-{[}24{]} L.
+{[}29{]} L.
 Yang, G.
@@
 \leavevmode\hypertarget{ref-beck1998}{}%
-{[}25{]} N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. (1998) 596--627.
+{[}30{]} N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. (1998) 596--627.
 
 \leavevmode\hypertarget{ref-wood2017}{}%
-{[}26{]} S.N. Wood, Generalized additive models: An introduction with r, CRC press, 2017.
+{[}31{]} S.N. Wood, Generalized additive models: An introduction with r, CRC press, 2017.
 
 \leavevmode\hypertarget{ref-wood2016}{}%
-{[}27{]} S.N. Wood, N., Pya, B. Säfken, Smoothing parameter and model selection for general smooth models (with discussion), Journal of the American Statistical Association. 111 (2016) 1548--1575.
+{[}32{]} S.N. Wood, N., Pya, B. Säfken, Smoothing parameter and model selection for general smooth models (with discussion), Journal of the American Statistical Association. 111 (2016) 1548--1575.
 
 \leavevmode\hypertarget{ref-haverkamp2017}{}%
-{[}28{]} N.
+{[}33{]} N.
 Haverkamp, A.
@@
 \leavevmode\hypertarget{ref-wolfinger1996}{}%
-{[}29{]} R.D. Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics. (1996) 205--230.
+{[}34{]} R.D. Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics. (1996) 205--230.
 
 \leavevmode\hypertarget{ref-west2014}{}%
-{[}30{]} B.T. West, K.B. Welch, A.T. Galecki, Linear mixed models: A practical guide using statistical software, CRC Press, 2014.
+{[}35{]} B.T. West, K.B. Welch, A.T. Galecki, Linear mixed models: A practical guide using statistical software, CRC Press, 2014.
 
 \leavevmode\hypertarget{ref-weiss2005}{}%
-{[}31{]} R.E. Weiss, Modeling longitudinal data, Springer Science \& Business Media, 2005.
+{[}36{]} R.E. Weiss, Modeling longitudinal data, Springer Science \& Business Media, 2005.
 
 \leavevmode\hypertarget{ref-geisser1958}{}%
-{[}32{]} S. Geisser, S.W. Greenhouse, others, An extension of box's results on the use of the \(F\) distribution in multivariate analysis, The Annals of Mathematical Statistics. 29 (1958) 885--891.
+{[}37{]} S. Geisser, S.W. Greenhouse, others, An extension of box's results on the use of the \(F\) distribution in multivariate analysis, The Annals of Mathematical Statistics. 29 (1958) 885--891.
 
 \leavevmode\hypertarget{ref-huynh1976}{}%
-{[}33{]} H. Huynh, L.S. Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics. 1 (1976) 69--82.
+{[}38{]} H. Huynh, L.S. Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics. 1 (1976) 69--82.
 
 \leavevmode\hypertarget{ref-maxwell2017}{}%
-{[}34{]} S.E. Maxwell, H.D. Delaney, K. Kelley, Designing experiments and analyzing data: A model comparison perspective, Routledge, 2017.
+{[}39{]} S.E. Maxwell, H.D. Delaney, K. Kelley, Designing experiments and analyzing data: A model comparison perspective, Routledge, 2017.
 
 \leavevmode\hypertarget{ref-keselman2001}{}%
-{[}35{]} H. Keselman, J. Algina, R.K. Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematical and Statistical Psychology. 54 (2001) 1--20.
+{[}40{]} H. Keselman, J. Algina, R.K. Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematical and Statistical Psychology. 54 (2001) 1--20.
 
 \leavevmode\hypertarget{ref-molenberghs2004}{}%
-{[}36{]} G. Molenberghs, H. Thijs, I. Jansen, C. Beunckens, M.G. Kenward, C. Mallinckrodt, R.J. Carroll, Analyzing incomplete longitudinal clinical trial data, Biostatistics. 5 (2004) 445--464.
+{[}41{]} G. Molenberghs, H. Thijs, I. Jansen, C. Beunckens, M.G. Kenward, C. Mallinckrodt, R.J. Carroll, Analyzing incomplete longitudinal clinical trial data, Biostatistics. 5 (2004) 445--464.
 
 \leavevmode\hypertarget{ref-scheffer2002}{}%
-{[}37{]} J. Scheffer, Dealing with missing data, (2002).
+{[}42{]} J. Scheffer, Dealing with missing data, (2002).
 
 \leavevmode\hypertarget{ref-potthoff2006}{}%
-{[}38{]} R.F.
+{[}43{]} R.F.
 Potthoff, G.E.
@@
 \leavevmode\hypertarget{ref-ma2012}{}%
-{[}39{]} Y.
+{[}44{]} Y.
 Ma, M.
@@
 \leavevmode\hypertarget{ref-hastie1987}{}%
-{[}40{]} T. Hastie, R. Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association. 82 (1987) 371--386.
+{[}45{]} T. Hastie, R. Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association. 82 (1987) 371--386.
 
 \leavevmode\hypertarget{ref-hefley2017}{}%
-{[}41{]} T.J.
+{[}46{]} T.J.
 Hefley, K.M.
@@
 \leavevmode\hypertarget{ref-wegman1983}{}%
-{[}42{]} E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365.
+{[}47{]} E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365.
 

### 2021-02-09 Manuscripts/Manuscript_AM_v2.tex
Finalized revision for now, updated pdf and docx

@@
 The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of more complex models beyond LMEMs.
-In particular, \(\textsf{R}\){[}{\textbf{???}}{]}has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}31,32{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}33{]}.
+In particular, \(\textsf{R}\){[}32{]}has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}31,33{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}34{]}.
 Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and can be also used for power calculations and study design questions.
@@
 When the data does not follow a linear trend, the fit that an rm-ANOVA or LMEM produces may not accurately represent the trends in the data.
-To demonstrate this inaccurate representation, longitudinal data where a normally distributed response of two groups of 10 subjects each is simulated, and a rm-ANOVA model (Equation \eqref{eq:linear-model}) is fitted to the data using \(\textsf{R}\){[}{\textbf{???}}{]} and the package \emph{nlme}{[}{\textbf{???}}{]} (code in the Appendix).
+To demonstrate this inaccurate representation, longitudinal data where a normally distributed response of two groups of 10 subjects each is simulated, and a rm-ANOVA model (Equation \eqref{eq:linear-model}) is fitted to the data using \(\textsf{R}\){[}32{]} and the package \emph{nlme}{[}35{]} (code in the Appendix).
 Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes (negative slope for Group 1 and positive slope for Group 2 ).
 In the second case, a quadratic function is used for the mean response per group (concave down for Group 1 and concave up for Group 2 ).
-From the mean response per group in the linear or quadratic case, individual responses for each subject in each group are simulated using covariance matrices for with compound symmetry (constant variance across time) or independent errors (errors that are generated from a a normal distribution but are not constant over time).
+From the mean response per group (linear or quadratic), individual responses for each subject within each group are simulated using covariance matrices for with compound symmetry (constant variance across time) or independent errors (errors that are generated from a a normal distribution but are not constant over time).
 Thus, the response per subject corresponds to the mean response per group plus the error (compound symmetry or independent).
-To this data, an rm-ANOVA model identical as Equation @ref:eqn(linear-model) is fitted.
+To this data, an rm-ANOVA model identical as Equation @ref:eq(linear-model) is fitted.
 The resulting simulated data, simulated errors, individual response and fitted parameters from the model are plotted.
@@
 \includegraphics{Manuscript_AM_v2_files/figure-latex/linear-models-1.pdf}
-\caption{\label{fig:linear-models}Simulated linear responses from two groups with a rm-ANOVA model fitted.
-Top row, simulated data, lines represent mean response.
-Bottom row, fitted model, thick lines represent predicted mean response}
+\caption{\label{fig:linear-models}\textbf{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
+B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
+C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
+D,H: Estimations from the rm-ANOVA model for the mean group response.
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.}}
 \end{figure}
 
-It is clear from Figure \ref{fig:linear-models} that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data.
+In Figure \ref{fig:linear-models} panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
+In panel G,because the errors are independent, there is no expectation that responses are likely to follow a trend.
+It is clear from \ref{fig:linear-models} fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data.In particular,
 
 However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:quadratic-response}.
-Here, the simulated data follows a quadratic behavior.
-The figure shows that changes in each group occur through the timeline, although the final mean value is the same as the initial value.
-Fitting an rm-ANOVA model \eqref{eq:linear-model} to this data produces the fit that appears at the bottom right of Figure \ref{fig:quadratic-response}.
+Here, the mean response per group was simulated using a quadratic function and errors, individual responses and the rm-ANOVA model were produced in the same manner as \ref{fig:linear-models} .
+The figure shows that changes in each group occur through the timeline, and that the mean value is the same as the initial value by the fifth time point.
+Fitting an rm-ANOVA model \eqref{eq:linear-model} to this data produces the fit that appears in panels D and H in Figure \ref{fig:quadratic-response}.
 
@@
 \includegraphics{Manuscript_AM_v2_files/figure-latex/quadratic-response-1.pdf}
-\caption{\label{fig:quadratic-response}Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-Top row, simulated data, lines represent mean response.
-Bottom row, fitted model, thick lines represent predicted mean response}
+\caption{\label{fig:quadratic-response}\textbf{Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
+A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
+B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
+C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
+D,H: Estimations from the rm-ANOVA model for the mean group response.
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.}}
 \end{figure}
 
-In this case, compare the predictions to the simulated data and it is noticeable that the model is not capturing the changes within each group throughout the timeline.
-This highlights the limitation of rm-ANOVA and LMEMs with longitudinal non-linear data, where the model is only flexible enough to allow different slopes per group, but is unable to follow any trend that is highly variable.
+In this case, in Figure \ref{fig:quadratic-response}, panel A shows that both groups start with the same mean value at the first time point.
+This is intentional, as this is the expected behavior in a fold change analysis of a physiological quantity (such as tumor response to chemotherapy).
+Therefore, the idea behind \ref{fig:quadratic-response} is to mimic a non-linear behavior from two groups that follow the inverse trajectory.
+A comparison of the predictions of the rm-ANOVA model to the the simulated data (\ref{fig:quadratic-response}, panels D and H) indicates that the model is not capturing the changes within each group in a good way.
+Specifically, note that in Figure \ref{fig:quadratic-response} the predictions of the rm-ANOVA model (panels D and H) show that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
+If such change has important physiological implications,the rm-ANOVA model omits it from the predictions, potentially limiting valuable inferences from the data.
 
@@
 
-In a longitudinal study there is an expected \emph{variance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}34{]} (For an in-depth analysis of the covariance matrix see {[}35,36{]}).
+In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
+The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}36{]} (For an in-depth analysis of the covariance matrix see {[}37,38{]}).
 
 In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}36--38{]}.
-However, it has been shown that this condition is frequently unjustified because the correlation between measurements tends to change over time {[}39{]}; and it is higher between consecutive measurements {[}12,14{]}.
-A visual representation of the
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser) the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}33{]}, and they are not robust if the group sizes are unbalanced {[}40{]}.
-In other words, if the data does not present constant correlation between repeated measurements, the assumptions required for an rm-ANOVA model are not met and the use of corrections may still not provide a reasonable adjustment that makes the model valid.
+Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}38--40{]}.
+However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}41{]}; and it is higher between consecutive measurements {[}12,14{]}.
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}40,42{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}34{]}, and they are not robust if the group sizes are unbalanced {[}43{]}.
+Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
 
@@
 Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}41{]}.
-Dropout from patients, attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}36{]}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}42{]}.
-If the data are MCAR, this means that the missigness is completely unrelated to the collected information {[}43{]}, and in the case of MNAR the missing values are dependent on their value.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}44{]}.
+Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
+Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}38{]}.
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}45{]}.
+If the data are MCAR, this means that the missigness is completely unrelated to the collected information {[}46{]}, and in the case of MNAR the missing values are dependent on their value.
 An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}44{]}.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}47{]}.
 
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}35{]}.
-The pattern of missing observations can be considered MAR if the missing observations are not related any of the other variables measured in the study {[}39{]}.
+In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}37{]}.
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
 
-This section has presented the assumptions of rm-ANOVA to analyze longitudinal information and its differences when compared to LMEMs regarding to missing data and the modeling of the covariance matrix.
+This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding missing data and modeling the covariance matrix.
 Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high variability, LMEMs fail to capture the non-linear trend of the data.
-To analyze such type of data, we present generalized additive models (GAMs) as an alternative in the following section.
+However, when the data presents high a non-linear behavior, LMEMs fail to capture the non-linear trend of the data.
+To analyze longitudinal non-linear data, we present generalized additive models (GAMs) as an alternative in the following section.
 
@@
 
-A GAM is a special case of the Generalized Linear Model (GLM), a framework that allows for response distributions that do not follow a normal distribution {[}31,45{]}.
-Following the notation by Simpson {[}28{]} A GAM model can be represented as:
+A GAM is regression-based method to estimate trends{[}28{]}, can be considered a broader class of which the Generalized Linear Model (GLM) is a special case.
+a framework that allows for response distributions that do not follow a normal distribution {[}31,48{]}.
+A GAM model can be represented as:
 
 \begin{equation}
-  y_{ijt}=\beta_0+f(x_t\mid \beta)+\varepsilon_{ijt}
+  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
   \label{eq:GAM}
@@
 
-Where \(y_{ijt}\) is the response at time \(t\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the function \(f(x_t\mid \beta)\) and \(\varepsilon_{ijt}\) represents the residuals.
+Where \(y_{ijt}\) is the response at time \(t\) of subject \$i \$in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the function \(f(x_t\mid \beta_j)\) and \(\varepsilon_{ijt}\) represents the residuals.
 
@@
 This approach is more advantageous as it does not restrict the model to a linear relationship.
-One possible function for \(f(x_t\mid \beta)\) that allows for non-linear responses is a polynomial, but a major limitation is that they create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}30{]}.In particular, as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta)\) goes to \(\pm \infty\) which is almost always unrealistic.
+One possible function for \(f(x_t\mid \beta_j)\) that allows for non-linear responses is a polynomial, but a major limitation is that they create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}30{]}.
+In particular, as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic, and causes bias at the endpoints of the time period.
 
-The model specification for GAMs requires that the \emph{smooth functions} are represented in a parametric way that fit within the GLM framework, and this step is achieved by using \emph{basis functions} to represent them.
-A \emph{basis} is a set of functions that capture the space where the smooths that approximate \(f(x_t \mid \beta)\) exist {[}28{]}.
-For the linear model in \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}46{]}, which in the case of \eqref{eq:linear-model} is restricted to a linear behavior.
-In the case of \eqref{eq:GAM}, the basis function is \(f(x_t\mid \beta)\), which means that the model allows relationships beyond linear for the covariates.
+The model specification for GAMs requires that the \emph{smooth functions} are represented in a parametric way that can be fit within the GLM framework, and this step is achieved by using \emph{basis functions} to represent them.
+A \emph{basis} is a set of functions that capture the space where the smooths that approximate \(f(x_t \mid \beta_j)\) exist {[}28{]}.
+For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}49{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+In the case of Equation \eqref{eq:GAM}, the basis function is \(f(x_t\mid \beta_j)\), which means that the model allows relationships beyond linear for the covariates.
 
-In GAMs the \emph{basis functions} are represented over evenly spaced ranges of the covariates known as \emph{knots}.
-A commonly used \emph{basis function} is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations{[}28,31{]}.
-Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness {[}47{]}.
+In GAMs the \emph{basis functions} are represented over the range of the covariates and are typically evenly spaced in \emph{knots}.
+A commonly used \emph{basis function} is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations{[}28,31{]}.
+Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness {[}50{]}.
 Therefore, GAMs overcome the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear, such as Figure \ref{fig:quadratic-response}.
-Regarding longitudinal data, Pedersen et al {[}27{]} demonstrated the capabilities of GAMs in this area using ecological data.
 
@@
 
+\leavevmode\hypertarget{ref-r}{}%
+{[}32{]} R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.
+
 \leavevmode\hypertarget{ref-wood2016}{}%
-{[}32{]} S.N. Wood, N., Pya, B. Säfken, Smoothing parameter and model selection for general smooth models (with discussion), Journal of the American Statistical Association. 111 (2016) 1548--1575.
+{[}33{]} S.N. Wood, N., Pya, B. Säfken, Smoothing parameter and model selection for general smooth models (with discussion), Journal of the American Statistical Association. 111 (2016) 1548--1575.
 
 \leavevmode\hypertarget{ref-haverkamp2017}{}%
-{[}33{]} N.
+{[}34{]} N.
 Haverkamp, A.
@@
 
+\leavevmode\hypertarget{ref-nlme}{}%
+{[}35{]} J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, nlme: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.
+
 \leavevmode\hypertarget{ref-wolfinger1996}{}%
-{[}34{]} R.D. Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics. (1996) 205--230.
+{[}36{]} R.D. Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics. (1996) 205--230.
 
 \leavevmode\hypertarget{ref-west2014}{}%
-{[}35{]} B.T. West, K.B. Welch, A.T. Galecki, Linear mixed models: A practical guide using statistical software, CRC Press, 2014.
+{[}37{]} B.T. West, K.B. Welch, A.T. Galecki, Linear mixed models: A practical guide using statistical software, CRC Press, 2014.
 
 \leavevmode\hypertarget{ref-weiss2005}{}%
-{[}36{]} R.E. Weiss, Modeling longitudinal data, Springer Science \& Business Media, 2005.
+{[}38{]} R.E. Weiss, Modeling longitudinal data, Springer Science \& Business Media, 2005.
 
 \leavevmode\hypertarget{ref-geisser1958}{}%
-{[}37{]} S. Geisser, S.W. Greenhouse, others, An extension of box's results on the use of the \(F\) distribution in multivariate analysis, The Annals of Mathematical Statistics. 29 (1958) 885--891.
+{[}39{]} S. Geisser, S.W. Greenhouse, others, An extension of box's results on the use of the \(F\) distribution in multivariate analysis, The Annals of Mathematical Statistics. 29 (1958) 885--891.
 
 \leavevmode\hypertarget{ref-huynh1976}{}%
-{[}38{]} H. Huynh, L.S. Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics. 1 (1976) 69--82.
+{[}40{]} H. Huynh, L.S. Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics. 1 (1976) 69--82.
 
 \leavevmode\hypertarget{ref-maxwell2017}{}%
-{[}39{]} S.E. Maxwell, H.D. Delaney, K. Kelley, Designing experiments and analyzing data: A model comparison perspective, Routledge, 2017.
+{[}41{]} S.E. Maxwell, H.D. Delaney, K. Kelley, Designing experiments and analyzing data: A model comparison perspective, Routledge, 2017.
+
+\leavevmode\hypertarget{ref-greenhouse1959}{}%
+{[}42{]} S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112.
 
 \leavevmode\hypertarget{ref-keselman2001}{}%
-{[}40{]} H. Keselman, J. Algina, R.K. Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematical and Statistical Psychology. 54 (2001) 1--20.
+{[}43{]} H. Keselman, J. Algina, R.K. Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematical and Statistical Psychology. 54 (2001) 1--20.
 
 \leavevmode\hypertarget{ref-molenberghs2004}{}%
-{[}41{]} G. Molenberghs, H. Thijs, I. Jansen, C. Beunckens, M.G. Kenward, C. Mallinckrodt, R.J. Carroll, Analyzing incomplete longitudinal clinical trial data, Biostatistics. 5 (2004) 445--464.
+{[}44{]} G. Molenberghs, H. Thijs, I. Jansen, C. Beunckens, M.G. Kenward, C. Mallinckrodt, R.J. Carroll, Analyzing incomplete longitudinal clinical trial data, Biostatistics. 5 (2004) 445--464.
 
 \leavevmode\hypertarget{ref-scheffer2002}{}%
-{[}42{]} J. Scheffer, Dealing with missing data, (2002).
+{[}45{]} J. Scheffer, Dealing with missing data, (2002).
 
 \leavevmode\hypertarget{ref-potthoff2006}{}%
-{[}43{]} R.F.
+{[}46{]} R.F.
 Potthoff, G.E.
@@
 \leavevmode\hypertarget{ref-ma2012}{}%
-{[}44{]} Y.
+{[}47{]} Y.
 Ma, M.
@@
 \leavevmode\hypertarget{ref-hastie1987}{}%
-{[}45{]} T. Hastie, R. Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association. 82 (1987) 371--386.
+{[}48{]} T. Hastie, R. Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association. 82 (1987) 371--386.
 
 \leavevmode\hypertarget{ref-hefley2017}{}%
-{[}46{]} T.J.
+{[}49{]} T.J.
 Hefley, K.M.
@@
 \leavevmode\hypertarget{ref-wegman1983}{}%
-{[}47{]} E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365.
+{[}50{]} E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365.
 

### 2021-02-25 Manuscripts/Manuscript_AM_v3.Rmd
Updated background section, need to work on the following parts

--- 2021-02-09 Manuscripts/Manuscript_AM_v2.Rmd
@@
 output:
-  bookdown::html_document2: default
-  bookdown::word_document2: default
+  bookdown::word_document2:
+    fig_caption: yes
+    keep_md: yes
   bookdown::pdf_document2:
@@
     extra_dependencies: ["subfig","breqn"]
+  bookdown::html_document2: default
   link-citations: yes
@@
 
-A longitudinal study is designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of the treatment effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Medical research commonly uses longitudinal studies to analyze the evolution of patients before, during and after treatment.
+A longitudinal study is designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
+Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect in patients before, during and after therapy.
 For example, Sio et.
 al.
-analyzed the evolution of dermatitis symptoms in breast cancer patients after radiotherapy (RT), by taking weekly measurements over the course of two months [@sio2016], and Kamstra et al measured mouth opening at regular monthly intervals after RT in neck cancer patients [@kamstra2015].
+analyzed the evolution of dermatitis symptoms in breast cancer patients following radiotherapy (RT), by taking weekly measurements over the course of two months [@sio2016], and Kamstra et al measured mouth opening at regular monthly intervals following RT in neck cancer patients [@kamstra2015].
 Other studies using longitudinal designs in biomedical research have tumor  analyzed tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
@@
 The frequentist paradigm regards probability as a limiting frequency[@wagenmakers2008] by assuming a null hypothesis under a statistical model that is often assumed to be an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model typically makes two key assumptions regarding longitudinal data: constant correlation across same-subject measurements, and observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].
+The rm-ANOVA model typically makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].
 
-The assumption of constant correlation (often known as the _compoun symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].
+Indeed, an expected linear behavior of the response through time is a requisite in rm-ANOVA [@pinheiro2006].
+This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
+In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
+These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
+Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model can only accommodate linear trends.
+
+Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric relies on how adequate the model fits the data.
+A _post hoc_ analysis is based on multiple repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
+The _p-value_ requires that the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019].
+For example, if a model that is only able to explain linear responses is fitted to data that follows a quadratic trend, the resulting _p-values_ and parameter estimates  are invalid[@wang2019].
+
+Additionally, the _p-value_ is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010;@halsey2015], consequently biasing the conclusions of the study.Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019].
+Therefore, the _post hoc_ comparisons commonly used in rm-ANOVA impose a tradeoff between Type I, II and III errors.
+
+The assumption of constant correlation (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].
+
 Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial voluntary withdrawal from one or multiple patients can occur, whereas attrition in preclinical animal studies due to injury or weight loss is possible.
+For example, in a clinical trial patients may voluntarily withdraw, whereas attrition  due to injury or weight loss in preclinical animal studies is a possible and indeed common occurrence.
 It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
 In each of these missing data scenarios,the _complete observations_ assumption of  classical rm-ANOVA is violated.
-
 When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis[@gueorguieva2004].
@@
 At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
-Additionally, rm-ANOVA uses a _post hoc_ analysis to assess differences between the measured response in different groups.
-A _post hoc_ analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance.
-Because the _p-value_ is highly variable, multiple comparisons can inflate the false positivity rate (Type I error) [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
-Although corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), they in turn reduce statistical power ($\alpha$)[@nakagawa2004], and lead to increased Type II error (fail to reject the null hypothesis although it is false [@gelman2012;@albers2019].
-Therefore, the _post hoc_ comparisons commonly used in rm-ANOVA impose a tradeoff between Type I and II errors.
 
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-The recognition on the limitations of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor data [@skala2010;@vishwanath2009].
+The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor data [@skala2010;@vishwanath2009].
 Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
 When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-On the other hand, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
+On the other hand, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013]; but more importantly, they also expect a linear relationship between the response and time [@pinheiro2006], making them also limited to analyze non-linear data.
 
-One final assumption that is not initially evident for both rm-ANOVA and LMEMs models is that the mean response is expected to change linearly through time[@pinheiro2006].
-The linearity assumption in both rm-ANOVA and LMEMs  implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, a particular case of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit LMEMs or rm-ANOVA model inconsistent with the observed variation.
-Therefore, when LMEMs and rm-ANOVA are used to draw inference of such highly-variable data the estimates are inevitably biased, because non-linear data is provided to a model which can only accomodate linear trends.
-
-Additionally,  although it is possible that a _post hoc_ analysis is able to find “significant” _p-values_( _p_<0.05) the validity of such metric relies on how adequate the model fits the data.
-In other words, the _p-value_ requires that the model and the data have good agreement and if that is not the case, a "Type 3" error (known as "model misspecification") occurs[@dennis2019].
-For example, a model that is only able to explain linear responses but is fitted to data that has a quadratic behavior results in _p-values_ and parameter estimates that are invalid[@wang2019].
-
-As  the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed behavior) to determine the trend in the data, which enables inference that is appropriate.
+As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropiate inference.
 
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend of the model, can easily accommodate missing observations, and can model non-constant correlation between repeated measurements [@wood2017].
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data.
+Although not frequently used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017] and 3) can easily accommodate missing observations.
+Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
 
-The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs.
+The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
 In particular, $\textsf{R}$[@r]has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
@@
 
-This work provides biomedical researchers with a clear understanding of the theory and the practical implementation of GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding missing observations, assumption of linearity of response and constant correlation structures is explained in detail.
+This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
+First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations is explained in detail.
 Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, simulated data that reproduces patterns in previously reported studies [@vishwanath2009] is used to illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
+Third, using simulated data that reproduces patterns in previously reported studies [@vishwanath2009] we illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
 The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
@@
 
-In summary, the exploration of modern statistical techniques to analyze longitudinal data may allow biomedical researchers to consider the use of GAMs instead of rm-ANOVA or LMEMs when the data does not follow a linear trend, and will also help to improve the standards for reproducibility  in biomedical research.
+In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models by improving the standards for reproducibility  in biomedical research.
 
@@
 Thus, the response per subject corresponds to the mean response per group plus the error (compound symmetry or independent).
-To this data, an rm-ANOVA model identical as Equation \@ref:eq(linear-model) is fitted.
+To this data, an rm-ANOVA model identical as Equation  \@ref(eq:linear-model) is fitted.
 The resulting simulated data, simulated errors, individual response and fitted parameters from the model are plotted.
@@
     # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +
+    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
     scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
@@
     ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='response')+
+    geom_line(show.legend=FALSE) +labs(y='errors')+
      theme_classic()+
@@
 
-```{r, linear-models, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+```{r, linear-models, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
 A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
@@
 D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data."}
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
 # linear response, correlated errors (subject effect)
@@
 
-```{r, quadratic-response, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
+```{r, quadratic-response, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
 A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
@@
 
-In this case, a comparison of the predictions of the rm-ANOVA model to the the simulated data and indicates that the model is not capturing the changes within each group in a good way.
-This is expected as the rm-ANOVA model is only able to accommodate linear changes, which in this case fail to capture the changes over the timeline.
-Specifically, note that in Figure \@ref(fig:quadratic-response) panels D and H the model is not capturing the increase in the response through time points 2 and 4.
-If such incrase may indicate a physiologically relevant change, the model will not be able to capture it, potentially limiting inferences from the data.
+In this case, in Figure \@ref(fig:quadratic-response), panel A shows that both groups start with the same mean value at the first time point.
+This is intentional, as this is the expected behavior in a fold change analysis of a physiological quantity (such as tumor response to chemotherapy).
+Therefore, the idea behind \@ref(fig:quadratic-response) is to mimic a non-linear behavior from two groups that follow the inverse trajectory.
+A comparison of the predictions of the rm-ANOVA model to the the simulated data (\@ref(fig:quadratic-response), panels D and H)  indicates that the model is not capturing the changes within each group in a good way.
+Specifically, note that in Figure \@ref(fig:quadratic-response) the predictions of the rm-ANOVA model (panels D and H) show that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
+If such change has important physiological implications,the rm-ANOVA model omits it from the predictions, potentially limiting valuable inferences from the data.
 
@@
 
-In a longitudinal study there is an expected _variance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
+In a longitudinal study there is an expected _covariance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
 The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
@@
 Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
-However, it has been shown that this condition is frequently unjustified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
-A visual representation of the
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser) the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
-In other words, if the data does not present constant correlation between repeated measurements, the assumptions required for an rm-ANOVA model are not met and the use of corrections may still not provide a reasonable adjustment that makes the model valid.
+However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)[@huynh1976;@greenhouse1959] the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
+Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
 
@@
 In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004].
-Dropout from patients, attrition or injury in animals are among the reasons for missing observations.
+Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
 Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
 In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002].
-If the data are MCAR, this means that the missigness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
+If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
 An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
@@
 In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014].
-The pattern of missing observations can be considered MAR if the missing observations are not related any of the other variables measured in the study [@maxwell2017].
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.
 
-This section has presented the assumptions of rm-ANOVA to analyze longitudinal information and  its differences when compared to LMEMs  regarding to missing data and the modeling of the covariance matrix.
+This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding missing data and modeling the covariance matrix.
 Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high variability, LMEMs fail to capture the non-linear trend of the data.
-To analyze such type of data, we present generalized additive models (GAMs) as an alternative in the following section.
+However, when the data presents high a non-linear behavior, LMEMs fail to capture the non-linear trend of the data.
+To analyze longitudinal non-linear data, we present generalized additive models (GAMs) as an alternative in the following section.
 
@@
 
-A GAM is a special case of the Generalized Linear Model (GLM), a framework that allows for response distributions that do not follow a normal distribution [@wood2017;@hastie1987].
-Following the notation by Simpson [@simpson2018] A GAM model can be represented as:
+A GAM is regression-based  method to estimate trends[@simpson2018], can be considered a broader class of which the Generalized Linear Model (GLM) is a special case.
+a framework that allows for response distributions that do not follow a normal distribution [@wood2017;@hastie1987].
+A GAM model can be represented as:
 
 \begin{equation}
-  y_{ijt}=\beta_0+f(x_t\mid \beta)+\varepsilon_{ijt}
+  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
   (\#eq:GAM)
@@
 
-Where $y_{ijt}$ is the response at time $t$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta)$ and $\varepsilon_{ijt}$ represents the residuals.
+Where $y_{ijt}$ is the response at time $t$ of subject  $i $in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ and $\varepsilon_{ijt}$ represents the residuals.
 
@@
 This approach is more advantageous as it does not restrict the model to a linear relationship.
-One possible function for $f(x_t\mid \beta)$ that allows for non-linear responses is a polynomial, but a major limitation is that they create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].In particular, as $t$ goes to $\pm \infty$, $f(x_t \mid \beta)$ goes to $\pm \infty$ which is almost always unrealistic.
+One possible function for $f(x_t\mid \beta_j)$ that allows for non-linear responses is a polynomial, but a major limitation is that they create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
+In particular, as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.
 
-The model specification for GAMs requires that the _smooth functions_ are represented in a parametric way that fit within the GLM framework, and this step is achieved by using _basis functions_ to represent them.
-A _basis_ is a set of functions that capture the space where the smooths that approximate $f(x_t \mid \beta)$ exist [@simpson2018].
-For the linear model in \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of \@ref(eq:linear-model) is restricted to a linear behavior.
-In the case of \@ref(eq:GAM), the basis function is $f(x_t\mid \beta)$, which means that the model allows relationships beyond linear for the covariates.
+The model specification for GAMs requires that the _smooth functions_ are represented in a parametric way that can be fit within the GLM framework, and this step is achieved by using _basis functions_ to represent them.
+A _basis_ is a set of functions that capture the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
+For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
+In the case of Equation \@ref(eq:GAM), the basis function is $f(x_t\mid \beta_j)$, which means that the model allows relationships beyond linear for the covariates.
 
-In GAMs the _basis functions_  are represented over  evenly spaced ranges of the covariates known as _knots_.
-A commonly used _basis function_ is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations[@wood2017;@simpson2018].
+In GAMs the _basis functions_  are represented over  the range of the covariates and are typically evenly spaced in _knots_.
+A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations[@wood2017;@simpson2018].
 Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness [@wegman1983].
 Therefore, GAMs overcome the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear, such as Figure \@ref(fig:quadratic-response).
-Regarding longitudinal data, Pedersen et al [@pedersen2019] demonstrated the capabilities of GAMs in this area using ecological data.
 

### 2021-03-01 Manuscripts/Manuscript_AM_v3.Rmd
Updated manuscript

@@
 
-In this model $y_{ijt}$ is the response by subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-The _random effect_ $\mu_{ij}$, accounts for variability in each subject within each group by allowing different values for each subject.
-Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ or _random_ effects.
+In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
+The _random effect_ $\mu_{ij}$ is assumed to be independently normally distributed with mean zero and variance $\sigma^2_\mu$, which can be expressed as $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
+This _random effect_ accounts for variability in each subject within each group by allowing different values for each subject.
+Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ or _random_ effects, and are assumed to be $\sim N(0,\sigma^2)$ .
 In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
@@
 "chemotherapy").
-Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $j=0$ representing the first treatment group (Group A) and $j=1$ representing the second treatment group (Group B).
+Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B).
 The linear models then can be expressed as
@@
 \beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_1 \times time_{t} + \beta_2 \times treatment_{j} +\beta_3 \times time_{t} \times treatment_{j}+\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\beta_0 + \beta_1 \times time_{t} + \beta_2 \times +\beta_3 \times time_{t} \times +\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
 \end{cases}
@@
 To further simplify the expression, substitute $\widetilde{\mu}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B.
-This substitution allows for a different intercept and slope for Group B.
-The model is written as
+This substitution allows for a different intercept and slope for Groups A and B.
+The model is then written as
 
@@
 Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, When it is stated that an rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear behavior, the rm-ANOVa model will approximate this behavior with non-parallel lines.
+In other words, the rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data.
+If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
 Because the construction of the model is essentially the same in LMEMs, the linearity assumption holds for these models, although it is possible to fit models with different slopes at the subject-level[@pinheiro2006].
 
-When the data does not follow a linear trend, the fit that an rm-ANOVA or LMEM produces may not accurately represent the trends in the data.
-To demonstrate this inaccurate representation, longitudinal data where a normally distributed response of two groups of 10 subjects each is simulated, and a rm-ANOVA model  (Equation \@ref(eq:linear-model)) is fitted to the data using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix).
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes (negative slope for Group 1 and positive slope for Group 2 ).
-In the second case, a quadratic function is used for the mean response per group (concave down for Group 1 and concave up for Group 2 ).
-From the mean response per group (linear or quadratic), individual responses for each subject within each group are simulated using covariance matrices for with compound symmetry (constant variance across time) or independent errors (errors that are generated from a a normal distribution but are not constant over time).
-Thus, the response per subject corresponds to the mean response per group plus the error (compound symmetry or independent).
+## Covariance in rm-ANOVA and LMEMs
+
+In a longitudinal study there is an expected _covariance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
+The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
+
+In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
+Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
+However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)[@huynh1976;@greenhouse1959] the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
+Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
+
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others [@pinheiro2006].
+Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself.
+Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
+
+## Missing observations
+
+Missing observations are an issue that arises frequently in longitudinal studies.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004].
+Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
+Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002].
+If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
+An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
+
+In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014].
+For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.
+
+This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding missing data and modeling the covariance matrix.
+Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
+However, when the data presents high a non-linear behavior, LMEMs fail to capture the non-linear trend of the data.
+To analyze longitudinal non-linear data, we present generalized additive models (GAMs) as an alternative in the following section.
+
+#Simulation experiment on linearity and covariance structures
+
+To demonstrate the limitations of rm-ANOVA for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
+To each group, an rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix).
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group (Figure \@ref(fig:linear-models, A)) .In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:quadratic-response),A).
+From the mean response per group (linear or quadratic), individual responses for each subject within each group are simulated using covariance matrices with compound symmetry (constant variance across time) or independent errors (errors that are generated from a a normal distribution that are not constant over time).
+Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (either compound symmetry or independent, which appear in panels A and E, respectively in both Figure \@ref(fig:linear-models) and Figure \@ref(fig:quadratic-response)) .
 To this data, an rm-ANOVA model identical as Equation  \@ref(eq:linear-model) is fitted.
@@
 
-In Figure \@ref(fig:linear-models)  panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a trend.
-It is clear from \@ref(fig:linear-models) fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the "truth" of the simulated data.In particular,
+In Figure \@ref(fig:linear-models)  panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon $ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
+In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
+It is clear from Figure \@ref(fig:linear-models) that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the "truth" of the simulated data.
 
@@
 If such change has important physiological implications,the rm-ANOVA model omits it from the predictions, potentially limiting valuable inferences from the data.
-
-## Covariance in rm-ANOVA and LMEMs
-
-In a longitudinal study there is an expected _covariance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
-The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
-
-In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
-Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)[@huynh1976;@greenhouse1959] the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others [@pinheiro2006].
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself.
-Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
-
-## Missing observations
-
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004].
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002].
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
-
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014].
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.
-
-This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding missing data and modeling the covariance matrix.
-Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high a non-linear behavior, LMEMs fail to capture the non-linear trend of the data.
-To analyze longitudinal non-linear data, we present generalized additive models (GAMs) as an alternative in the following section.
 

### 2021-03-02 Manuscripts/Manuscript_AM_v3.Rmd
Updated and reorganized manuscript, need to work on basis function plot

@@
 
-This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding missing data and modeling the covariance matrix.
+This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and  missing data.
 Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high a non-linear behavior, LMEMs fail to capture the non-linear trend of the data.
-To analyze longitudinal non-linear data, we present generalized additive models (GAMs) as an alternative in the following section.
+However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data.
+To better convey the issues of linearity and correlation  in linear models fitted to non-linear data, simulation is used in the next section.
 
-#Simulation experiment on linearity and covariance structures
+## Simulation experiment on linearity and covariance structures
 
 To demonstrate the limitations of rm-ANOVA for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-To each group, an rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix).
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group (Figure \@ref(fig:linear-models, A)) .In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:quadratic-response),A).
-From the mean response per group (linear or quadratic), individual responses for each subject within each group are simulated using covariance matrices with compound symmetry (constant variance across time) or independent errors (errors that are generated from a a normal distribution that are not constant over time).
+An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix).
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:linear-models, A)).
+In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:quadratic-response),A).
+
+The rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the response indicates a difference in the effect of each treatment.
+From the mean response per group (linear or quadratic), individual responses for each subject within each group are simulated using covariance matrices with compound symmetry (constant variance across time) or independent errors (errors generated from a a normal distribution that are not constant over time).
 Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (either compound symmetry or independent, which appear in panels A and E, respectively in both Figure \@ref(fig:linear-models) and Figure \@ref(fig:quadratic-response)) .
-To this data, an rm-ANOVA model identical as Equation  \@ref(eq:linear-model) is fitted.
 The resulting simulated data, simulated errors, individual response and fitted parameters from the model are plotted.
@@
 
-In Figure \@ref(fig:linear-models)  panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon $ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
+In Figure \@ref(fig:linear-models)  panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
 In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-It is clear from Figure \@ref(fig:linear-models) that the fit produced by the rm-ANOVA model is good as the predictions for the mean response are reasonably close to the "truth" of the simulated data.
+It is clear from Figure \@ref(fig:linear-models) that the fit produced by the rm-ANOVA model is good, as the predictions for the mean response are reasonably close to the "truth" of the simulated data.
 
@@
 Here, the mean response per group was simulated using a quadratic function and errors, individual responses and the rm-ANOVA model were produced in the same manner as \@ref(fig:linear-models) .
-The figure shows that changes in each group occur through the timeline, and that the mean value is the same as the initial value by the fifth time point.
+The figure shows that the response is changing in each group through the timeline, and that the mean value is the same as the initial value by the fifth time point.
 Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panels D and H in Figure \@ref(fig:quadratic-response).
@@
 
-In this case, in Figure \@ref(fig:quadratic-response), panel A shows that both groups start with the same mean value at the first time point.
-This is intentional, as this is the expected behavior in a fold change analysis of a physiological quantity (such as tumor response to chemotherapy).
-Therefore, the idea behind \@ref(fig:quadratic-response) is to mimic a non-linear behavior from two groups that follow the inverse trajectory.
-A comparison of the predictions of the rm-ANOVA model to the the simulated data (\@ref(fig:quadratic-response), panels D and H)  indicates that the model is not capturing the changes within each group in a good way.
-Specifically, note that in Figure \@ref(fig:quadratic-response) the predictions of the rm-ANOVA model (panels D and H) show that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
-If such change has important physiological implications,the rm-ANOVA model omits it from the predictions, potentially limiting valuable inferences from the data.
+In this case, Figure \@ref(fig:quadratic-response), panel A shows that both groups start with the same mean value at the first time point.
+Again, this is intentional in order to simulate the temporal behavior of some physiological quantity.
+Therefore, the idea behind \@ref(fig:quadratic-response) is to mimic a non-linear behavior from two groups that follow an inverse trajectory.
+A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data (\@ref(fig:quadratic-response), panels D and H)  indicates that the model is not capturing the changes within each group in a good way.
+Specifically, note that in Figure \@ref(fig:quadratic-response) the fitted mean response of the rm-ANOVA model (panels D and H) show that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model and moreover, the model is not being able to capture the initial value (which is the same in each group), instead fitting non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels A and D, and E and H).
+If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response, potentially limiting valuable inferences from the data.
+
+This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
+Although the model fitted to the simulated data was an rm_ANOVA, the main issue of an expected linear trend in the response is the same in the case of LMEMs.
+In the following section, we present generalized additive models (GAMs) as an alternative to analyze longitudinal non-linear data.
 
@@
 
-A GAM is regression-based  method to estimate trends[@simpson2018], can be considered a broader class of which the Generalized Linear Model (GLM) is a special case.
-a framework that allows for response distributions that do not follow a normal distribution [@wood2017;@hastie1987].
-A GAM model can be represented as:
+Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors[@nelder1972].
+In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case[@simpson2018;@wood2017;@hastie1987].
+A GAM model can be written as:
 
@@
 
-Where $y_{ijt}$ is the response at time $t$ of subject  $i $in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ and $\varepsilon_{ijt}$ represents the residuals.
+Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ and $\varepsilon_{ijt}$ represents the residual error.
 
-In contrast to rm-ANOVA or LMEMs, GAMs use _smooth functions_  to model the relationship between the covariates and the response.
-This approach is more advantageous as it does not restrict the model to a linear relationship.
-One possible function for $f(x_t\mid \beta_j)$ that allows for non-linear responses is a polynomial, but a major limitation is that they create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
-In particular, as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.
+In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_.
+This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
+One possible function for $f(x_t\mid \beta_j)$ that allows for non-linear responses is a polynomial, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
+In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.
 
-The model specification for GAMs requires that the _smooth functions_ are represented in a parametric way that can be fit within the GLM framework, and this step is achieved by using _basis functions_ to represent them.
-A _basis_ is a set of functions that capture the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
+The smooth functional relationship between the covariates and the response in GAMs is specified   using a parametric relationship that can be fit within the GLM framework, by using _basis functions_ expansions of the covariates and by estimating random coefficients for these basis functions and this step is achieved by using _basis functions_ to represent them.
+A _basis_ is a set of functions that spans the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
 For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
@@
 
-In GAMs the _basis functions_  are represented over  the range of the covariates and are typically evenly spaced in _knots_.
+In GAMs the _basis functions_ are typically evenly spaced in _knots_ over the range of the covariates.
 A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations[@wood2017;@simpson2018].
 Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness [@wegman1983].
-Therefore, GAMs overcome the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear, such as Figure \@ref(fig:quadratic-response).
+Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
 

### 2021-03-03 Manuscripts/Manuscript_AM_v3.Rmd
Working on figure of basis functions, need to polish it

@@
 
+```{r,basis-functions}
+
+n_time = 6
+ x <- seq(1,6, length.out = n_time)
+ mu <- matrix(0, length(x), 2)
+ mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
+ mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
+ y <- array(0, dim = c(length(x), 2, 10))
+ errors <- array(0, dim = c(length(x), 2, 10))
+ for (i in 1:2) {     # number of treatments
+     for (j in 1:10) {  # number of subjects
+         # compound symmetry errors
+         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
+         y[, i, j] <- mu[, i] + errors[, i, j]
+     }
+ }
+  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
+ dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
+ dimnames(mu) <- list(time = x, treatment = 1:2)
+ dat <- as.data.frame.table(y, responseName = "y")
+ dat_errors <- as.data.frame.table(errors, responseName = "errors")
+ dat_mu <- as.data.frame.table(mu, responseName = "mu")
+ dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
+ dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
+ dat$time <- as.numeric(as.character(dat$time))
+ dat <- dat %>%
+     mutate(subject = factor(paste(subject, treatment, sep = "-")))
+
+  dat<-subset(dat,treatment==1)
+  dat<-dat[,c('y','time')]
+gm<-gam(y~s(time,k=5),data=dat)
+model_matrix<-predict(gm,type='lpmatrix')
+
+x_new <- as.data.frame(seq(0, max(dat$time), length.out = 100))
+y_pred <- predict(gm, data.frame(x = x_new))
+
+matplot(dat$time, model_matrix[,-1], type = "l", lty = 2, add = T)
+
+```
+
 In GAMs the _basis functions_ are typically evenly spaced in _knots_ over the range of the covariates.

### 2021-03-03 Manuscripts/Manuscript_AM_v3.Rmd
Working on plot

--- 2021-03-02 Manuscripts/Manuscript_AM_v3.Rmd
@@
 
+```{r,basis-functions}
+library(mgcv)
+library(mvnfast)
+  library(nlme)
+  library(ggsci)
+n_time = 6
+ x <- seq(1,6, length.out = n_time)
+ mu <- matrix(0, length(x), 2)
+ mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
+ mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
+ y <- array(0, dim = c(length(x), 2, 10))
+ errors <- array(0, dim = c(length(x), 2, 10))
+ for (i in 1:2) {     # number of treatments
+     for (j in 1:10) {  # number of subjects
+         # compound symmetry errors
+         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
+         y[, i, j] <- mu[, i] + errors[, i, j]
+     }
+ }
+  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
+ dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
+ dimnames(mu) <- list(time = x, treatment = 1:2)
+ dat <- as.data.frame.table(y, responseName = "y")
+ dat_errors <- as.data.frame.table(errors, responseName = "errors")
+ dat_mu <- as.data.frame.table(mu, responseName = "mu")
+ dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
+ dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
+ dat$time <- as.numeric(as.character(dat$time))
+ dat <- dat %>%
+     mutate(subject = factor(paste(subject, treatment, sep = "-")))
+
+  dat<-subset(dat,treatment==1)
+  dat<-dat[,c('y','time')]
+gm<-gam(y~s(time,k=5),data=dat)
+model_matrix<-predict(gm,type='lpmatrix')
+
+x_new <- (seq(0, max(dat$time), length.out = 100))
+y_pred <- predict(gm, newdata=dat)
+
+plot(dat$y~dat$time)
+
+matplot(dat$time, model_matrix[,-1], type = "l", lty = 2, add = T)
+lines(y_pred ~ dat$time, col = "red", lwd = 2)
+
+```
+
 In GAMs the _basis functions_ are typically evenly spaced in _knots_ over the range of the covariates.

### 2021-03-05 Manuscripts/Manuscript_AM_v3.Rmd
Working on basis plot

@@
 
-plot(dat$y~dat$time)
+time<-c(1:6)
 
-matplot(dat$time, model_matrix[,-1], type = "l", lty = 2, add = T)
-lines(y_pred ~ dat$time, col = "red", lwd = 2)
+basis1<-model_matrix[1:6,2]
+basis2<-model_matrix[1:6,3]
+basis3<-model_matrix[1:6,4]
+basis4<-model_matrix[1:6,5]
+
+plot(time,basis1,type='l')
+plot(time,basis2,type='l')
+plot(time,basis3,type='l')
+plot(time,basis4,type='l')
+
+#plot(dat$y~dat$time)
+
+#matplot(dat$time, model_matrix[,-1], type = "l", lty = 2, add = T)
+
+#lines(y_pred ~ dat$time, col = "red", lwd = 2)
 

### 2021-03-07 Manuscripts/Manuscript_AM_v3.Rmd
working on plot for basis functions

--- 2021-03-03 Manuscripts/Manuscript_AM_v3.Rmd
@@
 ```{r,basis-functions}
+library(mgcv)
+library(mvnfast)
+  library(nlme)
+  library(ggsci)
+library(tidyverse)
 
@@
 gm<-gam(y~s(time,k=5),data=dat)
-model_matrix<-predict(gm,type='lpmatrix')
+model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
 
-x_new <- as.data.frame(seq(0, max(dat$time), length.out = 100))
-y_pred <- predict(gm, data.frame(x = x_new))
+x_new <- (seq(0, max(dat$time), length.out = 100))
+y_pred <- predict(gm, newdata=dat)
 
-matplot(dat$time, model_matrix[,-1], type = "l", lty = 2, add = T)
+time<-c(1:6)
+
+basis<-model_matrix[1:6,-1] #extracting basis
+basis<-basis %>% #pivoting to long format
+  pivot_longer(
+    cols=starts_with("s")
+  )%>%
+  arrange(name) #ordering
+
+#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
+ln<-length(basis1)*(length(coef(gm))-1)
+
+basis_plot<-data.frame(Basis=integer(ln),
+                       value=double(ln),
+                       time=integer(ln),
+)
+
+basis_plot$time<-rep(time) #pasting timepoints
+basis_plot$Basis<-rep(1:4,each=6) #pasting basis number values
+basis_plot$value<-basis$value #pasting basis values
+
+ggplot(basis_plot,aes(x=time,y=value))+geom_line()+facet_grid(~Basis) #plotting
+
+#plot(dat$y~dat$time)
+
+#matplot(dat$time, model_matrix[,-1], type = "l", lty = 2, add = T)
+
+#lines(y_pred ~ dat$time, col = "red", lwd = 2)
 

### 2021-03-08 Manuscripts/Manuscript_AM_v3.Rmd
Need to tweak the plot for the basis

@@
   library(nlme)
+library(patchwork)
   library(ggsci)
@@
 #length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-length(basis1)*(length(coef(gm))-1)
+ln<-6*(length(coef(gm))-1)
 
 basis_plot<-data.frame(Basis=integer(ln),
-                       value=double(ln),
+                       value_orig=double(ln),
                        time=integer(ln),
+                       cof=double(ln)
 )
@@
 basis_plot$Basis<-rep(1:4,each=6) #pasting basis number values
-basis_plot$value<-basis$value #pasting basis values
+basis_plot$value_orig<-basis$value #pasting basis values
+basis_plot$cof<-rep(coef(gm)[2:5],each=6) #pasting coefficients
+basis_plot<-basis_plot%>%
+  mutate(mod_val=value_orig*cof)
 
-ggplot(basis_plot,aes(x=time,y=value))+geom_line()+facet_grid(~Basis) #plotting
+#calculating the final spline by aggregating the basis functions
 
+spl<-basis_plot%>% group_by(time)%>%summarize(spl=sum(mod_val))
+
+#original bases
+p1<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=factor(Basis)))+
+  geom_line()+geom_point()+
+  facet_grid(~Basis)+
+  theme_classic()+
+  scale_color_aaas()
+
+#basis multiplied by the coefficient
+p2<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=factor(Basis)))+
+  geom_line()+geom_point()+
+  facet_grid(~Basis)+
+  theme_classic()+
+  scale_color_aaas()
+
+#heatmap of the coefficient
+p3<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
+  geom_tile(aes(color='black'),size=2)+
+  geom_tile(size=2)+
+  scale_fill_gradient(low = "white", high = "red")+ 
+  geom_text(aes(label=round(cof,1)))+
+  theme_classic()
+
+p4<-ggplot(data=dat,aes(x=time,y=y))+
+  geom_point()+
+  geom_line(data=spl,aes(x=time,y=spl))+
+  scale_color_aaas()+
+  theme_classic()
+
+p1+p2+p3+p4+plot_annotation(tag_levels='A')
+#plotting
 #plot(dat$y~dat$time)

### 2021-03-08 Manuscripts/Manuscript_AM_v3.Rmd
NEed to improve labels

@@
 #original bases
-p1<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=factor(Basis)))+
+p1<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=as.factor(Basis)))+
   geom_line()+geom_point()+
+  labs(y='Original basis function',fill="Basis Number")+
   facet_grid(~Basis)+
@@
 #basis multiplied by the coefficient
-p2<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=factor(Basis)))+
-  geom_line()+geom_point()+
+p2<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
+  geom_line(show.legend = FALSE)+geom_point(show.legend = FALSE)+
+  labs(y='Modified basis function')+
   facet_grid(~Basis)+
@@
 p3<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
-  geom_tile(aes(color='black'),size=2)+
+  geom_tile(aes(color='black'),size=2,show.legend = FALSE)+
   geom_tile(size=2)+
@@
 
-p1+p2+p3+p4+plot_annotation(tag_levels='A')
+p1+p3+p2+p4+plot_annotation(tag_levels='A')
 #plotting

### 2021-03-08 Manuscripts/Manuscript_AM_v3.Rmd
Work on basis plot

@@
   scale_fill_gradient(low = "white", high = "red")+ 
-  geom_text(aes(label=round(cof,1)))+
+  geom_text(aes(label=round(cof,2)))+
   theme_classic()

### 2021-03-08 Manuscripts/Manuscript_AM_v3.Rmd
Made graph for basis functions, need to change order of last paragraphs to fit the location of the plot

--- 2021-03-08 Manuscripts/Manuscript_AM_v3.Rmd
@@
 output:
-  bookdown::word_document2:
-    fig_caption: yes
-    keep_md: yes
+  bookdown::html_document2: default
   bookdown::pdf_document2:
@@
     extra_dependencies: ["subfig","breqn"]
-  bookdown::html_document2: default
+  bookdown::word_document2:
+    fig_caption: yes
+    keep_md: yes
   link-citations: yes
@@
 
-```{r,basis-functions}
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure @\ref(fig:quadratic-response).
+The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits the trend trough the timeline.
+Because there are six timepoints for this group, the use of five equally spaced knots ensures that the model will not have computational issues.
+Asking the model to use five knots to construct the smooth term means that it will have _four_ effective basis functions (as one will be used for the intercept).
+The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots.
+In Panel A of Figure \@ref(fig:basis-functions-plot), the four basis functions are shown.
+Each of the basis functions is composed of six different points (because there are six points on the timeline).
+To control the wigliness of the fit, each of the basis functions of Panel A is penalized, or multipled by a coefficient according to the penalty matrix of Panel B.
+In other words, the six points of each basis will be multiplied by the corresponding coefficient, thereby increasing or decreasing the original basis functions of Panel B.
+In Figure \@ref(fig:basis-functions-plot), Panel C shows the resulting penalized basis functions.
+Note that the penalizetion for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has increased its value roughly 2.5 times.
+Finally, the penalized basis functions are added at each timepoint to produce the final value of the smooth term.
+The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
+
+```{r,basis-functions-plot,fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Basis functions for a single smoother for time with five knots.
+A: Basis functions for a single smoother for _time_ for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
+B: Penalty matrix for the basis functions.
+Each basis function is penalized by a coefficient which can be positive or negative.
+The coefficient determines the overall effect of each basis in the final smoother.
+C: Penalized basis functions.
+Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
+D.
+Smoother for _time_ and original datapoints.
+The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.**"}
+
 library(mgcv)
-library(mvnfast)
-  library(nlme)
-library(patchwork)
-  library(ggsci)
-library(tidyverse)
 
@@
 #original bases
-p1<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=factor(Basis)))+
-  geom_line()+geom_point()+
-  facet_grid(~Basis)+
+sz<-1
+p11<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=as.factor(Basis)))+
+  geom_line(size=sz,show.legend=FALSE)+
+  geom_point(size=sz+1,show.legend = FALSE)+
+  labs(y='Basis functions')+
+  facet_wrap(~Basis)+
   theme_classic()+
@@
 #basis multiplied by the coefficient
-p2<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=factor(Basis)))+
-  geom_line()+geom_point()+
-  facet_grid(~Basis)+
+p12<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
+  geom_line(show.legend = FALSE,size=sz)+
+  geom_point(show.legend = FALSE,size=sz+1)+
+  labs(y='Penalized \n basis functions')+
+  scale_y_continuous(breaks=seq(-1,1,1))+
+  facet_wrap(~Basis)+
   theme_classic()+
@@
 #heatmap of the coefficient
-p3<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
-  geom_tile(aes(color='black'),size=2)+
-  geom_tile(size=2)+
+p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
+  geom_tile(aes(color='black'),size=sz+1,show.legend = FALSE)+
+  geom_tile(size=sz+1)+
   scale_fill_gradient(low = "white", high = "red")+
-  geom_text(aes(label=round(cof,1)))+
+  labs(x='Basis',y='Basis')+
+  geom_text(aes(label=round(cof,2),size=10),show.legend = FALSE)+
+  theme_classic()+
+  theme(legend.title = element_blank())
+
+#plotting simulated datapoints and spline  
+p14<-ggplot(data=dat,aes(x=time,y=y))+
+  geom_point(size=sz+1)+
+  scale_color_aaas()+
+  labs(y='Simulated \n response')+
+  geom_line(data=spl,aes(x=time,y=spl),color="#B15731",size=sz+1)+
   theme_classic()
 
-p4<-ggplot(data=dat,aes(x=time,y=y))+
-  geom_point()+
-  geom_line(data=spl,aes(x=time,y=spl))+
-  scale_color_aaas()+
-  theme_classic()
-
-p1+p2+p3+p4+plot_annotation(tag_levels='A')
-#plotting
-#plot(dat$y~dat$time)
-
-#matplot(dat$time, model_matrix[,-1], type = "l", lty = 2, add = T)
-
-#lines(y_pred ~ dat$time, col = "red", lwd = 2)
+#Combining all
+p11+p13+p12+p14+plot_annotation(tag_levels='A')&
+  theme(plot.title = element_text(size = 30, 
+                                face = "bold"),
+     text=element_text(size=30))
 

### 2021-03-08 Manuscripts/Manuscript_AM_v3.Rmd
Figure on basis function complete. Need to add sections on covariance and assessing significance

@@
 
+A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations[@wood2017;@simpson2018].
+Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness [@wegman1983].
+Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
+
 To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure @\ref(fig:quadratic-response).
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits the trend trough the timeline.
-Because there are six timepoints for this group, the use of five equally spaced knots ensures that the model will not have computational issues.
-Asking the model to use five knots to construct the smooth term means that it will have _four_ effective basis functions (as one will be used for the intercept).
+The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits the trend of the group trough the timeline.
+The timeline can be divided equally spaced _knots_, each knot being a region where a different basis function will be used.
+Because there are six timepoints for this group, five knots can be used.
+The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
 The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots.
-In Panel A of Figure \@ref(fig:basis-functions-plot), the four basis functions are shown.
+In Panel A of Figure \@ref(fig:basis-functions-plot), the four basis functions (and the intercept) are shown.
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the wigliness of the fit, each of the basis functions of Panel A is penalized, or multipled by a coefficient according to the penalty matrix of Panel B.
-In other words, the six points of each basis will be multiplied by the corresponding coefficient, thereby increasing or decreasing the original basis functions of Panel B.
+To control the wigliness of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
+
+In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
 In Figure \@ref(fig:basis-functions-plot), Panel C shows the resulting penalized basis functions.
-Note that the penalizetion for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has increased its value roughly 2.5 times.
-Finally, the penalized basis functions are added at each timepoint to produce the final value of the smooth term.
+Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
+Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
 The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
@@
 
-basis<-model_matrix[1:6,-1] #extracting basis
+basis<-model_matrix[1:6,] #extracting basis
+#basis<-model_matrix[1:6,-1] #extracting basis
+colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
 basis<-basis %>% #pivoting to long format
@@
 #length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm))-1)
+ln<-6*(length(coef(gm)))
 
@@
 basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-rep(1:4,each=6) #pasting basis number values
+basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
 basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[2:5],each=6) #pasting coefficients
+basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
 basis_plot<-basis_plot%>%
   mutate(mod_val=value_orig*cof)
+
+#creating labeller to change the labels in the basis plots
+
+basis_names<-c(
+  `1`="Intercept",
+  `2`="1",
+  `3`="2",
+  `4`="3",
+  `5`="4"
+)
 
@@
 
-#original bases
+#original basis
 sz<-1
@@
   labs(y='Basis functions')+
-  facet_wrap(~Basis)+
+  facet_wrap(~Basis,labeller = as_labeller(basis_names))+
   theme_classic()+
@@
 
-#basis multiplied by the coefficient
+#penalized basis
 p12<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
@@
   scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis)+
+  facet_wrap(~Basis,labeller=as_labeller(basis_names))+
   theme_classic()+
@@
 
-#heatmap of the coefficient
+#heatmap of the penalization coefficient
 p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
@@
   labs(x='Basis',y='Basis')+
-  geom_text(aes(label=round(cof,2),size=10),show.legend = FALSE)+
+  geom_text(aes(label=round(cof,2)),size=10,show.legend = FALSE)+
   theme_classic()+
@@
 
-#plotting simulated datapoints and spline  
+#plotting simulated datapoints and smooth term
 p14<-ggplot(data=dat,aes(x=time,y=y))+
@@
 p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(plot.title = element_text(size = 30, 
-                                face = "bold"),
+  theme(
      text=element_text(size=30))
@@
 ```
-
-In GAMs the _basis functions_ are typically evenly spaced in _knots_ over the range of the covariates.
-A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations[@wood2017;@simpson2018].
-Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness [@wegman1983].
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
 

### 2021-03-10 Manuscripts/Manuscript_AM_v3.Rmd
Need to work on new sections
Updated background need to work on rm-ANOVa model update

@@
 
-A longitudinal study is designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect in patients before, during and after therapy.
-For example, Sio et.
-al.
-analyzed the evolution of dermatitis symptoms in breast cancer patients following radiotherapy (RT), by taking weekly measurements over the course of two months [@sio2016], and Kamstra et al measured mouth opening at regular monthly intervals following RT in neck cancer patients [@kamstra2015].
-Other studies using longitudinal designs in biomedical research have tumor  analyzed tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
+Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
+Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect before, during and after therapy, and the subjects to be studied can range from animals (mice, rats, rabbits), human patients, cells, or blood samples, among many others.
+Studies using longitudinal designs in biomedical research have analyzed tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
+The frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study.
+The frequency of measurements can range from minutely in animals to analyze anesthesia effects[@greening2018], weekly in patients to study the evolution of dermatitis symptoms in breast cancer [@sio2016], to monthly to study mouth opening following RT in neck cancer patients [@kamstra2015].
 
 Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as a limiting frequency[@wagenmakers2008] by assuming a null hypothesis under a statistical model that is often assumed to be an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model typically makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].
+The frequentist paradigm regards probability as a limiting frequency [@wagenmakers2008] by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
+The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].
 
-Indeed, an expected linear behavior of the response through time is a requisite in rm-ANOVA [@pinheiro2006].
+The expected linear behavior of the response through time is a key requisite in rm-ANOVA [@pinheiro2006].
 This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
@@
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model can only accommodate linear trends.
+Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model is only able to accommodate linear trends that are far from adequately representing the biological phenomenon of interest.
 
-Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric relies on how adequate the model fits the data.
-A _post hoc_ analysis is based on multiple repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
-The _p-value_ requires that the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019].
-For example, if a model that is only able to explain linear responses is fitted to data that follows a quadratic trend, the resulting _p-values_ and parameter estimates  are invalid[@wang2019].
+A _post hoc_ analysis is the statistical test used in conjuction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
+Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
+In other words, _p-values_ are valid only if the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019].
+For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting _p-values_ and parameter estimates to be invalid [@wang2019].
 
-Additionally, the _p-value_ is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010;@halsey2015], consequently biasing the conclusions of the study.Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019].
-Therefore, the _post hoc_ comparisons commonly used in rm-ANOVA impose a tradeoff between Type I, II and III errors.
+Additionally, the _p-value_ itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019].
+Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to balance in a biomedical longitudinal study where a delicate balance between statistical power and sample size exist.
 
-The assumption of constant correlation (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].
+On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].
+Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976;@greenhouse1959], but the effectiveness the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001].
+In the case of biomedical research, where living subjects are frequently used, sample sizes are often not "large" due to ethical and budgetary reasons [@charan2013] which might cause the corrections for lack of compound symmetry to be ineffective.
 
 Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial patients may voluntarily withdraw, whereas attrition  due to injury or weight loss in preclinical animal studies is a possible and indeed common occurrence.
+For example, in a clinical trial patients may voluntarily withdraw, whereas attrition  due to injury or weight loss in preclinical animal studies is possible.
 It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
 In each of these missing data scenarios,the _complete observations_ assumption of  classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis[@gueorguieva2004].
+When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis [@gueorguieva2004].
 This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
@@
 When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-On the other hand, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013]; but more importantly, they also expect a linear relationship between the response and time [@pinheiro2006], making them also limited to analyze non-linear data.
+On the other hand, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
+And even more importantly, LMEMs also expect a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze non-linear data.
 
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropiate inference.
+As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
 
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
+Although not frequently used in the biomedical community, these non-parametric models are frequently used in other fields to analyze longitudinal data.
+Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
 There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017] and 3) can easily accommodate missing observations.
@@
 The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, $\textsf{R}$[@r]has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
+In particular, $\textsf{R}$[@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
 hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
 At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017].
-Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and can be also used for power calculations and study design questions.
+Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
 
@@
 
-The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research, but certain assumptions are necessary for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: a linear relationship between covariates and response, a constant correlation between measurements, and complete observations for all subjects.
+The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research.
+This statistical methodology requires certain assumptions for the model to be valid.
+From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
 Each one of these assumptions is discussed below.
@@
 
-In a biomedical longitudinal study, two or more groups of subjects (patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs.
+In a biomedical longitudinal study, two or more groups of subjects (e.g., patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs.
 a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected respnse is modeled with both  _fixed_ and  _random_ components.
-The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the average effect of the novel drug in the "treatment" group.
+The collected response is modeled with both  _fixed_ and  _random_ components.
+The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug in the "treatment" group.
 The _random_ component can be defined as "noise" caused by some inherent variability within the study.
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while others are measured in the afternoon, it is possible that the difference in the collection time of the measurement introduces some "noise" in the signal.
+For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
 As their name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
@@
 
+## GAMs and covariance
+
+## GAMs and significance
+
 The use of GAMs to analyze biomedical longitudinal data, and the impact of missing observations in the fit of the model will be examined in detail in the following section using simulation.
+
+https://stats.stackexchange.com/questions/364568/extracting-significance-of-gam-smoothing-terms
 

### 2021-03-12 Manuscripts/Manuscript_AM_v4.Rmd
New manuscript document created

--- 2021-03-10 Manuscripts/Manuscript_AM_v3.Rmd
@@
 Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect before, during and after therapy, and the subjects to be studied can range from animals (mice, rats, rabbits), human patients, cells, or blood samples, among many others.
-Studies using longitudinal designs in biomedical research have analyzed tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010].
-The frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study.
-The frequency of measurements can range from minutely in animals to analyze anesthesia effects[@greening2018], weekly in patients to study the evolution of dermatitis symptoms in breast cancer [@sio2016], to monthly to study mouth opening following RT in neck cancer patients [@kamstra2015].
+Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
+Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response.
+Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following RT in neck cancer patients [@kamstra2015].
 
@@
 
-A _post hoc_ analysis is the statistical test used in conjuction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
+A _post hoc_ analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
 Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
@@
 Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019].
-Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to balance in a biomedical longitudinal study where a delicate balance between statistical power and sample size exist.
+Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to balance in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
 
@@
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor data [@skala2010;@vishwanath2009].
+The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data [@skala2010;@vishwanath2009].
 Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
 When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-On the other hand, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
+However, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
 And even more importantly, LMEMs also expect a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze non-linear data.
@@
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used in the biomedical community, these non-parametric models are frequently used in other fields to analyze longitudinal data.
+Although not frequently used by the biomedical community, these non-parametric models are customarily used in other fields to analyze longitudinal data.
 Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
@@
 
+### The repeated measures ANOVA case
+
 In a biomedical longitudinal study, two or more groups of subjects (e.g., patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs.
 a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response is modeled with both  _fixed_ and  _random_ components.
+The collected response is modeled with _fixed_ components.
 The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug in the "treatment" group.
-The _random_ component can be defined as "noise" caused by some inherent variability within the study.
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
-As their name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
 
@@
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{tij}\\ 
+y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{tij}\\ 
 (\#eq:linear-model)
@@
 In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-The _random effect_ $\mu_{ij}$ is assumed to be independently normally distributed with mean zero and variance $\sigma^2_\mu$, which can be expressed as $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
-This _random effect_ accounts for variability in each subject within each group by allowing different values for each subject.
-Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ or _random_ effects, and are assumed to be $\sim N(0,\sigma^2)$ .
+Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$.
 In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
@@
 If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
-Because the construction of the model is essentially the same in LMEMs, the linearity assumption holds for these models, although it is possible to fit models with different slopes at the subject-level[@pinheiro2006].
+
+### The Linear Mixed Model Case
+
+A linear mixed model for a longitudinal study can be written as:
+
+\begin{equation}
+y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{tij}\\ 
+(\#eq:LMEM)
+\end{equation}
+
+When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is apparent that the LMEM and rm-ANOVA models have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
+This term is defined as a _random effect_, which accounts for variability in each subject within each group.
+The _random_ component can be defined as "noise" caused by some inherent variability within the study that is not captured by the term $\varepsilon_{tij}$.
+
+For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
+As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
+The _random effect_ $\mu_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be independently normally distributed with mean zero and variance $\sigma^2_\mu$, which can be expressed as $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
+In essence,the _random effect_ in a LMEM enables to fit models with different slopes at the subject-level[@pinheiro2006].
+However,the expected linear relationship of the covariates and the response in Equation \@ref(eq:linear-model) and in Equation \@ref(eq:LMEM) is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
 
@@
 
-To demonstrate the limitations of rm-ANOVA for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
+To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
 An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix).
@@
 
-The rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the response indicates a difference in the effect of each treatment.
+The rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates a difference in the effect of each treatment.
 From the mean response per group (linear or quadratic), individual responses for each subject within each group are simulated using covariance matrices with compound symmetry (constant variance across time) or independent errors (errors generated from a a normal distribution that are not constant over time).
@@
 #heatmap of the penalization coefficient
+x_labels<-c("Intercept","1","2","3","4")
 p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
@@
   labs(x='Basis',y='Basis')+
+  scale_x_discrete(labels=x_labels)+
   geom_text(aes(label=round(cof,2)),size=10,show.legend = FALSE)+

### 2021-03-16 Manuscripts/Manuscript_AM_v4.Rmd
Working on v4 of manuscript
Updated final two sections, need to work on figure of linear and quadratic response

@@
 
-A linear mixed model for a longitudinal study can be written as:
+A linear mixed model (LMEM) is a class of statistical model that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to distinguish [@pinheiro2006;@west2014].
+A LMEM with interaction between time and treatment for a longitudinal study  can be written as:
 
@@
 
-When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is apparent that the LMEM and rm-ANOVA models have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
-This term is defined as a _random effect_, which accounts for variability in each subject within each group.
-The _random_ component can be defined as "noise" caused by some inherent variability within the study that is not captured by the term $\varepsilon_{tij}$.
+When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
+This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject within each group.
+The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the "global noise" term $\varepsilon_{tij}$ from Equation \@ref(eq:linear-model).
 
@@
 
+Although the specific methods of how GAMs model covariance and correlation is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
+A detailed description on basis functions and correlations can be found in [@hefley2017].
+In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the _time_ effect to account for the repeated measures over time.
+An example where this is covered is in the Appendix.
+
 ## GAMs and significance
+
+At the core of a biomedical longitudinal study lies the question of a significant difference between two or more treatments in different groups.
+Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_, GAMs use a different approach to estimate significance.
+In essence, the idea behind the estimation of significance in GAMs by group is that if the _difference_ between the confidence intervals of the smooths is not equal to zero then a significant difference exists at that timepoint (or timepoints).
+The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
+The estimation of siginificant differences in GAMs is covered in detail in the Appendix.
 

### 2021-03-17 Manuscripts/Manuscript v1/Manuscript_AM_v1.Rmd
Re arranged files, need to figure out the position of the last figure and the appendix section

--- 2021-01-28 Manuscripts/Manuscript_AM_v1.Rmd
@@
 ---
-title: '**The statistical analysis of non-linear longitudinal data in biomedical research
-  using generalized additive models**'
+title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
 author:
@@
 - John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
+subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
+output:
+  bookdown::pdf_document2:
+    keep_tex: yes
+    fig_caption: yes
+    extra_dependencies: ["subfig","breqn"]
+  bookdown::word_document2: default
+  bookdown::html_document2: default
+  link-citations: yes
+  css: style.css
+  '': default
 csl: ieee.csl
-output:
-- default
-- default
-- yes
-- style.css
-- default
-- extra_dependencies:
-  - subfig
-  - breqn
-  fig_caption: yes
-  keep_tex: yes
-subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
 bibliography: refs.bib

### 2021-03-17 Manuscripts/Manuscript v2/Manuscript_AM_v2.Rmd
Re arranged files, need to figure out the position of the last figure and the appendix section

--- 2021-02-09 Manuscripts/Manuscript_AM_v2.Rmd
@@
 In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004].
-Dropout from patients, attrition or injury in animals are among the reasons for missing observations.
+Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
 Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].

### 2021-03-17 Manuscripts/Manuscript v3/Manuscript_AM_v3.Rmd
Re arranged files, need to figure out the position of the last figure and the appendix section

--- 2021-03-12 Manuscripts/Manuscript_AM_v4.Rmd
@@
 
-### The repeated measures ANOVA case
-
 In a biomedical longitudinal study, two or more groups of subjects (e.g., patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs.
 a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response is modeled with _fixed_ components.
+The collected response is modeled with both  _fixed_ and  _random_ components.
 The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug in the "treatment" group.
+The _random_ component can be defined as "noise" caused by some inherent variability within the study.
+For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
+As their name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
 
@@
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{tij}\\ 
+y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{tij}\\ 
 (\#eq:linear-model)
@@
 In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$.
+The _random effect_ $\mu_{ij}$ is assumed to be independently normally distributed with mean zero and variance $\sigma^2_\mu$, which can be expressed as $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
+This _random effect_ accounts for variability in each subject within each group by allowing different values for each subject.
+Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ or _random_ effects, and are assumed to be $\sim N(0,\sigma^2)$ .
 In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
@@
 If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
-
-### The Linear Mixed Model Case
-
-A linear mixed model for a longitudinal study can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{tij}\\ 
-(\#eq:LMEM)
-\end{equation}
-
-When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is apparent that the LMEM and rm-ANOVA models have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
-This term is defined as a _random effect_, which accounts for variability in each subject within each group.
-The _random_ component can be defined as "noise" caused by some inherent variability within the study that is not captured by the term $\varepsilon_{tij}$.
-
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
-As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
-The _random effect_ $\mu_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be independently normally distributed with mean zero and variance $\sigma^2_\mu$, which can be expressed as $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
-In essence,the _random effect_ in a LMEM enables to fit models with different slopes at the subject-level[@pinheiro2006].
-However,the expected linear relationship of the covariates and the response in Equation \@ref(eq:linear-model) and in Equation \@ref(eq:LMEM) is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
+Because the construction of the model is essentially the same in LMEMs, the linearity assumption holds for these models, although it is possible to fit models with different slopes at the subject-level[@pinheiro2006].
 
@@
 
-To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
+To demonstrate the limitations of rm-ANOVA for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
 An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix).
@@
 
-The rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates a difference in the effect of each treatment.
+The rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the response indicates a difference in the effect of each treatment.
 From the mean response per group (linear or quadratic), individual responses for each subject within each group are simulated using covariance matrices with compound symmetry (constant variance across time) or independent errors (errors generated from a a normal distribution that are not constant over time).

### 2021-03-17 Manuscripts/Manuscript v4/Manuscript_AM_v4.Rmd
Re arranged files, need to figure out the position of the last figure and the appendix section

--- 2021-03-16 Manuscripts/Manuscript_AM_v4.Rmd
@@
 output:
-  bookdown::html_document2: default
   bookdown::pdf_document2:
@@
     extra_dependencies: ["subfig","breqn"]
+  bookdown::html_document2: default
   bookdown::word_document2:
@@
 
-## Simulation experiment on linearity and covariance structures
+## Simulation experiment on linearity and covariance structures {#simulation}
 
 To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix).
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:linear-models, A)).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:quadratic-response),A).
+An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme].
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response, A)).
+In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C).
+In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
+This is intentional in order to simulate the expected temporal evolution of some physiological quantity.
 
-The rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates a difference in the effect of each treatment.
-From the mean response per group (linear or quadratic), individual responses for each subject within each group are simulated using covariance matrices with compound symmetry (constant variance across time) or independent errors (errors generated from a a normal distribution that are not constant over time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (either compound symmetry or independent, which appear in panels A and E, respectively in both Figure \@ref(fig:linear-models) and Figure \@ref(fig:quadratic-response)) .
-The resulting simulated data, simulated errors, individual response and fitted parameters from the model are plotted.
+Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates a difference in the effect of each treatment.
+In other words, Group 1 can be thought as a "Control" group and Group 2 as a "Treatment" group.
+From the mean response per group (linear or quadratic), the variability or "error" of individual responses  within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
+Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \@ref(fig:l-q-response) B,D).
+A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \@ref(fig:linear-cases) and Figure \@ref(fig:quadratic-cases) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
 
@@
 
-  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A')) 
-    # return((p1 + p3 + plot_layout(guides = "collect")) / (p2 + p4 + plot_layout(guides = "collect")))
+  #return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
+
+  return((p1+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
 
@@
 
-```{r, linear-models, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
+```{r, l-q-response, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
+B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
+The rm-ANOVA model does not pick the trend of the quadratic data**"}
 # linear response, correlated errors (subject effect)
-A/B+plot_annotation(tag_levels = 'A')
+A/C+plot_annotation(tag_levels = 'A')
 ```
 
-In Figure \@ref(fig:linear-models)  panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-It is clear from Figure \@ref(fig:linear-models) that the fit produced by the rm-ANOVA model is good, as the predictions for the mean response are reasonably close to the "truth" of the simulated data.
+The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B).
+When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.
 
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:quadratic-response).
-Here, the mean response per group was simulated using a quadratic function and errors, individual responses and the rm-ANOVA model were produced in the same manner as \@ref(fig:linear-models) .
-The figure shows that the response is changing in each group through the timeline, and that the mean value is the same as the initial value by the fifth time point.
-Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panels D and H in Figure \@ref(fig:quadratic-response).
+However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response), C.
+Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in \@ref(fig:l-q-response) A and B .
+The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
+Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panel D in Figure \@ref(fig:l-q-response).
 
-```{r, quadratic-response, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
-# linear response, correlated errors (subject effect)
-C/D+plot_annotation(tag_levels='A')
-```
-
-In this case, Figure \@ref(fig:quadratic-response), panel A shows that both groups start with the same mean value at the first time point.
-Again, this is intentional in order to simulate the temporal behavior of some physiological quantity.
-Therefore, the idea behind \@ref(fig:quadratic-response) is to mimic a non-linear behavior from two groups that follow an inverse trajectory.
-A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data (\@ref(fig:quadratic-response), panels D and H)  indicates that the model is not capturing the changes within each group in a good way.
-Specifically, note that in Figure \@ref(fig:quadratic-response) the fitted mean response of the rm-ANOVA model (panels D and H) show that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model and moreover, the model is not being able to capture the initial value (which is the same in each group), instead fitting non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels A and D, and E and H).
+A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure (\@ref(fig:l-q-response), D indicates that the model is not capturing the changes within each group in a good way.
+Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
+Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D).
 If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response, potentially limiting valuable inferences from the data.
@@
 This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-Although the model fitted to the simulated data was an rm_ANOVA, the main issue of an expected linear trend in the response is the same in the case of LMEMs.
+Although the model fitted to the simulated data was an rm_ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM.
 In the following section, we present generalized additive models (GAMs) as an alternative to analyze longitudinal non-linear data.
@@
 
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure @\ref(fig:quadratic-response).
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure @\ref(fig:l-q-response), C.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits the trend of the group trough the timeline.
@@
 The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots.
-In Panel A of Figure \@ref(fig:basis-functions-plot), the four basis functions (and the intercept) are shown.
+In Panel A of Figure \@ref(fig:basis-plot), the four basis functions (and the intercept) are shown.
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
@@
 In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \@ref(fig:basis-functions-plot), Panel C shows the resulting penalized basis functions.
+In Figure \@ref(fig:basis-plot), Panel C shows the resulting penalized basis functions.
 Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
@@
 
-```{r,basis-functions-plot,fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for _time_ for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
-D.
-Smoother for _time_ and original datapoints.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.**"}
+```{r,basis-functions-plot, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
 
@@
 #Combining all
-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
+b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
   theme(
      text=element_text(size=30))
+
+```
+
+```{r,basis-plot,fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Basis functions for a single smoother for time with five knots.
+A: Basis functions for a single smoother for _time_ for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
+B: Penalty matrix for the basis functions.
+Each basis function is penalized by a coefficient which can be positive or negative.
+The coefficient determines the overall effect of each basis in the final smoother.
+C: Penalized basis functions.
+Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
+D.
+Smoother for _time_ and original datapoints.
+The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.**"}
+
+b_plot
 
@@
 
-## GAMs and significance
+## Determination of significance in GAMs
 
@@
 
-https://stats.stackexchange.com/questions/364568/extracting-significance-of-gam-smoothing-terms
-
 ***
@@
 # References
+
+<div id="refs"></div>
+
+# (APPENDIX) Appendix {-}
+
+# Simulation
+
+## Compound symmetry and independent errors in linear and quadratic responses
+
+This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation).
+The linear simulations using Figure \@ref(fig:linear-cases) show in  panels A and D the simulated mean responses and individual datapoints.
+Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
+In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
+Panels D and H show the predictions from the rm-ANOVA model.
+
+```{r,include=TRUE,message=FALSE,echo=TRUE}
+## Example with linear response
+example <- function(n_time = 6, 
+                    fun_type = "linear", 
+                    error_type = "correlated") {
+
+  if (!(fun_type %in% c("linear", "quadratic")))
+    stop('fun_type must be either "linear", or "quadratic"')
+  if (!(error_type %in% c("correlated", "independent")))
+    stop('fun_type must be either "correlated", or "independent"')
+
+  library(tidyverse)
+  library(mvnfast)
+  library(nlme)
+  library(ggsci)
+
+  x <- seq(1,6, length.out = n_time)
+  mu <- matrix(0, length(x), 2)
+  # linear response
+  if (fun_type == "linear") {
+    mu[, 1] <- - (0.25*x)+2  
+    mu[, 2] <- 0.25*x+2
+  } else {
+    # nonlinear response
+
+    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
+    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
+  }
+  # matplot(mu, type = 'l')
+
+  y <- array(0, dim = c(length(x), 2, 10))
+  errors <- array(0, dim = c(length(x), 2, 10))
+
+  if (error_type == "independent") {
+    ## independent errors
+    for (i in 1:2) {
+      for (j in 1:10) {
+        errors[, i, j] <- rnorm(6, 0, 0.25)
+        y[, i, j] <- mu[, i] + errors[, i, j]
+      }
+    }
+  } else {
+    for (i in 1:2) {     # number of treatments
+      for (j in 1:10) {  # number of subjects
+        # compound symmetry errors
+        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
+        y[, i, j] <- mu[, i] + errors[, i, j]
+      }
+    }
+  }
+
+  ## subject random effects
+
+  ## visualizing the difference between independent errors and compound symmetry
+  ## why do we need to account for this -- overly confident inference
+
+  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
+  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
+  dimnames(mu) <- list(time = x, treatment = 1:2)
+  dat <- as.data.frame.table(y, responseName = "y")
+  dat_errors <- as.data.frame.table(errors, responseName = "errors")
+  dat_mu <- as.data.frame.table(mu, responseName = "mu")
+  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
+  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
+  dat$time <- as.numeric(as.character(dat$time))
+  dat <- dat %>%
+    mutate(subject = factor(paste(subject, treatment, sep = "-")))
+
+  ## repeated measures ANOVA in R
+  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
+  dat$preds_lm <- predict(fit_lm)
+
+  fit_lme <- lme(y ~ treatment + time + treatment:time,
+                 data = dat,
+                 random = ~ 1 | subject,
+                 correlation = corCompSymm(form = ~ 1 | subject)
+  )
+
+  pred_dat <- expand.grid(
+    treatment = factor(1:2), 
+    time = unique(dat$time)
+  )
+
+  dat$y_pred <- predict(fit_lme)
+
+  return(list(
+    dat = dat,
+    pred_dat = pred_dat,
+    fit_lm = fit_lm,
+    fit_lme = fit_lme
+
+  ))
+}
+
+plot_example <- function(sim_dat) {
+  library(patchwork)
+  ## Plot the simulated data
+  p1 <- sim_dat$dat %>%
+    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
+    geom_point(show.legend=FALSE) +labs(y='response')+
+    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
+    theme_classic() +
+    #ggtitle("Simulated data with true response function")+
+    theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
+    scale_color_aaas()
+
+  p2 <- sim_dat$dat %>%
+    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
+    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
+    # facet_wrap(~ treatment) +
+    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
+    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
+    #ggtitle("Simulated data\nIndividual responses with population mean") +
+    theme_classic()+
+     theme(plot.title = element_text(size = 30, 
+                                face = "bold"),
+     text=element_text(size=30))+
+    scale_color_aaas()
+
+   p3 <- sim_dat$dat %>%
+    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
+    geom_line(show.legend=FALSE) +labs(y='errors')+
+     theme_classic()+
+    # facet_wrap(~ treatment) +
+    #ggtitle("Simulated errors") +
+     theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
+    scale_color_aaas()
+
+  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
+    geom_point()+labs(y='response')+
+    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
+    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
+    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
+    theme_classic() +
+    #ggtitle("Fitted Model")+
+    theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
+    scale_color_aaas()
+
+  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
+
+}
+
+txt<-18
+A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
+
+B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
+
+C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
+
+D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
+
+```
+
+```{r, linear-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
+B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
+The rm-ANOVA model does not pick the trend of the quadratic data**"}
+# linear response, correlated errors (subject effect)
+A1/B1+plot_annotation(tag_levels = 'A')
+```
+
+For the quadratic response case, Figure \@ref(fig:quadratic-cases) shows the simulated responses using compound symmetry and independent errors.
+
+```{r, quadratic-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
+A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
+B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
+C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
+D,H: Estimations from the rm-ANOVA model for the mean group response.
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
+# linear response, correlated errors (subject effect)
+C1/D1+plot_annotation(tag_levels = 'A')
+```

### 2021-03-17 Manuscripts/Manuscript v4/Manuscript_AM_v4.tex
Re arranged files, need to figure out the position of the last figure and the appendix section

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models},
  pdfauthor={Ariel Mundo; Timothy J. Muldoon; John R. Tipton},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{subfig}
\usepackage{breqn}

\title{\textbf{The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
\author{Ariel Mundo\footnote{Department of Biomedical Engineering, University of Arkansas, Fayetteville} \and Timothy J. Muldoon\footnote{Department of Biomedical Engineering, University of Arkansas, Fayetteville} \and John R. Tipton\footnote{Department of Mathematical Sciences, University of Arkansas, Fayetteville}}
\date{}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{background}{%
\section{Background}\label{background}}

Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study). Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others. Tumor response {[}1--4{]}, antibody expression {[}5,6{]}, and cell metabolism {[}7,8{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response. Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}9{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}10{]}, to monthly measurements to study a long-term response such as mouth opening following RT in neck cancer patients {[}11{]}.

Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study. The frequentist paradigm regards probability as a limiting frequency {[}12{]} by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}13,14{]}.

The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}15{]}. This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings {[}1,8,16{]}. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model is only able to accommodate linear trends that are far from adequately representing the biological phenomenon of interest.

A \emph{post hoc} analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data. In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}17{]}. For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}18{]}.

Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}19,20{]}, consequently biasing the conclusions of the study. Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}21{]}), but they in turn reduce statistical power (1-\(\beta\)){[}22{]}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) {[}23,24{]}. Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to balance in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.

On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}25{]}. Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}26,27{]}, but the effectiveness the correction is limited by the size of the sample, the number of measurements{[}28{]}, and group sizes {[}29{]}. In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}30{]} which might cause the corrections for lack of compound symmetry to be ineffective.

Due to a variety of causes, the number of observations during a study can vary between all subjects. For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible. It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points. In each of these missing data scenarios,the \emph{complete observations} assumption of classical rm-ANOVA is violated. When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}13{]}. This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information. The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}8,16{]}. Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}15{]}. However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}13,31{]}. And even more importantly, LMEMs also expect a linear relationship between the response and time {[}15{]}, making them unsuitable to analyze non-linear data.

As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.

In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data. Although not frequently used by the biomedical community, these non-parametric models are customarily used in other fields to analyze longitudinal data. Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}32--34{]}, health-environment interactions {[}35{]} and the dynamics of government in political science {[}36{]} . There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}37{]} and 3) can easily accommodate missing observations. Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.

The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs. In particular, \(\textsf{R}\){[}38{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}37,39{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}28{]}. Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.

This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations is explained in detail. Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations. Third, using simulated data that reproduces patterns in previously reported studies {[}16{]} we illustrate the type of non-linear longitudinal data that often occurs in biomedical research. The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies. Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in \(\textsf{R}\), in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.

In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models by improving the standards for reproducibility in biomedical research.

\hypertarget{challenges-presented-by-longitudinal-studies}{%
\section{Challenges presented by longitudinal studies}\label{challenges-presented-by-longitudinal-studies}}

\hypertarget{the-repeated-measures-anova}{%
\subsection{The repeated measures ANOVA}\label{the-repeated-measures-anova}}

The \emph{repeated measures analysis of variance} (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research. This statistical methodology requires certain assumptions for the model to be valid. From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects. Each one of these assumptions is discussed below.

\hypertarget{linear-relationship}{%
\subsection{Linear relationship}\label{linear-relationship}}

\hypertarget{the-repeated-measures-anova-case}{%
\subsubsection{The repeated measures ANOVA case}\label{the-repeated-measures-anova-case}}

In a biomedical longitudinal study, two or more groups of subjects (e.g., patients, mice, samples) are subject to different treatments (e.g., a ``treatment'' group receives a novel drug vs.~a ``control'' group that receives a placebo), and measurements from each subject within each group are collected at specific time points. The collected response is modeled with \emph{fixed} components. The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug in the ``treatment'' group.

Mathematically speaking, a rm-ANOVA model with an interaction can be written as:

\begin{equation}
y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{tij}\\ 
\label{eq:linear-model}
\end{equation}

In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively. Independent errors \(\varepsilon_{tij}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\).
In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy''). Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B). The linear models then can be expressed as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\beta_0 + \beta_1 \times time_{t} + \beta_2 \times +\beta_3 \times time_{t} \times +\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
\label{eq:ANOVA-by-group}
\end{equation}

To further simplify the expression, substitute \(\widetilde{\mu}=\beta_0+\beta_{2}\) and \(\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}\) in the equation for Group B. This substitution allows for a different intercept and slope for Groups A and B. The model is then written as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\widetilde{\mu} + \widetilde{\beta_1} \times time_{t} +\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
\label{eq:ANOVA-lines}
\end{equation}

Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group). In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response, this means that either presented as Equation \eqref{eq:linear-model}, Equation \eqref{eq:ANOVA-by-group} or Equation \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data. If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.

\hypertarget{the-linear-mixed-model-case}{%
\subsubsection{The Linear Mixed Model Case}\label{the-linear-mixed-model-case}}

A linear mixed model (LMEM) is a class of statistical model that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish {[}15,40{]}. A LMEM with interaction between time and treatment for a longitudinal study can be written as:

\begin{equation}
y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{tij}\\ 
\label{eq:LMEM}
\end{equation}

When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)). This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject within each group. The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the ``global noise'' term \(\varepsilon_{tij}\) from Equation \eqref{eq:linear-model}.

For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data. As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value. The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be independently normally distributed with mean zero and variance \(\sigma^2_\mu\), which can be expressed as \(\mu_{ij} \sim N(0,\sigma^2_\mu)\). In essence,the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}15{]}. However,the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.

\hypertarget{covariance-in-rm-anova-and-lmems}{%
\subsection{Covariance in rm-ANOVA and LMEMs}\label{covariance-in-rm-anova-and-lmems}}

In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group. The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}41{]} (For an in-depth analysis of the covariance matrix see {[}40,42{]}).

In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity''). Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}26,42,43{]}. However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}44{]}; and it is higher between consecutive measurements {[}13,25{]}. Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}26,27{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}28{]}, and they are not robust if the group sizes are unbalanced {[}29{]}. Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.

In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}15{]}. Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself. Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.

\hypertarget{missing-observations}{%
\subsection{Missing observations}\label{missing-observations}}

Missing observations are an issue that arises frequently in longitudinal studies. In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}45{]}. Dropout from patients and attrition or injury in animals are among the reasons for missing observations. Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}42{]}. In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}46{]}. If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}47{]}, and in the case of MNAR the missing values are dependent on their value. An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis. This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}48{]}.

In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}40{]}. For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.

This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and missing data. Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study. However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data. To better convey the issues of linearity and correlation in linear models fitted to non-linear data, simulation is used in the next section.

\hypertarget{simulation}{%
\subsection{Simulation experiment on linearity and covariance structures}\label{simulation}}

To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each. An rm-ANOVA model (Equation \eqref{eq:linear-model}) is fitted to each group, using \(\textsf{R}\){[}38{]} and the package \emph{nlme}{[}49{]}.
Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure @ref(fig:l-q-response, A)). In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}, C). In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point. This is intentional in order to simulate the expected temporal evolution of some physiological quantity.

Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates a difference in the effect of each treatment. In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group. From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time). Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D). A more comprehensive exploration of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \ref{fig:linear-cases} and Figure \ref{fig:quadratic-cases} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.

\begin{figure}
\centering
\includegraphics{Manuscript_AM_v4_files/figure-latex/l-q-response-1.pdf}
\caption{\label{fig:l-q-response}\textbf{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model. A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data. B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic). Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data. The rm-ANOVA model does not pick the trend of the quadratic data}}
\end{figure}

The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response},B). When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.

However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}, C. Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in \ref{fig:l-q-response} A and B . The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group. Fitting an rm-ANOVA model \eqref{eq:linear-model} to this data produces the fit that appears in panel D in Figure \ref{fig:l-q-response}.

A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure (\ref{fig:l-q-response}, D indicates that the model is not capturing the changes within each group in a good way. Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model. Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare panels C and D). If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response, potentially limiting valuable inferences from the data.

This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data. Although the model fitted to the simulated data was an rm\_ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM. In the following section, we present generalized additive models (GAMs) as an alternative to analyze longitudinal non-linear data.

\hypertarget{gams-as-a-special-case-of-generalized-linear-models}{%
\section{GAMs as a special case of Generalized Linear Models}\label{gams-as-a-special-case-of-generalized-linear-models}}

\hypertarget{gams-and-basis-functions}{%
\subsection{GAMs and Basis Functions}\label{gams-and-basis-functions}}

Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors{[}50{]}. In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}34,37,51{]}. A GAM model can be written as:

\begin{equation}
  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
  \label{eq:GAM}
\end{equation}

Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the function \(f(x_t\mid \beta_j)\) and \(\varepsilon_{ijt}\) represents the residual error.

In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible \emph{smooth functions}. This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response. One possible function for \(f(x_t\mid \beta_j)\) that allows for non-linear responses is a polynomial, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}36{]}. In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic, and causes bias at the endpoints of the time period.

The smooth functional relationship between the covariates and the response in GAMs is specified using a parametric relationship that can be fit within the GLM framework, by using \emph{basis functions} expansions of the covariates and by estimating random coefficients for these basis functions and this step is achieved by using \emph{basis functions} to represent them. A \emph{basis} is a set of functions that spans the space where the smooths that approximate \(f(x_t \mid \beta_j)\) exist {[}34{]}. For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\). The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}52{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions. In the case of Equation \eqref{eq:GAM}, the basis function is \(f(x_t\mid \beta_j)\), which means that the model allows relationships beyond linear for the covariates.

A commonly used \emph{basis function} is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations{[}34,37{]}. Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness {[}53{]}. Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.

To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure @\ref(fig:l-q-response), C. The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits the trend of the group trough the timeline. The timeline can be divided equally spaced \emph{knots}, each knot being a region where a different basis function will be used. Because there are six timepoints for this group, five knots can be used. The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept). The choice of basis functions is already optimized in the package \emph{mgcv} depending on the number of knots. In Panel A of Figure \ref{fig:basis-plot}, the four basis functions (and the intercept) are shown. Each of the basis functions is composed of six different points (because there are six points on the timeline). To control the wigliness of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.

In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A. In Figure \ref{fig:basis-plot}, Panel C shows the resulting penalized basis functions. Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value. Finally, the penalized basis functions are added at each timepoint to produce the smooth term. The resulting smooth term for the effect of \emph{time} is shown in Panel D (orange line) along the simulated values per group, which appear as points.

\begin{figure}
\centering
\includegraphics{Manuscript_AM_v4_files/figure-latex/basis-plot-1.pdf}
\caption{\label{fig:basis-plot}\textbf{Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for \emph{time} for the simulated data of Group 1 from Figure 2, the intercept basis is not shown. B: Penalty matrix for the basis functions. Each basis function is penalized by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Penalized basis functions. Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis. D. Smoother for \emph{time} and original datapoints. The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}}
\end{figure}

\hypertarget{gams-and-covariance}{%
\subsection{GAMs and covariance}\label{gams-and-covariance}}

Although the specific methods of how GAMs model covariance and correlation is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in {[}52{]}. In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the \emph{time} effect to account for the repeated measures over time. An example where this is covered is in the Appendix.

\hypertarget{determination-of-significance-in-gams}{%
\subsection{Determination of significance in GAMs}\label{determination-of-significance-in-gams}}

At the core of a biomedical longitudinal study lies the question of a significant difference between two or more treatments in different groups. Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values}, GAMs use a different approach to estimate significance. In essence, the idea behind the estimation of significance in GAMs by group is that if the \emph{difference} between the confidence intervals of the smooths is not equal to zero then a significant difference exists at that timepoint (or timepoints). The absence of a \emph{p-value} in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent. The estimation of siginificant differences in GAMs is covered in detail in the Appendix.

The use of GAMs to analyze biomedical longitudinal data, and the impact of missing observations in the fit of the model will be examined in detail in the following section using simulation.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-roblyer2011}{}%
{[}1{]} D. Roblyer, S. Ueda, A. Cerussi, W. Tanamai, A. Durkin, R. Mehta, D. Hsiang, J.A. Butler, C. McLaren, W.-P. Chen, others, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences. 108 (2011) 14626--14631.

\leavevmode\hypertarget{ref-tank2020}{}%
{[}2{]} A. Tank, H.M. Peterson, V. Pera, S. Tabassum, A. Leproux, T. O'Sullivan, E. Jones, H. Cabral, N. Ko, R.S. Mehta, others, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research. 22 (2020) 1--10.

\leavevmode\hypertarget{ref-pavlov2018}{}%
{[}3{]} M.V. Pavlov, T.I. Kalganova, Y.S. Lyubimtseva, V.I. Plekhanov, G.Y. Golubyatnikov, O.Y. Ilyinskaya, A.G. Orlova, P.V. Subochev, D.V. Safonov, N.M. Shakhova, others, Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy, Journal of Biomedical Optics. 23 (2018) 091410.

\leavevmode\hypertarget{ref-demidov2018}{}%
{[}4{]} V. Demidov, A. Maeda, M. Sugita, V. Madge, S. Sadanand, C. Flueraru, I.A. Vitkin, Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography, Scientific Reports. 8 (2018) 1--12.

\leavevmode\hypertarget{ref-ritter2001}{}%
{[}5{]} G. Ritter, L.S. Cohen, C. Williams, E.C. Richards, L.J. Old, S. Welt, Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody a33, Cancer Research. 61 (2001) 6851--6859.

\leavevmode\hypertarget{ref-roth2017}{}%
{[}6{]} E.M. Roth, A.C. Goldberg, A.L. Catapano, A. Torri, G.D. Yancopoulos, N. Stahl, A. Brunet, G. Lecorps, H.M. Colhoun, Antidrug antibodies in patients treated with alirocumab, (2017).

\leavevmode\hypertarget{ref-jones2018}{}%
{[}7{]} J.D. Jones, H.E. Ramser, A.E. Woessner, K.P. Quinn, In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing, Communications Biology. 1 (2018) 1--8.

\leavevmode\hypertarget{ref-skala2010}{}%
{[}8{]} M.C. Skala, A.N. Fontanella, L. Lan, J.A. Izatt, M.W. Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics. 15 (2010) 011112.

\leavevmode\hypertarget{ref-greening2018}{}%
{[}9{]} G.J. Greening, K.P. Miller, C.R. Spainhour, M.D. Cato, T.J. Muldoon, Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy, Biomedical Optics Express. 9 (2018) 2871--2886.

\leavevmode\hypertarget{ref-sio2016}{}%
{[}10{]} T.T. Sio, P.J. Atherton, B.J. Birckhead, D.J. Schwartz, J.A. Sloan, D.K. Seisler, J.A. Martenson, C.L. Loprinzi, P.C. Griffin, R.F. Morton, others, Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}), Supportive Care in Cancer. 24 (2016) 3847--3855.

\leavevmode\hypertarget{ref-kamstra2015}{}%
{[}11{]} J. Kamstra, P. Dijkstra, M. Van Leeuwen, J. Roodenburg, J. Langendijk, Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study, Oral Oncology. 51 (2015) 548--555.

\leavevmode\hypertarget{ref-wagenmakers2008}{}%
{[}12{]} E.-J. Wagenmakers, M. Lee, T. Lodewyckx, G.J. Iverson, Bayesian versus frequentist inference, in: Bayesian Evaluation of Informative Hypotheses, Springer, 2008: pp. 181--207.

\leavevmode\hypertarget{ref-gueorguieva2004}{}%
{[}13{]} R. Gueorguieva, J.H. Krystal, Move over anova: Progress in analyzing repeated-measures data andits reflection in papers published in the archives of general psychiatry, Archives of General Psychiatry. 61 (2004) 310--317.

\leavevmode\hypertarget{ref-schober2018}{}%
{[}14{]} P. Schober, T.R. Vetter, Repeated measures designs and analysis of longitudinal data: If at first you do not succeed---try, try again, Anesthesia and Analgesia. 127 (2018) 569.

\leavevmode\hypertarget{ref-pinheiro2006}{}%
{[}15{]} J. Pinheiro, D. Bates, Mixed-effects models in s and s-plus, Springer Science \& Business Media, 2006.

\leavevmode\hypertarget{ref-vishwanath2009}{}%
{[}16{]} K. Vishwanath, H. Yuan, W.T. Barry, M.W. Dewhirst, N. Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia. 11 (2009) 889--900.

\leavevmode\hypertarget{ref-dennis2019}{}%
{[}17{]} B. Dennis, J.M. Ponciano, M.L. Taper, S.R. Lele, Errors in statistical inference under model misspecification: Evidence, hypothesis testing, and aic, Frontiers in Ecology and Evolution. 7 (2019) 372.

\leavevmode\hypertarget{ref-wang2019}{}%
{[}18{]} B. Wang, Z. Zhou, H. Wang, X.M. Tu, C. Feng, The p-value and model specification in statistics, General Psychiatry. 32 (2019).

\leavevmode\hypertarget{ref-liu2010}{}%
{[}19{]} C. Liu, T.P. Cripe, M.-O. Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy. 18 (2010) 1724--1730.

\leavevmode\hypertarget{ref-halsey2015}{}%
{[}20{]} L.G. Halsey, D. Curran-Everett, S.L. Vowler, G.B. Drummond, The fickle p value generates irreproducible results, Nature Methods. 12 (2015) 179--185.

\leavevmode\hypertarget{ref-abdi2010}{}%
{[}21{]} H. Abdi, Holm's sequential bonferroni procedure, Encyclopedia of Research Design. 1 (2010) 1--8.

\leavevmode\hypertarget{ref-nakagawa2004}{}%
{[}22{]} S. Nakagawa, A farewell to bonferroni: The problems of low statistical power and publication bias, Behavioral Ecology. 15 (2004) 1044--1045.

\leavevmode\hypertarget{ref-gelman2012}{}%
{[}23{]} A. Gelman, J. Hill, M. Yajima, Why we (usually) don't have to worry about multiple comparisons, Journal of Research on Educational Effectiveness. 5 (2012) 189--211.

\leavevmode\hypertarget{ref-albers2019}{}%
{[}24{]} C. Albers, The problem with unadjusted multiple and sequential statistical testing, Nature Communications. 10 (2019) 1--4.

\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
{[}25{]} C. Ugrinowitsch, G.W. Fellingham, M.D. Ricard, Limitations of ordinary least squares models in analyzing repeated measures data, Medicine and Science in Sports and Exercise. 36 (2004) 2144--2148.

\leavevmode\hypertarget{ref-huynh1976}{}%
{[}26{]} H. Huynh, L.S. Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics. 1 (1976) 69--82.

\leavevmode\hypertarget{ref-greenhouse1959}{}%
{[}27{]} S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112.

\leavevmode\hypertarget{ref-haverkamp2017}{}%
{[}28{]} N. Haverkamp, A. Beauducel, Violation of the sphericity assumption and its effect on type-i error rates in repeated measures anova and multi-level linear models (mlm), Frontiers in Psychology. 8 (2017) 1841.

\leavevmode\hypertarget{ref-keselman2001}{}%
{[}29{]} H. Keselman, J. Algina, R.K. Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematical and Statistical Psychology. 54 (2001) 1--20.

\leavevmode\hypertarget{ref-charan2013}{}%
{[}30{]} J. Charan, N. Kantharia, How to calculate sample size in animal studies?, Journal of Pharmacology \& Pharmacotherapeutics. 4 (2013) 303.

\leavevmode\hypertarget{ref-barr2013}{}%
{[}31{]} D.J. Barr, R. Levy, C. Scheepers, H.J. Tily, Random effects structure for confirmatory hypothesis testing: Keep it maximal, Journal of Memory and Language. 68 (2013) 255--278.

\leavevmode\hypertarget{ref-rose2012}{}%
{[}32{]} N.L. Rose, H. Yang, S.D. Turner, G.L. Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to Scotland, UK, Geochimica et Cosmochimica Acta. 82 (2012) 113--135.

\leavevmode\hypertarget{ref-pedersen2019}{}%
{[}33{]} E.J. Pedersen, D.L. Miller, G.L. Simpson, N. Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, PeerJ. 7 (2019) e6876.

\leavevmode\hypertarget{ref-simpson2018}{}%
{[}34{]} G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018) 149.

\leavevmode\hypertarget{ref-yang2012}{}%
{[}35{]} L. Yang, G. Qin, N. Zhao, C. Wang, G. Song, Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality, BMC Medical Research Methodology. 12 (2012) 165.

\leavevmode\hypertarget{ref-beck1998}{}%
{[}36{]} N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. (1998) 596--627.

\leavevmode\hypertarget{ref-wood2017}{}%
{[}37{]} S.N. Wood, Generalized additive models: An introduction with r, CRC press, 2017.

\leavevmode\hypertarget{ref-r}{}%
{[}38{]} R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.

\leavevmode\hypertarget{ref-wood2016}{}%
{[}39{]} S.N. Wood, N., Pya, B. Säfken, Smoothing parameter and model selection for general smooth models (with discussion), Journal of the American Statistical Association. 111 (2016) 1548--1575.

\leavevmode\hypertarget{ref-west2014}{}%
{[}40{]} B.T. West, K.B. Welch, A.T. Galecki, Linear mixed models: A practical guide using statistical software, CRC Press, 2014.

\leavevmode\hypertarget{ref-wolfinger1996}{}%
{[}41{]} R.D. Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics. (1996) 205--230.

\leavevmode\hypertarget{ref-weiss2005}{}%
{[}42{]} R.E. Weiss, Modeling longitudinal data, Springer Science \& Business Media, 2005.

\leavevmode\hypertarget{ref-geisser1958}{}%
{[}43{]} S. Geisser, S.W. Greenhouse, others, An extension of box's results on the use of the \(F\) distribution in multivariate analysis, The Annals of Mathematical Statistics. 29 (1958) 885--891.

\leavevmode\hypertarget{ref-maxwell2017}{}%
{[}44{]} S.E. Maxwell, H.D. Delaney, K. Kelley, Designing experiments and analyzing data: A model comparison perspective, Routledge, 2017.

\leavevmode\hypertarget{ref-molenberghs2004}{}%
{[}45{]} G. Molenberghs, H. Thijs, I. Jansen, C. Beunckens, M.G. Kenward, C. Mallinckrodt, R.J. Carroll, Analyzing incomplete longitudinal clinical trial data, Biostatistics. 5 (2004) 445--464.

\leavevmode\hypertarget{ref-scheffer2002}{}%
{[}46{]} J. Scheffer, Dealing with missing data, (2002).

\leavevmode\hypertarget{ref-potthoff2006}{}%
{[}47{]} R.F. Potthoff, G.E. Tudor, K.S. Pieper, V. Hasselblad, Can one assess whether missing data are missing at random in medical studies?, Statistical Methods in Medical Research. 15 (2006) 213--234.

\leavevmode\hypertarget{ref-ma2012}{}%
{[}48{]} Y. Ma, M. Mazumdar, S.G. Memtsoudis, Beyond repeated-measures analysis of variance: Advanced statistical methods for the analysis of longitudinal data in anesthesia research, Regional Anesthesia \& Pain Medicine. 37 (2012) 99--105.

\leavevmode\hypertarget{ref-nlme}{}%
{[}49{]} J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, nlme: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.

\leavevmode\hypertarget{ref-nelder1972}{}%
{[}50{]} J.A. Nelder, R.W. Wedderburn, Generalized linear models, Journal of the Royal Statistical Society: Series A (General). 135 (1972) 370--384.

\leavevmode\hypertarget{ref-hastie1987}{}%
{[}51{]} T. Hastie, R. Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association. 82 (1987) 371--386.

\leavevmode\hypertarget{ref-hefley2017}{}%
{[}52{]} T.J. Hefley, K.M. Broms, B.M. Brost, F.E. Buderman, S.L. Kay, H.R. Scharf, J.R. Tipton, P.J. Williams, M.B. Hooten, The basis function approach for modeling autocorrelation in ecological data, Ecology. 98 (2017) 632--646.

\leavevmode\hypertarget{ref-wegman1983}{}%
{[}53{]} E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365.

\hypertarget{appendix-appendix}{%
\appendix}

\hypertarget{simulation-1}{%
\section{Simulation}\label{simulation-1}}

\hypertarget{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}{%
\subsection{Compound symmetry and independent errors in linear and quadratic responses}\label{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}}

This section simulated linear and quadratic data in the same manner as in Section \ref{simulation}. The linear simulations using Figure \ref{fig:linear-cases} show in panels A and D the simulated mean responses and individual datapoints. Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response. In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern. Panels D and H show the predictions from the rm-ANOVA model.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Example with linear response}
\NormalTok{example <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{n_time =} \DecValTok{6}\NormalTok{, }
                    \DataTypeTok{fun_type =} \StringTok{"linear"}\NormalTok{, }
                    \DataTypeTok{error_type =} \StringTok{"correlated"}\NormalTok{) \{}

  \ControlFlowTok{if}\NormalTok{ (}\OperatorTok{!}\NormalTok{(fun_type }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"linear"}\NormalTok{, }\StringTok{"quadratic"}\NormalTok{)))}
    \KeywordTok{stop}\NormalTok{(}\StringTok{'fun_type must be either "linear", or "quadratic"'}\NormalTok{)}
  \ControlFlowTok{if}\NormalTok{ (}\OperatorTok{!}\NormalTok{(error_type }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"correlated"}\NormalTok{, }\StringTok{"independent"}\NormalTok{)))}
    \KeywordTok{stop}\NormalTok{(}\StringTok{'fun_type must be either "correlated", or "independent"'}\NormalTok{)}

  \KeywordTok{library}\NormalTok{(tidyverse)}
  \KeywordTok{library}\NormalTok{(mvnfast)}
  \KeywordTok{library}\NormalTok{(nlme)}
  \KeywordTok{library}\NormalTok{(ggsci)}

\NormalTok{  x <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{6}\NormalTok{, }\DataTypeTok{length.out =}\NormalTok{ n_time)}
\NormalTok{  mu <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{length}\NormalTok{(x), }\DecValTok{2}\NormalTok{)}
  \CommentTok{# linear response}
  \ControlFlowTok{if}\NormalTok{ (fun_type }\OperatorTok{==}\StringTok{ "linear"}\NormalTok{) \{}
\NormalTok{    mu[, }\DecValTok{1}\NormalTok{] <-}\StringTok{ }\OperatorTok{-}\StringTok{ }\NormalTok{(}\FloatTok{0.25}\OperatorTok{*}\NormalTok{x)}\OperatorTok{+}\DecValTok{2}  
\NormalTok{    mu[, }\DecValTok{2}\NormalTok{] <-}\StringTok{ }\FloatTok{0.25}\OperatorTok{*}\NormalTok{x}\OperatorTok{+}\DecValTok{2}
\NormalTok{  \} }\ControlFlowTok{else}\NormalTok{ \{}
    \CommentTok{# nonlinear response}

\NormalTok{    mu[, }\DecValTok{1}\NormalTok{] <-}\StringTok{  }\OperatorTok{-}\NormalTok{(}\FloatTok{0.25} \OperatorTok{*}\StringTok{ }\NormalTok{x}\OperatorTok{^}\DecValTok{2}\NormalTok{) }\FloatTok{+1.5}\OperatorTok{*}\NormalTok{x}\FloatTok{-1.25}
\NormalTok{    mu[, }\DecValTok{2}\NormalTok{] <-}\StringTok{ }\NormalTok{(}\FloatTok{0.25} \OperatorTok{*}\StringTok{ }\NormalTok{x}\OperatorTok{^}\DecValTok{2}\NormalTok{) }\FloatTok{-1.5}\OperatorTok{*}\NormalTok{x}\FloatTok{+1.25}
\NormalTok{  \}}
  \CommentTok{# matplot(mu, type = 'l')}

\NormalTok{  y <-}\StringTok{ }\KeywordTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DataTypeTok{dim =} \KeywordTok{c}\NormalTok{(}\KeywordTok{length}\NormalTok{(x), }\DecValTok{2}\NormalTok{, }\DecValTok{10}\NormalTok{))}
\NormalTok{  errors <-}\StringTok{ }\KeywordTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DataTypeTok{dim =} \KeywordTok{c}\NormalTok{(}\KeywordTok{length}\NormalTok{(x), }\DecValTok{2}\NormalTok{, }\DecValTok{10}\NormalTok{))}

  \ControlFlowTok{if}\NormalTok{ (error_type }\OperatorTok{==}\StringTok{ "independent"}\NormalTok{) \{}
    \CommentTok{## independent errors}
    \ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{) \{}
      \ControlFlowTok{for}\NormalTok{ (j }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{) \{}
\NormalTok{        errors[, i, j] <-}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(}\DecValTok{6}\NormalTok{, }\DecValTok{0}\NormalTok{, }\FloatTok{0.25}\NormalTok{)}
\NormalTok{        y[, i, j] <-}\StringTok{ }\NormalTok{mu[, i] }\OperatorTok{+}\StringTok{ }\NormalTok{errors[, i, j]}
\NormalTok{      \}}
\NormalTok{    \}}
\NormalTok{  \} }\ControlFlowTok{else}\NormalTok{ \{}
    \ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{) \{     }\CommentTok{# number of treatments}
      \ControlFlowTok{for}\NormalTok{ (j }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{) \{  }\CommentTok{# number of subjects}
        \CommentTok{# compound symmetry errors}
\NormalTok{        errors[, i, j] <-}\StringTok{ }\KeywordTok{rmvn}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{length}\NormalTok{(x)), }\FloatTok{0.1} \OperatorTok{*}\StringTok{ }\KeywordTok{diag}\NormalTok{(}\DecValTok{6}\NormalTok{) }\OperatorTok{+}\StringTok{ }\FloatTok{0.25} \OperatorTok{*}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{6}\NormalTok{))}
\NormalTok{        y[, i, j] <-}\StringTok{ }\NormalTok{mu[, i] }\OperatorTok{+}\StringTok{ }\NormalTok{errors[, i, j]}
\NormalTok{      \}}
\NormalTok{    \}}
\NormalTok{  \}    }

  \CommentTok{## subject random effects}

  \CommentTok{## visualizing the difference between independent errors and compound symmetry}
  \CommentTok{## why do we need to account for this -- overly confident inference}

  \KeywordTok{dimnames}\NormalTok{(y) <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{time =}\NormalTok{ x, }\DataTypeTok{treatment =} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{, }\DataTypeTok{subject =} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{)}
  \KeywordTok{dimnames}\NormalTok{(errors) <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{time =}\NormalTok{ x, }\DataTypeTok{treatment =} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{, }\DataTypeTok{subject =} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{)}
  \KeywordTok{dimnames}\NormalTok{(mu) <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{time =}\NormalTok{ x, }\DataTypeTok{treatment =} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{)}
\NormalTok{  dat <-}\StringTok{ }\KeywordTok{as.data.frame.table}\NormalTok{(y, }\DataTypeTok{responseName =} \StringTok{"y"}\NormalTok{)}
\NormalTok{  dat_errors <-}\StringTok{ }\KeywordTok{as.data.frame.table}\NormalTok{(errors, }\DataTypeTok{responseName =} \StringTok{"errors"}\NormalTok{)}
\NormalTok{  dat_mu <-}\StringTok{ }\KeywordTok{as.data.frame.table}\NormalTok{(mu, }\DataTypeTok{responseName =} \StringTok{"mu"}\NormalTok{)}
\NormalTok{  dat <-}\StringTok{ }\KeywordTok{left_join}\NormalTok{(dat, dat_errors, }\DataTypeTok{by =} \KeywordTok{c}\NormalTok{(}\StringTok{"time"}\NormalTok{, }\StringTok{"treatment"}\NormalTok{, }\StringTok{"subject"}\NormalTok{))}
\NormalTok{  dat <-}\StringTok{ }\KeywordTok{left_join}\NormalTok{(dat, dat_mu, }\DataTypeTok{by =} \KeywordTok{c}\NormalTok{(}\StringTok{"time"}\NormalTok{, }\StringTok{"treatment"}\NormalTok{))}
\NormalTok{  dat}\OperatorTok{$}\NormalTok{time <-}\StringTok{ }\KeywordTok{as.numeric}\NormalTok{(}\KeywordTok{as.character}\NormalTok{(dat}\OperatorTok{$}\NormalTok{time))}
\NormalTok{  dat <-}\StringTok{ }\NormalTok{dat }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{subject =} \KeywordTok{factor}\NormalTok{(}\KeywordTok{paste}\NormalTok{(subject, treatment, }\DataTypeTok{sep =} \StringTok{"-"}\NormalTok{)))}

  \CommentTok{## repeated measures ANOVA in R}
\NormalTok{  fit_lm <-}\StringTok{ }\KeywordTok{lm}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{time }\OperatorTok{+}\StringTok{ }\NormalTok{treatment }\OperatorTok{+}\StringTok{ }\NormalTok{time }\OperatorTok{*}\StringTok{ }\NormalTok{treatment, }\DataTypeTok{data =}\NormalTok{ dat)}
\NormalTok{  dat}\OperatorTok{$}\NormalTok{preds_lm <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(fit_lm)}

\NormalTok{  fit_lme <-}\StringTok{ }\KeywordTok{lme}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{treatment }\OperatorTok{+}\StringTok{ }\NormalTok{time }\OperatorTok{+}\StringTok{ }\NormalTok{treatment}\OperatorTok{:}\NormalTok{time,}
                 \DataTypeTok{data =}\NormalTok{ dat,}
                 \DataTypeTok{random =} \OperatorTok{~}\StringTok{ }\DecValTok{1} \OperatorTok{|}\StringTok{ }\NormalTok{subject,}
                 \DataTypeTok{correlation =} \KeywordTok{corCompSymm}\NormalTok{(}\DataTypeTok{form =} \OperatorTok{~}\StringTok{ }\DecValTok{1} \OperatorTok{|}\StringTok{ }\NormalTok{subject)}
\NormalTok{  )}

\NormalTok{  pred_dat <-}\StringTok{ }\KeywordTok{expand.grid}\NormalTok{(}
    \DataTypeTok{treatment =} \KeywordTok{factor}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{), }
    \DataTypeTok{time =} \KeywordTok{unique}\NormalTok{(dat}\OperatorTok{$}\NormalTok{time)}
\NormalTok{  )}

\NormalTok{  dat}\OperatorTok{$}\NormalTok{y_pred <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(fit_lme)}

  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}
    \DataTypeTok{dat =}\NormalTok{ dat,}
    \DataTypeTok{pred_dat =}\NormalTok{ pred_dat,}
    \DataTypeTok{fit_lm =}\NormalTok{ fit_lm,}
    \DataTypeTok{fit_lme =}\NormalTok{ fit_lme}

\NormalTok{  ))}
\NormalTok{\}}

\NormalTok{plot_example <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(sim_dat) \{}
  \KeywordTok{library}\NormalTok{(patchwork)}
  \CommentTok{## Plot the simulated data}
\NormalTok{  p1 <-}\StringTok{ }\NormalTok{sim_dat}\OperatorTok{$}\NormalTok{dat }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ y, }\DataTypeTok{group =}\NormalTok{ treatment, }\DataTypeTok{color =}\NormalTok{ treatment)) }\OperatorTok{+}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{show.legend=}\OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'response'}\NormalTok{)}\OperatorTok{+}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ mu, }\DataTypeTok{color =}\NormalTok{ treatment),}\DataTypeTok{show.legend=}\OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
\StringTok{    }\KeywordTok{theme_classic}\NormalTok{() }\OperatorTok{+}
\StringTok{    }\CommentTok{#ggtitle("Simulated data with true response function")+}
\StringTok{    }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{plot.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{30}\NormalTok{, }
                                  \DataTypeTok{face =} \StringTok{"bold"}\NormalTok{),}
        \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{30}\NormalTok{))}\OperatorTok{+}
\StringTok{    }\KeywordTok{scale_color_aaas}\NormalTok{()}

\NormalTok{  p2 <-}\StringTok{ }\NormalTok{sim_dat}\OperatorTok{$}\NormalTok{dat }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ y, }\DataTypeTok{group =}\NormalTok{ subject, }\DataTypeTok{color =}\NormalTok{ treatment)) }\OperatorTok{+}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{size =} \StringTok{"Subjects"}\NormalTok{),}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
\StringTok{    }\CommentTok{# facet_wrap(~ treatment) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ mu, }\DataTypeTok{color =}\NormalTok{ treatment, }\DataTypeTok{size =} \StringTok{"Simulated Truth"}\NormalTok{), }\DataTypeTok{lty =} \DecValTok{1}\NormalTok{,}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'response'}\NormalTok{)}\OperatorTok{+}
\StringTok{    }\KeywordTok{scale_size_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Type"}\NormalTok{, }\DataTypeTok{values=}\KeywordTok{c}\NormalTok{(}\StringTok{"Subjects"}\NormalTok{ =}\StringTok{ }\FloatTok{0.5}\NormalTok{, }\StringTok{"Simulated Truth"}\NormalTok{ =}\StringTok{ }\DecValTok{3}\NormalTok{)) }\OperatorTok{+}
\StringTok{    }\CommentTok{#ggtitle("Simulated data\textbackslash{}nIndividual responses with population mean") +}
\StringTok{    }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
\StringTok{     }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{plot.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{30}\NormalTok{, }
                                \DataTypeTok{face =} \StringTok{"bold"}\NormalTok{),}
     \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{30}\NormalTok{))}\OperatorTok{+}
\StringTok{    }\KeywordTok{scale_color_aaas}\NormalTok{()}

\NormalTok{   p3 <-}\StringTok{ }\NormalTok{sim_dat}\OperatorTok{$}\NormalTok{dat }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ errors, }\DataTypeTok{group =}\NormalTok{ subject, }\DataTypeTok{color =}\NormalTok{ treatment)) }\OperatorTok{+}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{show.legend=}\OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'errors'}\NormalTok{)}\OperatorTok{+}
\StringTok{     }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
\StringTok{    }\CommentTok{# facet_wrap(~ treatment) +}
\StringTok{    }\CommentTok{#ggtitle("Simulated errors") +}
\StringTok{     }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{plot.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{30}\NormalTok{, }
                                  \DataTypeTok{face =} \StringTok{"bold"}\NormalTok{),}
        \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{30}\NormalTok{))}\OperatorTok{+}
\StringTok{    }\KeywordTok{scale_color_aaas}\NormalTok{()}

\NormalTok{  p4 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(sim_dat}\OperatorTok{$}\NormalTok{dat, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ y, }\DataTypeTok{color =}\NormalTok{ treatment)) }\OperatorTok{+}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{()}\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'response'}\NormalTok{)}\OperatorTok{+}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \KeywordTok{predict}\NormalTok{(sim_dat}\OperatorTok{$}\NormalTok{fit_lme), }\DataTypeTok{group =}\NormalTok{ subject, }\DataTypeTok{size =} \StringTok{"Subjects"}\NormalTok{)) }\OperatorTok{+}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sim_dat}\OperatorTok{$}\NormalTok{pred_dat, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \KeywordTok{predict}\NormalTok{(sim_dat}\OperatorTok{$}\NormalTok{fit_lme, }\DataTypeTok{level =} \DecValTok{0}\NormalTok{, }\DataTypeTok{newdata =}\NormalTok{ sim_dat}\OperatorTok{$}\NormalTok{pred_dat), }\DataTypeTok{size =} \StringTok{"Population"}\NormalTok{)) }\OperatorTok{+}
\StringTok{    }\KeywordTok{scale_size_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Predictions"}\NormalTok{, }\DataTypeTok{values=}\KeywordTok{c}\NormalTok{(}\StringTok{"Subjects"}\NormalTok{ =}\StringTok{ }\FloatTok{0.5}\NormalTok{, }\StringTok{"Population"}\NormalTok{ =}\StringTok{ }\DecValTok{3}\NormalTok{)) }\OperatorTok{+}
\StringTok{    }\KeywordTok{theme_classic}\NormalTok{() }\OperatorTok{+}
\StringTok{    }\CommentTok{#ggtitle("Fitted Model")+}
\StringTok{    }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{plot.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{30}\NormalTok{, }
                                  \DataTypeTok{face =} \StringTok{"bold"}\NormalTok{),}
        \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{30}\NormalTok{))}\OperatorTok{+}
\StringTok{    }\KeywordTok{scale_color_aaas}\NormalTok{()}

  \KeywordTok{return}\NormalTok{((p1}\OperatorTok{+}\NormalTok{p3}\OperatorTok{+}\NormalTok{p2}\OperatorTok{+}\NormalTok{p4)}\OperatorTok{+}\KeywordTok{plot_layout}\NormalTok{(}\DataTypeTok{nrow=}\DecValTok{1}\NormalTok{)}\OperatorTok{+}\KeywordTok{plot_annotation}\NormalTok{(}\DataTypeTok{tag_levels =} \StringTok{'A'}\NormalTok{)) }

\NormalTok{\}}

\NormalTok{txt<-}\DecValTok{18}
\NormalTok{A1<-}\KeywordTok{plot_example}\NormalTok{(}\KeywordTok{example}\NormalTok{(}\DataTypeTok{fun_type =} \StringTok{"linear"}\NormalTok{, }\DataTypeTok{error_type =} \StringTok{"correlated"}\NormalTok{)) }

\NormalTok{B1<-}\KeywordTok{plot_example}\NormalTok{(}\KeywordTok{example}\NormalTok{(}\DataTypeTok{fun_type =} \StringTok{"linear"}\NormalTok{, }\DataTypeTok{error_type =} \StringTok{"independent"}\NormalTok{)) }

\NormalTok{C1<-}\KeywordTok{plot_example}\NormalTok{(}\KeywordTok{example}\NormalTok{(}\DataTypeTok{fun_type =} \StringTok{"quadratic"}\NormalTok{, }\DataTypeTok{error_type =} \StringTok{"correlated"}\NormalTok{)) }

\NormalTok{D1<-}\KeywordTok{plot_example}\NormalTok{(}\KeywordTok{example}\NormalTok{(}\DataTypeTok{fun_type =} \StringTok{"quadratic"}\NormalTok{, }\DataTypeTok{error_type =} \StringTok{"independent"}\NormalTok{)) }
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{Manuscript_AM_v4_files/figure-latex/linear-cases-1.pdf}
\caption{\label{fig:linear-cases}\textbf{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model. A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data. B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic). Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data. The rm-ANOVA model does not pick the trend of the quadratic data}}
\end{figure}

For the quadratic response case, Figure \ref{fig:quadratic-cases} shows the simulated responses using compound symmetry and independent errors.

\begin{figure}
\centering
\includegraphics{Manuscript_AM_v4_files/figure-latex/quadratic-cases-1.pdf}
\caption{\label{fig:quadratic-cases}\textbf{Simulated quadratic responses from two groups with a rm-ANOVA model fitted. A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data. B,F: Generated errors showing the difference in the behavior of correlated and independent errors. C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline. D,H: Estimations from the rm-ANOVA model for the mean group response. Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.}}
\end{figure}

\end{document}

### 2021-03-17 Manuscripts/Manuscript v4/Manuscript_AM_v4.Rmd
Need to work on the position of the basis figure and the appendix

@@
 output:
+  bookdown::html_document2: default
   bookdown::pdf_document2:
@@
     extra_dependencies: ["subfig","breqn"]
-  bookdown::html_document2: default
   bookdown::word_document2:
@@
 # References
-
-<div id="refs"></div>
-
-# (APPENDIX) Appendix {-}
-
-# Simulation
-
-## Compound symmetry and independent errors in linear and quadratic responses
-
-This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation).
-The linear simulations using Figure \@ref(fig:linear-cases) show in  panels A and D the simulated mean responses and individual datapoints.
-Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-```{r,include=TRUE,message=FALSE,echo=TRUE}
-## Example with linear response
-example <- function(n_time = 6, 
-                    fun_type = "linear", 
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  library(tidyverse)
-  library(mvnfast)
-  library(nlme)
-  library(ggsci)
-
-  x <- seq(1,6, length.out = n_time)
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # nonlinear response
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-  # matplot(mu, type = 'l')
-
-  y <- array(0, dim = c(length(x), 2, 10))
-  errors <- array(0, dim = c(length(x), 2, 10))
-
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(mu) <- list(time = x, treatment = 1:2)
-  dat <- as.data.frame.table(y, responseName = "y")
-  dat_errors <- as.data.frame.table(errors, responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, responseName = "mu")
-  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
-  dat$time <- as.numeric(as.character(dat$time))
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  ## repeated measures ANOVA in R
-  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
-  dat$preds_lm <- predict(fit_lm)
-
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  dat$y_pred <- predict(fit_lme)
-
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_lm = fit_lm,
-    fit_lme = fit_lme
-
-  ))
-}
-
-plot_example <- function(sim_dat) {
-  library(patchwork)
-  ## Plot the simulated data
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE) +labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
-    theme_classic() +
-    #ggtitle("Simulated data with true response function")+
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    #ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                face = "bold"),
-     text=element_text(size=30))+
-    scale_color_aaas()
-
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='errors')+
-     theme_classic()+
-    # facet_wrap(~ treatment) +
-    #ggtitle("Simulated errors") +
-     theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point()+labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    #ggtitle("Fitted Model")+
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
-
-```
-
-```{r, linear-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data**"}
-# linear response, correlated errors (subject effect)
-A1/B1+plot_annotation(tag_levels = 'A')
-```
-
-For the quadratic response case, Figure \@ref(fig:quadratic-cases) shows the simulated responses using compound symmetry and independent errors.
-
-```{r, quadratic-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
-# linear response, correlated errors (subject effect)
-C1/D1+plot_annotation(tag_levels = 'A')
-```

### 2021-03-17 Manuscripts/Manuscript v4/Manuscript_AM_v4.tex
Need to work on the position of the basis figure and the appendix

@@
 \usepackage[margin=1in]{geometry}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{longtable,booktabs}
@@
 \hypertarget{references}{%
-\section{References}\label{references}}
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
 
@@
 
-\hypertarget{appendix-appendix}{%
-\appendix}
-
-\hypertarget{simulation-1}{%
-\section{Simulation}\label{simulation-1}}
-
-\hypertarget{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}{%
-\subsection{Compound symmetry and independent errors in linear and quadratic responses}\label{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}}
-
-This section simulated linear and quadratic data in the same manner as in Section \ref{simulation}.
-The linear simulations using Figure \ref{fig:linear-cases} show in panels A and D the simulated mean responses and individual datapoints.
-Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## Example with linear response}
-\NormalTok{example <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{n_time =} \DecValTok{6}\NormalTok{, }
-                    \DataTypeTok{fun_type =} \StringTok{"linear"}\NormalTok{, }
-                    \DataTypeTok{error_type =} \StringTok{"correlated"}\NormalTok{) \{}
-
-  \ControlFlowTok{if}\NormalTok{ (}\OperatorTok{!}\NormalTok{(fun_type }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"linear"}\NormalTok{, }\StringTok{"quadratic"}\NormalTok{)))}
-    \KeywordTok{stop}\NormalTok{(}\StringTok{'fun_type must be either "linear", or "quadratic"'}\NormalTok{)}
-  \ControlFlowTok{if}\NormalTok{ (}\OperatorTok{!}\NormalTok{(error_type }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"correlated"}\NormalTok{, }\StringTok{"independent"}\NormalTok{)))}
-    \KeywordTok{stop}\NormalTok{(}\StringTok{'fun_type must be either "correlated", or "independent"'}\NormalTok{)}
-
-  \KeywordTok{library}\NormalTok{(tidyverse)}
-  \KeywordTok{library}\NormalTok{(mvnfast)}
-  \KeywordTok{library}\NormalTok{(nlme)}
-  \KeywordTok{library}\NormalTok{(ggsci)}
-
-\NormalTok{  x <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{6}\NormalTok{, }\DataTypeTok{length.out =}\NormalTok{ n_time)}
-\NormalTok{  mu <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{length}\NormalTok{(x), }\DecValTok{2}\NormalTok{)}
-  \CommentTok{# linear response}
-  \ControlFlowTok{if}\NormalTok{ (fun_type }\OperatorTok{==}\StringTok{ "linear"}\NormalTok{) \{}
-\NormalTok{    mu[, }\DecValTok{1}\NormalTok{] <-}\StringTok{ }\OperatorTok{-}\StringTok{ }\NormalTok{(}\FloatTok{0.25}\OperatorTok{*}\NormalTok{x)}\OperatorTok{+}\DecValTok{2}  
-\NormalTok{    mu[, }\DecValTok{2}\NormalTok{] <-}\StringTok{ }\FloatTok{0.25}\OperatorTok{*}\NormalTok{x}\OperatorTok{+}\DecValTok{2}
-\NormalTok{  \} }\ControlFlowTok{else}\NormalTok{ \{}
-    \CommentTok{# nonlinear response}
-
-\NormalTok{    mu[, }\DecValTok{1}\NormalTok{] <-}\StringTok{  }\OperatorTok{-}\NormalTok{(}\FloatTok{0.25} \OperatorTok{*}\StringTok{ }\NormalTok{x}\OperatorTok{^}\DecValTok{2}\NormalTok{) }\FloatTok{+1.5}\OperatorTok{*}\NormalTok{x}\FloatTok{-1.25}
-\NormalTok{    mu[, }\DecValTok{2}\NormalTok{] <-}\StringTok{ }\NormalTok{(}\FloatTok{0.25} \OperatorTok{*}\StringTok{ }\NormalTok{x}\OperatorTok{^}\DecValTok{2}\NormalTok{) }\FloatTok{-1.5}\OperatorTok{*}\NormalTok{x}\FloatTok{+1.25}
-\NormalTok{  \}}
-  \CommentTok{# matplot(mu, type = 'l')}
-
-\NormalTok{  y <-}\StringTok{ }\KeywordTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DataTypeTok{dim =} \KeywordTok{c}\NormalTok{(}\KeywordTok{length}\NormalTok{(x), }\DecValTok{2}\NormalTok{, }\DecValTok{10}\NormalTok{))}
-\NormalTok{  errors <-}\StringTok{ }\KeywordTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DataTypeTok{dim =} \KeywordTok{c}\NormalTok{(}\KeywordTok{length}\NormalTok{(x), }\DecValTok{2}\NormalTok{, }\DecValTok{10}\NormalTok{))}
-
-  \ControlFlowTok{if}\NormalTok{ (error_type }\OperatorTok{==}\StringTok{ "independent"}\NormalTok{) \{}
-    \CommentTok{## independent errors}
-    \ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{) \{}
-      \ControlFlowTok{for}\NormalTok{ (j }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{) \{}
-\NormalTok{        errors[, i, j] <-}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(}\DecValTok{6}\NormalTok{, }\DecValTok{0}\NormalTok{, }\FloatTok{0.25}\NormalTok{)}
-\NormalTok{        y[, i, j] <-}\StringTok{ }\NormalTok{mu[, i] }\OperatorTok{+}\StringTok{ }\NormalTok{errors[, i, j]}
-\NormalTok{      \}}
-\NormalTok{    \}}
-\NormalTok{  \} }\ControlFlowTok{else}\NormalTok{ \{}
-    \ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{) \{     }\CommentTok{# number of treatments}
-      \ControlFlowTok{for}\NormalTok{ (j }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{) \{  }\CommentTok{# number of subjects}
-        \CommentTok{# compound symmetry errors}
-\NormalTok{        errors[, i, j] <-}\StringTok{ }\KeywordTok{rmvn}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{length}\NormalTok{(x)), }\FloatTok{0.1} \OperatorTok{*}\StringTok{ }\KeywordTok{diag}\NormalTok{(}\DecValTok{6}\NormalTok{) }\OperatorTok{+}\StringTok{ }\FloatTok{0.25} \OperatorTok{*}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{6}\NormalTok{))}
-\NormalTok{        y[, i, j] <-}\StringTok{ }\NormalTok{mu[, i] }\OperatorTok{+}\StringTok{ }\NormalTok{errors[, i, j]}
-\NormalTok{      \}}
-\NormalTok{    \}}
-\NormalTok{  \}    }
-
-  \CommentTok{## subject random effects}
-
-  \CommentTok{## visualizing the difference between independent errors and compound symmetry}
-  \CommentTok{## why do we need to account for this -- overly confident inference}
-
-  \KeywordTok{dimnames}\NormalTok{(y) <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{time =}\NormalTok{ x, }\DataTypeTok{treatment =} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{, }\DataTypeTok{subject =} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{)}
-  \KeywordTok{dimnames}\NormalTok{(errors) <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{time =}\NormalTok{ x, }\DataTypeTok{treatment =} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{, }\DataTypeTok{subject =} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{)}
-  \KeywordTok{dimnames}\NormalTok{(mu) <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{time =}\NormalTok{ x, }\DataTypeTok{treatment =} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{)}
-\NormalTok{  dat <-}\StringTok{ }\KeywordTok{as.data.frame.table}\NormalTok{(y, }\DataTypeTok{responseName =} \StringTok{"y"}\NormalTok{)}
-\NormalTok{  dat_errors <-}\StringTok{ }\KeywordTok{as.data.frame.table}\NormalTok{(errors, }\DataTypeTok{responseName =} \StringTok{"errors"}\NormalTok{)}
-\NormalTok{  dat_mu <-}\StringTok{ }\KeywordTok{as.data.frame.table}\NormalTok{(mu, }\DataTypeTok{responseName =} \StringTok{"mu"}\NormalTok{)}
-\NormalTok{  dat <-}\StringTok{ }\KeywordTok{left_join}\NormalTok{(dat, dat_errors, }\DataTypeTok{by =} \KeywordTok{c}\NormalTok{(}\StringTok{"time"}\NormalTok{, }\StringTok{"treatment"}\NormalTok{, }\StringTok{"subject"}\NormalTok{))}
-\NormalTok{  dat <-}\StringTok{ }\KeywordTok{left_join}\NormalTok{(dat, dat_mu, }\DataTypeTok{by =} \KeywordTok{c}\NormalTok{(}\StringTok{"time"}\NormalTok{, }\StringTok{"treatment"}\NormalTok{))}
-\NormalTok{  dat}\OperatorTok{$}\NormalTok{time <-}\StringTok{ }\KeywordTok{as.numeric}\NormalTok{(}\KeywordTok{as.character}\NormalTok{(dat}\OperatorTok{$}\NormalTok{time))}
-\NormalTok{  dat <-}\StringTok{ }\NormalTok{dat }\OperatorTok{%>%}
-\StringTok{    }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{subject =} \KeywordTok{factor}\NormalTok{(}\KeywordTok{paste}\NormalTok{(subject, treatment, }\DataTypeTok{sep =} \StringTok{"-"}\NormalTok{)))}
-
-  \CommentTok{## repeated measures ANOVA in R}
-\NormalTok{  fit_lm <-}\StringTok{ }\KeywordTok{lm}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{time }\OperatorTok{+}\StringTok{ }\NormalTok{treatment }\OperatorTok{+}\StringTok{ }\NormalTok{time }\OperatorTok{*}\StringTok{ }\NormalTok{treatment, }\DataTypeTok{data =}\NormalTok{ dat)}
-\NormalTok{  dat}\OperatorTok{$}\NormalTok{preds_lm <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(fit_lm)}
-
-\NormalTok{  fit_lme <-}\StringTok{ }\KeywordTok{lme}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{treatment }\OperatorTok{+}\StringTok{ }\NormalTok{time }\OperatorTok{+}\StringTok{ }\NormalTok{treatment}\OperatorTok{:}\NormalTok{time,}
-                 \DataTypeTok{data =}\NormalTok{ dat,}
-                 \DataTypeTok{random =} \OperatorTok{~}\StringTok{ }\DecValTok{1} \OperatorTok{|}\StringTok{ }\NormalTok{subject,}
-                 \DataTypeTok{correlation =} \KeywordTok{corCompSymm}\NormalTok{(}\DataTypeTok{form =} \OperatorTok{~}\StringTok{ }\DecValTok{1} \OperatorTok{|}\StringTok{ }\NormalTok{subject)}
-\NormalTok{  )}
-
-\NormalTok{  pred_dat <-}\StringTok{ }\KeywordTok{expand.grid}\NormalTok{(}
-    \DataTypeTok{treatment =} \KeywordTok{factor}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{), }
-    \DataTypeTok{time =} \KeywordTok{unique}\NormalTok{(dat}\OperatorTok{$}\NormalTok{time)}
-\NormalTok{  )}
-
-\NormalTok{  dat}\OperatorTok{$}\NormalTok{y_pred <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(fit_lme)}
-
-  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}
-    \DataTypeTok{dat =}\NormalTok{ dat,}
-    \DataTypeTok{pred_dat =}\NormalTok{ pred_dat,}
-    \DataTypeTok{fit_lm =}\NormalTok{ fit_lm,}
-    \DataTypeTok{fit_lme =}\NormalTok{ fit_lme}
-
-\NormalTok{  ))}
-\NormalTok{\}}
-
-\NormalTok{plot_example <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(sim_dat) \{}
-  \KeywordTok{library}\NormalTok{(patchwork)}
-  \CommentTok{## Plot the simulated data}
-\NormalTok{  p1 <-}\StringTok{ }\NormalTok{sim_dat}\OperatorTok{$}\NormalTok{dat }\OperatorTok{%>%}
-\StringTok{    }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ y, }\DataTypeTok{group =}\NormalTok{ treatment, }\DataTypeTok{color =}\NormalTok{ treatment)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{show.legend=}\OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'response'}\NormalTok{)}\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ mu, }\DataTypeTok{color =}\NormalTok{ treatment),}\DataTypeTok{show.legend=}\OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{theme_classic}\NormalTok{() }\OperatorTok{+}
-\StringTok{    }\CommentTok{#ggtitle("Simulated data with true response function")+}
-\StringTok{    }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{plot.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{30}\NormalTok{, }
-                                  \DataTypeTok{face =} \StringTok{"bold"}\NormalTok{),}
-        \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{30}\NormalTok{))}\OperatorTok{+}
-\StringTok{    }\KeywordTok{scale_color_aaas}\NormalTok{()}
-
-\NormalTok{  p2 <-}\StringTok{ }\NormalTok{sim_dat}\OperatorTok{$}\NormalTok{dat }\OperatorTok{%>%}
-\StringTok{    }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ y, }\DataTypeTok{group =}\NormalTok{ subject, }\DataTypeTok{color =}\NormalTok{ treatment)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{size =} \StringTok{"Subjects"}\NormalTok{),}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
-\StringTok{    }\CommentTok{# facet_wrap(~ treatment) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ mu, }\DataTypeTok{color =}\NormalTok{ treatment, }\DataTypeTok{size =} \StringTok{"Simulated Truth"}\NormalTok{), }\DataTypeTok{lty =} \DecValTok{1}\NormalTok{,}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'response'}\NormalTok{)}\OperatorTok{+}
-\StringTok{    }\KeywordTok{scale_size_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Type"}\NormalTok{, }\DataTypeTok{values=}\KeywordTok{c}\NormalTok{(}\StringTok{"Subjects"}\NormalTok{ =}\StringTok{ }\FloatTok{0.5}\NormalTok{, }\StringTok{"Simulated Truth"}\NormalTok{ =}\StringTok{ }\DecValTok{3}\NormalTok{)) }\OperatorTok{+}
-\StringTok{    }\CommentTok{#ggtitle("Simulated data\textbackslash{}nIndividual responses with population mean") +}
-\StringTok{    }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
-\StringTok{     }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{plot.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{30}\NormalTok{, }
-                                \DataTypeTok{face =} \StringTok{"bold"}\NormalTok{),}
-     \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{30}\NormalTok{))}\OperatorTok{+}
-\StringTok{    }\KeywordTok{scale_color_aaas}\NormalTok{()}
-
-\NormalTok{   p3 <-}\StringTok{ }\NormalTok{sim_dat}\OperatorTok{$}\NormalTok{dat }\OperatorTok{%>%}
-\StringTok{    }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ errors, }\DataTypeTok{group =}\NormalTok{ subject, }\DataTypeTok{color =}\NormalTok{ treatment)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{show.legend=}\OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'errors'}\NormalTok{)}\OperatorTok{+}
-\StringTok{     }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
-\StringTok{    }\CommentTok{# facet_wrap(~ treatment) +}
-\StringTok{    }\CommentTok{#ggtitle("Simulated errors") +}
-\StringTok{     }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{plot.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{30}\NormalTok{, }
-                                  \DataTypeTok{face =} \StringTok{"bold"}\NormalTok{),}
-        \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{30}\NormalTok{))}\OperatorTok{+}
-\StringTok{    }\KeywordTok{scale_color_aaas}\NormalTok{()}
-
-\NormalTok{  p4 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(sim_dat}\OperatorTok{$}\NormalTok{dat, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ y, }\DataTypeTok{color =}\NormalTok{ treatment)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_point}\NormalTok{()}\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'response'}\NormalTok{)}\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \KeywordTok{predict}\NormalTok{(sim_dat}\OperatorTok{$}\NormalTok{fit_lme), }\DataTypeTok{group =}\NormalTok{ subject, }\DataTypeTok{size =} \StringTok{"Subjects"}\NormalTok{)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sim_dat}\OperatorTok{$}\NormalTok{pred_dat, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \KeywordTok{predict}\NormalTok{(sim_dat}\OperatorTok{$}\NormalTok{fit_lme, }\DataTypeTok{level =} \DecValTok{0}\NormalTok{, }\DataTypeTok{newdata =}\NormalTok{ sim_dat}\OperatorTok{$}\NormalTok{pred_dat), }\DataTypeTok{size =} \StringTok{"Population"}\NormalTok{)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{scale_size_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Predictions"}\NormalTok{, }\DataTypeTok{values=}\KeywordTok{c}\NormalTok{(}\StringTok{"Subjects"}\NormalTok{ =}\StringTok{ }\FloatTok{0.5}\NormalTok{, }\StringTok{"Population"}\NormalTok{ =}\StringTok{ }\DecValTok{3}\NormalTok{)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{theme_classic}\NormalTok{() }\OperatorTok{+}
-\StringTok{    }\CommentTok{#ggtitle("Fitted Model")+}
-\StringTok{    }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{plot.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{30}\NormalTok{, }
-                                  \DataTypeTok{face =} \StringTok{"bold"}\NormalTok{),}
-        \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{30}\NormalTok{))}\OperatorTok{+}
-\StringTok{    }\KeywordTok{scale_color_aaas}\NormalTok{()}
-
-  \KeywordTok{return}\NormalTok{((p1}\OperatorTok{+}\NormalTok{p3}\OperatorTok{+}\NormalTok{p2}\OperatorTok{+}\NormalTok{p4)}\OperatorTok{+}\KeywordTok{plot_layout}\NormalTok{(}\DataTypeTok{nrow=}\DecValTok{1}\NormalTok{)}\OperatorTok{+}\KeywordTok{plot_annotation}\NormalTok{(}\DataTypeTok{tag_levels =} \StringTok{'A'}\NormalTok{)) }
-
-\NormalTok{\}}
-
-\NormalTok{txt<-}\DecValTok{18}
-\NormalTok{A1<-}\KeywordTok{plot_example}\NormalTok{(}\KeywordTok{example}\NormalTok{(}\DataTypeTok{fun_type =} \StringTok{"linear"}\NormalTok{, }\DataTypeTok{error_type =} \StringTok{"correlated"}\NormalTok{)) }
-
-\NormalTok{B1<-}\KeywordTok{plot_example}\NormalTok{(}\KeywordTok{example}\NormalTok{(}\DataTypeTok{fun_type =} \StringTok{"linear"}\NormalTok{, }\DataTypeTok{error_type =} \StringTok{"independent"}\NormalTok{)) }
-
-\NormalTok{C1<-}\KeywordTok{plot_example}\NormalTok{(}\KeywordTok{example}\NormalTok{(}\DataTypeTok{fun_type =} \StringTok{"quadratic"}\NormalTok{, }\DataTypeTok{error_type =} \StringTok{"correlated"}\NormalTok{)) }
-
-\NormalTok{D1<-}\KeywordTok{plot_example}\NormalTok{(}\KeywordTok{example}\NormalTok{(}\DataTypeTok{fun_type =} \StringTok{"quadratic"}\NormalTok{, }\DataTypeTok{error_type =} \StringTok{"independent"}\NormalTok{)) }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}
-\centering
-\includegraphics{Manuscript_AM_v4_files/figure-latex/linear-cases-1.pdf}
-\caption{\label{fig:linear-cases}\textbf{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data}}
-\end{figure}
-
-For the quadratic response case, Figure \ref{fig:quadratic-cases} shows the simulated responses using compound symmetry and independent errors.
-
-\begin{figure}
-\centering
-\includegraphics{Manuscript_AM_v4_files/figure-latex/quadratic-cases-1.pdf}
-\caption{\label{fig:quadratic-cases}\textbf{Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.}}
-\end{figure}
-
 \end{document}

### 2021-03-18 Manuscripts/Manuscript v4/Manuscript_AM_v4.Rmd
Figure positions work now, need to adjust the font size for "Intercept" in basis plot

@@
     fig_caption: yes
-    extra_dependencies: ["subfig","breqn"]
+    extra_dependencies: 
+      subfig: null
+      breqn: null
+      caption: ["font={small}"]
+      float: null
+  bookdown::html_document2: default
   bookdown::word_document2:
@@
 
-## Simulation experiment on linearity and covariance structures {#simulation}
+## How does an rm-ANOVA fit looks like? A visual representation using simulation {#simulation}
 
@@
 library(tidyverse)
+knitr::opts_chunk$set(fig.pos = "!h", out.extra = "")
 ```
@@
 
-```{r, l-q-response, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap = "Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
 A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
@@
 Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data**"}
+The rm-ANOVA model does not pick the trend of the quadratic data."}
 # linear response, correlated errors (subject effect)
+par(mar = c(0.5, 0.5, 0.5, 0.5))
 A/C+plot_annotation(tag_levels = 'A')
@@
 To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure @\ref(fig:l-q-response), C.
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits the trend of the group trough the timeline.
+The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group trough the timeline.
 The timeline can be divided equally spaced _knots_, each knot being a region where a different basis function will be used.
@@
 To control the wigliness of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
-
-In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \@ref(fig:basis-plot), Panel C shows the resulting penalized basis functions.
-Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
-Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
 
@@
 
-```{r,basis-plot,fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for _time_ for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
+In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
+In Figure \@ref(fig:basis-plot), Panel C shows the resulting penalized basis functions.
+Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
+Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
+The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
+
+```{r,basis-plot,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap="Basis functions for a single smoother for time with five knots.
+A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
 B: Penalty matrix for the basis functions.
@@
 D.
-Smoother for _time_ and original datapoints.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.**"}
+Smoother for time and original datapoints.
+The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points."}
+
+par(mar = c(2, 2, 2, 2))
 
@@
 
-Although the specific methods of how GAMs model covariance and correlation is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
+Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
 A detailed description on basis functions and correlations can be found in [@hefley2017].
@@
 
-At the core of a biomedical longitudinal study lies the question of a significant difference between two or more treatments in different groups.
-Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_, GAMs use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs by group is that if the _difference_ between the confidence intervals of the smooths is not equal to zero then a significant difference exists at that timepoint (or timepoints).
+At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
+Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs use a different approach to estimate significance.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that timepoint (or timepoints).
 The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
-The estimation of siginificant differences in GAMs is covered in detail in the Appendix.
-
-The use of GAMs to analyze biomedical longitudinal data, and the impact of missing observations in the fit of the model will be examined in detail in the following section using simulation.
+The major advantage of the significance estimation in GAMs  is that they allow to determine significance at a specific timepoint, rather than providing a single _p-value_ for the overall treatment effect.
+The estimation of siginificant differences in GAMs is covered in detail in the Appendix using simulated longitudinal data that follows previously reported trends of tumor response between multiple treatment groups [@vishwanath2009].
 

### 2021-03-18 Manuscripts/Manuscript v4/Manuscript_AM_v4.tex
Figure positions work now, need to adjust the font size for "Intercept" in basis plot

@@
 \fi
-% Use upquote if available, for straight quotes in verbatim environments
+% use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
   pdfauthor={Ariel Mundo; Timothy J. Muldoon; John R. Tipton},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
+  pdfborder={0 0 0},
+  breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
 \usepackage[margin=1in]{geometry}
 \usepackage{longtable,booktabs}
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
 % Allow footnotes in longtable head/foot
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
@@
 \setcounter{secnumdepth}{5}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
 \usepackage{subfig}
 \usepackage{breqn}
+\usepackage[font={small}]{caption}
+\usepackage{float}
 
@@
 A \emph{post hoc} analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
-Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
+Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless{}0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
 In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}17{]}.
@@
 \hypertarget{simulation}{%
-\subsection{Simulation experiment on linearity and covariance structures}\label{simulation}}
+\subsection{How does an rm-ANOVA fit looks like? A visual representation using simulation}\label{simulation}}
 
@@
 
-\begin{figure}
-\centering
-\includegraphics{Manuscript_AM_v4_files/figure-latex/l-q-response-1.pdf}
-\caption{\label{fig:l-q-response}\textbf{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v4_files/figure-latex/l-q-response-1}
+
+}
+
+\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
 A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
@@
 Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data}}
+The rm-ANOVA model does not pick the trend of the quadratic data.}\label{fig:l-q-response}
 \end{figure}
@@
 To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure @\ref(fig:l-q-response), C.
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits the trend of the group trough the timeline.
+The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group trough the timeline.
 The timeline can be divided equally spaced \emph{knots}, each knot being a region where a different basis function will be used.
@@
 
-\begin{figure}
-\centering
-\includegraphics{Manuscript_AM_v4_files/figure-latex/basis-plot-1.pdf}
-\caption{\label{fig:basis-plot}\textbf{Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for \emph{time} for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v4_files/figure-latex/basis-plot-1}
+
+}
+
+\caption{Basis functions for a single smoother for time with five knots.
+A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
 B: Penalty matrix for the basis functions.
@@
 D.
-Smoother for \emph{time} and original datapoints.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}}
+Smoother for time and original datapoints.
+The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}\label{fig:basis-plot}
 \end{figure}
@@
 
-Although the specific methods of how GAMs model covariance and correlation is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
+Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
 A detailed description on basis functions and correlations can be found in {[}52{]}.
@@
 
-At the core of a biomedical longitudinal study lies the question of a significant difference between two or more treatments in different groups.
-Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values}, GAMs use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs by group is that if the \emph{difference} between the confidence intervals of the smooths is not equal to zero then a significant difference exists at that timepoint (or timepoints).
+At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
+Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs use a different approach to estimate significance.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the \emph{difference} between the confidence intervals of the fitted smooths for such groups is non-zero, then a significant difference exists at that timepoint (or timepoints).
 The absence of a \emph{p-value} in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
-The estimation of siginificant differences in GAMs is covered in detail in the Appendix.
-
-The use of GAMs to analyze biomedical longitudinal data, and the impact of missing observations in the fit of the model will be examined in detail in the following section using simulation.
+The major advantage of the significance estimation in GAMs is that they allow to determine significance at a specific timepoint, rather than providing a single \emph{p-value} for the overall treatment effect.
+The estimation of siginificant differences in GAMs is covered in detail in the Appendix using simulated longitudinal data that follows previously reported trends of tumor response between multiple treatment groups {[}16{]}.
 
@@
 \leavevmode\hypertarget{ref-wood2016}{}%
-{[}39{]} S.N. Wood, N., Pya, B. Säfken, Smoothing parameter and model selection for general smooth models (with discussion), Journal of the American Statistical Association. 111 (2016) 1548--1575.
+{[}39{]} S. Wood, N., Pya, B. Säfken, Smoothing parameter and model selection for general smooth models (with discussion), Journal of the American Statistical Association. 111 (2016) 1548--1575.
 

### 2021-03-18 Manuscripts/Bookdown testing/Appendix.Rmd
Manuscript updated with page breaks

--- 2021-03-17 Manuscripts/Manuscript v3/Manuscript_AM_v3.Rmd
@@
----
-title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
-subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
-author:
-- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
-- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
-  Fayetteville]
-- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
-output:
-  bookdown::html_document2: default
-  bookdown::pdf_document2:
-    keep_tex: yes
-    fig_caption: yes
-    extra_dependencies: ["subfig","breqn"]
-  bookdown::word_document2:
-    fig_caption: yes
-    keep_md: yes
-  link-citations: yes
-  css: style.css
-  '': default
-csl: Elsevier.csl
-bibliography: refs.bib
----
+# (APPENDIX) Appendix {-}
 
-# Background
+# Simulation
 
-Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following RT in neck cancer patients [@kamstra2015].
+## Compound symmetry and independent errors in linear and quadratic responses
 
-Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as a limiting frequency [@wagenmakers2008] by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].
+This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation).
+The linear simulations using Figure \@ref(fig:linear-cases) show in  panels A and D the simulated mean responses and individual datapoints.
+Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
+In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
+Panels D and H show the predictions from the rm-ANOVA model.
 
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA [@pinheiro2006].
-This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model is only able to accommodate linear trends that are far from adequately representing the biological phenomenon of interest.
-
-A _post hoc_ analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
-Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, _p-values_ are valid only if the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019].
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting _p-values_ and parameter estimates to be invalid [@wang2019].
-
-Additionally, the _p-value_ itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019].
-Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to balance in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].
-Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976;@greenhouse1959], but the effectiveness the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001].
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not "large" due to ethical and budgetary reasons [@charan2013] which might cause the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial patients may voluntarily withdraw, whereas attrition  due to injury or weight loss in preclinical animal studies is possible.
-It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios,the _complete observations_ assumption of  classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis [@gueorguieva2004].
-This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
-At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
-
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data [@skala2010;@vishwanath2009].
-Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-However, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
-And even more importantly, LMEMs also expect a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze non-linear data.
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
-
-In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used by the biomedical community, these non-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017] and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
-
-The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, $\textsf{R}$[@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
-hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017].
-Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
-
-This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations is explained in detail.
-Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, using simulated data that reproduces patterns in previously reported studies [@vishwanath2009] we illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
-The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in $\textsf{R}$, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
-
-In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models by improving the standards for reproducibility  in biomedical research.
-
-# Challenges presented by longitudinal studies
-
-## The repeated measures ANOVA
-
-The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research.
-This statistical methodology requires certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
-Each one of these assumptions is discussed below.
-
-## Linear relationship
-
-In a biomedical longitudinal study, two or more groups of subjects (e.g., patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs.
-a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response is modeled with both  _fixed_ and  _random_ components.
-The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug in the "treatment" group.
-The _random_ component can be defined as "noise" caused by some inherent variability within the study.
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
-As their name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
-
-Mathematically speaking, a rm-ANOVA model with an interaction can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{tij}\\ 
-(\#eq:linear-model)
-\end{equation}
-
-In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-The _random effect_ $\mu_{ij}$ is assumed to be independently normally distributed with mean zero and variance $\sigma^2_\mu$, which can be expressed as $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
-This _random effect_ accounts for variability in each subject within each group by allowing different values for each subject.
-Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ or _random_ effects, and are assumed to be $\sim N(0,\sigma^2)$ .
-In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
-"novel drug" or "saline" vs.
-"chemotherapy").
-Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B).
-The linear models then can be expressed as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_1 \times time_{t} + \beta_2 \times +\beta_3 \times time_{t} \times +\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-(\#eq:ANOVA-by-group)
-\end{equation}
-
-To further simplify the expression, substitute $\widetilde{\mu}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B.
-This substitution allows for a different intercept and slope for Groups A and B.
-The model is then written as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\widetilde{\mu} + \widetilde{\beta_1} \times time_{t} +\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-(\#eq:ANOVA-lines)
-\end{equation}
-
-Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
-Because the construction of the model is essentially the same in LMEMs, the linearity assumption holds for these models, although it is possible to fit models with different slopes at the subject-level[@pinheiro2006].
-
-## Covariance in rm-ANOVA and LMEMs
-
-In a longitudinal study there is an expected _covariance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
-The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
-
-In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
-Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)[@huynh1976;@greenhouse1959] the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others [@pinheiro2006].
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself.
-Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
-
-## Missing observations
-
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004].
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002].
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
-
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014].
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.
-
-This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and  missing data.
-Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data.
-To better convey the issues of linearity and correlation  in linear models fitted to non-linear data, simulation is used in the next section.
-
-## Simulation experiment on linearity and covariance structures
-
-To demonstrate the limitations of rm-ANOVA for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix).
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:linear-models, A)).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:quadratic-response),A).
-
-The rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the response indicates a difference in the effect of each treatment.
-From the mean response per group (linear or quadratic), individual responses for each subject within each group are simulated using covariance matrices with compound symmetry (constant variance across time) or independent errors (errors generated from a a normal distribution that are not constant over time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (either compound symmetry or independent, which appear in panels A and E, respectively in both Figure \@ref(fig:linear-models) and Figure \@ref(fig:quadratic-response)) .
-The resulting simulated data, simulated errors, individual response and fitted parameters from the model are plotted.
-
-```{r, message = FALSE,include=FALSE}
-set.seed(11)
-library(patchwork)
-library(tidyverse)
-```
-
-```{r,include=FALSE,message=FALSE,echo=FALSE}
+```{r,include=TRUE,message=FALSE,echo=TRUE}
 ## Example with linear response
@@
   return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-    # return((p1 + p3 + plot_layout(guides = "collect")) / (p2 + p4 + plot_layout(guides = "collect")))
 
@@
 txt<-18
-A<-plot_example(example(fun_type = "linear", error_type = "correlated"))
+A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
 
-B<-plot_example(example(fun_type = "linear", error_type = "independent"))
+B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
 
-C<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
+C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
 
-D<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
+D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
 
@@
 
-```{r, linear-models, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+```{r, linear-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
+B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
+The rm-ANOVA model does not pick the trend of the quadratic data**"}
+# linear response, correlated errors (subject effect)
+A1/B1+plot_annotation(tag_levels = 'A')
+```
+
+For the quadratic response case, Figure \@ref(fig:quadratic-cases) shows the simulated responses using compound symmetry and independent errors.
+
+```{r, quadratic-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
 A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
@@
 # linear response, correlated errors (subject effect)
-A/B+plot_annotation(tag_levels = 'A')
+C1/D1+plot_annotation(tag_levels = 'A')
 ```
-
-In Figure \@ref(fig:linear-models)  panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-It is clear from Figure \@ref(fig:linear-models) that the fit produced by the rm-ANOVA model is good, as the predictions for the mean response are reasonably close to the "truth" of the simulated data.
-
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:quadratic-response).
-Here, the mean response per group was simulated using a quadratic function and errors, individual responses and the rm-ANOVA model were produced in the same manner as \@ref(fig:linear-models) .
-The figure shows that the response is changing in each group through the timeline, and that the mean value is the same as the initial value by the fifth time point.
-Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panels D and H in Figure \@ref(fig:quadratic-response).
-
-```{r, quadratic-response, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
-# linear response, correlated errors (subject effect)
-C/D+plot_annotation(tag_levels='A')
-```
-
-In this case, Figure \@ref(fig:quadratic-response), panel A shows that both groups start with the same mean value at the first time point.
-Again, this is intentional in order to simulate the temporal behavior of some physiological quantity.
-Therefore, the idea behind \@ref(fig:quadratic-response) is to mimic a non-linear behavior from two groups that follow an inverse trajectory.
-A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data (\@ref(fig:quadratic-response), panels D and H)  indicates that the model is not capturing the changes within each group in a good way.
-Specifically, note that in Figure \@ref(fig:quadratic-response) the fitted mean response of the rm-ANOVA model (panels D and H) show that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model and moreover, the model is not being able to capture the initial value (which is the same in each group), instead fitting non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels A and D, and E and H).
-If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response, potentially limiting valuable inferences from the data.
-
-This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-Although the model fitted to the simulated data was an rm_ANOVA, the main issue of an expected linear trend in the response is the same in the case of LMEMs.
-In the following section, we present generalized additive models (GAMs) as an alternative to analyze longitudinal non-linear data.
-
-# GAMs as a special case of Generalized Linear Models
-
-## GAMs and Basis Functions
-
-Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors[@nelder1972].
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case[@simpson2018;@wood2017;@hastie1987].
-A GAM model can be written as:
-
-\begin{equation}
-  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
-  (\#eq:GAM)
-\end{equation}
-
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ and $\varepsilon_{ijt}$ represents the residual error.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
-One possible function for $f(x_t\mid \beta_j)$ that allows for non-linear responses is a polynomial, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
-In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.
-
-The smooth functional relationship between the covariates and the response in GAMs is specified   using a parametric relationship that can be fit within the GLM framework, by using _basis functions_ expansions of the covariates and by estimating random coefficients for these basis functions and this step is achieved by using _basis functions_ to represent them.
-A _basis_ is a set of functions that spans the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
-For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
-In the case of Equation \@ref(eq:GAM), the basis function is $f(x_t\mid \beta_j)$, which means that the model allows relationships beyond linear for the covariates.
-
-A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations[@wood2017;@simpson2018].
-Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness [@wegman1983].
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure @\ref(fig:quadratic-response).
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits the trend of the group trough the timeline.
-The timeline can be divided equally spaced _knots_, each knot being a region where a different basis function will be used.
-Because there are six timepoints for this group, five knots can be used.
-The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots.
-In Panel A of Figure \@ref(fig:basis-functions-plot), the four basis functions (and the intercept) are shown.
-Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the wigliness of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
-
-In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \@ref(fig:basis-functions-plot), Panel C shows the resulting penalized basis functions.
-Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
-Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
-
-```{r,basis-functions-plot,fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for _time_ for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
-D.
-Smoother for _time_ and original datapoints.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.**"}
-
-library(mgcv)
-
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  dat<-subset(dat,treatment==1)
-  dat<-dat[,c('y','time')]
-gm<-gam(y~s(time,k=5),data=dat)
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-x_new <- (seq(0, max(dat$time), length.out = 100))
-y_pred <- predict(gm, newdata=dat)
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof)
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final spline by aggregating the basis functions
-
-spl<-basis_plot%>% group_by(time)%>%summarize(spl=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=as.factor(Basis)))+
-  geom_line(size=sz,show.legend=FALSE)+
-  geom_point(size=sz+1,show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,labeller = as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#penalized basis
-p12<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
-  geom_line(show.legend = FALSE,size=sz)+
-  geom_point(show.legend = FALSE,size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,labeller=as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#heatmap of the penalization coefficient
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
-  geom_tile(aes(color='black'),size=sz+1,show.legend = FALSE)+
-  geom_tile(size=sz+1)+
-  scale_fill_gradient(low = "white", high = "red")+
-  labs(x='Basis',y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),size=10,show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  scale_color_aaas()+
-  labs(y='Simulated \n response')+
-  geom_line(data=spl,aes(x=time,y=spl),color="#B15731",size=sz+1)+
-  theme_classic()
-
-#Combining all
-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=30))
-
-```
-
-## GAMs and covariance
-
-## GAMs and significance
-
-The use of GAMs to analyze biomedical longitudinal data, and the impact of missing observations in the fit of the model will be examined in detail in the following section using simulation.
-
-https://stats.stackexchange.com/questions/364568/extracting-significance-of-gam-smoothing-terms
-
-***
-
-# References

### 2021-03-18 Manuscripts/Bookdown testing/Rendering.R

@@ -1,0 +1,4 @@
+render_book(input = ".", output_format = NULL, ..., clean = TRUE,
+            envir = parent.frame(), clean_envir = !interactive(),
+            output_dir = NULL, new_session = NA, preview = FALSE,
+            config_file = "_bookdown.yml")

### 2021-04-13 Manuscripts/Manuscript v4/Manuscript_AM_v4.Rmd
Updated manuscript with Dr. Tipton's commentaries

@@
 On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].
-Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976;@greenhouse1959], but the effectiveness the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001].
+Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976;@greenhouse1959], but the effectiveness of the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001].
 In the case of biomedical research, where living subjects are frequently used, sample sizes are often not "large" due to ethical and budgetary reasons [@charan2013] which might cause the corrections for lack of compound symmetry to be ineffective.
@@
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used by the biomedical community, these non-parametric models are customarily used in other fields to analyze longitudinal data.
+Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
 Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
@@
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{tij}\\ 
+y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
 (\#eq:linear-model)
@@
 y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_1 \times time_{t} + \beta_2 \times +\beta_3 \times time_{t} \times +\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
+\beta_0 + \beta_1 \times time_{t} + \beta_2 \times +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
 \end{cases}
@@
 
-To further simplify the expression, substitute $\widetilde{\mu}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B.
+To further simplify the expression, substitute $\widetilde{\beta_{0}}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B.
 This substitution allows for a different intercept and slope for Groups A and B.
@@
 y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\widetilde{\mu} + \widetilde{\beta_1} \times time_{t} +\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
+\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
 \end{cases}
@@
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{tij}\\ 
+y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
 (\#eq:LMEM)
@@
 This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject within each group.
-The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the "global noise" term $\varepsilon_{tij}$ from Equation \@ref(eq:linear-model).
+The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the "global noise" term $\varepsilon_{ijt}$ from Equation \@ref(eq:linear-model).
 
@@
 This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and  missing data.
-Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
+In particular, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
 However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data.
@@
 An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme].
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response, A)).
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A)).
 In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C).
@@
 Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \@ref(fig:l-q-response) B,D).
-A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \@ref(fig:linear-cases) and Figure \@ref(fig:quadratic-cases) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \@ref(fig:linear-cases) and Figure \@ref(fig:quadratic-cases) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
 
@@
 library(tidyverse)
+
 knitr::opts_chunk$set(fig.pos = "!h", out.extra = "")
@@
 ```{r,include=FALSE,message=FALSE,echo=FALSE}
+## Example with linear response
+example <- function(n_time = 6, 
+                    fun_type = "linear", 
+                    error_type = "correlated") {
+
+  if (!(fun_type %in% c("linear", "quadratic")))
+    stop('fun_type must be either "linear", or "quadratic"')
+  if (!(error_type %in% c("correlated", "independent")))
+    stop('fun_type must be either "correlated", or "independent"')
+
+  library(tidyverse)
+  library(mvnfast)
+  library(nlme)
+  library(ggsci)
+
+  x <- seq(1,6, length.out = n_time)
+  mu <- matrix(0, length(x), 2)
+  # linear response
+  if (fun_type == "linear") {
+    mu[, 1] <- - (0.25*x)+2  
+    mu[, 2] <- 0.25*x+2
+  } else {
+    # nonlinear response
+
+    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
+    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
+  }
+  # matplot(mu, type = 'l')
+
+  y <- array(0, dim = c(length(x), 2, 10))
+  errors <- array(0, dim = c(length(x), 2, 10))
+
+  if (error_type == "independent") {
+    ## independent errors
+    for (i in 1:2) {
+      for (j in 1:10) {
+        errors[, i, j] <- rnorm(6, 0, 0.25)
+        y[, i, j] <- mu[, i] + errors[, i, j]
+      }
+    }
+  } else {
+    for (i in 1:2) {     # number of treatments
+      for (j in 1:10) {  # number of subjects
+        # compound symmetry errors
+        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
+        y[, i, j] <- mu[, i] + errors[, i, j]
+      }
+    }
+  }
+
+  ## subject random effects
+
+  ## visualizing the difference between independent errors and compound symmetry
+  ## why do we need to account for this -- overly confident inference
+
+  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
+  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
+  dimnames(mu) <- list(time = x, treatment = 1:2)
+  dat <- as.data.frame.table(y, responseName = "y")
+  dat_errors <- as.data.frame.table(errors, responseName = "errors")
+  dat_mu <- as.data.frame.table(mu, responseName = "mu")
+  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
+  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
+  dat$time <- as.numeric(as.character(dat$time))
+  dat <- dat %>%
+    mutate(subject = factor(paste(subject, treatment, sep = "-")))
+
+  ## repeated measures ANOVA in R
+  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
+  dat$preds_lm <- predict(fit_lm)
+
+  fit_lme <- lme(y ~ treatment + time + treatment:time,
+                 data = dat,
+                 random = ~ 1 | subject,
+                 correlation = corCompSymm(form = ~ 1 | subject)
+  )
+
+  pred_dat <- expand.grid(
+    treatment = factor(1:2), 
+    time = unique(dat$time)
+  )
+
+  dat$y_pred <- predict(fit_lme)
+
+  return(list(
+    dat = dat,
+    pred_dat = pred_dat,
+    fit_lm = fit_lm,
+    fit_lme = fit_lme
+
+  ))
+}
+
+plot_example <- function(sim_dat) {
+  library(patchwork)
+  ## Plot the simulated data
+  p1 <- sim_dat$dat %>%
+    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
+    geom_point(show.legend=FALSE) +labs(y='response')+
+    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
+    theme_classic() +
+    #ggtitle("Simulated data with true response function")+
+    theme(plot.title = element_text(size = 18, 
+                                  face = "bold"),
+        text=element_text(size=18))+
+    scale_color_aaas()
+
+  p2 <- sim_dat$dat %>%
+    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
+    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
+    # facet_wrap(~ treatment) +
+    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
+    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
+    #ggtitle("Simulated data\nIndividual responses with population mean") +
+    theme_classic()+
+     theme(plot.title = element_text(size = 18, 
+                                face = "bold"),
+     text=element_text(size=18))+
+    scale_color_aaas()
+
+   p3 <- sim_dat$dat %>%
+    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
+    geom_line(show.legend=FALSE) +labs(y='errors')+
+     theme_classic()+
+    # facet_wrap(~ treatment) +
+    #ggtitle("Simulated errors") +
+     theme(plot.title = element_text(size = 18, 
+                                  face = "bold"),
+        text=element_text(size=18))+
+    scale_color_aaas()
+
+  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
+    geom_point()+labs(y='response')+
+    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
+    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
+    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
+    theme_classic() +
+    #ggtitle("Fitted Model")+
+    theme(plot.title = element_text(size = 18, 
+                                  face = "bold"),
+        text=element_text(size=18))+
+    scale_color_aaas()
+
+  #return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
+
+  return((p1+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
+
+}
+
+txt<-18
+A<-plot_example(example(fun_type = "linear", error_type = "correlated"))
+
+B<-plot_example(example(fun_type = "linear", error_type = "independent"))
+
+C<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
+
+D<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
+
+```
+
+```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap = "Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
+B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
+The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions."}
+# linear response, correlated errors (subject effect)
+par(mar = c(0.5, 0.5, 0.5, 0.5))
+A/C+plot_annotation(tag_levels = 'A')
+```
+
+The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B).
+When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.
+
+However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response), C.
+Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in Figure \@ref(fig:l-q-response) A and B .
+The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
+Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panel D in Figure \@ref(fig:l-q-response).
+
+A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\@ref(fig:l-q-response), D) indicates that the model is not capturing the changes within each group in a good way.
+Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
+Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D).
+If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response.
+Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
+
+This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
+Although the model fitted to the simulated data was an rm-ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM.
+In the following section, we present generalized additive models (GAMs) as a data-driven  alternative method to analyze longitudinal non-linear data.
+
+# GAMs as a special case of Generalized Linear Models
+
+## GAMs and Basis Functions
+
+Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors[@nelder1972].
+In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case[@simpson2018;@wood2017;@hastie1987].
+A GAM model can be written as:
+
+\begin{equation}
+  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
+  (\#eq:GAM)
+\end{equation}
+
+Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual error.
+
+In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_.
+This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
+One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
+In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.
+
+The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using _basis functions_ expansions of the covariates and by estimating random coefficients for these basis functions.
+A _basis_ is a set of functions that spans the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
+For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
+In the case of Equation \@ref(eq:GAM), the basis function is $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
+
+A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness [@wood2017;@simpson2018].
+Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
+Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
+
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\@ref(fig:l-q-response), C).
+The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
+The timeline can be divided in equally spaced _knots_, each knot being a region where a different basis function will be used.
+Because there are six timepoints for this group, five knots can be used.
+The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
+The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots.
+In Panel A of Figure \@ref(fig:basis-plot), the four basis functions (and the intercept) are shown.
+Each of the basis functions is composed of six different points (because there are six points on the timeline).
+To control the "wigliness" of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
+The penalty reduces the "wigliness" of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
+
+```{r,basis-functions-plot, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
+
+library(mgcv)
+
+n_time = 6
+ x <- seq(1,6, length.out = n_time)
+ mu <- matrix(0, length(x), 2)
+ mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
+ mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
+ y <- array(0, dim = c(length(x), 2, 10))
+ errors <- array(0, dim = c(length(x), 2, 10))
+ for (i in 1:2) {     # number of treatments
+     for (j in 1:10) {  # number of subjects
+         # compound symmetry errors
+         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
+         y[, i, j] <- mu[, i] + errors[, i, j]
+     }
+ }
+  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
+ dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
+ dimnames(mu) <- list(time = x, treatment = 1:2)
+ dat <- as.data.frame.table(y, responseName = "y")
+ dat_errors <- as.data.frame.table(errors, responseName = "errors")
+ dat_mu <- as.data.frame.table(mu, responseName = "mu")
+ dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
+ dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
+ dat$time <- as.numeric(as.character(dat$time))
+ dat <- dat %>%
+     mutate(subject = factor(paste(subject, treatment, sep = "-")))
+
+  dat<-subset(dat,treatment==1)
+  dat<-dat[,c('y','time')]
+gm<-gam(y~s(time,k=5),data=dat)
+model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
+
+x_new <- (seq(0, max(dat$time), length.out = 100))
+y_pred <- predict(gm, newdata=dat)
+
+time<-c(1:6)
+
+basis<-model_matrix[1:6,] #extracting basis
+#basis<-model_matrix[1:6,-1] #extracting basis
+colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
+basis<-basis %>% #pivoting to long format
+  pivot_longer(
+    cols=starts_with("s")
+  )%>%
+  arrange(name) #ordering
+
+#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
+ln<-6*(length(coef(gm)))
+
+basis_plot<-data.frame(Basis=integer(ln),
+                       value_orig=double(ln),
+                       time=integer(ln),
+                       cof=double(ln)
+)
+
+basis_plot$time<-rep(time) #pasting timepoints
+basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
+basis_plot$value_orig<-basis$value #pasting basis values
+basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
+basis_plot<-basis_plot%>%
+  mutate(mod_val=value_orig*cof)
+
+#creating labeller to change the labels in the basis plots
+
+basis_names<-c(
+  `1`="Intercept",
+  `2`="1",
+  `3`="2",
+  `4`="3",
+  `5`="4"
+)
+
+#calculating the final spline by aggregating the basis functions
+
+spl<-basis_plot%>% group_by(time)%>%summarize(spl=sum(mod_val))
+
+#original basis
+sz<-1
+p11<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=as.factor(Basis)))+
+  geom_line(size=sz,show.legend=FALSE)+
+  geom_point(size=sz+1,show.legend = FALSE)+
+  labs(y='Basis functions')+
+  facet_wrap(~Basis,labeller = as_labeller(basis_names))+
+  theme_classic()+
+  scale_color_aaas()
+
+#penalized basis
+p12<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
+  geom_line(show.legend = FALSE,size=sz)+
+  geom_point(show.legend = FALSE,size=sz+1)+
+  labs(y='Penalized \n basis functions')+
+  scale_y_continuous(breaks=seq(-1,1,1))+
+  facet_wrap(~Basis,labeller=as_labeller(basis_names))+
+  theme_classic()+
+  scale_color_aaas()
+
+#heatmap of the penalization coefficient
+x_labels<-c("Intercept","1","2","3","4")
+p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
+  geom_tile(aes(color='black'),size=sz+1,show.legend = FALSE)+
+  geom_tile(size=sz+1)+
+  scale_fill_gradient(low = "white", high = "red")+
+  labs(x='Basis',y='Basis')+
+  scale_x_discrete(labels=x_labels)+
+  geom_text(aes(label=round(cof,2)),size=10,show.legend = FALSE)+
+  theme_classic()+
+  theme(legend.title = element_blank())
+
+#plotting simulated datapoints and smooth term
+p14<-ggplot(data=dat,aes(x=time,y=y))+
+  geom_point(size=sz+1)+
+  scale_color_aaas()+
+  labs(y='Simulated \n response')+
+  geom_line(data=spl,aes(x=time,y=spl),color="#B15731",size=sz+1)+
+  theme_classic()
+
+#Combining all
+b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
+  theme(
+     text=element_text(size=18))
+
+```
+
+In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
+In Figure \@ref(fig:basis-plot), Panel C shows the resulting penalized basis functions.
+Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
+Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
+The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
+
+```{r,basis-plot,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap="Basis functions for a single smoother for time with five knots.
+A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
+B: Penalty matrix for the basis functions.
+Each basis function is penalized by a coefficient which can be positive or negative.
+The coefficient determines the overall effect of each basis in the final smoother.
+C: Penalized basis functions.
+Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
+D.
+Smoother for time and original datapoints.
+The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points."}
+
+par(mar = c(2, 2, 2, 2))
+
+b_plot
+
+```
+
+ \newpage
+
+## GAMs and covariance
+
+Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
+A detailed description on basis functions and correlations can be found in [@hefley2017].
+In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the _time_ effect to account for the repeated measures over time.
+An example where this is covered is in the Appendix.
+
+## Determination of significance in GAMs
+
+At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
+Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs use a different approach to estimate significance.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that timepoint (or timepoints).
+The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
+The major advantage of the significance estimation in GAMs  is that they allow to determine significance at a specific timepoint, rather than providing a single _p-value_ for the overall treatment effect.
+The estimation of siginificant differences in GAMs is covered in detail in the Appendix using simulated longitudinal data that follows previously reported trends of tumor response between multiple treatment groups [@vishwanath2009].
+
+***
+
+# References
+
+<div id="refs"></div>
+
+\newpage
+
+# (APPENDIX) Appendix {-}
+
+# Simulation
+
+## Compound symmetry and independent errors in linear and quadratic responses
+
+This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation).
+The linear simulations using Figure \@ref(fig:linear-cases) show in  panels A and D the simulated mean responses and individual datapoints.
+Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
+In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
+Panels D and H show the predictions from the rm-ANOVA model.
+
+```{r,include=TRUE,message=FALSE,echo=TRUE}
 ## Example with linear response
@@
 
-  #return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-  return((p1+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
+  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
 
@@
 
-```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap = "Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+```{r, linear-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
 A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
@@
 Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data."}
+The rm-ANOVA model does not pick the trend of the quadratic data**"}
 # linear response, correlated errors (subject effect)
-par(mar = c(0.5, 0.5, 0.5, 0.5))
-A/C+plot_annotation(tag_levels = 'A')
+A1/B1+plot_annotation(tag_levels = 'A')
 ```
 
-The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B).
-When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.
+For the quadratic response case, Figure \@ref(fig:quadratic-cases) shows the simulated responses using compound symmetry and independent errors.
 
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response), C.
-Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in \@ref(fig:l-q-response) A and B .
-The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panel D in Figure \@ref(fig:l-q-response).
-
-A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure (\@ref(fig:l-q-response), D indicates that the model is not capturing the changes within each group in a good way.
-Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
-Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D).
-If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response, potentially limiting valuable inferences from the data.
-
-This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-Although the model fitted to the simulated data was an rm_ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM.
-In the following section, we present generalized additive models (GAMs) as an alternative to analyze longitudinal non-linear data.
-
-# GAMs as a special case of Generalized Linear Models
-
-## GAMs and Basis Functions
-
-Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors[@nelder1972].
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case[@simpson2018;@wood2017;@hastie1987].
-A GAM model can be written as:
-
-\begin{equation}
-  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
-  (\#eq:GAM)
-\end{equation}
-
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ and $\varepsilon_{ijt}$ represents the residual error.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
-One possible function for $f(x_t\mid \beta_j)$ that allows for non-linear responses is a polynomial, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
-In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.
-
-The smooth functional relationship between the covariates and the response in GAMs is specified   using a parametric relationship that can be fit within the GLM framework, by using _basis functions_ expansions of the covariates and by estimating random coefficients for these basis functions and this step is achieved by using _basis functions_ to represent them.
-A _basis_ is a set of functions that spans the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
-For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
-In the case of Equation \@ref(eq:GAM), the basis function is $f(x_t\mid \beta_j)$, which means that the model allows relationships beyond linear for the covariates.
-
-A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations[@wood2017;@simpson2018].
-Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness [@wegman1983].
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure @\ref(fig:l-q-response), C.
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group trough the timeline.
-The timeline can be divided equally spaced _knots_, each knot being a region where a different basis function will be used.
-Because there are six timepoints for this group, five knots can be used.
-The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots.
-In Panel A of Figure \@ref(fig:basis-plot), the four basis functions (and the intercept) are shown.
-Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the wigliness of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
-
-```{r,basis-functions-plot, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-library(mgcv)
-
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  dat<-subset(dat,treatment==1)
-  dat<-dat[,c('y','time')]
-gm<-gam(y~s(time,k=5),data=dat)
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-x_new <- (seq(0, max(dat$time), length.out = 100))
-y_pred <- predict(gm, newdata=dat)
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof)
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final spline by aggregating the basis functions
-
-spl<-basis_plot%>% group_by(time)%>%summarize(spl=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=as.factor(Basis)))+
-  geom_line(size=sz,show.legend=FALSE)+
-  geom_point(size=sz+1,show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,labeller = as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#penalized basis
-p12<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
-  geom_line(show.legend = FALSE,size=sz)+
-  geom_point(show.legend = FALSE,size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,labeller=as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#heatmap of the penalization coefficient
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
-  geom_tile(aes(color='black'),size=sz+1,show.legend = FALSE)+
-  geom_tile(size=sz+1)+
-  scale_fill_gradient(low = "white", high = "red")+
-  labs(x='Basis',y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),size=10,show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  scale_color_aaas()+
-  labs(y='Simulated \n response')+
-  geom_line(data=spl,aes(x=time,y=spl),color="#B15731",size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=30))
-
+```{r, quadratic-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
+A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
+B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
+C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
+D,H: Estimations from the rm-ANOVA model for the mean group response.
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
+# linear response, correlated errors (subject effect)
+C1/D1+plot_annotation(tag_levels = 'A')
 ```
-
-In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \@ref(fig:basis-plot), Panel C shows the resulting penalized basis functions.
-Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
-Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
-
-```{r,basis-plot,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap="Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
-D.
-Smoother for time and original datapoints.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points."}
-
-par(mar = c(2, 2, 2, 2))
-
-b_plot
-
-```
-
-## GAMs and covariance
-
-Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in [@hefley2017].
-In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the _time_ effect to account for the repeated measures over time.
-An example where this is covered is in the Appendix.
-
-## Determination of significance in GAMs
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that timepoint (or timepoints).
-The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
-The major advantage of the significance estimation in GAMs  is that they allow to determine significance at a specific timepoint, rather than providing a single _p-value_ for the overall treatment effect.
-The estimation of siginificant differences in GAMs is covered in detail in the Appendix using simulated longitudinal data that follows previously reported trends of tumor response between multiple treatment groups [@vishwanath2009].
-
-***
-
-# References

### 2021-04-13 Manuscripts/Manuscript v4/Manuscript_AM_v4.Rmd
Need to work on the references to omit the "others" and properly write the Journal names

@@
 \beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_1 \times time_{t} + \beta_2 \times +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\beta_0 + \beta_1 \times time_{t} + \beta_2+\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
 \end{cases}
@@
 An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme].
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A)).
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A).
 In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C).
@@
 However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response), C.
-Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in Figure \@ref(fig:l-q-response) A and B .
+Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \@ref(fig:l-q-response) A and B) .
 The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.

### 2021-04-13 Manuscripts/Manuscript v4/Manuscript_AM_v4.tex
Need to work on the references to omit the "others" and properly write the Journal names

--- 2021-03-17 Manuscripts/Manuscript v4/Manuscript_AM_v4.tex
@@
 \usepackage{breqn}
+\usepackage[font={small}]{caption}
+\usepackage{float}
 
@@
 On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}25{]}.
-Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}26,27{]}, but the effectiveness the correction is limited by the size of the sample, the number of measurements{[}28{]}, and group sizes {[}29{]}.
+Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}26,27{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}28{]}, and group sizes {[}29{]}.
 In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}30{]} which might cause the corrections for lack of compound symmetry to be ineffective.
@@
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used by the biomedical community, these non-parametric models are customarily used in other fields to analyze longitudinal data.
+Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
 Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}32--34{]}, health-environment interactions {[}35{]} and the dynamics of government in political science {[}36{]} .
@@
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{tij}\\ 
+y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
 \label{eq:linear-model}
@@
 y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_1 \times time_{t} + \beta_2 \times +\beta_3 \times time_{t} \times +\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
+\beta_0 + \beta_1 \times time_{t} + \beta_2+\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
 \end{cases}
@@
 
-To further simplify the expression, substitute \(\widetilde{\mu}=\beta_0+\beta_{2}\) and \(\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}\) in the equation for Group B.
+To further simplify the expression, substitute \(\widetilde{\beta_{0}}=\beta_0+\beta_{2}\) and \(\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}\) in the equation for Group B.
 This substitution allows for a different intercept and slope for Groups A and B.
@@
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{tij}\\ 
+y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
 \label{eq:LMEM}
@@
 This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject within each group.
-The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the ``global noise'' term \(\varepsilon_{tij}\) from Equation \eqref{eq:linear-model}.
+The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the ``global noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.
 
@@
 This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and missing data.
-Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
+In particular, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
 However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data.
@@
 An rm-ANOVA model (Equation \eqref{eq:linear-model}) is fitted to each group, using \(\textsf{R}\){[}38{]} and the package \emph{nlme}{[}49{]}.
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure @ref(fig:l-q-response, A)).
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}, A)).
 In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}, C).
@@
 Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D).
-A more comprehensive exploration of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \ref{fig:linear-cases} and Figure \ref{fig:quadratic-cases} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+A more comprehensive exploration of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \ref{fig:linear-cases} and Figure \ref{fig:quadratic-cases} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
 
-\begin{figure}
-\centering
-\includegraphics{Manuscript_AM_v4_files/figure-latex/l-q-response-1.pdf}
-\caption{\label{fig:l-q-response}\textbf{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v4_files/figure-latex/l-q-response-1}
+
+}
+
+\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
 A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
+B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
 Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data}}
+The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.}\label{fig:l-q-response}
 \end{figure}
@@
 However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}, C.
-Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in \ref{fig:l-q-response} A and B .
+Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in Figure \ref{fig:l-q-response} A and B .
 The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
@@
 
-A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure (\ref{fig:l-q-response}, D indicates that the model is not capturing the changes within each group in a good way.
+A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\ref{fig:l-q-response}, D) indicates that the model is not capturing the changes within each group in a good way.
 Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
 Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare panels C and D).
-If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response, potentially limiting valuable inferences from the data.
+If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response.
+Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
 
 This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-Although the model fitted to the simulated data was an rm\_ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM.
-In the following section, we present generalized additive models (GAMs) as an alternative to analyze longitudinal non-linear data.
+Although the model fitted to the simulated data was an rm-ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM.
+In the following section, we present generalized additive models (GAMs) as a data-driven alternative method to analyze longitudinal non-linear data.
 
@@
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the function \(f(x_t\mid \beta_j)\) and \(\varepsilon_{ijt}\) represents the residual error.
+Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the function \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.
 
@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
-One possible function for \(f(x_t\mid \beta_j)\) that allows for non-linear responses is a polynomial, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}36{]}.
+One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}36{]}.
 In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic, and causes bias at the endpoints of the time period.
 
-The smooth functional relationship between the covariates and the response in GAMs is specified using a parametric relationship that can be fit within the GLM framework, by using \emph{basis functions} expansions of the covariates and by estimating random coefficients for these basis functions and this step is achieved by using \emph{basis functions} to represent them.
+The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis functions} expansions of the covariates and by estimating random coefficients for these basis functions.
 A \emph{basis} is a set of functions that spans the space where the smooths that approximate \(f(x_t \mid \beta_j)\) exist {[}34{]}.
@@
 The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}52{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
-In the case of Equation \eqref{eq:GAM}, the basis function is \(f(x_t\mid \beta_j)\), which means that the model allows relationships beyond linear for the covariates.
+In the case of Equation \eqref{eq:GAM}, the basis function is \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-A commonly used \emph{basis function} is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations{[}34,37{]}.
-Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness {[}53{]}.
+A commonly used \emph{basis function} is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness {[}34,37{]}.
+Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}53{]}.
 Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
 
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure @\ref(fig:l-q-response), C.
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits the trend of the group trough the timeline.
-The timeline can be divided equally spaced \emph{knots}, each knot being a region where a different basis function will be used.
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\ref{fig:l-q-response}, C).
+The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
+The timeline can be divided in equally spaced \emph{knots}, each knot being a region where a different basis function will be used.
 Because there are six timepoints for this group, five knots can be used.
@@
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the wigliness of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
+To control the ``wigliness'' of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
+The penalty reduces the ``wigliness'' of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
 
@@
 D.
-Smoother for \emph{time} and original datapoints.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}}
+Smoother for time and original datapoints.
+The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}\label{fig:basis-plot}
 \end{figure}
+
+\newpage
 
@@
 
+\newpage
+
 \hypertarget{appendix-appendix}{%
@@
 
-\begin{figure}
-\centering
-\includegraphics{Manuscript_AM_v4_files/figure-latex/linear-cases-1.pdf}
-\caption{\label{fig:linear-cases}\textbf{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+\begin{figure}[!h]
+\includegraphics{Manuscript_AM_v4_files/figure-latex/linear-cases-1} \caption{**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
 A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
@@
 Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data}}
+The rm-ANOVA model does not pick the trend of the quadratic data**}\label{fig:linear-cases}
 \end{figure}
@@
 
-\begin{figure}
-\centering
-\includegraphics{Manuscript_AM_v4_files/figure-latex/quadratic-cases-1.pdf}
-\caption{\label{fig:quadratic-cases}\textbf{Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
+\begin{figure}[!h]
+\includegraphics{Manuscript_AM_v4_files/figure-latex/quadratic-cases-1} \caption{**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
 A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
@@
 D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.}}
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**}\label{fig:quadratic-cases}
 \end{figure}

### 2021-04-14 Manuscripts/Manuscript v4/Manuscript_AM_v4.tex
References fixed, need to improve equations when rendering to Word for TM

@@
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
+y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times \textrm{treatment}_{j} +\beta_3 \times time_{t}\times \textrm{treatment}_{j}+\varepsilon_{ijt}\\ 
 \label{eq:linear-model}
@@
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
+y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times \textrm{treatment}_{j} + \beta_3 \times time_{t}\times \textrm{treatment}_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
 \label{eq:LMEM}
@@
 An rm-ANOVA model (Equation \eqref{eq:linear-model}) is fitted to each group, using \(\textsf{R}\){[}38{]} and the package \emph{nlme}{[}49{]}.
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}, A)).
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}, A).
 In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}, C).
@@
 However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}, C.
-Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in Figure \ref{fig:l-q-response} A and B .
+Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \ref{fig:l-q-response} A and B) .
 The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
@@
 Butler, C.
-McLaren, W.-P.
-Chen, others, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences.
+McLaren, W.P.
+Chen, B.
+Tromberg, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences of the United States of America.
 108 (2011) 14626--14631.
+\url{https://doi.org/10.1073/pnas.1013103108}.
 
@@
 Ko, R.S.
-Mehta, others, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research.
+Mehta, B.J.
+Tromberg, D.
+Roblyer, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research.
 22 (2020) 1--10.
+\url{https://doi.org/doi:10.1186/s13058-020-01262-1}.
 
@@
 Safonov, N.M.
-Shakhova, others, Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy, Journal of Biomedical Optics.
-23 (2018) 091410.
+Shakhova, A.V.
+Maslennikova, Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy, Journal of Biomedical Optics.
+23 (n.d.).
+\url{https://doi.org/\%7B10.1117/1.JBO.23.9.091410\%7D}.
 
@@
 Vitkin, Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography, Scientific Reports.
-8 (2018) 1--12.
+8 (n.d.).
+\url{https://doi.org/\%7B10.1038/s41598-017-18635-w\%7D}.
 
@@
 {[}5{]} G.
-Ritter, L.S.
+Ritter, L.
 Cohen, C.
-Williams, E.C.
-Richards, L.J.
+Williams, E.
+Richards, L.
 Old, S.
-Welt, Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody a33, Cancer Research.
-61 (2001) 6851--6859.
+Welt, Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33, Cancer Research.
+61 (n.d.) 6851--6859.
 
 \leavevmode\hypertarget{ref-roth2017}{}%
-{[}6{]} E.M. Roth, A.C. Goldberg, A.L. Catapano, A. Torri, G.D. Yancopoulos, N. Stahl, A. Brunet, G. Lecorps, H.M. Colhoun, Antidrug antibodies in patients treated with alirocumab, (2017).
+{[}6{]} E.M.
+Roth, A.C.
+Goldberg, A.L.
+Catapano, A.
+Torri, G.D.
+Yancopoulos, N.
+Stahl, A.
+Brunet, G.
+Lecorps, H.M.
+Colhoun, Antidrug Antibodies in Patients Treated with Alirocumab, New England Journal of Medicine.
+376 (n.d.) 1589--1590.
+\url{https://doi.org/\%7B10.1056/NEJMc1616623\%7D}.
 
@@
 Quinn, In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing, Communications Biology.
-1 (2018) 1--8.
+1 (n.d.).
+\url{https://doi.org/\%7B10.1038/s42003-018-0206-4\%7D}.
 
 \leavevmode\hypertarget{ref-skala2010}{}%
-{[}8{]} M.C. Skala, A.N. Fontanella, L. Lan, J.A. Izatt, M.W. Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics. 15 (2010) 011112.
+{[}8{]} M.C.
+Skala, A.
+Fontanella, L.
+Lan, J.A.
+Izatt, M.W.
+Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics.
+15 (2010).
+\url{https://doi.org/10.1117/1.3285584}.
 
@@
 Muldoon, Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy, Biomedical Optics Express.
-9 (2018) 2871--2886.
+9 (n.d.) 2871--2886.
+\url{https://doi.org/\%7B10.1364/BOE.9.002871\%7D}.
 
@@
 Griffin, R.F.
-Morton, others, Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}), Supportive Care in Cancer.
-24 (2016) 3847--3855.
+Morton, J.C.
+Anders, T.J.
+Stoffel, R.E.
+Haselow, R.B.
+Mowat, M.A.N.
+Wittich, J.D.
+Bearden III, R.C.
+Miller, Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}), Supportive Care in Cancer.
+24 (n.d.) 3847--3855.
+\url{https://doi.org/\%7B10.1007/s00520-016-3213-3\%7D}.
 
 \leavevmode\hypertarget{ref-kamstra2015}{}%
-{[}11{]} J.
-Kamstra, P.
+{[}11{]} J.I.
+Kamstra, P.U.
 Dijkstra, M.
-Van Leeuwen, J.
-Roodenburg, J.
+van Leeuwen, J.L.N.
+Roodenburg, J.A.
 Langendijk, Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study, Oral Oncology.
-51 (2015) 548--555.
+51 (n.d.) 548--555.
+\url{https://doi.org/\%7B10.1016/j.oraloncology.2015.01.016\%7D}.
 
 \leavevmode\hypertarget{ref-wagenmakers2008}{}%
-{[}12{]} E.-J. Wagenmakers, M. Lee, T. Lodewyckx, G.J. Iverson, Bayesian versus frequentist inference, in: Bayesian Evaluation of Informative Hypotheses, Springer, 2008: pp. 181--207.
+{[}12{]} E.-J.
+Wagenmakers, M.
+Lee, T.
+Lodewyckx, G.J.
+Iverson, Bayesian versus frequentist inference, in: H.
+Hoijtink, I.
+Klugkist, P.A.
+Boelen (Eds.), Bayesian Evaluation of Informative Hypotheses, Springer New York, New York, NY, 2008: pp.
+181--207.
+\url{https://doi.org/10.1007/978-0-387-09612-4_9}.
 
@@
 Gueorguieva, J.H.
-Krystal, Move over anova: Progress in analyzing repeated-measures data andits reflection in papers published in the archives of general psychiatry, Archives of General Psychiatry.
+Krystal, Move over ANOVA - Progress in analyzing repeated-measures data and its reflection in papers published in the archives of general psychiatry, Archives of General Psychiatry.
 61 (2004) 310--317.
+\url{https://doi.org/10.1001/archpsyc.61.3.310}.
 
 \leavevmode\hypertarget{ref-schober2018}{}%
-{[}14{]} P. Schober, T.R. Vetter, Repeated measures designs and analysis of longitudinal data: If at first you do not succeed---try, try again, Anesthesia and Analgesia. 127 (2018) 569.
+{[}14{]} P.
+Schober, T.R.
+Vetter, Repeated measures designs and analysis of longitudinal data: If at first you do not succeed-try, try again, Anesthesia and Analgesia.
+127 (2018) 569--575.
+\url{https://doi.org/10.1213/ane.0000000000003511}.
 
 \leavevmode\hypertarget{ref-pinheiro2006}{}%
-{[}15{]} J. Pinheiro, D. Bates, Mixed-effects models in s and s-plus, Springer Science \& Business Media, 2006.
+{[}15{]} J. Pinheiro, D. Bates, Mixed-effects models in S and S-PLUS, Springer Science \& Business Media, 2006. \url{https://doi.org/https://doi.org/10.1007/b98882}.
 
 \leavevmode\hypertarget{ref-vishwanath2009}{}%
-{[}16{]} K. Vishwanath, H. Yuan, W.T. Barry, M.W. Dewhirst, N. Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia. 11 (2009) 889--900.
+{[}16{]} K.
+Vishwanath, H.
+Yuan, W.T.
+Barry, M.W.
+Dewhirst, N.
+Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia.
+11 (2009) 889--900.
+\url{https://doi.org/10.1593/neo.09580}.
 
@@
 Taper, S.R.
-Lele, Errors in statistical inference under model misspecification: Evidence, hypothesis testing, and aic, Frontiers in Ecology and Evolution.
-7 (2019) 372.
+Lele, Errors in Statistical Inference Under Model Misspecification: Evidence, Hypothesis Testing, and AIC, Frontiers in Ecology and Evolution.
+7 (n.d.).
+\url{https://doi.org/\%7B10.3389/fevo.2019.00372\%7D}.
 
 \leavevmode\hypertarget{ref-wang2019}{}%
-{[}18{]} B. Wang, Z. Zhou, H. Wang, X.M. Tu, C. Feng, The p-value and model specification in statistics, General Psychiatry. 32 (2019).
+{[}18{]} B. Wang, Z. Zhou, H. Wang, X.M. Tu, C. Feng, The p-value and model specification in statistics, General Psychiatry. 32 (n.d.). \url{https://doi.org/\%7B10.1136/gpsych-2019-100081\%7D}.
 
 \leavevmode\hypertarget{ref-liu2010}{}%
-{[}19{]} C. Liu, T.P. Cripe, M.-O. Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy. 18 (2010) 1724--1730.
+{[}19{]} C.
+Liu, T.P.
+Cripe, M.-O.
+Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy.
+18 (2010) 1724--1730.
+\url{https://doi.org/10.1038/mt.2010.127}.
 
 \leavevmode\hypertarget{ref-halsey2015}{}%
-{[}20{]} L.G. Halsey, D. Curran-Everett, S.L. Vowler, G.B. Drummond, The fickle p value generates irreproducible results, Nature Methods. 12 (2015) 179--185.
+{[}20{]} L.G.
+Halsey, D.
+Curran-Everett, S.L.
+Vowler, G.B.
+Drummond, The fickle P value generates irreproducible results, Nature Methods.
+12 (n.d.) 179--185.
+\url{https://doi.org/\%7B10.1038/nmeth.3288\%7D}.
 
 \leavevmode\hypertarget{ref-abdi2010}{}%
-{[}21{]} H. Abdi, Holm's sequential bonferroni procedure, Encyclopedia of Research Design. 1 (2010) 1--8.
+{[}21{]} H. Abdi, Holm's Sequential Bonferroni Procedure, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.
 
 \leavevmode\hypertarget{ref-nakagawa2004}{}%
-{[}22{]} S. Nakagawa, A farewell to bonferroni: The problems of low statistical power and publication bias, Behavioral Ecology. 15 (2004) 1044--1045.
+{[}22{]} S.
+Nakagawa, A farewell to Bonferroni: the problems of low statistical power and publication bias, Behavioral Ecology.
+15 (n.d.) 1044--1045.
+\url{https://doi.org/\%7B10.1093/beheco/arh107\%7D}.
 
 \leavevmode\hypertarget{ref-gelman2012}{}%
-{[}23{]} A. Gelman, J. Hill, M. Yajima, Why we (usually) don't have to worry about multiple comparisons, Journal of Research on Educational Effectiveness. 5 (2012) 189--211.
+{[}23{]} A.
+Gelman, J.
+Hill, M.
+Yajima, Why We (Usually) Don't Have to Worry About Multiple Comparisons, Journal of Research on Educational Effectiveness.
+5 (n.d.) 189--211.
+\url{https://doi.org/\%7B10.1080/19345747.2011.618213\%7D}.
 
 \leavevmode\hypertarget{ref-albers2019}{}%
-{[}24{]} C. Albers, The problem with unadjusted multiple and sequential statistical testing, Nature Communications. 10 (2019) 1--4.
+{[}24{]} C. Albers, The problem with unadjusted multiple and sequential statistical testing, Nature Communications. 10 (n.d.). \url{https://doi.org/\%7B10.1038/s41467-019-09941-0\%7D}.
 
@@
 36 (2004) 2144--2148.
+\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.
 
 \leavevmode\hypertarget{ref-huynh1976}{}%
-{[}26{]} H. Huynh, L.S. Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics. 1 (1976) 69--82.
+{[}26{]} H.
+Huynh, L.S.
+Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics.
+1 (1976) 69--82.
+\url{https://doi.org/10.3102/10769986001001069}.
 
 \leavevmode\hypertarget{ref-greenhouse1959}{}%
-{[}27{]} S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112.
+{[}27{]} S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.
 
@@
 Haverkamp, A.
-Beauducel, Violation of the sphericity assumption and its effect on type-i error rates in repeated measures anova and multi-level linear models (mlm), Frontiers in Psychology.
-8 (2017) 1841.
+Beauducel, Violation of the Sphericity Assumption and Its Effect on Type-I Error Rates in Repeated Measures ANOVA and Multi-Level Linear Models (MLM), Frontiers in Psychology.
+8 (n.d.).
+\url{https://doi.org/\%7B10.3389/fpsyg.2017.01841\%7D}.
 
 \leavevmode\hypertarget{ref-keselman2001}{}%
-{[}29{]} H. Keselman, J. Algina, R.K. Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematical and Statistical Psychology. 54 (2001) 1--20.
+{[}29{]} H.
+Keselman, J.
+Algina, R.
+Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematica \& Statistical Psychology.
+54 (n.d.) 1--20.
+\url{https://doi.org/\%7B10.1348/000711001159357\%7D}.
 
 \leavevmode\hypertarget{ref-charan2013}{}%
-{[}30{]} J. Charan, N. Kantharia, How to calculate sample size in animal studies?, Journal of Pharmacology \& Pharmacotherapeutics. 4 (2013) 303.
+{[}30{]} J. Charan, N. Kantharia, How to calculate sample size in animal studies?, Journal of Pharmacology and Pharmacotherapeutics. 4 (2013) 303--306. \url{https://doi.org/10.4103/0976-500X.119726}.
 
 \leavevmode\hypertarget{ref-barr2013}{}%
-{[}31{]} D.J. Barr, R. Levy, C. Scheepers, H.J. Tily, Random effects structure for confirmatory hypothesis testing: Keep it maximal, Journal of Memory and Language. 68 (2013) 255--278.
+{[}31{]} D.J.
+Barr, R.
+Levy, C.
+Scheepers, H.J.
+Tily, Random effects structure for confirmatory hypothesis testing: Keep it maximal, JournalL of Memory and Language.
+68 (n.d.) 255--278.
+\url{https://doi.org/\%7B10.1016/j.jml.2012.11.001\%7D}.
 
@@
 {[}32{]} N.L.
-Rose, H.
+Rose, H.D.
 Yang, S.D.
 Turner, G.L.
-Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to Scotland, UK, Geochimica et Cosmochimica Acta.
+Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, uk, Geochimica et Cosmochimica Acta.
 82 (2012) 113--135.
+\url{https://doi.org/10.1016/j.gca.2010.12.026}.
 
 \leavevmode\hypertarget{ref-pedersen2019}{}%
-{[}33{]} E.J. Pedersen, D.L. Miller, G.L. Simpson, N. Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, PeerJ. 7 (2019) e6876.
+{[}33{]} E.J. Pedersen, D.L. Miller, G.L. Simpson, N. Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, Peerj. 7 (2019). \url{https://doi.org/10.7717/peerj.6876}.
 
 \leavevmode\hypertarget{ref-simpson2018}{}%
-{[}34{]} G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018) 149.
+{[}34{]} G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.
 
@@
 Song, Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality, BMC Medical Research Methodology.
-12 (2012) 165.
+12 (n.d.).
+\url{https://doi.org/\%7B10.1186/1471-2288-12-165\%7D}.
 
 \leavevmode\hypertarget{ref-beck1998}{}%
-{[}36{]} N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. (1998) 596--627.
+{[}36{]} N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. 42 (n.d.) 596--627. \url{https://doi.org/\%7B10.2307/2991772\%7D}.
 
 \leavevmode\hypertarget{ref-wood2017}{}%
-{[}37{]} S.N. Wood, Generalized additive models: An introduction with r, CRC press, 2017.
+{[}37{]} S.N. Wood, Generalized additive models: An introduction with , second edition, CRC Press LLC, Philadelphia, PA, 2017.
 
@@
 \leavevmode\hypertarget{ref-wood2016}{}%
-{[}39{]} S.N. Wood, N., Pya, B. Säfken, Smoothing parameter and model selection for general smooth models (with discussion), Journal of the American Statistical Association. 111 (2016) 1548--1575.
+{[}39{]} S.N.
+Wood, N.
+Pya, B.
+Saefken, Smoothing Parameter and Model Selection for General Smooth Models, Journal of the American Statistical Association.
+111 (n.d.) 1548--1563.
+\url{https://doi.org/\%7B10.1080/01621459.2016.1180986\%7D}.
 
 \leavevmode\hypertarget{ref-west2014}{}%
-{[}40{]} B.T. West, K.B. Welch, A.T. Galecki, Linear mixed models: A practical guide using statistical software, CRC Press, 2014.
+{[}40{]} B.T.
+West, K.B.
+Welch, A.T.
+Galecki, Linear mixed models: A practical guide using statistical software, second edition, Taylor \& Francis, 2014.
+\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.
 
 \leavevmode\hypertarget{ref-wolfinger1996}{}%
-{[}41{]} R.D. Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics. (1996) 205--230.
+{[}41{]} R.D.
+Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics.
+1 (1996) 205--230.
+\url{http://www.jstor.org/stable/1400366}.
 
 \leavevmode\hypertarget{ref-weiss2005}{}%
-{[}42{]} R.E. Weiss, Modeling longitudinal data, Springer Science \& Business Media, 2005.
+{[}42{]} R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://books.google.com/books?id=MQ/_bvWDPsEAC}.
 
 \leavevmode\hypertarget{ref-geisser1958}{}%
-{[}43{]} S. Geisser, S.W. Greenhouse, others, An extension of box's results on the use of the \(F\) distribution in multivariate analysis, The Annals of Mathematical Statistics. 29 (1958) 885--891.
+{[}43{]} S.
+Geisser, S.W.
+Greenhouse, An Extension of Box's Results on the Use of the \(F\) Distribution in Multivariate Analysis, The Annals of Mathematical Statistics.
+29 (1958) 885--891.
+\url{https://doi.org/10.1214/aoms/1177706545}.
 
 \leavevmode\hypertarget{ref-maxwell2017}{}%
-{[}44{]} S.E. Maxwell, H.D. Delaney, K. Kelley, Designing experiments and analyzing data: A model comparison perspective, Routledge, 2017.
+{[}44{]} S.E.
+Maxwell, H.D.
+Delaney, K.
+Kelley, Designing experiments and analyzing data: A model comparison perspective, third edition, Taylor \& Francis, 2017.
+\url{https://books.google.com/books?id=NmFQDwAAQBAJ}.
 
 \leavevmode\hypertarget{ref-molenberghs2004}{}%
-{[}45{]} G. Molenberghs, H. Thijs, I. Jansen, C. Beunckens, M.G. Kenward, C. Mallinckrodt, R.J. Carroll, Analyzing incomplete longitudinal clinical trial data, Biostatistics. 5 (2004) 445--464.
+{[}45{]} G.
+Molenberghs, H.
+Thijs, I.
+Jansen, C.
+Beunckens, M.
+Kenward, C.
+Mallinckrodt, R.
+Carroll, Analyzing incomplete longitudinal clinical trial data, Biostatistics.
+5 (n.d.) 445--464.
+\url{https://doi.org/\%7B10.1093/biostatistics/kxh001\%7D}.
 
 \leavevmode\hypertarget{ref-scheffer2002}{}%
-{[}46{]} J. Scheffer, Dealing with missing data, (2002).
+{[}46{]} J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.
 
@@
 Pieper, V.
-Hasselblad, Can one assess whether missing data are missing at random in medical studies?, Statistical Methods in Medical Research.
-15 (2006) 213--234.
+Hasselblad, Can one assess whether missing data are missing at random in medical studies?, STATISTICAL METHODS IN MEDICAL RESEARCH.
+15 (n.d.) 213--234.
+\url{https://doi.org/\%7B10.1191/0962280206sm448oa\%7D}.
 
@@
 Mazumdar, S.G.
-Memtsoudis, Beyond repeated-measures analysis of variance: Advanced statistical methods for the analysis of longitudinal data in anesthesia research, Regional Anesthesia \& Pain Medicine.
-37 (2012) 99--105.
+Memtsoudis, Beyond Repeated-Measures Analysis of Variance Advanced Statistical Methods for the Analysis of Longitudinal Data in Anesthesia Research, Regional Anesthesia and Pain Medicine.
+37 (n.d.) 99--105.
+\url{https://doi.org/\%7B10.1097/AAP.0b013e31823ebc74\%7D}.
 
@@
 \leavevmode\hypertarget{ref-nelder1972}{}%
-{[}50{]} J.A. Nelder, R.W. Wedderburn, Generalized linear models, Journal of the Royal Statistical Society: Series A (General). 135 (1972) 370--384.
+{[}50{]} J.A. Nelder, R.W.M. Wedderburn, Generalized linear models, Journal of the Royal Statistical Society. Series A (General). 135 (1972) 370--384. \url{http://www.jstor.org/stable/2344614}.
 
 \leavevmode\hypertarget{ref-hastie1987}{}%
-{[}51{]} T. Hastie, R. Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association. 82 (1987) 371--386.
+{[}51{]} T.
+Hastie, R.
+Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association.
+82 (1987) 371--386.
+\url{https://doi.org/10.1080/01621459.1987.10478440}.
 
@@
 Williams, M.B.
-Hooten, The basis function approach for modeling autocorrelation in ecological data, Ecology.
-98 (2017) 632--646.
+Hooten, The basis function approach for modeling autocorrelation in ecological data, ECOLOGY.
+98 (n.d.) 632--646.
+\url{https://doi.org/\%7B10.1002/ecy.1674\%7D}.
 
 \leavevmode\hypertarget{ref-wegman1983}{}%
-{[}53{]} E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365.
+{[}53{]} E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.
 

### 2021-04-15 Manuscripts/Manuscript v5/Manuscript_AM_v4.Rmd
Created folder for v5 of the manuscript

--- 2021-04-13 Manuscripts/Manuscript v4/Manuscript_AM_v4.Rmd
@@
       float: null
-  bookdown::html_document2: default
   bookdown::word_document2:
@@
     keep_md: yes
+  bookdown::html_document2: default
   link-citations: yes
@@
   '': default
-csl: Elsevier.csl
+csl: elsevier-with-titles.csl
 bibliography: refs.bib
@@
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
+y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
 (\#eq:linear-model)

### 2021-04-15 Manuscripts/Manuscript v5/Manuscript_AM_v4.tex
Created folder for v5 of the manuscript

--- 2021-04-14 Manuscripts/Manuscript v4/Manuscript_AM_v4.tex
@@
 \begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times \textrm{treatment}_{j} +\beta_3 \times time_{t}\times \textrm{treatment}_{j}+\varepsilon_{ijt}\\ 
+y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
 \label{eq:linear-model}
@@
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times \textrm{treatment}_{j} + \beta_3 \times time_{t}\times \textrm{treatment}_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
+y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
 \label{eq:LMEM}

### 2021-04-15 Manuscripts/Manuscript v5/Manuscript_AM_v5.tex
New v5 pdf created

--- 2021-04-15 Manuscripts/Manuscript v5/Manuscript_AM_v4.tex
@@
 \beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_1 \times time_{t} + \beta_2+\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
 \end{cases}
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v4_files/figure-latex/l-q-response-1}
+{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v5_files/figure-latex/l-q-response-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v4_files/figure-latex/basis-plot-1}
+{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v5_files/figure-latex/basis-plot-1}
 
@@
 \begin{figure}[!h]
-\includegraphics{Manuscript_AM_v4_files/figure-latex/linear-cases-1} \caption{**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+\includegraphics{Manuscript_AM_v5_files/figure-latex/linear-cases-1} \caption{**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
 A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
@@
 \begin{figure}[!h]
-\includegraphics{Manuscript_AM_v4_files/figure-latex/quadratic-cases-1} \caption{**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
+\includegraphics{Manuscript_AM_v5_files/figure-latex/quadratic-cases-1} \caption{**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
 A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.

### 2021-05-02 Code and miscellaneous/Tumor response GAM fit.R
Updated models for missing data and created plots, need to homogenize the aesthetics

@@ -25,7 +25,7 @@ ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
 
 #This function simulates data for the tumor data using a dafault number of
 #10 observations per time point,
-#and a standard deviation of 5% StO2
+#and a standard deviation of 10% StO2
 #Because physiologically StO2 cannot go below 0%, data is generated with a cutoff value of
 #0.0001 (the "observation_tobit")
 
@@ -43,7 +43,7 @@ simulate_data <- function(dat, n = 10, sd = 5) {
 
 
 n <- 10
-sd <- 5
+sd <- 10
 set.seed(11) #setting seed
 df <- 6
 dat_sim <- simulate_data(dat, n, sd)
@@ -92,23 +92,34 @@ gam.check(mod1)
 #plotting smooths and simulated data
 #full data
 #I am doing something wrong here but I can't figure it out!!!!
-f_predict<-with(dat_sim,expand.grid(StO2=seq(min(observation_tobit),
-                                                       max(observation_tobit),
-                                                        length=10),
-                                                    Group=Group,Day=Day))
+#f_predict<-with(dat_sim,expand.grid(StO2=seq(min(observation_tobit),
+#                                                        max(observation_tobit),
+#                                                         length=10),
+#                                                     Group=Group,Day=Day))
 
+ #f_predict<-cbind(f_predict,
+   #                       predict(mod2,f_predict,
+  #                                se.fit = TRUE,type='response'))
+#
 
-f_predict<-cbind(f_predict,
-                         predict(mod2,f_predict,
-                                 se.fit = TRUE,type='response'))
+#creates a dataframe wusing the lenght of the covariates
+f_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = seq(0, 10, by = 0.1))
+
+#adds the predictions to the grid and creates a confidence interval
+f_predict<-f_predict%>%
+    mutate(fit = predict(mod2,f_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(mod2, f_predict,se.fit = TRUE,type='response')$se.fit)
+
 
 ggplot(data=dat_sim, aes(x=Day, y=observation_tobit, group=Group)) +
     facet_wrap(~Group)+
[…]

### 2021-05-02 Code and miscellaneous/Tumor response GAM fit.R
Created plots and model for missing data, need to homogenize aesthetics

@@ -230,3 +230,5 @@ c2<-ggplot(comp_StO2_m, aes(x = Day, y = diff, group = pair)) +
 
 
 c2
+
+c1+c2

### 2021-05-02 Code and miscellaneous/Tumor response GAM fit.R
Working on understanding the model

@@ -61,19 +61,41 @@ n_treatment <- length(unique(dat_sim$Group))
 #if you have n timepoints can you do n knots or should you do n-1 knots? (I read somewhere that
 #the n-1 option was preferred for computational issues)?
 
-
-mod1 <- gam(observation_tobit ~ Group+s(Day, by = Group, k = 5), data  = dat_sim)
-appraise(mod1) #SIMPLEST MODEL: A SMOOTH FOR TIME AS THE COVARIATE, WHICH CAN VARY BY GROUP BUT... APPRAISING THE MODEL SHOWS A WEIRD TREND
-#ON THE RESIDUALS
-
-mod2 <- gam(observation_tobit ~ Group+s(Day,by=Group,k=5),data  = dat_sim)
+mod2 <- gam(observation_tobit ~ s(Day,by=Group,k=5),data  = dat_sim)
 appraise(mod2)
+#SIMPLEST MODEL: A SMOOTH FOR TIME AS THE COVARIATE, WHICH CAN VARY BY GROUP BUT... APPRAISING THE MODEL SHOWS A WEIRD TREND
+#ON THE RESIDUALS
+#this model has a separate smooth for each Group.
+
+mod1 <- gam(observation_tobit ~ Group+te(Day, by = Group, k = 5), data  = dat_sim)
+appraise(mod1)
+
+
+mod3 <- gam(observation_tobit ~ Group+s(Day, by = Group, k = 5), data  = dat_sim)
+appraise(mod3)
+
+#Ok, so mod1 and mod3 are doing essentially the same thing. We have a smooth of time
+#per group, and we take the main effect of each treatment is also taken into account.
+#So this model effectively takes into account treatment and time? and the simples model
+#only takes the effect of time.
+
 #THIS MODEL is accounting for a linear effect for each group, and a smooth for time by each group
 #Why is it fitting better than the "simple" model? What does this "linear" Group term mean?
 #The idea for this came from
 #https://stats.stackexchange.com/questions/486118/model-building-in-generalized-additive-mixed-models-gamms
 
 #mod2 <- gam(observation ~ s(Day, by = Group, k = 5), data  = dat_missing)
+
+# Reading through Simon Wood's book it seems that the linear term can be used when the
+# expected contribution of a factor is linear. The simples model then creates smooths by day
+# per group, but the updated model separates a linear trend within each group from the
[…]

### 2021-05-02 Code and miscellaneous/Tumor response GAM with comments.Rmd
Created RMarkdown document

---
title: "Tumor_Response_GAM_comments"
author: "Ariel Mundo"
date: "5/2/2021"
output: html_document
---

```{r,loading packages}
library(mgcv)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(gratia)
```

number of observations -- currently using 5
use measurement error of sd as 10 (rough estimate)

Data are simulated from Fig. 3 (C) of Vishwanath et. al:
Vishwanath, Karthik, et al. "Using optical spectroscopy to
longitudinally monitor physiological changes within solid tumors."
Neoplasia 11.9 (2009): 889-900.
https://www.sciencedirect.com/science/article/pii/S1476558609800420

```{r,simulating data}
dat <- read_csv(here::here("data", "Vishwanath StO2.csv"))
dat$Group<-as.factor(dat$Group)
## plot the mean response
ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
    geom_line()+
    geom_point()+
    scale_x_continuous(breaks=c(0,2,5,7,10))

#This function simulates data for the tumor data using a dafault number of
#10 observations per time point,
#and a standard deviation of 10% StO2
#Because physiologically StO2 cannot go below 0%, data is generated with a cutoff value of
#0.0001 (the "observation_tobit")

simulate_data <- function(dat, n = 10, sd = 5) {
    dat_sim <- dat %>%
        slice(rep(1:n(), each = n)) %>%
        group_by(Group, Day) %>%
        mutate(observation = rnorm(n, StO2, sd),
               observation_tobit = pmax(rnorm(n, StO2, sd), 0.0001)) %>%
        ungroup()

    return(dat_sim)
}

n <- 10
sd <- 10
set.seed(11) #setting seed
df <- 6
dat_sim <- simulate_data(dat, n, sd)

#plotting simulated data
ggplot(dat_sim, aes(x = Day, y = observation_tobit, color = Group)) +
    geom_point()+
    scale_x_continuous(breaks=c(0,2,5,7,10))

n_time <- length(unique(dat_sim$Day))
n_treatment <- length(unique(dat_sim$Group))

#if you have n timepoints can you do n knots or should you do n-1 knots? (I read somewhere that
#the n-1 option was preferred for computational issues)?

```

```{r,models}
mod_simple  <- gam(observation_tobit ~ s(Day,by=Group,k=5),data  = dat_sim)
appraise(mod2)
#SIMPLEST MODEL: A SMOOTH FOR TIME AS THE COVARIATE, WHICH CAN VARY BY GROUP BUT... APPRAISING THE MODEL SHOWS ISSUES ON THE RESIDUALS AND THE DATA DISTRIBUTION

#Model with interaction using tensor
mod_int1 <- gam(observation_tobit ~ Group+te(Day, by = Group, k = 5), data  = dat_sim)
appraise(mod1)

#Model with interaction using s
mod3 <- gam(observation_tobit ~ Group+s(Day, by = Group, k = 5), data  = dat_sim)
appraise(mod3)
```

Ok, so mod1 and mod3 are doing essentially the same thing. We have a smooth of time
per group, and we take into account the main effect of each treatment.
So this model effectively takes into account treatment and time? and the simples model only takes the effect of time.

"To avoid confounding problems the smooths are all subject to sum to zero constraints, which usually means that the main effect of g should also be included
in the model specification. For example, g + te(z,x,by=g)(pg.327)" (From Simon Wood's book)

The idea for this came from
https://stats.stackexchange.com/questions/486118/model-building-in-generalized-additive-mixed-models-gamms

```{r, plotting model}
#plotting smooths and simulated data
#full data
#I am doing something wrong here but I can't figure it out!!!!
#f_predict<-with(dat_sim,expand.grid(StO2=seq(min(observation_tobit),
#                                                        max(observation_tobit),
#                                                         length=10),
#                                                     Group=Group,Day=Day))

 #f_predict<-cbind(f_predict,
   #                       predict(mod2,f_predict,
  #                                se.fit = TRUE,type='response'))
#

#creates a dataframe wusing the lenght of the covariates
f_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = seq(0, 10, by = 0.1))

#adds the predictions to the grid and creates a confidence interval
f_predict<-f_predict%>%
    mutate(fit = predict(mod1,f_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(mod1, f_predict,se.fit = TRUE,type='response')$se.fit)

ggplot(data=dat_sim, aes(x=Day, y=observation_tobit, group=Group)) +
    facet_wrap(~Group)+
    geom_point(colour='black',size=1,alpha=0.5)+
    geom_ribbon(aes(ymin=(fit - 2*se.fit), ymax=(fit + 2*se.fit), x=Day),
                data=f_predict, alpha=0.5,
                inherit.aes=FALSE) +
    geom_line(aes(y=(fit),color=factor(Group)),
              size=1,data=f_predict,show.legend = FALSE)

#it works now, but why do the smooths still look so rough? (the connections at the knots
#don't look "smooth")

##trying the pairwise comparisons

pdat <- expand.grid(Day = seq(0, 10, length = 400),
                    Group = c('Treatment', 'Control'))
smooth_diff <- function(model, newdata, f1, f2, alpha = 0.05,
                        unconditional = FALSE) {
    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
    c1 <- grepl(f1, colnames(xp))
    c2 <- grepl(f2, colnames(xp))
    #r1 <- newdata[[var]] == f1
    #r2 <- newdata[[var]] == f2
    r1 <- with(newdata, Group == f1)
    r2 <- with(newdata, Group == f2)
    ## difference rows of xp for data from comparison
    X <- xp[r1, ] - xp[r2, ]
    ## zero out cols of X related to splines for other lochs
    X[, ! (c1 | c2)] <- 0
    ## zero out the parametric cols
    X[, !grepl('^s\\(', colnames(xp))] <- 0
    dif <- X %*% coef(model)
    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
    upr <- dif + (crit * se)
    lwr <- dif - (crit * se)
    data.frame(pair = paste(f1, f2, sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)
}
comp1<-smooth_diff(mod2,pdat,'Control','Treatment')
comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
                   rbind(comp1))

c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line(color='black',size=1) +
    facet_wrap(~ pair) +
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))

c1
    ##################################################################
############################################################
#missing data
#create a sequence of 40 random numbers between 1 and 100, these numbers will
#correspond to the row numbers to be randomly erased from the original dataset
missing <- sample(1:100, 40)

#create a new dataframe from the simulated data with 40 rows randomly removed
dat_missing <- dat_sim[-missing, ]

#Count the number of remaining observations per day (original dataset had 10 per group per day)
dat_missing %>%
    group_by(Day, Group) %>%
    summarize(count = n())

#plot the simulated data
ggplot(dat_missing, aes(x = Day, y = observation_tobit, color = Group)) +
    geom_point() +
    ggtitle("Simulated time-varying responses") +
    scale_color_viridis_d(end = 0.75) +
    geom_line(aes(y = StO2), alpha = 0.75) +
    theme_bw()

#the same model used for the full dataset
mod_m1 <- gam(observation_tobit ~ Group+s(Day,by=Group,k=5), data  = dat_missing)
#appraise the model
appraise(mod_m1)

m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = seq(0, 10, by = 0.1))

#adds the predictions to the grid and creates a confidence interval
m_predict<-m_predict%>%
    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)

ggplot(data=dat_missing, aes(x=Day, y=observation_tobit, group=Group)) +
    facet_wrap(~Group)+
    geom_point(colour='black',size=1,alpha=0.5)+
    geom_ribbon(aes(ymin=(fit - 2*se.fit), ymax=(fit + 2*se.fit), x=Day),
                data=m_predict, alpha=0.5,
                inherit.aes=FALSE) +
    geom_line(aes(y=(fit),color=factor(Group)),
              size=1,data=m_predict,show.legend = FALSE)

comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
comp_StO2_m <- cbind(Day = seq(0, 10, length = 400),
                   rbind(comp2))

c2<-ggplot(comp_StO2_m, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line(color='black',size=1) +
    facet_wrap(~ pair) +
    labs(x = 'Days', y = expression(paste('Difference in StO'[2],'(missing data)' )))

c2

c1+c2
```

### 2021-05-03 Code and miscellaneous/Tumor response GAM with comments.Rmd
Updated file, need to work on describing the pairwise interactions

@@
     geom_point()+
+    stat_summary(aes(y = observation_tobit,
+                     group=Group), 
+                 fun=mean, geom="line")+
     scale_x_continuous(breaks=c(0,2,5,7,10))
@@
 mod_simple  <- gam(observation_tobit ~ s(Day,by=Group,k=5),data  = dat_sim)
-appraise(mod2)
+appraise(mod_simple)
 #SIMPLEST MODEL: A SMOOTH FOR TIME AS THE COVARIATE, WHICH CAN VARY BY GROUP BUT... APPRAISING THE MODEL SHOWS ISSUES ON THE RESIDUALS AND THE DATA DISTRIBUTION
@@
 #Model with interaction using tensor
-mod_int1 <- gam(observation_tobit ~ Group+te(Day, by = Group, k = 5), data  = dat_sim)
+mod_int1 <- gam(observation_tobit ~ Group+te(Day,by=Group, k = 5), data  = dat_sim)
 appraise(mod1)
@@
 #Model with interaction using s
-mod3 <- gam(observation_tobit ~ Group+s(Day, by = Group, k = 5), data  = dat_sim)
-appraise(mod3)
+mod_int2 <- gam(observation_tobit ~ s(Group)+s(Day, by = Group, k = 5), data  = dat_sim)
+appraise(mod_int2)
 ```
 
-Ok, so mod1 and mod3 are doing essentially the same thing. We have a smooth of time
-per group, and we take into account the main effect of each treatment.
-So this model effectively takes into account treatment and time? and the simples model only takes the effect of time.
+The simplest model attempts to explain how StO2 changes over time, and gives each treatment group a separate smooth.
+However, from the diagnostics, it is clear that there are some issues with the assumptions, because the QQ plot, the residuals vs linear predictor and the histogram show that the assumptions of normality are not met.
+
+Next, we build a model that accounts for the _interaction_ of time and treatment.
+In this sense, mod_int1 and mod_int2 are doing essentially the same thing: We have a smooth to explain how StO2 changes per group, but we also account separately for the "main effect" of treatment.
+In other words, we are trying to explain the changes in StO2 by considering how it changes over time, but also disentangling the treatment effect from the response.
 

### 2021-05-03 Code and miscellaneous/Tumor response GAM with comments.Rmd
Updated model, seems the ~te option of gam has some different output that does not work with the function to make the pairwise-comparisons. Missing data example added as well

--- 2021-05-02 Code and miscellaneous/Tumor response GAM with comments.Rmd
@@
 library(gratia)
+library (ggsci)
 ```
@@
 Data are simulated from Fig. 3 (C) of Vishwanath et. al:
-Vishwanath, Karthik, et al. "Using optical spectroscopy to
-longitudinally monitor physiological changes within solid tumors."
-Neoplasia 11.9 (2009): 889-900.
+"Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors." Neoplasia 11.9 (2009): 889-900.
 https://www.sciencedirect.com/science/article/pii/S1476558609800420
+
+From the figure the maximum standard deviation for the measurements is approximately 10%.
 
@@
 dat$Group<-as.factor(dat$Group)
+
 ## plot the mean response
-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
-    geom_line()+
-    geom_point()+
+f1<-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
+    geom_line(size=1,show.legend = FALSE)+
+    geom_point(show.legend = FALSE)+
+  labs(y=expression(paste('Mean StO'[2],' (real)')))+
+  theme_classic()+
+  scale_color_aaas()+
     scale_x_continuous(breaks=c(0,2,5,7,10))
 
-#This function simulates data for the tumor data using a dafault number of
-#10 observations per time point,
-#and a standard deviation of 10% StO2
-#Because physiologically StO2 cannot go below 0%, data is generated with a cutoff value of
-#0.0001 (the "observation_tobit")
+#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
+#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
 
@@
         group_by(Group, Day) %>%
-        mutate(observation = rnorm(n, StO2, sd),
-               observation_tobit = pmax(rnorm(n, StO2, sd), 0.0001)) %>%
+        mutate(
+               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001)) %>%
         ungroup()
@@
 
-n <- 10
-sd <- 10
-set.seed(11) #setting seed
+n <- 10 #number of observations
+sd <- 10 #approximate sd from paper
+set.seed(1) #setting seed
 df <- 6
@@
 #plotting simulated data
-ggplot(dat_sim, aes(x = Day, y = observation_tobit, color = Group)) +
+f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group)) +
     geom_point()+
+    stat_summary(aes(y = StO2_sim,
+                     group=Group), 
+                 fun=mean, geom="line",size=1)+
+  labs(y=expression(paste('StO'[2],' (simulated)')))+
+  theme_classic()+
+  scale_color_aaas()+
     scale_x_continuous(breaks=c(0,2,5,7,10))
 
-n_time <- length(unique(dat_sim$Day))
-n_treatment <- length(unique(dat_sim$Group))
+#plot the "true" data with the simulated response
 
-#if you have n timepoints can you do n knots or should you do n-1 knots? (I read somewhere that
-#the n-1 option was preferred for computational issues)?
+f1+f2+plot_layout(guides = 'collect') &
+  ylim(c(0,75)) &
+  theme(
+     text=element_text(size=18),
+     legend.title=element_blank()
+     )&
+  scale_color_aaas()
 
@@
 ```{r,models}
-mod_simple  <- gam(observation_tobit ~ s(Day,by=Group,k=5),data  = dat_sim)
-appraise(mod2)
+mod_simple  <- gam(StO2_sim ~ s(Day,by=Group,k=5),data  = dat_sim)
+appraise(mod_simple)
 #SIMPLEST MODEL: A SMOOTH FOR TIME AS THE COVARIATE, WHICH CAN VARY BY GROUP BUT... APPRAISING THE MODEL SHOWS ISSUES ON THE RESIDUALS AND THE DATA DISTRIBUTION
 
-#Model with interaction using tensor
-mod_int1 <- gam(observation_tobit ~ Group+te(Day, by = Group, k = 5), data  = dat_sim)
-appraise(mod1)
+#Model with interaction using tensor: there is some different specification that does not allow to plot it easily
+mod_int1 <- gam(StO2_sim ~ Group+te(Day,by=Group, k = 5),
+                data  = dat_sim)
+appraise(mod_int1)
 
 #Model with interaction using s
-mod3 <- gam(observation_tobit ~ Group+s(Day, by = Group, k = 5), data  = dat_sim)
-appraise(mod3)
+mod_int2 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5), data  = dat_sim)
+appraise(mod_int2)
 ```
 
-Ok, so mod1 and mod3 are doing essentially the same thing. We have a smooth of time
-per group, and we take into account the main effect of each treatment.
-So this model effectively takes into account treatment and time? and the simples model only takes the effect of time.
+The simplest model attempts to explain how $StO_2$ changes over time, and gives each treatment group a separate smooth.
+However, from the diagnostics, it is clear that there are some issues because the QQ plot, the residuals vs linear predictor and the histogram show that the assumptions of normality are not met.
+
+Next, we build a model that accounts for the _interaction_ of time and treatment.
+In this sense, mod_int1 and mod_int2 are doing essentially the same thing: We have a smooth to explain how $StO_2$ changes per group, but we also account separately for the "main effect" of treatment.
+In other words, we are trying to explain the changes in $StO_2$ by considering how it changes over time, but also disentangling the treatment effect from the response.
 
@@
 #plotting smooths and simulated data
-#full data
-#I am doing something wrong here but I can't figure it out!!!!
-#f_predict<-with(dat_sim,expand.grid(StO2=seq(min(observation_tobit),
-#                                                        max(observation_tobit),
-#                                                         length=10),
-#                                                     Group=Group,Day=Day))
 
- #f_predict<-cbind(f_predict,
-   #                       predict(mod2,f_predict,
-  #                                se.fit = TRUE,type='response'))
-#
-
-#creates a dataframe wusing the lenght of the covariates
+#creates a dataframe using the length of the covariates
 f_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
@@
 f_predict<-f_predict%>%
-    mutate(fit = predict(mod1,f_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod1, f_predict,se.fit = TRUE,type='response')$se.fit)
+    mutate(fit = predict(mod_int2,f_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(mod_int2, f_predict,se.fit = TRUE,type='response')$se.fit)
 
-ggplot(data=dat_sim, aes(x=Day, y=observation_tobit, group=Group)) +
+#plot smooths and confidence interval
+f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
+    geom_point(aes(color=Group),size=1.5,alpha=0.5)+
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),fill=Group),
+              alpha=0.3,
+              data=f_predict,
+                inherit.aes=FALSE) +
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=f_predict,
+              show.legend = FALSE)+
   facet_wrap(~Group)+
-    geom_point(colour='black',size=1,alpha=0.5)+
-    geom_ribbon(aes(ymin=(fit - 2*se.fit), ymax=(fit + 2*se.fit), x=Day),
-                data=f_predict, alpha=0.5,
-                inherit.aes=FALSE) +
-    geom_line(aes(y=(fit),color=factor(Group)),
-              size=1,data=f_predict,show.legend = FALSE)
+  labs(y=expression(paste('StO'[2],' (simulated)')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+      scale_color_aaas()+
+  scale_fill_aaas()
 
-#it works now, but why do the smooths still look so rough? (the connections at the knots
-#don't look "smooth")
+f3 & theme(
+     text=element_text(size=18),
+     legend.title=element_blank()
+     )
 
-##trying the pairwise comparisons
+##Pairwise comparisons
 
@@
                     Group = c('Treatment', 'Control'))
+
 smooth_diff <- function(model, newdata, f1, f2, alpha = 0.05,
@@
 }
-comp1<-smooth_diff(mod2,pdat,'Control','Treatment')
+
+comp1<-smooth_diff(mod_int2,pdat,'Control','Treatment')
+
 comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
@@
 c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
+    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5,fill='#993336') +
     geom_line(color='black',size=1) +
     facet_wrap(~ pair) +
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))
+  theme_classic()+
+    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+  scale_x_continuous(breaks=c(0,2,5,7,10))+
+  theme(
+     text=element_text(size=18),
+     legend.title=element_blank()
+     )
 
 c1
+```
+
+GAMs do not require equally-spaced observations and can accommodate missing observations as well. In a case where 40% of the observations are missing:
+
+```{r,missing data}
     ##################################################################
@@
 
-#create a new dataframe from the simulated data with 40 rows randomly removed
-dat_missing <- dat_sim[-missing, ]
+#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
+
+ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
+
+#create a new dataframe, remove the StO2 column
+dat_missing <- dat_sim[,-1]
+
+#add NAs at the ind positions
+dat_missing$StO2_sim[ind]<-NA
 
@@
 dat_missing %>%
-    group_by(Day, Group) %>%
-    summarize(count = n())
+    group_by(Day,Group) %>%
+    filter(!is.na(StO2_sim))%>%
+  count(Day)
 
 #plot the simulated data
-ggplot(dat_missing, aes(x = Day, y = observation_tobit, color = Group)) +
+f4<-ggplot(dat_missing, aes(x = Day, y = StO2_sim, color = Group)) +
     geom_point() +
-    ggtitle("Simulated time-varying responses") +
-    scale_color_viridis_d(end = 0.75) +
-    geom_line(aes(y = StO2), alpha = 0.75) +
-    theme_bw()
+  stat_summary(aes(y = StO2_sim,
+                     group=Group), 
+                 fun=mean, geom="line",size=1)+
+    ggtitle("Missing data") +
+  labs(y='StO2')+
+  scale_color_aaas()+
+    theme_classic()
+
+f2+f4+plot_layout(guides = 'collect') &
+  ylim(c(0,75)) &
+  theme(
+     text=element_text(size=18),
+     legend.title=element_blank()
+     )&
+  scale_color_aaas()
 
 #the same model used for the full dataset
-mod_m1 <- gam(observation_tobit ~ Group+s(Day,by=Group,k=5), data  = dat_missing)
+mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing)
 #appraise the model
@@
 
-ggplot(data=dat_missing, aes(x=Day, y=observation_tobit, group=Group)) +
+ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
     facet_wrap(~Group)+
@@
     facet_wrap(~ pair) +
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2],'(missing data)' )))
+    labs(x = 'Days', y = expression(paste('Difference in StO'[2],'(missing data)' )))+
+  scale_x_continuous(breaks=c(0,2,5,7,10))+
+  theme_classic()+
+  theme(
+     text=element_text(size=18),
+     legend.title=element_blank()
+     )
 

### 2021-05-03 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
New v5 pdf created
Updated model, seems the ~te option of gam has some different output that does not work with the function to make the pairwise-comparisons. Missing data example added as well

--- 2021-04-15 Manuscripts/Manuscript v5/Manuscript_AM_v4.Rmd
@@
 
-spl<-basis_plot%>% group_by(time)%>%summarize(spl=sum(mod_val))
+spl<-basis_plot%>% 
+  group_by(time)%>%
+  summarize(spl=sum(mod_val))
 
@@
   theme(
-     text=element_text(size=18))
+     text=element_text(size=18)
+     )
 

### 2021-05-04 Code and miscellaneous/Tumor_response_GAM__comments.Rmd
Compiled pdf for RMarkdown document of GAM model for tumor data

--- 2021-05-03 Code and miscellaneous/Tumor response GAM with comments.Rmd
@@
 ---
-title: "Tumor_Response_GAM_comments"
+title: "Tumor Response GAM models with comments"
 author: "Ariel Mundo"
 date: "5/2/2021"
-output: html_document
+output:
+  pdf_document: default
+  html_document: default
 ---
 
-```{r,loading packages}
+```{r,loading packages,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
 library(mgcv)
@@
 
-```{r,simulating data}
+```{r,simulating data,cho=FALSE,include=FALSE,message=FALSE,warning=FALSE}
 dat <- read_csv(here::here("data", "Vishwanath StO2.csv"))
@@
 sd <- 10 #approximate sd from paper
-set.seed(1) #setting seed
+set.seed(1) #set seed for reproducibility
 df <- 6
@@
 #Model with interaction using s
-mod_int2 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5), data  = dat_sim)
+mod_int2 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),method='REML', data  = dat_sim)
 appraise(mod_int2)
@@
 "To avoid confounding problems the smooths are all subject to sum to zero constraints, which usually means that the main effect of g should also be included
-in the model specification. For example, g + te(z,x,by=g)(pg.327)" (From Simon Wood's book)
+in the model specification. For example, g + te(z,x,by=g)(pg.327)" (From Simon Wood's book). This gives a good explanation on why mod_int1 and mod_int2 work.
 
@@
 
-```{r, plotting model}
+```{r, plotting model,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
 #plotting smooths and simulated data
@@
 
-```{r,missing data}
+```{r,missing data,echo=FALSE,include=TRUE,message=FALSE,warning=FALSE}
     ##################################################################
@@
     facet_wrap(~ pair) +
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2],'(missing data)' )))+
+    labs(x = 'Days', 
+         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
+                              )))+
   scale_x_continuous(breaks=c(0,2,5,7,10))+
@@
 ```
+
+It is reassuring to see that we get the same diagnostics by either having a dataframe with missing observations as NA or a dataframe where the missing observations are not present at all.
+
+So two questions remains after this:
+-How do we want to present the pairwise smooth confidence interval comparison? 
+The GAMs are referred as "empirical Bayesian", but I could not find a straightforward reference that had information in detail of what the pairwise comparisons can tell at certain timepoints.
+Sure, they are different, but what else?
+
+-Do we want to address prediction? I recently saw a webinar from Gavin Simpson regarding this, wondered if it would be feasible to include in a concise manner.

### 2021-05-05 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
Working on last section

@@
 
-Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates a difference in the effect of each treatment.
+Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates an effect due to treatment.
 In other words, Group 1 can be thought as a "Control" group and Group 2 as a "Treatment" group.
@@
 
-## GAMs and covariance
+## The analyisis of longitudinal biomedical data using GAMs
 
+The previous section provided the basic theoretical framework to understand the advantages of GAMs to model non-linear data.
 Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
@@
 In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the _time_ effect to account for the repeated measures over time.
-An example where this is covered is in the Appendix.
+Finally, GAMs can handle missing observations that are MAR or MCAR.
+To demonstrate these points, this section will use simulation to generate data that follows previously reported non-linear trends in the biomedical literature.
+The simulated data takes as a starting point the reported longitudinal changes in oxygen saturation in subcutaneous that are reported in Figure 3(C) in [vishwanath2009].
+An increase in oxygen saturation in the treatment group is observed through time, while a decrease is seen in the Control group.
+Following the reported trend, 10 normally distributed observations at each time point with a standard deviation of 10% were simulated.
 

### 2021-05-06 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
Working on manuscript on the GAM section

@@
 The previous section provided the basic theoretical framework to understand the advantages of GAMs to model non-linear data.
-Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
+Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
 A detailed description on basis functions and correlations can be found in [@hefley2017].
@@
 To demonstrate these points, this section will use simulation to generate data that follows previously reported non-linear trends in the biomedical literature.
-The simulated data takes as a starting point the reported longitudinal changes in oxygen saturation in subcutaneous that are reported in Figure 3(C) in [vishwanath2009].
+
+The simulated data is based on the reported longitudinal changes in oxygen saturation ($StO_2$) in subcutaneous tumors under chemotherapy  that are reported in Figure 3(C) in [vishwanath2009].
 An increase in oxygen saturation in the treatment group is observed through time, while a decrease is seen in the Control group.
-Following the reported trend, 10 normally distributed observations at each time point with a standard deviation of 10% were simulated.
+Following the reported trend, 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (approximately matching the SD in the figure) are simulated.
+The "real" and simulated data appear in Figure \@ref(fig:sim-smooth-plot) A,B.
+
+```{r,sim-tumor-response,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
+
+dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
+            Day=rep(c(0,2,5,7,10),times=2),
+            Group=as.factor(rep(c("Control","Treatment"),each=5))
+            )
+
+## plot the mean response
+f1<-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
+    geom_line(size=1,show.legend = FALSE)+
+    geom_point(show.legend = FALSE)+
+  labs(y=expression(paste('Mean StO'[2],'\n (real)')))+
+  theme_classic()+
+  scale_color_aaas()+
+    scale_x_continuous(breaks=c(0,2,5,7,10))
+
+#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
+#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
+
+simulate_data <- function(dat, n = 10, sd = 5) {
+    dat_sim <- dat %>%
+        slice(rep(1:n(), each = n)) %>%
+        group_by(Group, Day) %>%
+        mutate(
+               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001)) %>%
+        ungroup()
+
+    return(dat_sim)
+}
+
+n <- 10 #number of observations
+sd <- 10 #approximate sd from paper
+set.seed(1) #set seed for reproducibility
+df <- 6
+dat_sim <- simulate_data(dat, n, sd)
+
+#plotting simulated data
+f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group)) +
+    geom_point(show.legend=FALSE)+
+    stat_summary(aes(y = StO2_sim,
+                     group=Group), 
+                 fun=mean, geom="line",size=1,show.legend = FALSE)+
+  labs(y=expression(paste('StO'[2],'\n (simulated)')))+
+  theme_classic()+
+  scale_color_aaas()+
+    scale_x_continuous(breaks=c(0,2,5,7,10))
+
+tumor_sim<-f1+f2+plot_layout(guides = 'collect')+
+  plot_annotation(tag_levels='A')&
+  ylim(c(0,75)) &
+  theme(
+     text=element_text(size=18),
+     legend.title=element_blank(),
+     legend.text = element_blank()
+     )&
+  scale_color_aaas()
+
+```
+
+For the data in Figure \@ref(fig:sim-smooth-plot),B it is of interest to know the effect that each treatment has in oxygen saturation.
+In a practical sense, this implies an interaction model where smooths are created for _Group_ and _Day_, respectively.
+The main thing to consider is that the interaction needs to be properly structured in the model syntax to account for the fact that one of the variables is numeric ( _Day_ ) and that the other is a factor ( _Group_ ).
+Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term.
+Using the package `mgcv` The model then becomes:
+
+`m1<-gam(StO2~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim`
+
+This syntax specifies that the object `m1` will store the model, and that the change in $StO_2$ is modeled by independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
+The parametric term _Group_ is added to separate the effect of treatment, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
+When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $StO_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),C).
+
+```{r,smooth-plot,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
+
+#GAM for StO2
+
+m_1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),method='REML', data  = dat_sim)
+
+#creates a dataframe using the length of the covariates
+f_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = seq(0, 10, by = 0.1))
+
+#adds the predictions to the grid and creates a confidence interval
+f_predict<-f_predict%>%
+    mutate(fit = predict(m_1,f_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(m_1, f_predict,se.fit = TRUE,type='response')$se.fit)
+
+#plot smooths and confidence interval
+f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
+    geom_point(aes(color=Group),size=1.5,alpha=0.5)+
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),fill=Group),
+              alpha=0.3,
+              data=f_predict,
+                inherit.aes=FALSE) +
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=f_predict,
+              show.legend = FALSE)+
+  #facet_wrap(~Group)+
+  labs(y=expression(paste('StO'[2],' (simulated)')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+      scale_color_aaas()+
+  scale_fill_aaas()
+
+sim_smooth_plot<-f3 & theme(
+     text=element_text(size=18),
+     legend.title=element_blank()
+     )
+```
+
+(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors.
+A: Real data that follows previously reported trends in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+B: Simulated data from  a normal distribution with standard deviation of 10% with 10 observations per time point.
+In A and B lines indicate mean oxygen saturation C: Smooths from the GAM model with interaction of Group and Treatment.Lines represent smooths for each group, shaded regions are 95% confidence intervals.
+
+```{r,sim-smooth-plot,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}
+
+tumor_sim+sim_smooth_plot
+
+```
+
+```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
+##Pairwise comparisons
+
+# pdat <- expand.grid(Day = seq(0, 10, length = 400),
+#                     Group = c('Treatment', 'Control'))
+# 
+# smooth_diff <- function(model, newdata, f1, f2, alpha = 0.05,
+#                         unconditional = FALSE) {
+#     xp <- predict(model, newdata = newdata, type = 'lpmatrix')
+#     c1 <- grepl(f1, colnames(xp))
+#     c2 <- grepl(f2, colnames(xp))
+#     #r1 <- newdata[[var]] == f1
+#     #r2 <- newdata[[var]] == f2
+#     r1 <- with(newdata, Group == f1)
+#     r2 <- with(newdata, Group == f2)
+#     ## difference rows of xp for data from comparison
+#     X <- xp[r1, ] - xp[r2, ]
+#     ## zero out cols of X related to splines for other lochs
+#     X[, ! (c1 | c2)] <- 0
+#     ## zero out the parametric cols
+#     X[, !grepl('^s\\(', colnames(xp))] <- 0
+#     dif <- X %*% coef(model)
+#     se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
+#     crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
+#     upr <- dif + (crit * se)
+#     lwr <- dif - (crit * se)
+#     data.frame(pair = paste(f1, f2, sep = '-'),
+#                diff = dif,
+#                se = se,
+#                upper = upr,
+#                lower = lwr)
+# }
+# 
+# comp1<-smooth_diff(mod_int2,pdat,'Control','Treatment')
+# 
+# comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
+#                    rbind(comp1))
+# 
+# c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
+#     geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5,fill='#993336') +
+#     geom_line(color='black',size=1) +
+#     facet_wrap(~ pair) +
+#   theme_classic()+
+#     labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+#   scale_x_continuous(breaks=c(0,2,5,7,10))+
+#   theme(
+#      text=element_text(size=18),
+#      legend.title=element_blank()
+#      )
+#   
+# 
+# 
+# c1
+
+```
 

### 2021-05-06 Manuscripts/Manuscript v5/Manuscript_AM_v5.tex
Working on manuscript on the GAM section

@@
 
-Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates a difference in the effect of each treatment.
+Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates an effect due to treatment.
 In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group.
@@
 
-\hypertarget{gams-and-covariance}{%
-\subsection{GAMs and covariance}\label{gams-and-covariance}}
-
-Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
+\hypertarget{the-analyisis-of-longitudinal-biomedical-data-using-gams}{%
+\subsection{The analyisis of longitudinal biomedical data using GAMs}\label{the-analyisis-of-longitudinal-biomedical-data-using-gams}}
+
+The previous section provided the basic theoretical framework to understand the advantages of GAMs to model non-linear data.
+Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
 A detailed description on basis functions and correlations can be found in {[}52{]}.
 In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the \emph{time} effect to account for the repeated measures over time.
-An example where this is covered is in the Appendix.
+Finally, GAMs can handle missing observations that are MAR or MCAR.
+To demonstrate these points, this section will use simulation to generate data that follows previously reported non-linear trends in the biomedical literature.
+
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(StO_2\)) in subcutaneous tumors under chemotherapy that are reported in Figure 3(C) in {[}vishwanath2009{]}.
+An increase in oxygen saturation in the treatment group is observed through time, while a decrease is seen in the Control group.
+Following the reported trend, 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (approximately matching the SD in the figure) are simulated.
+The ``real'' and simulated data appear in Figure \ref{fig:sim-smooth-plot} A,B.
+
+For the data in Figure \ref{fig:sim-smooth-plot},B it is of interest to know the effect that each treatment has in oxygen saturation.
+In a practical sense, this implies an interaction model where smooths are created for \emph{Group} and \emph{Day}, respectively.
+The main thing to consider is that the interaction needs to be properly structured in the model syntax to account for the fact that one of the variables is numeric ( \emph{Day} ) and that the other is a factor ( \emph{Group} ).
+Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term.
+Using the package \texttt{mgcv} The model then becomes:
+
+\texttt{m1\textless{}-gam(StO2\textasciitilde{}Group+s(Day,by=Group,k=5),\ method=\textquotesingle{}REML\textquotesingle{},data=dat\_sim}
+
+This syntax specifies that the object \texttt{m1} will store the model, and that the change in \(StO_2\) is modeled by independent smooths for \emph{Group} and \emph{Day} (the parenthesis preceded by \emph{s}) using 5 knots.
+The parametric term \emph{Group} is added to separate the effect of treatment, and the \texttt{method} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}37{]}.
+When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(StO_2\) for each group across time (Figure \ref{fig:sim-smooth-plot},C).
+
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v5_files/figure-latex/sim-smooth-plot-1}
+
+}
+
+\caption{Simulated data and smooths for oxygen saturation in tumors.
+A: Real data that follows previously reported trends in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+B: Simulated data from a normal distribution with standard deviation of 10\% with 10 observations per time point.
+In A and B lines indicate mean oxygen saturation C: Smooths from the GAM model with interaction of Group and Treatment.Lines represent smooths for each group, shaded regions are 95\% confidence intervals.}\label{fig:sim-smooth-plot}
+\end{figure}
 

### 2021-05-06 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
Wroking on last section

@@
 
-```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap = "Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+(ref:l-q-response-caption) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
 A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
@@
 Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions."}
+The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.
+
+```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='(l-q-response-caption)' }
 # linear response, correlated errors (subject effect)
@@
 
-```{r,basis-plot,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap="Basis functions for a single smoother for time with five knots.
+(ref:basis-plot-caption) Basis functions for a single smoother for time with five knots.
 A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
@@
 D.
-Smoother for time and original datapoints.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points."}
+Smoother for time and original data points.
+The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.
+
+```{r,basis-plot,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap='(ref:basis-plot-caption)'}
 
@@
 The main thing to consider is that the interaction needs to be properly structured in the model syntax to account for the fact that one of the variables is numeric ( _Day_ ) and that the other is a factor ( _Group_ ).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term.
-Using the package `mgcv` The model then becomes:
-
-`m1<-gam(StO2~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim`
-
-This syntax specifies that the object `m1` will store the model, and that the change in $StO_2$ is modeled by independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
+Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
+Using $\textsf{R}$ and the package `mgcv` the model syntax is:
+
+`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim`
+
+This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation `StO2_sim` is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
 The parametric term _Group_ is added to separate the effect of treatment, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $StO_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),C).
+Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package `gratia`[@gratia].
+A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].
 
@@
 
-m_1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),method='REML', data  = dat_sim)
+m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),method='REML', data  = dat_sim)
 
@@
 f_predict<-f_predict%>%
-    mutate(fit = predict(m_1,f_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(m_1, f_predict,se.fit = TRUE,type='response')$se.fit)
+    mutate(fit = predict(m1,f_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(m1, f_predict,se.fit = TRUE,type='response')$se.fit)
 
@@
 
+```
+
+Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data that has missing observations.
+Again, using the simulated $StO_2$ values from Figure (\@ref(fig:sim-smooth-plot), B), 40% of the total observations can be randomly deleted and a GAM model with interaction can be fitted to the remaining observations.
+
+```{r,missing-data,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
+#missing data
+#create a sequence of 40 random numbers between 1 and 100, these numbers will
+#correspond to the row numbers to be randomly erased from the original dataset
+missing <- sample(1:100, 40)
+
+#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
+
+ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
+
+#create a new dataframe, remove the StO2 column
+dat_missing <- dat_sim[,-1]
+
+#add NAs at the ind positions
+dat_missing$StO2_sim[ind]<-NA
+
+#Count the number of remaining observations per day (original dataset had 10 per group per day)
+dat_missing %>%
+    group_by(Day,Group) %>%
+    filter(!is.na(StO2_sim))%>%
+  count(Day)
+
+#plot the simulated data
+f4<-ggplot(dat_missing, aes(x = Day, y = StO2_sim, color = Group)) +
+    geom_point() +
+  stat_summary(aes(y = StO2_sim,
+                     group=Group), 
+                 fun=mean, geom="line",size=1)+
+    ggtitle("Missing data") +
+  labs(y='StO2')+
+  scale_color_aaas()+
+    theme_classic()
+
+f2+f4+plot_layout(guides = 'collect') &
+  ylim(c(0,75)) &
+  theme(
+     text=element_text(size=18),
+     legend.title=element_blank()
+     )&
+  scale_color_aaas()
+
+#the same model used for the full dataset
+mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing)
+#appraise the model
+appraise(mod_m1)
+
+m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = seq(0, 10, by = 0.1))
+
+#adds the predictions to the grid and creates a confidence interval
+m_predict<-m_predict%>%
+    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
+
+ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
+    facet_wrap(~Group)+
+    geom_point(colour='black',size=1,alpha=0.5)+
+    geom_ribbon(aes(ymin=(fit - 2*se.fit), ymax=(fit + 2*se.fit), x=Day),
+                data=m_predict, alpha=0.5,
+                inherit.aes=FALSE) +
+    geom_line(aes(y=(fit),color=factor(Group)),
+              size=1,data=m_predict,show.legend = FALSE)
+
+comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
+comp_StO2_m <- cbind(Day = seq(0, 10, length = 400),
+                   rbind(comp2))
+
+missing_plot<-ggplot(comp_StO2_m, aes(x = Day, y = diff, group = pair)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
+    geom_line(color='black',size=1) +
+    facet_wrap(~ pair) +
+    labs(x = 'Days', 
+         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
+                              )))+
+  scale_x_continuous(breaks=c(0,2,5,7,10))+
+  theme_classic()+
+  theme(
+     text=element_text(size=18),
+     legend.title=element_blank()
+     )
 ```

### 2021-05-07 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
Working on model for lme in simulated data

@@
 ```{r, message = FALSE,include=FALSE}
-set.seed(11)
 library(patchwork)
 library(tidyverse)
+library(mvnfast)
+library(nlme)
+library(ggsci)
+library(mgcv)
 
@@
     stop('fun_type must be either "correlated", or "independent"')
-
-  library(tidyverse)
-  library(mvnfast)
-  library(nlme)
-  library(ggsci)
 
@@
 
-library(mgcv)
-
 n_time = 6
@@
 
-## The analyisis of longitudinal biomedical data using GAMs
-
-The previous section provided the basic theoretical framework to understand the advantages of GAMs to model non-linear data.
-Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
+# The analyisis of longitudinal biomedical data using GAMs
+
+The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA.
+This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using $\textsf{R}$ and the package `mgcv`.
+The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
+
+## Simulated data
+
+The simulated data is based on the reported longitudinal changes in oxygen saturation ($StO_2$) in subcutaneous tumors that appear in Figure 3(C) in [vishwanath2009].
+An increase in oxygen saturation in the "Treatment" group (chemotherapy) is observed through time, while a decrease is seen in the "Control" (saline) group.
+Following the reported trend, 10 normally distributed observations at each time point are generated with a standard deviation (SD) of 10% (matching the SD in the original paper).
+The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot), A and the inlet, respectively.
+
+## An interaction GAM model for longitudinal data
+
+An interaction model is typically the main interest in longitudinal biomedical data, as it takes into account the effect of treatment, time, and their combination.
+In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the _time_ effect to account for the repeated measures over time.
+Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
 A detailed description on basis functions and correlations can be found in [@hefley2017].
-In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the _time_ effect to account for the repeated measures over time.
-Finally, GAMs can handle missing observations that are MAR or MCAR.
-To demonstrate these points, this section will use simulation to generate data that follows previously reported non-linear trends in the biomedical literature.
-
-The simulated data is based on the reported longitudinal changes in oxygen saturation ($StO_2$) in subcutaneous tumors under chemotherapy  that are reported in Figure 3(C) in [vishwanath2009].
-An increase in oxygen saturation in the treatment group is observed through time, while a decrease is seen in the Control group.
-Following the reported trend, 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (approximately matching the SD in the figure) are simulated.
-The "real" and simulated data appear in Figure \@ref(fig:sim-smooth-plot) A,B.
+
+For the data in Figure \@ref(fig:sim-smooth-plot),A it is of interest to know the effect that each treatment has in oxygen saturation.
+The model then needs to incorporate independent smooths for _Group_ and _Day_, respectively.
+The main thing to consider is the model syntax to account for the fact that one of the variables is numeric ( _Day_ ) and that the other is a factor ( _Group_ ).
+Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
+Using $\textsf{R}$ and the package `mgcv` the model syntax is:
+
+`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim`
+
+This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation `StO2_sim` is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
+The parametric term _Group_ is added to separate the effect of treatment, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
+When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $StO_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B).
+Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package `gratia`[@gratia].
+A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].
+
+One question might be "what is the fit that an rm-ANOVA model produces for the simulated data?".
+The rm-ANOVA model, which corresponds to \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot),C.
 
@@
   scale_color_aaas()+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+  plot_layout(tag_level = 'new')
 
@@
         mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001)) %>%
+               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
+               subject=rep(1:10),
+               subject=factor(paste(subject, Group, sep = "-"))
+               ) %>%
         ungroup()
@@
 }
+
+#subject = factor(paste(subject, treatment, sep = "-")))
 
@@
 
-tumor_sim<-f1+f2+plot_layout(guides = 'collect')+
-  plot_annotation(tag_levels='A')&
-  ylim(c(0,75)) &
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank(),
-     legend.text = element_blank()
-     )&
-  scale_color_aaas()
-
-```
-
-For the data in Figure \@ref(fig:sim-smooth-plot),B it is of interest to know the effect that each treatment has in oxygen saturation.
-In a practical sense, this implies an interaction model where smooths are created for _Group_ and _Day_, respectively.
-The main thing to consider is that the interaction needs to be properly structured in the model syntax to account for the fact that one of the variables is numeric ( _Day_ ) and that the other is a factor ( _Group_ ).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
-Using $\textsf{R}$ and the package `mgcv` the model syntax is:
-
-`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim`
-
-This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation `StO2_sim` is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
-The parametric term _Group_ is added to separate the effect of treatment, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $StO_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),C).
-Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package `gratia`[@gratia].
-A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].
+# tumor_sim<-f1+f2+plot_layout(guides = 'collect')+
+#   plot_annotation(tag_levels='A')&
+#   ylim(c(0,75)) &
+#   theme(
+#      text=element_text(size=18),
+#      legend.title=element_blank(),
+#      legend.text = element_blank()
+#      )&
+#   scale_color_aaas()
+
+```
 
@@
 
-m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),method='REML', data  = dat_sim)
-
-#creates a dataframe using the length of the covariates
-f_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
+gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),method='REML', data  = dat_sim)
+
+#linear model
+lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
+
+lm1<-lme(StO2_sim ~ Group + Day + Group:Day,
+                 data = dat_sim,
+                 random = ~ 1 | subject,
+                 correlation = corCompSymm(form = ~ 1 | subject)
+  )
+
+#creates a dataframe using the length of the covariates for the GAM
+gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
                          Day = seq(0, 10, by = 0.1))
 
-#adds the predictions to the grid and creates a confidence interval
-f_predict<-f_predict%>%
-    mutate(fit = predict(m1,f_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(m1, f_predict,se.fit = TRUE,type='response')$se.fit)
-
-#plot smooths and confidence interval
+#creates a dataframe using the length of the covariates for the rm-ANOVA
+lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = seq(0, 10, by = 0.1),
+                        subject=factor(rep(1:10)),
+                          )
+lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
+
+#adds the predictions to the grid and creates a confidence interval for GAM
+gam_predict<-gam_predict%>%
+    mutate(fit = predict(gam1,gam_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(gam1, gam_predict,se.fit = TRUE,type='response')$se.fit)
+
+lm_predict<-lm_predict%>%
+    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
+
+#plot smooths and confidence interval for GAM
 f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
@@
               alpha=0.3,
-              data=f_predict,
+              data=gam_predict,
                 inherit.aes=FALSE) +
@@
 
+#plot linear fit for rm-ANOVA
+f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
+    geom_point(aes(color=Group),size=1.5,alpha=0.5)+
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),fill=Group),
+              alpha=0.3,
+              data=lm_predict,
+                inherit.aes=FALSE) +
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=lm_predict,
+              show.legend = FALSE)+
+  #facet_wrap(~Group)+
+  labs(y=expression(paste('StO'[2],' (simulated)')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+      scale_color_aaas()+
+  scale_fill_aaas()
+
 sim_smooth_plot<-f3 & theme(
+     text=element_text(size=18),
+     legend.title=element_blank()
+     )
+
+sim_linear_plot<-f4 & theme(
      text=element_text(size=18),
@@
 
-tumor_sim+sim_smooth_plot
+f2+inset_element(f1, left = 0.01, bottom = 0.7, right = 0.5, top = 1.0)+f3+f4+
+  plot_layout(guides = 'collect')+
+   plot_annotation(tag_levels='A')&
+   ylim(c(0,75)) &
+   theme(
+      text=element_text(size=18),
+      legend.title=element_blank(),
+      legend.text = element_blank()
+      )&
+  scale_color_aaas()
 

### 2021-05-09 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
Working on post-hoc comparisons

@@
 
-lm1<-lme(StO2_sim ~ Group + Day + Group:Day,
-                 data = dat_sim,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
+#lm2<-lme(StO2_sim ~ Group + Day + Group:Day,
+#                 data = dat_sim,
+#                 random = ~ 1 | subject,
+#                 correlation = corCompSymm(form = ~ 1 | subject)
+#  )
 
@@
 lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
+                         Day = c(0,2,5,7,10),
                         subject=factor(rep(1:10)),
@@
 
+#using lm
 lm_predict<-lm_predict%>%
@@
            se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
+
+#using lme
+#lm_predict<-lm_predict%>%
+#   mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response',level=0),
+ #         se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response',level=0))
 
@@
                 color=Group),
-              size=1,data=f_predict,
+              size=1,data=gam_predict,
               show.legend = FALSE)+
@@
      )
+
+#posthoc comparisons for the linear model
+library(multcomp)
+
+summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
+
 ```

### 2021-05-09 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
working on Posthocs
Time as factor does follow the trend?

@@
 #linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
+lm1<-lm(StO2_sim ~ factor(Day) + Group + factor(Day) * Group, data = dat_sim)
 
@@
 
+summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
 summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
@@
 ```
+
+What does this mean?
+https://stats.stackexchange.com/questions/377609/rm-anova-time-as-continuous-or-categorical
+
+If you treat time as a factor then you do get the changes over time!
 

### 2021-05-11 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
working on LMEM vs GAM part
Working on confidence interval, need to center the comparison to zero

--- 2021-05-09 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
@@
 
+# Abstract
+
+In biomedical research, the outcome of longitudinal studies is frequently analyzed statistically using the repeated measures analysis or variance (rm-ANOVA).
+However, rm-ANOVA is restrictive as is only able to accommodate linear responses, requires equal correlation between each measurement (compound symmetry), and cannot work with missing observations.
+When the data fails to fullfill all the required assumptions by rm-ANOVA, the model provides biased estimates and inference.
+In contrast, generalized additive models (GAMs) are a class of models that overcome all of the limitations imposed by rm-ANOVA and that can be implemented in a straightforward manner using the _mgcv_ package in R.
+This paper summarizes the assumptions of rm-ANOVA and how their limit its use in longitudinal biomedical research, presents the basic theoretical framework for GAMs and provides examples of their implementation in R.
+The code and data used to generate this paper are available at:
+
 # Background
@@
 library(mgcv)
+library(gratia)
 
@@
 
-```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='(l-q-response-caption)' }
+```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='ref(l-q-response-caption)' }
 # linear response, correlated errors (subject effect)
@@
 Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
-D.
-Smoother for time and original data points.
+D: Smoother for time and original data points.
 The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.
@@
 The rm-ANOVA model, which corresponds to \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot),C.
+This is a typical case of model misspecification: The slopes of each group ar different, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \@ref(fig:sim-smooth-plot),B) .
 
@@
     geom_point(show.legend = FALSE)+
-  labs(y=expression(paste('Mean StO'[2],'\n (real)')))+
+  labs(y=expression(paste(StO[2],' (real)')))+
   theme_classic()+
   scale_color_aaas()+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-  plot_layout(tag_level = 'new')
+    scale_x_continuous(breaks=c(0,5,10))+
+    scale_y_continuous(breaks=c(0,40))+
+  plot_layout(tag_level = 'new')+
+  theme(
+    plot.background = element_rect(fill = "transparent", color = NA),
+    axis.text=element_text(size=14)
+  )
 
@@
                  fun=mean, geom="line",size=1,show.legend = FALSE)+
-  labs(y=expression(paste('StO'[2],'\n (simulated)')))+
-  theme_classic()+
+  labs(y=expression(atop(StO[2],'(simulated)')))+
+  theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
   scale_color_aaas()+
@@
 
-gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),method='REML', data  = dat_sim)
+gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
+            method='REML',
+            data  = dat_sim)
 
@@
 
-#lm2<-lme(StO2_sim ~ Group + Day + Group:Day,
-#                 data = dat_sim,
-#                 random = ~ 1 | subject,
-#                 correlation = corCompSymm(form = ~ 1 | subject)
-#  )
-
 #creates a dataframe using the length of the covariates for the GAM
 gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#creates a dataframe using the length of the covariates for the rm-ANOVA
+                         Day = seq(0, 10, by = 0.1),
+                         subject=factor(rep(1:10)))
+
+#creates a dataframe using the length of the covariates for rm-ANOVA
 lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0,2,5,7,10),
+                         Day = c(0:10),
                         subject=factor(rep(1:10)),
@@
 f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5)+
+    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
   geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
+                   ymax=(fit + 2*se.fit),
+                   fill=Group
+                   ),
               alpha=0.3,
               data=gam_predict,
+            show.legend=FALSE,
                 inherit.aes=FALSE) +
@@
   #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
+  labs(y=expression(atop(StO[2],'complete')))+
     scale_x_continuous(breaks=c(0,2,5,7,10))+
       theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
       scale_color_aaas()+
@@
 f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5)+
+    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
   geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
@@
               data=lm_predict,
+              show.legend = FALSE,
                 inherit.aes=FALSE) +
@@
       theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
       scale_color_aaas()+
@@
 
-sim_smooth_plot<-f3 & theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
-sim_linear_plot<-f4 & theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
 #posthoc comparisons for the linear model
@@
 
-(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors.
-A: Real data that follows previously reported trends in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-B: Simulated data from  a normal distribution with standard deviation of 10% with 10 observations per time point.
-In A and B lines indicate mean oxygen saturation C: Smooths from the GAM model with interaction of Group and Treatment.Lines represent smooths for each group, shaded regions are 95% confidence intervals.
-
-```{r,sim-smooth-plot,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}
-
-f2+inset_element(f1, left = 0.01, bottom = 0.7, right = 0.5, top = 1.0)+f3+f4+
-  plot_layout(guides = 'collect')+
-   plot_annotation(tag_levels='A')&
-   ylim(c(0,75)) &
-   theme(
-      text=element_text(size=18),
-      legend.title=element_blank(),
-      legend.text = element_blank()
-      )&
-  scale_color_aaas()
-
-```
-
-Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data that has missing observations.
-Again, using the simulated $StO_2$ values from Figure (\@ref(fig:sim-smooth-plot), B), 40% of the total observations can be randomly deleted and a GAM model with interaction can be fitted to the remaining observations.
+Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
+Using the simulated $StO_2$ values from Figure (\@ref(fig:sim-smooth-plot), B), 40% of the total observations can be randomly deleted and the same GAM model with interaction used for the full simulated data can be fitted to the remaining observations.
 
@@
 
-#plot the simulated data
-f4<-ggplot(dat_missing, aes(x = Day, y = StO2_sim, color = Group)) +
-    geom_point() +
-  stat_summary(aes(y = StO2_sim,
-                     group=Group), 
-                 fun=mean, geom="line",size=1)+
-    ggtitle("Missing data") +
-  labs(y='StO2')+
-  scale_color_aaas()+
-    theme_classic()
-
-f2+f4+plot_layout(guides = 'collect') &
-  ylim(c(0,75)) &
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )&
-  scale_color_aaas()
-
 #the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing)
+mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
 #appraise the model
@@
 
-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    facet_wrap(~Group)+
-    geom_point(colour='black',size=1,alpha=0.5)+
-    geom_ribbon(aes(ymin=(fit - 2*se.fit), ymax=(fit + 2*se.fit), x=Day),
-                data=m_predict, alpha=0.5,
+f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
+    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),
+                   fill=Group
+                   ),
+              alpha=0.3,
+              data=m_predict,
+            show.legend=FALSE,
                 inherit.aes=FALSE) +
-    geom_line(aes(y=(fit),color=factor(Group)),
-              size=1,data=m_predict,show.legend = FALSE)
-
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=m_predict,
+              show.legend = TRUE)+
+  #facet_wrap(~Group)+
+  labs(y=expression(atop(StO[2],'missing')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
+      scale_color_aaas()+
+  scale_fill_aaas()
+
+```
+
+(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors.
+A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point.
+Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
+Lines represent smooths for each group, shaded regions are 95% confidence intervals.C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
+D: Smooths for the GAM model for the simulated data with 40% of its observations missing.
+Lines represent smooths for each group, shaded regions are 95% confidence intervals
+
+```{r,sim-smooth-plot,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}
+
+mult_plot<-f2+inset_element(
+  f1, left = 0.01, 
+  bottom = 0.5, 
+  right = 0.5, 
+  top = 1.0)+
+  f3+f4+f6+
+  #plot_layout(guides = 'collect')+
+   plot_annotation(tag_levels='A')&
+   ylim(c(-5,75)) &
+  theme(
+     text=element_text(size=18)
+     )&
+  scale_color_aaas()
+
+mult_plot
+
+```
+
+## Determination of significance in GAMs
+
+At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
+Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs use a different approach to estimate significance.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that timepoint (or timepoints).
+The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
+The major advantage of the significance estimation in GAMs  is that they allow to determine significance at a specific timepoint, rather than providing a single _p-value_ for the overall treatment effect.
+The estimation of siginificant differences in GAMs is covered in detail in the Appendix using simulated longitudinal data that follows previously reported trends of tumor response between multiple treatment groups [@vishwanath2009].
+
+```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
+##Pairwise comparisons
+
+pdat <- expand.grid(Day = seq(0, 10, length = 400),
+                    Group = c('Treatment', 'Control'))
+
+smooth_diff <- function(model, newdata, f1, f2, alpha = 0.05,
+                        unconditional = FALSE) {
+    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
+    c1 <- grepl(f1, colnames(xp))
+    c2 <- grepl(f2, colnames(xp))
+    #r1 <- newdata[[var]] == f1
+    #r2 <- newdata[[var]] == f2
+    r1 <- with(newdata, Group == f1)
+    r2 <- with(newdata, Group == f2)
+    ## difference rows of xp for data from comparison
+    X <- xp[r1, ] - xp[r2, ]
+    ## zero out cols of X related to splines for other lochs
+    X[, ! (c1 | c2)] <- 0
+    ## zero out the parametric cols
+    X[, !grepl('^s\\(', colnames(xp))] <- 0
+    dif <- X %*% coef(model)
+    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
+    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
+    upr <- dif + (crit * se)
+    lwr <- dif - (crit * se)
+    data.frame(pair = paste(f1, f2, sep = '-'),
+               diff = dif,
+               se = se,
+               upper = upr,
+               lower = lwr)
+}
+
+comp1<-smooth_diff(gam1,pdat,'Control','Treatment')
+
+comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
+                   rbind(comp1)) %>%
+  mutate(interval=case_when(
+    upper>0 & lower<0~"no-diff",
+    upper<0~"less",
+    lower>0~"greater"
+  ))
+
+c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5,fill='#993336') +
+    geom_line(color='black',size=1) +
+    facet_wrap(~ pair) +
+  theme_classic()+
+    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+  scale_x_continuous(breaks=c(0,2,5,7,10))+
+  theme(
+     text=element_text(size=18),
+     legend.title=element_blank()
+     )
+#+
+ # geom_rect(aes(color=interval))
+
+c1
+
+###for missing data
 comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
@@
      )
+
 ```
-
-```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-##Pairwise comparisons
-
-# pdat <- expand.grid(Day = seq(0, 10, length = 400),
-#                     Group = c('Treatment', 'Control'))
-#
-# smooth_diff <- function(model, newdata, f1, f2, alpha = 0.05,
-#                         unconditional = FALSE) {
-#     xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-#     c1 <- grepl(f1, colnames(xp))
-#     c2 <- grepl(f2, colnames(xp))
-#     #r1 <- newdata[[var]] == f1
-#     #r2 <- newdata[[var]] == f2
-#     r1 <- with(newdata, Group == f1)
-#     r2 <- with(newdata, Group == f2)
-#     ## difference rows of xp for data from comparison
-#     X <- xp[r1, ] - xp[r2, ]
-#     ## zero out cols of X related to splines for other lochs
-#     X[, ! (c1 | c2)] <- 0
-#     ## zero out the parametric cols
-#     X[, !grepl('^s\\(', colnames(xp))] <- 0
-#     dif <- X %*% coef(model)
-#     se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-#     crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-#     upr <- dif + (crit * se)
-#     lwr <- dif - (crit * se)
-#     data.frame(pair = paste(f1, f2, sep = '-'),
-#                diff = dif,
-#                se = se,
-#                upper = upr,
-#                lower = lwr)
-# }
-#
-# comp1<-smooth_diff(mod_int2,pdat,'Control','Treatment')
-#
-# comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
-#                    rbind(comp1))
-#
-# c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
-#     geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5,fill='#993336') +
-#     geom_line(color='black',size=1) +
-#     facet_wrap(~ pair) +
-#   theme_classic()+
-#     labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-#   scale_x_continuous(breaks=c(0,2,5,7,10))+
-#   theme(
-#      text=element_text(size=18),
-#      legend.title=element_blank()
-#      )
-#
-#
-#
-# c1
-
-```
-
-## Determination of significance in GAMs
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that timepoint (or timepoints).
-The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
-The major advantage of the significance estimation in GAMs  is that they allow to determine significance at a specific timepoint, rather than providing a single _p-value_ for the overall treatment effect.
-The estimation of siginificant differences in GAMs is covered in detail in the Appendix using simulated longitudinal data that follows previously reported trends of tumor response between multiple treatment groups [@vishwanath2009].
 

### 2021-05-11 Manuscripts/Manuscript v5/Manuscript_AM_v5.tex
Working on confidence interval, need to center the comparison to zero

--- 2021-04-15 Manuscripts/Manuscript v5/Manuscript_AM_v5.tex
@@
 }
+\hypertarget{abstract}{%
+\section{Abstract}\label{abstract}}
+
+In biomedical research, the outcome of longitudinal studies is frequently analyzed statistically using the repeated measures analysis or variance (rm-ANOVA).
+However, rm-ANOVA is restrictive as is only able to accommodate linear responses, requires equal correlation between each measurement (compound symmetry), and cannot work with missing observations.
+When the data fails to fullfill all the required assumptions by rm-ANOVA, the model provides biased estimates and inference.
+In contrast, generalized additive models (GAMs) are a class of models that overcome all of the limitations imposed by rm-ANOVA and that can be implemented in a straightforward manner using the \emph{mgcv} package in R.
+This paper summarizes the assumptions of rm-ANOVA and how their limit its use in longitudinal biomedical research, presents the basic theoretical framework for GAMs and provides examples of their implementation in R.
+The code and data used to generate this paper are available at:
+
 \hypertarget{background}{%
@@
 
-\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.}\label{fig:l-q-response}
+\caption{(l-q-response-caption)}\label{fig:l-q-response}
 \end{figure}
@@
 D.
-Smoother for time and original datapoints.
+Smoother for time and original data points.
 The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}\label{fig:basis-plot}
@@
 
-\hypertarget{gams-and-covariance}{%
-\subsection{GAMs and covariance}\label{gams-and-covariance}}
-
-Although the specific methods of how GAMs model correlation structures is a topic beyond the scope of this work, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
+\hypertarget{the-analyisis-of-longitudinal-biomedical-data-using-gams}{%
+\section{The analyisis of longitudinal biomedical data using GAMs}\label{the-analyisis-of-longitudinal-biomedical-data-using-gams}}
+
+The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA.
+This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \texttt{mgcv}.
+The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
+
+\hypertarget{simulated-data}{%
+\subsection{Simulated data}\label{simulated-data}}
+
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(StO_2\)) in subcutaneous tumors that appear in Figure 3(C) in {[}vishwanath2009{]}.
+An increase in oxygen saturation in the ``Treatment'' group (chemotherapy) is observed through time, while a decrease is seen in the ``Control'' (saline) group.
+Following the reported trend, 10 normally distributed observations at each time point are generated with a standard deviation (SD) of 10\% (matching the SD in the original paper).
+The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}, A and the inlet, respectively.
+
+\hypertarget{an-interaction-gam-model-for-longitudinal-data}{%
+\subsection{An interaction GAM model for longitudinal data}\label{an-interaction-gam-model-for-longitudinal-data}}
+
+An interaction model is typically the main interest in longitudinal biomedical data, as it takes into account the effect of treatment, time, and their combination.
+In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time.
+Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
 A detailed description on basis functions and correlations can be found in {[}52{]}.
-In a practical sense, when a GAM is implemented for longitudinal data, a spline can be added to the model for the \emph{time} effect to account for the repeated measures over time.
-An example where this is covered is in the Appendix.
+
+For the data in Figure \ref{fig:sim-smooth-plot},A it is of interest to know the effect that each treatment has in oxygen saturation.
+The model then needs to incorporate independent smooths for \emph{Group} and \emph{Day}, respectively.
+The main thing to consider is the model syntax to account for the fact that one of the variables is numeric ( \emph{Day} ) and that the other is a factor ( \emph{Group} ).
+Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
+Using \(\textsf{R}\) and the package \texttt{mgcv} the model syntax is:
+
+\texttt{m1\textless{}-gam(StO2\_sim\textasciitilde{}Group+s(Day,by=Group,k=5),\ method=\textquotesingle{}REML\textquotesingle{},data=dat\_sim}
+
+This syntax specifies that \texttt{m1} will store the model, and that the change in the simulated oxygen saturation \texttt{StO2\_sim} is modeled using independent smooths for \emph{Group} and \emph{Day} (the parenthesis preceded by \emph{s}) using 5 knots.
+The parametric term \emph{Group} is added to separate the effect of treatment, and the \texttt{method} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}37{]}.
+When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(StO_2\) for each group across time (Figure \ref{fig:sim-smooth-plot},B).
+Model diagnostics can be obtained using the \texttt{gam.check} function, and the function \texttt{appraise} from the package \texttt{gratia}{[}54{]}.
+A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in {[}37{]} and {[}55{]}.
+
+One question might be ``what is the fit that an rm-ANOVA model produces for the simulated data?''.
+The rm-ANOVA model, which corresponds to \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot},C.
+This is a typical case of model misspecification: The slopes of each group ar different, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \ref{fig:sim-smooth-plot},B) .
+
+Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
+Using the simulated \(StO_2\) values from Figure (\ref{fig:sim-smooth-plot}, B), 40\% of the total observations can be randomly deleted and the same GAM model with interaction used for the full simulated data can be fitted to the remaining observations.
+
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v5_files/figure-latex/sim-smooth-plot-1}
+
+}
+
+\caption{Simulated data and smooths for oxygen saturation in tumors.
+A: Real data that follows previously reported trends in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+B: Simulated data from a normal distribution with standard deviation of 10\% with 10 observations per time point.
+In A and B lines indicate mean oxygen saturation C: Smooths from the GAM model with interaction of Group and Treatment.Lines represent smooths for each group, shaded regions are 95\% confidence intervals.}\label{fig:sim-smooth-plot}
+\end{figure}
 
@@
 
+\leavevmode\hypertarget{ref-gratia}{}%
+{[}54{]} G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for gams fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.
+
+\leavevmode\hypertarget{ref-harezlak2018}{}%
+{[}55{]} J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with r, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.
+
 \newpage

### 2021-05-11 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
Confidence interval figure works now with shading. Good explanation of implication of difference required.

@@
     ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE) +labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
+    geom_point(show.legend=FALSE,size=1.5,alpha=0.5) +
+    labs(y='response')+
+    geom_line(aes(x = time, y = mu, color = treatment,size=1),show.legend=FALSE) +
     theme_classic() +
@@
     # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
+    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"),
+              lty = 1,show.legend = FALSE) +labs(y='response')+
     scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
@@
      theme_classic()+
-    # facet_wrap(~ treatment) +
-    #ggtitle("Simulated errors") +
      theme(plot.title = element_text(size = 18, 
@@
   p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point()+labs(y='response')+
+    geom_point(size=1.5,alpha=0.5)+labs(y='response')+
     geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation ($StO_2$) in subcutaneous tumors that appear in Figure 3(C) in [vishwanath2009].
+The simulated data is based on the reported longitudinal changes in oxygen saturation ($StO_2$) in subcutaneous tumors that appear in Figure 3(C) in [@vishwanath2009].
 An increase in oxygen saturation in the "Treatment" group (chemotherapy) is observed through time, while a decrease is seen in the "Control" (saline) group.
@@
 
-This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation `StO2_sim` is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
+This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
 The parametric term _Group_ is added to separate the effect of treatment, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
@@
     geom_line(size=1,show.legend = FALSE)+
-    geom_point(show.legend = FALSE)+
+    geom_point(show.legend = FALSE,size=1.5,alpha=0.5)+
   labs(y=expression(paste(StO[2],' (real)')))+
@@
 f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group)) +
-    geom_point(show.legend=FALSE)+
+    geom_point(show.legend=FALSE,size=1.5,alpha=0.5)+
     stat_summary(aes(y = StO2_sim,
@@
 
-summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
+#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
+#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
 
@@
   f3+f4+f6+
-  #plot_layout(guides = 'collect')+
    plot_annotation(tag_levels='A')&
@@
 
-## Determination of significance in GAMs
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that timepoint (or timepoints).
-The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
-The major advantage of the significance estimation in GAMs  is that they allow to determine significance at a specific timepoint, rather than providing a single _p-value_ for the overall treatment effect.
-The estimation of siginificant differences in GAMs is covered in detail in the Appendix using simulated longitudinal data that follows previously reported trends of tumor response between multiple treatment groups [@vishwanath2009].
-
 ```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
@@
 pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Treatment', 'Control'))
-
-smooth_diff <- function(model, newdata, f1, f2, alpha = 0.05,
+                    Group = c('Control', 'Treatment'))
+
+#this function takes the model, grid and groups to be compared using the lpmatrix
+
+smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
                         unconditional = FALSE) {
     xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    c1 <- grepl(f1, colnames(xp))
-    c2 <- grepl(f2, colnames(xp))
+    #Find columns in xp where the name contains "Control"
+    col1 <- grepl(g1, colnames(xp))
+    #Find columns in xp where the name contains 'Treatment'
+    col2 <- grepl(g2, colnames(xp))
     #r1 <- newdata[[var]] == f1
     #r2 <- newdata[[var]] == f2
-    r1 <- with(newdata, Group == f1)
-    r2 <- with(newdata, Group == f2)
+    row1 <- with(newdata, Group == g1)
+    row2 <- with(newdata, Group == g2)
     ## difference rows of xp for data from comparison
-    X <- xp[r1, ] - xp[r2, ]
+    X <- xp[row1, ] - xp[row2, ]
     ## zero out cols of X related to splines for other lochs
-    X[, ! (c1 | c2)] <- 0
+    X[, ! (col1 | col2)] <- 0
     ## zero out the parametric cols
@@
     lwr <- dif - (crit * se)
-    data.frame(pair = paste(f1, f2, sep = '-'),
+    data.frame(pair = paste(g1, g2, sep = '-'),
                diff = dif,
@@
 
-comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
+comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
                    rbind(comp1)) %>%
@@
 
-c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5,fill='#993336') +
+c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper),
+                alpha = 0.5,
+                fill='black') +
     geom_line(color='black',size=1) +
+    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
+    geom_ribbon(data=comp_StO2_full%>%
+                    filter(lower>0),
+                aes(ymin =0, ymax =lower),
+                alpha = 0.5,
+                fill='blue') +
+    geom_ribbon(data=comp_StO2_full %>%
+                    filter(upper<0),
+                    aes(ymin =0, ymax =upper),
+                alpha = 0.5,
+                fill='red') +
     facet_wrap(~ pair) +
@@
     )
-#+
- # geom_rect(aes(color=interval))
-
-c1
 
@@
 comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
-comp_StO2_m <- cbind(Day = seq(0, 10, length = 400),
+comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
                    rbind(comp2))
 
-missing_plot<-ggplot(comp_StO2_m, aes(x = Day, y = diff, group = pair)) +
+missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
     geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
@@
 
-```
+c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper),
+                alpha = 0.5,
+                fill='black') +
+    geom_line(color='black',size=1) +
+    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
+    geom_ribbon(data=comp_StO2_missing%>%
+                    filter(lower>0),
+                aes(ymin =0, ymax =lower),
+                alpha = 0.5,
+                fill='blue') +
+    geom_ribbon(data=comp_StO2_missing %>%
+                    filter(upper<0),
+                    aes(ymin =0, ymax =upper),
+                alpha = 0.5,
+                fill='red') +
+    facet_wrap(~ pair) +
+    theme_classic()+
+    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    theme(
+        text=element_text(size=18),
+        legend.title=element_blank()
+    )
+
+pair_comp<-c1+c2
+```
+
+(ref:pairwise-comp) Pairwise comparisons for smooth terms.
+A: Pairwise comparisons for the full dataset.
+Significant differences exist where the interval does not cover 0.
+B: Pairwise comparisons for the full dataset.
+Significant differences exist where the interval does not cover 0.
+
+```{r,plot-pairwise-comp,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-comp)'}
+
+pair_comp+plot_annotation(tag_levels='A')&
+  theme(
+     text=element_text(size=18)
+     )
+
+```
+
+## Determination of significance in GAMs
+
+At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
+Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs can use a different approach to estimate significance.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that time point(s).
+The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
+The major advantage of the significance estimation in GAMs using the confidence interval approach is that they allow to determine significance at a specific time point, rather than providing a single _p-value_ for the overall treatment effect.
+
+Consider the simulated data used in the previous section.
+Figure (\@ref:sim-smooth-plot), B and D show the estimated smooths for the interaction model using the full dataset and with missing observations.
+If the confidence intervals are compared, it can be seen over time which Group has greater effect as therapy progresses.
+Specifically, Figure (\@ref:plot-pairwise-comp), shows the comparison between each treatment group.
+Notice that the shaded region between days 0 and 4 indicates that through that time, the "Control" group has higher $StO_2$, but as therapy progresses the effect is reversed and by day 5 it is the "Treatment" group the one that has greater contribution.
 
@@
 
+\newpage
 # References
@@
 <div id="refs"></div>
-
-\newpage
 

### 2021-05-11 Manuscripts/Manuscript v5/Manuscript_AM_v5.tex
Confidence interval figure works now with shading. Good explanation of implication of difference required.

@@
 
-\caption{(l-q-response-caption)}\label{fig:l-q-response}
+\caption{ref(l-q-response-caption)}\label{fig:l-q-response}
 \end{figure}
@@
 Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
-D.
-Smoother for time and original data points.
+D: Smoother for time and original data points.
 The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}\label{fig:basis-plot}
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(StO_2\)) in subcutaneous tumors that appear in Figure 3(C) in {[}vishwanath2009{]}.
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(StO_2\)) in subcutaneous tumors that appear in Figure 3(C) in {[}16{]}.
 An increase in oxygen saturation in the ``Treatment'' group (chemotherapy) is observed through time, while a decrease is seen in the ``Control'' (saline) group.
@@
 
-This syntax specifies that \texttt{m1} will store the model, and that the change in the simulated oxygen saturation \texttt{StO2\_sim} is modeled using independent smooths for \emph{Group} and \emph{Day} (the parenthesis preceded by \emph{s}) using 5 knots.
+This syntax specifies that \texttt{m1} will store the model, and that the change in the simulated oxygen saturation (\texttt{StO2\_sim}) is modeled using independent smooths for \emph{Group} and \emph{Day} (the parenthesis preceded by \emph{s}) using 5 knots.
 The parametric term \emph{Group} is added to separate the effect of treatment, and the \texttt{method} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}37{]}.
@@
 \caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Real data that follows previously reported trends in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-B: Simulated data from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-In A and B lines indicate mean oxygen saturation C: Smooths from the GAM model with interaction of Group and Treatment.Lines represent smooths for each group, shaded regions are 95\% confidence intervals.}\label{fig:sim-smooth-plot}
+A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
+Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
+Lines represent smooths for each group, shaded regions are 95\% confidence intervals.C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
+D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
+Lines represent smooths for each group, shaded regions are 95\% confidence intervals}\label{fig:sim-smooth-plot}
+\end{figure}
+
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v5_files/figure-latex/plot-pairwise-comp-1}
+
+}
+
+\caption{Pairwise comparisons for smooth terms.
+A: Pairwise comparisons for the full dataset.
+Significant differences exist where the interval does not cover 0.
+B: Pairwise comparisons for the full dataset.
+Significant differences exist where the interval does not cover 0.}\label{fig:plot-pairwise-comp}
 \end{figure}
@@
 At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the \emph{difference} between the confidence intervals of the fitted smooths for such groups is non-zero, then a significant difference exists at that timepoint (or timepoints).
+Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the \emph{difference} between the confidence intervals of the fitted smooths for such groups is non-zero, then a significant difference exists at that time point(s).
 The absence of a \emph{p-value} in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
-The major advantage of the significance estimation in GAMs is that they allow to determine significance at a specific timepoint, rather than providing a single \emph{p-value} for the overall treatment effect.
-The estimation of siginificant differences in GAMs is covered in detail in the Appendix using simulated longitudinal data that follows previously reported trends of tumor response between multiple treatment groups {[}16{]}.
+The major advantage of the significance estimation in GAMs using the confidence interval approach is that they allow to determine significance at a specific time point, rather than providing a single \emph{p-value} for the overall treatment effect.
+
+Consider the simulated data used in the previous section.
+Figure (@ref:sim-smooth-plot), B and D show the estimated smooths for the interaction model using the full dataset and with missing observations.
+If the confidence intervals are compared, it can be seen over time which Group has greater effect as therapy progresses.
+Specifically, Figure (@ref:plot-pairwise-comp), shows the comparison between each treatment group.
+Notice that the shaded region between days 0 and 4 indicates that through that time, the ``Control'' group has higher \(StO_2\), but as therapy progresses the effect is reversed and by day 5 it is the ``Treatment'' group the one that has greater contribution.
 
 \begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+\newpage
 
@@
 
-\newpage
-
 \hypertarget{appendix-appendix}{%

### 2021-05-11 Code and miscellaneous/Pairwise comparisons for smooths.R

@@ -1,0 +1,185 @@
+#This post contains how to perform calculations in a gam model, when testing for
+#significance
+#https://stats.stackexchange.com/questions/376237/correcting-for-multiple-pairwise-comparisons-with-gam-objects-mgcv-in-r
+
+
+dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
+            Day=rep(c(0,2,5,7,10),times=2),
+            Group=as.factor(rep(c("Control","Treatment"),each=5))
+)
+
+
+
+simulate_data <- function(dat, n = 10, sd = 5) {
+    dat_sim <- dat %>%
+        slice(rep(1:n(), each = n)) %>%
+        group_by(Group, Day) %>%
+        mutate(
+            StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
+            subject=rep(1:10),
+            subject=factor(paste(subject, Group, sep = "-"))
+        ) %>%
+        ungroup()
+
+    return(dat_sim)
+}
+
+
+#subject = factor(paste(subject, treatment, sep = "-")))
+
+n <- 10 #number of observations
+sd <- 10 #approximate sd from paper
+set.seed(1) #set seed for reproducibility
+df <- 6
+dat_sim <- simulate_data(dat, n, sd)
+
+
+gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
+            method='REML',
+            data  = dat_sim)
+
+#creates fine grid at the values of Day
+pdat <- expand.grid(Day = seq(0, 10, length = 400),
+                    Group = c('Control', 'Treatment'))
+
+##matrix that contains the basis functions evaluated at the points in pdat
+    xp <- predict(gam1, newdata = pdat, type = 'lpmatrix')
+
+    #remove intercept column and group identifying column
+    xp<-xp[,-c(1:2)]
+
+    control<-xp[1:400,]
+    treatment<-xp[401:800,]
+
+
+#Find columns in xp where the name contains "Control"
+    c1 <- grepl('Control', colnames(xp))
+
+#Find columns in xp where the name contains 'Treatment'
+    c2 <- grepl('Treatment', colnames(xp))
+
+#Find rows in pdat that correspond to either 'Control' or 'Treatment'
+    r1 <- with(pdat, Group == 'Control')
+    r2 <- with(pdat, Group == 'Treatment')
+
[…]

### 2021-05-12 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
Manuscript updated, need to work on re-structuring on a chapter basis and add code to the Appendix. TM comments for the previous version have been revised and changes made.

@@
 
-In biomedical research, the outcome of longitudinal studies is frequently analyzed statistically using the repeated measures analysis or variance (rm-ANOVA).
-However, rm-ANOVA is restrictive as is only able to accommodate linear responses, requires equal correlation between each measurement (compound symmetry), and cannot work with missing observations.
-When the data fails to fullfill all the required assumptions by rm-ANOVA, the model provides biased estimates and inference.
-In contrast, generalized additive models (GAMs) are a class of models that overcome all of the limitations imposed by rm-ANOVA and that can be implemented in a straightforward manner using the _mgcv_ package in R.
-This paper summarizes the assumptions of rm-ANOVA and how their limit its use in longitudinal biomedical research, presents the basic theoretical framework for GAMs and provides examples of their implementation in R.
-The code and data used to generate this paper are available at:
+In biomedical research, the outcome of a longitudinal study has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, a _linear mixed model_ (LMEM).
+Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
+In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+Although not commonly used in biomedical research, GAMs present an excellent choice to analyze non-linear longitudinal data.
+This paper summarizes the limitations of linear models (rm-ANOVA in particular), presents the basic theory of GAMs, and demonstrates their implementation in $\textsf{R}$ via the package _mgcv_ using simulated data that follows longitudinal trends reported in biomedical literature.
+To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:________.
 
@@
 Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following RT in neck cancer patients [@kamstra2015].
+Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients [@kamstra2015].
 
 Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as a limiting frequency [@wagenmakers2008] by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
+The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times [@wagenmakers2008], and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
 The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].
@@
 In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
@@
 Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019].
-Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to balance in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
+Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
 
@@
 It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios,the _complete observations_ assumption of  classical rm-ANOVA is violated.
+In each of these missing data scenarios, the _complete observations_ assumption of  classical rm-ANOVA is violated.
 When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis [@gueorguieva2004].
@@
 
-In a biomedical longitudinal study, two or more groups of subjects (e.g., patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs.
+In a longitudinal biomedical study, two or more groups of subjects (e.g., human subject, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug or intervention vs.
 a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
 The collected response is modeled with _fixed_ components.
-The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug in the "treatment" group.
+The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug/intervention in the "treatment" group.
 
@@
 In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$.
+Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$ (independently normally distributed with mean zero and variance $\sigma^2_\mu$).
 In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
@@
 As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
-The _random effect_ $\mu_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be independently normally distributed with mean zero and variance $\sigma^2_\mu$, which can be expressed as $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
+The _random effect_ $\mu_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
 In essence,the _random effect_ in a LMEM enables to fit models with different slopes at the subject-level[@pinheiro2006].
@@
 In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others [@pinheiro2006].
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself.
+Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
 Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
@@
 
-## How does an rm-ANOVA fit looks like? A visual representation using simulation {#simulation}
+## What does an rm-ANOVA fit looks like? A visual representation using simulated data {#simulation}
 
@@
 ```{r,include=FALSE,message=FALSE,echo=FALSE}
-## Example with linear response
+## Example with linear response. This function generates either linear or quadratic mean responses with correlated or uncorrelated errors and fits a linear model to the data.
 example <- function(n_time = 6, 
@@
 ```{r,basis-functions-plot, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
+#generate the response
 n_time = 6
@@
  mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
+ mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
+ mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
  y <- array(0, dim = c(length(x), 2, 10))
@@
  dat$time <- as.numeric(as.character(dat$time))
+
+ #add subject column
  dat <- dat %>%
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation ($StO_2$) in subcutaneous tumors that appear in Figure 3(C) in [@vishwanath2009].
-An increase in oxygen saturation in the "Treatment" group (chemotherapy) is observed through time, while a decrease is seen in the "Control" (saline) group.
-Following the reported trend, 10 normally distributed observations at each time point are generated with a standard deviation (SD) of 10% (matching the SD in the original paper).
+The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors that appear in Figure 3(C) in [@vishwanath2009].
+In the paper, diffuse reflectance spectroscopy was used to quantify $\mbox{StO}_2$ changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
+In the "Treatment" group (chemotherapy) an increase in $\mbox{StO}_2$ is observed through time, while a decrease is seen in the "Control" (saline) group.
+Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (matching the SD in the original paper).
 The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot), A and the inlet, respectively.
 
-## An interaction GAM model for longitudinal data
+## An interaction GAM for longitudinal data
 
@@
 In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the _time_ effect to account for the repeated measures over time.
-Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
+Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
 A detailed description on basis functions and correlations can be found in [@hefley2017].
 
-For the data in Figure \@ref(fig:sim-smooth-plot),A it is of interest to know the effect that each treatment has in oxygen saturation.
+For the data in Figure \@ref(fig:sim-smooth-plot),A it is of interest to know the effect that each treatment has in $\mbox{StO}_2$.
 The model then needs to incorporate independent smooths for _Group_ and _Day_, respectively.
-The main thing to consider is the model syntax to account for the fact that one of the variables is numeric ( _Day_ ) and that the other is a factor ( _Group_ ).
+The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric ( _Day_ ) and the other is a factor ( _Group_ ).
 Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
@@
 
-`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim`
+`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5,bs="gp"), method='REML',data=dat_sim`
 
 This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
-The parametric term _Group_ is added to separate the effect of treatment, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $StO_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B).
-Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package `gratia`[@gratia].
+The smooth is constructed using gaussian process smooths, which is indicated by `bs="gp"`.
+These splines are used to model temporal trends and might be particularly suited for long-term studies where the correlation between measurements changes as a function of the time intervals [@simpson2018].
+The parametric term _Group_ is added to quantify differences in the effect of treatment between groups, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
+When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B).
+Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_[@gratia].
 A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].
 
-One question might be "what is the fit that an rm-ANOVA model produces for the simulated data?".
-The rm-ANOVA model, which corresponds to \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot),C.
-This is a typical case of model misspecification: The slopes of each group ar different, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \@ref(fig:sim-smooth-plot),B) .
+One question that might arise at this point is "what is the fit that an rm-ANOVA model produces for the simulated data?".
+The rm-ANOVA model, which corresponds to Equation \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot),C.
+This is a typical case of model misspecification: The slopes of each group are different, which would lead to a _p-value_ indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \@ref(fig:sim-smooth-plot),B) .
 
@@
 
-# tumor_sim<-f1+f2+plot_layout(guides = 'collect')+
-#   plot_annotation(tag_levels='A')&
-#   ylim(c(0,75)) &
-#   theme(
-#      text=element_text(size=18),
-#      legend.title=element_blank(),
-#      legend.text = element_blank()
-#      )&
-#   scale_color_aaas()
-
 ```
@@
            se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lme
-#lm_predict<-lm_predict%>%
-#   mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response',level=0),
- #         se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response',level=0))
 
@@
 Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
-Using the simulated $StO_2$ values from Figure (\@ref(fig:sim-smooth-plot), B), 40% of the total observations can be randomly deleted and the same GAM model with interaction used for the full simulated data can be fitted to the remaining observations.
+The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
+However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
+Consider the  simulated $\mbox{StO}_2$ values from Figure (\@ref(fig:sim-smooth-plot), B).
+If 40% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but the "Treatment" smooth takes an overall more linear profile (\@ref(fig:sim-smooth-plot), D).
+Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.
 
@@
 Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent smooths for each group, shaded regions are 95% confidence intervals.C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
+Lines represent trends for each group, shaded regions are 95% confidence intervals.
+C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
 D: Smooths for the GAM model for the simulated data with 40% of its observations missing.
-Lines represent smooths for each group, shaded regions are 95% confidence intervals
+Lines represent trends for each group, shaded regions are 95% confidence intervals.
 
@@
     geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill='black') +
-    geom_line(color='black',size=1) +
+                alpha = 0.9,
+                fill='#fdf05d') +
+    geom_line(color='#755006',size=1) +
     geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
@@
                 alpha = 0.5,
-                fill='blue') +
+                fill='#003049') +
     geom_ribbon(data=comp_StO2_full %>%
@@
                 alpha = 0.5,
-                fill='red') +
+                fill='#c1121f') +
     facet_wrap(~ pair) +
@@
     geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill='black') +
-    geom_line(color='black',size=1) +
+                alpha = 0.9,
+                fill='#fdf05d') +
+    geom_line(color='#755006',size=1) +
     geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
@@
                 alpha = 0.5,
-                fill='blue') +
+                fill='#003049') +
     geom_ribbon(data=comp_StO2_missing %>%
@@
 A: Pairwise comparisons for the full dataset.
+B: Pairwise comparisons for the dataset with missing observations.
 Significant differences exist where the interval does not cover 0.
-B: Pairwise comparisons for the full dataset.
-Significant differences exist where the interval does not cover 0.
+In both cases the effect of treatment is significant after day 5.
 
@@
 
-## Determination of significance in GAMs
+## Determination of significance in GAMs for longitudinal data
 
@@
 In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that time point(s).
-The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
-The major advantage of the significance estimation in GAMs using the confidence interval approach is that they allow to determine significance at a specific time point, rather than providing a single _p-value_ for the overall treatment effect.
-
-Consider the simulated data used in the previous section.
-Figure (\@ref:sim-smooth-plot), B and D show the estimated smooths for the interaction model using the full dataset and with missing observations.
-If the confidence intervals are compared, it can be seen over time which Group has greater effect as therapy progresses.
-Specifically, Figure (\@ref:plot-pairwise-comp), shows the comparison between each treatment group.
-Notice that the shaded region between days 0 and 4 indicates that through that time, the "Control" group has higher $StO_2$, but as therapy progresses the effect is reversed and by day 5 it is the "Treatment" group the one that has greater contribution.
+The absence of a _p-value_ in this case might seem odd, but the confidence interval comparison can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
+This is what happens for the simulated data in Figure \@ref(fig:sim-smooth-plot), A where the chemotherapy causes $\mbox{StO}_2$ to increase over time.
+
+With this expectation of different trends in each group, computing the difference between the trends will identify if the observed change is significant.
+The difference between groups with similar trends is likely to yield zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
+
+Consider the calculation of pairwise differences for the smooths in \@ref(fig:sim-smooth-plot) B and D.
+Figure \@ref(fig:plot-pairwise-comp), shows the comparison between each treatment group.Here, the effect of the "Control" group is compared to that of the "Treatment" group.
+The shaded region under the curve highlights the interval where each group has a higher effect than the other.
+Notice that the shaded region between days 0 and 4 indicates that through that time, the "Control" group has higher $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by day 5 it is the "Treatment" group the one that has greater contribution.
+This would suggest that the effect of chemotherapy is significant after day 5 for the model used; which is noticeable in the data with missing observations albeit the missing data model has a wider confidence interval and is not able to pick the change between days 7 and 10 that the full dataset model is able to detect.
+
+# Conclusion
 
@@
 ```
+
+## Basis functions and GAMs
+
+## Data simulation and GAM models

### 2021-05-12 Manuscripts/Manuscript v5/Manuscript_AM_v5.tex
Manuscript updated, need to work on re-structuring on a chapter basis and add code to the Appendix. TM comments for the previous version have been revised and changes made.

@@
 
-In biomedical research, the outcome of longitudinal studies is frequently analyzed statistically using the repeated measures analysis or variance (rm-ANOVA).
-However, rm-ANOVA is restrictive as is only able to accommodate linear responses, requires equal correlation between each measurement (compound symmetry), and cannot work with missing observations.
-When the data fails to fullfill all the required assumptions by rm-ANOVA, the model provides biased estimates and inference.
-In contrast, generalized additive models (GAMs) are a class of models that overcome all of the limitations imposed by rm-ANOVA and that can be implemented in a straightforward manner using the \emph{mgcv} package in R.
-This paper summarizes the assumptions of rm-ANOVA and how their limit its use in longitudinal biomedical research, presents the basic theoretical framework for GAMs and provides examples of their implementation in R.
-The code and data used to generate this paper are available at:
+In biomedical research, the outcome of a longitudinal study has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, a \emph{linear mixed model} (LMEM).
+Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
+In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+Although not commonly used in biomedical research, GAMs present an excellent choice to analyze non-linear longitudinal data.
+This paper summarizes the limitations of linear models (rm-ANOVA in particular), presents the basic theory of GAMs, and demonstrates their implementation in \(\textsf{R}\) via the package \emph{mgcv} using simulated data that follows longitudinal trends reported in biomedical literature.
+To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:\_\_\_\_\_\_\_\_.
 
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(StO_2\)) in subcutaneous tumors that appear in Figure 3(C) in {[}16{]}.
-An increase in oxygen saturation in the ``Treatment'' group (chemotherapy) is observed through time, while a decrease is seen in the ``Control'' (saline) group.
-Following the reported trend, 10 normally distributed observations at each time point are generated with a standard deviation (SD) of 10\% (matching the SD in the original paper).
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3(C) in {[}16{]}.
+In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
+In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group.
+Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
 The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}, A and the inlet, respectively.
 
-\hypertarget{an-interaction-gam-model-for-longitudinal-data}{%
-\subsection{An interaction GAM model for longitudinal data}\label{an-interaction-gam-model-for-longitudinal-data}}
+\hypertarget{an-interaction-gam-for-longitudinal-data}{%
+\subsection{An interaction GAM for longitudinal data}\label{an-interaction-gam-for-longitudinal-data}}
 
@@
 In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time.
-Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
+Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
 A detailed description on basis functions and correlations can be found in {[}52{]}.
 
-For the data in Figure \ref{fig:sim-smooth-plot},A it is of interest to know the effect that each treatment has in oxygen saturation.
+For the data in Figure \ref{fig:sim-smooth-plot},A it is of interest to know the effect that each treatment has in \(\mbox{StO}_2\).
 The model then needs to incorporate independent smooths for \emph{Group} and \emph{Day}, respectively.
-The main thing to consider is the model syntax to account for the fact that one of the variables is numeric ( \emph{Day} ) and that the other is a factor ( \emph{Group} ).
+The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric ( \emph{Day} ) and the other is a factor ( \emph{Group} ).
 Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
@@
 
-\texttt{m1\textless{}-gam(StO2\_sim\textasciitilde{}Group+s(Day,by=Group,k=5),\ method=\textquotesingle{}REML\textquotesingle{},data=dat\_sim}
+\texttt{m1\textless{}-gam(StO2\_sim\textasciitilde{}Group+s(Day,by=Group,k=5,bs="gp"),\ method=\textquotesingle{}REML\textquotesingle{},data=dat\_sim}
 
 This syntax specifies that \texttt{m1} will store the model, and that the change in the simulated oxygen saturation (\texttt{StO2\_sim}) is modeled using independent smooths for \emph{Group} and \emph{Day} (the parenthesis preceded by \emph{s}) using 5 knots.
-The parametric term \emph{Group} is added to separate the effect of treatment, and the \texttt{method} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}37{]}.
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(StO_2\) for each group across time (Figure \ref{fig:sim-smooth-plot},B).
-Model diagnostics can be obtained using the \texttt{gam.check} function, and the function \texttt{appraise} from the package \texttt{gratia}{[}54{]}.
+The smooth is constructed using gaussian process smooths, which is indicated by \texttt{bs="gp"}.
+These splines are used to model temporal trends and might be particularly suited for long-term studies where the correlation between measurements changes as a function of the time intervals {[}34{]}.
+The parametric term \emph{Group} is added to quantify differences in the effect of treatment between groups, and the \texttt{method} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}37{]}.
+When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot},B).
+Model diagnostics can be obtained using the \texttt{gam.check} function, and the function \texttt{appraise} from the package \emph{gratia}{[}54{]}.
 A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in {[}37{]} and {[}55{]}.
 
-One question might be ``what is the fit that an rm-ANOVA model produces for the simulated data?''.
-The rm-ANOVA model, which corresponds to \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot},C.
-This is a typical case of model misspecification: The slopes of each group ar different, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \ref{fig:sim-smooth-plot},B) .
+One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?''.
+The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot},C.
+This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \ref{fig:sim-smooth-plot},B) .
 
 Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
-Using the simulated \(StO_2\) values from Figure (\ref{fig:sim-smooth-plot}, B), 40\% of the total observations can be randomly deleted and the same GAM model with interaction used for the full simulated data can be fitted to the remaining observations.
+The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
+However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
+Consider the simulated \(\mbox{StO}_2\) values from Figure (\ref{fig:sim-smooth-plot}, B).
+If 40\% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but the ``Treatment'' smooth takes an overall more linear profile (\ref{fig:sim-smooth-plot}, D).
+Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.
 
@@
 Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent smooths for each group, shaded regions are 95\% confidence intervals.C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
+Lines represent trends for each group, shaded regions are 95\% confidence intervals.
+C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
 D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent smooths for each group, shaded regions are 95\% confidence intervals}\label{fig:sim-smooth-plot}
+Lines represent trends for each group, shaded regions are 95\% confidence intervals.}\label{fig:sim-smooth-plot}
 \end{figure}
@@
 A: Pairwise comparisons for the full dataset.
+B: Pairwise comparisons for the dataset with missing observations.
 Significant differences exist where the interval does not cover 0.
-B: Pairwise comparisons for the full dataset.
-Significant differences exist where the interval does not cover 0.}\label{fig:plot-pairwise-comp}
+In both cases the effect of treatment is significant after day 5.}\label{fig:plot-pairwise-comp}
 \end{figure}
 
-\hypertarget{determination-of-significance-in-gams}{%
-\subsection{Determination of significance in GAMs}\label{determination-of-significance-in-gams}}
+\hypertarget{determination-of-significance-in-gams-for-longitudinal-data}{%
+\subsection{Determination of significance in GAMs for longitudinal data}\label{determination-of-significance-in-gams-for-longitudinal-data}}
 
@@
 In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the \emph{difference} between the confidence intervals of the fitted smooths for such groups is non-zero, then a significant difference exists at that time point(s).
-The absence of a \emph{p-value} in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
-The major advantage of the significance estimation in GAMs using the confidence interval approach is that they allow to determine significance at a specific time point, rather than providing a single \emph{p-value} for the overall treatment effect.
-
-Consider the simulated data used in the previous section.
-Figure (@ref:sim-smooth-plot), B and D show the estimated smooths for the interaction model using the full dataset and with missing observations.
-If the confidence intervals are compared, it can be seen over time which Group has greater effect as therapy progresses.
-Specifically, Figure (@ref:plot-pairwise-comp), shows the comparison between each treatment group.
-Notice that the shaded region between days 0 and 4 indicates that through that time, the ``Control'' group has higher \(StO_2\), but as therapy progresses the effect is reversed and by day 5 it is the ``Treatment'' group the one that has greater contribution.
+The absence of a \emph{p-value} in this case might seem odd, but the confidence interval comparison can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
+This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}, A where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
+
+With this expectation of different trends in each group, computing the difference between the trends will identify if the observed change is significant.
+The difference between groups with similar trends is likely to yield zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
+
+Consider the calculation of pairwise differences for the smooths in \ref{fig:sim-smooth-plot} B and D.
+Figure \ref{fig:plot-pairwise-comp}, shows the comparison between each treatment group.Here, the effect of the ``Control'' group is compared to that of the ``Treatment'' group.
+The shaded region under the curve highlights the interval where each group has a higher effect than the other.
+Notice that the shaded region between days 0 and 4 indicates that through that time, the ``Control'' group has higher \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by day 5 it is the ``Treatment'' group the one that has greater contribution.
+This would suggest that the effect of chemotherapy is significant after day 5 for the model used; which is noticeable in the data with missing observations albeit the missing data model has a wider confidence interval and is not able to pick the change between days 7 and 10 that the full dataset model is able to detect.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
@@
 
+\hypertarget{basis-functions-and-gams}{%
+\subsection{Basis functions and GAMs}\label{basis-functions-and-gams}}
+
+\hypertarget{data-simulation-and-gam-models}{%
+\subsection{Data simulation and GAM models}\label{data-simulation-and-gam-models}}
+
 \end{document}

### 2021-05-12 Code and miscellaneous/Pairwise comparisons for smooths.R

@@ -79,7 +79,7 @@ pdat <- expand.grid(Day = seq(0, 10, length = 400),
     #dimensions (n,1). The resulting matrix has dimensions (p,1)
     dif <- X %*% coef(gam1)
 
-    comp<-test %*% coef(gam1)[3:10]
+    #comp<-test %*% coef(gam1)[3:10]
 
 
     se <- sqrt(rowSums((X %*% vcov(gam1, unconditional = FALSE)) * X))
@@ -128,7 +128,34 @@ c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
 c1
 
 
+ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper),
+                alpha = 0.9,
+                fill="#f6d8ae") +
+    geom_line(color='#E79626',size=1) +
+    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
+    geom_ribbon(data=comp_StO2%>%
+                    filter(lower>0),
+                aes(ymin =0, ymax =lower),
+                alpha = 0.9,
+                fill="#083d77") +
+    geom_ribbon(data=comp_StO2 %>%
+                    filter(upper<0),
+                aes(ymin =0, ymax =upper),
+                alpha = 0.9,
+                fill="#da4167") +
+    facet_wrap(~ pair) +
+    theme_classic()+
+    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    theme(
+        text=element_text(size=18),
+        legend.title=element_blank()
+    )+
+    scale_color_aaas()+
+    scale_fill_aaas()
 
+c1
 #######################
 
 ##original code for the comparisons, from Gavin Simpson's blog.

### 2021-05-13 Manuscripts/Bookdown testing/02-Challenges.Rmd
Splitted the document in sections, author section now works. Need to work on Appendix, as there is some issue going

--- 2021-05-12 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
@@
----
-title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
-subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
-author:
-- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
-- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
-  Fayetteville]
-- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
-output:
-  bookdown::pdf_document2:
-    keep_tex: yes
-    fig_caption: yes
-    extra_dependencies:
-      subfig: null
-      breqn: null
-      caption: ["font={small}"]
-      float: null
-  bookdown::word_document2:
-    fig_caption: yes
-    keep_md: yes
-  bookdown::html_document2: default
-  link-citations: yes
-  css: style.css
-  '': default
-csl: elsevier-with-titles.csl
-bibliography: refs.bib
----
-
-# Abstract
-
-In biomedical research, the outcome of a longitudinal study has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, a _linear mixed model_ (LMEM).
-Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
-In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
-Although not commonly used in biomedical research, GAMs present an excellent choice to analyze non-linear longitudinal data.
-This paper summarizes the limitations of linear models (rm-ANOVA in particular), presents the basic theory of GAMs, and demonstrates their implementation in $\textsf{R}$ via the package _mgcv_ using simulated data that follows longitudinal trends reported in biomedical literature.
-To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:________.
-
-# Background
-
-Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients [@kamstra2015].
-
-Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times [@wagenmakers2008], and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA [@pinheiro2006].
-This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model is only able to accommodate linear trends that are far from adequately representing the biological phenomenon of interest.
-
-A _post hoc_ analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
-Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, _p-values_ are valid only if the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019].
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting _p-values_ and parameter estimates to be invalid [@wang2019].
-
-Additionally, the _p-value_ itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019].
-Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].
-Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976;@greenhouse1959], but the effectiveness of the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001].
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not "large" due to ethical and budgetary reasons [@charan2013] which might cause the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial patients may voluntarily withdraw, whereas attrition  due to injury or weight loss in preclinical animal studies is possible.
-It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios, the _complete observations_ assumption of  classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis [@gueorguieva2004].
-This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
-At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
-
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data [@skala2010;@vishwanath2009].
-Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-However, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
-And even more importantly, LMEMs also expect a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze non-linear data.
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
-
-In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017] and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
-
-The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, $\textsf{R}$[@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
-hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017].
-Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
-
-This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations is explained in detail.
-Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, using simulated data that reproduces patterns in previously reported studies [@vishwanath2009] we illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
-The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in $\textsf{R}$, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
-
-In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models by improving the standards for reproducibility  in biomedical research.
-
 # Challenges presented by longitudinal studies
@@
 A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \@ref(fig:linear-cases) and Figure \@ref(fig:quadratic-cases) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-
-```{r, message = FALSE,include=FALSE}
-library(patchwork)
-library(tidyverse)
-library(mvnfast)
-library(nlme)
-library(ggsci)
-library(mgcv)
-library(gratia)
-
-knitr::opts_chunk$set(fig.pos = "!h", out.extra = "")
-```
 
@@
 In the following section, we present generalized additive models (GAMs) as a data-driven  alternative method to analyze longitudinal non-linear data.
-
-# GAMs as a special case of Generalized Linear Models
-
-## GAMs and Basis Functions
-
-Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors[@nelder1972].
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case[@simpson2018;@wood2017;@hastie1987].
-A GAM model can be written as:
-
-\begin{equation}
-  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
-  (\#eq:GAM)
-\end{equation}
-
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual error.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
-In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.
-
-The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using _basis functions_ expansions of the covariates and by estimating random coefficients for these basis functions.
-A _basis_ is a set of functions that spans the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
-For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
-In the case of Equation \@ref(eq:GAM), the basis function is $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
-
-A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness [@wood2017;@simpson2018].
-Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\@ref(fig:l-q-response), C).
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-The timeline can be divided in equally spaced _knots_, each knot being a region where a different basis function will be used.
-Because there are six timepoints for this group, five knots can be used.
-The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots.
-In Panel A of Figure \@ref(fig:basis-plot), the four basis functions (and the intercept) are shown.
-Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the "wigliness" of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
-The penalty reduces the "wigliness" of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
-
-```{r,basis-functions-plot, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-#generate the response
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
-
- #add subject column
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  dat<-subset(dat,treatment==1)
-  dat<-dat[,c('y','time')]
-gm<-gam(y~s(time,k=5),data=dat)
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-x_new <- (seq(0, max(dat$time), length.out = 100))
-y_pred <- predict(gm, newdata=dat)
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof)
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final spline by aggregating the basis functions
-
-spl<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(spl=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=as.factor(Basis)))+
-  geom_line(size=sz,show.legend=FALSE)+
-  geom_point(size=sz+1,show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,labeller = as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#penalized basis
-p12<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
-  geom_line(show.legend = FALSE,size=sz)+
-  geom_point(show.legend = FALSE,size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,labeller=as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#heatmap of the penalization coefficient
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
-  geom_tile(aes(color='black'),size=sz+1,show.legend = FALSE)+
-  geom_tile(size=sz+1)+
-  scale_fill_gradient(low = "white", high = "red")+
-  labs(x='Basis',y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),size=10,show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  scale_color_aaas()+
-  labs(y='Simulated \n response')+
-  geom_line(data=spl,aes(x=time,y=spl),color="#B15731",size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-
-```
-
-In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \@ref(fig:basis-plot), Panel C shows the resulting penalized basis functions.
-Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
-Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
-
-(ref:basis-plot-caption) Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.
-
-```{r,basis-plot,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap='(ref:basis-plot-caption)'}
-
-par(mar = c(2, 2, 2, 2))
-
-b_plot
-
-```
-
- \newpage
-
-# The analyisis of longitudinal biomedical data using GAMs
-
-The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA.
-This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using $\textsf{R}$ and the package `mgcv`.
-The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
-
-## Simulated data
-
-The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors that appear in Figure 3(C) in [@vishwanath2009].
-In the paper, diffuse reflectance spectroscopy was used to quantify $\mbox{StO}_2$ changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
-In the "Treatment" group (chemotherapy) an increase in $\mbox{StO}_2$ is observed through time, while a decrease is seen in the "Control" (saline) group.
-Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (matching the SD in the original paper).
-The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot), A and the inlet, respectively.
-
-## An interaction GAM for longitudinal data
-
-An interaction model is typically the main interest in longitudinal biomedical data, as it takes into account the effect of treatment, time, and their combination.
-In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the _time_ effect to account for the repeated measures over time.
-Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in [@hefley2017].
-
-For the data in Figure \@ref(fig:sim-smooth-plot),A it is of interest to know the effect that each treatment has in $\mbox{StO}_2$.
-The model then needs to incorporate independent smooths for _Group_ and _Day_, respectively.
-The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric ( _Day_ ) and the other is a factor ( _Group_ ).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
-Using $\textsf{R}$ and the package `mgcv` the model syntax is:
-
-`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5,bs="gp"), method='REML',data=dat_sim`
-
-This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
-The smooth is constructed using gaussian process smooths, which is indicated by `bs="gp"`.
-These splines are used to model temporal trends and might be particularly suited for long-term studies where the correlation between measurements changes as a function of the time intervals [@simpson2018].
-The parametric term _Group_ is added to quantify differences in the effect of treatment between groups, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B).
-Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_[@gratia].
-A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].
-
-One question that might arise at this point is "what is the fit that an rm-ANOVA model produces for the simulated data?".
-The rm-ANOVA model, which corresponds to Equation \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot),C.
-This is a typical case of model misspecification: The slopes of each group are different, which would lead to a _p-value_ indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \@ref(fig:sim-smooth-plot),B) .
-
-```{r,sim-tumor-response,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
-
-## plot the mean response
-f1<-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
-    geom_line(size=1,show.legend = FALSE)+
-    geom_point(show.legend = FALSE,size=1.5,alpha=0.5)+
-  labs(y=expression(paste(StO[2],' (real)')))+
-  theme_classic()+
-  scale_color_aaas()+
-    scale_x_continuous(breaks=c(0,5,10))+
-    scale_y_continuous(breaks=c(0,40))+
-  plot_layout(tag_level = 'new')+
-  theme(
-    plot.background = element_rect(fill = "transparent", color = NA),
-    axis.text=element_text(size=14)
-  )
-
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-
-n <- 10 #number of observations
-sd <- 10 #approximate sd from paper
-set.seed(1) #set seed for reproducibility
-df <- 6
-dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group)) +
-    geom_point(show.legend=FALSE,size=1.5,alpha=0.5)+
-    stat_summary(aes(y = StO2_sim,
-                     group=Group), 
-                 fun=mean, geom="line",size=1,show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],'(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-  scale_color_aaas()+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
-
-```
-
-```{r,smooth-plot,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-#GAM for StO2
-
-gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
-            method='REML',
-            data  = dat_sim)
-
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-#creates a dataframe using the length of the covariates for the GAM
-gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
-                         subject=factor(rep(1:10)))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#adds the predictions to the grid and creates a confidence interval for GAM
-gam_predict<-gam_predict%>%
-    mutate(fit = predict(gam1,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(gam1, gam_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=gam_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'complete')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-#posthoc comparisons for the linear model
-library(multcomp)
-
-#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
-
-```
-
-Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
-The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
-However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
-Consider the  simulated $\mbox{StO}_2$ values from Figure (\@ref(fig:sim-smooth-plot), B).
-If 40% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but the "Treatment" smooth takes an overall more linear profile (\@ref(fig:sim-smooth-plot), D).
-Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.
-
-```{r,missing-data,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
-
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'missing')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-```
-
-(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95% confidence intervals.
-C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40% of its observations missing.
-Lines represent trends for each group, shaded regions are 95% confidence intervals.
-
-```{r,sim-smooth-plot,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}
-
-mult_plot<-f2+inset_element(
-  f1, left = 0.01, 
-  bottom = 0.5, 
-  right = 0.5, 
-  top = 1.0)+
-  f3+f4+f6+
-   plot_annotation(tag_levels='A')&
-   ylim(c(-5,75)) &
-  theme(
-     text=element_text(size=18)
-     )&
-  scale_color_aaas()
-
-mult_plot
-
-```
-
-```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-##Pairwise comparisons
-
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-#this function takes the model, grid and groups to be compared using the lpmatrix
-
-smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    #Find columns in xp where the name contains "Control"
-    col1 <- grepl(g1, colnames(xp))
-    #Find columns in xp where the name contains 'Treatment'
-    col2 <- grepl(g2, colnames(xp))
-    #r1 <- newdata[[var]] == f1
-    #r2 <- newdata[[var]] == f2
-    row1 <- with(newdata, Group == g1)
-    row2 <- with(newdata, Group == g2)
-    ## difference rows of xp for data from comparison
-    X <- xp[row1, ] - xp[row2, ]
-    ## zero out cols of X related to splines for other lochs
-    X[, ! (col1 | col2)] <- 0
-    ## zero out the parametric cols
-    X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(g1, g2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-}
-
-comp1<-smooth_diff(gam1,pdat,'Control','Treatment')
-
-comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
-
-c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.9,
-                fill='#fdf05d') +
-    geom_line(color='#755006',size=1) +
-    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
-    geom_ribbon(data=comp_StO2_full%>%
-                    filter(lower>0),
-                aes(ymin =0, ymax =lower),
-                alpha = 0.5,
-                fill='#003049') +
-    geom_ribbon(data=comp_StO2_full %>%
-                    filter(upper<0),
-                    aes(ymin =0, ymax =upper),
-                alpha = 0.5,
-                fill='#c1121f') +
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-###for missing data
-comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
-comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp2))
-
-missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    labs(x = 'Days', 
-         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
-                              )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme_classic()+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
-c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.9,
-                fill='#fdf05d') +
-    geom_line(color='#755006',size=1) +
-    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
-    geom_ribbon(data=comp_StO2_missing%>%
-                    filter(lower>0),
-                aes(ymin =0, ymax =lower),
-                alpha = 0.5,
-                fill='#003049') +
-    geom_ribbon(data=comp_StO2_missing %>%
-                    filter(upper<0),
-                    aes(ymin =0, ymax =upper),
-                alpha = 0.5,
-                fill='#c1121f') +
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-pair_comp<-c1+c2
-```
-
-(ref:pairwise-comp) Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-B: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the interval does not cover 0.
-In both cases the effect of treatment is significant after day 5.
-
-```{r,plot-pairwise-comp,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-comp)'}
-
-pair_comp+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-
-```
-
-## Determination of significance in GAMs for longitudinal data
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that time point(s).
-The absence of a _p-value_ in this case might seem odd, but the confidence interval comparison can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
-This is what happens for the simulated data in Figure \@ref(fig:sim-smooth-plot), A where the chemotherapy causes $\mbox{StO}_2$ to increase over time.
-
-With this expectation of different trends in each group, computing the difference between the trends will identify if the observed change is significant.
-The difference between groups with similar trends is likely to yield zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
-
-Consider the calculation of pairwise differences for the smooths in \@ref(fig:sim-smooth-plot) B and D.
-Figure \@ref(fig:plot-pairwise-comp), shows the comparison between each treatment group.Here, the effect of the "Control" group is compared to that of the "Treatment" group.
-The shaded region under the curve highlights the interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and 4 indicates that through that time, the "Control" group has higher $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by day 5 it is the "Treatment" group the one that has greater contribution.
-This would suggest that the effect of chemotherapy is significant after day 5 for the model used; which is noticeable in the data with missing observations albeit the missing data model has a wider confidence interval and is not able to pick the change between days 7 and 10 that the full dataset model is able to detect.
-
-# Conclusion
-
-***
-
-\newpage
-# References
-
-<div id="refs"></div>
-
-# (APPENDIX) Appendix {-}
-
-# Simulation
-
-## Compound symmetry and independent errors in linear and quadratic responses
-
-This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation).
-The linear simulations using Figure \@ref(fig:linear-cases) show in  panels A and D the simulated mean responses and individual datapoints.
-Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-```{r,include=TRUE,message=FALSE,echo=TRUE}
-## Example with linear response
-example <- function(n_time = 6, 
-                    fun_type = "linear", 
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  library(tidyverse)
-  library(mvnfast)
-  library(nlme)
-  library(ggsci)
-
-  x <- seq(1,6, length.out = n_time)
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # nonlinear response
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-  # matplot(mu, type = 'l')
-
-  y <- array(0, dim = c(length(x), 2, 10))
-  errors <- array(0, dim = c(length(x), 2, 10))
-
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(mu) <- list(time = x, treatment = 1:2)
-  dat <- as.data.frame.table(y, responseName = "y")
-  dat_errors <- as.data.frame.table(errors, responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, responseName = "mu")
-  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
-  dat$time <- as.numeric(as.character(dat$time))
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  ## repeated measures ANOVA in R
-  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
-  dat$preds_lm <- predict(fit_lm)
-
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  dat$y_pred <- predict(fit_lme)
-
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_lm = fit_lm,
-    fit_lme = fit_lme
-
-  ))
-}
-
-plot_example <- function(sim_dat) {
-  library(patchwork)
-  ## Plot the simulated data
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE) +labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
-    theme_classic() +
-    #ggtitle("Simulated data with true response function")+
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    #ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                face = "bold"),
-     text=element_text(size=30))+
-    scale_color_aaas()
-
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='errors')+
-     theme_classic()+
-    # facet_wrap(~ treatment) +
-    #ggtitle("Simulated errors") +
-     theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point()+labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    #ggtitle("Fitted Model")+
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
-
-```
-
-```{r, linear-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data**"}
-# linear response, correlated errors (subject effect)
-A1/B1+plot_annotation(tag_levels = 'A')
-```
-
-For the quadratic response case, Figure \@ref(fig:quadratic-cases) shows the simulated responses using compound symmetry and independent errors.
-
-```{r, quadratic-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
-# linear response, correlated errors (subject effect)
-C1/D1+plot_annotation(tag_levels = 'A')
-```
-
-## Basis functions and GAMs
-
-## Data simulation and GAM models

### 2021-05-13 Manuscripts/Bookdown testing/03-GAM Theory.Rmd
Splitted the document in sections, author section now works. Need to work on Appendix, as there is some issue going

--- 2021-05-11 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
@@
----
-title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
-subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
-author:
-- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
-- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
-  Fayetteville]
-- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
-output:
-  bookdown::pdf_document2:
-    keep_tex: yes
-    fig_caption: yes
-    extra_dependencies:
-      subfig: null
-      breqn: null
-      caption: ["font={small}"]
-      float: null
-  bookdown::word_document2:
-    fig_caption: yes
-    keep_md: yes
-  bookdown::html_document2: default
-  link-citations: yes
-  css: style.css
-  '': default
-csl: elsevier-with-titles.csl
-bibliography: refs.bib
----
-
-# Abstract
-
-In biomedical research, the outcome of longitudinal studies is frequently analyzed statistically using the repeated measures analysis or variance (rm-ANOVA).
-However, rm-ANOVA is restrictive as is only able to accommodate linear responses, requires equal correlation between each measurement (compound symmetry), and cannot work with missing observations.
-When the data fails to fullfill all the required assumptions by rm-ANOVA, the model provides biased estimates and inference.
-In contrast, generalized additive models (GAMs) are a class of models that overcome all of the limitations imposed by rm-ANOVA and that can be implemented in a straightforward manner using the _mgcv_ package in R.
-This paper summarizes the assumptions of rm-ANOVA and how their limit its use in longitudinal biomedical research, presents the basic theoretical framework for GAMs and provides examples of their implementation in R.
-The code and data used to generate this paper are available at:
-
-# Background
-
-Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following RT in neck cancer patients [@kamstra2015].
-
-Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as a limiting frequency [@wagenmakers2008] by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA [@pinheiro2006].
-This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model is only able to accommodate linear trends that are far from adequately representing the biological phenomenon of interest.
-
-A _post hoc_ analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
-Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, _p-values_ are valid only if the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019].
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting _p-values_ and parameter estimates to be invalid [@wang2019].
-
-Additionally, the _p-value_ itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019].
-Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to balance in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].
-Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976;@greenhouse1959], but the effectiveness of the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001].
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not "large" due to ethical and budgetary reasons [@charan2013] which might cause the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial patients may voluntarily withdraw, whereas attrition  due to injury or weight loss in preclinical animal studies is possible.
-It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios,the _complete observations_ assumption of  classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis [@gueorguieva2004].
-This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
-At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
-
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data [@skala2010;@vishwanath2009].
-Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-However, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
-And even more importantly, LMEMs also expect a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze non-linear data.
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
-
-In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017] and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
-
-The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, $\textsf{R}$[@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
-hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017].
-Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
-
-This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations is explained in detail.
-Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, using simulated data that reproduces patterns in previously reported studies [@vishwanath2009] we illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
-The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in $\textsf{R}$, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
-
-In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models by improving the standards for reproducibility  in biomedical research.
-
-# Challenges presented by longitudinal studies
-
-## The repeated measures ANOVA
-
-The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research.
-This statistical methodology requires certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
-Each one of these assumptions is discussed below.
-
-## Linear relationship
-
-### The repeated measures ANOVA case
-
-In a biomedical longitudinal study, two or more groups of subjects (e.g., patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs.
-a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response is modeled with _fixed_ components.
-The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug in the "treatment" group.
-
-Mathematically speaking, a rm-ANOVA model with an interaction can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
-(\#eq:linear-model)
-\end{equation}
-
-In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$.
-In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
-"novel drug" or "saline" vs.
-"chemotherapy").
-Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B).
-The linear models then can be expressed as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-(\#eq:ANOVA-by-group)
-\end{equation}
-
-To further simplify the expression, substitute $\widetilde{\beta_{0}}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B.
-This substitution allows for a different intercept and slope for Groups A and B.
-The model is then written as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-(\#eq:ANOVA-lines)
-\end{equation}
-
-Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
-
-### The Linear Mixed Model Case
-
-A linear mixed model (LMEM) is a class of statistical model that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to distinguish [@pinheiro2006;@west2014].
-A LMEM with interaction between time and treatment for a longitudinal study  can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
-(\#eq:LMEM)
-\end{equation}
-
-When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
-This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject within each group.
-The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the "global noise" term $\varepsilon_{ijt}$ from Equation \@ref(eq:linear-model).
-
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
-As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
-The _random effect_ $\mu_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be independently normally distributed with mean zero and variance $\sigma^2_\mu$, which can be expressed as $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
-In essence,the _random effect_ in a LMEM enables to fit models with different slopes at the subject-level[@pinheiro2006].
-However,the expected linear relationship of the covariates and the response in Equation \@ref(eq:linear-model) and in Equation \@ref(eq:LMEM) is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
-
-## Covariance in rm-ANOVA and LMEMs
-
-In a longitudinal study there is an expected _covariance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
-The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
-
-In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
-Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)[@huynh1976;@greenhouse1959] the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others [@pinheiro2006].
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself.
-Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
-
-## Missing observations
-
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004].
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002].
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
-
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014].
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.
-
-This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and  missing data.
-In particular, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data.
-To better convey the issues of linearity and correlation  in linear models fitted to non-linear data, simulation is used in the next section.
-
-## How does an rm-ANOVA fit looks like? A visual representation using simulation {#simulation}
-
-To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme].
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C).
-In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity.
-
-Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates an effect due to treatment.
-In other words, Group 1 can be thought as a "Control" group and Group 2 as a "Treatment" group.
-From the mean response per group (linear or quadratic), the variability or "error" of individual responses  within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \@ref(fig:l-q-response) B,D).
-A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \@ref(fig:linear-cases) and Figure \@ref(fig:quadratic-cases) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-
-```{r, message = FALSE,include=FALSE}
-library(patchwork)
-library(tidyverse)
-library(mvnfast)
-library(nlme)
-library(ggsci)
-library(mgcv)
-library(gratia)
-
-knitr::opts_chunk$set(fig.pos = "!h", out.extra = "")
-```
-
-```{r,include=FALSE,message=FALSE,echo=FALSE}
-## Example with linear response
-example <- function(n_time = 6, 
-                    fun_type = "linear", 
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  x <- seq(1,6, length.out = n_time)
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # nonlinear response
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-  # matplot(mu, type = 'l')
-
-  y <- array(0, dim = c(length(x), 2, 10))
-  errors <- array(0, dim = c(length(x), 2, 10))
-
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(mu) <- list(time = x, treatment = 1:2)
-  dat <- as.data.frame.table(y, responseName = "y")
-  dat_errors <- as.data.frame.table(errors, responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, responseName = "mu")
-  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
-  dat$time <- as.numeric(as.character(dat$time))
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  ## repeated measures ANOVA in R
-  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
-  dat$preds_lm <- predict(fit_lm)
-
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  dat$y_pred <- predict(fit_lme)
-
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_lm = fit_lm,
-    fit_lme = fit_lme
-
-  ))
-}
-
-plot_example <- function(sim_dat) {
-  library(patchwork)
-  ## Plot the simulated data
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE) +labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
-    theme_classic() +
-    #ggtitle("Simulated data with true response function")+
-    theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    #ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_classic()+
-     theme(plot.title = element_text(size = 18, 
-                                face = "bold"),
-     text=element_text(size=18))+
-    scale_color_aaas()
-
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='errors')+
-     theme_classic()+
-    # facet_wrap(~ treatment) +
-    #ggtitle("Simulated errors") +
-     theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point()+labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    #ggtitle("Fitted Model")+
-    theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  #return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-  return((p1+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-A<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
-
-```
-
-(ref:l-q-response-caption) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.
-
-```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='ref(l-q-response-caption)' }
-# linear response, correlated errors (subject effect)
-par(mar = c(0.5, 0.5, 0.5, 0.5))
-A/C+plot_annotation(tag_levels = 'A')
-```
-
-The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B).
-When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.
-
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response), C.
-Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \@ref(fig:l-q-response) A and B) .
-The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panel D in Figure \@ref(fig:l-q-response).
-
-A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\@ref(fig:l-q-response), D) indicates that the model is not capturing the changes within each group in a good way.
-Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
-Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D).
-If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response.
-Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
-
-This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-Although the model fitted to the simulated data was an rm-ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM.
-In the following section, we present generalized additive models (GAMs) as a data-driven  alternative method to analyze longitudinal non-linear data.
-
 # GAMs as a special case of Generalized Linear Models
@@
 ```
-
- \newpage
-
-# The analyisis of longitudinal biomedical data using GAMs
-
-The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA.
-This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using $\textsf{R}$ and the package `mgcv`.
-The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
-
-## Simulated data
-
-The simulated data is based on the reported longitudinal changes in oxygen saturation ($StO_2$) in subcutaneous tumors that appear in Figure 3(C) in [vishwanath2009].
-An increase in oxygen saturation in the "Treatment" group (chemotherapy) is observed through time, while a decrease is seen in the "Control" (saline) group.
-Following the reported trend, 10 normally distributed observations at each time point are generated with a standard deviation (SD) of 10% (matching the SD in the original paper).
-The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot), A and the inlet, respectively.
-
-## An interaction GAM model for longitudinal data
-
-An interaction model is typically the main interest in longitudinal biomedical data, as it takes into account the effect of treatment, time, and their combination.
-In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the _time_ effect to account for the repeated measures over time.
-Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in [@hefley2017].
-
-For the data in Figure \@ref(fig:sim-smooth-plot),A it is of interest to know the effect that each treatment has in oxygen saturation.
-The model then needs to incorporate independent smooths for _Group_ and _Day_, respectively.
-The main thing to consider is the model syntax to account for the fact that one of the variables is numeric ( _Day_ ) and that the other is a factor ( _Group_ ).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
-Using $\textsf{R}$ and the package `mgcv` the model syntax is:
-
-`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim`
-
-This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation `StO2_sim` is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
-The parametric term _Group_ is added to separate the effect of treatment, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $StO_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B).
-Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package `gratia`[@gratia].
-A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].
-
-One question might be "what is the fit that an rm-ANOVA model produces for the simulated data?".
-The rm-ANOVA model, which corresponds to \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot),C.
-This is a typical case of model misspecification: The slopes of each group ar different, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \@ref(fig:sim-smooth-plot),B) .
-
-```{r,sim-tumor-response,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
-
-## plot the mean response
-f1<-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
-    geom_line(size=1,show.legend = FALSE)+
-    geom_point(show.legend = FALSE)+
-  labs(y=expression(paste(StO[2],' (real)')))+
-  theme_classic()+
-  scale_color_aaas()+
-    scale_x_continuous(breaks=c(0,5,10))+
-    scale_y_continuous(breaks=c(0,40))+
-  plot_layout(tag_level = 'new')+
-  theme(
-    plot.background = element_rect(fill = "transparent", color = NA),
-    axis.text=element_text(size=14)
-  )
-
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-
-n <- 10 #number of observations
-sd <- 10 #approximate sd from paper
-set.seed(1) #set seed for reproducibility
-df <- 6
-dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group)) +
-    geom_point(show.legend=FALSE)+
-    stat_summary(aes(y = StO2_sim,
-                     group=Group), 
-                 fun=mean, geom="line",size=1,show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],'(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-  scale_color_aaas()+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
-
-# tumor_sim<-f1+f2+plot_layout(guides = 'collect')+
-#   plot_annotation(tag_levels='A')&
-#   ylim(c(0,75)) &
-#   theme(
-#      text=element_text(size=18),
-#      legend.title=element_blank(),
-#      legend.text = element_blank()
-#      )&
-#   scale_color_aaas()
-
-```
-
-```{r,smooth-plot,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-#GAM for StO2
-
-gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
-            method='REML',
-            data  = dat_sim)
-
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-#creates a dataframe using the length of the covariates for the GAM
-gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
-                         subject=factor(rep(1:10)))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#adds the predictions to the grid and creates a confidence interval for GAM
-gam_predict<-gam_predict%>%
-    mutate(fit = predict(gam1,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(gam1, gam_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lme
-#lm_predict<-lm_predict%>%
-#   mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response',level=0),
- #         se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response',level=0))
-
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=gam_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'complete')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-#posthoc comparisons for the linear model
-library(multcomp)
-
-summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
-
-```
-
-Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
-Using the simulated $StO_2$ values from Figure (\@ref(fig:sim-smooth-plot), B), 40% of the total observations can be randomly deleted and the same GAM model with interaction used for the full simulated data can be fitted to the remaining observations.
-
-```{r,missing-data,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
-
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'missing')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-```
-
-(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent smooths for each group, shaded regions are 95% confidence intervals.C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40% of its observations missing.
-Lines represent smooths for each group, shaded regions are 95% confidence intervals
-
-```{r,sim-smooth-plot,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}
-
-mult_plot<-f2+inset_element(
-  f1, left = 0.01, 
-  bottom = 0.5, 
-  right = 0.5, 
-  top = 1.0)+
-  f3+f4+f6+
-  #plot_layout(guides = 'collect')+
-   plot_annotation(tag_levels='A')&
-   ylim(c(-5,75)) &
-  theme(
-     text=element_text(size=18)
-     )&
-  scale_color_aaas()
-
-mult_plot
-
-```
-
-## Determination of significance in GAMs
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that timepoint (or timepoints).
-The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
-The major advantage of the significance estimation in GAMs  is that they allow to determine significance at a specific timepoint, rather than providing a single _p-value_ for the overall treatment effect.
-The estimation of siginificant differences in GAMs is covered in detail in the Appendix using simulated longitudinal data that follows previously reported trends of tumor response between multiple treatment groups [@vishwanath2009].
-
-```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-##Pairwise comparisons
-
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Treatment', 'Control'))
-
-smooth_diff <- function(model, newdata, f1, f2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    c1 <- grepl(f1, colnames(xp))
-    c2 <- grepl(f2, colnames(xp))
-    #r1 <- newdata[[var]] == f1
-    #r2 <- newdata[[var]] == f2
-    r1 <- with(newdata, Group == f1)
-    r2 <- with(newdata, Group == f2)
-    ## difference rows of xp for data from comparison
-    X <- xp[r1, ] - xp[r2, ]
-    ## zero out cols of X related to splines for other lochs
-    X[, ! (c1 | c2)] <- 0
-    ## zero out the parametric cols
-    X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(f1, f2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-}
-
-comp1<-smooth_diff(gam1,pdat,'Control','Treatment')
-
-comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
-
-c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5,fill='#993336') +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-  theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-#+
- # geom_rect(aes(color=interval))
-
-c1
-
-###for missing data
-comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
-comp_StO2_m <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp2))
-
-missing_plot<-ggplot(comp_StO2_m, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    labs(x = 'Days', 
-         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
-                              )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme_classic()+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
-```
-
-***
-
-# References
-
-<div id="refs"></div>
-
-\newpage
-
-# (APPENDIX) Appendix {-}
-
-# Simulation
-
-## Compound symmetry and independent errors in linear and quadratic responses
-
-This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation).
-The linear simulations using Figure \@ref(fig:linear-cases) show in  panels A and D the simulated mean responses and individual datapoints.
-Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-```{r,include=TRUE,message=FALSE,echo=TRUE}
-## Example with linear response
-example <- function(n_time = 6, 
-                    fun_type = "linear", 
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  library(tidyverse)
-  library(mvnfast)
-  library(nlme)
-  library(ggsci)
-
-  x <- seq(1,6, length.out = n_time)
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # nonlinear response
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-  # matplot(mu, type = 'l')
-
-  y <- array(0, dim = c(length(x), 2, 10))
-  errors <- array(0, dim = c(length(x), 2, 10))
-
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(mu) <- list(time = x, treatment = 1:2)
-  dat <- as.data.frame.table(y, responseName = "y")
-  dat_errors <- as.data.frame.table(errors, responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, responseName = "mu")
-  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
-  dat$time <- as.numeric(as.character(dat$time))
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  ## repeated measures ANOVA in R
-  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
-  dat$preds_lm <- predict(fit_lm)
-
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  dat$y_pred <- predict(fit_lme)
-
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_lm = fit_lm,
-    fit_lme = fit_lme
-
-  ))
-}
-
-plot_example <- function(sim_dat) {
-  library(patchwork)
-  ## Plot the simulated data
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE) +labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
-    theme_classic() +
-    #ggtitle("Simulated data with true response function")+
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    #ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                face = "bold"),
-     text=element_text(size=30))+
-    scale_color_aaas()
-
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='errors')+
-     theme_classic()+
-    # facet_wrap(~ treatment) +
-    #ggtitle("Simulated errors") +
-     theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point()+labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    #ggtitle("Fitted Model")+
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
-
-```
-
-```{r, linear-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data**"}
-# linear response, correlated errors (subject effect)
-A1/B1+plot_annotation(tag_levels = 'A')
-```
-
-For the quadratic response case, Figure \@ref(fig:quadratic-cases) shows the simulated responses using compound symmetry and independent errors.
-
-```{r, quadratic-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
-# linear response, correlated errors (subject effect)
-C1/D1+plot_annotation(tag_levels = 'A')
-```

### 2021-05-13 Manuscripts/Bookdown testing/04-Longitudinal analysis with GAMs.Rmd
Splitted the document in sections, author section now works. Need to work on Appendix, as there is some issue going

--- 2021-05-12 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
@@
----
-title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
-subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
-author:
-- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
-- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
-  Fayetteville]
-- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
-output:
-  bookdown::pdf_document2:
-    keep_tex: yes
-    fig_caption: yes
-    extra_dependencies:
-      subfig: null
-      breqn: null
-      caption: ["font={small}"]
-      float: null
-  bookdown::word_document2:
-    fig_caption: yes
-    keep_md: yes
-  bookdown::html_document2: default
-  link-citations: yes
-  css: style.css
-  '': default
-csl: elsevier-with-titles.csl
-bibliography: refs.bib
----
-
-# Abstract
-
-In biomedical research, the outcome of a longitudinal study has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, a _linear mixed model_ (LMEM).
-Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
-In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
-Although not commonly used in biomedical research, GAMs present an excellent choice to analyze non-linear longitudinal data.
-This paper summarizes the limitations of linear models (rm-ANOVA in particular), presents the basic theory of GAMs, and demonstrates their implementation in $\textsf{R}$ via the package _mgcv_ using simulated data that follows longitudinal trends reported in biomedical literature.
-To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:________.
-
-# Background
-
-Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients [@kamstra2015].
-
-Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times [@wagenmakers2008], and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA [@pinheiro2006].
-This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model is only able to accommodate linear trends that are far from adequately representing the biological phenomenon of interest.
-
-A _post hoc_ analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
-Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, _p-values_ are valid only if the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019].
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting _p-values_ and parameter estimates to be invalid [@wang2019].
-
-Additionally, the _p-value_ itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019].
-Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].
-Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976;@greenhouse1959], but the effectiveness of the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001].
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not "large" due to ethical and budgetary reasons [@charan2013] which might cause the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial patients may voluntarily withdraw, whereas attrition  due to injury or weight loss in preclinical animal studies is possible.
-It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios, the _complete observations_ assumption of  classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis [@gueorguieva2004].
-This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
-At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
-
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data [@skala2010;@vishwanath2009].
-Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-However, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
-And even more importantly, LMEMs also expect a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze non-linear data.
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
-
-In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017] and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
-
-The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, $\textsf{R}$[@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
-hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017].
-Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
-
-This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations is explained in detail.
-Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, using simulated data that reproduces patterns in previously reported studies [@vishwanath2009] we illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
-The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in $\textsf{R}$, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
-
-In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models by improving the standards for reproducibility  in biomedical research.
-
-# Challenges presented by longitudinal studies
-
-## The repeated measures ANOVA
-
-The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research.
-This statistical methodology requires certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
-Each one of these assumptions is discussed below.
-
-## Linear relationship
-
-### The repeated measures ANOVA case
-
-In a longitudinal biomedical study, two or more groups of subjects (e.g., human subject, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug or intervention vs.
-a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response is modeled with _fixed_ components.
-The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug/intervention in the "treatment" group.
-
-Mathematically speaking, a rm-ANOVA model with an interaction can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
-(\#eq:linear-model)
-\end{equation}
-
-In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$ (independently normally distributed with mean zero and variance $\sigma^2_\mu$).
-In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
-"novel drug" or "saline" vs.
-"chemotherapy").
-Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B).
-The linear models then can be expressed as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-(\#eq:ANOVA-by-group)
-\end{equation}
-
-To further simplify the expression, substitute $\widetilde{\beta_{0}}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B.
-This substitution allows for a different intercept and slope for Groups A and B.
-The model is then written as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-(\#eq:ANOVA-lines)
-\end{equation}
-
-Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
-
-### The Linear Mixed Model Case
-
-A linear mixed model (LMEM) is a class of statistical model that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to distinguish [@pinheiro2006;@west2014].
-A LMEM with interaction between time and treatment for a longitudinal study  can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
-(\#eq:LMEM)
-\end{equation}
-
-When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
-This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject within each group.
-The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the "global noise" term $\varepsilon_{ijt}$ from Equation \@ref(eq:linear-model).
-
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
-As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
-The _random effect_ $\mu_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
-In essence,the _random effect_ in a LMEM enables to fit models with different slopes at the subject-level[@pinheiro2006].
-However,the expected linear relationship of the covariates and the response in Equation \@ref(eq:linear-model) and in Equation \@ref(eq:LMEM) is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
-
-## Covariance in rm-ANOVA and LMEMs
-
-In a longitudinal study there is an expected _covariance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
-The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
-
-In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
-Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)[@huynh1976;@greenhouse1959] the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others [@pinheiro2006].
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
-Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
-
-## Missing observations
-
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004].
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002].
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
-
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014].
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.
-
-This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and  missing data.
-In particular, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data.
-To better convey the issues of linearity and correlation  in linear models fitted to non-linear data, simulation is used in the next section.
-
-## What does an rm-ANOVA fit looks like? A visual representation using simulated data {#simulation}
-
-To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme].
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C).
-In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity.
-
-Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates an effect due to treatment.
-In other words, Group 1 can be thought as a "Control" group and Group 2 as a "Treatment" group.
-From the mean response per group (linear or quadratic), the variability or "error" of individual responses  within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \@ref(fig:l-q-response) B,D).
-A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \@ref(fig:linear-cases) and Figure \@ref(fig:quadratic-cases) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-
-```{r, message = FALSE,include=FALSE}
-library(patchwork)
-library(tidyverse)
-library(mvnfast)
-library(nlme)
-library(ggsci)
-library(mgcv)
-library(gratia)
-
-knitr::opts_chunk$set(fig.pos = "!h", out.extra = "")
-```
-
-```{r,include=FALSE,message=FALSE,echo=FALSE}
-## Example with linear response. This function generates either linear or quadratic mean responses with correlated or uncorrelated errors and fits a linear model to the data.
-example <- function(n_time = 6, 
-                    fun_type = "linear", 
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  x <- seq(1,6, length.out = n_time)
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # nonlinear response
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-  # matplot(mu, type = 'l')
-
-  y <- array(0, dim = c(length(x), 2, 10))
-  errors <- array(0, dim = c(length(x), 2, 10))
-
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(mu) <- list(time = x, treatment = 1:2)
-  dat <- as.data.frame.table(y, responseName = "y")
-  dat_errors <- as.data.frame.table(errors, responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, responseName = "mu")
-  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
-  dat$time <- as.numeric(as.character(dat$time))
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  ## repeated measures ANOVA in R
-  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
-  dat$preds_lm <- predict(fit_lm)
-
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  dat$y_pred <- predict(fit_lme)
-
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_lm = fit_lm,
-    fit_lme = fit_lme
-
-  ))
-}
-
-plot_example <- function(sim_dat) {
-  library(patchwork)
-  ## Plot the simulated data
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE,size=1.5,alpha=0.5) +
-    labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment,size=1),show.legend=FALSE) +
-    theme_classic() +
-    #ggtitle("Simulated data with true response function")+
-    theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"),
-              lty = 1,show.legend = FALSE) +labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    #ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_classic()+
-     theme(plot.title = element_text(size = 18, 
-                                face = "bold"),
-     text=element_text(size=18))+
-    scale_color_aaas()
-
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='errors')+
-     theme_classic()+
-     theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point(size=1.5,alpha=0.5)+labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    #ggtitle("Fitted Model")+
-    theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  #return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-  return((p1+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-A<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
-
-```
-
-(ref:l-q-response-caption) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.
-
-```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='ref(l-q-response-caption)' }
-# linear response, correlated errors (subject effect)
-par(mar = c(0.5, 0.5, 0.5, 0.5))
-A/C+plot_annotation(tag_levels = 'A')
-```
-
-The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B).
-When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.
-
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response), C.
-Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \@ref(fig:l-q-response) A and B) .
-The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panel D in Figure \@ref(fig:l-q-response).
-
-A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\@ref(fig:l-q-response), D) indicates that the model is not capturing the changes within each group in a good way.
-Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
-Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D).
-If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response.
-Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
-
-This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-Although the model fitted to the simulated data was an rm-ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM.
-In the following section, we present generalized additive models (GAMs) as a data-driven  alternative method to analyze longitudinal non-linear data.
-
-# GAMs as a special case of Generalized Linear Models
-
-## GAMs and Basis Functions
-
-Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors[@nelder1972].
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case[@simpson2018;@wood2017;@hastie1987].
-A GAM model can be written as:
-
-\begin{equation}
-  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
-  (\#eq:GAM)
-\end{equation}
-
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual error.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
-In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.
-
-The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using _basis functions_ expansions of the covariates and by estimating random coefficients for these basis functions.
-A _basis_ is a set of functions that spans the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
-For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
-In the case of Equation \@ref(eq:GAM), the basis function is $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
-
-A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness [@wood2017;@simpson2018].
-Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\@ref(fig:l-q-response), C).
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-The timeline can be divided in equally spaced _knots_, each knot being a region where a different basis function will be used.
-Because there are six timepoints for this group, five knots can be used.
-The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots.
-In Panel A of Figure \@ref(fig:basis-plot), the four basis functions (and the intercept) are shown.
-Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the "wigliness" of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
-The penalty reduces the "wigliness" of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
-
-```{r,basis-functions-plot, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-#generate the response
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
-
- #add subject column
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  dat<-subset(dat,treatment==1)
-  dat<-dat[,c('y','time')]
-gm<-gam(y~s(time,k=5),data=dat)
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-x_new <- (seq(0, max(dat$time), length.out = 100))
-y_pred <- predict(gm, newdata=dat)
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof)
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final spline by aggregating the basis functions
-
-spl<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(spl=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=as.factor(Basis)))+
-  geom_line(size=sz,show.legend=FALSE)+
-  geom_point(size=sz+1,show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,labeller = as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#penalized basis
-p12<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
-  geom_line(show.legend = FALSE,size=sz)+
-  geom_point(show.legend = FALSE,size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,labeller=as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#heatmap of the penalization coefficient
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
-  geom_tile(aes(color='black'),size=sz+1,show.legend = FALSE)+
-  geom_tile(size=sz+1)+
-  scale_fill_gradient(low = "white", high = "red")+
-  labs(x='Basis',y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),size=10,show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  scale_color_aaas()+
-  labs(y='Simulated \n response')+
-  geom_line(data=spl,aes(x=time,y=spl),color="#B15731",size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-
-```
-
-In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \@ref(fig:basis-plot), Panel C shows the resulting penalized basis functions.
-Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
-Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
-
-(ref:basis-plot-caption) Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.
-
-```{r,basis-plot,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap='(ref:basis-plot-caption)'}
-
-par(mar = c(2, 2, 2, 2))
-
-b_plot
-
-```
-
 \newpage
@@
 This would suggest that the effect of chemotherapy is significant after day 5 for the model used; which is noticeable in the data with missing observations albeit the missing data model has a wider confidence interval and is not able to pick the change between days 7 and 10 that the full dataset model is able to detect.
-
-# Conclusion
-
-***
-
-\newpage
-# References
-
-<div id="refs"></div>
-
-# (APPENDIX) Appendix {-}
-
-# Simulation
-
-## Compound symmetry and independent errors in linear and quadratic responses
-
-This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation).
-The linear simulations using Figure \@ref(fig:linear-cases) show in  panels A and D the simulated mean responses and individual datapoints.
-Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-```{r,include=TRUE,message=FALSE,echo=TRUE}
-## Example with linear response
-example <- function(n_time = 6, 
-                    fun_type = "linear", 
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  library(tidyverse)
-  library(mvnfast)
-  library(nlme)
-  library(ggsci)
-
-  x <- seq(1,6, length.out = n_time)
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # nonlinear response
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-  # matplot(mu, type = 'l')
-
-  y <- array(0, dim = c(length(x), 2, 10))
-  errors <- array(0, dim = c(length(x), 2, 10))
-
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(mu) <- list(time = x, treatment = 1:2)
-  dat <- as.data.frame.table(y, responseName = "y")
-  dat_errors <- as.data.frame.table(errors, responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, responseName = "mu")
-  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
-  dat$time <- as.numeric(as.character(dat$time))
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  ## repeated measures ANOVA in R
-  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
-  dat$preds_lm <- predict(fit_lm)
-
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  dat$y_pred <- predict(fit_lme)
-
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_lm = fit_lm,
-    fit_lme = fit_lme
-
-  ))
-}
-
-plot_example <- function(sim_dat) {
-  library(patchwork)
-  ## Plot the simulated data
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE) +labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
-    theme_classic() +
-    #ggtitle("Simulated data with true response function")+
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    #ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                face = "bold"),
-     text=element_text(size=30))+
-    scale_color_aaas()
-
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='errors')+
-     theme_classic()+
-    # facet_wrap(~ treatment) +
-    #ggtitle("Simulated errors") +
-     theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point()+labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    #ggtitle("Fitted Model")+
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
-
-```
-
-```{r, linear-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data**"}
-# linear response, correlated errors (subject effect)
-A1/B1+plot_annotation(tag_levels = 'A')
-```
-
-For the quadratic response case, Figure \@ref(fig:quadratic-cases) shows the simulated responses using compound symmetry and independent errors.
-
-```{r, quadratic-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
-# linear response, correlated errors (subject effect)
-C1/D1+plot_annotation(tag_levels = 'A')
-```
-
-## Basis functions and GAMs
-
-## Data simulation and GAM models

### 2021-05-13 Manuscripts/Bookdown testing/06-Appendix.Rmd
Splitted the document in sections, author section now works. Need to work on Appendix, as there is some issue going

--- 2021-05-11 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
@@
----
-title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
-subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
-author:
-- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
-- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
-  Fayetteville]
-- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
-output:
-  bookdown::pdf_document2:
-    keep_tex: yes
-    fig_caption: yes
-    extra_dependencies:
-      subfig: null
-      breqn: null
-      caption: ["font={small}"]
-      float: null
-  bookdown::word_document2:
-    fig_caption: yes
-    keep_md: yes
-  bookdown::html_document2: default
-  link-citations: yes
-  css: style.css
-  '': default
-csl: elsevier-with-titles.csl
-bibliography: refs.bib
----
-
-# Abstract
-
-In biomedical research, the outcome of longitudinal studies is frequently analyzed statistically using the repeated measures analysis or variance (rm-ANOVA).
-However, rm-ANOVA is restrictive as is only able to accommodate linear responses, requires equal correlation between each measurement (compound symmetry), and cannot work with missing observations.
-When the data fails to fullfill all the required assumptions by rm-ANOVA, the model provides biased estimates and inference.
-In contrast, generalized additive models (GAMs) are a class of models that overcome all of the limitations imposed by rm-ANOVA and that can be implemented in a straightforward manner using the _mgcv_ package in R.
-This paper summarizes the assumptions of rm-ANOVA and how their limit its use in longitudinal biomedical research, presents the basic theoretical framework for GAMs and provides examples of their implementation in R.
-The code and data used to generate this paper are available at:
-
-# Background
-
-Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following RT in neck cancer patients [@kamstra2015].
-
-Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as a limiting frequency [@wagenmakers2008] by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA [@pinheiro2006].
-This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model is only able to accommodate linear trends that are far from adequately representing the biological phenomenon of interest.
-
-A _post hoc_ analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
-Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, _p-values_ are valid only if the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019].
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting _p-values_ and parameter estimates to be invalid [@wang2019].
-
-Additionally, the _p-value_ itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019].
-Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to balance in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].
-Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976;@greenhouse1959], but the effectiveness of the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001].
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not "large" due to ethical and budgetary reasons [@charan2013] which might cause the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial patients may voluntarily withdraw, whereas attrition  due to injury or weight loss in preclinical animal studies is possible.
-It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios,the _complete observations_ assumption of  classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis [@gueorguieva2004].
-This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
-At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
-
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data [@skala2010;@vishwanath2009].
-Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-However, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
-And even more importantly, LMEMs also expect a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze non-linear data.
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
-
-In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017] and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
-
-The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, $\textsf{R}$[@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
-hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017].
-Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
-
-This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations is explained in detail.
-Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, using simulated data that reproduces patterns in previously reported studies [@vishwanath2009] we illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
-The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in $\textsf{R}$, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
-
-In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models by improving the standards for reproducibility  in biomedical research.
-
-# Challenges presented by longitudinal studies
-
-## The repeated measures ANOVA
-
-The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research.
-This statistical methodology requires certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
-Each one of these assumptions is discussed below.
-
-## Linear relationship
-
-### The repeated measures ANOVA case
-
-In a biomedical longitudinal study, two or more groups of subjects (e.g., patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs.
-a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response is modeled with _fixed_ components.
-The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug in the "treatment" group.
-
-Mathematically speaking, a rm-ANOVA model with an interaction can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
-(\#eq:linear-model)
-\end{equation}
-
-In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$.
-In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
-"novel drug" or "saline" vs.
-"chemotherapy").
-Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B).
-The linear models then can be expressed as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-(\#eq:ANOVA-by-group)
-\end{equation}
-
-To further simplify the expression, substitute $\widetilde{\beta_{0}}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B.
-This substitution allows for a different intercept and slope for Groups A and B.
-The model is then written as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-(\#eq:ANOVA-lines)
-\end{equation}
-
-Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
-
-### The Linear Mixed Model Case
-
-A linear mixed model (LMEM) is a class of statistical model that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to distinguish [@pinheiro2006;@west2014].
-A LMEM with interaction between time and treatment for a longitudinal study  can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
-(\#eq:LMEM)
-\end{equation}
-
-When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
-This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject within each group.
-The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the "global noise" term $\varepsilon_{ijt}$ from Equation \@ref(eq:linear-model).
-
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
-As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
-The _random effect_ $\mu_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be independently normally distributed with mean zero and variance $\sigma^2_\mu$, which can be expressed as $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
-In essence,the _random effect_ in a LMEM enables to fit models with different slopes at the subject-level[@pinheiro2006].
-However,the expected linear relationship of the covariates and the response in Equation \@ref(eq:linear-model) and in Equation \@ref(eq:LMEM) is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
-
-## Covariance in rm-ANOVA and LMEMs
-
-In a longitudinal study there is an expected _covariance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
-The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
-
-In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
-Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)[@huynh1976;@greenhouse1959] the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others [@pinheiro2006].
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself.
-Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
-
-## Missing observations
-
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004].
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002].
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
-
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014].
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.
-
-This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and  missing data.
-In particular, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data.
-To better convey the issues of linearity and correlation  in linear models fitted to non-linear data, simulation is used in the next section.
-
-## How does an rm-ANOVA fit looks like? A visual representation using simulation {#simulation}
-
-To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme].
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C).
-In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity.
-
-Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates an effect due to treatment.
-In other words, Group 1 can be thought as a "Control" group and Group 2 as a "Treatment" group.
-From the mean response per group (linear or quadratic), the variability or "error" of individual responses  within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \@ref(fig:l-q-response) B,D).
-A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \@ref(fig:linear-cases) and Figure \@ref(fig:quadratic-cases) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-
-```{r, message = FALSE,include=FALSE}
-library(patchwork)
-library(tidyverse)
-library(mvnfast)
-library(nlme)
-library(ggsci)
-library(mgcv)
-library(gratia)
-
-knitr::opts_chunk$set(fig.pos = "!h", out.extra = "")
-```
-
-```{r,include=FALSE,message=FALSE,echo=FALSE}
-## Example with linear response
-example <- function(n_time = 6, 
-                    fun_type = "linear", 
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  x <- seq(1,6, length.out = n_time)
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # nonlinear response
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-  # matplot(mu, type = 'l')
-
-  y <- array(0, dim = c(length(x), 2, 10))
-  errors <- array(0, dim = c(length(x), 2, 10))
-
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(mu) <- list(time = x, treatment = 1:2)
-  dat <- as.data.frame.table(y, responseName = "y")
-  dat_errors <- as.data.frame.table(errors, responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, responseName = "mu")
-  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
-  dat$time <- as.numeric(as.character(dat$time))
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  ## repeated measures ANOVA in R
-  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
-  dat$preds_lm <- predict(fit_lm)
-
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  dat$y_pred <- predict(fit_lme)
-
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_lm = fit_lm,
-    fit_lme = fit_lme
-
-  ))
-}
-
-plot_example <- function(sim_dat) {
-  library(patchwork)
-  ## Plot the simulated data
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE,size=1.5,alpha=0.5) +
-    labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment,size=1),show.legend=FALSE) +
-    theme_classic() +
-    #ggtitle("Simulated data with true response function")+
-    theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"),
-              lty = 1,show.legend = FALSE) +labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    #ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_classic()+
-     theme(plot.title = element_text(size = 18, 
-                                face = "bold"),
-     text=element_text(size=18))+
-    scale_color_aaas()
-
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='errors')+
-     theme_classic()+
-     theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point(size=1.5,alpha=0.5)+labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    #ggtitle("Fitted Model")+
-    theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  #return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-  return((p1+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-A<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
-
-```
-
-(ref:l-q-response-caption) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.
-
-```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='ref(l-q-response-caption)' }
-# linear response, correlated errors (subject effect)
-par(mar = c(0.5, 0.5, 0.5, 0.5))
-A/C+plot_annotation(tag_levels = 'A')
-```
-
-The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B).
-When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.
-
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response), C.
-Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \@ref(fig:l-q-response) A and B) .
-The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panel D in Figure \@ref(fig:l-q-response).
-
-A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\@ref(fig:l-q-response), D) indicates that the model is not capturing the changes within each group in a good way.
-Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
-Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D).
-If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response.
-Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
-
-This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-Although the model fitted to the simulated data was an rm-ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM.
-In the following section, we present generalized additive models (GAMs) as a data-driven  alternative method to analyze longitudinal non-linear data.
-
-# GAMs as a special case of Generalized Linear Models
-
-## GAMs and Basis Functions
-
-Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors[@nelder1972].
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case[@simpson2018;@wood2017;@hastie1987].
-A GAM model can be written as:
-
-\begin{equation}
-  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
-  (\#eq:GAM)
-\end{equation}
-
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual error.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
-In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.
-
-The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using _basis functions_ expansions of the covariates and by estimating random coefficients for these basis functions.
-A _basis_ is a set of functions that spans the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
-For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
-In the case of Equation \@ref(eq:GAM), the basis function is $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
-
-A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness [@wood2017;@simpson2018].
-Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\@ref(fig:l-q-response), C).
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-The timeline can be divided in equally spaced _knots_, each knot being a region where a different basis function will be used.
-Because there are six timepoints for this group, five knots can be used.
-The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots.
-In Panel A of Figure \@ref(fig:basis-plot), the four basis functions (and the intercept) are shown.
-Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the "wigliness" of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
-The penalty reduces the "wigliness" of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
-
-```{r,basis-functions-plot, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  dat<-subset(dat,treatment==1)
-  dat<-dat[,c('y','time')]
-gm<-gam(y~s(time,k=5),data=dat)
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-x_new <- (seq(0, max(dat$time), length.out = 100))
-y_pred <- predict(gm, newdata=dat)
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof)
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final spline by aggregating the basis functions
-
-spl<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(spl=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=as.factor(Basis)))+
-  geom_line(size=sz,show.legend=FALSE)+
-  geom_point(size=sz+1,show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,labeller = as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#penalized basis
-p12<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
-  geom_line(show.legend = FALSE,size=sz)+
-  geom_point(show.legend = FALSE,size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,labeller=as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#heatmap of the penalization coefficient
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
-  geom_tile(aes(color='black'),size=sz+1,show.legend = FALSE)+
-  geom_tile(size=sz+1)+
-  scale_fill_gradient(low = "white", high = "red")+
-  labs(x='Basis',y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),size=10,show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  scale_color_aaas()+
-  labs(y='Simulated \n response')+
-  geom_line(data=spl,aes(x=time,y=spl),color="#B15731",size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-
-```
-
-In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \@ref(fig:basis-plot), Panel C shows the resulting penalized basis functions.
-Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
-Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
-
-(ref:basis-plot-caption) Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.
-
-```{r,basis-plot,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap='(ref:basis-plot-caption)'}
-
-par(mar = c(2, 2, 2, 2))
-
-b_plot
-
-```
-
- \newpage
-
-# The analyisis of longitudinal biomedical data using GAMs
-
-The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA.
-This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using $\textsf{R}$ and the package `mgcv`.
-The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
-
-## Simulated data
-
-The simulated data is based on the reported longitudinal changes in oxygen saturation ($StO_2$) in subcutaneous tumors that appear in Figure 3(C) in [@vishwanath2009].
-An increase in oxygen saturation in the "Treatment" group (chemotherapy) is observed through time, while a decrease is seen in the "Control" (saline) group.
-Following the reported trend, 10 normally distributed observations at each time point are generated with a standard deviation (SD) of 10% (matching the SD in the original paper).
-The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot), A and the inlet, respectively.
-
-## An interaction GAM model for longitudinal data
-
-An interaction model is typically the main interest in longitudinal biomedical data, as it takes into account the effect of treatment, time, and their combination.
-In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the _time_ effect to account for the repeated measures over time.
-Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in [@hefley2017].
-
-For the data in Figure \@ref(fig:sim-smooth-plot),A it is of interest to know the effect that each treatment has in oxygen saturation.
-The model then needs to incorporate independent smooths for _Group_ and _Day_, respectively.
-The main thing to consider is the model syntax to account for the fact that one of the variables is numeric ( _Day_ ) and that the other is a factor ( _Group_ ).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
-Using $\textsf{R}$ and the package `mgcv` the model syntax is:
-
-`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim`
-
-This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
-The parametric term _Group_ is added to separate the effect of treatment, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $StO_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B).
-Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package `gratia`[@gratia].
-A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].
-
-One question might be "what is the fit that an rm-ANOVA model produces for the simulated data?".
-The rm-ANOVA model, which corresponds to \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot),C.
-This is a typical case of model misspecification: The slopes of each group ar different, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \@ref(fig:sim-smooth-plot),B) .
-
-```{r,sim-tumor-response,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
-
-## plot the mean response
-f1<-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
-    geom_line(size=1,show.legend = FALSE)+
-    geom_point(show.legend = FALSE,size=1.5,alpha=0.5)+
-  labs(y=expression(paste(StO[2],' (real)')))+
-  theme_classic()+
-  scale_color_aaas()+
-    scale_x_continuous(breaks=c(0,5,10))+
-    scale_y_continuous(breaks=c(0,40))+
-  plot_layout(tag_level = 'new')+
-  theme(
-    plot.background = element_rect(fill = "transparent", color = NA),
-    axis.text=element_text(size=14)
-  )
-
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-
-n <- 10 #number of observations
-sd <- 10 #approximate sd from paper
-set.seed(1) #set seed for reproducibility
-df <- 6
-dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group)) +
-    geom_point(show.legend=FALSE,size=1.5,alpha=0.5)+
-    stat_summary(aes(y = StO2_sim,
-                     group=Group), 
-                 fun=mean, geom="line",size=1,show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],'(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-  scale_color_aaas()+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
-
-# tumor_sim<-f1+f2+plot_layout(guides = 'collect')+
-#   plot_annotation(tag_levels='A')&
-#   ylim(c(0,75)) &
-#   theme(
-#      text=element_text(size=18),
-#      legend.title=element_blank(),
-#      legend.text = element_blank()
-#      )&
-#   scale_color_aaas()
-
-```
-
-```{r,smooth-plot,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-#GAM for StO2
-
-gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
-            method='REML',
-            data  = dat_sim)
-
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-#creates a dataframe using the length of the covariates for the GAM
-gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
-                         subject=factor(rep(1:10)))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#adds the predictions to the grid and creates a confidence interval for GAM
-gam_predict<-gam_predict%>%
-    mutate(fit = predict(gam1,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(gam1, gam_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lme
-#lm_predict<-lm_predict%>%
-#   mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response',level=0),
- #         se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response',level=0))
-
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=gam_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'complete')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-#posthoc comparisons for the linear model
-library(multcomp)
-
-#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
-
-```
-
-Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
-Using the simulated $StO_2$ values from Figure (\@ref(fig:sim-smooth-plot), B), 40% of the total observations can be randomly deleted and the same GAM model with interaction used for the full simulated data can be fitted to the remaining observations.
-
-```{r,missing-data,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
-
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'missing')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-```
-
-(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent smooths for each group, shaded regions are 95% confidence intervals.C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40% of its observations missing.
-Lines represent smooths for each group, shaded regions are 95% confidence intervals
-
-```{r,sim-smooth-plot,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}
-
-mult_plot<-f2+inset_element(
-  f1, left = 0.01, 
-  bottom = 0.5, 
-  right = 0.5, 
-  top = 1.0)+
-  f3+f4+f6+
-   plot_annotation(tag_levels='A')&
-   ylim(c(-5,75)) &
-  theme(
-     text=element_text(size=18)
-     )&
-  scale_color_aaas()
-
-mult_plot
-
-```
-
-```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-##Pairwise comparisons
-
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-#this function takes the model, grid and groups to be compared using the lpmatrix
-
-smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    #Find columns in xp where the name contains "Control"
-    col1 <- grepl(g1, colnames(xp))
-    #Find columns in xp where the name contains 'Treatment'
-    col2 <- grepl(g2, colnames(xp))
-    #r1 <- newdata[[var]] == f1
-    #r2 <- newdata[[var]] == f2
-    row1 <- with(newdata, Group == g1)
-    row2 <- with(newdata, Group == g2)
-    ## difference rows of xp for data from comparison
-    X <- xp[row1, ] - xp[row2, ]
-    ## zero out cols of X related to splines for other lochs
-    X[, ! (col1 | col2)] <- 0
-    ## zero out the parametric cols
-    X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(g1, g2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-}
-
-comp1<-smooth_diff(gam1,pdat,'Control','Treatment')
-
-comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
-
-c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill='black') +
-    geom_line(color='black',size=1) +
-    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
-    geom_ribbon(data=comp_StO2_full%>%
-                    filter(lower>0),
-                aes(ymin =0, ymax =lower),
-                alpha = 0.5,
-                fill='blue') +
-    geom_ribbon(data=comp_StO2_full %>%
-                    filter(upper<0),
-                    aes(ymin =0, ymax =upper),
-                alpha = 0.5,
-                fill='red') +
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-###for missing data
-comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
-comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp2))
-
-missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    labs(x = 'Days', 
-         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
-                              )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme_classic()+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
-c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill='black') +
-    geom_line(color='black',size=1) +
-    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
-    geom_ribbon(data=comp_StO2_missing%>%
-                    filter(lower>0),
-                aes(ymin =0, ymax =lower),
-                alpha = 0.5,
-                fill='blue') +
-    geom_ribbon(data=comp_StO2_missing %>%
-                    filter(upper<0),
-                    aes(ymin =0, ymax =upper),
-                alpha = 0.5,
-                fill='red') +
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-pair_comp<-c1+c2
-```
-
-(ref:pairwise-comp) Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-Significant differences exist where the interval does not cover 0.
-B: Pairwise comparisons for the full dataset.
-Significant differences exist where the interval does not cover 0.
-
-```{r,plot-pairwise-comp,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-comp)'}
-
-pair_comp+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-
-```
-
-## Determination of significance in GAMs
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that time point(s).
-The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
-The major advantage of the significance estimation in GAMs using the confidence interval approach is that they allow to determine significance at a specific time point, rather than providing a single _p-value_ for the overall treatment effect.
-
-Consider the simulated data used in the previous section.
-Figure (\@ref:sim-smooth-plot), B and D show the estimated smooths for the interaction model using the full dataset and with missing observations.
-If the confidence intervals are compared, it can be seen over time which Group has greater effect as therapy progresses.
-Specifically, Figure (\@ref:plot-pairwise-comp), shows the comparison between each treatment group.
-Notice that the shaded region between days 0 and 4 indicates that through that time, the "Control" group has higher $StO_2$, but as therapy progresses the effect is reversed and by day 5 it is the "Treatment" group the one that has greater contribution.
-
-***
-
-\newpage
-# References
-
-<div id="refs"></div>
-
 # (APPENDIX) Appendix {-}
@@
 This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation).
-The linear simulations using Figure \@ref(fig:linear-cases) show in  panels A and D the simulated mean responses and individual datapoints.
+The linear simulations using Figure \@ref(fig:linear-cases-appendix) show in  panels A and D the simulated mean responses and individual datapoints.
 Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
@@
 
-```{r,include=TRUE,message=FALSE,echo=TRUE}
+```{r-responses-appendix,include=TRUE,message=FALSE,echo=TRUE}
 ## Example with linear response
@@
     ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
+    geom_line(aes(size = "Subjects"),
+              show.legend = FALSE) +
     # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
+    geom_line(aes(
+      x = time, 
+      y = mu, 
+      color = treatment, 
+      size = "Simulated Truth"
+      ), 
+      lty = 1,
+      show.legend = FALSE
+      ) +
+    labs(y='response')+
+    scale_size_manual(
+      name = "Type", 
+      values=c("Subjects" = 0.5, "Simulated Truth" = 3)
+      ) +
     #ggtitle("Simulated data\nIndividual responses with population mean") +
@@
    p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='errors')+
+    ggplot(aes(x = time, 
+               y = errors, 
+               group = subject, 
+               color = treatment)) +
+    geom_line(show.legend=FALSE) +
+     labs(y='errors')+
      theme_classic()+
@@
 
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point()+labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
+  p4 <- ggplot(sim_dat$dat, 
+               aes(x = time,
+                   y = y, 
+                   color = treatment)) +
+    geom_point()+
+    labs(y='response')+
+    geom_line(aes(
+      y = predict(sim_dat$fit_lme), 
+      group = subject, 
+      size = "Subjects")
+      ) +
+    geom_line(data = sim_dat$pred_dat, 
+              aes(y = predict(sim_dat$fit_lme, 
+                              level = 0, 
+                              newdata = sim_dat$pred_dat),
+                  size = "Population")) +
+    scale_size_manual(name = "Predictions", 
+                      values=c("Subjects" = 0.5, 
+                               "Population" = 3)) +
     theme_classic() +
@@
 
-```{r, linear-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
+```{r, linear-cases-appendix, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+A, C: Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
 B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
 Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data**"}
+The rm-ANOVA model does not pick the trend of the quadratic data."}
 # linear response, correlated errors (subject effect)
 A1/B1+plot_annotation(tag_levels = 'A')
+
 ```
 
-For the quadratic response case, Figure \@ref(fig:quadratic-cases) shows the simulated responses using compound symmetry and independent errors.
+For the quadratic response case, Figure \@ref(fig:quadratic-cases-appendix) shows the simulated responses using compound symmetry and independent errors.
 
-```{r, quadratic-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
+```{r, quadratic-cases-appendix, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
 A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
@@
 D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data."}
 # linear response, correlated errors (subject effect)
 C1/D1+plot_annotation(tag_levels = 'A')
+
 ```
+
+## Basis functions and GAMs
+
+```{r,basis-plot-Appendix, echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
+#generate the response
+n_time = 6
+ x <- seq(1,6, length.out = n_time)
+ mu <- matrix(0, length(x), 2)
+ mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
+ mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
+ y <- array(0, dim = c(length(x), 2, 10))
+ errors <- array(0, dim = c(length(x), 2, 10))
+ for (i in 1:2) {     # number of treatments
+     for (j in 1:10) {  # number of subjects
+         # compound symmetry errors
+         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
+         y[, i, j] <- mu[, i] + errors[, i, j]
+     }
+ }
+  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
+ dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
+ dimnames(mu) <- list(time = x, treatment = 1:2)
+ dat <- as.data.frame.table(y, responseName = "y")
+ dat_errors <- as.data.frame.table(errors, responseName = "errors")
+ dat_mu <- as.data.frame.table(mu, responseName = "mu")
+ dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
+ dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
+ dat$time <- as.numeric(as.character(dat$time))
+
+ #add subject column
+ dat <- dat %>%
+     mutate(subject = factor(paste(subject, treatment, sep = "-")))
+
+  dat<-subset(dat,treatment==1)
+  dat<-dat[,c('y','time')]
+gm<-gam(y~s(time,k=5),data=dat)
+model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
+
+x_new <- (seq(0, max(dat$time), length.out = 100))
+y_pred <- predict(gm, newdata=dat)
+
+time<-c(1:6)
+
+basis<-model_matrix[1:6,] #extracting basis
+#basis<-model_matrix[1:6,-1] #extracting basis
+colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
+basis<-basis %>% #pivoting to long format
+  pivot_longer(
+    cols=starts_with("s")
+  )%>%
+  arrange(name) #ordering
+
+#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
+ln<-6*(length(coef(gm)))
+
+basis_plot<-data.frame(Basis=integer(ln),
+                       value_orig=double(ln),
+                       time=integer(ln),
+                       cof=double(ln)
+)
+
+basis_plot$time<-rep(time) #pasting timepoints
+basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
+basis_plot$value_orig<-basis$value #pasting basis values
+basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
+basis_plot<-basis_plot%>%
+  mutate(mod_val=value_orig*cof)
+
+#creating labeller to change the labels in the basis plots
+
+basis_names<-c(
+  `1`="Intercept",
+  `2`="1",
+  `3`="2",
+  `4`="3",
+  `5`="4"
+)
+
+#calculating the final spline by aggregating the basis functions
+
+spl<-basis_plot%>% 
+  group_by(time)%>%
+  summarize(spl=sum(mod_val))
+
+#original basis
+sz<-1
+p11<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=as.factor(Basis)))+
+  geom_line(size=sz,show.legend=FALSE)+
+  geom_point(size=sz+1,show.legend = FALSE)+
+  labs(y='Basis functions')+
+  facet_wrap(~Basis,labeller = as_labeller(basis_names))+
+  theme_classic()+
+  scale_color_aaas()
+
+#penalized basis
+p12<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
+  geom_line(show.legend = FALSE,size=sz)+
+  geom_point(show.legend = FALSE,size=sz+1)+
+  labs(y='Penalized \n basis functions')+
+  scale_y_continuous(breaks=seq(-1,1,1))+
+  facet_wrap(~Basis,labeller=as_labeller(basis_names))+
+  theme_classic()+
+  scale_color_aaas()
+
+#heatmap of the penalization coefficient
+x_labels<-c("Intercept","1","2","3","4")
+p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
+  geom_tile(aes(color='black'),size=sz+1,show.legend = FALSE)+
+  geom_tile(size=sz+1)+
+  scale_fill_gradient(low = "white", high = "red")+
+  labs(x='Basis',y='Basis')+
+  scale_x_discrete(labels=x_labels)+
+  geom_text(aes(label=round(cof,2)),size=10,show.legend = FALSE)+
+  theme_classic()+
+  theme(legend.title = element_blank())
+
+#plotting simulated datapoints and smooth term
+p14<-ggplot(data=dat,aes(x=time,y=y))+
+  geom_point(size=sz+1)+
+  scale_color_aaas()+
+  labs(y='Simulated \n response')+
+  geom_line(data=spl,aes(x=time,y=spl),color="#B15731",size=sz+1)+
+  theme_classic()
+
+#Combining all
+b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
+  theme(
+     text=element_text(size=18)
+     )
+
+b_plot
+```
+
+## Data simulation and GAM models
+
+```{r,sim-tumor-response-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
+
+dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
+            Day=rep(c(0,2,5,7,10),times=2),
+            Group=as.factor(rep(c("Control","Treatment"),each=5))
+            )
+
+## plot the mean response
+f1<-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
+    geom_line(size=1,show.legend = FALSE)+
+    geom_point(show.legend = FALSE,size=1.5,alpha=0.5)+
+  labs(y=expression(paste(StO[2],' (real)')))+
+  theme_classic()+
+  scale_color_aaas()+
+    scale_x_continuous(breaks=c(0,5,10))+
+    scale_y_continuous(breaks=c(0,40))+
+  plot_layout(tag_level = 'new')+
+  theme(
+    plot.background = element_rect(fill = "transparent", color = NA),
+    axis.text=element_text(size=14)
+  )
+
+#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
+#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
+
+simulate_data <- function(dat, n = 10, sd = 5) {
+    dat_sim <- dat %>%
+        slice(rep(1:n(), each = n)) %>%
+        group_by(Group, Day) %>%
+        mutate(
+               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
+               subject=rep(1:10),
+               subject=factor(paste(subject, Group, sep = "-"))
+               ) %>%
+        ungroup()
+
+    return(dat_sim)
+}
+
+#subject = factor(paste(subject, treatment, sep = "-")))
+
+n <- 10 #number of observations
+sd <- 10 #approximate sd from paper
+set.seed(1) #set seed for reproducibility
+df <- 6
+dat_sim <- simulate_data(dat, n, sd)
+
+#plotting simulated data
+f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group)) +
+    geom_point(show.legend=FALSE,size=1.5,alpha=0.5)+
+    stat_summary(aes(y = StO2_sim,
+                     group=Group), 
+                 fun=mean, geom="line",size=1,show.legend = FALSE)+
+  labs(y=expression(atop(StO[2],'(simulated)')))+
+  theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
+  scale_color_aaas()+
+    scale_x_continuous(breaks=c(0,2,5,7,10))
+
+```
+
+```{r,smooth-plot-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
+
+#GAM for StO2
+
+gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
+            method='REML',
+            data  = dat_sim)
+
+#linear model
+lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
+
+#creates a dataframe using the length of the covariates for the GAM
+gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = seq(0, 10, by = 0.1),
+                         subject=factor(rep(1:10)))
+
+#creates a dataframe using the length of the covariates for rm-ANOVA
+lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = c(0:10),
+                        subject=factor(rep(1:10)),
+                          )
+lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
+
+#adds the predictions to the grid and creates a confidence interval for GAM
+gam_predict<-gam_predict%>%
+    mutate(fit = predict(gam1,gam_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(gam1, gam_predict,se.fit = TRUE,type='response')$se.fit)
+
+#using lm
+lm_predict<-lm_predict%>%
+    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
+
+#plot smooths and confidence interval for GAM
+f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
+    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),
+                   fill=Group
+                   ),
+              alpha=0.3,
+              data=gam_predict,
+            show.legend=FALSE,
+                inherit.aes=FALSE) +
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=gam_predict,
+              show.legend = FALSE)+
+  #facet_wrap(~Group)+
+  labs(y=expression(atop(StO[2],'complete')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
+      scale_color_aaas()+
+  scale_fill_aaas()
+
+#plot linear fit for rm-ANOVA
+f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
+    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),fill=Group),
+              alpha=0.3,
+              data=lm_predict,
+              show.legend = FALSE,
+                inherit.aes=FALSE) +
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=lm_predict,
+              show.legend = FALSE)+
+  #facet_wrap(~Group)+
+  labs(y=expression(paste('StO'[2],' (simulated)')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
+      scale_color_aaas()+
+  scale_fill_aaas()
+
+#posthoc comparisons for the linear model
+library(multcomp)
+
+#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
+#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
+
+```
+
+```{r,missing-data-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
+#missing data
+#create a sequence of 40 random numbers between 1 and 100, these numbers will
+#correspond to the row numbers to be randomly erased from the original dataset
+missing <- sample(1:100, 40)
+
+#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
+
+ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
+
+#create a new dataframe, remove the StO2 column
+dat_missing <- dat_sim[,-1]
+
+#add NAs at the ind positions
+dat_missing$StO2_sim[ind]<-NA
+
+#Count the number of remaining observations per day (original dataset had 10 per group per day)
+dat_missing %>%
+    group_by(Day,Group) %>%
+    filter(!is.na(StO2_sim))%>%
+  count(Day)
+
+#the same model used for the full dataset
+mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
+#appraise the model
+appraise(mod_m1)
+
+m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = seq(0, 10, by = 0.1))
+
+#adds the predictions to the grid and creates a confidence interval
+m_predict<-m_predict%>%
+    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
+
+f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
+    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),
+                   fill=Group
+                   ),
+              alpha=0.3,
+              data=m_predict,
+            show.legend=FALSE,
+                inherit.aes=FALSE) +
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=m_predict,
+              show.legend = TRUE)+
+  #facet_wrap(~Group)+
+  labs(y=expression(atop(StO[2],'missing')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
+      scale_color_aaas()+
+  scale_fill_aaas()
+
+```
+
+(ref:sim-plot-caption-Appendix) Simulated data and smooths for oxygen saturation in tumors.
+A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point.
+Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
+Lines represent trends for each group, shaded regions are 95% confidence intervals.
+C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
+D: Smooths for the GAM model for the simulated data with 40% of its observations missing.
+Lines represent trends for each group, shaded regions are 95% confidence intervals.
+
+```{r,sim-smooth-plot-Appendix,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=TRUE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption-Appendix)'}
+
+mult_plot<-f2+inset_element(
+  f1, left = 0.01, 
+  bottom = 0.5, 
+  right = 0.5, 
+  top = 1.0)+
+  f3+f4+f6+
+   plot_annotation(tag_levels='A')&
+   ylim(c(-5,75)) &
+  theme(
+     text=element_text(size=18)
+     )&
+  scale_color_aaas()
+
+mult_plot
+
+```

### 2021-05-13 Manuscripts/Bookdown testing/Full_document.Rmd
Splitted the document in sections, author section now works. Need to work on Appendix, as there is some issue going

--- 2021-05-11 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
@@
 subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
-author:
-- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
-- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
-  Fayetteville]
-- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
+header-includes:
+    \usepackage{placeins}
 output:  
@@
       float: null
+    includes:
+      in_header: preamble.sty
   bookdown::word_document2:
@@
 
-In biomedical research, the outcome of longitudinal studies is frequently analyzed statistically using the repeated measures analysis or variance (rm-ANOVA).
-However, rm-ANOVA is restrictive as is only able to accommodate linear responses, requires equal correlation between each measurement (compound symmetry), and cannot work with missing observations.
-When the data fails to fullfill all the required assumptions by rm-ANOVA, the model provides biased estimates and inference.
-In contrast, generalized additive models (GAMs) are a class of models that overcome all of the limitations imposed by rm-ANOVA and that can be implemented in a straightforward manner using the _mgcv_ package in R.
-This paper summarizes the assumptions of rm-ANOVA and how their limit its use in longitudinal biomedical research, presents the basic theoretical framework for GAMs and provides examples of their implementation in R.
-The code and data used to generate this paper are available at:
+In biomedical research, the outcome of a longitudinal study has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, a _linear mixed model_ (LMEM).
+Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
+In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+Although not commonly used in biomedical research, GAMs present an excellent choice to analyze non-linear longitudinal data.
+This paper summarizes the limitations of linear models (rm-ANOVA in particular), presents the basic theory of GAMs, and demonstrates their implementation in $\textsf{R}$ via the package _mgcv_ using simulated data that follows longitudinal trends reported in biomedical literature.
+To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:________.
 
-# Background
+```{r libraries, message = FALSE,include=FALSE}
+#the packages needed to run the code
 
-Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following RT in neck cancer patients [@kamstra2015].
-
-Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as a limiting frequency [@wagenmakers2008] by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA [@pinheiro2006].
-This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model is only able to accommodate linear trends that are far from adequately representing the biological phenomenon of interest.
-
-A _post hoc_ analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
-Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, _p-values_ are valid only if the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019].
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting _p-values_ and parameter estimates to be invalid [@wang2019].
-
-Additionally, the _p-value_ itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019].
-Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to balance in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].
-Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976;@greenhouse1959], but the effectiveness of the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001].
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not "large" due to ethical and budgetary reasons [@charan2013] which might cause the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial patients may voluntarily withdraw, whereas attrition  due to injury or weight loss in preclinical animal studies is possible.
-It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios,the _complete observations_ assumption of  classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis [@gueorguieva2004].
-This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
-At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
-
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data [@skala2010;@vishwanath2009].
-Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-However, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
-And even more importantly, LMEMs also expect a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze non-linear data.
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
-
-In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017] and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
-
-The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, $\textsf{R}$[@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
-hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017].
-Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
-
-This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations is explained in detail.
-Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, using simulated data that reproduces patterns in previously reported studies [@vishwanath2009] we illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
-The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in $\textsf{R}$, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
-
-In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models by improving the standards for reproducibility  in biomedical research.
-
-# Challenges presented by longitudinal studies
-
-## The repeated measures ANOVA
-
-The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research.
-This statistical methodology requires certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
-Each one of these assumptions is discussed below.
-
-## Linear relationship
-
-### The repeated measures ANOVA case
-
-In a biomedical longitudinal study, two or more groups of subjects (e.g., patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs.
-a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response is modeled with _fixed_ components.
-The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug in the "treatment" group.
-
-Mathematically speaking, a rm-ANOVA model with an interaction can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
-(\#eq:linear-model)
-\end{equation}
-
-In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$.
-In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
-"novel drug" or "saline" vs.
-"chemotherapy").
-Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B).
-The linear models then can be expressed as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-(\#eq:ANOVA-by-group)
-\end{equation}
-
-To further simplify the expression, substitute $\widetilde{\beta_{0}}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B.
-This substitution allows for a different intercept and slope for Groups A and B.
-The model is then written as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-(\#eq:ANOVA-lines)
-\end{equation}
-
-Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
-
-### The Linear Mixed Model Case
-
-A linear mixed model (LMEM) is a class of statistical model that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to distinguish [@pinheiro2006;@west2014].
-A LMEM with interaction between time and treatment for a longitudinal study  can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
-(\#eq:LMEM)
-\end{equation}
-
-When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
-This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject within each group.
-The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the "global noise" term $\varepsilon_{ijt}$ from Equation \@ref(eq:linear-model).
-
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
-As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
-The _random effect_ $\mu_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be independently normally distributed with mean zero and variance $\sigma^2_\mu$, which can be expressed as $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
-In essence,the _random effect_ in a LMEM enables to fit models with different slopes at the subject-level[@pinheiro2006].
-However,the expected linear relationship of the covariates and the response in Equation \@ref(eq:linear-model) and in Equation \@ref(eq:LMEM) is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
-
-## Covariance in rm-ANOVA and LMEMs
-
-In a longitudinal study there is an expected _covariance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
-The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
-
-In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
-Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)[@huynh1976;@greenhouse1959] the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others [@pinheiro2006].
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself.
-Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
-
-## Missing observations
-
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004].
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002].
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
-
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014].
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.
-
-This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and  missing data.
-In particular, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data.
-To better convey the issues of linearity and correlation  in linear models fitted to non-linear data, simulation is used in the next section.
-
-## How does an rm-ANOVA fit looks like? A visual representation using simulation {#simulation}
-
-To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme].
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C).
-In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity.
-
-Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates an effect due to treatment.
-In other words, Group 1 can be thought as a "Control" group and Group 2 as a "Treatment" group.
-From the mean response per group (linear or quadratic), the variability or "error" of individual responses  within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \@ref(fig:l-q-response) B,D).
-A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \@ref(fig:linear-cases) and Figure \@ref(fig:quadratic-cases) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-
-```{r, message = FALSE,include=FALSE}
 library(patchwork)
@@
 
-knitr::opts_chunk$set(fig.pos = "!h", out.extra = "")
+knitr::opts_chunk$set(fig.pos = "!h", out.extra = "",error=TRUE)
 ```
 
-```{r,include=FALSE,message=FALSE,echo=FALSE}
-## Example with linear response
-example <- function(n_time = 6, 
-                    fun_type = "linear", 
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  x <- seq(1,6, length.out = n_time)
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # nonlinear response
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-  # matplot(mu, type = 'l')
-
-  y <- array(0, dim = c(length(x), 2, 10))
-  errors <- array(0, dim = c(length(x), 2, 10))
-
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(mu) <- list(time = x, treatment = 1:2)
-  dat <- as.data.frame.table(y, responseName = "y")
-  dat_errors <- as.data.frame.table(errors, responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, responseName = "mu")
-  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
-  dat$time <- as.numeric(as.character(dat$time))
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  ## repeated measures ANOVA in R
-  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
-  dat$preds_lm <- predict(fit_lm)
-
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  dat$y_pred <- predict(fit_lme)
-
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_lm = fit_lm,
-    fit_lme = fit_lme
-
-  ))
-}
-
-plot_example <- function(sim_dat) {
-  library(patchwork)
-  ## Plot the simulated data
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE) +labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
-    theme_classic() +
-    #ggtitle("Simulated data with true response function")+
-    theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    #ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_classic()+
-     theme(plot.title = element_text(size = 18, 
-                                face = "bold"),
-     text=element_text(size=18))+
-    scale_color_aaas()
-
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='errors')+
-     theme_classic()+
-    # facet_wrap(~ treatment) +
-    #ggtitle("Simulated errors") +
-     theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point()+labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    #ggtitle("Fitted Model")+
-    theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  #return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-  return((p1+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-A<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
-
+```{r child='01-Background.Rmd'}
 ```
 
-(ref:l-q-response-caption) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.
+\FloatBarrier
 
-```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='ref(l-q-response-caption)' }
-# linear response, correlated errors (subject effect)
-par(mar = c(0.5, 0.5, 0.5, 0.5))
-A/C+plot_annotation(tag_levels = 'A')
+```{r child='02-Challenges.Rmd'}
 ```
 
-The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B).
-When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.
+\FloatBarrier
 
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response), C.
-Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \@ref(fig:l-q-response) A and B) .
-The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panel D in Figure \@ref(fig:l-q-response).
-
-A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\@ref(fig:l-q-response), D) indicates that the model is not capturing the changes within each group in a good way.
-Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
-Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D).
-If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response.
-Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
-
-This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-Although the model fitted to the simulated data was an rm-ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM.
-In the following section, we present generalized additive models (GAMs) as a data-driven  alternative method to analyze longitudinal non-linear data.
-
-# GAMs as a special case of Generalized Linear Models
-
-## GAMs and Basis Functions
-
-Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors[@nelder1972].
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case[@simpson2018;@wood2017;@hastie1987].
-A GAM model can be written as:
-
-\begin{equation}
-  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
-  (\#eq:GAM)
-\end{equation}
-
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual error.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
-In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.
-
-The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using _basis functions_ expansions of the covariates and by estimating random coefficients for these basis functions.
-A _basis_ is a set of functions that spans the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
-For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
-In the case of Equation \@ref(eq:GAM), the basis function is $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
-
-A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness [@wood2017;@simpson2018].
-Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\@ref(fig:l-q-response), C).
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-The timeline can be divided in equally spaced _knots_, each knot being a region where a different basis function will be used.
-Because there are six timepoints for this group, five knots can be used.
-The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots.
-In Panel A of Figure \@ref(fig:basis-plot), the four basis functions (and the intercept) are shown.
-Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the "wigliness" of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
-The penalty reduces the "wigliness" of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
-
-```{r,basis-functions-plot, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  dat<-subset(dat,treatment==1)
-  dat<-dat[,c('y','time')]
-gm<-gam(y~s(time,k=5),data=dat)
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-x_new <- (seq(0, max(dat$time), length.out = 100))
-y_pred <- predict(gm, newdata=dat)
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof)
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final spline by aggregating the basis functions
-
-spl<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(spl=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=as.factor(Basis)))+
-  geom_line(size=sz,show.legend=FALSE)+
-  geom_point(size=sz+1,show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,labeller = as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#penalized basis
-p12<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
-  geom_line(show.legend = FALSE,size=sz)+
-  geom_point(show.legend = FALSE,size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,labeller=as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#heatmap of the penalization coefficient
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
-  geom_tile(aes(color='black'),size=sz+1,show.legend = FALSE)+
-  geom_tile(size=sz+1)+
-  scale_fill_gradient(low = "white", high = "red")+
-  labs(x='Basis',y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),size=10,show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  scale_color_aaas()+
-  labs(y='Simulated \n response')+
-  geom_line(data=spl,aes(x=time,y=spl),color="#B15731",size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-
+```{r child='03-GAM Theory.Rmd'}
 ```
 
-In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \@ref(fig:basis-plot), Panel C shows the resulting penalized basis functions.
-Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
-Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
+\FloatBarrier
 
-(ref:basis-plot-caption) Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.
-
-```{r,basis-plot,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap='(ref:basis-plot-caption)'}
-
-par(mar = c(2, 2, 2, 2))
-
-b_plot
-
+```{r child='04-Longitudinal analysis with GAMs.Rmd'}
 ```
 
- \newpage
+\FloatBarrier
 
-# The analyisis of longitudinal biomedical data using GAMs
-
-The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA.
-This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using $\textsf{R}$ and the package `mgcv`.
-The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
-
-## Simulated data
-
-The simulated data is based on the reported longitudinal changes in oxygen saturation ($StO_2$) in subcutaneous tumors that appear in Figure 3(C) in [vishwanath2009].
-An increase in oxygen saturation in the "Treatment" group (chemotherapy) is observed through time, while a decrease is seen in the "Control" (saline) group.
-Following the reported trend, 10 normally distributed observations at each time point are generated with a standard deviation (SD) of 10% (matching the SD in the original paper).
-The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot), A and the inlet, respectively.
-
-## An interaction GAM model for longitudinal data
-
-An interaction model is typically the main interest in longitudinal biomedical data, as it takes into account the effect of treatment, time, and their combination.
-In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the _time_ effect to account for the repeated measures over time.
-Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in [@hefley2017].
-
-For the data in Figure \@ref(fig:sim-smooth-plot),A it is of interest to know the effect that each treatment has in oxygen saturation.
-The model then needs to incorporate independent smooths for _Group_ and _Day_, respectively.
-The main thing to consider is the model syntax to account for the fact that one of the variables is numeric ( _Day_ ) and that the other is a factor ( _Group_ ).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
-Using $\textsf{R}$ and the package `mgcv` the model syntax is:
-
-`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim`
-
-This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation `StO2_sim` is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
-The parametric term _Group_ is added to separate the effect of treatment, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $StO_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B).
-Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package `gratia`[@gratia].
-A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].
-
-One question might be "what is the fit that an rm-ANOVA model produces for the simulated data?".
-The rm-ANOVA model, which corresponds to \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot),C.
-This is a typical case of model misspecification: The slopes of each group ar different, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \@ref(fig:sim-smooth-plot),B) .
-
-```{r,sim-tumor-response,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
-
-## plot the mean response
-f1<-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
-    geom_line(size=1,show.legend = FALSE)+
-    geom_point(show.legend = FALSE)+
-  labs(y=expression(paste(StO[2],' (real)')))+
-  theme_classic()+
-  scale_color_aaas()+
-    scale_x_continuous(breaks=c(0,5,10))+
-    scale_y_continuous(breaks=c(0,40))+
-  plot_layout(tag_level = 'new')+
-  theme(
-    plot.background = element_rect(fill = "transparent", color = NA),
-    axis.text=element_text(size=14)
-  )
-
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-
-n <- 10 #number of observations
-sd <- 10 #approximate sd from paper
-set.seed(1) #set seed for reproducibility
-df <- 6
-dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group)) +
-    geom_point(show.legend=FALSE)+
-    stat_summary(aes(y = StO2_sim,
-                     group=Group), 
-                 fun=mean, geom="line",size=1,show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],'(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-  scale_color_aaas()+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
-
-# tumor_sim<-f1+f2+plot_layout(guides = 'collect')+
-#   plot_annotation(tag_levels='A')&
-#   ylim(c(0,75)) &
-#   theme(
-#      text=element_text(size=18),
-#      legend.title=element_blank(),
-#      legend.text = element_blank()
-#      )&
-#   scale_color_aaas()
-
+```{r child='05-Conclusion.Rmd'}
 ```
 
-```{r,smooth-plot,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-#GAM for StO2
-
-gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
-            method='REML',
-            data  = dat_sim)
-
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-#creates a dataframe using the length of the covariates for the GAM
-gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
-                         subject=factor(rep(1:10)))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#adds the predictions to the grid and creates a confidence interval for GAM
-gam_predict<-gam_predict%>%
-    mutate(fit = predict(gam1,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(gam1, gam_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lme
-#lm_predict<-lm_predict%>%
-#   mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response',level=0),
- #         se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response',level=0))
-
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=gam_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'complete')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-#posthoc comparisons for the linear model
-library(multcomp)
-
-summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
-
-```
-
-Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
-Using the simulated $StO_2$ values from Figure (\@ref(fig:sim-smooth-plot), B), 40% of the total observations can be randomly deleted and the same GAM model with interaction used for the full simulated data can be fitted to the remaining observations.
-
-```{r,missing-data,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
-
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'missing')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-```
-
-(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent smooths for each group, shaded regions are 95% confidence intervals.C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40% of its observations missing.
-Lines represent smooths for each group, shaded regions are 95% confidence intervals
-
-```{r,sim-smooth-plot,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}
-
-mult_plot<-f2+inset_element(
-  f1, left = 0.01, 
-  bottom = 0.5, 
-  right = 0.5, 
-  top = 1.0)+
-  f3+f4+f6+
-  #plot_layout(guides = 'collect')+
-   plot_annotation(tag_levels='A')&
-   ylim(c(-5,75)) &
-  theme(
-     text=element_text(size=18)
-     )&
-  scale_color_aaas()
-
-mult_plot
-
-```
-
-## Determination of significance in GAMs
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that timepoint (or timepoints).
-The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
-The major advantage of the significance estimation in GAMs  is that they allow to determine significance at a specific timepoint, rather than providing a single _p-value_ for the overall treatment effect.
-The estimation of siginificant differences in GAMs is covered in detail in the Appendix using simulated longitudinal data that follows previously reported trends of tumor response between multiple treatment groups [@vishwanath2009].
-
-```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-##Pairwise comparisons
-
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Treatment', 'Control'))
-
-smooth_diff <- function(model, newdata, f1, f2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    c1 <- grepl(f1, colnames(xp))
-    c2 <- grepl(f2, colnames(xp))
-    #r1 <- newdata[[var]] == f1
-    #r2 <- newdata[[var]] == f2
-    r1 <- with(newdata, Group == f1)
-    r2 <- with(newdata, Group == f2)
-    ## difference rows of xp for data from comparison
-    X <- xp[r1, ] - xp[r2, ]
-    ## zero out cols of X related to splines for other lochs
-    X[, ! (c1 | c2)] <- 0
-    ## zero out the parametric cols
-    X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(f1, f2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-}
-
-comp1<-smooth_diff(gam1,pdat,'Control','Treatment')
-
-comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
-
-c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5,fill='#993336') +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-  theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-#+
- # geom_rect(aes(color=interval))
-
-c1
-
-###for missing data
-comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
-comp_StO2_m <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp2))
-
-missing_plot<-ggplot(comp_StO2_m, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    labs(x = 'Days', 
-         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
-                              )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme_classic()+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
-```
+\FloatBarrier
 
@@
 
-\newpage
-
-# (APPENDIX) Appendix {-}
-
-# Simulation
-
-## Compound symmetry and independent errors in linear and quadratic responses
-
-This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation).
-The linear simulations using Figure \@ref(fig:linear-cases) show in  panels A and D the simulated mean responses and individual datapoints.
-Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-```{r,include=TRUE,message=FALSE,echo=TRUE}
-## Example with linear response
-example <- function(n_time = 6, 
-                    fun_type = "linear", 
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  library(tidyverse)
-  library(mvnfast)
-  library(nlme)
-  library(ggsci)
-
-  x <- seq(1,6, length.out = n_time)
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # nonlinear response
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-  # matplot(mu, type = 'l')
-
-  y <- array(0, dim = c(length(x), 2, 10))
-  errors <- array(0, dim = c(length(x), 2, 10))
-
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(mu) <- list(time = x, treatment = 1:2)
-  dat <- as.data.frame.table(y, responseName = "y")
-  dat_errors <- as.data.frame.table(errors, responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, responseName = "mu")
-  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
-  dat$time <- as.numeric(as.character(dat$time))
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  ## repeated measures ANOVA in R
-  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
-  dat$preds_lm <- predict(fit_lm)
-
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  dat$y_pred <- predict(fit_lme)
-
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_lm = fit_lm,
-    fit_lme = fit_lme
-
-  ))
-}
-
-plot_example <- function(sim_dat) {
-  library(patchwork)
-  ## Plot the simulated data
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE) +labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
-    theme_classic() +
-    #ggtitle("Simulated data with true response function")+
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    #ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                face = "bold"),
-     text=element_text(size=30))+
-    scale_color_aaas()
-
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='errors')+
-     theme_classic()+
-    # facet_wrap(~ treatment) +
-    #ggtitle("Simulated errors") +
-     theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point()+labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    #ggtitle("Fitted Model")+
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
-
+```{r child='06-Appendix.Rmd'}
 ```
-
-```{r, linear-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data**"}
-# linear response, correlated errors (subject effect)
-A1/B1+plot_annotation(tag_levels = 'A')
-```
-
-For the quadratic response case, Figure \@ref(fig:quadratic-cases) shows the simulated responses using compound symmetry and independent errors.
-
-```{r, quadratic-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
-# linear response, correlated errors (subject effect)
-C1/D1+plot_annotation(tag_levels = 'A')
-```

### 2021-05-13 Manuscripts/Bookdown testing/Full_document.tex
Splitted the document in sections, author section now works. Need to work on Appendix, as there is some issue going

--- 2021-05-12 Manuscripts/Manuscript v5/Manuscript_AM_v5.tex
@@
   pdftitle={The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models},
-  pdfauthor={Ariel Mundo; Timothy J. Muldoon; John R. Tipton},
   hidelinks,
@@
 \setcounter{secnumdepth}{5}
+\usepackage{lineno}
+\usepackage{authblk}
+
+\linenumbers
+
+\author[1]{Ariel I. Mundo}
+\author[2]{John R. Tipton}
+\author[1]{Timothy J. Muldoon}
+\renewcommand\Authands {} %added because an additional "and" appeared after last author
+\affil[1]{\footnotesize Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
+\affil[2]{\footnotesize Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
+\usepackage{placeins}
 \usepackage{subfig}
@@
 \subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
-\author{Ariel Mundo\footnote{Department of Biomedical Engineering, University of Arkansas, Fayetteville} \and Timothy J.
-Muldoon\footnote{Department of Biomedical Engineering, University of Arkansas, Fayetteville} \and John R.
-Tipton\footnote{Department of Mathematical Sciences, University of Arkansas, Fayetteville}}
-\date{}
+\author{}
+\date{\vspace{-2.5em}}
 
@@
 Tumor response {[}1--4{]}, antibody expression {[}5,6{]}, and cell metabolism {[}7,8{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}9{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}10{]}, to monthly measurements to study a long-term response such as mouth opening following RT in neck cancer patients {[}11{]}.
+Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}9{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}10{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}11{]}.
 
 Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as a limiting frequency {[}12{]} by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
+The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}12{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
 The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}13,14{]}.
@@
 In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings {[}1,8,16{]}.
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}1,8,16{]}.
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
@@
 Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}21{]}), but they in turn reduce statistical power (1-\(\beta\)){[}22{]}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) {[}23,24{]}.
-Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to balance in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
+Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
 
@@
 It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios,the \emph{complete observations} assumption of classical rm-ANOVA is violated.
+In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
 When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}13{]}.
@@
 
+\FloatBarrier
+
 \hypertarget{challenges-presented-by-longitudinal-studies}{%
@@
 
-In a biomedical longitudinal study, two or more groups of subjects (e.g., patients, mice, samples) are subject to different treatments (e.g., a ``treatment'' group receives a novel drug vs.~a ``control'' group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
+In a longitudinal biomedical study, two or more groups of subjects (e.g., human subject, mice, samples) are subject to different treatments (e.g., a ``treatment'' group receives a novel drug or intervention vs.~a ``control'' group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
 The collected response is modeled with \emph{fixed} components.
-The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug in the ``treatment'' group.
+The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug/intervention in the ``treatment'' group.
 
@@
 In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
-Independent errors \(\varepsilon_{tij}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\).
+Independent errors \(\varepsilon_{tij}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently normally distributed with mean zero and variance \(\sigma^2_\mu\)).
 In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy'').
@@
 As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
-The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be independently normally distributed with mean zero and variance \(\sigma^2_\mu\), which can be expressed as \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
+The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
 In essence,the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}15{]}.
@@
 In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}15{]}.
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself.
+Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
 Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
@@
 \hypertarget{simulation}{%
-\subsection{How does an rm-ANOVA fit looks like? A visual representation using simulation}\label{simulation}}
+\subsection{What does an rm-ANOVA fit looks like? A visual representation using simulated data}\label{simulation}}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v5_files/figure-latex/l-q-response-1}
+{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/l-q-response-1}
 
@@
 In the following section, we present generalized additive models (GAMs) as a data-driven alternative method to analyze longitudinal non-linear data.
+
+\FloatBarrier
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v5_files/figure-latex/basis-plot-1}
+{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/basis-plot-1}
 
@@
 
+\FloatBarrier
+
 \newpage
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v5_files/figure-latex/sim-smooth-plot-1}
+{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Manuscript_AM_v5_files/figure-latex/plot-pairwise-comp-1}
+{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/plot-pairwise-comp-1}
 
@@
 
+\FloatBarrier
+
 \hypertarget{conclusion}{%
 \section{Conclusion}\label{conclusion}}
+
+\FloatBarrier
 
@@
 This section simulated linear and quadratic data in the same manner as in Section \ref{simulation}.
-The linear simulations using Figure \ref{fig:linear-cases} show in panels A and D the simulated mean responses and individual datapoints.
+The linear simulations using Figure \ref{fig:linear-cases-appendix} show in panels A and D the simulated mean responses and individual datapoints.
 Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## Example with linear response}
-\NormalTok{example <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{n_time =} \DecValTok{6}\NormalTok{, }
-                    \DataTypeTok{fun_type =} \StringTok{"linear"}\NormalTok{, }
-                    \DataTypeTok{error_type =} \StringTok{"correlated"}\NormalTok{) \{}
-
-  \ControlFlowTok{if}\NormalTok{ (}\OperatorTok{!}\NormalTok{(fun_type }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"linear"}\NormalTok{, }\StringTok{"quadratic"}\NormalTok{)))}
-    \KeywordTok{stop}\NormalTok{(}\StringTok{'fun_type must be either "linear", or "quadratic"'}\NormalTok{)}
-  \ControlFlowTok{if}\NormalTok{ (}\OperatorTok{!}\NormalTok{(error_type }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"correlated"}\NormalTok{, }\StringTok{"independent"}\NormalTok{)))}
-    \KeywordTok{stop}\NormalTok{(}\StringTok{'fun_type must be either "correlated", or "independent"'}\NormalTok{)}
-
-  \KeywordTok{library}\NormalTok{(tidyverse)}
-  \KeywordTok{library}\NormalTok{(mvnfast)}
-  \KeywordTok{library}\NormalTok{(nlme)}
-  \KeywordTok{library}\NormalTok{(ggsci)}
-
-\NormalTok{  x <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{6}\NormalTok{, }\DataTypeTok{length.out =}\NormalTok{ n_time)}
-\NormalTok{  mu <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{length}\NormalTok{(x), }\DecValTok{2}\NormalTok{)}
-  \CommentTok{# linear response}
-  \ControlFlowTok{if}\NormalTok{ (fun_type }\OperatorTok{==}\StringTok{ "linear"}\NormalTok{) \{}
-\NormalTok{    mu[, }\DecValTok{1}\NormalTok{] <-}\StringTok{ }\OperatorTok{-}\StringTok{ }\NormalTok{(}\FloatTok{0.25}\OperatorTok{*}\NormalTok{x)}\OperatorTok{+}\DecValTok{2}  
-\NormalTok{    mu[, }\DecValTok{2}\NormalTok{] <-}\StringTok{ }\FloatTok{0.25}\OperatorTok{*}\NormalTok{x}\OperatorTok{+}\DecValTok{2}
-\NormalTok{  \} }\ControlFlowTok{else}\NormalTok{ \{}
-    \CommentTok{# nonlinear response}
-
-\NormalTok{    mu[, }\DecValTok{1}\NormalTok{] <-}\StringTok{  }\OperatorTok{-}\NormalTok{(}\FloatTok{0.25} \OperatorTok{*}\StringTok{ }\NormalTok{x}\OperatorTok{^}\DecValTok{2}\NormalTok{) }\FloatTok{+1.5}\OperatorTok{*}\NormalTok{x}\FloatTok{-1.25}
-\NormalTok{    mu[, }\DecValTok{2}\NormalTok{] <-}\StringTok{ }\NormalTok{(}\FloatTok{0.25} \OperatorTok{*}\StringTok{ }\NormalTok{x}\OperatorTok{^}\DecValTok{2}\NormalTok{) }\FloatTok{-1.5}\OperatorTok{*}\NormalTok{x}\FloatTok{+1.25}
-\NormalTok{  \}}
-  \CommentTok{# matplot(mu, type = 'l')}
-
-\NormalTok{  y <-}\StringTok{ }\KeywordTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DataTypeTok{dim =} \KeywordTok{c}\NormalTok{(}\KeywordTok{length}\NormalTok{(x), }\DecValTok{2}\NormalTok{, }\DecValTok{10}\NormalTok{))}
-\NormalTok{  errors <-}\StringTok{ }\KeywordTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DataTypeTok{dim =} \KeywordTok{c}\NormalTok{(}\KeywordTok{length}\NormalTok{(x), }\DecValTok{2}\NormalTok{, }\DecValTok{10}\NormalTok{))}
-
-  \ControlFlowTok{if}\NormalTok{ (error_type }\OperatorTok{==}\StringTok{ "independent"}\NormalTok{) \{}
-    \CommentTok{## independent errors}
-    \ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{) \{}
-      \ControlFlowTok{for}\NormalTok{ (j }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{) \{}
-\NormalTok{        errors[, i, j] <-}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(}\DecValTok{6}\NormalTok{, }\DecValTok{0}\NormalTok{, }\FloatTok{0.25}\NormalTok{)}
-\NormalTok{        y[, i, j] <-}\StringTok{ }\NormalTok{mu[, i] }\OperatorTok{+}\StringTok{ }\NormalTok{errors[, i, j]}
-\NormalTok{      \}}
-\NormalTok{    \}}
-\NormalTok{  \} }\ControlFlowTok{else}\NormalTok{ \{}
-    \ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{) \{     }\CommentTok{# number of treatments}
-      \ControlFlowTok{for}\NormalTok{ (j }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{) \{  }\CommentTok{# number of subjects}
-        \CommentTok{# compound symmetry errors}
-\NormalTok{        errors[, i, j] <-}\StringTok{ }\KeywordTok{rmvn}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{length}\NormalTok{(x)), }\FloatTok{0.1} \OperatorTok{*}\StringTok{ }\KeywordTok{diag}\NormalTok{(}\DecValTok{6}\NormalTok{) }\OperatorTok{+}\StringTok{ }\FloatTok{0.25} \OperatorTok{*}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{6}\NormalTok{))}
-\NormalTok{        y[, i, j] <-}\StringTok{ }\NormalTok{mu[, i] }\OperatorTok{+}\StringTok{ }\NormalTok{errors[, i, j]}
-\NormalTok{      \}}
-\NormalTok{    \}}
-\NormalTok{  \}    }
-
-  \CommentTok{## subject random effects}
-
-  \CommentTok{## visualizing the difference between independent errors and compound symmetry}
-  \CommentTok{## why do we need to account for this -- overly confident inference}
-
-  \KeywordTok{dimnames}\NormalTok{(y) <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{time =}\NormalTok{ x, }\DataTypeTok{treatment =} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{, }\DataTypeTok{subject =} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{)}
-  \KeywordTok{dimnames}\NormalTok{(errors) <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{time =}\NormalTok{ x, }\DataTypeTok{treatment =} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{, }\DataTypeTok{subject =} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{)}
-  \KeywordTok{dimnames}\NormalTok{(mu) <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{time =}\NormalTok{ x, }\DataTypeTok{treatment =} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{)}
-\NormalTok{  dat <-}\StringTok{ }\KeywordTok{as.data.frame.table}\NormalTok{(y, }\DataTypeTok{responseName =} \StringTok{"y"}\NormalTok{)}
-\NormalTok{  dat_errors <-}\StringTok{ }\KeywordTok{as.data.frame.table}\NormalTok{(errors, }\DataTypeTok{responseName =} \StringTok{"errors"}\NormalTok{)}
-\NormalTok{  dat_mu <-}\StringTok{ }\KeywordTok{as.data.frame.table}\NormalTok{(mu, }\DataTypeTok{responseName =} \StringTok{"mu"}\NormalTok{)}
-\NormalTok{  dat <-}\StringTok{ }\KeywordTok{left_join}\NormalTok{(dat, dat_errors, }\DataTypeTok{by =} \KeywordTok{c}\NormalTok{(}\StringTok{"time"}\NormalTok{, }\StringTok{"treatment"}\NormalTok{, }\StringTok{"subject"}\NormalTok{))}
-\NormalTok{  dat <-}\StringTok{ }\KeywordTok{left_join}\NormalTok{(dat, dat_mu, }\DataTypeTok{by =} \KeywordTok{c}\NormalTok{(}\StringTok{"time"}\NormalTok{, }\StringTok{"treatment"}\NormalTok{))}
-\NormalTok{  dat}\OperatorTok{$}\NormalTok{time <-}\StringTok{ }\KeywordTok{as.numeric}\NormalTok{(}\KeywordTok{as.character}\NormalTok{(dat}\OperatorTok{$}\NormalTok{time))}
-\NormalTok{  dat <-}\StringTok{ }\NormalTok{dat }\OperatorTok{%>%}
-\StringTok{    }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{subject =} \KeywordTok{factor}\NormalTok{(}\KeywordTok{paste}\NormalTok{(subject, treatment, }\DataTypeTok{sep =} \StringTok{"-"}\NormalTok{)))}
-
-  \CommentTok{## repeated measures ANOVA in R}
-\NormalTok{  fit_lm <-}\StringTok{ }\KeywordTok{lm}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{time }\OperatorTok{+}\StringTok{ }\NormalTok{treatment }\OperatorTok{+}\StringTok{ }\NormalTok{time }\OperatorTok{*}\StringTok{ }\NormalTok{treatment, }\DataTypeTok{data =}\NormalTok{ dat)}
-\NormalTok{  dat}\OperatorTok{$}\NormalTok{preds_lm <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(fit_lm)}
-
-\NormalTok{  fit_lme <-}\StringTok{ }\KeywordTok{lme}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{treatment }\OperatorTok{+}\StringTok{ }\NormalTok{time }\OperatorTok{+}\StringTok{ }\NormalTok{treatment}\OperatorTok{:}\NormalTok{time,}
-                 \DataTypeTok{data =}\NormalTok{ dat,}
-                 \DataTypeTok{random =} \OperatorTok{~}\StringTok{ }\DecValTok{1} \OperatorTok{|}\StringTok{ }\NormalTok{subject,}
-                 \DataTypeTok{correlation =} \KeywordTok{corCompSymm}\NormalTok{(}\DataTypeTok{form =} \OperatorTok{~}\StringTok{ }\DecValTok{1} \OperatorTok{|}\StringTok{ }\NormalTok{subject)}
-\NormalTok{  )}
-
-\NormalTok{  pred_dat <-}\StringTok{ }\KeywordTok{expand.grid}\NormalTok{(}
-    \DataTypeTok{treatment =} \KeywordTok{factor}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{), }
-    \DataTypeTok{time =} \KeywordTok{unique}\NormalTok{(dat}\OperatorTok{$}\NormalTok{time)}
-\NormalTok{  )}
-
-\NormalTok{  dat}\OperatorTok{$}\NormalTok{y_pred <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(fit_lme)}
-
-  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}
-    \DataTypeTok{dat =}\NormalTok{ dat,}
-    \DataTypeTok{pred_dat =}\NormalTok{ pred_dat,}
-    \DataTypeTok{fit_lm =}\NormalTok{ fit_lm,}
-    \DataTypeTok{fit_lme =}\NormalTok{ fit_lme}
-
-\NormalTok{  ))}
-\NormalTok{\}}
-
-\NormalTok{plot_example <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(sim_dat) \{}
-  \KeywordTok{library}\NormalTok{(patchwork)}
-  \CommentTok{## Plot the simulated data}
-\NormalTok{  p1 <-}\StringTok{ }\NormalTok{sim_dat}\OperatorTok{$}\NormalTok{dat }\OperatorTok{%>%}
-\StringTok{    }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ y, }\DataTypeTok{group =}\NormalTok{ treatment, }\DataTypeTok{color =}\NormalTok{ treatment)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{show.legend=}\OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'response'}\NormalTok{)}\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ mu, }\DataTypeTok{color =}\NormalTok{ treatment),}\DataTypeTok{show.legend=}\OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{theme_classic}\NormalTok{() }\OperatorTok{+}
-\StringTok{    }\CommentTok{#ggtitle("Simulated data with true response function")+}
-\StringTok{    }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{plot.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{30}\NormalTok{, }
-                                  \DataTypeTok{face =} \StringTok{"bold"}\NormalTok{),}
-        \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{30}\NormalTok{))}\OperatorTok{+}
-\StringTok{    }\KeywordTok{scale_color_aaas}\NormalTok{()}
-
-\NormalTok{  p2 <-}\StringTok{ }\NormalTok{sim_dat}\OperatorTok{$}\NormalTok{dat }\OperatorTok{%>%}
-\StringTok{    }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ y, }\DataTypeTok{group =}\NormalTok{ subject, }\DataTypeTok{color =}\NormalTok{ treatment)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{size =} \StringTok{"Subjects"}\NormalTok{),}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
-\StringTok{    }\CommentTok{# facet_wrap(~ treatment) +}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ mu, }\DataTypeTok{color =}\NormalTok{ treatment, }\DataTypeTok{size =} \StringTok{"Simulated Truth"}\NormalTok{), }\DataTypeTok{lty =} \DecValTok{1}\NormalTok{,}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'response'}\NormalTok{)}\OperatorTok{+}
-\StringTok{    }\KeywordTok{scale_size_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Type"}\NormalTok{, }\DataTypeTok{values=}\KeywordTok{c}\NormalTok{(}\StringTok{"Subjects"}\NormalTok{ =}\StringTok{ }\FloatTok{0.5}\NormalTok{, }\StringTok{"Simulated Truth"}\NormalTok{ =}\StringTok{ }\DecValTok{3}\NormalTok{)) }\OperatorTok{+}
-\StringTok{    }\CommentTok{#ggtitle("Simulated data\textbackslash{}nIndividual responses with population mean") +}
-\StringTok{    }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
-\StringTok{     }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{plot.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{30}\NormalTok{, }
-                                \DataTypeTok{face =} \StringTok{"bold"}\NormalTok{),}
-     \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{30}\NormalTok{))}\OperatorTok{+}
-\StringTok{    }\KeywordTok{scale_color_aaas}\NormalTok{()}
-
-\NormalTok{   p3 <-}\StringTok{ }\NormalTok{sim_dat}\OperatorTok{$}\NormalTok{dat }\OperatorTok{%>%}
-\StringTok{    }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ errors, }\DataTypeTok{group =}\NormalTok{ subject, }\DataTypeTok{color =}\NormalTok{ treatment)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{show.legend=}\OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'errors'}\NormalTok{)}\OperatorTok{+}
-\StringTok{     }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
-\StringTok{    }\CommentTok{# facet_wrap(~ treatment) +}
-\StringTok{    }\CommentTok{#ggtitle("Simulated errors") +}
-\StringTok{     }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{plot.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{30}\NormalTok{, }
-                                  \DataTypeTok{face =} \StringTok{"bold"}\NormalTok{),}
-        \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{30}\NormalTok{))}\OperatorTok{+}
-\StringTok{    }\KeywordTok{scale_color_aaas}\NormalTok{()}
-
-\NormalTok{  p4 <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(sim_dat}\OperatorTok{$}\NormalTok{dat, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ time, }\DataTypeTok{y =}\NormalTok{ y, }\DataTypeTok{color =}\NormalTok{ treatment)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_point}\NormalTok{()}\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'response'}\NormalTok{)}\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \KeywordTok{predict}\NormalTok{(sim_dat}\OperatorTok{$}\NormalTok{fit_lme), }\DataTypeTok{group =}\NormalTok{ subject, }\DataTypeTok{size =} \StringTok{"Subjects"}\NormalTok{)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sim_dat}\OperatorTok{$}\NormalTok{pred_dat, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =} \KeywordTok{predict}\NormalTok{(sim_dat}\OperatorTok{$}\NormalTok{fit_lme, }\DataTypeTok{level =} \DecValTok{0}\NormalTok{, }\DataTypeTok{newdata =}\NormalTok{ sim_dat}\OperatorTok{$}\NormalTok{pred_dat), }\DataTypeTok{size =} \StringTok{"Population"}\NormalTok{)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{scale_size_manual}\NormalTok{(}\DataTypeTok{name =} \StringTok{"Predictions"}\NormalTok{, }\DataTypeTok{values=}\KeywordTok{c}\NormalTok{(}\StringTok{"Subjects"}\NormalTok{ =}\StringTok{ }\FloatTok{0.5}\NormalTok{, }\StringTok{"Population"}\NormalTok{ =}\StringTok{ }\DecValTok{3}\NormalTok{)) }\OperatorTok{+}
-\StringTok{    }\KeywordTok{theme_classic}\NormalTok{() }\OperatorTok{+}
-\StringTok{    }\CommentTok{#ggtitle("Fitted Model")+}
-\StringTok{    }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{plot.title =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{30}\NormalTok{, }
-                                  \DataTypeTok{face =} \StringTok{"bold"}\NormalTok{),}
-        \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{30}\NormalTok{))}\OperatorTok{+}
-\StringTok{    }\KeywordTok{scale_color_aaas}\NormalTok{()}
-
-  \KeywordTok{return}\NormalTok{((p1}\OperatorTok{+}\NormalTok{p3}\OperatorTok{+}\NormalTok{p2}\OperatorTok{+}\NormalTok{p4)}\OperatorTok{+}\KeywordTok{plot_layout}\NormalTok{(}\DataTypeTok{nrow=}\DecValTok{1}\NormalTok{)}\OperatorTok{+}\KeywordTok{plot_annotation}\NormalTok{(}\DataTypeTok{tag_levels =} \StringTok{'A'}\NormalTok{)) }
-
-\NormalTok{\}}
-
-\NormalTok{txt<-}\DecValTok{18}
-\NormalTok{A1<-}\KeywordTok{plot_example}\NormalTok{(}\KeywordTok{example}\NormalTok{(}\DataTypeTok{fun_type =} \StringTok{"linear"}\NormalTok{, }\DataTypeTok{error_type =} \StringTok{"correlated"}\NormalTok{)) }
-
-\NormalTok{B1<-}\KeywordTok{plot_example}\NormalTok{(}\KeywordTok{example}\NormalTok{(}\DataTypeTok{fun_type =} \StringTok{"linear"}\NormalTok{, }\DataTypeTok{error_type =} \StringTok{"independent"}\NormalTok{)) }
-
-\NormalTok{C1<-}\KeywordTok{plot_example}\NormalTok{(}\KeywordTok{example}\NormalTok{(}\DataTypeTok{fun_type =} \StringTok{"quadratic"}\NormalTok{, }\DataTypeTok{error_type =} \StringTok{"correlated"}\NormalTok{)) }
-
-\NormalTok{D1<-}\KeywordTok{plot_example}\NormalTok{(}\KeywordTok{example}\NormalTok{(}\DataTypeTok{fun_type =} \StringTok{"quadratic"}\NormalTok{, }\DataTypeTok{error_type =} \StringTok{"independent"}\NormalTok{)) }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}[!h]
-\includegraphics{Manuscript_AM_v5_files/figure-latex/linear-cases-1} \caption{**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data**}\label{fig:linear-cases}
-\end{figure}
-
-For the quadratic response case, Figure \ref{fig:quadratic-cases} shows the simulated responses using compound symmetry and independent errors.
-
-\begin{figure}[!h]
-\includegraphics{Manuscript_AM_v5_files/figure-latex/quadratic-cases-1} \caption{**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**}\label{fig:quadratic-cases}
-\end{figure}
-
-\hypertarget{basis-functions-and-gams}{%
-\subsection{Basis functions and GAMs}\label{basis-functions-and-gams}}
-
-\hypertarget{data-simulation-and-gam-models}{%
-\subsection{Data simulation and GAM models}\label{data-simulation-and-gam-models}}
+\begin{verbatim}
+## Example with linear response
+example <- function(n_time = 6, 
+                    fun_type = "linear", 
+                    error_type = "correlated") {
+
+  if (!(fun_type %in% c("linear", "quadratic")))
+    stop('fun_type must be either "linear", or "quadratic"')
+  if (!(error_type %in% c("correlated", "independent")))
+    stop('fun_type must be either "correlated", or "independent"')
+
+  library(tidyverse)
+  library(mvnfast)
+  library(nlme)
+  library(ggsci)
+
+  x <- seq(1,6, length.out = n_time)
+  mu <- matrix(0, length(x), 2)
+  # linear response
+  if (fun_type == "linear") {
+    mu[, 1] <- - (0.25*x)+2  
+    mu[, 2] <- 0.25*x+2
+  } else {
+    # nonlinear response
+
+    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
+    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
+  }
+  # matplot(mu, type = 'l')
+
+  y <- array(0, dim = c(length(x), 2, 10))
+  errors <- array(0, dim = c(length(x), 2, 10))
+
+  if (error_type == "independent") {
+    ## independent errors
+    for (i in 1:2) {
+      for (j in 1:10) {
+        errors[, i, j] <- rnorm(6, 0, 0.25)
+        y[, i, j] <- mu[, i] + errors[, i, j]
+      }
+    }
+  } else {
+    for (i in 1:2) {     # number of treatments
+      for (j in 1:10) {  # number of subjects
+        # compound symmetry errors
+        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
+        y[, i, j] <- mu[, i] + errors[, i, j]
+      }
+    }
+  }
+
+  ## subject random effects
+
+  ## visualizing the difference between independent errors and compound symmetry
+  ## why do we need to account for this -- overly confident inference
+
+  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
+  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
+  dimnames(mu) <- list(time = x, treatment = 1:2)
+  dat <- as.data.frame.table(y, responseName = "y")
+  dat_errors <- as.data.frame.table(errors, responseName = "errors")
+  dat_mu <- as.data.frame.table(mu, responseName = "mu")
+  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
+  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
+  dat$time <- as.numeric(as.character(dat$time))
+  dat <- dat %>%
+    mutate(subject = factor(paste(subject, treatment, sep = "-")))
+
+  ## repeated measures ANOVA in R
+  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
+  dat$preds_lm <- predict(fit_lm)
+
+  fit_lme <- lme(y ~ treatment + time + treatment:time,
+                 data = dat,
+                 random = ~ 1 | subject,
+                 correlation = corCompSymm(form = ~ 1 | subject)
+  )
+
+  pred_dat <- expand.grid(
+    treatment = factor(1:2), 
+    time = unique(dat$time)
+  )
+
+  dat$y_pred <- predict(fit_lme)
+
+  return(list(
+    dat = dat,
+    pred_dat = pred_dat,
+    fit_lm = fit_lm,
+    fit_lme = fit_lme
+
+  ))
+}
+
+plot_example <- function(sim_dat) {
+  library(patchwork)
+  ## Plot the simulated data
+  p1 <- sim_dat$dat %>%
+    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
+    geom_point(show.legend=FALSE) +labs(y='response')+
+    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
+    theme_classic() +
+    #ggtitle("Simulated data with true response function")+
+    theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
+    scale_color_aaas()
+
+  p2 <- sim_dat$dat %>%
+    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
+    geom_line(aes(size = "Subjects"),
+              show.legend = FALSE) +
+    # facet_wrap(~ treatment) +
+    geom_line(aes(
+      x = time, 
+      y = mu, 
+      color = treatment, 
+      size = "Simulated Truth"
+      ), 
+      lty = 1,
+      show.legend = FALSE
+      ) +
+    labs(y='response')+
+    scale_size_manual(
+      name = "Type", 
+      values=c("Subjects" = 0.5, "Simulated Truth" = 3)
+      ) +
+    #ggtitle("Simulated data\nIndividual responses with population mean") +
+    theme_classic()+
+     theme(plot.title = element_text(size = 30, 
+                                face = "bold"),
+     text=element_text(size=30))+
+    scale_color_aaas()
+
+   p3 <- sim_dat$dat %>%
+    ggplot(aes(x = time, 
+               y = errors, 
+               group = subject, 
+               color = treatment)) +
+    geom_line(show.legend=FALSE) +
+     labs(y='errors')+
+     theme_classic()+
+    # facet_wrap(~ treatment) +
+    #ggtitle("Simulated errors") +
+     theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
+    scale_color_aaas()
+
+  p4 <- ggplot(sim_dat$dat, 
+               aes(x = time,
+                   y = y, 
+                   color = treatment)) +
+    geom_point()+
+    labs(y='response')+
+    geom_line(aes(
+      y = predict(sim_dat$fit_lme), 
+      group = subject, 
+      size = "Subjects")
+      ) +
+    geom_line(data = sim_dat$pred_dat, 
+              aes(y = predict(sim_dat$fit_lme, 
+                              level = 0, 
+                              newdata = sim_dat$pred_dat),
+                  size = "Population")) +
+    scale_size_manual(name = "Predictions", 
+                      values=c("Subjects" = 0.5, 
+                               "Population" = 3)) +
+    theme_classic() +
+    #ggtitle("Fitted Model")+
+    theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
+    scale_color_aaas()
+
+  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
+
+}
+
+txt<-18
+A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
+
+B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
+
+C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
+
+D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
+
+\end{verbatim}
+
+\begin{verbatim}
+## Error in eval(expr, envir, enclos): object 'A1' not found
+\end{verbatim}
+
+For the quadratic response case, Figure \ref{fig:quadratic-cases-appendix} shows the simulated responses using compound symmetry and independent errors.
+
+\begin{verbatim}
+## Error in eval(expr, envir, enclos): object 'C1' not found
+\end{verbatim}
 

### 2021-05-13 Manuscripts/Bookdown testing/Rendering.R

@@ -1,4 +1,0 @@
-render_book(input = ".", output_format = NULL, ..., clean = TRUE,
-            envir = parent.frame(), clean_envir = !interactive(),
-            output_dir = NULL, new_session = NA, preview = FALSE,
-            config_file = "_bookdown.yml")

### 2021-05-13 Manuscripts/Bookdown testing/06-Appendix.Rmd
Need to work on Appendix and Conclusion.

@@
 This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation).
-The linear simulations using Figure \@ref(fig:linear-cases-appendix) show in  panels A and D the simulated mean responses and individual datapoints.
+The linear simulations using Figure \@ref(fig:linear-cases) show in  panels A and D the simulated mean responses and individual datapoints.
 Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
@@
 
-```{r-responses-appendix,include=TRUE,message=FALSE,echo=TRUE}
+```{r ,include=TRUE,message=FALSE,echo=TRUE}
 ## Example with linear response
@@
     ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),
-              show.legend = FALSE) +
+    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
     # facet_wrap(~ treatment) +
-    geom_line(aes(
-      x = time, 
-      y = mu, 
-      color = treatment, 
-      size = "Simulated Truth"
-      ), 
-      lty = 1,
-      show.legend = FALSE
-      ) +
-    labs(y='response')+
-    scale_size_manual(
-      name = "Type", 
-      values=c("Subjects" = 0.5, "Simulated Truth" = 3)
-      ) +
+    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
+    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
     #ggtitle("Simulated data\nIndividual responses with population mean") +
@@
    p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = errors, 
-               group = subject, 
-               color = treatment)) +
-    geom_line(show.legend=FALSE) +
-     labs(y='errors')+
+    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
+    geom_line(show.legend=FALSE) +labs(y='errors')+
      theme_classic()+
@@
 
-  p4 <- ggplot(sim_dat$dat, 
-               aes(x = time,
-                   y = y, 
-                   color = treatment)) +
-    geom_point()+
-    labs(y='response')+
-    geom_line(aes(
-      y = predict(sim_dat$fit_lme), 
-      group = subject, 
-      size = "Subjects")
-      ) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_lme, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat),
-                  size = "Population")) +
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, 
-                               "Population" = 3)) +
+  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
+    geom_point()+labs(y='response')+
+    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
+    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
+    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
     theme_classic() +
@@
 
-```{r, linear-cases-appendix, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C: Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
+```{r, linear-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
 B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
 Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data."}
+The rm-ANOVA model does not pick the trend of the quadratic data**"}
 # linear response, correlated errors (subject effect)
 A1/B1+plot_annotation(tag_levels = 'A')
-
 ```
 
-For the quadratic response case, Figure \@ref(fig:quadratic-cases-appendix) shows the simulated responses using compound symmetry and independent errors.
+For the quadratic response case, Figure \@ref(fig:quadratic-cases) shows the simulated responses using compound symmetry and independent errors.
 
-```{r, quadratic-cases-appendix, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
+```{r, quadratic-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
 A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
@@
 D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data."}
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
 # linear response, correlated errors (subject effect)
 C1/D1+plot_annotation(tag_levels = 'A')
-
 ```

### 2021-05-13 Manuscripts/Bookdown testing/Full_document.tex
Need to work on Appendix and Conclusion.

--- 2021-05-12 Manuscripts/Manuscript v5/Manuscript_AM_v5.tex
@@
 \begin{figure}[!h]
-\includegraphics{Manuscript_AM_v5_files/figure-latex/linear-cases-1} \caption{**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+\includegraphics{Full_document_files/figure-latex/linear-cases-1} \caption{**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
 A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
@@
 \begin{figure}[!h]
-\includegraphics{Manuscript_AM_v5_files/figure-latex/quadratic-cases-1} \caption{**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
+\includegraphics{Full_document_files/figure-latex/quadratic-cases-1} \caption{**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
 A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
@@
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{#generate the response}
+\NormalTok{n_time =}\StringTok{ }\DecValTok{6}
+\NormalTok{ x <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{6}\NormalTok{, }\DataTypeTok{length.out =}\NormalTok{ n_time)}
+\NormalTok{ mu <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{length}\NormalTok{(x), }\DecValTok{2}\NormalTok{)}
+\NormalTok{ mu[, }\DecValTok{1}\NormalTok{] <-}\StringTok{  }\OperatorTok{-}\NormalTok{(}\FloatTok{0.25} \OperatorTok{*}\StringTok{ }\NormalTok{x}\OperatorTok{^}\DecValTok{2}\NormalTok{) }\FloatTok{+1.5}\OperatorTok{*}\NormalTok{x}\FloatTok{-1.25} \CommentTok{#mean response}
+\NormalTok{ mu[, }\DecValTok{2}\NormalTok{] <-}\StringTok{ }\NormalTok{(}\FloatTok{0.25} \OperatorTok{*}\StringTok{ }\NormalTok{x}\OperatorTok{^}\DecValTok{2}\NormalTok{) }\FloatTok{-1.5}\OperatorTok{*}\NormalTok{x}\FloatTok{+1.25} \CommentTok{#mean response}
+\NormalTok{ y <-}\StringTok{ }\KeywordTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DataTypeTok{dim =} \KeywordTok{c}\NormalTok{(}\KeywordTok{length}\NormalTok{(x), }\DecValTok{2}\NormalTok{, }\DecValTok{10}\NormalTok{))}
+\NormalTok{ errors <-}\StringTok{ }\KeywordTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DataTypeTok{dim =} \KeywordTok{c}\NormalTok{(}\KeywordTok{length}\NormalTok{(x), }\DecValTok{2}\NormalTok{, }\DecValTok{10}\NormalTok{))}
+ \ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{) \{     }\CommentTok{# number of treatments}
+     \ControlFlowTok{for}\NormalTok{ (j }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{) \{  }\CommentTok{# number of subjects}
+         \CommentTok{# compound symmetry errors}
+\NormalTok{         errors[, i, j] <-}\StringTok{ }\KeywordTok{rmvn}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{length}\NormalTok{(x)), }\FloatTok{0.1} \OperatorTok{*}\StringTok{ }\KeywordTok{diag}\NormalTok{(}\DecValTok{6}\NormalTok{) }\OperatorTok{+}\StringTok{ }\FloatTok{0.25} \OperatorTok{*}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{6}\NormalTok{))}
+\NormalTok{         y[, i, j] <-}\StringTok{ }\NormalTok{mu[, i] }\OperatorTok{+}\StringTok{ }\NormalTok{errors[, i, j]}
+\NormalTok{     \}}
+\NormalTok{ \}}
+  \KeywordTok{dimnames}\NormalTok{(y) <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{time =}\NormalTok{ x, }\DataTypeTok{treatment =} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{, }\DataTypeTok{subject =} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{)}
+ \KeywordTok{dimnames}\NormalTok{(errors) <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{time =}\NormalTok{ x, }\DataTypeTok{treatment =} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{, }\DataTypeTok{subject =} \DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{)}
+ \KeywordTok{dimnames}\NormalTok{(mu) <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{time =}\NormalTok{ x, }\DataTypeTok{treatment =} \DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{)}
+\NormalTok{ dat <-}\StringTok{ }\KeywordTok{as.data.frame.table}\NormalTok{(y, }\DataTypeTok{responseName =} \StringTok{"y"}\NormalTok{)}
+\NormalTok{ dat_errors <-}\StringTok{ }\KeywordTok{as.data.frame.table}\NormalTok{(errors, }\DataTypeTok{responseName =} \StringTok{"errors"}\NormalTok{)}
+\NormalTok{ dat_mu <-}\StringTok{ }\KeywordTok{as.data.frame.table}\NormalTok{(mu, }\DataTypeTok{responseName =} \StringTok{"mu"}\NormalTok{)}
+\NormalTok{ dat <-}\StringTok{ }\KeywordTok{left_join}\NormalTok{(dat, dat_errors, }\DataTypeTok{by =} \KeywordTok{c}\NormalTok{(}\StringTok{"time"}\NormalTok{, }\StringTok{"treatment"}\NormalTok{, }\StringTok{"subject"}\NormalTok{))}
+\NormalTok{ dat <-}\StringTok{ }\KeywordTok{left_join}\NormalTok{(dat, dat_mu, }\DataTypeTok{by =} \KeywordTok{c}\NormalTok{(}\StringTok{"time"}\NormalTok{, }\StringTok{"treatment"}\NormalTok{))}
+\NormalTok{ dat}\OperatorTok{$}\NormalTok{time <-}\StringTok{ }\KeywordTok{as.numeric}\NormalTok{(}\KeywordTok{as.character}\NormalTok{(dat}\OperatorTok{$}\NormalTok{time))}
+
+ \CommentTok{#add subject column}
+\NormalTok{ dat <-}\StringTok{ }\NormalTok{dat }\OperatorTok{%>%}
+\StringTok{     }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{subject =} \KeywordTok{factor}\NormalTok{(}\KeywordTok{paste}\NormalTok{(subject, treatment, }\DataTypeTok{sep =} \StringTok{"-"}\NormalTok{)))}
+
+\NormalTok{  dat<-}\KeywordTok{subset}\NormalTok{(dat,treatment}\OperatorTok{==}\DecValTok{1}\NormalTok{)}
+\NormalTok{  dat<-dat[,}\KeywordTok{c}\NormalTok{(}\StringTok{'y'}\NormalTok{,}\StringTok{'time'}\NormalTok{)]}
+\NormalTok{gm<-}\KeywordTok{gam}\NormalTok{(y}\OperatorTok{~}\KeywordTok{s}\NormalTok{(time,}\DataTypeTok{k=}\DecValTok{5}\NormalTok{),}\DataTypeTok{data=}\NormalTok{dat)}
+\NormalTok{model_matrix<-}\KeywordTok{as.data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(gm,}\DataTypeTok{type=}\StringTok{'lpmatrix'}\NormalTok{))}
+
+\NormalTok{x_new <-}\StringTok{ }\NormalTok{(}\KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{max}\NormalTok{(dat}\OperatorTok{$}\NormalTok{time), }\DataTypeTok{length.out =} \DecValTok{100}\NormalTok{))}
+\NormalTok{y_pred <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(gm, }\DataTypeTok{newdata=}\NormalTok{dat) }
+
+\NormalTok{time<-}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{6}\NormalTok{)}
+
+\NormalTok{basis<-model_matrix[}\DecValTok{1}\OperatorTok{:}\DecValTok{6}\NormalTok{,] }\CommentTok{#extracting basis}
+\CommentTok{#basis<-model_matrix[1:6,-1] #extracting basis}
+\KeywordTok{colnames}\NormalTok{(basis)[}\KeywordTok{colnames}\NormalTok{(basis)}\OperatorTok{==}\StringTok{"(Intercept)"}\NormalTok{]<-}\StringTok{"s(time).0"}
+\NormalTok{basis<-basis }\OperatorTok{%>%}\StringTok{ }\CommentTok{#pivoting to long format}
+\StringTok{  }\KeywordTok{pivot_longer}\NormalTok{(}
+    \DataTypeTok{cols=}\KeywordTok{starts_with}\NormalTok{(}\StringTok{"s"}\NormalTok{)}
+\NormalTok{  )}\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{arrange}\NormalTok{(name) }\CommentTok{#ordering}
+
+\CommentTok{#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)}
+\NormalTok{ln<-}\DecValTok{6}\OperatorTok{*}\NormalTok{(}\KeywordTok{length}\NormalTok{(}\KeywordTok{coef}\NormalTok{(gm))) }
+
+\NormalTok{basis_plot<-}\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Basis=}\KeywordTok{integer}\NormalTok{(ln),}
+                       \DataTypeTok{value_orig=}\KeywordTok{double}\NormalTok{(ln),}
+                       \DataTypeTok{time=}\KeywordTok{integer}\NormalTok{(ln),}
+                       \DataTypeTok{cof=}\KeywordTok{double}\NormalTok{(ln)}
+\NormalTok{)}
+
+\NormalTok{basis_plot}\OperatorTok{$}\NormalTok{time<-}\KeywordTok{rep}\NormalTok{(time) }\CommentTok{#pasting timepoints}
+\NormalTok{basis_plot}\OperatorTok{$}\NormalTok{Basis<-}\KeywordTok{factor}\NormalTok{(}\KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{),}\DataTypeTok{each=}\DecValTok{6}\NormalTok{)) }\CommentTok{#pasting basis number values}
+\NormalTok{basis_plot}\OperatorTok{$}\NormalTok{value_orig<-basis}\OperatorTok{$}\NormalTok{value }\CommentTok{#pasting basis values}
+\NormalTok{basis_plot}\OperatorTok{$}\NormalTok{cof<-}\KeywordTok{rep}\NormalTok{(}\KeywordTok{coef}\NormalTok{(gm)[}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{],}\DataTypeTok{each=}\DecValTok{6}\NormalTok{) }\CommentTok{#pasting coefficients}
+\NormalTok{basis_plot<-basis_plot}\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{mod_val=}\NormalTok{value_orig}\OperatorTok{*}\NormalTok{cof)}
+
+\CommentTok{#creating labeller to change the labels in the basis plots}
+
+\NormalTok{basis_names<-}\KeywordTok{c}\NormalTok{(}
+  \StringTok{`}\DataTypeTok{1}\StringTok{`}\NormalTok{=}\StringTok{"Intercept"}\NormalTok{,}
+  \StringTok{`}\DataTypeTok{2}\StringTok{`}\NormalTok{=}\StringTok{"1"}\NormalTok{,}
+  \StringTok{`}\DataTypeTok{3}\StringTok{`}\NormalTok{=}\StringTok{"2"}\NormalTok{,}
+  \StringTok{`}\DataTypeTok{4}\StringTok{`}\NormalTok{=}\StringTok{"3"}\NormalTok{,}
+  \StringTok{`}\DataTypeTok{5}\StringTok{`}\NormalTok{=}\StringTok{"4"}
+\NormalTok{)}
+
+\CommentTok{#calculating the final spline by aggregating the basis functions}
+
+\NormalTok{spl<-basis_plot}\OperatorTok{%>%}\StringTok{ }
+\StringTok{  }\KeywordTok{group_by}\NormalTok{(time)}\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{summarize}\NormalTok{(}\DataTypeTok{spl=}\KeywordTok{sum}\NormalTok{(mod_val))}
+
+\CommentTok{#original basis}
+\NormalTok{sz<-}\DecValTok{1}
+\NormalTok{p11<-}\KeywordTok{ggplot}\NormalTok{(basis_plot,}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{time,}\DataTypeTok{y=}\NormalTok{value_orig,}\DataTypeTok{colour=}\KeywordTok{as.factor}\NormalTok{(Basis)))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{size=}\NormalTok{sz,}\DataTypeTok{show.legend=}\OtherTok{FALSE}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{size=}\NormalTok{sz}\OperatorTok{+}\DecValTok{1}\NormalTok{,}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'Basis functions'}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{facet_wrap}\NormalTok{(}\OperatorTok{~}\NormalTok{Basis,}\DataTypeTok{labeller =} \KeywordTok{as_labeller}\NormalTok{(basis_names))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_color_aaas}\NormalTok{()}
+
+\CommentTok{#penalized basis}
+\NormalTok{p12<-}\KeywordTok{ggplot}\NormalTok{(basis_plot,}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{time,}\DataTypeTok{y=}\NormalTok{mod_val,}\DataTypeTok{colour=}\KeywordTok{as.factor}\NormalTok{(Basis)))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{,}\DataTypeTok{size=}\NormalTok{sz)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{,}\DataTypeTok{size=}\NormalTok{sz}\OperatorTok{+}\DecValTok{1}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'Penalized }\CharTok{\textbackslash{}n}\StringTok{ basis functions'}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_y_continuous}\NormalTok{(}\DataTypeTok{breaks=}\KeywordTok{seq}\NormalTok{(}\OperatorTok{-}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{facet_wrap}\NormalTok{(}\OperatorTok{~}\NormalTok{Basis,}\DataTypeTok{labeller=}\KeywordTok{as_labeller}\NormalTok{(basis_names))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_color_aaas}\NormalTok{()}
+
+\CommentTok{#heatmap of the penalization coefficient}
+\NormalTok{x_labels<-}\KeywordTok{c}\NormalTok{(}\StringTok{"Intercept"}\NormalTok{,}\StringTok{"1"}\NormalTok{,}\StringTok{"2"}\NormalTok{,}\StringTok{"3"}\NormalTok{,}\StringTok{"4"}\NormalTok{)}
+\NormalTok{p13<-}\KeywordTok{ggplot}\NormalTok{(basis_plot,}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{Basis,}\DataTypeTok{y=}\NormalTok{Basis,}\DataTypeTok{fill=}\NormalTok{cof))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_tile}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{color=}\StringTok{'black'}\NormalTok{),}\DataTypeTok{size=}\NormalTok{sz}\OperatorTok{+}\DecValTok{1}\NormalTok{,}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_tile}\NormalTok{(}\DataTypeTok{size=}\NormalTok{sz}\OperatorTok{+}\DecValTok{1}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_fill_gradient}\NormalTok{(}\DataTypeTok{low =} \StringTok{"white"}\NormalTok{, }\DataTypeTok{high =} \StringTok{"red"}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{x=}\StringTok{'Basis'}\NormalTok{,}\DataTypeTok{y=}\StringTok{'Basis'}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_x_discrete}\NormalTok{(}\DataTypeTok{labels=}\NormalTok{x_labels)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_text}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{label=}\KeywordTok{round}\NormalTok{(cof,}\DecValTok{2}\NormalTok{)),}\DataTypeTok{size=}\DecValTok{10}\NormalTok{,}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.title =} \KeywordTok{element_blank}\NormalTok{())}
+
+\CommentTok{#plotting simulated datapoints and smooth term}
+\NormalTok{p14<-}\KeywordTok{ggplot}\NormalTok{(}\DataTypeTok{data=}\NormalTok{dat,}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{time,}\DataTypeTok{y=}\NormalTok{y))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{size=}\NormalTok{sz}\OperatorTok{+}\DecValTok{1}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_color_aaas}\NormalTok{()}\OperatorTok{+}
+\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\StringTok{'Simulated }\CharTok{\textbackslash{}n}\StringTok{ response'}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{data=}\NormalTok{spl,}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{time,}\DataTypeTok{y=}\NormalTok{spl),}\DataTypeTok{color=}\StringTok{"#B15731"}\NormalTok{,}\DataTypeTok{size=}\NormalTok{sz}\OperatorTok{+}\DecValTok{1}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_classic}\NormalTok{()}
+
+\CommentTok{#Combining all}
+\NormalTok{b_plot<-p11}\OperatorTok{+}\NormalTok{p13}\OperatorTok{+}\NormalTok{p12}\OperatorTok{+}\NormalTok{p14}\OperatorTok{+}\KeywordTok{plot_annotation}\NormalTok{(}\DataTypeTok{tag_levels=}\StringTok{'A'}\NormalTok{)}\OperatorTok{&}
+\StringTok{  }\KeywordTok{theme}\NormalTok{(}
+     \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{18}\NormalTok{)}
+\NormalTok{     )}
+
+\NormalTok{b_plot}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Full_document_files/figure-latex/basis-plot-Appendix-1}
+
 \hypertarget{data-simulation-and-gam-models}{%
@@
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{dat<-}\KeywordTok{tibble}\NormalTok{(}\DataTypeTok{StO2=}\KeywordTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{,}\DecValTok{27}\NormalTok{,}\DecValTok{3}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.5}\NormalTok{,}\DecValTok{7}\NormalTok{,}\DecValTok{4}\NormalTok{,}\DecValTok{50}\NormalTok{,}\DecValTok{45}\NormalTok{,}\DecValTok{56}\NormalTok{),}
+            \DataTypeTok{Day=}\KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{2}\NormalTok{,}\DecValTok{5}\NormalTok{,}\DecValTok{7}\NormalTok{,}\DecValTok{10}\NormalTok{),}\DataTypeTok{times=}\DecValTok{2}\NormalTok{),}
+            \DataTypeTok{Group=}\KeywordTok{as.factor}\NormalTok{(}\KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"Control"}\NormalTok{,}\StringTok{"Treatment"}\NormalTok{),}\DataTypeTok{each=}\DecValTok{5}\NormalTok{))}
+\NormalTok{            )}
+
+\CommentTok{## plot the mean response}
+\NormalTok{f1<-}\KeywordTok{ggplot}\NormalTok{(dat, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Day, }\DataTypeTok{y =}\NormalTok{ StO2, }\DataTypeTok{color =}\NormalTok{ Group)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{size=}\DecValTok{1}\NormalTok{,}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{)}\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{,}\DataTypeTok{size=}\FloatTok{1.5}\NormalTok{,}\DataTypeTok{alpha=}\FloatTok{0.5}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\KeywordTok{expression}\NormalTok{(}\KeywordTok{paste}\NormalTok{(StO[}\DecValTok{2}\NormalTok{],}\StringTok{' (real)'}\NormalTok{)))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_color_aaas}\NormalTok{()}\OperatorTok{+}
+\StringTok{    }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks=}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{5}\NormalTok{,}\DecValTok{10}\NormalTok{))}\OperatorTok{+}
+\StringTok{    }\KeywordTok{scale_y_continuous}\NormalTok{(}\DataTypeTok{breaks=}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{40}\NormalTok{))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{plot_layout}\NormalTok{(}\DataTypeTok{tag_level =} \StringTok{'new'}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme}\NormalTok{(}
+    \DataTypeTok{plot.background =} \KeywordTok{element_rect}\NormalTok{(}\DataTypeTok{fill =} \StringTok{"transparent"}\NormalTok{, }\DataTypeTok{color =} \OtherTok{NA}\NormalTok{),}
+    \DataTypeTok{axis.text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{14}\NormalTok{)}
+\NormalTok{  )}
+
+\CommentTok{#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.}
+\CommentTok{#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")}
+
+\NormalTok{simulate_data <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(dat, }\DataTypeTok{n =} \DecValTok{10}\NormalTok{, }\DataTypeTok{sd =} \DecValTok{5}\NormalTok{) \{}
+\NormalTok{    dat_sim <-}\StringTok{ }\NormalTok{dat }\OperatorTok{%>%}
+\StringTok{        }\KeywordTok{slice}\NormalTok{(}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{n}\NormalTok{(), }\DataTypeTok{each =}\NormalTok{ n)) }\OperatorTok{%>%}
+\StringTok{        }\KeywordTok{group_by}\NormalTok{(Group, Day) }\OperatorTok{%>%}
+\StringTok{        }\KeywordTok{mutate}\NormalTok{(}
+               \DataTypeTok{StO2_sim =} \KeywordTok{pmax}\NormalTok{(}\KeywordTok{rnorm}\NormalTok{(n, StO2, sd), }\FloatTok{0.0001}\NormalTok{),}
+               \DataTypeTok{subject=}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{),}
+               \DataTypeTok{subject=}\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste}\NormalTok{(subject, Group, }\DataTypeTok{sep =} \StringTok{"-"}\NormalTok{))}
+\NormalTok{               ) }\OperatorTok{%>%}
+\StringTok{        }\KeywordTok{ungroup}\NormalTok{()}
+
+    \KeywordTok{return}\NormalTok{(dat_sim)}
+\NormalTok{\}}
+
+\CommentTok{#subject = factor(paste(subject, treatment, sep = "-")))}
+
+\NormalTok{n <-}\StringTok{ }\DecValTok{10} \CommentTok{#number of observations}
+\NormalTok{sd <-}\StringTok{ }\DecValTok{10} \CommentTok{#approximate sd from paper}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{) }\CommentTok{#set seed for reproducibility}
+\NormalTok{df <-}\StringTok{ }\DecValTok{6}
+\NormalTok{dat_sim <-}\StringTok{ }\KeywordTok{simulate_data}\NormalTok{(dat, n, sd)}
+
+\CommentTok{#plotting simulated data}
+\NormalTok{f2<-}\KeywordTok{ggplot}\NormalTok{(dat_sim, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Day, }\DataTypeTok{y =}\NormalTok{ StO2_sim, }\DataTypeTok{color =}\NormalTok{ Group)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{show.legend=}\OtherTok{FALSE}\NormalTok{,}\DataTypeTok{size=}\FloatTok{1.5}\NormalTok{,}\DataTypeTok{alpha=}\FloatTok{0.5}\NormalTok{)}\OperatorTok{+}
+\StringTok{    }\KeywordTok{stat_summary}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ StO2_sim,}
+                     \DataTypeTok{group=}\NormalTok{Group), }
+                 \DataTypeTok{fun=}\NormalTok{mean, }\DataTypeTok{geom=}\StringTok{"line"}\NormalTok{,}\DataTypeTok{size=}\DecValTok{1}\NormalTok{,}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\KeywordTok{expression}\NormalTok{(}\KeywordTok{atop}\NormalTok{(StO[}\DecValTok{2}\NormalTok{],}\StringTok{'(simulated)'}\NormalTok{)))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme}\NormalTok{(}
+    \DataTypeTok{axis.text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{22}\NormalTok{)}
+\NormalTok{  )}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_color_aaas}\NormalTok{()}\OperatorTok{+}
+\StringTok{    }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks=}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{2}\NormalTok{,}\DecValTok{5}\NormalTok{,}\DecValTok{7}\NormalTok{,}\DecValTok{10}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{#GAM for StO2}
+
+\NormalTok{gam1 <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(StO2_sim }\OperatorTok{~}\StringTok{ }\NormalTok{Group}\OperatorTok{+}\KeywordTok{s}\NormalTok{(Day, }\DataTypeTok{by =}\NormalTok{ Group, }\DataTypeTok{k =} \DecValTok{5}\NormalTok{,}\DataTypeTok{bs=}\StringTok{"gp"}\NormalTok{),}
+            \DataTypeTok{method=}\StringTok{'REML'}\NormalTok{,}
+            \DataTypeTok{data  =}\NormalTok{ dat_sim)}
+
+\CommentTok{#linear model}
+\NormalTok{lm1<-}\KeywordTok{lm}\NormalTok{(StO2_sim }\OperatorTok{~}\StringTok{ }\NormalTok{Day }\OperatorTok{+}\StringTok{ }\NormalTok{Group }\OperatorTok{+}\StringTok{ }\NormalTok{Day }\OperatorTok{*}\StringTok{ }\NormalTok{Group, }\DataTypeTok{data =}\NormalTok{ dat_sim)}
+
+\CommentTok{#creates a dataframe using the length of the covariates for the GAM}
+\NormalTok{gam_predict <-}\StringTok{ }\KeywordTok{expand_grid}\NormalTok{(}\DataTypeTok{Group =} \KeywordTok{factor}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"Control"}\NormalTok{, }\StringTok{"Treatment"}\NormalTok{)),}
+                         \DataTypeTok{Day =} \KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DataTypeTok{by =} \FloatTok{0.1}\NormalTok{),}
+                         \DataTypeTok{subject=}\KeywordTok{factor}\NormalTok{(}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{)))}
+
+\CommentTok{#creates a dataframe using the length of the covariates for rm-ANOVA}
+\NormalTok{lm_predict<-}\KeywordTok{expand_grid}\NormalTok{(}\DataTypeTok{Group =} \KeywordTok{factor}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"Control"}\NormalTok{, }\StringTok{"Treatment"}\NormalTok{)),}
+                         \DataTypeTok{Day =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\OperatorTok{:}\DecValTok{10}\NormalTok{),}
+                        \DataTypeTok{subject=}\KeywordTok{factor}\NormalTok{(}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{)),}
+\NormalTok{                          )}
+\NormalTok{lm_predict}\OperatorTok{$}\NormalTok{subject<-}\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste}\NormalTok{(lm_predict}\OperatorTok{$}\NormalTok{subject, lm_predict}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{sep =} \StringTok{"-"}\NormalTok{))}
+
+\CommentTok{#adds the predictions to the grid and creates a confidence interval for GAM}
+\NormalTok{gam_predict<-gam_predict}\OperatorTok{%>%}
+\StringTok{    }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{fit =} \KeywordTok{predict}\NormalTok{(gam1,gam_predict,}\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{,}\DataTypeTok{type=}\StringTok{'response'}\NormalTok{)}\OperatorTok{$}\NormalTok{fit,}
+           \DataTypeTok{se.fit =} \KeywordTok{predict}\NormalTok{(gam1, gam_predict,}\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{,}\DataTypeTok{type=}\StringTok{'response'}\NormalTok{)}\OperatorTok{$}\NormalTok{se.fit)}
+
+\CommentTok{#using lm}
+\NormalTok{lm_predict<-lm_predict}\OperatorTok{%>%}
+\StringTok{    }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{fit =} \KeywordTok{predict}\NormalTok{(lm1,lm_predict,}\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{,}\DataTypeTok{type=}\StringTok{'response'}\NormalTok{)}\OperatorTok{$}\NormalTok{fit,}
+           \DataTypeTok{se.fit =} \KeywordTok{predict}\NormalTok{(lm1, lm_predict,}\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{,}\DataTypeTok{type=}\StringTok{'response'}\NormalTok{)}\OperatorTok{$}\NormalTok{se.fit)}
+
+\CommentTok{#plot smooths and confidence interval for GAM}
+\NormalTok{f3<-}\KeywordTok{ggplot}\NormalTok{(}\DataTypeTok{data=}\NormalTok{dat_sim, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{Day, }\DataTypeTok{y=}\NormalTok{StO2_sim, }\DataTypeTok{group=}\NormalTok{Group)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{color=}\NormalTok{Group),}\DataTypeTok{size=}\FloatTok{1.5}\NormalTok{,}\DataTypeTok{alpha=}\FloatTok{0.5}\NormalTok{,}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{( }\DataTypeTok{x=}\NormalTok{Day,}\DataTypeTok{ymin=}\NormalTok{(fit }\OperatorTok{-}\StringTok{ }\DecValTok{2}\OperatorTok{*}\NormalTok{se.fit), }
+                   \DataTypeTok{ymax=}\NormalTok{(fit }\OperatorTok{+}\StringTok{ }\DecValTok{2}\OperatorTok{*}\NormalTok{se.fit),}
+                   \DataTypeTok{fill=}\NormalTok{Group}
+\NormalTok{                   ),}
+              \DataTypeTok{alpha=}\FloatTok{0.3}\NormalTok{,}
+              \DataTypeTok{data=}\NormalTok{gam_predict,}
+            \DataTypeTok{show.legend=}\OtherTok{FALSE}\NormalTok{,}
+                \DataTypeTok{inherit.aes=}\OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y=}\NormalTok{fit,}
+                \DataTypeTok{color=}\NormalTok{Group),}
+              \DataTypeTok{size=}\DecValTok{1}\NormalTok{,}\DataTypeTok{data=}\NormalTok{gam_predict,}
+              \DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\CommentTok{#facet_wrap(~Group)+}
+\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\KeywordTok{expression}\NormalTok{(}\KeywordTok{atop}\NormalTok{(StO[}\DecValTok{2}\NormalTok{],}\StringTok{'complete'}\NormalTok{)))}\OperatorTok{+}
+\StringTok{    }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks=}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{2}\NormalTok{,}\DecValTok{5}\NormalTok{,}\DecValTok{7}\NormalTok{,}\DecValTok{10}\NormalTok{))}\OperatorTok{+}
+\StringTok{      }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme}\NormalTok{(}
+    \DataTypeTok{axis.text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{22}\NormalTok{)}
+\NormalTok{  )}\OperatorTok{+}
+\StringTok{      }\KeywordTok{scale_color_aaas}\NormalTok{()}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_fill_aaas}\NormalTok{()}
+
+\CommentTok{#plot linear fit for rm-ANOVA}
+\NormalTok{f4<-}\KeywordTok{ggplot}\NormalTok{(}\DataTypeTok{data=}\NormalTok{dat_sim, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{Day, }\DataTypeTok{y=}\NormalTok{StO2_sim, }\DataTypeTok{group=}\NormalTok{Group)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{color=}\NormalTok{Group),}\DataTypeTok{size=}\FloatTok{1.5}\NormalTok{,}\DataTypeTok{alpha=}\FloatTok{0.5}\NormalTok{,}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{( }\DataTypeTok{x=}\NormalTok{Day,}\DataTypeTok{ymin=}\NormalTok{(fit }\OperatorTok{-}\StringTok{ }\DecValTok{2}\OperatorTok{*}\NormalTok{se.fit), }
+                   \DataTypeTok{ymax=}\NormalTok{(fit }\OperatorTok{+}\StringTok{ }\DecValTok{2}\OperatorTok{*}\NormalTok{se.fit),}\DataTypeTok{fill=}\NormalTok{Group),}
+              \DataTypeTok{alpha=}\FloatTok{0.3}\NormalTok{,}
+              \DataTypeTok{data=}\NormalTok{lm_predict,}
+              \DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{,}
+                \DataTypeTok{inherit.aes=}\OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y=}\NormalTok{fit,}
+                \DataTypeTok{color=}\NormalTok{Group),}
+              \DataTypeTok{size=}\DecValTok{1}\NormalTok{,}\DataTypeTok{data=}\NormalTok{lm_predict,}
+              \DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\CommentTok{#facet_wrap(~Group)+}
+\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\KeywordTok{expression}\NormalTok{(}\KeywordTok{paste}\NormalTok{(}\StringTok{'StO'}\NormalTok{[}\DecValTok{2}\NormalTok{],}\StringTok{' (simulated)'}\NormalTok{)))}\OperatorTok{+}
+\StringTok{    }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks=}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{2}\NormalTok{,}\DecValTok{5}\NormalTok{,}\DecValTok{7}\NormalTok{,}\DecValTok{10}\NormalTok{))}\OperatorTok{+}
+\StringTok{      }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme}\NormalTok{(}
+    \DataTypeTok{axis.text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{22}\NormalTok{)}
+\NormalTok{  )}\OperatorTok{+}
+\StringTok{      }\KeywordTok{scale_color_aaas}\NormalTok{()}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_fill_aaas}\NormalTok{()}
+
+\CommentTok{#posthoc comparisons for the linear model}
+\KeywordTok{library}\NormalTok{(multcomp)}
+
+\CommentTok{#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))}
+\CommentTok{#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{#missing data}
+\CommentTok{#create a sequence of 40 random numbers between 1 and 100, these numbers will}
+\CommentTok{#correspond to the row numbers to be randomly erased from the original dataset}
+\NormalTok{missing <-}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{100}\NormalTok{, }\DecValTok{40}\NormalTok{)}
+
+\CommentTok{#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA}
+
+\NormalTok{ind <-}\StringTok{ }\KeywordTok{which}\NormalTok{(dat_sim}\OperatorTok{$}\NormalTok{StO2_sim }\OperatorTok{%in%}\StringTok{ }\KeywordTok{sample}\NormalTok{(dat_sim}\OperatorTok{$}\NormalTok{StO2_sim, }\DecValTok{40}\NormalTok{))}
+
+\CommentTok{#create a new dataframe, remove the StO2 column}
+\NormalTok{dat_missing <-}\StringTok{ }\NormalTok{dat_sim[,}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+
+\CommentTok{#add NAs at the ind positions}
+\NormalTok{dat_missing}\OperatorTok{$}\NormalTok{StO2_sim[ind]<-}\OtherTok{NA}
+
+\CommentTok{#Count the number of remaining observations per day (original dataset had 10 per group per day)}
+\NormalTok{dat_missing }\OperatorTok{%>%}
+\StringTok{    }\KeywordTok{group_by}\NormalTok{(Day,Group) }\OperatorTok{%>%}
+\StringTok{    }\KeywordTok{filter}\NormalTok{(}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(StO2_sim))}\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{count}\NormalTok{(Day)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## # A tibble: 10 x 3
+## # Groups:   Day, Group [10]
+##      Day Group         n
+##    <dbl> <fct>     <int>
+##  1     0 Control       5
+##  2     0 Treatment     4
+##  3     2 Control       8
+##  4     2 Treatment     6
+##  5     5 Control       4
+##  6     5 Treatment     4
+##  7     7 Control       4
+##  8     7 Treatment     4
+##  9    10 Control       4
+## 10    10 Treatment     7
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{#the same model used for the full dataset}
+\NormalTok{mod_m1 <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(StO2_sim }\OperatorTok{~}\StringTok{ }\NormalTok{Group}\OperatorTok{+}\KeywordTok{s}\NormalTok{(Day,}\DataTypeTok{by=}\NormalTok{Group,}\DataTypeTok{k=}\DecValTok{5}\NormalTok{), }\DataTypeTok{data  =}\NormalTok{ dat_missing,}\DataTypeTok{family=}\NormalTok{scat)}
+\CommentTok{#appraise the model}
+\KeywordTok{appraise}\NormalTok{(mod_m1)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{Full_document_files/figure-latex/missing-data-Appendix-1}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{m_predict <-}\StringTok{ }\KeywordTok{expand_grid}\NormalTok{(}\DataTypeTok{Group =} \KeywordTok{factor}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"Control"}\NormalTok{, }\StringTok{"Treatment"}\NormalTok{)),}
+                         \DataTypeTok{Day =} \KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DataTypeTok{by =} \FloatTok{0.1}\NormalTok{))}
+
+\CommentTok{#adds the predictions to the grid and creates a confidence interval}
+\NormalTok{m_predict<-m_predict}\OperatorTok{%>%}
+\StringTok{    }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{fit =} \KeywordTok{predict}\NormalTok{(mod_m1,m_predict,}\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{,}\DataTypeTok{type=}\StringTok{'response'}\NormalTok{)}\OperatorTok{$}\NormalTok{fit,}
+           \DataTypeTok{se.fit =} \KeywordTok{predict}\NormalTok{(mod_m1, m_predict,}\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{,}\DataTypeTok{type=}\StringTok{'response'}\NormalTok{)}\OperatorTok{$}\NormalTok{se.fit)}
+
+\NormalTok{f6<-}\KeywordTok{ggplot}\NormalTok{(}\DataTypeTok{data=}\NormalTok{dat_missing, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{Day, }\DataTypeTok{y=}\NormalTok{StO2_sim, }\DataTypeTok{group=}\NormalTok{Group)) }\OperatorTok{+}
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{color=}\NormalTok{Group),}\DataTypeTok{size=}\FloatTok{1.5}\NormalTok{,}\DataTypeTok{alpha=}\FloatTok{0.5}\NormalTok{,}\DataTypeTok{show.legend =} \OtherTok{FALSE}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{( }\DataTypeTok{x=}\NormalTok{Day,}\DataTypeTok{ymin=}\NormalTok{(fit }\OperatorTok{-}\StringTok{ }\DecValTok{2}\OperatorTok{*}\NormalTok{se.fit), }
+                   \DataTypeTok{ymax=}\NormalTok{(fit }\OperatorTok{+}\StringTok{ }\DecValTok{2}\OperatorTok{*}\NormalTok{se.fit),}
+                   \DataTypeTok{fill=}\NormalTok{Group}
+\NormalTok{                   ),}
+              \DataTypeTok{alpha=}\FloatTok{0.3}\NormalTok{,}
+              \DataTypeTok{data=}\NormalTok{m_predict,}
+            \DataTypeTok{show.legend=}\OtherTok{FALSE}\NormalTok{,}
+                \DataTypeTok{inherit.aes=}\OtherTok{FALSE}\NormalTok{) }\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y=}\NormalTok{fit,}
+                \DataTypeTok{color=}\NormalTok{Group),}
+              \DataTypeTok{size=}\DecValTok{1}\NormalTok{,}\DataTypeTok{data=}\NormalTok{m_predict,}
+              \DataTypeTok{show.legend =} \OtherTok{TRUE}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\CommentTok{#facet_wrap(~Group)+}
+\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y=}\KeywordTok{expression}\NormalTok{(}\KeywordTok{atop}\NormalTok{(StO[}\DecValTok{2}\NormalTok{],}\StringTok{'missing'}\NormalTok{)))}\OperatorTok{+}
+\StringTok{    }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks=}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{2}\NormalTok{,}\DecValTok{5}\NormalTok{,}\DecValTok{7}\NormalTok{,}\DecValTok{10}\NormalTok{))}\OperatorTok{+}
+\StringTok{      }\KeywordTok{theme_classic}\NormalTok{()}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme}\NormalTok{(}
+    \DataTypeTok{axis.text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{22}\NormalTok{)}
+\NormalTok{  )}\OperatorTok{+}
+\StringTok{      }\KeywordTok{scale_color_aaas}\NormalTok{()}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_fill_aaas}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{mult_plot<-f2}\OperatorTok{+}\KeywordTok{inset_element}\NormalTok{(}
+\NormalTok{  f1, }\DataTypeTok{left =} \FloatTok{0.01}\NormalTok{, }
+  \DataTypeTok{bottom =} \FloatTok{0.5}\NormalTok{, }
+  \DataTypeTok{right =} \FloatTok{0.5}\NormalTok{, }
+  \DataTypeTok{top =} \FloatTok{1.0}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\NormalTok{f3}\OperatorTok{+}\NormalTok{f4}\OperatorTok{+}\NormalTok{f6}\OperatorTok{+}
+\StringTok{   }\KeywordTok{plot_annotation}\NormalTok{(}\DataTypeTok{tag_levels=}\StringTok{'A'}\NormalTok{)}\OperatorTok{&}
+\StringTok{   }\KeywordTok{ylim}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\OperatorTok{-}\DecValTok{5}\NormalTok{,}\DecValTok{75}\NormalTok{)) }\OperatorTok{&}
+\StringTok{  }\KeywordTok{theme}\NormalTok{(}
+     \DataTypeTok{text=}\KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size=}\DecValTok{18}\NormalTok{)}
+\NormalTok{     )}\OperatorTok{&}
+\StringTok{  }\KeywordTok{scale_color_aaas}\NormalTok{()}
+
+\NormalTok{mult_plot}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}[!h]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/sim-smooth-plot-Appendix-1}
+
+}
+
+\caption{Simulated data and smooths for oxygen saturation in tumors.
+A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
+Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
+Lines represent trends for each group, shaded regions are 95\% confidence intervals.
+C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
+D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
+Lines represent trends for each group, shaded regions are 95\% confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
+\end{figure}
+
 \end{document}

### 2021-05-14 Manuscripts/Manuscript by chapters -v1/01-Background.Rmd
Changed folder name, working on Appendix, need to create figures and captions for the GAMs and models

--- 2021-05-11 Manuscripts/Manuscript v5/Manuscript_AM_v5.Rmd
@@
----
-title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
-subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
-author:
-- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
-- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
-  Fayetteville]
-- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
-output:
-  bookdown::pdf_document2:
-    keep_tex: yes
-    fig_caption: yes
-    extra_dependencies:
-      subfig: null
-      breqn: null
-      caption: ["font={small}"]
-      float: null
-  bookdown::word_document2:
-    fig_caption: yes
-    keep_md: yes
-  bookdown::html_document2: default
-  link-citations: yes
-  css: style.css
-  '': default
-csl: elsevier-with-titles.csl
-bibliography: refs.bib
----
-
-# Abstract
-
-In biomedical research, the outcome of longitudinal studies is frequently analyzed statistically using the repeated measures analysis or variance (rm-ANOVA).
-However, rm-ANOVA is restrictive as is only able to accommodate linear responses, requires equal correlation between each measurement (compound symmetry), and cannot work with missing observations.
-When the data fails to fullfill all the required assumptions by rm-ANOVA, the model provides biased estimates and inference.
-In contrast, generalized additive models (GAMs) are a class of models that overcome all of the limitations imposed by rm-ANOVA and that can be implemented in a straightforward manner using the _mgcv_ package in R.
-This paper summarizes the assumptions of rm-ANOVA and how their limit its use in longitudinal biomedical research, presents the basic theoretical framework for GAMs and provides examples of their implementation in R.
-The code and data used to generate this paper are available at:
-
 # Background
@@
 In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models by improving the standards for reproducibility  in biomedical research.
-
-# Challenges presented by longitudinal studies
-
-## The repeated measures ANOVA
-
-The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research.
-This statistical methodology requires certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
-Each one of these assumptions is discussed below.
-
-## Linear relationship
-
-### The repeated measures ANOVA case
-
-In a biomedical longitudinal study, two or more groups of subjects (e.g., patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs.
-a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response is modeled with _fixed_ components.
-The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug in the "treatment" group.
-
-Mathematically speaking, a rm-ANOVA model with an interaction can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
-(\#eq:linear-model)
-\end{equation}
-
-In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$.
-In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
-"novel drug" or "saline" vs.
-"chemotherapy").
-Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B).
-The linear models then can be expressed as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-(\#eq:ANOVA-by-group)
-\end{equation}
-
-To further simplify the expression, substitute $\widetilde{\beta_{0}}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B.
-This substitution allows for a different intercept and slope for Groups A and B.
-The model is then written as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-(\#eq:ANOVA-lines)
-\end{equation}
-
-Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
-
-### The Linear Mixed Model Case
-
-A linear mixed model (LMEM) is a class of statistical model that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to distinguish [@pinheiro2006;@west2014].
-A LMEM with interaction between time and treatment for a longitudinal study  can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
-(\#eq:LMEM)
-\end{equation}
-
-When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
-This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject within each group.
-The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the "global noise" term $\varepsilon_{ijt}$ from Equation \@ref(eq:linear-model).
-
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
-As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
-The _random effect_ $\mu_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be independently normally distributed with mean zero and variance $\sigma^2_\mu$, which can be expressed as $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
-In essence,the _random effect_ in a LMEM enables to fit models with different slopes at the subject-level[@pinheiro2006].
-However,the expected linear relationship of the covariates and the response in Equation \@ref(eq:linear-model) and in Equation \@ref(eq:LMEM) is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
-
-## Covariance in rm-ANOVA and LMEMs
-
-In a longitudinal study there is an expected _covariance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
-The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
-
-In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity").
-Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)[@huynh1976;@greenhouse1959] the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others [@pinheiro2006].
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself.
-Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
-
-## Missing observations
-
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004].
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002].
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
-
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014].
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.
-
-This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and  missing data.
-In particular, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data.
-To better convey the issues of linearity and correlation  in linear models fitted to non-linear data, simulation is used in the next section.
-
-## How does an rm-ANOVA fit looks like? A visual representation using simulation {#simulation}
-
-To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme].
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C).
-In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity.
-
-Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates an effect due to treatment.
-In other words, Group 1 can be thought as a "Control" group and Group 2 as a "Treatment" group.
-From the mean response per group (linear or quadratic), the variability or "error" of individual responses  within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \@ref(fig:l-q-response) B,D).
-A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \@ref(fig:linear-cases) and Figure \@ref(fig:quadratic-cases) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-
-```{r, message = FALSE,include=FALSE}
-library(patchwork)
-library(tidyverse)
-library(mvnfast)
-library(nlme)
-library(ggsci)
-library(mgcv)
-library(gratia)
-
-knitr::opts_chunk$set(fig.pos = "!h", out.extra = "")
-```
-
-```{r,include=FALSE,message=FALSE,echo=FALSE}
-## Example with linear response
-example <- function(n_time = 6, 
-                    fun_type = "linear", 
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  x <- seq(1,6, length.out = n_time)
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # nonlinear response
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-  # matplot(mu, type = 'l')
-
-  y <- array(0, dim = c(length(x), 2, 10))
-  errors <- array(0, dim = c(length(x), 2, 10))
-
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(mu) <- list(time = x, treatment = 1:2)
-  dat <- as.data.frame.table(y, responseName = "y")
-  dat_errors <- as.data.frame.table(errors, responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, responseName = "mu")
-  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
-  dat$time <- as.numeric(as.character(dat$time))
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  ## repeated measures ANOVA in R
-  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
-  dat$preds_lm <- predict(fit_lm)
-
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  dat$y_pred <- predict(fit_lme)
-
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_lm = fit_lm,
-    fit_lme = fit_lme
-
-  ))
-}
-
-plot_example <- function(sim_dat) {
-  library(patchwork)
-  ## Plot the simulated data
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE) +labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
-    theme_classic() +
-    #ggtitle("Simulated data with true response function")+
-    theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    #ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_classic()+
-     theme(plot.title = element_text(size = 18, 
-                                face = "bold"),
-     text=element_text(size=18))+
-    scale_color_aaas()
-
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='errors')+
-     theme_classic()+
-    # facet_wrap(~ treatment) +
-    #ggtitle("Simulated errors") +
-     theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point()+labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    #ggtitle("Fitted Model")+
-    theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  #return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-  return((p1+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-A<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
-
-```
-
-(ref:l-q-response-caption) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.
-
-```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='ref(l-q-response-caption)' }
-# linear response, correlated errors (subject effect)
-par(mar = c(0.5, 0.5, 0.5, 0.5))
-A/C+plot_annotation(tag_levels = 'A')
-```
-
-The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B).
-When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.
-
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response), C.
-Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \@ref(fig:l-q-response) A and B) .
-The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panel D in Figure \@ref(fig:l-q-response).
-
-A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\@ref(fig:l-q-response), D) indicates that the model is not capturing the changes within each group in a good way.
-Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
-Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D).
-If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response.
-Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
-
-This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-Although the model fitted to the simulated data was an rm-ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM.
-In the following section, we present generalized additive models (GAMs) as a data-driven  alternative method to analyze longitudinal non-linear data.
-
-# GAMs as a special case of Generalized Linear Models
-
-## GAMs and Basis Functions
-
-Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors[@nelder1972].
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case[@simpson2018;@wood2017;@hastie1987].
-A GAM model can be written as:
-
-\begin{equation}
-  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
-  (\#eq:GAM)
-\end{equation}
-
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual error.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
-In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.
-
-The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using _basis functions_ expansions of the covariates and by estimating random coefficients for these basis functions.
-A _basis_ is a set of functions that spans the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
-For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
-In the case of Equation \@ref(eq:GAM), the basis function is $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
-
-A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness [@wood2017;@simpson2018].
-Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\@ref(fig:l-q-response), C).
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-The timeline can be divided in equally spaced _knots_, each knot being a region where a different basis function will be used.
-Because there are six timepoints for this group, five knots can be used.
-The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots.
-In Panel A of Figure \@ref(fig:basis-plot), the four basis functions (and the intercept) are shown.
-Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the "wigliness" of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
-The penalty reduces the "wigliness" of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
-
-```{r,basis-functions-plot, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  dat<-subset(dat,treatment==1)
-  dat<-dat[,c('y','time')]
-gm<-gam(y~s(time,k=5),data=dat)
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-x_new <- (seq(0, max(dat$time), length.out = 100))
-y_pred <- predict(gm, newdata=dat)
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof)
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final spline by aggregating the basis functions
-
-spl<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(spl=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=as.factor(Basis)))+
-  geom_line(size=sz,show.legend=FALSE)+
-  geom_point(size=sz+1,show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,labeller = as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#penalized basis
-p12<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
-  geom_line(show.legend = FALSE,size=sz)+
-  geom_point(show.legend = FALSE,size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,labeller=as_labeller(basis_names))+
-  theme_classic()+
-  scale_color_aaas()
-
-#heatmap of the penalization coefficient
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
-  geom_tile(aes(color='black'),size=sz+1,show.legend = FALSE)+
-  geom_tile(size=sz+1)+
-  scale_fill_gradient(low = "white", high = "red")+
-  labs(x='Basis',y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),size=10,show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  scale_color_aaas()+
-  labs(y='Simulated \n response')+
-  geom_line(data=spl,aes(x=time,y=spl),color="#B15731",size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-
-```
-
-In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \@ref(fig:basis-plot), Panel C shows the resulting penalized basis functions.
-Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
-Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
-
-(ref:basis-plot-caption) Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.
-
-```{r,basis-plot,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap='(ref:basis-plot-caption)'}
-
-par(mar = c(2, 2, 2, 2))
-
-b_plot
-
-```
-
- \newpage
-
-# The analyisis of longitudinal biomedical data using GAMs
-
-The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA.
-This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using $\textsf{R}$ and the package `mgcv`.
-The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
-
-## Simulated data
-
-The simulated data is based on the reported longitudinal changes in oxygen saturation ($StO_2$) in subcutaneous tumors that appear in Figure 3(C) in [vishwanath2009].
-An increase in oxygen saturation in the "Treatment" group (chemotherapy) is observed through time, while a decrease is seen in the "Control" (saline) group.
-Following the reported trend, 10 normally distributed observations at each time point are generated with a standard deviation (SD) of 10% (matching the SD in the original paper).
-The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot), A and the inlet, respectively.
-
-## An interaction GAM model for longitudinal data
-
-An interaction model is typically the main interest in longitudinal biomedical data, as it takes into account the effect of treatment, time, and their combination.
-In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the _time_ effect to account for the repeated measures over time.
-Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in [@hefley2017].
-
-For the data in Figure \@ref(fig:sim-smooth-plot),A it is of interest to know the effect that each treatment has in oxygen saturation.
-The model then needs to incorporate independent smooths for _Group_ and _Day_, respectively.
-The main thing to consider is the model syntax to account for the fact that one of the variables is numeric ( _Day_ ) and that the other is a factor ( _Group_ ).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
-Using $\textsf{R}$ and the package `mgcv` the model syntax is:
-
-`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim`
-
-This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation `StO2_sim` is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
-The parametric term _Group_ is added to separate the effect of treatment, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $StO_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B).
-Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package `gratia`[@gratia].
-A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].
-
-One question might be "what is the fit that an rm-ANOVA model produces for the simulated data?".
-The rm-ANOVA model, which corresponds to \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot),C.
-This is a typical case of model misspecification: The slopes of each group ar different, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \@ref(fig:sim-smooth-plot),B) .
-
-```{r,sim-tumor-response,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
-
-## plot the mean response
-f1<-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
-    geom_line(size=1,show.legend = FALSE)+
-    geom_point(show.legend = FALSE)+
-  labs(y=expression(paste(StO[2],' (real)')))+
-  theme_classic()+
-  scale_color_aaas()+
-    scale_x_continuous(breaks=c(0,5,10))+
-    scale_y_continuous(breaks=c(0,40))+
-  plot_layout(tag_level = 'new')+
-  theme(
-    plot.background = element_rect(fill = "transparent", color = NA),
-    axis.text=element_text(size=14)
-  )
-
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-
-n <- 10 #number of observations
-sd <- 10 #approximate sd from paper
-set.seed(1) #set seed for reproducibility
-df <- 6
-dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group)) +
-    geom_point(show.legend=FALSE)+
-    stat_summary(aes(y = StO2_sim,
-                     group=Group), 
-                 fun=mean, geom="line",size=1,show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],'(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-  scale_color_aaas()+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
-
-# tumor_sim<-f1+f2+plot_layout(guides = 'collect')+
-#   plot_annotation(tag_levels='A')&
-#   ylim(c(0,75)) &
-#   theme(
-#      text=element_text(size=18),
-#      legend.title=element_blank(),
-#      legend.text = element_blank()
-#      )&
-#   scale_color_aaas()
-
-```
-
-```{r,smooth-plot,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
-#GAM for StO2
-
-gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
-            method='REML',
-            data  = dat_sim)
-
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-#creates a dataframe using the length of the covariates for the GAM
-gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
-                         subject=factor(rep(1:10)))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#adds the predictions to the grid and creates a confidence interval for GAM
-gam_predict<-gam_predict%>%
-    mutate(fit = predict(gam1,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(gam1, gam_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lme
-#lm_predict<-lm_predict%>%
-#   mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response',level=0),
- #         se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response',level=0))
-
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=gam_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'complete')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-#posthoc comparisons for the linear model
-library(multcomp)
-
-summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
-
-```
-
-Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
-Using the simulated $StO_2$ values from Figure (\@ref(fig:sim-smooth-plot), B), 40% of the total observations can be randomly deleted and the same GAM model with interaction used for the full simulated data can be fitted to the remaining observations.
-
-```{r,missing-data,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
-
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'missing')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      scale_color_aaas()+
-  scale_fill_aaas()
-
-```
-
-(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent smooths for each group, shaded regions are 95% confidence intervals.C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40% of its observations missing.
-Lines represent smooths for each group, shaded regions are 95% confidence intervals
-
-```{r,sim-smooth-plot,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}
-
-mult_plot<-f2+inset_element(
-  f1, left = 0.01, 
-  bottom = 0.5, 
-  right = 0.5, 
-  top = 1.0)+
-  f3+f4+f6+
-  #plot_layout(guides = 'collect')+
-   plot_annotation(tag_levels='A')&
-   ylim(c(-5,75)) &
-  theme(
-     text=element_text(size=18)
-     )&
-  scale_color_aaas()
-
-mult_plot
-
-```
-
-## Determination of significance in GAMs
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that timepoint (or timepoints).
-The absence of a _p-value_ in this case might seem odd, but when a confidence interval comparison is put into context the validity of its rationale becomes apparent.
-The major advantage of the significance estimation in GAMs  is that they allow to determine significance at a specific timepoint, rather than providing a single _p-value_ for the overall treatment effect.
-The estimation of siginificant differences in GAMs is covered in detail in the Appendix using simulated longitudinal data that follows previously reported trends of tumor response between multiple treatment groups [@vishwanath2009].
-
-```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-##Pairwise comparisons
-
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Treatment', 'Control'))
-
-smooth_diff <- function(model, newdata, f1, f2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    c1 <- grepl(f1, colnames(xp))
-    c2 <- grepl(f2, colnames(xp))
-    #r1 <- newdata[[var]] == f1
-    #r2 <- newdata[[var]] == f2
-    r1 <- with(newdata, Group == f1)
-    r2 <- with(newdata, Group == f2)
-    ## difference rows of xp for data from comparison
-    X <- xp[r1, ] - xp[r2, ]
-    ## zero out cols of X related to splines for other lochs
-    X[, ! (c1 | c2)] <- 0
-    ## zero out the parametric cols
-    X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(f1, f2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-}
-
-comp1<-smooth_diff(gam1,pdat,'Control','Treatment')
-
-comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
-
-c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5,fill='#993336') +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-  theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-#+
- # geom_rect(aes(color=interval))
-
-c1
-
-###for missing data
-comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
-comp_StO2_m <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp2))
-
-missing_plot<-ggplot(comp_StO2_m, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    labs(x = 'Days', 
-         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
-                              )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme_classic()+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
-```
-
-***
-
-# References
-
-<div id="refs"></div>
-
-\newpage
-
-# (APPENDIX) Appendix {-}
-
-# Simulation
-
-## Compound symmetry and independent errors in linear and quadratic responses
-
-This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation).
-The linear simulations using Figure \@ref(fig:linear-cases) show in  panels A and D the simulated mean responses and individual datapoints.
-Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-```{r,include=TRUE,message=FALSE,echo=TRUE}
-## Example with linear response
-example <- function(n_time = 6, 
-                    fun_type = "linear", 
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  library(tidyverse)
-  library(mvnfast)
-  library(nlme)
-  library(ggsci)
-
-  x <- seq(1,6, length.out = n_time)
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # nonlinear response
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-  # matplot(mu, type = 'l')
-
-  y <- array(0, dim = c(length(x), 2, 10))
-  errors <- array(0, dim = c(length(x), 2, 10))
-
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(mu) <- list(time = x, treatment = 1:2)
-  dat <- as.data.frame.table(y, responseName = "y")
-  dat_errors <- as.data.frame.table(errors, responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, responseName = "mu")
-  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
-  dat$time <- as.numeric(as.character(dat$time))
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
-  ## repeated measures ANOVA in R
-  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
-  dat$preds_lm <- predict(fit_lm)
-
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  dat$y_pred <- predict(fit_lme)
-
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_lm = fit_lm,
-    fit_lme = fit_lme
-
-  ))
-}
-
-plot_example <- function(sim_dat) {
-  library(patchwork)
-  ## Plot the simulated data
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE) +labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
-    theme_classic() +
-    #ggtitle("Simulated data with true response function")+
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    #ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                face = "bold"),
-     text=element_text(size=30))+
-    scale_color_aaas()
-
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='errors')+
-     theme_classic()+
-    # facet_wrap(~ treatment) +
-    #ggtitle("Simulated errors") +
-     theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point()+labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    #ggtitle("Fitted Model")+
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    scale_color_aaas()
-
-  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
-
-```
-
-```{r, linear-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data**"}
-# linear response, correlated errors (subject effect)
-A1/B1+plot_annotation(tag_levels = 'A')
-```
-
-For the quadratic response case, Figure \@ref(fig:quadratic-cases) shows the simulated responses using compound symmetry and independent errors.
-
-```{r, quadratic-cases, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
-# linear response, correlated errors (subject effect)
-C1/D1+plot_annotation(tag_levels = 'A')
-```

### 2021-05-14 Manuscripts/Manuscript by chapters -v1/02-Challenges.Rmd
Changed folder name, working on Appendix, need to create figures and captions for the GAMs and models

--- 2021-05-13 Manuscripts/Bookdown testing/02-Challenges.Rmd
@@
 
-## What does an rm-ANOVA fit looks like? A visual representation using simulated data {#simulation}
+## What does an rm-ANOVA fit looks like? A visual representation using simulated data{#simulation}
 
@@
 Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \@ref(fig:l-q-response) B,D).
-A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \@ref(fig:linear-cases) and Figure \@ref(fig:quadratic-cases) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal appears in Figure \@ref(fig:linear-cases-Appendix) and Figure \@ref(fig:quadratic-cases-Appendix) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
 
@@
 ## Example with linear response. This function generates either linear or quadratic mean responses with correlated or uncorrelated errors and fits a linear model to the data.
-example <- function(n_time = 6, 
-                    fun_type = "linear", 
+example <- function(n_time = 6, #number of time points
+                    fun_type = "linear", #type of reponse
                     error_type = "correlated") {
@@
   x <- seq(1,6, length.out = n_time)
+
+  #Create mean response matrix: linear or quadratic
   mu <- matrix(0, length(x), 2)
@@
   } else {
-    # nonlinear response
+    # quadratic response (non-linear)
 
@@
   }
-  # matplot(mu, type = 'l')
 
+  #create an array where individual observations per each time point for each group are to be stored. Currently using 10 observations per timepoint
   y <- array(0, dim = c(length(x), 2, 10))
+
+  #Create array to store the "errors" for each group at each timepoint. The "errors" are the 
+  #between-group variability in the response.
   errors <- array(0, dim = c(length(x), 2, 10))
+  #create an array where 10 observations per each time point for each group are to be stored
 
+  #The following cycles create independent or correlated responses.
+To each value of mu (mean response per group) a randomly generated error (correlated or uncorrelated) is added and thus the individual response is created.
   if (error_type == "independent") {
@@
       for (j in 1:10) {  # number of subjects
-        # compound symmetry errors
+        # compound symmetry errors: variance covariance matrix
         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
@@
 
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
-  dimnames(mu) <- list(time = x, treatment = 1:2)
-  dat <- as.data.frame.table(y, responseName = "y")
-  dat_errors <- as.data.frame.table(errors, responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, responseName = "mu")
-  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
+#labelling y and errors  
+  dimnames(y) <- list(time = x, 
+                      treatment = 1:2, 
+                      subject = 1:10)
+
+  dimnames(errors) <- list(time = x, 
+                           treatment = 1:2, 
+                           subject = 1:10)
+
+  #labeling the mean response
+  dimnames(mu) <- list(time = x, 
+                       treatment = 1:2)
+
+  #convert y, mu and errors to  dataframes with time, treatment and subject columns
+  dat <- as.data.frame.table(y, 
+                             responseName = "y")
+  dat_errors <- as.data.frame.table(errors, 
+                                    responseName = "errors")
+  dat_mu <- as.data.frame.table(mu, 
+                                responseName = "mu")
+
+  #join the dataframes to show mean response and errors per subject
+  dat <- left_join(dat, dat_errors, 
+                   by = c("time", "treatment", "subject"))
+  dat <- left_join(dat, dat_mu, 
+                   by = c("time", "treatment"))
+  #add time
   dat$time <- as.numeric(as.character(dat$time))
+  #label subjects per group
   dat <- dat %>%
-    mutate(subject = factor(paste(subject, treatment, sep = "-")))
+    mutate(subject = factor(paste(subject, 
+                                  treatment, 
+                                  sep = "-")))
 
   ## repeated measures ANOVA in R
-  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
-  dat$preds_lm <- predict(fit_lm)
-
+#time and treatment interaction model, compound symmetry required by the model
   fit_lme <- lme(y ~ treatment + time + treatment:time,
@@
 
+  #create a prediction frame where the model can be used for plotting purposes
   pred_dat <- expand.grid(
@@
 
+  #add model predictions to the dataframe that has the simulated data
   dat$y_pred <- predict(fit_lme)
 
+  #return everything in a list
   return(list(
@@
     pred_dat = pred_dat,
-    fit_lm = fit_lm,
     fit_lme = fit_lme
@@
 }
+##################Section for plotting#################################
+#######################################################################
+#This function will create the plots for either a "linear" or "quadratic" response
 
 plot_example <- function(sim_dat) {
-  library(patchwork)
-  ## Plot the simulated data
+  ## Plot the simulated data (scatterplot)
   p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE,size=1.5,alpha=0.5) +
+    ggplot(aes(x = time, 
+               y = y, 
+               group = treatment, 
+               color = treatment)
+           ) +
+    geom_point(show.legend=FALSE) +
     labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment,size=1),show.legend=FALSE) +
+    geom_line(aes(x = time, 
+                  y = mu, 
+                  color = treatment),
+              show.legend=FALSE) +
     theme_classic() +
-    #ggtitle("Simulated data with true response function")+
-    theme(plot.title = element_text(size = 18, 
+    theme(plot.title = element_text(size = 30, 
                                   face = "bold"),
-        text=element_text(size=18))+
+        text=element_text(size=30))+
     scale_color_aaas()
 
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"),
-              lty = 1,show.legend = FALSE) +labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    #ggtitle("Simulated data\nIndividual responses with population mean") +
-    theme_classic()+
-     theme(plot.title = element_text(size = 18, 
+   #plot the model predictions
+  p4 <- ggplot(sim_dat$dat, 
+               aes(x = time, 
+                   y = y, 
+                   color = treatment)) +
+    geom_point()+
+    labs(y='response')+
+    geom_line(aes(y = predict(sim_dat$fit_lme), 
+                  group = subject, size = "Subjects")) +
+    geom_line(data = sim_dat$pred_dat, 
+              aes(y = predict(sim_dat$fit_lme, 
+                              level = 0, 
+                              newdata = sim_dat$pred_dat), 
+                  size = "Population")) +
+    scale_size_manual(name = "Predictions", 
+                      values=c("Subjects" = 0.5, "Population" = 3)) +
+    theme_classic() +
+    theme(plot.title = element_text(size = 30, 
                                   face = "bold"),
-     text=element_text(size=18))+
+        text=element_text(size=30))+
     scale_color_aaas()
-
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend=FALSE) +labs(y='errors')+
-     theme_classic()+
-     theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point(size=1.5,alpha=0.5)+labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
-    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    #ggtitle("Fitted Model")+
-    theme(plot.title = element_text(size = 18, 
-                                  face = "bold"),
-        text=element_text(size=18))+
-    scale_color_aaas()
-
-  #return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
 
@@
 
-B<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
 C<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
 
@@
 
-```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='ref(l-q-response-caption)' }
+```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='(ref:l-q-response-caption)'}
 # linear response, correlated errors (subject effect)

### 2021-05-14 Manuscripts/Manuscript by chapters -v1/03-GAM Theory.Rmd
Changed folder name, working on Appendix, need to create figures and captions for the GAMs and models

--- 2021-05-13 Manuscripts/Bookdown testing/03-GAM Theory.Rmd
@@
 ```{r,basis-functions-plot, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-#generate the response
 n_time = 6
@@
  }
+
+ #label each table
   dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
@@
  dimnames(mu) <- list(time = x, treatment = 1:2)
+
+ #Convert to dataframes with subject, time and group columns
  dat <- as.data.frame.table(y, responseName = "y")
@@
 
- #add subject column
+ #label subject per group
  dat <- dat %>%
@@
 
+ #extract  "Group 1" to fit the GAM
   dat<-subset(dat,treatment==1)
+ #keep just the response and timepoint columns
    dat<-dat[,c('y','time')]
+
+   #GAM model of time, 5 knots
 gm<-gam(y~s(time,k=5),data=dat)
+
+#model_matrix (also known as) 'design matrix'
+#will contain the smooths used to create  model 'gm'
 model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-x_new <- (seq(0, max(dat$time), length.out = 100))
-y_pred <- predict(gm, newdata=dat)
 
@@
 
-basis<-model_matrix[1:6,] #extracting basis
+basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
 #basis<-model_matrix[1:6,-1] #extracting basis
@@
 basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof)
+  mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be 
+#multiplied by the coefficients
 
@@
 
-#calculating the final spline by aggregating the basis functions
+#calculating the final smooth by aggregating the basis functions
 
-spl<-basis_plot%>% 
+smooth<-basis_plot%>% 
   group_by(time)%>%
-  summarize(spl=sum(mod_val))
+  summarize(smooth=sum(mod_val))
 
@@
 sz<-1
-p11<-ggplot(basis_plot,aes(x=time,y=value_orig,colour=as.factor(Basis)))+
-  geom_line(size=sz,show.legend=FALSE)+
-  geom_point(size=sz+1,show.legend = FALSE)+
+p11<-ggplot(basis_plot,
+            aes(x=time,
+                y=value_orig,
+                colour=as.factor(Basis)
+                )
+            )+
+  geom_line(size=sz,
+            show.legend=FALSE)+
+  geom_point(size=sz+1,
+             show.legend = FALSE)+
   labs(y='Basis functions')+
-  facet_wrap(~Basis,labeller = as_labeller(basis_names))+
+  facet_wrap(~Basis,
+             labeller = as_labeller(basis_names)
+             )+
   theme_classic()+
@@
 #penalized basis
-p12<-ggplot(basis_plot,aes(x=time,y=mod_val,colour=as.factor(Basis)))+
-  geom_line(show.legend = FALSE,size=sz)+
-  geom_point(show.legend = FALSE,size=sz+1)+
+p12<-ggplot(basis_plot,
+            aes(x=time,
+                y=mod_val,
+                colour=as.factor(Basis)
+                )
+            )+
+  geom_line(show.legend = FALSE,
+            size=sz)+
+  geom_point(show.legend = FALSE,
+             size=sz+1)+
   labs(y='Penalized \n basis functions')+
   scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,labeller=as_labeller(basis_names))+
+  facet_wrap(~Basis,
+             labeller=as_labeller(basis_names)
+             )+
   theme_classic()+
@@
 x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,aes(x=Basis,y=Basis,fill=cof))+
-  geom_tile(aes(color='black'),size=sz+1,show.legend = FALSE)+
+p13<-ggplot(basis_plot,
+            aes(x=Basis,
+                y=Basis,
+                fill=cof))+
+  geom_tile(aes(color='black'),
+            size=sz+1,
+            show.legend = FALSE)+
   geom_tile(size=sz+1)+
   scale_fill_gradient(low = "white", high = "red")+
-  labs(x='Basis',y='Basis')+
+  labs(x='Basis',
+       y='Basis')+
   scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),size=10,show.legend = FALSE)+
+  geom_text(aes(label=round(cof,2)),
+            size=10,
+            show.legend = FALSE)+
   theme_classic()+
@@
 #plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,aes(x=time,y=y))+
+p14<-ggplot(data=dat,
+            aes(x=time,y=y))+
   geom_point(size=sz+1)+
@@
   labs(y='Simulated \n response')+
-  geom_line(data=spl,aes(x=time,y=spl),color="#B15731",size=sz+1)+
+  geom_line(data=smooth,
+            aes(x=time,
+                y=smooth),
+            color="#B15731",
+            size=sz+1)+
   theme_classic()

### 2021-05-14 Manuscripts/Manuscript by chapters -v1/04-Longitudinal analysis with GAMs.Rmd
Changed folder name, working on Appendix, need to create figures and captions for the GAMs and models

--- 2021-05-13 Manuscripts/Bookdown testing/04-Longitudinal analysis with GAMs.Rmd
@@
 
-# The analyisis of longitudinal biomedical data using GAMs
+# The analyisis of longitudinal biomedical data using GAMs{#longitudinal-GAMs}
 
@@
 ## plot the mean response
-f1<-ggplot(dat, aes(x = Day, y = StO2, color = Group)) +
-    geom_line(size=1,show.legend = FALSE)+
-    geom_point(show.legend = FALSE,size=1.5,alpha=0.5)+
-  labs(y=expression(paste(StO[2],' (real)')))+
+f1<-ggplot(dat, 
+           aes(x = Day, 
+               y = StO2, 
+               color = Group)) +
+    geom_line(size=1,
+              show.legend = FALSE)+
+    geom_point(show.legend = FALSE,
+               size=1.5,
+               alpha=0.5)+
+  labs(y=expression(paste(StO[2],
+                          ' (real)')))+
   theme_classic()+
@@
   theme(
-    plot.background = element_rect(fill = "transparent", color = NA),
+    plot.background = element_rect(fill = "transparent", 
+                                   color = NA),
     axis.text=element_text(size=14)

### 2021-05-14 Manuscripts/Manuscript by chapters -v1/05-Conclusion.Rmd
Changed folder name, working on Appendix, need to create figures and captions for the GAMs and models

# Conclusion

The traditional methods to analyze biomedical longitudinal data (rm-ANOVA, LMEMs) are restrictive when the data does not follow a linear pattern and their use in such situations leads to unreliable inference and biased estimates. GAMs present a suitable alternative to analyze non-linear longitudinal biomedical data, as they overcome the limitations of linearity, compound symmetry, and missing observations imposed by traditional methods. By presenting the implementation of GAMs using simulated data that follows previously reported trends in the literature, we aim at encouraging biomedical researchers to consider different statistical tools to analyze longitudinal data. Finally, by providing the data and code used in this paper we hope to address the need of creating and sharing reproducible work in biomedical research.

### 2021-05-14 Manuscripts/Manuscript by chapters -v1/06-Appendix.Rmd
Changed folder name, working on Appendix, need to create figures and captions for the GAMs and models

# (APPENDIX) Appendix {-}

# Simulation

## Compound symmetry and independent errors in linear and quadratic responses

This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation). The linear simulations using Figure \@ref(fig:linear-cases-Appendix) show in  panels A and D the simulated mean responses and individual data points.  Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response. In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.  Panels D and H show the predictions from the rm-ANOVA model.

The following code produces a more comprehensive exploration of Figure \@ref(fig:l-q-response) in the main manuscript.

```{r, responses-Appendix,include=TRUE,message=FALSE,echo=TRUE}

####################Section for calculations######################################
################################################################################

## Example with linear response

#This function simulates data using a linear or quadratic mean response and each with correlated
#or uncorrelated errors. Each group has a different slope/concavity.
example <- function(n_time = 6, #number of time points
                    fun_type = "linear", #type of response
                    error_type = "correlated") {

  if (!(fun_type %in% c("linear", "quadratic")))
    stop('fun_type must be either "linear", or "quadratic"')
  if (!(error_type %in% c("correlated", "independent")))
    stop('fun_type must be either "correlated", or "independent"')

  x <- seq(1,6, length.out = n_time)

  #Create mean response matrix: linear or quadratic
  mu <- matrix(0, length(x), 2)
  # linear response
  if (fun_type == "linear") {
    mu[, 1] <- - (0.25*x)+2  
    mu[, 2] <- 0.25*x+2
  } else {
    # quadratic response (non-linear)

    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
  }

  #create an array where individual observations per each time point for each group are to be stored. Currently using 10 observations per timepoint
  y <- array(0, dim = c(length(x), 2, 10))

  #Create array to store the "errors" for each group at each timepoint. The "errors" are the 
  #between-group variability in the response.
  errors <- array(0, dim = c(length(x), 2, 10))
  #create an array where 10 observations per each time point for each group are to be stored

  #The following cycles create independent or correlated responses. To each value of mu (mean response per group) a randomly generated error (correlated or uncorrelated) is added and thus the individual response is created.
  if (error_type == "independent") {
    ## independent errors
    for (i in 1:2) {
      for (j in 1:10) {
        errors[, i, j] <- rnorm(6, 0, 0.25)
        y[, i, j] <- mu[, i] + errors[, i, j]
      }
    }
  } else {
    for (i in 1:2) {     # number of treatments
      for (j in 1:10) {  # number of subjects
        # compound symmetry errors: variance covariance matrix
        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
        y[, i, j] <- mu[, i] + errors[, i, j]
      }
    }
  }

  ## subject random effects

  ## visualizing the difference between independent errors and compound symmetry
  ## why do we need to account for this -- overly confident inference

#labelling y and errors  
  dimnames(y) <- list(time = x, 
                      treatment = 1:2, 
                      subject = 1:10)

  dimnames(errors) <- list(time = x, 
                           treatment = 1:2, 
                           subject = 1:10)

  #labeling the mean response
  dimnames(mu) <- list(time = x, 
                       treatment = 1:2)

  #convert y, mu and errors to  dataframes with time, treatment and subject columns
  dat <- as.data.frame.table(y, 
                             responseName = "y")
  dat_errors <- as.data.frame.table(errors, 
                                    responseName = "errors")
  dat_mu <- as.data.frame.table(mu, 
                                responseName = "mu")

  #join the dataframes to show mean response and errors per subject
  dat <- left_join(dat, dat_errors, 
                   by = c("time", "treatment", "subject"))
  dat <- left_join(dat, dat_mu, 
                   by = c("time", "treatment"))
  #add time
  dat$time <- as.numeric(as.character(dat$time))
  #label subjects per group
  dat <- dat %>%
    mutate(subject = factor(paste(subject, 
                                  treatment, 
                                  sep = "-")))

  ## repeated measures ANOVA in R
#time and treatment interaction model, compound symmetry required by the model
  fit_lme <- lme(y ~ treatment + time + treatment:time,
                 data = dat,
                 random = ~ 1 | subject,
                 correlation = corCompSymm(form = ~ 1 | subject)
  )

  #create a prediction frame where the model can be used for plotting purposes
  pred_dat <- expand.grid(
    treatment = factor(1:2), 
    time = unique(dat$time)
  )

  #add model predictions to the dataframe that has the simulated data
  dat$y_pred <- predict(fit_lme)

  #return everything in a list
  return(list(
    dat = dat,
    pred_dat = pred_dat,
    fit_lme = fit_lme

  ))
}
##################Section for plotting#################################
#######################################################################
#This function will create the plots for either a "linear" or "quadratic" response

plot_example <- function(sim_dat) {
  ## Plot the simulated data (scatterplot)
  p1 <- sim_dat$dat %>%
    ggplot(aes(x = time, 
               y = y, 
               group = treatment, 
               color = treatment)
           ) +
    geom_point(show.legend=FALSE) +
    labs(y='response')+
    geom_line(aes(x = time, 
                  y = mu, 
                  color = treatment),
              show.legend=FALSE) +
    theme_classic() +
    theme(plot.title = element_text(size = 30, 
                                  face = "bold"),
        text=element_text(size=30))+
    scale_color_aaas()

  #plot the simulated data with trajectories per each subject
  p2 <- sim_dat$dat %>%
    ggplot(aes(x = time, 
               y = y, 
               group = subject, 
               color = treatment)
           ) +
    geom_line(aes(size = "Subjects"),
              show.legend = FALSE) +
    # facet_wrap(~ treatment) +
    geom_line(aes(x = time, 
                  y = mu, 
                  color = treatment,
                  size = "Simulated Truth"), 
              lty = 1,show.legend = FALSE) +
    labs(y='response')+
    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
    theme_classic()+
     theme(plot.title = element_text(size = 30, 
                                face = "bold"),
     text=element_text(size=30))+
    scale_color_aaas()

  #plot the errors
   p3 <- sim_dat$dat %>%
    ggplot(aes(x = time, 
               y = errors,
               group = subject, 
               color = treatment)) +
    geom_line(show.legend=FALSE) +
     labs(y='errors')+
     theme_classic()+
     theme(plot.title = element_text(size = 30, 
                                  face = "bold"),
        text=element_text(size=30))+
    scale_color_aaas()

   #plot the model predictions
  p4 <- ggplot(sim_dat$dat, 
               aes(x = time, 
                   y = y, 
                   color = treatment)) +
    geom_point()+
    labs(y='response')+
    geom_line(aes(y = predict(sim_dat$fit_lme), 
                  group = subject, size = "Subjects")) +
    geom_line(data = sim_dat$pred_dat, 
              aes(y = predict(sim_dat$fit_lme, 
                              level = 0, 
                              newdata = sim_dat$pred_dat), 
                  size = "Population")) +
    scale_size_manual(name = "Predictions", 
                      values=c("Subjects" = 0.5, "Population" = 3)) +
    theme_classic() +
    theme(plot.title = element_text(size = 30, 
                                  face = "bold"),
        text=element_text(size=30))+
    scale_color_aaas()

  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))

}

txt<-18

#Store each plot in a separate object
A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))

B1<-plot_example(example(fun_type = "linear", error_type = "independent"))

C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))

D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"))

```

(ref:linear-cases-appendix) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model. A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data. B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic). Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data. The rm-ANOVA model does not pick the trend of the quadratic data.

```{r, linear-cases-Appendix, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:linear-cases-appendix)'}
# linear response, correlated errors (subject effect)
A1/B1+plot_annotation(tag_levels = 'A')
```

For the quadratic response case, Figure \@ref(fig:quadratic-cases-Appendix) shows the simulated responses using compound symmetry and independent errors.

(ref:quadratic-caption-appendix) Simulated quadratic responses from two groups with a rm-ANOVA model fitted. A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data. B,F: Generated errors showing the difference in the behavior of correlated and independent errors.  C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline. D,H: Estimations from the rm-ANOVA model for the mean group response. Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.

```{r, quadratic-cases-Appendix, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:quadratic-caption-appendix)'}
# linear response, correlated errors (subject effect)
C1/D1+plot_annotation(tag_levels = 'A')
```

## Basis functions and GAMs

This code produces Figure \@ref(fig:basis-plot) from the main manuscript. Briefly, a non-linear (quadratic) response is simulated a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed. The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.

```{r,basis-Appendix, echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
#generate the response: the same initial procedure from the previous section to simulate
#the response
n_time = 6
 x <- seq(1,6, length.out = n_time)
 mu <- matrix(0, length(x), 2)
 mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
 mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
 y <- array(0, dim = c(length(x), 2, 10))
 errors <- array(0, dim = c(length(x), 2, 10))
 for (i in 1:2) {     # number of treatments
     for (j in 1:10) {  # number of subjects
         # compound symmetry errors
         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
         y[, i, j] <- mu[, i] + errors[, i, j]
     }
 }

 #label each table
  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
 dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
 dimnames(mu) <- list(time = x, treatment = 1:2)

 #Convert to dataframes with subject, time and group columns
 dat <- as.data.frame.table(y, responseName = "y")
 dat_errors <- as.data.frame.table(errors, responseName = "errors")
 dat_mu <- as.data.frame.table(mu, responseName = "mu")
 dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
 dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
 dat$time <- as.numeric(as.character(dat$time))

 #label subject per group
 dat <- dat %>%
     mutate(subject = factor(paste(subject, treatment, sep = "-")))

 #extract  "Group 1" to fit the GAM
  dat<-subset(dat,treatment==1)
 #keep just the response and timepoint columns
   dat<-dat[,c('y','time')]

   #GAM model of time, 5 knots
gm<-gam(y~s(time,k=5),data=dat)

#model_matrix (also known as) 'design matrix'
#will contain the smooths used to create  model 'gm'
model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))

time<-c(1:6)

basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
#basis<-model_matrix[1:6,-1] #extracting basis
colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
basis<-basis %>% #pivoting to long format
  pivot_longer(
    cols=starts_with("s")
  )%>%
  arrange(name) #ordering

#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
ln<-6*(length(coef(gm)))

basis_plot<-data.frame(Basis=integer(ln),
                       value_orig=double(ln),
                       time=integer(ln),
                       cof=double(ln)
)

basis_plot$time<-rep(time) #pasting timepoints
basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
basis_plot$value_orig<-basis$value #pasting basis values
basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
basis_plot<-basis_plot%>%
  mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be 
#multiplied by the coefficients

#creating labeller to change the labels in the basis plots

basis_names<-c(
  `1`="Intercept",
  `2`="1",
  `3`="2",
  `4`="3",
  `5`="4"
)

#calculating the final smooth by aggregating the basis functions

smooth<-basis_plot%>% 
  group_by(time)%>%
  summarize(smooth=sum(mod_val))

#original basis
sz<-1
p11<-ggplot(basis_plot,
            aes(x=time,
                y=value_orig,
                colour=as.factor(Basis)
                )
            )+
  geom_line(size=sz,
            show.legend=FALSE)+
  geom_point(size=sz+1,
             show.legend = FALSE)+
  labs(y='Basis functions')+
  facet_wrap(~Basis,
             labeller = as_labeller(basis_names)
             )+
  theme_classic()+
  scale_color_aaas()

#penalized basis
p12<-ggplot(basis_plot,
            aes(x=time,
                y=mod_val,
                colour=as.factor(Basis)
                )
            )+
  geom_line(show.legend = FALSE,
            size=sz)+
  geom_point(show.legend = FALSE,
             size=sz+1)+
  labs(y='Penalized \n basis functions')+
  scale_y_continuous(breaks=seq(-1,1,1))+
  facet_wrap(~Basis,
             labeller=as_labeller(basis_names)
             )+
  theme_classic()+
  scale_color_aaas()

#heatmap of the penalization coefficient
x_labels<-c("Intercept","1","2","3","4")
p13<-ggplot(basis_plot,
            aes(x=Basis,
                y=Basis,
                fill=cof))+
  geom_tile(aes(color='black'),
            size=sz+1,
            show.legend = FALSE)+
  geom_tile(size=sz+1)+
  scale_fill_gradient(low = "white", high = "red")+
  labs(x='Basis',
       y='Basis')+
  scale_x_discrete(labels=x_labels)+
  geom_text(aes(label=round(cof,2)),
            size=10,
            show.legend = FALSE)+
  theme_classic()+
  theme(legend.title = element_blank())

#plotting simulated datapoints and smooth term
p14<-ggplot(data=dat,
            aes(x=time,y=y))+
  geom_point(size=sz+1)+
  scale_color_aaas()+
  labs(y='Simulated \n response')+
  geom_line(data=smooth,
            aes(x=time,
                y=smooth),
            color="#B15731",
            size=sz+1)+
  theme_classic()

#Combining all
b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
  theme(
     text=element_text(size=18)
     )
```

(ref:basis-caption-appendix) Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown. B: Penalty matrix for the basis functions. Each basis function is penalized by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Penalized basis functions. Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis. D: Smoother for time and original data points. The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.

```{r,basis-plot-appendix,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap='(ref:basis-caption-appendix)'}

par(mar = c(2, 2, 2, 2))

b_plot

```

## Data simulation and GAM models

This section describes how to fit GAMs to longitudinal data using simulated data. First, data is simulated according to Section \@ref(longitudinal-GAMs).

```{r,sim-tumor-response-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}

dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
            Day=rep(c(0,2,5,7,10),times=2),
            Group=as.factor(rep(c("Control","Treatment"),each=5))
            )

## plot the mean response
f1<-ggplot(dat, 
           aes(x = Day, 
               y = StO2, 
               color = Group)) +
    geom_line(size=1,
              show.legend = FALSE)+
    geom_point(show.legend = FALSE,
               size=1.5,
               alpha=0.5)+
  labs(y=expression(paste(StO[2],
                          ' (real)')))+
  theme_classic()+
  scale_color_aaas()+
    scale_x_continuous(breaks=c(0,5,10))+
    scale_y_continuous(breaks=c(0,40))+
  plot_layout(tag_level = 'new')+
  theme(
    plot.background = element_rect(fill = "transparent", 
                                   color = NA),
    axis.text=element_text(size=14)
  )

#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")

simulate_data <- function(dat, n = 10, sd = 5) {
    dat_sim <- dat %>%
        slice(rep(1:n(), each = n)) %>%
        group_by(Group, Day) %>%
        mutate(
               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
               subject=rep(1:10),
               subject=factor(paste(subject, Group, sep = "-"))
               ) %>%
        ungroup()

    return(dat_sim)
}

#subject = factor(paste(subject, treatment, sep = "-")))

n <- 10 #number of observations
sd <- 10 #approximate sd from paper
set.seed(1) #set seed for reproducibility
df <- 6
dat_sim <- simulate_data(dat, n, sd)

#plotting simulated data
f2<-ggplot(dat_sim, 
           aes(x = Day, 
               y = StO2_sim, 
               color = Group)) +
    geom_point(show.legend=FALSE,
               size=1.5,
               alpha=0.5)+
    stat_summary(aes(y = StO2_sim,
                     group=Group), 
                 fun=mean, geom="line",
                 size=1,
                 show.legend = FALSE)+
  labs(y=expression(atop(StO[2],
                         '(simulated)')))+
  theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
  scale_color_aaas()+
    scale_x_continuous(breaks=c(0,2,5,7,10))

```

### Workflow for GAMs

Next, a series of increasingly complex GAMs are fitted to the simulated data. The first model is a model that only accounts for different smooths by day. The model syntax specifies that `gam_00` is the object that will contain all the model information, and that the model attempts to explain changes in `StO2_sim` (simulated $\mbox{StO}_2$) using a smooth per `Day`. The model will use 5 knots (`k=5`) for the smooth. And that the smooth is constructed using gaussian process basis (`bs="gp"`). The smoothing parameter estimation method used is the restricted maximum likelihood (`REML`).

```{r, first-GAM, echo=TRUE, include=TRUE, message=FALSE,warning=FALSE}

gam_00<-gam(StO2_sim ~ s(Day, k = 5,bs="gp"),
            method='REML',
            data  = dat_sim)

```

To obtain model diagnostics, two methodologies can be used: 1) graphical diagnostics, and 2) a model check. In the first case, the function `appraise` from the package _gratia_ can be used to obtain a single plot will all the diagnostic information. For model check, the function `gam.check` from  _mgcv_ provides detailed information about the model.

#### Graphical diagnostics

```{r, first-GAM-diag, echo=TRUE, include=TRUE, message=FALSE,warning=FALSE}
#need to add figure number and caption
appr1<-appraise(gam_00)
sm1<-draw(gam_00)

visual_check<-appr1+sm1

visual_check
```

From the `appraise` plot, the major indicators concern about the model are the QQ plot of residuals and the histogram of residuals. The QQ plot shows that the errors are not reasonably located along the 45$^{\circ}$ line, as there are multiple points that deviate from the trend, which is more noticeable in the tails. The histogram also shows that the variation (residuals) is not following the assumption of a normal distribution.

The `draw` function permits to plot the smooths as `ggplot2` objects. Because this model specifies only one smooth for the time covariate (Day) the plot only contains only one smooth. Note that the smooth show an almost linear profile.

#### Model check
```{r, first-GAM-check, echo=TRUE, include=TRUE, message=FALSE,warning=FALSE}
#need to add figure number and caption
gam.check(gam_00)

summary(gam_00)

```

Special attention must be paid to the 'k-index' from `gam.check`. This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data, which is indicated by a low (<1) value. From the model, it can be seen that the k-index is effectively <0.3, which indicates that the model is not capturing the variability in the data. The 'edf' (effective degrees of freedom) is an indicator of the complexity of the smooth. Here the complexity of the smooth is comparable to that of a 4th degree polynomial.

From the `summary` function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained. The link function is 'identity' as the model does not make any transformation on the predictors. The 'significance of smooth terms' _p-value_ indicates if each smooth is adding significance to the model. Here, the _p-value_ is low but we have seen that there are issues with the model from the previous outputs. Finally, the 'deviance explained' indicates how much of the data the model is able to capture.

The major flaw of `gam_00` is that is not taking into account the fact that the data is nested in groups. The next iteration is a model where a different smooth of time (Day) is assigned for each group using `by=Group` in the model syntax.

```{r, second-GAM, echo=TRUE, include=TRUE, message=FALSE,warning=FALSE}

gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5,bs="gp"),
            method='REML',
            data  = dat_sim)
```

Diagnostics for this model indicate that the k-index is still below 1, and that the residuals are still not normally distributed. Moreover, if the smooths are plotted via the `draw()` function, it can be seen that their profile is fairly linear, which is not capturing the trends observed in the data. From `summary()`, the deviance explained by the model is ~43%.

As indicated in Section \@ref(longitudinal-GAMs), in order to differentiate between each group, a parametric term needs to be added to the model in order to account for the interaction of _Day_ and _Group_.

The resulting model is model `gam1` from the main manuscript. By running `appraise()` we see that the trend on the QQ plot has improved and that the histogram of the residuals appears to be reasonably distributed. From running `gam.check`, the k-index is now at an acceptable value (~1.02), and `summary` now indicates that the model is able to capture 87% of the data.

**Need to add the AIC comparison for the three models**

```{r,smooth-plot-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}

#GAM for StO2

gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
            method='REML',
            data  = dat_sim)

#linear model
lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)

#creates a dataframe using the length of the covariates for the GAM
gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = seq(0, 10, by = 0.1),
                         subject=factor(rep(1:10)))

#creates a dataframe using the length of the covariates for rm-ANOVA
lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = c(0:10),
                        subject=factor(rep(1:10)),
                          )
lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))

#adds the predictions to the grid and creates a confidence interval for GAM
gam_predict<-gam_predict%>%
    mutate(fit = predict(gam1,gam_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(gam1, gam_predict,se.fit = TRUE,type='response')$se.fit)

#using lm
lm_predict<-lm_predict%>%
    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)

#plot smooths and confidence interval for GAM
f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=0.3,
              data=gam_predict,
            show.legend=FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=gam_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'complete')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
      scale_color_aaas()+
  scale_fill_aaas()

#plot linear fit for rm-ANOVA
f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),fill=Group),
              alpha=0.3,
              data=lm_predict,
              show.legend = FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=lm_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(paste('StO'[2],' (simulated)')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
      scale_color_aaas()+
  scale_fill_aaas()

#posthoc comparisons for the linear model
library(multcomp)

#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))

```

```{r,missing-data-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
#missing data
#create a sequence of 40 random numbers between 1 and 100, these numbers will
#correspond to the row numbers to be randomly erased from the original dataset
missing <- sample(1:100, 40)

#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA

ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))

#create a new dataframe, remove the StO2 column
dat_missing <- dat_sim[,-1]

#add NAs at the ind positions
dat_missing$StO2_sim[ind]<-NA

#Count the number of remaining observations per day (original dataset had 10 per group per day)
dat_missing %>%
    group_by(Day,Group) %>%
    filter(!is.na(StO2_sim))%>%
  count(Day)

#the same model used for the full dataset
mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
#appraise the model
appraise(mod_m1)

m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = seq(0, 10, by = 0.1))

#adds the predictions to the grid and creates a confidence interval
m_predict<-m_predict%>%
    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)

f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=0.3,
              data=m_predict,
            show.legend=FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=m_predict,
              show.legend = TRUE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'missing')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
      scale_color_aaas()+
  scale_fill_aaas()

```

(ref:sim-plot-caption-Appendix) Simulated data and smooths for oxygen saturation in tumors. A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment. Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point. Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment. Lines represent trends for each group, shaded regions are 95% confidence intervals. C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time. D: Smooths for the GAM model for the simulated data with 40% of its observations missing. Lines represent trends for each group, shaded regions are 95% confidence intervals.

```{r,sim-smooth-plot-Appendix,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=TRUE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption-Appendix)'}

mult_plot<-f2+inset_element(
  f1, left = 0.01, 
  bottom = 0.5, 
  right = 0.5, 
  top = 1.0)+
  f3+f4+f6+
   plot_annotation(tag_levels='A')&
   ylim(c(-5,75)) &
  theme(
     text=element_text(size=18)
     )&
  scale_color_aaas()

mult_plot

```

### 2021-05-14 Manuscripts/Manuscript by chapters -v1/Full_document.Rmd
Changed folder name, working on Appendix, need to create figures and captions for the GAMs and models

--- 2021-05-13 Manuscripts/Bookdown testing/Full_document.Rmd
@@
   bookdown::pdf_document2:
+    pandoc_args: --listings
     keep_tex: yes
@@
 
-knitr::opts_chunk$set(fig.pos = "!h", out.extra = "",error=TRUE)
+knitr::opts_chunk$set(fig.pos = 'H', out.extra = "",error=TRUE)
 ```

### 2021-05-14 Manuscripts/Manuscript by chapters -v1/Full_document.tex
Changed folder name, working on Appendix, need to create figures and captions for the GAMs and models

--- 2021-05-13 Manuscripts/Bookdown testing/Full_document.tex
@@
 \usepackage[margin=1in]{geometry}
+\usepackage{listings}
+\newcommand{\passthrough}[1]{#1}
+\lstset{defaultdialect=[5.3]Lua}
+\lstset{defaultdialect=[x86masm]Assembler}
 \usepackage{longtable,booktabs}
@@
 \affil[2]{\footnotesize Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
+
+\lstset{
+  breaklines=true
+}
 \usepackage{placeins}
@@
 Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D).
-A more comprehensive exploration of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \ref{fig:linear-cases} and Figure \ref{fig:quadratic-cases} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-
-\begin{figure}[!h]
+A more comprehensive exploration of the fit of rm-ANOVA for linear and non-linear longitudinal appears in Figure \ref{fig:linear-cases-Appendix} and Figure \ref{fig:quadratic-cases-Appendix} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+
+\begin{figure}[H]
 
@@
 
-\caption{ref(l-q-response-caption)}\label{fig:l-q-response}
+\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
+B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
+The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.}\label{fig:l-q-response}
 \end{figure}
@@
 
-\begin{figure}[!h]
+\begin{figure}[H]
 
@@
 
-\hypertarget{the-analyisis-of-longitudinal-biomedical-data-using-gams}{%
-\section{The analyisis of longitudinal biomedical data using GAMs}\label{the-analyisis-of-longitudinal-biomedical-data-using-gams}}
+\hypertarget{longitudinal-GAMs}{%
+\section{The analyisis of longitudinal biomedical data using GAMs}\label{longitudinal-GAMs}}
 
 The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA.
-This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \texttt{mgcv}.
+This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}.
 The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
@@
 Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
-Using \(\textsf{R}\) and the package \texttt{mgcv} the model syntax is:
-
-\texttt{m1\textless{}-gam(StO2\_sim\textasciitilde{}Group+s(Day,by=Group,k=5,bs="gp"),\ method=\textquotesingle{}REML\textquotesingle{},data=dat\_sim}
-
-This syntax specifies that \texttt{m1} will store the model, and that the change in the simulated oxygen saturation (\texttt{StO2\_sim}) is modeled using independent smooths for \emph{Group} and \emph{Day} (the parenthesis preceded by \emph{s}) using 5 knots.
-The smooth is constructed using gaussian process smooths, which is indicated by \texttt{bs="gp"}.
+Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:
+
+\passthrough{\lstinline!m1<-gam(StO2\_sim\~Group+s(Day,by=Group,k=5,bs="gp"), method='REML',data=dat\_sim!}
+
+This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths for \emph{Group} and \emph{Day} (the parenthesis preceded by \emph{s}) using 5 knots.
+The smooth is constructed using gaussian process smooths, which is indicated by \passthrough{\lstinline!bs="gp"!}.
 These splines are used to model temporal trends and might be particularly suited for long-term studies where the correlation between measurements changes as a function of the time intervals {[}34{]}.
-The parametric term \emph{Group} is added to quantify differences in the effect of treatment between groups, and the \texttt{method} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}37{]}.
+The parametric term \emph{Group} is added to quantify differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}37{]}.
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot},B).
-Model diagnostics can be obtained using the \texttt{gam.check} function, and the function \texttt{appraise} from the package \emph{gratia}{[}54{]}.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}{[}54{]}.
 A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in {[}37{]} and {[}55{]}.
@@
 \section{Conclusion}\label{conclusion}}
+
+The traditional methods to analyze biomedical longitudinal data (rm-ANOVA, LMEMs) are restrictive when the data does not follow a linear pattern and their use in such situations leads to unreliable inference and biased estimates.
+GAMs present a suitable alternative to analyze non-linear longitudinal biomedical data, as they overcome the limitations of linearity, compound symmetry, and missing observations imposed by traditional methods.
+By presenting the implementation of GAMs using simulated data that follows previously reported trends in the literature, we aim at encouraging biomedical researchers to consider different statistical tools to analyze longitudinal data.
+Finally, by providing the data and code used in this paper we hope to address the need of creating and sharing reproducible work in biomedical research.
 
@@
 This section simulated linear and quadratic data in the same manner as in Section \ref{simulation}.
-The linear simulations using Figure \ref{fig:linear-cases-appendix} show in panels A and D the simulated mean responses and individual datapoints.
+The linear simulations using Figure \ref{fig:linear-cases-Appendix} show in panels A and D the simulated mean responses and individual data points.
 Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
@@
 
-\begin{verbatim}
+The following code produces a more comprehensive exploration of Figure \ref{fig:l-q-response} in the main manuscript.
+
+\begin{lstlisting}[language=R]
+####################Section for calculations######################################
+################################################################################
+
 ## Example with linear response
-example <- function(n_time = 6, 
-                    fun_type = "linear", 
+
+#This function simulates data using a linear or quadratic mean response and each with correlated
+#or uncorrelated errors. Each group has a different slope/concavity.
+example <- function(n_time = 6, #number of time points
+                    fun_type = "linear", #type of response
                     error_type = "correlated") {
@@
 
-  library(tidyverse)
-  library(mvnfast)
-  library(nlme)
-  library(ggsci)
-
   x <- seq(1,6, length.out = n_time)
+
+  #Create mean response matrix: linear or quadratic
   mu <- matrix(0, length(x), 2)
@@
   }
-  # matplot(mu, type = 'l')
-
+
+  #create an array where individual observations per each time point for each group are to be stored. Currently using 10 observations per timepoint
   y <- array(0, dim = c(length(x), 2, 10))
+
+  #Create array to store the "errors" for each group at each timepoint. The "errors" are the 
+  #between-group variability in the response.
   errors <- array(0, dim = c(length(x), 2, 10))
-
+  #create an array where 10 observations per each time point for each group are to be stored
+
+  #The following cycles create independent or correlated responses.
+To each value of mu (mean response per group) a randomly generated error (correlated or uncorrelated) is added and thus the individual response is created.
   if (error_type == "independent") {
@@
 }
+##################Section for plotting#################################
+#######################################################################
+#This function will create the plots for either a "linear" or "quadratic" response
 
 plot_example <- function(sim_dat) {
-  library(patchwork)
-  ## Plot the simulated data
+  ## Plot the simulated data (scatterplot)
   p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend=FALSE) +labs(y='response')+
-    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
+    ggplot(aes(x = time, 
+               y = y, 
+               group = treatment, 
+               color = treatment)
+           ) +
+    geom_point(show.legend=FALSE) +
+    labs(y='response')+
+    geom_line(aes(x = time, 
+                  y = mu, 
+                  color = treatment),
+              show.legend=FALSE) +
     theme_classic() +
-    #ggtitle("Simulated data with true response function")+
     theme(plot.title = element_text(size = 30, 
@@
 
+  #plot the simulated data with trajectories per each subject
   p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
+    ggplot(aes(x = time, 
+               y = y, 
+               group = subject, 
+               color = treatment)
+           ) +
     geom_line(aes(size = "Subjects"),
@@
     # facet_wrap(~ treatment) +
-    geom_line(aes(
-      x = time, 
+    geom_line(aes(x = time, 
                   y = mu, 
                   color = treatment,
-      size = "Simulated Truth"
-      ), 
-      lty = 1,
-      show.legend = FALSE
-      ) +
+                  size = "Simulated Truth"), 
+              lty = 1,show.legend = FALSE) +
     labs(y='response')+
-    scale_size_manual(
-      name = "Type", 
-      values=c("Subjects" = 0.5, "Simulated Truth" = 3)
-      ) +
-    #ggtitle("Simulated data\nIndividual responses with population mean") +
+    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
     theme_classic()+
@@
 
+  #plot the errors
    p3 <- sim_dat$dat %>%
@@
      theme_classic()+
-    # facet_wrap(~ treatment) +
-    #ggtitle("Simulated errors") +
      theme(plot.title = element_text(size = 30, 
@@
 
+   #plot the model predictions
   p4 <- ggplot(sim_dat$dat, 
@@
     labs(y='response')+
-    geom_line(aes(
-      y = predict(sim_dat$fit_lme), 
-      group = subject, 
-      size = "Subjects")
-      ) +
+    geom_line(aes(y = predict(sim_dat$fit_lme), 
+                  group = subject, size = "Subjects")) +
     geom_line(data = sim_dat$pred_dat, 
@@
     scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, 
-                               "Population" = 3)) +
+                      values=c("Subjects" = 0.5, "Population" = 3)) +
     theme_classic() +
-    #ggtitle("Fitted Model")+
     theme(plot.title = element_text(size = 30, 
@@
 txt<-18
+
+#Store each plot in a separate object
 A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
@@
 D1<-plot_example(example(fun_type = "quadratic", error_type = "independent")) 
-
-\end{verbatim}
-
-\begin{verbatim}
-## Error in eval(expr, envir, enclos): object 'A1' not found
-\end{verbatim}
-
-For the quadratic response case, Figure \ref{fig:quadratic-cases-appendix} shows the simulated responses using compound symmetry and independent errors.
-
-\begin{verbatim}
-## Error in eval(expr, envir, enclos): object 'C1' not found
-\end{verbatim}
+\end{lstlisting}
+
+\begin{figure}[H]
+\includegraphics{Full_document_files/figure-latex/linear-cases-Appendix-1} \caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
+A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
+B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
+The rm-ANOVA model does not pick the trend of the quadratic data.}\label{fig:linear-cases-Appendix}
+\end{figure}
+
+For the quadratic response case, Figure \ref{fig:quadratic-cases-Appendix} shows the simulated responses using compound symmetry and independent errors.
+
+\begin{figure}[H]
+\includegraphics{Full_document_files/figure-latex/quadratic-cases-Appendix-1} \caption{Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
+A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
+B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
+C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
+D,H: Estimations from the rm-ANOVA model for the mean group response.
+Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.}\label{fig:quadratic-cases-Appendix}
+\end{figure}
+
+\hypertarget{basis-functions-and-gams}{%
+\subsection{Basis functions and GAMs}\label{basis-functions-and-gams}}
+
+This code produces Figure \ref{fig:basis-plot} from the main manuscript.
+Briefly, a non-linear (quadratic) response is simulated a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
+The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
+
+\begin{lstlisting}[language=R]
+#generate the response: the same initial procedure from the previous section to simulate
+#the response
+n_time = 6
+ x <- seq(1,6, length.out = n_time)
+ mu <- matrix(0, length(x), 2)
+ mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
+ mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
+ y <- array(0, dim = c(length(x), 2, 10))
+ errors <- array(0, dim = c(length(x), 2, 10))
+ for (i in 1:2) {     # number of treatments
+     for (j in 1:10) {  # number of subjects
+         # compound symmetry errors
+         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
+         y[, i, j] <- mu[, i] + errors[, i, j]
+     }
+ }
+
+ #label each table
+  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
+ dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
+ dimnames(mu) <- list(time = x, treatment = 1:2)
+
+ #Convert to dataframes with subject, time and group columns
+ dat <- as.data.frame.table(y, responseName = "y")
+ dat_errors <- as.data.frame.table(errors, responseName = "errors")
+ dat_mu <- as.data.frame.table(mu, responseName = "mu")
+ dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
+ dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
+ dat$time <- as.numeric(as.character(dat$time))
+
+ #label subject per group
+ dat <- dat %>%
+     mutate(subject = factor(paste(subject, treatment, sep = "-")))
+
+ #extract  "Group 1" to fit the GAM
+  dat<-subset(dat,treatment==1)
+ #keep just the response and timepoint columns
+   dat<-dat[,c('y','time')]
+
+   #GAM model of time, 5 knots
+gm<-gam(y~s(time,k=5),data=dat)
+
+#model_matrix (also known as) 'design matrix'
+#will contain the smooths used to create  model 'gm'
+model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
+
+time<-c(1:6)
+
+basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
+#basis<-model_matrix[1:6,-1] #extracting basis
+colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
+basis<-basis %>% #pivoting to long format
+  pivot_longer(
+    cols=starts_with("s")
+  )%>%
+  arrange(name) #ordering
+
+#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
+ln<-6*(length(coef(gm)))
+
+basis_plot<-data.frame(Basis=integer(ln),
+                       value_orig=double(ln),
+                       time=integer(ln),
+                       cof=double(ln)
+)
+
+basis_plot$time<-rep(time) #pasting timepoints
+basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
+basis_plot$value_orig<-basis$value #pasting basis values
+basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
+basis_plot<-basis_plot%>%
+  mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be 
+#multiplied by the coefficients
+
+#creating labeller to change the labels in the basis plots
+
+basis_names<-c(
+  `1`="Intercept",
+  `2`="1",
+  `3`="2",
+  `4`="3",
+  `5`="4"
+)
+
+#calculating the final smooth by aggregating the basis functions
+
+smooth<-basis_plot%>% 
+  group_by(time)%>%
+  summarize(smooth=sum(mod_val))
+
+#original basis
+sz<-1
+p11<-ggplot(basis_plot,
+            aes(x=time,
+                y=value_orig,
+                colour=as.factor(Basis)
+                )
+            )+
+  geom_line(size=sz,
+            show.legend=FALSE)+
+  geom_point(size=sz+1,
+             show.legend = FALSE)+
+  labs(y='Basis functions')+
+  facet_wrap(~Basis,
+             labeller = as_labeller(basis_names)
+             )+
+  theme_classic()+
+  scale_color_aaas()
+
+#penalized basis
+p12<-ggplot(basis_plot,
+            aes(x=time,
+                y=mod_val,
+                colour=as.factor(Basis)
+                )
+            )+
+  geom_line(show.legend = FALSE,
+            size=sz)+
+  geom_point(show.legend = FALSE,
+             size=sz+1)+
+  labs(y='Penalized \n basis functions')+
+  scale_y_continuous(breaks=seq(-1,1,1))+
+  facet_wrap(~Basis,
+             labeller=as_labeller(basis_names)
+             )+
+  theme_classic()+
+  scale_color_aaas()
+
+#heatmap of the penalization coefficient
+x_labels<-c("Intercept","1","2","3","4")
+p13<-ggplot(basis_plot,
+            aes(x=Basis,
+                y=Basis,
+                fill=cof))+
+  geom_tile(aes(color='black'),
+            size=sz+1,
+            show.legend = FALSE)+
+  geom_tile(size=sz+1)+
+  scale_fill_gradient(low = "white", high = "red")+
+  labs(x='Basis',
+       y='Basis')+
+  scale_x_discrete(labels=x_labels)+
+  geom_text(aes(label=round(cof,2)),
+            size=10,
+            show.legend = FALSE)+
+  theme_classic()+
+  theme(legend.title = element_blank())
+
+#plotting simulated datapoints and smooth term
+p14<-ggplot(data=dat,
+            aes(x=time,y=y))+
+  geom_point(size=sz+1)+
+  scale_color_aaas()+
+  labs(y='Simulated \n response')+
+  geom_line(data=smooth,
+            aes(x=time,
+                y=smooth),
+            color="#B15731",
+            size=sz+1)+
+  theme_classic()
+
+#Combining all
+b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
+  theme(
+     text=element_text(size=18)
+     )
+\end{lstlisting}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/basis-plot-appendix-1}
+
+}
+
+\caption{Basis functions for a single smoother for time with five knots.
+A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
+B: Penalty matrix for the basis functions.
+Each basis function is penalized by a coefficient which can be positive or negative.
+The coefficient determines the overall effect of each basis in the final smoother.
+C: Penalized basis functions.
+Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
+D: Smoother for time and original data points.
+The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}\label{fig:basis-plot-appendix}
+\end{figure}
+
+\hypertarget{data-simulation-and-gam-models}{%
+\subsection{Data simulation and GAM models}\label{data-simulation-and-gam-models}}
+
+This section describes how to fit GAMs to longitudinal data using simulated data. First, data is simulated according to Section \ref{longitudinal-GAMs}.
+
+\begin{lstlisting}[language=R]
+dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
+            Day=rep(c(0,2,5,7,10),times=2),
+            Group=as.factor(rep(c("Control","Treatment"),each=5))
+            )
+
+## plot the mean response
+f1<-ggplot(dat, 
+           aes(x = Day, 
+               y = StO2, 
+               color = Group)) +
+    geom_line(size=1,
+              show.legend = FALSE)+
+    geom_point(show.legend = FALSE,
+               size=1.5,
+               alpha=0.5)+
+  labs(y=expression(paste(StO[2],
+                          ' (real)')))+
+  theme_classic()+
+  scale_color_aaas()+
+    scale_x_continuous(breaks=c(0,5,10))+
+    scale_y_continuous(breaks=c(0,40))+
+  plot_layout(tag_level = 'new')+
+  theme(
+    plot.background = element_rect(fill = "transparent", 
+                                   color = NA),
+    axis.text=element_text(size=14)
+  )
+
+#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
+#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
+
+simulate_data <- function(dat, n = 10, sd = 5) {
+    dat_sim <- dat %>%
+        slice(rep(1:n(), each = n)) %>%
+        group_by(Group, Day) %>%
+        mutate(
+               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
+               subject=rep(1:10),
+               subject=factor(paste(subject, Group, sep = "-"))
+               ) %>%
+        ungroup()
+
+    return(dat_sim)
+}
+
+#subject = factor(paste(subject, treatment, sep = "-")))
+
+n <- 10 #number of observations
+sd <- 10 #approximate sd from paper
+set.seed(1) #set seed for reproducibility
+df <- 6
+dat_sim <- simulate_data(dat, n, sd)
+
+#plotting simulated data
+f2<-ggplot(dat_sim, 
+           aes(x = Day, 
+               y = StO2_sim, 
+               color = Group)) +
+    geom_point(show.legend=FALSE,
+               size=1.5,
+               alpha=0.5)+
+    stat_summary(aes(y = StO2_sim,
+                     group=Group), 
+                 fun=mean, geom="line",
+                 size=1,
+                 show.legend = FALSE)+
+  labs(y=expression(atop(StO[2],
+                         '(simulated)')))+
+  theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
+  scale_color_aaas()+
+    scale_x_continuous(breaks=c(0,2,5,7,10))
+\end{lstlisting}
+
+\hypertarget{workflow-for-gams}{%
+\subsubsection{Workflow for GAMs}\label{workflow-for-gams}}
+
+Next, a series of increasingly complex GAMs are fitted to the simulated data.
+The first model is a model that only accounts for different smooths by day.
+The model syntax specifies that \passthrough{\lstinline!gam\_00!} is the object that will contain all the model information, and that the model attempts to explain changes in \passthrough{\lstinline!StO2\_sim!} (simulated \(\mbox{StO}_2\)) using a smooth per \passthrough{\lstinline!Day!}.
+The model will use 5 knots (\passthrough{\lstinline!k=5!}) for the smooth.
+And that the smooth is constructed using gaussian process basis (\passthrough{\lstinline!bs="gp"!}).
+The smoothing parameter estimation method used is the restricted maximum likelihood (\passthrough{\lstinline!REML!}).
+
+\begin{lstlisting}[language=R]
+gam_00<-gam(StO2_sim ~ s(Day, k = 5,bs="gp"),
+            method='REML',
+            data  = dat_sim)
+\end{lstlisting}
+
+To obtain model diagnostics, two methodologies can be used: 1) graphical diagnostics, and 2) a model check.
+In the first case, the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} can be used to obtain a single plot will all the diagnostic information.
+For model check, the function \passthrough{\lstinline!gam.check!} from \emph{mgcv} provides detailed information about the model.
+
+\hypertarget{graphical-diagnostics}{%
+\paragraph{Graphical diagnostics}\label{graphical-diagnostics}}
+
+\begin{lstlisting}[language=R]
+#need to add figure number and caption
+appr1<-appraise(gam_00)
+sm1<-draw(gam_00)
+
+visual_check<-appr1+sm1
+
+visual_check
+\end{lstlisting}
+
+\includegraphics{Full_document_files/figure-latex/first-GAM-diag-1}
+
+From the \passthrough{\lstinline!appraise!} plot, the major indicators concern about the model are the QQ plot of residuals and the histogram of residuals.
+The QQ plot shows that the errors are not reasonably located along the 45\(^{\circ}\) line, as there are multiple points that deviate from the trend, which is more noticeable in the tails.
+The histogram also shows that the variation (residuals) is not following the assumption of a normal distribution.
+
+The \passthrough{\lstinline!draw!} function permits to plot the smooths as \passthrough{\lstinline!ggplot2!} objects.
+Because this model specifies only one smooth for the time covariate (Day) the plot only contains only one smooth.
+Note that the smooth show an almost linear profile.
+
+\hypertarget{model-check}{%
+\paragraph{Model check}\label{model-check}}
+
+\begin{lstlisting}[language=R]
+#need to add figure number and caption
+gam.check(gam_00)
+\end{lstlisting}
+
+\includegraphics{Full_document_files/figure-latex/first-GAM-check-1}
+
+\begin{lstlisting}
+## 
+## Method: REML   Optimizer: outer newton
+## full convergence after 6 iterations.
+## Gradient range [-4.142968e-08,2.799316e-12]
+## (score 440.4108 & scale 414.2575).
+## Hessian positive definite, eigenvalue range [0.04576008,49.0005].
+## Model rank =  5 / 5 
+## 
+## Basis dimension (k) checking results. Low p-value (k-index<1) may
+## indicate that k is too low, especially if edf is close to k'.
+## 
+##          k'  edf k-index p-value    
+## s(Day) 4.00 1.31    0.26  <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{lstlisting}
+
+\begin{lstlisting}[language=R]
+summary(gam_00)
+\end{lstlisting}
+
+\begin{lstlisting}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## StO2_sim ~ s(Day, k = 5, bs = "gp")
+## 
+## Parametric coefficients:
+##             Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)   21.929      2.035   10.77   <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##          edf Ref.df     F p-value   
+## s(Day) 1.314  1.536 9.151 0.00253 **
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.105   Deviance explained = 11.7%
+## -REML = 440.41  Scale est. = 414.26    n = 100
+\end{lstlisting}
+
+Special attention must be paid to the `k-index' from \passthrough{\lstinline!gam.check!}.
+This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data, which is indicated by a low (\textless1) value.
+From the model, it can be seen that the k-index is effectively \textless0.3, which indicates that the model is not capturing the variability in the data.
+The `edf' (effective degrees of freedom) is an indicator of the complexity of the smooth.
+Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
+
+From the \passthrough{\lstinline!summary!} function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
+The link function is `identity' as the model does not make any transformation on the predictors.
+The `significance of smooth terms' \emph{p-value} indicates if each smooth is adding significance to the model.
+Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
+Finally, the `deviance explained' indicates how much of the data the model is able to capture.
+
+The major flaw of \passthrough{\lstinline!gam\_00!} is that is not taking into account the fact that the data is nested in groups.
+The next iteration is a model where a different smooth of time (Day) is assigned for each group using \passthrough{\lstinline!by=Group!} in the model syntax.
+
+\begin{lstlisting}[language=R]
+gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5,bs="gp"),
+            method='REML',
+            data  = dat_sim)
+\end{lstlisting}
+
+Diagnostics for this model indicate that the k-index is still below 1, and that the residuals are still not normally distributed.
+Moreover, if the smooths are plotted via the \passthrough{\lstinline!draw()!} function, it can be seen that their profile is fairly linear, which is not capturing the trends observed in the data.
+From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \textasciitilde43\%.
+
+As indicated in Section \ref{longitudinal-GAMs}, in order to differentiate between each group, a parametric term needs to be added to the model in order to account for the interaction of \emph{Day} and \emph{Group}.
+
+The resulting model is model \passthrough{\lstinline!gam1!} from the main manuscript.
+By running \passthrough{\lstinline!appraise()!} we see that the trend on the QQ plot has improved and that the histogram of the residuals appears to be reasonably distributed.
+From running \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\textasciitilde1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 87\% of the data.
+
+\textbf{Need to add the AIC comparison for the three models}
+
+\begin{lstlisting}[language=R]
+#GAM for StO2
+
+gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
+            method='REML',
+            data  = dat_sim)
+
+#linear model
+lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
+
+#creates a dataframe using the length of the covariates for the GAM
+gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = seq(0, 10, by = 0.1),
+                         subject=factor(rep(1:10)))
+
+#creates a dataframe using the length of the covariates for rm-ANOVA
+lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = c(0:10),
+                        subject=factor(rep(1:10)),
+                          )
+lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
+
+#adds the predictions to the grid and creates a confidence interval for GAM
+gam_predict<-gam_predict%>%
+    mutate(fit = predict(gam1,gam_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(gam1, gam_predict,se.fit = TRUE,type='response')$se.fit)
+
+#using lm
+lm_predict<-lm_predict%>%
+    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
+
+#plot smooths and confidence interval for GAM
+f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
+    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),
+                   fill=Group
+                   ),
+              alpha=0.3,
+              data=gam_predict,
+            show.legend=FALSE,
+                inherit.aes=FALSE) +
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=gam_predict,
+              show.legend = FALSE)+
+  #facet_wrap(~Group)+
+  labs(y=expression(atop(StO[2],'complete')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
+      scale_color_aaas()+
+  scale_fill_aaas()
+
+#plot linear fit for rm-ANOVA
+f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
+    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),fill=Group),
+              alpha=0.3,
+              data=lm_predict,
+              show.legend = FALSE,
+                inherit.aes=FALSE) +
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=lm_predict,
+              show.legend = FALSE)+
+  #facet_wrap(~Group)+
+  labs(y=expression(paste('StO'[2],' (simulated)')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
+      scale_color_aaas()+
+  scale_fill_aaas()
+
+#posthoc comparisons for the linear model
+library(multcomp)
+
+#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
+#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
+\end{lstlisting}
+
+\begin{lstlisting}[language=R]
+#missing data
+#create a sequence of 40 random numbers between 1 and 100, these numbers will
+#correspond to the row numbers to be randomly erased from the original dataset
+missing <- sample(1:100, 40)
+
+#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
+
+ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
+
+#create a new dataframe, remove the StO2 column
+dat_missing <- dat_sim[,-1]
+
+#add NAs at the ind positions
+dat_missing$StO2_sim[ind]<-NA
+
+#Count the number of remaining observations per day (original dataset had 10 per group per day)
+dat_missing %>%
+    group_by(Day,Group) %>%
+    filter(!is.na(StO2_sim))%>%
+  count(Day)
+\end{lstlisting}
+
+\begin{lstlisting}
+## # A tibble: 10 x 3
+## # Groups:   Day, Group [10]
+##      Day Group         n
+##    <dbl> <fct>     <int>
+##  1     0 Control       4
+##  2     0 Treatment     3
+##  3     2 Control       9
+##  4     2 Treatment     3
+##  5     5 Control       5
+##  6     5 Treatment     7
+##  7     7 Control       1
+##  8     7 Treatment     7
+##  9    10 Control       3
+## 10    10 Treatment     6
+\end{lstlisting}
+
+\begin{lstlisting}[language=R]
+#the same model used for the full dataset
+mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
+#appraise the model
+appraise(mod_m1)
+\end{lstlisting}
+
+\includegraphics{Full_document_files/figure-latex/missing-data-Appendix-1}
+
+\begin{lstlisting}[language=R]
+m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = seq(0, 10, by = 0.1))
+
+#adds the predictions to the grid and creates a confidence interval
+m_predict<-m_predict%>%
+    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
+
+f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
+    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),
+                   fill=Group
+                   ),
+              alpha=0.3,
+              data=m_predict,
+            show.legend=FALSE,
+                inherit.aes=FALSE) +
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=m_predict,
+              show.legend = TRUE)+
+  #facet_wrap(~Group)+
+  labs(y=expression(atop(StO[2],'missing')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
+      scale_color_aaas()+
+  scale_fill_aaas()
+\end{lstlisting}
+
+\begin{lstlisting}[language=R]
+mult_plot<-f2+inset_element(
+  f1, left = 0.01, 
+  bottom = 0.5, 
+  right = 0.5, 
+  top = 1.0)+
+  f3+f4+f6+
+   plot_annotation(tag_levels='A')&
+   ylim(c(-5,75)) &
+  theme(
+     text=element_text(size=18)
+     )&
+  scale_color_aaas()
+
+mult_plot
+\end{lstlisting}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/sim-smooth-plot-Appendix-1}
+
+}
+
+\caption{Simulated data and smooths for oxygen saturation in tumors.
+A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
+Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
+Lines represent trends for each group, shaded regions are 95\% confidence intervals.
+C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
+D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
+Lines represent trends for each group, shaded regions are 95\% confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
+\end{figure}
 

### 2021-05-18 Manuscripts/Manuscript by chapters -v1/06-Appendix.Rmd
Working on code chunks in PDF

@@
 
-# Simulation
+# Code for Manuscript data
+
+This section presents the code used to generate figures, models and simulated data from Sections 3 and 4 from the main manuscript.
 
@@
 
-## Data simulation and GAM models
+# Longitudinal biomedical data simulation and GAMs{#tumor-data-simulation}
 
-This section describes how to fit GAMs to longitudinal data using simulated data. First, data is simulated according to Section \@ref(longitudinal-GAMs).
+This section describes how to fit GAMs to longitudinal data using simulated data.
+First, data is simulated according to Section \@ref(longitudinal-GAMs), where reported data of oxygen saturation ($\mbox{StO}_2$) in tumors under either chemotherapy or saline control is used as a starting point to generate individual responses in each group.
 
@@
 
+#Dataframe that contains the original reported trends
 dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
@@
 
-### Workflow for GAMs
+## A basic Workflow for GAMs
 
-Next, a series of increasingly complex GAMs are fitted to the simulated data.
-The first model is a model that only accounts for different smooths by day.
+This section shows a basic workflow to fit a series of increasingly complex GAMs to the simulated data from the previous section.
+Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
+
+### First model
+
+The first model fitted to the data is one that only accounts for different smooths by day.
 The model syntax specifies that `gam_00` is the object that will contain all the model information, and that the model attempts to explain changes in `StO2_sim` (simulated $\mbox{StO}_2$) using a smooth per `Day`.
@@
 
-To obtain model diagnostics, two methodologies can be used: 1) graphical diagnostics, and 2) a model check.
-In the first case, the function `appraise` from the package _gratia_ can be used to obtain a single plot will all the diagnostic information.
-For model check, the function `gam.check` from  _mgcv_ provides detailed information about the model.
+To obtain model diagnostics, two methodologies are to be used: 1) graphical diagnostics, and 2) a model check.
+In the first case, the functions `appraise` and `draw` from the package _gratia_ can be used to obtain a single output with all the graphical diagnostics.
+For model check, the functions `gam.check` and `summary` from  _mgcv_ provide detailed information about the model fit and its parameters.
+
+(ref:first-GAM-caption) Graphical diagnostics for the first GAM model.
+Left: Graphical diagnostics provided by the function `appraise` from the package _gratia_.
+Right: Fitted smooth for the model, provided by the function `draw`.
 
@@
 
-```{r, first-GAM-diag, echo=TRUE, include=TRUE, message=FALSE,warning=FALSE}
+```{r, first-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:first-GAM-caption)'}
 #need to add figure number and caption
@@
 sm1<-draw(gam_00)
-
 visual_check<-appr1+sm1
@@
 
-From the `appraise` plot, the major indicators concern about the model are the QQ plot of residuals and the histogram of residuals.
-The QQ plot shows that the errors are not reasonably located along the 45$^{\circ}$ line, as there are multiple points that deviate from the trend, which is more noticeable in the tails.
+From the output of the function `appraise` in Figure \@ref(fig:first-GAM-diag), the major indicators of concern about the model are the QQ plot of residuals and the histogram of residuals.
+The QQ plot shows that the errors are not reasonably located along the 45$^{\circ}$ line (which indicates normality), as there are multiple points that deviate from the trend, specially in the tails.
 The histogram also shows that the variation (residuals) is not following the assumption of a normal distribution.
 
-The `draw` function permits to plot the smooths as `ggplot2` objects.
-Because this model specifies only one smooth for the time covariate (Day) the plot only contains only one smooth.
-Note that the smooth show an almost linear profile.
+The `draw` function permits to plot the smooths as `ggplot2` objects, which eases subsequent manipulation, if desired.
+Because model `gam_00` specifies only one smooth for the time covariate (Day), the plot only contains only one smooth.
+Note that the smooth shows an almost linear profile.
 
 #### Model check
+
 ```{r, first-GAM-check, echo=TRUE, include=TRUE, message=FALSE,warning=FALSE}
@@
 Special attention must be paid to the 'k-index' from `gam.check`.
-This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data, which is indicated by a low (<1) value.
-From the model, it can be seen that the k-index is effectively <0.3, which indicates that the model is not capturing the variability in the data.
+This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
+If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (<1).
+From the output, it can be seen that the k-index is effectively <0.3, which indicates that the model is not capturing the variability in the data.
 The 'edf' (effective degrees of freedom) is an indicator of the complexity of the smooth.
@@
 Here, the _p-value_ is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the 'deviance explained' indicates how much of the data the model is able to capture.
+Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to ~12%.
 
-The major flaw of `gam_00` is that is not taking into account the fact that the data is nested in groups.
+### Second model
+
+The major flaw of `gam_00` is that this model is not taking into account the fact that the data is nested in groups.
 The next iteration is a model where a different smooth of time (Day) is assigned for each group using `by=Group` in the model syntax.
@@
             data  = dat_sim)
+
+gam.check(gam_01)
+summary(gam_01)
 ```
 
-Diagnostics for this model indicate that the k-index is still below 1, and that the residuals are still not normally distributed.
-Moreover, if the smooths are plotted via the `draw()` function, it can be seen that their profile is fairly linear, which is not capturing the trends observed in the data.
+Diagnostics for this model indicate that the k-index is still below 1 (0.32 from `gam.check`), and that the residuals are still not following a normal distribution (Figure  \@ref(fig:second-GAM-diag)).
+Moreover, the smooths (plotted via the `draw()` function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
 From `summary()`, the deviance explained by the model is ~43%.
 
-As indicated in Section \@ref(longitudinal-GAMs), in order to differentiate between each group, a parametric term needs to be added to the model in order to account for the interaction of _Day_ and _Group_.
+(ref:second-GAM-caption) Graphical diagnostics for the second GAM model.
+Left: Graphical diagnostics provided by the function `appraise` from the package _gratia_.
+Right: Fitted smooth for the model, provided by the function `draw`.
 
-The resulting model is model `gam1` from the main manuscript.
-By running `appraise()` we see that the trend on the QQ plot has improved and that the histogram of the residuals appears to be reasonably distributed.
-From running `gam.check`, the k-index is now at an acceptable value (~1.02), and `summary` now indicates that the model is able to capture 87% of the data.
+```{r, second-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:second-GAM-caption)'}
+#need to add figure number and caption
+appr2<-appraise(gam_01)
+sm2<-draw(gam_01)
 
-**Need to add the AIC comparison for the three models**
+visual_check2<-appr2+sm2
 
-```{r,smooth-plot-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
+visual_check2
+```
+
+### Third model
+
+Model `gam_00` was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted, the model diagnostics, and, the low variance explained by the model.
+On the other hand, `gam_01` takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \@ref(longitudinal-GAMs), in order to differentiate between each group a parametric term needs to be added to the model for the interaction of _Day_ and _Group_.
+
+```{r,final-model-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
 
@@
 
+gam.check(gam1)
+summary(gam1)
+```
+
+The resulting model is model `gam1`, which is the model fitted in the main manuscript.
+By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group .
+From `gam.check`, the k-index is now at an acceptable value (~1.02), and `summary` now indicates that the model is able to capture 87% of the variance data.
+
+(ref:final-GAM-caption) Graphical diagnostics for the final GAM model.
+Left: Graphical diagnostics provided by the function `appraise` from the package _gratia_.
+Right: Fitted smooths for the model, provided by the function `draw`.
+
+```{r, final-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:final-GAM-caption)'}
+#need to add figure number and caption
+appr3<-appraise(gam1)
+sm3<-draw(gam1)
+
+visual_check3<-appr3+sm3
+
+visual_check3
+```
+
+### Comparing models via AIC
+
+One final  comparison that can be made for model selection involves the use of the Aikake Information Criterion (AIC).
+This metric is used to estimate information loss, which we want to minimize with an appropriate model.
+Therefore, when 2 or more models are compared, the model with lower AIC is preferred.
+In R, the comparison is done using the `AIC` function.
+
+```{r,model-selection-AIC}
+
+AIC(gam_00,gam_01,gam1)
+
+```
+
+The output in this case is expected: model `gam1` has a lower AIC (711.46) whereas the initial two models have higher AICs (891 and 850).
+The AIC should not be considered as the only estimator of model goodness, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
+
+# GAM and Linear model plots and Missing data
+
+This section covers the code used to generate Figure \@ref(fig:sim-smooth-plot), where the simulated data, fit of the "final" GAM (`gam1`), linear model  and GAM on data with missing observations are presented.
+Note that panel A in Figure \@ref(fig:sim-smooth-plot) and the inlet are generated in the code chunk where the data is simulated in Section \@ref(tumor-data-simulation), and are called later to build the figure.
+
+## GAM and Linear model plots
+
+This code chunk creates panels B and D in Figure \@ref(fig:sim-smooth-plot).
+Note that this code uses the final GAM from the previous section (`gam1`), so the simulated data and the model should be generated before running this section.
+
+```{r, plotting-appendix, echo=TRUE, include=TRUE, message=FALSE, warning=FALSE}
 #linear model
@@
 ```
+
+## Working with Missing data in GAMs
+
+This code chunk randomly deletes 40% of the total observations in the original simulated data, and then an interaction GAM is fitted.
+Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
 

### 2021-05-18 Manuscripts/Manuscript by chapters -v1/Full_document.Rmd
Working on code chunks in PDF

@@
     \usepackage{placeins}
+    \usepackage{solarized-light}
 output:  
@@
 
-In biomedical research, the outcome of a longitudinal study has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, a _linear mixed model_ (LMEM).
+In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, a _linear mixed model_ (LMEM).
 Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
-In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
-Although not commonly used in biomedical research, GAMs present an excellent choice to analyze non-linear longitudinal data.
-This paper summarizes the limitations of linear models (rm-ANOVA in particular), presents the basic theory of GAMs, and demonstrates their implementation in $\textsf{R}$ via the package _mgcv_ using simulated data that follows longitudinal trends reported in biomedical literature.
+In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures, thereby being an excellent choice to analyze non-linear longitudinal data.
+This paper summarizes the limitations of LMEMs and rm-ANOVA, presents the basic theory of GAMs, and demonstrates their implementation in $\textsf{R}$ via the package _mgcv_ using simulated data that follows longitudinal trends reported in biomedical literature.
 To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:________.
@@
 library(gratia)
+set.seed(1) #set seed for reproducibility
 

### 2021-05-18 Manuscripts/Manuscript by chapters -v1/Full_document.tex
Working on code chunks in PDF

@@
 }
-\usepackage{placeins}
+\usepackage{placeins} \usepackage{solarized-light}
 \usepackage{subfig}
@@
 
-In biomedical research, the outcome of a longitudinal study has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, a \emph{linear mixed model} (LMEM).
+In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, a \emph{linear mixed model} (LMEM).
 Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
-In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
-Although not commonly used in biomedical research, GAMs present an excellent choice to analyze non-linear longitudinal data.
-This paper summarizes the limitations of linear models (rm-ANOVA in particular), presents the basic theory of GAMs, and demonstrates their implementation in \(\textsf{R}\) via the package \emph{mgcv} using simulated data that follows longitudinal trends reported in biomedical literature.
+In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures, thereby being an excellent choice to analyze non-linear longitudinal data.
+This paper summarizes the limitations of LMEMs and rm-ANOVA, presents the basic theory of GAMs, and demonstrates their implementation in \(\textsf{R}\) via the package \emph{mgcv} using simulated data that follows longitudinal trends reported in biomedical literature.
 To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:\_\_\_\_\_\_\_\_.
@@
 
-\hypertarget{simulation-1}{%
-\section{Simulation}\label{simulation-1}}
+\hypertarget{code-for-manuscript-data}{%
+\section{Code for Manuscript data}\label{code-for-manuscript-data}}
+
+This section presents the code used to generate figures, models and simulated data from Sections 3 and 4 from the main manuscript.
 
@@
 
-\hypertarget{data-simulation-and-gam-models}{%
-\subsection{Data simulation and GAM models}\label{data-simulation-and-gam-models}}
-
-This section describes how to fit GAMs to longitudinal data using simulated data. First, data is simulated according to Section \ref{longitudinal-GAMs}.
+\hypertarget{tumor-data-simulation}{%
+\section{Longitudinal biomedical data simulation and GAMs}\label{tumor-data-simulation}}
+
+This section describes how to fit GAMs to longitudinal data using simulated data.
+First, data is simulated according to Section \ref{longitudinal-GAMs}, where reported data of oxygen saturation (\(\mbox{StO}_2\)) in tumors under either chemotherapy or saline control is used as a starting point to generate individual responses in each group.
 
 \begin{lstlisting}[language=R]
+#Dataframe that contains the original reported trends
 dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
@@
 
-\hypertarget{workflow-for-gams}{%
-\subsubsection{Workflow for GAMs}\label{workflow-for-gams}}
-
-Next, a series of increasingly complex GAMs are fitted to the simulated data.
-The first model is a model that only accounts for different smooths by day.
+\hypertarget{a-basic-workflow-for-gams}{%
+\subsection{A basic Workflow for GAMs}\label{a-basic-workflow-for-gams}}
+
+This section shows a basic workflow to fit a series of increasingly complex GAMs to the simulated data from the previous section.
+Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
+
+\hypertarget{first-model}{%
+\subsubsection{First model}\label{first-model}}
+
+The first model fitted to the data is one that only accounts for different smooths by day.
 The model syntax specifies that \passthrough{\lstinline!gam\_00!} is the object that will contain all the model information, and that the model attempts to explain changes in \passthrough{\lstinline!StO2\_sim!} (simulated \(\mbox{StO}_2\)) using a smooth per \passthrough{\lstinline!Day!}.
@@
 
-To obtain model diagnostics, two methodologies can be used: 1) graphical diagnostics, and 2) a model check.
-In the first case, the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} can be used to obtain a single plot will all the diagnostic information.
-For model check, the function \passthrough{\lstinline!gam.check!} from \emph{mgcv} provides detailed information about the model.
+To obtain model diagnostics, two methodologies are to be used: 1) graphical diagnostics, and 2) a model check.
+In the first case, the functions \passthrough{\lstinline!appraise!} and \passthrough{\lstinline!draw!} from the package \emph{gratia} can be used to obtain a single output with all the graphical diagnostics.
+For model check, the functions \passthrough{\lstinline!gam.check!} and \passthrough{\lstinline!summary!} from \emph{mgcv} provide detailed information about the model fit and its parameters.
 
@@
 
-\begin{lstlisting}[language=R]
-#need to add figure number and caption
-appr1<-appraise(gam_00)
-sm1<-draw(gam_00)
-
-visual_check<-appr1+sm1
-
-visual_check
-\end{lstlisting}
-
-\includegraphics{Full_document_files/figure-latex/first-GAM-diag-1}
-
-From the \passthrough{\lstinline!appraise!} plot, the major indicators concern about the model are the QQ plot of residuals and the histogram of residuals.
-The QQ plot shows that the errors are not reasonably located along the 45\(^{\circ}\) line, as there are multiple points that deviate from the trend, which is more noticeable in the tails.
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/first-GAM-diag-1}
+
+}
+
+\caption{Graphical diagnostics for the first GAM model.
+Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
+Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:first-GAM-diag}
+\end{figure}
+
+From the output of the function \passthrough{\lstinline!appraise!} in Figure \ref{fig:first-GAM-diag}, the major indicators of concern about the model are the QQ plot of residuals and the histogram of residuals.
+The QQ plot shows that the errors are not reasonably located along the 45\(^{\circ}\) line (which indicates normality), as there are multiple points that deviate from the trend, specially in the tails.
 The histogram also shows that the variation (residuals) is not following the assumption of a normal distribution.
 
-The \passthrough{\lstinline!draw!} function permits to plot the smooths as \passthrough{\lstinline!ggplot2!} objects.
-Because this model specifies only one smooth for the time covariate (Day) the plot only contains only one smooth.
-Note that the smooth show an almost linear profile.
+The \passthrough{\lstinline!draw!} function permits to plot the smooths as \passthrough{\lstinline!ggplot2!} objects, which eases subsequent manipulation, if desired.
+Because model \passthrough{\lstinline!gam\_00!} specifies only one smooth for the time covariate (Day), the plot only contains only one smooth.
+Note that the smooth shows an almost linear profile.
 
@@
 Special attention must be paid to the `k-index' from \passthrough{\lstinline!gam.check!}.
-This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data, which is indicated by a low (\textless1) value.
-From the model, it can be seen that the k-index is effectively \textless0.3, which indicates that the model is not capturing the variability in the data.
+This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
+If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless1).
+From the output, it can be seen that the k-index is effectively \textless0.3, which indicates that the model is not capturing the variability in the data.
 The `edf' (effective degrees of freedom) is an indicator of the complexity of the smooth.
@@
 Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the `deviance explained' indicates how much of the data the model is able to capture.
-
-The major flaw of \passthrough{\lstinline!gam\_00!} is that is not taking into account the fact that the data is nested in groups.
+Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \textasciitilde12\%.
+
+\hypertarget{second-model}{%
+\subsubsection{Second model}\label{second-model}}
+
+The major flaw of \passthrough{\lstinline!gam\_00!} is that this model is not taking into account the fact that the data is nested in groups.
 The next iteration is a model where a different smooth of time (Day) is assigned for each group using \passthrough{\lstinline!by=Group!} in the model syntax.
@@
             data  = dat_sim)
-\end{lstlisting}
-
-Diagnostics for this model indicate that the k-index is still below 1, and that the residuals are still not normally distributed.
-Moreover, if the smooths are plotted via the \passthrough{\lstinline!draw()!} function, it can be seen that their profile is fairly linear, which is not capturing the trends observed in the data.
+
+gam.check(gam_01)
+\end{lstlisting}
+
+\includegraphics{Full_document_files/figure-latex/second-GAM-1}
+
+\begin{lstlisting}
+## 
+## Method: REML   Optimizer: outer newton
+## full convergence after 10 iterations.
+## Gradient range [-0.0001703751,9.561998e-05]
+## (score 418.612 & scale 270.7177).
+## Hessian positive definite, eigenvalue range [0.0001702821,48.50255].
+## Model rank =  9 / 9 
+## 
+## Basis dimension (k) checking results. Low p-value (k-index<1) may
+## indicate that k is too low, especially if edf is close to k'.
+## 
+##                         k'  edf k-index p-value    
+## s(Day):GroupControl   4.00 1.00    0.32  <2e-16 ***
+## s(Day):GroupTreatment 4.00 1.72    0.32  <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{lstlisting}
+
+\begin{lstlisting}[language=R]
+summary(gam_01)
+\end{lstlisting}
+
+\begin{lstlisting}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## StO2_sim ~ s(Day, by = Group, k = 5, bs = "gp")
+## 
+## Parametric coefficients:
+##             Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)   21.929      1.645   13.33   <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##                         edf Ref.df      F p-value    
+## s(Day):GroupControl   1.001  1.001  4.099  0.0456 *  
+## s(Day):GroupTreatment 1.715  1.979 35.551  <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.415   Deviance explained = 43.1%
+## -REML = 418.61  Scale est. = 270.72    n = 100
+\end{lstlisting}
+
+Diagnostics for this model indicate that the k-index is still below 1 (0.32 from \passthrough{\lstinline!gam.check!}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
+Moreover, the smooths (plotted via the \passthrough{\lstinline!draw()!} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
 From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \textasciitilde43\%.
 
-As indicated in Section \ref{longitudinal-GAMs}, in order to differentiate between each group, a parametric term needs to be added to the model in order to account for the interaction of \emph{Day} and \emph{Group}.
-
-The resulting model is model \passthrough{\lstinline!gam1!} from the main manuscript.
-By running \passthrough{\lstinline!appraise()!} we see that the trend on the QQ plot has improved and that the histogram of the residuals appears to be reasonably distributed.
-From running \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\textasciitilde1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 87\% of the data.
-
-\textbf{Need to add the AIC comparison for the three models}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/second-GAM-diag-1}
+
+}
+
+\caption{Graphical diagnostics for the second GAM model.
+Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
+Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:second-GAM-diag}
+\end{figure}
+
+\hypertarget{third-model}{%
+\subsubsection{Third model}\label{third-model}}
+
+Model \passthrough{\lstinline!gam\_00!} was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted, the model diagnostics, and, the low variance explained by the model.
+On the other hand, \passthrough{\lstinline!gam\_01!} takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \ref{longitudinal-GAMs}, in order to differentiate between each group a parametric term needs to be added to the model for the interaction of \emph{Day} and \emph{Group}.
 
@@
 
+gam.check(gam1)
+\end{lstlisting}
+
+\includegraphics{Full_document_files/figure-latex/final-model-Appendix-1}
+
+\begin{lstlisting}
+## 
+## Method: REML   Optimizer: outer newton
+## full convergence after 9 iterations.
+## Gradient range [-1.003557e-07,3.562136e-08]
+## (score 362.7587 & scale 64.03804).
+## Hessian positive definite, eigenvalue range [0.9494021,48.08513].
+## Model rank =  10 / 10 
+## 
+## Basis dimension (k) checking results. Low p-value (k-index<1) may
+## indicate that k is too low, especially if edf is close to k'.
+## 
+##                         k'  edf k-index p-value
+## s(Day):GroupControl   4.00 3.83    1.02    0.52
+## s(Day):GroupTreatment 4.00 3.84    1.02    0.59
+\end{lstlisting}
+
+\begin{lstlisting}[language=R]
+summary(gam1)
+\end{lstlisting}
+
+\begin{lstlisting}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## StO2_sim ~ Group + s(Day, by = Group, k = 5, bs = "gp")
+## 
+## Parametric coefficients:
+##                Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)       9.781      1.132   8.643 1.85e-13 ***
+## GroupTreatment   24.296      1.600  15.181  < 2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##                         edf Ref.df     F p-value    
+## s(Day):GroupControl   3.825  3.971 16.81  <2e-16 ***
+## s(Day):GroupTreatment 3.835  3.974 78.84  <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.862   Deviance explained = 87.4%
+## -REML = 362.76  Scale est. = 64.038    n = 100
+\end{lstlisting}
+
+The resulting model is model \passthrough{\lstinline!gam1!}, which is the model fitted in the main manuscript.
+By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group .
+From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\textasciitilde1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 87\% of the variance data.
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/final-GAM-diag-1}
+
+}
+
+\caption{Graphical diagnostics for the final GAM model.
+Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
+Right: Fitted smooths for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:final-GAM-diag}
+\end{figure}
+
+\hypertarget{comparing-models-via-aic}{%
+\subsubsection{Comparing models via AIC}\label{comparing-models-via-aic}}
+
+One final comparison that can be made for model selection involves the use of the Aikake Information Criterion (AIC).
+This metric is used to estimate information loss, which we want to minimize with an appropriate model.
+Therefore, when 2 or more models are compared, the model with lower AIC is preferred.
+In R, the comparison is done using the \passthrough{\lstinline!AIC!} function.
+
+\begin{lstlisting}[language=R]
+AIC(gam_00,gam_01,gam1)
+\end{lstlisting}
+
+\begin{lstlisting}
+##               df      AIC
+## gam_00  3.536147 891.1671
+## gam_01  4.980481 850.0698
+## gam1   10.945191 711.4662
+\end{lstlisting}
+
+The output in this case is expected: model \passthrough{\lstinline!gam1!} has a lower AIC (711.46) whereas the initial two models have higher AICs (891 and 850).
+The AIC should not be considered as the only estimator of model goodness, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
+
+\hypertarget{gam-and-linear-model-plots-and-missing-data}{%
+\section{GAM and Linear model plots and Missing data}\label{gam-and-linear-model-plots-and-missing-data}}
+
+This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\passthrough{\lstinline!gam1!}), linear model and GAM on data with missing observations are presented.
+Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inlet are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
+
+\hypertarget{gam-and-linear-model-plots}{%
+\subsection{GAM and Linear model plots}\label{gam-and-linear-model-plots}}
+
+This code chunk creates panels B and D in Figure \ref{fig:sim-smooth-plot}.
+Note that this code uses the final GAM from the previous section (\passthrough{\lstinline!gam1!}), so the simulated data and the model should be generated before running this section.
+
+\begin{lstlisting}[language=R]
 #linear model
@@
 
+\hypertarget{working-with-missing-data-in-gams}{%
+\subsection{Working with Missing data in GAMs}\label{working-with-missing-data-in-gams}}
+
+This code chunk randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted.
+Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
+
 \begin{lstlisting}[language=R]
@@
 ##    <dbl> <fct>     <int>
-##  1     0 Control       4
-##  2     0 Treatment     3
-##  3     2 Control       9
-##  4     2 Treatment     3
-##  5     5 Control       5
-##  6     5 Treatment     7
-##  7     7 Control       1
-##  8     7 Treatment     7
+##  1     0 Control       2
+##  2     0 Treatment     4
+##  3     2 Control       6
+##  4     2 Treatment     5
+##  5     5 Control       6
+##  6     5 Treatment     4
+##  7     7 Control       3
+##  8     7 Treatment     5
 ##  9    10 Control       3
-## 10    10 Treatment     6
+## 10    10 Treatment     8
 \end{lstlisting}

### 2021-05-18 Manuscripts/Manuscript by chapters -v1/Full_document.Rmd
Themes for code chunks finally work!

@@
     \usepackage{placeins}
-    \usepackage{solarized-light}
 output:  

### 2021-05-18 Manuscripts/Manuscript by chapters -v1/Full_document.tex
Themes for code chunks finally work!

@@
 
+%command for the package lineno
 \linenumbers
 
+%authors
 \author[1]{Ariel I. Mundo}
@@
 
+\usepackage{xcolor} %theme colors for the code chunks (originally from latex-solarized on GitHub)
+\definecolor{sbase03}{HTML}{002B36}
+\definecolor{sbase02}{HTML}{073642}
+\definecolor{sbase01}{HTML}{586E75}
+\definecolor{sbase00}{HTML}{657B83}
+\definecolor{sbase0}{HTML}{839496}
+\definecolor{sbase1}{HTML}{93A1A1}
+\definecolor{sbase2}{HTML}{EEE8D5}
+\definecolor{sbase3}{HTML}{FDF6E3}
+\definecolor{syellow}{HTML}{B58900}
+\definecolor{sorange}{HTML}{CB4B16}
+\definecolor{sred}{HTML}{DC322F}
+\definecolor{smagenta}{HTML}{D33682}
+\definecolor{sviolet}{HTML}{6C71C4}
+\definecolor{sblue}{HTML}{268BD2}
+\definecolor{scyan}{HTML}{2AA198}
+\definecolor{sgreen}{HTML}{859900}
+%command to set parameter(s) in package listings
 \lstset{
-  breaklines=true
-}
-\usepackage{placeins} \usepackage{solarized-light}
+    % How/what to match
+    sensitive=true,
+    % Border (above and below)
+    frame=lines,
+    % Extra margin on line (align with paragraph)
+    xleftmargin=\parindent,
+    % Put extra space under caption
+    belowcaptionskip=1\baselineskip,
+    % Colors
+    backgroundcolor=\color{sbase3},
+    basicstyle=\color{sbase00}\ttfamily,
+    keywordstyle=\color{scyan},
+    commentstyle=\color{sbase1},
+    stringstyle=\color{sblue},
+    numberstyle=\color{sviolet},
+    identifierstyle=\color{sbase00},
+    % Break long lines into multiple lines?
+    breaklines=true,
+    % Show a character for spaces?
+    showstringspaces=false,
+    tabsize=2
+}
+
+%\lstset{
+%  breaklines=true,
+%  stringstyle=\ttfamily,
+%  backgroundcolor=\color{gray}
+%}
+\usepackage{placeins}
 \usepackage{subfig}

### 2021-05-18 Manuscripts/Manuscript by chapters -v1/Full_document.tex
First version ready

@@
 
-The traditional methods to analyze biomedical longitudinal data (rm-ANOVA, LMEMs) are restrictive when the data does not follow a linear pattern and their use in such situations leads to unreliable inference and biased estimates.
+The traditional methods to analyze biomedical longitudinal data (rm-ANOVA, LMEMs) are restricted to cases when the data follows a linear pattern.
+When the trend of the data is non-linear (as it is frequently the case in biomedical research), the use of rm-ANOVA and LMEMs leads to unreliable inference and biased estimates.
 GAMs present a suitable alternative to analyze non-linear longitudinal biomedical data, as they overcome the limitations of linearity, compound symmetry, and missing observations imposed by traditional methods.
-By presenting the implementation of GAMs using simulated data that follows previously reported trends in the literature, we aim at encouraging biomedical researchers to consider different statistical tools to analyze longitudinal data.
+By presenting in a clear and visual manner the limitations of rm-ANOVA and LMEMs, and the implementation of GAMs using simulated data that follows previously reported trends in the literature, we aim at encouraging biomedical researchers to consider different statistical tools to analyze longitudinal data.
 Finally, by providing the data and code used in this paper we hope to address the need of creating and sharing reproducible work in biomedical research.

### 2021-05-19 Manuscripts/Manuscript by chapters -v1/02-Challenges.Rmd
Updated Fig 1

@@
                   color = treatment),
+              size=3,
               show.legend=FALSE) +

### 2021-05-19 Manuscripts/Manuscript by chapters -v1/Full_document.tex
Updated Fig 1

@@
 ]{article}
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
@@
 \lstset{defaultdialect=[x86masm]Assembler}
-\usepackage{longtable,booktabs}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
 % Correct order of tables after \paragraph or \subparagraph
@@
 \makesavenoteenv{longtable}
-\usepackage{graphicx,grffile}
+\usepackage{graphicx}
 \makeatletter
@@
 \usepackage{float}
+\ifluatex
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?''.
-The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot},C.
+One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot},C.
 This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \ref{fig:sim-smooth-plot},B) .
@@
 \hypertarget{refs}{}
+\begin{CSLReferences}{0}{0}
 \leavevmode\hypertarget{ref-roblyer2011}{}%
-{[}1{]} D.
+\CSLLeftMargin{{[}1{]} }
+\CSLRightInline{D.
 Roblyer, S.
@@
 108 (2011) 14626--14631.
-\url{https://doi.org/10.1073/pnas.1013103108}.
+\url{https://doi.org/10.1073/pnas.1013103108}.}
 
 \leavevmode\hypertarget{ref-tank2020}{}%
-{[}2{]} A.
+\CSLLeftMargin{{[}2{]} }
+\CSLRightInline{A.
 Tank, H.M.
@@
 22 (2020) 1--10.
-\url{https://doi.org/doi:10.1186/s13058-020-01262-1}.
+\url{https://doi.org/doi:10.1186/s13058-020-01262-1}.}
 
 \leavevmode\hypertarget{ref-pavlov2018}{}%
-{[}3{]} M.V.
+\CSLLeftMargin{{[}3{]} }
+\CSLRightInline{M.V.
 Pavlov, T.I.
@@
 Shakhova, A.V.
-Maslennikova, Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy, Journal of Biomedical Optics.
-23 (n.d.).
-\url{https://doi.org/\%7B10.1117/1.JBO.23.9.091410\%7D}.
+Maslennikova, {Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy}, {Journal of Biomedical Optics}.
+{23} (2018).
+\url{https://doi.org/\%7B10.1117/1.JBO.23.9.091410\%7D}.}
 
 \leavevmode\hypertarget{ref-demidov2018}{}%
-{[}4{]} V.
+\CSLLeftMargin{{[}4{]} }
+\CSLRightInline{V.
 Demidov, A.
@@
 Flueraru, I.A.
-Vitkin, Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography, Scientific Reports.
-8 (n.d.).
-\url{https://doi.org/\%7B10.1038/s41598-017-18635-w\%7D}.
+Vitkin, {Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography}, {Scientific Reports}.
+{8} (2018).
+\url{https://doi.org/\%7B10.1038/s41598-017-18635-w\%7D}.}
 
 \leavevmode\hypertarget{ref-ritter2001}{}%
-{[}5{]} G.
+\CSLLeftMargin{{[}5{]} }
+\CSLRightInline{G.
 Ritter, L.
@@
 Old, S.
-Welt, Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33, Cancer Research.
-61 (n.d.) 6851--6859.
+Welt, {Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}, {Cancer Research}.
+{61} (2001) 6851--6859.}
 
 \leavevmode\hypertarget{ref-roth2017}{}%
-{[}6{]} E.M.
+\CSLLeftMargin{{[}6{]} }
+\CSLRightInline{E.M.
 Roth, A.C.
@@
 Lecorps, H.M.
-Colhoun, Antidrug Antibodies in Patients Treated with Alirocumab, New England Journal of Medicine.
-376 (n.d.) 1589--1590.
-\url{https://doi.org/\%7B10.1056/NEJMc1616623\%7D}.
+Colhoun, {Antidrug Antibodies in Patients Treated with Alirocumab}, {New England Journal of Medicine}.
+{376} (2017) 1589--1590.
+\url{https://doi.org/\%7B10.1056/NEJMc1616623\%7D}.}
 
 \leavevmode\hypertarget{ref-jones2018}{}%
-{[}7{]} J.D.
+\CSLLeftMargin{{[}7{]} }
+\CSLRightInline{J.D.
 Jones, H.E.
@@
 Woessner, K.P.
-Quinn, In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing, Communications Biology.
-1 (n.d.).
-\url{https://doi.org/\%7B10.1038/s42003-018-0206-4\%7D}.
+Quinn, {In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing}, {Communications Biology}.
+{1} (2018).
+\url{https://doi.org/\%7B10.1038/s42003-018-0206-4\%7D}.}
 
 \leavevmode\hypertarget{ref-skala2010}{}%
-{[}8{]} M.C.
+\CSLLeftMargin{{[}8{]} }
+\CSLRightInline{M.C.
 Skala, A.
@@
 15 (2010).
-\url{https://doi.org/10.1117/1.3285584}.
+\url{https://doi.org/10.1117/1.3285584}.}
 
 \leavevmode\hypertarget{ref-greening2018}{}%
-{[}9{]} G.J.
+\CSLLeftMargin{{[}9{]} }
+\CSLRightInline{G.J.
 Greening, K.P.
@@
 Cato, T.J.
-Muldoon, Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy, Biomedical Optics Express.
-9 (n.d.) 2871--2886.
-\url{https://doi.org/\%7B10.1364/BOE.9.002871\%7D}.
+Muldoon, {Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy}, {Biomedical Optics Express}.
+{9} (2018) 2871--2886.
+\url{https://doi.org/\%7B10.1364/BOE.9.002871\%7D}.}
 
 \leavevmode\hypertarget{ref-sio2016}{}%
-{[}10{]} T.T.
+\CSLLeftMargin{{[}10{]} }
+\CSLRightInline{T.T.
 Sio, P.J.
@@
 Bearden III, R.C.
-Miller, Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}), Supportive Care in Cancer.
-24 (n.d.) 3847--3855.
-\url{https://doi.org/\%7B10.1007/s00520-016-3213-3\%7D}.
+Miller, {Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {{[}}alliance{]})}, {Supportive Care in Cancer}.
+{24} (2016) 3847--3855.
+\url{https://doi.org/\%7B10.1007/s00520-016-3213-3\%7D}.}
 
 \leavevmode\hypertarget{ref-kamstra2015}{}%
-{[}11{]} J.I.
+\CSLLeftMargin{{[}11{]} }
+\CSLRightInline{J.I.
 Kamstra, P.U.
@@
 Roodenburg, J.A.
-Langendijk, Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study, Oral Oncology.
-51 (n.d.) 548--555.
-\url{https://doi.org/\%7B10.1016/j.oraloncology.2015.01.016\%7D}.
+Langendijk, {Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study}, {Oral Oncology}.
+{51} (2015) 548--555.
+\url{https://doi.org/\%7B10.1016/j.oraloncology.2015.01.016\%7D}.}
 
 \leavevmode\hypertarget{ref-wagenmakers2008}{}%
-{[}12{]} E.-J.
+\CSLLeftMargin{{[}12{]} }
+\CSLRightInline{E.-J.
 Wagenmakers, M.
@@
 181--207.
-\url{https://doi.org/10.1007/978-0-387-09612-4_9}.
+\url{https://doi.org/10.1007/978-0-387-09612-4_9}.}
 
 \leavevmode\hypertarget{ref-gueorguieva2004}{}%
-{[}13{]} R.
+\CSLLeftMargin{{[}13{]} }
+\CSLRightInline{R.
 Gueorguieva, J.H.
-Krystal, Move over ANOVA - Progress in analyzing repeated-measures data and its reflection in papers published in the archives of general psychiatry, Archives of General Psychiatry.
+Krystal, {Move over ANOVA - Progress in analyzing repeated-measures data and its reflection in papers published in the archives of general psychiatry}, Archives of General Psychiatry.
 61 (2004) 310--317.
-\url{https://doi.org/10.1001/archpsyc.61.3.310}.
+\url{https://doi.org/10.1001/archpsyc.61.3.310}.}
 
 \leavevmode\hypertarget{ref-schober2018}{}%
-{[}14{]} P.
+\CSLLeftMargin{{[}14{]} }
+\CSLRightInline{P.
 Schober, T.R.
@@
 127 (2018) 569--575.
-\url{https://doi.org/10.1213/ane.0000000000003511}.
+\url{https://doi.org/10.1213/ane.0000000000003511}.}
 
 \leavevmode\hypertarget{ref-pinheiro2006}{}%
-{[}15{]} J. Pinheiro, D. Bates, Mixed-effects models in S and S-PLUS, Springer Science \& Business Media, 2006. \url{https://doi.org/https://doi.org/10.1007/b98882}.
+\CSLLeftMargin{{[}15{]} }
+\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science \& Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}
 
 \leavevmode\hypertarget{ref-vishwanath2009}{}%
-{[}16{]} K.
+\CSLLeftMargin{{[}16{]} }
+\CSLRightInline{K.
 Vishwanath, H.
@@
 11 (2009) 889--900.
-\url{https://doi.org/10.1593/neo.09580}.
+\url{https://doi.org/10.1593/neo.09580}.}
 
 \leavevmode\hypertarget{ref-dennis2019}{}%
-{[}17{]} B.
+\CSLLeftMargin{{[}17{]} }
+\CSLRightInline{B.
 Dennis, J.M.
@@
 Taper, S.R.
-Lele, Errors in Statistical Inference Under Model Misspecification: Evidence, Hypothesis Testing, and AIC, Frontiers in Ecology and Evolution.
-7 (n.d.).
-\url{https://doi.org/\%7B10.3389/fevo.2019.00372\%7D}.
+Lele, {Errors in Statistical Inference Under Model Misspecification: Evidence, Hypothesis Testing, and AIC}, {Frontiers in Ecology and Evolution}.
+{7} (2019).
+\url{https://doi.org/\%7B10.3389/fevo.2019.00372\%7D}.}
 
 \leavevmode\hypertarget{ref-wang2019}{}%
-{[}18{]} B. Wang, Z. Zhou, H. Wang, X.M. Tu, C. Feng, The p-value and model specification in statistics, General Psychiatry. 32 (n.d.). \url{https://doi.org/\%7B10.1136/gpsych-2019-100081\%7D}.
+\CSLLeftMargin{{[}18{]} }
+\CSLRightInline{B.
+Wang, Z.
+Zhou, H.
+Wang, X.M.
+Tu, C.
+Feng, {The p-value and model specification in statistics}, {General Psychiatry}.
+{32} (2019).
+\url{https://doi.org/\%7B10.1136/gpsych-2019-100081\%7D}.}
 
 \leavevmode\hypertarget{ref-liu2010}{}%
-{[}19{]} C.
+\CSLLeftMargin{{[}19{]} }
+\CSLRightInline{C.
 Liu, T.P.
@@
 18 (2010) 1724--1730.
-\url{https://doi.org/10.1038/mt.2010.127}.
+\url{https://doi.org/10.1038/mt.2010.127}.}
 
 \leavevmode\hypertarget{ref-halsey2015}{}%
-{[}20{]} L.G.
+\CSLLeftMargin{{[}20{]} }
+\CSLRightInline{L.G.
 Halsey, D.
@@
 Vowler, G.B.
-Drummond, The fickle P value generates irreproducible results, Nature Methods.
-12 (n.d.) 179--185.
-\url{https://doi.org/\%7B10.1038/nmeth.3288\%7D}.
+Drummond, {The fickle P value generates irreproducible results}, {Nature Methods}.
+{12} (2015) 179--185.
+\url{https://doi.org/\%7B10.1038/nmeth.3288\%7D}.}
 
 \leavevmode\hypertarget{ref-abdi2010}{}%
-{[}21{]} H. Abdi, Holm's Sequential Bonferroni Procedure, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.
+\CSLLeftMargin{{[}21{]} }
+\CSLRightInline{H. Abdi, {Holm's Sequential Bonferroni Procedure}, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.}
 
 \leavevmode\hypertarget{ref-nakagawa2004}{}%
-{[}22{]} S.
-Nakagawa, A farewell to Bonferroni: the problems of low statistical power and publication bias, Behavioral Ecology.
-15 (n.d.) 1044--1045.
-\url{https://doi.org/\%7B10.1093/beheco/arh107\%7D}.
+\CSLLeftMargin{{[}22{]} }
+\CSLRightInline{S.
+Nakagawa, {A farewell to Bonferroni: the problems of low statistical power and publication bias}, {Behavioral Ecology}.
+{15} (2004) 1044--1045.
+\url{https://doi.org/\%7B10.1093/beheco/arh107\%7D}.}
 
 \leavevmode\hypertarget{ref-gelman2012}{}%
-{[}23{]} A.
+\CSLLeftMargin{{[}23{]} }
+\CSLRightInline{A.
 Gelman, J.
 Hill, M.
-Yajima, Why We (Usually) Don't Have to Worry About Multiple Comparisons, Journal of Research on Educational Effectiveness.
-5 (n.d.) 189--211.
-\url{https://doi.org/\%7B10.1080/19345747.2011.618213\%7D}.
+Yajima, {Why We (Usually) Don't Have to Worry About Multiple Comparisons}, {Journal of Research on Educational Effectiveness}.
+{5} (2012) 189--211.
+\url{https://doi.org/\%7B10.1080/19345747.2011.618213\%7D}.}
 
 \leavevmode\hypertarget{ref-albers2019}{}%
-{[}24{]} C. Albers, The problem with unadjusted multiple and sequential statistical testing, Nature Communications. 10 (n.d.). \url{https://doi.org/\%7B10.1038/s41467-019-09941-0\%7D}.
+\CSLLeftMargin{{[}24{]} }
+\CSLRightInline{C. Albers, {The problem with unadjusted multiple and sequential statistical testing}, {Nature Communications}. {10} (2019). \url{https://doi.org/\%7B10.1038/s41467-019-09941-0\%7D}.}
 
 \leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
-{[}25{]} C.
+\CSLLeftMargin{{[}25{]} }
+\CSLRightInline{C.
 Ugrinowitsch, G.W.
@@
 36 (2004) 2144--2148.
-\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.
+\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.}
 
 \leavevmode\hypertarget{ref-huynh1976}{}%
-{[}26{]} H.
+\CSLLeftMargin{{[}26{]} }
+\CSLRightInline{H.
 Huynh, L.S.
@@
 1 (1976) 69--82.
-\url{https://doi.org/10.3102/10769986001001069}.
+\url{https://doi.org/10.3102/10769986001001069}.}
 
 \leavevmode\hypertarget{ref-greenhouse1959}{}%
-{[}27{]} S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.
+\CSLLeftMargin{{[}27{]} }
+\CSLRightInline{S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.}
 
 \leavevmode\hypertarget{ref-haverkamp2017}{}%
-{[}28{]} N.
+\CSLLeftMargin{{[}28{]} }
+\CSLRightInline{N.
 Haverkamp, A.
-Beauducel, Violation of the Sphericity Assumption and Its Effect on Type-I Error Rates in Repeated Measures ANOVA and Multi-Level Linear Models (MLM), Frontiers in Psychology.
-8 (n.d.).
-\url{https://doi.org/\%7B10.3389/fpsyg.2017.01841\%7D}.
+Beauducel, {Violation of the Sphericity Assumption and Its Effect on Type-I Error Rates in Repeated Measures ANOVA and Multi-Level Linear Models (MLM)}, {Frontiers in Psychology}.
+{8} (2017).
+\url{https://doi.org/\%7B10.3389/fpsyg.2017.01841\%7D}.}
 
 \leavevmode\hypertarget{ref-keselman2001}{}%
-{[}29{]} H.
+\CSLLeftMargin{{[}29{]} }
+\CSLRightInline{H.
 Keselman, J.
 Algina, R.
-Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematica \& Statistical Psychology.
-54 (n.d.) 1--20.
-\url{https://doi.org/\%7B10.1348/000711001159357\%7D}.
+Kowalchuk, {The analysis of repeated measures designs: A review}, {British Journal of Mathematica \& Statistical Psychology}.
+{54} (2001) 1--20.
+\url{https://doi.org/\%7B10.1348/000711001159357\%7D}.}
 
 \leavevmode\hypertarget{ref-charan2013}{}%
-{[}30{]} J. Charan, N. Kantharia, How to calculate sample size in animal studies?, Journal of Pharmacology and Pharmacotherapeutics. 4 (2013) 303--306. \url{https://doi.org/10.4103/0976-500X.119726}.
+\CSLLeftMargin{{[}30{]} }
+\CSLRightInline{Jaykaran.
+Charan, N.
+Kantharia, {How to calculate sample size in animal studies?}, Journal of Pharmacology and Pharmacotherapeutics.
+4 (2013) 303--306.
+\url{https://doi.org/10.4103/0976-500X.119726}.}
 
 \leavevmode\hypertarget{ref-barr2013}{}%
-{[}31{]} D.J.
+\CSLLeftMargin{{[}31{]} }
+\CSLRightInline{D.J.
 Barr, R.
@@
 Scheepers, H.J.
-Tily, Random effects structure for confirmatory hypothesis testing: Keep it maximal, JournalL of Memory and Language.
-68 (n.d.) 255--278.
-\url{https://doi.org/\%7B10.1016/j.jml.2012.11.001\%7D}.
+Tily, {Random effects structure for confirmatory hypothesis testing: Keep it maximal}, {JournalL of Memory and Language}.
+{68} (2013) 255--278.
+\url{https://doi.org/\%7B10.1016/j.jml.2012.11.001\%7D}.}
 
 \leavevmode\hypertarget{ref-rose2012}{}%
-{[}32{]} N.L.
+\CSLLeftMargin{{[}32{]} }
+\CSLRightInline{N.L.
 Rose, H.D.
@@
 Turner, G.L.
-Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, uk, Geochimica et Cosmochimica Acta.
+Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, UK, Geochimica Et Cosmochimica Acta.
 82 (2012) 113--135.
-\url{https://doi.org/10.1016/j.gca.2010.12.026}.
+\url{https://doi.org/10.1016/j.gca.2010.12.026}.}
 
 \leavevmode\hypertarget{ref-pedersen2019}{}%
-{[}33{]} E.J. Pedersen, D.L. Miller, G.L. Simpson, N. Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, Peerj. 7 (2019). \url{https://doi.org/10.7717/peerj.6876}.
+\CSLLeftMargin{{[}33{]} }
+\CSLRightInline{E.J.
+Pedersen, D.L.
+Miller, G.L.
+Simpson, N.
+Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, Peerj.
+7 (2019).
+\url{https://doi.org/10.7717/peerj.6876}.}
 
 \leavevmode\hypertarget{ref-simpson2018}{}%
-{[}34{]} G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.
+\CSLLeftMargin{{[}34{]} }
+\CSLRightInline{G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.}
 
 \leavevmode\hypertarget{ref-yang2012}{}%
-{[}35{]} L.
+\CSLLeftMargin{{[}35{]} }
+\CSLRightInline{L.
 Yang, G.
@@
 Wang, G.
-Song, Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality, BMC Medical Research Methodology.
-12 (n.d.).
-\url{https://doi.org/\%7B10.1186/1471-2288-12-165\%7D}.
+Song, {Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality}, {BMC Medical Research Methodology}.
+{12} (2012).
+\url{https://doi.org/\%7B10.1186/1471-2288-12-165\%7D}.}
 
 \leavevmode\hypertarget{ref-beck1998}{}%
-{[}36{]} N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. 42 (n.d.) 596--627. \url{https://doi.org/\%7B10.2307/2991772\%7D}.
+\CSLLeftMargin{{[}36{]} }
+\CSLRightInline{N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. (1998) 596--627.}
 
 \leavevmode\hypertarget{ref-wood2017}{}%
-{[}37{]} S.N. Wood, Generalized additive models: An introduction with , second edition, CRC Press LLC, Philadelphia, PA, 2017.
+\CSLLeftMargin{{[}37{]} }
+\CSLRightInline{S.N. Wood, Generalized additive models: An introduction with {}, second edition, CRC Press LLC, Philadelphia, PA, 2017.}
 
 \leavevmode\hypertarget{ref-r}{}%
-{[}38{]} R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.
+\CSLLeftMargin{{[}38{]} }
+\CSLRightInline{R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.}
 
 \leavevmode\hypertarget{ref-wood2016}{}%
-{[}39{]} S.N.
+\CSLLeftMargin{{[}39{]} }
+\CSLRightInline{S.N.
 Wood, N.
 Pya, B.
-Saefken, Smoothing Parameter and Model Selection for General Smooth Models, Journal of the American Statistical Association.
-111 (n.d.) 1548--1563.
-\url{https://doi.org/\%7B10.1080/01621459.2016.1180986\%7D}.
+Saefken, {Smoothing Parameter and Model Selection for General Smooth Models}, {Journal of the American Statistical Association}.
+{111} (2016) 1548--1563.
+\url{https://doi.org/\%7B10.1080/01621459.2016.1180986\%7D}.}
 
 \leavevmode\hypertarget{ref-west2014}{}%
-{[}40{]} B.T.
+\CSLLeftMargin{{[}40{]} }
+\CSLRightInline{B.T.
 West, K.B.
@@
 Galecki, Linear mixed models: A practical guide using statistical software, second edition, Taylor \& Francis, 2014.
-\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.
+\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.}
 
 \leavevmode\hypertarget{ref-wolfinger1996}{}%
-{[}41{]} R.D.
+\CSLLeftMargin{{[}41{]} }
+\CSLRightInline{R.D.
 Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics.
 1 (1996) 205--230.
-\url{http://www.jstor.org/stable/1400366}.
+\url{http://www.jstor.org/stable/1400366}.}
 
 \leavevmode\hypertarget{ref-weiss2005}{}%
-{[}42{]} R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://books.google.com/books?id=MQ/_bvWDPsEAC}.
+\CSLLeftMargin{{[}42{]} }
+\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://books.google.com/books?id=MQ/_bvWDPsEAC}.}
 
 \leavevmode\hypertarget{ref-geisser1958}{}%
-{[}43{]} S.
+\CSLLeftMargin{{[}43{]} }
+\CSLRightInline{S.
 Geisser, S.W.
-Greenhouse, An Extension of Box's Results on the Use of the \(F\) Distribution in Multivariate Analysis, The Annals of Mathematical Statistics.
+Greenhouse, {An Extension of Box's Results on the Use of the \(F\) Distribution in Multivariate Analysis}, The Annals of Mathematical Statistics.
 29 (1958) 885--891.
-\url{https://doi.org/10.1214/aoms/1177706545}.
+\url{https://doi.org/10.1214/aoms/1177706545}.}
 
 \leavevmode\hypertarget{ref-maxwell2017}{}%
-{[}44{]} S.E.
+\CSLLeftMargin{{[}44{]} }
+\CSLRightInline{S.E.
 Maxwell, H.D.
@@
 Kelley, Designing experiments and analyzing data: A model comparison perspective, third edition, Taylor \& Francis, 2017.
-\url{https://books.google.com/books?id=NmFQDwAAQBAJ}.
+\url{https://books.google.com/books?id=NmFQDwAAQBAJ}.}
 
 \leavevmode\hypertarget{ref-molenberghs2004}{}%
-{[}45{]} G.
+\CSLLeftMargin{{[}45{]} }
+\CSLRightInline{G.
 Molenberghs, H.
@@
 Mallinckrodt, R.
-Carroll, Analyzing incomplete longitudinal clinical trial data, Biostatistics.
-5 (n.d.) 445--464.
-\url{https://doi.org/\%7B10.1093/biostatistics/kxh001\%7D}.
+Carroll, {Analyzing incomplete longitudinal clinical trial data}, {Biostatistics}.
+{5} (2004) 445--464.
+\url{https://doi.org/\%7B10.1093/biostatistics/kxh001\%7D}.}
 
 \leavevmode\hypertarget{ref-scheffer2002}{}%
-{[}46{]} J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.
+\CSLLeftMargin{{[}46{]} }
+\CSLRightInline{J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.}
 
 \leavevmode\hypertarget{ref-potthoff2006}{}%
-{[}47{]} R.F.
+\CSLLeftMargin{{[}47{]} }
+\CSLRightInline{R.F.
 Potthoff, G.E.
@@
 Pieper, V.
-Hasselblad, Can one assess whether missing data are missing at random in medical studies?, STATISTICAL METHODS IN MEDICAL RESEARCH.
-15 (n.d.) 213--234.
-\url{https://doi.org/\%7B10.1191/0962280206sm448oa\%7D}.
+Hasselblad, {Can one assess whether missing data are missing at random in medical studies?}, {STATISTICAL METHODS IN MEDICAL RESEARCH}.
+{15} (2006) 213--234.
+\url{https://doi.org/\%7B10.1191/0962280206sm448oa\%7D}.}
 
 \leavevmode\hypertarget{ref-ma2012}{}%
-{[}48{]} Y.
+\CSLLeftMargin{{[}48{]} }
+\CSLRightInline{Y.
 Ma, M.
 Mazumdar, S.G.
-Memtsoudis, Beyond Repeated-Measures Analysis of Variance Advanced Statistical Methods for the Analysis of Longitudinal Data in Anesthesia Research, Regional Anesthesia and Pain Medicine.
-37 (n.d.) 99--105.
-\url{https://doi.org/\%7B10.1097/AAP.0b013e31823ebc74\%7D}.
+Memtsoudis, {Beyond Repeated-Measures Analysis of Variance Advanced Statistical Methods for the Analysis of Longitudinal Data in Anesthesia Research}, {Regional Anesthesia and Pain Medicine}.
+{37} (2012) 99--105.
+\url{https://doi.org/\%7B10.1097/AAP.0b013e31823ebc74\%7D}.}
 
 \leavevmode\hypertarget{ref-nlme}{}%
-{[}49{]} J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, nlme: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.
+\CSLLeftMargin{{[}49{]} }
+\CSLRightInline{J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, {nlme}: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.}
 
 \leavevmode\hypertarget{ref-nelder1972}{}%
-{[}50{]} J.A. Nelder, R.W.M. Wedderburn, Generalized linear models, Journal of the Royal Statistical Society. Series A (General). 135 (1972) 370--384. \url{http://www.jstor.org/stable/2344614}.
+\CSLLeftMargin{{[}50{]} }
+\CSLRightInline{J.A.
+Nelder, R.W.M.
+Wedderburn, Generalized linear models, Journal of the Royal Statistical Society.
+Series A (General).
+135 (1972) 370--384.
+\url{http://www.jstor.org/stable/2344614}.}
 
 \leavevmode\hypertarget{ref-hastie1987}{}%
-{[}51{]} T.
+\CSLLeftMargin{{[}51{]} }
+\CSLRightInline{T.
 Hastie, R.
@@
 82 (1987) 371--386.
-\url{https://doi.org/10.1080/01621459.1987.10478440}.
+\url{https://doi.org/10.1080/01621459.1987.10478440}.}
 
 \leavevmode\hypertarget{ref-hefley2017}{}%
-{[}52{]} T.J.
+\CSLLeftMargin{{[}52{]} }
+\CSLRightInline{T.J.
 Hefley, K.M.
@@
 Williams, M.B.
-Hooten, The basis function approach for modeling autocorrelation in ecological data, ECOLOGY.
-98 (n.d.) 632--646.
-\url{https://doi.org/\%7B10.1002/ecy.1674\%7D}.
+Hooten, {The basis function approach for modeling autocorrelation in ecological data}, {ECOLOGY}.
+{98} (2017) 632--646.
+\url{https://doi.org/\%7B10.1002/ecy.1674\%7D}.}
 
 \leavevmode\hypertarget{ref-wegman1983}{}%
-{[}53{]} E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.
+\CSLLeftMargin{{[}53{]} }
+\CSLRightInline{E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.}
 
 \leavevmode\hypertarget{ref-gratia}{}%
-{[}54{]} G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for gams fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.
+\CSLLeftMargin{{[}54{]} }
+\CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
 
 \leavevmode\hypertarget{ref-harezlak2018}{}%
-{[}55{]} J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with r, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.
+\CSLLeftMargin{{[}55{]} }
+\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with r, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
+
+\end{CSLReferences}
 
@@
 
+\begin{lstlisting}
+## Error: The given dimensions cannot hold all panels. Please increase `ncol` or `nrow`
+\end{lstlisting}
+
 \begin{figure}[H]
@@
 From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\textasciitilde1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 87\% of the variance data.
+
+\begin{lstlisting}
+## Error: The given dimensions cannot hold all panels. Please increase `ncol` or `nrow`
+\end{lstlisting}
 

### 2021-05-20 Manuscripts/Manuscript by chapters -v1/Full_document.Rmd
Styling upgraded, author issue fixed after putting an issue on github

@@
 output:  
+  bookdown::html_document2:
+    css: "style.css" #style for the HTML document
+  bookdown::word_document2:
+    fig_caption: yes #figure caption
+    keep_md: yes
   bookdown::pdf_document2:
-    pandoc_args: --listings
-    keep_tex: yes
-    fig_caption: yes
+    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
+    pandoc_args: --listings #calls the listings package to fit code in the page margins
+    keep_tex: yes #keep LaTeX file for submission
+    fig_caption: yes #allows captions in figures
     extra_dependencies:
-      subfig: null
-      breqn: null
-      caption: ["font={small}"]
-      float: null
+      subfig: null #allows for subfigures
+      breqn: null #line breaks for long equations
+      caption: ["font={small}"] #size of the figure captions
+      float: null #allows for control of placement of figures
     includes:
-      in_header: preamble.sty
-  bookdown::word_document2:
-    fig_caption: yes
-    keep_md: yes
-  bookdown::html_document2: default
-  link-citations: yes
-  css: style.css
+      in_header: preamble.sty #additional LaTeX formatting
+csl: elsevier-with-titles.csl #style for references
+bibliography: refs.bib #references
+link-citations: yes #adds links to the citations
 '': default
-csl: elsevier-with-titles.csl
-bibliography: refs.bib
 ---
@@
 knitr::opts_chunk$set(fig.pos = 'H', out.extra = "",error=TRUE)
+
+#this function will remove the default "author information" that is the Pandoc Latex template.
+Otherwise, and additional "and" will appear at the end of the author list.Or, the custom template that is in the YAML header and that has the "author" line deleted may be used as well.
+An explanation in detail is provided at:
+#https://github.com/rstudio/bookdown/issues/1158
+remove_author <- function(x) {
+  # identify empty author line
+  i <- grep("^\\\\author\\{\\}$", x)
+  # be sure it is the one pandoc inserts
+  if(length(i) != 0 && grepl('^\\\\date\\{', x[i+1])) x <- x[-i]
+  x
+}
+options(bookdown.post.latex = remove_author)
 ```

### 2021-05-20 Manuscripts/Manuscript by chapters -v1/Full_document.tex
Styling upgraded, author issue fixed after putting an issue on github

@@
 \author[1]{Timothy J. Muldoon}
-\renewcommand\Authands {} %added because an additional "and" appeared after last author
 \affil[1]{\footnotesize Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
@@
 \subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
-\author{}
 \date{\vspace{-2.5em}}
@@
 Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response {[}1--4{]}, antibody expression {[}5,6{]}, and cell metabolism {[}7,8{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}9{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}10{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}11{]}.
+Tumor response {[}\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}{]}, antibody expression {[}\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}{]}, and cell metabolism {[}\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
+Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}\protect\hyperlink{ref-greening2018}{9}{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}\protect\hyperlink{ref-sio2016}{10}{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}\protect\hyperlink{ref-kamstra2015}{11}{]}.
 
 Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}12{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}13,14{]}.
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}15{]}.
+The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}\protect\hyperlink{ref-wagenmakers2008}{12}{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
+The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}{]}.
+
+The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
 This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
 In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}1,8,16{]}.
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
@@
 Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}17{]}.
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}18{]}.
-
-Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}19,20{]}, consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}21{]}), but they in turn reduce statistical power (1-\(\beta\)){[}22{]}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) {[}23,24{]}.
+In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}.
+For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}\protect\hyperlink{ref-wang2019}{18}{]}.
+
+Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}{]}, consequently biasing the conclusions of the study.
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}\protect\hyperlink{ref-abdi2010}{21}{]}), but they in turn reduce statistical power (1-\(\beta\)){[}\protect\hyperlink{ref-nakagawa2004}{22}{]}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) {[}\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}{]}.
 Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
 
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}25{]}.
-Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}26,27{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}28{]}, and group sizes {[}29{]}.
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}30{]} which might cause the corrections for lack of compound symmetry to be ineffective.
+On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
+Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and group sizes {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
+In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}\protect\hyperlink{ref-charan2013}{30}{]} which might cause the corrections for lack of compound symmetry to be ineffective.
 
@@
 In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}13{]}.
+When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}\protect\hyperlink{ref-gueorguieva2004}{13}{]}.
 This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
@@
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}8,16{]}.
+The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
 Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}15{]}.
-However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}13,31{]}.
-And even more importantly, LMEMs also expect a linear relationship between the response and time {[}15{]}, making them unsuitable to analyze non-linear data.
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
+However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}.
+And even more importantly, LMEMs also expect a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}, making them unsuitable to analyze non-linear data.
 
@@
 Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}32--34{]}, health-environment interactions {[}35{]} and the dynamics of government in political science {[}36{]} .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}37{]} and 3) can easily accommodate missing observations.
+Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]} .
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]} and 3) can easily accommodate missing observations.
 Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
@@
 The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, \(\textsf{R}\){[}38{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}37,39{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}28{]}.
+In particular, \(\textsf{R}\){[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
 Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
@@
 Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, using simulated data that reproduces patterns in previously reported studies {[}16{]} we illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
+Third, using simulated data that reproduces patterns in previously reported studies {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} we illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
 The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
@@
 
-A linear mixed model (LMEM) is a class of statistical model that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish {[}15,40{]}.
+A linear mixed model (LMEM) is a class of statistical model that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish {[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}.
 A LMEM with interaction between time and treatment for a longitudinal study can be written as:
@@
 The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence,the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}15{]}.
+In essence,the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
 However,the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
@@
 In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}41{]} (For an in-depth analysis of the covariance matrix see {[}40,42{]}).
+The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{41}{]} (For an in-depth analysis of the covariance matrix see {[}\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}{]}).
 
 In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}26,42,43{]}.
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}44{]}; and it is higher between consecutive measurements {[}13,25{]}.
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}26,27{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}28{]}, and they are not robust if the group sizes are unbalanced {[}29{]}.
+Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}{]}.
+However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}\protect\hyperlink{ref-maxwell2017}{44}{]}; and it is higher between consecutive measurements {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and they are not robust if the group sizes are unbalanced {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
 Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
 
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}15{]}.
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
 Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
@@
 Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}45{]}.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
 Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}42{]}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}46{]}.
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}47{]}, and in the case of MNAR the missing values are dependent on their value.
+Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}\protect\hyperlink{ref-weiss2005}{42}{]}.
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{46}{]}.
+If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{47}{]}, and in the case of MNAR the missing values are dependent on their value.
 An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}48{]}.
-
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}40{]}.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{48}{]}.
+
+In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}\protect\hyperlink{ref-west2014}{40}{]}.
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
@@
 To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}) is fitted to each group, using \(\textsf{R}\){[}38{]} and the package \emph{nlme}{[}49{]}.
+An rm-ANOVA model (Equation \eqref{eq:linear-model}) is fitted to each group, using \(\textsf{R}\){[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme}{[}\protect\hyperlink{ref-nlme}{49}{]}.
 Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}, A).
@@
 
-Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors{[}50{]}.
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}34,37,51{]}.
+Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors{[}\protect\hyperlink{ref-nelder1972}{50}{]}.
+In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{51}{]}.
 A GAM model can be written as:
@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}36{]}.
+One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}\protect\hyperlink{ref-beck1998}{36}{]}.
 In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic, and causes bias at the endpoints of the time period.
@@
 The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis functions} expansions of the covariates and by estimating random coefficients for these basis functions.
-A \emph{basis} is a set of functions that spans the space where the smooths that approximate \(f(x_t \mid \beta_j)\) exist {[}34{]}.
+A \emph{basis} is a set of functions that spans the space where the smooths that approximate \(f(x_t \mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
 For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}52{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{52}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
 In the case of Equation \eqref{eq:GAM}, the basis function is \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-A commonly used \emph{basis function} is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness {[}34,37{]}.
-Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}53{]}.
+A commonly used \emph{basis function} is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
+Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
 Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3(C) in {[}16{]}.
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3(C) in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
 In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
@@
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in {[}52{]}.
+A detailed description on basis functions and correlations can be found in {[}\protect\hyperlink{ref-hefley2017}{52}{]}.
 
@@
 The smooth is constructed using gaussian process smooths, which is indicated by \passthrough{\lstinline!bs="gp"!}.
-These splines are used to model temporal trends and might be particularly suited for long-term studies where the correlation between measurements changes as a function of the time intervals {[}34{]}.
-The parametric term \emph{Group} is added to quantify differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}37{]}.
+These splines are used to model temporal trends and might be particularly suited for long-term studies where the correlation between measurements changes as a function of the time intervals {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
+The parametric term \emph{Group} is added to quantify differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}.
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot},B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}{[}54{]}.
-A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in {[}37{]} and {[}55{]}.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}{[}\protect\hyperlink{ref-gratia}{54}{]}.
+A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{55}{]}.
 

### 2021-05-21 Manuscripts/Manuscript by chapters -v1/02-Challenges.Rmd
Updated color aesthetics

--- 2021-05-14 Manuscripts/Manuscript by chapters -v1/02-Challenges.Rmd
@@
 
-Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates an effect due to treatment.
+Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect.
 In other words, Group 1 can be thought as a "Control" group and Group 2 as a "Treatment" group.
@@
         text=element_text(size=30))+
-    scale_color_aaas()
+    thm
 
@@
                   size = "Population")) +
+    guides(color = guide_legend(override.aes = list(size = 2)))+
     scale_size_manual(name = "Predictions", 

### 2021-05-21 Manuscripts/Manuscript by chapters -v1/03-GAM Theory.Rmd
Updated color aesthetics

@@
   theme_classic()+
-  scale_color_aaas()
+  thm
 
@@
             aes(x=Basis,
-                y=Basis,
-                fill=cof))+
-  geom_tile(aes(color='black'),
-            size=sz+1,
-            show.legend = FALSE)+
-  geom_tile(size=sz+1)+
-  scale_fill_gradient(low = "white", high = "red")+
+                y=Basis))+
+  geom_tile(aes(fill = cof), 
+            colour = "black") +
+    scale_fill_gradient(low = "white",
+                        high = "#B50A2AFF")+ #color picked from KikiMedium
   labs(x='Basis',
@@
   geom_text(aes(label=round(cof,2)),
-            size=10,
+            size=7,
             show.legend = FALSE)+
@@
                 y=smooth),
-            color="#B15731",
+            color="#6C581DFF",
             size=sz+1)+

### 2021-05-21 Manuscripts/Manuscript by chapters -v1/04-Longitudinal analysis with GAMs.Rmd
Working on code chunks in PDF
Updated color aesthetics

@@
   theme_classic()+
-  scale_color_aaas()+
+  thm+
     scale_x_continuous(breaks=c(0,5,10))+
@@
 sd <- 10 #approximate sd from paper
-set.seed(1) #set seed for reproducibility
 df <- 6
@@
   )+
-  scale_color_aaas()+
+  thm+
     scale_x_continuous(breaks=c(0,2,5,7,10))
@@
   )+
-      scale_color_aaas()+
-  scale_fill_aaas()
+      thm+
+  thm1
 
@@
   )+
-      scale_color_aaas()+
-  scale_fill_aaas()
+  thm+
+  thm1
 
@@
      )&
-  scale_color_aaas()
+  thm
 
@@
                 alpha = 0.9,
-                fill='#fdf05d') +
-    geom_line(color='#755006',size=1) +
+                fill='#278B9AFF') +
+    geom_line(color='#E75B64FF',size=1) +
     geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
@@
                 aes(ymin =0, ymax =lower),
-                alpha = 0.5,
-                fill='#003049') +
+                alpha = 0.9,
+                fill='#4C413FFF') +
     geom_ribbon(data=comp_StO2_full %>%
@@
                     aes(ymin =0, ymax =upper),
-                alpha = 0.5,
-                fill='#c1121f') +
+                alpha = 0.9,
+                fill='#5A6F80FF') +
     facet_wrap(~ pair) +
@@
                 alpha = 0.9,
-                fill='#fdf05d') +
-    geom_line(color='#755006',size=1) +
+                fill='#278B9AFF') +
+    geom_line(color='#E75B64FF',size=1) +
     geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
@@
                 aes(ymin =0, ymax =lower),
-                alpha = 0.5,
-                fill='#003049') +
+                alpha = 0.9,
+                fill='#4C413FFF') +
     geom_ribbon(data=comp_StO2_missing %>%

### 2021-05-21 Manuscripts/Manuscript by chapters -v1/06-Appendix.Rmd
Updated color aesthetics

@@
      text=element_text(size=30))+
-    scale_color_aaas()
+    thm
 
@@
      )
+
 ```

### 2021-05-21 Manuscripts/Manuscript by chapters -v1/Full_document.Rmd
Updated color aesthetics

@@
 output:  
-  bookdown::html_document2:
-    css: "style.css" #style for the HTML document
-  bookdown::word_document2:
-    fig_caption: yes #figure caption
-    keep_md: yes
   bookdown::pdf_document2:
@@
       in_header: preamble.sty #additional LaTeX formatting
+  bookdown::word_document2:
+    fig_caption: yes #figure caption
+    keep_md: yes
+  bookdown::html_document2:
+    css: "style.css" #style for the HTML document
 csl: elsevier-with-titles.csl #style for references
@@
 library(gratia)
+library(ghibli)
 set.seed(1) #set seed for reproducibility
@@
 knitr::opts_chunk$set(fig.pos = 'H', out.extra = "",error=TRUE)
+
+thm<-scale_color_ghibli_d("PonyoMedium",direction=1) #global theme for plotting
+thm1<-scale_fill_ghibli_d("PonyoMedium",direction=1) #global theme for plotting
 

### 2021-05-21 Manuscripts/Manuscript by chapters -v1/Full_document.tex
Updated color aesthetics

@@
 
-Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates an effect due to treatment.
+Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect.
 In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group.
@@
 ## full convergence after 6 iterations.
-## Gradient range [-4.142968e-08,2.799316e-12]
-## (score 440.4108 & scale 414.2575).
-## Hessian positive definite, eigenvalue range [0.04576008,49.0005].
+## Gradient range [-0.0001585015,0.0008415702]
+## (score 436.939 & scale 387.386).
+## Hessian positive definite, eigenvalue range [0.0001600441,48.99916].
 ## Model rank =  5 / 5 
@@
 ##        k' edf k-index p-value    
-## s(Day) 4.00 1.31    0.26  <2e-16 ***
+## s(Day)  4   1    0.37  <2e-16 ***
 ## ---
@@
 ##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   21.929      2.035   10.77   <2e-16 ***
+## (Intercept)   21.578      1.968   10.96   <2e-16 ***
 ## ---
@@
 ##          edf Ref.df     F  p-value    
-## s(Day) 1.314  1.536 9.151 0.00253 **
+## s(Day) 1.002  1.004 21.54 1.09e-05 ***
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.105   Deviance explained = 11.7%
-## -REML = 440.41  Scale est. = 414.26    n = 100
+## R-sq.(adj) =  0.172   Deviance explained = 18.1%
+## -REML = 436.94  Scale est. = 387.39    n = 100
 \end{lstlisting}
@@
 ## Method: REML   Optimizer: outer newton
-## full convergence after 10 iterations.
-## Gradient range [-0.0001703751,9.561998e-05]
-## (score 418.612 & scale 270.7177).
-## Hessian positive definite, eigenvalue range [0.0001702821,48.50255].
+## full convergence after 8 iterations.
+## Gradient range [-0.0001335263,0.001505324]
+## (score 415.0769 & scale 254.693).
+## Hessian positive definite, eigenvalue range [9.414522e-05,48.49849].
 ## Model rank =  9 / 9 
@@
 ##                       k' edf k-index p-value    
-## s(Day):GroupControl   4.00 1.00    0.32  <2e-16 ***
-## s(Day):GroupTreatment 4.00 1.72    0.32  <2e-16 ***
+## s(Day):GroupControl    4   1    0.46  <2e-16 ***
+## s(Day):GroupTreatment  4   1    0.46  <2e-16 ***
 ## ---
@@
 ##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   21.929      1.645   13.33   <2e-16 ***
+## (Intercept)   21.578      1.596   13.52   <2e-16 ***
 ## ---
@@
 ##                         edf Ref.df      F p-value    
-## s(Day):GroupControl   1.001  1.001  4.099  0.0456 *  
-## s(Day):GroupTreatment 1.715  1.979 35.551  <2e-16 ***
+## s(Day):GroupControl   1.004  1.008  1.083   0.299    
+## s(Day):GroupTreatment 1.000  1.001 83.768  <2e-16 ***
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.415   Deviance explained = 43.1%
-## -REML = 418.61  Scale est. = 270.72    n = 100
+## R-sq.(adj) =  0.456   Deviance explained = 46.7%
+## -REML = 415.08  Scale est. = 254.69    n = 100
 \end{lstlisting}
@@
 ## full convergence after 9 iterations.
-## Gradient range [-1.003557e-07,3.562136e-08]
-## (score 362.7587 & scale 64.03804).
-## Hessian positive definite, eigenvalue range [0.9494021,48.08513].
+## Gradient range [-1.291119e-06,1.637752e-06]
+## (score 374.509 & scale 96.64781).
+## Hessian positive definite, eigenvalue range [0.02491842,48.04328].
 ## Model rank =  10 / 10 
@@
 ##                         k'  edf k-index p-value  
-## s(Day):GroupControl   4.00 3.83    1.02    0.52
-## s(Day):GroupTreatment 4.00 3.84    1.02    0.59
+## s(Day):GroupControl   4.00 3.84    0.85   0.075 .
+## s(Day):GroupTreatment 4.00 1.24    0.85   0.085 .
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
 \end{lstlisting}
@@
 ##                Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)       9.781      1.132   8.643 1.85e-13 ***
-## GroupTreatment   24.296      1.600  15.181  < 2e-16 ***
+## (Intercept)      10.503      1.390   7.554 2.87e-11 ***
+## GroupTreatment   22.150      1.966  11.266  < 2e-16 ***
 ## ---
@@
 ##                         edf Ref.df       F  p-value    
-## s(Day):GroupControl   3.825  3.971 16.81  <2e-16 ***
-## s(Day):GroupTreatment 3.835  3.974 78.84  <2e-16 ***
+## s(Day):GroupControl   3.843  3.976   7.975 8.99e-06 ***
+## s(Day):GroupTreatment 1.236  1.419 159.701  < 2e-16 ***
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.862   Deviance explained = 87.4%
-## -REML = 362.76  Scale est. = 64.038    n = 100
+## R-sq.(adj) =  0.794   Deviance explained = 80.6%
+## -REML = 374.51  Scale est. = 96.648    n = 100
 \end{lstlisting}
@@
 ##              df      AIC
-## gam_00  3.536147 891.1671
-## gam_01  4.980481 850.0698
-## gam1   10.945191 711.4662
+## gam_00 3.004277 883.7160
+## gam_01 4.008467 842.7602
+## gam1   8.395441 750.3446
 \end{lstlisting}
@@
 ##  2     0 Treatment     4
-##  3     2 Control       6
-##  4     2 Treatment     5
-##  5     5 Control       6
-##  6     5 Treatment     4
-##  7     7 Control       3
-##  8     7 Treatment     5
-##  9    10 Control       3
-## 10    10 Treatment     8
+##  3     2 Control       7
+##  4     2 Treatment     4
+##  5     5 Control       2
+##  6     5 Treatment     8
+##  7     7 Control       5
+##  8     7 Treatment     8
+##  9    10 Control       4
+## 10    10 Treatment     5
 \end{lstlisting}
@@
   )+
-      scale_color_aaas()+
-  scale_fill_aaas()
+      thm+
+  thm1
 \end{lstlisting}

### 2021-05-23 Manuscripts/Manuscript by chapters -v1/Full_document.Rmd
Aesthetics, need to add pairwise comparison code to the Appendix

@@
 library(ghibli)
+library(viridis)
 set.seed(1) #set seed for reproducibility
@@
 
-thm<-scale_color_ghibli_d("PonyoMedium",direction=1) #global theme for plotting
-thm1<-scale_fill_ghibli_d("PonyoMedium",direction=1) #global theme for plotting
+#global themes for plotting
+thm<-scale_color_viridis_d(option="magma",begin=0.1,end=0.7)
+thm1<-scale_fill_viridis_d(option="magma",begin=0.1,end=0.7)#global theme for plotting
 
-#this function will remove the default "author information" that is the Pandoc Latex template.
+#thm<-scale_color_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting
+#thm1<-scale_fill_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting
+
+#this function will remove the default "author information" that is the Pandoc LaTeX template.
 Otherwise, and additional "and" will appear at the end of the author list.Or, the custom template that is in the YAML header and that has the "author" line deleted may be used as well.

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/04-Longitudinal analysis with GAMs.Rmd
Working on pairwise comparisons section in the Appendix

@@
 #posthoc comparisons for the linear model
-library(multcomp)
+#library(multcomp)
 
@@
     geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.9,
-                fill='#278B9AFF') +
+                alpha = 0.5,
+                fill='#E1DD37FF') +
     geom_line(color='#E75B64FF',size=1) +
@@
                 aes(ymin =0, ymax =lower),
-                alpha = 0.9,
-                fill='#4C413FFF') +
+                alpha = 0.5,
+                fill='#30123BFF') +
     geom_ribbon(data=comp_StO2_full %>%
@@
                     aes(ymin =0, ymax =upper),
-                alpha = 0.9,
-                fill='#5A6F80FF') +
+                alpha = 0.5,
+                fill='#7A0403FF') +
     facet_wrap(~ pair) +
@@
                 aes(ymin =0, ymax =lower),
-                alpha = 0.9,
-                fill='#4C413FFF') +
+                alpha = 0.5,
+                fill='#30123BFF') +
     geom_ribbon(data=comp_StO2_missing %>%
@@
 
-## Determination of significance in GAMs for longitudinal data
+## Determination of significance in GAMs for longitudinal data{#GAM-sig}
 

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/06-Appendix.Rmd
Working on pairwise comparisons section in the Appendix

@@
 
+#### Pairwise comparisons of smooth confidence intervals
+
+The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \@ref(GAM-significance).
+In this case, the "design matrix" is used to estimate the pairwise comparisons (see Manuscript for details and associated references).
+Briefly, the "design matrix" from `gam1` is used to estimate a 95% confidence interval of the difference between the smooth terms for each group.
+This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
+
+```{r,pairwise-comp-workflow,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
+##Pairwise comparisons
+
+pdat <- expand.grid(Day = seq(0, 10, length = 400),
+                    Group = c('Control', 'Treatment'))
+
+#this function takes the model, grid and groups to be compared using the lpmatrix
+
+smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
+                        unconditional = FALSE) {
+    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
+    #Find columns in xp where the name contains "Control"
+    col1 <- grepl(g1, colnames(xp))
+    #Find columns in xp where the name contains 'Treatment'
+    col2 <- grepl(g2, colnames(xp))
+    #r1 <- newdata[[var]] == f1
+    #r2 <- newdata[[var]] == f2
+    row1 <- with(newdata, Group == g1)
+    row2 <- with(newdata, Group == g2)
+    ## difference rows of xp for data from comparison
+    X <- xp[row1, ] - xp[row2, ]
+    ## zero out cols of X related to splines for other lochs
+    X[, ! (col1 | col2)] <- 0
+    ## zero out the parametric cols
+    X[, !grepl('^s\\(', colnames(xp))] <- 0
+    dif <- X %*% coef(model)
+    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
+    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
+    upr <- dif + (crit * se)
+    lwr <- dif - (crit * se)
+    data.frame(pair = paste(g1, g2, sep = '-'),
+               diff = dif,
+               se = se,
+               upper = upr,
+               lower = lwr)
+}
+
+comp1<-smooth_diff(gam1,pdat,'Control','Treatment')
+
+comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
+                   rbind(comp1)) %>%
+  mutate(interval=case_when(
+    upper>0 & lower<0~"no-diff",
+    upper<0~"less",
+    lower>0~"greater"
+  ))
+
+c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper),
+                alpha = 0.5,
+                fill='#E1DD37FF') +
+    geom_line(color='#E75B64FF',size=1) +
+    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
+    geom_ribbon(data=comp_StO2_full%>%
+                    filter(lower>0),
+                aes(ymin =0, ymax =lower),
+                alpha = 0.5,
+                fill='#30123BFF') +
+    geom_ribbon(data=comp_StO2_full %>%
+                    filter(upper<0),
+                    aes(ymin =0, ymax =upper),
+                alpha = 0.5,
+                fill='#7A0403FF') +
+    facet_wrap(~ pair) +
+    theme_classic()+
+    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    theme(
+        text=element_text(size=18),
+        legend.title=element_blank()
+    )
+
+```
+
 # GAM and Linear model plots and Missing data
@@
 ```
+
+## Pairwise comparisons in GAMs: full and missing data cases
+
+The next code chunk reproduces Figure \@ref(fig:plot-pairwise-comp). Here pairwise comparisons
+
+```{r,pairwise-comp-Appendix,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
+##Pairwise comparisons
+
+pdat <- expand.grid(Day = seq(0, 10, length = 400),
+                    Group = c('Control', 'Treatment'))
+
+#this function takes the model, grid and groups to be compared using the lpmatrix
+
+smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
+                        unconditional = FALSE) {
+    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
+    #Find columns in xp where the name contains "Control"
+    col1 <- grepl(g1, colnames(xp))
+    #Find columns in xp where the name contains 'Treatment'
+    col2 <- grepl(g2, colnames(xp))
+    #r1 <- newdata[[var]] == f1
+    #r2 <- newdata[[var]] == f2
+    row1 <- with(newdata, Group == g1)
+    row2 <- with(newdata, Group == g2)
+    ## difference rows of xp for data from comparison
+    X <- xp[row1, ] - xp[row2, ]
+    ## zero out cols of X related to splines for other lochs
+    X[, ! (col1 | col2)] <- 0
+    ## zero out the parametric cols
+    X[, !grepl('^s\\(', colnames(xp))] <- 0
+    dif <- X %*% coef(model)
+    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
+    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
+    upr <- dif + (crit * se)
+    lwr <- dif - (crit * se)
+    data.frame(pair = paste(g1, g2, sep = '-'),
+               diff = dif,
+               se = se,
+               upper = upr,
+               lower = lwr)
+}
+
+comp1<-smooth_diff(gam1,pdat,'Control','Treatment')
+
+comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
+                   rbind(comp1)) %>%
+  mutate(interval=case_when(
+    upper>0 & lower<0~"no-diff",
+    upper<0~"less",
+    lower>0~"greater"
+  ))
+
+c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper),
+                alpha = 0.5,
+                fill='#E1DD37FF') +
+    geom_line(color='#E75B64FF',size=1) +
+    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
+    geom_ribbon(data=comp_StO2_full%>%
+                    filter(lower>0),
+                aes(ymin =0, ymax =lower),
+                alpha = 0.5,
+                fill='#30123BFF') +
+    geom_ribbon(data=comp_StO2_full %>%
+                    filter(upper<0),
+                    aes(ymin =0, ymax =upper),
+                alpha = 0.5,
+                fill='#7A0403FF') +
+    facet_wrap(~ pair) +
+    theme_classic()+
+    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    theme(
+        text=element_text(size=18),
+        legend.title=element_blank()
+    )
+
+###for missing data
+comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
+comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
+                   rbind(comp2))
+
+missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
+    geom_line(color='black',size=1) +
+    facet_wrap(~ pair) +
+    labs(x = 'Days', 
+         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
+                              )))+
+  scale_x_continuous(breaks=c(0,2,5,7,10))+
+  theme_classic()+
+  theme(
+     text=element_text(size=18),
+     legend.title=element_blank()
+     )
+
+c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper),
+                alpha = 0.5,
+                fill='#E1DD37FF') +
+    geom_line(color='#E75B64FF',size=1) +
+    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
+    geom_ribbon(data=comp_StO2_missing%>%
+                    filter(lower>0),
+                aes(ymin =0, ymax =lower),
+                alpha = 0.5,
+                fill='#30123BFF') +
+    geom_ribbon(data=comp_StO2_missing %>%
+                    filter(upper<0),
+                    aes(ymin =0, ymax =upper),
+                alpha = 0.5,
+                fill='#7A0403FF') +
+    facet_wrap(~ pair) +
+    theme_classic()+
+    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    theme(
+        text=element_text(size=18),
+        legend.title=element_blank()
+    )
+
+pair_comp<-c1+c2
+```
+
+(ref:pairwise-comp-caption) Pairwise comparisons for smooth terms.
+A: Pairwise comparisons for the full dataset.
+B: Pairwise comparisons for the dataset with missing observations.
+Significant differences exist where the interval does not cover 0.
+In both cases the effect of treatment is significant after day 5.
+
+```{r,plot-pairwise-comp-Appendix,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-comp-caption)'}
+
+pair_comp+plot_annotation(tag_levels='A')&
+  theme(
+     text=element_text(size=18)
+     )
+
+```

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/Full_document.Rmd
Working on pairwise comparisons section in the Appendix

--- 2021-05-20 Manuscripts/Manuscript by chapters -v1/Full_document.Rmd
@@
 library(gratia)
+library(ghibli)
+library(viridis)
 set.seed(1) #set seed for reproducibility
@@
 
-#this function will remove the default "author information" that is the Pandoc Latex template.
+#global themes for plotting
+thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
+thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
+
+#thm<-scale_color_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting
+#thm1<-scale_fill_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting
+
+#this function will remove the default "author information" that is the Pandoc LaTeX template.
 Otherwise, and additional "and" will appear at the end of the author list.Or, the custom template that is in the YAML header and that has the "author" line deleted may be used as well.

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/06-Appendix.Rmd
Added pairwise comparison code to the workflow, need to tidy aesthetics and section reference.

@@
 ```{r, final-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:final-GAM-caption)'}
-#need to add figure number and caption
+
 appr3<-appraise(gam1)
@@
 The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \@ref(GAM-significance).
-In this case, the "design matrix" is used to estimate the pairwise comparisons (see Manuscript for details and associated references).
-Briefly, the "design matrix" from `gam1` is used to estimate a 95% confidence interval of the difference between the smooth terms for each group.
+In this case, the "design matrix" is used to estimate the pairwise comparisons (see Manuscript for details and references that discuss this).
+Briefly, the "design matrix" (also known as the "Xp matrix") from the selected model (`gam1`) is used to calculate a 95% confidence interval of the difference between the smooth terms for each group.
 This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
 
-```{r,pairwise-comp-workflow,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
+```{r,pairwise-comp-workflow,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
+##Pairwise comparisons
+
+##matrix that contains the basis functions evaluated at the points in pdat
+    xp <- predict(gam1, newdata = pdat, type = 'lpmatrix')
+
+#Find columns in xp where the name contains "Control"
+    c1 <- grepl('Control', colnames(xp))
+
+#Find columns in xp where the name contains 'Treatment'
+    c2 <- grepl('Treatment', colnames(xp))
+
+#Find rows in pdat that correspond to either 'Control' or 'Treatment'
+    r1 <- with(pdat, Group == 'Control')
+    r2 <- with(pdat, Group == 'Treatment')
+
+# In xp: find the rows that correspond to Control or Treatment, those that do not match will be
+    #set to zero. Then, substract the values from the rows corresponding to 'Control' from those that correspond
+    #to 'Treatment'
+    X <- xp[r1, ] - xp[r2, ]
+
+    ## remove columns that do not contain name 'Control' or 'Treatment'
+    X[, ! (c1 | c2)] <- 0
+    ## zero out the parametric cols, those that do not contain in the characters 's('
+    X[, !grepl('^s\\(', colnames(xp))] <- 0
+
+    #Multiply matrix by model coefficients. X has (p,n) (rows, columns) and the coefficient matrix has
+    #dimensions (n,1). The resulting matrix has dimensions (p,1)
+    dif <- X %*% coef(gam1)
+
+    #comp<-test %*% coef(gam1)[3:10]
+
+#Calculate standard error for the computed differences using the variance-covariance matrix
+    #of the model
+    se <- sqrt(rowSums((X %*% vcov(gam1, unconditional = FALSE)) * X))
+    crit <- qt(0.05/2, df.residual(gam1), lower.tail = FALSE)
+    #upper  limits
+    upr <- dif + (crit * se)
+    #lower limits
+    lwr <- dif - (crit * se)
+    #put all components in a dataframe for plotting
+    comp1<-data.frame(pair = paste('Control', 'Treatment', sep = '-'),
+               diff = dif,
+               se = se,
+               upper = upr,
+               lower = lwr)
+
+#add time point sequence
+comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
+                   rbind(comp1))
+
+#plot the difference
+c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
+  #ribbon for difference confidence interval  
+  geom_ribbon(aes(ymin = lower, ymax = upper),
+                alpha = 0.5,
+                fill='black') +
+    geom_line(color='black',size=1) +
+    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
+  #highlight area under the curve where "Control" is higher
+  geom_ribbon(data=comp_StO2%>%
+                    filter(lower>0),
+                aes(ymin =0, ymax =lower),
+                alpha = 0.5,
+                fill='blue') +
+  #highlight area under the curve where "Treatment" is higher
+  geom_ribbon(data=comp_StO2 %>%
+                    filter(upper<0),
+                    aes(ymin =0, ymax =upper),
+                alpha = 0.5,
+                fill='red') +
+    facet_wrap(~ pair) +
+    theme_classic()+
+    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    theme(
+        text=element_text(size=18),
+        legend.title=element_blank()
+    )
+
+```
+
+(ref:pairwise-comp-caption) Smooth pairwise comparisons using a 95% confidence interval for the difference between smooths.
+
+```{r, pairwise-comp-workflow-fig, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pariwise-comp-caption)'}
+c1
+```
+
+Of notice, a conveniente wrapper for the function described above exists as `difference_smooths` in the package `gratia`.
+It produces the same plot when the function is used on a fitted model.
+The function syntax can be checked at:
+
+https://cran.r-project.org/web/packages/gratia/gratia.pdf
+
+# GAM and Linear model plots and Missing data
+
+This section covers the code used to generate Figure \@ref(fig:sim-smooth-plot), where the simulated data, fit of the "final" GAM (`gam1`), linear model  and GAM on data with missing observations are presented.
+Note that panel A in Figure \@ref(fig:sim-smooth-plot) and the inlet are generated in the code chunk where the data is simulated in Section \@ref(tumor-data-simulation), and are called later to build the figure.
+
+## GAM and Linear model plots
+
+This code chunk creates panels B and D in Figure \@ref(fig:sim-smooth-plot).
+Note that this code uses the final GAM from the previous section (`gam1`), so the simulated data and the model should be generated before running this section.
+
+```{r, plotting-appendix, echo=TRUE, include=TRUE, message=FALSE, warning=FALSE}
+#linear model
+lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
+
+#creates a dataframe using the length of the covariates for the GAM
+gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = seq(0, 10, by = 0.1),
+                         subject=factor(rep(1:10)))
+
+#creates a dataframe using the length of the covariates for rm-ANOVA
+lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = c(0:10),
+                        subject=factor(rep(1:10)),
+                          )
+lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
+
+#adds the predictions to the grid and creates a confidence interval for GAM
+gam_predict<-gam_predict%>%
+    mutate(fit = predict(gam1,gam_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(gam1, gam_predict,se.fit = TRUE,type='response')$se.fit)
+
+#using lm
+lm_predict<-lm_predict%>%
+    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
+
+#plot smooths and confidence interval for GAM
+f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
+    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),
+                   fill=Group
+                   ),
+              alpha=0.3,
+              data=gam_predict,
+            show.legend=FALSE,
+                inherit.aes=FALSE) +
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=gam_predict,
+              show.legend = FALSE)+
+  #facet_wrap(~Group)+
+  labs(y=expression(atop(StO[2],'complete')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
+      thm+
+  thm1
+
+#plot linear fit for rm-ANOVA
+f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
+    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),fill=Group),
+              alpha=0.3,
+              data=lm_predict,
+              show.legend = FALSE,
+                inherit.aes=FALSE) +
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=lm_predict,
+              show.legend = FALSE)+
+  #facet_wrap(~Group)+
+  labs(y=expression(paste('StO'[2],' (simulated)')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
+      thm+
+  thm1
+
+#posthoc comparisons for the linear model
+#library(multcomp)
+
+#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
+#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
+
+```
+
+## Working with Missing data in GAMs
+
+This code chunk randomly deletes 40% of the total observations in the original simulated data, and then an interaction GAM is fitted.
+Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
+
+```{r,missing-data-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
+#missing data
+#create a sequence of 40 random numbers between 1 and 100, these numbers will
+#correspond to the row numbers to be randomly erased from the original dataset
+missing <- sample(1:100, 40)
+
+#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
+
+ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
+
+#create a new dataframe, remove the StO2 column
+dat_missing <- dat_sim[,-1]
+
+#add NAs at the ind positions
+dat_missing$StO2_sim[ind]<-NA
+
+#Count the number of remaining observations per day (original dataset had 10 per group per day)
+dat_missing %>%
+    group_by(Day,Group) %>%
+    filter(!is.na(StO2_sim))%>%
+  count(Day)
+
+#the same model used for the full dataset
+mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
+#appraise the model
+appraise(mod_m1)
+
+m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = seq(0, 10, by = 0.1))
+
+#adds the predictions to the grid and creates a confidence interval
+m_predict<-m_predict%>%
+    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
+
+f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
+    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),
+                   fill=Group
+                   ),
+              alpha=0.3,
+              data=m_predict,
+            show.legend=FALSE,
+                inherit.aes=FALSE) +
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=m_predict,
+              show.legend = TRUE)+
+  #facet_wrap(~Group)+
+  labs(y=expression(atop(StO[2],'missing')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+  theme(
+    axis.text=element_text(size=22)
+  )+
+      thm+
+  thm1
+
+```
+
+(ref:sim-plot-caption-Appendix) Simulated data and smooths for oxygen saturation in tumors.
+A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point.
+Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
+Lines represent trends for each group, shaded regions are 95% confidence intervals.
+C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
+D: Smooths for the GAM model for the simulated data with 40% of its observations missing.
+Lines represent trends for each group, shaded regions are 95% confidence intervals.
+
+```{r,sim-smooth-plot-Appendix,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=TRUE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption-Appendix)'}
+
+mult_plot<-f2+inset_element(
+  f1, left = 0.01, 
+  bottom = 0.5, 
+  right = 0.5, 
+  top = 1.0)+
+  f3+f4+f6+
+   plot_annotation(tag_levels='A')&
+   ylim(c(-5,75)) &
+  theme(
+     text=element_text(size=18)
+     )&
+  thm
+
+mult_plot
+
+```
+
+## Pairwise comparisons in GAMs: full and missing data cases
+
+The next code chunk reproduces Figure \@ref(fig:plot-pairwise-comp). Here pairwise comparisons are made for the full and missing datasets.
+
+```{r,pairwise-comp-Appendix,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
 ##Pairwise comparisons
@@
 
-```
-
-# GAM and Linear model plots and Missing data
-
-This section covers the code used to generate Figure \@ref(fig:sim-smooth-plot), where the simulated data, fit of the "final" GAM (`gam1`), linear model  and GAM on data with missing observations are presented.
-Note that panel A in Figure \@ref(fig:sim-smooth-plot) and the inlet are generated in the code chunk where the data is simulated in Section \@ref(tumor-data-simulation), and are called later to build the figure.
-
-## GAM and Linear model plots
-
-This code chunk creates panels B and D in Figure \@ref(fig:sim-smooth-plot).
-Note that this code uses the final GAM from the previous section (`gam1`), so the simulated data and the model should be generated before running this section.
-
-```{r, plotting-appendix, echo=TRUE, include=TRUE, message=FALSE, warning=FALSE}
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-#creates a dataframe using the length of the covariates for the GAM
-gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
-                         subject=factor(rep(1:10)))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#adds the predictions to the grid and creates a confidence interval for GAM
-gam_predict<-gam_predict%>%
-    mutate(fit = predict(gam1,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(gam1, gam_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=gam_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'complete')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-
-#posthoc comparisons for the linear model
-#library(multcomp)
-
-#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
-
-```
-
-## Working with Missing data in GAMs
-
-This code chunk randomly deletes 40% of the total observations in the original simulated data, and then an interaction GAM is fitted.
-Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
-
-```{r,missing-data-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
-
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'missing')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-
-```
-
-(ref:sim-plot-caption-Appendix) Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95% confidence intervals.
-C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40% of its observations missing.
-Lines represent trends for each group, shaded regions are 95% confidence intervals.
-
-```{r,sim-smooth-plot-Appendix,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=TRUE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption-Appendix)'}
-
-mult_plot<-f2+inset_element(
-  f1, left = 0.01, 
-  bottom = 0.5, 
-  right = 0.5, 
-  top = 1.0)+
-  f3+f4+f6+
-   plot_annotation(tag_levels='A')&
-   ylim(c(-5,75)) &
-  theme(
-     text=element_text(size=18)
-     )&
-  thm
-
-mult_plot
-
-```
-
-## Pairwise comparisons in GAMs: full and missing data cases
-
-The next code chunk reproduces Figure \@ref(fig:plot-pairwise-comp). Here pairwise comparisons
-
-```{r,pairwise-comp-Appendix,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-##Pairwise comparisons
-
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-#this function takes the model, grid and groups to be compared using the lpmatrix
-
-smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    #Find columns in xp where the name contains "Control"
-    col1 <- grepl(g1, colnames(xp))
-    #Find columns in xp where the name contains 'Treatment'
-    col2 <- grepl(g2, colnames(xp))
-    #r1 <- newdata[[var]] == f1
-    #r2 <- newdata[[var]] == f2
-    row1 <- with(newdata, Group == g1)
-    row2 <- with(newdata, Group == g2)
-    ## difference rows of xp for data from comparison
-    X <- xp[row1, ] - xp[row2, ]
-    ## zero out cols of X related to splines for other lochs
-    X[, ! (col1 | col2)] <- 0
-    ## zero out the parametric cols
-    X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(g1, g2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-}
-
-comp1<-smooth_diff(gam1,pdat,'Control','Treatment')
-
-comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
-
-c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill='#E1DD37FF') +
-    geom_line(color='#E75B64FF',size=1) +
-    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
-    geom_ribbon(data=comp_StO2_full%>%
-                    filter(lower>0),
-                aes(ymin =0, ymax =lower),
-                alpha = 0.5,
-                fill='#30123BFF') +
-    geom_ribbon(data=comp_StO2_full %>%
-                    filter(upper<0),
-                    aes(ymin =0, ymax =upper),
-                alpha = 0.5,
-                fill='#7A0403FF') +
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
 ###for missing data

### 2021-05-24 Code and miscellaneous/Pairwise comparisons for smooths.R

@@ -45,12 +45,6 @@ pdat <- expand.grid(Day = seq(0, 10, length = 400),
 ##matrix that contains the basis functions evaluated at the points in pdat
     xp <- predict(gam1, newdata = pdat, type = 'lpmatrix')
 
-    #remove intercept column and group identifying column
-    xp<-xp[,-c(1:2)]
-
-    control<-xp[1:400,]
-    treatment<-xp[401:800,]
-
 
 #Find columns in xp where the name contains "Control"
     c1 <- grepl('Control', colnames(xp))
@@ -72,7 +66,7 @@ pdat <- expand.grid(Day = seq(0, 10, length = 400),
 
     ## remove columns that do not contain name 'Control' or 'Treatment'
     X[, ! (c1 | c2)] <- 0
-    ## zero out the parametric cols, those that do not contain in the name 's()'
+    ## zero out the parametric cols, those that do not contain in the characters 's('
     X[, !grepl('^s\\(', colnames(xp))] <- 0
 
     #Multiply matrix by model coefficients. X has (p,n) (rows, columns) and the coefficient matrix has
@@ -81,11 +75,15 @@ pdat <- expand.grid(Day = seq(0, 10, length = 400),
 
     #comp<-test %*% coef(gam1)[3:10]
 
-
+#Calculate standard error for the computed differences using the variance-covariance matrix
+    #of the model
     se <- sqrt(rowSums((X %*% vcov(gam1, unconditional = FALSE)) * X))
     crit <- qt(0.05/2, df.residual(gam1), lower.tail = FALSE)
+    #upper  limits
     upr <- dif + (crit * se)
+    #lower limits
     lwr <- dif - (crit * se)
+    #put all components in a dataframe for plotting
     comp1<-data.frame(pair = paste('Control', 'Treatment', sep = '-'),
                diff = dif,
                se = se,
@@ -94,11 +92,11 @@ pdat <- expand.grid(Day = seq(0, 10, length = 400),
 
 
 
-
+#add time point sequence
 comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
                    rbind(comp1))
 
-
+#plot the difference
 c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
     geom_ribbon(aes(ymin = lower, ymax = upper),
                 alpha = 0.5,

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/02-Challenges.Rmd
Need to improve aesthetics of combined gam plots in the Appendix

@@
 # linear response, correlated errors (subject effect)
-par(mar = c(0.5, 0.5, 0.5, 0.5))
+#par(mar = c(0.5, 0.5, 0.5, 0.5))
 A/C+plot_annotation(tag_levels = 'A')

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/04-Longitudinal analysis with GAMs.Rmd
Need to improve aesthetics of combined gam plots in the Appendix

@@
                 alpha = 0.5,
-                fill='#E1DD37FF') +
+                fill='#DB3A07FF') +
     geom_line(color='#E75B64FF',size=1) +
@@
 
-## Determination of significance in GAMs for longitudinal data{#GAM-sig}
+## Determination of significance in GAMs for longitudinal data{#GAM-significance}
 

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/06-Appendix.Rmd
Need to improve aesthetics of combined gam plots in the Appendix

@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
 # (APPENDIX) Appendix {-}
@@
 
+```{r, plotting layouts,echo=TRUE, include=TRUE}
+layout1 <- c(
+  area(1, 1),
+  area( 1, 2),
+  area(2, 1),
+  area(2, 2),
+  area(1, 3, 2)
+)
+
+layout2 <- c(
+  area(1, 1),
+  area( 1, 2),
+  area(2, 1),
+  area(2, 2),
+  area(1, 3, 2),
+  area(1,4,2),
+  area(1,5,1)
+)
+
+```
+
 #### Graphical diagnostics
@@
 ```{r, first-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:first-GAM-caption)'}
-#need to add figure number and caption
+
 appr1<-appraise(gam_00)
@@
 
-visual_check 
+visual_check+plot_layout(design=layout1)
 ```
@@
 
-visual_check2
+visual_check2+plot_layout(design=layout1)
 ```
@@
 
-visual_check3
+visual_check3+plot_layout(design=layout2)
 ```
@@
 The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \@ref(GAM-significance).
-In this case, the "design matrix" is used to estimate the pairwise comparisons (see Manuscript for details and references that discuss this).
+In this case, the "design matrix" is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
 Briefly, the "design matrix" (also known as the "Xp matrix") from the selected model (`gam1`) is used to calculate a 95% confidence interval of the difference between the smooth terms for each group.
@@
                 alpha = 0.5,
-                fill='black') +
+                fill='#DB3A07FF') +
     geom_line(color='black',size=1) +
@@
                 alpha = 0.5,
-                fill='blue') +
+                fill='#30123BFF') +
   #highlight area under the curve where "Treatment" is higher
@@
                 alpha = 0.5,
-                fill='red') +
+                fill='#7A0403FF') +
     facet_wrap(~ pair) +
@@
 
-(ref:pairwise-comp-caption) Smooth pairwise comparisons using a 95% confidence interval for the difference between smooths.
-
-```{r, pairwise-comp-workflow-fig, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pariwise-comp-caption)'}
+(ref:pairwise-comp-caption) Smooth pairwise comparisons for model `gam1` using a 95% confidence interval for the difference between smooths.
+
+```{r, pairwise-comp-workflow-fig, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-comp-caption)'}
 c1
@@
 
-Of notice, a conveniente wrapper for the function described above exists as `difference_smooths` in the package `gratia`.
-It produces the same plot when the function is used on a fitted model.
-The function syntax can be checked at:
+Of notice, a convenient wrapper for the function described above exists in the package `gratia`.
+In this package, `difference_smooths` is a function that makes the comparisons and produces Figure \@ref(fig:pairwise-comp-workflow-fig) when is used on a fitted model.
+The function syntax and an example can be found at:
 
@@
 
-This code chunk randomly deletes 40% of the total observations in the original simulated data, and then an interaction GAM is fitted.
+This code chunk first randomly deletes 40% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data.
 Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
@@
 
-```{r,pairwise-comp-Appendix,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
+```{r,pairwise-comp-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
 ##Pairwise comparisons

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/Full_document.Rmd
Need to improve aesthetics of combined gam plots in the Appendix

--- 2021-05-23 Manuscripts/Manuscript by chapters -v1/Full_document.Rmd
@@
 #global themes for plotting
-thm<-scale_color_viridis_d(option="magma",begin=0.1,end=0.7)
-thm1<-scale_fill_viridis_d(option="magma",begin=0.1,end=0.7)#global theme for plotting
+thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
+thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
 

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/Full_document.tex
Need to improve aesthetics of combined gam plots in the Appendix

@@
 
-\hypertarget{determination-of-significance-in-gams-for-longitudinal-data}{%
-\subsection{Determination of significance in GAMs for longitudinal data}\label{determination-of-significance-in-gams-for-longitudinal-data}}
+\hypertarget{GAM-significance}{%
+\subsection{Determination of significance in GAMs for longitudinal data}\label{GAM-significance}}
 
@@
 
+\begin{lstlisting}[language=R]
+layout1 <- c(
+  area(1, 1),
+  area( 1, 2),
+  area(2, 1),
+  area(2, 2),
+  area(1, 3, 2)
+)
+
+layout2 <- c(
+  area(1, 1),
+  area( 1, 2),
+  area(2, 1),
+  area(2, 2),
+  area(1, 3, 2),
+  area(1,4,2),
+  area(1,5,1)
+)
+\end{lstlisting}
+
 \hypertarget{graphical-diagnostics}{%
 \paragraph{Graphical diagnostics}\label{graphical-diagnostics}}
-
-\begin{lstlisting}
-## Error: The given dimensions cannot hold all panels. Please increase `ncol` or `nrow`
-\end{lstlisting}
 
@@
 From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \textasciitilde43\%.
-
-\begin{lstlisting}
-## Error: The given dimensions cannot hold all panels. Please increase `ncol` or `nrow`
-\end{lstlisting}
 
@@
 
-\begin{lstlisting}
-## Error: The given dimensions cannot hold all panels. Please increase `ncol` or `nrow`
-\end{lstlisting}
-
 \begin{figure}[H]
@@
 The AIC should not be considered as the only estimator of model goodness, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
+
+\hypertarget{pairwise-comparisons-of-smooth-confidence-intervals}{%
+\paragraph{Pairwise comparisons of smooth confidence intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
+
+The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \ref{GAM-significance}.
+In this case, the ``design matrix'' is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
+Briefly, the ``design matrix'' (also known as the ``Xp matrix'') from the selected model (\passthrough{\lstinline!gam1!}) is used to calculate a 95\% confidence interval of the difference between the smooth terms for each group.
+This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
+
+\begin{lstlisting}[language=R]
+##Pairwise comparisons
+
+##matrix that contains the basis functions evaluated at the points in pdat
+    xp <- predict(gam1, newdata = pdat, type = 'lpmatrix')
+
+#Find columns in xp where the name contains "Control"
+    c1 <- grepl('Control', colnames(xp))
+
+#Find columns in xp where the name contains 'Treatment'
+    c2 <- grepl('Treatment', colnames(xp))
+
+#Find rows in pdat that correspond to either 'Control' or 'Treatment'
+    r1 <- with(pdat, Group == 'Control')
+    r2 <- with(pdat, Group == 'Treatment')
+
+# In xp: find the rows that correspond to Control or Treatment, those that do not match will be
+    #set to zero. Then, substract the values from the rows corresponding to 'Control' from those that correspond
+    #to 'Treatment'
+    X <- xp[r1, ] - xp[r2, ]
+
+    ## remove columns that do not contain name 'Control' or 'Treatment'
+    X[, ! (c1 | c2)] <- 0
+    ## zero out the parametric cols, those that do not contain in the characters 's('
+    X[, !grepl('^s\\(', colnames(xp))] <- 0
+
+    #Multiply matrix by model coefficients. X has (p,n) (rows, columns) and the coefficient matrix has
+    #dimensions (n,1). The resulting matrix has dimensions (p,1)
+    dif <- X %*% coef(gam1)
+
+    #comp<-test %*% coef(gam1)[3:10]
+
+#Calculate standard error for the computed differences using the variance-covariance matrix
+    #of the model
+    se <- sqrt(rowSums((X %*% vcov(gam1, unconditional = FALSE)) * X))
+    crit <- qt(0.05/2, df.residual(gam1), lower.tail = FALSE)
+    #upper  limits
+    upr <- dif + (crit * se)
+    #lower limits
+    lwr <- dif - (crit * se)
+    #put all components in a dataframe for plotting
+    comp1<-data.frame(pair = paste('Control', 'Treatment', sep = '-'),
+               diff = dif,
+               se = se,
+               upper = upr,
+               lower = lwr)
+
+#add time point sequence
+comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
+                   rbind(comp1))
+
+#plot the difference
+c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
+  #ribbon for difference confidence interval  
+  geom_ribbon(aes(ymin = lower, ymax = upper),
+                alpha = 0.5,
+                fill='#DB3A07FF') +
+    geom_line(color='black',size=1) +
+    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
+  #highlight area under the curve where "Control" is higher
+  geom_ribbon(data=comp_StO2%>%
+                    filter(lower>0),
+                aes(ymin =0, ymax =lower),
+                alpha = 0.5,
+                fill='#30123BFF') +
+  #highlight area under the curve where "Treatment" is higher
+  geom_ribbon(data=comp_StO2 %>%
+                    filter(upper<0),
+                    aes(ymin =0, ymax =upper),
+                alpha = 0.5,
+                fill='#7A0403FF') +
+    facet_wrap(~ pair) +
+    theme_classic()+
+    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    theme(
+        text=element_text(size=18),
+        legend.title=element_blank()
+    )
+\end{lstlisting}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/pairwise-comp-workflow-fig-1}
+
+}
+
+\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!gam1!} using a 95\% confidence interval for the difference between smooths.}\label{fig:pairwise-comp-workflow-fig}
+\end{figure}
+
+Of notice, a convenient wrapper for the function described above exists in the package \passthrough{\lstinline!gratia!}.
+In this package, \passthrough{\lstinline!difference\_smooths!} is a function that makes the comparisons and produces Figure \ref{fig:pairwise-comp-workflow-fig} when is used on a fitted model.
+The function syntax and an example can be found at:
+
+\url{https://cran.r-project.org/web/packages/gratia/gratia.pdf}
 
@@
 
-This code chunk randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted.
+This code chunk first randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data.
 Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
@@
 
+\hypertarget{pairwise-comparisons-in-gams-full-and-missing-data-cases}{%
+\subsection{Pairwise comparisons in GAMs: full and missing data cases}\label{pairwise-comparisons-in-gams-full-and-missing-data-cases}}
+
+The next code chunk reproduces Figure \ref{fig:plot-pairwise-comp}. Here pairwise comparisons are made for the full and missing datasets.
+
+\begin{lstlisting}[language=R]
+##Pairwise comparisons
+
+pdat <- expand.grid(Day = seq(0, 10, length = 400),
+                    Group = c('Control', 'Treatment'))
+
+#this function takes the model, grid and groups to be compared using the lpmatrix
+
+smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
+                        unconditional = FALSE) {
+    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
+    #Find columns in xp where the name contains "Control"
+    col1 <- grepl(g1, colnames(xp))
+    #Find columns in xp where the name contains 'Treatment'
+    col2 <- grepl(g2, colnames(xp))
+    #r1 <- newdata[[var]] == f1
+    #r2 <- newdata[[var]] == f2
+    row1 <- with(newdata, Group == g1)
+    row2 <- with(newdata, Group == g2)
+    ## difference rows of xp for data from comparison
+    X <- xp[row1, ] - xp[row2, ]
+    ## zero out cols of X related to splines for other lochs
+    X[, ! (col1 | col2)] <- 0
+    ## zero out the parametric cols
+    X[, !grepl('^s\\(', colnames(xp))] <- 0
+    dif <- X %*% coef(model)
+    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
+    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
+    upr <- dif + (crit * se)
+    lwr <- dif - (crit * se)
+    data.frame(pair = paste(g1, g2, sep = '-'),
+               diff = dif,
+               se = se,
+               upper = upr,
+               lower = lwr)
+}
+
+comp1<-smooth_diff(gam1,pdat,'Control','Treatment')
+
+comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
+                   rbind(comp1)) %>%
+  mutate(interval=case_when(
+    upper>0 & lower<0~"no-diff",
+    upper<0~"less",
+    lower>0~"greater"
+  ))
+
+c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper),
+                alpha = 0.5,
+                fill='#DB3A07FF') +
+    geom_line(color='#E75B64FF',size=1) +
+    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
+    geom_ribbon(data=comp_StO2_full%>%
+                    filter(lower>0),
+                aes(ymin =0, ymax =lower),
+                alpha = 0.5,
+                fill='#30123BFF') +
+    geom_ribbon(data=comp_StO2_full %>%
+                    filter(upper<0),
+                    aes(ymin =0, ymax =upper),
+                alpha = 0.5,
+                fill='#7A0403FF') +
+    facet_wrap(~ pair) +
+    theme_classic()+
+    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    theme(
+        text=element_text(size=18),
+        legend.title=element_blank()
+    )
+
+###for missing data
+comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
+comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
+                   rbind(comp2))
+
+missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
+    geom_line(color='black',size=1) +
+    facet_wrap(~ pair) +
+    labs(x = 'Days', 
+         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
+                              )))+
+  scale_x_continuous(breaks=c(0,2,5,7,10))+
+  theme_classic()+
+  theme(
+     text=element_text(size=18),
+     legend.title=element_blank()
+     )
+
+c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
+    geom_ribbon(aes(ymin = lower, ymax = upper),
+                alpha = 0.5,
+                fill='#DB3A07FF') +
+    geom_line(color='#E75B64FF',size=1) +
+    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
+    geom_ribbon(data=comp_StO2_missing%>%
+                    filter(lower>0),
+                aes(ymin =0, ymax =lower),
+                alpha = 0.5,
+                fill='#30123BFF') +
+    geom_ribbon(data=comp_StO2_missing %>%
+                    filter(upper<0),
+                    aes(ymin =0, ymax =upper),
+                alpha = 0.5,
+                fill='#7A0403FF') +
+    facet_wrap(~ pair) +
+    theme_classic()+
+    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    theme(
+        text=element_text(size=18),
+        legend.title=element_blank()
+    )
+
+pair_comp<-c1+c2
+\end{lstlisting}
+
+Smooth pairwise comparisons for model \passthrough{\lstinline!gam1!} using a 95\% confidence interval for the difference between smooths.
+Pairwise comparisons for smooth terms.
+A: Pairwise comparisons for the full dataset.
+B: Pairwise comparisons for the dataset with missing observations.
+Significant differences exist where the interval does not cover 0.
+In both cases the effect of treatment is significant after day 5.
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/plot-pairwise-comp-Appendix-1}
+
+}
+
+\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!gam1!} using a 95\% confidence interval for the difference between smooths.}\label{fig:plot-pairwise-comp-Appendix}
+\end{figure}
+
 \end{document}

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/03-GAM Theory.Rmd
Changed aesthetics, updated manuscript, added missing sections on Appendix.

@@
   geom_point(size=sz+1)+
-  scale_color_aaas()+
+  thm1+
   labs(y='Simulated \n response')+

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/06-Appendix.Rmd
Changed aesthetics, updated manuscript, added missing sections on Appendix.

@@
 ```{r, responses-Appendix,include=TRUE,message=FALSE,echo=TRUE}
-
+set.seed(1)
 ####################Section for calculations######################################
@@
 ```{r, plotting layouts,echo=TRUE, include=TRUE}
+
+#see https://patchwork.data-imaginist.com/reference/area.html
 layout1 <- c(
@@
   area(2, 2),
-  area(1, 3, 2),
-  area(1,4,2),
-  area(1,5,1)
-)
-
+  area(1,3,2,5)
+)
+
+#plot(layout2)
 ```
@@
 
-gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
+m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
             method='REML',
@@
 
-gam.check(gam1)
-summary(gam1)
-```
-
-The resulting model is model `gam1`, which is the model fitted in the main manuscript.
+gam.check(m1)
+summary(m1)
+```
+
+The resulting model is`m1`, which is the model fitted in the main manuscript.
 By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group .
@@
 
-appr3<-appraise(gam1)
-sm3<-draw(gam1)
+appr3<-appraise(m1)
+sm3<-draw(m1)
 
@@
 #correspond to the row numbers to be randomly erased from the original dataset
+set.seed(1)
 missing <- sample(1:100, 40)
@@
 
-comp1<-smooth_diff(gam1,pdat,'Control','Treatment')
+comp1<-smooth_diff(m1,pdat,'Control','Treatment')
 

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/Full_document.tex
Changed aesthetics, updated manuscript, added missing sections on Appendix.

@@
 ]{article}
-\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
+\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
@@
   pdftitle={The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
+  pdfborder={0 0 0},
+  breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
 \usepackage[margin=1in]{geometry}
@@
 \lstset{defaultdialect=[x86masm]Assembler}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
+\usepackage{longtable,booktabs}
 % Allow footnotes in longtable head/foot
@@
 \makesavenoteenv{longtable}
-\usepackage{graphicx}
+\usepackage{graphicx,grffile}
 \makeatletter
@@
 \setcounter{secnumdepth}{5}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
 \usepackage{lineno}
@@
 \usepackage{float}
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 A \emph{post hoc} analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
-Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
+Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless{}0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
 In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}.
@@
 
-\passthrough{\lstinline!m1<-gam(StO2\_sim\~Group+s(Day,by=Group,k=5,bs="gp"), method='REML',data=dat\_sim!}
+\passthrough{\lstinline!m1<-gam(StO2\_sim\~Group+s(Day,by=Group,k=5,bs="gp"), method='REML',data=dat\_sim)!}
 
@@
 
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot},C.
+One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?''.
+The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot},C.
 This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \ref{fig:sim-smooth-plot},B) .
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
 \leavevmode\hypertarget{ref-roblyer2011}{}%
-\CSLLeftMargin{{[}1{]} }
-\CSLRightInline{D.
+{[}1{]} D.
 Roblyer, S.
@@
 108 (2011) 14626--14631.
-\url{https://doi.org/10.1073/pnas.1013103108}.}
+\url{https://doi.org/10.1073/pnas.1013103108}.
 
 \leavevmode\hypertarget{ref-tank2020}{}%
-\CSLLeftMargin{{[}2{]} }
-\CSLRightInline{A.
+{[}2{]} A.
 Tank, H.M.
@@
 22 (2020) 1--10.
-\url{https://doi.org/doi:10.1186/s13058-020-01262-1}.}
+\url{https://doi.org/doi:10.1186/s13058-020-01262-1}.
 
 \leavevmode\hypertarget{ref-pavlov2018}{}%
-\CSLLeftMargin{{[}3{]} }
-\CSLRightInline{M.V.
+{[}3{]} M.V.
 Pavlov, T.I.
@@
 Shakhova, A.V.
-Maslennikova, {Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy}, {Journal of Biomedical Optics}.
-{23} (2018).
-\url{https://doi.org/\%7B10.1117/1.JBO.23.9.091410\%7D}.}
+Maslennikova, Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy, Journal of Biomedical Optics.
+23 (n.d.).
+\href{https://doi.org/\%7B10.1117/1.JBO.23.9.091410\%7D}{https://doi.org/\{10.1117/1.JBO.23.9.091410\}}.
 
 \leavevmode\hypertarget{ref-demidov2018}{}%
-\CSLLeftMargin{{[}4{]} }
-\CSLRightInline{V.
+{[}4{]} V.
 Demidov, A.
@@
 Flueraru, I.A.
-Vitkin, {Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography}, {Scientific Reports}.
-{8} (2018).
-\url{https://doi.org/\%7B10.1038/s41598-017-18635-w\%7D}.}
+Vitkin, Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography, Scientific Reports.
+8 (n.d.).
+\href{https://doi.org/\%7B10.1038/s41598-017-18635-w\%7D}{https://doi.org/\{10.1038/s41598-017-18635-w\}}.
 
 \leavevmode\hypertarget{ref-ritter2001}{}%
-\CSLLeftMargin{{[}5{]} }
-\CSLRightInline{G.
+{[}5{]} G.
 Ritter, L.
@@
 Old, S.
-Welt, {Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}, {Cancer Research}.
-{61} (2001) 6851--6859.}
+Welt, Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33, Cancer Research.
+61 (n.d.) 6851--6859.
 
 \leavevmode\hypertarget{ref-roth2017}{}%
-\CSLLeftMargin{{[}6{]} }
-\CSLRightInline{E.M.
+{[}6{]} E.M.
 Roth, A.C.
@@
 Lecorps, H.M.
-Colhoun, {Antidrug Antibodies in Patients Treated with Alirocumab}, {New England Journal of Medicine}.
-{376} (2017) 1589--1590.
-\url{https://doi.org/\%7B10.1056/NEJMc1616623\%7D}.}
+Colhoun, Antidrug Antibodies in Patients Treated with Alirocumab, New England Journal of Medicine.
+376 (n.d.) 1589--1590.
+\href{https://doi.org/\%7B10.1056/NEJMc1616623\%7D}{https://doi.org/\{10.1056/NEJMc1616623\}}.
 
 \leavevmode\hypertarget{ref-jones2018}{}%
-\CSLLeftMargin{{[}7{]} }
-\CSLRightInline{J.D.
+{[}7{]} J.D.
 Jones, H.E.
@@
 Woessner, K.P.
-Quinn, {In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing}, {Communications Biology}.
-{1} (2018).
-\url{https://doi.org/\%7B10.1038/s42003-018-0206-4\%7D}.}
+Quinn, In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing, Communications Biology.
+1 (n.d.).
+\href{https://doi.org/\%7B10.1038/s42003-018-0206-4\%7D}{https://doi.org/\{10.1038/s42003-018-0206-4\}}.
 
 \leavevmode\hypertarget{ref-skala2010}{}%
-\CSLLeftMargin{{[}8{]} }
-\CSLRightInline{M.C.
+{[}8{]} M.C.
 Skala, A.
@@
 15 (2010).
-\url{https://doi.org/10.1117/1.3285584}.}
+\url{https://doi.org/10.1117/1.3285584}.
 
 \leavevmode\hypertarget{ref-greening2018}{}%
-\CSLLeftMargin{{[}9{]} }
-\CSLRightInline{G.J.
+{[}9{]} G.J.
 Greening, K.P.
@@
 Cato, T.J.
-Muldoon, {Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy}, {Biomedical Optics Express}.
-{9} (2018) 2871--2886.
-\url{https://doi.org/\%7B10.1364/BOE.9.002871\%7D}.}
+Muldoon, Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy, Biomedical Optics Express.
+9 (n.d.) 2871--2886.
+\href{https://doi.org/\%7B10.1364/BOE.9.002871\%7D}{https://doi.org/\{10.1364/BOE.9.002871\}}.
 
 \leavevmode\hypertarget{ref-sio2016}{}%
-\CSLLeftMargin{{[}10{]} }
-\CSLRightInline{T.T.
+{[}10{]} T.T.
 Sio, P.J.
@@
 Bearden III, R.C.
-Miller, {Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {{[}}alliance{]})}, {Supportive Care in Cancer}.
-{24} (2016) 3847--3855.
-\url{https://doi.org/\%7B10.1007/s00520-016-3213-3\%7D}.}
+Miller, Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}), Supportive Care in Cancer.
+24 (n.d.) 3847--3855.
+\href{https://doi.org/\%7B10.1007/s00520-016-3213-3\%7D}{https://doi.org/\{10.1007/s00520-016-3213-3\}}.
 
 \leavevmode\hypertarget{ref-kamstra2015}{}%
-\CSLLeftMargin{{[}11{]} }
-\CSLRightInline{J.I.
+{[}11{]} J.I.
 Kamstra, P.U.
@@
 Roodenburg, J.A.
-Langendijk, {Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study}, {Oral Oncology}.
-{51} (2015) 548--555.
-\url{https://doi.org/\%7B10.1016/j.oraloncology.2015.01.016\%7D}.}
+Langendijk, Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study, Oral Oncology.
+51 (n.d.) 548--555.
+\href{https://doi.org/\%7B10.1016/j.oraloncology.2015.01.016\%7D}{https://doi.org/\{10.1016/j.oraloncology.2015.01.016\}}.
 
 \leavevmode\hypertarget{ref-wagenmakers2008}{}%
-\CSLLeftMargin{{[}12{]} }
-\CSLRightInline{E.-J.
+{[}12{]} E.-J.
 Wagenmakers, M.
@@
 181--207.
-\url{https://doi.org/10.1007/978-0-387-09612-4_9}.}
+\url{https://doi.org/10.1007/978-0-387-09612-4_9}.
 
 \leavevmode\hypertarget{ref-gueorguieva2004}{}%
-\CSLLeftMargin{{[}13{]} }
-\CSLRightInline{R.
+{[}13{]} R.
 Gueorguieva, J.H.
-Krystal, {Move over ANOVA - Progress in analyzing repeated-measures data and its reflection in papers published in the archives of general psychiatry}, Archives of General Psychiatry.
+Krystal, Move over ANOVA - Progress in analyzing repeated-measures data and its reflection in papers published in the archives of general psychiatry, Archives of General Psychiatry.
 61 (2004) 310--317.
-\url{https://doi.org/10.1001/archpsyc.61.3.310}.}
+\url{https://doi.org/10.1001/archpsyc.61.3.310}.
 
 \leavevmode\hypertarget{ref-schober2018}{}%
-\CSLLeftMargin{{[}14{]} }
-\CSLRightInline{P.
+{[}14{]} P.
 Schober, T.R.
@@
 127 (2018) 569--575.
-\url{https://doi.org/10.1213/ane.0000000000003511}.}
+\url{https://doi.org/10.1213/ane.0000000000003511}.
 
 \leavevmode\hypertarget{ref-pinheiro2006}{}%
-\CSLLeftMargin{{[}15{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science \& Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}
+{[}15{]} J. Pinheiro, D. Bates, Mixed-effects models in S and S-PLUS, Springer Science \& Business Media, 2006. \url{https://doi.org/https://doi.org/10.1007/b98882}.
 
 \leavevmode\hypertarget{ref-vishwanath2009}{}%
-\CSLLeftMargin{{[}16{]} }
-\CSLRightInline{K.
+{[}16{]} K.
 Vishwanath, H.
@@
 11 (2009) 889--900.
-\url{https://doi.org/10.1593/neo.09580}.}
+\url{https://doi.org/10.1593/neo.09580}.
 
 \leavevmode\hypertarget{ref-dennis2019}{}%
-\CSLLeftMargin{{[}17{]} }
-\CSLRightInline{B.
+{[}17{]} B.
 Dennis, J.M.
@@
 Taper, S.R.
-Lele, {Errors in Statistical Inference Under Model Misspecification: Evidence, Hypothesis Testing, and AIC}, {Frontiers in Ecology and Evolution}.
-{7} (2019).
-\url{https://doi.org/\%7B10.3389/fevo.2019.00372\%7D}.}
+Lele, Errors in Statistical Inference Under Model Misspecification: Evidence, Hypothesis Testing, and AIC, Frontiers in Ecology and Evolution.
+7 (n.d.).
+\href{https://doi.org/\%7B10.3389/fevo.2019.00372\%7D}{https://doi.org/\{10.3389/fevo.2019.00372\}}.
 
 \leavevmode\hypertarget{ref-wang2019}{}%
-\CSLLeftMargin{{[}18{]} }
-\CSLRightInline{B.
+{[}18{]} B.
 Wang, Z.
@@
 Tu, C.
-Feng, {The p-value and model specification in statistics}, {General Psychiatry}.
-{32} (2019).
-\url{https://doi.org/\%7B10.1136/gpsych-2019-100081\%7D}.}
+Feng, The p-value and model specification in statistics, General Psychiatry.
+32 (n.d.).
+\href{https://doi.org/\%7B10.1136/gpsych-2019-100081\%7D}{https://doi.org/\{10.1136/gpsych-2019-100081\}}.
 
 \leavevmode\hypertarget{ref-liu2010}{}%
-\CSLLeftMargin{{[}19{]} }
-\CSLRightInline{C.
+{[}19{]} C.
 Liu, T.P.
@@
 18 (2010) 1724--1730.
-\url{https://doi.org/10.1038/mt.2010.127}.}
+\url{https://doi.org/10.1038/mt.2010.127}.
 
 \leavevmode\hypertarget{ref-halsey2015}{}%
-\CSLLeftMargin{{[}20{]} }
-\CSLRightInline{L.G.
+{[}20{]} L.G.
 Halsey, D.
@@
 Vowler, G.B.
-Drummond, {The fickle P value generates irreproducible results}, {Nature Methods}.
-{12} (2015) 179--185.
-\url{https://doi.org/\%7B10.1038/nmeth.3288\%7D}.}
+Drummond, The fickle P value generates irreproducible results, Nature Methods.
+12 (n.d.) 179--185.
+\href{https://doi.org/\%7B10.1038/nmeth.3288\%7D}{https://doi.org/\{10.1038/nmeth.3288\}}.
 
 \leavevmode\hypertarget{ref-abdi2010}{}%
-\CSLLeftMargin{{[}21{]} }
-\CSLRightInline{H. Abdi, {Holm's Sequential Bonferroni Procedure}, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.}
+{[}21{]} H. Abdi, Holm's Sequential Bonferroni Procedure, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.
 
 \leavevmode\hypertarget{ref-nakagawa2004}{}%
-\CSLLeftMargin{{[}22{]} }
-\CSLRightInline{S.
-Nakagawa, {A farewell to Bonferroni: the problems of low statistical power and publication bias}, {Behavioral Ecology}.
-{15} (2004) 1044--1045.
-\url{https://doi.org/\%7B10.1093/beheco/arh107\%7D}.}
+{[}22{]} S.
+Nakagawa, A farewell to Bonferroni: the problems of low statistical power and publication bias, Behavioral Ecology.
+15 (n.d.) 1044--1045.
+\href{https://doi.org/\%7B10.1093/beheco/arh107\%7D}{https://doi.org/\{10.1093/beheco/arh107\}}.
 
 \leavevmode\hypertarget{ref-gelman2012}{}%
-\CSLLeftMargin{{[}23{]} }
-\CSLRightInline{A.
+{[}23{]} A.
 Gelman, J.
 Hill, M.
-Yajima, {Why We (Usually) Don't Have to Worry About Multiple Comparisons}, {Journal of Research on Educational Effectiveness}.
-{5} (2012) 189--211.
-\url{https://doi.org/\%7B10.1080/19345747.2011.618213\%7D}.}
+Yajima, Why We (Usually) Don't Have to Worry About Multiple Comparisons, Journal of Research on Educational Effectiveness.
+5 (n.d.) 189--211.
+\href{https://doi.org/\%7B10.1080/19345747.2011.618213\%7D}{https://doi.org/\{10.1080/19345747.2011.618213\}}.
 
 \leavevmode\hypertarget{ref-albers2019}{}%
-\CSLLeftMargin{{[}24{]} }
-\CSLRightInline{C. Albers, {The problem with unadjusted multiple and sequential statistical testing}, {Nature Communications}. {10} (2019). \url{https://doi.org/\%7B10.1038/s41467-019-09941-0\%7D}.}
+{[}24{]} C.
+Albers, The problem with unadjusted multiple and sequential statistical testing, Nature Communications.
+10 (n.d.).
+\href{https://doi.org/\%7B10.1038/s41467-019-09941-0\%7D}{https://doi.org/\{10.1038/s41467-019-09941-0\}}.
 
 \leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
-\CSLLeftMargin{{[}25{]} }
-\CSLRightInline{C.
+{[}25{]} C.
 Ugrinowitsch, G.W.
@@
 36 (2004) 2144--2148.
-\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.}
+\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.
 
 \leavevmode\hypertarget{ref-huynh1976}{}%
-\CSLLeftMargin{{[}26{]} }
-\CSLRightInline{H.
+{[}26{]} H.
 Huynh, L.S.
@@
 1 (1976) 69--82.
-\url{https://doi.org/10.3102/10769986001001069}.}
+\url{https://doi.org/10.3102/10769986001001069}.
 
 \leavevmode\hypertarget{ref-greenhouse1959}{}%
-\CSLLeftMargin{{[}27{]} }
-\CSLRightInline{S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.}
+{[}27{]} S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.
 
 \leavevmode\hypertarget{ref-haverkamp2017}{}%
-\CSLLeftMargin{{[}28{]} }
-\CSLRightInline{N.
+{[}28{]} N.
 Haverkamp, A.
-Beauducel, {Violation of the Sphericity Assumption and Its Effect on Type-I Error Rates in Repeated Measures ANOVA and Multi-Level Linear Models (MLM)}, {Frontiers in Psychology}.
-{8} (2017).
-\url{https://doi.org/\%7B10.3389/fpsyg.2017.01841\%7D}.}
+Beauducel, Violation of the Sphericity Assumption and Its Effect on Type-I Error Rates in Repeated Measures ANOVA and Multi-Level Linear Models (MLM), Frontiers in Psychology.
+8 (n.d.).
+\href{https://doi.org/\%7B10.3389/fpsyg.2017.01841\%7D}{https://doi.org/\{10.3389/fpsyg.2017.01841\}}.
 
 \leavevmode\hypertarget{ref-keselman2001}{}%
-\CSLLeftMargin{{[}29{]} }
-\CSLRightInline{H.
+{[}29{]} H.
 Keselman, J.
 Algina, R.
-Kowalchuk, {The analysis of repeated measures designs: A review}, {British Journal of Mathematica \& Statistical Psychology}.
-{54} (2001) 1--20.
-\url{https://doi.org/\%7B10.1348/000711001159357\%7D}.}
+Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematica \& Statistical Psychology.
+54 (n.d.) 1--20.
+\href{https://doi.org/\%7B10.1348/000711001159357\%7D}{https://doi.org/\{10.1348/000711001159357\}}.
 
 \leavevmode\hypertarget{ref-charan2013}{}%
-\CSLLeftMargin{{[}30{]} }
-\CSLRightInline{Jaykaran.
-Charan, N.
-Kantharia, {How to calculate sample size in animal studies?}, Journal of Pharmacology and Pharmacotherapeutics.
-4 (2013) 303--306.
-\url{https://doi.org/10.4103/0976-500X.119726}.}
+{[}30{]} J. Charan, N. Kantharia, How to calculate sample size in animal studies?, Journal of Pharmacology and Pharmacotherapeutics. 4 (2013) 303--306. \url{https://doi.org/10.4103/0976-500X.119726}.
 
 \leavevmode\hypertarget{ref-barr2013}{}%
-\CSLLeftMargin{{[}31{]} }
-\CSLRightInline{D.J.
+{[}31{]} D.J.
 Barr, R.
@@
 Scheepers, H.J.
-Tily, {Random effects structure for confirmatory hypothesis testing: Keep it maximal}, {JournalL of Memory and Language}.
-{68} (2013) 255--278.
-\url{https://doi.org/\%7B10.1016/j.jml.2012.11.001\%7D}.}
+Tily, Random effects structure for confirmatory hypothesis testing: Keep it maximal, JournalL of Memory and Language.
+68 (n.d.) 255--278.
+\href{https://doi.org/\%7B10.1016/j.jml.2012.11.001\%7D}{https://doi.org/\{10.1016/j.jml.2012.11.001\}}.
 
 \leavevmode\hypertarget{ref-rose2012}{}%
-\CSLLeftMargin{{[}32{]} }
-\CSLRightInline{N.L.
+{[}32{]} N.L.
 Rose, H.D.
@@
 Turner, G.L.
-Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, UK, Geochimica Et Cosmochimica Acta.
+Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, uk, Geochimica et Cosmochimica Acta.
 82 (2012) 113--135.
-\url{https://doi.org/10.1016/j.gca.2010.12.026}.}
+\url{https://doi.org/10.1016/j.gca.2010.12.026}.
 
 \leavevmode\hypertarget{ref-pedersen2019}{}%
-\CSLLeftMargin{{[}33{]} }
-\CSLRightInline{E.J.
-Pedersen, D.L.
-Miller, G.L.
-Simpson, N.
-Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, Peerj.
-7 (2019).
-\url{https://doi.org/10.7717/peerj.6876}.}
+{[}33{]} E.J. Pedersen, D.L. Miller, G.L. Simpson, N. Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, Peerj. 7 (2019). \url{https://doi.org/10.7717/peerj.6876}.
 
 \leavevmode\hypertarget{ref-simpson2018}{}%
-\CSLLeftMargin{{[}34{]} }
-\CSLRightInline{G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.}
+{[}34{]} G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.
 
 \leavevmode\hypertarget{ref-yang2012}{}%
-\CSLLeftMargin{{[}35{]} }
-\CSLRightInline{L.
+{[}35{]} L.
 Yang, G.
@@
 Wang, G.
-Song, {Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality}, {BMC Medical Research Methodology}.
-{12} (2012).
-\url{https://doi.org/\%7B10.1186/1471-2288-12-165\%7D}.}
+Song, Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality, BMC Medical Research Methodology.
+12 (n.d.).
+\href{https://doi.org/\%7B10.1186/1471-2288-12-165\%7D}{https://doi.org/\{10.1186/1471-2288-12-165\}}.
 
 \leavevmode\hypertarget{ref-beck1998}{}%
-\CSLLeftMargin{{[}36{]} }
-\CSLRightInline{N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. (1998) 596--627.}
+{[}36{]} N.
+Beck, S.
+Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science.
+42 (n.d.) 596--627.
+\href{https://doi.org/\%7B10.2307/2991772\%7D}{https://doi.org/\{10.2307/2991772\}}.
 
 \leavevmode\hypertarget{ref-wood2017}{}%
-\CSLLeftMargin{{[}37{]} }
-\CSLRightInline{S.N. Wood, Generalized additive models: An introduction with {}, second edition, CRC Press LLC, Philadelphia, PA, 2017.}
+{[}37{]} S.N. Wood, Generalized additive models: An introduction with , second edition, CRC Press LLC, Philadelphia, PA, 2017.
 
 \leavevmode\hypertarget{ref-r}{}%
-\CSLLeftMargin{{[}38{]} }
-\CSLRightInline{R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.}
+{[}38{]} R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.
 
 \leavevmode\hypertarget{ref-wood2016}{}%
-\CSLLeftMargin{{[}39{]} }
-\CSLRightInline{S.N.
+{[}39{]} S.N.
 Wood, N.
 Pya, B.
-Saefken, {Smoothing Parameter and Model Selection for General Smooth Models}, {Journal of the American Statistical Association}.
-{111} (2016) 1548--1563.
-\url{https://doi.org/\%7B10.1080/01621459.2016.1180986\%7D}.}
+Saefken, Smoothing Parameter and Model Selection for General Smooth Models, Journal of the American Statistical Association.
+111 (n.d.) 1548--1563.
+\href{https://doi.org/\%7B10.1080/01621459.2016.1180986\%7D}{https://doi.org/\{10.1080/01621459.2016.1180986\}}.
 
 \leavevmode\hypertarget{ref-west2014}{}%
-\CSLLeftMargin{{[}40{]} }
-\CSLRightInline{B.T.
-West, K.B.
-Welch, A.T.
+{[}40{]} B.
+West, K.
+Welch, A.
 Galecki, Linear mixed models: A practical guide using statistical software, second edition, Taylor \& Francis, 2014.
-\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.}
+\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.
 
 \leavevmode\hypertarget{ref-wolfinger1996}{}%
-\CSLLeftMargin{{[}41{]} }
-\CSLRightInline{R.D.
+{[}41{]} R.D.
 Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics.
 1 (1996) 205--230.
-\url{http://www.jstor.org/stable/1400366}.}
+\url{http://www.jstor.org/stable/1400366}.
 
 \leavevmode\hypertarget{ref-weiss2005}{}%
-\CSLLeftMargin{{[}42{]} }
-\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://books.google.com/books?id=MQ/_bvWDPsEAC}.}
+{[}42{]} R. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://books.google.com/books?id=MQ/_bvWDPsEAC}.
 
 \leavevmode\hypertarget{ref-geisser1958}{}%
-\CSLLeftMargin{{[}43{]} }
-\CSLRightInline{S.
+{[}43{]} S.
 Geisser, S.W.
-Greenhouse, {An Extension of Box's Results on the Use of the \(F\) Distribution in Multivariate Analysis}, The Annals of Mathematical Statistics.
+Greenhouse, An Extension of Box's Results on the Use of the \(F\) Distribution in Multivariate Analysis, The Annals of Mathematical Statistics.
 29 (1958) 885--891.
-\url{https://doi.org/10.1214/aoms/1177706545}.}
+\url{https://doi.org/10.1214/aoms/1177706545}.
 
 \leavevmode\hypertarget{ref-maxwell2017}{}%
-\CSLLeftMargin{{[}44{]} }
-\CSLRightInline{S.E.
-Maxwell, H.D.
+{[}44{]} S.
+Maxwell, H.
 Delaney, K.
 Kelley, Designing experiments and analyzing data: A model comparison perspective, third edition, Taylor \& Francis, 2017.
-\url{https://books.google.com/books?id=NmFQDwAAQBAJ}.}
+\url{https://books.google.com/books?id=NmFQDwAAQBAJ}.
 
 \leavevmode\hypertarget{ref-molenberghs2004}{}%
-\CSLLeftMargin{{[}45{]} }
-\CSLRightInline{G.
+{[}45{]} G.
 Molenberghs, H.
@@
 Mallinckrodt, R.
-Carroll, {Analyzing incomplete longitudinal clinical trial data}, {Biostatistics}.
-{5} (2004) 445--464.
-\url{https://doi.org/\%7B10.1093/biostatistics/kxh001\%7D}.}
+Carroll, Analyzing incomplete longitudinal clinical trial data, Biostatistics.
+5 (n.d.) 445--464.
+\href{https://doi.org/\%7B10.1093/biostatistics/kxh001\%7D}{https://doi.org/\{10.1093/biostatistics/kxh001\}}.
 
 \leavevmode\hypertarget{ref-scheffer2002}{}%
-\CSLLeftMargin{{[}46{]} }
-\CSLRightInline{J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.}
+{[}46{]} J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.
 
 \leavevmode\hypertarget{ref-potthoff2006}{}%
-\CSLLeftMargin{{[}47{]} }
-\CSLRightInline{R.F.
+{[}47{]} R.F.
 Potthoff, G.E.
@@
 Pieper, V.
-Hasselblad, {Can one assess whether missing data are missing at random in medical studies?}, {STATISTICAL METHODS IN MEDICAL RESEARCH}.
-{15} (2006) 213--234.
-\url{https://doi.org/\%7B10.1191/0962280206sm448oa\%7D}.}
+Hasselblad, Can one assess whether missing data are missing at random in medical studies?, STATISTICAL METHODS IN MEDICAL RESEARCH.
+15 (n.d.) 213--234.
+\href{https://doi.org/\%7B10.1191/0962280206sm448oa\%7D}{https://doi.org/\{10.1191/0962280206sm448oa\}}.
 
 \leavevmode\hypertarget{ref-ma2012}{}%
-\CSLLeftMargin{{[}48{]} }
-\CSLRightInline{Y.
+{[}48{]} Y.
 Ma, M.
 Mazumdar, S.G.
-Memtsoudis, {Beyond Repeated-Measures Analysis of Variance Advanced Statistical Methods for the Analysis of Longitudinal Data in Anesthesia Research}, {Regional Anesthesia and Pain Medicine}.
-{37} (2012) 99--105.
-\url{https://doi.org/\%7B10.1097/AAP.0b013e31823ebc74\%7D}.}
+Memtsoudis, Beyond Repeated-Measures Analysis of Variance Advanced Statistical Methods for the Analysis of Longitudinal Data in Anesthesia Research, Regional Anesthesia and Pain Medicine.
+37 (n.d.) 99--105.
+\href{https://doi.org/\%7B10.1097/AAP.0b013e31823ebc74\%7D}{https://doi.org/\{10.1097/AAP.0b013e31823ebc74\}}.
 
 \leavevmode\hypertarget{ref-nlme}{}%
-\CSLLeftMargin{{[}49{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, {nlme}: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.}
+{[}49{]} J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, nlme: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.
 
 \leavevmode\hypertarget{ref-nelder1972}{}%
-\CSLLeftMargin{{[}50{]} }
-\CSLRightInline{J.A.
-Nelder, R.W.M.
-Wedderburn, Generalized linear models, Journal of the Royal Statistical Society.
-Series A (General).
-135 (1972) 370--384.
-\url{http://www.jstor.org/stable/2344614}.}
+{[}50{]} J.A. Nelder, R.W.M. Wedderburn, Generalized linear models, Journal of the Royal Statistical Society. Series A (General). 135 (1972) 370--384. \url{http://www.jstor.org/stable/2344614}.
 
 \leavevmode\hypertarget{ref-hastie1987}{}%
-\CSLLeftMargin{{[}51{]} }
-\CSLRightInline{T.
+{[}51{]} T.
 Hastie, R.
@@
 82 (1987) 371--386.
-\url{https://doi.org/10.1080/01621459.1987.10478440}.}
+\url{https://doi.org/10.1080/01621459.1987.10478440}.
 
 \leavevmode\hypertarget{ref-hefley2017}{}%
-\CSLLeftMargin{{[}52{]} }
-\CSLRightInline{T.J.
+{[}52{]} T.J.
 Hefley, K.M.
@@
 Williams, M.B.
-Hooten, {The basis function approach for modeling autocorrelation in ecological data}, {ECOLOGY}.
-{98} (2017) 632--646.
-\url{https://doi.org/\%7B10.1002/ecy.1674\%7D}.}
+Hooten, The basis function approach for modeling autocorrelation in ecological data, ECOLOGY.
+98 (n.d.) 632--646.
+\href{https://doi.org/\%7B10.1002/ecy.1674\%7D}{https://doi.org/\{10.1002/ecy.1674\}}.
 
 \leavevmode\hypertarget{ref-wegman1983}{}%
-\CSLLeftMargin{{[}53{]} }
-\CSLRightInline{E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.}
+{[}53{]} E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.
 
 \leavevmode\hypertarget{ref-gratia}{}%
-\CSLLeftMargin{{[}54{]} }
-\CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
+{[}54{]} G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for gams fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.
 
 \leavevmode\hypertarget{ref-harezlak2018}{}%
-\CSLLeftMargin{{[}55{]} }
-\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with r, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
-
-\end{CSLReferences}
+{[}55{]} J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with r, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.
 
@@
 \begin{lstlisting}[language=R]
+set.seed(1)
 ####################Section for calculations######################################
@@
   theme_classic()
-
+\end{lstlisting}
+
+\begin{lstlisting}
+## Error in scale_color_aaas(): could not find function "scale_color_aaas"
+\end{lstlisting}
+
+\begin{lstlisting}[language=R]
 #Combining all
@@
 \begin{lstlisting}[language=R]
+#see https://patchwork.data-imaginist.com/reference/area.html
 layout1 <- c(
@@
   area(2, 2),
-  area(1, 3, 2),
-  area(1,4,2),
-  area(1,5,1)
+  area(1,3,2,5)
 )
+
+#plot(layout2)
 \end{lstlisting}
@@
 ## full convergence after 6 iterations.
-## Gradient range [-0.0001585015,0.0008415702]
-## (score 436.939 & scale 387.386).
-## Hessian positive definite, eigenvalue range [0.0001600441,48.99916].
+## Gradient range [-4.297642e-08,-3.237913e-09]
+## (score 437.636 & scale 390.173).
+## Hessian positive definite, eigenvalue range [0.1251014,49.00133].
 ## Model rank =  5 / 5 
@@
 ##          k'  edf k-index p-value    
-## s(Day)  4   1    0.37  <2e-16 ***
+## s(Day) 4.00 1.51    0.36  <2e-16 ***
 ## ---
@@
 ##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   21.578      1.968   10.96   <2e-16 ***
+## (Intercept)   20.111      1.975   10.18   <2e-16 ***
 ## ---
@@
 ##          edf Ref.df     F p-value   
-## s(Day) 1.002  1.004 21.54 1.09e-05 ***
+## s(Day) 1.511  1.782 8.116 0.00349 **
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.172   Deviance explained = 18.1%
-## -REML = 436.94  Scale est. = 387.39    n = 100
+## R-sq.(adj) =  0.106   Deviance explained = 11.9%
+## -REML = 437.64  Scale est. = 390.17    n = 100
 \end{lstlisting}
@@
 This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
-If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless1).
-From the output, it can be seen that the k-index is effectively \textless0.3, which indicates that the model is not capturing the variability in the data.
+If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless{}1).
+From the output, it can be seen that the k-index is effectively \textless{}0.3, which indicates that the model is not capturing the variability in the data.
 The `edf' (effective degrees of freedom) is an indicator of the complexity of the smooth.
@@
 Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \textasciitilde12\%.
+Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \textasciitilde{}12\%.
 
@@
 ## Method: REML   Optimizer: outer newton
-## full convergence after 8 iterations.
-## Gradient range [-0.0001335263,0.001505324]
-## (score 415.0769 & scale 254.693).
-## Hessian positive definite, eigenvalue range [9.414522e-05,48.49849].
+## full convergence after 9 iterations.
+## Gradient range [-0.0001383864,0.0002277102]
+## (score 414.2119 & scale 246.4112).
+## Hessian positive definite, eigenvalue range [0.0001383659,48.50328].
 ## Model rank =  9 / 9 
@@
 ##                         k'  edf k-index p-value    
-## s(Day):GroupControl    4   1    0.46  <2e-16 ***
-## s(Day):GroupTreatment  4   1    0.46  <2e-16 ***
+## s(Day):GroupControl   4.00 1.00    0.47  <2e-16 ***
+## s(Day):GroupTreatment 4.00 1.82    0.47  <2e-16 ***
 ## ---
@@
 ##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   21.578      1.596   13.52   <2e-16 ***
+## (Intercept)    20.11       1.57   12.81   <2e-16 ***
 ## ---
@@
 ##                         edf Ref.df      F p-value    
-## s(Day):GroupControl   1.004  1.008  1.083   0.299    
-## s(Day):GroupTreatment 1.000  1.001 83.768  <2e-16 ***
+## s(Day):GroupControl   1.001  1.001  5.243  0.0242 *  
+## s(Day):GroupTreatment 1.823  2.071 35.305  <2e-16 ***
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.456   Deviance explained = 46.7%
-## -REML = 415.08  Scale est. = 254.69    n = 100
+## R-sq.(adj) =  0.435   Deviance explained = 45.1%
+## -REML = 414.21  Scale est. = 246.41    n = 100
 \end{lstlisting}
@@
 Moreover, the smooths (plotted via the \passthrough{\lstinline!draw()!} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \textasciitilde43\%.
+From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \textasciitilde{}43\%.
 
@@
 
-gam.check(gam1)
+gam.check(m1)
 \end{lstlisting}
@@
 ## Method: REML   Optimizer: outer newton
-## full convergence after 9 iterations.
-## Gradient range [-1.291119e-06,1.637752e-06]
-## (score 374.509 & scale 96.64781).
-## Hessian positive definite, eigenvalue range [0.02491842,48.04328].
+## full convergence after 7 iterations.
+## Gradient range [-2.60407e-05,5.064354e-06]
+## (score 368.6968 & scale 78.46844).
+## Hessian positive definite, eigenvalue range [0.4967554,48.06648].
 ## Model rank =  10 / 10 
@@
 ##                         k'  edf k-index p-value
-## s(Day):GroupControl   4.00 3.84    0.85   0.075 .
-## s(Day):GroupTreatment 4.00 1.24    0.85   0.085 .
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## s(Day):GroupControl   4.00 3.19    1.17    0.94
+## s(Day):GroupTreatment 4.00 3.79    1.17    0.95
 \end{lstlisting}
@@
 \begin{lstlisting}[language=R]
-summary(gam1)
+summary(m1)
 \end{lstlisting}
@@
 ##                Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)      10.503      1.390   7.554 2.87e-11 ***
-## GroupTreatment   22.150      1.966  11.266  < 2e-16 ***
+## (Intercept)       8.964      1.253   7.156 2.06e-10 ***
+## GroupTreatment   22.293      1.772  12.583  < 2e-16 ***
 ## ---
@@
 ##                         edf Ref.df     F  p-value    
-## s(Day):GroupControl   3.843  3.976   7.975 8.99e-06 ***
-## s(Day):GroupTreatment 1.236  1.419 159.701  < 2e-16 ***
+## s(Day):GroupControl   3.191  3.437 11.05 1.01e-06 ***
+## s(Day):GroupTreatment 3.788  3.957 63.23  < 2e-16 ***
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.794   Deviance explained = 80.6%
-## -REML = 374.51  Scale est. = 96.648    n = 100
-\end{lstlisting}
-
-The resulting model is model \passthrough{\lstinline!gam1!}, which is the model fitted in the main manuscript.
+## R-sq.(adj) =   0.82   Deviance explained = 83.5%
+## -REML =  368.7  Scale est. = 78.468    n = 100
+\end{lstlisting}
+
+The resulting model is\passthrough{\lstinline!m1!}, which is the model fitted in the main manuscript.
 By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group .
-From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\textasciitilde1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 87\% of the variance data.
+From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\textasciitilde{}1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 87\% of the variance data.
 
@@
 ##               df      AIC
-## gam_00 3.004277 883.7160
-## gam_01 4.008467 842.7602
-## gam1   8.395441 750.3446
+## gam_00  3.781952 885.4678
+## gam_01  5.072091 840.7331
+## gam1   10.936093 732.3416
 \end{lstlisting}
@@
 ##    <dbl> <fct>     <int>
-##  1     0 Control       2
-##  2     0 Treatment     4
-##  3     2 Control       7
-##  4     2 Treatment     4
-##  5     5 Control       2
-##  6     5 Treatment     8
-##  7     7 Control       5
-##  8     7 Treatment     8
-##  9    10 Control       4
-## 10    10 Treatment     5
+##  1     0 Control       5
+##  2     0 Treatment     3
+##  3     2 Control       5
+##  4     2 Treatment     5
+##  5     5 Control       4
+##  6     5 Treatment     6
+##  7     7 Control       2
+##  8     7 Treatment     7
+##  9    10 Control       1
+## 10    10 Treatment     6
 \end{lstlisting}

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/Full_document.Rmd
Updated manuscript

@@
 library(nlme)
-library(ggsci)
 library(mgcv)
 library(gratia)
-library(ghibli)
 library(viridis)

### 2021-05-27 Manuscripts/Manuscript by chapters -v2/01-Background.Rmd
Making proposed changes in the manuscript

--- 2021-05-14 Manuscripts/Manuscript by chapters -v1/01-Background.Rmd
@@
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model is only able to accommodate linear trends that are far from adequately representing the biological phenomenon of interest.
+Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
 
-A _post hoc_ analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
+A _post hoc_ analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
 Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
@@
 
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
+During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
 The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data [@skala2010;@vishwanath2009].
@@
 However, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
-And even more importantly, LMEMs also expect a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze non-linear data.
+And even more importantly, LMEMs also assume a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze non-linear data.
 

### 2021-05-27 Manuscripts/Manuscript by chapters -v2/02-Challenges.Rmd
Making proposed changes in the manuscript

--- 2021-05-24 Manuscripts/Manuscript by chapters -v1/02-Challenges.Rmd
@@
 In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$ (independently normally distributed with mean zero and variance $\sigma^2_\mu$).
+Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$ (independently and identically normally distributed with mean zero and variance $\sigma^2$).
 In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
@@
 If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
+
 An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
 This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
-
 In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014].
@@
 
-This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and  missing data.
-In particular, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data.
-To better convey the issues of linearity and correlation  in linear models fitted to non-linear data, simulation is used in the next section.
-
 ## What does an rm-ANOVA fit looks like? A visual representation using simulated data{#simulation}
 
-To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
+To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
 An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme].
@@
 In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity.
+This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments.
 
@@
 Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \@ref(fig:l-q-response) B,D).
+
 A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal appears in Figure \@ref(fig:linear-cases-Appendix) and Figure \@ref(fig:quadratic-cases-Appendix) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs.
+Of notice, Section \@ref(longitudinal-GAMs) uses simulated data that does follow reported trends in the biomedical literature.
 
@@
 
-A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\@ref(fig:l-q-response), D) indicates that the model is not capturing the changes within each group in a good way.
+A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\@ref(fig:l-q-response), D) indicates that the model is not capturing the changes within each group.
 Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
 Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D).
-If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response.
+If such a change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response.
 Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.

### 2021-05-27 Manuscripts/Manuscript by chapters -v2/03-GAM Theory.Rmd
Making proposed changes in the manuscript

--- 2021-05-24 Manuscripts/Manuscript by chapters -v1/03-GAM Theory.Rmd
@@
 
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual error.
+Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the _smooth function_ $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual error.
 
@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].
+One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with inference [@beck1998].
 In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.
 
-The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using _basis functions_ expansions of the covariates and by estimating random coefficients for these basis functions.
-A _basis_ is a set of functions that spans the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
+The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using _basis function_ expansions of the covariates and by estimating random coefficients associated with these basis functions.
+A _basis_ is a set of functions that spans the mathematical space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
 For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.

### 2021-05-27 Manuscripts/Manuscript by chapters -v2/05-Conclusion.Rmd
Making proposed changes in the manuscript

--- 2021-05-21 Manuscripts/Manuscript by chapters -v1/Full_document.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
-\usepackage{listings}
-\newcommand{\passthrough}[1]{#1}
-\lstset{defaultdialect=[5.3]Lua}
-\lstset{defaultdialect=[x86masm]Assembler}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-\usepackage{lineno}
-\usepackage{authblk}
-
-%command for the package lineno
-\linenumbers
-
-%authors
-\author[1]{Ariel I. Mundo}
-\author[2]{John R. Tipton}
-\author[1]{Timothy J. Muldoon}
-\affil[1]{\footnotesize Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
-\affil[2]{\footnotesize Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
-
-\usepackage{xcolor} %theme colors for the code chunks (originally from latex-solarized on GitHub)
-\definecolor{sbase03}{HTML}{002B36}
-\definecolor{sbase02}{HTML}{073642}
-\definecolor{sbase01}{HTML}{586E75}
-\definecolor{sbase00}{HTML}{657B83}
-\definecolor{sbase0}{HTML}{839496}
-\definecolor{sbase1}{HTML}{93A1A1}
-\definecolor{sbase2}{HTML}{EEE8D5}
-\definecolor{sbase3}{HTML}{FDF6E3}
-\definecolor{syellow}{HTML}{B58900}
-\definecolor{sorange}{HTML}{CB4B16}
-\definecolor{sred}{HTML}{DC322F}
-\definecolor{smagenta}{HTML}{D33682}
-\definecolor{sviolet}{HTML}{6C71C4}
-\definecolor{sblue}{HTML}{268BD2}
-\definecolor{scyan}{HTML}{2AA198}
-\definecolor{sgreen}{HTML}{859900}
-%command to set parameter(s) in package listings
-\lstset{
-    % How/what to match
-    sensitive=true,
-    % Border (above and below)
-    frame=lines,
-    % Extra margin on line (align with paragraph)
-    xleftmargin=\parindent,
-    % Put extra space under caption
-    belowcaptionskip=1\baselineskip,
-    % Colors
-    backgroundcolor=\color{sbase3},
-    basicstyle=\color{sbase00}\ttfamily,
-    keywordstyle=\color{scyan},
-    commentstyle=\color{sbase1},
-    stringstyle=\color{sblue},
-    numberstyle=\color{sviolet},
-    identifierstyle=\color{sbase00},
-    % Break long lines into multiple lines?
-    breaklines=true,
-    % Show a character for spaces?
-    showstringspaces=false,
-    tabsize=2
-}
-
-%\lstset{
-%  breaklines=true,
-%  stringstyle=\ttfamily,
-%  backgroundcolor=\color{gray}
-%}
-\usepackage{placeins}
-\usepackage{subfig}
-\usepackage{breqn}
-\usepackage[font={small}]{caption}
-\usepackage{float}
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{\textbf{The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models}}
-\usepackage{etoolbox}
-\makeatletter
-\providecommand{\subtitle}[1]{% add subtitle to \maketitle
-  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
-}
-\makeatother
-\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
-\date{\vspace{-2.5em}}
-
-\begin{document}
-\maketitle
-
-{
-\setcounter{tocdepth}{2}
-\tableofcontents
-}
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
-
-In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, a \emph{linear mixed model} (LMEM).
-Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
-In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures, thereby being an excellent choice to analyze non-linear longitudinal data.
-This paper summarizes the limitations of LMEMs and rm-ANOVA, presents the basic theory of GAMs, and demonstrates their implementation in \(\textsf{R}\) via the package \emph{mgcv} using simulated data that follows longitudinal trends reported in biomedical literature.
-To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:\_\_\_\_\_\_\_\_.
-
-\hypertarget{background}{%
-\section{Background}\label{background}}
-
-Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response {[}\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}{]}, antibody expression {[}\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}{]}, and cell metabolism {[}\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}\protect\hyperlink{ref-greening2018}{9}{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}\protect\hyperlink{ref-sio2016}{10}{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}\protect\hyperlink{ref-kamstra2015}{11}{]}.
-
-Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}\protect\hyperlink{ref-wagenmakers2008}{12}{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}{]}.
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model is only able to accommodate linear trends that are far from adequately representing the biological phenomenon of interest.
-
-A \emph{post hoc} analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
-Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}.
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}\protect\hyperlink{ref-wang2019}{18}{]}.
-
-Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}{]}, consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}\protect\hyperlink{ref-abdi2010}{21}{]}), but they in turn reduce statistical power (1-\(\beta\)){[}\protect\hyperlink{ref-nakagawa2004}{22}{]}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) {[}\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}{]}.
-Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and group sizes {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}\protect\hyperlink{ref-charan2013}{30}{]} which might cause the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible.
-It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}\protect\hyperlink{ref-gueorguieva2004}{13}{]}.
-This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
-At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
-
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}.
-And even more importantly, LMEMs also expect a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}, making them unsuitable to analyze non-linear data.
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
-
-In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]} .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]} and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
-
-The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, \(\textsf{R}\){[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
-Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
-
-This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations is explained in detail.
-Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, using simulated data that reproduces patterns in previously reported studies {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} we illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
-The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in \(\textsf{R}\), in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
-
-In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models by improving the standards for reproducibility in biomedical research.
-
-\FloatBarrier
-
-\hypertarget{challenges-presented-by-longitudinal-studies}{%
-\section{Challenges presented by longitudinal studies}\label{challenges-presented-by-longitudinal-studies}}
-
-\hypertarget{the-repeated-measures-anova}{%
-\subsection{The repeated measures ANOVA}\label{the-repeated-measures-anova}}
-
-The \emph{repeated measures analysis of variance} (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research.
-This statistical methodology requires certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
-Each one of these assumptions is discussed below.
-
-\hypertarget{linear-relationship}{%
-\subsection{Linear relationship}\label{linear-relationship}}
-
-\hypertarget{the-repeated-measures-anova-case}{%
-\subsubsection{The repeated measures ANOVA case}\label{the-repeated-measures-anova-case}}
-
-In a longitudinal biomedical study, two or more groups of subjects (e.g., human subject, mice, samples) are subject to different treatments (e.g., a ``treatment'' group receives a novel drug or intervention vs.~a ``control'' group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response is modeled with \emph{fixed} components.
-The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug/intervention in the ``treatment'' group.
-
-Mathematically speaking, a rm-ANOVA model with an interaction can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
-\label{eq:linear-model}
-\end{equation}
-
-In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
-Independent errors \(\varepsilon_{tij}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently normally distributed with mean zero and variance \(\sigma^2_\mu\)).
-In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy'').
-Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B).
-The linear models then can be expressed as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-\label{eq:ANOVA-by-group}
-\end{equation}
-
-To further simplify the expression, substitute \(\widetilde{\beta_{0}}=\beta_0+\beta_{2}\) and \(\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}\) in the equation for Group B.
-This substitution allows for a different intercept and slope for Groups A and B.
-The model is then written as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-\label{eq:ANOVA-lines}
-\end{equation}
-
-Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response, this means that either presented as Equation \eqref{eq:linear-model}, Equation \eqref{eq:ANOVA-by-group} or Equation \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
-
-\hypertarget{the-linear-mixed-model-case}{%
-\subsubsection{The Linear Mixed Model Case}\label{the-linear-mixed-model-case}}
-
-A linear mixed model (LMEM) is a class of statistical model that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish {[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}.
-A LMEM with interaction between time and treatment for a longitudinal study can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
-\label{eq:LMEM}
-\end{equation}
-
-When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)).
-This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject within each group.
-The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the ``global noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.
-
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data.
-As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
-The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence,the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However,the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
-
-\hypertarget{covariance-in-rm-anova-and-lmems}{%
-\subsection{Covariance in rm-ANOVA and LMEMs}\label{covariance-in-rm-anova-and-lmems}}
-
-In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{41}{]} (For an in-depth analysis of the covariance matrix see {[}\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}{]}).
-
-In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}{]}.
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}\protect\hyperlink{ref-maxwell2017}{44}{]}; and it is higher between consecutive measurements {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and they are not robust if the group sizes are unbalanced {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
-Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
-
-\hypertarget{missing-observations}{%
-\subsection{Missing observations}\label{missing-observations}}
-
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}\protect\hyperlink{ref-weiss2005}{42}{]}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{46}{]}.
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{47}{]}, and in the case of MNAR the missing values are dependent on their value.
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{48}{]}.
-
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}\protect\hyperlink{ref-west2014}{40}{]}.
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
-
-This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and missing data.
-In particular, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data.
-To better convey the issues of linearity and correlation in linear models fitted to non-linear data, simulation is used in the next section.
-
-\hypertarget{simulation}{%
-\subsection{What does an rm-ANOVA fit looks like? A visual representation using simulated data}\label{simulation}}
-
-To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}) is fitted to each group, using \(\textsf{R}\){[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme}{[}\protect\hyperlink{ref-nlme}{49}{]}.
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}, A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}, C).
-In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity.
-
-Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect.
-In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group.
-From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D).
-A more comprehensive exploration of the fit of rm-ANOVA for linear and non-linear longitudinal appears in Figure \ref{fig:linear-cases-Appendix} and Figure \ref{fig:quadratic-cases-Appendix} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/l-q-response-1}
-
-}
-
-\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.}\label{fig:l-q-response}
-\end{figure}
-
-The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response},B).
-When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.
-
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}, C.
-Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \ref{fig:l-q-response} A and B) .
-The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model \eqref{eq:linear-model} to this data produces the fit that appears in panel D in Figure \ref{fig:l-q-response}.
-
-A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\ref{fig:l-q-response}, D) indicates that the model is not capturing the changes within each group in a good way.
-Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
-Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare panels C and D).
-If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response.
-Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
-
-This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-Although the model fitted to the simulated data was an rm-ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM.
-In the following section, we present generalized additive models (GAMs) as a data-driven alternative method to analyze longitudinal non-linear data.
-
-\FloatBarrier
-
-\hypertarget{gams-as-a-special-case-of-generalized-linear-models}{%
-\section{GAMs as a special case of Generalized Linear Models}\label{gams-as-a-special-case-of-generalized-linear-models}}
-
-\hypertarget{gams-and-basis-functions}{%
-\subsection{GAMs and Basis Functions}\label{gams-and-basis-functions}}
-
-Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors{[}\protect\hyperlink{ref-nelder1972}{50}{]}.
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{51}{]}.
-A GAM model can be written as:
-
-\begin{equation}
-  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
-  \label{eq:GAM}
-\end{equation}
-
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the function \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible \emph{smooth functions}.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}\protect\hyperlink{ref-beck1998}{36}{]}.
-In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic, and causes bias at the endpoints of the time period.
-
-The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis functions} expansions of the covariates and by estimating random coefficients for these basis functions.
-A \emph{basis} is a set of functions that spans the space where the smooths that approximate \(f(x_t \mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{52}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
-In the case of Equation \eqref{eq:GAM}, the basis function is \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
-
-A commonly used \emph{basis function} is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
-Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\ref{fig:l-q-response}, C).
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-The timeline can be divided in equally spaced \emph{knots}, each knot being a region where a different basis function will be used.
-Because there are six timepoints for this group, five knots can be used.
-The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is already optimized in the package \emph{mgcv} depending on the number of knots.
-In Panel A of Figure \ref{fig:basis-plot}, the four basis functions (and the intercept) are shown.
-Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the ``wigliness'' of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
-The penalty reduces the ``wigliness'' of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
-
-In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \ref{fig:basis-plot}, Panel C shows the resulting penalized basis functions.
-Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
-Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of \emph{time} is shown in Panel D (orange line) along the simulated values per group, which appear as points.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/basis-plot-1}
-
-}
-
-\caption{Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}\label{fig:basis-plot}
-\end{figure}
-
-\FloatBarrier
-
-\newpage
-
-\hypertarget{longitudinal-GAMs}{%
-\section{The analyisis of longitudinal biomedical data using GAMs}\label{longitudinal-GAMs}}
-
-The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA.
-This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}.
-The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
-
-\hypertarget{simulated-data}{%
-\subsection{Simulated data}\label{simulated-data}}
-
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3(C) in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
-In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group.
-Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
-The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}, A and the inlet, respectively.
-
-\hypertarget{an-interaction-gam-for-longitudinal-data}{%
-\subsection{An interaction GAM for longitudinal data}\label{an-interaction-gam-for-longitudinal-data}}
-
-An interaction model is typically the main interest in longitudinal biomedical data, as it takes into account the effect of treatment, time, and their combination.
-In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time.
-Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in {[}\protect\hyperlink{ref-hefley2017}{52}{]}.
-
-For the data in Figure \ref{fig:sim-smooth-plot},A it is of interest to know the effect that each treatment has in \(\mbox{StO}_2\).
-The model then needs to incorporate independent smooths for \emph{Group} and \emph{Day}, respectively.
-The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric ( \emph{Day} ) and the other is a factor ( \emph{Group} ).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
-Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:
-
-\passthrough{\lstinline!m1<-gam(StO2\_sim\~Group+s(Day,by=Group,k=5,bs="gp"), method='REML',data=dat\_sim!}
-
-This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths for \emph{Group} and \emph{Day} (the parenthesis preceded by \emph{s}) using 5 knots.
-The smooth is constructed using gaussian process smooths, which is indicated by \passthrough{\lstinline!bs="gp"!}.
-These splines are used to model temporal trends and might be particularly suited for long-term studies where the correlation between measurements changes as a function of the time intervals {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-The parametric term \emph{Group} is added to quantify differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot},B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}{[}\protect\hyperlink{ref-gratia}{54}{]}.
-A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{55}{]}.
-
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot},C.
-This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \ref{fig:sim-smooth-plot},B) .
-
-Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
-The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
-However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
-Consider the simulated \(\mbox{StO}_2\) values from Figure (\ref{fig:sim-smooth-plot}, B).
-If 40\% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but the ``Treatment'' smooth takes an overall more linear profile (\ref{fig:sim-smooth-plot}, D).
-Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/sim-smooth-plot-1}
-
-}
-
-\caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.}\label{fig:sim-smooth-plot}
-\end{figure}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/plot-pairwise-comp-1}
-
-}
-
-\caption{Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-B: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the interval does not cover 0.
-In both cases the effect of treatment is significant after day 5.}\label{fig:plot-pairwise-comp}
-\end{figure}
-
-\hypertarget{determination-of-significance-in-gams-for-longitudinal-data}{%
-\subsection{Determination of significance in GAMs for longitudinal data}\label{determination-of-significance-in-gams-for-longitudinal-data}}
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the \emph{difference} between the confidence intervals of the fitted smooths for such groups is non-zero, then a significant difference exists at that time point(s).
-The absence of a \emph{p-value} in this case might seem odd, but the confidence interval comparison can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
-This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}, A where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
-
-With this expectation of different trends in each group, computing the difference between the trends will identify if the observed change is significant.
-The difference between groups with similar trends is likely to yield zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
-
-Consider the calculation of pairwise differences for the smooths in \ref{fig:sim-smooth-plot} B and D.
-Figure \ref{fig:plot-pairwise-comp}, shows the comparison between each treatment group.Here, the effect of the ``Control'' group is compared to that of the ``Treatment'' group.
-The shaded region under the curve highlights the interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and 4 indicates that through that time, the ``Control'' group has higher \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by day 5 it is the ``Treatment'' group the one that has greater contribution.
-This would suggest that the effect of chemotherapy is significant after day 5 for the model used; which is noticeable in the data with missing observations albeit the missing data model has a wider confidence interval and is not able to pick the change between days 7 and 10 that the full dataset model is able to detect.
-
-\FloatBarrier
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
+# Conclusion
 
@@
 Finally, by providing the data and code used in this paper we hope to address the need of creating and sharing reproducible work in biomedical research.
-
-\FloatBarrier
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\leavevmode\hypertarget{ref-roblyer2011}{}%
-\CSLLeftMargin{{[}1{]} }
-\CSLRightInline{D.
-Roblyer, S.
-Ueda, A.
-Cerussi, W.
-Tanamai, A.
-Durkin, R.
-Mehta, D.
-Hsiang, J.A.
-Butler, C.
-McLaren, W.P.
-Chen, B.
-Tromberg, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences of the United States of America.
-108 (2011) 14626--14631.
-\url{https://doi.org/10.1073/pnas.1013103108}.}
-
-\leavevmode\hypertarget{ref-tank2020}{}%
-\CSLLeftMargin{{[}2{]} }
-\CSLRightInline{A.
-Tank, H.M.
-Peterson, V.
-Pera, S.
-Tabassum, A.
-Leproux, T.
-O'Sullivan, E.
-Jones, H.
-Cabral, N.
-Ko, R.S.
-Mehta, B.J.
-Tromberg, D.
-Roblyer, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research.
-22 (2020) 1--10.
-\url{https://doi.org/doi:10.1186/s13058-020-01262-1}.}
-
-\leavevmode\hypertarget{ref-pavlov2018}{}%
-\CSLLeftMargin{{[}3{]} }
-\CSLRightInline{M.V.
-Pavlov, T.I.
-Kalganova, Y.S.
-Lyubimtseva, V.I.
-Plekhanov, G.Y.
-Golubyatnikov, O.Y.
-Ilyinskaya, A.G.
-Orlova, P.V.
-Subochev, D.V.
-Safonov, N.M.
-Shakhova, A.V.
-Maslennikova, {Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy}, {Journal of Biomedical Optics}.
-{23} (2018).
-\url{https://doi.org/\%7B10.1117/1.JBO.23.9.091410\%7D}.}
-
-\leavevmode\hypertarget{ref-demidov2018}{}%
-\CSLLeftMargin{{[}4{]} }
-\CSLRightInline{V.
-Demidov, A.
-Maeda, M.
-Sugita, V.
-Madge, S.
-Sadanand, C.
-Flueraru, I.A.
-Vitkin, {Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography}, {Scientific Reports}.
-{8} (2018).
-\url{https://doi.org/\%7B10.1038/s41598-017-18635-w\%7D}.}
-
-\leavevmode\hypertarget{ref-ritter2001}{}%
-\CSLLeftMargin{{[}5{]} }
-\CSLRightInline{G.
-Ritter, L.
-Cohen, C.
-Williams, E.
-Richards, L.
-Old, S.
-Welt, {Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}, {Cancer Research}.
-{61} (2001) 6851--6859.}
-
-\leavevmode\hypertarget{ref-roth2017}{}%
-\CSLLeftMargin{{[}6{]} }
-\CSLRightInline{E.M.
-Roth, A.C.
-Goldberg, A.L.
-Catapano, A.
-Torri, G.D.
-Yancopoulos, N.
-Stahl, A.
-Brunet, G.
-Lecorps, H.M.
-Colhoun, {Antidrug Antibodies in Patients Treated with Alirocumab}, {New England Journal of Medicine}.
-{376} (2017) 1589--1590.
-\url{https://doi.org/\%7B10.1056/NEJMc1616623\%7D}.}
-
-\leavevmode\hypertarget{ref-jones2018}{}%
-\CSLLeftMargin{{[}7{]} }
-\CSLRightInline{J.D.
-Jones, H.E.
-Ramser, A.E.
-Woessner, K.P.
-Quinn, {In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing}, {Communications Biology}.
-{1} (2018).
-\url{https://doi.org/\%7B10.1038/s42003-018-0206-4\%7D}.}
-
-\leavevmode\hypertarget{ref-skala2010}{}%
-\CSLLeftMargin{{[}8{]} }
-\CSLRightInline{M.C.
-Skala, A.
-Fontanella, L.
-Lan, J.A.
-Izatt, M.W.
-Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics.
-15 (2010).
-\url{https://doi.org/10.1117/1.3285584}.}
-
-\leavevmode\hypertarget{ref-greening2018}{}%
-\CSLLeftMargin{{[}9{]} }
-\CSLRightInline{G.J.
-Greening, K.P.
-Miller, C.R.
-Spainhour, M.D.
-Cato, T.J.
-Muldoon, {Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy}, {Biomedical Optics Express}.
-{9} (2018) 2871--2886.
-\url{https://doi.org/\%7B10.1364/BOE.9.002871\%7D}.}
-
-\leavevmode\hypertarget{ref-sio2016}{}%
-\CSLLeftMargin{{[}10{]} }
-\CSLRightInline{T.T.
-Sio, P.J.
-Atherton, B.J.
-Birckhead, D.J.
-Schwartz, J.A.
-Sloan, D.K.
-Seisler, J.A.
-Martenson, C.L.
-Loprinzi, P.C.
-Griffin, R.F.
-Morton, J.C.
-Anders, T.J.
-Stoffel, R.E.
-Haselow, R.B.
-Mowat, M.A.N.
-Wittich, J.D.
-Bearden III, R.C.
-Miller, {Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {{[}}alliance{]})}, {Supportive Care in Cancer}.
-{24} (2016) 3847--3855.
-\url{https://doi.org/\%7B10.1007/s00520-016-3213-3\%7D}.}
-
-\leavevmode\hypertarget{ref-kamstra2015}{}%
-\CSLLeftMargin{{[}11{]} }
-\CSLRightInline{J.I.
-Kamstra, P.U.
-Dijkstra, M.
-van Leeuwen, J.L.N.
-Roodenburg, J.A.
-Langendijk, {Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study}, {Oral Oncology}.
-{51} (2015) 548--555.
-\url{https://doi.org/\%7B10.1016/j.oraloncology.2015.01.016\%7D}.}
-
-\leavevmode\hypertarget{ref-wagenmakers2008}{}%
-\CSLLeftMargin{{[}12{]} }
-\CSLRightInline{E.-J.
-Wagenmakers, M.
-Lee, T.
-Lodewyckx, G.J.
-Iverson, Bayesian versus frequentist inference, in: H.
-Hoijtink, I.
-Klugkist, P.A.
-Boelen (Eds.), Bayesian Evaluation of Informative Hypotheses, Springer New York, New York, NY, 2008: pp.
-181--207.
-\url{https://doi.org/10.1007/978-0-387-09612-4_9}.}
-
-\leavevmode\hypertarget{ref-gueorguieva2004}{}%
-\CSLLeftMargin{{[}13{]} }
-\CSLRightInline{R.
-Gueorguieva, J.H.
-Krystal, {Move over ANOVA - Progress in analyzing repeated-measures data and its reflection in papers published in the archives of general psychiatry}, Archives of General Psychiatry.
-61 (2004) 310--317.
-\url{https://doi.org/10.1001/archpsyc.61.3.310}.}
-
-\leavevmode\hypertarget{ref-schober2018}{}%
-\CSLLeftMargin{{[}14{]} }
-\CSLRightInline{P.
-Schober, T.R.
-Vetter, Repeated measures designs and analysis of longitudinal data: If at first you do not succeed-try, try again, Anesthesia and Analgesia.
-127 (2018) 569--575.
-\url{https://doi.org/10.1213/ane.0000000000003511}.}
-
-\leavevmode\hypertarget{ref-pinheiro2006}{}%
-\CSLLeftMargin{{[}15{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science \& Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}
-
-\leavevmode\hypertarget{ref-vishwanath2009}{}%
-\CSLLeftMargin{{[}16{]} }
-\CSLRightInline{K.
-Vishwanath, H.
-Yuan, W.T.
-Barry, M.W.
-Dewhirst, N.
-Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia.
-11 (2009) 889--900.
-\url{https://doi.org/10.1593/neo.09580}.}
-
-\leavevmode\hypertarget{ref-dennis2019}{}%
-\CSLLeftMargin{{[}17{]} }
-\CSLRightInline{B.
-Dennis, J.M.
-Ponciano, M.L.
-Taper, S.R.
-Lele, {Errors in Statistical Inference Under Model Misspecification: Evidence, Hypothesis Testing, and AIC}, {Frontiers in Ecology and Evolution}.
-{7} (2019).
-\url{https://doi.org/\%7B10.3389/fevo.2019.00372\%7D}.}
-
-\leavevmode\hypertarget{ref-wang2019}{}%
-\CSLLeftMargin{{[}18{]} }
-\CSLRightInline{B.
-Wang, Z.
-Zhou, H.
-Wang, X.M.
-Tu, C.
-Feng, {The p-value and model specification in statistics}, {General Psychiatry}.
-{32} (2019).
-\url{https://doi.org/\%7B10.1136/gpsych-2019-100081\%7D}.}
-
-\leavevmode\hypertarget{ref-liu2010}{}%
-\CSLLeftMargin{{[}19{]} }
-\CSLRightInline{C.
-Liu, T.P.
-Cripe, M.-O.
-Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy.
-18 (2010) 1724--1730.
-\url{https://doi.org/10.1038/mt.2010.127}.}
-
-\leavevmode\hypertarget{ref-halsey2015}{}%
-\CSLLeftMargin{{[}20{]} }
-\CSLRightInline{L.G.
-Halsey, D.
-Curran-Everett, S.L.
-Vowler, G.B.
-Drummond, {The fickle P value generates irreproducible results}, {Nature Methods}.
-{12} (2015) 179--185.
-\url{https://doi.org/\%7B10.1038/nmeth.3288\%7D}.}
-
-\leavevmode\hypertarget{ref-abdi2010}{}%
-\CSLLeftMargin{{[}21{]} }
-\CSLRightInline{H. Abdi, {Holm's Sequential Bonferroni Procedure}, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.}
-
-\leavevmode\hypertarget{ref-nakagawa2004}{}%
-\CSLLeftMargin{{[}22{]} }
-\CSLRightInline{S.
-Nakagawa, {A farewell to Bonferroni: the problems of low statistical power and publication bias}, {Behavioral Ecology}.
-{15} (2004) 1044--1045.
-\url{https://doi.org/\%7B10.1093/beheco/arh107\%7D}.}
-
-\leavevmode\hypertarget{ref-gelman2012}{}%
-\CSLLeftMargin{{[}23{]} }
-\CSLRightInline{A.
-Gelman, J.
-Hill, M.
-Yajima, {Why We (Usually) Don't Have to Worry About Multiple Comparisons}, {Journal of Research on Educational Effectiveness}.
-{5} (2012) 189--211.
-\url{https://doi.org/\%7B10.1080/19345747.2011.618213\%7D}.}
-
-\leavevmode\hypertarget{ref-albers2019}{}%
-\CSLLeftMargin{{[}24{]} }
-\CSLRightInline{C. Albers, {The problem with unadjusted multiple and sequential statistical testing}, {Nature Communications}. {10} (2019). \url{https://doi.org/\%7B10.1038/s41467-019-09941-0\%7D}.}
-
-\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
-\CSLLeftMargin{{[}25{]} }
-\CSLRightInline{C.
-Ugrinowitsch, G.W.
-Fellingham, M.D.
-Ricard, Limitations of ordinary least squares models in analyzing repeated measures data, Medicine and Science in Sports and Exercise.
-36 (2004) 2144--2148.
-\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.}
-
-\leavevmode\hypertarget{ref-huynh1976}{}%
-\CSLLeftMargin{{[}26{]} }
-\CSLRightInline{H.
-Huynh, L.S.
-Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics.
-1 (1976) 69--82.
-\url{https://doi.org/10.3102/10769986001001069}.}
-
-\leavevmode\hypertarget{ref-greenhouse1959}{}%
-\CSLLeftMargin{{[}27{]} }
-\CSLRightInline{S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.}
-
-\leavevmode\hypertarget{ref-haverkamp2017}{}%
-\CSLLeftMargin{{[}28{]} }
-\CSLRightInline{N.
-Haverkamp, A.
-Beauducel, {Violation of the Sphericity Assumption and Its Effect on Type-I Error Rates in Repeated Measures ANOVA and Multi-Level Linear Models (MLM)}, {Frontiers in Psychology}.
-{8} (2017).
-\url{https://doi.org/\%7B10.3389/fpsyg.2017.01841\%7D}.}
-
-\leavevmode\hypertarget{ref-keselman2001}{}%
-\CSLLeftMargin{{[}29{]} }
-\CSLRightInline{H.
-Keselman, J.
-Algina, R.
-Kowalchuk, {The analysis of repeated measures designs: A review}, {British Journal of Mathematica \& Statistical Psychology}.
-{54} (2001) 1--20.
-\url{https://doi.org/\%7B10.1348/000711001159357\%7D}.}
-
-\leavevmode\hypertarget{ref-charan2013}{}%
-\CSLLeftMargin{{[}30{]} }
-\CSLRightInline{Jaykaran.
-Charan, N.
-Kantharia, {How to calculate sample size in animal studies?}, Journal of Pharmacology and Pharmacotherapeutics.
-4 (2013) 303--306.
-\url{https://doi.org/10.4103/0976-500X.119726}.}
-
-\leavevmode\hypertarget{ref-barr2013}{}%
-\CSLLeftMargin{{[}31{]} }
-\CSLRightInline{D.J.
-Barr, R.
-Levy, C.
-Scheepers, H.J.
-Tily, {Random effects structure for confirmatory hypothesis testing: Keep it maximal}, {JournalL of Memory and Language}.
-{68} (2013) 255--278.
-\url{https://doi.org/\%7B10.1016/j.jml.2012.11.001\%7D}.}
-
-\leavevmode\hypertarget{ref-rose2012}{}%
-\CSLLeftMargin{{[}32{]} }
-\CSLRightInline{N.L.
-Rose, H.D.
-Yang, S.D.
-Turner, G.L.
-Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, UK, Geochimica Et Cosmochimica Acta.
-82 (2012) 113--135.
-\url{https://doi.org/10.1016/j.gca.2010.12.026}.}
-
-\leavevmode\hypertarget{ref-pedersen2019}{}%
-\CSLLeftMargin{{[}33{]} }
-\CSLRightInline{E.J.
-Pedersen, D.L.
-Miller, G.L.
-Simpson, N.
-Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, Peerj.
-7 (2019).
-\url{https://doi.org/10.7717/peerj.6876}.}
-
-\leavevmode\hypertarget{ref-simpson2018}{}%
-\CSLLeftMargin{{[}34{]} }
-\CSLRightInline{G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.}
-
-\leavevmode\hypertarget{ref-yang2012}{}%
-\CSLLeftMargin{{[}35{]} }
-\CSLRightInline{L.
-Yang, G.
-Qin, N.
-Zhao, C.
-Wang, G.
-Song, {Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality}, {BMC Medical Research Methodology}.
-{12} (2012).
-\url{https://doi.org/\%7B10.1186/1471-2288-12-165\%7D}.}
-
-\leavevmode\hypertarget{ref-beck1998}{}%
-\CSLLeftMargin{{[}36{]} }
-\CSLRightInline{N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. (1998) 596--627.}
-
-\leavevmode\hypertarget{ref-wood2017}{}%
-\CSLLeftMargin{{[}37{]} }
-\CSLRightInline{S.N. Wood, Generalized additive models: An introduction with {}, second edition, CRC Press LLC, Philadelphia, PA, 2017.}
-
-\leavevmode\hypertarget{ref-r}{}%
-\CSLLeftMargin{{[}38{]} }
-\CSLRightInline{R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.}
-
-\leavevmode\hypertarget{ref-wood2016}{}%
-\CSLLeftMargin{{[}39{]} }
-\CSLRightInline{S.N.
-Wood, N.
-Pya, B.
-Saefken, {Smoothing Parameter and Model Selection for General Smooth Models}, {Journal of the American Statistical Association}.
-{111} (2016) 1548--1563.
-\url{https://doi.org/\%7B10.1080/01621459.2016.1180986\%7D}.}
-
-\leavevmode\hypertarget{ref-west2014}{}%
-\CSLLeftMargin{{[}40{]} }
-\CSLRightInline{B.T.
-West, K.B.
-Welch, A.T.
-Galecki, Linear mixed models: A practical guide using statistical software, second edition, Taylor \& Francis, 2014.
-\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.}
-
-\leavevmode\hypertarget{ref-wolfinger1996}{}%
-\CSLLeftMargin{{[}41{]} }
-\CSLRightInline{R.D.
-Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics.
-1 (1996) 205--230.
-\url{http://www.jstor.org/stable/1400366}.}
-
-\leavevmode\hypertarget{ref-weiss2005}{}%
-\CSLLeftMargin{{[}42{]} }
-\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://books.google.com/books?id=MQ/_bvWDPsEAC}.}
-
-\leavevmode\hypertarget{ref-geisser1958}{}%
-\CSLLeftMargin{{[}43{]} }
-\CSLRightInline{S.
-Geisser, S.W.
-Greenhouse, {An Extension of Box's Results on the Use of the \(F\) Distribution in Multivariate Analysis}, The Annals of Mathematical Statistics.
-29 (1958) 885--891.
-\url{https://doi.org/10.1214/aoms/1177706545}.}
-
-\leavevmode\hypertarget{ref-maxwell2017}{}%
-\CSLLeftMargin{{[}44{]} }
-\CSLRightInline{S.E.
-Maxwell, H.D.
-Delaney, K.
-Kelley, Designing experiments and analyzing data: A model comparison perspective, third edition, Taylor \& Francis, 2017.
-\url{https://books.google.com/books?id=NmFQDwAAQBAJ}.}
-
-\leavevmode\hypertarget{ref-molenberghs2004}{}%
-\CSLLeftMargin{{[}45{]} }
-\CSLRightInline{G.
-Molenberghs, H.
-Thijs, I.
-Jansen, C.
-Beunckens, M.
-Kenward, C.
-Mallinckrodt, R.
-Carroll, {Analyzing incomplete longitudinal clinical trial data}, {Biostatistics}.
-{5} (2004) 445--464.
-\url{https://doi.org/\%7B10.1093/biostatistics/kxh001\%7D}.}
-
-\leavevmode\hypertarget{ref-scheffer2002}{}%
-\CSLLeftMargin{{[}46{]} }
-\CSLRightInline{J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.}
-
-\leavevmode\hypertarget{ref-potthoff2006}{}%
-\CSLLeftMargin{{[}47{]} }
-\CSLRightInline{R.F.
-Potthoff, G.E.
-Tudor, K.S.
-Pieper, V.
-Hasselblad, {Can one assess whether missing data are missing at random in medical studies?}, {STATISTICAL METHODS IN MEDICAL RESEARCH}.
-{15} (2006) 213--234.
-\url{https://doi.org/\%7B10.1191/0962280206sm448oa\%7D}.}
-
-\leavevmode\hypertarget{ref-ma2012}{}%
-\CSLLeftMargin{{[}48{]} }
-\CSLRightInline{Y.
-Ma, M.
-Mazumdar, S.G.
-Memtsoudis, {Beyond Repeated-Measures Analysis of Variance Advanced Statistical Methods for the Analysis of Longitudinal Data in Anesthesia Research}, {Regional Anesthesia and Pain Medicine}.
-{37} (2012) 99--105.
-\url{https://doi.org/\%7B10.1097/AAP.0b013e31823ebc74\%7D}.}
-
-\leavevmode\hypertarget{ref-nlme}{}%
-\CSLLeftMargin{{[}49{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, {nlme}: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.}
-
-\leavevmode\hypertarget{ref-nelder1972}{}%
-\CSLLeftMargin{{[}50{]} }
-\CSLRightInline{J.A.
-Nelder, R.W.M.
-Wedderburn, Generalized linear models, Journal of the Royal Statistical Society.
-Series A (General).
-135 (1972) 370--384.
-\url{http://www.jstor.org/stable/2344614}.}
-
-\leavevmode\hypertarget{ref-hastie1987}{}%
-\CSLLeftMargin{{[}51{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association.
-82 (1987) 371--386.
-\url{https://doi.org/10.1080/01621459.1987.10478440}.}
-
-\leavevmode\hypertarget{ref-hefley2017}{}%
-\CSLLeftMargin{{[}52{]} }
-\CSLRightInline{T.J.
-Hefley, K.M.
-Broms, B.M.
-Brost, F.E.
-Buderman, S.L.
-Kay, H.R.
-Scharf, J.R.
-Tipton, P.J.
-Williams, M.B.
-Hooten, {The basis function approach for modeling autocorrelation in ecological data}, {ECOLOGY}.
-{98} (2017) 632--646.
-\url{https://doi.org/\%7B10.1002/ecy.1674\%7D}.}
-
-\leavevmode\hypertarget{ref-wegman1983}{}%
-\CSLLeftMargin{{[}53{]} }
-\CSLRightInline{E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.}
-
-\leavevmode\hypertarget{ref-gratia}{}%
-\CSLLeftMargin{{[}54{]} }
-\CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
-
-\leavevmode\hypertarget{ref-harezlak2018}{}%
-\CSLLeftMargin{{[}55{]} }
-\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with r, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
-
-\end{CSLReferences}
-
-\hypertarget{appendix-appendix}{%
-\appendix}
-
-\hypertarget{code-for-manuscript-data}{%
-\section{Code for Manuscript data}\label{code-for-manuscript-data}}
-
-This section presents the code used to generate figures, models and simulated data from Sections 3 and 4 from the main manuscript.
-
-\hypertarget{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}{%
-\subsection{Compound symmetry and independent errors in linear and quadratic responses}\label{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}}
-
-This section simulated linear and quadratic data in the same manner as in Section \ref{simulation}.
-The linear simulations using Figure \ref{fig:linear-cases-Appendix} show in panels A and D the simulated mean responses and individual data points.
-Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-The following code produces a more comprehensive exploration of Figure \ref{fig:l-q-response} in the main manuscript.
-
-\begin{lstlisting}[language=R]
-####################Section for calculations######################################
-################################################################################
-
-## Example with linear response
-
-#This function simulates data using a linear or quadratic mean response and each with correlated
-#or uncorrelated errors. Each group has a different slope/concavity.
-example <- function(n_time = 6, #number of time points
-                    fun_type = "linear", #type of response
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  x <- seq(1,6, length.out = n_time)
-
-  #Create mean response matrix: linear or quadratic
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # quadratic response (non-linear)
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-
-  #create an array where individual observations per each time point for each group are to be stored. Currently using 10 observations per timepoint
-  y <- array(0, dim = c(length(x), 2, 10))
-
-  #Create array to store the "errors" for each group at each timepoint. The "errors" are the 
-  #between-group variability in the response.
-  errors <- array(0, dim = c(length(x), 2, 10))
-  #create an array where 10 observations per each time point for each group are to be stored
-
-  #The following cycles create independent or correlated responses.
-To each value of mu (mean response per group) a randomly generated error (correlated or uncorrelated) is added and thus the individual response is created.
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors: variance covariance matrix
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-#labelling y and errors  
-  dimnames(y) <- list(time = x, 
-                      treatment = 1:2, 
-                      subject = 1:10)
-
-  dimnames(errors) <- list(time = x, 
-                           treatment = 1:2, 
-                           subject = 1:10)
-
-  #labeling the mean response
-  dimnames(mu) <- list(time = x, 
-                       treatment = 1:2)
-
-  #convert y, mu and errors to  dataframes with time, treatment and subject columns
-  dat <- as.data.frame.table(y, 
-                             responseName = "y")
-  dat_errors <- as.data.frame.table(errors, 
-                                    responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, 
-                                responseName = "mu")
-
-  #join the dataframes to show mean response and errors per subject
-  dat <- left_join(dat, dat_errors, 
-                   by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, 
-                   by = c("time", "treatment"))
-  #add time
-  dat$time <- as.numeric(as.character(dat$time))
-  #label subjects per group
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, 
-                                  treatment, 
-                                  sep = "-")))
-
-  ## repeated measures ANOVA in R
-#time and treatment interaction model, compound symmetry required by the model
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  #create a prediction frame where the model can be used for plotting purposes
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  #add model predictions to the dataframe that has the simulated data
-  dat$y_pred <- predict(fit_lme)
-
-  #return everything in a list
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_lme = fit_lme
-
-  ))
-}
-##################Section for plotting#################################
-#######################################################################
-#This function will create the plots for either a "linear" or "quadratic" response
-
-plot_example <- function(sim_dat) {
-  ## Plot the simulated data (scatterplot)
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = treatment, 
-               color = treatment)
-           ) +
-    geom_point(show.legend=FALSE) +
-    labs(y='response')+
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment),
-              show.legend=FALSE) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-  #plot the simulated data with trajectories per each subject
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = subject, 
-               color = treatment)
-           ) +
-    geom_line(aes(size = "Subjects"),
-              show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment,
-                  size = "Simulated Truth"), 
-              lty = 1,show.legend = FALSE) +
-    labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                face = "bold"),
-     text=element_text(size=30))+
-    thm
-
-  #plot the errors
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = errors,
-               group = subject, 
-               color = treatment)) +
-    geom_line(show.legend=FALSE) +
-     labs(y='errors')+
-     theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-   #plot the model predictions
-  p4 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point()+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), 
-                  group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_lme, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population")) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-
-#Store each plot in a separate object
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent")) 
-\end{lstlisting}
-
-\begin{figure}[H]
-\includegraphics{Full_document_files/figure-latex/linear-cases-Appendix-1} \caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data.}\label{fig:linear-cases-Appendix}
-\end{figure}
-
-For the quadratic response case, Figure \ref{fig:quadratic-cases-Appendix} shows the simulated responses using compound symmetry and independent errors.
-
-\begin{figure}[H]
-\includegraphics{Full_document_files/figure-latex/quadratic-cases-Appendix-1} \caption{Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.}\label{fig:quadratic-cases-Appendix}
-\end{figure}
-
-\hypertarget{basis-functions-and-gams}{%
-\subsection{Basis functions and GAMs}\label{basis-functions-and-gams}}
-
-This code produces Figure \ref{fig:basis-plot} from the main manuscript.
-Briefly, a non-linear (quadratic) response is simulated a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
-The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
-
-\begin{lstlisting}[language=R]
-#generate the response: the same initial procedure from the previous section to simulate
-#the response
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-
- #label each table
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
-
- #Convert to dataframes with subject, time and group columns
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
-
- #label subject per group
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
- #extract  "Group 1" to fit the GAM
-  dat<-subset(dat,treatment==1)
- #keep just the response and timepoint columns
-   dat<-dat[,c('y','time')]
-
-   #GAM model of time, 5 knots
-gm<-gam(y~s(time,k=5),data=dat)
-
-#model_matrix (also known as) 'design matrix'
-#will contain the smooths used to create  model 'gm'
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be 
-#multiplied by the coefficients
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final smooth by aggregating the basis functions
-
-smooth<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(smooth=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,
-            aes(x=time,
-                y=value_orig,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(size=sz,
-            show.legend=FALSE)+
-  geom_point(size=sz+1,
-             show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,
-             labeller = as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#penalized basis
-p12<-ggplot(basis_plot,
-            aes(x=time,
-                y=mod_val,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(show.legend = FALSE,
-            size=sz)+
-  geom_point(show.legend = FALSE,
-             size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,
-             labeller=as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#heatmap of the penalization coefficient
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,
-            aes(x=Basis,
-                y=Basis))+
-  geom_tile(aes(fill = cof), 
-            colour = "black") +
-    scale_fill_gradient(low = "white",
-                        high = "#B50A2AFF")+ #color picked from KikiMedium
-  labs(x='Basis',
-       y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),
-            size=7,
-            show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,
-            aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  scale_color_aaas()+
-  labs(y='Simulated \n response')+
-  geom_line(data=smooth,
-            aes(x=time,
-                y=smooth),
-            color="#6C581DFF",
-            size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/basis-plot-appendix-1}
-
-}
-
-\caption{Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}\label{fig:basis-plot-appendix}
-\end{figure}
-
-\hypertarget{tumor-data-simulation}{%
-\section{Longitudinal biomedical data simulation and GAMs}\label{tumor-data-simulation}}
-
-This section describes how to fit GAMs to longitudinal data using simulated data.
-First, data is simulated according to Section \ref{longitudinal-GAMs}, where reported data of oxygen saturation (\(\mbox{StO}_2\)) in tumors under either chemotherapy or saline control is used as a starting point to generate individual responses in each group.
-
-\begin{lstlisting}[language=R]
-#Dataframe that contains the original reported trends
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
-
-## plot the mean response
-f1<-ggplot(dat, 
-           aes(x = Day, 
-               y = StO2, 
-               color = Group)) +
-    geom_line(size=1,
-              show.legend = FALSE)+
-    geom_point(show.legend = FALSE,
-               size=1.5,
-               alpha=0.5)+
-  labs(y=expression(paste(StO[2],
-                          ' (real)')))+
-  theme_classic()+
-  thm+
-    scale_x_continuous(breaks=c(0,5,10))+
-    scale_y_continuous(breaks=c(0,40))+
-  plot_layout(tag_level = 'new')+
-  theme(
-    plot.background = element_rect(fill = "transparent", 
-                                   color = NA),
-    axis.text=element_text(size=14)
-  )
-
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-
-n <- 10 #number of observations
-sd <- 10 #approximate sd from paper
-df <- 6
-dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, 
-           aes(x = Day, 
-               y = StO2_sim, 
-               color = Group)) +
-    geom_point(show.legend=FALSE,
-               size=1.5,
-               alpha=0.5)+
-    stat_summary(aes(y = StO2_sim,
-                     group=Group), 
-                 fun=mean, geom="line",
-                 size=1,
-                 show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],
-                         '(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-  thm+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
-\end{lstlisting}
-
-\hypertarget{a-basic-workflow-for-gams}{%
-\subsection{A basic Workflow for GAMs}\label{a-basic-workflow-for-gams}}
-
-This section shows a basic workflow to fit a series of increasingly complex GAMs to the simulated data from the previous section.
-Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
-
-\hypertarget{first-model}{%
-\subsubsection{First model}\label{first-model}}
-
-The first model fitted to the data is one that only accounts for different smooths by day.
-The model syntax specifies that \passthrough{\lstinline!gam\_00!} is the object that will contain all the model information, and that the model attempts to explain changes in \passthrough{\lstinline!StO2\_sim!} (simulated \(\mbox{StO}_2\)) using a smooth per \passthrough{\lstinline!Day!}.
-The model will use 5 knots (\passthrough{\lstinline!k=5!}) for the smooth.
-And that the smooth is constructed using gaussian process basis (\passthrough{\lstinline!bs="gp"!}).
-The smoothing parameter estimation method used is the restricted maximum likelihood (\passthrough{\lstinline!REML!}).
-
-\begin{lstlisting}[language=R]
-gam_00<-gam(StO2_sim ~ s(Day, k = 5,bs="gp"),
-            method='REML',
-            data  = dat_sim)
-\end{lstlisting}
-
-To obtain model diagnostics, two methodologies are to be used: 1) graphical diagnostics, and 2) a model check.
-In the first case, the functions \passthrough{\lstinline!appraise!} and \passthrough{\lstinline!draw!} from the package \emph{gratia} can be used to obtain a single output with all the graphical diagnostics.
-For model check, the functions \passthrough{\lstinline!gam.check!} and \passthrough{\lstinline!summary!} from \emph{mgcv} provide detailed information about the model fit and its parameters.
-
-\hypertarget{graphical-diagnostics}{%
-\paragraph{Graphical diagnostics}\label{graphical-diagnostics}}
-
-\begin{lstlisting}
-## Error: The given dimensions cannot hold all panels. Please increase `ncol` or `nrow`
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/first-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the first GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:first-GAM-diag}
-\end{figure}
-
-From the output of the function \passthrough{\lstinline!appraise!} in Figure \ref{fig:first-GAM-diag}, the major indicators of concern about the model are the QQ plot of residuals and the histogram of residuals.
-The QQ plot shows that the errors are not reasonably located along the 45\(^{\circ}\) line (which indicates normality), as there are multiple points that deviate from the trend, specially in the tails.
-The histogram also shows that the variation (residuals) is not following the assumption of a normal distribution.
-
-The \passthrough{\lstinline!draw!} function permits to plot the smooths as \passthrough{\lstinline!ggplot2!} objects, which eases subsequent manipulation, if desired.
-Because model \passthrough{\lstinline!gam\_00!} specifies only one smooth for the time covariate (Day), the plot only contains only one smooth.
-Note that the smooth shows an almost linear profile.
-
-\hypertarget{model-check}{%
-\paragraph{Model check}\label{model-check}}
-
-\begin{lstlisting}[language=R]
-#need to add figure number and caption
-gam.check(gam_00)
-\end{lstlisting}
-
-\includegraphics{Full_document_files/figure-latex/first-GAM-check-1}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 6 iterations.
-## Gradient range [-0.0001585015,0.0008415702]
-## (score 436.939 & scale 387.386).
-## Hessian positive definite, eigenvalue range [0.0001600441,48.99916].
-## Model rank =  5 / 5 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##        k' edf k-index p-value    
-## s(Day)  4   1    0.37  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-summary(gam_00)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ s(Day, k = 5, bs = "gp")
-## 
-## Parametric coefficients:
-##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   21.578      1.968   10.96   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##          edf Ref.df     F  p-value    
-## s(Day) 1.002  1.004 21.54 1.09e-05 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.172   Deviance explained = 18.1%
-## -REML = 436.94  Scale est. = 387.39    n = 100
-\end{lstlisting}
-
-Special attention must be paid to the `k-index' from \passthrough{\lstinline!gam.check!}.
-This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
-If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless1).
-From the output, it can be seen that the k-index is effectively \textless0.3, which indicates that the model is not capturing the variability in the data.
-The `edf' (effective degrees of freedom) is an indicator of the complexity of the smooth.
-Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
-
-From the \passthrough{\lstinline!summary!} function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
-The link function is `identity' as the model does not make any transformation on the predictors.
-The `significance of smooth terms' \emph{p-value} indicates if each smooth is adding significance to the model.
-Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \textasciitilde12\%.
-
-\hypertarget{second-model}{%
-\subsubsection{Second model}\label{second-model}}
-
-The major flaw of \passthrough{\lstinline!gam\_00!} is that this model is not taking into account the fact that the data is nested in groups.
-The next iteration is a model where a different smooth of time (Day) is assigned for each group using \passthrough{\lstinline!by=Group!} in the model syntax.
-
-\begin{lstlisting}[language=R]
-gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5,bs="gp"),
-            method='REML',
-            data  = dat_sim)
-
-gam.check(gam_01)
-\end{lstlisting}
-
-\includegraphics{Full_document_files/figure-latex/second-GAM-1}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 8 iterations.
-## Gradient range [-0.0001335263,0.001505324]
-## (score 415.0769 & scale 254.693).
-## Hessian positive definite, eigenvalue range [9.414522e-05,48.49849].
-## Model rank =  9 / 9 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                       k' edf k-index p-value    
-## s(Day):GroupControl    4   1    0.46  <2e-16 ***
-## s(Day):GroupTreatment  4   1    0.46  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-summary(gam_01)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ s(Day, by = Group, k = 5, bs = "gp")
-## 
-## Parametric coefficients:
-##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   21.578      1.596   13.52   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                         edf Ref.df      F p-value    
-## s(Day):GroupControl   1.004  1.008  1.083   0.299    
-## s(Day):GroupTreatment 1.000  1.001 83.768  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.456   Deviance explained = 46.7%
-## -REML = 415.08  Scale est. = 254.69    n = 100
-\end{lstlisting}
-
-Diagnostics for this model indicate that the k-index is still below 1 (0.32 from \passthrough{\lstinline!gam.check!}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
-Moreover, the smooths (plotted via the \passthrough{\lstinline!draw()!} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \textasciitilde43\%.
-
-\begin{lstlisting}
-## Error: The given dimensions cannot hold all panels. Please increase `ncol` or `nrow`
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/second-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the second GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:second-GAM-diag}
-\end{figure}
-
-\hypertarget{third-model}{%
-\subsubsection{Third model}\label{third-model}}
-
-Model \passthrough{\lstinline!gam\_00!} was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted, the model diagnostics, and, the low variance explained by the model.
-On the other hand, \passthrough{\lstinline!gam\_01!} takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \ref{longitudinal-GAMs}, in order to differentiate between each group a parametric term needs to be added to the model for the interaction of \emph{Day} and \emph{Group}.
-
-\begin{lstlisting}[language=R]
-#GAM for StO2
-
-gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
-            method='REML',
-            data  = dat_sim)
-
-gam.check(gam1)
-\end{lstlisting}
-
-\includegraphics{Full_document_files/figure-latex/final-model-Appendix-1}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 9 iterations.
-## Gradient range [-1.291119e-06,1.637752e-06]
-## (score 374.509 & scale 96.64781).
-## Hessian positive definite, eigenvalue range [0.02491842,48.04328].
-## Model rank =  10 / 10 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                         k'  edf k-index p-value  
-## s(Day):GroupControl   4.00 3.84    0.85   0.075 .
-## s(Day):GroupTreatment 4.00 1.24    0.85   0.085 .
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-summary(gam1)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ Group + s(Day, by = Group, k = 5, bs = "gp")
-## 
-## Parametric coefficients:
-##                Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)      10.503      1.390   7.554 2.87e-11 ***
-## GroupTreatment   22.150      1.966  11.266  < 2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                         edf Ref.df       F  p-value    
-## s(Day):GroupControl   3.843  3.976   7.975 8.99e-06 ***
-## s(Day):GroupTreatment 1.236  1.419 159.701  < 2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.794   Deviance explained = 80.6%
-## -REML = 374.51  Scale est. = 96.648    n = 100
-\end{lstlisting}
-
-The resulting model is model \passthrough{\lstinline!gam1!}, which is the model fitted in the main manuscript.
-By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group .
-From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\textasciitilde1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 87\% of the variance data.
-
-\begin{lstlisting}
-## Error: The given dimensions cannot hold all panels. Please increase `ncol` or `nrow`
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/final-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the final GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooths for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:final-GAM-diag}
-\end{figure}
-
-\hypertarget{comparing-models-via-aic}{%
-\subsubsection{Comparing models via AIC}\label{comparing-models-via-aic}}
-
-One final comparison that can be made for model selection involves the use of the Aikake Information Criterion (AIC).
-This metric is used to estimate information loss, which we want to minimize with an appropriate model.
-Therefore, when 2 or more models are compared, the model with lower AIC is preferred.
-In R, the comparison is done using the \passthrough{\lstinline!AIC!} function.
-
-\begin{lstlisting}[language=R]
-AIC(gam_00,gam_01,gam1)
-\end{lstlisting}
-
-\begin{lstlisting}
-##              df      AIC
-## gam_00 3.004277 883.7160
-## gam_01 4.008467 842.7602
-## gam1   8.395441 750.3446
-\end{lstlisting}
-
-The output in this case is expected: model \passthrough{\lstinline!gam1!} has a lower AIC (711.46) whereas the initial two models have higher AICs (891 and 850).
-The AIC should not be considered as the only estimator of model goodness, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
-
-\hypertarget{gam-and-linear-model-plots-and-missing-data}{%
-\section{GAM and Linear model plots and Missing data}\label{gam-and-linear-model-plots-and-missing-data}}
-
-This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\passthrough{\lstinline!gam1!}), linear model and GAM on data with missing observations are presented.
-Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inlet are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
-
-\hypertarget{gam-and-linear-model-plots}{%
-\subsection{GAM and Linear model plots}\label{gam-and-linear-model-plots}}
-
-This code chunk creates panels B and D in Figure \ref{fig:sim-smooth-plot}.
-Note that this code uses the final GAM from the previous section (\passthrough{\lstinline!gam1!}), so the simulated data and the model should be generated before running this section.
-
-\begin{lstlisting}[language=R]
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-#creates a dataframe using the length of the covariates for the GAM
-gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
-                         subject=factor(rep(1:10)))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#adds the predictions to the grid and creates a confidence interval for GAM
-gam_predict<-gam_predict%>%
-    mutate(fit = predict(gam1,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(gam1, gam_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=gam_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'complete')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-
-#posthoc comparisons for the linear model
-library(multcomp)
-
-#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
-\end{lstlisting}
-
-\hypertarget{working-with-missing-data-in-gams}{%
-\subsection{Working with Missing data in GAMs}\label{working-with-missing-data-in-gams}}
-
-This code chunk randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted.
-Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
-
-\begin{lstlisting}[language=R]
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-\end{lstlisting}
-
-\begin{lstlisting}
-## # A tibble: 10 x 3
-## # Groups:   Day, Group [10]
-##      Day Group         n
-##    <dbl> <fct>     <int>
-##  1     0 Control       2
-##  2     0 Treatment     4
-##  3     2 Control       7
-##  4     2 Treatment     4
-##  5     5 Control       2
-##  6     5 Treatment     8
-##  7     7 Control       5
-##  8     7 Treatment     8
-##  9    10 Control       4
-## 10    10 Treatment     5
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-\end{lstlisting}
-
-\includegraphics{Full_document_files/figure-latex/missing-data-Appendix-1}
-
-\begin{lstlisting}[language=R]
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
-
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'missing')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-mult_plot<-f2+inset_element(
-  f1, left = 0.01, 
-  bottom = 0.5, 
-  right = 0.5, 
-  top = 1.0)+
-  f3+f4+f6+
-   plot_annotation(tag_levels='A')&
-   ylim(c(-5,75)) &
-  theme(
-     text=element_text(size=18)
-     )&
-  thm
-
-mult_plot
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/sim-smooth-plot-Appendix-1}
-
-}
-
-\caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
-\end{figure}
-
-\end{document}

### 2021-05-27 Manuscripts/Manuscript by chapters -v2/06-Appendix.Rmd
Making proposed changes in the manuscript

--- 2021-05-24 Manuscripts/Manuscript by chapters -v1/06-Appendix.Rmd
@@
 set.seed(1)
-####################Section for calculations######################################
-################################################################################
+##########Section for calculations###########
 
@@
   geom_point(size=sz+1)+
-  scale_color_aaas()+
   labs(y='Simulated \n response')+
@@
 Here, the _p-value_ is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to ~12%.
+Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to $\approx$12%.
 
@@
 Moreover, the smooths (plotted via the `draw()` function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-From `summary()`, the deviance explained by the model is ~43%.
+From `summary()`, the deviance explained by the model is $\approx$43%.
 
@@
 By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group .
-From `gam.check`, the k-index is now at an acceptable value (~1.02), and `summary` now indicates that the model is able to capture 87% of the variance data.
+From `gam.check`, the k-index is now at an acceptable value ($\approx$1.02), and `summary` now indicates that the model is able to capture 87% of the variance data.
 

### 2021-05-27 Manuscripts/Manuscript by chapters -v2/Full_document.Rmd
Making proposed changes in the manuscript

--- 2021-05-24 Manuscripts/Manuscript by chapters -v1/Full_document.Rmd
@@
 In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures, thereby being an excellent choice to analyze non-linear longitudinal data.
-This paper summarizes the limitations of LMEMs and rm-ANOVA, presents the basic theory of GAMs, and demonstrates their implementation in $\textsf{R}$ via the package _mgcv_ using simulated data that follows longitudinal trends reported in biomedical literature.
+This paper summarizes the limitations of rm-ANOVA and LEMEs, presents the basic theory of GAMs, and demonstrates their implementation in $\textsf{R}$ via the package _mgcv_ using simulated data that follows longitudinal trends reported in biomedical literature.
 To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:________.

### 2021-05-28 Manuscripts/Manuscript by chapters -v2/03-GAM Theory.Rmd
Updated files

@@
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the "wigliness" of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
-The penalty reduces the "wigliness" of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
+To control the "wigliness" of the fit, each of the basis functions of Panel A is weighted by multiplying it by a coefficient according to the matrix of Panel B.
+The parameter estimates are penalized where the penalty reduces the "wigliness" of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
 
@@
 
-#heatmap of the penalization coefficient
+#heatmap of the  coefficients
 x_labels<-c("Intercept","1","2","3","4")
@@
 A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
+B: Matrix for basis function penalization.
+Each basis function is multiplied by a coefficient which can be positive or negative.
 The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
+C: Weighted basis functions.
+Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
 D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.
+The smoother (line) is the result of the sum of each penalized basis function at each time point, with simulated values for the group shown as points.
 

### 2021-05-28 Manuscripts/Manuscript by chapters -v2/04-Longitudinal analysis with GAMs.Rmd
Making proposed changes in the manuscript
Updated files

--- 2021-05-24 Manuscripts/Manuscript by chapters -v1/04-Longitudinal analysis with GAMs.Rmd
@@
 
-The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA.
+The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs.
 This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using $\textsf{R}$ and the package `mgcv`.
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors that appear in Figure 3(C) in [@vishwanath2009].
+The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors that appear in Figure 3, C in [@vishwanath2009].
 In the paper, diffuse reflectance spectroscopy was used to quantify $\mbox{StO}_2$ changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
@@
 
-An interaction model is typically the main interest in longitudinal biomedical data, as it takes into account the effect of treatment, time, and their combination.
+An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination.
 In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the _time_ effect to account for the repeated measures over time.
@@
 
-For the data in Figure \@ref(fig:sim-smooth-plot),A it is of interest to know the effect that each treatment has in $\mbox{StO}_2$.
-The model then needs to incorporate independent smooths for _Group_ and _Day_, respectively.
+For the data in Figure \@ref(fig:sim-smooth-plot), A the main effect of interest is how $\mbox{StO}_2$ changes over time for each treatment.
+To estimate this, the model needs to incorporate independent smooths for _Group_ and _Day_, respectively.
 The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric ( _Day_ ) and the other is a factor ( _Group_ ).
@@
 
-`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5,bs="gp"), method='REML',data=dat_sim`
+`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5,bs="gp"), method='REML',data=dat_sim)`
 
@@
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B).
-Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_[@gratia].
+Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_ [@gratia].
 A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].
@@
    plot_annotation(tag_levels='A')&
-   ylim(c(-5,75)) &
+   ylim(c(-7,75)) &
   theme(

### 2021-05-28 Manuscripts/Manuscript by chapters -v2/06-Appendix.Rmd
Updated files

@@
 
-#heatmap of the penalization coefficient
+#heatmap of the coefficients
 x_labels<-c("Intercept","1","2","3","4")

### 2021-05-28 Manuscripts/Manuscript by chapters -v1/02-Challenges.Rmd
Updated, added LMEM and rm-ANOVA models in text and in appendix, corrected cubic spline to thin plate spline, changed to default in gam models instead of gp smooths

@@
   ## repeated measures ANOVA in R
-#time and treatment interaction model, compound symmetry required by the model
+#time and treatment interaction model
+  fit_anova <- lm(y ~ time + treatment + time * treatment, data = dat)
+
+  #LMEM with compound symmetry
+
   fit_lme <- lme(y ~ treatment + time + treatment:time,
@@
   #add model predictions to the dataframe that has the simulated data
-  dat$y_pred <- predict(fit_lme)
+  dat$pred_anova <- predict(fit_anova)
+  dat$pred_lmem <- predict(fit_lme)
 
@@
     pred_dat = pred_dat,
-    fit_lme = fit_lme
+    fit_lme = fit_lme,
+    fit_anova=fit_anova
 
@@
 
-   #plot the model predictions
+   #plot the model predictions for rm-ANOVA
+  p2 <- ggplot(sim_dat$dat, 
+               aes(x = time, 
+                   y = y, 
+                   color = treatment)) +
+    geom_point()+
+    labs(y='response')+
+    geom_line(aes(y = predict(sim_dat$fit_anova), 
+                  group = subject, size = "Subjects")) +
+    geom_line(data = sim_dat$pred_dat, 
+              aes(y = predict(sim_dat$fit_anova, 
+                              level = 0, 
+                              newdata = sim_dat$pred_dat), 
+                  size = "Population")) +
+    guides(color = guide_legend(override.aes = list(size = 2)))+
+    scale_size_manual(name = "Predictions", 
+                      values=c("Subjects" = 0.5, "Population" = 3)) +
+    theme_classic() +
+    theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
+    thm
+
+   #plot the model predictions for LMEM
   p4 <- ggplot(sim_dat$dat, 
@@
 
-  return((p1+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
+  return((p1+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
 

### 2021-05-28 Manuscripts/Manuscript by chapters -v1/Full_document.tex
Updated, added LMEM and rm-ANOVA models in text and in appendix, corrected cubic spline to thin plate spline, changed to default in gam models instead of gp smooths

--- 2021-05-24 Manuscripts/Manuscript by chapters -v1/Full_document.tex
@@
 ## 
-## R-sq.(adj) =  0.794   Deviance explained = 80.6%
-## -REML = 374.51  Scale est. = 96.648    n = 100
-\end{lstlisting}
-
-The resulting model is model \passthrough{\lstinline!gam1!}, which is the model fitted in the main manuscript.
+## R-sq.(adj) =   0.82   Deviance explained = 83.5%
+## -REML =  368.7  Scale est. = 78.468    n = 100
+\end{lstlisting}
+
+The resulting model is\passthrough{\lstinline!m1!}, which is the model fitted in the main manuscript.
 By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group .
@@
 ##matrix that contains the basis functions evaluated at the points in pdat
-    xp <- predict(gam1, newdata = pdat, type = 'lpmatrix')
+    xp <- predict(m1, newdata = pdat, type = 'lpmatrix')
 

### 2021-05-28 Manuscripts/Manuscript by chapters -v2/02-Challenges.Rmd
Updated, added LMEM and rm-ANOVA models in text and in appendix, corrected cubic spline to thin plate spline, changed to default in gam models instead of gp smooths

@@
 To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme].
+An rm-ANOVA model (Equation \@ref(eq:linear-model)), and a LMEM  (Equation \@ref(eq:LMEM)) are fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme].
+
 Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A).
@@
 
-A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal appears in Figure \@ref(fig:linear-cases-Appendix) and Figure \@ref(fig:quadratic-cases-Appendix) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+A more comprehensive exploration  of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal appears in Figure \@ref(fig:linear-cases-Appendix) and Figure \@ref(fig:quadratic-cases-Appendix) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
 We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs.
@@
    ## repeated measures ANOVA in R
-#time and treatment interaction model, compound symmetry required by the model
+#time and treatment interaction model
+  fit_anova <- lm(y ~ time + treatment + time * treatment, data = dat)
+
+  #LMEM with compound symmetry
+
   fit_lme <- lme(y ~ treatment + time + treatment:time,
@@
 
-   #plot the model predictions
+   #plot the model predictions for rm-ANOVA
+  p2 <- ggplot(sim_dat$dat, 
+               aes(x = time, 
+                   y = y, 
+                   color = treatment)) +
+    geom_point(show.legend=FALSE)+
+    labs(y='response')+
+    geom_line(aes(y = predict(sim_dat$fit_anova), 
+                  group = subject, size = "Subjects"),show.legend = FALSE) +
+    geom_line(data = sim_dat$pred_dat, 
+              aes(y = predict(sim_dat$fit_anova, 
+                              level = 0, 
+                              newdata = sim_dat$pred_dat), 
+                  size = "Population"),
+              show.legend=FALSE) +
+    guides(color = guide_legend(override.aes = list(size = 2)))+
+    scale_size_manual(name = "Predictions", 
+                      values=c("Subjects" = 0.5, "Population" = 3)) +
+    theme_classic() +
+    theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
+    thm
+
+   #plot the model predictions for LMEM
   p4 <- ggplot(sim_dat$dat, 
@@
 
-(ref:l-q-response-caption) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.
+(ref:l-q-response-caption) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a LMEM and a rm-ANOVA model.
+A, D: Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
+B,E: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
+Points represent the original raw data.
+The rm-ANOVA model not only fails to pick the trend of the quadratic data but also assigns a global estimate that does not take between-subject variation.
+C, F: Estimates from the LMEM model in the linear and quadratic case.The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data in each group and grossly bias the initial estimates for each group.
 
@@
 
-The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B).
+The simulation shows that the fit produced by the LMEM and the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B, E).
 When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.
@@
 The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panel D in Figure \@ref(fig:l-q-response).
+Fitting an rm-ANOVA model \@ref(eq:linear-model) or a LMEM \@ref(eq:LMEM) to this data  produces the fit that appears in panels E and F in Figure \@ref(fig:l-q-response).
 
-A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\@ref(fig:l-q-response), D) indicates that the model is not capturing the changes within each group.
-Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
-Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D).
-If such a change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response.
-Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
+A comparison of the fitted mean response of the LMEM and the rm-ANOVA model to the simulated data in Figure ((\@ref(fig:l-q-response), E, F) indicates that the models are not capturing the changes within each group.
+Specifically, note that the fitted mean response of both models (panel E, F) show that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.
+The LMEM is only able to account for between-subject variation by providing different intercepts to each subject, but both models are not able to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panel D with panels E and F).
+If such a change has important physiological implications,both rm-ANOVA and LMEMs omit it from the fitted mean response.
+Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
 

### 2021-05-28 Manuscripts/Manuscript by chapters -v2/03-GAM Theory.Rmd
Updated, added LMEM and rm-ANOVA models in text and in appendix, corrected cubic spline to thin plate spline, changed to default in gam models instead of gp smooths

@@
 
-A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness [@wood2017;@simpson2018].
-Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
+Commonly used _basis functions_ are splines (cubic, thin plate regression among others).
+A cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data [@wood2017;@simpson2018].
+Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
 Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.

### 2021-05-28 Manuscripts/Manuscript by chapters -v2/04-Longitudinal analysis with GAMs.Rmd
Updated, added LMEM and rm-ANOVA models in text and in appendix, corrected cubic spline to thin plate spline, changed to default in gam models instead of gp smooths

@@
 
-`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5,bs="gp"), method='REML',data=dat_sim)`
+`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim)`
 
 This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
-The smooth is constructed using gaussian process smooths, which is indicated by `bs="gp"`.
-These splines are used to model temporal trends and might be particularly suited for long-term studies where the correlation between measurements changes as a function of the time intervals [@simpson2018].
+The smooth is constructed by default using thin plate regression splines.
+Other splines can be used if desired, including gaussian process smooths [@simpson2018].
 The parametric term _Group_ is added to quantify differences in the effect of treatment between groups, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
@@
 
-gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
+gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
             method='REML',

### 2021-05-28 Manuscripts/Manuscript by chapters -v2/06-Appendix.Rmd
Updated, added LMEM and rm-ANOVA models in text and in appendix, corrected cubic spline to thin plate spline, changed to default in gam models instead of gp smooths

@@
 
-#labelling y and errors  
+#labeling y and errors  
   dimnames(y) <- list(time = x, 
@@
 
-  ## repeated measures ANOVA in R
-#time and treatment interaction model, compound symmetry required by the model
+  ## repeated measures ANOVA
+
+  fit_anova <- lm(y ~ time + treatment + time * treatment, data = dat)
+
+#LMEM: time and treatment interaction model, compound symmetry 
   fit_lme <- lme(y ~ treatment + time + treatment:time,
@@
     pred_dat = pred_dat,
+    fit_anova=fit_anova,
     fit_lme = fit_lme
-
   ))
@@
   ## Plot the simulated data (scatterplot)
+
   p1 <- sim_dat$dat %>%
@@
 
-   #plot the model predictions
+   #plot the model predictions for rm-ANOVA
   p4 <- ggplot(sim_dat$dat, 
+               aes(x = time, 
+                   y = y, 
+                   color = treatment)) +
+    geom_point(show.legend=FALSE)+
+    labs(y='response')+
+    geom_line(aes(y = predict(sim_dat$fit_anova), 
+                  group = subject, size = "Subjects"),show.legend = FALSE) +
+    geom_line(data = sim_dat$pred_dat, 
+              aes(y = predict(sim_dat$fit_anova, 
+                              level = 0, 
+                              newdata = sim_dat$pred_dat), 
+                  size = "Population"),
+              show.legend=FALSE) +
+    guides(color = guide_legend(override.aes = list(size = 2)))+
+    scale_size_manual(name = "Predictions", 
+                      values=c("Subjects" = 0.5, "Population" = 3)) +
+    theme_classic() +
+    theme(plot.title = element_text(size = 30, 
+                                  face = "bold"),
+        text=element_text(size=30))+
+    thm
+
+   #plot the LMEM predictions
+  p5 <- ggplot(sim_dat$dat, 
                aes(x = time, 
@@
 
-  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
+  return((p1+p3+p2+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
 
@@
 
-(ref:linear-cases-appendix) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data.
+(ref:linear-cases-appendix) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
+A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
+B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
+C,H: Simulated data with thin lines representing individual trajectories.
+D,I: Estimations from the rm-ANOVA model for the mean group response.
+E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
+In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
+Both rm-ANOVA and the LMEM are able to capture the trend of the data.
 
@@
 
-(ref:quadratic-caption-appendix) Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
+(ref:quadratic-caption-appendix) Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
+A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
+B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
+C,H: Simulated data with thin lines representing individual trajectories.
+D,I: Estimations from the rm-ANOVA model for the mean group response.
+E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
+In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
+Both rm-ANOVA and the LMEM are not able to capture the changes in each group over time.
 
@@
 The model will use 5 knots (`k=5`) for the smooth.
-And that the smooth is constructed using gaussian process basis (`bs="gp"`).
+The smooth is constructed by default using thin plate regression splines.
 The smoothing parameter estimation method used is the restricted maximum likelihood (`REML`).
@@
 
-gam_00<-gam(StO2_sim ~ s(Day, k = 5,bs="gp"),
+gam_00<-gam(StO2_sim ~ s(Day, k = 5),
             method='REML',
@@
 
-gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5,bs="gp"),
+gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5),
             method='REML',
@@
 
-m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
+m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
             method='REML',
@@
 
-(ref:pairwise-comp-caption) Pairwise comparisons for smooth terms.
+(ref:pairwise-caption-both) Pairwise comparisons for smooth terms.
 A: Pairwise comparisons for the full dataset.
@@
 
-```{r,plot-pairwise-comp-Appendix,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-comp-caption)'}
+```{r,plot-pairwise-comp-Appendix,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-caption-both)'}
 

### 2021-05-28 Manuscripts/Manuscript by chapters -v2/Full_document.Rmd
Updated, added LMEM and rm-ANOVA models in text and in appendix, corrected cubic spline to thin plate spline, changed to default in gam models instead of gp smooths

@@
 output:  
+  bookdown::html_document2:
+    css: "style.css" #style for the HTML document
+  bookdown::word_document2:
+    fig_caption: yes #figure caption
+    keep_md: yes
   bookdown::pdf_document2:
@@
       in_header: preamble.sty #additional LaTeX formatting
-  bookdown::word_document2:
-    fig_caption: yes #figure caption
-    keep_md: yes
-  bookdown::html_document2:
-    css: "style.css" #style for the HTML document
 csl: elsevier-with-titles.csl #style for references

### 2021-05-31 Manuscripts/Manuscript by chapters -v2/05-Discussion.Rmd
Added discussion

# Discussion

Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data. We have visually demonstrated in Section \@ref(simulation) that rm-ANOVA models and LMEMs yield biased estimates when they are used to fit non-linear data, in what is known as "Type III" error [@dennis2019]. This is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model. Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) [@lang2015] but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of simply reporting an $R^2$ value.

In this paper we have introduced GAMs as a method to analyze non-linear longitudinal data.  
Reproducibility is still an issue in biomedical research [@begley2015] [@weissgerber2018], and

### 2021-05-31 Manuscripts/Manuscript by chapters -v2/Full_document.Rmd
Added discussion

@@
 
-In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, a _linear mixed model_ (LMEM).
-Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
-In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures, thereby being an excellent choice to analyze non-linear longitudinal data.
-This paper summarizes the limitations of rm-ANOVA and LEMEs, presents the basic theory of GAMs, and demonstrates their implementation in $\textsf{R}$ via the package _mgcv_ using simulated data that follows longitudinal trends reported in biomedical literature.
-To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:________.
+In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs).
+Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
+In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
+This paper summarizes the limitations of rm-ANOVA and LMEMs, presents the basic theory of GAMs, and uses simulated data that follows trends reported in the biomedical literature to demonstrate their implementation in $\textsf{R}$ via the package _mgcv_.
+To make this work reproducible, the code and data used in this paper are available at:________.
 
@@
 
-```{r child='05-Conclusion.Rmd'}
+```{r child='05-Discussion.Rmd'}
+```
+
+\FloatBarrier
+
+```{r child='06-Conclusion.Rmd'}
 ```
@@
 
-```{r child='06-Appendix.Rmd'}
+```{r child='07-Appendix.Rmd'}
 ```

### 2021-06-01 Manuscripts/Manuscript by chapters -v2/05-Discussion.Rmd
Working on discussion

@@
 
-Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data.
-We have visually demonstrated in Section \@ref(simulation) that rm-ANOVA models and LMEMs yield biased estimates when they are used to fit non-linear data, in what is known as "Type III" error [@dennis2019].
+Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
+Although LMEMs have started to replace rm-ANOVA models to analyze biomedical data, we have visually demonstrated in Section \@ref(simulation) that both methods yield biased estimates when they are used to fit non-linear data, in what is known as "Type III" error [@dennis2019].
 This is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
@@
 
-In this paper we have introduced GAMs as a method to analyze non-linear longitudinal data.  
-Reproducibility is still an issue in biomedical research [@begley2015] [@weissgerber2018], and
+In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) [@hefley2017; @pedersen2019] they are not routinely used in biomedical research despite a publication from Trevor and Hastie that demonstrated their use in medical research [@hastie1995].
+This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \@ref(GAM-theory) is to demonstrate that at its core it is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+
+However, from a practical standpoint is equally important to demonstrate how GAMs are implemented.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature [@vishwanath2009] using $\textsf{R}$ and the package _mgcv_[@wood2017] in Section \@ref(longitudinal-GAMs), while a basic workflow for model selection is in the Appendix.
+One of the features of GAMs is that they go beyond a mere _p-value_ to estimate differences between groups and can provide a time-based estimate of shifts in the response that can be directly tied to biological values.
+Because the analysis of GAMs is more computationally intensive when compared to traditional linear models, $\textsf{R}$ is the software of choice because it not only provides a fully developed environment to fit thse models, but eases simulation and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
+
+In this regard, reproducibility is still an issue in biomedical research [@begley2015;@weissgerber2018], but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact soon rather than later this field.
+Researchers need to plan on how they will make their data, code, and any other materials open and accesible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
+In this paper, we have shared all the data and code in order to make our work reproducible, and we hope that this will encourage other researchers to

### 2021-06-01 Manuscripts/Manuscript by chapters -v2/03-GAM Theory.Rmd
Working on discussion
Updated with Discussion and conclusions

@@
-# GAMs as a special case of Generalized Linear Models
+# GAMs as a special case of Generalized Linear Models{#GAM-theory}
 
@@
 ```{r,basis-functions-plot, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
+set.seed(1)
 n_time = 6

### 2021-06-01 Manuscripts/Manuscript by chapters -v2/04-Longitudinal analysis with GAMs.Rmd
Updated with Discussion and conclusions

@@
 ```{r,sim-tumor-response,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-
+set.seed(1)
 dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
@@
 
-gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
+m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
             method='REML',
@@
 gam_predict<-gam_predict%>%
-    mutate(fit = predict(gam1,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(gam1, gam_predict,se.fit = TRUE,type='response')$se.fit)
+    mutate(fit = predict(m1,gam_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(m1, gam_predict,se.fit = TRUE,type='response')$se.fit)
 
@@
 Consider the  simulated $\mbox{StO}_2$ values from Figure (\@ref(fig:sim-smooth-plot), B).
-If 40% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but the "Treatment" smooth takes an overall more linear profile (\@ref(fig:sim-smooth-plot), D).
+If 40% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but it can be seen that the smooths overlap during the first 3 days because with less data points, the trend is less pronounced than in the full dataset (\@ref(fig:sim-smooth-plot), D).
 Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.
@@
     xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    #Find columns in xp where the name contains "Control"
+    #Find columns in xp where the name contains "Control" and "Treatment"
     col1 <- grepl(g1, colnames(xp))
-    #Find columns in xp where the name contains 'Treatment'
     col2 <- grepl(g2, colnames(xp))
-    #r1 <- newdata[[var]] == f1
-    #r2 <- newdata[[var]] == f2
+    #Find rows in xp that correspond to each treatment
     row1 <- with(newdata, Group == g1)
@@
     ## zero out the parametric cols
-    X[, !grepl('^s\\(', colnames(xp))] <- 0
+    #X[, !grepl('^s\\(', colnames(xp))] <- 0
     dif <- X %*% coef(model)
@@
     geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
-    geom_ribbon(data=comp_StO2_missing%>%
-                    filter(lower>0),
-                aes(ymin =0, ymax =lower),
-                alpha = 0.5,
-                fill='#30123BFF') +
+    # geom_ribbon(data=comp_StO2_missing%>%
+    #                 filter(lower>0),
+    #             aes(ymin =0, ymax =lower),
+    #             alpha = 0.5,
+    #             fill='#30123BFF') +
     geom_ribbon(data=comp_StO2_missing %>%
@@
 
-Consider the calculation of pairwise differences for the smooths in \@ref(fig:sim-smooth-plot) B and D.
-Figure \@ref(fig:plot-pairwise-comp), shows the comparison between each treatment group.Here, the effect of the "Control" group is compared to that of the "Treatment" group.
-The shaded region under the curve highlights the interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and 4 indicates that through that time, the "Control" group has higher $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by day 5 it is the "Treatment" group the one that has greater contribution.
-This would suggest that the effect of chemotherapy is significant after day 5 for the model used; which is noticeable in the data with missing observations albeit the missing data model has a wider confidence interval and is not able to pick the change between days 7 and 10 that the full dataset model is able to detect.
+Consider the calculation of pairwise differences for the smooths in Figure \@ref(fig:sim-smooth-plot), B and D.
+Figure \@ref(fig:plot-pairwise-comp), shows the comparison between each treatment group for the full and missing datasets.
+Here, the  "Control" group is used as the reference to which "Treatment" group is being compared.
+Of notice, the pairwise comparison has been set on the response scale (see Appendix for code details), because otherwise the comparison appears shifted and is not intuitively easy to relate to the original data.
+
+With this correction in mind, the shaded region under or above the confidence interval (that does not cover 0) indicates the interval where each group has a higher effect than the other.
+Notice that the shaded region between days 0 and 3 for the full dataset indicates that through that time, the "Control" group has higher $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by day 4 it is the "Treatment" group the one that has greater $\mbox{StO}_2$.
+This would suggest that the effect of chemotherapy in the "Treatment" group becomes significant after day 4 for the model used.
+Moreover, notice that although there is no actual measurement at day 4, the model is capable of providing an estimate of when the shift in $\mbox{StO}_2$ occurs.
+
+On the data with missing observations (Figure \@ref(fig:sim-smooth-plot), D), the confidence intervals of the smooths overlap between days 0 and 3.
+Consequently, the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp), B) shows that there is not a significant difference between the groups during that period, but is still able to pick the change on day 4 as the full dataset smooth pairwise comparison.
+
+In a sense, the pairwise smooth comparison is more informative than a _post-hoc_ _p-value_.
+For biomedical studies, it is able to provide an estimate of _when_ a biological process becomes significant.
+This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change is expected.

### 2021-06-01 Manuscripts/Manuscript by chapters -v2/05-Discussion.Rmd
Updated with Discussion and conclusions

--- 2021-05-31 Manuscripts/Manuscript by chapters -v2/05-Discussion.Rmd
@@
 
-Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data.
-We have visually demonstrated in Section \@ref(simulation) that rm-ANOVA models and LMEMs yield biased estimates when they are used to fit non-linear data, in what is known as "Type III" error [@dennis2019].
-This is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) [@lang2015] but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of simply reporting an $R^2$ value.
+Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
+Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \@ref(simulation).
+This "model misspecification" error, also is known as"" "Type III" error [@dennis2019] is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) [@lang2015] but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a $R^2$ value.
 
-In this paper we have introduced GAMs as a method to analyze non-linear longitudinal data.  
-Reproducibility is still an issue in biomedical research [@begley2015] [@weissgerber2018], and
+In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) [@hefley2017; @pedersen2019] they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research [@hastie1995].
+This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \@ref(GAM-theory) is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+
+However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature [@vishwanath2009] using $\textsf{R}$ and the package _mgcv_[@wood2017] in Section \@ref(longitudinal-GAMs), while a basic workflow for model selection is in the Appendix.
+One of the features of GAMs is that they go beyond a mere _p-value_ to indicate differences between groups, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure  \@ref(fig:plot-pairwise-comp) indicate.
+The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( $\approx$ day 4 in Figure \@ref(fig:plot-pairwise-comp) A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
+
+We have used $\textsf{R}$ as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
+In this regard, reproducibility is still an issue in biomedical research [@begley2015;@weissgerber2018], but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact soon rather than later this field.
+Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
+We have made all the data and code used in this paper accessible, and we hope that this will encourage other researchers to do the same with future projects.

### 2021-06-01 Manuscripts/Manuscript by chapters -v2/06-Conclusion.Rmd
Added discussion
Updated with Discussion and conclusions

--- 2021-05-27 Manuscripts/Manuscript by chapters -v2/05-Conclusion.Rmd
@@
 
-The traditional methods to analyze biomedical longitudinal data (rm-ANOVA, LMEMs) are restricted to cases when the data follows a linear pattern.
-When the trend of the data is non-linear (as it is frequently the case in biomedical research), the use of rm-ANOVA and LMEMs leads to unreliable inference and biased estimates.
-GAMs present a suitable alternative to analyze non-linear longitudinal biomedical data, as they overcome the limitations of linearity, compound symmetry, and missing observations imposed by traditional methods.
-By  presenting in a clear and visual manner the limitations of rm-ANOVA and LMEMs, and the implementation of GAMs using simulated data that follows previously reported trends in the literature, we aim at encouraging biomedical researchers to consider different statistical tools to analyze longitudinal data.
-Finally, by providing the data and code used in this paper we hope to address the need of creating and sharing reproducible work in biomedical research.
+We have presented GAMs as a method to analyze longitudinal biomedical data.
+Future directions of this work will include simulation-based estimations of statistical power using GAMs, as well as demonstrating the prediction capabilities of these models using large datasets.
+By making the data and code used in this paper accessible, we hope to address the need of creating and sharing reproducible work in biomedical research.
+
+# Acknowledgements
+
+This work was supported by the National Science Foundation Career Award (CBET 1751554) and the Arkansas Biosciences Institute.

### 2021-06-01 Manuscripts/Manuscript by chapters -v2/07-Appendix.Rmd
Updated with Discussion and conclusions

--- 2021-05-28 Manuscripts/Manuscript by chapters -v2/06-Appendix.Rmd
@@
 ```{r, responses-Appendix,include=TRUE,message=FALSE,echo=TRUE}
+
 set.seed(1)
@@
 #the response
+set.seed(1)
 n_time = 6
@@
 b_plot
-
 ```
@@
 ```{r,sim-tumor-response-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
-
+set.seed(1)
 #Dataframe that contains the original reported trends
@@
 #subject = factor(paste(subject, treatment, sep = "-")))
-
 n <- 10 #number of observations
@@
 Here, the _p-value_ is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to $\approx$12%.
+Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to $\approx$ 12%.
 
@@
 Moreover, the smooths (plotted via the `draw()` function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-From `summary()`, the deviance explained by the model is $\approx$43%.
+From `summary()`, the deviance explained by the model is $\approx$ 43%.
 
@@
 By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group .
-From `gam.check`, the k-index is now at an acceptable value ($\approx$1.02), and `summary` now indicates that the model is able to capture 87% of the variance data.
+From `gam.check`, the k-index is now at an acceptable value ($\approx$ 1.02), and `summary` now indicates that the model is able to capture 87% of the variance data.
 
@@
 
-AIC(gam_00,gam_01,gam1)
+AIC(gam_00,gam_01,m1)
 
@@
 In this case, the "design matrix" is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
-Briefly, the "design matrix" (also known as the "Xp matrix") from the selected model (`gam1`) is used to calculate a 95% confidence interval of the difference between the smooth terms for each group.
+Briefly, the "design matrix" (also known as the "Xp matrix") from the selected model (`m1`) is used to calculate a 95% confidence interval of the difference between the smooth terms for each group.
 This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
+**All pairwise comparisons in this paper have been centered at the response scale to ease interpretation **.
 
@@
 ##Pairwise comparisons
+pdat <- expand.grid(Day = seq(0, 10, length = 400),
+                    Group = c('Control', 'Treatment'))
 
 ##matrix that contains the basis functions evaluated at the points in pdat
-    xp <- predict(gam1, newdata = pdat, type = 'lpmatrix')
+    xp <- predict(m1, newdata = pdat, type = 'lpmatrix')
 
@@
     ## zero out the parametric cols, those that do not contain in the characters 's('
-    X[, !grepl('^s\\(', colnames(xp))] <- 0
+    #X[, !grepl('^s\\(', colnames(xp))] <- 0
 
@@
     #dimensions (n,1). The resulting matrix has dimensions (p,1)
-    dif <- X %*% coef(gam1)
+    dif <- X %*% coef(m1)
 
@@
     #of the model
-    se <- sqrt(rowSums((X %*% vcov(gam1, unconditional = FALSE)) * X))
-    crit <- qt(0.05/2, df.residual(gam1), lower.tail = FALSE)
+    se <- sqrt(rowSums((X %*% vcov(m1, unconditional = FALSE)) * X))
+    crit <- qt(0.05/2, df.residual(m1), lower.tail = FALSE)
     #upper  limits
@@
 
-(ref:pairwise-comp-caption) Smooth pairwise comparisons for model `gam1` using a 95% confidence interval for the difference between smooths.
+(ref:pairwise-comp-caption) Smooth pairwise comparisons for model `m1` using a 95% confidence interval for the difference between smooths.
 
@@
 
+Keep in mind that this function **does not** center the pairwise comparison at the response scale, so it has to be shifted in order to be compared to the raw data.
+
 # GAM and Linear model plots and Missing data
 
-This section covers the code used to generate Figure \@ref(fig:sim-smooth-plot), where the simulated data, fit of the "final" GAM (`gam1`), linear model  and GAM on data with missing observations are presented.
+This section covers the code used to generate Figure \@ref(fig:sim-smooth-plot), where the simulated data, fit of the "final" GAM (`m1`), linear model  and GAM on data with missing observations are presented.
 Note that panel A in Figure \@ref(fig:sim-smooth-plot) and the inlet are generated in the code chunk where the data is simulated in Section \@ref(tumor-data-simulation), and are called later to build the figure.
@@
 This code chunk creates panels B and D in Figure \@ref(fig:sim-smooth-plot).
-Note that this code uses the final GAM from the previous section (`gam1`), so the simulated data and the model should be generated before running this section.
+Note that this code uses the final GAM from the previous section (`m1`), so the simulated data and the model should be generated before running this section.
 
@@
 #correspond to the row numbers to be randomly erased from the original dataset
-set.seed(1)
+
 missing <- sample(1:100, 40)

### 2021-06-01 Manuscripts/Manuscript by chapters -v2/Full_document.Rmd
Updated with Discussion and conclusions

@@
 output:  
-  bookdown::html_document2:
-    css: "style.css" #style for the HTML document
   bookdown::word_document2:
@@
     keep_md: yes
+  bookdown::html_document2:
+    css: "style.css" #style for the HTML document
   bookdown::pdf_document2:
     #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
-    pandoc_args: --listings #calls the listings package to fit code in the page margins
+    pandoc_args: --listings #calls the listings package to fit code within the page margins
     keep_tex: yes #keep LaTeX file for submission
@@
 ```{r child='07-Appendix.Rmd'}
+
 ```

### 2021-06-01 Manuscripts/Manuscript by chapters -v2/Full_document.tex
Updated with Discussion and conclusions

--- 2021-05-28 Manuscripts/Manuscript by chapters -v1/Full_document.tex
@@
 
-In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, a \emph{linear mixed model} (LMEM).
-Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
-In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures, thereby being an excellent choice to analyze non-linear longitudinal data.
-This paper summarizes the limitations of LMEMs and rm-ANOVA, presents the basic theory of GAMs, and demonstrates their implementation in \(\textsf{R}\) via the package \emph{mgcv} using simulated data that follows longitudinal trends reported in biomedical literature.
-To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:\_\_\_\_\_\_\_\_.
+In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs).
+Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
+In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
+This paper summarizes the limitations of rm-ANOVA and LMEMs, presents the basic theory of GAMs, and uses simulated data that follows trends reported in the biomedical literature to demonstrate their implementation in \(\textsf{R}\) via the package \emph{mgcv}.
+To make this work reproducible, the code and data used in this paper are available at:\_\_\_\_\_\_\_\_.
 
@@
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model is only able to accommodate linear trends that are far from adequately representing the biological phenomenon of interest.
-
-A \emph{post hoc} analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
+Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
+
+A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
 Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
@@
 
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information.
+During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
 The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
@@
 However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}.
-And even more importantly, LMEMs also expect a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}, making them unsuitable to analyze non-linear data.
+And even more importantly, LMEMs also assume a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}, making them unsuitable to analyze non-linear data.
 
@@
 In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
-Independent errors \(\varepsilon_{tij}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently normally distributed with mean zero and variance \(\sigma^2_\mu\)).
+Independent errors \(\varepsilon_{tij}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
 In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy'').
@@
 If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{47}{]}, and in the case of MNAR the missing values are dependent on their value.
+
 An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
 This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{48}{]}.
-
 In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}\protect\hyperlink{ref-west2014}{40}{]}.
@@
 
-This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and missing data.
-In particular, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study.
-However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data.
-To better convey the issues of linearity and correlation in linear models fitted to non-linear data, simulation is used in the next section.
-
 \hypertarget{simulation}{%
@@
 
-To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}) is fitted to each group, using \(\textsf{R}\){[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme}{[}\protect\hyperlink{ref-nlme}{49}{]}.
+To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\){[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme}{[}\protect\hyperlink{ref-nlme}{49}{]}.
+
 Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}, A).
@@
 Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D).
-A more comprehensive exploration of the fit of rm-ANOVA for linear and non-linear longitudinal appears in Figure \ref{fig:linear-cases-Appendix} and Figure \ref{fig:quadratic-cases-Appendix} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+
+A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal appears in Figure \ref{fig:linear-cases-Appendix} and Figure \ref{fig:quadratic-cases-Appendix} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs.
+Of notice, Section \ref{longitudinal-GAMs} uses simulated data that does follow reported trends in the biomedical literature.
 
@@
 
-\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.}\label{fig:l-q-response}
+\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a LMEM and a rm-ANOVA model.
+A, D: Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
+B,E: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
+Points represent the original raw data.
+The rm-ANOVA model not only fails to pick the trend of the quadratic data but also assigns a global estimate that does not take between-subject variation.
+C, F: Estimates from the LMEM model in the linear and quadratic case.The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data in each group and grossly bias the initial estimates for each group.}\label{fig:l-q-response}
 \end{figure}
 
-The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response},B).
+The simulation shows that the fit produced by the LMEM and the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response},B, E).
 When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.
@@
 The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model \eqref{eq:linear-model} to this data produces the fit that appears in panel D in Figure \ref{fig:l-q-response}.
-
-A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\ref{fig:l-q-response}, D) indicates that the model is not capturing the changes within each group in a good way.
-Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model.
-Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare panels C and D).
-If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response.
-Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
+Fitting an rm-ANOVA model \eqref{eq:linear-model} or a LMEM \eqref{eq:LMEM} to this data produces the fit that appears in panels E and F in Figure \ref{fig:l-q-response}.
+
+A comparison of the fitted mean response of the LMEM and the rm-ANOVA model to the simulated data in Figure ((\ref{fig:l-q-response}, E, F) indicates that the models are not capturing the changes within each group.
+Specifically, note that the fitted mean response of both models (panel E, F) show that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.
+The LMEM is only able to account for between-subject variation by providing different intercepts to each subject, but both models are not able to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare panel D with panels E and F).
+If such a change has important physiological implications,both rm-ANOVA and LMEMs omit it from the fitted mean response.
+Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
 
@@
 
-\hypertarget{gams-as-a-special-case-of-generalized-linear-models}{%
-\section{GAMs as a special case of Generalized Linear Models}\label{gams-as-a-special-case-of-generalized-linear-models}}
+\hypertarget{GAM-theory}{%
+\section{GAMs as a special case of Generalized Linear Models}\label{GAM-theory}}
 
@@
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the function \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.
+Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.
 
@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}\protect\hyperlink{ref-beck1998}{36}{]}.
+One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
 In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic, and causes bias at the endpoints of the time period.
 
-The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis functions} expansions of the covariates and by estimating random coefficients for these basis functions.
-A \emph{basis} is a set of functions that spans the space where the smooths that approximate \(f(x_t \mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
+The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
+A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t \mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
 For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
@@
 
-A commonly used \emph{basis function} is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
-Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
+Commonly used \emph{basis functions} are splines (cubic, thin plate regression among others).
+A cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
+Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
 Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
@@
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the ``wigliness'' of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B.
-The penalty reduces the ``wigliness'' of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
+To control the ``wigliness'' of the fit, each of the basis functions of Panel A is weighted by multiplying it by a coefficient according to the matrix of Panel B.
+The parameter estimates are penalized where the penalty reduces the ``wigliness'' of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
 
@@
 A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
+B: Matrix for basis function penalization.
+Each basis function is multiplied by a coefficient which can be positive or negative.
 The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
+C: Weighted basis functions.
+Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
 D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}\label{fig:basis-plot}
+The smoother (line) is the result of the sum of each penalized basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
 \end{figure}
@@
 
-The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA.
+The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs.
 This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}.
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3(C) in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3, C in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
 In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
@@
 
-An interaction model is typically the main interest in longitudinal biomedical data, as it takes into account the effect of treatment, time, and their combination.
+An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination.
 In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time.
@@
 
-For the data in Figure \ref{fig:sim-smooth-plot},A it is of interest to know the effect that each treatment has in \(\mbox{StO}_2\).
-The model then needs to incorporate independent smooths for \emph{Group} and \emph{Day}, respectively.
+For the data in Figure \ref{fig:sim-smooth-plot}, A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
+To estimate this, the model needs to incorporate independent smooths for \emph{Group} and \emph{Day}, respectively.
 The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric ( \emph{Day} ) and the other is a factor ( \emph{Group} ).
@@
 
-\passthrough{\lstinline!m1<-gam(StO2\_sim\~Group+s(Day,by=Group,k=5,bs="gp"), method='REML',data=dat\_sim)!}
+\passthrough{\lstinline!m1<-gam(StO2\_sim\~Group+s(Day,by=Group,k=5), method='REML',data=dat\_sim)!}
 
 This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths for \emph{Group} and \emph{Day} (the parenthesis preceded by \emph{s}) using 5 knots.
-The smooth is constructed using gaussian process smooths, which is indicated by \passthrough{\lstinline!bs="gp"!}.
-These splines are used to model temporal trends and might be particularly suited for long-term studies where the correlation between measurements changes as a function of the time intervals {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
+The smooth is constructed by default using thin plate regression splines.
+Other splines can be used if desired, including gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
 The parametric term \emph{Group} is added to quantify differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}.
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot},B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}{[}\protect\hyperlink{ref-gratia}{54}{]}.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{54}{]}.
 A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{55}{]}.
@@
 Consider the simulated \(\mbox{StO}_2\) values from Figure (\ref{fig:sim-smooth-plot}, B).
-If 40\% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but the ``Treatment'' smooth takes an overall more linear profile (\ref{fig:sim-smooth-plot}, D).
+If 40\% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but it can be seen that the smooths overlap during the first 3 days because with less data points, the trend is less pronounced than in the full dataset (\ref{fig:sim-smooth-plot}, D).
 Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.
@@
 
-Consider the calculation of pairwise differences for the smooths in \ref{fig:sim-smooth-plot} B and D.
-Figure \ref{fig:plot-pairwise-comp}, shows the comparison between each treatment group.Here, the effect of the ``Control'' group is compared to that of the ``Treatment'' group.
-The shaded region under the curve highlights the interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and 4 indicates that through that time, the ``Control'' group has higher \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by day 5 it is the ``Treatment'' group the one that has greater contribution.
-This would suggest that the effect of chemotherapy is significant after day 5 for the model used; which is noticeable in the data with missing observations albeit the missing data model has a wider confidence interval and is not able to pick the change between days 7 and 10 that the full dataset model is able to detect.
+Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}, B and D.
+Figure \ref{fig:plot-pairwise-comp}, shows the comparison between each treatment group for the full and missing datasets.
+Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared.
+Of notice, the pairwise comparison has been set on the response scale (see Appendix for code details), because otherwise the comparison appears shifted and is not intuitively easy to relate to the original data.
+
+With this correction in mind, the shaded region under or above the confidence interval (that does not cover 0) indicates the interval where each group has a higher effect than the other.
+Notice that the shaded region between days 0 and 3 for the full dataset indicates that through that time, the ``Control'' group has higher \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by day 4 it is the ``Treatment'' group the one that has greater \(\mbox{StO}_2\).
+This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 4 for the model used.
+Moreover, notice that although there is no actual measurement at day 4, the model is capable of providing an estimate of when the shift in \(\mbox{StO}_2\) occurs.
+
+On the data with missing observations (Figure \ref{fig:sim-smooth-plot}, D), the confidence intervals of the smooths overlap between days 0 and 3.
+Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}, B) shows that there is not a significant difference between the groups during that period, but is still able to pick the change on day 4 as the full dataset smooth pairwise comparison.
+
+In a sense, the pairwise smooth comparison is more informative than a \emph{post-hoc} \emph{p-value}.
+For biomedical studies, it is able to provide an estimate of \emph{when} a biological process becomes significant.
+This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change is expected.
+
+\FloatBarrier
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
+Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}.
+This ``model misspecification'' error, also is known as"" ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{56}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+
+In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{57}{]}.
+This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+
+However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} using \(\textsf{R}\) and the package \emph{mgcv}{[}\protect\hyperlink{ref-wood2017}{37}{]} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the Appendix.
+One of the features of GAMs is that they go beyond a mere \emph{p-value} to indicate differences between groups, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
+The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 4 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
+
+We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
+In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{58},\protect\hyperlink{ref-weissgerber2018}{59}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact soon rather than later this field.
+Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
+We have made all the data and code used in this paper accessible, and we hope that this will encourage other researchers to do the same with future projects.
 
@@
 
-The traditional methods to analyze biomedical longitudinal data (rm-ANOVA, LMEMs) are restricted to cases when the data follows a linear pattern.
-When the trend of the data is non-linear (as it is frequently the case in biomedical research), the use of rm-ANOVA and LMEMs leads to unreliable inference and biased estimates.
-GAMs present a suitable alternative to analyze non-linear longitudinal biomedical data, as they overcome the limitations of linearity, compound symmetry, and missing observations imposed by traditional methods.
-By presenting in a clear and visual manner the limitations of rm-ANOVA and LMEMs, and the implementation of GAMs using simulated data that follows previously reported trends in the literature, we aim at encouraging biomedical researchers to consider different statistical tools to analyze longitudinal data.
-Finally, by providing the data and code used in this paper we hope to address the need of creating and sharing reproducible work in biomedical research.
+We have presented GAMs as a method to analyze longitudinal biomedical data.
+Future directions of this work will include simulation-based estimations of statistical power using GAMs, as well as demonstrating the prediction capabilities of these models using large datasets.
+By making the data and code used in this paper accessible, we hope to address the need of creating and sharing reproducible work in biomedical research.
+
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+This work was supported by the National Science Foundation Career Award (CBET 1751554) and the Arkansas Biosciences Institute.
 
@@
 Lecorps, H.M.
-Colhoun, {Antidrug Antibodies in Patients Treated with Alirocumab}, {New England Journal of Medicine}.
+Colhoun, {Antidrug antibodies in atients treated with alirocumab}, {New England Journal of Medicine}.
 {376} (2017) 1589--1590.
@@
 Taper, S.R.
-Lele, {Errors in Statistical Inference Under Model Misspecification: Evidence, Hypothesis Testing, and AIC}, {Frontiers in Ecology and Evolution}.
+Lele, {Errors in statistical inference under model misspecification: evidence, hypothesis testing, and AIC}, {Frontiers in Ecology and Evolution}.
 {7} (2019).
@@
 Vowler, G.B.
-Drummond, {The fickle P value generates irreproducible results}, {Nature Methods}.
+Drummond, {The fickle \(p\) value generates irreproducible results}, {Nature Methods}.
 {12} (2015) 179--185.
@@
 \CSLLeftMargin{{[}21{]} }
-\CSLRightInline{H. Abdi, {Holm's Sequential Bonferroni Procedure}, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.}
+\CSLRightInline{H. Abdi, {Holm's sequential Bonferroni procedure}, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.}
 
@@
 Hill, M.
-Yajima, {Why We (Usually) Don't Have to Worry About Multiple Comparisons}, {Journal of Research on Educational Effectiveness}.
+Yajima, {Why we (usually) don't have to worry about multiple comparisons}, {Journal of Research on Educational Effectiveness}.
 {5} (2012) 189--211.
@@
 Huynh, L.S.
-Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics.
+Feldt, {Estimation of the Box correction for degrees of freedom from sample data in randomized block and split-Plot designs}, Journal of Educational Statistics.
 1 (1976) 69--82.
@@
 Haverkamp, A.
-Beauducel, {Violation of the Sphericity Assumption and Its Effect on Type-I Error Rates in Repeated Measures ANOVA and Multi-Level Linear Models (MLM)}, {Frontiers in Psychology}.
+Beauducel, {Violation of the sphericity assumption and its effect on type-I error rates in repeated measures ANOVA and multi-level linear models (MLM)}, {Frontiers in Psychology}.
 {8} (2017).
@@
 Turner, G.L.
-Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, UK, Geochimica Et Cosmochimica Acta.
+Simpson, {An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to Scotland, UK}, Geochimica Et Cosmochimica Acta.
 82 (2012) 113--135.
@@
 \CSLLeftMargin{{[}37{]} }
-\CSLRightInline{S.N. Wood, Generalized additive models: An introduction with {}, second edition, CRC Press LLC, Philadelphia, PA, 2017.}
+\CSLRightInline{S.N. Wood, {Generalized additive models: An introduction with R, Second Edition}, CRC Press LLC, Philadelphia, PA, 2017.}
 
@@
 Pya, B.
-Saefken, {Smoothing Parameter and Model Selection for General Smooth Models}, {Journal of the American Statistical Association}.
+Saefken, Smoothing parameter and model selection for general smooth models, {Journal of the American Statistical Association}.
 {111} (2016) 1548--1563.
@@
 Geisser, S.W.
-Greenhouse, {An Extension of Box's Results on the Use of the \(F\) Distribution in Multivariate Analysis}, The Annals of Mathematical Statistics.
+Greenhouse, {An extension of Box's results on the use of the \(F\) distribution in multivariate analysis}, The Annals of Mathematical Statistics.
 29 (1958) 885--891.
@@
 Pieper, V.
-Hasselblad, {Can one assess whether missing data are missing at random in medical studies?}, {STATISTICAL METHODS IN MEDICAL RESEARCH}.
+Hasselblad, {Can one assess whether missing data are missing at random in medical studies?}, {Statistical Methods in Medical Research}.
 {15} (2006) 213--234.
@@
 Mazumdar, S.G.
-Memtsoudis, {Beyond Repeated-Measures Analysis of Variance Advanced Statistical Methods for the Analysis of Longitudinal Data in Anesthesia Research}, {Regional Anesthesia and Pain Medicine}.
+Memtsoudis, Beyond repeated-measures analysis of variance advanced statistical methods for the analysis of longitudinal data in anesthesia research, {Regional Anesthesia and Pain Medicine}.
 {37} (2012) 99--105.
@@
 Williams, M.B.
-Hooten, {The basis function approach for modeling autocorrelation in ecological data}, {ECOLOGY}.
+Hooten, {The basis function approach for modeling autocorrelation in ecological data}, {Ecology}.
 {98} (2017) 632--646.
@@
 \CSLLeftMargin{{[}55{]} }
-\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with r, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
+\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, {Semiparametric Regression with R}, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
+
+\leavevmode\hypertarget{ref-lang2015}{}%
+\CSLLeftMargin{{[}56{]} }
+\CSLRightInline{T.A.
+Lang, D.G.
+Altman, {Basic statistical reporting for articles published in Biomedical Journals: The ``Statistical Analyses and Methods in the Published Literature{''} or the SAMPL Guidelines}, {INTERNATIONAL JOURNAL OF NURSING STUDIES}.
+{52} (2015) 5--9.
+\url{https://doi.org/\%7B10.1016/j.ijnurstu.2014.09.006\%7D}.}
+
+\leavevmode\hypertarget{ref-hastie1995}{}%
+\CSLLeftMargin{{[}57{]} }
+\CSLRightInline{T.
+Hastie, R.
+Tibshirani, Generalized additive models for medical research, {Statistical Methods in Medical Research}.
+4 (1995) 187--196.
+\url{https://doi.org/10.1177/096228029500400302}.}
+
+\leavevmode\hypertarget{ref-begley2015}{}%
+\CSLLeftMargin{{[}58{]} }
+\CSLRightInline{C.G.
+Begley, J.P.A.
+Ioannidis, {Reproducibility in Science Improving the Standard for Basic and Preclinical Research}, {Circulation Research}.
+{116} (2015) 116--126.
+\url{https://doi.org/\%7B10.1161/CIRCRESAHA.114.303819\%7D}.}
+
+\leavevmode\hypertarget{ref-weissgerber2018}{}%
+\CSLLeftMargin{{[}59{]} }
+\CSLRightInline{T.L.
+Weissgerber, O.
+Garcia-Valencia, V.D.
+Garovic, N.M.
+Milic, S.J.
+Winham, {Meta-Research: Why we need to report more than 'Data were Analyzed by t-tests or ANOVA'}, Elife.
+7 (2018) e36163.
+\url{https://doi.org/10.7554/eLife.36163}.}
 
@@
 \begin{figure}[H]
-\includegraphics{Full_document_files/figure-latex/linear-cases-Appendix-1} \caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model.
-A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
-B,D: Estimations from the rm-ANOVA model for the mean group response (linear of quadratic).
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-The rm-ANOVA model does not pick the trend of the quadratic data.}\label{fig:linear-cases-Appendix}
+\includegraphics{Full_document_files/figure-latex/linear-cases-Appendix-1} \caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
+A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
+B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
+C,H: Simulated data with thin lines representing individual trajectories.
+D,I: Estimations from the rm-ANOVA model for the mean group response.
+E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
+In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
+Both rm-ANOVA and the LMEM are able to capture the trend of the data.}\label{fig:linear-cases-Appendix}
 \end{figure}
@@
 \begin{figure}[H]
-\includegraphics{Full_document_files/figure-latex/quadratic-cases-Appendix-1} \caption{Simulated quadratic responses from two groups with a rm-ANOVA model fitted.
-A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data.
-B,F: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline.
-D,H: Estimations from the rm-ANOVA model for the mean group response.
-Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.}\label{fig:quadratic-cases-Appendix}
+\includegraphics{Full_document_files/figure-latex/quadratic-cases-Appendix-1} \caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
+A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
+B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
+C,H: Simulated data with thin lines representing individual trajectories.
+D,I: Estimations from the rm-ANOVA model for the mean group response.
+E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
+In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
+Both rm-ANOVA and the LMEM are not able to capture the changes in each group over time.}\label{fig:quadratic-cases-Appendix}
 \end{figure}
@@
   theme_classic()
-\end{lstlisting}
-
-\begin{lstlisting}
-## Error in scale_color_aaas(): could not find function "scale_color_aaas"
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
+
 #Combining all
@@
 \begin{lstlisting}[language=R]
+set.seed(1)
 #Dataframe that contains the original reported trends
@@
 The model will use 5 knots (\passthrough{\lstinline!k=5!}) for the smooth.
-And that the smooth is constructed using gaussian process basis (\passthrough{\lstinline!bs="gp"!}).
+The smooth is constructed by default using thin plate regression splines.
 The smoothing parameter estimation method used is the restricted maximum likelihood (\passthrough{\lstinline!REML!}).
@@
 \begin{lstlisting}[language=R]
-gam_00<-gam(StO2_sim ~ s(Day, k = 5,bs="gp"),
+gam_00<-gam(StO2_sim ~ s(Day, k = 5),
             method='REML',
@@
 ## Method: REML   Optimizer: outer newton
-## full convergence after 6 iterations.
-## Gradient range [-4.297642e-08,-3.237913e-09]
-## (score 437.636 & scale 390.173).
-## Hessian positive definite, eigenvalue range [0.1251014,49.00133].
+## full convergence after 5 iterations.
+## Gradient range [-6.11034e-07,-1.169842e-07]
+## (score 439.1428 & scale 414.047).
+## Hessian positive definite, eigenvalue range [0.05006795,49.00066].
 ## Model rank =  5 / 5 
@@
 ##          k'  edf k-index p-value    
-## s(Day) 4.00 1.51    0.36  <2e-16 ***
+## s(Day) 4.00 1.36    0.26  <2e-16 ***
 ## ---
@@
 ## Formula:
-## StO2_sim ~ s(Day, k = 5, bs = "gp")
+## StO2_sim ~ s(Day, k = 5)
 ## 
@@
 ##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   20.111      1.975   10.18   <2e-16 ***
+## (Intercept)   21.929      2.035   10.78   <2e-16 ***
 ## ---
@@
 ##          edf Ref.df     F p-value   
-## s(Day) 1.511  1.782 8.116 0.00349 **
+## s(Day) 1.359  1.624 6.695 0.00273 **
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.106   Deviance explained = 11.9%
-## -REML = 437.64  Scale est. = 390.17    n = 100
+## R-sq.(adj) =  0.106   Deviance explained = 11.8%
+## -REML = 439.14  Scale est. = 414.05    n = 100
 \end{lstlisting}
@@
 Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \textasciitilde12\%.
+Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \(\approx\) 12\%.
 
@@
 \begin{lstlisting}[language=R]
-gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5,bs="gp"),
+gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5),
             method='REML',
@@
 ## Method: REML   Optimizer: outer newton
-## full convergence after 9 iterations.
-## Gradient range [-0.0001383864,0.0002277102]
-## (score 414.2119 & scale 246.4112).
-## Hessian positive definite, eigenvalue range [0.0001383659,48.50328].
+## full convergence after 11 iterations.
+## Gradient range [-3.57434e-06,1.383186e-06]
+## (score 413.523 & scale 230.4732).
+## Hessian positive definite, eigenvalue range [0.1532335,48.55232].
 ## Model rank =  9 / 9 
@@
 ##                         k'  edf k-index p-value    
-## s(Day):GroupControl   4.00 1.00    0.47  <2e-16 ***
-## s(Day):GroupTreatment 4.00 1.82    0.47  <2e-16 ***
+## s(Day):GroupControl   4.00 3.49    0.38  <2e-16 ***
+## s(Day):GroupTreatment 4.00 2.96    0.38  <2e-16 ***
 ## ---
@@
 ## Formula:
-## StO2_sim ~ s(Day, by = Group, k = 5, bs = "gp")
+## StO2_sim ~ s(Day, by = Group, k = 5)
 ## 
@@
 ##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)    20.11       1.57   12.81   <2e-16 ***
+## (Intercept)   21.929      1.518   14.45   <2e-16 ***
 ## ---
@@
 ##                         edf Ref.df      F p-value    
-## s(Day):GroupControl   1.001  1.001  5.243  0.0242 *  
-## s(Day):GroupTreatment 1.823  2.071 35.305  <2e-16 ***
+## s(Day):GroupControl   3.488  3.851  5.244 0.00442 ** 
+## s(Day):GroupTreatment 2.962  3.461 24.272 < 2e-16 ***
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.435   Deviance explained = 45.1%
-## -REML = 414.21  Scale est. = 246.41    n = 100
+## R-sq.(adj) =  0.502   Deviance explained = 53.5%
+## -REML = 413.52  Scale est. = 230.47    n = 100
 \end{lstlisting}
@@
 Moreover, the smooths (plotted via the \passthrough{\lstinline!draw()!} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \textasciitilde43\%.
+From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \(\approx\) 43\%.
 
@@
 ## Method: REML   Optimizer: outer newton
-## full convergence after 7 iterations.
-## Gradient range [-2.60407e-05,5.064354e-06]
-## (score 368.6968 & scale 78.46844).
-## Hessian positive definite, eigenvalue range [0.4967554,48.06648].
+## full convergence after 9 iterations.
+## Gradient range [-2.780424e-08,2.076237e-08]
+## (score 354.6068 & scale 63.7304).
+## Hessian positive definite, eigenvalue range [1.095531,48.08644].
 ## Model rank =  10 / 10 
@@
 ##                         k'  edf k-index p-value
-## s(Day):GroupControl   4.00 3.19    1.17    0.94
-## s(Day):GroupTreatment 4.00 3.79    1.17    0.95
+## s(Day):GroupControl   4.00 3.87    1.02    0.52
+## s(Day):GroupTreatment 4.00 3.83    1.02    0.58
 \end{lstlisting}
@@
 ## Formula:
-## StO2_sim ~ Group + s(Day, by = Group, k = 5, bs = "gp")
+## StO2_sim ~ Group + s(Day, by = Group, k = 5)
 ## 
@@
 ##                Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)       8.964      1.253   7.156 2.06e-10 ***
-## GroupTreatment   22.293      1.772  12.583  < 2e-16 ***
+## (Intercept)       9.781      1.129   8.664 1.68e-13 ***
+## GroupTreatment   24.296      1.597  15.217  < 2e-16 ***
 ## ---
@@
 ##                         edf Ref.df     F p-value    
-## s(Day):GroupControl   3.191  3.437 11.05 1.01e-06 ***
-## s(Day):GroupTreatment 3.788  3.957 63.23  < 2e-16 ***
+## s(Day):GroupControl   3.867  3.989 19.38  <2e-16 ***
+## s(Day):GroupTreatment 3.826  3.981 80.29  <2e-16 ***
 ## ---
@@
 ## 
-## R-sq.(adj) =   0.82   Deviance explained = 83.5%
-## -REML =  368.7  Scale est. = 78.468    n = 100
+## R-sq.(adj) =  0.862   Deviance explained = 87.4%
+## -REML = 354.61  Scale est. = 63.73     n = 100
 \end{lstlisting}
@@
 By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group .
-From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\textasciitilde1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 87\% of the variance data.
+From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\(\approx\) 1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 87\% of the variance data.
 
@@
 \begin{lstlisting}[language=R]
-AIC(gam_00,gam_01,gam1)
+AIC(gam_00,gam_01,m1)
 \end{lstlisting}
@@
 ##               df      AIC
-## gam_00  3.781952 885.4678
-## gam_01  5.072091 840.7331
-## gam1   10.936093 732.3416
+## gam_00  3.623938 891.2466
+## gam_01  9.312053 838.6825
+## m1     10.970436 710.9994
 \end{lstlisting}
@@
 In this case, the ``design matrix'' is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
-Briefly, the ``design matrix'' (also known as the ``Xp matrix'') from the selected model (\passthrough{\lstinline!gam1!}) is used to calculate a 95\% confidence interval of the difference between the smooth terms for each group.
+Briefly, the ``design matrix'' (also known as the ``Xp matrix'') from the selected model (\passthrough{\lstinline!m1!}) is used to calculate a 95\% confidence interval of the difference between the smooth terms for each group.
 This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
+\textbf{All pairwise comparisons in this paper have been centered at the response scale to ease interpretation }.
 
@@
 ##Pairwise comparisons
+pdat <- expand.grid(Day = seq(0, 10, length = 400),
+                    Group = c('Control', 'Treatment'))
 
@@
 
-\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!gam1!} using a 95\% confidence interval for the difference between smooths.}\label{fig:pairwise-comp-workflow-fig}
+\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!m1!} using a 95\% confidence interval for the difference between smooths.}\label{fig:pairwise-comp-workflow-fig}
 \end{figure}
@@
 
+Keep in mind that this function \textbf{does not} center the pairwise comparison at the response scale, so it has to be shifted in order to be compared to the raw data.
+
 \hypertarget{gam-and-linear-model-plots-and-missing-data}{%
@@
 
-This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\passthrough{\lstinline!gam1!}), linear model and GAM on data with missing observations are presented.
+This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\passthrough{\lstinline!m1!}), linear model and GAM on data with missing observations are presented.
 Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inlet are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
@@
 This code chunk creates panels B and D in Figure \ref{fig:sim-smooth-plot}.
-Note that this code uses the final GAM from the previous section (\passthrough{\lstinline!gam1!}), so the simulated data and the model should be generated before running this section.
+Note that this code uses the final GAM from the previous section (\passthrough{\lstinline!m1!}), so the simulated data and the model should be generated before running this section.
 
@@
 ##    <dbl> <fct>     <int>
-##  1     0 Control       5
-##  2     0 Treatment     3
-##  3     2 Control       5
+##  1     0 Control       2
+##  2     0 Treatment     4
+##  3     2 Control       6
 ##  4     2 Treatment     5
-##  5     5 Control       4
-##  6     5 Treatment     6
-##  7     7 Control       2
-##  8     7 Treatment     7
-##  9    10 Control       1
-## 10    10 Treatment     6
+##  5     5 Control       6
+##  6     5 Treatment     4
+##  7     7 Control       3
+##  8     7 Treatment     5
+##  9    10 Control       3
+## 10    10 Treatment     8
 \end{lstlisting}
@@
 
-Smooth pairwise comparisons for model \passthrough{\lstinline!gam1!} using a 95\% confidence interval for the difference between smooths.
-Pairwise comparisons for smooth terms.
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/plot-pairwise-comp-Appendix-1}
+
+}
+
+\caption{Pairwise comparisons for smooth terms.
 A: Pairwise comparisons for the full dataset.
@@
 Significant differences exist where the interval does not cover 0.
-In both cases the effect of treatment is significant after day 5.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/plot-pairwise-comp-Appendix-1}
-
-}
-
-\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!gam1!} using a 95\% confidence interval for the difference between smooths.}\label{fig:plot-pairwise-comp-Appendix}
+In both cases the effect of treatment is significant after day 5.}\label{fig:plot-pairwise-comp-Appendix}
 \end{figure}

### 2021-06-01 Code and miscellaneous/comparing-differences-in-smooths.R

@@ -1,0 +1,258 @@
+set.seed(2021)
+library(readr)
+library(dplyr)
+library(ggplot2)
+theme_set(theme_bw())
+library(mgcv)
+library(patchwork)
+
+
+## Example with linear response
+example <- function(n_time = 6, 
+                    fun_type = "linear", 
+                    error_type = "correlated") {
+  
+  if (!(fun_type %in% c("linear", "quadratic")))
+    stop('fun_type must be either "linear", or "quadratic"')
+  if (!(error_type %in% c("correlated", "independent")))
+    stop('fun_type must be either "correlated", or "independent"')
+  
+  library(tidyverse)
+  library(mvnfast)
+  library(nlme)
+  
+  x <- seq(0, 1, length.out = n_time)
+  mu <- matrix(0, length(x), 2)
+  # linear response
+  if (fun_type == "linear") {
+    mu[, 1] <- - (x - .5) + 0.25
+    mu[, 2] <- (x - .3)
+  } else {
+    # nonlinear response
+    mu[, 1] <- - 4 * (x - .5)^2 + 0.25
+    mu[, 2] <- 4 * (x - .5)^2
+  }
+  # matplot(mu, type = 'l')
+  
+  y <- array(0, dim = c(length(x), 2, 10))
+  errors <- array(0, dim = c(length(x), 2, 10))
+  
+  if (error_type == "independent") {
+    ## independent errors
+    for (i in 1:2) {
+      for (j in 1:10) {
+        errors[, i, j] <- rnorm(6, 0, 0.25)
+        y[, i, j] <- mu[, i] + errors[, i, j]
+      }
+    }
+  } else {
+    for (i in 1:2) {     # number of treatments
+      for (j in 1:10) {  # number of subjects
+        # compound symmetry errors
+        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
+        y[, i, j] <- mu[, i] + errors[, i, j]
+      }
+    }
+  }    
+  
+  
+  ## subject random effects
+  
+  ## visualizing the difference between independent errors and compound symmetry
+  ## why do we need to account for this -- overly confident inference
+  
+  
+  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
[…]

### 2021-06-02 Code and miscellaneous/Pairwise comparisons for smooths.R
geom_rect works but alpha seems not to work

@@ -126,6 +126,36 @@ c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
 c1
 
 
+c2<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
+    geom_rect(data=comp_StO2%>%
+                    filter(lower>0),
+              fill='blue',
+              alpha = 0.1,
+                aes(xmin =Day[1], xmax =Day[153],ymin=-Inf,ymax=Inf),
+                ) +
+    geom_rect(data=comp_StO2 %>%
+                  filter(upper<0),
+              fill='red',
+              alpha = 0.1,
+              aes(xmin=Day[1],xmax=Day[212],ymin =-Inf, ymax =Inf),
+
+              ) +
+    geom_ribbon(aes(ymin = lower, ymax = upper),
+                alpha = 0.9,
+                fill='black') +
+    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
+    geom_line(color='black',size=1) +
+
+    facet_wrap(~ pair) +
+    theme_classic()+
+    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    theme(
+        text=element_text(size=18),
+        legend.title=element_blank()
+    )
+
+c2
 ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
     geom_ribbon(aes(ymin = lower, ymax = upper),
                 alpha = 0.9,
@@ -149,9 +179,7 @@ ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
     theme(
         text=element_text(size=18),
         legend.title=element_blank()
-    )+
-    scale_color_aaas()+
-    scale_fill_aaas()
+    )
 
 c1
 #######################
@@ -208,3 +236,4 @@ c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
         text=element_text(size=18),
         legend.title=element_blank()
     )
+

### 2021-06-02 Manuscripts/Manuscript by chapters -v2/02-Challenges.Rmd
Manuscript with final changes from JT added. Need to tweak the shades in the pairwise comparison plots

@@
 
-## What does an rm-ANOVA fit looks like? A visual representation using simulated data{#simulation}
+## What do an rm-ANOVA and LMEM fit look like? A visual representation using simulated data{#simulation}
 
@@
 
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A).
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A).
 In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C).
@@
 
-Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect.
+Specifically, the rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect.
 In other words, Group 1 can be thought as a "Control" group and Group 2 as a "Treatment" group.

### 2021-06-02 Manuscripts/Manuscript by chapters -v2/03-GAM Theory.Rmd
Manuscript with final changes from JT added. Need to tweak the shades in the pairwise comparison plots

@@
 One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with inference [@beck1998].
-In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.
+In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic and causes bias at the endpoints of the time period.
 
 The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using _basis function_ expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A _basis_ is a set of functions that spans the mathematical space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018].
+A _basis_ is a set of functions that spans the mathematical space where the smooths that approximate $f(x_t\mid \beta_j)$ exist [@simpson2018].
 For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
 The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
-In the case of Equation \@ref(eq:GAM), the basis function is $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
+In the case of Equation \@ref(eq:GAM), the basis functions are $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
 
-Commonly used _basis functions_ are splines (cubic, thin plate regression among others).
-A cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data [@wood2017;@simpson2018].
+Splines (cubic, thin plate, etc.) are commonly used _basis functions_; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data [@wood2017;@simpson2018].
 Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
@@
 The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is already optimized in the package _mgcv_ depending on the number of knots.
+The choice of basis functions is set using default values in the package _mgcv_ depending on the number of knots.
 In Panel A of Figure \@ref(fig:basis-plot), the four basis functions (and the intercept) are shown.
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the "wigliness" of the fit, each of the basis functions of Panel A is weighted by multiplying it by a coefficient according to the matrix of Panel B.
-The parameter estimates are penalized where the penalty reduces the "wigliness" of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
+To control the "wiggliness" of the fit, each of the basis functions of Panel A is weighted by multiplying it by a coefficient according to the matrix of Panel B.
+The parameter estimates are penalized where the penalty reduces the "wiggliness" of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
 
@@
 In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \@ref(fig:basis-plot), Panel C shows the resulting penalized basis functions.
-Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
-Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
+In Figure \@ref(fig:basis-plot), Panel C shows the resulting weighted basis functions.
+Note that the magnitude of the weighting for basis function 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1).
+On the other hand, basis function 3 has roughly doubled its value.
+Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
 The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
@@
 A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Matrix for basis function penalization.
+B: Matrix for basis function weights.
 Each basis function is multiplied by a coefficient which can be positive or negative.

### 2021-06-02 Manuscripts/Manuscript by chapters -v2/04-Longitudinal analysis with GAMs.Rmd
Manuscript with final changes from JT added. Need to tweak the shades in the pairwise comparison plots

@@
 For the data in Figure \@ref(fig:sim-smooth-plot), A the main effect of interest is how $\mbox{StO}_2$ changes over time for each treatment.
-To estimate this, the model needs to incorporate independent smooths for _Group_ and _Day_, respectively.
-The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric ( _Day_ ) and the other is a factor ( _Group_ ).
+To estimate this, the model incorporates independent smooths for _Group_ and _Day_, respectively.
+The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (*Day*) and the other is a factor (*Group*).
 Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
@@
 
-`m1<-gam(StO2_sim~Group+s(Day,by=Group,k=5), method='REML',data=dat_sim)`
+`m1 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML',data=dat_sim)`
 
-This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths for  _Group_ and _Day_ (the parenthesis preceded by _s_) using 5 knots.
-The smooth is constructed by default using thin plate regression splines.
-Other splines can be used if desired, including gaussian process smooths [@simpson2018].
-The parametric term _Group_ is added to quantify differences in the effect of treatment between groups, and the `method` chosen to select the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
+This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths for `Group` and `Day` (the parenthesis preceded by `s`) using 5 knots.
+The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths [@simpson2018].
+The parametric term `Group` is added to quantify differences in the effect of treatment between groups, and the `method` chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B).
@@
 One question that might arise at this point is "what is the fit that an rm-ANOVA model produces for the simulated data?".
-The rm-ANOVA model, which corresponds to Equation \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot),C.
-This is a typical case of model misspecification: The slopes of each group are different, which would lead to a _p-value_ indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \@ref(fig:sim-smooth-plot),B) .
+The rm-ANOVA model, which corresponds to Equation \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot), C.
+This is a typical case of model misspecification: The slopes of each group are different, which would lead to a _p-value_ indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \@ref(fig:sim-smooth-plot), B) .
 
@@
 Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that time point(s).
-The absence of a _p-value_ in this case might seem odd, but the confidence interval comparison can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the empirical Bayesian confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that time point(s).
+The absence of a _p-value_ in this case might seem odd, but the empirical Bayesian confidence interval interpretation can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
 This is what happens for the simulated data in Figure \@ref(fig:sim-smooth-plot), A where the chemotherapy causes $\mbox{StO}_2$ to increase over time.
@@
 In a sense, the pairwise smooth comparison is more informative than a _post-hoc_ _p-value_.
-For biomedical studies, it is able to provide an estimate of _when_ a biological process becomes significant.
-This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change is expected.
+For biomedical studies, the smooth comparison is able to provide an estimate of _when_ and by _how much_ a biological process becomes significant.
+This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.

### 2021-06-02 Manuscripts/Manuscript by chapters -v2/05-Discussion.Rmd
Manuscript with final changes from JT added. Need to tweak the shades in the pairwise comparison plots

@@
 Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \@ref(simulation).
-This "model misspecification" error, also is known as"" "Type III" error [@dennis2019] is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
+This "model misspecification" error, also is known as a "Type III" error [@dennis2019] is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
 Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) [@lang2015] but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a $R^2$ value.
@@
 We have used $\textsf{R}$ as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research [@begley2015;@weissgerber2018], but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact soon rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research [@begley2015;@weissgerber2018], but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.

### 2021-06-02 Manuscripts/Manuscript by chapters -v2/07-Appendix.Rmd
Manuscript with final changes from JT added. Need to tweak the shades in the pairwise comparison plots

@@
 A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
+B: Coefficient matrix for the basis functions.
+Each basis function is weighted by a coefficient which can be positive or negative.
 The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
+C: Weighted basis functions.
+Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
 D: Smoother for time and original data points.
@@
 
-The resulting model is`m1`, which is the model fitted in the main manuscript.
-By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group .
+The resulting model is `m1`, which is the model fitted in the main manuscript.
+By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
 From `gam.check`, the k-index is now at an acceptable value ($\approx$ 1.02), and `summary` now indicates that the model is able to capture 87% of the variance data.
@@
 The output in this case is expected: model `gam1` has a lower AIC (711.46) whereas the initial two models have higher AICs (891 and 850).
-The AIC should not be considered as the only estimator of model goodness, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
+The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
 

### 2021-06-03 Code and miscellaneous/Pairwise comparisons for smooths.R
annotate works, need to implement in Manuscript

@@ -125,21 +125,35 @@ c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
 
 c1
 
+annotate("rect",
+         xmin=100,
+         xmax=200,
+         ymin=0,
+         ymax=Inf,
+         alpha=0.2,
+         fill="red")
+
+#find interval where Control is higher, get the value of Day
+
+v1<-comp_StO2%>%
+    filter(lower>0)%>%
+    select(Day)
+
+v2<-comp_StO2%>%
+    filter(upper<0)%>%
+    select(Day)
 
 c2<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
-    geom_rect(data=comp_StO2%>%
-                    filter(lower>0),
-              fill='blue',
-              alpha = 0.1,
-                aes(xmin =Day[1], xmax =Day[153],ymin=-Inf,ymax=Inf),
+    annotate("rect",
+                xmin =0, xmax =v1$Day[[nrow(v1)]],ymin=-Inf,ymax=Inf,
+                fill='blue',
+                alpha = 0.1,
                 ) +
-    geom_rect(data=comp_StO2 %>%
-                  filter(upper<0),
-              fill='red',
-              alpha = 0.1,
-              aes(xmin=Day[1],xmax=Day[212],ymin =-Inf, ymax =Inf),
-
-              ) +
+    annotate("rect",
+             xmin =v2$Day[[1]], xmax =v2$Day[[nrow(v2)]],ymin=-Inf,ymax=Inf,
+             fill='red',
+             alpha = 0.1,
+    ) +
     geom_ribbon(aes(ymin = lower, ymax = upper),
                 alpha = 0.9,
                 fill='black') +
@@ -156,6 +170,8 @@ c2<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
     )
 
 c2
+
+
 ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
     geom_ribbon(aes(ymin = lower, ymax = upper),
                 alpha = 0.9,

### 2021-06-04 Manuscripts/Manuscript by chapters -v2/00-Full_document.Rmd
Document seems to be ready (for now)

--- 2021-06-01 Manuscripts/Manuscript by chapters -v2/Full_document.Rmd
@@
 output:  
-  bookdown::word_document2:
-    fig_caption: yes #figure caption
-    keep_md: yes
-  bookdown::html_document2:
-    css: "style.css" #style for the HTML document
   bookdown::pdf_document2:
@@
 This paper summarizes the limitations of rm-ANOVA and LMEMs, presents the basic theory of GAMs, and uses simulated data that follows trends reported in the biomedical literature to demonstrate their implementation in $\textsf{R}$ via the package _mgcv_.
-To make this work reproducible, the code and data used in this paper are available at:________.
+To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research.
 
@@
 library(viridis)
-set.seed(1) #set seed for reproducibility
+set.seed(2021) #set seed for reproducibility
 

### 2021-06-04 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
Document seems to be ready (for now)

--- 2021-06-01 Manuscripts/Manuscript by chapters -v2/Full_document.tex
@@
 This paper summarizes the limitations of rm-ANOVA and LMEMs, presents the basic theory of GAMs, and uses simulated data that follows trends reported in the biomedical literature to demonstrate their implementation in \(\textsf{R}\) via the package \emph{mgcv}.
-To make this work reproducible, the code and data used in this paper are available at:\_\_\_\_\_\_\_\_.
+To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
 
@@
 \hypertarget{simulation}{%
-\subsection{What does an rm-ANOVA fit looks like? A visual representation using simulated data}\label{simulation}}
+\subsection{What do an rm-ANOVA and LMEM fit look like? A visual representation using simulated data}\label{simulation}}
 
@@
 
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}, A).
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}, A).
 In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}, C).
@@
 
-Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect.
+Specifically, the rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect.
 In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group.
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/l-q-response-1}
+{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/l-q-response-1}
 
@@
 One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
-In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic, and causes bias at the endpoints of the time period.
+In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
 
 The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t \mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
+A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
 For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
 The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{52}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
-In the case of Equation \eqref{eq:GAM}, the basis function is \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
-
-Commonly used \emph{basis functions} are splines (cubic, thin plate regression among others).
-A cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
+In the case of Equation \eqref{eq:GAM}, the basis functions are \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
+
+Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
 Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
@@
 The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is already optimized in the package \emph{mgcv} depending on the number of knots.
+The choice of basis functions is set using default values in the package \emph{mgcv} depending on the number of knots.
 In Panel A of Figure \ref{fig:basis-plot}, the four basis functions (and the intercept) are shown.
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the ``wigliness'' of the fit, each of the basis functions of Panel A is weighted by multiplying it by a coefficient according to the matrix of Panel B.
-The parameter estimates are penalized where the penalty reduces the ``wigliness'' of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
+To control the ``wiggliness'' of the fit, each of the basis functions of Panel A is weighted by multiplying it by a coefficient according to the matrix of Panel B.
+The parameter estimates are penalized where the penalty reduces the ``wiggliness'' of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
 
 In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \ref{fig:basis-plot}, Panel C shows the resulting penalized basis functions.
-Note that the penalization for basis 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is negative and less than 1); on the other hand, basis 3 has roughly doubled its value.
-Finally, the penalized basis functions are added at each timepoint to produce the smooth term.
+In Figure \ref{fig:basis-plot}, Panel C shows the resulting weighted basis functions.
+Note that the magnitude of the weighting for basis function 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1).
+On the other hand, basis function 3 has roughly doubled its value.
+Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
 The resulting smooth term for the effect of \emph{time} is shown in Panel D (orange line) along the simulated values per group, which appear as points.
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/basis-plot-1}
+{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-1}
 
@@
 For the data in Figure \ref{fig:sim-smooth-plot}, A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
-To estimate this, the model needs to incorporate independent smooths for \emph{Group} and \emph{Day}, respectively.
-The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric ( \emph{Day} ) and the other is a factor ( \emph{Group} ).
+To estimate this, the model incorporates independent smooths for \emph{Group} and \emph{Day}, respectively.
+The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}).
 Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
@@
 
-\passthrough{\lstinline!m1<-gam(StO2\_sim\~Group+s(Day,by=Group,k=5), method='REML',data=dat\_sim)!}
-
-This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths for \emph{Group} and \emph{Day} (the parenthesis preceded by \emph{s}) using 5 knots.
-The smooth is constructed by default using thin plate regression splines.
-Other splines can be used if desired, including gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-The parametric term \emph{Group} is added to quantify differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to select the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+\passthrough{\lstinline!m1 <- gam(StO2\_sim \~ Group + s(Day, by=Group, k=5), method='REML',data=dat\_sim)!}
+
+This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths for \passthrough{\lstinline!Group!} and \passthrough{\lstinline!Day!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
+The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
+The parametric term \passthrough{\lstinline!Group!} is added to quantify differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}.
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot},B).
@@
 
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot},C.
-This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \ref{fig:sim-smooth-plot},B) .
+One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}, C.
+This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \ref{fig:sim-smooth-plot}, B) .
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/sim-smooth-plot-1}
+{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/plot-pairwise-comp-1}
+{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-1}
 
@@
 Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the \emph{difference} between the confidence intervals of the fitted smooths for such groups is non-zero, then a significant difference exists at that time point(s).
-The absence of a \emph{p-value} in this case might seem odd, but the confidence interval comparison can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the \emph{difference} between the empirical Bayesian confidence intervals of the fitted smooths for such groups is non-zero, then a significant difference exists at that time point(s).
+The absence of a \emph{p-value} in this case might seem odd, but the empirical Bayesian confidence interval interpretation can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
 This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}, A where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
@@
 
-With this correction in mind, the shaded region under or above the confidence interval (that does not cover 0) indicates the interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and 3 for the full dataset indicates that through that time, the ``Control'' group has higher \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by day 4 it is the ``Treatment'' group the one that has greater \(\mbox{StO}_2\).
+With this correction in mind, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
+Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by day 4 it is the ``Treatment'' group the one that has greater \(\mbox{StO}_2\).
 This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 4 for the model used.
@@
 In a sense, the pairwise smooth comparison is more informative than a \emph{post-hoc} \emph{p-value}.
-For biomedical studies, it is able to provide an estimate of \emph{when} a biological process becomes significant.
-This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change is expected.
+For biomedical studies, the smooth comparison is able to provide an estimate of \emph{when} and by \emph{how much} a biological process becomes significant.
+This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.
 
@@
 Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}.
-This ``model misspecification'' error, also is known as"" ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
+This ``model misspecification'' error, also is known as a ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
 Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{56}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{58},\protect\hyperlink{ref-weissgerber2018}{59}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact soon rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{58},\protect\hyperlink{ref-weissgerber2018}{59}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 Huynh, L.S.
-Feldt, {Estimation of the Box correction for degrees of freedom from sample data in randomized block and split-Plot designs}, Journal of Educational Statistics.
+Feldt, {Estimation of the Box correction for degrees of freedom from sample data in randomized block and split-plot designs}, Journal of Educational Statistics.
 1 (1976) 69--82.
@@
 Scheepers, H.J.
-Tily, {Random effects structure for confirmatory hypothesis testing: Keep it maximal}, {JournalL of Memory and Language}.
+Tily, {Random effects structure for confirmatory hypothesis testing: Keep it maximal}, {Journal of Memory and Language}.
 {68} (2013) 255--278.
@@
 Simpson, N.
-Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, Peerj.
+Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, {PeerJ}.
 7 (2019).
@@
 Lang, D.G.
-Altman, {Basic statistical reporting for articles published in Biomedical Journals: The ``Statistical Analyses and Methods in the Published Literature{''} or the SAMPL Guidelines}, {INTERNATIONAL JOURNAL OF NURSING STUDIES}.
+Altman, {Basic statistical reporting for articles published in Biomedical Journals: The ``Statistical Analyses and Methods in the Published Literature{''} or the SAMPL Guidelines}, INTERNATIONAL JOURNAL OF NURSING STUDIES.
 {52} (2015) 5--9.
@@
 \begin{lstlisting}[language=R]
-set.seed(1)
 ##########Section for calculations###########
@@
 \begin{figure}[H]
-\includegraphics{Full_document_files/figure-latex/linear-cases-Appendix-1} \caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
+\includegraphics{00-Full_document_files/figure-latex/linear-cases-Appendix-1} \caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
 A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
@@
 \begin{figure}[H]
-\includegraphics{Full_document_files/figure-latex/quadratic-cases-Appendix-1} \caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
+\includegraphics{00-Full_document_files/figure-latex/quadratic-cases-Appendix-1} \caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
 A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/basis-plot-appendix-1}
+{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-appendix-1}
 
@@
 A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Penalty matrix for the basis functions.
-Each basis function is penalized by a coefficient which can be positive or negative.
+B: Matrix for basis function weights.
+Each basis function is multiplied by a coefficient which can be positive or negative.
 The coefficient determines the overall effect of each basis in the final smoother.
-C: Penalized basis functions.
-Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
+C: Weighted basis functions.
+Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
 D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}\label{fig:basis-plot-appendix}
+The smoother (line) is the result of the sum of each penalized basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot-appendix}
 \end{figure}
@@
 
-\begin{lstlisting}[language=R]
-#see https://patchwork.data-imaginist.com/reference/area.html
-layout1 <- c(
-  area(1, 1),
-  area( 1, 2),
-  area(2, 1),
-  area(2, 2),
-  area(1, 3, 2)
-)
-
-layout2 <- c(
-  area(1, 1),
-  area( 1, 2),
-  area(2, 1),
-  area(2, 2),
-  area(1,3,2,5)
-)
-
-#plot(layout2)
-\end{lstlisting}
-
 \hypertarget{graphical-diagnostics}{%
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/first-GAM-diag-1}
+{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/first-GAM-diag-1}
 
@@
 
-\includegraphics{Full_document_files/figure-latex/first-GAM-check-1}
+\includegraphics{00-Full_document_files/figure-latex/first-GAM-check-1}
 
@@
 ## full convergence after 5 iterations.
-## Gradient range [-6.11034e-07,-1.169842e-07]
-## (score 439.1428 & scale 414.047).
-## Hessian positive definite, eigenvalue range [0.05006795,49.00066].
+## Gradient range [-0.0003727881,-6.621452e-07]
+## (score 444.0118 & scale 450.6638).
+## Hessian positive definite, eigenvalue range [0.3881695,49.00676].
 ## Model rank =  5 / 5 
@@
 ##          k'  edf k-index p-value    
-## s(Day) 4.00 1.36    0.26  <2e-16 ***
+## s(Day) 4.00 2.11    0.36  <2e-16 ***
 ## ---
@@
 ##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   21.929      2.035   10.78   <2e-16 ***
+## (Intercept)   22.967      2.123   10.82   <2e-16 ***
 ## ---
@@
 ##          edf Ref.df     F  p-value    
-## s(Day) 1.359  1.624 6.695 0.00273 **
+## s(Day) 2.114  2.565 7.633 0.000517 ***
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.106   Deviance explained = 11.8%
-## -REML = 439.14  Scale est. = 414.05    n = 100
+## R-sq.(adj) =  0.153   Deviance explained = 17.2%
+## -REML = 444.01  Scale est. = 450.66    n = 100
 \end{lstlisting}
@@
 If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless1).
-From the output, it can be seen that the k-index is effectively \textless0.3, which indicates that the model is not capturing the variability in the data.
-The `edf' (effective degrees of freedom) is an indicator of the complexity of the smooth.
+From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data.
+The \passthrough{\lstinline!edf!} (effective degrees of freedom) is an indicator of the complexity of the smooth.
 Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
@@
 Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \(\approx\) 12\%.
+Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \(\approx\) 17\%.
 
@@
 
-\includegraphics{Full_document_files/figure-latex/second-GAM-1}
+\includegraphics{00-Full_document_files/figure-latex/second-GAM-1}
 
@@
 ## Method: REML   Optimizer: outer newton
-## full convergence after 11 iterations.
-## Gradient range [-3.57434e-06,1.383186e-06]
-## (score 413.523 & scale 230.4732).
-## Hessian positive definite, eigenvalue range [0.1532335,48.55232].
+## full convergence after 7 iterations.
+## Gradient range [-5.51754e-05,2.671715e-06]
+## (score 423.3916 & scale 280.8777).
+## Hessian positive definite, eigenvalue range [0.3162258,48.5557].
 ## Model rank =  9 / 9 
@@
 ##                         k'  edf k-index p-value    
-## s(Day):GroupControl   4.00 3.49    0.38  <2e-16 ***
-## s(Day):GroupTreatment 4.00 2.96    0.38  <2e-16 ***
+## s(Day):GroupControl   4.00 3.39    0.43  <2e-16 ***
+## s(Day):GroupTreatment 4.00 3.23    0.43  <2e-16 ***
 ## ---
@@
 ##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   21.929      1.518   14.45   <2e-16 ***
+## (Intercept)   22.967      1.676    13.7   <2e-16 ***
 ## ---
@@
 ##                         edf Ref.df      F p-value    
-## s(Day):GroupControl   3.488  3.851  5.244 0.00442 ** 
-## s(Day):GroupTreatment 2.962  3.461 24.272 < 2e-16 ***
+## s(Day):GroupControl   3.392  3.794  3.817  0.0304 *  
+## s(Day):GroupTreatment 3.229  3.682 21.174  <2e-16 ***
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.502   Deviance explained = 53.5%
-## -REML = 413.52  Scale est. = 230.47    n = 100
+## R-sq.(adj) =  0.472   Deviance explained = 50.8%
+## -REML = 423.39  Scale est. = 280.88    n = 100
 \end{lstlisting}
 
-Diagnostics for this model indicate that the k-index is still below 1 (0.32 from \passthrough{\lstinline!gam.check!}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
+Diagnostics for this model indicate that the k-index is still below 1 (0.43 from \passthrough{\lstinline!gam.check!}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
 Moreover, the smooths (plotted via the \passthrough{\lstinline!draw()!} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \(\approx\) 43\%.
+From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \(\approx\) 51\%.
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/second-GAM-diag-1}
+{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/second-GAM-diag-1}
 
@@
 
-\includegraphics{Full_document_files/figure-latex/final-model-Appendix-1}
+\includegraphics{00-Full_document_files/figure-latex/final-model-Appendix-1}
 
@@
 ## Method: REML   Optimizer: outer newton
-## full convergence after 9 iterations.
-## Gradient range [-2.780424e-08,2.076237e-08]
-## (score 354.6068 & scale 63.7304).
-## Hessian positive definite, eigenvalue range [1.095531,48.08644].
+## full convergence after 10 iterations.
+## Gradient range [-8.164307e-08,1.500338e-08]
+## (score 355.8554 & scale 64.53344).
+## Hessian positive definite, eigenvalue range [1.174841,48.08834].
 ## Model rank =  10 / 10 
@@
 ##                         k'  edf k-index p-value
-## s(Day):GroupControl   4.00 3.87    1.02    0.52
-## s(Day):GroupTreatment 4.00 3.83    1.02    0.58
+## s(Day):GroupControl   4.00 3.87    1.02    0.59
+## s(Day):GroupTreatment 4.00 3.88    1.02    0.54
 \end{lstlisting}
@@
 ##                Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)       9.781      1.129   8.664 1.68e-13 ***
-## GroupTreatment   24.296      1.597  15.217  < 2e-16 ***
+## (Intercept)       9.084      1.136   7.996 4.09e-12 ***
+## GroupTreatment   27.766      1.607  17.282  < 2e-16 ***
 ## ---
@@
 ##                         edf Ref.df     F p-value    
-## s(Day):GroupControl   3.867  3.989 19.38  <2e-16 ***
-## s(Day):GroupTreatment 3.826  3.981 80.29  <2e-16 ***
+## s(Day):GroupControl   3.873  3.990 17.57  <2e-16 ***
+## s(Day):GroupTreatment 3.879  3.991 89.33  <2e-16 ***
 ## ---
@@
 ## 
-## R-sq.(adj) =  0.862   Deviance explained = 87.4%
-## -REML = 354.61  Scale est. = 63.73     n = 100
+## R-sq.(adj) =  0.879   Deviance explained = 88.9%
+## -REML = 355.86  Scale est. = 64.533    n = 100
 \end{lstlisting}
 
-The resulting model is\passthrough{\lstinline!m1!}, which is the model fitted in the main manuscript.
-By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group .
-From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\(\approx\) 1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 87\% of the variance data.
+The resulting model is \passthrough{\lstinline!m1!}, which is the model fitted in the main manuscript.
+By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
+From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\(\approx\) 1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 89\% of the variance in the data.
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/final-GAM-diag-1}
+{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/final-GAM-diag-1}
 
@@
 ##               df      AIC
-## gam_00  3.623938 891.2466
-## gam_01  9.312053 838.6825
-## m1     10.970436 710.9994
+## gam_00  4.564893 900.8257
+## gam_01  9.476137 858.6051
+## m1     10.980983 712.2067
 \end{lstlisting}
 
-The output in this case is expected: model \passthrough{\lstinline!gam1!} has a lower AIC (711.46) whereas the initial two models have higher AICs (891 and 850).
-The AIC should not be considered as the only estimator of model goodness, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
+The output in this case is expected: model \passthrough{\lstinline!m1!} has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
+The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
 
@@
 
+#use function from the pairwise comparison plot in the manuscript to get the shaded regions
+
+    my_list<-pairwise_limits(comp_StO2)
+
 #plot the difference
 c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
+  #shaded region
+  annotate("rect",
+                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
+                fill='#30123BFF',
+                alpha = 0.5,
+                ) +
+  annotate("text",
+             x=1.5,
+             y=-10,
+             label="Control",size=10
+           )+
+  #shaded region  
+  annotate("rect",
+             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
+             fill='#7A0403FF',
+             alpha = 0.5
+           ) +
+  annotate("text",
+             x=6,
+             y=-10,
+             label="Treatment",
+             size=10
+           )+
   #ribbon for difference confidence interval  
@@
     geom_line(data=comp_StO2,aes(y=0),size=0.5)+
-  #highlight area under the curve where "Control" is higher
-  geom_ribbon(data=comp_StO2%>%
-                    filter(lower>0),
-                aes(ymin =0, ymax =lower),
-                alpha = 0.5,
-                fill='#30123BFF') +
-  #highlight area under the curve where "Treatment" is higher
-  geom_ribbon(data=comp_StO2 %>%
-                    filter(upper<0),
-                    aes(ymin =0, ymax =upper),
-                alpha = 0.5,
-                fill='#7A0403FF') +
     facet_wrap(~ pair) +
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/pairwise-comp-workflow-fig-1}
-
-}
-
-\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!m1!} using a 95\% confidence interval for the difference between smooths.}\label{fig:pairwise-comp-workflow-fig}
+{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/pairwise-comp-workflow-fig-1}
+
+}
+
+\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!m1!} using a 95\% confidence interval for the difference between smooths.
+The comparison is centered at the response scale.
+Shaded regions indicate time intervals where each treatment group has a non-zero effect.}\label{fig:pairwise-comp-workflow-fig}
 \end{figure}
@@
   count(Day)
-\end{lstlisting}
-
-\begin{lstlisting}
-## # A tibble: 10 x 3
-## # Groups:   Day, Group [10]
-##      Day Group         n
-##    <dbl> <fct>     <int>
-##  1     0 Control       2
-##  2     0 Treatment     4
-##  3     2 Control       6
-##  4     2 Treatment     5
-##  5     5 Control       6
-##  6     5 Treatment     4
-##  7     7 Control       3
-##  8     7 Treatment     5
-##  9    10 Control       3
-## 10    10 Treatment     8
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
+
 #the same model used for the full dataset
@@
 appraise(mod_m1)
-\end{lstlisting}
-
-\includegraphics{Full_document_files/figure-latex/missing-data-Appendix-1}
-
-\begin{lstlisting}[language=R]
+
 m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
@@
 
-\begin{lstlisting}[language=R]
-mult_plot<-f2+inset_element(
-  f1, left = 0.01, 
-  bottom = 0.5, 
-  right = 0.5, 
-  top = 1.0)+
-  f3+f4+f6+
-   plot_annotation(tag_levels='A')&
-   ylim(c(-7,75)) &
-  theme(
-     text=element_text(size=18)
-     )&
-  thm
-
-mult_plot
-\end{lstlisting}
-
 \begin{figure}[H]
 
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/sim-smooth-plot-Appendix-1}
+{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-Appendix-1}
 
@@
 #this function takes the model, grid and groups to be compared using the lpmatrix
+#originally developed by G. Simpson:
+#https://fromthebottomoftheheap.net/2017/10/10/difference-splines-i/
 
@@
     X[, ! (col1 | col2)] <- 0
+
     ## zero out the parametric cols
+    #This line has been commented to keep the comparison at the response level,
+    #otherwise it gives the marginal change between smooths
     #X[, !grepl('^s\\(', colnames(xp))] <- 0
     dif <- X %*% coef(model)
+    #get standard error, critical value and boundaries
     se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
@@
 
+#use the function to calculate the difference in smooths
 comp1<-smooth_diff(m1,pdat,'Control','Treatment')
 
+#Create a dataframe with time, comparisons and labels for regions where difference exists
 comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
@@
 
+pairwise_limits<-function(dataframe){
+    #extract values where the lower limit of the ribbon is greater than zero
+    #this is the region where the control group effect is greater
+    v1<-dataframe%>%
+        filter(lower>0)%>%
+        select(Day)
+    #get day  initial value
+    init1=v1$Day[[1]]
+    #get day final value
+    final1=v1$Day[[nrow(v1)]]
+
+    #extract values where the value of the upper limit of the ribbon is lower than zero
+    #this corresponds to the region where the treatment group effect is greater
+    v2<-comp_StO2_full%>%
+        filter(upper<0)%>%
+        select(Day)
+
+    init2=v2$Day[[1]]
+    final2=v2$Day[[nrow(v2)]]
+    #store values
+    my_list<-list(init1=init1,
+                  final1=final1,
+                  init2=init2,
+                  final2=final2)
+return(my_list)
+}
+
+my_list<-pairwise_limits(comp_StO2_full)
+
 c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
+    annotate("rect",
+                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
+                fill='#30123BFF',
+                alpha = 0.5,
+                ) +
+  annotate("text",
+             x=1.5,
+             y=-15,
+             label="Control",
+           size=10,
+           angle=90
+           )+
+    annotate("rect",
+             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
+             fill='#7A0403FF',
+             alpha = 0.5,
+    ) +
+  annotate("text",
+             x=6,
+             y=-15,
+             label="Treatment",
+             size=10,
+           angle=90
+           )+
     geom_ribbon(aes(ymin = lower, ymax = upper),
@@
                 fill='#DB3A07FF') +
-    geom_line(color='#E75B64FF',size=1) +
     geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
-    geom_ribbon(data=comp_StO2_full%>%
-                    filter(lower>0),
-                aes(ymin =0, ymax =lower),
-                alpha = 0.5,
-                fill='#30123BFF') +
-    geom_ribbon(data=comp_StO2_full %>%
-                    filter(upper<0),
-                    aes(ymin =0, ymax =upper),
-                alpha = 0.5,
-                fill='#7A0403FF') +
+    geom_line(color='black',size=1) +
+
     facet_wrap(~ pair) +
@@
 
+my_list<-pairwise_limits(comp_StO2_missing)
+
 c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
+    annotate("rect",
+             xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
+             fill='#30123BFF',
+             alpha = 0.5,
+    ) +
+  annotate("text",
+             x=1.5,
+             y=-15,
+             label="Control",size=10
+           )+
+    annotate("rect",
+             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
+             fill='#7A0403FF',
+             alpha = 0.5,
+    ) +
+  annotate("text",
+             x=6,
+             y=-15,
+             label="Treatment",
+             size=10)+
     geom_ribbon(aes(ymin = lower, ymax = upper),
@@
                 fill='#DB3A07FF') +
-    geom_line(color='#E75B64FF',size=1) +
     geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
-    # geom_ribbon(data=comp_StO2_missing%>%
-    #                 filter(lower>0),
-    #             aes(ymin =0, ymax =lower),
-    #             alpha = 0.5,
-    #             fill='#30123BFF') +
-    geom_ribbon(data=comp_StO2_missing %>%
-                    filter(upper<0),
-                    aes(ymin =0, ymax =upper),
-                alpha = 0.5,
-                fill='#7A0403FF') +
+    geom_line(color='black',size=1) +
     facet_wrap(~ pair) +
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/plot-pairwise-comp-Appendix-1}
+{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-Appendix-1}
 

### 2021-06-04 Manuscripts/Manuscript by chapters -v2/03-GAM Theory.Rmd
Document seems to be ready (for now)

@@
 ```{r,basis-functions-plot, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-set.seed(1)
 n_time = 6

### 2021-06-04 Manuscripts/Manuscript by chapters -v2/04-Longitudinal analysis with GAMs.Rmd
Document seems to be ready (for now)

@@
 ```{r,sim-tumor-response,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-set.seed(1)
 dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
@@
 
+#function to obtain values for the shading regions
+pairwise_limits<-function(dataframe){
+    #extract values where the lower limit of the ribbon is greater than zero
+    #this is the region where the control group effect is greater
+    v1<-dataframe%>%
+        filter(lower>0)%>%
+        select(Day)
+    #get day  initial value
+    init1=v1$Day[[1]]
+    #get day final value
+    final1=v1$Day[[nrow(v1)]]
+
+    #extract values where the value of the upper limit of the ribbon is lower than zero
+    #this corresponds to the region where the treatment group effect is greater
+    v2<-comp_StO2_full%>%
+        filter(upper<0)%>%
+        select(Day)
+
+    init2=v2$Day[[1]]
+    final2=v2$Day[[nrow(v2)]]
+    #store values
+    my_list<-list(init1=init1,
+                  final1=final1,
+                  init2=init2,
+                  final2=final2)
+return(my_list)
+}
+
+my_list<-pairwise_limits(comp_StO2_full)
+
 c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
+    annotate("rect",
+                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
+                fill='#30123BFF',
+                alpha = 0.5,
+                ) +
+  annotate("text",
+             x=1.5,
+             y=-15,
+             label="Control",
+           size=10,
+           angle=90
+           )+
+    annotate("rect",
+             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
+             fill='#7A0403FF',
+             alpha = 0.5,
+    ) +
+  annotate("text",
+             x=6,
+             y=-15,
+             label="Treatment",
+             size=10,
+           angle=90
+           )+
     geom_ribbon(aes(ymin = lower, ymax = upper),
@@
 
+my_list<-pairwise_limits(comp_StO2_missing)
+
 c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
+    annotate("rect",
+             xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
+             fill='#30123BFF',
+             alpha = 0.5
+             ) +
+  annotate("text",
+             x=1.5,
+             y=-15,
+             label="Control",
+           size=10,
+           angle=90
+           )+
+    annotate("rect",
+             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
+             fill='#7A0403FF',
+             alpha = 0.5,
+    ) +
+  annotate("text",
+             x=6,
+             y=-15,
+             label="Treatment",
+             size=10,
+           angle=90
+           )+
     geom_ribbon(aes(ymin = lower, ymax = upper),
@@
 
-With this correction in mind, the shaded region under or above the confidence interval (that does not cover 0) indicates the interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and 3 for the full dataset indicates that through that time, the "Control" group has higher $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by day 4 it is the "Treatment" group the one that has greater $\mbox{StO}_2$.
+With this correction in mind, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
+Notice that the shaded region between days 0 and $\approx$ 2 for the full dataset indicates that through that time, the "Control" group has higher $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by day 4 it is the "Treatment" group the one that has greater $\mbox{StO}_2$.
 This would suggest that the effect of chemotherapy in the "Treatment" group becomes significant after day 4 for the model used.

### 2021-06-04 Manuscripts/Manuscript by chapters -v2/07-Appendix.Rmd
Document seems to be ready (for now)

@@
 
-set.seed(1)
 ##########Section for calculations###########
@@
 
-```{r,basis-Appendix, echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
+```{r,basis-Appendix, echo=TRUE,include=TRUE,message=FALSE,warning=FALSE,eval=FALSE}
 #generate the response: the same initial procedure from the previous section to simulate
@@
 
-(ref:basis-caption-appendix) Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Coefficient matrix for the basis functions.
-Each basis function is weighted by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.
-
-```{r,basis-plot-appendix,fig.width=10, fig.height=10, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, fig.show='hold', fig.cap='(ref:basis-caption-appendix)'}
-
-par(mar = c(2, 2, 2, 2))
-
-b_plot
+```{r,basis-plot-appendix, ref.label=I('basis-plot')}
+
 ```
@@
 
-```{r,sim-tumor-response-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
+```{r,sim-tumor-response-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE,eval=FALSE}
 set.seed(1)
@@
 
-```{r, plotting layouts,echo=TRUE, include=TRUE}
+```{r, plotting layouts,echo=TRUE, include=FALSE}
 
@@
 If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (<1).
-From the output, it can be seen that the k-index is effectively <0.3, which indicates that the model is not capturing the variability in the data.
-The 'edf' (effective degrees of freedom) is an indicator of the complexity of the smooth.
+From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data.
+The `edf` (effective degrees of freedom) is an indicator of the complexity of the smooth.
 Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
@@
 Here, the _p-value_ is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to $\approx$ 12%.
+Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to $\approx$ 17%.
 
@@
 
-Diagnostics for this model indicate that the k-index is still below 1 (0.32 from `gam.check`), and that the residuals are still not following a normal distribution (Figure  \@ref(fig:second-GAM-diag)).
+Diagnostics for this model indicate that the k-index is still below 1 (0.43 from `gam.check`), and that the residuals are still not following a normal distribution (Figure  \@ref(fig:second-GAM-diag)).
 Moreover, the smooths (plotted via the `draw()` function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-From `summary()`, the deviance explained by the model is $\approx$ 43%.
+From `summary()`, the deviance explained by the model is $\approx$ 51%.
 
@@
 By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From `gam.check`, the k-index is now at an acceptable value ($\approx$ 1.02), and `summary` now indicates that the model is able to capture 87% of the variance data.
+From `gam.check`, the k-index is now at an acceptable value ($\approx$ 1.02), and `summary` now indicates that the model is able to capture 89% of the variance in the data.
 
@@
 
-The output in this case is expected: model `gam1` has a lower AIC (711.46) whereas the initial two models have higher AICs (891 and 850).
+The output in this case is expected: model `m1` has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
 The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
@@
 (ref:pairwise-comp-caption) Smooth pairwise comparisons for model `m1` using a 95% confidence interval for the difference between smooths.
+The comparison is centered at the response scale.
+Shaded regions indicate time intervals where each treatment group has a non-zero effect.
 
@@
 
-```{r, plotting-appendix, echo=TRUE, include=TRUE, message=FALSE, warning=FALSE}
+```{r, plotting-appendix, echo=TRUE, include=TRUE, message=FALSE, warning=FALSE,eval=FALSE}
 #linear model
@@
 
-```{r,missing-data-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
+```{r,missing-data-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE,eval=FALSE}
 #missing data
@@
 
-(ref:sim-plot-caption-Appendix) Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95% confidence intervals.
-C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40% of its observations missing.
-Lines represent trends for each group, shaded regions are 95% confidence intervals.
-
-```{r,sim-smooth-plot-Appendix,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=TRUE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption-Appendix)'}
-
-mult_plot<-f2+inset_element(
-  f1, left = 0.01, 
-  bottom = 0.5, 
-  right = 0.5, 
-  top = 1.0)+
-  f3+f4+f6+
-   plot_annotation(tag_levels='A')&
-   ylim(c(-7,75)) &
-  theme(
-     text=element_text(size=18)
-     )&
-  thm
-
-mult_plot
+```{r,sim-smooth-plot-Appendix, ref.label=I('sim-smooth-plot')}
 
@@
 
-```{r,pairwise-comp-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
+```{r,pairwise-comp-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE, eval=FALSE}
 ##Pairwise comparisons
@@
 
-(ref:pairwise-caption-both) Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-B: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the interval does not cover 0.
-In both cases the effect of treatment is significant after day 5.
-
-```{r,plot-pairwise-comp-Appendix,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-caption-both)'}
-
-pair_comp+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-
-```
+```{r,plot-pairwise-comp-Appendix,ref.label=I('plot-pairwise-comp')}
+
+```

### 2021-06-04 Code and miscellaneous/Pairwise comparisons for smooths.R

@@ -2,7 +2,7 @@
 #significance
 #https://stats.stackexchange.com/questions/376237/correcting-for-multiple-pairwise-comparisons-with-gam-objects-mgcv-in-r
 
-
+set.seed(123)
 dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
             Day=rep(c(0,2,5,7,10),times=2),
             Group=as.factor(rep(c("Control","Treatment"),each=5))
@@ -34,132 +34,143 @@ df <- 6
 dat_sim <- simulate_data(dat, n, sd)
 
 
-gam1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
+m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5,bs="gp"),
             method='REML',
             data  = dat_sim)
 
-#creates fine grid at the values of Day
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
+######missing data###########
+#############################
+missing <- sample(1:100, 40)
 
-##matrix that contains the basis functions evaluated at the points in pdat
-    xp <- predict(gam1, newdata = pdat, type = 'lpmatrix')
+#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
 
+ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
 
-#Find columns in xp where the name contains "Control"
-    c1 <- grepl('Control', colnames(xp))
+#create a new dataframe, remove the StO2 column
+dat_missing <- dat_sim[,-1]
 
-#Find columns in xp where the name contains 'Treatment'
-    c2 <- grepl('Treatment', colnames(xp))
+#add NAs at the ind positions
+dat_missing$StO2_sim[ind]<-NA
 
-#Find rows in pdat that correspond to either 'Control' or 'Treatment'
-    r1 <- with(pdat, Group == 'Control')
-    r2 <- with(pdat, Group == 'Treatment')
-
-# In xp: find the rows that correspond to Control or Treatment, those that do not match will be
-    #set to zero. Then, substract the values from the rows corresponding to 'Control' from those that correspond
-    #to 'Treatment'
-    X <- xp[r1, ] - xp[r2, ]
-
-    test<-control-treatment
[…]

### 2021-06-07 Manuscripts/Manuscript by chapters -v2/00-Full_document.Rmd
Updated files with minor changes.

@@
 Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
-This paper summarizes the limitations of rm-ANOVA and LMEMs, presents the basic theory of GAMs, and uses simulated data that follows trends reported in the biomedical literature to demonstrate their implementation in $\textsf{R}$ via the package _mgcv_.
+This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
+We also present the basic theory of GAMs, and use simulated data that follows trends reported in the biomedical literature to demonstrate how these models are implemented in $\textsf{R}$ via the package _mgcv_, showing that GAMs are able to produce estimates that are consistent with the trends of non-linear data even if the case when missing observations exist.
 To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research.

### 2021-06-07 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
Updated files with minor changes.

@@
 Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
-This paper summarizes the limitations of rm-ANOVA and LMEMs, presents the basic theory of GAMs, and uses simulated data that follows trends reported in the biomedical literature to demonstrate their implementation in \(\textsf{R}\) via the package \emph{mgcv}.
+This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
+We also present the basic theory of GAMs, and use simulated data that follows trends reported in the biomedical literature to demonstrate how these models are implemented in \(\textsf{R}\) via the package \emph{mgcv}, showing that GAMs are able to produce estimates that are consistent with the trends of non-linear data even if the case when missing observations exist.
 To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
@@
 Significant differences exist where the interval does not cover 0.
-In both cases the effect of treatment is significant after day 5.}\label{fig:plot-pairwise-comp}
+In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp}
 \end{figure}
@@
 With this correction in mind, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by day 4 it is the ``Treatment'' group the one that has greater \(\mbox{StO}_2\).
-This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 4 for the model used.
-Moreover, notice that although there is no actual measurement at day 4, the model is capable of providing an estimate of when the shift in \(\mbox{StO}_2\) occurs.
+Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by \(\approx\) 3 day it is the ``Treatment'' group the one that has greater \(\mbox{StO}_2\).
+This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the model used.
+Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in \(\mbox{StO}_2\) occurs.
 
 On the data with missing observations (Figure \ref{fig:sim-smooth-plot}, D), the confidence intervals of the smooths overlap between days 0 and 3.
-Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}, B) shows that there is not a significant difference between the groups during that period, but is still able to pick the change on day 4 as the full dataset smooth pairwise comparison.
+Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}, B) shows that there is not a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.
 
@@
 One of the features of GAMs is that they go beyond a mere \emph{p-value} to indicate differences between groups, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
-The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 4 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
+The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
 
@@
 
-This work was supported by the National Science Foundation Career Award (CBET 1751554) and the Arkansas Biosciences Institute.
+This work was supported by the National Science Foundation Career Award (CBET 1751554, TJM) and the Arkansas Biosciences Institute.
 
@@
              x=1.5,
-             y=-15,
-             label="Control",
-           size=10,
+             y=-18,
+             label="Control>Treatment",
+           size=8,
            angle=90
@@
              x=6,
-             y=-15,
-             label="Treatment",
-             size=10,
+             y=-18,
+             label="Treatment>Control",
+             size=8,
            angle=90
@@
              x=1.5,
-             y=-15,
-             label="Control",size=10
+             y=-18,
+             label="Control>Treatment",
+           size=8
            )+
@@
              x=6,
-             y=-15,
-             label="Treatment",
-             size=10)+
+             y=-18,
+             label="Treatment>Control",
+             size=8)+
     geom_ribbon(aes(ymin = lower, ymax = upper),
@@
 Significant differences exist where the interval does not cover 0.
-In both cases the effect of treatment is significant after day 5.}\label{fig:plot-pairwise-comp-Appendix}
+In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp-Appendix}
 \end{figure}

### 2021-06-07 Manuscripts/Manuscript by chapters -v2/04-Longitudinal analysis with GAMs.Rmd
Updated files with minor changes.

@@
 Significant differences exist where the interval does not cover 0.
-In both cases the effect of treatment is significant after day 5.
+In both cases the effect of treatment is significant after day 3.
 
@@
 With this correction in mind, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and $\approx$ 2 for the full dataset indicates that through that time, the "Control" group has higher $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by day 4 it is the "Treatment" group the one that has greater $\mbox{StO}_2$.
-This would suggest that the effect of chemotherapy in the "Treatment" group becomes significant after day 4 for the model used.
-Moreover, notice that although there is no actual measurement at day 4, the model is capable of providing an estimate of when the shift in $\mbox{StO}_2$ occurs.
+Notice that the shaded region between days 0 and $\approx$ 2 for the full dataset indicates that through that time, the "Control" group has higher $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by $\approx$ 3 day  it is the "Treatment" group the one that has greater $\mbox{StO}_2$.
+This would suggest that the effect of chemotherapy in the "Treatment" group becomes significant after day 3 for the model used.
+Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in $\mbox{StO}_2$ occurs.
 
 On the data with missing observations (Figure \@ref(fig:sim-smooth-plot), D), the confidence intervals of the smooths overlap between days 0 and 3.
-Consequently, the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp), B) shows that there is not a significant difference between the groups during that period, but is still able to pick the change on day 4 as the full dataset smooth pairwise comparison.
+Consequently, the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp), B) shows that there is not a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.
 

### 2021-06-07 Manuscripts/Manuscript by chapters -v2/05-Discussion.Rmd
Updated files with minor changes.

@@
 One of the features of GAMs is that they go beyond a mere _p-value_ to indicate differences between groups, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure  \@ref(fig:plot-pairwise-comp) indicate.
-The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( $\approx$ day 4 in Figure \@ref(fig:plot-pairwise-comp) A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
+The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( $\approx$ day 3 in Figure \@ref(fig:plot-pairwise-comp) A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
 

### 2021-06-07 Manuscripts/Manuscript by chapters -v2/00-Full_document.Rmd
Updated title.

@@
 ---
-title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
+title: '**Using generalized additive models to analyze biomedical non-linear longitudinal data**'
 subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_

### 2021-06-07 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
Updated title.

@@
 \hypersetup{
-  pdftitle={The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models},
+  pdftitle={Using generalized additive models to analyze biomedical non-linear longitudinal data},
   hidelinks,
@@
 
-\title{\textbf{The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models}}
+\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
 \usepackage{etoolbox}

### 2021-06-09 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
Updated title name, removed table of contents
Changed preamble.sty
Added ORCID in author info., corresponding author info.
Made changes suggested by JRT

@@
 \usepackage{authblk}
+\usepackage{hyperref}
+\usepackage{graphicx}
+
+%set a box to put the ORCID logo
+\newbox{\myorcidaffilbox}
+\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}
+
+%add hyperlink to the box
+\newcommand{\orcidaffila}[1]{%
+  \href{https://orcid.org/0000-0002-6014-4538}{\usebox{\myorcidaffilbox}}}
 
@@
 %authors
-\author[1]{Ariel I. Mundo}
+\author[1]{Ariel I. Mundo \orcidaffila{}}
 \author[2]{John R. Tipton}
-\author[1]{Timothy J. Muldoon}
+\author[1]{Timothy J. Muldoon \thanks{Corresponding author, tmuldoon@uark.edu}}
 \affil[1]{\footnotesize Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
@@
 
-\usepackage{xcolor} %theme colors for the code chunks (originally from latex-solarized on GitHub)
+%theme colors for the code chunks (originally from latex-solarized on GitHub)
+%https://github.com/jez/latex-solarized
+\usepackage{xcolor}
 \definecolor{sbase03}{HTML}{002B36}
@@
 
-{
-\setcounter{tocdepth}{2}
-\tableofcontents
-}
 \hypertarget{abstract}{%
@@
 As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
-
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
@@
 The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, \(\textsf{R}\){[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+In particular, \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
 At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
@@
 This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations is explained in detail.
+First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations are explained in detail.
 Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, using simulated data that reproduces patterns in previously reported studies {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} we illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
+Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
 The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
@@
 
-In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models by improving the standards for reproducibility in biomedical research.
+In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models to improve the standards for reproducibility in biomedical research.
 
@@
 
-\hypertarget{the-repeated-measures-anova}{%
-\subsection{The repeated measures ANOVA}\label{the-repeated-measures-anova}}
-
-The \emph{repeated measures analysis of variance} (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research.
-This statistical methodology requires certain assumptions for the model to be valid.
+\hypertarget{the-repeated-measures-anova-and-linear-mixed-model}{%
+\subsection{The repeated measures ANOVA and Linear Mixed Model}\label{the-repeated-measures-anova-and-linear-mixed-model}}
+
+The \emph{repeated measures analysis of variance} (rm-ANOVA) and the \emph{linear mixed model} (LMEM) are the most commonly used statistical analysis for longitudinal data in biomedical research.
+These statistical methodologies require certain assumptions for the model to be valid.
 From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
@@
 In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
-Independent errors \(\varepsilon_{tij}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
+Independent errors \(\varepsilon_{ijt}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
 In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy'').
 Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B).
-The linear models then can be expressed as
+With this notation, the linear model then can be expressed as
 
@@
 
-\hypertarget{the-linear-mixed-model-case}{%
-\subsubsection{The Linear Mixed Model Case}\label{the-linear-mixed-model-case}}
-
-A linear mixed model (LMEM) is a class of statistical model that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish {[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}.
+\hypertarget{the-linear-mixed-model-case-lmem}{%
+\subsubsection{The Linear Mixed Model Case (LMEM)}\label{the-linear-mixed-model-case-lmem}}
+
+A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish {[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}.
 A LMEM with interaction between time and treatment for a longitudinal study can be written as:
@@
 When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)).
-This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject within each group.
+This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject (\(subject_i\)) within each group (\(group_j\)).
 The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the ``global noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.
@@
 The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence,the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However,the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
+In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
+However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
 
@@
 To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\){[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme}{[}\protect\hyperlink{ref-nlme}{49}{]}.
-
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}, A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}, C).
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{49}{]}.
+
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}A).
+In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}C).
 In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments.
+This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments where a strong non-linear trend is present.
 
@@
 We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs.
-Of notice, Section \ref{longitudinal-GAMs} uses simulated data that does follow reported trends in the biomedical literature.
+Of notice, in Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
 
@@
 
-\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a LMEM and a rm-ANOVA model.
-A, D: Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
+\caption{Simulated responses from two groups with correlated errors using a LMEM and a rm-ANOVA model.
+Top row: linear response, bottom row: quadratic response.
+A: Simulated linear data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data.
+D: Simulated quadratic data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data.
 B,E: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
 Points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data but also assigns a global estimate that does not take between-subject variation.
-C, F: Estimates from the LMEM model in the linear and quadratic case.The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data in each group and grossly bias the initial estimates for each group.}\label{fig:l-q-response}
+The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation.
+C, F: Estimates from the LMEM in the linear and quadratic case.
+The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data and grossly bias the initial estimates for each group in the quadratic case (bottom row).}\label{fig:l-q-response}
 \end{figure}
 
-The simulation shows that the fit produced by the LMEM and the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response},B, E).
-When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.
-
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}, C.
-Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \ref{fig:l-q-response} A and B) .
-The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model \eqref{eq:linear-model} or a LMEM \eqref{eq:LMEM} to this data produces the fit that appears in panels E and F in Figure \ref{fig:l-q-response}.
-
-A comparison of the fitted mean response of the LMEM and the rm-ANOVA model to the simulated data in Figure ((\ref{fig:l-q-response}, E, F) indicates that the models are not capturing the changes within each group.
-Specifically, note that the fitted mean response of both models (panel E, F) show that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.
-The LMEM is only able to account for between-subject variation by providing different intercepts to each subject, but both models are not able to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare panel D with panels E and F).
-If such a change has important physiological implications,both rm-ANOVA and LMEMs omit it from the fitted mean response.
+The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response}A).
+When the linearity and compound symmetry assumptions are met, the rm\_ANOVA model approximates well the global trend by group (Figure \ref{fig:l-q-response}B).
+Note that because the LMEM incorporates \emph{random effects}, is able to provide estimates for each subject and a ``global'' estimate (Figure \ref{fig:l-q-response}C).
+
+However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}D.
+Here, the mean response per group was simulated using a quadratic function, and errors and individual responses were produced as in Figure \ref{fig:l-q-response}A.
+The mean response in the simulated data with quadratic behavior changes in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
+Fitting an rm-ANOVA model (Equation \eqref{eq:linear-model}) or a LMEM (Equation \eqref{eq:LMEM}) to this data produces the fit that appears in Figure \ref{fig:l-q-response}E, F.
+
+Comparing the fitted responses of the LMEM and the rm-ANOVA models used in the simulated quadratic data (Figure \ref{fig:l-q-response}E, F) indicates that the models are not capturing the changes within each group.
+Specifically, note that the fitted mean response of both models shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.
+The LMEM is only able to account for between-subject variation by providing estimates for each subject (Figure \ref{fig:l-q-response}F), but both models are unable to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare Figure \ref{fig:l-q-response}D with Figure \ref{fig:l-q-response}E, F).
+If such a change has important physiological implications, both rm-ANOVA and LMEMs omit it from the fitted mean response.
 Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
@@
 This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-Although the model fitted to the simulated data was an rm-ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM.
-In the following section, we present generalized additive models (GAMs) as a data-driven alternative method to analyze longitudinal non-linear data.
+The models fitted to the simulated data were an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
+In the following section, we present generalized additive models (GAMs) as a data-driven alternative method to analyze longitudinal non-linear data that overcomes the linearity assumption.
 
@@
 
-Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors{[}\protect\hyperlink{ref-nelder1972}{50}{]}.
+Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors {[}\protect\hyperlink{ref-nelder1972}{50}{]}.
 In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{51}{]}.
@@
 In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible \emph{smooth functions}.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
+This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
 One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
@@
 The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{52}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
-In the case of Equation \eqref{eq:GAM}, the basis functions are \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
+In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
@@
 
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\ref{fig:l-q-response}, C).
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-The timeline can be divided in equally spaced \emph{knots}, each knot being a region where a different basis function will be used.
+The timeline can be divided in equally spaced \emph{knots}, each knot being a region where a different set of basis functions will be used.
 Because there are six timepoints for this group, five knots can be used.
@@
 The choice of basis functions is set using default values in the package \emph{mgcv} depending on the number of knots.
-In Panel A of Figure \ref{fig:basis-plot}, the four basis functions (and the intercept) are shown.
+In Figure \ref{fig:basis-plot}A, the four basis functions (and the intercept) are shown.
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the ``wiggliness'' of the fit, each of the basis functions of Panel A is weighted by multiplying it by a coefficient according to the matrix of Panel B.
-The parameter estimates are penalized where the penalty reduces the ``wiggliness'' of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
-
-In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \ref{fig:basis-plot}, Panel C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for basis function 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1).
-On the other hand, basis function 3 has roughly doubled its value.
+To control the ``wiggliness'' of the fit, each of the basis functions of Figure \ref{fig:basis-plot}A is weighted by multiplying it by a coefficient according to the matrix of Figure \ref{fig:basis-plot}B.
+The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the ``wiggliness'' of the smooth fit to prevent overfitting.
+A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
+
+To get the weighted basis functions, each basis (from Figure Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
+Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions.
+Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1).
+On the other hand, the third basis function has roughly doubled its value.
 Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of \emph{time} is shown in Panel D (orange line) along the simulated values per group, which appear as points.
+The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (orange line), along the simulated values per group, which appear as points.
 
@@
 \caption{Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Matrix for basis function weights.
+A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
+B: Matrix of basis function weights.
 Each basis function is multiplied by a coefficient which can be positive or negative.
@@
 C: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
+Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
+Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
 D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
+The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
 \end{figure}
+
+\hypertarget{a-bayesian-interpretation-of-gams}{%
+\section{A Bayesian interpretation of GAMs}\label{a-bayesian-interpretation-of-gams}}
+
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}\protect\hyperlink{ref-mcelreath2018}{54}{]}.
+In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) {[}\protect\hyperlink{ref-miller2019}{55}{]}.
+Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{56}{]}.
+Therefore, the confidence intervals calculated for the smooth terms calculated by the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (pointwise coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+This last part means that the estimated confidence intervals for the smooths will contain the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated).
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}\protect\hyperlink{ref-marra2012}{57}{]}.
+Because in the next section we refer to the confidence intervals for the GAMs as ``empirical Bayesian,'' this brief introduction to the Bayesian interpretation of GAMs has been provided as a necessary step to understand the subsequent terminology.
 
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3, C in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
 In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
@@
 Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
-The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}, A and the inlet, respectively.
+The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}A and the inset, respectively.
 
@@
 
-\passthrough{\lstinline!m1 <- gam(StO2\_sim \~ Group + s(Day, by=Group, k=5), method='REML',data=dat\_sim)!}
-
-This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths for \passthrough{\lstinline!Group!} and \passthrough{\lstinline!Day!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
+\passthrough{\lstinline!m1 <- gam(StO2\_sim \~ Group + s(Day, by=Group, k=5), method='REML, data = dat\_sim)!}
+
+This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
 The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-The parametric term \passthrough{\lstinline!Group!} is added to quantify differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot},B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{54}{]}.
-A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{55}{]}.
-
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}, C.
-This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \ref{fig:sim-smooth-plot}, B) .
+The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{58}{]}.
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{59}{]}.
+
+One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
+This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
 
@@
 However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
-Consider the simulated \(\mbox{StO}_2\) values from Figure (\ref{fig:sim-smooth-plot}, B).
-If 40\% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but it can be seen that the smooths overlap during the first 3 days because with less data points, the trend is less pronounced than in the full dataset (\ref{fig:sim-smooth-plot}, D).
+Consider the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}B.
+If 40\% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \ref{fig:sim-smooth-plot}D).
 Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.
@@
 Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
+C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
 D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.}\label{fig:sim-smooth-plot}
+Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot}
 \end{figure}
@@
 B: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the interval does not cover 0.
+Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
 In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp}
@@
 The absence of a \emph{p-value} in this case might seem odd, but the empirical Bayesian confidence interval interpretation can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
-This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}, A where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
+This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
 
@@
 
-Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}, B and D.
-Figure \ref{fig:plot-pairwise-comp}, shows the comparison between each treatment group for the full and missing datasets.
+Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}B and Figure \ref{fig:sim-smooth-plot}D.
+Figure \ref{fig:plot-pairwise-comp} shows the comparison between each treatment group for the full and missing datasets.
 Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared.
@@
 With this correction in mind, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by \(\approx\) 3 day it is the ``Treatment'' group the one that has greater \(\mbox{StO}_2\).
-This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the model used.
-Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in \(\mbox{StO}_2\) occurs.
-
-On the data with missing observations (Figure \ref{fig:sim-smooth-plot}, D), the confidence intervals of the smooths overlap between days 0 and 3.
-Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}, B) shows that there is not a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.
+Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by \(\approx\) 3 day it is the ``Treatment'' group the one that on average, has greater \(\mbox{StO}_2\).
+This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the given model.
+Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in mean \(\mbox{StO}_2\) occurs.
+
+On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the empirical Bayesian credible confidence intervals of the smooths overlap between days 0 and 3.
+Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that there is no evidence of a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.
 
@@
 This ``model misspecification'' error, also is known as a ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{56}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{60}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
 
 In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{57}{]}.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{61}{]}.
 This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{58},\protect\hyperlink{ref-weissgerber2018}{59}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{62},\protect\hyperlink{ref-weissgerber2018}{63}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 
+\leavevmode\hypertarget{ref-mcelreath2018}{}%
+\CSLLeftMargin{{[}54{]} }
+\CSLRightInline{R. McElreath, {Statistical rethinking: A Bayesian course with examples in R and Stan}, {Chapman and Hall/CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
+
+\leavevmode\hypertarget{ref-miller2019}{}%
+\CSLLeftMargin{{[}55{]} }
+\CSLRightInline{D.L. Miller, Bayesian views of generalized additive modelling, arXiv Preprint arXiv:1902.01330. (2019).}
+
+\leavevmode\hypertarget{ref-laird1982}{}%
+\CSLLeftMargin{{[}56{]} }
+\CSLRightInline{N.M. Laird, J.H. Ware, Random-effects models for longitudinal data, Biometrics. 38 (1982) 963--974. \url{http://www.jstor.org/stable/2529876}.}
+
+\leavevmode\hypertarget{ref-marra2012}{}%
+\CSLLeftMargin{{[}57{]} }
+\CSLRightInline{G.
+Marra, S.N.
+Wood, Coverage properties of confidence intervals for generalized additive model components, {Scandinavian Journal of Statistics}.
+{39} (2012) 53--74.
+\url{https://doi.org/\%7B10.1111/j.1467-9469.2011.00760.x\%7D}.}
+
 \leavevmode\hypertarget{ref-gratia}{}%
-\CSLLeftMargin{{[}54{]} }
+\CSLLeftMargin{{[}58{]} }
 \CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
@@
 \leavevmode\hypertarget{ref-harezlak2018}{}%
-\CSLLeftMargin{{[}55{]} }
+\CSLLeftMargin{{[}59{]} }
 \CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, {Semiparametric Regression with R}, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
@@
 \leavevmode\hypertarget{ref-lang2015}{}%
-\CSLLeftMargin{{[}56{]} }
+\CSLLeftMargin{{[}60{]} }
 \CSLRightInline{T.A.
@@
 \leavevmode\hypertarget{ref-hastie1995}{}%
-\CSLLeftMargin{{[}57{]} }
+\CSLLeftMargin{{[}61{]} }
 \CSLRightInline{T.
@@
 \leavevmode\hypertarget{ref-begley2015}{}%
-\CSLLeftMargin{{[}58{]} }
+\CSLLeftMargin{{[}62{]} }
 \CSLRightInline{C.G.
@@
 \leavevmode\hypertarget{ref-weissgerber2018}{}%
-\CSLLeftMargin{{[}59{]} }
+\CSLLeftMargin{{[}63{]} }
 \CSLRightInline{T.L.
@@
 C: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
+Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
+Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
 D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot-appendix}
+The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot-appendix}
 \end{figure}
@@
 
-\hypertarget{a-basic-workflow-for-gams}{%
-\subsection{A basic Workflow for GAMs}\label{a-basic-workflow-for-gams}}
+\hypertarget{workflow}{%
+\subsection{A basic Workflow for GAMs}\label{workflow}}
 
@@
     my_list<-pairwise_limits(comp_StO2)
-
+  rib_col<-'#EDD03AFF' #color for the ribbon
 #plot the difference
@@
                 alpha = 0.5,
-                fill='#DB3A07FF') +
+                fill=rib_col) +
     geom_line(color='black',size=1) +
@@
 This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\passthrough{\lstinline!m1!}), linear model and GAM on data with missing observations are presented.
-Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inlet are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
+Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inset are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
 
@@
 Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
+C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
 D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
+Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
 \end{figure}
@@
 my_list<-pairwise_limits(comp_StO2_full)
+rib_col<-'#EDD03AFF'
 
@@
                 alpha = 0.5,
-                fill='#DB3A07FF') +
+                fill=rib_col) +
     geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
@@
                 alpha = 0.5,
-                fill='#DB3A07FF') +
+                fill=rib_col) +
     geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
@@
 B: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the interval does not cover 0.
+Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
 In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp-Appendix}

### 2021-06-09 Manuscripts/Manuscript by chapters -v2/02-Challenges.Rmd
Made changes suggested by JRT

@@
 
-## The repeated measures ANOVA
+## The repeated measures ANOVA and Linear Mixed Model
 
-The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research.
-This statistical methodology requires certain assumptions for the model to be valid.
+The _repeated measures analysis of variance_ (rm-ANOVA) and the _linear mixed model_ (LMEM) are the most commonly used statistical analysis for longitudinal data in biomedical research.
+These statistical methodologies require certain assumptions for the model to be valid.
 From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
@@
 In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$ (independently and identically normally distributed with mean zero and variance $\sigma^2$).
+Independent errors $\varepsilon_{ijt}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$ (independently and identically normally distributed with mean zero and variance $\sigma^2$).
 In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
@@
 Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B).
-The linear models then can be expressed as
+With this notation, the linear model then can be expressed as
 
@@
 
-### The Linear Mixed Model Case
+### The Linear Mixed Model Case (LMEM)
 
-A linear mixed model (LMEM) is a class of statistical model that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to distinguish [@pinheiro2006;@west2014].
+A LMEM is a class of statistical models that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to distinguish [@pinheiro2006;@west2014].
 A LMEM with interaction between time and treatment for a longitudinal study  can be written as:
@@
 When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
-This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject within each group.
+This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject ($subject_i$) within each group ($group_j$).
 The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the "global noise" term $\varepsilon_{ijt}$ from Equation \@ref(eq:linear-model).
@@
 The _random effect_ $\mu_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
-In essence,the _random effect_ in a LMEM enables to fit models with different slopes at the subject-level[@pinheiro2006].
-However,the expected linear relationship of the covariates and the response in Equation \@ref(eq:linear-model) and in Equation \@ref(eq:LMEM) is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
+In essence, the _random effect_ in a LMEM enables to fit models with different slopes at the subject-level[@pinheiro2006].
+However, the expected linear relationship of the covariates and the response in Equation \@ref(eq:linear-model) and in Equation \@ref(eq:LMEM) is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
 
@@
 To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \@ref(eq:linear-model)), and a LMEM  (Equation \@ref(eq:LMEM)) are fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme].
+An rm-ANOVA model (Equation \@ref(eq:linear-model)), and a LMEM  (Equation \@ref(eq:LMEM)) are fitted to each group, using $\textsf{R}$ [@r] and the package _nlme_ [@nlme].
 
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C).
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response)A).
+In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response)C).
 In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments.
+This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments where a strong non-linear trend is present.
 
@@
 We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs.
-Of notice, Section \@ref(longitudinal-GAMs) uses simulated data that does follow reported trends in the biomedical literature.
+Of notice, in Section \@ref(longitudinal-GAMs) we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
 
@@
 
-(ref:l-q-response-caption) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a LMEM and a rm-ANOVA model.
-A, D: Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data.
+(ref:l-q-response-caption) Simulated responses from two groups with correlated errors using a LMEM and a rm-ANOVA model.
+Top row: linear response, bottom row: quadratic response.
+A: Simulated linear data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data.
+D: Simulated quadratic data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data.
 B,E: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
 Points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data but also assigns a global estimate that does not take between-subject variation.
-C, F: Estimates from the LMEM model in the linear and quadratic case.The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data in each group and grossly bias the initial estimates for each group.
+The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation.
+C, F: Estimates from the LMEM in the linear and quadratic case.
+The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data and grossly bias the initial estimates for each group in the quadratic case (bottom row).
 
@@
 
-The simulation shows that the fit produced by the LMEM and the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B, E).
-When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.
+The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response)A).
+When the linearity and compound symmetry assumptions are met, the rm_ANOVA model approximates well the global trend by group (Figure \@ref(fig:l-q-response)B).
+Note that because the LMEM incorporates _random effects_, is able to provide estimates for each subject and a "global" estimate (Figure \@ref(fig:l-q-response)C).
 
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response), C.
-Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \@ref(fig:l-q-response) A and B) .
-The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model \@ref(eq:linear-model) or a LMEM \@ref(eq:LMEM) to this data  produces the fit that appears in panels E and F in Figure \@ref(fig:l-q-response).
+However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response)D.
+Here, the mean response per group was simulated using a quadratic function, and errors and individual responses were produced as in Figure \@ref(fig:l-q-response)A.
+The mean response in the simulated data with quadratic behavior changes in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
+Fitting an rm-ANOVA model (Equation \@ref(eq:linear-model)) or a LMEM (Equation \@ref(eq:LMEM)) to this data  produces the fit that appears in Figure \@ref(fig:l-q-response)E, F.
 
-A comparison of the fitted mean response of the LMEM and the rm-ANOVA model to the simulated data in Figure ((\@ref(fig:l-q-response), E, F) indicates that the models are not capturing the changes within each group.
-Specifically, note that the fitted mean response of both models (panel E, F) show that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.
-The LMEM is only able to account for between-subject variation by providing different intercepts to each subject, but both models are not able to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panel D with panels E and F).
-If such a change has important physiological implications,both rm-ANOVA and LMEMs omit it from the fitted mean response.
+Comparing the fitted responses of the LMEM and the rm-ANOVA models used in the simulated quadratic data (Figure \@ref(fig:l-q-response)E, F) indicates that the models are not capturing the changes within each group.
+Specifically, note that the fitted mean response of both models shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.
+The LMEM is only able to account for between-subject variation by providing estimates for each subject (Figure \@ref(fig:l-q-response)F), but both models are unable to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare Figure \@ref(fig:l-q-response)D  with Figure \@ref(fig:l-q-response)E, F).
+If such a change has important physiological implications, both rm-ANOVA and LMEMs omit it from the fitted mean response.
 Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
@@
 This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-Although the model fitted to the simulated data was an rm-ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM.
-In the following section, we present generalized additive models (GAMs) as a data-driven  alternative method to analyze longitudinal non-linear data.
+The models fitted to the simulated data were an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
+In the following section, we present generalized additive models (GAMs) as a data-driven  alternative method to analyze longitudinal non-linear data that overcomes the linearity assumption.

### 2021-06-09 Manuscripts/Manuscript by chapters -v2/03-GAM Theory.Rmd
Made changes suggested by JRT

@@
 
-Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors[@nelder1972].
+Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors [@nelder1972].
 In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case[@simpson2018;@wood2017;@hastie1987].
@@
 In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response.
+This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
 One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with inference [@beck1998].
@@
 The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
-In the case of Equation \@ref(eq:GAM), the basis functions are $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
+In the case of Equation \@ref(eq:GAM), the basis functions are contained in the expression $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
 
@@
 
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\@ref(fig:l-q-response), C).
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \@ref(fig:l-q-response)C.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-The timeline can be divided in equally spaced _knots_, each knot being a region where a different basis function will be used.
+The timeline can be divided in equally spaced _knots_, each knot being a region where a different set of basis functions will be used.
 Because there are six timepoints for this group, five knots can be used.
@@
 The choice of basis functions is set using default values in the package _mgcv_ depending on the number of knots.
-In Panel A of Figure \@ref(fig:basis-plot), the four basis functions (and the intercept) are shown.
+In Figure \@ref(fig:basis-plot)A, the four basis functions (and the intercept) are shown.
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the "wiggliness" of the fit, each of the basis functions of Panel A is weighted by multiplying it by a coefficient according to the matrix of Panel B.
-The parameter estimates are penalized where the penalty reduces the "wiggliness" of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
+To control the "wiggliness" of the fit, each of the basis functions of Figure \@ref(fig:basis-plot)A is weighted by multiplying it by a coefficient according to the matrix of Figure \@ref(fig:basis-plot)B.
+The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the "wiggliness" of the smooth fit to prevent overfitting.
+A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
 
@@
     scale_fill_gradient(low = "white",
-                        high = "#B50A2AFF")+ #color picked from KikiMedium
+                        high = "#B50A2AFF")+ 
   labs(x='Basis',
@@
 
-In other words, the six points of each basis are multiplied by the corresponding coefficient in panel B, thereby increasing or decreasing the original basis functions of Panel A.
-In Figure \@ref(fig:basis-plot), Panel C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for basis function 1 has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1).
-On the other hand, basis function 3 has roughly doubled its value.
+To get the weighted basis functions, each basis (from Figure Figure \@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients in Figure \@ref(fig:basis-plot)B, thereby increasing or decreasing the original basis functions.
+Figure \@ref(fig:basis-plot)C shows the resulting weighted basis functions.
+Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1).
+On the other hand, the third basis function has roughly doubled its value.
 Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of _time_ is shown in Panel D (orange line) along the simulated values per group, which appear as points.
+The resulting smooth term for the effect of _time_ is shown in Figure \@ref(fig:basis-plot)D (orange line), along the simulated values per group, which appear as points.
 
 (ref:basis-plot-caption) Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown.
-B: Matrix for basis function weights.
+A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
+B: Matrix of basis function weights.
 Each basis function is multiplied by a coefficient which can be positive or negative.
@@
 C: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis.
+Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
+Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
 D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each penalized basis function at each time point, with simulated values for the group shown as points.
+The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.
 
@@
 ```
+
+# A Bayesian interpretation of GAMs
+
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief [@mcelreath2018].
+In the case of non-linear data, the belief that the _true_ trend of the data is likely to be smooth rather than "wiggly" introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs [@wood2017].
+GAMs are considered "empirical" Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the "Full Bayes" case) [@miller2019].
+Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model [@laird1982;@pedersen2019].
+Therefore, the confidence intervals calculated for the smooth terms calculated by the package _mgcv_ are considered empirical Bayesian posterior credible intervals [@pedersen2019], which have good "frequentist" coverage (pointwise coverage), and _across the function_ coverage [@wood2017].
+This last part means that the estimated confidence intervals for the smooths will contain the true function of the data 95% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated).
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in [@miller2019;@wood2017 and @marra2012].
+Because in the next section we refer to the confidence intervals for the GAMs as "empirical Bayesian", this brief introduction to the Bayesian interpretation of GAMs has been provided as a necessary step to understand the subsequent terminology.

### 2021-06-09 Manuscripts/Manuscript by chapters -v2/04-Longitudinal analysis with GAMs.Rmd
Made changes suggested by JRT

@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors that appear in Figure 3, C in [@vishwanath2009].
+The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors that appear in Figure 3C in [@vishwanath2009].
 In the paper, diffuse reflectance spectroscopy was used to quantify $\mbox{StO}_2$ changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
@@
 Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (matching the SD in the original paper).
-The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot), A and the inlet, respectively.
+The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot)A and the inset, respectively.
 
@@
 
-`m1 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML',data=dat_sim)`
+`m1 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML, data = dat_sim)`
 
-This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths for `Group` and `Day` (the parenthesis preceded by `s`) using 5 knots.
+This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using 5 knots.
 The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths [@simpson2018].
-The parametric term `Group` is added to quantify differences in the effect of treatment between groups, and the `method` chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot),B).
+The parametric term `Group` is added to quantify overall mean differences in the effect of treatment between groups, and the `method` chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
+When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot)B).
 Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_ [@gratia].
-A guide for model selection and diagnostics is in the Appendix, and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].
+A guide for model selection and diagnostics is in the [Appendix](#workflow), and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].
 
 One question that might arise at this point is "what is the fit that an rm-ANOVA model produces for the simulated data?".
-The rm-ANOVA model, which corresponds to Equation \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot), C.
-This is a typical case of model misspecification: The slopes of each group are different, which would lead to a _p-value_ indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to do so (Figure \@ref(fig:sim-smooth-plot), B) .
+The rm-ANOVA model, which corresponds to Equation \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot)C.
+This is a typical case of model misspecification: The slopes of each group are different, which would lead to a _p-value_ indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \@ref(fig:sim-smooth-plot)B) .
 
@@
 However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
-Consider the  simulated $\mbox{StO}_2$ values from Figure (\@ref(fig:sim-smooth-plot), B).
-If 40% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but it can be seen that the smooths overlap during the first 3 days because with less data points, the trend is less pronounced than in the full dataset (\@ref(fig:sim-smooth-plot), D).
+Consider the  simulated $\mbox{StO}_2$ values from Figure \@ref(fig:sim-smooth-plot)B.
+If 40% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \@ref(fig:sim-smooth-plot)D).
 Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.
@@
 Lines represent trends for each group, shaded regions are 95% confidence intervals.
-C: rm-ANOVA model for the simulated data, the model does not capture the changes in each group over time.
+C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
 D: Smooths for the GAM model for the simulated data with 40% of its observations missing.
-Lines represent trends for each group, shaded regions are 95% confidence intervals.
+Lines represent trends for each group, shaded regions are 95% empirical Bayesian confidence intervals.
 
@@
 my_list<-pairwise_limits(comp_StO2_full)
+rib_col<-'#EDD03AFF' #color for ribbon
 
@@
 B: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the interval does not cover 0.
+Significant differences exist where the 95% empirical Bayesian credible interval does not cover 0.
 In both cases the effect of treatment is significant after day 3.
@@
 The absence of a _p-value_ in this case might seem odd, but the empirical Bayesian confidence interval interpretation can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
-This is what happens for the simulated data in Figure \@ref(fig:sim-smooth-plot), A where the chemotherapy causes $\mbox{StO}_2$ to increase over time.
+This is what happens for the simulated data in Figure \@ref(fig:sim-smooth-plot)A, where the chemotherapy causes $\mbox{StO}_2$ to increase over time.
 
@@
 
-Consider the calculation of pairwise differences for the smooths in Figure \@ref(fig:sim-smooth-plot), B and D.
-Figure \@ref(fig:plot-pairwise-comp), shows the comparison between each treatment group for the full and missing datasets.
+Consider the calculation of pairwise differences for the smooths in Figure \@ref(fig:sim-smooth-plot)B and Figure \@ref(fig:sim-smooth-plot)D.
+Figure \@ref(fig:plot-pairwise-comp) shows the comparison between each treatment group for the full and missing datasets.
 Here, the  "Control" group is used as the reference to which "Treatment" group is being compared.
@@
 With this correction in mind, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and $\approx$ 2 for the full dataset indicates that through that time, the "Control" group has higher $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by $\approx$ 3 day  it is the "Treatment" group the one that has greater $\mbox{StO}_2$.
-This would suggest that the effect of chemotherapy in the "Treatment" group becomes significant after day 3 for the model used.
-Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in $\mbox{StO}_2$ occurs.
+Notice that the shaded region between days 0 and $\approx$ 2 for the full dataset indicates that through that time, the "Control" group has higher mean $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by $\approx$ 3 day  it is the "Treatment" group the one that on average, has greater $\mbox{StO}_2$.
+This would suggest that the effect of chemotherapy in the "Treatment" group becomes significant after day 3 for the given model.
+Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in mean $\mbox{StO}_2$ occurs.
 
-On the data with missing observations (Figure \@ref(fig:sim-smooth-plot), D), the confidence intervals of the smooths overlap between days 0 and 3.
-Consequently, the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp), B) shows that there is not a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.
+On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the empirical Bayesian credible confidence intervals of the smooths overlap between days 0 and 3.
+Consequently, the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that there is no evidence of a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.
 

### 2021-06-09 Manuscripts/Manuscript by chapters -v2/07-Appendix.Rmd
Made changes suggested by JRT

--- 2021-06-04 Manuscripts/Manuscript by chapters -v2/07-Appendix.Rmd
@@
 
-## A basic Workflow for GAMs
+## A basic Workflow for GAMs{#workflow}
 
@@
 This section covers the code used to generate Figure \@ref(fig:sim-smooth-plot), where the simulated data, fit of the "final" GAM (`m1`), linear model  and GAM on data with missing observations are presented.
-Note that panel A in Figure \@ref(fig:sim-smooth-plot) and the inlet are generated in the code chunk where the data is simulated in Section \@ref(tumor-data-simulation), and are called later to build the figure.
+Note that panel A in Figure \@ref(fig:sim-smooth-plot) and the inset are generated in the code chunk where the data is simulated in Section \@ref(tumor-data-simulation), and are called later to build the figure.
 
@@
              x=6,
-             y=-15,
-             label="Treatment",
-             size=10)+
+             y=-18,
+             label="Treatment>Control",
+             size=8)+
     geom_ribbon(aes(ymin = lower, ymax = upper),
                 alpha = 0.5,
-                fill='#DB3A07FF') +
+                fill=rib_col) +
     geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+

### 2021-06-09 Code and miscellaneous/Pairwise comparisons for smooths.R

@@ -146,7 +146,7 @@ return(my_list)
 
 
 my_list<-pairwise_limits(comp_StO2_full)
-
+rib_col<-'#EDD03AFF'
 c2<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
     annotate("rect",
                 xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
@@ -168,7 +168,7 @@ c2<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
              size=10)+
     geom_ribbon(aes(ymin = lower, ymax = upper),
                 alpha = 0.9,
-                fill='black') +
+                fill=rib_col) +
     geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
     geom_line(color='black',size=1) +
     facet_wrap(~ pair) +

### 2021-06-10 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
Final changes added, ready for bioRxiv

@@
 
+\newcommand{\orcidaffilb}[1]{%
+  \href{https://orcid.org/0000-0002-6135-8191}{\usebox{\myorcidaffilbox}}}
+
 %command for the package lineno
@@
 \author[1]{Ariel I. Mundo \orcidaffila{}}
-\author[2]{John R. Tipton}
+\author[2]{John R. Tipton \orcidaffilb{}}
 \author[1]{Timothy J. Muldoon \thanks{Corresponding author, tmuldoon@uark.edu}}
@@
 When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)).
-This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject (\(subject_i\)) within each group (\(group_j\)).
+This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)).
 The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the ``global noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.
@@
 The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response}A).
-When the linearity and compound symmetry assumptions are met, the rm\_ANOVA model approximates well the global trend by group (Figure \ref{fig:l-q-response}B).
+When the linearity and compound symmetry assumptions are met, the rm-ANOVA model approximates well the global trend by group (Figure \ref{fig:l-q-response}B).
 Note that because the LMEM incorporates \emph{random effects}, is able to provide estimates for each subject and a ``global'' estimate (Figure \ref{fig:l-q-response}C).
@@
 Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{56}{]}.
-Therefore, the confidence intervals calculated for the smooth terms calculated by the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (pointwise coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-This last part means that the estimated confidence intervals for the smooths will contain the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated).
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}\protect\hyperlink{ref-marra2012}{57}{]}.
-Because in the next section we refer to the confidence intervals for the GAMs as ``empirical Bayesian,'' this brief introduction to the Bayesian interpretation of GAMs has been provided as a necessary step to understand the subsequent terminology.
+Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{55}{]} and {[}\protect\hyperlink{ref-marra2012}{57}{]}.
+With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
 
 \FloatBarrier
-
-\newpage
 
@@
 
-On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the empirical Bayesian credible confidence intervals of the smooths overlap between days 0 and 3.
+On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the empirical Bayesian credible intervals of the smooths overlap between days 0 and 3.
 Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that there is no evidence of a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.
@@
 However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} using \(\textsf{R}\) and the package \emph{mgcv}{[}\protect\hyperlink{ref-wood2017}{37}{]} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the Appendix.
-One of the features of GAMs is that they go beyond a mere \emph{p-value} to indicate differences between groups, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} using \(\textsf{R}\) and the package \emph{mgcv}{[}\protect\hyperlink{ref-wood2017}{37}{]} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
+One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
 The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.

### 2021-06-10 Manuscripts/Manuscript by chapters -v2/02-Challenges.Rmd
Final changes added, ready for bioRxiv

@@
 When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
-This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject ($subject_i$) within each group ($group_j$).
+This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject (subject$_i$) within each group (group$_j$).
 The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the "global noise" term $\varepsilon_{ijt}$ from Equation \@ref(eq:linear-model).
@@
 The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response)A).
-When the linearity and compound symmetry assumptions are met, the rm_ANOVA model approximates well the global trend by group (Figure \@ref(fig:l-q-response)B).
+When the linearity and compound symmetry assumptions are met, the rm-ANOVA model approximates well the global trend by group (Figure \@ref(fig:l-q-response)B).
 Note that because the LMEM incorporates _random effects_, is able to provide estimates for each subject and a "global" estimate (Figure \@ref(fig:l-q-response)C).

### 2021-06-10 Manuscripts/Manuscript by chapters -v2/03-GAM Theory.Rmd
Final changes added, ready for bioRxiv

@@
 Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model [@laird1982;@pedersen2019].
-Therefore, the confidence intervals calculated for the smooth terms calculated by the package _mgcv_ are considered empirical Bayesian posterior credible intervals [@pedersen2019], which have good "frequentist" coverage (pointwise coverage), and _across the function_ coverage [@wood2017].
-This last part means that the estimated confidence intervals for the smooths will contain the true function of the data 95% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated).
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in [@miller2019;@wood2017 and @marra2012].
-Because in the next section we refer to the confidence intervals for the GAMs as "empirical Bayesian", this brief introduction to the Bayesian interpretation of GAMs has been provided as a necessary step to understand the subsequent terminology.
+Therefore, the confidence intervals calculated for the smooth terms using the package _mgcv_ are considered empirical Bayesian posterior credible intervals [@pedersen2019], which have good "frequentist" coverage (pointwise coverage or "single point" coverage), and _across the function_ coverage [@wood2017].
+This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain _on average_ the true function of the data 95% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in [@miller2019;@wood2017;@simpson2018] and [@marra2012].
+With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as "empirical Bayesian" through the rest of this paper.

### 2021-06-10 Manuscripts/Manuscript by chapters -v2/04-Longitudinal analysis with GAMs.Rmd
Final changes added, ready for bioRxiv

@@
-\newpage
-
 # The analyisis of longitudinal biomedical data using GAMs{#longitudinal-GAMs}
@@
 
-On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the empirical Bayesian credible confidence intervals of the smooths overlap between days 0 and 3.
+On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the empirical Bayesian credible intervals of the smooths overlap between days 0 and 3.
 Consequently, the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that there is no evidence of a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.

### 2021-06-10 Manuscripts/Manuscript by chapters -v2/00-Full_document.Rmd
Updated abstract

@@
     fig_caption: yes #allows captions in figures
+    toc: false #do not include table of contents
     extra_dependencies:
@@
 In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs).
-Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
+Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
+
 In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
@@
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and use simulated data that follows trends reported in the biomedical literature to demonstrate how these models are implemented in $\textsf{R}$ via the package _mgcv_, showing that GAMs are able to produce estimates that are consistent with the trends of non-linear data even if the case when missing observations exist.
+We also present the basic theory of GAMs, and using trends of oxygen saturation reported in the biomedical literature we simulate longitudinal data (2 treatment groups, 10 subjects per group, 6 repeated measures for each group) to demonstrate how these models are implemented in $\textsf{R}$ via the package _mgcv_.
+We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the observations missing), allowing reliable inference from the data.
 To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research.

### 2021-06-10 Manuscripts/Manuscript by chapters -v2/00-Full_document.Rmd
Updated abstract.

@@
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and using trends of oxygen saturation reported in the biomedical literature we simulate longitudinal data (2 treatment groups, 10 subjects per group, 6 repeated measures for each group) to demonstrate how these models are implemented in $\textsf{R}$ via the package _mgcv_.
+We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 6 repeated measures for each group) to demonstrate how these models can be computationally implemented.
 We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the observations missing), allowing reliable inference from the data.

### 2021-06-10 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
Updated abstract.

@@
 In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs).
-Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
+Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
+
 In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
@@
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and use simulated data that follows trends reported in the biomedical literature to demonstrate how these models are implemented in \(\textsf{R}\) via the package \emph{mgcv}, showing that GAMs are able to produce estimates that are consistent with the trends of non-linear data even if the case when missing observations exist.
+We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 6 repeated measures for each group) to demonstrate how these models can be computationally implemented.
+We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40\% of the observations missing), allowing reliable inference from the data.
 To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.

### 2021-06-11 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
Fixed minor error in Abstract.

@@
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 6 repeated measures for each group) to demonstrate how these models can be computationally implemented.
-We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40\% of the observations missing), allowing reliable inference from the data.
+We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented.
+We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing), allowing reliable inference from the data.
 To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.

### 2021-06-18 Manuscripts/Manuscript by chapters -for submission/01-Background.Rmd
References do not work

--- 2021-05-27 Manuscripts/Manuscript by chapters -v2/01-Background.Rmd
@@
 The current advances in programming languages designed for statistical analysis (specifically  $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, $\textsf{R}$[@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
+In particular, $\textsf{R}$ [@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
 hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
@@
 This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations is explained in detail.
+First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations are explained in detail.
 Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, using simulated data that reproduces patterns in previously reported studies [@vishwanath2009] we illustrate the type of non-linear longitudinal data that often occurs in biomedical research.
+Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies [@vishwanath2009].
 The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
@@
 
-In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models by improving the standards for reproducibility  in biomedical research.
+In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models to improve the standards for reproducibility  in biomedical research.

### 2021-06-18 Manuscripts/Manuscript by chapters -for submission/06-Conclusion.Rmd
References do not work

--- 2021-06-10 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Using generalized additive models to analyze biomedical non-linear longitudinal data},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
-\usepackage{listings}
-\newcommand{\passthrough}[1]{#1}
-\lstset{defaultdialect=[5.3]Lua}
-\lstset{defaultdialect=[x86masm]Assembler}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-\usepackage{lineno}
-\usepackage{authblk}
-\usepackage{hyperref}
-\usepackage{graphicx}
-
-%set a box to put the ORCID logo
-\newbox{\myorcidaffilbox}
-\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}
-
-%add hyperlink to the box
-\newcommand{\orcidaffila}[1]{%
-  \href{https://orcid.org/0000-0002-6014-4538}{\usebox{\myorcidaffilbox}}}
-
-\newcommand{\orcidaffilb}[1]{%
-  \href{https://orcid.org/0000-0002-6135-8191}{\usebox{\myorcidaffilbox}}}
-
-%command for the package lineno
-\linenumbers
-
-%authors
-\author[1]{Ariel I. Mundo \orcidaffila{}}
-\author[2]{John R. Tipton \orcidaffilb{}}
-\author[1]{Timothy J. Muldoon \thanks{Corresponding author, tmuldoon@uark.edu}}
-\affil[1]{\footnotesize Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
-\affil[2]{\footnotesize Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
-
-%theme colors for the code chunks (originally from latex-solarized on GitHub)
-%https://github.com/jez/latex-solarized
-\usepackage{xcolor}
-\definecolor{sbase03}{HTML}{002B36}
-\definecolor{sbase02}{HTML}{073642}
-\definecolor{sbase01}{HTML}{586E75}
-\definecolor{sbase00}{HTML}{657B83}
-\definecolor{sbase0}{HTML}{839496}
-\definecolor{sbase1}{HTML}{93A1A1}
-\definecolor{sbase2}{HTML}{EEE8D5}
-\definecolor{sbase3}{HTML}{FDF6E3}
-\definecolor{syellow}{HTML}{B58900}
-\definecolor{sorange}{HTML}{CB4B16}
-\definecolor{sred}{HTML}{DC322F}
-\definecolor{smagenta}{HTML}{D33682}
-\definecolor{sviolet}{HTML}{6C71C4}
-\definecolor{sblue}{HTML}{268BD2}
-\definecolor{scyan}{HTML}{2AA198}
-\definecolor{sgreen}{HTML}{859900}
-%command to set parameter(s) in package listings
-\lstset{
-    % How/what to match
-    sensitive=true,
-    % Border (above and below)
-    frame=lines,
-    % Extra margin on line (align with paragraph)
-    xleftmargin=\parindent,
-    % Put extra space under caption
-    belowcaptionskip=1\baselineskip,
-    % Colors
-    backgroundcolor=\color{sbase3},
-    basicstyle=\color{sbase00}\ttfamily,
-    keywordstyle=\color{scyan},
-    commentstyle=\color{sbase1},
-    stringstyle=\color{sblue},
-    numberstyle=\color{sviolet},
-    identifierstyle=\color{sbase00},
-    % Break long lines into multiple lines?
-    breaklines=true,
-    % Show a character for spaces?
-    showstringspaces=false,
-    tabsize=2
-}
-
-%\lstset{
-%  breaklines=true,
-%  stringstyle=\ttfamily,
-%  backgroundcolor=\color{gray}
-%}
-\usepackage{placeins}
-\usepackage{subfig}
-\usepackage{breqn}
-\usepackage[font={small}]{caption}
-\usepackage{float}
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
-\usepackage{etoolbox}
-\makeatletter
-\providecommand{\subtitle}[1]{% add subtitle to \maketitle
-  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
-}
-\makeatother
-\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
-\date{\vspace{-2.5em}}
-
-\begin{document}
-\maketitle
-
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
-
-In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs).
-Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear.
-In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
-Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
-This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and use simulated data that follows trends reported in the biomedical literature to demonstrate how these models are implemented in \(\textsf{R}\) via the package \emph{mgcv}, showing that GAMs are able to produce estimates that are consistent with the trends of non-linear data even if the case when missing observations exist.
-To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
-
-\hypertarget{background}{%
-\section{Background}\label{background}}
-
-Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response {[}\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}{]}, antibody expression {[}\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}{]}, and cell metabolism {[}\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}\protect\hyperlink{ref-greening2018}{9}{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}\protect\hyperlink{ref-sio2016}{10}{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}\protect\hyperlink{ref-kamstra2015}{11}{]}.
-
-Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}\protect\hyperlink{ref-wagenmakers2008}{12}{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}{]}.
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
-
-A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
-Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}.
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}\protect\hyperlink{ref-wang2019}{18}{]}.
-
-Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}{]}, consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}\protect\hyperlink{ref-abdi2010}{21}{]}), but they in turn reduce statistical power (1-\(\beta\)){[}\protect\hyperlink{ref-nakagawa2004}{22}{]}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) {[}\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}{]}.
-Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and group sizes {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}\protect\hyperlink{ref-charan2013}{30}{]} which might cause the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible.
-It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}\protect\hyperlink{ref-gueorguieva2004}{13}{]}.
-This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
-At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
-
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}.
-And even more importantly, LMEMs also assume a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}, making them unsuitable to analyze non-linear data.
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
-In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]} .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]} and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
-
-The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
-Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
-
-This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations are explained in detail.
-Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in \(\textsf{R}\), in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
-
-In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models to improve the standards for reproducibility in biomedical research.
-
-\FloatBarrier
-
-\hypertarget{challenges-presented-by-longitudinal-studies}{%
-\section{Challenges presented by longitudinal studies}\label{challenges-presented-by-longitudinal-studies}}
-
-\hypertarget{the-repeated-measures-anova-and-linear-mixed-model}{%
-\subsection{The repeated measures ANOVA and Linear Mixed Model}\label{the-repeated-measures-anova-and-linear-mixed-model}}
-
-The \emph{repeated measures analysis of variance} (rm-ANOVA) and the \emph{linear mixed model} (LMEM) are the most commonly used statistical analysis for longitudinal data in biomedical research.
-These statistical methodologies require certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
-Each one of these assumptions is discussed below.
-
-\hypertarget{linear-relationship}{%
-\subsection{Linear relationship}\label{linear-relationship}}
-
-\hypertarget{the-repeated-measures-anova-case}{%
-\subsubsection{The repeated measures ANOVA case}\label{the-repeated-measures-anova-case}}
-
-In a longitudinal biomedical study, two or more groups of subjects (e.g., human subject, mice, samples) are subject to different treatments (e.g., a ``treatment'' group receives a novel drug or intervention vs.~a ``control'' group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response is modeled with \emph{fixed} components.
-The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug/intervention in the ``treatment'' group.
-
-Mathematically speaking, a rm-ANOVA model with an interaction can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
-\label{eq:linear-model}
-\end{equation}
-
-In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
-Independent errors \(\varepsilon_{ijt}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
-In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy'').
-Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B).
-With this notation, the linear model then can be expressed as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-\label{eq:ANOVA-by-group}
-\end{equation}
-
-To further simplify the expression, substitute \(\widetilde{\beta_{0}}=\beta_0+\beta_{2}\) and \(\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}\) in the equation for Group B.
-This substitution allows for a different intercept and slope for Groups A and B.
-The model is then written as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
-\end{cases}
-\label{eq:ANOVA-lines}
-\end{equation}
-
-Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response, this means that either presented as Equation \eqref{eq:linear-model}, Equation \eqref{eq:ANOVA-by-group} or Equation \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
-
-\hypertarget{the-linear-mixed-model-case-lmem}{%
-\subsubsection{The Linear Mixed Model Case (LMEM)}\label{the-linear-mixed-model-case-lmem}}
-
-A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish {[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}.
-A LMEM with interaction between time and treatment for a longitudinal study can be written as:
-
-\begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
-\label{eq:LMEM}
-\end{equation}
-
-When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)).
-This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)).
-The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the ``global noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.
-
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data.
-As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
-The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
-
-\hypertarget{covariance-in-rm-anova-and-lmems}{%
-\subsection{Covariance in rm-ANOVA and LMEMs}\label{covariance-in-rm-anova-and-lmems}}
-
-In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{41}{]} (For an in-depth analysis of the covariance matrix see {[}\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}{]}).
-
-In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}{]}.
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}\protect\hyperlink{ref-maxwell2017}{44}{]}; and it is higher between consecutive measurements {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and they are not robust if the group sizes are unbalanced {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
-Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
-
-\hypertarget{missing-observations}{%
-\subsection{Missing observations}\label{missing-observations}}
-
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}\protect\hyperlink{ref-weiss2005}{42}{]}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{46}{]}.
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{47}{]}, and in the case of MNAR the missing values are dependent on their value.
-
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{48}{]}.
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}\protect\hyperlink{ref-west2014}{40}{]}.
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
-
-\hypertarget{simulation}{%
-\subsection{What do an rm-ANOVA and LMEM fit look like? A visual representation using simulated data}\label{simulation}}
-
-To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{49}{]}.
-
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}C).
-In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments where a strong non-linear trend is present.
-
-Specifically, the rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect.
-In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group.
-From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D).
-
-A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal appears in Figure \ref{fig:linear-cases-Appendix} and Figure \ref{fig:quadratic-cases-Appendix} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs.
-Of notice, in Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/l-q-response-1}
-
-}
-
-\caption{Simulated responses from two groups with correlated errors using a LMEM and a rm-ANOVA model.
-Top row: linear response, bottom row: quadratic response.
-A: Simulated linear data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data.
-D: Simulated quadratic data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data.
-B,E: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
-Points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation.
-C, F: Estimates from the LMEM in the linear and quadratic case.
-The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data and grossly bias the initial estimates for each group in the quadratic case (bottom row).}\label{fig:l-q-response}
-\end{figure}
-
-The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response}A).
-When the linearity and compound symmetry assumptions are met, the rm-ANOVA model approximates well the global trend by group (Figure \ref{fig:l-q-response}B).
-Note that because the LMEM incorporates \emph{random effects}, is able to provide estimates for each subject and a ``global'' estimate (Figure \ref{fig:l-q-response}C).
-
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}D.
-Here, the mean response per group was simulated using a quadratic function, and errors and individual responses were produced as in Figure \ref{fig:l-q-response}A.
-The mean response in the simulated data with quadratic behavior changes in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model (Equation \eqref{eq:linear-model}) or a LMEM (Equation \eqref{eq:LMEM}) to this data produces the fit that appears in Figure \ref{fig:l-q-response}E, F.
-
-Comparing the fitted responses of the LMEM and the rm-ANOVA models used in the simulated quadratic data (Figure \ref{fig:l-q-response}E, F) indicates that the models are not capturing the changes within each group.
-Specifically, note that the fitted mean response of both models shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.
-The LMEM is only able to account for between-subject variation by providing estimates for each subject (Figure \ref{fig:l-q-response}F), but both models are unable to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare Figure \ref{fig:l-q-response}D with Figure \ref{fig:l-q-response}E, F).
-If such a change has important physiological implications, both rm-ANOVA and LMEMs omit it from the fitted mean response.
-Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
-
-This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-The models fitted to the simulated data were an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
-In the following section, we present generalized additive models (GAMs) as a data-driven alternative method to analyze longitudinal non-linear data that overcomes the linearity assumption.
-
-\FloatBarrier
-
-\hypertarget{GAM-theory}{%
-\section{GAMs as a special case of Generalized Linear Models}\label{GAM-theory}}
-
-\hypertarget{gams-and-basis-functions}{%
-\subsection{GAMs and Basis Functions}\label{gams-and-basis-functions}}
-
-Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors {[}\protect\hyperlink{ref-nelder1972}{50}{]}.
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{51}{]}.
-A GAM model can be written as:
-
-\begin{equation}
-  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
-  \label{eq:GAM}
-\end{equation}
-
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible \emph{smooth functions}.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
-In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
-
-The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{52}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
-In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
-
-Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
-Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-The timeline can be divided in equally spaced \emph{knots}, each knot being a region where a different set of basis functions will be used.
-Because there are six timepoints for this group, five knots can be used.
-The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is set using default values in the package \emph{mgcv} depending on the number of knots.
-In Figure \ref{fig:basis-plot}A, the four basis functions (and the intercept) are shown.
-Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the ``wiggliness'' of the fit, each of the basis functions of Figure \ref{fig:basis-plot}A is weighted by multiplying it by a coefficient according to the matrix of Figure \ref{fig:basis-plot}B.
-The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the ``wiggliness'' of the smooth fit to prevent overfitting.
-A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
-
-To get the weighted basis functions, each basis (from Figure Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
-Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1).
-On the other hand, the third basis function has roughly doubled its value.
-Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (orange line), along the simulated values per group, which appear as points.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-1}
-
-}
-
-\caption{Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-B: Matrix of basis function weights.
-Each basis function is multiplied by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
-Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
-\end{figure}
-
-\hypertarget{a-bayesian-interpretation-of-gams}{%
-\section{A Bayesian interpretation of GAMs}\label{a-bayesian-interpretation-of-gams}}
-
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}\protect\hyperlink{ref-mcelreath2018}{54}{]}.
-In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) {[}\protect\hyperlink{ref-miller2019}{55}{]}.
-Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{56}{]}.
-Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{55}{]} and {[}\protect\hyperlink{ref-marra2012}{57}{]}.
-With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
-
-\FloatBarrier
-
-\hypertarget{longitudinal-GAMs}{%
-\section{The analyisis of longitudinal biomedical data using GAMs}\label{longitudinal-GAMs}}
-
-The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs.
-This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}.
-The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
-
-\hypertarget{simulated-data}{%
-\subsection{Simulated data}\label{simulated-data}}
-
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
-In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group.
-Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
-The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}A and the inset, respectively.
-
-\hypertarget{an-interaction-gam-for-longitudinal-data}{%
-\subsection{An interaction GAM for longitudinal data}\label{an-interaction-gam-for-longitudinal-data}}
-
-An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination.
-In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time.
-Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in {[}\protect\hyperlink{ref-hefley2017}{52}{]}.
-
-For the data in Figure \ref{fig:sim-smooth-plot}, A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
-To estimate this, the model incorporates independent smooths for \emph{Group} and \emph{Day}, respectively.
-The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
-Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:
-
-\passthrough{\lstinline!m1 <- gam(StO2\_sim \~ Group + s(Day, by=Group, k=5), method='REML, data = dat\_sim)!}
-
-This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
-The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{58}{]}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{59}{]}.
-
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
-This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
-
-Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
-The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
-However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
-Consider the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}B.
-If 40\% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \ref{fig:sim-smooth-plot}D).
-Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-1}
-
-}
-
-\caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot}
-\end{figure}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-1}
-
-}
-
-\caption{Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-B: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
-In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp}
-\end{figure}
-
-\hypertarget{GAM-significance}{%
-\subsection{Determination of significance in GAMs for longitudinal data}\label{GAM-significance}}
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the \emph{difference} between the empirical Bayesian confidence intervals of the fitted smooths for such groups is non-zero, then a significant difference exists at that time point(s).
-The absence of a \emph{p-value} in this case might seem odd, but the empirical Bayesian confidence interval interpretation can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
-This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
-
-With this expectation of different trends in each group, computing the difference between the trends will identify if the observed change is significant.
-The difference between groups with similar trends is likely to yield zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
-
-Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}B and Figure \ref{fig:sim-smooth-plot}D.
-Figure \ref{fig:plot-pairwise-comp} shows the comparison between each treatment group for the full and missing datasets.
-Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared.
-Of notice, the pairwise comparison has been set on the response scale (see Appendix for code details), because otherwise the comparison appears shifted and is not intuitively easy to relate to the original data.
-
-With this correction in mind, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by \(\approx\) 3 day it is the ``Treatment'' group the one that on average, has greater \(\mbox{StO}_2\).
-This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the given model.
-Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in mean \(\mbox{StO}_2\) occurs.
-
-On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the empirical Bayesian credible intervals of the smooths overlap between days 0 and 3.
-Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that there is no evidence of a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.
-
-In a sense, the pairwise smooth comparison is more informative than a \emph{post-hoc} \emph{p-value}.
-For biomedical studies, the smooth comparison is able to provide an estimate of \emph{when} and by \emph{how much} a biological process becomes significant.
-This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.
-
-\FloatBarrier
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
-Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}.
-This ``model misspecification'' error, also is known as a ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{60}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
-
-In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{61}{]}.
-This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
-
-However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} using \(\textsf{R}\) and the package \emph{mgcv}{[}\protect\hyperlink{ref-wood2017}{37}{]} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
-One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
-The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
-
-We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{62},\protect\hyperlink{ref-weissgerber2018}{63}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
-Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
-We have made all the data and code used in this paper accessible, and we hope that this will encourage other researchers to do the same with future projects.
-
-\FloatBarrier
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
+# Conclusion
 
@@
 
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
+# Acknowledgements
 
 This work was supported by the National Science Foundation Career Award (CBET 1751554, TJM) and the Arkansas Biosciences Institute.
-
-\FloatBarrier
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\leavevmode\hypertarget{ref-roblyer2011}{}%
-\CSLLeftMargin{{[}1{]} }
-\CSLRightInline{D.
-Roblyer, S.
-Ueda, A.
-Cerussi, W.
-Tanamai, A.
-Durkin, R.
-Mehta, D.
-Hsiang, J.A.
-Butler, C.
-McLaren, W.P.
-Chen, B.
-Tromberg, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences of the United States of America.
-108 (2011) 14626--14631.
-\url{https://doi.org/10.1073/pnas.1013103108}.}
-
-\leavevmode\hypertarget{ref-tank2020}{}%
-\CSLLeftMargin{{[}2{]} }
-\CSLRightInline{A.
-Tank, H.M.
-Peterson, V.
-Pera, S.
-Tabassum, A.
-Leproux, T.
-O'Sullivan, E.
-Jones, H.
-Cabral, N.
-Ko, R.S.
-Mehta, B.J.
-Tromberg, D.
-Roblyer, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research.
-22 (2020) 1--10.
-\url{https://doi.org/doi:10.1186/s13058-020-01262-1}.}
-
-\leavevmode\hypertarget{ref-pavlov2018}{}%
-\CSLLeftMargin{{[}3{]} }
-\CSLRightInline{M.V.
-Pavlov, T.I.
-Kalganova, Y.S.
-Lyubimtseva, V.I.
-Plekhanov, G.Y.
-Golubyatnikov, O.Y.
-Ilyinskaya, A.G.
-Orlova, P.V.
-Subochev, D.V.
-Safonov, N.M.
-Shakhova, A.V.
-Maslennikova, {Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy}, {Journal of Biomedical Optics}.
-{23} (2018).
-\url{https://doi.org/\%7B10.1117/1.JBO.23.9.091410\%7D}.}
-
-\leavevmode\hypertarget{ref-demidov2018}{}%
-\CSLLeftMargin{{[}4{]} }
-\CSLRightInline{V.
-Demidov, A.
-Maeda, M.
-Sugita, V.
-Madge, S.
-Sadanand, C.
-Flueraru, I.A.
-Vitkin, {Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography}, {Scientific Reports}.
-{8} (2018).
-\url{https://doi.org/\%7B10.1038/s41598-017-18635-w\%7D}.}
-
-\leavevmode\hypertarget{ref-ritter2001}{}%
-\CSLLeftMargin{{[}5{]} }
-\CSLRightInline{G.
-Ritter, L.
-Cohen, C.
-Williams, E.
-Richards, L.
-Old, S.
-Welt, {Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}, {Cancer Research}.
-{61} (2001) 6851--6859.}
-
-\leavevmode\hypertarget{ref-roth2017}{}%
-\CSLLeftMargin{{[}6{]} }
-\CSLRightInline{E.M.
-Roth, A.C.
-Goldberg, A.L.
-Catapano, A.
-Torri, G.D.
-Yancopoulos, N.
-Stahl, A.
-Brunet, G.
-Lecorps, H.M.
-Colhoun, {Antidrug antibodies in atients treated with alirocumab}, {New England Journal of Medicine}.
-{376} (2017) 1589--1590.
-\url{https://doi.org/\%7B10.1056/NEJMc1616623\%7D}.}
-
-\leavevmode\hypertarget{ref-jones2018}{}%
-\CSLLeftMargin{{[}7{]} }
-\CSLRightInline{J.D.
-Jones, H.E.
-Ramser, A.E.
-Woessner, K.P.
-Quinn, {In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing}, {Communications Biology}.
-{1} (2018).
-\url{https://doi.org/\%7B10.1038/s42003-018-0206-4\%7D}.}
-
-\leavevmode\hypertarget{ref-skala2010}{}%
-\CSLLeftMargin{{[}8{]} }
-\CSLRightInline{M.C.
-Skala, A.
-Fontanella, L.
-Lan, J.A.
-Izatt, M.W.
-Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics.
-15 (2010).
-\url{https://doi.org/10.1117/1.3285584}.}
-
-\leavevmode\hypertarget{ref-greening2018}{}%
-\CSLLeftMargin{{[}9{]} }
-\CSLRightInline{G.J.
-Greening, K.P.
-Miller, C.R.
-Spainhour, M.D.
-Cato, T.J.
-Muldoon, {Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy}, {Biomedical Optics Express}.
-{9} (2018) 2871--2886.
-\url{https://doi.org/\%7B10.1364/BOE.9.002871\%7D}.}
-
-\leavevmode\hypertarget{ref-sio2016}{}%
-\CSLLeftMargin{{[}10{]} }
-\CSLRightInline{T.T.
-Sio, P.J.
-Atherton, B.J.
-Birckhead, D.J.
-Schwartz, J.A.
-Sloan, D.K.
-Seisler, J.A.
-Martenson, C.L.
-Loprinzi, P.C.
-Griffin, R.F.
-Morton, J.C.
-Anders, T.J.
-Stoffel, R.E.
-Haselow, R.B.
-Mowat, M.A.N.
-Wittich, J.D.
-Bearden III, R.C.
-Miller, {Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {{[}}alliance{]})}, {Supportive Care in Cancer}.
-{24} (2016) 3847--3855.
-\url{https://doi.org/\%7B10.1007/s00520-016-3213-3\%7D}.}
-
-\leavevmode\hypertarget{ref-kamstra2015}{}%
-\CSLLeftMargin{{[}11{]} }
-\CSLRightInline{J.I.
-Kamstra, P.U.
-Dijkstra, M.
-van Leeuwen, J.L.N.
-Roodenburg, J.A.
-Langendijk, {Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study}, {Oral Oncology}.
-{51} (2015) 548--555.
-\url{https://doi.org/\%7B10.1016/j.oraloncology.2015.01.016\%7D}.}
-
-\leavevmode\hypertarget{ref-wagenmakers2008}{}%
-\CSLLeftMargin{{[}12{]} }
-\CSLRightInline{E.-J.
-Wagenmakers, M.
-Lee, T.
-Lodewyckx, G.J.
-Iverson, Bayesian versus frequentist inference, in: H.
-Hoijtink, I.
-Klugkist, P.A.
-Boelen (Eds.), Bayesian Evaluation of Informative Hypotheses, Springer New York, New York, NY, 2008: pp.
-181--207.
-\url{https://doi.org/10.1007/978-0-387-09612-4_9}.}
-
-\leavevmode\hypertarget{ref-gueorguieva2004}{}%
-\CSLLeftMargin{{[}13{]} }
-\CSLRightInline{R.
-Gueorguieva, J.H.
-Krystal, {Move over ANOVA - Progress in analyzing repeated-measures data and its reflection in papers published in the archives of general psychiatry}, Archives of General Psychiatry.
-61 (2004) 310--317.
-\url{https://doi.org/10.1001/archpsyc.61.3.310}.}
-
-\leavevmode\hypertarget{ref-schober2018}{}%
-\CSLLeftMargin{{[}14{]} }
-\CSLRightInline{P.
-Schober, T.R.
-Vetter, Repeated measures designs and analysis of longitudinal data: If at first you do not succeed-try, try again, Anesthesia and Analgesia.
-127 (2018) 569--575.
-\url{https://doi.org/10.1213/ane.0000000000003511}.}
-
-\leavevmode\hypertarget{ref-pinheiro2006}{}%
-\CSLLeftMargin{{[}15{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science \& Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}
-
-\leavevmode\hypertarget{ref-vishwanath2009}{}%
-\CSLLeftMargin{{[}16{]} }
-\CSLRightInline{K.
-Vishwanath, H.
-Yuan, W.T.
-Barry, M.W.
-Dewhirst, N.
-Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia.
-11 (2009) 889--900.
-\url{https://doi.org/10.1593/neo.09580}.}
-
-\leavevmode\hypertarget{ref-dennis2019}{}%
-\CSLLeftMargin{{[}17{]} }
-\CSLRightInline{B.
-Dennis, J.M.
-Ponciano, M.L.
-Taper, S.R.
-Lele, {Errors in statistical inference under model misspecification: evidence, hypothesis testing, and AIC}, {Frontiers in Ecology and Evolution}.
-{7} (2019).
-\url{https://doi.org/\%7B10.3389/fevo.2019.00372\%7D}.}
-
-\leavevmode\hypertarget{ref-wang2019}{}%
-\CSLLeftMargin{{[}18{]} }
-\CSLRightInline{B.
-Wang, Z.
-Zhou, H.
-Wang, X.M.
-Tu, C.
-Feng, {The p-value and model specification in statistics}, {General Psychiatry}.
-{32} (2019).
-\url{https://doi.org/\%7B10.1136/gpsych-2019-100081\%7D}.}
-
-\leavevmode\hypertarget{ref-liu2010}{}%
-\CSLLeftMargin{{[}19{]} }
-\CSLRightInline{C.
-Liu, T.P.
-Cripe, M.-O.
-Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy.
-18 (2010) 1724--1730.
-\url{https://doi.org/10.1038/mt.2010.127}.}
-
-\leavevmode\hypertarget{ref-halsey2015}{}%
-\CSLLeftMargin{{[}20{]} }
-\CSLRightInline{L.G.
-Halsey, D.
-Curran-Everett, S.L.
-Vowler, G.B.
-Drummond, {The fickle \(p\) value generates irreproducible results}, {Nature Methods}.
-{12} (2015) 179--185.
-\url{https://doi.org/\%7B10.1038/nmeth.3288\%7D}.}
-
-\leavevmode\hypertarget{ref-abdi2010}{}%
-\CSLLeftMargin{{[}21{]} }
-\CSLRightInline{H. Abdi, {Holm's sequential Bonferroni procedure}, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.}
-
-\leavevmode\hypertarget{ref-nakagawa2004}{}%
-\CSLLeftMargin{{[}22{]} }
-\CSLRightInline{S.
-Nakagawa, {A farewell to Bonferroni: the problems of low statistical power and publication bias}, {Behavioral Ecology}.
-{15} (2004) 1044--1045.
-\url{https://doi.org/\%7B10.1093/beheco/arh107\%7D}.}
-
-\leavevmode\hypertarget{ref-gelman2012}{}%
-\CSLLeftMargin{{[}23{]} }
-\CSLRightInline{A.
-Gelman, J.
-Hill, M.
-Yajima, {Why we (usually) don't have to worry about multiple comparisons}, {Journal of Research on Educational Effectiveness}.
-{5} (2012) 189--211.
-\url{https://doi.org/\%7B10.1080/19345747.2011.618213\%7D}.}
-
-\leavevmode\hypertarget{ref-albers2019}{}%
-\CSLLeftMargin{{[}24{]} }
-\CSLRightInline{C. Albers, {The problem with unadjusted multiple and sequential statistical testing}, {Nature Communications}. {10} (2019). \url{https://doi.org/\%7B10.1038/s41467-019-09941-0\%7D}.}
-
-\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
-\CSLLeftMargin{{[}25{]} }
-\CSLRightInline{C.
-Ugrinowitsch, G.W.
-Fellingham, M.D.
-Ricard, Limitations of ordinary least squares models in analyzing repeated measures data, Medicine and Science in Sports and Exercise.
-36 (2004) 2144--2148.
-\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.}
-
-\leavevmode\hypertarget{ref-huynh1976}{}%
-\CSLLeftMargin{{[}26{]} }
-\CSLRightInline{H.
-Huynh, L.S.
-Feldt, {Estimation of the Box correction for degrees of freedom from sample data in randomized block and split-plot designs}, Journal of Educational Statistics.
-1 (1976) 69--82.
-\url{https://doi.org/10.3102/10769986001001069}.}
-
-\leavevmode\hypertarget{ref-greenhouse1959}{}%
-\CSLLeftMargin{{[}27{]} }
-\CSLRightInline{S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.}
-
-\leavevmode\hypertarget{ref-haverkamp2017}{}%
-\CSLLeftMargin{{[}28{]} }
-\CSLRightInline{N.
-Haverkamp, A.
-Beauducel, {Violation of the sphericity assumption and its effect on type-I error rates in repeated measures ANOVA and multi-level linear models (MLM)}, {Frontiers in Psychology}.
-{8} (2017).
-\url{https://doi.org/\%7B10.3389/fpsyg.2017.01841\%7D}.}
-
-\leavevmode\hypertarget{ref-keselman2001}{}%
-\CSLLeftMargin{{[}29{]} }
-\CSLRightInline{H.
-Keselman, J.
-Algina, R.
-Kowalchuk, {The analysis of repeated measures designs: A review}, {British Journal of Mathematica \& Statistical Psychology}.
-{54} (2001) 1--20.
-\url{https://doi.org/\%7B10.1348/000711001159357\%7D}.}
-
-\leavevmode\hypertarget{ref-charan2013}{}%
-\CSLLeftMargin{{[}30{]} }
-\CSLRightInline{Jaykaran.
-Charan, N.
-Kantharia, {How to calculate sample size in animal studies?}, Journal of Pharmacology and Pharmacotherapeutics.
-4 (2013) 303--306.
-\url{https://doi.org/10.4103/0976-500X.119726}.}
-
-\leavevmode\hypertarget{ref-barr2013}{}%
-\CSLLeftMargin{{[}31{]} }
-\CSLRightInline{D.J.
-Barr, R.
-Levy, C.
-Scheepers, H.J.
-Tily, {Random effects structure for confirmatory hypothesis testing: Keep it maximal}, {Journal of Memory and Language}.
-{68} (2013) 255--278.
-\url{https://doi.org/\%7B10.1016/j.jml.2012.11.001\%7D}.}
-
-\leavevmode\hypertarget{ref-rose2012}{}%
-\CSLLeftMargin{{[}32{]} }
-\CSLRightInline{N.L.
-Rose, H.D.
-Yang, S.D.
-Turner, G.L.
-Simpson, {An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to Scotland, UK}, Geochimica Et Cosmochimica Acta.
-82 (2012) 113--135.
-\url{https://doi.org/10.1016/j.gca.2010.12.026}.}
-
-\leavevmode\hypertarget{ref-pedersen2019}{}%
-\CSLLeftMargin{{[}33{]} }
-\CSLRightInline{E.J.
-Pedersen, D.L.
-Miller, G.L.
-Simpson, N.
-Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, {PeerJ}.
-7 (2019).
-\url{https://doi.org/10.7717/peerj.6876}.}
-
-\leavevmode\hypertarget{ref-simpson2018}{}%
-\CSLLeftMargin{{[}34{]} }
-\CSLRightInline{G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.}
-
-\leavevmode\hypertarget{ref-yang2012}{}%
-\CSLLeftMargin{{[}35{]} }
-\CSLRightInline{L.
-Yang, G.
-Qin, N.
-Zhao, C.
-Wang, G.
-Song, {Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality}, {BMC Medical Research Methodology}.
-{12} (2012).
-\url{https://doi.org/\%7B10.1186/1471-2288-12-165\%7D}.}
-
-\leavevmode\hypertarget{ref-beck1998}{}%
-\CSLLeftMargin{{[}36{]} }
-\CSLRightInline{N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. (1998) 596--627.}
-
-\leavevmode\hypertarget{ref-wood2017}{}%
-\CSLLeftMargin{{[}37{]} }
-\CSLRightInline{S.N. Wood, {Generalized additive models: An introduction with R, Second Edition}, CRC Press LLC, Philadelphia, PA, 2017.}
-
-\leavevmode\hypertarget{ref-r}{}%
-\CSLLeftMargin{{[}38{]} }
-\CSLRightInline{R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.}
-
-\leavevmode\hypertarget{ref-wood2016}{}%
-\CSLLeftMargin{{[}39{]} }
-\CSLRightInline{S.N.
-Wood, N.
-Pya, B.
-Saefken, Smoothing parameter and model selection for general smooth models, {Journal of the American Statistical Association}.
-{111} (2016) 1548--1563.
-\url{https://doi.org/\%7B10.1080/01621459.2016.1180986\%7D}.}
-
-\leavevmode\hypertarget{ref-west2014}{}%
-\CSLLeftMargin{{[}40{]} }
-\CSLRightInline{B.T.
-West, K.B.
-Welch, A.T.
-Galecki, Linear mixed models: A practical guide using statistical software, second edition, Taylor \& Francis, 2014.
-\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.}
-
-\leavevmode\hypertarget{ref-wolfinger1996}{}%
-\CSLLeftMargin{{[}41{]} }
-\CSLRightInline{R.D.
-Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics.
-1 (1996) 205--230.
-\url{http://www.jstor.org/stable/1400366}.}
-
-\leavevmode\hypertarget{ref-weiss2005}{}%
-\CSLLeftMargin{{[}42{]} }
-\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://books.google.com/books?id=MQ/_bvWDPsEAC}.}
-
-\leavevmode\hypertarget{ref-geisser1958}{}%
-\CSLLeftMargin{{[}43{]} }
-\CSLRightInline{S.
-Geisser, S.W.
-Greenhouse, {An extension of Box's results on the use of the \(F\) distribution in multivariate analysis}, The Annals of Mathematical Statistics.
-29 (1958) 885--891.
-\url{https://doi.org/10.1214/aoms/1177706545}.}
-
-\leavevmode\hypertarget{ref-maxwell2017}{}%
-\CSLLeftMargin{{[}44{]} }
-\CSLRightInline{S.E.
-Maxwell, H.D.
-Delaney, K.
-Kelley, Designing experiments and analyzing data: A model comparison perspective, third edition, Taylor \& Francis, 2017.
-\url{https://books.google.com/books?id=NmFQDwAAQBAJ}.}
-
-\leavevmode\hypertarget{ref-molenberghs2004}{}%
-\CSLLeftMargin{{[}45{]} }
-\CSLRightInline{G.
-Molenberghs, H.
-Thijs, I.
-Jansen, C.
-Beunckens, M.
-Kenward, C.
-Mallinckrodt, R.
-Carroll, {Analyzing incomplete longitudinal clinical trial data}, {Biostatistics}.
-{5} (2004) 445--464.
-\url{https://doi.org/\%7B10.1093/biostatistics/kxh001\%7D}.}
-
-\leavevmode\hypertarget{ref-scheffer2002}{}%
-\CSLLeftMargin{{[}46{]} }
-\CSLRightInline{J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.}
-
-\leavevmode\hypertarget{ref-potthoff2006}{}%
-\CSLLeftMargin{{[}47{]} }
-\CSLRightInline{R.F.
-Potthoff, G.E.
-Tudor, K.S.
-Pieper, V.
-Hasselblad, {Can one assess whether missing data are missing at random in medical studies?}, {Statistical Methods in Medical Research}.
-{15} (2006) 213--234.
-\url{https://doi.org/\%7B10.1191/0962280206sm448oa\%7D}.}
-
-\leavevmode\hypertarget{ref-ma2012}{}%
-\CSLLeftMargin{{[}48{]} }
-\CSLRightInline{Y.
-Ma, M.
-Mazumdar, S.G.
-Memtsoudis, Beyond repeated-measures analysis of variance advanced statistical methods for the analysis of longitudinal data in anesthesia research, {Regional Anesthesia and Pain Medicine}.
-{37} (2012) 99--105.
-\url{https://doi.org/\%7B10.1097/AAP.0b013e31823ebc74\%7D}.}
-
-\leavevmode\hypertarget{ref-nlme}{}%
-\CSLLeftMargin{{[}49{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, {nlme}: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.}
-
-\leavevmode\hypertarget{ref-nelder1972}{}%
-\CSLLeftMargin{{[}50{]} }
-\CSLRightInline{J.A.
-Nelder, R.W.M.
-Wedderburn, Generalized linear models, Journal of the Royal Statistical Society.
-Series A (General).
-135 (1972) 370--384.
-\url{http://www.jstor.org/stable/2344614}.}
-
-\leavevmode\hypertarget{ref-hastie1987}{}%
-\CSLLeftMargin{{[}51{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association.
-82 (1987) 371--386.
-\url{https://doi.org/10.1080/01621459.1987.10478440}.}
-
-\leavevmode\hypertarget{ref-hefley2017}{}%
-\CSLLeftMargin{{[}52{]} }
-\CSLRightInline{T.J.
-Hefley, K.M.
-Broms, B.M.
-Brost, F.E.
-Buderman, S.L.
-Kay, H.R.
-Scharf, J.R.
-Tipton, P.J.
-Williams, M.B.
-Hooten, {The basis function approach for modeling autocorrelation in ecological data}, {Ecology}.
-{98} (2017) 632--646.
-\url{https://doi.org/\%7B10.1002/ecy.1674\%7D}.}
-
-\leavevmode\hypertarget{ref-wegman1983}{}%
-\CSLLeftMargin{{[}53{]} }
-\CSLRightInline{E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.}
-
-\leavevmode\hypertarget{ref-mcelreath2018}{}%
-\CSLLeftMargin{{[}54{]} }
-\CSLRightInline{R. McElreath, {Statistical rethinking: A Bayesian course with examples in R and Stan}, {Chapman and Hall/CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
-
-\leavevmode\hypertarget{ref-miller2019}{}%
-\CSLLeftMargin{{[}55{]} }
-\CSLRightInline{D.L. Miller, Bayesian views of generalized additive modelling, arXiv Preprint arXiv:1902.01330. (2019).}
-
-\leavevmode\hypertarget{ref-laird1982}{}%
-\CSLLeftMargin{{[}56{]} }
-\CSLRightInline{N.M. Laird, J.H. Ware, Random-effects models for longitudinal data, Biometrics. 38 (1982) 963--974. \url{http://www.jstor.org/stable/2529876}.}
-
-\leavevmode\hypertarget{ref-marra2012}{}%
-\CSLLeftMargin{{[}57{]} }
-\CSLRightInline{G.
-Marra, S.N.
-Wood, Coverage properties of confidence intervals for generalized additive model components, {Scandinavian Journal of Statistics}.
-{39} (2012) 53--74.
-\url{https://doi.org/\%7B10.1111/j.1467-9469.2011.00760.x\%7D}.}
-
-\leavevmode\hypertarget{ref-gratia}{}%
-\CSLLeftMargin{{[}58{]} }
-\CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
-
-\leavevmode\hypertarget{ref-harezlak2018}{}%
-\CSLLeftMargin{{[}59{]} }
-\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, {Semiparametric Regression with R}, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
-
-\leavevmode\hypertarget{ref-lang2015}{}%
-\CSLLeftMargin{{[}60{]} }
-\CSLRightInline{T.A.
-Lang, D.G.
-Altman, {Basic statistical reporting for articles published in Biomedical Journals: The ``Statistical Analyses and Methods in the Published Literature{''} or the SAMPL Guidelines}, INTERNATIONAL JOURNAL OF NURSING STUDIES.
-{52} (2015) 5--9.
-\url{https://doi.org/\%7B10.1016/j.ijnurstu.2014.09.006\%7D}.}
-
-\leavevmode\hypertarget{ref-hastie1995}{}%
-\CSLLeftMargin{{[}61{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models for medical research, {Statistical Methods in Medical Research}.
-4 (1995) 187--196.
-\url{https://doi.org/10.1177/096228029500400302}.}
-
-\leavevmode\hypertarget{ref-begley2015}{}%
-\CSLLeftMargin{{[}62{]} }
-\CSLRightInline{C.G.
-Begley, J.P.A.
-Ioannidis, {Reproducibility in Science Improving the Standard for Basic and Preclinical Research}, {Circulation Research}.
-{116} (2015) 116--126.
-\url{https://doi.org/\%7B10.1161/CIRCRESAHA.114.303819\%7D}.}
-
-\leavevmode\hypertarget{ref-weissgerber2018}{}%
-\CSLLeftMargin{{[}63{]} }
-\CSLRightInline{T.L.
-Weissgerber, O.
-Garcia-Valencia, V.D.
-Garovic, N.M.
-Milic, S.J.
-Winham, {Meta-Research: Why we need to report more than 'Data were Analyzed by t-tests or ANOVA'}, Elife.
-7 (2018) e36163.
-\url{https://doi.org/10.7554/eLife.36163}.}
-
-\end{CSLReferences}
-
-\hypertarget{appendix-appendix}{%
-\appendix}
-
-\hypertarget{code-for-manuscript-data}{%
-\section{Code for Manuscript data}\label{code-for-manuscript-data}}
-
-This section presents the code used to generate figures, models and simulated data from Sections 3 and 4 from the main manuscript.
-
-\hypertarget{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}{%
-\subsection{Compound symmetry and independent errors in linear and quadratic responses}\label{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}}
-
-This section simulated linear and quadratic data in the same manner as in Section \ref{simulation}.
-The linear simulations using Figure \ref{fig:linear-cases-Appendix} show in panels A and D the simulated mean responses and individual data points.
-Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-The following code produces a more comprehensive exploration of Figure \ref{fig:l-q-response} in the main manuscript.
-
-\begin{lstlisting}[language=R]
-##########Section for calculations###########
-
-## Example with linear response
-
-#This function simulates data using a linear or quadratic mean response and each with correlated
-#or uncorrelated errors. Each group has a different slope/concavity.
-example <- function(n_time = 6, #number of time points
-                    fun_type = "linear", #type of response
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  x <- seq(1,6, length.out = n_time)
-
-  #Create mean response matrix: linear or quadratic
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # quadratic response (non-linear)
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-
-  #create an array where individual observations per each time point for each group are to be stored. Currently using 10 observations per timepoint
-  y <- array(0, dim = c(length(x), 2, 10))
-
-  #Create array to store the "errors" for each group at each timepoint. The "errors" are the 
-  #between-group variability in the response.
-  errors <- array(0, dim = c(length(x), 2, 10))
-  #create an array where 10 observations per each time point for each group are to be stored
-
-  #The following cycles create independent or correlated responses.
-To each value of mu (mean response per group) a randomly generated error (correlated or uncorrelated) is added and thus the individual response is created.
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors: variance covariance matrix
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-#labeling y and errors  
-  dimnames(y) <- list(time = x, 
-                      treatment = 1:2, 
-                      subject = 1:10)
-
-  dimnames(errors) <- list(time = x, 
-                           treatment = 1:2, 
-                           subject = 1:10)
-
-  #labeling the mean response
-  dimnames(mu) <- list(time = x, 
-                       treatment = 1:2)
-
-  #convert y, mu and errors to  dataframes with time, treatment and subject columns
-  dat <- as.data.frame.table(y, 
-                             responseName = "y")
-  dat_errors <- as.data.frame.table(errors, 
-                                    responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, 
-                                responseName = "mu")
-
-  #join the dataframes to show mean response and errors per subject
-  dat <- left_join(dat, dat_errors, 
-                   by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, 
-                   by = c("time", "treatment"))
-  #add time
-  dat$time <- as.numeric(as.character(dat$time))
-  #label subjects per group
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, 
-                                  treatment, 
-                                  sep = "-")))
-
-  ## repeated measures ANOVA
-
-  fit_anova <- lm(y ~ time + treatment + time * treatment, data = dat)
-
-#LMEM: time and treatment interaction model, compound symmetry 
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  #create a prediction frame where the model can be used for plotting purposes
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  #add model predictions to the dataframe that has the simulated data
-  dat$pred_anova <- predict(fit_anova)
-  dat$pred_lmem <- predict(fit_lme)
-
-  #return everything in a list
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_anova=fit_anova,
-    fit_lme = fit_lme
-  ))
-}
-##################Section for plotting#################################
-#######################################################################
-#This function will create the plots for either a "linear" or "quadratic" response
-
-plot_example <- function(sim_dat) {
-  ## Plot the simulated data (scatterplot)
-
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = treatment, 
-               color = treatment)
-           ) +
-    geom_point(show.legend=FALSE) +
-    labs(y='response')+
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment),
-              show.legend=FALSE) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-  #plot the simulated data with trajectories per each subject
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = subject, 
-               color = treatment)
-           ) +
-    geom_line(aes(size = "Subjects"),
-              show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment,
-                  size = "Simulated Truth"), 
-              lty = 1,show.legend = FALSE) +
-    labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                face = "bold"),
-     text=element_text(size=30))+
-    thm
-
-  #plot the errors
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = errors,
-               group = subject, 
-               color = treatment)) +
-    geom_line(show.legend=FALSE) +
-     labs(y='errors')+
-     theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-   #plot the model predictions for rm-ANOVA
-  p4 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point(show.legend=FALSE)+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_anova), 
-                  group = subject, size = "Subjects"),show.legend = FALSE) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_anova, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population"),
-              show.legend=FALSE) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-   #plot the LMEM predictions
-  p5 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point()+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), 
-                  group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_lme, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population")) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-  return((p1+p3+p2+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-
-#Store each plot in a separate object
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent")) 
-\end{lstlisting}
-
-\begin{figure}[H]
-\includegraphics{00-Full_document_files/figure-latex/linear-cases-Appendix-1} \caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are able to capture the trend of the data.}\label{fig:linear-cases-Appendix}
-\end{figure}
-
-For the quadratic response case, Figure \ref{fig:quadratic-cases-Appendix} shows the simulated responses using compound symmetry and independent errors.
-
-\begin{figure}[H]
-\includegraphics{00-Full_document_files/figure-latex/quadratic-cases-Appendix-1} \caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are not able to capture the changes in each group over time.}\label{fig:quadratic-cases-Appendix}
-\end{figure}
-
-\hypertarget{basis-functions-and-gams}{%
-\subsection{Basis functions and GAMs}\label{basis-functions-and-gams}}
-
-This code produces Figure \ref{fig:basis-plot} from the main manuscript.
-Briefly, a non-linear (quadratic) response is simulated a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
-The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
-
-\begin{lstlisting}[language=R]
-#generate the response: the same initial procedure from the previous section to simulate
-#the response
-set.seed(1)
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-
- #label each table
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
-
- #Convert to dataframes with subject, time and group columns
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
-
- #label subject per group
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
- #extract  "Group 1" to fit the GAM
-  dat<-subset(dat,treatment==1)
- #keep just the response and timepoint columns
-   dat<-dat[,c('y','time')]
-
-   #GAM model of time, 5 knots
-gm<-gam(y~s(time,k=5),data=dat)
-
-#model_matrix (also known as) 'design matrix'
-#will contain the smooths used to create  model 'gm'
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be 
-#multiplied by the coefficients
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final smooth by aggregating the basis functions
-
-smooth<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(smooth=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,
-            aes(x=time,
-                y=value_orig,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(size=sz,
-            show.legend=FALSE)+
-  geom_point(size=sz+1,
-             show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,
-             labeller = as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#penalized basis
-p12<-ggplot(basis_plot,
-            aes(x=time,
-                y=mod_val,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(show.legend = FALSE,
-            size=sz)+
-  geom_point(show.legend = FALSE,
-             size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,
-             labeller=as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#heatmap of the coefficients
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,
-            aes(x=Basis,
-                y=Basis))+
-  geom_tile(aes(fill = cof), 
-            colour = "black") +
-    scale_fill_gradient(low = "white",
-                        high = "#B50A2AFF")+ #color picked from KikiMedium
-  labs(x='Basis',
-       y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),
-            size=7,
-            show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,
-            aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  labs(y='Simulated \n response')+
-  geom_line(data=smooth,
-            aes(x=time,
-                y=smooth),
-            color="#6C581DFF",
-            size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-appendix-1}
-
-}
-
-\caption{Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-B: Matrix of basis function weights.
-Each basis function is multiplied by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
-Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot-appendix}
-\end{figure}
-
-\hypertarget{tumor-data-simulation}{%
-\section{Longitudinal biomedical data simulation and GAMs}\label{tumor-data-simulation}}
-
-This section describes how to fit GAMs to longitudinal data using simulated data.
-First, data is simulated according to Section \ref{longitudinal-GAMs}, where reported data of oxygen saturation (\(\mbox{StO}_2\)) in tumors under either chemotherapy or saline control is used as a starting point to generate individual responses in each group.
-
-\begin{lstlisting}[language=R]
-set.seed(1)
-#Dataframe that contains the original reported trends
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
-
-## plot the mean response
-f1<-ggplot(dat, 
-           aes(x = Day, 
-               y = StO2, 
-               color = Group)) +
-    geom_line(size=1,
-              show.legend = FALSE)+
-    geom_point(show.legend = FALSE,
-               size=1.5,
-               alpha=0.5)+
-  labs(y=expression(paste(StO[2],
-                          ' (real)')))+
-  theme_classic()+
-  thm+
-    scale_x_continuous(breaks=c(0,5,10))+
-    scale_y_continuous(breaks=c(0,40))+
-  plot_layout(tag_level = 'new')+
-  theme(
-    plot.background = element_rect(fill = "transparent", 
-                                   color = NA),
-    axis.text=element_text(size=14)
-  )
-
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-n <- 10 #number of observations
-sd <- 10 #approximate sd from paper
-df <- 6
-dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, 
-           aes(x = Day, 
-               y = StO2_sim, 
-               color = Group)) +
-    geom_point(show.legend=FALSE,
-               size=1.5,
-               alpha=0.5)+
-    stat_summary(aes(y = StO2_sim,
-                     group=Group), 
-                 fun=mean, geom="line",
-                 size=1,
-                 show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],
-                         '(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-  thm+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
-\end{lstlisting}
-
-\hypertarget{workflow}{%
-\subsection{A basic Workflow for GAMs}\label{workflow}}
-
-This section shows a basic workflow to fit a series of increasingly complex GAMs to the simulated data from the previous section.
-Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
-
-\hypertarget{first-model}{%
-\subsubsection{First model}\label{first-model}}
-
-The first model fitted to the data is one that only accounts for different smooths by day.
-The model syntax specifies that \passthrough{\lstinline!gam\_00!} is the object that will contain all the model information, and that the model attempts to explain changes in \passthrough{\lstinline!StO2\_sim!} (simulated \(\mbox{StO}_2\)) using a smooth per \passthrough{\lstinline!Day!}.
-The model will use 5 knots (\passthrough{\lstinline!k=5!}) for the smooth.
-The smooth is constructed by default using thin plate regression splines.
-The smoothing parameter estimation method used is the restricted maximum likelihood (\passthrough{\lstinline!REML!}).
-
-\begin{lstlisting}[language=R]
-gam_00<-gam(StO2_sim ~ s(Day, k = 5),
-            method='REML',
-            data  = dat_sim)
-\end{lstlisting}
-
-To obtain model diagnostics, two methodologies are to be used: 1) graphical diagnostics, and 2) a model check.
-In the first case, the functions \passthrough{\lstinline!appraise!} and \passthrough{\lstinline!draw!} from the package \emph{gratia} can be used to obtain a single output with all the graphical diagnostics.
-For model check, the functions \passthrough{\lstinline!gam.check!} and \passthrough{\lstinline!summary!} from \emph{mgcv} provide detailed information about the model fit and its parameters.
-
-\hypertarget{graphical-diagnostics}{%
-\paragraph{Graphical diagnostics}\label{graphical-diagnostics}}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/first-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the first GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:first-GAM-diag}
-\end{figure}
-
-From the output of the function \passthrough{\lstinline!appraise!} in Figure \ref{fig:first-GAM-diag}, the major indicators of concern about the model are the QQ plot of residuals and the histogram of residuals.
-The QQ plot shows that the errors are not reasonably located along the 45\(^{\circ}\) line (which indicates normality), as there are multiple points that deviate from the trend, specially in the tails.
-The histogram also shows that the variation (residuals) is not following the assumption of a normal distribution.
-
-The \passthrough{\lstinline!draw!} function permits to plot the smooths as \passthrough{\lstinline!ggplot2!} objects, which eases subsequent manipulation, if desired.
-Because model \passthrough{\lstinline!gam\_00!} specifies only one smooth for the time covariate (Day), the plot only contains only one smooth.
-Note that the smooth shows an almost linear profile.
-
-\hypertarget{model-check}{%
-\paragraph{Model check}\label{model-check}}
-
-\begin{lstlisting}[language=R]
-#need to add figure number and caption
-gam.check(gam_00)
-\end{lstlisting}
-
-\includegraphics{00-Full_document_files/figure-latex/first-GAM-check-1}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 5 iterations.
-## Gradient range [-0.0003727881,-6.621452e-07]
-## (score 444.0118 & scale 450.6638).
-## Hessian positive definite, eigenvalue range [0.3881695,49.00676].
-## Model rank =  5 / 5 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##          k'  edf k-index p-value    
-## s(Day) 4.00 2.11    0.36  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-summary(gam_00)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ s(Day, k = 5)
-## 
-## Parametric coefficients:
-##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   22.967      2.123   10.82   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##          edf Ref.df     F  p-value    
-## s(Day) 2.114  2.565 7.633 0.000517 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.153   Deviance explained = 17.2%
-## -REML = 444.01  Scale est. = 450.66    n = 100
-\end{lstlisting}
-
-Special attention must be paid to the `k-index' from \passthrough{\lstinline!gam.check!}.
-This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
-If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless1).
-From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data.
-The \passthrough{\lstinline!edf!} (effective degrees of freedom) is an indicator of the complexity of the smooth.
-Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
-
-From the \passthrough{\lstinline!summary!} function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
-The link function is `identity' as the model does not make any transformation on the predictors.
-The `significance of smooth terms' \emph{p-value} indicates if each smooth is adding significance to the model.
-Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \(\approx\) 17\%.
-
-\hypertarget{second-model}{%
-\subsubsection{Second model}\label{second-model}}
-
-The major flaw of \passthrough{\lstinline!gam\_00!} is that this model is not taking into account the fact that the data is nested in groups.
-The next iteration is a model where a different smooth of time (Day) is assigned for each group using \passthrough{\lstinline!by=Group!} in the model syntax.
-
-\begin{lstlisting}[language=R]
-gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5),
-            method='REML',
-            data  = dat_sim)
-
-gam.check(gam_01)
-\end{lstlisting}
-
-\includegraphics{00-Full_document_files/figure-latex/second-GAM-1}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 7 iterations.
-## Gradient range [-5.51754e-05,2.671715e-06]
-## (score 423.3916 & scale 280.8777).
-## Hessian positive definite, eigenvalue range [0.3162258,48.5557].
-## Model rank =  9 / 9 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                         k'  edf k-index p-value    
-## s(Day):GroupControl   4.00 3.39    0.43  <2e-16 ***
-## s(Day):GroupTreatment 4.00 3.23    0.43  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-summary(gam_01)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ s(Day, by = Group, k = 5)
-## 
-## Parametric coefficients:
-##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   22.967      1.676    13.7   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                         edf Ref.df      F p-value    
-## s(Day):GroupControl   3.392  3.794  3.817  0.0304 *  
-## s(Day):GroupTreatment 3.229  3.682 21.174  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.472   Deviance explained = 50.8%
-## -REML = 423.39  Scale est. = 280.88    n = 100
-\end{lstlisting}
-
-Diagnostics for this model indicate that the k-index is still below 1 (0.43 from \passthrough{\lstinline!gam.check!}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
-Moreover, the smooths (plotted via the \passthrough{\lstinline!draw()!} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \(\approx\) 51\%.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/second-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the second GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:second-GAM-diag}
-\end{figure}
-
-\hypertarget{third-model}{%
-\subsubsection{Third model}\label{third-model}}
-
-Model \passthrough{\lstinline!gam\_00!} was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted, the model diagnostics, and, the low variance explained by the model.
-On the other hand, \passthrough{\lstinline!gam\_01!} takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \ref{longitudinal-GAMs}, in order to differentiate between each group a parametric term needs to be added to the model for the interaction of \emph{Day} and \emph{Group}.
-
-\begin{lstlisting}[language=R]
-#GAM for StO2
-
-m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
-            method='REML',
-            data  = dat_sim)
-
-gam.check(m1)
-\end{lstlisting}
-
-\includegraphics{00-Full_document_files/figure-latex/final-model-Appendix-1}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 10 iterations.
-## Gradient range [-8.164307e-08,1.500338e-08]
-## (score 355.8554 & scale 64.53344).
-## Hessian positive definite, eigenvalue range [1.174841,48.08834].
-## Model rank =  10 / 10 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                         k'  edf k-index p-value
-## s(Day):GroupControl   4.00 3.87    1.02    0.59
-## s(Day):GroupTreatment 4.00 3.88    1.02    0.54
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-summary(m1)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ Group + s(Day, by = Group, k = 5)
-## 
-## Parametric coefficients:
-##                Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)       9.084      1.136   7.996 4.09e-12 ***
-## GroupTreatment   27.766      1.607  17.282  < 2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                         edf Ref.df     F p-value    
-## s(Day):GroupControl   3.873  3.990 17.57  <2e-16 ***
-## s(Day):GroupTreatment 3.879  3.991 89.33  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.879   Deviance explained = 88.9%
-## -REML = 355.86  Scale est. = 64.533    n = 100
-\end{lstlisting}
-
-The resulting model is \passthrough{\lstinline!m1!}, which is the model fitted in the main manuscript.
-By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\(\approx\) 1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 89\% of the variance in the data.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/final-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the final GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooths for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:final-GAM-diag}
-\end{figure}
-
-\hypertarget{comparing-models-via-aic}{%
-\subsubsection{Comparing models via AIC}\label{comparing-models-via-aic}}
-
-One final comparison that can be made for model selection involves the use of the Aikake Information Criterion (AIC).
-This metric is used to estimate information loss, which we want to minimize with an appropriate model.
-Therefore, when 2 or more models are compared, the model with lower AIC is preferred.
-In R, the comparison is done using the \passthrough{\lstinline!AIC!} function.
-
-\begin{lstlisting}[language=R]
-AIC(gam_00,gam_01,m1)
-\end{lstlisting}
-
-\begin{lstlisting}
-##               df      AIC
-## gam_00  4.564893 900.8257
-## gam_01  9.476137 858.6051
-## m1     10.980983 712.2067
-\end{lstlisting}
-
-The output in this case is expected: model \passthrough{\lstinline!m1!} has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
-The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
-
-\hypertarget{pairwise-comparisons-of-smooth-confidence-intervals}{%
-\paragraph{Pairwise comparisons of smooth confidence intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
-
-The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \ref{GAM-significance}.
-In this case, the ``design matrix'' is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
-Briefly, the ``design matrix'' (also known as the ``Xp matrix'') from the selected model (\passthrough{\lstinline!m1!}) is used to calculate a 95\% confidence interval of the difference between the smooth terms for each group.
-This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
-\textbf{All pairwise comparisons in this paper have been centered at the response scale to ease interpretation }.
-
-\begin{lstlisting}[language=R]
-##Pairwise comparisons
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-##matrix that contains the basis functions evaluated at the points in pdat
-    xp <- predict(m1, newdata = pdat, type = 'lpmatrix')
-
-#Find columns in xp where the name contains "Control"
-    c1 <- grepl('Control', colnames(xp))
-
-#Find columns in xp where the name contains 'Treatment'
-    c2 <- grepl('Treatment', colnames(xp))
-
-#Find rows in pdat that correspond to either 'Control' or 'Treatment'
-    r1 <- with(pdat, Group == 'Control')
-    r2 <- with(pdat, Group == 'Treatment')
-
-# In xp: find the rows that correspond to Control or Treatment, those that do not match will be
-    #set to zero. Then, substract the values from the rows corresponding to 'Control' from those that correspond
-    #to 'Treatment'
-    X <- xp[r1, ] - xp[r2, ]
-
-    ## remove columns that do not contain name 'Control' or 'Treatment'
-    X[, ! (c1 | c2)] <- 0
-    ## zero out the parametric cols, those that do not contain in the characters 's('
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
-
-    #Multiply matrix by model coefficients. X has (p,n) (rows, columns) and the coefficient matrix has
-    #dimensions (n,1). The resulting matrix has dimensions (p,1)
-    dif <- X %*% coef(m1)
-
-    #comp<-test %*% coef(gam1)[3:10]
-
-#Calculate standard error for the computed differences using the variance-covariance matrix
-    #of the model
-    se <- sqrt(rowSums((X %*% vcov(m1, unconditional = FALSE)) * X))
-    crit <- qt(0.05/2, df.residual(m1), lower.tail = FALSE)
-    #upper  limits
-    upr <- dif + (crit * se)
-    #lower limits
-    lwr <- dif - (crit * se)
-    #put all components in a dataframe for plotting
-    comp1<-data.frame(pair = paste('Control', 'Treatment', sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-
-#add time point sequence
-comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1))
-
-#use function from the pairwise comparison plot in the manuscript to get the shaded regions
-
-    my_list<-pairwise_limits(comp_StO2)
-  rib_col<-'#EDD03AFF' #color for the ribbon
-#plot the difference
-c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
-  #shaded region
-  annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill='#30123BFF',
-                alpha = 0.5,
-                ) +
-  annotate("text",
-             x=1.5,
-             y=-10,
-             label="Control",size=10
-           )+
-  #shaded region  
-  annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
-             alpha = 0.5
-           ) +
-  annotate("text",
-             x=6,
-             y=-10,
-             label="Treatment",
-             size=10
-           )+
-  #ribbon for difference confidence interval  
-  geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(color='black',size=1) +
-    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/pairwise-comp-workflow-fig-1}
-
-}
-
-\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!m1!} using a 95\% confidence interval for the difference between smooths.
-The comparison is centered at the response scale.
-Shaded regions indicate time intervals where each treatment group has a non-zero effect.}\label{fig:pairwise-comp-workflow-fig}
-\end{figure}
-
-Of notice, a convenient wrapper for the function described above exists in the package \passthrough{\lstinline!gratia!}.
-In this package, \passthrough{\lstinline!difference\_smooths!} is a function that makes the comparisons and produces Figure \ref{fig:pairwise-comp-workflow-fig} when is used on a fitted model.
-The function syntax and an example can be found at:
-
-\url{https://cran.r-project.org/web/packages/gratia/gratia.pdf}
-
-Keep in mind that this function \textbf{does not} center the pairwise comparison at the response scale, so it has to be shifted in order to be compared to the raw data.
-
-\hypertarget{gam-and-linear-model-plots-and-missing-data}{%
-\section{GAM and Linear model plots and Missing data}\label{gam-and-linear-model-plots-and-missing-data}}
-
-This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\passthrough{\lstinline!m1!}), linear model and GAM on data with missing observations are presented.
-Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inset are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
-
-\hypertarget{gam-and-linear-model-plots}{%
-\subsection{GAM and Linear model plots}\label{gam-and-linear-model-plots}}
-
-This code chunk creates panels B and D in Figure \ref{fig:sim-smooth-plot}.
-Note that this code uses the final GAM from the previous section (\passthrough{\lstinline!m1!}), so the simulated data and the model should be generated before running this section.
-
-\begin{lstlisting}[language=R]
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-#creates a dataframe using the length of the covariates for the GAM
-gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
-                         subject=factor(rep(1:10)))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#adds the predictions to the grid and creates a confidence interval for GAM
-gam_predict<-gam_predict%>%
-    mutate(fit = predict(m1,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(m1, gam_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=gam_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'complete')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-
-#posthoc comparisons for the linear model
-#library(multcomp)
-
-#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
-\end{lstlisting}
-
-\hypertarget{working-with-missing-data-in-gams}{%
-\subsection{Working with Missing data in GAMs}\label{working-with-missing-data-in-gams}}
-
-This code chunk first randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data.
-Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
-
-\begin{lstlisting}[language=R]
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
-
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'missing')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-Appendix-1}
-
-}
-
-\caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
-\end{figure}
-
-\hypertarget{pairwise-comparisons-in-gams-full-and-missing-data-cases}{%
-\subsection{Pairwise comparisons in GAMs: full and missing data cases}\label{pairwise-comparisons-in-gams-full-and-missing-data-cases}}
-
-The next code chunk reproduces Figure \ref{fig:plot-pairwise-comp}. Here pairwise comparisons are made for the full and missing datasets.
-
-\begin{lstlisting}[language=R]
-##Pairwise comparisons
-
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-#this function takes the model, grid and groups to be compared using the lpmatrix
-#originally developed by G. Simpson:
-#https://fromthebottomoftheheap.net/2017/10/10/difference-splines-i/
-
-smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    #Find columns in xp where the name contains "Control" and "Treatment"
-    col1 <- grepl(g1, colnames(xp))
-    col2 <- grepl(g2, colnames(xp))
-    #Find rows in xp that correspond to each treatment
-    row1 <- with(newdata, Group == g1)
-    row2 <- with(newdata, Group == g2)
-    ## difference rows of xp for data from comparison
-    X <- xp[row1, ] - xp[row2, ]
-    ## zero out cols of X related to splines for other lochs
-    X[, ! (col1 | col2)] <- 0
-
-    ## zero out the parametric cols
-    #This line has been commented to keep the comparison at the response level,
-    #otherwise it gives the marginal change between smooths
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    #get standard error, critical value and boundaries
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(g1, g2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-}
-
-#use the function to calculate the difference in smooths
-comp1<-smooth_diff(m1,pdat,'Control','Treatment')
-
-#Create a dataframe with time, comparisons and labels for regions where difference exists
-comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
-
-pairwise_limits<-function(dataframe){
-    #extract values where the lower limit of the ribbon is greater than zero
-    #this is the region where the control group effect is greater
-    v1<-dataframe%>%
-        filter(lower>0)%>%
-        select(Day)
-    #get day  initial value
-    init1=v1$Day[[1]]
-    #get day final value
-    final1=v1$Day[[nrow(v1)]]
-
-    #extract values where the value of the upper limit of the ribbon is lower than zero
-    #this corresponds to the region where the treatment group effect is greater
-    v2<-comp_StO2_full%>%
-        filter(upper<0)%>%
-        select(Day)
-
-    init2=v2$Day[[1]]
-    final2=v2$Day[[nrow(v2)]]
-    #store values
-    my_list<-list(init1=init1,
-                  final1=final1,
-                  init2=init2,
-                  final2=final2)
-return(my_list)
-}
-
-my_list<-pairwise_limits(comp_StO2_full)
-rib_col<-'#EDD03AFF'
-
-c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
-    annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill='#30123BFF',
-                alpha = 0.5,
-                ) +
-  annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=8,
-           angle=90
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=8,
-           angle=90
-           )+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-###for missing data
-comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
-comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp2))
-
-missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    labs(x = 'Days', 
-         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
-                              )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme_classic()+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
-my_list<-pairwise_limits(comp_StO2_missing)
-
-c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    annotate("rect",
-             xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-             fill='#30123BFF',
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=8
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=8)+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-pair_comp<-c1+c2
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-Appendix-1}
-
-}
-
-\caption{Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-B: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
-In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp-Appendix}
-\end{figure}
-
-\end{document}

### 2021-06-18 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.Rmd
References do not work

--- 2021-06-10 Manuscripts/Manuscript by chapters -v2/00-Full_document.Rmd
@@
 ---
-title: '**Using generalized additive models to analyze biomedical non-linear longitudinal data**'
-subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
-header-includes:
-    \usepackage{placeins}
-output:  
-  bookdown::pdf_document2:
-    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
-    pandoc_args: --listings #calls the listings package to fit code within the page margins
-    keep_tex: yes #keep LaTeX file for submission
-    fig_caption: yes #allows captions in figures
-    toc: false #do not include table of contents
-    extra_dependencies:
-      subfig: null #allows for subfigures
-      breqn: null #line breaks for long equations
-      caption: ["font={small}"] #size of the figure captions
-      float: null #allows for control of placement of figures
-    includes:
-      in_header: preamble.sty #additional LaTeX formatting
-  bookdown::word_document2:
-    fig_caption: yes #figure caption
-    keep_md: yes
-  bookdown::html_document2:
-    css: "style.css" #style for the HTML document
-csl: elsevier-with-titles.csl #style for references
-bibliography: refs.bib #references
-link-citations: yes #adds links to the citations
-'': default
----
-
-# Abstract
-
-In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs).
+title: Using generalized additive models to analyze biomedical non-linear longitudinal data
+runninghead: Mundo \emph{et al}.
+author:
+- name: Ariel I. Mundo
+  num: 1
+- name: John R. Tipton
+  num: 2
+- name: Timothy J. Muldoon*
+  num: 1
+address:
+- num: 1
+  org: Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, US
+- num: 2
+  org:  Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, US
+corrauth: "Corresponding author"
+email: tmuldoon@uark.edu
+abstract: "In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs).
 Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
-
-In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+In contrast, _generalized additive models_ (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
 Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 6 repeated measures for each group) to demonstrate how these models can be computationally implemented.
-We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the observations missing), allowing reliable inference from the data.
-To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research.
+We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented.
+We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing), allowing reliable inference from the data.
+To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research."
+keywords: Cancer biology; Tumor response; generalized additive models; simulation; R
+classoption:
+  - Royal
+  - times
+bibliography: refs1.bib
+bibliographystyle: sagev
+output:
+  rticles::sage_article:
+    keep_tex: yes
+    #pandoc_args: --listings
+    fig_caption: yes
+    #extra_dependencies:
+      #subfig: null #allows for subfigures
+      #breqn: null #line breaks for long equations
+      #caption: ["font={small}"] #size of the figure captions
+      #float: null #allows for control of placement of figures
+---
 
@@
 
-knitr::opts_chunk$set(fig.pos = 'H', out.extra = "",error=TRUE)
+#knitr::opts_chunk$set(fig.pos = 'H', out.extra = "",error=TRUE)
 
@@
 #https://github.com/rstudio/bookdown/issues/1158
-remove_author <- function(x) {
-  # identify empty author line
-  i <- grep("^\\\\author\\{\\}$", x)
-  # be sure it is the one pandoc inserts
-  if(length(i) != 0 && grepl('^\\\\date\\{', x[i+1])) x <- x[-i]
-  x
-}
-options(bookdown.post.latex = remove_author)
+# remove_author <- function(x) {
+#   # identify empty author line
+#   i <- grep("^\\\\author\\{\\}$", x)
+#   # be sure it is the one pandoc inserts
+#   if(length(i) != 0 && grepl('^\\\\date\\{', x[i+1])) x <- x[-i]
+#   x
+# }
+# options(bookdown.post.latex = remove_author)
 ```
@@
 
-\FloatBarrier
+<!-- \FloatBarrier -->
 
@@
 
-\newpage
-# References
+# Bibliography
 
-<div id="refs"></div>
+<!-- <div id="refs"></div> -->
 

### 2021-06-18 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.tex
References do not work

--- 2021-06-11 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Using generalized additive models to analyze biomedical non-linear longitudinal data},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
-\usepackage{listings}
-\newcommand{\passthrough}[1]{#1}
-\lstset{defaultdialect=[5.3]Lua}
-\lstset{defaultdialect=[x86masm]Assembler}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-\usepackage{lineno}
-\usepackage{authblk}
-\usepackage{hyperref}
-\usepackage{graphicx}
-
-%set a box to put the ORCID logo
-\newbox{\myorcidaffilbox}
-\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}
-
-%add hyperlink to the box
-\newcommand{\orcidaffila}[1]{%
-  \href{https://orcid.org/0000-0002-6014-4538}{\usebox{\myorcidaffilbox}}}
-
-\newcommand{\orcidaffilb}[1]{%
-  \href{https://orcid.org/0000-0002-6135-8191}{\usebox{\myorcidaffilbox}}}
-
-%command for the package lineno
-\linenumbers
-
-%authors
-\author[1]{Ariel I. Mundo \orcidaffila{}}
-\author[2]{John R. Tipton \orcidaffilb{}}
-\author[1]{Timothy J. Muldoon \thanks{Corresponding author, tmuldoon@uark.edu}}
-\affil[1]{\footnotesize Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
-\affil[2]{\footnotesize Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
-
-%theme colors for the code chunks (originally from latex-solarized on GitHub)
-%https://github.com/jez/latex-solarized
-\usepackage{xcolor}
-\definecolor{sbase03}{HTML}{002B36}
-\definecolor{sbase02}{HTML}{073642}
-\definecolor{sbase01}{HTML}{586E75}
-\definecolor{sbase00}{HTML}{657B83}
-\definecolor{sbase0}{HTML}{839496}
-\definecolor{sbase1}{HTML}{93A1A1}
-\definecolor{sbase2}{HTML}{EEE8D5}
-\definecolor{sbase3}{HTML}{FDF6E3}
-\definecolor{syellow}{HTML}{B58900}
-\definecolor{sorange}{HTML}{CB4B16}
-\definecolor{sred}{HTML}{DC322F}
-\definecolor{smagenta}{HTML}{D33682}
-\definecolor{sviolet}{HTML}{6C71C4}
-\definecolor{sblue}{HTML}{268BD2}
-\definecolor{scyan}{HTML}{2AA198}
-\definecolor{sgreen}{HTML}{859900}
-%command to set parameter(s) in package listings
-\lstset{
-    % How/what to match
-    sensitive=true,
-    % Border (above and below)
-    frame=lines,
-    % Extra margin on line (align with paragraph)
-    xleftmargin=\parindent,
-    % Put extra space under caption
-    belowcaptionskip=1\baselineskip,
-    % Colors
-    backgroundcolor=\color{sbase3},
-    basicstyle=\color{sbase00}\ttfamily,
-    keywordstyle=\color{scyan},
-    commentstyle=\color{sbase1},
-    stringstyle=\color{sblue},
-    numberstyle=\color{sviolet},
-    identifierstyle=\color{sbase00},
-    % Break long lines into multiple lines?
-    breaklines=true,
-    % Show a character for spaces?
-    showstringspaces=false,
-    tabsize=2
-}
-
-%\lstset{
-%  breaklines=true,
-%  stringstyle=\ttfamily,
-%  backgroundcolor=\color{gray}
-%}
-\usepackage{placeins}
-\usepackage{subfig}
-\usepackage{breqn}
-\usepackage[font={small}]{caption}
-\usepackage{float}
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
-\usepackage{etoolbox}
-\makeatletter
-\providecommand{\subtitle}[1]{% add subtitle to \maketitle
-  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
-}
-\makeatother
-\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
-\date{\vspace{-2.5em}}
+\documentclass[Royal,times,sagev]{sagej}
+
+\usepackage{moreverb,url,natbib, multirow, tabularx}
+\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
+
+% Pandoc citation processing
 
 \begin{document}
+
+\title{Using generalized additive models to analyze biomedical
+non-linear longitudinal data}
+
+\runninghead{Mundo \emph{et al}.}
+
+\author{Ariel I. Mundo\affilnum{1}, John R. Tipton\affilnum{2}, Timothy
+J. Muldoon*\affilnum{1}}
+
+\affiliation{\affilnum{1}{Department of Biomedical Engineering,
+University of Arkansas, Fayetteville, AR, US}\\\affilnum{2}{Department
+of Mathematical Sciences, University of Arkansas, Fayetteville, AR, US}}
+
+\corrauth{Corresponding author}
+
+\email{\href{mailto:tmuldoon@uark.edu}{\nolinkurl{tmuldoon@uark.edu}}}
+
+\begin{abstract}
+In biomedical research, the outcome of longitudinal studies has been
+traditionally analyzed using the \emph{repeated measures analysis of
+variance} (rm-ANOVA) or more recently, \emph{linear mixed models}
+(LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of
+correlation and missing observations, both methodologies share an
+assumption of linearity in the measured response, which results in
+biased estimates and unreliable inference when they are used to analyze
+data where the trends are non-linear, which is a common occurrence in
+biomedical research. In contrast, \emph{generalized additive models}
+(GAMs) relax the linearity assumption, and allow the data to determine
+the fit of the model while permitting missing observations and different
+correlation structures. Therefore, GAMs present an excellent choice to
+analyze non-linear longitudinal data in the context of biomedical
+research. This paper summarizes the limitations of rm-ANOVA and LMEMs
+and uses simulated data to visually show how both methods produce biased
+estimates when used on non-linear data. We also present the basic theory
+of GAMs, and using trends of oxygen saturation in tumors reported in the
+biomedical literature, we simulate example longitudinal data (2
+treatment groups, 10 subjects per group, 5 repeated measures for each
+group) to demonstrate how these models can be computationally
+implemented. We show that GAMs are able to produce estimates that are
+consistent with the trends of biomedical non-linear data even in the
+case when missing observations exist (with 40\% of the simulated
+observations missing), allowing reliable inference from the data. To
+make this work reproducible, the code and data used in this paper are
+available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
+\end{abstract}
+
+\keywords{Cancer biology; Tumor response; generalized additive models;
+simulation; R}
+
 \maketitle
-
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
-
-In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs).
-Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
-
-In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
-Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
-This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented.
-We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing), allowing reliable inference from the data.
-To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
 
@@
 
-Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response {[}\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}{]}, antibody expression {[}\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}{]}, and cell metabolism {[}\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}\protect\hyperlink{ref-greening2018}{9}{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}\protect\hyperlink{ref-sio2016}{10}{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}\protect\hyperlink{ref-kamstra2015}{11}{]}.
-
-Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}\protect\hyperlink{ref-wagenmakers2008}{12}{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}{]}.
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
-
-A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
-Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}.
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}\protect\hyperlink{ref-wang2019}{18}{]}.
-
-Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}{]}, consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}\protect\hyperlink{ref-abdi2010}{21}{]}), but they in turn reduce statistical power (1-\(\beta\)){[}\protect\hyperlink{ref-nakagawa2004}{22}{]}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) {[}\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}{]}.
-Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and group sizes {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}\protect\hyperlink{ref-charan2013}{30}{]} which might cause the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible.
-It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}\protect\hyperlink{ref-gueorguieva2004}{13}{]}.
-This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
-At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
-
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}.
-And even more importantly, LMEMs also assume a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}, making them unsuitable to analyze non-linear data.
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
-In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]} .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]} and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
-
-The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
-Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
-
-This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations are explained in detail.
-Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in \(\textsf{R}\), in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
-
-In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models to improve the standards for reproducibility in biomedical research.
-
-\FloatBarrier
+Longitudinal studies are designed to repeatedly measure a variable of
+interest in a group (or groups) of subjects, with the intention of
+observing the evolution of effect across time rather than analyzing a
+single time point (e.g., a cross-sectional study). Biomedical research
+frequently uses longitudinal studies to analyze the evolution of a
+``treatment'' effect across multiple time points; and in such studies
+the subjects of analysis range from animals (mice, rats, rabbits), to
+human patients, cells, or blood samples, among many others. Tumor
+response \citep{roblyer2011, tank2020, pavlov2018, demidov2018},
+antibody expression \citep{ritter2001, roth2017}, and cell metabolism
+\citep{jones2018, skala2010} are examples of the different situations
+where researchers have used longitudinal designs to study some
+physiological response. Because the frequency of the measurements in a
+longitudinal study is dependent on the biological phenomena of interest
+and the experimental design of the study, the frequency of such
+measurements can range from minute intervals to study a short-term
+response such as anesthesia effects in animals\citep{greening2018}, to
+weekly measurements to analyze a mid-term response like the evolution of
+dermatitis symptoms in breast cancer patients \citep{sio2016}, to
+monthly measurements to study a long-term response such as mouth opening
+following radiotherapy (RT) in neck cancer patients \citep{kamstra2015}.
+
+Traditionally, a ``frequentist'' or ``classical'' statistical paradigm
+is used in biomedical research to derive inferences from a longitudinal
+study. The frequentist paradigm regards probability as the limit of the
+expected outcome when an experiment is repeated a large number of times
+\citep{wagenmakers2008}, and such view is applied to the analysis of
+longitudinal data by assuming a null hypothesis under a statistical
+model that is often an \emph{analysis of variance over repeated
+measures} (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model
+makes three key assumptions regarding longitudinal data: 1) linearity of
+the response across time, 2) constant correlation across same-subject
+measurements, and 3) observations from each subject are obtained at all
+time points through the study (a condition also known as \emph{complete
+observations}) \citep{gueorguieva2004, schober2018}.
+
+The expected linear behavior of the response through time is a key
+requisite in rm-ANOVA \citep{pinheiro2006}. This ``linearity
+assumption'' in rm-ANOVA implies that the model is misspecified when the
+data does not follow a linear trend, which results in unreliable
+inference. In biomedical research, non-linear trends are the norm rather
+than the exception in longitudinal studies. A particular example of this
+non-linear behavior in longitudinal data arises in measurements of tumor
+response to chemo and/or radiotherapy in preclinical and clinical
+settings \citep{roblyer2011, skala2010, vishwanath2009}. These studies
+have shown that the collected signal does not follow a linear trend over
+time, and presents extreme variability at different time points, making
+the fit of rm-ANOVA model inconsistent with the observed variation.
+Therefore, when rm-ANOVA is used to draw inference of such data the
+estimates are inevitably biased, because the model is only able to
+accommodate linear trends that fail to adequately represent the
+biological phenomenon of interest.
+
+A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to
+perform repeated comparisons to estimate a \emph{p-value}, which in turn
+is used as a measure of significance. Although it is possible that a
+\emph{post hoc} analysis of rm-ANOVA is able to find ``significant''
+\emph{p-values}( \emph{p}\textless0.05) from non-linear data, the
+validity of such metric is dependent on how adequate the model fits the
+data. In other words, \emph{p-values} are valid only if the model and
+the data have good agreement; if that is not the case, a ``Type III''
+error (known as ``model misspecification'') occurs\citep{dennis2019}.
+For example, model misspecification will occur when a model that is only
+able to explain linear responses (such as rm-ANOVA) is fitted to data
+that follows a quadratic trend, thereby causing the resulting
+\emph{p-values} and parameter estimates to be invalid \citep{wang2019}.
+
+Additionally, the \emph{p-value} itself is highly variable, and multiple
+comparisons can inflate the false positivity rate (Type I error or
+\(\alpha\)) \citep{liu2010, halsey2015}, consequently biasing the
+conclusions of the study. Corrections exist to address the Type I error
+issue of multiple comparisons (such as Bonferroni \citep{abdi2010}), but
+they in turn reduce statistical power (1-\(\beta\))\citep{nakagawa2004},
+and lead to increased Type II error (failing to reject the null
+hypothesis when the null hypothesis is false)
+\citep{gelman2012, albers2019}. Therefore, the tradeoff of \emph{post
+hoc} comparisons in rm-ANOVA between Type I, II and III errors might be
+difficult to resolve in a biomedical longitudinal study where a delicate
+balance exists between statistical power and sample size.
+
+On the other hand, the assumption of constant correlation in rm-ANOVA
+(often known as the \emph{compound symmetry assumption}) is typically
+unreasonable because correlation between the measured responses often
+diminishes as the time interval between the observation increases
+\citep{ugrinowitsch2004}. Corrections can be made in rm-ANOVA in the
+absence of compound symmetry \citep{huynh1976, greenhouse1959}, but the
+effectiveness of the correction is limited by the size of the sample,
+the number of measurements\citep{haverkamp2017}, and group sizes
+\citep{keselman2001}. In the case of biomedical research, where living
+subjects are frequently used, sample sizes are often not ``large'' due
+to ethical and budgetary reasons \citep{charan2013} which might cause
+the corrections for lack of compound symmetry to be ineffective.
+
+Due to a variety of causes, the number of observations during a study
+can vary between all subjects. For example, in a clinical trial patients
+may voluntarily withdraw, whereas attrition due to injury or weight loss
+in preclinical animal studies is possible. It is even plausible that
+unexpected complications with equipment or supplies arise that prevent
+the researcher from collecting measurements at certain time points. In
+each of these missing data scenarios, the \emph{complete observations}
+assumption of classical rm-ANOVA is violated. When incomplete
+observations occur, a rm-ANOVA model is fit by excluding all subjects
+with missing observations from the analysis \citep{gueorguieva2004}.
+This elimination of partially missing data from the analysis can result
+in increased costs if the desired statistical power is not met with the
+remaining observations, because it would be necessary to enroll more
+subjects. At the same time, if the excluded observations contain
+insightful information that is not used, their elimination from the
+analysis may limit the demonstration of significant differences between
+groups.
+
+During the last decade, the biomedical community has started to
+recognize the limitations of rm-ANOVA in the analysis of longitudinal
+data. The recognition on the shortcomings of rm-ANOVA is exemplified by
+the use of linear mixed effects models (LMEMs) by certain groups to
+analyze longitudinal tumor response data
+\citep{skala2010, vishwanath2009}. Briefly, LMEMs incorporate
+\emph{fixed effects}, which correspond to the levels of experimental
+factors in the study (e.g., the different drug regimens in a clinical
+trial), and \emph{random effects}, which account for random variation
+within the population (e.g., the individual-level differences not due to
+treatment such as weight or age). When compared to the traditional
+rm-ANOVA, LMEMs are more flexible as they can accommodate missing
+observations for multiple subjects and allow different modeling
+strategies for the variability within each measure in every subject
+\citep{pinheiro2006}. However, LMEMs impose restrictions in the
+distribution of the errors of the random effects, which need to be
+normally distributed and independent \citep{gueorguieva2004, barr2013}.
+And even more importantly, LMEMs also assume a linear relationship
+between the response and time \citep{pinheiro2006}, making them
+unsuitable to analyze non-linear data.
+
+As the rm-ANOVA and the more flexible LMEM approaches make overly
+restrictive assumptions regarding the linearity of the response, there
+is a need for biomedical researchers to explore the use of additional
+statistical tools that allow the data (and not an assumption in trend)
+to determine the trend of the fitted model, to enable appropriate
+inference. In this regard, generalized additive models (GAMs) present an
+alternative approach to analyze longitudinal data. Although not
+frequently used by the biomedical community, these semi-parametric
+models are customarily used in other fields to analyze longitudinal
+data. Examples of the use of GAMs include the analysis of temporal
+variations in geochemical and palaeoecological data
+\citep{rose2012, pedersen2019, simpson2018}, health-environment
+interactions \citep{yang2012} and the dynamics of government in
+political science \citep{beck1998} . There are several advantages of
+GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible
+class of smooth responses that enable the data to dictate the trend in
+the fit of the model, 2) they can model non-constant correlation between
+repeated measurements \citep{wood2017} and 3) can easily accommodate
+missing observations. Therefore, GAMs can provide a more flexible
+statistical approach to analyze non-linear biomedical longitudinal data
+than LMEMs and rm-ANOVA.
+
+The current advances in programming languages designed for statistical
+analysis (specifically \(\textsf{R}\)), have eased the computational
+implementation of traditional models such as rm-ANOVA and more complex
+approaches such as LMEMs and GAMs. In particular, \(\textsf{R}\)
+\citep{r} has an extensive collection of documentation and functions to
+fit GAMs in the package \emph{mgcv} \citep{wood2016, wood2017} that not
+only speed up the initial stages of the analysis but also enable the use
+of advanced modeling structures (e.g.~hierarchical models, confidence
+interval comparisons) without requiring advanced programming skills from
+the user. At the same time, \(\textsf{R}\) has many tools that simplify
+data simulation, an emerging strategy used to test statistical models
+\citep{haverkamp2017}. Data simulation methods allow the researcher to
+create and explore different alternatives for analysis without
+collecting information in the field, reducing the time window between
+experiment design and its implementation, and simulation can be also
+used for power calculations and study design questions.
+
+This work provides biomedical researchers with a clear understanding of
+the theory and the practice of using GAMs to analyze longitudinal data
+using by focusing on four areas. First, the limitations of LMEMs and
+rm-ANOVA regarding linearity of response, constant correlation
+structures and missing observations are explained in detail. Second, the
+key theoretical elements of GAMs are presented using clear and simple
+mathematical notation while explaining the context and interpretation of
+the equations. Third, we illustrate the type of non-linear longitudinal
+data that often occurs in biomedical research using simulated data that
+reproduces patterns in previously reported studies
+\citep{vishwanath2009}. The simulated data experiments highlight the
+differences in inference between rm-ANOVA, LMEMs and GAMs on data
+similar to what is commonly observed in biomedical studies. Finally,
+reproducibility is emphasized by providing the code to generate the
+simulated data and the implementation of different models in
+\(\textsf{R}\), in conjunction with a step-by-step guide demonstrating
+how to fit models of increasing complexity.
+
+In summary, this work will allow biomedical researchers to identify when
+the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze
+longitudinal data, and provide guidance on the implementation of these
+models to improve the standards for reproducibility in biomedical
+research.
 
 \hypertarget{challenges-presented-by-longitudinal-studies}{%
-\section{Challenges presented by longitudinal studies}\label{challenges-presented-by-longitudinal-studies}}
+\section{Challenges presented by longitudinal
+studies}\label{challenges-presented-by-longitudinal-studies}}
 
 \hypertarget{the-repeated-measures-anova-and-linear-mixed-model}{%
-\subsection{The repeated measures ANOVA and Linear Mixed Model}\label{the-repeated-measures-anova-and-linear-mixed-model}}
-
-The \emph{repeated measures analysis of variance} (rm-ANOVA) and the \emph{linear mixed model} (LMEM) are the most commonly used statistical analysis for longitudinal data in biomedical research.
-These statistical methodologies require certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
-Each one of these assumptions is discussed below.
+\subsection{The repeated measures ANOVA and Linear Mixed
+Model}\label{the-repeated-measures-anova-and-linear-mixed-model}}
+
+The \emph{repeated measures analysis of variance} (rm-ANOVA) and the
+\emph{linear mixed model} (LMEM) are the most commonly used statistical
+analysis for longitudinal data in biomedical research. These statistical
+methodologies require certain assumptions for the model to be valid.
+From a practical view, the assumptions can be divided in three areas: 1)
+linear relationship between covariates and response, 2) a constant
+correlation between measurements, and, 3) complete observations for all
+subjects. Each one of these assumptions is discussed below.
 
@@
 \hypertarget{the-repeated-measures-anova-case}{%
-\subsubsection{The repeated measures ANOVA case}\label{the-repeated-measures-anova-case}}
-
-In a longitudinal biomedical study, two or more groups of subjects (e.g., human subject, mice, samples) are subject to different treatments (e.g., a ``treatment'' group receives a novel drug or intervention vs.~a ``control'' group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response is modeled with \emph{fixed} components.
-The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug/intervention in the ``treatment'' group.
-
-Mathematically speaking, a rm-ANOVA model with an interaction can be written as:
+\subsubsection{The repeated measures ANOVA
+case}\label{the-repeated-measures-anova-case}}
+
+In a longitudinal biomedical study, two or more groups of subjects
+(e.g., human subject, mice, samples) are subject to different treatments
+(e.g., a ``treatment'' group receives a novel drug or intervention vs.~a
+``control'' group that receives a placebo), and measurements from each
+subject within each group are collected at specific time points. The
+collected response is modeled with \emph{fixed} components. The
+\emph{fixed} component can be understood as a constant value in the
+response which the researcher is interested in measuring, i.e., the
+average effect of the novel drug/intervention in the ``treatment''
+group.
+
+Mathematically speaking, a rm-ANOVA model with an interaction can be
+written as:
 
@@
 y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
-\label{eq:linear-model}
+(\#eq:linear-model)
 \end{equation}
 
-In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
-Independent errors \(\varepsilon_{ijt}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
-In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy'').
-Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B).
-With this notation, the linear model then can be expressed as
+In this model \(y_{ijt}\) is the response for subject \(i\), in
+treatment group \(j\) at time \(t\), which can be decomposed in a mean
+value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment
+(\(treatment_j\)) and their interaction \(time_t*treatment_j\) which
+have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\),
+respectively. Independent errors \(\varepsilon_{ijt}\) represent random
+variation not explained by the \emph{fixed} effects, and are assumed to
+be \(\sim N(0,\sigma^2)\) (independently and identically normally
+distributed with mean zero and variance \(\sigma^2\)). In a biomedical
+research context, suppose two treatments groups are used in a study
+(e.g., ``placebo'' vs.~``novel drug'' or ``saline''
+vs.~``chemotherapy''). Then, the group terms in Equation
+@ref(eq:linear-model) can be written as below with \(treatment_j=0\)
+representing the first treatment group (Group A) and \(treatment_j=1\)
+representing the second treatment group (Group B). With this notation,
+the linear model then can be expressed as
 
@@
 \end{cases}
-\label{eq:ANOVA-by-group}
+(\#eq:ANOVA-by-group)
 \end{equation}
 
-To further simplify the expression, substitute \(\widetilde{\beta_{0}}=\beta_0+\beta_{2}\) and \(\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}\) in the equation for Group B.
-This substitution allows for a different intercept and slope for Groups A and B.
-The model is then written as
+To further simplify the expression, substitute
+\(\widetilde{\beta_{0}}=\beta_0+\beta_{2}\) and
+\(\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}\) in the equation for Group
+B. This substitution allows for a different intercept and slope for
+Groups A and B. The model is then written as
 
@@
 \end{cases}
-\label{eq:ANOVA-lines}
+(\#eq:ANOVA-lines)
 \end{equation}
 
-Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response, this means that either presented as Equation \eqref{eq:linear-model}, Equation \eqref{eq:ANOVA-by-group} or Equation \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
+Presenting the model in this manner makes clear that when treating
+different groups, an rm-ANOVA model is able to accommodate non-parallel
+lines in each case (different intercepts and slopes per group). In other
+words, the rm-ANOVA model ``expects'' a linear relationship between the
+covariates and the response, this means that either presented as
+Equation @ref(eq:linear-model), Equation @ref(eq:ANOVA-by-group) or
+Equation @ref(eq:ANOVA-lines), an rm-ANOVA model is only able to
+accommodate linear patterns in the data. If the data show non-linear
+behavior, the rm-ANOVA model will approximate this behavior with
+non-parallel lines.
 
 \hypertarget{the-linear-mixed-model-case-lmem}{%
-\subsubsection{The Linear Mixed Model Case (LMEM)}\label{the-linear-mixed-model-case-lmem}}
-
-A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish {[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}.
-A LMEM with interaction between time and treatment for a longitudinal study can be written as:
+\subsubsection{The Linear Mixed Model Case
+(LMEM)}\label{the-linear-mixed-model-case-lmem}}
+
+A LMEM is a class of statistical models that incorporates \emph{fixed
+effects} to model the relationship between the covariates and the
+response, and \emph{random effects} to model subject variability that is
+not the primary focus of the study but that might be important to
+distinguish \citep{pinheiro2006, west2014}. A LMEM with interaction
+between time and treatment for a longitudinal study can be written as:
 
@@
 y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
-\label{eq:LMEM}
+(\#eq:LMEM)
 \end{equation}
 
-When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)).
-This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)).
-The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the ``global noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.
-
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data.
-As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
-The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
+When Equation @ref(eq:linear-model) and Equation @ref(eq:LMEM) are
+compared, it is easily noticeable that LMEM and rm-ANOVA have the same
+construction regarding the \emph{fixed effects} of time and treatment,
+but that the LMEM incorporates an additional source of variation (the
+term \(\mu_{ij}\)). This term \(\mu_{ij}\) is the one that corresponds
+to the \emph{random effect}, accounting for variability in each subject
+(subject\(_i\)) within each group (group\(_j\)). The \emph{random}
+component can also be understood as used to model some ``noise'' in the
+response, but that is intended to be analyzed and disentangled from the
+``global noise'' term \(\varepsilon_{ijt}\) from Equation
+@ref(eq:linear-model).
+
+For example, if the blood concentration of the drug is measured in
+certain subjects in the early hours of the morning while other subjects
+are measured in the afternoon, it is possible that the difference in the
+collection time introduces some ``noise'' in the data. As the name
+suggests, this ``random'' variability needs to be modeled as a variable
+rather than as a constant value. The \emph{random effect} \(\mu_{ij}\)
+in Equation @ref(eq:LMEM) is assumed to be
+\(\mu_{ij} \sim N(0,\sigma^2_\mu)\). In essence, the \emph{random
+effect} in a LMEM enables to fit models with different slopes at the
+subject-level\citep{pinheiro2006}. However, the expected linear
+relationship of the covariates and the response in Equation
+@ref(eq:linear-model) and in Equation @ref(eq:LMEM) is essentially the
+same, representing a major limitation of LMEMs to fit a non-linear
+response.
 
 \hypertarget{covariance-in-rm-anova-and-lmems}{%
-\subsection{Covariance in rm-ANOVA and LMEMs}\label{covariance-in-rm-anova-and-lmems}}
-
-In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{41}{]} (For an in-depth analysis of the covariance matrix see {[}\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}{]}).
-
-In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}{]}.
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}\protect\hyperlink{ref-maxwell2017}{44}{]}; and it is higher between consecutive measurements {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and they are not robust if the group sizes are unbalanced {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
-Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
+\subsection{Covariance in rm-ANOVA and
+LMEMs}\label{covariance-in-rm-anova-and-lmems}}
+
+In a longitudinal study there is an expected \emph{covariance} between
+repeated measurements on the same subject, and because repeated measures
+occur in the subjects within each group, there is a \emph{covariance}
+between measurements at each time point within each group. The
+\emph{covariance matrix} (also known as the variance-covariance matrix)
+is a matrix that captures the variation between and within subjects in a
+longitudinal study\citep{wolfinger1996} (For an in-depth analysis of the
+covariance matrix see \citep{west2014, weiss2005}).
+
+In the case of an rm-ANOVA analysis, it is typically assumed that the
+covariance matrix has a specific construction known as \emph{compound
+symmetry} (also known as ``sphericity'' or ``circularity''). Under this
+assumption, the between-subject variance and within-subject correlation
+are constant across time \citep{weiss2005, geisser1958, huynh1976}.
+However, it has been shown that this condition is frequently not
+justified because the correlation between measurements tends to change
+over time \citep{maxwell2017}; and it is higher between consecutive
+measurements \citep{gueorguieva2004, ugrinowitsch2004}. Although
+corrections can be made (such as Huyhn-Feldt or
+Greenhouse-Geisser)\citep{huynh1976, greenhouse1959} the effectiveness
+of each correction is limited because it depends on the size of the
+sample,the number of repeated measurements\citep{haverkamp2017}, and
+they are not robust if the group sizes are unbalanced
+\citep{keselman2001}. Because biomedical longitudinal studies are often
+limited in sample size and can have an imbalanced design, the
+corrections required to use an rm-ANOVA model may not be able to provide
+a reasonable adjustment that makes the model valid.
+
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow
+different structures for the variance-covariance matrix including
+exponential, autoregressive of order 1, rational quadratic and others
+\citep{pinheiro2006}. Nevertheless, the analysis required to determine
+an appropriate variance-covariance structure for the data can be a
+challenging process by itself. Overall, the spherical assumption for
+rm-ANOVA may not capture the natural variations of the correlation in
+the data, and can bias the inferences from the analysis.
 
@@
 
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}\protect\hyperlink{ref-weiss2005}{42}{]}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{46}{]}.
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{47}{]}, and in the case of MNAR the missing values are dependent on their value.
-
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{48}{]}.
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}\protect\hyperlink{ref-west2014}{40}{]}.
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
+Missing observations are an issue that arises frequently in longitudinal
+studies. In biomedical research, this situation can be caused by reasons
+beyond the control of the investigator \citep{molenberghs2004}. Dropout
+from patients and attrition or injury in animals are among the reasons
+for missing observations. Statistically, missing information can be
+classified as \emph{missing at random} (MAR), \emph{missing completely
+at random} (MCAR), and \emph{missing not at random} (MNAR)
+\citep{weiss2005}. In a MAR scenario, the pattern of the missing
+information is related to some variable in the data, but it is not
+related to the variable of interest \citep{scheffer2002}. If the data
+are MCAR, this means that the missingness is completely unrelated to the
+collected information \citep{potthoff2006}, and in the case of MNAR the
+missing values are dependent on their value.
+
+An rm-ANOVA model assumes complete observations for all subjects, and
+therefore subjects with one or more missing observations are excluded
+from the analysis. This is inconvenient because the remaining subjects
+might not accurately represent the population, and statistical power is
+affected by this reduction in sample size \citep{ma2012}. In the case of
+LMEMs, inferences from the model are valid when missing observations in
+the data exist that are MAR or MCAR \citep{west2014}. For example, if
+attrition occurs in all mice that had lower weights at the beginning of
+a chemotherapy response study, the missing data can be considered MAR
+because the missigness is unrelated to other variables of interest.
 
 \hypertarget{simulation}{%
-\subsection{What do an rm-ANOVA and LMEM fit look like? A visual representation using simulated data}\label{simulation}}
-
-To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{49}{]}.
-
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}C).
-In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments where a strong non-linear trend is present.
-
-Specifically, the rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect.
-In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group.
-From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D).
-
-A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal appears in Figure \ref{fig:linear-cases-Appendix} and Figure \ref{fig:quadratic-cases-Appendix} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs.
-Of notice, in Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/l-q-response-1}
+\subsection{What do an rm-ANOVA and LMEM fit look like? A visual
+representation using simulated data}\label{simulation}}
+
+To visually demonstrate the limitations of rm-ANOVA an LMEMs for
+non-linear longitudinal data, this section presents a simulation
+experiment of a normally distributed response of two groups of 10
+subjects each. An rm-ANOVA model (Equation @ref(eq:linear-model)), and a
+LMEM (Equation @ref(eq:LMEM)) are fitted to each group, using
+\(\textsf{R}\) \citep{r} and the package \emph{nlme} \citep{nlme}.
+
+Briefly, two cases for the mean responses for each group are considered:
+in the first case, the mean response in each group is a linear function
+over time with different intercepts and slopes; a negative slope is used
+for Group 1 and a positive slope is used for Group 2 (Figure
+@ref(fig:l-q-response)A). In the second case, a second-degree polynomial
+(quadratic) function is used for the mean response per group: the
+quadratic function is concave down for Group 1 and it is concave up for
+Group 2 (Figure @ref(fig:l-q-response)C). In both the linear and
+quadratic simulated data, the groups start with the same mean value at
+the first time point. This is intentional in order to simulate the
+expected temporal evolution of some physiological quantity, which is
+typical in biomedical experiments where a strong non-linear trend is
+present.
+
+Specifically, the rationale for the chosen linear and quadratic
+functions is the expectation that a measured response in two treatment
+groups is similar in the initial phase of the study, but as therapy
+progresses a divergence in the trend of the response indicates a
+treatment effect. In other words, Group 1 can be thought as a
+``Control'' group and Group 2 as a ``Treatment'' group. From the mean
+response per group (linear or quadratic), the variability or ``error''
+of individual responses within each group is simulated using a
+covariance matrix with compound symmetry (constant variance across
+time). Thus, the response per subject in both the linear and quadratic
+simulation corresponds to the mean response per group plus the error
+(Figure @ref(fig:l-q-response) B,D).
+
+A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for
+linear and non-linear longitudinal appears in Figure
+@ref(fig:linear-cases-Appendix) and Figure
+@ref(fig:quadratic-cases-Appendix) in the Appendix, where simulation
+with compound symmetry and independent errors (errors generated from a
+normal distribution that are not constant over time) and the plot of
+simulated errors, and fitted parameters in presented. We are aware that
+the simulated data used in this section present an extreme case that
+might not occur frequently in biomedical research, but they are used as
+a representation of the consequences of modeling non-linear data with a
+linear model such as rm-ANOVA or LMEMs. Of notice, in Section
+@ref(longitudinal-GAMs) we use simulated data that does follow reported
+trends in the biomedical literature to implement GAMs.
+
+(ref:l-q-response-caption) Simulated responses from two groups with
+correlated errors using a LMEM and a rm-ANOVA model. Top row: linear
+response, bottom row: quadratic response. A: Simulated linear data with
+known mean response (thin lines) and individual responses (points)
+showing the dispersion of the data. D: Simulated quadratic data with
+known mean response (thin lines) and individual responses (points)
+showing the dispersion of the data. B,E: Estimates from the rm-ANOVA
+model for the mean group response (linear of quadratic). Points
+represent the original raw data. The rm-ANOVA model not only fails to
+pick the trend of the quadratic data (D) but also assigns a global
+estimate that does not take between-subject variation. C, F: Estimates
+from the LMEM in the linear and quadratic case. The LMEM incorporates a
+random effect for each subject, but this model and the rm-ANOVA model
+are unable to follow the trend of the data and grossly bias the initial
+estimates for each group in the quadratic case (bottom row).
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/l-q-response-1}
 
@@
 
-\caption{Simulated responses from two groups with correlated errors using a LMEM and a rm-ANOVA model.
-Top row: linear response, bottom row: quadratic response.
-A: Simulated linear data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data.
-D: Simulated quadratic data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data.
-B,E: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
-Points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation.
-C, F: Estimates from the LMEM in the linear and quadratic case.
-The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data and grossly bias the initial estimates for each group in the quadratic case (bottom row).}\label{fig:l-q-response}
+\caption{(ref:l-q-response-caption)}\label{fig:l-q-response}
 \end{figure}
 
-The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response}A).
-When the linearity and compound symmetry assumptions are met, the rm-ANOVA model approximates well the global trend by group (Figure \ref{fig:l-q-response}B).
-Note that because the LMEM incorporates \emph{random effects}, is able to provide estimates for each subject and a ``global'' estimate (Figure \ref{fig:l-q-response}C).
-
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}D.
-Here, the mean response per group was simulated using a quadratic function, and errors and individual responses were produced as in Figure \ref{fig:l-q-response}A.
-The mean response in the simulated data with quadratic behavior changes in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model (Equation \eqref{eq:linear-model}) or a LMEM (Equation \eqref{eq:LMEM}) to this data produces the fit that appears in Figure \ref{fig:l-q-response}E, F.
-
-Comparing the fitted responses of the LMEM and the rm-ANOVA models used in the simulated quadratic data (Figure \ref{fig:l-q-response}E, F) indicates that the models are not capturing the changes within each group.
-Specifically, note that the fitted mean response of both models shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.
-The LMEM is only able to account for between-subject variation by providing estimates for each subject (Figure \ref{fig:l-q-response}F), but both models are unable to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare Figure \ref{fig:l-q-response}D with Figure \ref{fig:l-q-response}E, F).
-If such a change has important physiological implications, both rm-ANOVA and LMEMs omit it from the fitted mean response.
-Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
-
-This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-The models fitted to the simulated data were an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
-In the following section, we present generalized additive models (GAMs) as a data-driven alternative method to analyze longitudinal non-linear data that overcomes the linearity assumption.
-
-\FloatBarrier
+The simulation shows that the fits produced by the LMEM and the rm-ANOVA
+model are good for linear data, as the predictions for the mean response
+are reasonably close to the ``truth'' of the simulated data (Figure
+@ref(fig:l-q-response)A). When the linearity and compound symmetry
+assumptions are met, the rm-ANOVA model approximates well the global
+trend by group (Figure @ref(fig:l-q-response)B). Note that because the
+LMEM incorporates \emph{random effects}, is able to provide estimates
+for each subject and a ``global'' estimate (Figure
+@ref(fig:l-q-response)C).
+
+However, consider the case when the data follows a non-linear trend,
+such as the simulated data in Figure @ref(fig:l-q-response)D. Here, the
+mean response per group was simulated using a quadratic function, and
+errors and individual responses were produced as in Figure
+@ref(fig:l-q-response)A. The mean response in the simulated data with
+quadratic behavior changes in each group through the timeline, and the
+mean value is the same as the initial value by the fifth time point for
+each group. Fitting an rm-ANOVA model (Equation @ref(eq:linear-model))
+or a LMEM (Equation @ref(eq:LMEM)) to this data produces the fit that
+appears in Figure @ref(fig:l-q-response)E, F.
+
+Comparing the fitted responses of the LMEM and the rm-ANOVA models used
+in the simulated quadratic data (Figure @ref(fig:l-q-response)E, F)
+indicates that the models are not capturing the changes within each
+group. Specifically, note that the fitted mean response of both models
+shows that the change (increase for Treatment 1 or decrease for
+Treatment 2) in the response through time points 2 and 4 is not being
+captured. The LMEM is only able to account for between-subject variation
+by providing estimates for each subject (Figure
+@ref(fig:l-q-response)F), but both models are unable to capture the fact
+that the initial values are the same in each group, and instead fit
+non-parallel lines that have initial values that are markedly different
+from the ``true'' initial values in each case (compare Figure
+@ref(fig:l-q-response)D with Figure @ref(fig:l-q-response)E, F). If such
+a change has important physiological implications, both rm-ANOVA and
+LMEMs omit it from the fitted mean response. Thus, even though the model
+correctly detects a divergence between treatment groups, the exact
+nature of this difference is not correctly identified, limiting valuable
+inferences from the data.
+
+This section has used simulation to better convey the limitations of
+linearity and correlation in the response in non-linear data. The models
+fitted to the simulated data were an rm-ANOVA model and a LMEM, where
+the main issue is the expected linear trend in the response. In the
+following section, we present generalized additive models (GAMs) as a
+data-driven alternative method to analyze longitudinal non-linear data
+that overcomes the linearity assumption.
 
 \hypertarget{GAM-theory}{%
-\section{GAMs as a special case of Generalized Linear Models}\label{GAM-theory}}
+\section{GAMs as a special case of Generalized Linear
+Models}\label{GAM-theory}}
 
@@
 
-Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors {[}\protect\hyperlink{ref-nelder1972}{50}{]}.
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{51}{]}.
-A GAM model can be written as:
+Generalized linear models (GLMs) are a family of models (which include
+rm-ANOVA and LMEMs) that fit a linear response function to data that may
+not have normally distributed errors \citep{nelder1972}. In contrast,
+GAMs are a family of regression-based methods for estimating smoothly
+varying trends and are a broader class of models that contain the GLM
+family as a special case\citep{simpson2018, wood2017, hastie1987}. A GAM
+model can be written as:
 
@@
   y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
-  \label{eq:GAM}
+  (\#eq:GAM)
 \end{equation}
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible \emph{smooth functions}.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
-In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
-
-The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{52}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
-In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
-
-Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
-Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-The timeline can be divided in equally spaced \emph{knots}, each knot being a region where a different set of basis functions will be used.
-Because there are six timepoints for this group, five knots can be used.
-The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is set using default values in the package \emph{mgcv} depending on the number of knots.
-In Figure \ref{fig:basis-plot}A, the four basis functions (and the intercept) are shown.
-Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the ``wiggliness'' of the fit, each of the basis functions of Figure \ref{fig:basis-plot}A is weighted by multiplying it by a coefficient according to the matrix of Figure \ref{fig:basis-plot}B.
-The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the ``wiggliness'' of the smooth fit to prevent overfitting.
-A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
-
-To get the weighted basis functions, each basis (from Figure Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
-Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1).
-On the other hand, the third basis function has roughly doubled its value.
-Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (orange line), along the simulated values per group, which appear as points.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-1}
+Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in
+group \(j\), \(\beta_0\) is the expected value at time 0, the change of
+\(y_{ijt}\) over time is represented by the \emph{smooth function}
+\(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and
+parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the
+residual error.
+
+In contrast to the linear functions used to model the relationship
+between the covariates and the response in rm-ANOVA or LMEM, GAMs use
+more flexible \emph{smooth functions}. This approach is advantageous as
+it does not restrict the model to a linear relationship, although a GAM
+can estimate a linear relationship if the data is consistent with a
+linear response. One possible set of functions for
+\(f(x_t\mid \beta_j)\) that allow for non-linear responses are
+polynomials, but a major limitation is that polynomials create a
+``global'' fit as they assume that the same relationship exists
+everywhere, which can cause problems with inference \citep{beck1998}. In
+particular, polynomial fits are known to show boundary effects because
+as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to
+\(\pm \infty\) which is almost always unrealistic and causes bias at the
+endpoints of the time period.
+
+The smooth functional relationship between the covariates and the
+response in GAMs is specified using a semi-parametric relationship that
+can be fit within the GLM framework, by using \emph{basis function}
+expansions of the covariates and by estimating random coefficients
+associated with these basis functions. A \emph{basis} is a set of
+functions that spans the mathematical space where the smooths that
+approximate \(f(x_t\mid \beta_j)\) exist \citep{simpson2018}. For the
+linear model in Equation @ref(eq:linear-model), the basis coefficients
+are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are
+\(time_t\), \(treatment_j\) and \(time_t \times treatment_j\). The basis
+function then, is the combination of basis coefficients and basis
+vectors that map the possible relationship between the covariates and
+the response \citep{hefley2017}, which in the case of Equation
+@ref(eq:linear-model) is restricted to a linear family of functions. In
+the case of Equation @ref(eq:GAM), the basis functions are contained in
+the expression \(f(x_t\mid \beta_j)\), which means that the model allows
+for non-linear relationships among the covariates.
+
+Splines (cubic, thin plate, etc.) are commonly used \emph{basis
+functions}; a cubic spline is a smooth curve constructed from cubic
+polynomials joined together in a manner that enforces smoothness, and
+thin plate regression splines are an optimized version that work well
+with noisy data \citep{wood2017, simpson2018}. Splines have a long
+history in solving semi-parametric statistical problems and are often a
+default choice to fit GAMs as they are a simple, flexible and powerful
+option to obtain smoothness \citep{wegman1983}. Therefore, this
+data-driven flexibility in GAMs overcomes the limitation that occurs in
+LMEMs and rm-ANOVA when the data is non linear.
+
+To further clarify the concept of basis functions and smooth functions,
+consider the simulated response for Group 1 in Figure
+@ref(fig:l-q-response)C. The simplest GAM model that can be used to
+estimate such response is that of a single smooth term for the time
+effect; i.e., a model that fits a smooth to the trend of the group
+through time. The timeline can be divided in equally spaced
+\emph{knots}, each knot being a region where a different set of basis
+functions will be used. Because there are six timepoints for this group,
+five knots can be used. The model with five knots to construct the
+smooth term means that it will have four basis functions (plus one that
+corresponds to the intercept). The choice of basis functions is set
+using default values in the package \emph{mgcv} depending on the number
+of knots. In Figure @ref(fig:basis-plot)A, the four basis functions (and
+the intercept) are shown. Each of the basis functions is composed of six
+different points (because there are six points on the timeline). To
+control the ``wiggliness'' of the fit, each of the basis functions of
+Figure @ref(fig:basis-plot)A is weighted by multiplying it by a
+coefficient according to the matrix of Figure @ref(fig:basis-plot)B. The
+parameter estimates are penalized (shrunk towards 0) where the penalty
+reduces the ``wiggliness'' of the smooth fit to prevent overfitting. A
+weak penalty estimate will result in wiggly functions whereas a strong
+penalty estimate provides evidence that a linear response is
+appropriate.
+
+To get the weighted basis functions, each basis (from Figure Figure
+@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients
+in Figure @ref(fig:basis-plot)B, thereby increasing or decreasing the
+original basis functions. Figure @ref(fig:basis-plot)C shows the
+resulting weighted basis functions. Note that the magnitude of the
+weighting for the first basis function has resulted in a decrease of its
+overall value (because the coefficient for that basis function is less
+than 1). On the other hand, the third basis function has roughly doubled
+its value. Finally, the weighted basis functions are added at each
+timepoint to produce the smooth term. The resulting smooth term for the
+effect of \emph{time} is shown in Figure @ref(fig:basis-plot)D (orange
+line), along the simulated values per group, which appear as points.
+
+(ref:basis-plot-caption) Basis functions for a single smoother for time
+with five knots. A: Basis functions for a single smoother for time for
+the simulated data of Group 1 from Figure 2. B: Matrix of basis function
+weights. Each basis function is multiplied by a coefficient which can be
+positive or negative. The coefficient determines the overall effect of
+each basis in the final smoother. C: Weighted basis functions. Each of
+the four basis functions of panel A has been weighted by the
+corresponding coefficient shown in Panel B. Note the corresponding
+increase (or decrease) in magnitude of each weighted basis function. D:
+Smoother for time and original data points. The smoother (line) is the
+result of the sum of each weighted basis function at each time point,
+with simulated values for the group shown as points.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/basis-plot-1}
 
@@
 
-\caption{Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-B: Matrix of basis function weights.
-Each basis function is multiplied by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
-Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
+\caption{(ref:basis-plot-caption)}\label{fig:basis-plot}
 \end{figure}
@@
 \hypertarget{a-bayesian-interpretation-of-gams}{%
-\section{A Bayesian interpretation of GAMs}\label{a-bayesian-interpretation-of-gams}}
-
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}\protect\hyperlink{ref-mcelreath2018}{54}{]}.
-In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) {[}\protect\hyperlink{ref-miller2019}{55}{]}.
-Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{56}{]}.
-Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{55}{]} and {[}\protect\hyperlink{ref-marra2012}{57}{]}.
-With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
-
-\FloatBarrier
+\section{A Bayesian interpretation of
+GAMs}\label{a-bayesian-interpretation-of-gams}}
+
+Bayes' theorem states that the probability of an event can be calculated
+using prior knowledge or belief \citep{mcelreath2018}. In the case of
+non-linear data, the belief that the \emph{true} trend of the data is
+likely to be smooth rather than ``wiggly'' introduces the concept of a
+prior distribution for wiggliness (and therefore a Bayesian view) of
+GAMs \citep{wood2017}. GAMs are considered ``empirical'' Bayesian models
+because the smoothing parameters are estimated from the data (and not
+from a prior distribution as in the ``Full Bayes'' case)
+\citep{miller2019}. Moreover, the use of the restricted maximum
+likelihood (REML) to estimate the smoothing parameters gives an
+empirical estimate of the smooth model \citep{laird1982, pedersen2019}.
+Therefore, the confidence intervals calculated for the smooth terms
+using the package \emph{mgcv} are considered empirical Bayesian
+posterior credible intervals \citep{pedersen2019}, which have good
+``frequentist'' coverage (pointwise coverage or ``single point''
+coverage), and \emph{across the function} coverage \citep{wood2017}.
+This last part means that contrary to a pointwise coverage (where the
+coverage of the interval is correct for a single point) the estimated
+confidence intervals for the smooths will contain \emph{on average} the
+true function of the data 95\% of the time across the entire timeline
+(in the case of longitudinal data for which smooths are calculated),
+which allows to obtain better inference from the model. In-depth theory
+of the Bayesian interpretation of GAMs is beyond the scope of this
+paper, but can be found in \citep{miller2019, wood2017, simpson2018} and
+\citep{marra2012}. With this brief introduction to the Bayesian
+interpretation of GAMs, we henceforth refer to the confidence intervals
+for the smooths in GAMs as ``empirical Bayesian'' through the rest of
+this paper.
 
 \hypertarget{longitudinal-GAMs}{%
-\section{The analyisis of longitudinal biomedical data using GAMs}\label{longitudinal-GAMs}}
-
-The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs.
-This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}.
-The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
+\section{The analyisis of longitudinal biomedical data using
+GAMs}\label{longitudinal-GAMs}}
+
+The previous sections provided the basic framework to understand the GAM
+framework and how these models are more advantageous to analyze
+non-linear longitudinal data when compared to rm-ANOVA or LMEMs. This
+section will use simulation to present the practical implementation of
+GAMs for longitudinal biomedical data using \(\textsf{R}\) and the
+package \texttt{mgcv}. The code for the simulated data and figures, and
+a brief guide for model selection and diagnostics appear in the
+Appendix.
 
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
-In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group.
-Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
-The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}A and the inset, respectively.
+The simulated data is based on the reported longitudinal changes in
+oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear
+in Figure 3C in \citep{vishwanath2009}. In the paper, diffuse
+reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes
+in both groups at the same time points (days 0, 2, 5, 7 and 10). In the
+``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is
+observed through time, while a decrease is seen in the ``Control''
+(saline) group. Following the reported trend, we simulated 10 normally
+distributed observations at each time point with a standard deviation
+(SD) of 10\% (matching the SD in the original paper). The simulated and
+real data appear in Figure @ref(fig:sim-smooth-plot)A and the inset,
+respectively.
 
 \hypertarget{an-interaction-gam-for-longitudinal-data}{%
-\subsection{An interaction GAM for longitudinal data}\label{an-interaction-gam-for-longitudinal-data}}
-
-An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination.
-In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time.
-Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in {[}\protect\hyperlink{ref-hefley2017}{52}{]}.
-
-For the data in Figure \ref{fig:sim-smooth-plot}, A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
-To estimate this, the model incorporates independent smooths for \emph{Group} and \emph{Day}, respectively.
-The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
-Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:
-
-\passthrough{\lstinline!m1 <- gam(StO2\_sim \~ Group + s(Day, by=Group, k=5), method='REML, data = dat\_sim)!}
-
-This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
-The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{58}{]}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{59}{]}.
-
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
-This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
-
-Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
-The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
-However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
-Consider the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}B.
-If 40\% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \ref{fig:sim-smooth-plot}D).
-Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-1}
+\subsection{An interaction GAM for longitudinal
+data}\label{an-interaction-gam-for-longitudinal-data}}
+
+An interaction effect is typically the main interest in longitudinal
+biomedical data, as it takes into account treatment, time, and their
+combination. In a practical sense, when a GAM is implemented for
+longitudinal data, a smooth can be added to the model for the
+\emph{time} effect to account for the repeated measures over time.
+Although specific methods of how GAMs model correlation structures is a
+topic beyond the scope of this paper, it suffices to say that GAMs are
+flexible and can handle correlation structures beyond compound symmetry.
+A detailed description on basis functions and correlations can be found
+in \citep{hefley2017}.
+
+For the data in Figure @ref(fig:sim-smooth-plot), A the main effect of
+interest is how \(\mbox{StO}_2\) changes over time for each treatment.
+To estimate this, the model incorporates independent smooths for
+\emph{Group} and \emph{Day}, respectively. The main thing to consider is
+that model syntax accounts for the fact that one of the variables is
+numeric (\emph{Day}) and the other is a factor (\emph{Group}). Because
+the smooths are centered at 0, the factor variable needs to be specified
+as a parametric term in order to identify any differences between the
+groups. Using \(\textsf{R}\) and the package \texttt{mgcv} the model
+syntax is:
+
+\texttt{m1\ \textless{}-\ gam(StO2\_sim\ \textasciitilde{}\ Group\ +\ s(Day,\ by=Group,\ k=5),\ method=\textquotesingle{}REML,\ data\ =\ dat\_sim)}
+
+This syntax specifies that \texttt{m1} will store the model, and that
+the change in the simulated oxygen saturation (\texttt{StO2\_sim}) is
+modeled using independent smooths over \texttt{Day} for each
+\texttt{Group} (the parenthesis preceded by \texttt{s}) using 5 knots.
+The smooth is constructed by default using thin plate regression
+splines, but other splines can be used if desired, including Gaussian
+process smooths \citep{simpson2018}. The parametric term \texttt{Group}
+is added to quantify overall mean differences in the effect of treatment
+between groups, and the \texttt{method} chosen to estimate the smoothing
+parameters is the restricted maximum likelihood (REML) \citep{wood2017}.
+When the smooths are plotted over the raw data, it is clear that the
+model has been able to capture the trend of the change of
+\(\mbox{StO}_2\) for each group across time (Figure
+@ref(fig:sim-smooth-plot)B). Model diagnostics can be obtained using the
+\texttt{gam.check} function, and the function \texttt{appraise} from the
+package \emph{gratia} \citep{gratia}. A guide for model selection and
+diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an
+in-depth analysis can be found in \citep{wood2017} and
+\citep{harezlak2018}.
+
+One question that might arise at this point is ``what is the fit that an
+rm-ANOVA model produces for the simulated data?''. The rm-ANOVA model,
+which corresponds to Equation @ref(eq:linear-model) is presented in
+Figure @ref(fig:sim-smooth-plot)C. This is a typical case of model
+misspecification: The slopes of each group are different, which would
+lead to a \emph{p-value} indicating significance for the treatment and
+time effects, but the model is not capturing the changes that occur at
+days 2 and between days 5 and 7, whereas the GAM model is able to
+reliably estimate the trend over all timepoints (Figure
+@ref(fig:sim-smooth-plot)B) .
+
+Because GAMs do not require equally-spaced or complete observations for
+all subjects, they are advantageous to analyze longitudinal data where
+missingness exists. The rationale behind this is that GAMs are able to
+pick the trend in the data even when some observations are missing.
+However, this usually causes the resulting smooths to have wider
+confidence intervals and less ability to pick certain trends. Consider
+the simulated \(\mbox{StO}_2\) values from Figure
+@ref(fig:sim-smooth-plot)B. If 40\% of the total observations are
+randomly deleted and the same interaction GAM fitted for the complete
+dataset is used, the resulting smooths are still able to show a
+different trend for each group, but because the empirical Bayesian
+credible intervals for the smooths overlap during the first 3 days with
+fewer data points, the trend is less pronounced than in the full dataset
+(Figure @ref(fig:sim-smooth-plot)D). Although the confidence intervals
+have increased for both smooths, the model still shows different trends
+with as little as 4 observations per group at certain time points.
+
+(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation
+in tumors. A: Simulated data that follows previously reported trends
+(inset) in tumors under chemotherapy (Treatment) or saline (Control)
+treatment. Simulated data is from a normal distribution with standard
+deviation of 10\% with 10 observations per time point. Lines indicate
+mean oxygen saturation B: Smooths from the GAM model for the full
+simulated data with interaction of Group and Treatment. Lines represent
+trends for each group, shaded regions are 95\% confidence intervals. C:
+The rm-ANOVA model for the simulated data, which does not capture the
+changes in each group over time. D: Smooths for the GAM model for the
+simulated data with 40\% of its observations missing. Lines represent
+trends for each group, shaded regions are 95\% empirical Bayesian
+confidence intervals.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-\caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot}
+\caption{(ref:sim-plot-caption)}\label{fig:sim-smooth-plot}
 \end{figure}
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-1}
+(ref:pairwise-comp) Pairwise comparisons for smooth terms. A: Pairwise
+comparisons for the full dataset. B: Pairwise comparisons for the
+dataset with missing observations. Significant differences exist where
+the 95\% empirical Bayesian credible interval does not cover 0. In both
+cases the effect of treatment is significant after day 3.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/plot-pairwise-comp-1}
 
@@
 
-\caption{Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-B: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
-In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp}
+\caption{(ref:pairwise-comp)}\label{fig:plot-pairwise-comp}
 \end{figure}
@@
 \hypertarget{GAM-significance}{%
-\subsection{Determination of significance in GAMs for longitudinal data}\label{GAM-significance}}
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the \emph{difference} between the empirical Bayesian confidence intervals of the fitted smooths for such groups is non-zero, then a significant difference exists at that time point(s).
-The absence of a \emph{p-value} in this case might seem odd, but the empirical Bayesian confidence interval interpretation can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
-This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
-
-With this expectation of different trends in each group, computing the difference between the trends will identify if the observed change is significant.
-The difference between groups with similar trends is likely to yield zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
-
-Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}B and Figure \ref{fig:sim-smooth-plot}D.
-Figure \ref{fig:plot-pairwise-comp} shows the comparison between each treatment group for the full and missing datasets.
-Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared.
-Of notice, the pairwise comparison has been set on the response scale (see Appendix for code details), because otherwise the comparison appears shifted and is not intuitively easy to relate to the original data.
-
-With this correction in mind, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by \(\approx\) 3 day it is the ``Treatment'' group the one that on average, has greater \(\mbox{StO}_2\).
-This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the given model.
-Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in mean \(\mbox{StO}_2\) occurs.
-
-On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the empirical Bayesian credible intervals of the smooths overlap between days 0 and 3.
-Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that there is no evidence of a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.
-
-In a sense, the pairwise smooth comparison is more informative than a \emph{post-hoc} \emph{p-value}.
-For biomedical studies, the smooth comparison is able to provide an estimate of \emph{when} and by \emph{how much} a biological process becomes significant.
-This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.
-
-\FloatBarrier
+\subsection{Determination of significance in GAMs for longitudinal
+data}\label{GAM-significance}}
+
+At the core of a biomedical longitudinal study lies the question of a
+significant difference between the effect of two or more treatments in
+different groups. Whereas in rm-ANOVA a \emph{post-hoc} analysis is
+required to answer such question by calculating some \emph{p-values}
+after multiple comparisons, GAMs can use a different approach to
+estimate significance. In essence, the idea behind the estimation of
+significance in GAMs across different treatment groups is that if the
+\emph{difference} between the empirical Bayesian confidence intervals of
+the fitted smooths for such groups is non-zero, then a significant
+difference exists at that time point(s). The absence of a \emph{p-value}
+in this case might seem odd, but the empirical Bayesian confidence
+interval interpretation can be conceptualized in the following manner:
+Different trends in each group are an indication of an effect by the
+treatment. This is what happens for the simulated data in Figure
+@ref(fig:sim-smooth-plot)A, where the chemotherapy causes
+\(\mbox{StO}_2\) to increase over time.
+
+With this expectation of different trends in each group, computing the
+difference between the trends will identify if the observed change is
+significant. The difference between groups with similar trends is likely
+to yield zero, which would indicate that the treatment is not causing a
+change in the response in one of the groups (assuming the other group is
+a Control or Reference group).
+
+Consider the calculation of pairwise differences for the smooths in
+Figure @ref(fig:sim-smooth-plot)B and Figure @ref(fig:sim-smooth-plot)D.
+Figure @ref(fig:plot-pairwise-comp) shows the comparison between each
+treatment group for the full and missing datasets. Here, the ``Control''
+group is used as the reference to which ``Treatment'' group is being
+compared. Of notice, the pairwise comparison has been set on the
+response scale (see Appendix for code details), because otherwise the
+comparison appears shifted and is not intuitively easy to relate to the
+original data.
+
+With this correction in mind, the shaded regions over the confidence
+interval (where it does not cover 0) indicate the time interval where
+each group has a higher effect than the other. Notice that the shaded
+region between days 0 and \(\approx\) 2 for the full dataset indicates
+that through that time, the ``Control'' group has higher mean
+\(\mbox{StO}_2\), but as therapy progresses the effect is reversed and
+by \(\approx\) 3 day it is the ``Treatment'' group the one that on
+average, has greater \(\mbox{StO}_2\). This would suggest that the
+effect of chemotherapy in the ``Treatment'' group becomes significant
+after day 3 for the given model. Moreover, notice that although there is
+no actual measurement at day 3, the model is capable of providing an
+estimate of when the shift in mean \(\mbox{StO}_2\) occurs.
+
+On the data with missing observations (Figure
+@ref(fig:sim-smooth-plot)D), the empirical Bayesian credible intervals
+of the smooths overlap between days 0 and 3. Consequently, the smooth
+pairwise comparison (Figure @ref(fig:plot-pairwise-comp)B) shows that
+there is no evidence of a significant difference between the groups
+during that period, but is still able to pick the change on day 3 as the
+full dataset smooth pairwise comparison.
+
+In a sense, the pairwise smooth comparison is more informative than a
+\emph{post-hoc} \emph{p-value}. For biomedical studies, the smooth
+comparison is able to provide an estimate of \emph{when} and by
+\emph{how much} a biological process becomes significant. This is
+advantageous because it can help researchers gain insight on metabolic
+changes and other biological processes that can be worth examining, and
+can help refine the experimental design of future studies in order to
+obtain measurements at time points where a significant change might be
+expected.
 
@@
 
-Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
-Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}.
-This ``model misspecification'' error, also is known as a ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{60}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
-
-In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{61}{]}.
-This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
-
-However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} using \(\textsf{R}\) and the package \emph{mgcv}{[}\protect\hyperlink{ref-wood2017}{37}{]} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
-One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
-The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
-
-We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{62},\protect\hyperlink{ref-weissgerber2018}{63}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
-Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
-We have made all the data and code used in this paper accessible, and we hope that this will encourage other researchers to do the same with future projects.
-
-\FloatBarrier
+Biomedical longitudinal non-linear data is particularly challenging to
+analyze due to the likelihood of missing observations and different
+correlation structures in the data, which limit the use of rm-ANOVA.
+Although LMEMs have started to replace rm-ANOVA as the choice to analyze
+biomedical data, both methods yield biased estimates when they are used
+to fit non-linear data as we have visually demonstrated in Section
+@ref(simulation). This ``model misspecification'' error, also is known
+as a ``Type III'' error \citep{dennis2019} is particularly important
+because although the \emph{p-value} is the common measure of statistical
+significance, the validity of its interpretation is determined by the
+agreement of the data and the model. Guidelines for statistical
+reporting in biomedical journals exist (the SAMPL guidelines)
+\citep{lang2015} but they have not been widely adopted and in the case
+of longitudinal data, we consider that researchers would benefit from
+reporting a visual assessment of the correspondence between the model
+fit and the data, instead of merely relying on a \(R^2\) value.
+
+In this paper we have presented GAMs as a suitable method to analyze
+non-linear longitudinal data. It is interesting to note that although
+GAMs are a well established method to analyze temporal data in different
+fields (among which are palaeoecology, geochemistry, and ecology)
+\citep{hefley2017, pedersen2019} they are not routinely used in
+biomedical research despite an early publication from Hastie and
+Tibshirani that demonstrated their use in medical research
+\citep{hastie1995}. This is possibly due to the fact that the theory
+behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but
+the purpose of Section @ref(GAM-theory) is to demonstrate that at its
+core the theory quite simple: Instead of using a linear relationship to
+model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions
+to build smooths that are capable of following non-linear trends in the
+data.
+
+However, from a practical standpoint is equally important to demonstrate
+how GAMs are computationally implemented. We have provided an example on
+how GAMs can be fitted using simulated data that follows trends reported
+in biomedical literature \citep{vishwanath2009} using \(\textsf{R}\) and
+the package \emph{mgcv}\citep{wood2017} in Section
+@ref(longitudinal-GAMs), while a basic workflow for model selection is
+in the \protect\hyperlink{workflow}{Appendix}. One of the features of
+GAMs is that their Bayesian interpretation allows to indicate
+differences between groups without the need of a \emph{p-value}, and in
+turn provide a time-based estimate of shifts in the response that can be
+directly tied to biological values as the pairwise smooth comparisons in
+Figure @ref(fig:plot-pairwise-comp) indicate. The model is therefore
+able to provide an estimate of significant change between the groups at
+time points were data was not directly measured even with missing data
+exists ( \(\approx\) day 3 in Figure @ref(fig:plot-pairwise-comp) A, B
+), which can be used by researchers as feedback on experiment design and
+to further evaluate important biological changes in future studies.
+
+We have used \(\textsf{R}\) as the software of choice for this paper
+because not only provides a fully developed environment to fit GAMs, but
+also eases simulation (which is becoming increasingly used for
+exploratory statistical analysis and power calculations) and provides
+powerful and convenient methods of visualization, which are key aspects
+that biomedical researchers might need to consider to make their work
+reproducible. In this regard, reproducibility is still an issue in
+biomedical research \citep{begley2015, weissgerber2018}, but it is
+becoming apparent that what other disciplines have experienced in this
+aspect is likely to impact sooner rather than later this field.
+Researchers need to plan on how they will make their data, code, and any
+other materials open and accessible as more journals and funding
+agencies recognize the importance and benefits of open science in
+biomedical research. We have made all the data and code used in this
+paper accessible, and we hope that this will encourage other researchers
+to do the same with future projects.
 
@@
 
-We have presented GAMs as a method to analyze longitudinal biomedical data.
-Future directions of this work will include simulation-based estimations of statistical power using GAMs, as well as demonstrating the prediction capabilities of these models using large datasets.
-By making the data and code used in this paper accessible, we hope to address the need of creating and sharing reproducible work in biomedical research.
+We have presented GAMs as a method to analyze longitudinal biomedical
+data. Future directions of this work will include simulation-based
+estimations of statistical power using GAMs, as well as demonstrating
+the prediction capabilities of these models using large datasets. By
+making the data and code used in this paper accessible, we hope to
+address the need of creating and sharing reproducible work in biomedical
+research.
 
@@
 
-This work was supported by the National Science Foundation Career Award (CBET 1751554, TJM) and the Arkansas Biosciences Institute.
-
-\FloatBarrier
+This work was supported by the National Science Foundation Career Award
+(CBET 1751554, TJM) and the Arkansas Biosciences Institute.
 
@@
 
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\leavevmode\hypertarget{ref-roblyer2011}{}%
-\CSLLeftMargin{{[}1{]} }
-\CSLRightInline{D.
-Roblyer, S.
-Ueda, A.
-Cerussi, W.
-Tanamai, A.
-Durkin, R.
-Mehta, D.
-Hsiang, J.A.
-Butler, C.
-McLaren, W.P.
-Chen, B.
-Tromberg, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences of the United States of America.
-108 (2011) 14626--14631.
-\url{https://doi.org/10.1073/pnas.1013103108}.}
-
-\leavevmode\hypertarget{ref-tank2020}{}%
-\CSLLeftMargin{{[}2{]} }
-\CSLRightInline{A.
-Tank, H.M.
-Peterson, V.
-Pera, S.
-Tabassum, A.
-Leproux, T.
-O'Sullivan, E.
-Jones, H.
-Cabral, N.
-Ko, R.S.
-Mehta, B.J.
-Tromberg, D.
-Roblyer, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research.
-22 (2020) 1--10.
-\url{https://doi.org/doi:10.1186/s13058-020-01262-1}.}
-
-\leavevmode\hypertarget{ref-pavlov2018}{}%
-\CSLLeftMargin{{[}3{]} }
-\CSLRightInline{M.V.
-Pavlov, T.I.
-Kalganova, Y.S.
-Lyubimtseva, V.I.
-Plekhanov, G.Y.
-Golubyatnikov, O.Y.
-Ilyinskaya, A.G.
-Orlova, P.V.
-Subochev, D.V.
-Safonov, N.M.
-Shakhova, A.V.
-Maslennikova, {Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy}, {Journal of Biomedical Optics}.
-{23} (2018).
-\url{https://doi.org/\%7B10.1117/1.JBO.23.9.091410\%7D}.}
-
-\leavevmode\hypertarget{ref-demidov2018}{}%
-\CSLLeftMargin{{[}4{]} }
-\CSLRightInline{V.
-Demidov, A.
-Maeda, M.
-Sugita, V.
-Madge, S.
-Sadanand, C.
-Flueraru, I.A.
-Vitkin, {Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography}, {Scientific Reports}.
-{8} (2018).
-\url{https://doi.org/\%7B10.1038/s41598-017-18635-w\%7D}.}
-
-\leavevmode\hypertarget{ref-ritter2001}{}%
-\CSLLeftMargin{{[}5{]} }
-\CSLRightInline{G.
-Ritter, L.
-Cohen, C.
-Williams, E.
-Richards, L.
-Old, S.
-Welt, {Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}, {Cancer Research}.
-{61} (2001) 6851--6859.}
-
-\leavevmode\hypertarget{ref-roth2017}{}%
-\CSLLeftMargin{{[}6{]} }
-\CSLRightInline{E.M.
-Roth, A.C.
-Goldberg, A.L.
-Catapano, A.
-Torri, G.D.
-Yancopoulos, N.
-Stahl, A.
-Brunet, G.
-Lecorps, H.M.
-Colhoun, {Antidrug antibodies in atients treated with alirocumab}, {New England Journal of Medicine}.
-{376} (2017) 1589--1590.
-\url{https://doi.org/\%7B10.1056/NEJMc1616623\%7D}.}
-
-\leavevmode\hypertarget{ref-jones2018}{}%
-\CSLLeftMargin{{[}7{]} }
-\CSLRightInline{J.D.
-Jones, H.E.
-Ramser, A.E.
-Woessner, K.P.
-Quinn, {In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing}, {Communications Biology}.
-{1} (2018).
-\url{https://doi.org/\%7B10.1038/s42003-018-0206-4\%7D}.}
-
-\leavevmode\hypertarget{ref-skala2010}{}%
-\CSLLeftMargin{{[}8{]} }
-\CSLRightInline{M.C.
-Skala, A.
-Fontanella, L.
-Lan, J.A.
-Izatt, M.W.
-Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics.
-15 (2010).
-\url{https://doi.org/10.1117/1.3285584}.}
-
-\leavevmode\hypertarget{ref-greening2018}{}%
-\CSLLeftMargin{{[}9{]} }
-\CSLRightInline{G.J.
-Greening, K.P.
-Miller, C.R.
-Spainhour, M.D.
-Cato, T.J.
-Muldoon, {Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy}, {Biomedical Optics Express}.
-{9} (2018) 2871--2886.
-\url{https://doi.org/\%7B10.1364/BOE.9.002871\%7D}.}
-
-\leavevmode\hypertarget{ref-sio2016}{}%
-\CSLLeftMargin{{[}10{]} }
-\CSLRightInline{T.T.
-Sio, P.J.
-Atherton, B.J.
-Birckhead, D.J.
-Schwartz, J.A.
-Sloan, D.K.
-Seisler, J.A.
-Martenson, C.L.
-Loprinzi, P.C.
-Griffin, R.F.
-Morton, J.C.
-Anders, T.J.
-Stoffel, R.E.
-Haselow, R.B.
-Mowat, M.A.N.
-Wittich, J.D.
-Bearden III, R.C.
-Miller, {Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {{[}}alliance{]})}, {Supportive Care in Cancer}.
-{24} (2016) 3847--3855.
-\url{https://doi.org/\%7B10.1007/s00520-016-3213-3\%7D}.}
-
-\leavevmode\hypertarget{ref-kamstra2015}{}%
-\CSLLeftMargin{{[}11{]} }
-\CSLRightInline{J.I.
-Kamstra, P.U.
-Dijkstra, M.
-van Leeuwen, J.L.N.
-Roodenburg, J.A.
-Langendijk, {Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study}, {Oral Oncology}.
-{51} (2015) 548--555.
-\url{https://doi.org/\%7B10.1016/j.oraloncology.2015.01.016\%7D}.}
-
-\leavevmode\hypertarget{ref-wagenmakers2008}{}%
-\CSLLeftMargin{{[}12{]} }
-\CSLRightInline{E.-J.
-Wagenmakers, M.
-Lee, T.
-Lodewyckx, G.J.
-Iverson, Bayesian versus frequentist inference, in: H.
-Hoijtink, I.
-Klugkist, P.A.
-Boelen (Eds.), Bayesian Evaluation of Informative Hypotheses, Springer New York, New York, NY, 2008: pp.
-181--207.
-\url{https://doi.org/10.1007/978-0-387-09612-4_9}.}
-
-\leavevmode\hypertarget{ref-gueorguieva2004}{}%
-\CSLLeftMargin{{[}13{]} }
-\CSLRightInline{R.
-Gueorguieva, J.H.
-Krystal, {Move over ANOVA - Progress in analyzing repeated-measures data and its reflection in papers published in the archives of general psychiatry}, Archives of General Psychiatry.
-61 (2004) 310--317.
-\url{https://doi.org/10.1001/archpsyc.61.3.310}.}
-
-\leavevmode\hypertarget{ref-schober2018}{}%
-\CSLLeftMargin{{[}14{]} }
-\CSLRightInline{P.
-Schober, T.R.
-Vetter, Repeated measures designs and analysis of longitudinal data: If at first you do not succeed-try, try again, Anesthesia and Analgesia.
-127 (2018) 569--575.
-\url{https://doi.org/10.1213/ane.0000000000003511}.}
-
-\leavevmode\hypertarget{ref-pinheiro2006}{}%
-\CSLLeftMargin{{[}15{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science \& Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}
-
-\leavevmode\hypertarget{ref-vishwanath2009}{}%
-\CSLLeftMargin{{[}16{]} }
-\CSLRightInline{K.
-Vishwanath, H.
-Yuan, W.T.
-Barry, M.W.
-Dewhirst, N.
-Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia.
-11 (2009) 889--900.
-\url{https://doi.org/10.1593/neo.09580}.}
-
-\leavevmode\hypertarget{ref-dennis2019}{}%
-\CSLLeftMargin{{[}17{]} }
-\CSLRightInline{B.
-Dennis, J.M.
-Ponciano, M.L.
-Taper, S.R.
-Lele, {Errors in statistical inference under model misspecification: evidence, hypothesis testing, and AIC}, {Frontiers in Ecology and Evolution}.
-{7} (2019).
-\url{https://doi.org/\%7B10.3389/fevo.2019.00372\%7D}.}
-
-\leavevmode\hypertarget{ref-wang2019}{}%
-\CSLLeftMargin{{[}18{]} }
-\CSLRightInline{B.
-Wang, Z.
-Zhou, H.
-Wang, X.M.
-Tu, C.
-Feng, {The p-value and model specification in statistics}, {General Psychiatry}.
-{32} (2019).
-\url{https://doi.org/\%7B10.1136/gpsych-2019-100081\%7D}.}
-
-\leavevmode\hypertarget{ref-liu2010}{}%
-\CSLLeftMargin{{[}19{]} }
-\CSLRightInline{C.
-Liu, T.P.
-Cripe, M.-O.
-Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy.
-18 (2010) 1724--1730.
-\url{https://doi.org/10.1038/mt.2010.127}.}
-
-\leavevmode\hypertarget{ref-halsey2015}{}%
-\CSLLeftMargin{{[}20{]} }
-\CSLRightInline{L.G.
-Halsey, D.
-Curran-Everett, S.L.
-Vowler, G.B.
-Drummond, {The fickle \(p\) value generates irreproducible results}, {Nature Methods}.
-{12} (2015) 179--185.
-\url{https://doi.org/\%7B10.1038/nmeth.3288\%7D}.}
-
-\leavevmode\hypertarget{ref-abdi2010}{}%
-\CSLLeftMargin{{[}21{]} }
-\CSLRightInline{H. Abdi, {Holm's sequential Bonferroni procedure}, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.}
-
-\leavevmode\hypertarget{ref-nakagawa2004}{}%
-\CSLLeftMargin{{[}22{]} }
-\CSLRightInline{S.
-Nakagawa, {A farewell to Bonferroni: the problems of low statistical power and publication bias}, {Behavioral Ecology}.
-{15} (2004) 1044--1045.
-\url{https://doi.org/\%7B10.1093/beheco/arh107\%7D}.}
-
-\leavevmode\hypertarget{ref-gelman2012}{}%
-\CSLLeftMargin{{[}23{]} }
-\CSLRightInline{A.
-Gelman, J.
-Hill, M.
-Yajima, {Why we (usually) don't have to worry about multiple comparisons}, {Journal of Research on Educational Effectiveness}.
-{5} (2012) 189--211.
-\url{https://doi.org/\%7B10.1080/19345747.2011.618213\%7D}.}
-
-\leavevmode\hypertarget{ref-albers2019}{}%
-\CSLLeftMargin{{[}24{]} }
-\CSLRightInline{C. Albers, {The problem with unadjusted multiple and sequential statistical testing}, {Nature Communications}. {10} (2019). \url{https://doi.org/\%7B10.1038/s41467-019-09941-0\%7D}.}
-
-\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
-\CSLLeftMargin{{[}25{]} }
-\CSLRightInline{C.
-Ugrinowitsch, G.W.
-Fellingham, M.D.
-Ricard, Limitations of ordinary least squares models in analyzing repeated measures data, Medicine and Science in Sports and Exercise.
-36 (2004) 2144--2148.
-\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.}
-
-\leavevmode\hypertarget{ref-huynh1976}{}%
-\CSLLeftMargin{{[}26{]} }
-\CSLRightInline{H.
-Huynh, L.S.
-Feldt, {Estimation of the Box correction for degrees of freedom from sample data in randomized block and split-plot designs}, Journal of Educational Statistics.
-1 (1976) 69--82.
-\url{https://doi.org/10.3102/10769986001001069}.}
-
-\leavevmode\hypertarget{ref-greenhouse1959}{}%
-\CSLLeftMargin{{[}27{]} }
-\CSLRightInline{S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.}
-
-\leavevmode\hypertarget{ref-haverkamp2017}{}%
-\CSLLeftMargin{{[}28{]} }
-\CSLRightInline{N.
-Haverkamp, A.
-Beauducel, {Violation of the sphericity assumption and its effect on type-I error rates in repeated measures ANOVA and multi-level linear models (MLM)}, {Frontiers in Psychology}.
-{8} (2017).
-\url{https://doi.org/\%7B10.3389/fpsyg.2017.01841\%7D}.}
-
-\leavevmode\hypertarget{ref-keselman2001}{}%
-\CSLLeftMargin{{[}29{]} }
-\CSLRightInline{H.
-Keselman, J.
-Algina, R.
-Kowalchuk, {The analysis of repeated measures designs: A review}, {British Journal of Mathematica \& Statistical Psychology}.
-{54} (2001) 1--20.
-\url{https://doi.org/\%7B10.1348/000711001159357\%7D}.}
-
-\leavevmode\hypertarget{ref-charan2013}{}%
-\CSLLeftMargin{{[}30{]} }
-\CSLRightInline{Jaykaran.
-Charan, N.
-Kantharia, {How to calculate sample size in animal studies?}, Journal of Pharmacology and Pharmacotherapeutics.
-4 (2013) 303--306.
-\url{https://doi.org/10.4103/0976-500X.119726}.}
-
-\leavevmode\hypertarget{ref-barr2013}{}%
-\CSLLeftMargin{{[}31{]} }
-\CSLRightInline{D.J.
-Barr, R.
-Levy, C.
-Scheepers, H.J.
-Tily, {Random effects structure for confirmatory hypothesis testing: Keep it maximal}, {Journal of Memory and Language}.
-{68} (2013) 255--278.
-\url{https://doi.org/\%7B10.1016/j.jml.2012.11.001\%7D}.}
-
-\leavevmode\hypertarget{ref-rose2012}{}%
-\CSLLeftMargin{{[}32{]} }
-\CSLRightInline{N.L.
-Rose, H.D.
-Yang, S.D.
-Turner, G.L.
-Simpson, {An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to Scotland, UK}, Geochimica Et Cosmochimica Acta.
-82 (2012) 113--135.
-\url{https://doi.org/10.1016/j.gca.2010.12.026}.}
-
-\leavevmode\hypertarget{ref-pedersen2019}{}%
-\CSLLeftMargin{{[}33{]} }
-\CSLRightInline{E.J.
-Pedersen, D.L.
-Miller, G.L.
-Simpson, N.
-Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, {PeerJ}.
-7 (2019).
-\url{https://doi.org/10.7717/peerj.6876}.}
-
-\leavevmode\hypertarget{ref-simpson2018}{}%
-\CSLLeftMargin{{[}34{]} }
-\CSLRightInline{G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.}
-
-\leavevmode\hypertarget{ref-yang2012}{}%
-\CSLLeftMargin{{[}35{]} }
-\CSLRightInline{L.
-Yang, G.
-Qin, N.
-Zhao, C.
-Wang, G.
-Song, {Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality}, {BMC Medical Research Methodology}.
-{12} (2012).
-\url{https://doi.org/\%7B10.1186/1471-2288-12-165\%7D}.}
-
-\leavevmode\hypertarget{ref-beck1998}{}%
-\CSLLeftMargin{{[}36{]} }
-\CSLRightInline{N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. (1998) 596--627.}
-
-\leavevmode\hypertarget{ref-wood2017}{}%
-\CSLLeftMargin{{[}37{]} }
-\CSLRightInline{S.N. Wood, {Generalized additive models: An introduction with R, Second Edition}, CRC Press LLC, Philadelphia, PA, 2017.}
-
-\leavevmode\hypertarget{ref-r}{}%
-\CSLLeftMargin{{[}38{]} }
-\CSLRightInline{R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.}
-
-\leavevmode\hypertarget{ref-wood2016}{}%
-\CSLLeftMargin{{[}39{]} }
-\CSLRightInline{S.N.
-Wood, N.
-Pya, B.
-Saefken, Smoothing parameter and model selection for general smooth models, {Journal of the American Statistical Association}.
-{111} (2016) 1548--1563.
-\url{https://doi.org/\%7B10.1080/01621459.2016.1180986\%7D}.}
-
-\leavevmode\hypertarget{ref-west2014}{}%
-\CSLLeftMargin{{[}40{]} }
-\CSLRightInline{B.T.
-West, K.B.
-Welch, A.T.
-Galecki, Linear mixed models: A practical guide using statistical software, second edition, Taylor \& Francis, 2014.
-\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.}
-
-\leavevmode\hypertarget{ref-wolfinger1996}{}%
-\CSLLeftMargin{{[}41{]} }
-\CSLRightInline{R.D.
-Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics.
-1 (1996) 205--230.
-\url{http://www.jstor.org/stable/1400366}.}
-
-\leavevmode\hypertarget{ref-weiss2005}{}%
-\CSLLeftMargin{{[}42{]} }
-\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://books.google.com/books?id=MQ/_bvWDPsEAC}.}
-
-\leavevmode\hypertarget{ref-geisser1958}{}%
-\CSLLeftMargin{{[}43{]} }
-\CSLRightInline{S.
-Geisser, S.W.
-Greenhouse, {An extension of Box's results on the use of the \(F\) distribution in multivariate analysis}, The Annals of Mathematical Statistics.
-29 (1958) 885--891.
-\url{https://doi.org/10.1214/aoms/1177706545}.}
-
-\leavevmode\hypertarget{ref-maxwell2017}{}%
-\CSLLeftMargin{{[}44{]} }
-\CSLRightInline{S.E.
-Maxwell, H.D.
-Delaney, K.
-Kelley, Designing experiments and analyzing data: A model comparison perspective, third edition, Taylor \& Francis, 2017.
-\url{https://books.google.com/books?id=NmFQDwAAQBAJ}.}
-
-\leavevmode\hypertarget{ref-molenberghs2004}{}%
-\CSLLeftMargin{{[}45{]} }
-\CSLRightInline{G.
-Molenberghs, H.
-Thijs, I.
-Jansen, C.
-Beunckens, M.
-Kenward, C.
-Mallinckrodt, R.
-Carroll, {Analyzing incomplete longitudinal clinical trial data}, {Biostatistics}.
-{5} (2004) 445--464.
-\url{https://doi.org/\%7B10.1093/biostatistics/kxh001\%7D}.}
-
-\leavevmode\hypertarget{ref-scheffer2002}{}%
-\CSLLeftMargin{{[}46{]} }
-\CSLRightInline{J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.}
-
-\leavevmode\hypertarget{ref-potthoff2006}{}%
-\CSLLeftMargin{{[}47{]} }
-\CSLRightInline{R.F.
-Potthoff, G.E.
-Tudor, K.S.
-Pieper, V.
-Hasselblad, {Can one assess whether missing data are missing at random in medical studies?}, {Statistical Methods in Medical Research}.
-{15} (2006) 213--234.
-\url{https://doi.org/\%7B10.1191/0962280206sm448oa\%7D}.}
-
-\leavevmode\hypertarget{ref-ma2012}{}%
-\CSLLeftMargin{{[}48{]} }
-\CSLRightInline{Y.
-Ma, M.
-Mazumdar, S.G.
-Memtsoudis, Beyond repeated-measures analysis of variance advanced statistical methods for the analysis of longitudinal data in anesthesia research, {Regional Anesthesia and Pain Medicine}.
-{37} (2012) 99--105.
-\url{https://doi.org/\%7B10.1097/AAP.0b013e31823ebc74\%7D}.}
-
-\leavevmode\hypertarget{ref-nlme}{}%
-\CSLLeftMargin{{[}49{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, {nlme}: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.}
-
-\leavevmode\hypertarget{ref-nelder1972}{}%
-\CSLLeftMargin{{[}50{]} }
-\CSLRightInline{J.A.
-Nelder, R.W.M.
-Wedderburn, Generalized linear models, Journal of the Royal Statistical Society.
-Series A (General).
-135 (1972) 370--384.
-\url{http://www.jstor.org/stable/2344614}.}
-
-\leavevmode\hypertarget{ref-hastie1987}{}%
-\CSLLeftMargin{{[}51{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association.
-82 (1987) 371--386.
-\url{https://doi.org/10.1080/01621459.1987.10478440}.}
-
-\leavevmode\hypertarget{ref-hefley2017}{}%
-\CSLLeftMargin{{[}52{]} }
-\CSLRightInline{T.J.
-Hefley, K.M.
-Broms, B.M.
-Brost, F.E.
-Buderman, S.L.
-Kay, H.R.
-Scharf, J.R.
-Tipton, P.J.
-Williams, M.B.
-Hooten, {The basis function approach for modeling autocorrelation in ecological data}, {Ecology}.
-{98} (2017) 632--646.
-\url{https://doi.org/\%7B10.1002/ecy.1674\%7D}.}
-
-\leavevmode\hypertarget{ref-wegman1983}{}%
-\CSLLeftMargin{{[}53{]} }
-\CSLRightInline{E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.}
-
-\leavevmode\hypertarget{ref-mcelreath2018}{}%
-\CSLLeftMargin{{[}54{]} }
-\CSLRightInline{R. McElreath, {Statistical rethinking: A Bayesian course with examples in R and Stan}, {Chapman and Hall/CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
-
-\leavevmode\hypertarget{ref-miller2019}{}%
-\CSLLeftMargin{{[}55{]} }
-\CSLRightInline{D.L. Miller, Bayesian views of generalized additive modelling, arXiv Preprint arXiv:1902.01330. (2019).}
-
-\leavevmode\hypertarget{ref-laird1982}{}%
-\CSLLeftMargin{{[}56{]} }
-\CSLRightInline{N.M. Laird, J.H. Ware, Random-effects models for longitudinal data, Biometrics. 38 (1982) 963--974. \url{http://www.jstor.org/stable/2529876}.}
-
-\leavevmode\hypertarget{ref-marra2012}{}%
-\CSLLeftMargin{{[}57{]} }
-\CSLRightInline{G.
-Marra, S.N.
-Wood, Coverage properties of confidence intervals for generalized additive model components, {Scandinavian Journal of Statistics}.
-{39} (2012) 53--74.
-\url{https://doi.org/\%7B10.1111/j.1467-9469.2011.00760.x\%7D}.}
-
-\leavevmode\hypertarget{ref-gratia}{}%
-\CSLLeftMargin{{[}58{]} }
-\CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
-
-\leavevmode\hypertarget{ref-harezlak2018}{}%
-\CSLLeftMargin{{[}59{]} }
-\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, {Semiparametric Regression with R}, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
-
-\leavevmode\hypertarget{ref-lang2015}{}%
-\CSLLeftMargin{{[}60{]} }
-\CSLRightInline{T.A.
-Lang, D.G.
-Altman, {Basic statistical reporting for articles published in Biomedical Journals: The ``Statistical Analyses and Methods in the Published Literature{''} or the SAMPL Guidelines}, INTERNATIONAL JOURNAL OF NURSING STUDIES.
-{52} (2015) 5--9.
-\url{https://doi.org/\%7B10.1016/j.ijnurstu.2014.09.006\%7D}.}
-
-\leavevmode\hypertarget{ref-hastie1995}{}%
-\CSLLeftMargin{{[}61{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models for medical research, {Statistical Methods in Medical Research}.
-4 (1995) 187--196.
-\url{https://doi.org/10.1177/096228029500400302}.}
-
-\leavevmode\hypertarget{ref-begley2015}{}%
-\CSLLeftMargin{{[}62{]} }
-\CSLRightInline{C.G.
-Begley, J.P.A.
-Ioannidis, {Reproducibility in Science Improving the Standard for Basic and Preclinical Research}, {Circulation Research}.
-{116} (2015) 116--126.
-\url{https://doi.org/\%7B10.1161/CIRCRESAHA.114.303819\%7D}.}
-
-\leavevmode\hypertarget{ref-weissgerber2018}{}%
-\CSLLeftMargin{{[}63{]} }
-\CSLRightInline{T.L.
-Weissgerber, O.
-Garcia-Valencia, V.D.
-Garovic, N.M.
-Milic, S.J.
-Winham, {Meta-Research: Why we need to report more than 'Data were Analyzed by t-tests or ANOVA'}, Elife.
-7 (2018) e36163.
-\url{https://doi.org/10.7554/eLife.36163}.}
-
-\end{CSLReferences}
-
-\hypertarget{appendix-appendix}{%
-\appendix}
+\hypertarget{bibliography}{%
+\section{Bibliography}\label{bibliography}}
+
+\hypertarget{appendix}{%
+\section{(APPENDIX)}\label{appendix}}
 
@@
 
-This section presents the code used to generate figures, models and simulated data from Sections 3 and 4 from the main manuscript.
+This section presents the code used to generate figures, models and
+simulated data from Sections 3 and 4 from the main manuscript.
 
 \hypertarget{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}{%
-\subsection{Compound symmetry and independent errors in linear and quadratic responses}\label{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}}
-
-This section simulated linear and quadratic data in the same manner as in Section \ref{simulation}.
-The linear simulations using Figure \ref{fig:linear-cases-Appendix} show in panels A and D the simulated mean responses and individual data points.
-Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
+\subsection{Compound symmetry and independent errors in linear and
+quadratic
+responses}\label{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}}
+
+This section simulated linear and quadratic data in the same manner as
+in Section @ref(simulation). The linear simulations using Figure
+@ref(fig:linear-cases-Appendix) show in panels A and D the simulated
+mean responses and individual data points. Panels C and G show a visual
+interpretation of ``correlation'' in the responses: In panel C, subjects
+that have a value of the random error \(\varepsilon\) either above or
+below the mean group response are more likely to have other observations
+that follow the same trajectory, thereby demonstrating correlation in
+the response. In panel G,because the errors are independent, there is no
+expectation that responses are likely to follow a similar pattern.
 Panels D and H show the predictions from the rm-ANOVA model.
 
-The following code produces a more comprehensive exploration of Figure \ref{fig:l-q-response} in the main manuscript.
-
-\begin{lstlisting}[language=R]
+The following code produces a more comprehensive exploration of Figure
+@ref(fig:l-q-response) in the main manuscript.
+
+\begin{verbatim}
 ##########Section for calculations###########
@@
 D1<-plot_example(example(fun_type = "quadratic", error_type = "independent")) 
-\end{lstlisting}
-
-\begin{figure}[H]
-\includegraphics{00-Full_document_files/figure-latex/linear-cases-Appendix-1} \caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are able to capture the trend of the data.}\label{fig:linear-cases-Appendix}
+\end{verbatim}
+
+(ref:linear-cases-appendix) Simulated linear responses from two groups
+with correlated (top row) or independent (bottom row) errors using a
+rm-ANOVA model and a LMEM. A, F:Simulated data with known mean response
+and individual responses (points) showing the dispersion of the data.
+B,G: Generated errors showing the difference in the behavior of
+correlated and independent errors. C,H: Simulated data with thin lines
+representing individual trajectories. D,I: Estimations from the rm-ANOVA
+model for the mean group response. E, J: Estimations from the LMEM for
+the mean group response and individual responses (thin lines). In all
+panels, thick lines are the predicted mean response per group, thin
+lines are the random effects for each subject and points represent the
+original raw data. Both rm-ANOVA and the LMEM are able to capture the
+trend of the data.
+
+\begin{figure}
+\centering
+\includegraphics{Full_document_SAGE_files/figure-latex/linear-cases-Appendix-1.pdf}
+\caption{(ref:linear-cases-appendix)}
 \end{figure}
 
-For the quadratic response case, Figure \ref{fig:quadratic-cases-Appendix} shows the simulated responses using compound symmetry and independent errors.
-
-\begin{figure}[H]
-\includegraphics{00-Full_document_files/figure-latex/quadratic-cases-Appendix-1} \caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are not able to capture the changes in each group over time.}\label{fig:quadratic-cases-Appendix}
+For the quadratic response case, Figure
+@ref(fig:quadratic-cases-Appendix) shows the simulated responses using
+compound symmetry and independent errors.
+
+(ref:quadratic-caption-appendix) Simulated quadratic responses from two
+groups with correlated (top row) or independent (bottom row) errors
+using a rm-ANOVA model and a LMEM. A, F:Simulated data with known mean
+response and individual responses (points) showing the dispersion of the
+data. B,G: Generated errors showing the difference in the behavior of
+correlated and independent errors. C,H: Simulated data with thin lines
+representing individual trajectories. D,I: Estimations from the rm-ANOVA
+model for the mean group response. E, J: Estimations from the LMEM for
+the mean group response and individual responses (thin lines). In all
+panels, thick lines are the predicted mean response per group, thin
+lines are the random effects for each subject and points represent the
+original raw data. Both rm-ANOVA and the LMEM are not able to capture
+the changes in each group over time.
+
+\begin{figure}
+\centering
+\includegraphics{Full_document_SAGE_files/figure-latex/quadratic-cases-Appendix-1.pdf}
+\caption{(ref:quadratic-caption-appendix)}
 \end{figure}
@@
 
-This code produces Figure \ref{fig:basis-plot} from the main manuscript.
-Briefly, a non-linear (quadratic) response is simulated a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
-The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
-
-\begin{lstlisting}[language=R]
+This code produces Figure @ref(fig:basis-plot) from the main manuscript.
+Briefly, a non-linear (quadratic) response is simulated a gam model is
+fitted and the basis are extracted in order to explain how the smooth is
+constructed. The code for data simulation is used again here for the
+sake of keeping the same structure, although the data can be simulated
+in a more simple fashion.
+
+\begin{verbatim}
 #generate the response: the same initial procedure from the previous section to simulate
@@
      )
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-appendix-1}
+\end{verbatim}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/basis-plot-appendix-1}
 
@@
 
-\caption{Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-B: Matrix of basis function weights.
-Each basis function is multiplied by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
-Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot-appendix}
+\caption{(ref:basis-plot-caption)}\label{fig:basis-plot-appendix}
 \end{figure}
@@
 \hypertarget{tumor-data-simulation}{%
-\section{Longitudinal biomedical data simulation and GAMs}\label{tumor-data-simulation}}
-
-This section describes how to fit GAMs to longitudinal data using simulated data.
-First, data is simulated according to Section \ref{longitudinal-GAMs}, where reported data of oxygen saturation (\(\mbox{StO}_2\)) in tumors under either chemotherapy or saline control is used as a starting point to generate individual responses in each group.
-
-\begin{lstlisting}[language=R]
+\section{Longitudinal biomedical data simulation and
+GAMs}\label{tumor-data-simulation}}
+
+This section describes how to fit GAMs to longitudinal data using
+simulated data. First, data is simulated according to Section
+@ref(longitudinal-GAMs), where reported data of oxygen saturation
+(\(\mbox{StO}_2\)) in tumors under either chemotherapy or saline control
+is used as a starting point to generate individual responses in each
+group.
+
+\begin{verbatim}
 set.seed(1)
@@
     scale_x_continuous(breaks=c(0,2,5,7,10))
-\end{lstlisting}
+\end{verbatim}
 
@@
 
-This section shows a basic workflow to fit a series of increasingly complex GAMs to the simulated data from the previous section.
-Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
+This section shows a basic workflow to fit a series of increasingly
+complex GAMs to the simulated data from the previous section. Graphical
+and parameter diagnostics for goodness of fit are discussed, as well as
+model comparison via AIC (Aikake Information Criterion).
 
@@
 
-The first model fitted to the data is one that only accounts for different smooths by day.
-The model syntax specifies that \passthrough{\lstinline!gam\_00!} is the object that will contain all the model information, and that the model attempts to explain changes in \passthrough{\lstinline!StO2\_sim!} (simulated \(\mbox{StO}_2\)) using a smooth per \passthrough{\lstinline!Day!}.
-The model will use 5 knots (\passthrough{\lstinline!k=5!}) for the smooth.
-The smooth is constructed by default using thin plate regression splines.
-The smoothing parameter estimation method used is the restricted maximum likelihood (\passthrough{\lstinline!REML!}).
-
-\begin{lstlisting}[language=R]
+The first model fitted to the data is one that only accounts for
+different smooths by day. The model syntax specifies that
+\texttt{gam\_00} is the object that will contain all the model
+information, and that the model attempts to explain changes in
+\texttt{StO2\_sim} (simulated \(\mbox{StO}_2\)) using a smooth per
+\texttt{Day}. The model will use 5 knots (\texttt{k=5}) for the smooth.
+The smooth is constructed by default using thin plate regression
+splines. The smoothing parameter estimation method used is the
+restricted maximum likelihood (\texttt{REML}).
+
+\begin{verbatim}
 gam_00<-gam(StO2_sim ~ s(Day, k = 5),
@@
             data  = dat_sim)
-\end{lstlisting}
-
-To obtain model diagnostics, two methodologies are to be used: 1) graphical diagnostics, and 2) a model check.
-In the first case, the functions \passthrough{\lstinline!appraise!} and \passthrough{\lstinline!draw!} from the package \emph{gratia} can be used to obtain a single output with all the graphical diagnostics.
-For model check, the functions \passthrough{\lstinline!gam.check!} and \passthrough{\lstinline!summary!} from \emph{mgcv} provide detailed information about the model fit and its parameters.
+\end{verbatim}
+
+To obtain model diagnostics, two methodologies are to be used: 1)
+graphical diagnostics, and 2) a model check. In the first case, the
+functions \texttt{appraise} and \texttt{draw} from the package
+\emph{gratia} can be used to obtain a single output with all the
+graphical diagnostics. For model check, the functions \texttt{gam.check}
+and \texttt{summary} from \emph{mgcv} provide detailed information about
+the model fit and its parameters.
+
+(ref:first-GAM-caption) Graphical diagnostics for the first GAM model.
+Left: Graphical diagnostics provided by the function \texttt{appraise}
+from the package \emph{gratia}. Right: Fitted smooth for the model,
+provided by the function \texttt{draw}.
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/first-GAM-diag-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/first-GAM-diag-1}
 
@@
 
-\caption{Graphical diagnostics for the first GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:first-GAM-diag}
+\caption{(ref:first-GAM-caption)}\label{fig:first-GAM-diag}
 \end{figure}
 
-From the output of the function \passthrough{\lstinline!appraise!} in Figure \ref{fig:first-GAM-diag}, the major indicators of concern about the model are the QQ plot of residuals and the histogram of residuals.
-The QQ plot shows that the errors are not reasonably located along the 45\(^{\circ}\) line (which indicates normality), as there are multiple points that deviate from the trend, specially in the tails.
-The histogram also shows that the variation (residuals) is not following the assumption of a normal distribution.
-
-The \passthrough{\lstinline!draw!} function permits to plot the smooths as \passthrough{\lstinline!ggplot2!} objects, which eases subsequent manipulation, if desired.
-Because model \passthrough{\lstinline!gam\_00!} specifies only one smooth for the time covariate (Day), the plot only contains only one smooth.
-Note that the smooth shows an almost linear profile.
+From the output of the function \texttt{appraise} in Figure
+@ref(fig:first-GAM-diag), the major indicators of concern about the
+model are the QQ plot of residuals and the histogram of residuals. The
+QQ plot shows that the errors are not reasonably located along the
+45\(^{\circ}\) line (which indicates normality), as there are multiple
+points that deviate from the trend, specially in the tails. The
+histogram also shows that the variation (residuals) is not following the
+assumption of a normal distribution.
+
+The \texttt{draw} function permits to plot the smooths as
+\texttt{ggplot2} objects, which eases subsequent manipulation, if
+desired. Because model \texttt{gam\_00} specifies only one smooth for
+the time covariate (Day), the plot only contains only one smooth. Note
+that the smooth shows an almost linear profile.
 
@@
 
-\begin{lstlisting}[language=R]
+\begin{verbatim}
 #need to add figure number and caption
 gam.check(gam_00)
-\end{lstlisting}
-
-\includegraphics{00-Full_document_files/figure-latex/first-GAM-check-1}
-
-\begin{lstlisting}
+\end{verbatim}
+
+\includegraphics{Full_document_SAGE_files/figure-latex/first-GAM-check-1.pdf}
+
+\begin{verbatim}
 ## 
@@
 ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
+\end{verbatim}
+
+\begin{verbatim}
 summary(gam_00)
-\end{lstlisting}
-
-\begin{lstlisting}
+\end{verbatim}
+
+\begin{verbatim}
 ## 
@@
 ## -REML = 444.01  Scale est. = 450.66    n = 100
-\end{lstlisting}
-
-Special attention must be paid to the `k-index' from \passthrough{\lstinline!gam.check!}.
-This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
-If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless1).
-From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data.
-The \passthrough{\lstinline!edf!} (effective degrees of freedom) is an indicator of the complexity of the smooth.
-Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
-
-From the \passthrough{\lstinline!summary!} function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
-The link function is `identity' as the model does not make any transformation on the predictors.
-The `significance of smooth terms' \emph{p-value} indicates if each smooth is adding significance to the model.
-Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \(\approx\) 17\%.
+\end{verbatim}
+
+Special attention must be paid to the `k-index' from \texttt{gam.check}.
+This parameter indicates if the basis dimension of the smooth is
+adequate, i.e., it checks that the basis used to create the smooth are
+adequate to capture the trends in the data. If the model is not
+adequately capturing the trens in the data, this is indicated by a low
+k-index value (\textless1). From the output, it can be seen that the
+k-index is 0.36, which indicates that the model is not capturing the
+variability in the data. The \texttt{edf} (effective degrees of freedom)
+is an indicator of the complexity of the smooth. Here the complexity of
+the smooth is comparable to that of a 4th degree polynomial.
+
+From the \texttt{summary} function, information about the assumed
+distribution of the errors (Gaussian in this case) and the link function
+can be obtained. The link function is `identity' as the model does not
+make any transformation on the predictors. The `significance of smooth
+terms' \emph{p-value} indicates if each smooth is adding significance to
+the model. Here, the \emph{p-value} is low but we have seen that there
+are issues with the model from the previous outputs. Finally, the
+`deviance explained' indicates how much of the data the model is able to
+capture, which in this case corresponds to \(\approx\) 17\%.
 
@@
 
-The major flaw of \passthrough{\lstinline!gam\_00!} is that this model is not taking into account the fact that the data is nested in groups.
-The next iteration is a model where a different smooth of time (Day) is assigned for each group using \passthrough{\lstinline!by=Group!} in the model syntax.
-
-\begin{lstlisting}[language=R]
+The major flaw of \texttt{gam\_00} is that this model is not taking into
+account the fact that the data is nested in groups. The next iteration
+is a model where a different smooth of time (Day) is assigned for each
+group using \texttt{by=Group} in the model syntax.
+
+\begin{verbatim}
 gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5),
@@
 gam.check(gam_01)
-\end{lstlisting}
-
-\includegraphics{00-Full_document_files/figure-latex/second-GAM-1}
-
-\begin{lstlisting}
+\end{verbatim}
+
+\includegraphics{Full_document_SAGE_files/figure-latex/second-GAM-1.pdf}
+
+\begin{verbatim}
 ## 
@@
 ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
+\end{verbatim}
+
+\begin{verbatim}
 summary(gam_01)
-\end{lstlisting}
-
-\begin{lstlisting}
+\end{verbatim}
+
+\begin{verbatim}
 ## 
@@
 ## -REML = 423.39  Scale est. = 280.88    n = 100
-\end{lstlisting}
-
-Diagnostics for this model indicate that the k-index is still below 1 (0.43 from \passthrough{\lstinline!gam.check!}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
-Moreover, the smooths (plotted via the \passthrough{\lstinline!draw()!} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \(\approx\) 51\%.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/second-GAM-diag-1}
+\end{verbatim}
+
+Diagnostics for this model indicate that the k-index is still below 1
+(0.43 from \texttt{gam.check}), and that the residuals are still not
+following a normal distribution (Figure @ref(fig:second-GAM-diag)).
+Moreover, the smooths (plotted via the \texttt{draw()} function) appear
+with a fairly linear profile, which indicates they are still not
+capturing the trends observed in the data. From \texttt{summary()}, the
+deviance explained by the model is \(\approx\) 51\%.
+
+(ref:second-GAM-caption) Graphical diagnostics for the second GAM model.
+Left: Graphical diagnostics provided by the function \texttt{appraise}
+from the package \emph{gratia}. Right: Fitted smooth for the model,
+provided by the function \texttt{draw}.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/second-GAM-diag-1}
 
@@
 
-\caption{Graphical diagnostics for the second GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:second-GAM-diag}
+\caption{(ref:second-GAM-caption)}\label{fig:second-GAM-diag}
 \end{figure}
@@
 
-Model \passthrough{\lstinline!gam\_00!} was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted, the model diagnostics, and, the low variance explained by the model.
-On the other hand, \passthrough{\lstinline!gam\_01!} takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \ref{longitudinal-GAMs}, in order to differentiate between each group a parametric term needs to be added to the model for the interaction of \emph{Day} and \emph{Group}.
-
-\begin{lstlisting}[language=R]
+Model \texttt{gam\_00} was built for didactic purposes to cover the
+simplest case, but it does not account for the nesting of the data by
+Group, which is apparent from the type of smooth fitted, the model
+diagnostics, and, the low variance explained by the model. On the other
+hand, \texttt{gam\_01} takes into account the nesting within each group
+and provides better variance explanation, but as indicated in Section
+@ref(longitudinal-GAMs), in order to differentiate between each group a
+parametric term needs to be added to the model for the interaction of
+\emph{Day} and \emph{Group}.
+
+\begin{verbatim}
 #GAM for StO2
@@
 gam.check(m1)
-\end{lstlisting}
-
-\includegraphics{00-Full_document_files/figure-latex/final-model-Appendix-1}
-
-\begin{lstlisting}
+\end{verbatim}
+
+\includegraphics{Full_document_SAGE_files/figure-latex/final-model-Appendix-1.pdf}
+
+\begin{verbatim}
 ## 
@@
 ## s(Day):GroupTreatment 4.00 3.88    1.02    0.54
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
+\end{verbatim}
+
+\begin{verbatim}
 summary(m1)
-\end{lstlisting}
-
-\begin{lstlisting}
+\end{verbatim}
+
+\begin{verbatim}
 ## 
@@
 ## -REML = 355.86  Scale est. = 64.533    n = 100
-\end{lstlisting}
-
-The resulting model is \passthrough{\lstinline!m1!}, which is the model fitted in the main manuscript.
-By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\(\approx\) 1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 89\% of the variance in the data.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/final-GAM-diag-1}
+\end{verbatim}
+
+The resulting model is \texttt{m1}, which is the model fitted in the
+main manuscript. By using \texttt{appraise()} and \texttt{draw} on this
+model (Figure @ref(fig:final-GAM-diag)) we see that the trend on the QQ
+plot has improved, the histogram of the residuals appears to be
+reasonably distributed, and the smooths are capturing the trend of the
+data within each group. From \texttt{gam.check}, the k-index is now at
+an acceptable value (\(\approx\) 1.02), and \texttt{summary} now
+indicates that the model is able to capture 89\% of the variance in the
+data.
+
+(ref:final-GAM-caption) Graphical diagnostics for the final GAM model.
+Left: Graphical diagnostics provided by the function \texttt{appraise}
+from the package \emph{gratia}. Right: Fitted smooths for the model,
+provided by the function \texttt{draw}.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/final-GAM-diag-1}
 
@@
 
-\caption{Graphical diagnostics for the final GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooths for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:final-GAM-diag}
+\caption{(ref:final-GAM-caption)}\label{fig:final-GAM-diag}
 \end{figure}
@@
 \hypertarget{comparing-models-via-aic}{%
-\subsubsection{Comparing models via AIC}\label{comparing-models-via-aic}}
-
-One final comparison that can be made for model selection involves the use of the Aikake Information Criterion (AIC).
-This metric is used to estimate information loss, which we want to minimize with an appropriate model.
-Therefore, when 2 or more models are compared, the model with lower AIC is preferred.
-In R, the comparison is done using the \passthrough{\lstinline!AIC!} function.
-
-\begin{lstlisting}[language=R]
+\subsubsection{Comparing models via
+AIC}\label{comparing-models-via-aic}}
+
+One final comparison that can be made for model selection involves the
+use of the Aikake Information Criterion (AIC). This metric is used to
+estimate information loss, which we want to minimize with an appropriate
+model. Therefore, when 2 or more models are compared, the model with
+lower AIC is preferred. In R, the comparison is done using the
+\texttt{AIC} function.
+
+\begin{verbatim}
 AIC(gam_00,gam_01,m1)
-\end{lstlisting}
-
-\begin{lstlisting}
+\end{verbatim}
+
+\begin{verbatim}
 ##               df      AIC
@@
 ## m1     10.980983 712.2067
-\end{lstlisting}
-
-The output in this case is expected: model \passthrough{\lstinline!m1!} has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
-The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
+\end{verbatim}
+
+The output in this case is expected: model \texttt{m1} has a lower AIC
+(712.46) whereas the initial two models have higher AICs (900 and 858).
+The AIC should not be considered as the only estimator of model quality,
+instead to be used as complimentary information to the graphical
+diagnostics and model checks described above.
 
 \hypertarget{pairwise-comparisons-of-smooth-confidence-intervals}{%
-\paragraph{Pairwise comparisons of smooth confidence intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
-
-The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \ref{GAM-significance}.
-In this case, the ``design matrix'' is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
-Briefly, the ``design matrix'' (also known as the ``Xp matrix'') from the selected model (\passthrough{\lstinline!m1!}) is used to calculate a 95\% confidence interval of the difference between the smooth terms for each group.
-This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
-\textbf{All pairwise comparisons in this paper have been centered at the response scale to ease interpretation }.
-
-\begin{lstlisting}[language=R]
+\paragraph{Pairwise comparisons of smooth confidence
+intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
+
+The estimation of significant differences between each treatment group
+can be achieved via pairwise comparisons of the smooth confidence
+intervals as described in section @ref(GAM-significance). In this case,
+the ``design matrix'' is used to estimate the pairwise comparisons (see
+main manuscript for details and associated references). Briefly, the
+``design matrix'' (also known as the ``Xp matrix'') from the selected
+model (\texttt{m1}) is used to calculate a 95\% confidence interval of
+the difference between the smooth terms for each group. This approach
+allows to estimate the time intervals where a significant difference
+exists between the groups (confidence interval above or below 0).
+\textbf{All pairwise comparisons in this paper have been centered at the
+response scale to ease interpretation }.
+
+\begin{verbatim}
 ##Pairwise comparisons
@@
     )
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/pairwise-comp-workflow-fig-1}
+\end{verbatim}
+
+(ref:pairwise-comp-caption) Smooth pairwise comparisons for model
+\texttt{m1} using a 95\% confidence interval for the difference between
+smooths. The comparison is centered at the response scale. Shaded
+regions indicate time intervals where each treatment group has a
+non-zero effect.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/pairwise-comp-workflow-fig-1}
 
@@
 
-\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!m1!} using a 95\% confidence interval for the difference between smooths.
-The comparison is centered at the response scale.
-Shaded regions indicate time intervals where each treatment group has a non-zero effect.}\label{fig:pairwise-comp-workflow-fig}
+\caption{(ref:pairwise-comp-caption)}\label{fig:pairwise-comp-workflow-fig}
 \end{figure}
 
-Of notice, a convenient wrapper for the function described above exists in the package \passthrough{\lstinline!gratia!}.
-In this package, \passthrough{\lstinline!difference\_smooths!} is a function that makes the comparisons and produces Figure \ref{fig:pairwise-comp-workflow-fig} when is used on a fitted model.
-The function syntax and an example can be found at:
+Of notice, a convenient wrapper for the function described above exists
+in the package \texttt{gratia}. In this package,
+\texttt{difference\_smooths} is a function that makes the comparisons
+and produces Figure @ref(fig:pairwise-comp-workflow-fig) when is used on
+a fitted model. The function syntax and an example can be found at:
 
@@
 
-Keep in mind that this function \textbf{does not} center the pairwise comparison at the response scale, so it has to be shifted in order to be compared to the raw data.
+Keep in mind that this function \textbf{does not} center the pairwise
+comparison at the response scale, so it has to be shifted in order to be
+compared to the raw data.
 
 \hypertarget{gam-and-linear-model-plots-and-missing-data}{%
-\section{GAM and Linear model plots and Missing data}\label{gam-and-linear-model-plots-and-missing-data}}
-
-This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\passthrough{\lstinline!m1!}), linear model and GAM on data with missing observations are presented.
-Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inset are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
+\section{GAM and Linear model plots and Missing
+data}\label{gam-and-linear-model-plots-and-missing-data}}
+
+This section covers the code used to generate Figure
+@ref(fig:sim-smooth-plot), where the simulated data, fit of the
+``final'' GAM (\texttt{m1}), linear model and GAM on data with missing
+observations are presented. Note that panel A in Figure
+@ref(fig:sim-smooth-plot) and the inset are generated in the code chunk
+where the data is simulated in Section @ref(tumor-data-simulation), and
+are called later to build the figure.
 
 \hypertarget{gam-and-linear-model-plots}{%
-\subsection{GAM and Linear model plots}\label{gam-and-linear-model-plots}}
-
-This code chunk creates panels B and D in Figure \ref{fig:sim-smooth-plot}.
-Note that this code uses the final GAM from the previous section (\passthrough{\lstinline!m1!}), so the simulated data and the model should be generated before running this section.
-
-\begin{lstlisting}[language=R]
+\subsection{GAM and Linear model
+plots}\label{gam-and-linear-model-plots}}
+
+This code chunk creates panels B and D in Figure
+@ref(fig:sim-smooth-plot). Note that this code uses the final GAM from
+the previous section (\texttt{m1}), so the simulated data and the model
+should be generated before running this section.
+
+\begin{verbatim}
 #linear model
@@
 #summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
-\end{lstlisting}
+\end{verbatim}
 
 \hypertarget{working-with-missing-data-in-gams}{%
-\subsection{Working with Missing data in GAMs}\label{working-with-missing-data-in-gams}}
-
-This code chunk first randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data.
-Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
-
-\begin{lstlisting}[language=R]
+\subsection{Working with Missing data in
+GAMs}\label{working-with-missing-data-in-gams}}
+
+This code chunk first randomly deletes 40\% of the total observations in
+the original simulated data, and then an interaction GAM is fitted to
+the remaining data. Model diagnostics are presented, and an object that
+stores the fitted smooths is saved to be called in the final code chunk
+to build the figure.
+
+\begin{verbatim}
 #missing data
@@
   thm1
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-Appendix-1}
+\end{verbatim}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/sim-smooth-plot-Appendix-1}
 
@@
 
-\caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
+\caption{(ref:sim-plot-caption)}\label{fig:sim-smooth-plot-Appendix}
 \end{figure}
@@
 \hypertarget{pairwise-comparisons-in-gams-full-and-missing-data-cases}{%
-\subsection{Pairwise comparisons in GAMs: full and missing data cases}\label{pairwise-comparisons-in-gams-full-and-missing-data-cases}}
-
-The next code chunk reproduces Figure \ref{fig:plot-pairwise-comp}. Here pairwise comparisons are made for the full and missing datasets.
-
-\begin{lstlisting}[language=R]
+\subsection{Pairwise comparisons in GAMs: full and missing data
+cases}\label{pairwise-comparisons-in-gams-full-and-missing-data-cases}}
+
+The next code chunk reproduces Figure @ref(fig:plot-pairwise-comp). Here
+pairwise comparisons are made for the full and missing datasets.
+
+\begin{verbatim}
 ##Pairwise comparisons
@@
 pair_comp<-c1+c2
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-Appendix-1}
+\end{verbatim}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/plot-pairwise-comp-Appendix-1}
 
@@
 
-\caption{Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-B: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
-In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp-Appendix}
+\caption{(ref:pairwise-comp)}\label{fig:plot-pairwise-comp-Appendix}
 \end{figure}
 
+\bibliographystyle{sagev}
+\bibliography{refs1.bib}
+
 \end{document}

### 2021-06-18 Untitled/Untitled.tex
References do not work

\documentclass[Royal,times,sageh]{sagej}

\usepackage{moreverb,url,natbib, multirow, tabularx}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

% Pandoc citation processing

\begin{document}

\title{A demonstration of the \LaTeX class file for Statistical Methods
in Medical Research with Rmarkdown}

\runninghead{Uthor \emph{et al}.}

\author{A. Uthor*\affilnum{1,2}, O. Tro\affilnum{2}, O.
Vriga\affilnum{3}}

\affiliation{\affilnum{1}{Department of Incredible Research, University
A, City A, Country A}\\\affilnum{2}{Department of Applied Things,
University B, City B, Country B}\\\affilnum{3}{Very Important Stuff
Committee, Institute C, City C, Country C}}

\corrauth{Corresponding author name, This is sample corresponding
address.}

\email{\href{mailto:correspondingauthor@protonmail.com}{\nolinkurl{correspondingauthor@protonmail.com}}}

\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Aenean ut elit
odio. Donec fermentum tellus neque, vitae fringilla orci pretium vitae.
Fusce maximus finibus facilisis. Donec ut ullamcorper turpis. Donec ut
porta ipsum. Nullam cursus mauris a sapien ornare pulvinar. Aenean
malesuada molestie erat quis mattis. Praesent scelerisque posuere
faucibus. Praesent nunc nulla, ullamcorper ut ullamcorper sed, molestie
ut est. Donec consequat libero nisi, non semper velit vulputate et.
Quisque eleifend tincidunt ligula, bibendum finibus massa cursus eget.
Curabitur aliquet vehicula quam non pulvinar. Aliquam facilisis tortor
nec purus finibus, sit amet elementum eros sodales. Ut porta porttitor
vestibulum. Integer molestie, leo ut maximus aliquam, velit dui iaculis
nibh, eget hendrerit purus risus sit amet dolor. Sed sed tincidunt ex.
Curabitur imperdiet egestas tellus in iaculis. Maecenas ante neque,
pretium vel nisl at, lobortis lacinia neque. In gravida elit vel
volutpat imperdiet. Sed ut nulla arcu. Proin blandit interdum ex sit
amet laoreet. Phasellus efficitur, sem hendrerit mattis dapibus, nunc
tellus ornare nisi, nec eleifend enim nibh ac ipsum. Aenean tincidunt
nisl sit amet facilisis faucibus. Donec odio erat, bibendum eu imperdiet
sed, gravida luctus turpis.
\end{abstract}

\keywords{Class file; \LaTeX; SMMR; Rmarkdown;}

\maketitle

\hypertarget{the-article-header-information}{%
\section{The Article Header
Information}\label{the-article-header-information}}

YAML header:

\begin{verbatim}
output:
  rticles::sim_article:
    keep_tex: TRUE
\end{verbatim}

Configure the YAML header including the following elements:

\begin{itemize}
\item
  \texttt{title}: Title
\item
  \texttt{runninghead}: Author last names, use \emph{et al.} if there
  are three or more authors.
\item
  \texttt{author}: List of author(s) containing \texttt{name} and
  \texttt{num}
\item
  \texttt{corrauth}: Corresponding author's name and address.
\item
  \texttt{email}: Correspondence email
\item
  \texttt{abstract}: Limited to 200 words
\item
  \texttt{keywords}: Keywords for the article
\item
  \texttt{bibliography}: BibTeX \texttt{.bib} file
\item
  \texttt{bibliographystyle}: sageh or sagev
\item
  \texttt{classoption}: options of the \texttt{sagej} class
\end{itemize}

\hypertarget{remarks}{%
\subsection{Remarks}\label{remarks}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\item
  \texttt{bibliographystyle}
\item
  \texttt{classoption}
\item
  Keywords are separated by commas.
\end{enumerate}

\hypertarget{the-body-of-the-article}{%
\section{The Body of the Article}\label{the-body-of-the-article}}

\hypertarget{mathematics}{%
\subsection{Mathematics}\label{mathematics}}

Use mathematics in Rmarkdown as usual.

\hypertarget{figures-and-tables}{%
\subsection{Figures and Tables}\label{figures-and-tables}}

Figures are supported from R code:

\begin{verbatim}
x = rnorm(10)
y = rnorm(10)
plot(x, y)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{Untitled_files/figure-latex/plot-ref-1} \caption{Fancy Caption\label{fig:plot}}\label{fig:plot-ref}
\end{figure}

\ldots and can be referenced (Figure \ref{fig:plot}) by including the
\texttt{\textbackslash{}\textbackslash{}label\{\}} tag in the
\texttt{fig.cap} attribute of the R chunk:
\texttt{fig.cap\ =\ "Fancy\ Caption\textbackslash{}\textbackslash{}label\{fig:plot\}"}.
It is a quirky hack at the moment, see
\href{https://github.com/yihui/knitr/issues/323}{here}.

Analogously, use Rmarkdown to produce tables as usual:

\begin{verbatim}
if (!require("xtable")) install.packages("xtable")
\end{verbatim}

\begin{verbatim}
## Loading required package: xtable
\end{verbatim}

\begin{verbatim}
xt <- xtable(head(cars), caption = "A table", label = "tab:table")
print(xt, comment = FALSE)
\end{verbatim}

\begin{table}[ht]
\centering
\begin{tabular}{rrr}
  \hline
 & speed & dist \\ 
  \hline
1 & 4.00 & 2.00 \\ 
  2 & 4.00 & 10.00 \\ 
  3 & 7.00 & 4.00 \\ 
  4 & 7.00 & 22.00 \\ 
  5 & 8.00 & 16.00 \\ 
  6 & 9.00 & 10.00 \\ 
   \hline
\end{tabular}
\caption{A table} 
\label{tab:table}
\end{table}

Referenced via \ref{tab:table}. You can also use the YAML option
\texttt{header-includes} to includes custom \LaTeX packages for tables
(keep in mind that \texttt{pandoc} uses \texttt{longtables} by default,
and it is hardcoded; some things may require including the package
\texttt{longtable}). E.g., using \texttt{ctable}:

\begin{verbatim}
header-includes:
- \usepackage{ctable}
\end{verbatim}

Then, just write straight-up \LaTeX code and reference is as usual
(\texttt{\textbackslash{}ref\{tab:ctable\}}):

\begin{verbatim}
\ctable[cap = {Short caption},
        caption = {A caption for this table.},
        label={tab:ctable},]
        {cc}
        {
        \tnote[$\ast$]{Footnote 1}
        \tnote[$\dagger$]{Other footnote}
        \tnote[b]{Mistakes are possible.}
        }{
        \FL
        COL 1\tmark[a] & COL 2\tmark[$\ast$]
        \ML
        6.92\tmark[$\dagger$] & 0.09781 \\
        6.93\tmark[$\dagger$] & 0.09901 \\
        97 & 2000
        \LL
}
\end{verbatim}

It is also possible to set the \texttt{YAML} option
\texttt{longtable:\ true} and use markdown tables (or the
\texttt{knitr::kable} function): \texttt{knitr::kable(head(cars))}
produces the same table as the \texttt{xtable} example presented before.

\hypertarget{cross-referencing}{%
\subsection{Cross-referencing}\label{cross-referencing}}

The use of the Rmarkdown equivalent of the \LaTeX cross-reference system
for figures, tables, equations, etc., is encouraged (using
\texttt{{[}@\textless{}name\textgreater{}{]}}, equivalent of
\texttt{\textbackslash{}ref\{\textless{}name\textgreater{}\}} and
\texttt{\textbackslash{}label\{\textless{}name\textgreater{}\}}). That
works well for citations in Rmarkdown, not so well for figures and
tables. In that case, it is possible to revert to standard
\LaTeX syntax.

\hypertarget{double-spacing}{%
\subsection{Double Spacing}\label{double-spacing}}

If you need to double space your document for submission please use the
\texttt{doublespace} option in the header.

\hypertarget{bibliography}{%
\section{Bibliography}\label{bibliography}}

Link a \texttt{.bib} document via the YAML header, and bibliography will
be printed at the very end (as usual). The default bibliography style is
provided by Wiley as in \texttt{WileyNJD-AMA.bst}, do not delete that
file.

Use the Rmarkdown equivalent of the \LaTeX citation system using
\texttt{{[}@\textless{}name\textgreater{}{]}}. Example:
\citep{Taylor1937}, \citep{Knupp1999, Kamm2000}.

To include all citation from the \texttt{.bib} file, add
\texttt{\textbackslash{}nocite\{*\}} before the end of the document.

\hypertarget{further-information}{%
\section{Further information}\label{further-information}}

All \LaTeX enviroments supported by the main template are supported here
as well; see the \texttt{.tex} sample file
\href{http://onlinelibrary.wiley.com/journal/10.1002/(ISSN)1097-0258/homepage/la_tex_class_file.htm}{here}
for more details and example.

\bibliographystyle{sageh}
\bibliography{bibfile}

\end{document}

### 2021-06-18 Manuscripts/Manuscript by chapters -for submission/01-Background.Rmd
Need to check references. refs1 works but refs does not.

@@
 Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response.
+Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001] [@roth2017], and cell metabolism [@jones2018] [@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response.
 Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients [@kamstra2015].
@@
 The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times [@wagenmakers2008], and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].
+The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004] [@schober2018].
 
@@
 In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009].
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011] [@skala2010] [@vishwanath2009].
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
@@
 
-Additionally, the _p-value_ itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010;@halsey2015], consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019].
+Additionally, the _p-value_ itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010; @halsey2015], consequently biasing the conclusions of the study.
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012; @albers2019].
 Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
@@
 On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].
-Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976;@greenhouse1959], but the effectiveness of the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001].
+Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976; @greenhouse1959], but the effectiveness of the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001].
 In the case of biomedical research, where living subjects are frequently used, sample sizes are often not "large" due to ethical and budgetary reasons [@charan2013] which might cause the corrections for lack of compound symmetry to be ineffective.
@@
 Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] .
+Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012; @pedersen2019; @simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998].
 There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017] and 3) can easily accommodate missing observations.
@@
 The current advances in programming languages designed for statistical analysis (specifically  $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, $\textsf{R}$ [@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
+In particular, $\textsf{R}$ [@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016; @wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
 hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.

### 2021-06-18 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.Rmd
Need to check references. refs1 works but refs does not.

@@
   org:  Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, US
-corrauth: "Corresponding author"
+corrauth: "Timothy J. Muldoon"
 email: tmuldoon@uark.edu
@@
   - times
-bibliography: refs1.bib
+bibliography: refs.bib
 bibliographystyle: sagev

### 2021-06-18 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.tex
Need to check references. refs1 works but refs does not.

@@
 
-\corrauth{Corresponding author}
+\corrauth{Timothy J. Muldoon}
 
@@
 response \citep{roblyer2011, tank2020, pavlov2018, demidov2018},
-antibody expression \citep{ritter2001, roth2017}, and cell metabolism
-\citep{jones2018, skala2010} are examples of the different situations
-where researchers have used longitudinal designs to study some
-physiological response. Because the frequency of the measurements in a
-longitudinal study is dependent on the biological phenomena of interest
-and the experimental design of the study, the frequency of such
-measurements can range from minute intervals to study a short-term
-response such as anesthesia effects in animals\citep{greening2018}, to
-weekly measurements to analyze a mid-term response like the evolution of
-dermatitis symptoms in breast cancer patients \citep{sio2016}, to
-monthly measurements to study a long-term response such as mouth opening
-following radiotherapy (RT) in neck cancer patients \citep{kamstra2015}.
+antibody expression \citep{ritter2001} \citep{roth2017}, and cell
+metabolism \citep{jones2018} \citep{skala2010} are examples of the
+different situations where researchers have used longitudinal designs to
+study some physiological response. Because the frequency of the
+measurements in a longitudinal study is dependent on the biological
+phenomena of interest and the experimental design of the study, the
+frequency of such measurements can range from minute intervals to study
+a short-term response such as anesthesia effects in
+animals\citep{greening2018}, to weekly measurements to analyze a
+mid-term response like the evolution of dermatitis symptoms in breast
+cancer patients \citep{sio2016}, to monthly measurements to study a
+long-term response such as mouth opening following radiotherapy (RT) in
+neck cancer patients \citep{kamstra2015}.
 
@@
 time points through the study (a condition also known as \emph{complete
-observations}) \citep{gueorguieva2004, schober2018}.
+observations}) \citep{gueorguieva2004} \citep{schober2018}.
 
@@
 response to chemo and/or radiotherapy in preclinical and clinical
-settings \citep{roblyer2011, skala2010, vishwanath2009}. These studies
-have shown that the collected signal does not follow a linear trend over
-time, and presents extreme variability at different time points, making
-the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such data the
-estimates are inevitably biased, because the model is only able to
-accommodate linear trends that fail to adequately represent the
-biological phenomenon of interest.
+settings \citep{roblyer2011} \citep{skala2010} \citep{vishwanath2009}.
+These studies have shown that the collected signal does not follow a
+linear trend over time, and presents extreme variability at different
+time points, making the fit of rm-ANOVA model inconsistent with the
+observed variation. Therefore, when rm-ANOVA is used to draw inference
+of such data the estimates are inevitably biased, because the model is
+only able to accommodate linear trends that fail to adequately represent
+the biological phenomenon of interest.
 
@@
 interactions \citep{yang2012} and the dynamics of government in
-political science \citep{beck1998} . There are several advantages of
-GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible
-class of smooth responses that enable the data to dictate the trend in
-the fit of the model, 2) they can model non-constant correlation between
-repeated measurements \citep{wood2017} and 3) can easily accommodate
-missing observations. Therefore, GAMs can provide a more flexible
-statistical approach to analyze non-linear biomedical longitudinal data
-than LMEMs and rm-ANOVA.
+political science \citep{beck1998}. There are several advantages of GAMs
+over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of
+smooth responses that enable the data to dictate the trend in the fit of
+the model, 2) they can model non-constant correlation between repeated
+measurements \citep{wood2017} and 3) can easily accommodate missing
+observations. Therefore, GAMs can provide a more flexible statistical
+approach to analyze non-linear biomedical longitudinal data than LMEMs
+and rm-ANOVA.
 
@@
 \bibliographystyle{sagev}
-\bibliography{refs1.bib}
+\bibliography{refs.bib}
 

### 2021-06-21 Manuscripts/Manuscript by chapters -for submission/00-Full_document.tex
Need to fix the code highlighting. How to add the listing option for all chunks?

--- 2021-06-11 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\documentclass[,]{sagej}
+
+\usepackage{moreverb,url,natbib, multirow, tabularx}
+\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
+
+% Pandoc citation processing
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
   \fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Using generalized additive models to analyze biomedical non-linear longitudinal data},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
-\usepackage{listings}
-\newcommand{\passthrough}[1]{#1}
-\lstset{defaultdialect=[5.3]Lua}
-\lstset{defaultdialect=[x86masm]Assembler}
-\usepackage{longtable,booktabs,array}
+ }%
+ {}
 \usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
 \usepackage{lineno}
@@
 \usepackage{float}
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\begin{document}
 
 \title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
-\usepackage{etoolbox}
-\makeatletter
-\providecommand{\subtitle}[1]{% add subtitle to \maketitle
-  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
-}
-\makeatother
-\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
-\date{\vspace{-2.5em}}
-
-\begin{document}
+
+\runninghead{}
+
+\author{}
+
+\affiliation{}
+
 \maketitle
@@
 Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response {[}\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}{]}, antibody expression {[}\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}{]}, and cell metabolism {[}\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}\protect\hyperlink{ref-greening2018}{9}{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}\protect\hyperlink{ref-sio2016}{10}{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}\protect\hyperlink{ref-kamstra2015}{11}{]}.
+Tumor response \citep{roblyer2011, tank2020, pavlov2018, demidov2018}, antibody expression \citep{ritter2001, roth2017}, and cell metabolism \citep{jones2018, skala2010} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
+Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals\citep{greening2018}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients \citep{sio2016}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients \citep{kamstra2015}.
 
 Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}\protect\hyperlink{ref-wagenmakers2008}{12}{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}{]}.
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
+The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times \citep{wagenmakers2008}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
+The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) \citep{gueorguieva2004, schober2018}.
+
+The expected linear behavior of the response through time is a key requisite in rm-ANOVA \citep{pinheiro2006}.
 This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
 In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings \citep{roblyer2011, skala2010, vishwanath2009}.
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
@@
 Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}.
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}\protect\hyperlink{ref-wang2019}{18}{]}.
-
-Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}{]}, consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}\protect\hyperlink{ref-abdi2010}{21}{]}), but they in turn reduce statistical power (1-\(\beta\)){[}\protect\hyperlink{ref-nakagawa2004}{22}{]}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) {[}\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}{]}.
+In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\citep{dennis2019}.
+For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid \citep{wang2019}.
+
+Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) \citep{liu2010, halsey2015}, consequently biasing the conclusions of the study.
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni \citep{abdi2010}), but they in turn reduce statistical power (1-\(\beta\))\citep{nakagawa2004}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) \citep{gelman2012, albers2019}.
 Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
 
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and group sizes {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}\protect\hyperlink{ref-charan2013}{30}{]} which might cause the corrections for lack of compound symmetry to be ineffective.
+On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases \citep{ugrinowitsch2004}.
+Corrections can be made in rm-ANOVA in the absence of compound symmetry \citep{huynh1976, greenhouse1959}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements\citep{haverkamp2017}, and group sizes \citep{keselman2001}.
+In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons \citep{charan2013} which might cause the corrections for lack of compound symmetry to be ineffective.
 
@@
 In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}\protect\hyperlink{ref-gueorguieva2004}{13}{]}.
+When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis \citep{gueorguieva2004}.
 This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
@@
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data \citep{skala2010, vishwanath2009}.
 Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}.
-And even more importantly, LMEMs also assume a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}, making them unsuitable to analyze non-linear data.
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject \citep{pinheiro2006}.
+However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent \citep{gueorguieva2004, barr2013}.
+And even more importantly, LMEMs also assume a linear relationship between the response and time \citep{pinheiro2006}, making them unsuitable to analyze non-linear data.
 
@@
 Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]} .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]} and 3) can easily accommodate missing observations.
+Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data \citep{rose2012, pedersen2019, simpson2018}, health-environment interactions \citep{yang2012} and the dynamics of government in political science \citep{beck1998}.
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements \citep{wood2017} and 3) can easily accommodate missing observations.
 Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
@@
 The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
+In particular, \(\textsf{R}\) \citep{r} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} \citep{wood2016, wood2017} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models \citep{haverkamp2017}.
 Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
@@
 Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies \citep{vishwanath2009}.
 The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
@@
 
-A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish {[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}.
+A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish \citep{pinheiro2006, west2014}.
 A LMEM with interaction between time and treatment for a longitudinal study can be written as:
@@
 The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
+In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level\citep{pinheiro2006}.
 However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
@@
 In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{41}{]} (For an in-depth analysis of the covariance matrix see {[}\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}{]}).
+The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study\citep{wolfinger1996} (For an in-depth analysis of the covariance matrix see \citep{west2014, weiss2005}).
 
 In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}{]}.
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}\protect\hyperlink{ref-maxwell2017}{44}{]}; and it is higher between consecutive measurements {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and they are not robust if the group sizes are unbalanced {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
+Under this assumption, the between-subject variance and within-subject correlation are constant across time \citep{weiss2005, geisser1958, huynh1976}.
+However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time \citep{maxwell2017}; and it is higher between consecutive measurements \citep{gueorguieva2004, ugrinowitsch2004}.
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)\citep{huynh1976, greenhouse1959} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements\citep{haverkamp2017}, and they are not robust if the group sizes are unbalanced \citep{keselman2001}.
 Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
 
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others \citep{pinheiro2006}.
 Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
@@
 Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator \citep{molenberghs2004}.
 Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}\protect\hyperlink{ref-weiss2005}{42}{]}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{46}{]}.
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{47}{]}, and in the case of MNAR the missing values are dependent on their value.
+Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) \citep{weiss2005}.
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest \citep{scheffer2002}.
+If the data are MCAR, this means that the missingness is completely unrelated to the collected information \citep{potthoff2006}, and in the case of MNAR the missing values are dependent on their value.
 
 An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{48}{]}.
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}\protect\hyperlink{ref-west2014}{40}{]}.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size \citep{ma2012}.
+In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR \citep{west2014}.
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
@@
 To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{49}{]}.
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) \citep{r} and the package \emph{nlme} \citep{nlme}.
 
@@
 
-Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors {[}\protect\hyperlink{ref-nelder1972}{50}{]}.
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{51}{]}.
+Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors \citep{nelder1972}.
+In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case\citep{simpson2018, wood2017, hastie1987}.
 A GAM model can be written as:
@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
+One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference \citep{beck1998}.
 In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
@@
 The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
+A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist \citep{simpson2018}.
 For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{52}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response \citep{hefley2017}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
 In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
-Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
+Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data \citep{wood2017, simpson2018}.
+Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness \citep{wegman1983}.
 Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}\protect\hyperlink{ref-mcelreath2018}{54}{]}.
-In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) {[}\protect\hyperlink{ref-miller2019}{55}{]}.
-Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{56}{]}.
-Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief \citep{mcelreath2018}.
+In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs \citep{wood2017}.
+GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) \citep{miller2019}.
+Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model \citep{laird1982, pedersen2019}.
+Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals \citep{pedersen2019}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage \citep{wood2017}.
 This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{55}{]} and {[}\protect\hyperlink{ref-marra2012}{57}{]}.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in \citep{miller2019, wood2017, simpson2018} and \citep{marra2012}.
 With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in \citep{vishwanath2009}.
 In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
@@
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in {[}\protect\hyperlink{ref-hefley2017}{52}{]}.
+A detailed description on basis functions and correlations can be found in \citep{hefley2017}.
 
@@
 This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
-The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths \citep{simpson2018}.
+The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) \citep{wood2017}.
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{58}{]}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{59}{]}.
-
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} \citep{gratia}.
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in \citep{wood2017} and \citep{harezlak2018}.
+
+One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?''.
+The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
 This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
@@
 Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}.
-This ``model misspecification'' error, also is known as a ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{60}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+This ``model misspecification'' error, also is known as a ``Type III'' error \citep{dennis2019} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) \citep{lang2015} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
 
 In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{61}{]}.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) \citep{hefley2017, pedersen2019} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research \citep{hastie1995}.
 This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
@@
 However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} using \(\textsf{R}\) and the package \emph{mgcv}{[}\protect\hyperlink{ref-wood2017}{37}{]} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature \citep{vishwanath2009} using \(\textsf{R}\) and the package \emph{mgcv}\citep{wood2017} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
 One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{62},\protect\hyperlink{ref-weissgerber2018}{63}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research \citep{begley2015, weissgerber2018}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 
+\hypertarget{declaration-of-conflicting-interests}{%
+\section{Declaration of Conflicting Interests}\label{declaration-of-conflicting-interests}}
+
+The Authors declare that there is no conflict of interest.
+
 \FloatBarrier
@@
 \begin{CSLReferences}{0}{0}
-\leavevmode\hypertarget{ref-roblyer2011}{}%
-\CSLLeftMargin{{[}1{]} }
-\CSLRightInline{D.
-Roblyer, S.
-Ueda, A.
-Cerussi, W.
-Tanamai, A.
-Durkin, R.
-Mehta, D.
-Hsiang, J.A.
-Butler, C.
-McLaren, W.P.
-Chen, B.
-Tromberg, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences of the United States of America.
-108 (2011) 14626--14631.
-\url{https://doi.org/10.1073/pnas.1013103108}.}
-
-\leavevmode\hypertarget{ref-tank2020}{}%
-\CSLLeftMargin{{[}2{]} }
-\CSLRightInline{A.
-Tank, H.M.
-Peterson, V.
-Pera, S.
-Tabassum, A.
-Leproux, T.
-O'Sullivan, E.
-Jones, H.
-Cabral, N.
-Ko, R.S.
-Mehta, B.J.
-Tromberg, D.
-Roblyer, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research.
-22 (2020) 1--10.
-\url{https://doi.org/doi:10.1186/s13058-020-01262-1}.}
-
-\leavevmode\hypertarget{ref-pavlov2018}{}%
-\CSLLeftMargin{{[}3{]} }
-\CSLRightInline{M.V.
-Pavlov, T.I.
-Kalganova, Y.S.
-Lyubimtseva, V.I.
-Plekhanov, G.Y.
-Golubyatnikov, O.Y.
-Ilyinskaya, A.G.
-Orlova, P.V.
-Subochev, D.V.
-Safonov, N.M.
-Shakhova, A.V.
-Maslennikova, {Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy}, {Journal of Biomedical Optics}.
-{23} (2018).
-\url{https://doi.org/\%7B10.1117/1.JBO.23.9.091410\%7D}.}
-
-\leavevmode\hypertarget{ref-demidov2018}{}%
-\CSLLeftMargin{{[}4{]} }
-\CSLRightInline{V.
-Demidov, A.
-Maeda, M.
-Sugita, V.
-Madge, S.
-Sadanand, C.
-Flueraru, I.A.
-Vitkin, {Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography}, {Scientific Reports}.
-{8} (2018).
-\url{https://doi.org/\%7B10.1038/s41598-017-18635-w\%7D}.}
-
-\leavevmode\hypertarget{ref-ritter2001}{}%
-\CSLLeftMargin{{[}5{]} }
-\CSLRightInline{G.
-Ritter, L.
-Cohen, C.
-Williams, E.
-Richards, L.
-Old, S.
-Welt, {Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}, {Cancer Research}.
-{61} (2001) 6851--6859.}
-
-\leavevmode\hypertarget{ref-roth2017}{}%
-\CSLLeftMargin{{[}6{]} }
-\CSLRightInline{E.M.
-Roth, A.C.
-Goldberg, A.L.
-Catapano, A.
-Torri, G.D.
-Yancopoulos, N.
-Stahl, A.
-Brunet, G.
-Lecorps, H.M.
-Colhoun, {Antidrug antibodies in atients treated with alirocumab}, {New England Journal of Medicine}.
-{376} (2017) 1589--1590.
-\url{https://doi.org/\%7B10.1056/NEJMc1616623\%7D}.}
-
-\leavevmode\hypertarget{ref-jones2018}{}%
-\CSLLeftMargin{{[}7{]} }
-\CSLRightInline{J.D.
-Jones, H.E.
-Ramser, A.E.
-Woessner, K.P.
-Quinn, {In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing}, {Communications Biology}.
-{1} (2018).
-\url{https://doi.org/\%7B10.1038/s42003-018-0206-4\%7D}.}
-
-\leavevmode\hypertarget{ref-skala2010}{}%
-\CSLLeftMargin{{[}8{]} }
-\CSLRightInline{M.C.
-Skala, A.
-Fontanella, L.
-Lan, J.A.
-Izatt, M.W.
-Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics.
-15 (2010).
-\url{https://doi.org/10.1117/1.3285584}.}
-
-\leavevmode\hypertarget{ref-greening2018}{}%
-\CSLLeftMargin{{[}9{]} }
-\CSLRightInline{G.J.
-Greening, K.P.
-Miller, C.R.
-Spainhour, M.D.
-Cato, T.J.
-Muldoon, {Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy}, {Biomedical Optics Express}.
-{9} (2018) 2871--2886.
-\url{https://doi.org/\%7B10.1364/BOE.9.002871\%7D}.}
-
-\leavevmode\hypertarget{ref-sio2016}{}%
-\CSLLeftMargin{{[}10{]} }
-\CSLRightInline{T.T.
-Sio, P.J.
-Atherton, B.J.
-Birckhead, D.J.
-Schwartz, J.A.
-Sloan, D.K.
-Seisler, J.A.
-Martenson, C.L.
-Loprinzi, P.C.
-Griffin, R.F.
-Morton, J.C.
-Anders, T.J.
-Stoffel, R.E.
-Haselow, R.B.
-Mowat, M.A.N.
-Wittich, J.D.
-Bearden III, R.C.
-Miller, {Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {{[}}alliance{]})}, {Supportive Care in Cancer}.
-{24} (2016) 3847--3855.
-\url{https://doi.org/\%7B10.1007/s00520-016-3213-3\%7D}.}
-
-\leavevmode\hypertarget{ref-kamstra2015}{}%
-\CSLLeftMargin{{[}11{]} }
-\CSLRightInline{J.I.
-Kamstra, P.U.
-Dijkstra, M.
-van Leeuwen, J.L.N.
-Roodenburg, J.A.
-Langendijk, {Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study}, {Oral Oncology}.
-{51} (2015) 548--555.
-\url{https://doi.org/\%7B10.1016/j.oraloncology.2015.01.016\%7D}.}
-
-\leavevmode\hypertarget{ref-wagenmakers2008}{}%
-\CSLLeftMargin{{[}12{]} }
-\CSLRightInline{E.-J.
-Wagenmakers, M.
-Lee, T.
-Lodewyckx, G.J.
-Iverson, Bayesian versus frequentist inference, in: H.
-Hoijtink, I.
-Klugkist, P.A.
-Boelen (Eds.), Bayesian Evaluation of Informative Hypotheses, Springer New York, New York, NY, 2008: pp.
-181--207.
-\url{https://doi.org/10.1007/978-0-387-09612-4_9}.}
-
-\leavevmode\hypertarget{ref-gueorguieva2004}{}%
-\CSLLeftMargin{{[}13{]} }
-\CSLRightInline{R.
-Gueorguieva, J.H.
-Krystal, {Move over ANOVA - Progress in analyzing repeated-measures data and its reflection in papers published in the archives of general psychiatry}, Archives of General Psychiatry.
-61 (2004) 310--317.
-\url{https://doi.org/10.1001/archpsyc.61.3.310}.}
-
-\leavevmode\hypertarget{ref-schober2018}{}%
-\CSLLeftMargin{{[}14{]} }
-\CSLRightInline{P.
-Schober, T.R.
-Vetter, Repeated measures designs and analysis of longitudinal data: If at first you do not succeed-try, try again, Anesthesia and Analgesia.
-127 (2018) 569--575.
-\url{https://doi.org/10.1213/ane.0000000000003511}.}
-
-\leavevmode\hypertarget{ref-pinheiro2006}{}%
-\CSLLeftMargin{{[}15{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science \& Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}
-
-\leavevmode\hypertarget{ref-vishwanath2009}{}%
-\CSLLeftMargin{{[}16{]} }
-\CSLRightInline{K.
-Vishwanath, H.
-Yuan, W.T.
-Barry, M.W.
-Dewhirst, N.
-Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia.
-11 (2009) 889--900.
-\url{https://doi.org/10.1593/neo.09580}.}
-
-\leavevmode\hypertarget{ref-dennis2019}{}%
-\CSLLeftMargin{{[}17{]} }
-\CSLRightInline{B.
-Dennis, J.M.
-Ponciano, M.L.
-Taper, S.R.
-Lele, {Errors in statistical inference under model misspecification: evidence, hypothesis testing, and AIC}, {Frontiers in Ecology and Evolution}.
-{7} (2019).
-\url{https://doi.org/\%7B10.3389/fevo.2019.00372\%7D}.}
-
-\leavevmode\hypertarget{ref-wang2019}{}%
-\CSLLeftMargin{{[}18{]} }
-\CSLRightInline{B.
-Wang, Z.
-Zhou, H.
-Wang, X.M.
-Tu, C.
-Feng, {The p-value and model specification in statistics}, {General Psychiatry}.
-{32} (2019).
-\url{https://doi.org/\%7B10.1136/gpsych-2019-100081\%7D}.}
-
-\leavevmode\hypertarget{ref-liu2010}{}%
-\CSLLeftMargin{{[}19{]} }
-\CSLRightInline{C.
-Liu, T.P.
-Cripe, M.-O.
-Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy.
-18 (2010) 1724--1730.
-\url{https://doi.org/10.1038/mt.2010.127}.}
-
-\leavevmode\hypertarget{ref-halsey2015}{}%
-\CSLLeftMargin{{[}20{]} }
-\CSLRightInline{L.G.
-Halsey, D.
-Curran-Everett, S.L.
-Vowler, G.B.
-Drummond, {The fickle \(p\) value generates irreproducible results}, {Nature Methods}.
-{12} (2015) 179--185.
-\url{https://doi.org/\%7B10.1038/nmeth.3288\%7D}.}
-
-\leavevmode\hypertarget{ref-abdi2010}{}%
-\CSLLeftMargin{{[}21{]} }
-\CSLRightInline{H. Abdi, {Holm's sequential Bonferroni procedure}, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.}
-
-\leavevmode\hypertarget{ref-nakagawa2004}{}%
-\CSLLeftMargin{{[}22{]} }
-\CSLRightInline{S.
-Nakagawa, {A farewell to Bonferroni: the problems of low statistical power and publication bias}, {Behavioral Ecology}.
-{15} (2004) 1044--1045.
-\url{https://doi.org/\%7B10.1093/beheco/arh107\%7D}.}
-
-\leavevmode\hypertarget{ref-gelman2012}{}%
-\CSLLeftMargin{{[}23{]} }
-\CSLRightInline{A.
-Gelman, J.
-Hill, M.
-Yajima, {Why we (usually) don't have to worry about multiple comparisons}, {Journal of Research on Educational Effectiveness}.
-{5} (2012) 189--211.
-\url{https://doi.org/\%7B10.1080/19345747.2011.618213\%7D}.}
-
-\leavevmode\hypertarget{ref-albers2019}{}%
-\CSLLeftMargin{{[}24{]} }
-\CSLRightInline{C. Albers, {The problem with unadjusted multiple and sequential statistical testing}, {Nature Communications}. {10} (2019). \url{https://doi.org/\%7B10.1038/s41467-019-09941-0\%7D}.}
-
-\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
-\CSLLeftMargin{{[}25{]} }
-\CSLRightInline{C.
-Ugrinowitsch, G.W.
-Fellingham, M.D.
-Ricard, Limitations of ordinary least squares models in analyzing repeated measures data, Medicine and Science in Sports and Exercise.
-36 (2004) 2144--2148.
-\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.}
-
-\leavevmode\hypertarget{ref-huynh1976}{}%
-\CSLLeftMargin{{[}26{]} }
-\CSLRightInline{H.
-Huynh, L.S.
-Feldt, {Estimation of the Box correction for degrees of freedom from sample data in randomized block and split-plot designs}, Journal of Educational Statistics.
-1 (1976) 69--82.
-\url{https://doi.org/10.3102/10769986001001069}.}
-
-\leavevmode\hypertarget{ref-greenhouse1959}{}%
-\CSLLeftMargin{{[}27{]} }
-\CSLRightInline{S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.}
-
-\leavevmode\hypertarget{ref-haverkamp2017}{}%
-\CSLLeftMargin{{[}28{]} }
-\CSLRightInline{N.
-Haverkamp, A.
-Beauducel, {Violation of the sphericity assumption and its effect on type-I error rates in repeated measures ANOVA and multi-level linear models (MLM)}, {Frontiers in Psychology}.
-{8} (2017).
-\url{https://doi.org/\%7B10.3389/fpsyg.2017.01841\%7D}.}
-
-\leavevmode\hypertarget{ref-keselman2001}{}%
-\CSLLeftMargin{{[}29{]} }
-\CSLRightInline{H.
-Keselman, J.
-Algina, R.
-Kowalchuk, {The analysis of repeated measures designs: A review}, {British Journal of Mathematica \& Statistical Psychology}.
-{54} (2001) 1--20.
-\url{https://doi.org/\%7B10.1348/000711001159357\%7D}.}
-
-\leavevmode\hypertarget{ref-charan2013}{}%
-\CSLLeftMargin{{[}30{]} }
-\CSLRightInline{Jaykaran.
-Charan, N.
-Kantharia, {How to calculate sample size in animal studies?}, Journal of Pharmacology and Pharmacotherapeutics.
-4 (2013) 303--306.
-\url{https://doi.org/10.4103/0976-500X.119726}.}
-
-\leavevmode\hypertarget{ref-barr2013}{}%
-\CSLLeftMargin{{[}31{]} }
-\CSLRightInline{D.J.
-Barr, R.
-Levy, C.
-Scheepers, H.J.
-Tily, {Random effects structure for confirmatory hypothesis testing: Keep it maximal}, {Journal of Memory and Language}.
-{68} (2013) 255--278.
-\url{https://doi.org/\%7B10.1016/j.jml.2012.11.001\%7D}.}
-
-\leavevmode\hypertarget{ref-rose2012}{}%
-\CSLLeftMargin{{[}32{]} }
-\CSLRightInline{N.L.
-Rose, H.D.
-Yang, S.D.
-Turner, G.L.
-Simpson, {An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to Scotland, UK}, Geochimica Et Cosmochimica Acta.
-82 (2012) 113--135.
-\url{https://doi.org/10.1016/j.gca.2010.12.026}.}
-
-\leavevmode\hypertarget{ref-pedersen2019}{}%
-\CSLLeftMargin{{[}33{]} }
-\CSLRightInline{E.J.
-Pedersen, D.L.
-Miller, G.L.
-Simpson, N.
-Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, {PeerJ}.
-7 (2019).
-\url{https://doi.org/10.7717/peerj.6876}.}
-
-\leavevmode\hypertarget{ref-simpson2018}{}%
-\CSLLeftMargin{{[}34{]} }
-\CSLRightInline{G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.}
-
-\leavevmode\hypertarget{ref-yang2012}{}%
-\CSLLeftMargin{{[}35{]} }
-\CSLRightInline{L.
-Yang, G.
-Qin, N.
-Zhao, C.
-Wang, G.
-Song, {Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality}, {BMC Medical Research Methodology}.
-{12} (2012).
-\url{https://doi.org/\%7B10.1186/1471-2288-12-165\%7D}.}
-
-\leavevmode\hypertarget{ref-beck1998}{}%
-\CSLLeftMargin{{[}36{]} }
-\CSLRightInline{N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. (1998) 596--627.}
-
-\leavevmode\hypertarget{ref-wood2017}{}%
-\CSLLeftMargin{{[}37{]} }
-\CSLRightInline{S.N. Wood, {Generalized additive models: An introduction with R, Second Edition}, CRC Press LLC, Philadelphia, PA, 2017.}
-
-\leavevmode\hypertarget{ref-r}{}%
-\CSLLeftMargin{{[}38{]} }
-\CSLRightInline{R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.}
-
-\leavevmode\hypertarget{ref-wood2016}{}%
-\CSLLeftMargin{{[}39{]} }
-\CSLRightInline{S.N.
-Wood, N.
-Pya, B.
-Saefken, Smoothing parameter and model selection for general smooth models, {Journal of the American Statistical Association}.
-{111} (2016) 1548--1563.
-\url{https://doi.org/\%7B10.1080/01621459.2016.1180986\%7D}.}
-
-\leavevmode\hypertarget{ref-west2014}{}%
-\CSLLeftMargin{{[}40{]} }
-\CSLRightInline{B.T.
-West, K.B.
-Welch, A.T.
-Galecki, Linear mixed models: A practical guide using statistical software, second edition, Taylor \& Francis, 2014.
-\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.}
-
-\leavevmode\hypertarget{ref-wolfinger1996}{}%
-\CSLLeftMargin{{[}41{]} }
-\CSLRightInline{R.D.
-Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics.
-1 (1996) 205--230.
-\url{http://www.jstor.org/stable/1400366}.}
-
-\leavevmode\hypertarget{ref-weiss2005}{}%
-\CSLLeftMargin{{[}42{]} }
-\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://books.google.com/books?id=MQ/_bvWDPsEAC}.}
-
-\leavevmode\hypertarget{ref-geisser1958}{}%
-\CSLLeftMargin{{[}43{]} }
-\CSLRightInline{S.
-Geisser, S.W.
-Greenhouse, {An extension of Box's results on the use of the \(F\) distribution in multivariate analysis}, The Annals of Mathematical Statistics.
-29 (1958) 885--891.
-\url{https://doi.org/10.1214/aoms/1177706545}.}
-
-\leavevmode\hypertarget{ref-maxwell2017}{}%
-\CSLLeftMargin{{[}44{]} }
-\CSLRightInline{S.E.
-Maxwell, H.D.
-Delaney, K.
-Kelley, Designing experiments and analyzing data: A model comparison perspective, third edition, Taylor \& Francis, 2017.
-\url{https://books.google.com/books?id=NmFQDwAAQBAJ}.}
-
-\leavevmode\hypertarget{ref-molenberghs2004}{}%
-\CSLLeftMargin{{[}45{]} }
-\CSLRightInline{G.
-Molenberghs, H.
-Thijs, I.
-Jansen, C.
-Beunckens, M.
-Kenward, C.
-Mallinckrodt, R.
-Carroll, {Analyzing incomplete longitudinal clinical trial data}, {Biostatistics}.
-{5} (2004) 445--464.
-\url{https://doi.org/\%7B10.1093/biostatistics/kxh001\%7D}.}
-
-\leavevmode\hypertarget{ref-scheffer2002}{}%
-\CSLLeftMargin{{[}46{]} }
-\CSLRightInline{J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.}
-
-\leavevmode\hypertarget{ref-potthoff2006}{}%
-\CSLLeftMargin{{[}47{]} }
-\CSLRightInline{R.F.
-Potthoff, G.E.
-Tudor, K.S.
-Pieper, V.
-Hasselblad, {Can one assess whether missing data are missing at random in medical studies?}, {Statistical Methods in Medical Research}.
-{15} (2006) 213--234.
-\url{https://doi.org/\%7B10.1191/0962280206sm448oa\%7D}.}
-
-\leavevmode\hypertarget{ref-ma2012}{}%
-\CSLLeftMargin{{[}48{]} }
-\CSLRightInline{Y.
-Ma, M.
-Mazumdar, S.G.
-Memtsoudis, Beyond repeated-measures analysis of variance advanced statistical methods for the analysis of longitudinal data in anesthesia research, {Regional Anesthesia and Pain Medicine}.
-{37} (2012) 99--105.
-\url{https://doi.org/\%7B10.1097/AAP.0b013e31823ebc74\%7D}.}
-
-\leavevmode\hypertarget{ref-nlme}{}%
-\CSLLeftMargin{{[}49{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, {nlme}: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.}
-
-\leavevmode\hypertarget{ref-nelder1972}{}%
-\CSLLeftMargin{{[}50{]} }
-\CSLRightInline{J.A.
-Nelder, R.W.M.
-Wedderburn, Generalized linear models, Journal of the Royal Statistical Society.
-Series A (General).
-135 (1972) 370--384.
-\url{http://www.jstor.org/stable/2344614}.}
-
-\leavevmode\hypertarget{ref-hastie1987}{}%
-\CSLLeftMargin{{[}51{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association.
-82 (1987) 371--386.
-\url{https://doi.org/10.1080/01621459.1987.10478440}.}
-
-\leavevmode\hypertarget{ref-hefley2017}{}%
-\CSLLeftMargin{{[}52{]} }
-\CSLRightInline{T.J.
-Hefley, K.M.
-Broms, B.M.
-Brost, F.E.
-Buderman, S.L.
-Kay, H.R.
-Scharf, J.R.
-Tipton, P.J.
-Williams, M.B.
-Hooten, {The basis function approach for modeling autocorrelation in ecological data}, {Ecology}.
-{98} (2017) 632--646.
-\url{https://doi.org/\%7B10.1002/ecy.1674\%7D}.}
-
-\leavevmode\hypertarget{ref-wegman1983}{}%
-\CSLLeftMargin{{[}53{]} }
-\CSLRightInline{E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.}
-
-\leavevmode\hypertarget{ref-mcelreath2018}{}%
-\CSLLeftMargin{{[}54{]} }
-\CSLRightInline{R. McElreath, {Statistical rethinking: A Bayesian course with examples in R and Stan}, {Chapman and Hall/CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
-
-\leavevmode\hypertarget{ref-miller2019}{}%
-\CSLLeftMargin{{[}55{]} }
-\CSLRightInline{D.L. Miller, Bayesian views of generalized additive modelling, arXiv Preprint arXiv:1902.01330. (2019).}
-
-\leavevmode\hypertarget{ref-laird1982}{}%
-\CSLLeftMargin{{[}56{]} }
-\CSLRightInline{N.M. Laird, J.H. Ware, Random-effects models for longitudinal data, Biometrics. 38 (1982) 963--974. \url{http://www.jstor.org/stable/2529876}.}
-
-\leavevmode\hypertarget{ref-marra2012}{}%
-\CSLLeftMargin{{[}57{]} }
-\CSLRightInline{G.
-Marra, S.N.
-Wood, Coverage properties of confidence intervals for generalized additive model components, {Scandinavian Journal of Statistics}.
-{39} (2012) 53--74.
-\url{https://doi.org/\%7B10.1111/j.1467-9469.2011.00760.x\%7D}.}
-
-\leavevmode\hypertarget{ref-gratia}{}%
-\CSLLeftMargin{{[}58{]} }
-\CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
-
-\leavevmode\hypertarget{ref-harezlak2018}{}%
-\CSLLeftMargin{{[}59{]} }
-\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, {Semiparametric Regression with R}, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
-
-\leavevmode\hypertarget{ref-lang2015}{}%
-\CSLLeftMargin{{[}60{]} }
-\CSLRightInline{T.A.
-Lang, D.G.
-Altman, {Basic statistical reporting for articles published in Biomedical Journals: The ``Statistical Analyses and Methods in the Published Literature{''} or the SAMPL Guidelines}, INTERNATIONAL JOURNAL OF NURSING STUDIES.
-{52} (2015) 5--9.
-\url{https://doi.org/\%7B10.1016/j.ijnurstu.2014.09.006\%7D}.}
-
-\leavevmode\hypertarget{ref-hastie1995}{}%
-\CSLLeftMargin{{[}61{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models for medical research, {Statistical Methods in Medical Research}.
-4 (1995) 187--196.
-\url{https://doi.org/10.1177/096228029500400302}.}
-
-\leavevmode\hypertarget{ref-begley2015}{}%
-\CSLLeftMargin{{[}62{]} }
-\CSLRightInline{C.G.
-Begley, J.P.A.
-Ioannidis, {Reproducibility in Science Improving the Standard for Basic and Preclinical Research}, {Circulation Research}.
-{116} (2015) 116--126.
-\url{https://doi.org/\%7B10.1161/CIRCRESAHA.114.303819\%7D}.}
-
-\leavevmode\hypertarget{ref-weissgerber2018}{}%
-\CSLLeftMargin{{[}63{]} }
-\CSLRightInline{T.L.
-Weissgerber, O.
-Garcia-Valencia, V.D.
-Garovic, N.M.
-Milic, S.J.
-Winham, {Meta-Research: Why we need to report more than 'Data were Analyzed by t-tests or ANOVA'}, Elife.
-7 (2018) e36163.
-\url{https://doi.org/10.7554/eLife.36163}.}
-
 \end{CSLReferences}
 
-\hypertarget{appendix-appendix}{%
-\appendix}
+\hypertarget{appendix}{%
+\section{(APPENDIX)}\label{appendix}}
 
@@
 
+\bibliographystyle{}
+\bibliography{refs.bib}
+
 \end{document}

### 2021-06-21 Manuscripts/Manuscript by chapters -for submission/07-Appendix.Rmd
References do not work
Need to fix the code highlighting. How to add the listing option for all chunks?

--- 2021-06-09 Manuscripts/Manuscript by chapters -v2/07-Appendix.Rmd
@@
----
-output:
-  pdf_document: default
-  html_document: default
----
 # (APPENDIX) Appendix {-}
 
-# Code for Manuscript data
-
-This section presents the code used to generate figures, models and simulated data from Sections 3 and 4 from the main manuscript.
+<!-- allows for page numbering using letters for the Appendix -->
+\appendixpagenumbering
+
+<!-- Allows for figure numbers with letters for Appendix -->
+\counterwithin{figure}{section} 
+# APPENDIX
+
+This section presents the code used to generate figures, models and simulated data from the main manuscript.
 
@@
 
-This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation).
+This section simulates linear and quadratic data in the same manner as in Section \@ref(simulation).
 The linear simulations using Figure \@ref(fig:linear-cases-Appendix) show in panels A and D the simulated mean responses and individual data points.
@@
 ```{r, responses-Appendix,include=TRUE,message=FALSE,echo=TRUE}
-
 ##########Section for calculations###########
@@
 
-```{r, linear-cases-Appendix, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:linear-cases-appendix)'}
+```{r, linear-cases-Appendix, fig.width=14, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:linear-cases-appendix)'}
 # linear response, correlated errors (subject effect)
@@
 
-```{r, quadratic-cases-Appendix, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:quadratic-caption-appendix)'}
+```{r, quadratic-cases-Appendix, fig.width=14, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:quadratic-caption-appendix)'}
 # linear response, correlated errors (subject effect)
@@
 
-# Longitudinal biomedical data simulation and GAMs{#tumor-data-simulation}
+## Longitudinal biomedical data simulation and GAMs{#tumor-data-simulation}
 
@@
 
-# GAM and Linear model plots and Missing data
+## GAM and Linear model plots and Missing data
 
@@
 
-## GAM and Linear model plots
+### GAM and Linear model plots
 

### 2021-06-21 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.Rmd
Need to fix the code highlighting. How to add the listing option for all chunks?

@@
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented.
+We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (two treatment groups, 10 subjects per group, five repeated measures for each group) to demonstrate how these models can be computationally implemented.
 We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing), allowing reliable inference from the data.
@@
 output:
-  rticles::sage_article:
+  bookdown::pdf_book:
+    base_format: rticles::sage_article
     keep_tex: yes
-    #pandoc_args: --listings
     fig_caption: yes
-    #extra_dependencies:
-      #subfig: null #allows for subfigures
-      #breqn: null #line breaks for long equations
-      #caption: ["font={small}"] #size of the figure captions
-      #float: null #allows for control of placement of figures
+    includes:
+      in_header: preamble_SAGE.sty #LaTeX formatting options
 ---
@@
 #knitr::opts_chunk$set(fig.pos = 'H', out.extra = "",error=TRUE)
-
 #global themes for plotting
@@
 thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
-
-#thm<-scale_color_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting
-#thm1<-scale_fill_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting
-
-#this function will remove the default "author information" that is the Pandoc LaTeX template.
-Otherwise, and additional "and" will appear at the end of the author list.Or, the custom template that is in the YAML header and that has the "author" line deleted may be used as well.
-An explanation in detail is provided at:
-#https://github.com/rstudio/bookdown/issues/1158
-# remove_author <- function(x) {
-#   # identify empty author line
-#   i <- grep("^\\\\author\\{\\}$", x)
-#   # be sure it is the one pandoc inserts
-#   if(length(i) != 0 && grepl('^\\\\date\\{', x[i+1])) x <- x[-i]
-#   x
-# }
-# options(bookdown.post.latex = remove_author)
 ```
@@
 
-# Bibliography
-
-<!-- <div id="refs"></div> -->
-
 ```{r child='07-Appendix.Rmd'}
-
 ```

### 2021-06-22 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.tex
code highlighting

--- 2021-06-18 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.tex
@@
 
+\usepackage{float}
+\usepackage{listings}
+\setcounter{secnumdepth}{2} %to add section numbers
+\usepackage{chngcntr} %package to enable figure numbers by sections
+%\counterwithin{figure}{section} %command for chngcntr in order to provide numbers per section that match with the letter for the Appendix
+
+%to have page numbers with letters in the Appendix:
+%solution from https://tex.stackexchange.com/questions/59572/custom-page-numbering-for-appendix
+\newcommand{\appendixpagenumbering}{
+  \break
+  \pagenumbering{arabic}
+  \renewcommand{\thepage}{\thesection-\arabic{page}}
+}
+
+%newcommand{\passthrough}[1]{#1}
+
+%theme colors for the code chunks (originally from latex-solarized on GitHub)
+%https://github.com/jez/latex-solarized
+ \usepackage{xcolor}
+\definecolor{sbase03}{HTML}{002B36}
+\definecolor{sbase02}{HTML}{073642}
+\definecolor{sbase01}{HTML}{586E75}
+\definecolor{sbase00}{HTML}{657B83}
+\definecolor{sbase0}{HTML}{839496}
+\definecolor{sbase1}{HTML}{93A1A1}
+\definecolor{sbase2}{HTML}{EEE8D5}
+\definecolor{sbase3}{HTML}{FDF6E3}
+\definecolor{syellow}{HTML}{B58900}
+\definecolor{sorange}{HTML}{CB4B16}
+\definecolor{sred}{HTML}{DC322F}
+\definecolor{smagenta}{HTML}{D33682}
+\definecolor{sviolet}{HTML}{6C71C4}
+\definecolor{sblue}{HTML}{268BD2}
+\definecolor{scyan}{HTML}{2AA198}
+\definecolor{sgreen}{HTML}{859900}
+%command to set parameter(s) in package listings
+\lstdefinestyle{astyle}{
+    % How/what to match
+    language=R,
+    sensitive=true,
+    % Border (above and below)
+    frame=lines,
+    % Extra margin on line (align with paragraph)
+    xleftmargin=\parindent,
+    % Put extra space under caption
+    belowcaptionskip=1\baselineskip,
+    % Colors
+    backgroundcolor=\color{sbase3},
+    basicstyle=\color{sbase00}\ttfamily,
+    keywordstyle=\color{scyan},
+    commentstyle=\color{sbase1},
+    stringstyle=\color{sblue},
+    numberstyle=\color{sviolet},
+    identifierstyle=\color{sbase00},
+    % Break long lines into multiple lines?
+    breaklines=true,
+    % Show a character for spaces?
+    showstringspaces=false,
+    tabsize=2
+}
+
+\lstset{style=astyle}
+
 \begin{document}
 
-\title{Using generalized additive models to analyze biomedical
-non-linear longitudinal data}
+\title{Using generalized additive models to analyze biomedical non-linear longitudinal data}
 
@@
 
-\author{Ariel I. Mundo\affilnum{1}, John R. Tipton\affilnum{2}, Timothy
-J. Muldoon*\affilnum{1}}
-
-\affiliation{\affilnum{1}{Department of Biomedical Engineering,
-University of Arkansas, Fayetteville, AR, US}\\\affilnum{2}{Department
-of Mathematical Sciences, University of Arkansas, Fayetteville, AR, US}}
-
-\corrauth{Corresponding author}
+\author{Ariel I. Mundo\affilnum{1}, John R. Tipton\affilnum{2}, Timothy J. Muldoon*\affilnum{1}}
+
+\affiliation{\affilnum{1}{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, US}\\\affilnum{2}{Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, US}}
+
+\corrauth{Timothy J. Muldoon}
 
@@
 \begin{abstract}
-In biomedical research, the outcome of longitudinal studies has been
-traditionally analyzed using the \emph{repeated measures analysis of
-variance} (rm-ANOVA) or more recently, \emph{linear mixed models}
-(LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of
-correlation and missing observations, both methodologies share an
-assumption of linearity in the measured response, which results in
-biased estimates and unreliable inference when they are used to analyze
-data where the trends are non-linear, which is a common occurrence in
-biomedical research. In contrast, \emph{generalized additive models}
-(GAMs) relax the linearity assumption, and allow the data to determine
-the fit of the model while permitting missing observations and different
-correlation structures. Therefore, GAMs present an excellent choice to
-analyze non-linear longitudinal data in the context of biomedical
-research. This paper summarizes the limitations of rm-ANOVA and LMEMs
-and uses simulated data to visually show how both methods produce biased
-estimates when used on non-linear data. We also present the basic theory
-of GAMs, and using trends of oxygen saturation in tumors reported in the
-biomedical literature, we simulate example longitudinal data (2
-treatment groups, 10 subjects per group, 5 repeated measures for each
-group) to demonstrate how these models can be computationally
-implemented. We show that GAMs are able to produce estimates that are
-consistent with the trends of biomedical non-linear data even in the
-case when missing observations exist (with 40\% of the simulated
-observations missing), allowing reliable inference from the data. To
-make this work reproducible, the code and data used in this paper are
-available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
+In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs).
+Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
+In contrast, \emph{generalized additive models} (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
+This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
+We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (two treatment groups, 10 subjects per group, five repeated measures for each group) to demonstrate how these models can be computationally implemented.
+We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing), allowing reliable inference from the data.
+To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
 \end{abstract}
 
-\keywords{Cancer biology; Tumor response; generalized additive models;
-simulation; R}
+\keywords{Cancer biology; Tumor response; generalized additive models; simulation; R}
 
@@
 
-Longitudinal studies are designed to repeatedly measure a variable of
-interest in a group (or groups) of subjects, with the intention of
-observing the evolution of effect across time rather than analyzing a
-single time point (e.g., a cross-sectional study). Biomedical research
-frequently uses longitudinal studies to analyze the evolution of a
-``treatment'' effect across multiple time points; and in such studies
-the subjects of analysis range from animals (mice, rats, rabbits), to
-human patients, cells, or blood samples, among many others. Tumor
-response \citep{roblyer2011, tank2020, pavlov2018, demidov2018},
-antibody expression \citep{ritter2001, roth2017}, and cell metabolism
-\citep{jones2018, skala2010} are examples of the different situations
-where researchers have used longitudinal designs to study some
-physiological response. Because the frequency of the measurements in a
-longitudinal study is dependent on the biological phenomena of interest
-and the experimental design of the study, the frequency of such
-measurements can range from minute intervals to study a short-term
-response such as anesthesia effects in animals\citep{greening2018}, to
-weekly measurements to analyze a mid-term response like the evolution of
-dermatitis symptoms in breast cancer patients \citep{sio2016}, to
-monthly measurements to study a long-term response such as mouth opening
-following radiotherapy (RT) in neck cancer patients \citep{kamstra2015}.
-
-Traditionally, a ``frequentist'' or ``classical'' statistical paradigm
-is used in biomedical research to derive inferences from a longitudinal
-study. The frequentist paradigm regards probability as the limit of the
-expected outcome when an experiment is repeated a large number of times
-\citep{wagenmakers2008}, and such view is applied to the analysis of
-longitudinal data by assuming a null hypothesis under a statistical
-model that is often an \emph{analysis of variance over repeated
-measures} (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model
-makes three key assumptions regarding longitudinal data: 1) linearity of
-the response across time, 2) constant correlation across same-subject
-measurements, and 3) observations from each subject are obtained at all
-time points through the study (a condition also known as \emph{complete
-observations}) \citep{gueorguieva2004, schober2018}.
-
-The expected linear behavior of the response through time is a key
-requisite in rm-ANOVA \citep{pinheiro2006}. This ``linearity
-assumption'' in rm-ANOVA implies that the model is misspecified when the
-data does not follow a linear trend, which results in unreliable
-inference. In biomedical research, non-linear trends are the norm rather
-than the exception in longitudinal studies. A particular example of this
-non-linear behavior in longitudinal data arises in measurements of tumor
-response to chemo and/or radiotherapy in preclinical and clinical
-settings \citep{roblyer2011, skala2010, vishwanath2009}. These studies
-have shown that the collected signal does not follow a linear trend over
-time, and presents extreme variability at different time points, making
-the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such data the
-estimates are inevitably biased, because the model is only able to
-accommodate linear trends that fail to adequately represent the
-biological phenomenon of interest.
-
-A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to
-perform repeated comparisons to estimate a \emph{p-value}, which in turn
-is used as a measure of significance. Although it is possible that a
-\emph{post hoc} analysis of rm-ANOVA is able to find ``significant''
-\emph{p-values}( \emph{p}\textless0.05) from non-linear data, the
-validity of such metric is dependent on how adequate the model fits the
-data. In other words, \emph{p-values} are valid only if the model and
-the data have good agreement; if that is not the case, a ``Type III''
-error (known as ``model misspecification'') occurs\citep{dennis2019}.
-For example, model misspecification will occur when a model that is only
-able to explain linear responses (such as rm-ANOVA) is fitted to data
-that follows a quadratic trend, thereby causing the resulting
-\emph{p-values} and parameter estimates to be invalid \citep{wang2019}.
-
-Additionally, the \emph{p-value} itself is highly variable, and multiple
-comparisons can inflate the false positivity rate (Type I error or
-\(\alpha\)) \citep{liu2010, halsey2015}, consequently biasing the
-conclusions of the study. Corrections exist to address the Type I error
-issue of multiple comparisons (such as Bonferroni \citep{abdi2010}), but
-they in turn reduce statistical power (1-\(\beta\))\citep{nakagawa2004},
-and lead to increased Type II error (failing to reject the null
-hypothesis when the null hypothesis is false)
-\citep{gelman2012, albers2019}. Therefore, the tradeoff of \emph{post
-hoc} comparisons in rm-ANOVA between Type I, II and III errors might be
-difficult to resolve in a biomedical longitudinal study where a delicate
-balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA
-(often known as the \emph{compound symmetry assumption}) is typically
-unreasonable because correlation between the measured responses often
-diminishes as the time interval between the observation increases
-\citep{ugrinowitsch2004}. Corrections can be made in rm-ANOVA in the
-absence of compound symmetry \citep{huynh1976, greenhouse1959}, but the
-effectiveness of the correction is limited by the size of the sample,
-the number of measurements\citep{haverkamp2017}, and group sizes
-\citep{keselman2001}. In the case of biomedical research, where living
-subjects are frequently used, sample sizes are often not ``large'' due
-to ethical and budgetary reasons \citep{charan2013} which might cause
-the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study
-can vary between all subjects. For example, in a clinical trial patients
-may voluntarily withdraw, whereas attrition due to injury or weight loss
-in preclinical animal studies is possible. It is even plausible that
-unexpected complications with equipment or supplies arise that prevent
-the researcher from collecting measurements at certain time points. In
-each of these missing data scenarios, the \emph{complete observations}
-assumption of classical rm-ANOVA is violated. When incomplete
-observations occur, a rm-ANOVA model is fit by excluding all subjects
-with missing observations from the analysis \citep{gueorguieva2004}.
-This elimination of partially missing data from the analysis can result
-in increased costs if the desired statistical power is not met with the
-remaining observations, because it would be necessary to enroll more
-subjects. At the same time, if the excluded observations contain
-insightful information that is not used, their elimination from the
-analysis may limit the demonstration of significant differences between
-groups.
-
-During the last decade, the biomedical community has started to
-recognize the limitations of rm-ANOVA in the analysis of longitudinal
-data. The recognition on the shortcomings of rm-ANOVA is exemplified by
-the use of linear mixed effects models (LMEMs) by certain groups to
-analyze longitudinal tumor response data
-\citep{skala2010, vishwanath2009}. Briefly, LMEMs incorporate
-\emph{fixed effects}, which correspond to the levels of experimental
-factors in the study (e.g., the different drug regimens in a clinical
-trial), and \emph{random effects}, which account for random variation
-within the population (e.g., the individual-level differences not due to
-treatment such as weight or age). When compared to the traditional
-rm-ANOVA, LMEMs are more flexible as they can accommodate missing
-observations for multiple subjects and allow different modeling
-strategies for the variability within each measure in every subject
-\citep{pinheiro2006}. However, LMEMs impose restrictions in the
-distribution of the errors of the random effects, which need to be
-normally distributed and independent \citep{gueorguieva2004, barr2013}.
-And even more importantly, LMEMs also assume a linear relationship
-between the response and time \citep{pinheiro2006}, making them
-unsuitable to analyze non-linear data.
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly
-restrictive assumptions regarding the linearity of the response, there
-is a need for biomedical researchers to explore the use of additional
-statistical tools that allow the data (and not an assumption in trend)
-to determine the trend of the fitted model, to enable appropriate
-inference. In this regard, generalized additive models (GAMs) present an
-alternative approach to analyze longitudinal data. Although not
-frequently used by the biomedical community, these semi-parametric
-models are customarily used in other fields to analyze longitudinal
-data. Examples of the use of GAMs include the analysis of temporal
-variations in geochemical and palaeoecological data
-\citep{rose2012, pedersen2019, simpson2018}, health-environment
-interactions \citep{yang2012} and the dynamics of government in
-political science \citep{beck1998} . There are several advantages of
-GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible
-class of smooth responses that enable the data to dictate the trend in
-the fit of the model, 2) they can model non-constant correlation between
-repeated measurements \citep{wood2017} and 3) can easily accommodate
-missing observations. Therefore, GAMs can provide a more flexible
-statistical approach to analyze non-linear biomedical longitudinal data
-than LMEMs and rm-ANOVA.
-
-The current advances in programming languages designed for statistical
-analysis (specifically \(\textsf{R}\)), have eased the computational
-implementation of traditional models such as rm-ANOVA and more complex
-approaches such as LMEMs and GAMs. In particular, \(\textsf{R}\)
-\citep{r} has an extensive collection of documentation and functions to
-fit GAMs in the package \emph{mgcv} \citep{wood2016, wood2017} that not
-only speed up the initial stages of the analysis but also enable the use
-of advanced modeling structures (e.g.~hierarchical models, confidence
-interval comparisons) without requiring advanced programming skills from
-the user. At the same time, \(\textsf{R}\) has many tools that simplify
-data simulation, an emerging strategy used to test statistical models
-\citep{haverkamp2017}. Data simulation methods allow the researcher to
-create and explore different alternatives for analysis without
-collecting information in the field, reducing the time window between
-experiment design and its implementation, and simulation can be also
-used for power calculations and study design questions.
-
-This work provides biomedical researchers with a clear understanding of
-the theory and the practice of using GAMs to analyze longitudinal data
-using by focusing on four areas. First, the limitations of LMEMs and
-rm-ANOVA regarding linearity of response, constant correlation
-structures and missing observations are explained in detail. Second, the
-key theoretical elements of GAMs are presented using clear and simple
-mathematical notation while explaining the context and interpretation of
-the equations. Third, we illustrate the type of non-linear longitudinal
-data that often occurs in biomedical research using simulated data that
-reproduces patterns in previously reported studies
-\citep{vishwanath2009}. The simulated data experiments highlight the
-differences in inference between rm-ANOVA, LMEMs and GAMs on data
-similar to what is commonly observed in biomedical studies. Finally,
-reproducibility is emphasized by providing the code to generate the
-simulated data and the implementation of different models in
-\(\textsf{R}\), in conjunction with a step-by-step guide demonstrating
-how to fit models of increasing complexity.
-
-In summary, this work will allow biomedical researchers to identify when
-the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze
-longitudinal data, and provide guidance on the implementation of these
-models to improve the standards for reproducibility in biomedical
-research.
+Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
+Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
+Tumor response \citep{roblyer2011, tank2020, pavlov2018, demidov2018}, antibody expression \citep{ritter2001, roth2017}, and cell metabolism \citep{jones2018, skala2010} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
+Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals\citep{greening2018}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients \citep{sio2016}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients \citep{kamstra2015}.
+
+Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
+The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times \citep{wagenmakers2008}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
+The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) \citep{gueorguieva2004, schober2018}.
+
+The expected linear behavior of the response through time is a key requisite in rm-ANOVA \citep{pinheiro2006}.
+This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
+In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings \citep{roblyer2011, skala2010, vishwanath2009}.
+These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
+Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
+
+A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
+Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless{}0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
+In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\citep{dennis2019}.
+For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid \citep{wang2019}.
+
+Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) \citep{liu2010, halsey2015}, consequently biasing the conclusions of the study.
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni \citep{abdi2010}), but they in turn reduce statistical power (1-\(\beta\))\citep{nakagawa2004}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) \citep{gelman2012, albers2019}.
+Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
+
+On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases \citep{ugrinowitsch2004}.
+Corrections can be made in rm-ANOVA in the absence of compound symmetry \citep{huynh1976, greenhouse1959}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements\citep{haverkamp2017}, and group sizes \citep{keselman2001}.
+In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons \citep{charan2013} which might cause the corrections for lack of compound symmetry to be ineffective.
+
+Due to a variety of causes, the number of observations during a study can vary between all subjects.
+For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible.
+It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
+In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
+When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis \citep{gueorguieva2004}.
+This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
+At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
+
+During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
+The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data \citep{skala2010, vishwanath2009}.
+Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject \citep{pinheiro2006}.
+However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent \citep{gueorguieva2004, barr2013}.
+And even more importantly, LMEMs also assume a linear relationship between the response and time \citep{pinheiro2006}, making them unsuitable to analyze non-linear data.
+
+As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
+In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
+Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
+Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data \citep{rose2012, pedersen2019, simpson2018}, health-environment interactions \citep{yang2012} and the dynamics of government in political science \citep{beck1998}.
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements \citep{wood2017} and 3) can easily accommodate missing observations.
+Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
+
+The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
+In particular, \(\textsf{R}\) \citep{r} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} \citep{wood2016, wood2017} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models \citep{haverkamp2017}.
+Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
+
+This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
+First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations are explained in detail.
+Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
+Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies \citep{vishwanath2009}.
+The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
+Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in \(\textsf{R}\), in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
+
+In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models to improve the standards for reproducibility in biomedical research.
 
 \hypertarget{challenges-presented-by-longitudinal-studies}{%
-\section{Challenges presented by longitudinal
-studies}\label{challenges-presented-by-longitudinal-studies}}
+\section{Challenges presented by longitudinal studies}\label{challenges-presented-by-longitudinal-studies}}
 
 \hypertarget{the-repeated-measures-anova-and-linear-mixed-model}{%
-\subsection{The repeated measures ANOVA and Linear Mixed
-Model}\label{the-repeated-measures-anova-and-linear-mixed-model}}
-
-The \emph{repeated measures analysis of variance} (rm-ANOVA) and the
-\emph{linear mixed model} (LMEM) are the most commonly used statistical
-analysis for longitudinal data in biomedical research. These statistical
-methodologies require certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1)
-linear relationship between covariates and response, 2) a constant
-correlation between measurements, and, 3) complete observations for all
-subjects. Each one of these assumptions is discussed below.
+\subsection{The repeated measures ANOVA and Linear Mixed Model}\label{the-repeated-measures-anova-and-linear-mixed-model}}
+
+The \emph{repeated measures analysis of variance} (rm-ANOVA) and the \emph{linear mixed model} (LMEM) are the most commonly used statistical analysis for longitudinal data in biomedical research.
+These statistical methodologies require certain assumptions for the model to be valid.
+From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
+Each one of these assumptions is discussed below.
 
@@
 \hypertarget{the-repeated-measures-anova-case}{%
-\subsubsection{The repeated measures ANOVA
-case}\label{the-repeated-measures-anova-case}}
-
-In a longitudinal biomedical study, two or more groups of subjects
-(e.g., human subject, mice, samples) are subject to different treatments
-(e.g., a ``treatment'' group receives a novel drug or intervention vs.~a
-``control'' group that receives a placebo), and measurements from each
-subject within each group are collected at specific time points. The
-collected response is modeled with \emph{fixed} components. The
-\emph{fixed} component can be understood as a constant value in the
-response which the researcher is interested in measuring, i.e., the
-average effect of the novel drug/intervention in the ``treatment''
-group.
-
-Mathematically speaking, a rm-ANOVA model with an interaction can be
-written as:
+\subsubsection{The repeated measures ANOVA case}\label{the-repeated-measures-anova-case}}
+
+In a longitudinal biomedical study, two or more groups of subjects (e.g., human subject, mice, samples) are subject to different treatments (e.g., a ``treatment'' group receives a novel drug or intervention vs.~a ``control'' group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
+The collected response is modeled with \emph{fixed} components.
+The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug/intervention in the ``treatment'' group.
+
+Mathematically speaking, a rm-ANOVA model with an interaction can be written as:
 
@@
 y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
-(\#eq:linear-model)
+\label{eq:linear-model}
 \end{equation}
 
-In this model \(y_{ijt}\) is the response for subject \(i\), in
-treatment group \(j\) at time \(t\), which can be decomposed in a mean
-value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment
-(\(treatment_j\)) and their interaction \(time_t*treatment_j\) which
-have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\),
-respectively. Independent errors \(\varepsilon_{ijt}\) represent random
-variation not explained by the \emph{fixed} effects, and are assumed to
-be \(\sim N(0,\sigma^2)\) (independently and identically normally
-distributed with mean zero and variance \(\sigma^2\)). In a biomedical
-research context, suppose two treatments groups are used in a study
-(e.g., ``placebo'' vs.~``novel drug'' or ``saline''
-vs.~``chemotherapy''). Then, the group terms in Equation
-@ref(eq:linear-model) can be written as below with \(treatment_j=0\)
-representing the first treatment group (Group A) and \(treatment_j=1\)
-representing the second treatment group (Group B). With this notation,
-the linear model then can be expressed as
+In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
+Independent errors \(\varepsilon_{ijt}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
+In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy'').
+Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B).
+With this notation, the linear model then can be expressed as
 
@@
 \end{cases}
-(\#eq:ANOVA-by-group)
+\label{eq:ANOVA-by-group}
 \end{equation}
 
-To further simplify the expression, substitute
-\(\widetilde{\beta_{0}}=\beta_0+\beta_{2}\) and
-\(\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}\) in the equation for Group
-B. This substitution allows for a different intercept and slope for
-Groups A and B. The model is then written as
+To further simplify the expression, substitute \(\widetilde{\beta_{0}}=\beta_0+\beta_{2}\) and \(\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}\) in the equation for Group B.
+This substitution allows for a different intercept and slope for Groups A and B.
+The model is then written as
 
@@
 \end{cases}
-(\#eq:ANOVA-lines)
+\label{eq:ANOVA-lines}
 \end{equation}
 
-Presenting the model in this manner makes clear that when treating
-different groups, an rm-ANOVA model is able to accommodate non-parallel
-lines in each case (different intercepts and slopes per group). In other
-words, the rm-ANOVA model ``expects'' a linear relationship between the
-covariates and the response, this means that either presented as
-Equation @ref(eq:linear-model), Equation @ref(eq:ANOVA-by-group) or
-Equation @ref(eq:ANOVA-lines), an rm-ANOVA model is only able to
-accommodate linear patterns in the data. If the data show non-linear
-behavior, the rm-ANOVA model will approximate this behavior with
-non-parallel lines.
+Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
+In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response, this means that either presented as Equation \eqref{eq:linear-model}, Equation \eqref{eq:ANOVA-by-group} or Equation \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data.
+If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
 
 \hypertarget{the-linear-mixed-model-case-lmem}{%
-\subsubsection{The Linear Mixed Model Case
-(LMEM)}\label{the-linear-mixed-model-case-lmem}}
-
-A LMEM is a class of statistical models that incorporates \emph{fixed
-effects} to model the relationship between the covariates and the
-response, and \emph{random effects} to model subject variability that is
-not the primary focus of the study but that might be important to
-distinguish \citep{pinheiro2006, west2014}. A LMEM with interaction
-between time and treatment for a longitudinal study can be written as:
+\subsubsection{The Linear Mixed Model Case (LMEM)}\label{the-linear-mixed-model-case-lmem}}
+
+A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish \citep{pinheiro2006, west2014}.
+A LMEM with interaction between time and treatment for a longitudinal study can be written as:
 
@@
 y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
-(\#eq:LMEM)
+\label{eq:LMEM}
 \end{equation}
 
-When Equation @ref(eq:linear-model) and Equation @ref(eq:LMEM) are
-compared, it is easily noticeable that LMEM and rm-ANOVA have the same
-construction regarding the \emph{fixed effects} of time and treatment,
-but that the LMEM incorporates an additional source of variation (the
-term \(\mu_{ij}\)). This term \(\mu_{ij}\) is the one that corresponds
-to the \emph{random effect}, accounting for variability in each subject
-(subject\(_i\)) within each group (group\(_j\)). The \emph{random}
-component can also be understood as used to model some ``noise'' in the
-response, but that is intended to be analyzed and disentangled from the
-``global noise'' term \(\varepsilon_{ijt}\) from Equation
-@ref(eq:linear-model).
-
-For example, if the blood concentration of the drug is measured in
-certain subjects in the early hours of the morning while other subjects
-are measured in the afternoon, it is possible that the difference in the
-collection time introduces some ``noise'' in the data. As the name
-suggests, this ``random'' variability needs to be modeled as a variable
-rather than as a constant value. The \emph{random effect} \(\mu_{ij}\)
-in Equation @ref(eq:LMEM) is assumed to be
-\(\mu_{ij} \sim N(0,\sigma^2_\mu)\). In essence, the \emph{random
-effect} in a LMEM enables to fit models with different slopes at the
-subject-level\citep{pinheiro2006}. However, the expected linear
-relationship of the covariates and the response in Equation
-@ref(eq:linear-model) and in Equation @ref(eq:LMEM) is essentially the
-same, representing a major limitation of LMEMs to fit a non-linear
-response.
+When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)).
+This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)).
+The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the ``global noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.
+
+For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data.
+As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
+The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
+In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level\citep{pinheiro2006}.
+However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
 
 \hypertarget{covariance-in-rm-anova-and-lmems}{%
-\subsection{Covariance in rm-ANOVA and
-LMEMs}\label{covariance-in-rm-anova-and-lmems}}
-
-In a longitudinal study there is an expected \emph{covariance} between
-repeated measurements on the same subject, and because repeated measures
-occur in the subjects within each group, there is a \emph{covariance}
-between measurements at each time point within each group. The
-\emph{covariance matrix} (also known as the variance-covariance matrix)
-is a matrix that captures the variation between and within subjects in a
-longitudinal study\citep{wolfinger1996} (For an in-depth analysis of the
-covariance matrix see \citep{west2014, weiss2005}).
-
-In the case of an rm-ANOVA analysis, it is typically assumed that the
-covariance matrix has a specific construction known as \emph{compound
-symmetry} (also known as ``sphericity'' or ``circularity''). Under this
-assumption, the between-subject variance and within-subject correlation
-are constant across time \citep{weiss2005, geisser1958, huynh1976}.
-However, it has been shown that this condition is frequently not
-justified because the correlation between measurements tends to change
-over time \citep{maxwell2017}; and it is higher between consecutive
-measurements \citep{gueorguieva2004, ugrinowitsch2004}. Although
-corrections can be made (such as Huyhn-Feldt or
-Greenhouse-Geisser)\citep{huynh1976, greenhouse1959} the effectiveness
-of each correction is limited because it depends on the size of the
-sample,the number of repeated measurements\citep{haverkamp2017}, and
-they are not robust if the group sizes are unbalanced
-\citep{keselman2001}. Because biomedical longitudinal studies are often
-limited in sample size and can have an imbalanced design, the
-corrections required to use an rm-ANOVA model may not be able to provide
-a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow
-different structures for the variance-covariance matrix including
-exponential, autoregressive of order 1, rational quadratic and others
-\citep{pinheiro2006}. Nevertheless, the analysis required to determine
-an appropriate variance-covariance structure for the data can be a
-challenging process by itself. Overall, the spherical assumption for
-rm-ANOVA may not capture the natural variations of the correlation in
-the data, and can bias the inferences from the analysis.
+\subsection{Covariance in rm-ANOVA and LMEMs}\label{covariance-in-rm-anova-and-lmems}}
+
+In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
+The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study\citep{wolfinger1996} (For an in-depth analysis of the covariance matrix see \citep{west2014, weiss2005}).
+
+In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
+Under this assumption, the between-subject variance and within-subject correlation are constant across time \citep{weiss2005, geisser1958, huynh1976}.
+However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time \citep{maxwell2017}; and it is higher between consecutive measurements \citep{gueorguieva2004, ugrinowitsch2004}.
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)\citep{huynh1976, greenhouse1959} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements\citep{haverkamp2017}, and they are not robust if the group sizes are unbalanced \citep{keselman2001}.
+Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
+
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others \citep{pinheiro2006}.
+Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
+Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
 
@@
 
-Missing observations are an issue that arises frequently in longitudinal
-studies. In biomedical research, this situation can be caused by reasons
-beyond the control of the investigator \citep{molenberghs2004}. Dropout
-from patients and attrition or injury in animals are among the reasons
-for missing observations. Statistically, missing information can be
-classified as \emph{missing at random} (MAR), \emph{missing completely
-at random} (MCAR), and \emph{missing not at random} (MNAR)
-\citep{weiss2005}. In a MAR scenario, the pattern of the missing
-information is related to some variable in the data, but it is not
-related to the variable of interest \citep{scheffer2002}. If the data
-are MCAR, this means that the missingness is completely unrelated to the
-collected information \citep{potthoff2006}, and in the case of MNAR the
-missing values are dependent on their value.
-
-An rm-ANOVA model assumes complete observations for all subjects, and
-therefore subjects with one or more missing observations are excluded
-from the analysis. This is inconvenient because the remaining subjects
-might not accurately represent the population, and statistical power is
-affected by this reduction in sample size \citep{ma2012}. In the case of
-LMEMs, inferences from the model are valid when missing observations in
-the data exist that are MAR or MCAR \citep{west2014}. For example, if
-attrition occurs in all mice that had lower weights at the beginning of
-a chemotherapy response study, the missing data can be considered MAR
-because the missigness is unrelated to other variables of interest.
+Missing observations are an issue that arises frequently in longitudinal studies.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator \citep{molenberghs2004}.
+Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
+Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) \citep{weiss2005}.
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest \citep{scheffer2002}.
+If the data are MCAR, this means that the missingness is completely unrelated to the collected information \citep{potthoff2006}, and in the case of MNAR the missing values are dependent on their value.
+
+An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size \citep{ma2012}.
+In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR \citep{west2014}.
+For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
 
 \hypertarget{simulation}{%
-\subsection{What do an rm-ANOVA and LMEM fit look like? A visual
-representation using simulated data}\label{simulation}}
-
-To visually demonstrate the limitations of rm-ANOVA an LMEMs for
-non-linear longitudinal data, this section presents a simulation
-experiment of a normally distributed response of two groups of 10
-subjects each. An rm-ANOVA model (Equation @ref(eq:linear-model)), and a
-LMEM (Equation @ref(eq:LMEM)) are fitted to each group, using
-\(\textsf{R}\) \citep{r} and the package \emph{nlme} \citep{nlme}.
-
-Briefly, two cases for the mean responses for each group are considered:
-in the first case, the mean response in each group is a linear function
-over time with different intercepts and slopes; a negative slope is used
-for Group 1 and a positive slope is used for Group 2 (Figure
-@ref(fig:l-q-response)A). In the second case, a second-degree polynomial
-(quadratic) function is used for the mean response per group: the
-quadratic function is concave down for Group 1 and it is concave up for
-Group 2 (Figure @ref(fig:l-q-response)C). In both the linear and
-quadratic simulated data, the groups start with the same mean value at
-the first time point. This is intentional in order to simulate the
-expected temporal evolution of some physiological quantity, which is
-typical in biomedical experiments where a strong non-linear trend is
-present.
-
-Specifically, the rationale for the chosen linear and quadratic
-functions is the expectation that a measured response in two treatment
-groups is similar in the initial phase of the study, but as therapy
-progresses a divergence in the trend of the response indicates a
-treatment effect. In other words, Group 1 can be thought as a
-``Control'' group and Group 2 as a ``Treatment'' group. From the mean
-response per group (linear or quadratic), the variability or ``error''
-of individual responses within each group is simulated using a
-covariance matrix with compound symmetry (constant variance across
-time). Thus, the response per subject in both the linear and quadratic
-simulation corresponds to the mean response per group plus the error
-(Figure @ref(fig:l-q-response) B,D).
-
-A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for
-linear and non-linear longitudinal appears in Figure
-@ref(fig:linear-cases-Appendix) and Figure
-@ref(fig:quadratic-cases-Appendix) in the Appendix, where simulation
-with compound symmetry and independent errors (errors generated from a
-normal distribution that are not constant over time) and the plot of
-simulated errors, and fitted parameters in presented. We are aware that
-the simulated data used in this section present an extreme case that
-might not occur frequently in biomedical research, but they are used as
-a representation of the consequences of modeling non-linear data with a
-linear model such as rm-ANOVA or LMEMs. Of notice, in Section
-@ref(longitudinal-GAMs) we use simulated data that does follow reported
-trends in the biomedical literature to implement GAMs.
-
-(ref:l-q-response-caption) Simulated responses from two groups with
-correlated errors using a LMEM and a rm-ANOVA model. Top row: linear
-response, bottom row: quadratic response. A: Simulated linear data with
-known mean response (thin lines) and individual responses (points)
-showing the dispersion of the data. D: Simulated quadratic data with
-known mean response (thin lines) and individual responses (points)
-showing the dispersion of the data. B,E: Estimates from the rm-ANOVA
-model for the mean group response (linear of quadratic). Points
-represent the original raw data. The rm-ANOVA model not only fails to
-pick the trend of the quadratic data (D) but also assigns a global
-estimate that does not take between-subject variation. C, F: Estimates
-from the LMEM in the linear and quadratic case. The LMEM incorporates a
-random effect for each subject, but this model and the rm-ANOVA model
-are unable to follow the trend of the data and grossly bias the initial
-estimates for each group in the quadratic case (bottom row).
+\subsection{What do an rm-ANOVA and LMEM fit look like? A visual representation using simulated data}\label{simulation}}
+
+To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) \citep{r} and the package \emph{nlme} \citep{nlme}.
+
+Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}A).
+In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}C).
+In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
+This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments where a strong non-linear trend is present.
+
+Specifically, the rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect.
+In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group.
+From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
+Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D).
+
+A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal appears in Figure \ref{fig:linear-cases-Appendix} and Figure \ref{fig:quadratic-cases-Appendix} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs.
+Of notice, in Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
 
@@
 
-\caption{(ref:l-q-response-caption)}\label{fig:l-q-response}
+\caption{Simulated responses from two groups with correlated errors using a LMEM and a rm-ANOVA model.
+Top row: linear response, bottom row: quadratic response.
+A: Simulated linear data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data.
+D: Simulated quadratic data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data.
+B,E: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
+Points represent the original raw data.
+The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation.
+C, F: Estimates from the LMEM in the linear and quadratic case.
+The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data and grossly bias the initial estimates for each group in the quadratic case (bottom row).}\label{fig:l-q-response}
 \end{figure}
 
-The simulation shows that the fits produced by the LMEM and the rm-ANOVA
-model are good for linear data, as the predictions for the mean response
-are reasonably close to the ``truth'' of the simulated data (Figure
-@ref(fig:l-q-response)A). When the linearity and compound symmetry
-assumptions are met, the rm-ANOVA model approximates well the global
-trend by group (Figure @ref(fig:l-q-response)B). Note that because the
-LMEM incorporates \emph{random effects}, is able to provide estimates
-for each subject and a ``global'' estimate (Figure
-@ref(fig:l-q-response)C).
-
-However, consider the case when the data follows a non-linear trend,
-such as the simulated data in Figure @ref(fig:l-q-response)D. Here, the
-mean response per group was simulated using a quadratic function, and
-errors and individual responses were produced as in Figure
-@ref(fig:l-q-response)A. The mean response in the simulated data with
-quadratic behavior changes in each group through the timeline, and the
-mean value is the same as the initial value by the fifth time point for
-each group. Fitting an rm-ANOVA model (Equation @ref(eq:linear-model))
-or a LMEM (Equation @ref(eq:LMEM)) to this data produces the fit that
-appears in Figure @ref(fig:l-q-response)E, F.
-
-Comparing the fitted responses of the LMEM and the rm-ANOVA models used
-in the simulated quadratic data (Figure @ref(fig:l-q-response)E, F)
-indicates that the models are not capturing the changes within each
-group. Specifically, note that the fitted mean response of both models
-shows that the change (increase for Treatment 1 or decrease for
-Treatment 2) in the response through time points 2 and 4 is not being
-captured. The LMEM is only able to account for between-subject variation
-by providing estimates for each subject (Figure
-@ref(fig:l-q-response)F), but both models are unable to capture the fact
-that the initial values are the same in each group, and instead fit
-non-parallel lines that have initial values that are markedly different
-from the ``true'' initial values in each case (compare Figure
-@ref(fig:l-q-response)D with Figure @ref(fig:l-q-response)E, F). If such
-a change has important physiological implications, both rm-ANOVA and
-LMEMs omit it from the fitted mean response. Thus, even though the model
-correctly detects a divergence between treatment groups, the exact
-nature of this difference is not correctly identified, limiting valuable
-inferences from the data.
-
-This section has used simulation to better convey the limitations of
-linearity and correlation in the response in non-linear data. The models
-fitted to the simulated data were an rm-ANOVA model and a LMEM, where
-the main issue is the expected linear trend in the response. In the
-following section, we present generalized additive models (GAMs) as a
-data-driven alternative method to analyze longitudinal non-linear data
-that overcomes the linearity assumption.
+The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response}A).
+When the linearity and compound symmetry assumptions are met, the rm-ANOVA model approximates well the global trend by group (Figure \ref{fig:l-q-response}B).
+Note that because the LMEM incorporates \emph{random effects}, is able to provide estimates for each subject and a ``global'' estimate (Figure \ref{fig:l-q-response}C).
+
+However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}D.
+Here, the mean response per group was simulated using a quadratic function, and errors and individual responses were produced as in Figure \ref{fig:l-q-response}A.
+The mean response in the simulated data with quadratic behavior changes in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
+Fitting an rm-ANOVA model (Equation \eqref{eq:linear-model}) or a LMEM (Equation \eqref{eq:LMEM}) to this data produces the fit that appears in Figure \ref{fig:l-q-response}E, F.
+
+Comparing the fitted responses of the LMEM and the rm-ANOVA models used in the simulated quadratic data (Figure \ref{fig:l-q-response}E, F) indicates that the models are not capturing the changes within each group.
+Specifically, note that the fitted mean response of both models shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.
+The LMEM is only able to account for between-subject variation by providing estimates for each subject (Figure \ref{fig:l-q-response}F), but both models are unable to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare Figure \ref{fig:l-q-response}D with Figure \ref{fig:l-q-response}E, F).
+If such a change has important physiological implications, both rm-ANOVA and LMEMs omit it from the fitted mean response.
+Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
+
+This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
+The models fitted to the simulated data were an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
+In the following section, we present generalized additive models (GAMs) as a data-driven alternative method to analyze longitudinal non-linear data that overcomes the linearity assumption.
 
 \hypertarget{GAM-theory}{%
-\section{GAMs as a special case of Generalized Linear
-Models}\label{GAM-theory}}
+\section{GAMs as a special case of Generalized Linear Models}\label{GAM-theory}}
 
@@
 
-Generalized linear models (GLMs) are a family of models (which include
-rm-ANOVA and LMEMs) that fit a linear response function to data that may
-not have normally distributed errors \citep{nelder1972}. In contrast,
-GAMs are a family of regression-based methods for estimating smoothly
-varying trends and are a broader class of models that contain the GLM
-family as a special case\citep{simpson2018, wood2017, hastie1987}. A GAM
-model can be written as:
+Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors \citep{nelder1972}.
+In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case\citep{simpson2018, wood2017, hastie1987}.
+A GAM model can be written as:
 
@@
   y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
-  (\#eq:GAM)
+  \label{eq:GAM}
 \end{equation}
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in
-group \(j\), \(\beta_0\) is the expected value at time 0, the change of
-\(y_{ijt}\) over time is represented by the \emph{smooth function}
-\(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and
-parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the
-residual error.
-
-In contrast to the linear functions used to model the relationship
-between the covariates and the response in rm-ANOVA or LMEM, GAMs use
-more flexible \emph{smooth functions}. This approach is advantageous as
-it does not restrict the model to a linear relationship, although a GAM
-can estimate a linear relationship if the data is consistent with a
-linear response. One possible set of functions for
-\(f(x_t\mid \beta_j)\) that allow for non-linear responses are
-polynomials, but a major limitation is that polynomials create a
-``global'' fit as they assume that the same relationship exists
-everywhere, which can cause problems with inference \citep{beck1998}. In
-particular, polynomial fits are known to show boundary effects because
-as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to
-\(\pm \infty\) which is almost always unrealistic and causes bias at the
-endpoints of the time period.
-
-The smooth functional relationship between the covariates and the
-response in GAMs is specified using a semi-parametric relationship that
-can be fit within the GLM framework, by using \emph{basis function}
-expansions of the covariates and by estimating random coefficients
-associated with these basis functions. A \emph{basis} is a set of
-functions that spans the mathematical space where the smooths that
-approximate \(f(x_t\mid \beta_j)\) exist \citep{simpson2018}. For the
-linear model in Equation @ref(eq:linear-model), the basis coefficients
-are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are
-\(time_t\), \(treatment_j\) and \(time_t \times treatment_j\). The basis
-function then, is the combination of basis coefficients and basis
-vectors that map the possible relationship between the covariates and
-the response \citep{hefley2017}, which in the case of Equation
-@ref(eq:linear-model) is restricted to a linear family of functions. In
-the case of Equation @ref(eq:GAM), the basis functions are contained in
-the expression \(f(x_t\mid \beta_j)\), which means that the model allows
-for non-linear relationships among the covariates.
-
-Splines (cubic, thin plate, etc.) are commonly used \emph{basis
-functions}; a cubic spline is a smooth curve constructed from cubic
-polynomials joined together in a manner that enforces smoothness, and
-thin plate regression splines are an optimized version that work well
-with noisy data \citep{wood2017, simpson2018}. Splines have a long
-history in solving semi-parametric statistical problems and are often a
-default choice to fit GAMs as they are a simple, flexible and powerful
-option to obtain smoothness \citep{wegman1983}. Therefore, this
-data-driven flexibility in GAMs overcomes the limitation that occurs in
-LMEMs and rm-ANOVA when the data is non linear.
-
-To further clarify the concept of basis functions and smooth functions,
-consider the simulated response for Group 1 in Figure
-@ref(fig:l-q-response)C. The simplest GAM model that can be used to
-estimate such response is that of a single smooth term for the time
-effect; i.e., a model that fits a smooth to the trend of the group
-through time. The timeline can be divided in equally spaced
-\emph{knots}, each knot being a region where a different set of basis
-functions will be used. Because there are six timepoints for this group,
-five knots can be used. The model with five knots to construct the
-smooth term means that it will have four basis functions (plus one that
-corresponds to the intercept). The choice of basis functions is set
-using default values in the package \emph{mgcv} depending on the number
-of knots. In Figure @ref(fig:basis-plot)A, the four basis functions (and
-the intercept) are shown. Each of the basis functions is composed of six
-different points (because there are six points on the timeline). To
-control the ``wiggliness'' of the fit, each of the basis functions of
-Figure @ref(fig:basis-plot)A is weighted by multiplying it by a
-coefficient according to the matrix of Figure @ref(fig:basis-plot)B. The
-parameter estimates are penalized (shrunk towards 0) where the penalty
-reduces the ``wiggliness'' of the smooth fit to prevent overfitting. A
-weak penalty estimate will result in wiggly functions whereas a strong
-penalty estimate provides evidence that a linear response is
-appropriate.
-
-To get the weighted basis functions, each basis (from Figure Figure
-@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients
-in Figure @ref(fig:basis-plot)B, thereby increasing or decreasing the
-original basis functions. Figure @ref(fig:basis-plot)C shows the
-resulting weighted basis functions. Note that the magnitude of the
-weighting for the first basis function has resulted in a decrease of its
-overall value (because the coefficient for that basis function is less
-than 1). On the other hand, the third basis function has roughly doubled
-its value. Finally, the weighted basis functions are added at each
-timepoint to produce the smooth term. The resulting smooth term for the
-effect of \emph{time} is shown in Figure @ref(fig:basis-plot)D (orange
-line), along the simulated values per group, which appear as points.
-
-(ref:basis-plot-caption) Basis functions for a single smoother for time
-with five knots. A: Basis functions for a single smoother for time for
-the simulated data of Group 1 from Figure 2. B: Matrix of basis function
-weights. Each basis function is multiplied by a coefficient which can be
-positive or negative. The coefficient determines the overall effect of
-each basis in the final smoother. C: Weighted basis functions. Each of
-the four basis functions of panel A has been weighted by the
-corresponding coefficient shown in Panel B. Note the corresponding
-increase (or decrease) in magnitude of each weighted basis function. D:
-Smoother for time and original data points. The smoother (line) is the
-result of the sum of each weighted basis function at each time point,
-with simulated values for the group shown as points.
+Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.
+
+In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible \emph{smooth functions}.
+This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
+One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference \citep{beck1998}.
+In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
+
+The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
+A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist \citep{simpson2018}.
+For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response \citep{hefley2017}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
+
+Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data \citep{wood2017, simpson2018}.
+Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness \citep{wegman1983}.
+Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
+
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
+The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
+The timeline can be divided in equally spaced \emph{knots}, each knot being a region where a different set of basis functions will be used.
+Because there are six timepoints for this group, five knots can be used.
+The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
+The choice of basis functions is set using default values in the package \emph{mgcv} depending on the number of knots.
+In Figure \ref{fig:basis-plot}A, the four basis functions (and the intercept) are shown.
+Each of the basis functions is composed of six different points (because there are six points on the timeline).
+To control the ``wiggliness'' of the fit, each of the basis functions of Figure \ref{fig:basis-plot}A is weighted by multiplying it by a coefficient according to the matrix of Figure \ref{fig:basis-plot}B.
+The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the ``wiggliness'' of the smooth fit to prevent overfitting.
+A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
+
+To get the weighted basis functions, each basis (from Figure Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
+Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions.
+Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1).
+On the other hand, the third basis function has roughly doubled its value.
+Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
+The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (orange line), along the simulated values per group, which appear as points.
 
@@
 
-\caption{(ref:basis-plot-caption)}\label{fig:basis-plot}
+\caption{Basis functions for a single smoother for time with five knots.
+A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
+B: Matrix of basis function weights.
+Each basis function is multiplied by a coefficient which can be positive or negative.
+The coefficient determines the overall effect of each basis in the final smoother.
+C: Weighted basis functions.
+Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
+Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
+D: Smoother for time and original data points.
+The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
 \end{figure}
@@
 \hypertarget{a-bayesian-interpretation-of-gams}{%
-\section{A Bayesian interpretation of
-GAMs}\label{a-bayesian-interpretation-of-gams}}
-
-Bayes' theorem states that the probability of an event can be calculated
-using prior knowledge or belief \citep{mcelreath2018}. In the case of
-non-linear data, the belief that the \emph{true} trend of the data is
-likely to be smooth rather than ``wiggly'' introduces the concept of a
-prior distribution for wiggliness (and therefore a Bayesian view) of
-GAMs \citep{wood2017}. GAMs are considered ``empirical'' Bayesian models
-because the smoothing parameters are estimated from the data (and not
-from a prior distribution as in the ``Full Bayes'' case)
-\citep{miller2019}. Moreover, the use of the restricted maximum
-likelihood (REML) to estimate the smoothing parameters gives an
-empirical estimate of the smooth model \citep{laird1982, pedersen2019}.
-Therefore, the confidence intervals calculated for the smooth terms
-using the package \emph{mgcv} are considered empirical Bayesian
-posterior credible intervals \citep{pedersen2019}, which have good
-``frequentist'' coverage (pointwise coverage or ``single point''
-coverage), and \emph{across the function} coverage \citep{wood2017}.
-This last part means that contrary to a pointwise coverage (where the
-coverage of the interval is correct for a single point) the estimated
-confidence intervals for the smooths will contain \emph{on average} the
-true function of the data 95\% of the time across the entire timeline
-(in the case of longitudinal data for which smooths are calculated),
-which allows to obtain better inference from the model. In-depth theory
-of the Bayesian interpretation of GAMs is beyond the scope of this
-paper, but can be found in \citep{miller2019, wood2017, simpson2018} and
-\citep{marra2012}. With this brief introduction to the Bayesian
-interpretation of GAMs, we henceforth refer to the confidence intervals
-for the smooths in GAMs as ``empirical Bayesian'' through the rest of
-this paper.
+\section{A Bayesian interpretation of GAMs}\label{a-bayesian-interpretation-of-gams}}
+
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief \citep{mcelreath2018}.
+In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs \citep{wood2017}.
+GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) \citep{miller2019}.
+Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model \citep{laird1982, pedersen2019}.
+Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals \citep{pedersen2019}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage \citep{wood2017}.
+This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in \citep{miller2019, wood2017, simpson2018} and \citep{marra2012}.
+With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
 
 \hypertarget{longitudinal-GAMs}{%
-\section{The analyisis of longitudinal biomedical data using
-GAMs}\label{longitudinal-GAMs}}
-
-The previous sections provided the basic framework to understand the GAM
-framework and how these models are more advantageous to analyze
-non-linear longitudinal data when compared to rm-ANOVA or LMEMs. This
-section will use simulation to present the practical implementation of
-GAMs for longitudinal biomedical data using \(\textsf{R}\) and the
-package \texttt{mgcv}. The code for the simulated data and figures, and
-a brief guide for model selection and diagnostics appear in the
-Appendix.
+\section{The analyisis of longitudinal biomedical data using GAMs}\label{longitudinal-GAMs}}
+
+The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs.
+This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \texttt{mgcv}.
+The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
 
@@
 
-The simulated data is based on the reported longitudinal changes in
-oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear
-in Figure 3C in \citep{vishwanath2009}. In the paper, diffuse
-reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes
-in both groups at the same time points (days 0, 2, 5, 7 and 10). In the
-``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is
-observed through time, while a decrease is seen in the ``Control''
-(saline) group. Following the reported trend, we simulated 10 normally
-distributed observations at each time point with a standard deviation
-(SD) of 10\% (matching the SD in the original paper). The simulated and
-real data appear in Figure @ref(fig:sim-smooth-plot)A and the inset,
-respectively.
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in \citep{vishwanath2009}.
+In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
+In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group.
+Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
+The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}A and the inset, respectively.
 
 \hypertarget{an-interaction-gam-for-longitudinal-data}{%
-\subsection{An interaction GAM for longitudinal
-data}\label{an-interaction-gam-for-longitudinal-data}}
-
-An interaction effect is typically the main interest in longitudinal
-biomedical data, as it takes into account treatment, time, and their
-combination. In a practical sense, when a GAM is implemented for
-longitudinal data, a smooth can be added to the model for the
-\emph{time} effect to account for the repeated measures over time.
-Although specific methods of how GAMs model correlation structures is a
-topic beyond the scope of this paper, it suffices to say that GAMs are
-flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found
-in \citep{hefley2017}.
-
-For the data in Figure @ref(fig:sim-smooth-plot), A the main effect of
-interest is how \(\mbox{StO}_2\) changes over time for each treatment.
-To estimate this, the model incorporates independent smooths for
-\emph{Group} and \emph{Day}, respectively. The main thing to consider is
-that model syntax accounts for the fact that one of the variables is
-numeric (\emph{Day}) and the other is a factor (\emph{Group}). Because
-the smooths are centered at 0, the factor variable needs to be specified
-as a parametric term in order to identify any differences between the
-groups. Using \(\textsf{R}\) and the package \texttt{mgcv} the model
-syntax is:
+\subsection{An interaction GAM for longitudinal data}\label{an-interaction-gam-for-longitudinal-data}}
+
+An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination.
+In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time.
+Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
+A detailed description on basis functions and correlations can be found in \citep{hefley2017}.
+
+For the data in Figure \ref{fig:sim-smooth-plot}, A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
+To estimate this, the model incorporates independent smooths for \emph{Group} and \emph{Day}, respectively.
+The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}).
+Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
+Using \(\textsf{R}\) and the package \texttt{mgcv} the model syntax is:
 
@@
 
-This syntax specifies that \texttt{m1} will store the model, and that
-the change in the simulated oxygen saturation (\texttt{StO2\_sim}) is
-modeled using independent smooths over \texttt{Day} for each
-\texttt{Group} (the parenthesis preceded by \texttt{s}) using 5 knots.
-The smooth is constructed by default using thin plate regression
-splines, but other splines can be used if desired, including Gaussian
-process smooths \citep{simpson2018}. The parametric term \texttt{Group}
-is added to quantify overall mean differences in the effect of treatment
-between groups, and the \texttt{method} chosen to estimate the smoothing
-parameters is the restricted maximum likelihood (REML) \citep{wood2017}.
-When the smooths are plotted over the raw data, it is clear that the
-model has been able to capture the trend of the change of
-\(\mbox{StO}_2\) for each group across time (Figure
-@ref(fig:sim-smooth-plot)B). Model diagnostics can be obtained using the
-\texttt{gam.check} function, and the function \texttt{appraise} from the
-package \emph{gratia} \citep{gratia}. A guide for model selection and
-diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an
-in-depth analysis can be found in \citep{wood2017} and
-\citep{harezlak2018}.
-
-One question that might arise at this point is ``what is the fit that an
-rm-ANOVA model produces for the simulated data?''. The rm-ANOVA model,
-which corresponds to Equation @ref(eq:linear-model) is presented in
-Figure @ref(fig:sim-smooth-plot)C. This is a typical case of model
-misspecification: The slopes of each group are different, which would
-lead to a \emph{p-value} indicating significance for the treatment and
-time effects, but the model is not capturing the changes that occur at
-days 2 and between days 5 and 7, whereas the GAM model is able to
-reliably estimate the trend over all timepoints (Figure
-@ref(fig:sim-smooth-plot)B) .
-
-Because GAMs do not require equally-spaced or complete observations for
-all subjects, they are advantageous to analyze longitudinal data where
-missingness exists. The rationale behind this is that GAMs are able to
-pick the trend in the data even when some observations are missing.
-However, this usually causes the resulting smooths to have wider
-confidence intervals and less ability to pick certain trends. Consider
-the simulated \(\mbox{StO}_2\) values from Figure
-@ref(fig:sim-smooth-plot)B. If 40\% of the total observations are
-randomly deleted and the same interaction GAM fitted for the complete
-dataset is used, the resulting smooths are still able to show a
-different trend for each group, but because the empirical Bayesian
-credible intervals for the smooths overlap during the first 3 days with
-fewer data points, the trend is less pronounced than in the full dataset
-(Figure @ref(fig:sim-smooth-plot)D). Although the confidence intervals
-have increased for both smooths, the model still shows different trends
-with as little as 4 observations per group at certain time points.
-
-(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation
-in tumors. A: Simulated data that follows previously reported trends
-(inset) in tumors under chemotherapy (Treatment) or saline (Control)
-treatment. Simulated data is from a normal distribution with standard
-deviation of 10\% with 10 observations per time point. Lines indicate
-mean oxygen saturation B: Smooths from the GAM model for the full
-simulated data with interaction of Group and Treatment. Lines represent
-trends for each group, shaded regions are 95\% confidence intervals. C:
-The rm-ANOVA model for the simulated data, which does not capture the
-changes in each group over time. D: Smooths for the GAM model for the
-simulated data with 40\% of its observations missing. Lines represent
-trends for each group, shaded regions are 95\% empirical Bayesian
-confidence intervals.
+This syntax specifies that \texttt{m1} will store the model, and that the change in the simulated oxygen saturation (\texttt{StO2\_sim}) is modeled using independent smooths over \texttt{Day} for each \texttt{Group} (the parenthesis preceded by \texttt{s}) using 5 knots.
+The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths \citep{simpson2018}.
+The parametric term \texttt{Group} is added to quantify overall mean differences in the effect of treatment between groups, and the \texttt{method} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) \citep{wood2017}.
+When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
+Model diagnostics can be obtained using the \texttt{gam.check} function, and the function \texttt{appraise} from the package \emph{gratia} \citep{gratia}.
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in \citep{wood2017} and \citep{harezlak2018}.
+
+One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?''.
+The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
+This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
+
+Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
+The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
+However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
+Consider the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}B.
+If 40\% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \ref{fig:sim-smooth-plot}D).
+Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.
 
@@
 
-\caption{(ref:sim-plot-caption)}\label{fig:sim-smooth-plot}
+\caption{Simulated data and smooths for oxygen saturation in tumors.
+A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
+Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
+Lines represent trends for each group, shaded regions are 95\% confidence intervals.
+C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
+D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
+Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot}
 \end{figure}
 
-(ref:pairwise-comp) Pairwise comparisons for smooth terms. A: Pairwise
-comparisons for the full dataset. B: Pairwise comparisons for the
-dataset with missing observations. Significant differences exist where
-the 95\% empirical Bayesian credible interval does not cover 0. In both
-cases the effect of treatment is significant after day 3.
-
 \begin{figure}
@@
 
-\caption{(ref:pairwise-comp)}\label{fig:plot-pairwise-comp}
+\caption{Pairwise comparisons for smooth terms.
+A: Pairwise comparisons for the full dataset.
+B: Pairwise comparisons for the dataset with missing observations.
+Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
+In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp}
 \end{figure}
@@
 \hypertarget{GAM-significance}{%
-\subsection{Determination of significance in GAMs for longitudinal
-data}\label{GAM-significance}}
-
-At the core of a biomedical longitudinal study lies the question of a
-significant difference between the effect of two or more treatments in
-different groups. Whereas in rm-ANOVA a \emph{post-hoc} analysis is
-required to answer such question by calculating some \emph{p-values}
-after multiple comparisons, GAMs can use a different approach to
-estimate significance. In essence, the idea behind the estimation of
-significance in GAMs across different treatment groups is that if the
-\emph{difference} between the empirical Bayesian confidence intervals of
-the fitted smooths for such groups is non-zero, then a significant
-difference exists at that time point(s). The absence of a \emph{p-value}
-in this case might seem odd, but the empirical Bayesian confidence
-interval interpretation can be conceptualized in the following manner:
-Different trends in each group are an indication of an effect by the
-treatment. This is what happens for the simulated data in Figure
-@ref(fig:sim-smooth-plot)A, where the chemotherapy causes
-\(\mbox{StO}_2\) to increase over time.
-
-With this expectation of different trends in each group, computing the
-difference between the trends will identify if the observed change is
-significant. The difference between groups with similar trends is likely
-to yield zero, which would indicate that the treatment is not causing a
-change in the response in one of the groups (assuming the other group is
-a Control or Reference group).
-
-Consider the calculation of pairwise differences for the smooths in
-Figure @ref(fig:sim-smooth-plot)B and Figure @ref(fig:sim-smooth-plot)D.
-Figure @ref(fig:plot-pairwise-comp) shows the comparison between each
-treatment group for the full and missing datasets. Here, the ``Control''
-group is used as the reference to which ``Treatment'' group is being
-compared. Of notice, the pairwise comparison has been set on the
-response scale (see Appendix for code details), because otherwise the
-comparison appears shifted and is not intuitively easy to relate to the
-original data.
-
-With this correction in mind, the shaded regions over the confidence
-interval (where it does not cover 0) indicate the time interval where
-each group has a higher effect than the other. Notice that the shaded
-region between days 0 and \(\approx\) 2 for the full dataset indicates
-that through that time, the ``Control'' group has higher mean
-\(\mbox{StO}_2\), but as therapy progresses the effect is reversed and
-by \(\approx\) 3 day it is the ``Treatment'' group the one that on
-average, has greater \(\mbox{StO}_2\). This would suggest that the
-effect of chemotherapy in the ``Treatment'' group becomes significant
-after day 3 for the given model. Moreover, notice that although there is
-no actual measurement at day 3, the model is capable of providing an
-estimate of when the shift in mean \(\mbox{StO}_2\) occurs.
-
-On the data with missing observations (Figure
-@ref(fig:sim-smooth-plot)D), the empirical Bayesian credible intervals
-of the smooths overlap between days 0 and 3. Consequently, the smooth
-pairwise comparison (Figure @ref(fig:plot-pairwise-comp)B) shows that
-there is no evidence of a significant difference between the groups
-during that period, but is still able to pick the change on day 3 as the
-full dataset smooth pairwise comparison.
-
-In a sense, the pairwise smooth comparison is more informative than a
-\emph{post-hoc} \emph{p-value}. For biomedical studies, the smooth
-comparison is able to provide an estimate of \emph{when} and by
-\emph{how much} a biological process becomes significant. This is
-advantageous because it can help researchers gain insight on metabolic
-changes and other biological processes that can be worth examining, and
-can help refine the experimental design of future studies in order to
-obtain measurements at time points where a significant change might be
-expected.
+\subsection{Determination of significance in GAMs for longitudinal data}\label{GAM-significance}}
+
+At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
+Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the \emph{difference} between the empirical Bayesian confidence intervals of the fitted smooths for such groups is non-zero, then a significant difference exists at that time point(s).
+The absence of a \emph{p-value} in this case might seem odd, but the empirical Bayesian confidence interval interpretation can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
+This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
+
+With this expectation of different trends in each group, computing the difference between the trends will identify if the observed change is significant.
+The difference between groups with similar trends is likely to yield zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
+
+Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}B and Figure \ref{fig:sim-smooth-plot}D.
+Figure \ref{fig:plot-pairwise-comp} shows the comparison between each treatment group for the full and missing datasets.
+Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared.
+Of notice, the pairwise comparison has been set on the response scale (see Appendix for code details), because otherwise the comparison appears shifted and is not intuitively easy to relate to the original data.
+
+With this correction in mind, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
+Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by \(\approx\) 3 day it is the ``Treatment'' group the one that on average, has greater \(\mbox{StO}_2\).
+This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the given model.
+Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in mean \(\mbox{StO}_2\) occurs.
+
+On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the empirical Bayesian credible intervals of the smooths overlap between days 0 and 3.
+Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that there is no evidence of a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.
+
+In a sense, the pairwise smooth comparison is more informative than a \emph{post-hoc} \emph{p-value}.
+For biomedical studies, the smooth comparison is able to provide an estimate of \emph{when} and by \emph{how much} a biological process becomes significant.
+This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.
 
@@
 
-Biomedical longitudinal non-linear data is particularly challenging to
-analyze due to the likelihood of missing observations and different
-correlation structures in the data, which limit the use of rm-ANOVA.
-Although LMEMs have started to replace rm-ANOVA as the choice to analyze
-biomedical data, both methods yield biased estimates when they are used
-to fit non-linear data as we have visually demonstrated in Section
-@ref(simulation). This ``model misspecification'' error, also is known
-as a ``Type III'' error \citep{dennis2019} is particularly important
-because although the \emph{p-value} is the common measure of statistical
-significance, the validity of its interpretation is determined by the
-agreement of the data and the model. Guidelines for statistical
-reporting in biomedical journals exist (the SAMPL guidelines)
-\citep{lang2015} but they have not been widely adopted and in the case
-of longitudinal data, we consider that researchers would benefit from
-reporting a visual assessment of the correspondence between the model
-fit and the data, instead of merely relying on a \(R^2\) value.
-
-In this paper we have presented GAMs as a suitable method to analyze
-non-linear longitudinal data. It is interesting to note that although
-GAMs are a well established method to analyze temporal data in different
-fields (among which are palaeoecology, geochemistry, and ecology)
-\citep{hefley2017, pedersen2019} they are not routinely used in
-biomedical research despite an early publication from Hastie and
-Tibshirani that demonstrated their use in medical research
-\citep{hastie1995}. This is possibly due to the fact that the theory
-behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but
-the purpose of Section @ref(GAM-theory) is to demonstrate that at its
-core the theory quite simple: Instead of using a linear relationship to
-model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions
-to build smooths that are capable of following non-linear trends in the
-data.
-
-However, from a practical standpoint is equally important to demonstrate
-how GAMs are computationally implemented. We have provided an example on
-how GAMs can be fitted using simulated data that follows trends reported
-in biomedical literature \citep{vishwanath2009} using \(\textsf{R}\) and
-the package \emph{mgcv}\citep{wood2017} in Section
-@ref(longitudinal-GAMs), while a basic workflow for model selection is
-in the \protect\hyperlink{workflow}{Appendix}. One of the features of
-GAMs is that their Bayesian interpretation allows to indicate
-differences between groups without the need of a \emph{p-value}, and in
-turn provide a time-based estimate of shifts in the response that can be
-directly tied to biological values as the pairwise smooth comparisons in
-Figure @ref(fig:plot-pairwise-comp) indicate. The model is therefore
-able to provide an estimate of significant change between the groups at
-time points were data was not directly measured even with missing data
-exists ( \(\approx\) day 3 in Figure @ref(fig:plot-pairwise-comp) A, B
-), which can be used by researchers as feedback on experiment design and
-to further evaluate important biological changes in future studies.
-
-We have used \(\textsf{R}\) as the software of choice for this paper
-because not only provides a fully developed environment to fit GAMs, but
-also eases simulation (which is becoming increasingly used for
-exploratory statistical analysis and power calculations) and provides
-powerful and convenient methods of visualization, which are key aspects
-that biomedical researchers might need to consider to make their work
-reproducible. In this regard, reproducibility is still an issue in
-biomedical research \citep{begley2015, weissgerber2018}, but it is
-becoming apparent that what other disciplines have experienced in this
-aspect is likely to impact sooner rather than later this field.
-Researchers need to plan on how they will make their data, code, and any
-other materials open and accessible as more journals and funding
-agencies recognize the importance and benefits of open science in
-biomedical research. We have made all the data and code used in this
-paper accessible, and we hope that this will encourage other researchers
-to do the same with future projects.
+Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
+Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}.
+This ``model misspecification'' error, also is known as a ``Type III'' error \citep{dennis2019} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) \citep{lang2015} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+
+In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) \citep{hefley2017, pedersen2019} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research \citep{hastie1995}.
+This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+
+However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature \citep{vishwanath2009} using \(\textsf{R}\) and the package \emph{mgcv}\citep{wood2017} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
+One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
+The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
+
+We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
+In this regard, reproducibility is still an issue in biomedical research \citep{begley2015, weissgerber2018}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
+We have made all the data and code used in this paper accessible, and we hope that this will encourage other researchers to do the same with future projects.
 
@@
 
-We have presented GAMs as a method to analyze longitudinal biomedical
-data. Future directions of this work will include simulation-based
-estimations of statistical power using GAMs, as well as demonstrating
-the prediction capabilities of these models using large datasets. By
-making the data and code used in this paper accessible, we hope to
-address the need of creating and sharing reproducible work in biomedical
-research.
+We have presented GAMs as a method to analyze longitudinal biomedical data.
+Future directions of this work will include simulation-based estimations of statistical power using GAMs, as well as demonstrating the prediction capabilities of these models using large datasets.
+By making the data and code used in this paper accessible, we hope to address the need of creating and sharing reproducible work in biomedical research.
 
@@
 
-This work was supported by the National Science Foundation Career Award
-(CBET 1751554, TJM) and the Arkansas Biosciences Institute.
+This work was supported by the National Science Foundation Career Award (CBET 1751554, TJM) and the Arkansas Biosciences Institute.
+
+\hypertarget{declaration-of-conflicting-interests}{%
+\section{Declaration of Conflicting Interests}\label{declaration-of-conflicting-interests}}
+
+The Authors declare that there is no conflict of interest.
 
@@
 
-\hypertarget{bibliography}{%
-\section{Bibliography}\label{bibliography}}
+\hypertarget{appendix-appendix}{%
+\appendix}
+
+\appendixpagenumbering
+
+\counterwithin{figure}{section}
 
 \hypertarget{appendix}{%
-\section{(APPENDIX)}\label{appendix}}
-
-\hypertarget{code-for-manuscript-data}{%
-\section{Code for Manuscript data}\label{code-for-manuscript-data}}
-
-This section presents the code used to generate figures, models and
-simulated data from Sections 3 and 4 from the main manuscript.
+\section{APPENDIX}\label{appendix}}
+
+This section presents the code used to generate figures, models and simulated data from the main manuscript.
 
 \hypertarget{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}{%
-\subsection{Compound symmetry and independent errors in linear and
-quadratic
-responses}\label{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}}
-
-This section simulated linear and quadratic data in the same manner as
-in Section @ref(simulation). The linear simulations using Figure
-@ref(fig:linear-cases-Appendix) show in panels A and D the simulated
-mean responses and individual data points. Panels C and G show a visual
-interpretation of ``correlation'' in the responses: In panel C, subjects
-that have a value of the random error \(\varepsilon\) either above or
-below the mean group response are more likely to have other observations
-that follow the same trajectory, thereby demonstrating correlation in
-the response. In panel G,because the errors are independent, there is no
-expectation that responses are likely to follow a similar pattern.
+\subsection{Compound symmetry and independent errors in linear and quadratic responses}\label{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}}
+
+This section simulates linear and quadratic data in the same manner as in Section \ref{simulation}.
+The linear simulations using Figure \ref{fig:linear-cases-Appendix} show in panels A and D the simulated mean responses and individual data points.
+Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
+In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
 Panels D and H show the predictions from the rm-ANOVA model.
 
-The following code produces a more comprehensive exploration of Figure
-@ref(fig:l-q-response) in the main manuscript.
+The following code produces a more comprehensive exploration of Figure \ref{fig:l-q-response} in the main manuscript.
 
@@
 
-(ref:linear-cases-appendix) Simulated linear responses from two groups
-with correlated (top row) or independent (bottom row) errors using a
-rm-ANOVA model and a LMEM. A, F:Simulated data with known mean response
-and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of
-correlated and independent errors. C,H: Simulated data with thin lines
-representing individual trajectories. D,I: Estimations from the rm-ANOVA
-model for the mean group response. E, J: Estimations from the LMEM for
-the mean group response and individual responses (thin lines). In all
-panels, thick lines are the predicted mean response per group, thin
-lines are the random effects for each subject and points represent the
-original raw data. Both rm-ANOVA and the LMEM are able to capture the
-trend of the data.
-
 \begin{figure}
-\centering
-\includegraphics{Full_document_SAGE_files/figure-latex/linear-cases-Appendix-1.pdf}
-\caption{(ref:linear-cases-appendix)}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/linear-cases-Appendix-1}
+
+}
+
+\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
+A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
+B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
+C,H: Simulated data with thin lines representing individual trajectories.
+D,I: Estimations from the rm-ANOVA model for the mean group response.
+E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
+In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
+Both rm-ANOVA and the LMEM are able to capture the trend of the data.}\label{fig:linear-cases-Appendix}
 \end{figure}
 
-For the quadratic response case, Figure
-@ref(fig:quadratic-cases-Appendix) shows the simulated responses using
-compound symmetry and independent errors.
-
-(ref:quadratic-caption-appendix) Simulated quadratic responses from two
-groups with correlated (top row) or independent (bottom row) errors
-using a rm-ANOVA model and a LMEM. A, F:Simulated data with known mean
-response and individual responses (points) showing the dispersion of the
-data. B,G: Generated errors showing the difference in the behavior of
-correlated and independent errors. C,H: Simulated data with thin lines
-representing individual trajectories. D,I: Estimations from the rm-ANOVA
-model for the mean group response. E, J: Estimations from the LMEM for
-the mean group response and individual responses (thin lines). In all
-panels, thick lines are the predicted mean response per group, thin
-lines are the random effects for each subject and points represent the
-original raw data. Both rm-ANOVA and the LMEM are not able to capture
-the changes in each group over time.
+For the quadratic response case, Figure \ref{fig:quadratic-cases-Appendix} shows the simulated responses using compound symmetry and independent errors.
 
 \begin{figure}
-\centering
-\includegraphics{Full_document_SAGE_files/figure-latex/quadratic-cases-Appendix-1.pdf}
-\caption{(ref:quadratic-caption-appendix)}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/quadratic-cases-Appendix-1}
+
+}
+
+\caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
+A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
+B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
+C,H: Simulated data with thin lines representing individual trajectories.
+D,I: Estimations from the rm-ANOVA model for the mean group response.
+E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
+In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
+Both rm-ANOVA and the LMEM are not able to capture the changes in each group over time.}\label{fig:quadratic-cases-Appendix}
 \end{figure}
@@
 
-This code produces Figure @ref(fig:basis-plot) from the main manuscript.
-Briefly, a non-linear (quadratic) response is simulated a gam model is
-fitted and the basis are extracted in order to explain how the smooth is
-constructed. The code for data simulation is used again here for the
-sake of keeping the same structure, although the data can be simulated
-in a more simple fashion.
+This code produces Figure \ref{fig:basis-plot} from the main manuscript.
+Briefly, a non-linear (quadratic) response is simulated a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
+The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
 
@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/basis-plot-appendix-1}
-
-}
-
-\caption{(ref:basis-plot-caption)}\label{fig:basis-plot-appendix}
-\end{figure}
+\begin{verbatim}
+par(mar = c(2, 2, 2, 2))
+
+b_plot
+\end{verbatim}
+
+\includegraphics{Full_document_SAGE_files/figure-latex/basis-plot-appendix-1.pdf}
 
 \hypertarget{tumor-data-simulation}{%
-\section{Longitudinal biomedical data simulation and
-GAMs}\label{tumor-data-simulation}}
-
-This section describes how to fit GAMs to longitudinal data using
-simulated data. First, data is simulated according to Section
-@ref(longitudinal-GAMs), where reported data of oxygen saturation
-(\(\mbox{StO}_2\)) in tumors under either chemotherapy or saline control
-is used as a starting point to generate individual responses in each
-group.
+\subsection{Longitudinal biomedical data simulation and GAMs}\label{tumor-data-simulation}}
+
+This section describes how to fit GAMs to longitudinal data using simulated data.
+First, data is simulated according to Section \ref{longitudinal-GAMs}, where reported data of oxygen saturation (\(\mbox{StO}_2\)) in tumors under either chemotherapy or saline control is used as a starting point to generate individual responses in each group.
 
@@
 
-This section shows a basic workflow to fit a series of increasingly
-complex GAMs to the simulated data from the previous section. Graphical
-and parameter diagnostics for goodness of fit are discussed, as well as
-model comparison via AIC (Aikake Information Criterion).
+This section shows a basic workflow to fit a series of increasingly complex GAMs to the simulated data from the previous section.
+Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
 
@@
 
-The first model fitted to the data is one that only accounts for
-different smooths by day. The model syntax specifies that
-\texttt{gam\_00} is the object that will contain all the model
-information, and that the model attempts to explain changes in
-\texttt{StO2\_sim} (simulated \(\mbox{StO}_2\)) using a smooth per
-\texttt{Day}. The model will use 5 knots (\texttt{k=5}) for the smooth.
-The smooth is constructed by default using thin plate regression
-splines. The smoothing parameter estimation method used is the
-restricted maximum likelihood (\texttt{REML}).
+The first model fitted to the data is one that only accounts for different smooths by day.
+The model syntax specifies that \texttt{gam\_00} is the object that will contain all the model information, and that the model attempts to explain changes in \texttt{StO2\_sim} (simulated \(\mbox{StO}_2\)) using a smooth per \texttt{Day}.
+The model will use 5 knots (\texttt{k=5}) for the smooth.
+The smooth is constructed by default using thin plate regression splines.
+The smoothing parameter estimation method used is the restricted maximum likelihood (\texttt{REML}).
 
@@
 
-To obtain model diagnostics, two methodologies are to be used: 1)
-graphical diagnostics, and 2) a model check. In the first case, the
-functions \texttt{appraise} and \texttt{draw} from the package
-\emph{gratia} can be used to obtain a single output with all the
-graphical diagnostics. For model check, the functions \texttt{gam.check}
-and \texttt{summary} from \emph{mgcv} provide detailed information about
-the model fit and its parameters.
-
-(ref:first-GAM-caption) Graphical diagnostics for the first GAM model.
-Left: Graphical diagnostics provided by the function \texttt{appraise}
-from the package \emph{gratia}. Right: Fitted smooth for the model,
-provided by the function \texttt{draw}.
+To obtain model diagnostics, two methodologies are to be used: 1) graphical diagnostics, and 2) a model check.
+In the first case, the functions \texttt{appraise} and \texttt{draw} from the package \emph{gratia} can be used to obtain a single output with all the graphical diagnostics.
+For model check, the functions \texttt{gam.check} and \texttt{summary} from \emph{mgcv} provide detailed information about the model fit and its parameters.
 
@@
 
-\caption{(ref:first-GAM-caption)}\label{fig:first-GAM-diag}
+\caption{Graphical diagnostics for the first GAM model.
+Left: Graphical diagnostics provided by the function \texttt{appraise} from the package \emph{gratia}.
+Right: Fitted smooth for the model, provided by the function \texttt{draw}.}\label{fig:first-GAM-diag}
 \end{figure}
 
-From the output of the function \texttt{appraise} in Figure
-@ref(fig:first-GAM-diag), the major indicators of concern about the
-model are the QQ plot of residuals and the histogram of residuals. The
-QQ plot shows that the errors are not reasonably located along the
-45\(^{\circ}\) line (which indicates normality), as there are multiple
-points that deviate from the trend, specially in the tails. The
-histogram also shows that the variation (residuals) is not following the
-assumption of a normal distribution.
-
-The \texttt{draw} function permits to plot the smooths as
-\texttt{ggplot2} objects, which eases subsequent manipulation, if
-desired. Because model \texttt{gam\_00} specifies only one smooth for
-the time covariate (Day), the plot only contains only one smooth. Note
-that the smooth shows an almost linear profile.
+From the output of the function \texttt{appraise} in Figure \ref{fig:first-GAM-diag}, the major indicators of concern about the model are the QQ plot of residuals and the histogram of residuals.
+The QQ plot shows that the errors are not reasonably located along the 45\(^{\circ}\) line (which indicates normality), as there are multiple points that deviate from the trend, specially in the tails.
+The histogram also shows that the variation (residuals) is not following the assumption of a normal distribution.
+
+The \texttt{draw} function permits to plot the smooths as \texttt{ggplot2} objects, which eases subsequent manipulation, if desired.
+Because model \texttt{gam\_00} specifies only one smooth for the time covariate (Day), the plot only contains only one smooth.
+Note that the smooth shows an almost linear profile.
 
@@
 Special attention must be paid to the `k-index' from \texttt{gam.check}.
-This parameter indicates if the basis dimension of the smooth is
-adequate, i.e., it checks that the basis used to create the smooth are
-adequate to capture the trends in the data. If the model is not
-adequately capturing the trens in the data, this is indicated by a low
-k-index value (\textless1). From the output, it can be seen that the
-k-index is 0.36, which indicates that the model is not capturing the
-variability in the data. The \texttt{edf} (effective degrees of freedom)
-is an indicator of the complexity of the smooth. Here the complexity of
-the smooth is comparable to that of a 4th degree polynomial.
-
-From the \texttt{summary} function, information about the assumed
-distribution of the errors (Gaussian in this case) and the link function
-can be obtained. The link function is `identity' as the model does not
-make any transformation on the predictors. The `significance of smooth
-terms' \emph{p-value} indicates if each smooth is adding significance to
-the model. Here, the \emph{p-value} is low but we have seen that there
-are issues with the model from the previous outputs. Finally, the
-`deviance explained' indicates how much of the data the model is able to
-capture, which in this case corresponds to \(\approx\) 17\%.
+This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
+If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless{}1).
+From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data.
+The \texttt{edf} (effective degrees of freedom) is an indicator of the complexity of the smooth.
+Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
+
+From the \texttt{summary} function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
+The link function is `identity' as the model does not make any transformation on the predictors.
+The `significance of smooth terms' \emph{p-value} indicates if each smooth is adding significance to the model.
+Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
+Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \(\approx\) 17\%.
 
@@
 
-The major flaw of \texttt{gam\_00} is that this model is not taking into
-account the fact that the data is nested in groups. The next iteration
-is a model where a different smooth of time (Day) is assigned for each
-group using \texttt{by=Group} in the model syntax.
+The major flaw of \texttt{gam\_00} is that this model is not taking into account the fact that the data is nested in groups.
+The next iteration is a model where a different smooth of time (Day) is assigned for each group using \texttt{by=Group} in the model syntax.
 
@@
 
-Diagnostics for this model indicate that the k-index is still below 1
-(0.43 from \texttt{gam.check}), and that the residuals are still not
-following a normal distribution (Figure @ref(fig:second-GAM-diag)).
-Moreover, the smooths (plotted via the \texttt{draw()} function) appear
-with a fairly linear profile, which indicates they are still not
-capturing the trends observed in the data. From \texttt{summary()}, the
-deviance explained by the model is \(\approx\) 51\%.
-
-(ref:second-GAM-caption) Graphical diagnostics for the second GAM model.
-Left: Graphical diagnostics provided by the function \texttt{appraise}
-from the package \emph{gratia}. Right: Fitted smooth for the model,
-provided by the function \texttt{draw}.
+Diagnostics for this model indicate that the k-index is still below 1 (0.43 from \texttt{gam.check}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
+Moreover, the smooths (plotted via the \texttt{draw()} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
+From \texttt{summary()}, the deviance explained by the model is \(\approx\) 51\%.
 
@@
 
-\caption{(ref:second-GAM-caption)}\label{fig:second-GAM-diag}
+\caption{Graphical diagnostics for the second GAM model.
+Left: Graphical diagnostics provided by the function \texttt{appraise} from the package \emph{gratia}.
+Right: Fitted smooth for the model, provided by the function \texttt{draw}.}\label{fig:second-GAM-diag}
 \end{figure}
@@
 
-Model \texttt{gam\_00} was built for didactic purposes to cover the
-simplest case, but it does not account for the nesting of the data by
-Group, which is apparent from the type of smooth fitted, the model
-diagnostics, and, the low variance explained by the model. On the other
-hand, \texttt{gam\_01} takes into account the nesting within each group
-and provides better variance explanation, but as indicated in Section
-@ref(longitudinal-GAMs), in order to differentiate between each group a
-parametric term needs to be added to the model for the interaction of
-\emph{Day} and \emph{Group}.
+Model \texttt{gam\_00} was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted, the model diagnostics, and, the low variance explained by the model.
+On the other hand, \texttt{gam\_01} takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \ref{longitudinal-GAMs}, in order to differentiate between each group a parametric term needs to be added to the model for the interaction of \emph{Day} and \emph{Group}.
 
@@
 
-The resulting model is \texttt{m1}, which is the model fitted in the
-main manuscript. By using \texttt{appraise()} and \texttt{draw} on this
-model (Figure @ref(fig:final-GAM-diag)) we see that the trend on the QQ
-plot has improved, the histogram of the residuals appears to be
-reasonably distributed, and the smooths are capturing the trend of the
-data within each group. From \texttt{gam.check}, the k-index is now at
-an acceptable value (\(\approx\) 1.02), and \texttt{summary} now
-indicates that the model is able to capture 89\% of the variance in the
-data.
-
-(ref:final-GAM-caption) Graphical diagnostics for the final GAM model.
-Left: Graphical diagnostics provided by the function \texttt{appraise}
-from the package \emph{gratia}. Right: Fitted smooths for the model,
-provided by the function \texttt{draw}.
+The resulting model is \texttt{m1}, which is the model fitted in the main manuscript.
+By using \texttt{appraise()} and \texttt{draw} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
+From \texttt{gam.check}, the k-index is now at an acceptable value (\(\approx\) 1.02), and \texttt{summary} now indicates that the model is able to capture 89\% of the variance in the data.
 
@@
 
-\caption{(ref:final-GAM-caption)}\label{fig:final-GAM-diag}
+\caption{Graphical diagnostics for the final GAM model.
+Left: Graphical diagnostics provided by the function \texttt{appraise} from the package \emph{gratia}.
+Right: Fitted smooths for the model, provided by the function \texttt{draw}.}\label{fig:final-GAM-diag}
 \end{figure}
@@
 \hypertarget{comparing-models-via-aic}{%
-\subsubsection{Comparing models via
-AIC}\label{comparing-models-via-aic}}
-
-One final comparison that can be made for model selection involves the
-use of the Aikake Information Criterion (AIC). This metric is used to
-estimate information loss, which we want to minimize with an appropriate
-model. Therefore, when 2 or more models are compared, the model with
-lower AIC is preferred. In R, the comparison is done using the
-\texttt{AIC} function.
+\subsubsection{Comparing models via AIC}\label{comparing-models-via-aic}}
+
+One final comparison that can be made for model selection involves the use of the Aikake Information Criterion (AIC).
+This metric is used to estimate information loss, which we want to minimize with an appropriate model.
+Therefore, when 2 or more models are compared, the model with lower AIC is preferred.
+In R, the comparison is done using the \texttt{AIC} function.
 
@@
 
-The output in this case is expected: model \texttt{m1} has a lower AIC
-(712.46) whereas the initial two models have higher AICs (900 and 858).
-The AIC should not be considered as the only estimator of model quality,
-instead to be used as complimentary information to the graphical
-diagnostics and model checks described above.
+The output in this case is expected: model \texttt{m1} has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
+The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
 
 \hypertarget{pairwise-comparisons-of-smooth-confidence-intervals}{%
-\paragraph{Pairwise comparisons of smooth confidence
-intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
-
-The estimation of significant differences between each treatment group
-can be achieved via pairwise comparisons of the smooth confidence
-intervals as described in section @ref(GAM-significance). In this case,
-the ``design matrix'' is used to estimate the pairwise comparisons (see
-main manuscript for details and associated references). Briefly, the
-``design matrix'' (also known as the ``Xp matrix'') from the selected
-model (\texttt{m1}) is used to calculate a 95\% confidence interval of
-the difference between the smooth terms for each group. This approach
-allows to estimate the time intervals where a significant difference
-exists between the groups (confidence interval above or below 0).
-\textbf{All pairwise comparisons in this paper have been centered at the
-response scale to ease interpretation }.
+\paragraph{Pairwise comparisons of smooth confidence intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
+
+The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \ref{GAM-significance}.
+In this case, the ``design matrix'' is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
+Briefly, the ``design matrix'' (also known as the ``Xp matrix'') from the selected model (\texttt{m1}) is used to calculate a 95\% confidence interval of the difference between the smooth terms for each group.
+This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
+\textbf{All pairwise comparisons in this paper have been centered at the response scale to ease interpretation }.
 
@@
 
-(ref:pairwise-comp-caption) Smooth pairwise comparisons for model
-\texttt{m1} using a 95\% confidence interval for the difference between
-smooths. The comparison is centered at the response scale. Shaded
-regions indicate time intervals where each treatment group has a
-non-zero effect.
-
 \begin{figure}
@@
 
-\caption{(ref:pairwise-comp-caption)}\label{fig:pairwise-comp-workflow-fig}
+\caption{Smooth pairwise comparisons for model \texttt{m1} using a 95\% confidence interval for the difference between smooths.
+The comparison is centered at the response scale.
+Shaded regions indicate time intervals where each treatment group has a non-zero effect.}\label{fig:pairwise-comp-workflow-fig}
 \end{figure}
 
-Of notice, a convenient wrapper for the function described above exists
-in the package \texttt{gratia}. In this package,
-\texttt{difference\_smooths} is a function that makes the comparisons
-and produces Figure @ref(fig:pairwise-comp-workflow-fig) when is used on
-a fitted model. The function syntax and an example can be found at:
+Of notice, a convenient wrapper for the function described above exists in the package \texttt{gratia}.
+In this package, \texttt{difference\_smooths} is a function that makes the comparisons and produces Figure \ref{fig:pairwise-comp-workflow-fig} when is used on a fitted model.
+The function syntax and an example can be found at:
 
@@
 
-Keep in mind that this function \textbf{does not} center the pairwise
-comparison at the response scale, so it has to be shifted in order to be
-compared to the raw data.
+Keep in mind that this function \textbf{does not} center the pairwise comparison at the response scale, so it has to be shifted in order to be compared to the raw data.
 
 \hypertarget{gam-and-linear-model-plots-and-missing-data}{%
-\section{GAM and Linear model plots and Missing
-data}\label{gam-and-linear-model-plots-and-missing-data}}
-
-This section covers the code used to generate Figure
-@ref(fig:sim-smooth-plot), where the simulated data, fit of the
-``final'' GAM (\texttt{m1}), linear model and GAM on data with missing
-observations are presented. Note that panel A in Figure
-@ref(fig:sim-smooth-plot) and the inset are generated in the code chunk
-where the data is simulated in Section @ref(tumor-data-simulation), and
-are called later to build the figure.
+\subsection{GAM and Linear model plots and Missing data}\label{gam-and-linear-model-plots-and-missing-data}}
+
+This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\texttt{m1}), linear model and GAM on data with missing observations are presented.
+Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inset are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
 
 \hypertarget{gam-and-linear-model-plots}{%
-\subsection{GAM and Linear model
-plots}\label{gam-and-linear-model-plots}}
-
-This code chunk creates panels B and D in Figure
-@ref(fig:sim-smooth-plot). Note that this code uses the final GAM from
-the previous section (\texttt{m1}), so the simulated data and the model
-should be generated before running this section.
+\subsubsection{GAM and Linear model plots}\label{gam-and-linear-model-plots}}
+
+This code chunk creates panels B and D in Figure \ref{fig:sim-smooth-plot}.
+Note that this code uses the final GAM from the previous section (\texttt{m1}), so the simulated data and the model should be generated before running this section.
 
@@
 \hypertarget{working-with-missing-data-in-gams}{%
-\subsection{Working with Missing data in
-GAMs}\label{working-with-missing-data-in-gams}}
-
-This code chunk first randomly deletes 40\% of the total observations in
-the original simulated data, and then an interaction GAM is fitted to
-the remaining data. Model diagnostics are presented, and an object that
-stores the fitted smooths is saved to be called in the final code chunk
-to build the figure.
+\subsection{Working with Missing data in GAMs}\label{working-with-missing-data-in-gams}}
+
+This code chunk first randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data.
+Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
 
@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/sim-smooth-plot-Appendix-1}
-
-}
-
-\caption{(ref:sim-plot-caption)}\label{fig:sim-smooth-plot-Appendix}
-\end{figure}
+\begin{verbatim}
+mult_plot<-f2+inset_element(
+  f1, left = 0.01, 
+  bottom = 0.5, 
+  right = 0.5, 
+  top = 1.0)+
+  f3+f4+f6+
+   plot_annotation(tag_levels='A')&
+   ylim(c(-7,75)) &
+  theme(
+     text=element_text(size=18)
+     )&
+  thm
+\end{verbatim}
+
+\begin{verbatim}
+## Scale for 'y' is already present. Adding another scale for 'y', which will
+## replace the existing scale.
+\end{verbatim}
+
+\begin{verbatim}
+## Scale for 'colour' is already present. Adding another scale for 'colour',
+## which will replace the existing scale.
+## Scale for 'colour' is already present. Adding another scale for 'colour',
+## which will replace the existing scale.
+## Scale for 'colour' is already present. Adding another scale for 'colour',
+## which will replace the existing scale.
+## Scale for 'colour' is already present. Adding another scale for 'colour',
+## which will replace the existing scale.
+## Scale for 'colour' is already present. Adding another scale for 'colour',
+## which will replace the existing scale.
+\end{verbatim}
+
+\begin{verbatim}
+mult_plot
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 1 rows containing non-finite values (stat_summary).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 1 rows containing missing values (geom_point).
+
+## Warning: Removed 1 rows containing missing values (geom_point).
+
+## Warning: Removed 1 rows containing missing values (geom_point).
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Removed 53 rows containing missing values (geom_point).
+\end{verbatim}
+
+\includegraphics{Full_document_SAGE_files/figure-latex/sim-smooth-plot-Appendix-1.pdf}
 
 \hypertarget{pairwise-comparisons-in-gams-full-and-missing-data-cases}{%
-\subsection{Pairwise comparisons in GAMs: full and missing data
-cases}\label{pairwise-comparisons-in-gams-full-and-missing-data-cases}}
-
-The next code chunk reproduces Figure @ref(fig:plot-pairwise-comp). Here
-pairwise comparisons are made for the full and missing datasets.
+\subsection{Pairwise comparisons in GAMs: full and missing data cases}\label{pairwise-comparisons-in-gams-full-and-missing-data-cases}}
+
+The next code chunk reproduces Figure \ref{fig:plot-pairwise-comp}. Here pairwise comparisons are made for the full and missing datasets.
 
@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/plot-pairwise-comp-Appendix-1}
-
-}
-
-\caption{(ref:pairwise-comp)}\label{fig:plot-pairwise-comp-Appendix}
-\end{figure}
+\begin{verbatim}
+pair_comp+plot_annotation(tag_levels='A')&
+  theme(
+     text=element_text(size=18)
+     )
+\end{verbatim}
+
+\includegraphics{Full_document_SAGE_files/figure-latex/plot-pairwise-comp-Appendix-1.pdf}
 
 \bibliographystyle{sagev}
-\bibliography{refs1.bib}
+\bibliography{refs.bib}
 

### 2021-06-22 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.Rmd
Issue with code highlighting, manually adding \begin{lstlisting}[language=R] before the chunk and \end{lstlisting} after the chunk sort of works, but automatic formatting is not possible, and the "```r" header appears. Pandoc_args option does not work, I suspect because the Pandoc LaTeX template incorporates the option, but the Sage template does not. Adding the part of the Pandoc template that I think has the listings part did not work, though.

@@
 To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research."
-keywords: Cancer biology; Tumor response; generalized additive models; simulation; R
+keywords: cancer biology; tumor response; generalized additive models; simulation; R
 classoption:
@@
 
-***
-
-```{r child='07-Appendix.Rmd'}
+```{r child= '07-Appendix.Rmd'}
 ```

### 2021-06-22 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.tex
Issue with code highlighting, manually adding \begin{lstlisting}[language=R] before the chunk and \end{lstlisting} after the chunk sort of works, but automatic formatting is not possible, and the "```r" header appears. Pandoc_args option does not work, I suspect because the Pandoc LaTeX template incorporates the option, but the Sage template does not. Adding the part of the Pandoc template that I think has the listings part did not work, though.

@@
 \usepackage{float}
-\usepackage{listings}
 \setcounter{secnumdepth}{2} %to add section numbers
@@
 
-%newcommand{\passthrough}[1]{#1}
+\usepackage{listings}
+\newcommand{\passthrough}[1]{#1}
+\lstset{defaultdialect=[5.3]Lua}
+\lstset{defaultdialect=[x86masm]Assembler}
+
+%to control that figures from R output do not overflow
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
 
@@
 %command to set parameter(s) in package listings
-\lstdefinestyle{astyle}{
+\lstset{
     % How/what to match
-    language=R,
     sensitive=true,
@@
 
-\lstset{style=astyle}
-
 \begin{document}
@@
 
-\keywords{Cancer biology; Tumor response; generalized additive models; simulation; R}
+\keywords{cancer biology; tumor response; generalized additive models; simulation; R}
 
@@
 A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
-Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless{}0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
+Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
 In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\citep{dennis2019}.
@@
 
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
 \hypertarget{appendix-appendix}{%
@@
 
-\begin{verbatim}
+\begin{lstlisting}[language=R]
+
+```r
 ##########Section for calculations###########
@@
 D1<-plot_example(example(fun_type = "quadratic", error_type = "independent")) 
-\end{verbatim}
+```
+\end{lstlisting}
 
@@
 
-\begin{verbatim}
-par(mar = c(2, 2, 2, 2))
-
-b_plot
-\end{verbatim}
-
-\includegraphics{Full_document_SAGE_files/figure-latex/basis-plot-appendix-1.pdf}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/basis-plot-appendix-1}
+
+}
+
+\caption{Basis functions for a single smoother for time with five knots.
+A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
+B: Matrix of basis function weights.
+Each basis function is multiplied by a coefficient which can be positive or negative.
+The coefficient determines the overall effect of each basis in the final smoother.
+C: Weighted basis functions.
+Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
+Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
+D: Smoother for time and original data points.
+The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot-appendix}
+\end{figure}
 
@@
 This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
-If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless{}1).
+If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless1).
 From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data.
@@
 
-\begin{verbatim}
-mult_plot<-f2+inset_element(
-  f1, left = 0.01, 
-  bottom = 0.5, 
-  right = 0.5, 
-  top = 1.0)+
-  f3+f4+f6+
-   plot_annotation(tag_levels='A')&
-   ylim(c(-7,75)) &
-  theme(
-     text=element_text(size=18)
-     )&
-  thm
-\end{verbatim}
-
-\begin{verbatim}
-## Scale for 'y' is already present. Adding another scale for 'y', which will
-## replace the existing scale.
-\end{verbatim}
-
-\begin{verbatim}
-## Scale for 'colour' is already present. Adding another scale for 'colour',
-## which will replace the existing scale.
-## Scale for 'colour' is already present. Adding another scale for 'colour',
-## which will replace the existing scale.
-## Scale for 'colour' is already present. Adding another scale for 'colour',
-## which will replace the existing scale.
-## Scale for 'colour' is already present. Adding another scale for 'colour',
-## which will replace the existing scale.
-## Scale for 'colour' is already present. Adding another scale for 'colour',
-## which will replace the existing scale.
-\end{verbatim}
-
-\begin{verbatim}
-mult_plot
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 1 rows containing non-finite values (stat_summary).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 1 rows containing missing values (geom_point).
-
-## Warning: Removed 1 rows containing missing values (geom_point).
-
-## Warning: Removed 1 rows containing missing values (geom_point).
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Removed 53 rows containing missing values (geom_point).
-\end{verbatim}
-
-\includegraphics{Full_document_SAGE_files/figure-latex/sim-smooth-plot-Appendix-1.pdf}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/sim-smooth-plot-Appendix-1}
+
+}
+
+\caption{Simulated data and smooths for oxygen saturation in tumors.
+A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
+Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
+Lines represent trends for each group, shaded regions are 95\% confidence intervals.
+C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
+D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
+Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
+\end{figure}
 
@@
 
-\begin{verbatim}
-pair_comp+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-\end{verbatim}
-
-\includegraphics{Full_document_SAGE_files/figure-latex/plot-pairwise-comp-Appendix-1.pdf}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/plot-pairwise-comp-Appendix-1}
+
+}
+
+\caption{Pairwise comparisons for smooth terms.
+A: Pairwise comparisons for the full dataset.
+B: Pairwise comparisons for the dataset with missing observations.
+Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
+In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp-Appendix}
+\end{figure}
 

### 2021-06-22 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
Issue with code highlighting, manually adding \begin{lstlisting}[language=R] before the chunk and \end{lstlisting} after the chunk sort of works, but automatic formatting is not possible, and the "```r" header appears. Pandoc_args option does not work, I suspect because the Pandoc LaTeX template incorporates the option, but the Sage template does not. Adding the part of the Pandoc template that I think has the listings part did not work, though.

@@
 \CSLLeftMargin{{[}25{]} }
-\CSLRightInline{C.
-Ugrinowitsch, G.W.
-Fellingham, M.D.
-Ricard, Limitations of ordinary least squares models in analyzing repeated measures data, Medicine and Science in Sports and Exercise.
+\CSLRightInline{Ugrinowitsch, C.
+and Fellingham, G.
+W.
+and Ricard, M.
+D., Limitations of ordinary least squares models in analyzing repeated measures data, Medicine and Science in Sports and Exercise.
 36 (2004) 2144--2148.
@@
 Geisser, S.W.
-Greenhouse, {An extension of Box's results on the use of the \(F\) distribution in multivariate analysis}, The Annals of Mathematical Statistics.
+Greenhouse, An extension of box{{}}s results on the use of the {{}}f{{}} distribution in multivariate analysis, The Annals of Mathematical Statistics.
 29 (1958) 885--891.

### 2021-06-22 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.Rmd
updated to show code highlighting

@@
   bookdown::pdf_book:
+    pandoc_args: [ "--listings" ]
     base_format: rticles::sage_article

### 2021-06-22 Manuscripts/Manuscript by chapters -for submission/02-Challenges.Rmd
Manuscript without appendix created (TeX file), PDF file of Appendix created

--- 2021-06-10 Manuscripts/Manuscript by chapters -v2/02-Challenges.Rmd
@@
 
-A more comprehensive exploration  of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal appears in Figure \@ref(fig:linear-cases-Appendix) and Figure \@ref(fig:quadratic-cases-Appendix) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+A more comprehensive exploration  of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
 We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs.

### 2021-06-22 Manuscripts/Manuscript by chapters -for submission/07-Appendix.Rmd
Issue with code highlighting, manually adding \begin{lstlisting}[language=R] before the chunk and \end{lstlisting} after the chunk sort of works, but automatic formatting is not possible, and the "```r" header appears. Pandoc_args option does not work, I suspect because the Pandoc LaTeX template incorporates the option, but the Sage template does not. Adding the part of the Pandoc template that I think has the listings part did not work, though.
Manuscript without appendix created (TeX file), PDF file of Appendix created

@@
 
-This section simulates linear and quadratic data in the same manner as in Section \@ref(simulation).
+This section simulates linear and quadratic data in the same manner as in Section \@ref(simulation) in the main manuscript.
 The linear simulations using Figure \@ref(fig:linear-cases-Appendix) show in panels A and D the simulated mean responses and individual data points.
@@
 ##################Section for plotting#################################
-#######################################################################
+
 #This function will create the plots for either a "linear" or "quadratic" response
@@
 
-```{r, first-GAM-check, echo=TRUE, include=TRUE, message=FALSE,warning=FALSE}
-#need to add figure number and caption
-gam.check(gam_00)
-
-summary(gam_00)
-
-```
-
 Special attention must be paid to the 'k-index' from `gam.check`.
@@
 If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (<1).
+
+```{r, first-GAM-check, fig.width=10, fig.height=7, out.width='75%', fig.align='center', echo=TRUE, include=TRUE, message=FALSE,warning=FALSE}
+
+gam.check(gam_00)
+
+summary(gam_00)
+
+```
+
 From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data.
@@
 
-```{r, second-GAM, echo=TRUE, include=TRUE, message=FALSE,warning=FALSE}
+```{r, second-GAM, echo=TRUE, fig.width=10, fig.height=7, out.width='75%', fig.align='center', include=TRUE, message=FALSE,warning=FALSE}
 
@@
 
-```{r,final-model-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
+```{r,final-model-Appendix,fig.width=10, fig.height=7, out.width='75%', fig.align='center', echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
 

### 2021-06-22 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE_No_Appendix.Rmd
Manuscript without appendix created (TeX file), PDF file of Appendix created

--- 2021-06-22 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.Rmd
@@
   - times
-bibliography: refs.bib
+  - doublespace
+bibliography: refs.bib #references
 bibliographystyle: sagev
@@
       in_header: preamble_SAGE.sty #LaTeX formatting options
+
 ---
@@
 
-```{r child= '07-Appendix.Rmd'}
-```
+<!-- ```{r child= '07-Appendix.Rmd'} -->
+<!-- ``` -->

### 2021-06-22 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE_No_Appendix.tex
Manuscript without appendix created (TeX file), PDF file of Appendix created

--- 2021-06-22 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.tex
@@
-\documentclass[Royal,times,sagev]{sagej}
+\documentclass[Royal,times,doublespace,sagev]{sagej}
 
@@
 \usepackage{float}
+\usepackage{listings}
 \setcounter{secnumdepth}{2} %to add section numbers
@@
 
-\usepackage{listings}
+%so the code has listings
 \newcommand{\passthrough}[1]{#1}
-\lstset{defaultdialect=[5.3]Lua}
-\lstset{defaultdialect=[x86masm]Assembler}
-
-%to control that figures from R output do not overflow
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
 
@@
 %command to set parameter(s) in package listings
-\lstset{
+\lstdefinestyle{astyle}{
     % How/what to match
+    language=R,
     sensitive=true,
@@
 
+\lstset{style=astyle}
+
 \begin{document}
@@
 
-A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal appears in Figure \ref{fig:linear-cases-Appendix} and Figure \ref{fig:quadratic-cases-Appendix} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
 We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs.
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/l-q-response-1}
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_No_Appendix_files/figure-latex/l-q-response-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/basis-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_No_Appendix_files/figure-latex/basis-plot-1}
 
@@
 The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs.
-This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \texttt{mgcv}.
+This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}.
 The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
@@
 Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
-Using \(\textsf{R}\) and the package \texttt{mgcv} the model syntax is:
-
-\texttt{m1\ \textless{}-\ gam(StO2\_sim\ \textasciitilde{}\ Group\ +\ s(Day,\ by=Group,\ k=5),\ method=\textquotesingle{}REML,\ data\ =\ dat\_sim)}
-
-This syntax specifies that \texttt{m1} will store the model, and that the change in the simulated oxygen saturation (\texttt{StO2\_sim}) is modeled using independent smooths over \texttt{Day} for each \texttt{Group} (the parenthesis preceded by \texttt{s}) using 5 knots.
+Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:
+
+\begin{lstlisting}[language=R]
+m1 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
+\end{lstlisting}
+
+This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
 The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths \citep{simpson2018}.
-The parametric term \texttt{Group} is added to quantify overall mean differences in the effect of treatment between groups, and the \texttt{method} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) \citep{wood2017}.
+The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) \citep{wood2017}.
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \texttt{gam.check} function, and the function \texttt{appraise} from the package \emph{gratia} \citep{gratia}.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} \citep{gratia}.
 A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in \citep{wood2017} and \citep{harezlak2018}.
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/sim-smooth-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_No_Appendix_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/plot-pairwise-comp-1}
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_No_Appendix_files/figure-latex/plot-pairwise-comp-1}
 
@@
 
-\hypertarget{appendix-appendix}{%
-\appendix}
-
-\appendixpagenumbering
-
-\counterwithin{figure}{section}
-
-\hypertarget{appendix}{%
-\section{APPENDIX}\label{appendix}}
-
-This section presents the code used to generate figures, models and simulated data from the main manuscript.
-
-\hypertarget{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}{%
-\subsection{Compound symmetry and independent errors in linear and quadratic responses}\label{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}}
-
-This section simulates linear and quadratic data in the same manner as in Section \ref{simulation}.
-The linear simulations using Figure \ref{fig:linear-cases-Appendix} show in panels A and D the simulated mean responses and individual data points.
-Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-The following code produces a more comprehensive exploration of Figure \ref{fig:l-q-response} in the main manuscript.
-
-\begin{lstlisting}[language=R]
-
-```r
-##########Section for calculations###########
-
-## Example with linear response
-
-#This function simulates data using a linear or quadratic mean response and each with correlated
-#or uncorrelated errors. Each group has a different slope/concavity.
-example <- function(n_time = 6, #number of time points
-                    fun_type = "linear", #type of response
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  x <- seq(1,6, length.out = n_time)
-
-  #Create mean response matrix: linear or quadratic
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # quadratic response (non-linear)
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-
-  #create an array where individual observations per each time point for each group are to be stored. Currently using 10 observations per timepoint
-  y <- array(0, dim = c(length(x), 2, 10))
-
-  #Create array to store the "errors" for each group at each timepoint. The "errors" are the 
-  #between-group variability in the response.
-  errors <- array(0, dim = c(length(x), 2, 10))
-  #create an array where 10 observations per each time point for each group are to be stored
-
-  #The following cycles create independent or correlated responses.
-To each value of mu (mean response per group) a randomly generated error (correlated or uncorrelated) is added and thus the individual response is created.
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors: variance covariance matrix
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-#labeling y and errors  
-  dimnames(y) <- list(time = x, 
-                      treatment = 1:2, 
-                      subject = 1:10)
-
-  dimnames(errors) <- list(time = x, 
-                           treatment = 1:2, 
-                           subject = 1:10)
-
-  #labeling the mean response
-  dimnames(mu) <- list(time = x, 
-                       treatment = 1:2)
-
-  #convert y, mu and errors to  dataframes with time, treatment and subject columns
-  dat <- as.data.frame.table(y, 
-                             responseName = "y")
-  dat_errors <- as.data.frame.table(errors, 
-                                    responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, 
-                                responseName = "mu")
-
-  #join the dataframes to show mean response and errors per subject
-  dat <- left_join(dat, dat_errors, 
-                   by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, 
-                   by = c("time", "treatment"))
-  #add time
-  dat$time <- as.numeric(as.character(dat$time))
-  #label subjects per group
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, 
-                                  treatment, 
-                                  sep = "-")))
-
-  ## repeated measures ANOVA
-
-  fit_anova <- lm(y ~ time + treatment + time * treatment, data = dat)
-
-#LMEM: time and treatment interaction model, compound symmetry 
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  #create a prediction frame where the model can be used for plotting purposes
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  #add model predictions to the dataframe that has the simulated data
-  dat$pred_anova <- predict(fit_anova)
-  dat$pred_lmem <- predict(fit_lme)
-
-  #return everything in a list
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_anova=fit_anova,
-    fit_lme = fit_lme
-  ))
-}
-##################Section for plotting#################################
-#######################################################################
-#This function will create the plots for either a "linear" or "quadratic" response
-
-plot_example <- function(sim_dat) {
-  ## Plot the simulated data (scatterplot)
-
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = treatment, 
-               color = treatment)
-           ) +
-    geom_point(show.legend=FALSE) +
-    labs(y='response')+
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment),
-              show.legend=FALSE) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-  #plot the simulated data with trajectories per each subject
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = subject, 
-               color = treatment)
-           ) +
-    geom_line(aes(size = "Subjects"),
-              show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment,
-                  size = "Simulated Truth"), 
-              lty = 1,show.legend = FALSE) +
-    labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                face = "bold"),
-     text=element_text(size=30))+
-    thm
-
-  #plot the errors
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = errors,
-               group = subject, 
-               color = treatment)) +
-    geom_line(show.legend=FALSE) +
-     labs(y='errors')+
-     theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-   #plot the model predictions for rm-ANOVA
-  p4 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point(show.legend=FALSE)+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_anova), 
-                  group = subject, size = "Subjects"),show.legend = FALSE) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_anova, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population"),
-              show.legend=FALSE) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-   #plot the LMEM predictions
-  p5 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point()+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), 
-                  group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_lme, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population")) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-  return((p1+p3+p2+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-
-#Store each plot in a separate object
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent")) 
-```
-\end{lstlisting}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/linear-cases-Appendix-1}
-
-}
-
-\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are able to capture the trend of the data.}\label{fig:linear-cases-Appendix}
-\end{figure}
-
-For the quadratic response case, Figure \ref{fig:quadratic-cases-Appendix} shows the simulated responses using compound symmetry and independent errors.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/quadratic-cases-Appendix-1}
-
-}
-
-\caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are not able to capture the changes in each group over time.}\label{fig:quadratic-cases-Appendix}
-\end{figure}
-
-\hypertarget{basis-functions-and-gams}{%
-\subsection{Basis functions and GAMs}\label{basis-functions-and-gams}}
-
-This code produces Figure \ref{fig:basis-plot} from the main manuscript.
-Briefly, a non-linear (quadratic) response is simulated a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
-The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
-
-\begin{verbatim}
-#generate the response: the same initial procedure from the previous section to simulate
-#the response
-set.seed(1)
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-
- #label each table
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
-
- #Convert to dataframes with subject, time and group columns
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
-
- #label subject per group
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
- #extract  "Group 1" to fit the GAM
-  dat<-subset(dat,treatment==1)
- #keep just the response and timepoint columns
-   dat<-dat[,c('y','time')]
-
-   #GAM model of time, 5 knots
-gm<-gam(y~s(time,k=5),data=dat)
-
-#model_matrix (also known as) 'design matrix'
-#will contain the smooths used to create  model 'gm'
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be 
-#multiplied by the coefficients
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final smooth by aggregating the basis functions
-
-smooth<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(smooth=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,
-            aes(x=time,
-                y=value_orig,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(size=sz,
-            show.legend=FALSE)+
-  geom_point(size=sz+1,
-             show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,
-             labeller = as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#penalized basis
-p12<-ggplot(basis_plot,
-            aes(x=time,
-                y=mod_val,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(show.legend = FALSE,
-            size=sz)+
-  geom_point(show.legend = FALSE,
-             size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,
-             labeller=as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#heatmap of the coefficients
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,
-            aes(x=Basis,
-                y=Basis))+
-  geom_tile(aes(fill = cof), 
-            colour = "black") +
-    scale_fill_gradient(low = "white",
-                        high = "#B50A2AFF")+ #color picked from KikiMedium
-  labs(x='Basis',
-       y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),
-            size=7,
-            show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,
-            aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  labs(y='Simulated \n response')+
-  geom_line(data=smooth,
-            aes(x=time,
-                y=smooth),
-            color="#6C581DFF",
-            size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-\end{verbatim}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/basis-plot-appendix-1}
-
-}
-
-\caption{Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-B: Matrix of basis function weights.
-Each basis function is multiplied by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
-Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot-appendix}
-\end{figure}
-
-\hypertarget{tumor-data-simulation}{%
-\subsection{Longitudinal biomedical data simulation and GAMs}\label{tumor-data-simulation}}
-
-This section describes how to fit GAMs to longitudinal data using simulated data.
-First, data is simulated according to Section \ref{longitudinal-GAMs}, where reported data of oxygen saturation (\(\mbox{StO}_2\)) in tumors under either chemotherapy or saline control is used as a starting point to generate individual responses in each group.
-
-\begin{verbatim}
-set.seed(1)
-#Dataframe that contains the original reported trends
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
-
-## plot the mean response
-f1<-ggplot(dat, 
-           aes(x = Day, 
-               y = StO2, 
-               color = Group)) +
-    geom_line(size=1,
-              show.legend = FALSE)+
-    geom_point(show.legend = FALSE,
-               size=1.5,
-               alpha=0.5)+
-  labs(y=expression(paste(StO[2],
-                          ' (real)')))+
-  theme_classic()+
-  thm+
-    scale_x_continuous(breaks=c(0,5,10))+
-    scale_y_continuous(breaks=c(0,40))+
-  plot_layout(tag_level = 'new')+
-  theme(
-    plot.background = element_rect(fill = "transparent", 
-                                   color = NA),
-    axis.text=element_text(size=14)
-  )
-
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-n <- 10 #number of observations
-sd <- 10 #approximate sd from paper
-df <- 6
-dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, 
-           aes(x = Day, 
-               y = StO2_sim, 
-               color = Group)) +
-    geom_point(show.legend=FALSE,
-               size=1.5,
-               alpha=0.5)+
-    stat_summary(aes(y = StO2_sim,
-                     group=Group), 
-                 fun=mean, geom="line",
-                 size=1,
-                 show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],
-                         '(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-  thm+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
-\end{verbatim}
-
-\hypertarget{workflow}{%
-\subsection{A basic Workflow for GAMs}\label{workflow}}
-
-This section shows a basic workflow to fit a series of increasingly complex GAMs to the simulated data from the previous section.
-Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
-
-\hypertarget{first-model}{%
-\subsubsection{First model}\label{first-model}}
-
-The first model fitted to the data is one that only accounts for different smooths by day.
-The model syntax specifies that \texttt{gam\_00} is the object that will contain all the model information, and that the model attempts to explain changes in \texttt{StO2\_sim} (simulated \(\mbox{StO}_2\)) using a smooth per \texttt{Day}.
-The model will use 5 knots (\texttt{k=5}) for the smooth.
-The smooth is constructed by default using thin plate regression splines.
-The smoothing parameter estimation method used is the restricted maximum likelihood (\texttt{REML}).
-
-\begin{verbatim}
-gam_00<-gam(StO2_sim ~ s(Day, k = 5),
-            method='REML',
-            data  = dat_sim)
-\end{verbatim}
-
-To obtain model diagnostics, two methodologies are to be used: 1) graphical diagnostics, and 2) a model check.
-In the first case, the functions \texttt{appraise} and \texttt{draw} from the package \emph{gratia} can be used to obtain a single output with all the graphical diagnostics.
-For model check, the functions \texttt{gam.check} and \texttt{summary} from \emph{mgcv} provide detailed information about the model fit and its parameters.
-
-\hypertarget{graphical-diagnostics}{%
-\paragraph{Graphical diagnostics}\label{graphical-diagnostics}}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/first-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the first GAM model.
-Left: Graphical diagnostics provided by the function \texttt{appraise} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \texttt{draw}.}\label{fig:first-GAM-diag}
-\end{figure}
-
-From the output of the function \texttt{appraise} in Figure \ref{fig:first-GAM-diag}, the major indicators of concern about the model are the QQ plot of residuals and the histogram of residuals.
-The QQ plot shows that the errors are not reasonably located along the 45\(^{\circ}\) line (which indicates normality), as there are multiple points that deviate from the trend, specially in the tails.
-The histogram also shows that the variation (residuals) is not following the assumption of a normal distribution.
-
-The \texttt{draw} function permits to plot the smooths as \texttt{ggplot2} objects, which eases subsequent manipulation, if desired.
-Because model \texttt{gam\_00} specifies only one smooth for the time covariate (Day), the plot only contains only one smooth.
-Note that the smooth shows an almost linear profile.
-
-\hypertarget{model-check}{%
-\paragraph{Model check}\label{model-check}}
-
-\begin{verbatim}
-#need to add figure number and caption
-gam.check(gam_00)
-\end{verbatim}
-
-\includegraphics{Full_document_SAGE_files/figure-latex/first-GAM-check-1.pdf}
-
-\begin{verbatim}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 5 iterations.
-## Gradient range [-0.0003727881,-6.621452e-07]
-## (score 444.0118 & scale 450.6638).
-## Hessian positive definite, eigenvalue range [0.3881695,49.00676].
-## Model rank =  5 / 5 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##          k'  edf k-index p-value    
-## s(Day) 4.00 2.11    0.36  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{verbatim}
-
-\begin{verbatim}
-summary(gam_00)
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ s(Day, k = 5)
-## 
-## Parametric coefficients:
-##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   22.967      2.123   10.82   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##          edf Ref.df     F  p-value    
-## s(Day) 2.114  2.565 7.633 0.000517 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.153   Deviance explained = 17.2%
-## -REML = 444.01  Scale est. = 450.66    n = 100
-\end{verbatim}
-
-Special attention must be paid to the `k-index' from \texttt{gam.check}.
-This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
-If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless1).
-From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data.
-The \texttt{edf} (effective degrees of freedom) is an indicator of the complexity of the smooth.
-Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
-
-From the \texttt{summary} function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
-The link function is `identity' as the model does not make any transformation on the predictors.
-The `significance of smooth terms' \emph{p-value} indicates if each smooth is adding significance to the model.
-Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \(\approx\) 17\%.
-
-\hypertarget{second-model}{%
-\subsubsection{Second model}\label{second-model}}
-
-The major flaw of \texttt{gam\_00} is that this model is not taking into account the fact that the data is nested in groups.
-The next iteration is a model where a different smooth of time (Day) is assigned for each group using \texttt{by=Group} in the model syntax.
-
-\begin{verbatim}
-gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5),
-            method='REML',
-            data  = dat_sim)
-
-gam.check(gam_01)
-\end{verbatim}
-
-\includegraphics{Full_document_SAGE_files/figure-latex/second-GAM-1.pdf}
-
-\begin{verbatim}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 7 iterations.
-## Gradient range [-5.51754e-05,2.671715e-06]
-## (score 423.3916 & scale 280.8777).
-## Hessian positive definite, eigenvalue range [0.3162258,48.5557].
-## Model rank =  9 / 9 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                         k'  edf k-index p-value    
-## s(Day):GroupControl   4.00 3.39    0.43  <2e-16 ***
-## s(Day):GroupTreatment 4.00 3.23    0.43  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{verbatim}
-
-\begin{verbatim}
-summary(gam_01)
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ s(Day, by = Group, k = 5)
-## 
-## Parametric coefficients:
-##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   22.967      1.676    13.7   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                         edf Ref.df      F p-value    
-## s(Day):GroupControl   3.392  3.794  3.817  0.0304 *  
-## s(Day):GroupTreatment 3.229  3.682 21.174  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.472   Deviance explained = 50.8%
-## -REML = 423.39  Scale est. = 280.88    n = 100
-\end{verbatim}
-
-Diagnostics for this model indicate that the k-index is still below 1 (0.43 from \texttt{gam.check}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
-Moreover, the smooths (plotted via the \texttt{draw()} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-From \texttt{summary()}, the deviance explained by the model is \(\approx\) 51\%.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/second-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the second GAM model.
-Left: Graphical diagnostics provided by the function \texttt{appraise} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \texttt{draw}.}\label{fig:second-GAM-diag}
-\end{figure}
-
-\hypertarget{third-model}{%
-\subsubsection{Third model}\label{third-model}}
-
-Model \texttt{gam\_00} was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted, the model diagnostics, and, the low variance explained by the model.
-On the other hand, \texttt{gam\_01} takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \ref{longitudinal-GAMs}, in order to differentiate between each group a parametric term needs to be added to the model for the interaction of \emph{Day} and \emph{Group}.
-
-\begin{verbatim}
-#GAM for StO2
-
-m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
-            method='REML',
-            data  = dat_sim)
-
-gam.check(m1)
-\end{verbatim}
-
-\includegraphics{Full_document_SAGE_files/figure-latex/final-model-Appendix-1.pdf}
-
-\begin{verbatim}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 10 iterations.
-## Gradient range [-8.164307e-08,1.500338e-08]
-## (score 355.8554 & scale 64.53344).
-## Hessian positive definite, eigenvalue range [1.174841,48.08834].
-## Model rank =  10 / 10 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                         k'  edf k-index p-value
-## s(Day):GroupControl   4.00 3.87    1.02    0.59
-## s(Day):GroupTreatment 4.00 3.88    1.02    0.54
-\end{verbatim}
-
-\begin{verbatim}
-summary(m1)
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ Group + s(Day, by = Group, k = 5)
-## 
-## Parametric coefficients:
-##                Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)       9.084      1.136   7.996 4.09e-12 ***
-## GroupTreatment   27.766      1.607  17.282  < 2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                         edf Ref.df     F p-value    
-## s(Day):GroupControl   3.873  3.990 17.57  <2e-16 ***
-## s(Day):GroupTreatment 3.879  3.991 89.33  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.879   Deviance explained = 88.9%
-## -REML = 355.86  Scale est. = 64.533    n = 100
-\end{verbatim}
-
-The resulting model is \texttt{m1}, which is the model fitted in the main manuscript.
-By using \texttt{appraise()} and \texttt{draw} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From \texttt{gam.check}, the k-index is now at an acceptable value (\(\approx\) 1.02), and \texttt{summary} now indicates that the model is able to capture 89\% of the variance in the data.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/final-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the final GAM model.
-Left: Graphical diagnostics provided by the function \texttt{appraise} from the package \emph{gratia}.
-Right: Fitted smooths for the model, provided by the function \texttt{draw}.}\label{fig:final-GAM-diag}
-\end{figure}
-
-\hypertarget{comparing-models-via-aic}{%
-\subsubsection{Comparing models via AIC}\label{comparing-models-via-aic}}
-
-One final comparison that can be made for model selection involves the use of the Aikake Information Criterion (AIC).
-This metric is used to estimate information loss, which we want to minimize with an appropriate model.
-Therefore, when 2 or more models are compared, the model with lower AIC is preferred.
-In R, the comparison is done using the \texttt{AIC} function.
-
-\begin{verbatim}
-AIC(gam_00,gam_01,m1)
-\end{verbatim}
-
-\begin{verbatim}
-##               df      AIC
-## gam_00  4.564893 900.8257
-## gam_01  9.476137 858.6051
-## m1     10.980983 712.2067
-\end{verbatim}
-
-The output in this case is expected: model \texttt{m1} has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
-The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
-
-\hypertarget{pairwise-comparisons-of-smooth-confidence-intervals}{%
-\paragraph{Pairwise comparisons of smooth confidence intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
-
-The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \ref{GAM-significance}.
-In this case, the ``design matrix'' is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
-Briefly, the ``design matrix'' (also known as the ``Xp matrix'') from the selected model (\texttt{m1}) is used to calculate a 95\% confidence interval of the difference between the smooth terms for each group.
-This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
-\textbf{All pairwise comparisons in this paper have been centered at the response scale to ease interpretation }.
-
-\begin{verbatim}
-##Pairwise comparisons
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-##matrix that contains the basis functions evaluated at the points in pdat
-    xp <- predict(m1, newdata = pdat, type = 'lpmatrix')
-
-#Find columns in xp where the name contains "Control"
-    c1 <- grepl('Control', colnames(xp))
-
-#Find columns in xp where the name contains 'Treatment'
-    c2 <- grepl('Treatment', colnames(xp))
-
-#Find rows in pdat that correspond to either 'Control' or 'Treatment'
-    r1 <- with(pdat, Group == 'Control')
-    r2 <- with(pdat, Group == 'Treatment')
-
-# In xp: find the rows that correspond to Control or Treatment, those that do not match will be
-    #set to zero. Then, substract the values from the rows corresponding to 'Control' from those that correspond
-    #to 'Treatment'
-    X <- xp[r1, ] - xp[r2, ]
-
-    ## remove columns that do not contain name 'Control' or 'Treatment'
-    X[, ! (c1 | c2)] <- 0
-    ## zero out the parametric cols, those that do not contain in the characters 's('
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
-
-    #Multiply matrix by model coefficients. X has (p,n) (rows, columns) and the coefficient matrix has
-    #dimensions (n,1). The resulting matrix has dimensions (p,1)
-    dif <- X %*% coef(m1)
-
-    #comp<-test %*% coef(gam1)[3:10]
-
-#Calculate standard error for the computed differences using the variance-covariance matrix
-    #of the model
-    se <- sqrt(rowSums((X %*% vcov(m1, unconditional = FALSE)) * X))
-    crit <- qt(0.05/2, df.residual(m1), lower.tail = FALSE)
-    #upper  limits
-    upr <- dif + (crit * se)
-    #lower limits
-    lwr <- dif - (crit * se)
-    #put all components in a dataframe for plotting
-    comp1<-data.frame(pair = paste('Control', 'Treatment', sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-
-#add time point sequence
-comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1))
-
-#use function from the pairwise comparison plot in the manuscript to get the shaded regions
-
-    my_list<-pairwise_limits(comp_StO2)
-  rib_col<-'#EDD03AFF' #color for the ribbon
-#plot the difference
-c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
-  #shaded region
-  annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill='#30123BFF',
-                alpha = 0.5,
-                ) +
-  annotate("text",
-             x=1.5,
-             y=-10,
-             label="Control",size=10
-           )+
-  #shaded region  
-  annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
-             alpha = 0.5
-           ) +
-  annotate("text",
-             x=6,
-             y=-10,
-             label="Treatment",
-             size=10
-           )+
-  #ribbon for difference confidence interval  
-  geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(color='black',size=1) +
-    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-\end{verbatim}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/pairwise-comp-workflow-fig-1}
-
-}
-
-\caption{Smooth pairwise comparisons for model \texttt{m1} using a 95\% confidence interval for the difference between smooths.
-The comparison is centered at the response scale.
-Shaded regions indicate time intervals where each treatment group has a non-zero effect.}\label{fig:pairwise-comp-workflow-fig}
-\end{figure}
-
-Of notice, a convenient wrapper for the function described above exists in the package \texttt{gratia}.
-In this package, \texttt{difference\_smooths} is a function that makes the comparisons and produces Figure \ref{fig:pairwise-comp-workflow-fig} when is used on a fitted model.
-The function syntax and an example can be found at:
-
-\url{https://cran.r-project.org/web/packages/gratia/gratia.pdf}
-
-Keep in mind that this function \textbf{does not} center the pairwise comparison at the response scale, so it has to be shifted in order to be compared to the raw data.
-
-\hypertarget{gam-and-linear-model-plots-and-missing-data}{%
-\subsection{GAM and Linear model plots and Missing data}\label{gam-and-linear-model-plots-and-missing-data}}
-
-This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\texttt{m1}), linear model and GAM on data with missing observations are presented.
-Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inset are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
-
-\hypertarget{gam-and-linear-model-plots}{%
-\subsubsection{GAM and Linear model plots}\label{gam-and-linear-model-plots}}
-
-This code chunk creates panels B and D in Figure \ref{fig:sim-smooth-plot}.
-Note that this code uses the final GAM from the previous section (\texttt{m1}), so the simulated data and the model should be generated before running this section.
-
-\begin{verbatim}
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-#creates a dataframe using the length of the covariates for the GAM
-gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
-                         subject=factor(rep(1:10)))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#adds the predictions to the grid and creates a confidence interval for GAM
-gam_predict<-gam_predict%>%
-    mutate(fit = predict(m1,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(m1, gam_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=gam_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'complete')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-
-#posthoc comparisons for the linear model
-#library(multcomp)
-
-#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
-\end{verbatim}
-
-\hypertarget{working-with-missing-data-in-gams}{%
-\subsection{Working with Missing data in GAMs}\label{working-with-missing-data-in-gams}}
-
-This code chunk first randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data.
-Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
-
-\begin{verbatim}
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
-
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'missing')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-\end{verbatim}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/sim-smooth-plot-Appendix-1}
-
-}
-
-\caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
-\end{figure}
-
-\hypertarget{pairwise-comparisons-in-gams-full-and-missing-data-cases}{%
-\subsection{Pairwise comparisons in GAMs: full and missing data cases}\label{pairwise-comparisons-in-gams-full-and-missing-data-cases}}
-
-The next code chunk reproduces Figure \ref{fig:plot-pairwise-comp}. Here pairwise comparisons are made for the full and missing datasets.
-
-\begin{verbatim}
-##Pairwise comparisons
-
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-#this function takes the model, grid and groups to be compared using the lpmatrix
-#originally developed by G. Simpson:
-#https://fromthebottomoftheheap.net/2017/10/10/difference-splines-i/
-
-smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    #Find columns in xp where the name contains "Control" and "Treatment"
-    col1 <- grepl(g1, colnames(xp))
-    col2 <- grepl(g2, colnames(xp))
-    #Find rows in xp that correspond to each treatment
-    row1 <- with(newdata, Group == g1)
-    row2 <- with(newdata, Group == g2)
-    ## difference rows of xp for data from comparison
-    X <- xp[row1, ] - xp[row2, ]
-    ## zero out cols of X related to splines for other lochs
-    X[, ! (col1 | col2)] <- 0
-
-    ## zero out the parametric cols
-    #This line has been commented to keep the comparison at the response level,
-    #otherwise it gives the marginal change between smooths
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    #get standard error, critical value and boundaries
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(g1, g2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-}
-
-#use the function to calculate the difference in smooths
-comp1<-smooth_diff(m1,pdat,'Control','Treatment')
-
-#Create a dataframe with time, comparisons and labels for regions where difference exists
-comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
-
-pairwise_limits<-function(dataframe){
-    #extract values where the lower limit of the ribbon is greater than zero
-    #this is the region where the control group effect is greater
-    v1<-dataframe%>%
-        filter(lower>0)%>%
-        select(Day)
-    #get day  initial value
-    init1=v1$Day[[1]]
-    #get day final value
-    final1=v1$Day[[nrow(v1)]]
-
-    #extract values where the value of the upper limit of the ribbon is lower than zero
-    #this corresponds to the region where the treatment group effect is greater
-    v2<-comp_StO2_full%>%
-        filter(upper<0)%>%
-        select(Day)
-
-    init2=v2$Day[[1]]
-    final2=v2$Day[[nrow(v2)]]
-    #store values
-    my_list<-list(init1=init1,
-                  final1=final1,
-                  init2=init2,
-                  final2=final2)
-return(my_list)
-}
-
-my_list<-pairwise_limits(comp_StO2_full)
-rib_col<-'#EDD03AFF'
-
-c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
-    annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill='#30123BFF',
-                alpha = 0.5,
-                ) +
-  annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=8,
-           angle=90
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=8,
-           angle=90
-           )+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-###for missing data
-comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
-comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp2))
-
-missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    labs(x = 'Days', 
-         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
-                              )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme_classic()+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
-my_list<-pairwise_limits(comp_StO2_missing)
-
-c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    annotate("rect",
-             xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-             fill='#30123BFF',
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=8
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=8)+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-pair_comp<-c1+c2
-\end{verbatim}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SAGE_files/figure-latex/plot-pairwise-comp-Appendix-1}
-
-}
-
-\caption{Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-B: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
-In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp-Appendix}
-\end{figure}
-
 \bibliographystyle{sagev}

### 2021-06-22 Manuscripts/Manuscript by chapters -for submission/SAGE_Appendix.Rmd
Manuscript without appendix created (TeX file), PDF file of Appendix created

---
title: |
    | **Using generalized additive models to analyze biomedical non-linear longitudinal data**
    | _Supplementary Material: Appendix_
subtitle: Mundo et. al.
header-includes:
    \usepackage{placeins}
output:  
  bookdown::pdf_document2:
    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
    pandoc_args: --listings #calls the listings package to fit code within the page margins
    keep_tex: yes #keep LaTeX file for submission
    fig_caption: yes #allows captions in figures
    toc: false #do not include table of contents
    extra_dependencies:
      subfig: null #allows for subfigures
      breqn: null #line breaks for long equations
      caption: ["font={small}"] #size of the figure captions
      float: null #allows for control of placement of figures
    includes:
      in_header: preamble_Appendix.sty #additional LaTeX formatting
  bookdown::word_document2:
    fig_caption: yes #figure caption
    keep_md: yes
  bookdown::html_document2:
    css: "style.css" #style for the HTML document
#csl: elsevier-with-titles.csl #style for references
#bibliography: refs.bib #references
#link-citations: yes #adds links to the citations
'': default
---

```{r libraries, message = FALSE,include=FALSE}
#the packages needed to run the code

library(patchwork)
library(tidyverse)
library(mvnfast)
library(nlme)
library(mgcv)
library(gratia)
library(viridis)
set.seed(2021) #set seed for reproducibility

knitr::opts_chunk$set(fig.pos = 'H', out.extra = "",error=TRUE)
#global themes for plotting
thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
```

\newpage
```{r child='01-Background.Rmd'}
```

<!-- \FloatBarrier -->

```{r child='02-Challenges.Rmd'}
```

<!-- \FloatBarrier -->

```{r child='03-GAM Theory.Rmd'}
```

<!-- \FloatBarrier -->

```{r child='04-Longitudinal analysis with GAMs.Rmd'}
```

<!-- \FloatBarrier -->

```{r child='05-Discussion.Rmd'}
```

<!-- \FloatBarrier -->

```{r child='06-Conclusion.Rmd'}
```

<!-- \FloatBarrier -->

```{r child= '07-Appendix.Rmd'}
```

### 2021-06-22 Manuscripts/Manuscript by chapters -for submission/SAGE_Appendix.tex
Manuscript without appendix created (TeX file), PDF file of Appendix created

--- 2021-06-10 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
@@
 \hypersetup{
-  pdftitle={Using generalized additive models to analyze biomedical non-linear longitudinal data},
   hidelinks,
@@
 \setcounter{secnumdepth}{5}
-\usepackage{lineno}
-\usepackage{authblk}
-\usepackage{hyperref}
-\usepackage{graphicx}
-
-%set a box to put the ORCID logo
-\newbox{\myorcidaffilbox}
-\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}
-
-%add hyperlink to the box
-\newcommand{\orcidaffila}[1]{%
-  \href{https://orcid.org/0000-0002-6014-4538}{\usebox{\myorcidaffilbox}}}
-
-\newcommand{\orcidaffilb}[1]{%
-  \href{https://orcid.org/0000-0002-6135-8191}{\usebox{\myorcidaffilbox}}}
-
-%command for the package lineno
-\linenumbers
-
-%authors
-\author[1]{Ariel I. Mundo \orcidaffila{}}
-\author[2]{John R. Tipton \orcidaffilb{}}
-\author[1]{Timothy J. Muldoon \thanks{Corresponding author, tmuldoon@uark.edu}}
-\affil[1]{\footnotesize Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
-\affil[2]{\footnotesize Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
-
-%theme colors for the code chunks (originally from latex-solarized on GitHub)
-%https://github.com/jez/latex-solarized
+\usepackage{float}
+\usepackage{listings}
+\setcounter{secnumdepth}{2} %to add section numbers
+\usepackage{chngcntr} %package to enable figure numbers by sections
+%\counterwithin{figure}{section} %command for chngcntr in order to provide numbers per section that match with the letter for the Appendix
+
+%to have page numbers with letters in the Appendix:
+%solution from https://tex.stackexchange.com/questions/59572/custom-page-numbering-for-appendix
+\newcommand{\appendixpagenumbering}{
+  \break
+  \pagenumbering{arabic}
+  \renewcommand{\thepage}{\thesection-\arabic{page}}
+}
+
 \usepackage{xcolor}
@@
 
-%\lstset{
-%  breaklines=true,
-%  stringstyle=\ttfamily,
-%  backgroundcolor=\color{gray}
-%}
+\lstset{style=astyle}
+
 \usepackage{placeins}
@@
 \fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
+
+\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}\\
+\emph{Supplementary Material: Appendix}}
 \usepackage{etoolbox}
@@
 \makeatother
-\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
+\subtitle{Mundo et. al.}
+\author{}
 \date{\vspace{-2.5em}}
@@
 
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
-
-In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs).
-Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
-
-In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
-Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
-This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 6 repeated measures for each group) to demonstrate how these models can be computationally implemented.
-We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40\% of the observations missing), allowing reliable inference from the data.
-To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
+\newpage
 
@@
 Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response {[}\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}{]}, antibody expression {[}\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}{]}, and cell metabolism {[}\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}\protect\hyperlink{ref-greening2018}{9}{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}\protect\hyperlink{ref-sio2016}{10}{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}\protect\hyperlink{ref-kamstra2015}{11}{]}.
+Tumor response {[}@roblyer2011;@tank2020;@pavlov2018;@demidov2018{]}, antibody expression {[}@ritter2001; @roth2017{]}, and cell metabolism {[}@jones2018; @skala2010{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
+Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}@greening2018{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}@sio2016{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}@kamstra2015{]}.
 
 Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}\protect\hyperlink{ref-wagenmakers2008}{12}{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}{]}.
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
+The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}@wagenmakers2008{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
+The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}@gueorguieva2004; @schober2018{]}.
+
+The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}@pinheiro2006{]}.
 This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
 In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}@roblyer2011; @skala2010; @vishwanath2009{]}.
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
@@
 Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}.
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}\protect\hyperlink{ref-wang2019}{18}{]}.
-
-Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}{]}, consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}\protect\hyperlink{ref-abdi2010}{21}{]}), but they in turn reduce statistical power (1-\(\beta\)){[}\protect\hyperlink{ref-nakagawa2004}{22}{]}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) {[}\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}{]}.
+In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}@dennis2019{]}.
+For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}@wang2019{]}.
+
+Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}@liu2010; @halsey2015{]}, consequently biasing the conclusions of the study.
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}@abdi2010{]}), but they in turn reduce statistical power (1-\(\beta\)){[}@nakagawa2004{]}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) {[}@gelman2012; @albers2019{]}.
 Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
 
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and group sizes {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}\protect\hyperlink{ref-charan2013}{30}{]} which might cause the corrections for lack of compound symmetry to be ineffective.
+On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}@ugrinowitsch2004{]}.
+Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}@huynh1976; @greenhouse1959{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}@haverkamp2017{]}, and group sizes {[}@keselman2001{]}.
+In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}@charan2013{]} which might cause the corrections for lack of compound symmetry to be ineffective.
 
@@
 In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}\protect\hyperlink{ref-gueorguieva2004}{13}{]}.
+When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}@gueorguieva2004{]}.
 This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
@@
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}@skala2010;@vishwanath2009{]}.
 Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}.
-And even more importantly, LMEMs also assume a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}, making them unsuitable to analyze non-linear data.
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}@pinheiro2006{]}.
+However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}@gueorguieva2004;@barr2013{]}.
+And even more importantly, LMEMs also assume a linear relationship between the response and time {[}@pinheiro2006{]}, making them unsuitable to analyze non-linear data.
 
@@
 Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]} .
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]} and 3) can easily accommodate missing observations.
+Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}@rose2012; @pedersen2019; @simpson2018{]}, health-environment interactions {[}@yang2012{]} and the dynamics of government in political science {[}@beck1998{]}.
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}@wood2017{]} and 3) can easily accommodate missing observations.
 Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
@@
 The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
+In particular, \(\textsf{R}\) {[}@r{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}@wood2016; @wood2017{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}@haverkamp2017{]}.
 Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
@@
 Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies {[}@vishwanath2009{]}.
 The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
@@
 In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models to improve the standards for reproducibility in biomedical research.
-
-\FloatBarrier
 
@@
 
-A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish {[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}.
+A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish {[}@pinheiro2006;@west2014{]}.
 A LMEM with interaction between time and treatment for a longitudinal study can be written as:
@@
 The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
+In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}@pinheiro2006{]}.
 However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
@@
 In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{41}{]} (For an in-depth analysis of the covariance matrix see {[}\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}{]}).
+The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}@wolfinger1996{]} (For an in-depth analysis of the covariance matrix see {[}@west2014;@weiss2005{]}).
 
 In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}{]}.
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}\protect\hyperlink{ref-maxwell2017}{44}{]}; and it is higher between consecutive measurements {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and they are not robust if the group sizes are unbalanced {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
+Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}@weiss2005;@geisser1958;@huynh1976{]}.
+However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}@maxwell2017{]}; and it is higher between consecutive measurements {[}@gueorguieva2004;@ugrinowitsch2004{]}.
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}@huynh1976;@greenhouse1959{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}@haverkamp2017{]}, and they are not robust if the group sizes are unbalanced {[}@keselman2001{]}.
 Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
 
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}@pinheiro2006{]}.
 Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
@@
 Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}@molenberghs2004{]}.
 Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}\protect\hyperlink{ref-weiss2005}{42}{]}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{46}{]}.
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{47}{]}, and in the case of MNAR the missing values are dependent on their value.
+Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}@weiss2005{]}.
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}@scheffer2002{]}.
+If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}@potthoff2006{]}, and in the case of MNAR the missing values are dependent on their value.
 
 An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{48}{]}.
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}\protect\hyperlink{ref-west2014}{40}{]}.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}@ma2012{]}.
+In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}@west2014{]}.
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
@@
 To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{49}{]}.
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) {[}@r{]} and the package \emph{nlme} {[}@nlme{]}.
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/l-q-response-1}
+{\centering \includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/l-q-response-1}
 
@@
 
-\FloatBarrier
-
 \hypertarget{GAM-theory}{%
@@
 
-Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors {[}\protect\hyperlink{ref-nelder1972}{50}{]}.
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{51}{]}.
+Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors {[}@nelder1972{]}.
+In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}@simpson2018;@wood2017;@hastie1987{]}.
 A GAM model can be written as:
@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
+One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}@beck1998{]}.
 In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
@@
 The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
+A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist {[}@simpson2018{]}.
 For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{52}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}@hefley2017{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
 In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
-Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
+Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data {[}@wood2017;@simpson2018{]}.
+Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}@wegman1983{]}.
 Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-1}
+{\centering \includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/basis-plot-1}
 
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}\protect\hyperlink{ref-mcelreath2018}{54}{]}.
-In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) {[}\protect\hyperlink{ref-miller2019}{55}{]}.
-Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{56}{]}.
-Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}@mcelreath2018{]}.
+In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}@wood2017{]}.
+GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) {[}@miller2019{]}.
+Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model {[}@laird1982;@pedersen2019{]}.
+Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}@pedersen2019{]}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage {[}@wood2017{]}.
 This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{55}{]} and {[}\protect\hyperlink{ref-marra2012}{57}{]}.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}@miller2019;@wood2017;@simpson2018{]} and {[}@marra2012{]}.
 With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
-
-\FloatBarrier
 
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in {[}@vishwanath2009{]}.
 In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
@@
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in {[}\protect\hyperlink{ref-hefley2017}{52}{]}.
+A detailed description on basis functions and correlations can be found in {[}@hefley2017{]}.
 
@@
 
-\passthrough{\lstinline!m1 <- gam(StO2\_sim \~ Group + s(Day, by=Group, k=5), method='REML, data = dat\_sim)!}
+\begin{lstlisting}[language=R]
+m1 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
+\end{lstlisting}
 
 This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
-The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths {[}@simpson2018{]}.
+The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) {[}@wood2017{]}.
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{58}{]}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{59}{]}.
-
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}@gratia{]}.
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}@wood2017{]} and {[}@harezlak2018{]}.
+
+One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?''.
+The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
 This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-1}
+{\centering \includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-1}
+{\centering \includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/plot-pairwise-comp-1}
 
@@
 
-\FloatBarrier
-
 \hypertarget{discussion}{%
@@
 Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}.
-This ``model misspecification'' error, also is known as a ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{60}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+This ``model misspecification'' error, also is known as a ``Type III'' error {[}@dennis2019{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}@lang2015{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
 
 In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{61}{]}.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}@hefley2017; @pedersen2019{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}@hastie1995{]}.
 This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
@@
 However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} using \(\textsf{R}\) and the package \emph{mgcv}{[}\protect\hyperlink{ref-wood2017}{37}{]} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature {[}@vishwanath2009{]} using \(\textsf{R}\) and the package \emph{mgcv}{[}@wood2017{]} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
 One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{62},\protect\hyperlink{ref-weissgerber2018}{63}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research {[}@begley2015;@weissgerber2018{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
 We have made all the data and code used in this paper accessible, and we hope that this will encourage other researchers to do the same with future projects.
-
-\FloatBarrier
 
@@
 
-\FloatBarrier
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\leavevmode\hypertarget{ref-roblyer2011}{}%
-\CSLLeftMargin{{[}1{]} }
-\CSLRightInline{D.
-Roblyer, S.
-Ueda, A.
-Cerussi, W.
-Tanamai, A.
-Durkin, R.
-Mehta, D.
-Hsiang, J.A.
-Butler, C.
-McLaren, W.P.
-Chen, B.
-Tromberg, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences of the United States of America.
-108 (2011) 14626--14631.
-\url{https://doi.org/10.1073/pnas.1013103108}.}
-
-\leavevmode\hypertarget{ref-tank2020}{}%
-\CSLLeftMargin{{[}2{]} }
-\CSLRightInline{A.
-Tank, H.M.
-Peterson, V.
-Pera, S.
-Tabassum, A.
-Leproux, T.
-O'Sullivan, E.
-Jones, H.
-Cabral, N.
-Ko, R.S.
-Mehta, B.J.
-Tromberg, D.
-Roblyer, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research.
-22 (2020) 1--10.
-\url{https://doi.org/doi:10.1186/s13058-020-01262-1}.}
-
-\leavevmode\hypertarget{ref-pavlov2018}{}%
-\CSLLeftMargin{{[}3{]} }
-\CSLRightInline{M.V.
-Pavlov, T.I.
-Kalganova, Y.S.
-Lyubimtseva, V.I.
-Plekhanov, G.Y.
-Golubyatnikov, O.Y.
-Ilyinskaya, A.G.
-Orlova, P.V.
-Subochev, D.V.
-Safonov, N.M.
-Shakhova, A.V.
-Maslennikova, {Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy}, {Journal of Biomedical Optics}.
-{23} (2018).
-\url{https://doi.org/\%7B10.1117/1.JBO.23.9.091410\%7D}.}
-
-\leavevmode\hypertarget{ref-demidov2018}{}%
-\CSLLeftMargin{{[}4{]} }
-\CSLRightInline{V.
-Demidov, A.
-Maeda, M.
-Sugita, V.
-Madge, S.
-Sadanand, C.
-Flueraru, I.A.
-Vitkin, {Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography}, {Scientific Reports}.
-{8} (2018).
-\url{https://doi.org/\%7B10.1038/s41598-017-18635-w\%7D}.}
-
-\leavevmode\hypertarget{ref-ritter2001}{}%
-\CSLLeftMargin{{[}5{]} }
-\CSLRightInline{G.
-Ritter, L.
-Cohen, C.
-Williams, E.
-Richards, L.
-Old, S.
-Welt, {Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}, {Cancer Research}.
-{61} (2001) 6851--6859.}
-
-\leavevmode\hypertarget{ref-roth2017}{}%
-\CSLLeftMargin{{[}6{]} }
-\CSLRightInline{E.M.
-Roth, A.C.
-Goldberg, A.L.
-Catapano, A.
-Torri, G.D.
-Yancopoulos, N.
-Stahl, A.
-Brunet, G.
-Lecorps, H.M.
-Colhoun, {Antidrug antibodies in atients treated with alirocumab}, {New England Journal of Medicine}.
-{376} (2017) 1589--1590.
-\url{https://doi.org/\%7B10.1056/NEJMc1616623\%7D}.}
-
-\leavevmode\hypertarget{ref-jones2018}{}%
-\CSLLeftMargin{{[}7{]} }
-\CSLRightInline{J.D.
-Jones, H.E.
-Ramser, A.E.
-Woessner, K.P.
-Quinn, {In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing}, {Communications Biology}.
-{1} (2018).
-\url{https://doi.org/\%7B10.1038/s42003-018-0206-4\%7D}.}
-
-\leavevmode\hypertarget{ref-skala2010}{}%
-\CSLLeftMargin{{[}8{]} }
-\CSLRightInline{M.C.
-Skala, A.
-Fontanella, L.
-Lan, J.A.
-Izatt, M.W.
-Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics.
-15 (2010).
-\url{https://doi.org/10.1117/1.3285584}.}
-
-\leavevmode\hypertarget{ref-greening2018}{}%
-\CSLLeftMargin{{[}9{]} }
-\CSLRightInline{G.J.
-Greening, K.P.
-Miller, C.R.
-Spainhour, M.D.
-Cato, T.J.
-Muldoon, {Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy}, {Biomedical Optics Express}.
-{9} (2018) 2871--2886.
-\url{https://doi.org/\%7B10.1364/BOE.9.002871\%7D}.}
-
-\leavevmode\hypertarget{ref-sio2016}{}%
-\CSLLeftMargin{{[}10{]} }
-\CSLRightInline{T.T.
-Sio, P.J.
-Atherton, B.J.
-Birckhead, D.J.
-Schwartz, J.A.
-Sloan, D.K.
-Seisler, J.A.
-Martenson, C.L.
-Loprinzi, P.C.
-Griffin, R.F.
-Morton, J.C.
-Anders, T.J.
-Stoffel, R.E.
-Haselow, R.B.
-Mowat, M.A.N.
-Wittich, J.D.
-Bearden III, R.C.
-Miller, {Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {{[}}alliance{]})}, {Supportive Care in Cancer}.
-{24} (2016) 3847--3855.
-\url{https://doi.org/\%7B10.1007/s00520-016-3213-3\%7D}.}
-
-\leavevmode\hypertarget{ref-kamstra2015}{}%
-\CSLLeftMargin{{[}11{]} }
-\CSLRightInline{J.I.
-Kamstra, P.U.
-Dijkstra, M.
-van Leeuwen, J.L.N.
-Roodenburg, J.A.
-Langendijk, {Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study}, {Oral Oncology}.
-{51} (2015) 548--555.
-\url{https://doi.org/\%7B10.1016/j.oraloncology.2015.01.016\%7D}.}
-
-\leavevmode\hypertarget{ref-wagenmakers2008}{}%
-\CSLLeftMargin{{[}12{]} }
-\CSLRightInline{E.-J.
-Wagenmakers, M.
-Lee, T.
-Lodewyckx, G.J.
-Iverson, Bayesian versus frequentist inference, in: H.
-Hoijtink, I.
-Klugkist, P.A.
-Boelen (Eds.), Bayesian Evaluation of Informative Hypotheses, Springer New York, New York, NY, 2008: pp.
-181--207.
-\url{https://doi.org/10.1007/978-0-387-09612-4_9}.}
-
-\leavevmode\hypertarget{ref-gueorguieva2004}{}%
-\CSLLeftMargin{{[}13{]} }
-\CSLRightInline{R.
-Gueorguieva, J.H.
-Krystal, {Move over ANOVA - Progress in analyzing repeated-measures data and its reflection in papers published in the archives of general psychiatry}, Archives of General Psychiatry.
-61 (2004) 310--317.
-\url{https://doi.org/10.1001/archpsyc.61.3.310}.}
-
-\leavevmode\hypertarget{ref-schober2018}{}%
-\CSLLeftMargin{{[}14{]} }
-\CSLRightInline{P.
-Schober, T.R.
-Vetter, Repeated measures designs and analysis of longitudinal data: If at first you do not succeed-try, try again, Anesthesia and Analgesia.
-127 (2018) 569--575.
-\url{https://doi.org/10.1213/ane.0000000000003511}.}
-
-\leavevmode\hypertarget{ref-pinheiro2006}{}%
-\CSLLeftMargin{{[}15{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science \& Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}
-
-\leavevmode\hypertarget{ref-vishwanath2009}{}%
-\CSLLeftMargin{{[}16{]} }
-\CSLRightInline{K.
-Vishwanath, H.
-Yuan, W.T.
-Barry, M.W.
-Dewhirst, N.
-Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia.
-11 (2009) 889--900.
-\url{https://doi.org/10.1593/neo.09580}.}
-
-\leavevmode\hypertarget{ref-dennis2019}{}%
-\CSLLeftMargin{{[}17{]} }
-\CSLRightInline{B.
-Dennis, J.M.
-Ponciano, M.L.
-Taper, S.R.
-Lele, {Errors in statistical inference under model misspecification: evidence, hypothesis testing, and AIC}, {Frontiers in Ecology and Evolution}.
-{7} (2019).
-\url{https://doi.org/\%7B10.3389/fevo.2019.00372\%7D}.}
-
-\leavevmode\hypertarget{ref-wang2019}{}%
-\CSLLeftMargin{{[}18{]} }
-\CSLRightInline{B.
-Wang, Z.
-Zhou, H.
-Wang, X.M.
-Tu, C.
-Feng, {The p-value and model specification in statistics}, {General Psychiatry}.
-{32} (2019).
-\url{https://doi.org/\%7B10.1136/gpsych-2019-100081\%7D}.}
-
-\leavevmode\hypertarget{ref-liu2010}{}%
-\CSLLeftMargin{{[}19{]} }
-\CSLRightInline{C.
-Liu, T.P.
-Cripe, M.-O.
-Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy.
-18 (2010) 1724--1730.
-\url{https://doi.org/10.1038/mt.2010.127}.}
-
-\leavevmode\hypertarget{ref-halsey2015}{}%
-\CSLLeftMargin{{[}20{]} }
-\CSLRightInline{L.G.
-Halsey, D.
-Curran-Everett, S.L.
-Vowler, G.B.
-Drummond, {The fickle \(p\) value generates irreproducible results}, {Nature Methods}.
-{12} (2015) 179--185.
-\url{https://doi.org/\%7B10.1038/nmeth.3288\%7D}.}
-
-\leavevmode\hypertarget{ref-abdi2010}{}%
-\CSLLeftMargin{{[}21{]} }
-\CSLRightInline{H. Abdi, {Holm's sequential Bonferroni procedure}, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.}
-
-\leavevmode\hypertarget{ref-nakagawa2004}{}%
-\CSLLeftMargin{{[}22{]} }
-\CSLRightInline{S.
-Nakagawa, {A farewell to Bonferroni: the problems of low statistical power and publication bias}, {Behavioral Ecology}.
-{15} (2004) 1044--1045.
-\url{https://doi.org/\%7B10.1093/beheco/arh107\%7D}.}
-
-\leavevmode\hypertarget{ref-gelman2012}{}%
-\CSLLeftMargin{{[}23{]} }
-\CSLRightInline{A.
-Gelman, J.
-Hill, M.
-Yajima, {Why we (usually) don't have to worry about multiple comparisons}, {Journal of Research on Educational Effectiveness}.
-{5} (2012) 189--211.
-\url{https://doi.org/\%7B10.1080/19345747.2011.618213\%7D}.}
-
-\leavevmode\hypertarget{ref-albers2019}{}%
-\CSLLeftMargin{{[}24{]} }
-\CSLRightInline{C. Albers, {The problem with unadjusted multiple and sequential statistical testing}, {Nature Communications}. {10} (2019). \url{https://doi.org/\%7B10.1038/s41467-019-09941-0\%7D}.}
-
-\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
-\CSLLeftMargin{{[}25{]} }
-\CSLRightInline{C.
-Ugrinowitsch, G.W.
-Fellingham, M.D.
-Ricard, Limitations of ordinary least squares models in analyzing repeated measures data, Medicine and Science in Sports and Exercise.
-36 (2004) 2144--2148.
-\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.}
-
-\leavevmode\hypertarget{ref-huynh1976}{}%
-\CSLLeftMargin{{[}26{]} }
-\CSLRightInline{H.
-Huynh, L.S.
-Feldt, {Estimation of the Box correction for degrees of freedom from sample data in randomized block and split-plot designs}, Journal of Educational Statistics.
-1 (1976) 69--82.
-\url{https://doi.org/10.3102/10769986001001069}.}
-
-\leavevmode\hypertarget{ref-greenhouse1959}{}%
-\CSLLeftMargin{{[}27{]} }
-\CSLRightInline{S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.}
-
-\leavevmode\hypertarget{ref-haverkamp2017}{}%
-\CSLLeftMargin{{[}28{]} }
-\CSLRightInline{N.
-Haverkamp, A.
-Beauducel, {Violation of the sphericity assumption and its effect on type-I error rates in repeated measures ANOVA and multi-level linear models (MLM)}, {Frontiers in Psychology}.
-{8} (2017).
-\url{https://doi.org/\%7B10.3389/fpsyg.2017.01841\%7D}.}
-
-\leavevmode\hypertarget{ref-keselman2001}{}%
-\CSLLeftMargin{{[}29{]} }
-\CSLRightInline{H.
-Keselman, J.
-Algina, R.
-Kowalchuk, {The analysis of repeated measures designs: A review}, {British Journal of Mathematica \& Statistical Psychology}.
-{54} (2001) 1--20.
-\url{https://doi.org/\%7B10.1348/000711001159357\%7D}.}
-
-\leavevmode\hypertarget{ref-charan2013}{}%
-\CSLLeftMargin{{[}30{]} }
-\CSLRightInline{Jaykaran.
-Charan, N.
-Kantharia, {How to calculate sample size in animal studies?}, Journal of Pharmacology and Pharmacotherapeutics.
-4 (2013) 303--306.
-\url{https://doi.org/10.4103/0976-500X.119726}.}
-
-\leavevmode\hypertarget{ref-barr2013}{}%
-\CSLLeftMargin{{[}31{]} }
-\CSLRightInline{D.J.
-Barr, R.
-Levy, C.
-Scheepers, H.J.
-Tily, {Random effects structure for confirmatory hypothesis testing: Keep it maximal}, {Journal of Memory and Language}.
-{68} (2013) 255--278.
-\url{https://doi.org/\%7B10.1016/j.jml.2012.11.001\%7D}.}
-
-\leavevmode\hypertarget{ref-rose2012}{}%
-\CSLLeftMargin{{[}32{]} }
-\CSLRightInline{N.L.
-Rose, H.D.
-Yang, S.D.
-Turner, G.L.
-Simpson, {An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to Scotland, UK}, Geochimica Et Cosmochimica Acta.
-82 (2012) 113--135.
-\url{https://doi.org/10.1016/j.gca.2010.12.026}.}
-
-\leavevmode\hypertarget{ref-pedersen2019}{}%
-\CSLLeftMargin{{[}33{]} }
-\CSLRightInline{E.J.
-Pedersen, D.L.
-Miller, G.L.
-Simpson, N.
-Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, {PeerJ}.
-7 (2019).
-\url{https://doi.org/10.7717/peerj.6876}.}
-
-\leavevmode\hypertarget{ref-simpson2018}{}%
-\CSLLeftMargin{{[}34{]} }
-\CSLRightInline{G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.}
-
-\leavevmode\hypertarget{ref-yang2012}{}%
-\CSLLeftMargin{{[}35{]} }
-\CSLRightInline{L.
-Yang, G.
-Qin, N.
-Zhao, C.
-Wang, G.
-Song, {Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality}, {BMC Medical Research Methodology}.
-{12} (2012).
-\url{https://doi.org/\%7B10.1186/1471-2288-12-165\%7D}.}
-
-\leavevmode\hypertarget{ref-beck1998}{}%
-\CSLLeftMargin{{[}36{]} }
-\CSLRightInline{N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. (1998) 596--627.}
-
-\leavevmode\hypertarget{ref-wood2017}{}%
-\CSLLeftMargin{{[}37{]} }
-\CSLRightInline{S.N. Wood, {Generalized additive models: An introduction with R, Second Edition}, CRC Press LLC, Philadelphia, PA, 2017.}
-
-\leavevmode\hypertarget{ref-r}{}%
-\CSLLeftMargin{{[}38{]} }
-\CSLRightInline{R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.}
-
-\leavevmode\hypertarget{ref-wood2016}{}%
-\CSLLeftMargin{{[}39{]} }
-\CSLRightInline{S.N.
-Wood, N.
-Pya, B.
-Saefken, Smoothing parameter and model selection for general smooth models, {Journal of the American Statistical Association}.
-{111} (2016) 1548--1563.
-\url{https://doi.org/\%7B10.1080/01621459.2016.1180986\%7D}.}
-
-\leavevmode\hypertarget{ref-west2014}{}%
-\CSLLeftMargin{{[}40{]} }
-\CSLRightInline{B.T.
-West, K.B.
-Welch, A.T.
-Galecki, Linear mixed models: A practical guide using statistical software, second edition, Taylor \& Francis, 2014.
-\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.}
-
-\leavevmode\hypertarget{ref-wolfinger1996}{}%
-\CSLLeftMargin{{[}41{]} }
-\CSLRightInline{R.D.
-Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics.
-1 (1996) 205--230.
-\url{http://www.jstor.org/stable/1400366}.}
-
-\leavevmode\hypertarget{ref-weiss2005}{}%
-\CSLLeftMargin{{[}42{]} }
-\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://books.google.com/books?id=MQ/_bvWDPsEAC}.}
-
-\leavevmode\hypertarget{ref-geisser1958}{}%
-\CSLLeftMargin{{[}43{]} }
-\CSLRightInline{S.
-Geisser, S.W.
-Greenhouse, {An extension of Box's results on the use of the \(F\) distribution in multivariate analysis}, The Annals of Mathematical Statistics.
-29 (1958) 885--891.
-\url{https://doi.org/10.1214/aoms/1177706545}.}
-
-\leavevmode\hypertarget{ref-maxwell2017}{}%
-\CSLLeftMargin{{[}44{]} }
-\CSLRightInline{S.E.
-Maxwell, H.D.
-Delaney, K.
-Kelley, Designing experiments and analyzing data: A model comparison perspective, third edition, Taylor \& Francis, 2017.
-\url{https://books.google.com/books?id=NmFQDwAAQBAJ}.}
-
-\leavevmode\hypertarget{ref-molenberghs2004}{}%
-\CSLLeftMargin{{[}45{]} }
-\CSLRightInline{G.
-Molenberghs, H.
-Thijs, I.
-Jansen, C.
-Beunckens, M.
-Kenward, C.
-Mallinckrodt, R.
-Carroll, {Analyzing incomplete longitudinal clinical trial data}, {Biostatistics}.
-{5} (2004) 445--464.
-\url{https://doi.org/\%7B10.1093/biostatistics/kxh001\%7D}.}
-
-\leavevmode\hypertarget{ref-scheffer2002}{}%
-\CSLLeftMargin{{[}46{]} }
-\CSLRightInline{J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.}
-
-\leavevmode\hypertarget{ref-potthoff2006}{}%
-\CSLLeftMargin{{[}47{]} }
-\CSLRightInline{R.F.
-Potthoff, G.E.
-Tudor, K.S.
-Pieper, V.
-Hasselblad, {Can one assess whether missing data are missing at random in medical studies?}, {Statistical Methods in Medical Research}.
-{15} (2006) 213--234.
-\url{https://doi.org/\%7B10.1191/0962280206sm448oa\%7D}.}
-
-\leavevmode\hypertarget{ref-ma2012}{}%
-\CSLLeftMargin{{[}48{]} }
-\CSLRightInline{Y.
-Ma, M.
-Mazumdar, S.G.
-Memtsoudis, Beyond repeated-measures analysis of variance advanced statistical methods for the analysis of longitudinal data in anesthesia research, {Regional Anesthesia and Pain Medicine}.
-{37} (2012) 99--105.
-\url{https://doi.org/\%7B10.1097/AAP.0b013e31823ebc74\%7D}.}
-
-\leavevmode\hypertarget{ref-nlme}{}%
-\CSLLeftMargin{{[}49{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, {nlme}: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.}
-
-\leavevmode\hypertarget{ref-nelder1972}{}%
-\CSLLeftMargin{{[}50{]} }
-\CSLRightInline{J.A.
-Nelder, R.W.M.
-Wedderburn, Generalized linear models, Journal of the Royal Statistical Society.
-Series A (General).
-135 (1972) 370--384.
-\url{http://www.jstor.org/stable/2344614}.}
-
-\leavevmode\hypertarget{ref-hastie1987}{}%
-\CSLLeftMargin{{[}51{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association.
-82 (1987) 371--386.
-\url{https://doi.org/10.1080/01621459.1987.10478440}.}
-
-\leavevmode\hypertarget{ref-hefley2017}{}%
-\CSLLeftMargin{{[}52{]} }
-\CSLRightInline{T.J.
-Hefley, K.M.
-Broms, B.M.
-Brost, F.E.
-Buderman, S.L.
-Kay, H.R.
-Scharf, J.R.
-Tipton, P.J.
-Williams, M.B.
-Hooten, {The basis function approach for modeling autocorrelation in ecological data}, {Ecology}.
-{98} (2017) 632--646.
-\url{https://doi.org/\%7B10.1002/ecy.1674\%7D}.}
-
-\leavevmode\hypertarget{ref-wegman1983}{}%
-\CSLLeftMargin{{[}53{]} }
-\CSLRightInline{E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.}
-
-\leavevmode\hypertarget{ref-mcelreath2018}{}%
-\CSLLeftMargin{{[}54{]} }
-\CSLRightInline{R. McElreath, {Statistical rethinking: A Bayesian course with examples in R and Stan}, {Chapman and Hall/CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
-
-\leavevmode\hypertarget{ref-miller2019}{}%
-\CSLLeftMargin{{[}55{]} }
-\CSLRightInline{D.L. Miller, Bayesian views of generalized additive modelling, arXiv Preprint arXiv:1902.01330. (2019).}
-
-\leavevmode\hypertarget{ref-laird1982}{}%
-\CSLLeftMargin{{[}56{]} }
-\CSLRightInline{N.M. Laird, J.H. Ware, Random-effects models for longitudinal data, Biometrics. 38 (1982) 963--974. \url{http://www.jstor.org/stable/2529876}.}
-
-\leavevmode\hypertarget{ref-marra2012}{}%
-\CSLLeftMargin{{[}57{]} }
-\CSLRightInline{G.
-Marra, S.N.
-Wood, Coverage properties of confidence intervals for generalized additive model components, {Scandinavian Journal of Statistics}.
-{39} (2012) 53--74.
-\url{https://doi.org/\%7B10.1111/j.1467-9469.2011.00760.x\%7D}.}
-
-\leavevmode\hypertarget{ref-gratia}{}%
-\CSLLeftMargin{{[}58{]} }
-\CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
-
-\leavevmode\hypertarget{ref-harezlak2018}{}%
-\CSLLeftMargin{{[}59{]} }
-\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, {Semiparametric Regression with R}, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
-
-\leavevmode\hypertarget{ref-lang2015}{}%
-\CSLLeftMargin{{[}60{]} }
-\CSLRightInline{T.A.
-Lang, D.G.
-Altman, {Basic statistical reporting for articles published in Biomedical Journals: The ``Statistical Analyses and Methods in the Published Literature{''} or the SAMPL Guidelines}, INTERNATIONAL JOURNAL OF NURSING STUDIES.
-{52} (2015) 5--9.
-\url{https://doi.org/\%7B10.1016/j.ijnurstu.2014.09.006\%7D}.}
-
-\leavevmode\hypertarget{ref-hastie1995}{}%
-\CSLLeftMargin{{[}61{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models for medical research, {Statistical Methods in Medical Research}.
-4 (1995) 187--196.
-\url{https://doi.org/10.1177/096228029500400302}.}
-
-\leavevmode\hypertarget{ref-begley2015}{}%
-\CSLLeftMargin{{[}62{]} }
-\CSLRightInline{C.G.
-Begley, J.P.A.
-Ioannidis, {Reproducibility in Science Improving the Standard for Basic and Preclinical Research}, {Circulation Research}.
-{116} (2015) 116--126.
-\url{https://doi.org/\%7B10.1161/CIRCRESAHA.114.303819\%7D}.}
-
-\leavevmode\hypertarget{ref-weissgerber2018}{}%
-\CSLLeftMargin{{[}63{]} }
-\CSLRightInline{T.L.
-Weissgerber, O.
-Garcia-Valencia, V.D.
-Garovic, N.M.
-Milic, S.J.
-Winham, {Meta-Research: Why we need to report more than 'Data were Analyzed by t-tests or ANOVA'}, Elife.
-7 (2018) e36163.
-\url{https://doi.org/10.7554/eLife.36163}.}
-
-\end{CSLReferences}
+\hypertarget{declaration-of-conflicting-interests}{%
+\section{Declaration of Conflicting Interests}\label{declaration-of-conflicting-interests}}
+
+The Authors declare that there is no conflict of interest.
 
@@
 
-\hypertarget{code-for-manuscript-data}{%
-\section{Code for Manuscript data}\label{code-for-manuscript-data}}
-
-This section presents the code used to generate figures, models and simulated data from Sections 3 and 4 from the main manuscript.
+\appendixpagenumbering
+
+\counterwithin{figure}{section}
+
+\hypertarget{appendix}{%
+\section{APPENDIX}\label{appendix}}
+
+This section presents the code used to generate figures, models and simulated data from the main manuscript.
 
@@
 
-This section simulated linear and quadratic data in the same manner as in Section \ref{simulation}.
+This section simulates linear and quadratic data in the same manner as in Section \ref{simulation} in the main manuscript.
 The linear simulations using Figure \ref{fig:linear-cases-Appendix} show in panels A and D the simulated mean responses and individual data points.
@@
 \begin{figure}[H]
-\includegraphics{00-Full_document_files/figure-latex/linear-cases-Appendix-1} \caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
+
+{\centering \includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/linear-cases-Appendix-1}
+
+}
+
+\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
 A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
@@
 \begin{figure}[H]
-\includegraphics{00-Full_document_files/figure-latex/quadratic-cases-Appendix-1} \caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
+
+{\centering \includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/quadratic-cases-Appendix-1}
+
+}
+
+\caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
 A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-appendix-1}
+{\centering \includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/basis-plot-appendix-1}
 
@@
 \hypertarget{tumor-data-simulation}{%
-\section{Longitudinal biomedical data simulation and GAMs}\label{tumor-data-simulation}}
+\subsection{Longitudinal biomedical data simulation and GAMs}\label{tumor-data-simulation}}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/first-GAM-diag-1}
+{\centering \includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/first-GAM-diag-1}
 
@@
 
+Special attention must be paid to the `k-index' from \passthrough{\lstinline!gam.check!}.
+This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
+If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless1).
+
 \begin{lstlisting}[language=R]
-#need to add figure number and caption
 gam.check(gam_00)
@@
 
-\includegraphics{00-Full_document_files/figure-latex/first-GAM-check-1}
+\begin{center}\includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/first-GAM-check-1} \end{center}
 
@@
 
-Special attention must be paid to the `k-index' from \passthrough{\lstinline!gam.check!}.
-This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
-If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless1).
 From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data.
@@
 
-\includegraphics{00-Full_document_files/figure-latex/second-GAM-1}
+\begin{center}\includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/second-GAM-1} \end{center}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/second-GAM-diag-1}
+{\centering \includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/second-GAM-diag-1}
 
@@
 
-\includegraphics{00-Full_document_files/figure-latex/final-model-Appendix-1}
+\begin{center}\includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/final-model-Appendix-1} \end{center}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/final-GAM-diag-1}
+{\centering \includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/final-GAM-diag-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/pairwise-comp-workflow-fig-1}
+{\centering \includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/pairwise-comp-workflow-fig-1}
 
@@
 \hypertarget{gam-and-linear-model-plots-and-missing-data}{%
-\section{GAM and Linear model plots and Missing data}\label{gam-and-linear-model-plots-and-missing-data}}
+\subsection{GAM and Linear model plots and Missing data}\label{gam-and-linear-model-plots-and-missing-data}}
 
@@
 \hypertarget{gam-and-linear-model-plots}{%
-\subsection{GAM and Linear model plots}\label{gam-and-linear-model-plots}}
+\subsubsection{GAM and Linear model plots}\label{gam-and-linear-model-plots}}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-Appendix-1}
+{\centering \includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/sim-smooth-plot-Appendix-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-Appendix-1}
+{\centering \includegraphics[width=0.75\linewidth,]{SAGE_Appendix_files/figure-latex/plot-pairwise-comp-Appendix-1}
 

### 2021-06-23 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE_No_Appendix_Bibliography.tex
Cover letters uploaded

--- 2021-06-22 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE_No_Appendix.tex
@@
 Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response \citep{roblyer2011, tank2020, pavlov2018, demidov2018}, antibody expression \citep{ritter2001, roth2017}, and cell metabolism \citep{jones2018, skala2010} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals\citep{greening2018}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients \citep{sio2016}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients \citep{kamstra2015}.
+Tumor response \cite{roblyer2011, tank2020, pavlov2018, demidov2018}, antibody expression \cite{ritter2001, roth2017}, and cell metabolism \cite{jones2018, skala2010} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
+Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals\cite{greening2018}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients \cite{sio2016}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients \cite{kamstra2015}.
 
 Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times \citep{wagenmakers2008}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) \citep{gueorguieva2004, schober2018}.
+The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times \cite{wagenmakers2008}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
+The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) \cite{gueorguieva2004, schober2018}.
 
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA \citep{pinheiro2006}.
+The expected linear behavior of the response through time is a key requisite in rm-ANOVA \cite{pinheiro2006}.
 This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
 In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings \citep{roblyer2011, skala2010, vishwanath2009}.
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings \cite{roblyer2011, skala2010, vishwanath2009}.
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
@@
 Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\citep{dennis2019}.
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid \citep{wang2019}.
+In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\cite{dennis2019}.
+For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid \cite{wang2019}.
 
-Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) \citep{liu2010, halsey2015}, consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni \citep{abdi2010}), but they in turn reduce statistical power (1-\(\beta\))\citep{nakagawa2004}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) \citep{gelman2012, albers2019}.
+Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) \cite{liu2010, halsey2015}, consequently biasing the conclusions of the study.
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni \cite{abdi2010}), but they in turn reduce statistical power (1-\(\beta\))\cite{nakagawa2004}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) \cite{gelman2012, albers2019}.
 Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
 
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases \citep{ugrinowitsch2004}.
-Corrections can be made in rm-ANOVA in the absence of compound symmetry \citep{huynh1976, greenhouse1959}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements\citep{haverkamp2017}, and group sizes \citep{keselman2001}.
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons \citep{charan2013} which might cause the corrections for lack of compound symmetry to be ineffective.
+On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases \cite{ugrinowitsch2004}.
+Corrections can be made in rm-ANOVA in the absence of compound symmetry \cite{huynh1976, greenhouse1959}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements\cite{haverkamp2017}, and group sizes \cite{keselman2001}.
+In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons \cite{charan2013} which might cause the corrections for lack of compound symmetry to be ineffective.
 
@@
 In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis \citep{gueorguieva2004}.
+When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis \cite{gueorguieva2004}.
 This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
@@
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data \citep{skala2010, vishwanath2009}.
+The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data \cite{skala2010, vishwanath2009}.
 Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject \citep{pinheiro2006}.
-However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent \citep{gueorguieva2004, barr2013}.
-And even more importantly, LMEMs also assume a linear relationship between the response and time \citep{pinheiro2006}, making them unsuitable to analyze non-linear data.
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject \cite{pinheiro2006}.
+However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent \cite{gueorguieva2004, barr2013}.
+And even more importantly, LMEMs also assume a linear relationship between the response and time \cite{pinheiro2006}, making them unsuitable to analyze non-linear data.
 
@@
 Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data \citep{rose2012, pedersen2019, simpson2018}, health-environment interactions \citep{yang2012} and the dynamics of government in political science \citep{beck1998}.
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements \citep{wood2017} and 3) can easily accommodate missing observations.
+Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data \cite{rose2012, pedersen2019, simpson2018}, health-environment interactions \cite{yang2012} and the dynamics of government in political science \cite{beck1998}.
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements \cite{wood2017} and 3) can easily accommodate missing observations.
 Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
@@
 The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, \(\textsf{R}\) \citep{r} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} \citep{wood2016, wood2017} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models \citep{haverkamp2017}.
+In particular, \(\textsf{R}\) \cite{r} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} \cite{wood2016, wood2017} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models \cite{haverkamp2017}.
 Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
@@
 Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies \citep{vishwanath2009}.
+Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies \cite{vishwanath2009}.
 The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
@@
 
-A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish \citep{pinheiro2006, west2014}.
+A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish \cite{pinheiro2006, west2014}.
 A LMEM with interaction between time and treatment for a longitudinal study can be written as:
@@
 The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level\citep{pinheiro2006}.
+In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level\cite{pinheiro2006}.
 However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
@@
 In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study\citep{wolfinger1996} (For an in-depth analysis of the covariance matrix see \citep{west2014, weiss2005}).
+The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study\cite{wolfinger1996} (For an in-depth analysis of the covariance matrix see \cite{west2014, weiss2005}).
 
 In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time \citep{weiss2005, geisser1958, huynh1976}.
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time \citep{maxwell2017}; and it is higher between consecutive measurements \citep{gueorguieva2004, ugrinowitsch2004}.
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)\citep{huynh1976, greenhouse1959} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements\citep{haverkamp2017}, and they are not robust if the group sizes are unbalanced \citep{keselman2001}.
+Under this assumption, the between-subject variance and within-subject correlation are constant across time \cite{weiss2005, geisser1958, huynh1976}.
+However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time \cite{maxwell2017}; and it is higher between consecutive measurements \cite{gueorguieva2004, ugrinowitsch2004}.
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)\cite{huynh1976, greenhouse1959} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements\cite{haverkamp2017}, and they are not robust if the group sizes are unbalanced \cite{keselman2001}.
 Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
 
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others \citep{pinheiro2006}.
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others \cite{pinheiro2006}.
 Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
@@
 Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator \citep{molenberghs2004}.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator \cite{molenberghs2004}.
 Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) \citep{weiss2005}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest \citep{scheffer2002}.
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information \citep{potthoff2006}, and in the case of MNAR the missing values are dependent on their value.
+Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) \cite{weiss2005}.
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest \cite{scheffer2002}.
+If the data are MCAR, this means that the missingness is completely unrelated to the collected information \cite{potthoff2006}, and in the case of MNAR the missing values are dependent on their value.
 
 An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size \citep{ma2012}.
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR \citep{west2014}.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size \cite{ma2012}.
+In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR \cite{west2014}.
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
@@
 To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) \citep{r} and the package \emph{nlme} \citep{nlme}.
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) \cite{r} and the package \emph{nlme} \cite{nlme}.
 
@@
 
-Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors \citep{nelder1972}.
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case\citep{simpson2018, wood2017, hastie1987}.
+Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors \cite{nelder1972}.
+In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case\cite{simpson2018, wood2017, hastie1987}.
 A GAM model can be written as:
@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference \citep{beck1998}.
+One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference \cite{beck1998}.
 In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
@@
 The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist \citep{simpson2018}.
+A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist \cite{simpson2018}.
 For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response \citep{hefley2017}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response \cite{hefley2017}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
 In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data \citep{wood2017, simpson2018}.
-Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness \citep{wegman1983}.
+Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data \cite{wood2017, simpson2018}.
+Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness \cite{wegman1983}.
 Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief \citep{mcelreath2018}.
-In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs \citep{wood2017}.
-GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) \citep{miller2019}.
-Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model \citep{laird1982, pedersen2019}.
-Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals \citep{pedersen2019}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage \citep{wood2017}.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief \cite{mcelreath2018}.
+In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs \cite{wood2017}.
+GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) \cite{miller2019}.
+Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model \cite{laird1982, pedersen2019}.
+Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals \cite{pedersen2019}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage \cite{wood2017}.
 This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in \citep{miller2019, wood2017, simpson2018} and \citep{marra2012}.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in \cite{miller2019, wood2017, simpson2018} and \cite{marra2012}.
 With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in \citep{vishwanath2009}.
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in \cite{vishwanath2009}.
 In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
@@
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in \citep{hefley2017}.
+A detailed description on basis functions and correlations can be found in \cite{hefley2017}.
 
@@
 This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
-The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths \citep{simpson2018}.
-The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) \citep{wood2017}.
+The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths \cite{simpson2018}.
+The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) \cite{wood2017}.
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} \citep{gratia}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in \citep{wood2017} and \citep{harezlak2018}.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} \cite{gratia}.
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in \cite{wood2017} and \cite{harezlak2018}.
 
@@
 Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}.
-This ``model misspecification'' error, also is known as a ``Type III'' error \citep{dennis2019} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) \citep{lang2015} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+This ``model misspecification'' error, also is known as a ``Type III'' error \cite{dennis2019} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) \cite{lang2015} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
 
 In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) \citep{hefley2017, pedersen2019} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research \citep{hastie1995}.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) \cite{hefley2017, pedersen2019} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research \cite{hastie1995}.
 This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
@@
 However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature \citep{vishwanath2009} using \(\textsf{R}\) and the package \emph{mgcv}\citep{wood2017} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature \cite{vishwanath2009} using \(\textsf{R}\) and the package \emph{mgcv}\cite{wood2017} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
 One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research \citep{begley2015, weissgerber2018}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research \cite{begley2015, weissgerber2018}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 \bibliographystyle{sagev}
-\bibliography{refs.bib}
+
+\begin{thebibliography}{10}
+\providecommand{\url}[1]{\texttt{#1}}
+\providecommand{\urlprefix}{URL }
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{DOI:\discretionary{}{}{}#1}\else
+  \providecommand{\doi}{DOI:\discretionary{}{}{}\begingroup
+  \urlstyle{rm}\Url}\fi
+\providecommand{\eprint}[2][]{\url{#2}}
+
+\bibitem{roblyer2011}
+Roblyer D, Ueda S, Cerussi A et~al.
+\newblock Optical imaging of breast cancer oxyhemoglobin flare correlates with
+  neoadjuvant chemotherapy response one day after starting treatment.
+\newblock \emph{Proceedings of the National Academy of Sciences} 2011; 108(35):
+  14626--14631.
+\newblock \doi{10.1073/pnas.1013103108}.
+\newblock \urlprefix\url{https://doi.org/10.1073/pnas.1013103108}.
+
+\bibitem{tank2020}
+Tank A, Peterson HM, Pera V et~al.
+\newblock Diffuse optical spectroscopic imaging reveals distinct early breast
+  tumor hemodynamic responses to metronomic and maximum tolerated dose
+  regimens.
+\newblock \emph{Breast Cancer Research} 2020; 22(1).
+\newblock \doi{10.1186/s13058-020-01262-1}.
+\newblock \urlprefix\url{https://doi.org/10.1186/s13058-020-01262-1}.
+
+\bibitem{pavlov2018}
+Pavlov MV, Kalganova TI and Lyubimtseva YS.
+\newblock Multimodal approach in assessment of the response of breast cancer to
+  neoadjuvant chemotherapy.
+\newblock \emph{Journal of Biomedical Optics} 2018; 23(09): 1.
+\newblock \doi{10.1117/1.jbo.23.9.091410}.
+\newblock \urlprefix\url{https://doi.org/10.1117/1.jbo.23.9.091410}.
+
+\bibitem{demidov2018}
+Demidov V, Maeda A, Sugita M et~al.
+\newblock Preclinical longitudinal imaging of tumor microvascular
+  radiobiological response with functional optical coherence tomography.
+\newblock \emph{Scientific Reports} 2018; 8(1).
+\newblock \doi{10.1038/s41598-017-18635-w}.
+\newblock \urlprefix\url{https://doi.org/10.1038/s41598-017-18635-w}.
+
+\bibitem{ritter2001}
+Ritter G, Cohen L, Williams C et~al.
+\newblock {Serological analysis of human anti-human antibody responses in colon
+  cancer patients treated with repeated doses of humanized monoclonal antibody
+  A33}.
+\newblock \emph{{Cancer Research}} {2001}; {61}({18}): {6851--6859}.
+
+\bibitem{roth2017}
+Roth EM, Goldberg AC, Catapano AL et~al.
+\newblock Antidrug antibodies in patients treated with alirocumab.
+\newblock \emph{New England Journal of Medicine} 2017; 376(16): 1589--1590.
+\newblock \doi{10.1056/nejmc1616623}.
+\newblock \urlprefix\url{https://doi.org/10.1056/nejmc1616623}.
+
+\bibitem{jones2018}
+Jones JD, Ramser HE, Woessner AE et~al.
+\newblock In vivo multiphoton microscopy detects longitudinal metabolic changes
+  associated with delayed skin wound healing.
+\newblock \emph{Communications Biology} 2018; 1(1).
+\newblock \doi{10.1038/s42003-018-0206-4}.
+\newblock \urlprefix\url{https://doi.org/10.1038/s42003-018-0206-4}.
+
+\bibitem{skala2010}
+Skala MC, Fontanella A, Lan L et~al.
+\newblock Longitudinal optical imaging of tumor metabolism and hemodynamics.
+\newblock \emph{Journal of Biomedical Optics} 2010; 15(1): 011112.
+\newblock \doi{10.1117/1.3285584}.
+\newblock \urlprefix\url{https://doi.org/10.1117/1.3285584}.
+
+\bibitem{greening2018}
+Greening GJ, Miller KP, Spainhour CR et~al.
+\newblock Effects of isoflurane anesthesia on physiological parameters in
+  murine subcutaneous tumor allografts measured via diffuse reflectance
+  spectroscopy.
+\newblock \emph{Biomedical Optics Express} 2018; 9(6): 2871.
+\newblock \doi{10.1364/boe.9.002871}.
+\newblock \urlprefix\url{https://doi.org/10.1364/boe.9.002871}.
+
+\bibitem{sio2016}
+Sio TT, Atherton PJ, Birckhead BJ et~al.
+\newblock Repeated measures analyses of dermatitis symptom evolution in breast
+  cancer patients receiving radiotherapy in a phase 3 randomized trial of
+  mometasone furoate vs placebo (n06c4 [alliance]).
+\newblock \emph{Supportive Care in Cancer} 2016; 24(9): 3847--3855.
+\newblock \doi{10.1007/s00520-016-3213-3}.
+\newblock \urlprefix\url{https://doi.org/10.1007/s00520-016-3213-3}.
+
+\bibitem{kamstra2015}
+Kamstra J, Dijkstra P, van Leeuwen M et~al.
+\newblock Mouth opening in patients irradiated for head and neck cancer: A
+  prospective repeated measures study.
+\newblock \emph{Oral Oncology} 2015; 51(5): 548--555.
+\newblock \doi{10.1016/j.oraloncology.2015.01.016}.
+\newblock \urlprefix\url{https://doi.org/10.1016/j.oraloncology.2015.01.016}.
+
+\bibitem{wagenmakers2008}
+Wagenmakers EJ, Lee M, Lodewyckx T et~al.
+\newblock Bayesian versus frequentist inference.
+\newblock In \emph{Bayesian Evaluation of Informative Hypotheses}. Springer New
+  York, 2008.
+\newblock pp. 181--207.
+\newblock \doi{10.1007/978-0-387-09612-4_9}.
+\newblock \urlprefix\url{https://doi.org/10.1007/978-0-387-09612-4_9}.
+
+\bibitem{gueorguieva2004}
+Gueorguieva R and Krystal JH.
+\newblock Move over {ANOVA}.
+\newblock \emph{Archives of General Psychiatry} 2004; 61(3): 310.
+\newblock \doi{10.1001/archpsyc.61.3.310}.
+\newblock \urlprefix\url{https://doi.org/10.1001/archpsyc.61.3.310}.
+
+\bibitem{schober2018}
+Schober P and Vetter TR.
+\newblock Repeated measures designs and analysis of longitudinal data.
+\newblock \emph{Anesthesia {\&} Analgesia} 2018; 127(2): 569--575.
+\newblock \doi{10.1213/ane.0000000000003511}.
+\newblock \urlprefix\url{https://doi.org/10.1213/ane.0000000000003511}.
+
+\bibitem{pinheiro2006}
+Pinheiro J and Bates D.
+\newblock \emph{{Mixed-effects models in S and S-PLUS}}.
+\newblock Springer Science and Business Media, 2006.
+\newblock ISBN 0387227474.
+\newblock \doi{https://doi.org/10.1007/b98882}.
+
+\bibitem{vishwanath2009}
+Vishwanath K, Yuan H, Barry WT et~al.
+\newblock Using optical spectroscopy to longitudinally monitor physiological
+  changes within solid tumors.
+\newblock \emph{Neoplasia} 2009; 11(9): 889--900.
+\newblock \doi{10.1593/neo.09580}.
+\newblock \urlprefix\url{https://doi.org/10.1593/neo.09580}.
+
+\bibitem{dennis2019}
+Dennis B, Ponciano JM, Taper ML et~al.
+\newblock Errors in statistical inference under model misspecification:
+  Evidence, hypothesis testing, and {AIC}.
+\newblock \emph{Frontiers in Ecology and Evolution} 2019; 7.
+\newblock \doi{10.3389/fevo.2019.00372}.
+\newblock \urlprefix\url{https://doi.org/10.3389/fevo.2019.00372}.
+
+\bibitem{wang2019}
+Wang B, Zhou Z, Wang H et~al.
+\newblock The p-value and model specification in statistics.
+\newblock \emph{General Psychiatry} 2019; 32(3): e100081.
+\newblock \doi{10.1136/gpsych-2019-100081}.
+\newblock \urlprefix\url{https://doi.org/10.1136/gpsych-2019-100081}.
+
+\bibitem{liu2010}
+Liu C, Cripe TP and Kim MO.
+\newblock Statistical issues in longitudinal data analysis for treatment
+  efficacy studies in the biomedical sciences.
+\newblock \emph{Molecular Therapy} 2010; 18(9): 1724--1730.
+\newblock \doi{10.1038/mt.2010.127}.
+\newblock \urlprefix\url{https://doi.org/10.1038/mt.2010.127}.
+
+\bibitem{halsey2015}
+Halsey LG, Curran-Everett D, Vowler SL et~al.
+\newblock The fickle p value generates irreproducible results.
+\newblock \emph{Nature Methods} 2015; 12(3): 179--185.
+\newblock \doi{10.1038/nmeth.3288}.
+\newblock \urlprefix\url{https://doi.org/10.1038/nmeth.3288}.
+
+\bibitem{abdi2010}
+Abdi H.
+\newblock {Holm’s sequential Bonferroni procedure}.
+\newblock \emph{Encyclopedia of Research Design} 2010; 1(8): 1--8.
+\newblock \doi{10.4135/9781412961288.n178}.
+
+\bibitem{nakagawa2004}
+Nakagawa S.
+\newblock A farewell to bonferroni: the problems of low statistical power and
+  publication bias.
+\newblock \emph{Behavioral Ecology} 2004; 15(6): 1044--1045.
+\newblock \doi{10.1093/beheco/arh107}.
+\newblock \urlprefix\url{https://doi.org/10.1093/beheco/arh107}.
+
+\bibitem{gelman2012}
+Gelman A, Hill J and Yajima M.
+\newblock Why we (usually) don{\textquotesingle}t have to worry about multiple
+  comparisons.
+\newblock \emph{Journal of Research on Educational Effectiveness} 2012; 5(2):
+  189--211.
+\newblock \doi{10.1080/19345747.2011.618213}.
+\newblock \urlprefix\url{https://doi.org/10.1080/19345747.2011.618213}.
+
+\bibitem{albers2019}
+Albers C.
+\newblock The problem with unadjusted multiple and sequential statistical
+  testing.
+\newblock \emph{Nature Communications} 2019; 10(1).
+\newblock \doi{10.1038/s41467-019-09941-0}.
+\newblock \urlprefix\url{https://doi.org/10.1038/s41467-019-09941-0}.
+
+\bibitem{ugrinowitsch2004}
+Ugrinowitsch C, Fellingham GW and Ricard MD.
+\newblock Limitations of ordinary least squares models in analyzing repeated
+  measures data.
+\newblock \emph{Medicine {\&} Science in Sports {\&} Exercise} 2004; :
+  2144--2148\doi{10.1249/01.mss.0000147580.40591.75}.
+\newblock \urlprefix\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.
+
+\bibitem{huynh1976}
+Huynh H and Feldt LS.
+\newblock Estimation of the box correction for degrees of freedom from sample
+  data in randomized block and split-plot designs.
+\newblock \emph{Journal of Educational Statistics} 1976; 1(1): 69--82.
+\newblock \doi{10.3102/10769986001001069}.
+\newblock \urlprefix\url{https://doi.org/10.3102/10769986001001069}.
+
+\bibitem{greenhouse1959}
+Greenhouse SW and Geisser S.
+\newblock On methods in the analysis of profile data.
+\newblock \emph{Psychometrika} 1959; 24(2): 95--112.
+\newblock \doi{10.1007/bf02289823}.
+\newblock \urlprefix\url{https://doi.org/10.1007/bf02289823}.
+
+\bibitem{haverkamp2017}
+Haverkamp N and Beauducel A.
+\newblock Violation of the sphericity assumption and its effect on type-i error
+  rates in repeated measures {ANOVA} and multi-level linear models ({MLM}).
+\newblock \emph{Frontiers in Psychology} 2017; 8.
+\newblock \doi{10.3389/fpsyg.2017.01841}.
+\newblock \urlprefix\url{https://doi.org/10.3389/fpsyg.2017.01841}.
+
+\bibitem{keselman2001}
+Keselman HJ, Algina J and Kowalchuk RK.
+\newblock The analysis of repeated measures designs: A review.
+\newblock \emph{British Journal of Mathematical and Statistical Psychology}
+  2001; 54(1): 1--20.
+\newblock \doi{10.1348/000711001159357}.
+\newblock \urlprefix\url{https://doi.org/10.1348/000711001159357}.
+
+\bibitem{charan2013}
+Charan J and Kantharia N.
+\newblock How to calculate sample size in animal studies?
+\newblock \emph{Journal of Pharmacology and Pharmacotherapeutics} 2013; 4(4):
+  303.
+\newblock \doi{10.4103/0976-500x.119726}.
+\newblock \urlprefix\url{https://doi.org/10.4103/0976-500x.119726}.
+
+\bibitem{barr2013}
+Barr DJ, Levy R, Scheepers C et~al.
+\newblock Random effects structure for confirmatory hypothesis testing: Keep it
+  maximal.
+\newblock \emph{Journal of Memory and Language} 2013; 68(3): 255--278.
+\newblock \doi{10.1016/j.jml.2012.11.001}.
+\newblock \urlprefix\url{https://doi.org/10.1016/j.jml.2012.11.001}.
+
+\bibitem{rose2012}
+Rose NL, Yang H, Turner SD et~al.
+\newblock An assessment of the mechanisms for the transfer of lead and mercury
+  from atmospherically contaminated organic soils to lake sediments with
+  particular reference to scotland, {UK}.
+\newblock \emph{Geochimica et Cosmochimica Acta} 2012; 82: 113--135.
+\newblock \doi{10.1016/j.gca.2010.12.026}.
+\newblock \urlprefix\url{https://doi.org/10.1016/j.gca.2010.12.026}.
+
+\bibitem{pedersen2019}
+Pedersen EJ, Miller DL, Simpson GL et~al.
+\newblock Hierarchical generalized additive models in ecology: an introduction
+  with mgcv.
+\newblock \emph{{PeerJ}} 2019; 7: e6876.
+\newblock \doi{10.7717/peerj.6876}.
+\newblock \urlprefix\url{https://doi.org/10.7717/peerj.6876}.
+
+\bibitem{simpson2018}
+Simpson GL.
+\newblock Modelling palaeoecological time series using generalised additive
+  models.
+\newblock \emph{Frontiers in Ecology and Evolution} 2018; 6.
+\newblock \doi{10.3389/fevo.2018.00149}.
+\newblock \urlprefix\url{https://doi.org/10.3389/fevo.2018.00149}.
+
+\bibitem{yang2012}
+Yang L, Qin G, Zhao N et~al.
+\newblock Using a generalized additive model with autoregressive terms to study
+  the effects of daily temperature on mortality.
+\newblock \emph{{BMC} Medical Research Methodology} 2012; 12(1).
+\newblock \doi{10.1186/1471-2288-12-165}.
+\newblock \urlprefix\url{https://doi.org/10.1186/1471-2288-12-165}.
+
+\bibitem{beck1998}
+Beck N and Jackman S.
+\newblock Beyond linearity by default: Generalized additive models.
+\newblock \emph{American Journal of Political Science} 1998; 42(2): 596.
+\newblock \doi{10.2307/2991772}.
+\newblock \urlprefix\url{https://doi.org/10.2307/2991772}.
+
+\bibitem{wood2017}
+Wood SN.
+\newblock \emph{Generalized Additive Models}.
+\newblock Chapman and Hall/{CRC}, 2017.
+\newblock \doi{10.1201/9781315370279}.
+\newblock \urlprefix\url{https://doi.org/10.1201/9781315370279}.
+
+\bibitem{r}
+{R Core Team}.
+\newblock \emph{R: A Language and Environment for Statistical Computing}.
+\newblock R Foundation for Statistical Computing, Vienna, Austria, 2020.
+\newblock \urlprefix\url{https://www.R-project.org/}.
+
+\bibitem{wood2016}
+Wood SN, Pya N and S\"{a}fken B.
+\newblock Smoothing parameter and model selection for general smooth models.
+\newblock \emph{Journal of the American Statistical Association} 2016;
+  111(516): 1548--1563.
+\newblock \doi{10.1080/01621459.2016.1180986}.
+\newblock \urlprefix\url{https://doi.org/10.1080/01621459.2016.1180986}.
+
+\bibitem{west2014}
+West B, Welch K and Galecki A.
+\newblock \emph{Linear Mixed Models: A Practical Guide Using Statistical
+  Software, Second Edition}.
+\newblock Taylor \& Francis, 2014.
+\newblock ISBN 9781466560994.
+\newblock \urlprefix\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.
+
+\bibitem{wolfinger1996}
+Wolfinger RD.
+\newblock Heterogeneous variance: Covariance structures for repeated measures.
+\newblock \emph{Journal of Agricultural, Biological, and Environmental
+  Statistics} 1996; 1(2): 205.
+\newblock \doi{10.2307/1400366}.
+\newblock \urlprefix\url{https://doi.org/10.2307/1400366}.
+
+\bibitem{weiss2005}
+Weiss R.
+\newblock \emph{Modeling Longitudinal Data}.
+\newblock Springer Texts in Statistics, Springer New York, 2005.
+\newblock ISBN 9780387402710.
+\newblock \doi{10.1007/0-387-28314-5}.
+\newblock \urlprefix\url{https://books.google.com/books?id=MQ\_bvWDPsEAC}.
+
+\bibitem{geisser1958}
+Geisser S and Greenhouse SW.
+\newblock An extension of box{\textquotesingle}s results on the use of the
+  {\textdollar}f{\textdollar} distribution in multivariate analysis.
+\newblock \emph{The Annals of Mathematical Statistics} 1958; 29(3): 885--891.
+\newblock \doi{10.1214/aoms/1177706545}.
+\newblock \urlprefix\url{https://doi.org/10.1214/aoms/1177706545}.
+
+\bibitem{maxwell2017}
+Maxwell SE, Delaney HD and Kelley K.
+\newblock \emph{Designing Experiments and Analyzing Data}.
+\newblock Routledge, 2017.
+\newblock \doi{10.4324/9781315642956}.
+\newblock \urlprefix\url{https://doi.org/10.4324/9781315642956}.
+
+\bibitem{molenberghs2004}
+Molenberghs G.
+\newblock Analyzing incomplete longitudinal clinical trial data.
+\newblock \emph{Biostatistics} 2004; 5(3): 445--464.
+\newblock \doi{10.1093/biostatistics/kxh001}.
+\newblock \urlprefix\url{https://doi.org/10.1093/biostatistics/kxh001}.
+
+\bibitem{scheffer2002}
+Scheffer J.
+\newblock Dealing with missing data.
+\newblock \emph{Research Letters in the Information and Mathematical Sciences}
+  2002; 3: 153--160.
+
+\bibitem{potthoff2006}
+Potthoff RF, Tudor GE, Pieper KS et~al.
+\newblock Can one assess whether missing data are missing at random in medical
+  studies?
+\newblock \emph{Statistical Methods in Medical Research} 2006; 15(3): 213--234.
+\newblock \doi{10.1191/0962280206sm448oa}.
+\newblock \urlprefix\url{https://doi.org/10.1191/0962280206sm448oa}.
+
+\bibitem{ma2012}
+Ma Y, Mazumdar M and Memtsoudis SG.
+\newblock Beyond repeated-measures analysis of variance.
+\newblock \emph{Regional Anesthesia and Pain Medicine} 2012; 37(1): 99--105.
+\newblock \doi{10.1097/aap.0b013e31823ebc74}.
+\newblock \urlprefix\url{https://doi.org/10.1097/aap.0b013e31823ebc74}.
+
+\bibitem{nlme}
+Pinheiro J, Bates D, DebRoy S et~al.
+\newblock \emph{{nlme}: Linear and Nonlinear Mixed Effects Models}, 2020.
+\newblock \urlprefix\url{https://CRAN.R-project.org/package=nlme}.
+\newblock R package version 3.1-150.
+
+\bibitem{nelder1972}
+Nelder JA and Wedderburn RWM.
+\newblock Generalized linear models.
+\newblock \emph{Journal of the Royal Statistical Society Series A (General)}
+  1972; 135(3): 370.
+\newblock \doi{10.2307/2344614}.
+\newblock \urlprefix\url{https://doi.org/10.2307/2344614}.
+
+\bibitem{hastie1987}
+Hastie T and Tibshirani R.
+\newblock Generalized additive models: Some applications.
+\newblock \emph{Journal of the American Statistical Association} 1987; 82(398):
+  371--386.
+\newblock \doi{10.1080/01621459.1987.10478440}.
+\newblock \urlprefix\url{https://doi.org/10.1080/01621459.1987.10478440}.
+
+\bibitem{hefley2017}
+Hefley TJ, Broms KM, Brost BM et~al.
+\newblock The basis function approach for modeling autocorrelation in
+  ecological data.
+\newblock \emph{Ecology} 2017; 98(3): 632--646.
+\newblock \doi{10.1002/ecy.1674}.
+\newblock \urlprefix\url{https://doi.org/10.1002/ecy.1674}.
+
+\bibitem{wegman1983}
+Wegman EJ and Wright IW.
+\newblock Splines in statistics.
+\newblock \emph{Journal of the American Statistical Association} 1983; 78(382):
+  351--365.
+\newblock \doi{10.1080/01621459.1983.10477977}.
+\newblock \urlprefix\url{https://doi.org/10.1080/01621459.1983.10477977}.
+
+\bibitem{mcelreath2018}
+McElreath R.
+\newblock \emph{Statistical Rethinking}.
+\newblock Chapman and Hall/{CRC}, 2018.
+\newblock \doi{10.1201/9781315372495}.
+\newblock \urlprefix\url{https://doi.org/10.1201/9781315372495}.
+
+\bibitem{miller2019}
+Miller DL.
+\newblock Bayesian views of generalized additive modelling.
+\newblock \emph{arXiv preprint arXiv:190201330} 2019; .
+
+\bibitem{laird1982}
+Laird NM and Ware JH.
+\newblock Random-effects models for longitudinal data.
+\newblock \emph{Biometrics} 1982; 38(4): 963.
+\newblock \doi{10.2307/2529876}.
+\newblock \urlprefix\url{https://doi.org/10.2307/2529876}.
+
+\bibitem{marra2012}
+Marra G and Wood SN.
+\newblock Coverage properties of confidence intervals for generalized additive
+  model components.
+\newblock \emph{Scandinavian Journal of Statistics} 2012; 39(1): 53--74.
+\newblock \doi{10.1111/j.1467-9469.2011.00760.x}.
+\newblock \urlprefix\url{https://doi.org/10.1111/j.1467-9469.2011.00760.x}.
+
+\bibitem{gratia}
+Simpson GL.
+\newblock \emph{gratia: Graceful 'ggplot'-Based Graphics and Other Functions
+  for GAMs Fitted Using 'mgcv'}, 2020.
+\newblock \urlprefix\url{https://CRAN.R-project.org/package=gratia}.
+\newblock R package version 0.4.1.
+
+\bibitem{harezlak2018}
+Harezlak J, Ruppert D and Wand MP.
+\newblock \emph{Semiparametric Regression with R}.
+\newblock Springer New York, 2018.
+\newblock \doi{10.1007/978-1-4939-8853-2}.
+\newblock \urlprefix\url{https://doi.org/10.1007/978-1-4939-8853-2}.
+
+\bibitem{lang2015}
+Lang TA and Altman DG.
+\newblock Basic statistical reporting for articles published in biomedical
+  journals: The {\textquotedblleft}statistical analyses and methods in the
+  published literature{\textquotedblright} or the {SAMPL} guidelines.
+\newblock \emph{International Journal of Nursing Studies} 2015; 52(1): 5--9.
+\newblock \doi{10.1016/j.ijnurstu.2014.09.006}.
+\newblock \urlprefix\url{https://doi.org/10.1016/j.ijnurstu.2014.09.006}.
+
+\bibitem{hastie1995}
+Hastie T and Tibshirani R.
+\newblock Generalized additive models for medical research.
+\newblock \emph{Statistical Methods in Medical Research} 1995; 4(3): 187--196.
+\newblock \doi{10.1177/096228029500400302}.
+\newblock \urlprefix\url{https://doi.org/10.1177/096228029500400302}.
+
+\bibitem{begley2015}
+Begley CG and Ioannidis JP.
+\newblock Reproducibility in science.
+\newblock \emph{Circulation Research} 2015; 116(1): 116--126.
+\newblock \doi{10.1161/circresaha.114.303819}.
+\newblock \urlprefix\url{https://doi.org/10.1161/circresaha.114.303819}.
+
+\bibitem{weissgerber2018}
+Weissgerber TL, Garcia-Valencia O, Garovic VD et~al.
+\newblock Why we need to report more than {\textquotesingle}data were analyzed
+  by t-tests or {ANOVA}{\textquotesingle}.
+\newblock \emph{{eLife}} 2018; 7.
+\newblock \doi{10.7554/elife.36163}.
+\newblock \urlprefix\url{https://doi.org/10.7554/elife.36163}.
+
+\end{thebibliography}
 

### 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/00-Full_document_OUP.Rmd
New folder for OUP Submission (Biostatistics)

--- 2021-06-10 Manuscripts/Manuscript by chapters -v2/00-Full_document.Rmd
@@
       caption: ["font={small}"] #size of the figure captions
-      float: null #allows for control of placement of figures
+      #float: null #allows for control of placement of figures
     includes:
-      in_header: preamble.sty #additional LaTeX formatting
+      in_header: preamble_OUP.sty #additional LaTeX formatting
   bookdown::word_document2:
@@
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 6 repeated measures for each group) to demonstrate how these models can be computationally implemented.
-We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the observations missing), allowing reliable inference from the data.
+We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented.
+We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing), allowing reliable inference from the data.
 To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research.
+
+# Keywords {-}
+
+cancer biology; tumor response; generalized additive models; simulation; R
 
@@
 
-\FloatBarrier
+<!-- \FloatBarrier -->
+
+# Supplementary Materials {-}
+
+An Appendix which contains all the code used to create this manuscript, along with a basic workflow to implement GAMs in R is available as Supplementary Material in PDF.
+A GitHub repository containing all the code used for this paper along with detailed instructions for its use is available at https://github.com/aimundo/GAMs-biomedical-research.
 
@@
 
-```{r child='07-Appendix.Rmd'}
+<!-- ```{r child='07-Appendix.Rmd'} -->
 
-```
+<!-- ``` -->

### 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/00-Full_document_OUP.tex
New folder for OUP Submission (Biostatistics)

--- 2021-06-22 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
@@
 \usepackage{graphicx}
-
+\usepackage[nomarkers,figuresonly]{endfloat}
 %set a box to put the ORCID logo
@@
   \href{https://orcid.org/0000-0002-6135-8191}{\usebox{\myorcidaffilbox}}}
+
+%to have page numbers with letters in the Appendix:
+% %solution from https://tex.stackexchange.com/questions/59572/custom-page-numbering-for-appendix
+\newcommand{\appendixpagenumbering}{
+  \break
+  \pagenumbering{arabic}
+  \renewcommand{\thepage}{\thesection-\arabic{page}}
+}
 
@@
 \usepackage[font={small}]{caption}
-\usepackage{float}
 \ifluatex
@@
 
+\hypertarget{keywords}{%
+\section*{Keywords}\label{keywords}}
+\addcontentsline{toc}{section}{Keywords}
+
+cancer biology; tumor response; generalized additive models; simulation; R
+
 \hypertarget{background}{%
@@
 Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]} .
+Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]}.
 There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]} and 3) can easily accommodate missing observations.
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/l-q-response-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_files/figure-latex/l-q-response-1}
 
@@
 In the following section, we present generalized additive models (GAMs) as a data-driven alternative method to analyze longitudinal non-linear data that overcomes the linearity assumption.
-
-\FloatBarrier
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_files/figure-latex/basis-plot-1}
 
@@
 
-\FloatBarrier
-
 \hypertarget{longitudinal-GAMs}{%
@@
 
-\passthrough{\lstinline!m1 <- gam(StO2\_sim \~ Group + s(Day, by=Group, k=5), method='REML, data = dat\_sim)!}
+\begin{lstlisting}[language=R]
+m1 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
+\end{lstlisting}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_files/figure-latex/plot-pairwise-comp-1}
 
@@
 
-\FloatBarrier
-
 \hypertarget{conclusion}{%
@@
 
-\FloatBarrier
+\hypertarget{declaration-of-conflicting-interests}{%
+\section{Declaration of Conflicting Interests}\label{declaration-of-conflicting-interests}}
+
+The Authors declare that there is no conflict of interest.
+
+\hypertarget{supplementary-materials}{%
+\section*{Supplementary Materials}\label{supplementary-materials}}
+\addcontentsline{toc}{section}{Supplementary Materials}
+
+An Appendix which contains all the code used to create this manuscript, along with a basic workflow to implement GAMs in R is available as Supplementary Material in PDF.
+A GitHub repository containing all the code used for this paper along with detailed instructions for its use is available at \url{https://github.com/aimundo/GAMs-biomedical-research}.
 
@@
 Butler, C.
-McLaren, W.P.
+McLaren, W.-P.
 Chen, B.
-Tromberg, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences of the United States of America.
+Tromberg, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences.
 108 (2011) 14626--14631.
@@
 Roblyer, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research.
-22 (2020) 1--10.
-\url{https://doi.org/doi:10.1186/s13058-020-01262-1}.}
+22 (2020).
+\url{https://doi.org/10.1186/s13058-020-01262-1}.}
 
@@
 Kalganova, Y.S.
-Lyubimtseva, V.I.
-Plekhanov, G.Y.
-Golubyatnikov, O.Y.
-Ilyinskaya, A.G.
-Orlova, P.V.
-Subochev, D.V.
-Safonov, N.M.
-Shakhova, A.V.
-Maslennikova, {Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy}, {Journal of Biomedical Optics}.
-{23} (2018).
-\url{https://doi.org/\%7B10.1117/1.JBO.23.9.091410\%7D}.}
+Lyubimtseva, Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy, Journal of Biomedical Optics.
+23 (2018) 1.
+\url{https://doi.org/10.1117/1.jbo.23.9.091410}.}
 
@@
 Flueraru, I.A.
-Vitkin, {Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography}, {Scientific Reports}.
-{8} (2018).
-\url{https://doi.org/\%7B10.1038/s41598-017-18635-w\%7D}.}
+Vitkin, Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography, Scientific Reports.
+8 (2018).
+\url{https://doi.org/10.1038/s41598-017-18635-w}.}
 
@@
 Lecorps, H.M.
-Colhoun, {Antidrug antibodies in atients treated with alirocumab}, {New England Journal of Medicine}.
-{376} (2017) 1589--1590.
-\url{https://doi.org/\%7B10.1056/NEJMc1616623\%7D}.}
+Colhoun, Antidrug antibodies in patients treated with alirocumab, New England Journal of Medicine.
+376 (2017) 1589--1590.
+\url{https://doi.org/10.1056/nejmc1616623}.}
 
@@
 Woessner, K.P.
-Quinn, {In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing}, {Communications Biology}.
-{1} (2018).
-\url{https://doi.org/\%7B10.1038/s42003-018-0206-4\%7D}.}
+Quinn, In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing, Communications Biology.
+1 (2018).
+\url{https://doi.org/10.1038/s42003-018-0206-4}.}
 
@@
 Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics.
-15 (2010).
+15 (2010) 011112.
 \url{https://doi.org/10.1117/1.3285584}.}
@@
 Cato, T.J.
-Muldoon, {Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy}, {Biomedical Optics Express}.
-{9} (2018) 2871--2886.
-\url{https://doi.org/\%7B10.1364/BOE.9.002871\%7D}.}
+Muldoon, Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy, Biomedical Optics Express.
+9 (2018) 2871.
+\url{https://doi.org/10.1364/boe.9.002871}.}
 
@@
 Wittich, J.D.
-Bearden III, R.C.
-Miller, {Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {{[}}alliance{]})}, {Supportive Care in Cancer}.
-{24} (2016) 3847--3855.
-\url{https://doi.org/\%7B10.1007/s00520-016-3213-3\%7D}.}
+Bearden, R.C.
+Miller, Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}), Supportive Care in Cancer.
+24 (2016) 3847--3855.
+\url{https://doi.org/10.1007/s00520-016-3213-3}.}
 
@@
 Roodenburg, J.A.
-Langendijk, {Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study}, {Oral Oncology}.
-{51} (2015) 548--555.
-\url{https://doi.org/\%7B10.1016/j.oraloncology.2015.01.016\%7D}.}
+Langendijk, Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study, Oral Oncology.
+51 (2015) 548--555.
+\url{https://doi.org/10.1016/j.oraloncology.2015.01.016}.}
 
@@
 Lodewyckx, G.J.
-Iverson, Bayesian versus frequentist inference, in: H.
-Hoijtink, I.
-Klugkist, P.A.
-Boelen (Eds.), Bayesian Evaluation of Informative Hypotheses, Springer New York, New York, NY, 2008: pp.
+Iverson, Bayesian versus frequentist inference, in: Bayesian Evaluation of Informative Hypotheses, Springer New York, 2008: pp.
 181--207.
@@
 \CSLLeftMargin{{[}13{]} }
-\CSLRightInline{R.
-Gueorguieva, J.H.
-Krystal, {Move over ANOVA - Progress in analyzing repeated-measures data and its reflection in papers published in the archives of general psychiatry}, Archives of General Psychiatry.
-61 (2004) 310--317.
-\url{https://doi.org/10.1001/archpsyc.61.3.310}.}
+\CSLRightInline{R. Gueorguieva, J.H. Krystal, Move over {ANOVA}, Archives of General Psychiatry. 61 (2004) 310. \url{https://doi.org/10.1001/archpsyc.61.3.310}.}
 
@@
 Schober, T.R.
-Vetter, Repeated measures designs and analysis of longitudinal data: If at first you do not succeed-try, try again, Anesthesia and Analgesia.
+Vetter, Repeated measures designs and analysis of longitudinal data, Anesthesia {\&} Analgesia.
 127 (2018) 569--575.
@@
 \CSLLeftMargin{{[}15{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science \& Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}
+\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science; Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}
 
@@
 Taper, S.R.
-Lele, {Errors in statistical inference under model misspecification: evidence, hypothesis testing, and AIC}, {Frontiers in Ecology and Evolution}.
-{7} (2019).
-\url{https://doi.org/\%7B10.3389/fevo.2019.00372\%7D}.}
+Lele, Errors in statistical inference under model misspecification: Evidence, hypothesis testing, and {AIC}, Frontiers in Ecology and Evolution.
+7 (2019).
+\url{https://doi.org/10.3389/fevo.2019.00372}.}
 
@@
 Tu, C.
-Feng, {The p-value and model specification in statistics}, {General Psychiatry}.
-{32} (2019).
-\url{https://doi.org/\%7B10.1136/gpsych-2019-100081\%7D}.}
+Feng, The p-value and model specification in statistics, General Psychiatry.
+32 (2019) e100081.
+\url{https://doi.org/10.1136/gpsych-2019-100081}.}
 
@@
 Vowler, G.B.
-Drummond, {The fickle \(p\) value generates irreproducible results}, {Nature Methods}.
-{12} (2015) 179--185.
-\url{https://doi.org/\%7B10.1038/nmeth.3288\%7D}.}
+Drummond, The fickle p value generates irreproducible results, Nature Methods.
+12 (2015) 179--185.
+\url{https://doi.org/10.1038/nmeth.3288}.}
 
@@
 \CSLRightInline{S.
-Nakagawa, {A farewell to Bonferroni: the problems of low statistical power and publication bias}, {Behavioral Ecology}.
-{15} (2004) 1044--1045.
-\url{https://doi.org/\%7B10.1093/beheco/arh107\%7D}.}
+Nakagawa, A farewell to bonferroni: The problems of low statistical power and publication bias, Behavioral Ecology.
+15 (2004) 1044--1045.
+\url{https://doi.org/10.1093/beheco/arh107}.}
 
@@
 Hill, M.
-Yajima, {Why we (usually) don't have to worry about multiple comparisons}, {Journal of Research on Educational Effectiveness}.
-{5} (2012) 189--211.
-\url{https://doi.org/\%7B10.1080/19345747.2011.618213\%7D}.}
+Yajima, Why we (usually) don{{}}t have to worry about multiple comparisons, Journal of Research on Educational Effectiveness.
+5 (2012) 189--211.
+\url{https://doi.org/10.1080/19345747.2011.618213}.}
 
@@
 \CSLLeftMargin{{[}24{]} }
-\CSLRightInline{C. Albers, {The problem with unadjusted multiple and sequential statistical testing}, {Nature Communications}. {10} (2019). \url{https://doi.org/\%7B10.1038/s41467-019-09941-0\%7D}.}
+\CSLRightInline{C. Albers, The problem with unadjusted multiple and sequential statistical testing, Nature Communications. 10 (2019). \url{https://doi.org/10.1038/s41467-019-09941-0}.}
 
@@
 \CSLLeftMargin{{[}25{]} }
-\CSLRightInline{Ugrinowitsch, C.
-and Fellingham, G.
-W.
-and Ricard, M.
-D., Limitations of ordinary least squares models in analyzing repeated measures data, Medicine and Science in Sports and Exercise.
-36 (2004) 2144--2148.
+\CSLRightInline{C.
+Ugrinowitsch, G.W.
+Fellingham, M.D.
+Ricard, Limitations of ordinary least squares models in analyzing repeated measures data, Medicine {\&} Science in Sports {\&} Exercise.
+(2004) 2144--2148.
 \url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.}
@@
 Huynh, L.S.
-Feldt, {Estimation of the Box correction for degrees of freedom from sample data in randomized block and split-plot designs}, Journal of Educational Statistics.
+Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics.
 1 (1976) 69--82.
@@
 Haverkamp, A.
-Beauducel, {Violation of the sphericity assumption and its effect on type-I error rates in repeated measures ANOVA and multi-level linear models (MLM)}, {Frontiers in Psychology}.
-{8} (2017).
-\url{https://doi.org/\%7B10.3389/fpsyg.2017.01841\%7D}.}
+Beauducel, Violation of the sphericity assumption and its effect on type-i error rates in repeated measures {ANOVA} and multi-level linear models ({MLM}), Frontiers in Psychology.
+8 (2017).
+\url{https://doi.org/10.3389/fpsyg.2017.01841}.}
 
@@
 \CSLLeftMargin{{[}29{]} }
-\CSLRightInline{H.
+\CSLRightInline{H.J.
 Keselman, J.
-Algina, R.
-Kowalchuk, {The analysis of repeated measures designs: A review}, {British Journal of Mathematica \& Statistical Psychology}.
-{54} (2001) 1--20.
-\url{https://doi.org/\%7B10.1348/000711001159357\%7D}.}
+Algina, R.K.
+Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematical and Statistical Psychology.
+54 (2001) 1--20.
+\url{https://doi.org/10.1348/000711001159357}.}
 
@@
 \CSLLeftMargin{{[}30{]} }
-\CSLRightInline{Jaykaran.
+\CSLRightInline{J.
 Charan, N.
-Kantharia, {How to calculate sample size in animal studies?}, Journal of Pharmacology and Pharmacotherapeutics.
-4 (2013) 303--306.
-\url{https://doi.org/10.4103/0976-500X.119726}.}
+Kantharia, How to calculate sample size in animal studies?, Journal of Pharmacology and Pharmacotherapeutics.
+4 (2013) 303.
+\url{https://doi.org/10.4103/0976-500x.119726}.}
 
@@
 Scheepers, H.J.
-Tily, {Random effects structure for confirmatory hypothesis testing: Keep it maximal}, {Journal of Memory and Language}.
-{68} (2013) 255--278.
-\url{https://doi.org/\%7B10.1016/j.jml.2012.11.001\%7D}.}
+Tily, Random effects structure for confirmatory hypothesis testing: Keep it maximal, Journal of Memory and Language.
+68 (2013) 255--278.
+\url{https://doi.org/10.1016/j.jml.2012.11.001}.}
 
@@
 \CSLRightInline{N.L.
-Rose, H.D.
+Rose, H.
 Yang, S.D.
 Turner, G.L.
-Simpson, {An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to Scotland, UK}, Geochimica Et Cosmochimica Acta.
+Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, {UK}, Geochimica Et Cosmochimica Acta.
 82 (2012) 113--135.
@@
 Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, {PeerJ}.
-7 (2019).
+7 (2019) e6876.
 \url{https://doi.org/10.7717/peerj.6876}.}
@@
 Wang, G.
-Song, {Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality}, {BMC Medical Research Methodology}.
-{12} (2012).
-\url{https://doi.org/\%7B10.1186/1471-2288-12-165\%7D}.}
+Song, Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality, {BMC} Medical Research Methodology.
+12 (2012).
+\url{https://doi.org/10.1186/1471-2288-12-165}.}
 
@@
 \CSLLeftMargin{{[}36{]} }
-\CSLRightInline{N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. (1998) 596--627.}
+\CSLRightInline{N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. 42 (1998) 596. \url{https://doi.org/10.2307/2991772}.}
 
@@
 \CSLLeftMargin{{[}37{]} }
-\CSLRightInline{S.N. Wood, {Generalized additive models: An introduction with R, Second Edition}, CRC Press LLC, Philadelphia, PA, 2017.}
+\CSLRightInline{S.N. Wood, Generalized additive models, Chapman; Hall/{CRC}, 2017. \url{https://doi.org/10.1201/9781315370279}.}
 
@@
 Pya, B.
-Saefken, Smoothing parameter and model selection for general smooth models, {Journal of the American Statistical Association}.
-{111} (2016) 1548--1563.
-\url{https://doi.org/\%7B10.1080/01621459.2016.1180986\%7D}.}
+Säfken, Smoothing parameter and model selection for general smooth models, Journal of the American Statistical Association.
+111 (2016) 1548--1563.
+\url{https://doi.org/10.1080/01621459.2016.1180986}.}
 
@@
 Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics.
-1 (1996) 205--230.
-\url{http://www.jstor.org/stable/1400366}.}
+1 (1996) 205.
+\url{https://doi.org/10.2307/1400366}.}
 
@@
 \CSLLeftMargin{{[}42{]} }
-\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://books.google.com/books?id=MQ/_bvWDPsEAC}.}
+\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://doi.org/10.1007/0-387-28314-5}.}
 
@@
 \CSLLeftMargin{{[}44{]} }
-\CSLRightInline{S.E.
-Maxwell, H.D.
-Delaney, K.
-Kelley, Designing experiments and analyzing data: A model comparison perspective, third edition, Taylor \& Francis, 2017.
-\url{https://books.google.com/books?id=NmFQDwAAQBAJ}.}
+\CSLRightInline{S.E. Maxwell, H.D. Delaney, K. Kelley, Designing experiments and analyzing data, Routledge, 2017. \url{https://doi.org/10.4324/9781315642956}.}
 
@@
 \CSLLeftMargin{{[}45{]} }
-\CSLRightInline{G.
-Molenberghs, H.
-Thijs, I.
-Jansen, C.
-Beunckens, M.
-Kenward, C.
-Mallinckrodt, R.
-Carroll, {Analyzing incomplete longitudinal clinical trial data}, {Biostatistics}.
-{5} (2004) 445--464.
-\url{https://doi.org/\%7B10.1093/biostatistics/kxh001\%7D}.}
+\CSLRightInline{G. Molenberghs, Analyzing incomplete longitudinal clinical trial data, Biostatistics. 5 (2004) 445--464. \url{https://doi.org/10.1093/biostatistics/kxh001}.}
 
@@
 Pieper, V.
-Hasselblad, {Can one assess whether missing data are missing at random in medical studies?}, {Statistical Methods in Medical Research}.
-{15} (2006) 213--234.
-\url{https://doi.org/\%7B10.1191/0962280206sm448oa\%7D}.}
+Hasselblad, Can one assess whether missing data are missing at random in medical studies?, Statistical Methods in Medical Research.
+15 (2006) 213--234.
+\url{https://doi.org/10.1191/0962280206sm448oa}.}
 
@@
 Mazumdar, S.G.
-Memtsoudis, Beyond repeated-measures analysis of variance advanced statistical methods for the analysis of longitudinal data in anesthesia research, {Regional Anesthesia and Pain Medicine}.
-{37} (2012) 99--105.
-\url{https://doi.org/\%7B10.1097/AAP.0b013e31823ebc74\%7D}.}
+Memtsoudis, Beyond repeated-measures analysis of variance, Regional Anesthesia and Pain Medicine.
+37 (2012) 99--105.
+\url{https://doi.org/10.1097/aap.0b013e31823ebc74}.}
 
@@
 \CSLLeftMargin{{[}50{]} }
-\CSLRightInline{J.A.
-Nelder, R.W.M.
-Wedderburn, Generalized linear models, Journal of the Royal Statistical Society.
-Series A (General).
-135 (1972) 370--384.
-\url{http://www.jstor.org/stable/2344614}.}
+\CSLRightInline{J.A. Nelder, R.W.M. Wedderburn, Generalized linear models, Journal of the Royal Statistical Society. Series A (General). 135 (1972) 370. \url{https://doi.org/10.2307/2344614}.}
 
@@
 Williams, M.B.
-Hooten, {The basis function approach for modeling autocorrelation in ecological data}, {Ecology}.
-{98} (2017) 632--646.
-\url{https://doi.org/\%7B10.1002/ecy.1674\%7D}.}
+Hooten, The basis function approach for modeling autocorrelation in ecological data, Ecology.
+98 (2017) 632--646.
+\url{https://doi.org/10.1002/ecy.1674}.}
 
@@
 \CSLLeftMargin{{[}54{]} }
-\CSLRightInline{R. McElreath, {Statistical rethinking: A Bayesian course with examples in R and Stan}, {Chapman and Hall/CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
+\CSLRightInline{R. McElreath, Statistical rethinking, Chapman; Hall/{CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
 
@@
 \CSLLeftMargin{{[}56{]} }
-\CSLRightInline{N.M. Laird, J.H. Ware, Random-effects models for longitudinal data, Biometrics. 38 (1982) 963--974. \url{http://www.jstor.org/stable/2529876}.}
+\CSLRightInline{N.M. Laird, J.H. Ware, Random-effects models for longitudinal data, Biometrics. 38 (1982) 963. \url{https://doi.org/10.2307/2529876}.}
 
@@
 Marra, S.N.
-Wood, Coverage properties of confidence intervals for generalized additive model components, {Scandinavian Journal of Statistics}.
-{39} (2012) 53--74.
-\url{https://doi.org/\%7B10.1111/j.1467-9469.2011.00760.x\%7D}.}
+Wood, Coverage properties of confidence intervals for generalized additive model components, Scandinavian Journal of Statistics.
+39 (2012) 53--74.
+\url{https://doi.org/10.1111/j.1467-9469.2011.00760.x}.}
 
@@
 \CSLLeftMargin{{[}59{]} }
-\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, {Semiparametric Regression with R}, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
+\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with r, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
 
@@
 Lang, D.G.
-Altman, {Basic statistical reporting for articles published in Biomedical Journals: The ``Statistical Analyses and Methods in the Published Literature{''} or the SAMPL Guidelines}, INTERNATIONAL JOURNAL OF NURSING STUDIES.
-{52} (2015) 5--9.
-\url{https://doi.org/\%7B10.1016/j.ijnurstu.2014.09.006\%7D}.}
+Altman, Basic statistical reporting for articles published in biomedical journals: The {``}statistical analyses and methods in the published literature{''} or the {SAMPL} guidelines, International Journal of Nursing Studies.
+52 (2015) 5--9.
+\url{https://doi.org/10.1016/j.ijnurstu.2014.09.006}.}
 
@@
 Hastie, R.
-Tibshirani, Generalized additive models for medical research, {Statistical Methods in Medical Research}.
+Tibshirani, Generalized additive models for medical research, Statistical Methods in Medical Research.
 4 (1995) 187--196.
@@
 \CSLLeftMargin{{[}62{]} }
-\CSLRightInline{C.G.
-Begley, J.P.A.
-Ioannidis, {Reproducibility in Science Improving the Standard for Basic and Preclinical Research}, {Circulation Research}.
-{116} (2015) 116--126.
-\url{https://doi.org/\%7B10.1161/CIRCRESAHA.114.303819\%7D}.}
+\CSLRightInline{C.G. Begley, J.P.A. Ioannidis, Reproducibility in science, Circulation Research. 116 (2015) 116--126. \url{https://doi.org/10.1161/circresaha.114.303819}.}
 
@@
 Milic, S.J.
-Winham, {Meta-Research: Why we need to report more than 'Data were Analyzed by t-tests or ANOVA'}, Elife.
-7 (2018) e36163.
-\url{https://doi.org/10.7554/eLife.36163}.}
+Winham, Why we need to report more than {{}}data were analyzed by t-tests or {ANOVA}{{}}, {eLife}.
+7 (2018).
+\url{https://doi.org/10.7554/elife.36163}.}
 
@@
 
-\hypertarget{appendix-appendix}{%
-\appendix}
-
-\hypertarget{code-for-manuscript-data}{%
-\section{Code for Manuscript data}\label{code-for-manuscript-data}}
-
-This section presents the code used to generate figures, models and simulated data from Sections 3 and 4 from the main manuscript.
-
-\hypertarget{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}{%
-\subsection{Compound symmetry and independent errors in linear and quadratic responses}\label{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}}
-
-This section simulated linear and quadratic data in the same manner as in Section \ref{simulation}.
-The linear simulations using Figure \ref{fig:linear-cases-Appendix} show in panels A and D the simulated mean responses and individual data points.
-Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-The following code produces a more comprehensive exploration of Figure \ref{fig:l-q-response} in the main manuscript.
-
-\begin{lstlisting}[language=R]
-##########Section for calculations###########
-
-## Example with linear response
-
-#This function simulates data using a linear or quadratic mean response and each with correlated
-#or uncorrelated errors. Each group has a different slope/concavity.
-example <- function(n_time = 6, #number of time points
-                    fun_type = "linear", #type of response
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  x <- seq(1,6, length.out = n_time)
-
-  #Create mean response matrix: linear or quadratic
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # quadratic response (non-linear)
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-
-  #create an array where individual observations per each time point for each group are to be stored. Currently using 10 observations per timepoint
-  y <- array(0, dim = c(length(x), 2, 10))
-
-  #Create array to store the "errors" for each group at each timepoint. The "errors" are the 
-  #between-group variability in the response.
-  errors <- array(0, dim = c(length(x), 2, 10))
-  #create an array where 10 observations per each time point for each group are to be stored
-
-  #The following cycles create independent or correlated responses.
-To each value of mu (mean response per group) a randomly generated error (correlated or uncorrelated) is added and thus the individual response is created.
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors: variance covariance matrix
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-#labeling y and errors  
-  dimnames(y) <- list(time = x, 
-                      treatment = 1:2, 
-                      subject = 1:10)
-
-  dimnames(errors) <- list(time = x, 
-                           treatment = 1:2, 
-                           subject = 1:10)
-
-  #labeling the mean response
-  dimnames(mu) <- list(time = x, 
-                       treatment = 1:2)
-
-  #convert y, mu and errors to  dataframes with time, treatment and subject columns
-  dat <- as.data.frame.table(y, 
-                             responseName = "y")
-  dat_errors <- as.data.frame.table(errors, 
-                                    responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, 
-                                responseName = "mu")
-
-  #join the dataframes to show mean response and errors per subject
-  dat <- left_join(dat, dat_errors, 
-                   by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, 
-                   by = c("time", "treatment"))
-  #add time
-  dat$time <- as.numeric(as.character(dat$time))
-  #label subjects per group
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, 
-                                  treatment, 
-                                  sep = "-")))
-
-  ## repeated measures ANOVA
-
-  fit_anova <- lm(y ~ time + treatment + time * treatment, data = dat)
-
-#LMEM: time and treatment interaction model, compound symmetry 
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  #create a prediction frame where the model can be used for plotting purposes
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  #add model predictions to the dataframe that has the simulated data
-  dat$pred_anova <- predict(fit_anova)
-  dat$pred_lmem <- predict(fit_lme)
-
-  #return everything in a list
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_anova=fit_anova,
-    fit_lme = fit_lme
-  ))
-}
-##################Section for plotting#################################
-#######################################################################
-#This function will create the plots for either a "linear" or "quadratic" response
-
-plot_example <- function(sim_dat) {
-  ## Plot the simulated data (scatterplot)
-
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = treatment, 
-               color = treatment)
-           ) +
-    geom_point(show.legend=FALSE) +
-    labs(y='response')+
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment),
-              show.legend=FALSE) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-  #plot the simulated data with trajectories per each subject
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = subject, 
-               color = treatment)
-           ) +
-    geom_line(aes(size = "Subjects"),
-              show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment,
-                  size = "Simulated Truth"), 
-              lty = 1,show.legend = FALSE) +
-    labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                face = "bold"),
-     text=element_text(size=30))+
-    thm
-
-  #plot the errors
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = errors,
-               group = subject, 
-               color = treatment)) +
-    geom_line(show.legend=FALSE) +
-     labs(y='errors')+
-     theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-   #plot the model predictions for rm-ANOVA
-  p4 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point(show.legend=FALSE)+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_anova), 
-                  group = subject, size = "Subjects"),show.legend = FALSE) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_anova, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population"),
-              show.legend=FALSE) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-   #plot the LMEM predictions
-  p5 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point()+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), 
-                  group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_lme, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population")) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-  return((p1+p3+p2+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-
-#Store each plot in a separate object
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent")) 
-\end{lstlisting}
-
-\begin{figure}[H]
-\includegraphics{00-Full_document_files/figure-latex/linear-cases-Appendix-1} \caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are able to capture the trend of the data.}\label{fig:linear-cases-Appendix}
-\end{figure}
-
-For the quadratic response case, Figure \ref{fig:quadratic-cases-Appendix} shows the simulated responses using compound symmetry and independent errors.
-
-\begin{figure}[H]
-\includegraphics{00-Full_document_files/figure-latex/quadratic-cases-Appendix-1} \caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are not able to capture the changes in each group over time.}\label{fig:quadratic-cases-Appendix}
-\end{figure}
-
-\hypertarget{basis-functions-and-gams}{%
-\subsection{Basis functions and GAMs}\label{basis-functions-and-gams}}
-
-This code produces Figure \ref{fig:basis-plot} from the main manuscript.
-Briefly, a non-linear (quadratic) response is simulated a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
-The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
-
-\begin{lstlisting}[language=R]
-#generate the response: the same initial procedure from the previous section to simulate
-#the response
-set.seed(1)
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-
- #label each table
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
-
- #Convert to dataframes with subject, time and group columns
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
-
- #label subject per group
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
- #extract  "Group 1" to fit the GAM
-  dat<-subset(dat,treatment==1)
- #keep just the response and timepoint columns
-   dat<-dat[,c('y','time')]
-
-   #GAM model of time, 5 knots
-gm<-gam(y~s(time,k=5),data=dat)
-
-#model_matrix (also known as) 'design matrix'
-#will contain the smooths used to create  model 'gm'
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be 
-#multiplied by the coefficients
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final smooth by aggregating the basis functions
-
-smooth<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(smooth=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,
-            aes(x=time,
-                y=value_orig,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(size=sz,
-            show.legend=FALSE)+
-  geom_point(size=sz+1,
-             show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,
-             labeller = as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#penalized basis
-p12<-ggplot(basis_plot,
-            aes(x=time,
-                y=mod_val,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(show.legend = FALSE,
-            size=sz)+
-  geom_point(show.legend = FALSE,
-             size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,
-             labeller=as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#heatmap of the coefficients
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,
-            aes(x=Basis,
-                y=Basis))+
-  geom_tile(aes(fill = cof), 
-            colour = "black") +
-    scale_fill_gradient(low = "white",
-                        high = "#B50A2AFF")+ #color picked from KikiMedium
-  labs(x='Basis',
-       y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),
-            size=7,
-            show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,
-            aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  labs(y='Simulated \n response')+
-  geom_line(data=smooth,
-            aes(x=time,
-                y=smooth),
-            color="#6C581DFF",
-            size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-appendix-1}
-
-}
-
-\caption{Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-B: Matrix of basis function weights.
-Each basis function is multiplied by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
-Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot-appendix}
-\end{figure}
-
-\hypertarget{tumor-data-simulation}{%
-\section{Longitudinal biomedical data simulation and GAMs}\label{tumor-data-simulation}}
-
-This section describes how to fit GAMs to longitudinal data using simulated data.
-First, data is simulated according to Section \ref{longitudinal-GAMs}, where reported data of oxygen saturation (\(\mbox{StO}_2\)) in tumors under either chemotherapy or saline control is used as a starting point to generate individual responses in each group.
-
-\begin{lstlisting}[language=R]
-set.seed(1)
-#Dataframe that contains the original reported trends
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
-
-## plot the mean response
-f1<-ggplot(dat, 
-           aes(x = Day, 
-               y = StO2, 
-               color = Group)) +
-    geom_line(size=1,
-              show.legend = FALSE)+
-    geom_point(show.legend = FALSE,
-               size=1.5,
-               alpha=0.5)+
-  labs(y=expression(paste(StO[2],
-                          ' (real)')))+
-  theme_classic()+
-  thm+
-    scale_x_continuous(breaks=c(0,5,10))+
-    scale_y_continuous(breaks=c(0,40))+
-  plot_layout(tag_level = 'new')+
-  theme(
-    plot.background = element_rect(fill = "transparent", 
-                                   color = NA),
-    axis.text=element_text(size=14)
-  )
-
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-n <- 10 #number of observations
-sd <- 10 #approximate sd from paper
-df <- 6
-dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, 
-           aes(x = Day, 
-               y = StO2_sim, 
-               color = Group)) +
-    geom_point(show.legend=FALSE,
-               size=1.5,
-               alpha=0.5)+
-    stat_summary(aes(y = StO2_sim,
-                     group=Group), 
-                 fun=mean, geom="line",
-                 size=1,
-                 show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],
-                         '(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-  thm+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
-\end{lstlisting}
-
-\hypertarget{workflow}{%
-\subsection{A basic Workflow for GAMs}\label{workflow}}
-
-This section shows a basic workflow to fit a series of increasingly complex GAMs to the simulated data from the previous section.
-Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
-
-\hypertarget{first-model}{%
-\subsubsection{First model}\label{first-model}}
-
-The first model fitted to the data is one that only accounts for different smooths by day.
-The model syntax specifies that \passthrough{\lstinline!gam\_00!} is the object that will contain all the model information, and that the model attempts to explain changes in \passthrough{\lstinline!StO2\_sim!} (simulated \(\mbox{StO}_2\)) using a smooth per \passthrough{\lstinline!Day!}.
-The model will use 5 knots (\passthrough{\lstinline!k=5!}) for the smooth.
-The smooth is constructed by default using thin plate regression splines.
-The smoothing parameter estimation method used is the restricted maximum likelihood (\passthrough{\lstinline!REML!}).
-
-\begin{lstlisting}[language=R]
-gam_00<-gam(StO2_sim ~ s(Day, k = 5),
-            method='REML',
-            data  = dat_sim)
-\end{lstlisting}
-
-To obtain model diagnostics, two methodologies are to be used: 1) graphical diagnostics, and 2) a model check.
-In the first case, the functions \passthrough{\lstinline!appraise!} and \passthrough{\lstinline!draw!} from the package \emph{gratia} can be used to obtain a single output with all the graphical diagnostics.
-For model check, the functions \passthrough{\lstinline!gam.check!} and \passthrough{\lstinline!summary!} from \emph{mgcv} provide detailed information about the model fit and its parameters.
-
-\hypertarget{graphical-diagnostics}{%
-\paragraph{Graphical diagnostics}\label{graphical-diagnostics}}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/first-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the first GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:first-GAM-diag}
-\end{figure}
-
-From the output of the function \passthrough{\lstinline!appraise!} in Figure \ref{fig:first-GAM-diag}, the major indicators of concern about the model are the QQ plot of residuals and the histogram of residuals.
-The QQ plot shows that the errors are not reasonably located along the 45\(^{\circ}\) line (which indicates normality), as there are multiple points that deviate from the trend, specially in the tails.
-The histogram also shows that the variation (residuals) is not following the assumption of a normal distribution.
-
-The \passthrough{\lstinline!draw!} function permits to plot the smooths as \passthrough{\lstinline!ggplot2!} objects, which eases subsequent manipulation, if desired.
-Because model \passthrough{\lstinline!gam\_00!} specifies only one smooth for the time covariate (Day), the plot only contains only one smooth.
-Note that the smooth shows an almost linear profile.
-
-\hypertarget{model-check}{%
-\paragraph{Model check}\label{model-check}}
-
-\begin{lstlisting}[language=R]
-#need to add figure number and caption
-gam.check(gam_00)
-\end{lstlisting}
-
-\includegraphics{00-Full_document_files/figure-latex/first-GAM-check-1}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 5 iterations.
-## Gradient range [-0.0003727881,-6.621452e-07]
-## (score 444.0118 & scale 450.6638).
-## Hessian positive definite, eigenvalue range [0.3881695,49.00676].
-## Model rank =  5 / 5 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##          k'  edf k-index p-value    
-## s(Day) 4.00 2.11    0.36  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-summary(gam_00)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ s(Day, k = 5)
-## 
-## Parametric coefficients:
-##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   22.967      2.123   10.82   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##          edf Ref.df     F  p-value    
-## s(Day) 2.114  2.565 7.633 0.000517 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.153   Deviance explained = 17.2%
-## -REML = 444.01  Scale est. = 450.66    n = 100
-\end{lstlisting}
-
-Special attention must be paid to the `k-index' from \passthrough{\lstinline!gam.check!}.
-This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
-If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless1).
-From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data.
-The \passthrough{\lstinline!edf!} (effective degrees of freedom) is an indicator of the complexity of the smooth.
-Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
-
-From the \passthrough{\lstinline!summary!} function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
-The link function is `identity' as the model does not make any transformation on the predictors.
-The `significance of smooth terms' \emph{p-value} indicates if each smooth is adding significance to the model.
-Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \(\approx\) 17\%.
-
-\hypertarget{second-model}{%
-\subsubsection{Second model}\label{second-model}}
-
-The major flaw of \passthrough{\lstinline!gam\_00!} is that this model is not taking into account the fact that the data is nested in groups.
-The next iteration is a model where a different smooth of time (Day) is assigned for each group using \passthrough{\lstinline!by=Group!} in the model syntax.
-
-\begin{lstlisting}[language=R]
-gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5),
-            method='REML',
-            data  = dat_sim)
-
-gam.check(gam_01)
-\end{lstlisting}
-
-\includegraphics{00-Full_document_files/figure-latex/second-GAM-1}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 7 iterations.
-## Gradient range [-5.51754e-05,2.671715e-06]
-## (score 423.3916 & scale 280.8777).
-## Hessian positive definite, eigenvalue range [0.3162258,48.5557].
-## Model rank =  9 / 9 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                         k'  edf k-index p-value    
-## s(Day):GroupControl   4.00 3.39    0.43  <2e-16 ***
-## s(Day):GroupTreatment 4.00 3.23    0.43  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-summary(gam_01)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ s(Day, by = Group, k = 5)
-## 
-## Parametric coefficients:
-##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   22.967      1.676    13.7   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                         edf Ref.df      F p-value    
-## s(Day):GroupControl   3.392  3.794  3.817  0.0304 *  
-## s(Day):GroupTreatment 3.229  3.682 21.174  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.472   Deviance explained = 50.8%
-## -REML = 423.39  Scale est. = 280.88    n = 100
-\end{lstlisting}
-
-Diagnostics for this model indicate that the k-index is still below 1 (0.43 from \passthrough{\lstinline!gam.check!}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
-Moreover, the smooths (plotted via the \passthrough{\lstinline!draw()!} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \(\approx\) 51\%.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/second-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the second GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:second-GAM-diag}
-\end{figure}
-
-\hypertarget{third-model}{%
-\subsubsection{Third model}\label{third-model}}
-
-Model \passthrough{\lstinline!gam\_00!} was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted, the model diagnostics, and, the low variance explained by the model.
-On the other hand, \passthrough{\lstinline!gam\_01!} takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \ref{longitudinal-GAMs}, in order to differentiate between each group a parametric term needs to be added to the model for the interaction of \emph{Day} and \emph{Group}.
-
-\begin{lstlisting}[language=R]
-#GAM for StO2
-
-m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
-            method='REML',
-            data  = dat_sim)
-
-gam.check(m1)
-\end{lstlisting}
-
-\includegraphics{00-Full_document_files/figure-latex/final-model-Appendix-1}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 10 iterations.
-## Gradient range [-8.164307e-08,1.500338e-08]
-## (score 355.8554 & scale 64.53344).
-## Hessian positive definite, eigenvalue range [1.174841,48.08834].
-## Model rank =  10 / 10 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                         k'  edf k-index p-value
-## s(Day):GroupControl   4.00 3.87    1.02    0.59
-## s(Day):GroupTreatment 4.00 3.88    1.02    0.54
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-summary(m1)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ Group + s(Day, by = Group, k = 5)
-## 
-## Parametric coefficients:
-##                Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)       9.084      1.136   7.996 4.09e-12 ***
-## GroupTreatment   27.766      1.607  17.282  < 2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                         edf Ref.df     F p-value    
-## s(Day):GroupControl   3.873  3.990 17.57  <2e-16 ***
-## s(Day):GroupTreatment 3.879  3.991 89.33  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.879   Deviance explained = 88.9%
-## -REML = 355.86  Scale est. = 64.533    n = 100
-\end{lstlisting}
-
-The resulting model is \passthrough{\lstinline!m1!}, which is the model fitted in the main manuscript.
-By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\(\approx\) 1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 89\% of the variance in the data.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/final-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the final GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooths for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:final-GAM-diag}
-\end{figure}
-
-\hypertarget{comparing-models-via-aic}{%
-\subsubsection{Comparing models via AIC}\label{comparing-models-via-aic}}
-
-One final comparison that can be made for model selection involves the use of the Aikake Information Criterion (AIC).
-This metric is used to estimate information loss, which we want to minimize with an appropriate model.
-Therefore, when 2 or more models are compared, the model with lower AIC is preferred.
-In R, the comparison is done using the \passthrough{\lstinline!AIC!} function.
-
-\begin{lstlisting}[language=R]
-AIC(gam_00,gam_01,m1)
-\end{lstlisting}
-
-\begin{lstlisting}
-##               df      AIC
-## gam_00  4.564893 900.8257
-## gam_01  9.476137 858.6051
-## m1     10.980983 712.2067
-\end{lstlisting}
-
-The output in this case is expected: model \passthrough{\lstinline!m1!} has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
-The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
-
-\hypertarget{pairwise-comparisons-of-smooth-confidence-intervals}{%
-\paragraph{Pairwise comparisons of smooth confidence intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
-
-The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \ref{GAM-significance}.
-In this case, the ``design matrix'' is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
-Briefly, the ``design matrix'' (also known as the ``Xp matrix'') from the selected model (\passthrough{\lstinline!m1!}) is used to calculate a 95\% confidence interval of the difference between the smooth terms for each group.
-This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
-\textbf{All pairwise comparisons in this paper have been centered at the response scale to ease interpretation }.
-
-\begin{lstlisting}[language=R]
-##Pairwise comparisons
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-##matrix that contains the basis functions evaluated at the points in pdat
-    xp <- predict(m1, newdata = pdat, type = 'lpmatrix')
-
-#Find columns in xp where the name contains "Control"
-    c1 <- grepl('Control', colnames(xp))
-
-#Find columns in xp where the name contains 'Treatment'
-    c2 <- grepl('Treatment', colnames(xp))
-
-#Find rows in pdat that correspond to either 'Control' or 'Treatment'
-    r1 <- with(pdat, Group == 'Control')
-    r2 <- with(pdat, Group == 'Treatment')
-
-# In xp: find the rows that correspond to Control or Treatment, those that do not match will be
-    #set to zero. Then, substract the values from the rows corresponding to 'Control' from those that correspond
-    #to 'Treatment'
-    X <- xp[r1, ] - xp[r2, ]
-
-    ## remove columns that do not contain name 'Control' or 'Treatment'
-    X[, ! (c1 | c2)] <- 0
-    ## zero out the parametric cols, those that do not contain in the characters 's('
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
-
-    #Multiply matrix by model coefficients. X has (p,n) (rows, columns) and the coefficient matrix has
-    #dimensions (n,1). The resulting matrix has dimensions (p,1)
-    dif <- X %*% coef(m1)
-
-    #comp<-test %*% coef(gam1)[3:10]
-
-#Calculate standard error for the computed differences using the variance-covariance matrix
-    #of the model
-    se <- sqrt(rowSums((X %*% vcov(m1, unconditional = FALSE)) * X))
-    crit <- qt(0.05/2, df.residual(m1), lower.tail = FALSE)
-    #upper  limits
-    upr <- dif + (crit * se)
-    #lower limits
-    lwr <- dif - (crit * se)
-    #put all components in a dataframe for plotting
-    comp1<-data.frame(pair = paste('Control', 'Treatment', sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-
-#add time point sequence
-comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1))
-
-#use function from the pairwise comparison plot in the manuscript to get the shaded regions
-
-    my_list<-pairwise_limits(comp_StO2)
-  rib_col<-'#EDD03AFF' #color for the ribbon
-#plot the difference
-c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
-  #shaded region
-  annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill='#30123BFF',
-                alpha = 0.5,
-                ) +
-  annotate("text",
-             x=1.5,
-             y=-10,
-             label="Control",size=10
-           )+
-  #shaded region  
-  annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
-             alpha = 0.5
-           ) +
-  annotate("text",
-             x=6,
-             y=-10,
-             label="Treatment",
-             size=10
-           )+
-  #ribbon for difference confidence interval  
-  geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(color='black',size=1) +
-    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/pairwise-comp-workflow-fig-1}
-
-}
-
-\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!m1!} using a 95\% confidence interval for the difference between smooths.
-The comparison is centered at the response scale.
-Shaded regions indicate time intervals where each treatment group has a non-zero effect.}\label{fig:pairwise-comp-workflow-fig}
-\end{figure}
-
-Of notice, a convenient wrapper for the function described above exists in the package \passthrough{\lstinline!gratia!}.
-In this package, \passthrough{\lstinline!difference\_smooths!} is a function that makes the comparisons and produces Figure \ref{fig:pairwise-comp-workflow-fig} when is used on a fitted model.
-The function syntax and an example can be found at:
-
-\url{https://cran.r-project.org/web/packages/gratia/gratia.pdf}
-
-Keep in mind that this function \textbf{does not} center the pairwise comparison at the response scale, so it has to be shifted in order to be compared to the raw data.
-
-\hypertarget{gam-and-linear-model-plots-and-missing-data}{%
-\section{GAM and Linear model plots and Missing data}\label{gam-and-linear-model-plots-and-missing-data}}
-
-This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\passthrough{\lstinline!m1!}), linear model and GAM on data with missing observations are presented.
-Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inset are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
-
-\hypertarget{gam-and-linear-model-plots}{%
-\subsection{GAM and Linear model plots}\label{gam-and-linear-model-plots}}
-
-This code chunk creates panels B and D in Figure \ref{fig:sim-smooth-plot}.
-Note that this code uses the final GAM from the previous section (\passthrough{\lstinline!m1!}), so the simulated data and the model should be generated before running this section.
-
-\begin{lstlisting}[language=R]
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-#creates a dataframe using the length of the covariates for the GAM
-gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
-                         subject=factor(rep(1:10)))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#adds the predictions to the grid and creates a confidence interval for GAM
-gam_predict<-gam_predict%>%
-    mutate(fit = predict(m1,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(m1, gam_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=gam_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'complete')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-
-#posthoc comparisons for the linear model
-#library(multcomp)
-
-#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
-\end{lstlisting}
-
-\hypertarget{working-with-missing-data-in-gams}{%
-\subsection{Working with Missing data in GAMs}\label{working-with-missing-data-in-gams}}
-
-This code chunk first randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data.
-Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
-
-\begin{lstlisting}[language=R]
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
-
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'missing')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-Appendix-1}
-
-}
-
-\caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
-\end{figure}
-
-\hypertarget{pairwise-comparisons-in-gams-full-and-missing-data-cases}{%
-\subsection{Pairwise comparisons in GAMs: full and missing data cases}\label{pairwise-comparisons-in-gams-full-and-missing-data-cases}}
-
-The next code chunk reproduces Figure \ref{fig:plot-pairwise-comp}. Here pairwise comparisons are made for the full and missing datasets.
-
-\begin{lstlisting}[language=R]
-##Pairwise comparisons
-
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-#this function takes the model, grid and groups to be compared using the lpmatrix
-#originally developed by G. Simpson:
-#https://fromthebottomoftheheap.net/2017/10/10/difference-splines-i/
-
-smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    #Find columns in xp where the name contains "Control" and "Treatment"
-    col1 <- grepl(g1, colnames(xp))
-    col2 <- grepl(g2, colnames(xp))
-    #Find rows in xp that correspond to each treatment
-    row1 <- with(newdata, Group == g1)
-    row2 <- with(newdata, Group == g2)
-    ## difference rows of xp for data from comparison
-    X <- xp[row1, ] - xp[row2, ]
-    ## zero out cols of X related to splines for other lochs
-    X[, ! (col1 | col2)] <- 0
-
-    ## zero out the parametric cols
-    #This line has been commented to keep the comparison at the response level,
-    #otherwise it gives the marginal change between smooths
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    #get standard error, critical value and boundaries
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(g1, g2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-}
-
-#use the function to calculate the difference in smooths
-comp1<-smooth_diff(m1,pdat,'Control','Treatment')
-
-#Create a dataframe with time, comparisons and labels for regions where difference exists
-comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
-
-pairwise_limits<-function(dataframe){
-    #extract values where the lower limit of the ribbon is greater than zero
-    #this is the region where the control group effect is greater
-    v1<-dataframe%>%
-        filter(lower>0)%>%
-        select(Day)
-    #get day  initial value
-    init1=v1$Day[[1]]
-    #get day final value
-    final1=v1$Day[[nrow(v1)]]
-
-    #extract values where the value of the upper limit of the ribbon is lower than zero
-    #this corresponds to the region where the treatment group effect is greater
-    v2<-comp_StO2_full%>%
-        filter(upper<0)%>%
-        select(Day)
-
-    init2=v2$Day[[1]]
-    final2=v2$Day[[nrow(v2)]]
-    #store values
-    my_list<-list(init1=init1,
-                  final1=final1,
-                  init2=init2,
-                  final2=final2)
-return(my_list)
-}
-
-my_list<-pairwise_limits(comp_StO2_full)
-rib_col<-'#EDD03AFF'
-
-c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
-    annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill='#30123BFF',
-                alpha = 0.5,
-                ) +
-  annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=8,
-           angle=90
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=8,
-           angle=90
-           )+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-###for missing data
-comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
-comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp2))
-
-missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    labs(x = 'Days', 
-         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
-                              )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme_classic()+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
-my_list<-pairwise_limits(comp_StO2_missing)
-
-c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    annotate("rect",
-             xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-             fill='#30123BFF',
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=8
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=8)+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-pair_comp<-c1+c2
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-Appendix-1}
-
-}
-
-\caption{Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-B: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
-In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp-Appendix}
-\end{figure}
-
 \end{document}

### 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/00-Full_document_OUP_No_Appendix.Rmd
New folder for OUP Submission (Biostatistics)

--- 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/00-Full_document_OUP.Rmd
@@
     css: "style.css" #style for the HTML document
-csl: elsevier-with-titles.csl #style for references
+csl: biostatistics.csl #style for references
 bibliography: refs.bib #references
@@
 
-# Abstract
+# Summary
 
@@
 
-In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+In contrast, _generalized additive models_ (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
 Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.

### 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/00-Full_document_OUP_No_Appendix.tex
New folder for OUP Submission (Biostatistics)

--- 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/00-Full_document_OUP.tex
@@
 \usepackage{lineno}
-\usepackage{authblk}
+\usepackage[affil-it,blocks]{authblk}
 \usepackage{hyperref}
 \usepackage{graphicx}
+\usepackage{setspace}
 \usepackage[nomarkers,figuresonly]{endfloat}
@@
 %authors
-\author[1]{Ariel I. Mundo \orcidaffila{}}
-\author[2]{John R. Tipton \orcidaffilb{}}
-\author[1]{Timothy J. Muldoon \thanks{Corresponding author, tmuldoon@uark.edu}}
-\affil[1]{\footnotesize Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
-\affil[2]{\footnotesize Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
+\author{Ariel I. Mundo \orcidaffila{}, Timothy J. Muldoon*}
+\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
+\affil{\normalfont {tmuldoon@uark.edu}}
+\author{John R. Tipton \orcidaffilb{}}
+\affil{Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
 
@@
 %}
+
 \usepackage{placeins}
@@
 
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
+\doublespacing
+
+\hypertarget{summary}{%
+\section{Summary}\label{summary}}
 
@@
 
-In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+In contrast, \emph{generalized additive models} (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
 Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
@@
 Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response {[}\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}{]}, antibody expression {[}\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}{]}, and cell metabolism {[}\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}\protect\hyperlink{ref-greening2018}{9}{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}\protect\hyperlink{ref-sio2016}{10}{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}\protect\hyperlink{ref-kamstra2015}{11}{]}.
+Tumor response (\protect\hyperlink{ref-roblyer2011}{Roblyer \emph{and others} 2011}; \protect\hyperlink{ref-demidov2018}{Demidov \emph{and others} 2018}; \protect\hyperlink{ref-pavlov2018}{Pavlov \emph{and others} 2018}; \protect\hyperlink{ref-tank2020}{Tank \emph{and others} 2020}), antibody expression (\protect\hyperlink{ref-ritter2001}{Ritter \emph{and others} 2001}; \protect\hyperlink{ref-roth2017}{Roth \emph{and others} 2017}), and cell metabolism (\protect\hyperlink{ref-skala2010}{Skala \emph{and others} 2010}; \protect\hyperlink{ref-jones2018}{Jones \emph{and others} 2018}) are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
+Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals(\protect\hyperlink{ref-greening2018}{Greening \emph{and others} 2018}), to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients (\protect\hyperlink{ref-sio2016}{Sio \emph{and others} 2016}), to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients (\protect\hyperlink{ref-kamstra2015}{Kamstra \emph{and others} 2015}).
 
 Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}\protect\hyperlink{ref-wagenmakers2008}{12}{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}{]}.
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
+The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times (\protect\hyperlink{ref-wagenmakers2008}{Wagenmakers \emph{and others} 2008}), and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
+The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) (\protect\hyperlink{ref-gueorguieva2004}{Gueorguieva and Krystal 2004}; \protect\hyperlink{ref-schober2018}{Schober and Vetter 2018}).
+
+The expected linear behavior of the response through time is a key requisite in rm-ANOVA (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}).
 This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
 In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings (\protect\hyperlink{ref-vishwanath2009}{Vishwanath \emph{and others} 2009}; \protect\hyperlink{ref-skala2010}{Skala \emph{and others} 2010}; \protect\hyperlink{ref-roblyer2011}{Roblyer \emph{and others} 2011}).
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
@@
 Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}.
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}\protect\hyperlink{ref-wang2019}{18}{]}.
-
-Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}{]}, consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}\protect\hyperlink{ref-abdi2010}{21}{]}), but they in turn reduce statistical power (1-\(\beta\)){[}\protect\hyperlink{ref-nakagawa2004}{22}{]}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) {[}\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}{]}.
+In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs(\protect\hyperlink{ref-dennis2019}{Dennis \emph{and others} 2019}).
+For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid (\protect\hyperlink{ref-wang2019}{Wang \emph{and others} 2019}).
+
+Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) (\protect\hyperlink{ref-liu2010}{Liu \emph{and others} 2010}; \protect\hyperlink{ref-halsey2015}{Halsey \emph{and others} 2015}), consequently biasing the conclusions of the study.
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni (\protect\hyperlink{ref-abdi2010}{Abdi 2010})), but they in turn reduce statistical power (1-\(\beta\))(\protect\hyperlink{ref-nakagawa2004}{Nakagawa 2004}), and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) (\protect\hyperlink{ref-gelman2012}{Gelman \emph{and others} 2012}; \protect\hyperlink{ref-albers2019}{Albers 2019}).
 Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
 
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and group sizes {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}\protect\hyperlink{ref-charan2013}{30}{]} which might cause the corrections for lack of compound symmetry to be ineffective.
+On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases (\protect\hyperlink{ref-ugrinowitsch2004}{Ugrinowitsch \emph{and others} 2004}).
+Corrections can be made in rm-ANOVA in the absence of compound symmetry (\protect\hyperlink{ref-greenhouse1959}{Greenhouse and Geisser 1959}; \protect\hyperlink{ref-huynh1976}{Huynh and Feldt 1976}), but the effectiveness of the correction is limited by the size of the sample, the number of measurements(\protect\hyperlink{ref-haverkamp2017}{Haverkamp and Beauducel 2017}), and group sizes (\protect\hyperlink{ref-keselman2001}{Keselman \emph{and others} 2001}).
+In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons (\protect\hyperlink{ref-charan2013}{Charan and Kantharia 2013}) which might cause the corrections for lack of compound symmetry to be ineffective.
 
@@
 In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}\protect\hyperlink{ref-gueorguieva2004}{13}{]}.
+When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis (\protect\hyperlink{ref-gueorguieva2004}{Gueorguieva and Krystal 2004}).
 This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
@@
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data (\protect\hyperlink{ref-vishwanath2009}{Vishwanath \emph{and others} 2009}; \protect\hyperlink{ref-skala2010}{Skala \emph{and others} 2010}).
 Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}.
-And even more importantly, LMEMs also assume a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}, making them unsuitable to analyze non-linear data.
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}).
+However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent (\protect\hyperlink{ref-gueorguieva2004}{Gueorguieva and Krystal 2004}; \protect\hyperlink{ref-barr2013}{Barr \emph{and others} 2013}).
+And even more importantly, LMEMs also assume a linear relationship between the response and time (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}), making them unsuitable to analyze non-linear data.
 
@@
 Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]}.
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]} and 3) can easily accommodate missing observations.
+Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data (\protect\hyperlink{ref-rose2012}{Rose \emph{and others} 2012}; \protect\hyperlink{ref-simpson2018}{Simpson 2018}; \protect\hyperlink{ref-pedersen2019}{Pedersen \emph{and others} 2019}), health-environment interactions (\protect\hyperlink{ref-yang2012}{Yang \emph{and others} 2012}) and the dynamics of government in political science (\protect\hyperlink{ref-beck1998}{Beck and Jackman 1998}).
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements (\protect\hyperlink{ref-wood2017}{Wood 2017}) and 3) can easily accommodate missing observations.
 Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
@@
 The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
+In particular, \(\textsf{R}\) (\protect\hyperlink{ref-r}{R Core Team 2020}) has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} (\protect\hyperlink{ref-wood2016}{Wood \emph{and others} 2016}; \protect\hyperlink{ref-wood2017}{Wood 2017}) that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models (\protect\hyperlink{ref-haverkamp2017}{Haverkamp and Beauducel 2017}).
 Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
@@
 Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies (\protect\hyperlink{ref-vishwanath2009}{Vishwanath \emph{and others} 2009}).
 The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
@@
 
-A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish {[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}.
+A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}; \protect\hyperlink{ref-west2014}{West \emph{and others} 2014}).
 A LMEM with interaction between time and treatment for a longitudinal study can be written as:
@@
 The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
+In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level(\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}).
 However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
@@
 In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{41}{]} (For an in-depth analysis of the covariance matrix see {[}\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}{]}).
+The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study(\protect\hyperlink{ref-wolfinger1996}{Wolfinger 1996}) (For an in-depth analysis of the covariance matrix see (\protect\hyperlink{ref-weiss2005}{Weiss 2005}; \protect\hyperlink{ref-west2014}{West \emph{and others} 2014})).
 
 In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}{]}.
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}\protect\hyperlink{ref-maxwell2017}{44}{]}; and it is higher between consecutive measurements {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and they are not robust if the group sizes are unbalanced {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
+Under this assumption, the between-subject variance and within-subject correlation are constant across time (\protect\hyperlink{ref-geisser1958}{Geisser and Greenhouse 1958}; \protect\hyperlink{ref-huynh1976}{Huynh and Feldt 1976}; \protect\hyperlink{ref-weiss2005}{Weiss 2005}).
+However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time (\protect\hyperlink{ref-maxwell2017}{Maxwell \emph{and others} 2017}); and it is higher between consecutive measurements (\protect\hyperlink{ref-gueorguieva2004}{Gueorguieva and Krystal 2004}; \protect\hyperlink{ref-ugrinowitsch2004}{Ugrinowitsch \emph{and others} 2004}).
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)(\protect\hyperlink{ref-greenhouse1959}{Greenhouse and Geisser 1959}; \protect\hyperlink{ref-huynh1976}{Huynh and Feldt 1976}) the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements(\protect\hyperlink{ref-haverkamp2017}{Haverkamp and Beauducel 2017}), and they are not robust if the group sizes are unbalanced (\protect\hyperlink{ref-keselman2001}{Keselman \emph{and others} 2001}).
 Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
 
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}).
 Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
@@
 Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator (\protect\hyperlink{ref-molenberghs2004}{Molenberghs 2004}).
 Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}\protect\hyperlink{ref-weiss2005}{42}{]}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{46}{]}.
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{47}{]}, and in the case of MNAR the missing values are dependent on their value.
+Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) (\protect\hyperlink{ref-weiss2005}{Weiss 2005}).
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest (\protect\hyperlink{ref-scheffer2002}{Scheffer 2002}).
+If the data are MCAR, this means that the missingness is completely unrelated to the collected information (\protect\hyperlink{ref-potthoff2006}{Potthoff \emph{and others} 2006}), and in the case of MNAR the missing values are dependent on their value.
 
 An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{48}{]}.
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}\protect\hyperlink{ref-west2014}{40}{]}.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size (\protect\hyperlink{ref-ma2012}{Ma \emph{and others} 2012}).
+In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR (\protect\hyperlink{ref-west2014}{West \emph{and others} 2014}).
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
@@
 To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{49}{]}.
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) (\protect\hyperlink{ref-r}{R Core Team 2020}) and the package \emph{nlme} (\protect\hyperlink{ref-nlme}{Pinheiro \emph{and others} 2020}).
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_files/figure-latex/l-q-response-1}
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_No_Appendix_files/figure-latex/l-q-response-1}
 
@@
 
-Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors {[}\protect\hyperlink{ref-nelder1972}{50}{]}.
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{51}{]}.
+Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors (\protect\hyperlink{ref-nelder1972}{Nelder and Wedderburn 1972}).
+In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case(\protect\hyperlink{ref-hastie1987}{Hastie and Tibshirani 1987}; \protect\hyperlink{ref-wood2017}{Wood 2017}; \protect\hyperlink{ref-simpson2018}{Simpson 2018}).
 A GAM model can be written as:
@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
+One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference (\protect\hyperlink{ref-beck1998}{Beck and Jackman 1998}).
 In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
@@
 The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
+A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist (\protect\hyperlink{ref-simpson2018}{Simpson 2018}).
 For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{52}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response (\protect\hyperlink{ref-hefley2017}{Hefley \emph{and others} 2017}), which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
 In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
-Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
+Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data (\protect\hyperlink{ref-wood2017}{Wood 2017}; \protect\hyperlink{ref-simpson2018}{Simpson 2018}).
+Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness (\protect\hyperlink{ref-wegman1983}{Wegman and Wright 1983}).
 Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_files/figure-latex/basis-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_No_Appendix_files/figure-latex/basis-plot-1}
 
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}\protect\hyperlink{ref-mcelreath2018}{54}{]}.
-In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) {[}\protect\hyperlink{ref-miller2019}{55}{]}.
-Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{56}{]}.
-Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief (\protect\hyperlink{ref-mcelreath2018}{McElreath 2018}).
+In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs (\protect\hyperlink{ref-wood2017}{Wood 2017}).
+GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) (\protect\hyperlink{ref-miller2019}{Miller 2019}).
+Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model (\protect\hyperlink{ref-laird1982}{Laird and Ware 1982}; \protect\hyperlink{ref-pedersen2019}{Pedersen \emph{and others} 2019}).
+Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals (\protect\hyperlink{ref-pedersen2019}{Pedersen \emph{and others} 2019}), which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage (\protect\hyperlink{ref-wood2017}{Wood 2017}).
 This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{55}{]} and {[}\protect\hyperlink{ref-marra2012}{57}{]}.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in (\protect\hyperlink{ref-wood2017}{Wood 2017}; \protect\hyperlink{ref-simpson2018}{Simpson 2018}; \protect\hyperlink{ref-miller2019}{Miller 2019}) and (\protect\hyperlink{ref-marra2012}{Marra and Wood 2012}).
 With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in (\protect\hyperlink{ref-vishwanath2009}{Vishwanath \emph{and others} 2009}).
 In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
@@
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in {[}\protect\hyperlink{ref-hefley2017}{52}{]}.
+A detailed description on basis functions and correlations can be found in (\protect\hyperlink{ref-hefley2017}{Hefley \emph{and others} 2017}).
 
@@
 This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
-The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths (\protect\hyperlink{ref-simpson2018}{Simpson 2018}).
+The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) (\protect\hyperlink{ref-wood2017}{Wood 2017}).
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{58}{]}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{59}{]}.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} (\protect\hyperlink{ref-gratia}{Simpson 2020}).
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in (\protect\hyperlink{ref-wood2017}{Wood 2017}) and (\protect\hyperlink{ref-harezlak2018}{Harezlak \emph{and others} 2018}).
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_files/figure-latex/sim-smooth-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_No_Appendix_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_files/figure-latex/plot-pairwise-comp-1}
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_No_Appendix_files/figure-latex/plot-pairwise-comp-1}
 
@@
 Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}.
-This ``model misspecification'' error, also is known as a ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{60}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+This ``model misspecification'' error, also is known as a ``Type III'' error (\protect\hyperlink{ref-dennis2019}{Dennis \emph{and others} 2019}) is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) (\protect\hyperlink{ref-lang2015}{Lang and Altman 2015}) but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
 
 In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{61}{]}.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) (\protect\hyperlink{ref-hefley2017}{Hefley \emph{and others} 2017}; \protect\hyperlink{ref-pedersen2019}{Pedersen \emph{and others} 2019}) they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research (\protect\hyperlink{ref-hastie1995}{Hastie and Tibshirani 1995}).
 This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
@@
 However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} using \(\textsf{R}\) and the package \emph{mgcv}{[}\protect\hyperlink{ref-wood2017}{37}{]} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature (\protect\hyperlink{ref-vishwanath2009}{Vishwanath \emph{and others} 2009}) using \(\textsf{R}\) and the package \emph{mgcv}(\protect\hyperlink{ref-wood2017}{Wood 2017}) in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
 One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{62},\protect\hyperlink{ref-weissgerber2018}{63}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research (\protect\hyperlink{ref-begley2015}{Begley and Ioannidis 2015}; \protect\hyperlink{ref-weissgerber2018}{Weissgerber \emph{and others} 2018}), but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-abdi2010}{}%
+\textsc{Abdi, H.} (2010). {Holm's sequential Bonferroni procedure}. \emph{Encyclopedia of Research Design} \textbf{1}, 1--8.
+
+\leavevmode\hypertarget{ref-albers2019}{}%
+\textsc{Albers, C.} (2019).
+The problem with unadjusted multiple and sequential statistical testing.
+\emph{Nature Communications} \textbf{10}.
+doi:\href{https://doi.org/10.1038/s41467-019-09941-0}{10.1038/s41467-019-09941-0}.
+
+\leavevmode\hypertarget{ref-barr2013}{}%
+\textsc{Barr, D.
+J., Levy, R., Scheepers, C.
+and Tily, H.
+J.} (2013).
+Random effects structure for confirmatory hypothesis testing: Keep it maximal.
+\emph{Journal of Memory and Language} \textbf{68}, 255--278.
+
+\leavevmode\hypertarget{ref-beck1998}{}%
+\textsc{Beck, N. and Jackman, S.} (1998). Beyond linearity by default: Generalized additive models. \emph{American Journal of Political Science} \textbf{42}, 596.
+
+\leavevmode\hypertarget{ref-begley2015}{}%
+\textsc{Begley, C. G. and Ioannidis, J. P. A.} (2015). Reproducibility in science. \emph{Circulation Research} \textbf{116}, 116--126.
+
+\leavevmode\hypertarget{ref-charan2013}{}%
+\textsc{Charan, J. and Kantharia, N.} (2013). How to calculate sample size in animal studies? \emph{Journal of Pharmacology and Pharmacotherapeutics} \textbf{4}, 303.
+
+\leavevmode\hypertarget{ref-demidov2018}{}%
+\textsc{Demidov, V., Maeda, A., Sugita, M., Madge, V., Sadanand, S., Flueraru, C.
+and Vitkin, I.
+A.} (2018).
+Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography.
+\emph{Scientific Reports} \textbf{8}.
+doi:\href{https://doi.org/10.1038/s41598-017-18635-w}{10.1038/s41598-017-18635-w}.
+
+\leavevmode\hypertarget{ref-dennis2019}{}%
+\textsc{Dennis, B., Ponciano, J.
+M., Taper, M.
+L.
+and Lele, S.
+R.} (2019).
+Errors in statistical inference under model misspecification: Evidence, hypothesis testing, and {AIC}.
+\emph{Frontiers in Ecology and Evolution} \textbf{7}.
+doi:\href{https://doi.org/10.3389/fevo.2019.00372}{10.3389/fevo.2019.00372}.
+
+\leavevmode\hypertarget{ref-geisser1958}{}%
+\textsc{Geisser, S.
+and Greenhouse, S.
+W.} (1958).
+An extension of box{{}}s results on the use of the {{}}f{{}} distribution in multivariate analysis.
+\emph{The Annals of Mathematical Statistics} \textbf{29}, 885--891.
+
+\leavevmode\hypertarget{ref-gelman2012}{}%
+\textsc{Gelman, A., Hill, J. and Yajima, M.} (2012). Why we (usually) don{{}}t have to worry about multiple comparisons. \emph{Journal of Research on Educational Effectiveness} \textbf{5}, 189--211.
+
+\leavevmode\hypertarget{ref-greenhouse1959}{}%
+\textsc{Greenhouse, S. W. and Geisser, S.} (1959). On methods in the analysis of profile data. \emph{Psychometrika} \textbf{24}, 95--112.
+
+\leavevmode\hypertarget{ref-greening2018}{}%
+\textsc{Greening, G.
+J., Miller, K.
+P., Spainhour, C.
+R., Cato, M.
+D.
+and Muldoon, T.
+J.} (2018).
+Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy.
+\emph{Biomedical Optics Express} \textbf{9}, 2871.
+
+\leavevmode\hypertarget{ref-gueorguieva2004}{}%
+\textsc{Gueorguieva, R. and Krystal, J. H.} (2004). Move over {ANOVA}. \emph{Archives of General Psychiatry} \textbf{61}, 310.
+
+\leavevmode\hypertarget{ref-halsey2015}{}%
+\textsc{Halsey, L. G., Curran-Everett, D., Vowler, S. L. and Drummond, G. B.} (2015). The fickle p value generates irreproducible results. \emph{Nature Methods} \textbf{12}, 179--185.
+
+\leavevmode\hypertarget{ref-harezlak2018}{}%
+\textsc{Harezlak, J., Ruppert, D. and Wand, M. P.} (2018). Semiparametric regression with r. Springer New York. doi:\href{https://doi.org/10.1007/978-1-4939-8853-2}{10.1007/978-1-4939-8853-2}.
+
+\leavevmode\hypertarget{ref-hastie1995}{}%
+\textsc{Hastie, T. and Tibshirani, R.} (1995). Generalized additive models for medical research. \emph{Statistical Methods in Medical Research} \textbf{4}, 187--196.
+
+\leavevmode\hypertarget{ref-hastie1987}{}%
+\textsc{Hastie, T. and Tibshirani, R.} (1987). Generalized additive models: Some applications. \emph{Journal of the American Statistical Association} \textbf{82}, 371--386.
+
+\leavevmode\hypertarget{ref-haverkamp2017}{}%
+\textsc{Haverkamp, N.
+and Beauducel, A.} (2017).
+Violation of the sphericity assumption and its effect on type-i error rates in repeated measures {ANOVA} and multi-level linear models ({MLM}).
+\emph{Frontiers in Psychology} \textbf{8}.
+doi:\href{https://doi.org/10.3389/fpsyg.2017.01841}{10.3389/fpsyg.2017.01841}.
+
+\leavevmode\hypertarget{ref-hefley2017}{}%
+\textsc{Hefley, T.
+J., Broms, K.
+M., Brost, B.
+M., Buderman, F.
+E., Kay, S.
+L., Scharf, H.
+R., Tipton, J.
+R., Williams, P.
+J.
+and Hooten, M.
+B.} (2017).
+The basis function approach for modeling autocorrelation in ecological data.
+\emph{Ecology} \textbf{98}, 632--646.
+
+\leavevmode\hypertarget{ref-huynh1976}{}%
+\textsc{Huynh, H.
+and Feldt, L.
+S.} (1976).
+Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs.
+\emph{Journal of Educational Statistics} \textbf{1}, 69--82.
+
+\leavevmode\hypertarget{ref-jones2018}{}%
+\textsc{Jones, J.
+D., Ramser, H.
+E., Woessner, A.
+E.
+and Quinn, K.
+P.} (2018).
+In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing.
+\emph{Communications Biology} \textbf{1}.
+doi:\href{https://doi.org/10.1038/s42003-018-0206-4}{10.1038/s42003-018-0206-4}.
+
+\leavevmode\hypertarget{ref-kamstra2015}{}%
+\textsc{Kamstra, J.
+I., Dijkstra, P.
+U., Leeuwen, M.
+van, Roodenburg, J.
+L.
+N.
+and Langendijk, J.
+A.} (2015).
+Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study.
+\emph{Oral Oncology} \textbf{51}, 548--555.
+
+\leavevmode\hypertarget{ref-keselman2001}{}%
+\textsc{Keselman, H.
+J., Algina, J.
+and Kowalchuk, R.
+K.} (2001).
+The analysis of repeated measures designs: A review.
+\emph{British Journal of Mathematical and Statistical Psychology} \textbf{54}, 1--20.
+
+\leavevmode\hypertarget{ref-laird1982}{}%
+\textsc{Laird, N. M. and Ware, J. H.} (1982). Random-effects models for longitudinal data. \emph{Biometrics} \textbf{38}, 963.
+
+\leavevmode\hypertarget{ref-lang2015}{}%
+\textsc{Lang, T.
+A.
+and Altman, D.
+G.} (2015).
+Basic statistical reporting for articles published in biomedical journals: The {``}statistical analyses and methods in the published literature{''} or the {SAMPL} guidelines.
+\emph{International Journal of Nursing Studies} \textbf{52}, 5--9.
+
+\leavevmode\hypertarget{ref-liu2010}{}%
+\textsc{Liu, C., Cripe, T.
+P.
+and Kim, M.-O.} (2010).
+Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences.
+\emph{Molecular Therapy} \textbf{18}, 1724--1730.
+
+\leavevmode\hypertarget{ref-ma2012}{}%
+\textsc{Ma, Y., Mazumdar, M. and Memtsoudis, S. G.} (2012). Beyond repeated-measures analysis of variance. \emph{Regional Anesthesia and Pain Medicine} \textbf{37}, 99--105.
+
+\leavevmode\hypertarget{ref-marra2012}{}%
+\textsc{Marra, G. and Wood, S. N.} (2012). Coverage properties of confidence intervals for generalized additive model components. \emph{Scandinavian Journal of Statistics} \textbf{39}, 53--74.
+
+\leavevmode\hypertarget{ref-maxwell2017}{}%
+\textsc{Maxwell, S. E., Delaney, H. D. and Kelley, K.} (2017). Designing experiments and analyzing data. Routledge. doi:\href{https://doi.org/10.4324/9781315642956}{10.4324/9781315642956}.
+
+\leavevmode\hypertarget{ref-mcelreath2018}{}%
+\textsc{McElreath, R.} (2018). Statistical rethinking. Chapman; Hall/{CRC}. doi:\href{https://doi.org/10.1201/9781315372495}{10.1201/9781315372495}.
+
+\leavevmode\hypertarget{ref-miller2019}{}%
+\textsc{Miller, D. L.} (2019). Bayesian views of generalized additive modelling. \emph{arXiv preprint arXiv:1902.01330}.
+
+\leavevmode\hypertarget{ref-molenberghs2004}{}%
+\textsc{Molenberghs, G.} (2004). Analyzing incomplete longitudinal clinical trial data. \emph{Biostatistics} \textbf{5}, 445--464.
+
+\leavevmode\hypertarget{ref-nakagawa2004}{}%
+\textsc{Nakagawa, S.} (2004). A farewell to bonferroni: The problems of low statistical power and publication bias. \emph{Behavioral Ecology} \textbf{15}, 1044--1045.
+
+\leavevmode\hypertarget{ref-nelder1972}{}%
+\textsc{Nelder, J. A. and Wedderburn, R. W. M.} (1972). Generalized linear models. \emph{Journal of the Royal Statistical Society. Series A (General)} \textbf{135}, 370.
+
+\leavevmode\hypertarget{ref-pavlov2018}{}%
+\textsc{Pavlov, M.
+V., Kalganova, T.
+I.
+and Lyubimtseva, Y.
+S.} (2018).
+Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy.
+\emph{Journal of Biomedical Optics} \textbf{23}, 1.
+
+\leavevmode\hypertarget{ref-pedersen2019}{}%
+\textsc{Pedersen, E. J., Miller, D. L., Simpson, G. L. and Ross, N.} (2019). Hierarchical generalized additive models in ecology: An introduction with mgcv. \emph{{PeerJ}} \textbf{7}, e6876.
+
+\leavevmode\hypertarget{ref-pinheiro2006}{}%
+\textsc{Pinheiro, J. and Bates, D.} (2006). {Mixed-effects models in S and S-PLUS}. Springer Science; Business Media. doi:\url{https://doi.org/10.1007/b98882}.
+
+\leavevmode\hypertarget{ref-nlme}{}%
+\textsc{Pinheiro, J., Bates, D., DebRoy, S., Sarkar, D. and R Core Team}. (2020). {nlme}: Linear and nonlinear mixed effects models. \url{https://CRAN.R-project.org/package=nlme}.
+
+\leavevmode\hypertarget{ref-potthoff2006}{}%
+\textsc{Potthoff, R.
+F., Tudor, G.
+E., Pieper, K.
+S.
+and Hasselblad, V.} (2006).
+Can one assess whether missing data are missing at random in medical studies?
+\emph{Statistical Methods in Medical Research} \textbf{15}, 213--234.
+
+\leavevmode\hypertarget{ref-r}{}%
+\textsc{R Core Team}. (2020). R: A language and environment for statistical computing. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.
+
+\leavevmode\hypertarget{ref-ritter2001}{}%
+\textsc{Ritter, G., Cohen, L., Williams, C., Richards, E., Old, L.
+and Welt, S.} (2001).
+{Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}.
+\emph{{Cancer Research}} \textbf{{61}}, 6851--6859.
+
 \leavevmode\hypertarget{ref-roblyer2011}{}%
-\CSLLeftMargin{{[}1{]} }
-\CSLRightInline{D.
-Roblyer, S.
-Ueda, A.
-Cerussi, W.
-Tanamai, A.
-Durkin, R.
-Mehta, D.
-Hsiang, J.A.
-Butler, C.
-McLaren, W.-P.
-Chen, B.
-Tromberg, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences.
-108 (2011) 14626--14631.
-\url{https://doi.org/10.1073/pnas.1013103108}.}
+\textsc{Roblyer, D., Ueda, S., Cerussi, A., Tanamai, W., Durkin, A., Mehta, R., Hsiang, D., Butler, J.
+A., McLaren, C., Chen, W.-P., et al.} (2011).
+Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment.
+\emph{Proceedings of the National Academy of Sciences} \textbf{108}, 14626--14631.
+
+\leavevmode\hypertarget{ref-rose2012}{}%
+\textsc{Rose, N.
+L., Yang, H., Turner, S.
+D.
+and Simpson, G.
+L.} (2012).
+An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, {UK}.
+\emph{Geochimica et Cosmochimica Acta} \textbf{82}, 113--135.
+
+\leavevmode\hypertarget{ref-roth2017}{}%
+\textsc{Roth, E.
+M., Goldberg, A.
+C., Catapano, A.
+L., Torri, A., Yancopoulos, G.
+D., Stahl, N., Brunet, A., Lecorps, G.
+and Colhoun, H.
+M.} (2017).
+Antidrug antibodies in patients treated with alirocumab.
+\emph{New England Journal of Medicine} \textbf{376}, 1589--1590.
+
+\leavevmode\hypertarget{ref-scheffer2002}{}%
+\textsc{Scheffer, J.} (2002). Dealing with missing data. \emph{Research Letters in the Information and Mathematical Sciences} \textbf{3}, 153--160.
+
+\leavevmode\hypertarget{ref-schober2018}{}%
+\textsc{Schober, P. and Vetter, T. R.} (2018). Repeated measures designs and analysis of longitudinal data. \emph{Anesthesia {\&} Analgesia} \textbf{127}, 569--575.
+
+\leavevmode\hypertarget{ref-gratia}{}%
+\textsc{Simpson, G. L.} (2020). Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv'. \url{https://CRAN.R-project.org/package=gratia}.
+
+\leavevmode\hypertarget{ref-simpson2018}{}%
+\textsc{Simpson, G.
+L.} (2018).
+Modelling palaeoecological time series using generalised additive models.
+\emph{Frontiers in Ecology and Evolution} \textbf{6}.
+doi:\href{https://doi.org/10.3389/fevo.2018.00149}{10.3389/fevo.2018.00149}.
+
+\leavevmode\hypertarget{ref-sio2016}{}%
+\textsc{Sio, T.
+T., Atherton, P.
+J., Birckhead, B.
+J., Schwartz, D.
+J., Sloan, J.
+A., Seisler, D.
+K., Martenson, J.
+A., Loprinzi, C.
+L., Griffin, P.
+C., Morton, R.
+F., et al.} (2016).
+Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}).
+\emph{Supportive Care in Cancer} \textbf{24}, 3847--3855.
+
+\leavevmode\hypertarget{ref-skala2010}{}%
+\textsc{Skala, M.
+C., Fontanella, A., Lan, L., Izatt, J.
+A.
+and Dewhirst, M.
+W.} (2010).
+Longitudinal optical imaging of tumor metabolism and hemodynamics.
+\emph{Journal of Biomedical Optics} \textbf{15}, 011112.
 
 \leavevmode\hypertarget{ref-tank2020}{}%
-\CSLLeftMargin{{[}2{]} }
-\CSLRightInline{A.
-Tank, H.M.
-Peterson, V.
-Pera, S.
-Tabassum, A.
-Leproux, T.
-O'Sullivan, E.
-Jones, H.
-Cabral, N.
-Ko, R.S.
-Mehta, B.J.
-Tromberg, D.
-Roblyer, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research.
-22 (2020).
-\url{https://doi.org/10.1186/s13058-020-01262-1}.}
-
-\leavevmode\hypertarget{ref-pavlov2018}{}%
-\CSLLeftMargin{{[}3{]} }
-\CSLRightInline{M.V.
-Pavlov, T.I.
-Kalganova, Y.S.
-Lyubimtseva, Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy, Journal of Biomedical Optics.
-23 (2018) 1.
-\url{https://doi.org/10.1117/1.jbo.23.9.091410}.}
-
-\leavevmode\hypertarget{ref-demidov2018}{}%
-\CSLLeftMargin{{[}4{]} }
-\CSLRightInline{V.
-Demidov, A.
-Maeda, M.
-Sugita, V.
-Madge, S.
-Sadanand, C.
-Flueraru, I.A.
-Vitkin, Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography, Scientific Reports.
-8 (2018).
-\url{https://doi.org/10.1038/s41598-017-18635-w}.}
-
-\leavevmode\hypertarget{ref-ritter2001}{}%
-\CSLLeftMargin{{[}5{]} }
-\CSLRightInline{G.
-Ritter, L.
-Cohen, C.
-Williams, E.
-Richards, L.
-Old, S.
-Welt, {Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}, {Cancer Research}.
-{61} (2001) 6851--6859.}
-
-\leavevmode\hypertarget{ref-roth2017}{}%
-\CSLLeftMargin{{[}6{]} }
-\CSLRightInline{E.M.
-Roth, A.C.
-Goldberg, A.L.
-Catapano, A.
-Torri, G.D.
-Yancopoulos, N.
-Stahl, A.
-Brunet, G.
-Lecorps, H.M.
-Colhoun, Antidrug antibodies in patients treated with alirocumab, New England Journal of Medicine.
-376 (2017) 1589--1590.
-\url{https://doi.org/10.1056/nejmc1616623}.}
-
-\leavevmode\hypertarget{ref-jones2018}{}%
-\CSLLeftMargin{{[}7{]} }
-\CSLRightInline{J.D.
-Jones, H.E.
-Ramser, A.E.
-Woessner, K.P.
-Quinn, In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing, Communications Biology.
-1 (2018).
-\url{https://doi.org/10.1038/s42003-018-0206-4}.}
-
-\leavevmode\hypertarget{ref-skala2010}{}%
-\CSLLeftMargin{{[}8{]} }
-\CSLRightInline{M.C.
-Skala, A.
-Fontanella, L.
-Lan, J.A.
-Izatt, M.W.
-Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics.
-15 (2010) 011112.
-\url{https://doi.org/10.1117/1.3285584}.}
-
-\leavevmode\hypertarget{ref-greening2018}{}%
-\CSLLeftMargin{{[}9{]} }
-\CSLRightInline{G.J.
-Greening, K.P.
-Miller, C.R.
-Spainhour, M.D.
-Cato, T.J.
-Muldoon, Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy, Biomedical Optics Express.
-9 (2018) 2871.
-\url{https://doi.org/10.1364/boe.9.002871}.}
-
-\leavevmode\hypertarget{ref-sio2016}{}%
-\CSLLeftMargin{{[}10{]} }
-\CSLRightInline{T.T.
-Sio, P.J.
-Atherton, B.J.
-Birckhead, D.J.
-Schwartz, J.A.
-Sloan, D.K.
-Seisler, J.A.
-Martenson, C.L.
-Loprinzi, P.C.
-Griffin, R.F.
-Morton, J.C.
-Anders, T.J.
-Stoffel, R.E.
-Haselow, R.B.
-Mowat, M.A.N.
-Wittich, J.D.
-Bearden, R.C.
-Miller, Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}), Supportive Care in Cancer.
-24 (2016) 3847--3855.
-\url{https://doi.org/10.1007/s00520-016-3213-3}.}
-
-\leavevmode\hypertarget{ref-kamstra2015}{}%
-\CSLLeftMargin{{[}11{]} }
-\CSLRightInline{J.I.
-Kamstra, P.U.
-Dijkstra, M.
-van Leeuwen, J.L.N.
-Roodenburg, J.A.
-Langendijk, Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study, Oral Oncology.
-51 (2015) 548--555.
-\url{https://doi.org/10.1016/j.oraloncology.2015.01.016}.}
+\textsc{Tank, A., Peterson, H.
+M., Pera, V., Tabassum, S., Leproux, A., O'Sullivan, T., Jones, E., Cabral, H., Ko, N., Mehta, R.
+S., et al.} (2020).
+Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens.
+\emph{Breast Cancer Research} \textbf{22}.
+doi:\href{https://doi.org/10.1186/s13058-020-01262-1}{10.1186/s13058-020-01262-1}.
+
+\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
+\textsc{Ugrinowitsch, C., Fellingham, G.
+W.
+and Ricard, M.
+D.} (2004).
+Limitations of ordinary least squares models in analyzing repeated measures data.
+\emph{Medicine {\&} Science in Sports {\&} Exercise}, 2144--2148.
+
+\leavevmode\hypertarget{ref-vishwanath2009}{}%
+\textsc{Vishwanath, K., Yuan, H., Barry, W.
+T., Dewhirst, M.
+W.
+and Ramanujam, N.} (2009).
+Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors.
+\emph{Neoplasia} \textbf{11}, 889--900.
 
 \leavevmode\hypertarget{ref-wagenmakers2008}{}%
-\CSLLeftMargin{{[}12{]} }
-\CSLRightInline{E.-J.
-Wagenmakers, M.
-Lee, T.
-Lodewyckx, G.J.
-Iverson, Bayesian versus frequentist inference, in: Bayesian Evaluation of Informative Hypotheses, Springer New York, 2008: pp.
-181--207.
-\url{https://doi.org/10.1007/978-0-387-09612-4_9}.}
-
-\leavevmode\hypertarget{ref-gueorguieva2004}{}%
-\CSLLeftMargin{{[}13{]} }
-\CSLRightInline{R. Gueorguieva, J.H. Krystal, Move over {ANOVA}, Archives of General Psychiatry. 61 (2004) 310. \url{https://doi.org/10.1001/archpsyc.61.3.310}.}
-
-\leavevmode\hypertarget{ref-schober2018}{}%
-\CSLLeftMargin{{[}14{]} }
-\CSLRightInline{P.
-Schober, T.R.
-Vetter, Repeated measures designs and analysis of longitudinal data, Anesthesia {\&} Analgesia.
-127 (2018) 569--575.
-\url{https://doi.org/10.1213/ane.0000000000003511}.}
-
-\leavevmode\hypertarget{ref-pinheiro2006}{}%
-\CSLLeftMargin{{[}15{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science; Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}
-
-\leavevmode\hypertarget{ref-vishwanath2009}{}%
-\CSLLeftMargin{{[}16{]} }
-\CSLRightInline{K.
-Vishwanath, H.
-Yuan, W.T.
-Barry, M.W.
-Dewhirst, N.
-Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia.
-11 (2009) 889--900.
-\url{https://doi.org/10.1593/neo.09580}.}
-
-\leavevmode\hypertarget{ref-dennis2019}{}%
-\CSLLeftMargin{{[}17{]} }
-\CSLRightInline{B.
-Dennis, J.M.
-Ponciano, M.L.
-Taper, S.R.
-Lele, Errors in statistical inference under model misspecification: Evidence, hypothesis testing, and {AIC}, Frontiers in Ecology and Evolution.
-7 (2019).
-\url{https://doi.org/10.3389/fevo.2019.00372}.}
+\textsc{Wagenmakers, E.-J., Lee, M., Lodewyckx, T. and Iverson, G. J.} (2008). Bayesian versus frequentist inference. In Bayesian evaluation of informative hypotheses. Springer New York. pp 181--207.
 
 \leavevmode\hypertarget{ref-wang2019}{}%
-\CSLLeftMargin{{[}18{]} }
-\CSLRightInline{B.
-Wang, Z.
-Zhou, H.
-Wang, X.M.
-Tu, C.
-Feng, The p-value and model specification in statistics, General Psychiatry.
-32 (2019) e100081.
-\url{https://doi.org/10.1136/gpsych-2019-100081}.}
-
-\leavevmode\hypertarget{ref-liu2010}{}%
-\CSLLeftMargin{{[}19{]} }
-\CSLRightInline{C.
-Liu, T.P.
-Cripe, M.-O.
-Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy.
-18 (2010) 1724--1730.
-\url{https://doi.org/10.1038/mt.2010.127}.}
-
-\leavevmode\hypertarget{ref-halsey2015}{}%
-\CSLLeftMargin{{[}20{]} }
-\CSLRightInline{L.G.
-Halsey, D.
-Curran-Everett, S.L.
-Vowler, G.B.
-Drummond, The fickle p value generates irreproducible results, Nature Methods.
-12 (2015) 179--185.
-\url{https://doi.org/10.1038/nmeth.3288}.}
-
-\leavevmode\hypertarget{ref-abdi2010}{}%
-\CSLLeftMargin{{[}21{]} }
-\CSLRightInline{H. Abdi, {Holm's sequential Bonferroni procedure}, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.}
-
-\leavevmode\hypertarget{ref-nakagawa2004}{}%
-\CSLLeftMargin{{[}22{]} }
-\CSLRightInline{S.
-Nakagawa, A farewell to bonferroni: The problems of low statistical power and publication bias, Behavioral Ecology.
-15 (2004) 1044--1045.
-\url{https://doi.org/10.1093/beheco/arh107}.}
-
-\leavevmode\hypertarget{ref-gelman2012}{}%
-\CSLLeftMargin{{[}23{]} }
-\CSLRightInline{A.
-Gelman, J.
-Hill, M.
-Yajima, Why we (usually) don{{}}t have to worry about multiple comparisons, Journal of Research on Educational Effectiveness.
-5 (2012) 189--211.
-\url{https://doi.org/10.1080/19345747.2011.618213}.}
-
-\leavevmode\hypertarget{ref-albers2019}{}%
-\CSLLeftMargin{{[}24{]} }
-\CSLRightInline{C. Albers, The problem with unadjusted multiple and sequential statistical testing, Nature Communications. 10 (2019). \url{https://doi.org/10.1038/s41467-019-09941-0}.}
-
-\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
-\CSLLeftMargin{{[}25{]} }
-\CSLRightInline{C.
-Ugrinowitsch, G.W.
-Fellingham, M.D.
-Ricard, Limitations of ordinary least squares models in analyzing repeated measures data, Medicine {\&} Science in Sports {\&} Exercise.
-(2004) 2144--2148.
-\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.}
-
-\leavevmode\hypertarget{ref-huynh1976}{}%
-\CSLLeftMargin{{[}26{]} }
-\CSLRightInline{H.
-Huynh, L.S.
-Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics.
-1 (1976) 69--82.
-\url{https://doi.org/10.3102/10769986001001069}.}
-
-\leavevmode\hypertarget{ref-greenhouse1959}{}%
-\CSLLeftMargin{{[}27{]} }
-\CSLRightInline{S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.}
-
-\leavevmode\hypertarget{ref-haverkamp2017}{}%
-\CSLLeftMargin{{[}28{]} }
-\CSLRightInline{N.
-Haverkamp, A.
-Beauducel, Violation of the sphericity assumption and its effect on type-i error rates in repeated measures {ANOVA} and multi-level linear models ({MLM}), Frontiers in Psychology.
-8 (2017).
-\url{https://doi.org/10.3389/fpsyg.2017.01841}.}
-
-\leavevmode\hypertarget{ref-keselman2001}{}%
-\CSLLeftMargin{{[}29{]} }
-\CSLRightInline{H.J.
-Keselman, J.
-Algina, R.K.
-Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematical and Statistical Psychology.
-54 (2001) 1--20.
-\url{https://doi.org/10.1348/000711001159357}.}
-
-\leavevmode\hypertarget{ref-charan2013}{}%
-\CSLLeftMargin{{[}30{]} }
-\CSLRightInline{J.
-Charan, N.
-Kantharia, How to calculate sample size in animal studies?, Journal of Pharmacology and Pharmacotherapeutics.
-4 (2013) 303.
-\url{https://doi.org/10.4103/0976-500x.119726}.}
-
-\leavevmode\hypertarget{ref-barr2013}{}%
-\CSLLeftMargin{{[}31{]} }
-\CSLRightInline{D.J.
-Barr, R.
-Levy, C.
-Scheepers, H.J.
-Tily, Random effects structure for confirmatory hypothesis testing: Keep it maximal, Journal of Memory and Language.
-68 (2013) 255--278.
-\url{https://doi.org/10.1016/j.jml.2012.11.001}.}
-
-\leavevmode\hypertarget{ref-rose2012}{}%
-\CSLLeftMargin{{[}32{]} }
-\CSLRightInline{N.L.
-Rose, H.
-Yang, S.D.
-Turner, G.L.
-Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, {UK}, Geochimica Et Cosmochimica Acta.
-82 (2012) 113--135.
-\url{https://doi.org/10.1016/j.gca.2010.12.026}.}
-
-\leavevmode\hypertarget{ref-pedersen2019}{}%
-\CSLLeftMargin{{[}33{]} }
-\CSLRightInline{E.J.
-Pedersen, D.L.
-Miller, G.L.
-Simpson, N.
-Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, {PeerJ}.
-7 (2019) e6876.
-\url{https://doi.org/10.7717/peerj.6876}.}
-
-\leavevmode\hypertarget{ref-simpson2018}{}%
-\CSLLeftMargin{{[}34{]} }
-\CSLRightInline{G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.}
+\textsc{Wang, B., Zhou, Z., Wang, H., Tu, X. M. and Feng, C.} (2019). The p-value and model specification in statistics. \emph{General Psychiatry} \textbf{32}, e100081.
+
+\leavevmode\hypertarget{ref-wegman1983}{}%
+\textsc{Wegman, E. J. and Wright, I. W.} (1983). Splines in statistics. \emph{Journal of the American Statistical Association} \textbf{78}, 351--365.
+
+\leavevmode\hypertarget{ref-weiss2005}{}%
+\textsc{Weiss, R. E.} (2005). Modeling longitudinal data. Springer New York. doi:\href{https://doi.org/10.1007/0-387-28314-5}{10.1007/0-387-28314-5}.
+
+\leavevmode\hypertarget{ref-weissgerber2018}{}%
+\textsc{Weissgerber, T.
+L., Garcia-Valencia, O., Garovic, V.
+D., Milic, N.
+M.
+and Winham, S.
+J.} (2018).
+Why we need to report more than {{}}data were analyzed by t-tests or {ANOVA}{{}}.
+\emph{{eLife}} \textbf{7}.
+doi:\href{https://doi.org/10.7554/elife.36163}{10.7554/elife.36163}.
+
+\leavevmode\hypertarget{ref-west2014}{}%
+\textsc{West, B.
+T., Welch, K.
+B.
+and Galecki, A.
+T.} (2014).
+Linear mixed models: A practical guide using statistical software, second edition.
+Taylor \& Francis.
+\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.
+
+\leavevmode\hypertarget{ref-wolfinger1996}{}%
+\textsc{Wolfinger, R. D.} (1996). Heterogeneous variance: Covariance structures for repeated measures. \emph{Journal of Agricultural, Biological, and Environmental Statistics} \textbf{1}, 205.
+
+\leavevmode\hypertarget{ref-wood2017}{}%
+\textsc{Wood, S. N.} (2017). Generalized additive models. Chapman; Hall/{CRC}. doi:\href{https://doi.org/10.1201/9781315370279}{10.1201/9781315370279}.
+
+\leavevmode\hypertarget{ref-wood2016}{}%
+\textsc{Wood, S. N., Pya, N. and Säfken, B.} (2016). Smoothing parameter and model selection for general smooth models. \emph{Journal of the American Statistical Association} \textbf{111}, 1548--1563.
 
 \leavevmode\hypertarget{ref-yang2012}{}%
-\CSLLeftMargin{{[}35{]} }
-\CSLRightInline{L.
-Yang, G.
-Qin, N.
-Zhao, C.
-Wang, G.
-Song, Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality, {BMC} Medical Research Methodology.
-12 (2012).
-\url{https://doi.org/10.1186/1471-2288-12-165}.}
-
-\leavevmode\hypertarget{ref-beck1998}{}%
-\CSLLeftMargin{{[}36{]} }
-\CSLRightInline{N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. 42 (1998) 596. \url{https://doi.org/10.2307/2991772}.}
-
-\leavevmode\hypertarget{ref-wood2017}{}%
-\CSLLeftMargin{{[}37{]} }
-\CSLRightInline{S.N. Wood, Generalized additive models, Chapman; Hall/{CRC}, 2017. \url{https://doi.org/10.1201/9781315370279}.}
-
-\leavevmode\hypertarget{ref-r}{}%
-\CSLLeftMargin{{[}38{]} }
-\CSLRightInline{R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.}
-
-\leavevmode\hypertarget{ref-wood2016}{}%
-\CSLLeftMargin{{[}39{]} }
-\CSLRightInline{S.N.
-Wood, N.
-Pya, B.
-Säfken, Smoothing parameter and model selection for general smooth models, Journal of the American Statistical Association.
-111 (2016) 1548--1563.
-\url{https://doi.org/10.1080/01621459.2016.1180986}.}
-
-\leavevmode\hypertarget{ref-west2014}{}%
-\CSLLeftMargin{{[}40{]} }
-\CSLRightInline{B.T.
-West, K.B.
-Welch, A.T.
-Galecki, Linear mixed models: A practical guide using statistical software, second edition, Taylor \& Francis, 2014.
-\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.}
-
-\leavevmode\hypertarget{ref-wolfinger1996}{}%
-\CSLLeftMargin{{[}41{]} }
-\CSLRightInline{R.D.
-Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics.
-1 (1996) 205.
-\url{https://doi.org/10.2307/1400366}.}
-
-\leavevmode\hypertarget{ref-weiss2005}{}%
-\CSLLeftMargin{{[}42{]} }
-\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://doi.org/10.1007/0-387-28314-5}.}
-
-\leavevmode\hypertarget{ref-geisser1958}{}%
-\CSLLeftMargin{{[}43{]} }
-\CSLRightInline{S.
-Geisser, S.W.
-Greenhouse, An extension of box{{}}s results on the use of the {{}}f{{}} distribution in multivariate analysis, The Annals of Mathematical Statistics.
-29 (1958) 885--891.
-\url{https://doi.org/10.1214/aoms/1177706545}.}
-
-\leavevmode\hypertarget{ref-maxwell2017}{}%
-\CSLLeftMargin{{[}44{]} }
-\CSLRightInline{S.E. Maxwell, H.D. Delaney, K. Kelley, Designing experiments and analyzing data, Routledge, 2017. \url{https://doi.org/10.4324/9781315642956}.}
-
-\leavevmode\hypertarget{ref-molenberghs2004}{}%
-\CSLLeftMargin{{[}45{]} }
-\CSLRightInline{G. Molenberghs, Analyzing incomplete longitudinal clinical trial data, Biostatistics. 5 (2004) 445--464. \url{https://doi.org/10.1093/biostatistics/kxh001}.}
-
-\leavevmode\hypertarget{ref-scheffer2002}{}%
-\CSLLeftMargin{{[}46{]} }
-\CSLRightInline{J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.}
-
-\leavevmode\hypertarget{ref-potthoff2006}{}%
-\CSLLeftMargin{{[}47{]} }
-\CSLRightInline{R.F.
-Potthoff, G.E.
-Tudor, K.S.
-Pieper, V.
-Hasselblad, Can one assess whether missing data are missing at random in medical studies?, Statistical Methods in Medical Research.
-15 (2006) 213--234.
-\url{https://doi.org/10.1191/0962280206sm448oa}.}
-
-\leavevmode\hypertarget{ref-ma2012}{}%
-\CSLLeftMargin{{[}48{]} }
-\CSLRightInline{Y.
-Ma, M.
-Mazumdar, S.G.
-Memtsoudis, Beyond repeated-measures analysis of variance, Regional Anesthesia and Pain Medicine.
-37 (2012) 99--105.
-\url{https://doi.org/10.1097/aap.0b013e31823ebc74}.}
-
-\leavevmode\hypertarget{ref-nlme}{}%
-\CSLLeftMargin{{[}49{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, {nlme}: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.}
-
-\leavevmode\hypertarget{ref-nelder1972}{}%
-\CSLLeftMargin{{[}50{]} }
-\CSLRightInline{J.A. Nelder, R.W.M. Wedderburn, Generalized linear models, Journal of the Royal Statistical Society. Series A (General). 135 (1972) 370. \url{https://doi.org/10.2307/2344614}.}
-
-\leavevmode\hypertarget{ref-hastie1987}{}%
-\CSLLeftMargin{{[}51{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association.
-82 (1987) 371--386.
-\url{https://doi.org/10.1080/01621459.1987.10478440}.}
-
-\leavevmode\hypertarget{ref-hefley2017}{}%
-\CSLLeftMargin{{[}52{]} }
-\CSLRightInline{T.J.
-Hefley, K.M.
-Broms, B.M.
-Brost, F.E.
-Buderman, S.L.
-Kay, H.R.
-Scharf, J.R.
-Tipton, P.J.
-Williams, M.B.
-Hooten, The basis function approach for modeling autocorrelation in ecological data, Ecology.
-98 (2017) 632--646.
-\url{https://doi.org/10.1002/ecy.1674}.}
-
-\leavevmode\hypertarget{ref-wegman1983}{}%
-\CSLLeftMargin{{[}53{]} }
-\CSLRightInline{E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.}
-
-\leavevmode\hypertarget{ref-mcelreath2018}{}%
-\CSLLeftMargin{{[}54{]} }
-\CSLRightInline{R. McElreath, Statistical rethinking, Chapman; Hall/{CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
-
-\leavevmode\hypertarget{ref-miller2019}{}%
-\CSLLeftMargin{{[}55{]} }
-\CSLRightInline{D.L. Miller, Bayesian views of generalized additive modelling, arXiv Preprint arXiv:1902.01330. (2019).}
-
-\leavevmode\hypertarget{ref-laird1982}{}%
-\CSLLeftMargin{{[}56{]} }
-\CSLRightInline{N.M. Laird, J.H. Ware, Random-effects models for longitudinal data, Biometrics. 38 (1982) 963. \url{https://doi.org/10.2307/2529876}.}
-
-\leavevmode\hypertarget{ref-marra2012}{}%
-\CSLLeftMargin{{[}57{]} }
-\CSLRightInline{G.
-Marra, S.N.
-Wood, Coverage properties of confidence intervals for generalized additive model components, Scandinavian Journal of Statistics.
-39 (2012) 53--74.
-\url{https://doi.org/10.1111/j.1467-9469.2011.00760.x}.}
-
-\leavevmode\hypertarget{ref-gratia}{}%
-\CSLLeftMargin{{[}58{]} }
-\CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
-
-\leavevmode\hypertarget{ref-harezlak2018}{}%
-\CSLLeftMargin{{[}59{]} }
-\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with r, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
-
-\leavevmode\hypertarget{ref-lang2015}{}%
-\CSLLeftMargin{{[}60{]} }
-\CSLRightInline{T.A.
-Lang, D.G.
-Altman, Basic statistical reporting for articles published in biomedical journals: The {``}statistical analyses and methods in the published literature{''} or the {SAMPL} guidelines, International Journal of Nursing Studies.
-52 (2015) 5--9.
-\url{https://doi.org/10.1016/j.ijnurstu.2014.09.006}.}
-
-\leavevmode\hypertarget{ref-hastie1995}{}%
-\CSLLeftMargin{{[}61{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models for medical research, Statistical Methods in Medical Research.
-4 (1995) 187--196.
-\url{https://doi.org/10.1177/096228029500400302}.}
-
-\leavevmode\hypertarget{ref-begley2015}{}%
-\CSLLeftMargin{{[}62{]} }
-\CSLRightInline{C.G. Begley, J.P.A. Ioannidis, Reproducibility in science, Circulation Research. 116 (2015) 116--126. \url{https://doi.org/10.1161/circresaha.114.303819}.}
-
-\leavevmode\hypertarget{ref-weissgerber2018}{}%
-\CSLLeftMargin{{[}63{]} }
-\CSLRightInline{T.L.
-Weissgerber, O.
-Garcia-Valencia, V.D.
-Garovic, N.M.
-Milic, S.J.
-Winham, Why we need to report more than {{}}data were analyzed by t-tests or {ANOVA}{{}}, {eLife}.
-7 (2018).
-\url{https://doi.org/10.7554/elife.36163}.}
+\textsc{Yang, L., Qin, G., Zhao, N., Wang, C.
+and Song, G.} (2012).
+Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality.
+\emph{{BMC} Medical Research Methodology} \textbf{12}.
+doi:\href{https://doi.org/10.1186/1471-2288-12-165}{10.1186/1471-2288-12-165}.
 

### 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/00-OUP_Appendix.tex
New folder for OUP Submission (Biostatistics)

--- 2021-06-22 Manuscripts/Manuscript by chapters -v2/00-Full_document.tex
@@
 \setcounter{secnumdepth}{5}
-\usepackage{lineno}
-\usepackage{authblk}
+%\usepackage{lineno}
+\usepackage[affil-it,blocks]{authblk}
 \usepackage{hyperref}
 \usepackage{graphicx}
-
+%\usepackage[nomarkers,figuresonly]{endfloat}
 %set a box to put the ORCID logo
@@
 
+%to have page numbers with letters in the Appendix:
+% %solution from https://tex.stackexchange.com/questions/59572/custom-page-numbering-for-appendix
+\newcommand{\appendixpagenumbering}{
+  \break
+  \pagenumbering{arabic}
+  \renewcommand{\thepage}{\thesection-\arabic{page}}
+}
+
 %command for the package lineno
-\linenumbers
+%\linenumbers
 
 %authors
-\author[1]{Ariel I. Mundo \orcidaffila{}}
-\author[2]{John R. Tipton \orcidaffilb{}}
-\author[1]{Timothy J. Muldoon \thanks{Corresponding author, tmuldoon@uark.edu}}
-\affil[1]{\footnotesize Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
-\affil[2]{\footnotesize Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
+\author{Ariel I. Mundo \orcidaffila{}}
+\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
+\author{John R. Tipton \orcidaffilb{}}
+\affil{Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
+\author{Timothy J. Muldoon*}
+\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
+\affil{tmuldoon@uark.edu}
 
@@
 
-\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
-\usepackage{etoolbox}
-\makeatletter
-\providecommand{\subtitle}[1]{% add subtitle to \maketitle
-  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
-}
-\makeatother
-\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
+\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}\\
+\emph{Beyond repeated measures ANOVA and Linear Mixed Models}\\
+~\\
+SUPPLEMENTARY MATERIALS: APPENDIX}
 \date{\vspace{-2.5em}}
@@
 \maketitle
+
+\newpage
 
@@
 To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
+
+\hypertarget{keywords}{%
+\section*{Keywords}\label{keywords}}
+\addcontentsline{toc}{section}{Keywords}
+
+cancer biology; tumor response; generalized additive models; simulation; R
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/l-q-response-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/l-q-response-1}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/basis-plot-1}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/plot-pairwise-comp-1}
 
@@
 
-\begin{figure}[H]
-\includegraphics{00-Full_document_files/figure-latex/linear-cases-Appendix-1} \caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/linear-cases-Appendix-1}
+
+}
+
+\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
 A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
@@
 
-\begin{figure}[H]
-\includegraphics{00-Full_document_files/figure-latex/quadratic-cases-Appendix-1} \caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/quadratic-cases-Appendix-1}
+
+}
+
+\caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
 A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-appendix-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/basis-plot-appendix-1}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/first-GAM-diag-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/first-GAM-diag-1}
 
@@
 
-\includegraphics{00-Full_document_files/figure-latex/first-GAM-check-1}
+\begin{center}\includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/first-GAM-check-1} \end{center}
 
@@
 
-\includegraphics{00-Full_document_files/figure-latex/second-GAM-1}
+\begin{center}\includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/second-GAM-1} \end{center}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/second-GAM-diag-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/second-GAM-diag-1}
 
@@
 
-\includegraphics{00-Full_document_files/figure-latex/final-model-Appendix-1}
+\begin{center}\includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/final-model-Appendix-1} \end{center}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/final-GAM-diag-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/final-GAM-diag-1}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/pairwise-comp-workflow-fig-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/pairwise-comp-workflow-fig-1}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-Appendix-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/sim-smooth-plot-Appendix-1}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-Appendix-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/plot-pairwise-comp-Appendix-1}
 

### 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/01-Background.Rmd
New folder for OUP Submission (Biostatistics)

--- 2021-06-18 Manuscripts/Manuscript by chapters -for submission/01-Background.Rmd
@@
 Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001] [@roth2017], and cell metabolism [@jones2018] [@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response.
+Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001; @roth2017], and cell metabolism [@jones2018; @skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response.
 Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients [@kamstra2015].
@@
 The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times [@wagenmakers2008], and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004] [@schober2018].
+The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004; @schober2018].
 
@@
 In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011] [@skala2010] [@vishwanath2009].
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011; @skala2010; @vishwanath2009].
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.

### 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/00-Full_document_No_Appendix.tex
Created folder for Wiley submission

--- 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/00-Full_document_OUP_No_Appendix.tex
@@
 \setcounter{secnumdepth}{5}
+\usepackage{helvet}
 \usepackage{lineno}
@@
 
+%line spacing
+\doublespacing
 \usepackage{placeins}
@@
 
-\doublespacing
-
-\hypertarget{summary}{%
-\section{Summary}\label{summary}}
+\hypertarget{abstract}{%
+\section{Abstract}\label{abstract}}
 
@@
 
-cancer biology; tumor response; generalized additive models; simulation; R
+longitudinal data; biomedical data; generalized additive models; simulation; R
 
@@
 Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response (\protect\hyperlink{ref-roblyer2011}{Roblyer \emph{and others} 2011}; \protect\hyperlink{ref-demidov2018}{Demidov \emph{and others} 2018}; \protect\hyperlink{ref-pavlov2018}{Pavlov \emph{and others} 2018}; \protect\hyperlink{ref-tank2020}{Tank \emph{and others} 2020}), antibody expression (\protect\hyperlink{ref-ritter2001}{Ritter \emph{and others} 2001}; \protect\hyperlink{ref-roth2017}{Roth \emph{and others} 2017}), and cell metabolism (\protect\hyperlink{ref-skala2010}{Skala \emph{and others} 2010}; \protect\hyperlink{ref-jones2018}{Jones \emph{and others} 2018}) are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals(\protect\hyperlink{ref-greening2018}{Greening \emph{and others} 2018}), to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients (\protect\hyperlink{ref-sio2016}{Sio \emph{and others} 2016}), to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients (\protect\hyperlink{ref-kamstra2015}{Kamstra \emph{and others} 2015}).
+Tumor response,\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}} antibody expression,\textsuperscript{\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}} and cell metabolism\textsuperscript{\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
+Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals\textsuperscript{\protect\hyperlink{ref-greening2018}{9}}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients,\textsuperscript{\protect\hyperlink{ref-sio2016}{10}} to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients.\textsuperscript{\protect\hyperlink{ref-kamstra2015}{11}}
 
 Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times (\protect\hyperlink{ref-wagenmakers2008}{Wagenmakers \emph{and others} 2008}), and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) (\protect\hyperlink{ref-gueorguieva2004}{Gueorguieva and Krystal 2004}; \protect\hyperlink{ref-schober2018}{Schober and Vetter 2018}).
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}).
-This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
+The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times,\textsuperscript{\protect\hyperlink{ref-wagenmakers2008}{12}} and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
+The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}).\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}}
+
+The expected linear behavior of the response through time is a key requisite in rm-ANOVA.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
 In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings (\protect\hyperlink{ref-vishwanath2009}{Vishwanath \emph{and others} 2009}; \protect\hyperlink{ref-skala2010}{Skala \emph{and others} 2010}; \protect\hyperlink{ref-roblyer2011}{Roblyer \emph{and others} 2011}).
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings.\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}} These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
 Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
@@
 Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs(\protect\hyperlink{ref-dennis2019}{Dennis \emph{and others} 2019}).
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid (\protect\hyperlink{ref-wang2019}{Wang \emph{and others} 2019}).
-
-Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) (\protect\hyperlink{ref-liu2010}{Liu \emph{and others} 2010}; \protect\hyperlink{ref-halsey2015}{Halsey \emph{and others} 2015}), consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni (\protect\hyperlink{ref-abdi2010}{Abdi 2010})), but they in turn reduce statistical power (1-\(\beta\))(\protect\hyperlink{ref-nakagawa2004}{Nakagawa 2004}), and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) (\protect\hyperlink{ref-gelman2012}{Gelman \emph{and others} 2012}; \protect\hyperlink{ref-albers2019}{Albers 2019}).
-Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases (\protect\hyperlink{ref-ugrinowitsch2004}{Ugrinowitsch \emph{and others} 2004}).
-Corrections can be made in rm-ANOVA in the absence of compound symmetry (\protect\hyperlink{ref-greenhouse1959}{Greenhouse and Geisser 1959}; \protect\hyperlink{ref-huynh1976}{Huynh and Feldt 1976}), but the effectiveness of the correction is limited by the size of the sample, the number of measurements(\protect\hyperlink{ref-haverkamp2017}{Haverkamp and Beauducel 2017}), and group sizes (\protect\hyperlink{ref-keselman2001}{Keselman \emph{and others} 2001}).
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons (\protect\hyperlink{ref-charan2013}{Charan and Kantharia 2013}) which might cause the corrections for lack of compound symmetry to be ineffective.
+In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}}.
+For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid.\textsuperscript{\protect\hyperlink{ref-wang2019}{18}}
+
+Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)),\textsuperscript{\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}} consequently biasing the conclusions of the study.
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni),\textsuperscript{\protect\hyperlink{ref-abdi2010}{21}} but they in turn reduce statistical power (1-\(\beta\))\textsuperscript{\protect\hyperlink{ref-nakagawa2004}{22}}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false).\textsuperscript{\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}} Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
+
+On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases.\textsuperscript{\protect\hyperlink{ref-ugrinowitsch2004}{25}} Corrections can be made in rm-ANOVA in the absence of compound symmetry,\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}} but the effectiveness of the correction is limited by the size of the sample, the number of measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and group sizes.\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}} In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons\textsuperscript{\protect\hyperlink{ref-charan2013}{30}} which might cause the corrections for lack of compound symmetry to be ineffective.
 
@@
 In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis (\protect\hyperlink{ref-gueorguieva2004}{Gueorguieva and Krystal 2004}).
-This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
+When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13}} This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
 At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
@@
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data (\protect\hyperlink{ref-vishwanath2009}{Vishwanath \emph{and others} 2009}; \protect\hyperlink{ref-skala2010}{Skala \emph{and others} 2010}).
-Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}).
-However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent (\protect\hyperlink{ref-gueorguieva2004}{Gueorguieva and Krystal 2004}; \protect\hyperlink{ref-barr2013}{Barr \emph{and others} 2013}).
-And even more importantly, LMEMs also assume a linear relationship between the response and time (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}), making them unsuitable to analyze non-linear data.
+The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data.\textsuperscript{\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}} Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}} And even more importantly, LMEMs also assume a linear relationship between the response and time,\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} making them unsuitable to analyze non-linear data.
 
@@
 Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data (\protect\hyperlink{ref-rose2012}{Rose \emph{and others} 2012}; \protect\hyperlink{ref-simpson2018}{Simpson 2018}; \protect\hyperlink{ref-pedersen2019}{Pedersen \emph{and others} 2019}), health-environment interactions (\protect\hyperlink{ref-yang2012}{Yang \emph{and others} 2012}) and the dynamics of government in political science (\protect\hyperlink{ref-beck1998}{Beck and Jackman 1998}).
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements (\protect\hyperlink{ref-wood2017}{Wood 2017}) and 3) can easily accommodate missing observations.
+Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data,\textsuperscript{\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}} health-environment interactions\textsuperscript{\protect\hyperlink{ref-yang2012}{35}} and the dynamics of government in political science.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and 3) can easily accommodate missing observations.
 Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
@@
 The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, \(\textsf{R}\) (\protect\hyperlink{ref-r}{R Core Team 2020}) has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} (\protect\hyperlink{ref-wood2016}{Wood \emph{and others} 2016}; \protect\hyperlink{ref-wood2017}{Wood 2017}) that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models (\protect\hyperlink{ref-haverkamp2017}{Haverkamp and Beauducel 2017}).
-Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
+In particular, \(\textsf{R}\)\textsuperscript{\protect\hyperlink{ref-r}{38}} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models.\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}} Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
 
@@
 Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies (\protect\hyperlink{ref-vishwanath2009}{Vishwanath \emph{and others} 2009}).
-The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
+Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies.\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
 Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in \(\textsf{R}\), in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
@@
 
-A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}; \protect\hyperlink{ref-west2014}{West \emph{and others} 2014}).
-A LMEM with interaction between time and treatment for a longitudinal study can be written as:
+A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}} A LMEM with interaction between time and treatment for a longitudinal study can be written as:
 
@@
 The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level(\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}).
+In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
 However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
@@
 In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study(\protect\hyperlink{ref-wolfinger1996}{Wolfinger 1996}) (For an in-depth analysis of the covariance matrix see (\protect\hyperlink{ref-weiss2005}{Weiss 2005}; \protect\hyperlink{ref-west2014}{West \emph{and others} 2014})).
+The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study\textsuperscript{\protect\hyperlink{ref-wolfinger1996}{41}} (For an in-depth analysis of the covariance matrix see).\textsuperscript{\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}}
 
 In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time (\protect\hyperlink{ref-geisser1958}{Geisser and Greenhouse 1958}; \protect\hyperlink{ref-huynh1976}{Huynh and Feldt 1976}; \protect\hyperlink{ref-weiss2005}{Weiss 2005}).
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time (\protect\hyperlink{ref-maxwell2017}{Maxwell \emph{and others} 2017}); and it is higher between consecutive measurements (\protect\hyperlink{ref-gueorguieva2004}{Gueorguieva and Krystal 2004}; \protect\hyperlink{ref-ugrinowitsch2004}{Ugrinowitsch \emph{and others} 2004}).
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)(\protect\hyperlink{ref-greenhouse1959}{Greenhouse and Geisser 1959}; \protect\hyperlink{ref-huynh1976}{Huynh and Feldt 1976}) the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements(\protect\hyperlink{ref-haverkamp2017}{Haverkamp and Beauducel 2017}), and they are not robust if the group sizes are unbalanced (\protect\hyperlink{ref-keselman2001}{Keselman \emph{and others} 2001}).
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others (\protect\hyperlink{ref-pinheiro2006}{Pinheiro and Bates 2006}).
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
+Under this assumption, the between-subject variance and within-subject correlation are constant across time.\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}} However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time;\textsuperscript{\protect\hyperlink{ref-maxwell2017}{44}} and it is higher between consecutive measurements.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}} Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and they are not robust if the group sizes are unbalanced.\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}} Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
+
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
 Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
@@
 Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator (\protect\hyperlink{ref-molenberghs2004}{Molenberghs 2004}).
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) (\protect\hyperlink{ref-weiss2005}{Weiss 2005}).
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest (\protect\hyperlink{ref-scheffer2002}{Scheffer 2002}).
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information (\protect\hyperlink{ref-potthoff2006}{Potthoff \emph{and others} 2006}), and in the case of MNAR the missing values are dependent on their value.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator.\textsuperscript{\protect\hyperlink{ref-molenberghs2004}{45}} Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
+Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR).\textsuperscript{\protect\hyperlink{ref-weiss2005}{42}} In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest.\textsuperscript{\protect\hyperlink{ref-scheffer2002}{46}} If the data are MCAR, this means that the missingness is completely unrelated to the collected information,\textsuperscript{\protect\hyperlink{ref-potthoff2006}{47}} and in the case of MNAR the missing values are dependent on their value.
 
 An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size (\protect\hyperlink{ref-ma2012}{Ma \emph{and others} 2012}).
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR (\protect\hyperlink{ref-west2014}{West \emph{and others} 2014}).
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size.\textsuperscript{\protect\hyperlink{ref-ma2012}{48}} In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR.\textsuperscript{\protect\hyperlink{ref-west2014}{40}} For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
 
@@
 To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) (\protect\hyperlink{ref-r}{R Core Team 2020}) and the package \emph{nlme} (\protect\hyperlink{ref-nlme}{Pinheiro \emph{and others} 2020}).
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\)\textsuperscript{\protect\hyperlink{ref-r}{38}} and the package \emph{nlme}.\textsuperscript{\protect\hyperlink{ref-nlme}{49}}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_No_Appendix_files/figure-latex/l-q-response-1}
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/l-q-response-1}
 
@@
 
-Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors (\protect\hyperlink{ref-nelder1972}{Nelder and Wedderburn 1972}).
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case(\protect\hyperlink{ref-hastie1987}{Hastie and Tibshirani 1987}; \protect\hyperlink{ref-wood2017}{Wood 2017}; \protect\hyperlink{ref-simpson2018}{Simpson 2018}).
+Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors.\textsuperscript{\protect\hyperlink{ref-nelder1972}{50}} In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{51}}.
 A GAM model can be written as:
@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference (\protect\hyperlink{ref-beck1998}{Beck and Jackman 1998}).
-In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
+One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
 
 The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist (\protect\hyperlink{ref-simpson2018}{Simpson 2018}).
-For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response (\protect\hyperlink{ref-hefley2017}{Hefley \emph{and others} 2017}), which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{52}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
 In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data (\protect\hyperlink{ref-wood2017}{Wood 2017}; \protect\hyperlink{ref-simpson2018}{Simpson 2018}).
-Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness (\protect\hyperlink{ref-wegman1983}{Wegman and Wright 1983}).
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
+Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{53}} Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_No_Appendix_files/figure-latex/basis-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/basis-plot-1}
 
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief (\protect\hyperlink{ref-mcelreath2018}{McElreath 2018}).
-In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs (\protect\hyperlink{ref-wood2017}{Wood 2017}).
-GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) (\protect\hyperlink{ref-miller2019}{Miller 2019}).
-Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model (\protect\hyperlink{ref-laird1982}{Laird and Ware 1982}; \protect\hyperlink{ref-pedersen2019}{Pedersen \emph{and others} 2019}).
-Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals (\protect\hyperlink{ref-pedersen2019}{Pedersen \emph{and others} 2019}), which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage (\protect\hyperlink{ref-wood2017}{Wood 2017}).
-This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in (\protect\hyperlink{ref-wood2017}{Wood 2017}; \protect\hyperlink{ref-simpson2018}{Simpson 2018}; \protect\hyperlink{ref-miller2019}{Miller 2019}) and (\protect\hyperlink{ref-marra2012}{Marra and Wood 2012}).
-With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{54}} In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case).\textsuperscript{\protect\hyperlink{ref-miller2019}{55}} Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model.\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{56}} Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{55}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{57}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
 
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in (\protect\hyperlink{ref-vishwanath2009}{Vishwanath \emph{and others} 2009}).
-In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in.\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
 In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group.
@@
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in (\protect\hyperlink{ref-hefley2017}{Hefley \emph{and others} 2017}).
+A detailed description on basis functions and correlations can be found in.\textsuperscript{\protect\hyperlink{ref-hefley2017}{52}}
 
@@
 This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
-The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths (\protect\hyperlink{ref-simpson2018}{Simpson 2018}).
-The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) (\protect\hyperlink{ref-wood2017}{Wood 2017}).
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} (\protect\hyperlink{ref-gratia}{Simpson 2020}).
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in (\protect\hyperlink{ref-wood2017}{Wood 2017}) and (\protect\hyperlink{ref-harezlak2018}{Harezlak \emph{and others} 2018}).
+The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{58}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{59}}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_No_Appendix_files/figure-latex/sim-smooth-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_OUP_No_Appendix_files/figure-latex/plot-pairwise-comp-1}
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/plot-pairwise-comp-1}
 
@@
 Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}.
-This ``model misspecification'' error, also is known as a ``Type III'' error (\protect\hyperlink{ref-dennis2019}{Dennis \emph{and others} 2019}) is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) (\protect\hyperlink{ref-lang2015}{Lang and Altman 2015}) but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{60}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
 
 In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) (\protect\hyperlink{ref-hefley2017}{Hefley \emph{and others} 2017}; \protect\hyperlink{ref-pedersen2019}{Pedersen \emph{and others} 2019}) they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research (\protect\hyperlink{ref-hastie1995}{Hastie and Tibshirani 1995}).
-This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{61}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
 
 However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature (\protect\hyperlink{ref-vishwanath2009}{Vishwanath \emph{and others} 2009}) using \(\textsf{R}\) and the package \emph{mgcv}(\protect\hyperlink{ref-wood2017}{Wood 2017}) in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} using \(\textsf{R}\) and the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
 One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research (\protect\hyperlink{ref-begley2015}{Begley and Ioannidis 2015}; \protect\hyperlink{ref-weissgerber2018}{Weissgerber \emph{and others} 2018}), but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{62},\protect\hyperlink{ref-weissgerber2018}{63}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
+\begin{CSLReferences}{0}{0}
+\leavevmode\hypertarget{ref-roblyer2011}{}%
+\CSLLeftMargin{1. }
+\CSLRightInline{Roblyer D, Ueda S, Cerussi A, et al.
+Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment.
+\emph{Proceedings of the National Academy of Sciences}.
+2011;108(35):14626-14631.
+doi:\href{https://doi.org/10.1073/pnas.1013103108}{10.1073/pnas.1013103108}}
+
+\leavevmode\hypertarget{ref-tank2020}{}%
+\CSLLeftMargin{2. }
+\CSLRightInline{Tank A, Peterson HM, Pera V, et al.
+Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens.
+\emph{Breast Cancer Research}.
+2020;22(1).
+doi:\href{https://doi.org/10.1186/s13058-020-01262-1}{10.1186/s13058-020-01262-1}}
+
+\leavevmode\hypertarget{ref-pavlov2018}{}%
+\CSLLeftMargin{3. }
+\CSLRightInline{Pavlov MV, Kalganova TI, Lyubimtseva YS.
+Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy.
+\emph{Journal of Biomedical Optics}.
+2018;23(09):1.
+doi:\href{https://doi.org/10.1117/1.jbo.23.9.091410}{10.1117/1.jbo.23.9.091410}}
+
+\leavevmode\hypertarget{ref-demidov2018}{}%
+\CSLLeftMargin{4. }
+\CSLRightInline{Demidov V, Maeda A, Sugita M, et al.
+Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography.
+\emph{Scientific Reports}.
+2018;8(1).
+doi:\href{https://doi.org/10.1038/s41598-017-18635-w}{10.1038/s41598-017-18635-w}}
+
+\leavevmode\hypertarget{ref-ritter2001}{}%
+\CSLLeftMargin{5. }
+\CSLRightInline{Ritter G, Cohen L, Williams C, Richards E, Old L, Welt S.
+{Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}.
+\emph{{Cancer Research}}.
+2001;{61}({18}):6851-6859.}
+
+\leavevmode\hypertarget{ref-roth2017}{}%
+\CSLLeftMargin{6. }
+\CSLRightInline{Roth EM, Goldberg AC, Catapano AL, et al.
+Antidrug antibodies in patients treated with alirocumab.
+\emph{New England Journal of Medicine}.
+2017;376(16):1589-1590.
+doi:\href{https://doi.org/10.1056/nejmc1616623}{10.1056/nejmc1616623}}
+
+\leavevmode\hypertarget{ref-jones2018}{}%
+\CSLLeftMargin{7. }
+\CSLRightInline{Jones JD, Ramser HE, Woessner AE, Quinn KP.
+In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing.
+\emph{Communications Biology}.
+2018;1(1).
+doi:\href{https://doi.org/10.1038/s42003-018-0206-4}{10.1038/s42003-018-0206-4}}
+
+\leavevmode\hypertarget{ref-skala2010}{}%
+\CSLLeftMargin{8. }
+\CSLRightInline{Skala MC, Fontanella A, Lan L, Izatt JA, Dewhirst MW.
+Longitudinal optical imaging of tumor metabolism and hemodynamics.
+\emph{Journal of Biomedical Optics}.
+2010;15(1):011112.
+doi:\href{https://doi.org/10.1117/1.3285584}{10.1117/1.3285584}}
+
+\leavevmode\hypertarget{ref-greening2018}{}%
+\CSLLeftMargin{9. }
+\CSLRightInline{Greening GJ, Miller KP, Spainhour CR, Cato MD, Muldoon TJ.
+Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy.
+\emph{Biomedical Optics Express}.
+2018;9(6):2871.
+doi:\href{https://doi.org/10.1364/boe.9.002871}{10.1364/boe.9.002871}}
+
+\leavevmode\hypertarget{ref-sio2016}{}%
+\CSLLeftMargin{10. }
+\CSLRightInline{Sio TT, Atherton PJ, Birckhead BJ, et al.
+Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}).
+\emph{Supportive Care in Cancer}.
+2016;24(9):3847-3855.
+doi:\href{https://doi.org/10.1007/s00520-016-3213-3}{10.1007/s00520-016-3213-3}}
+
+\leavevmode\hypertarget{ref-kamstra2015}{}%
+\CSLLeftMargin{11. }
+\CSLRightInline{Kamstra JI, Dijkstra PU, Leeuwen M van, Roodenburg JLN, Langendijk JA.
+Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study.
+\emph{Oral Oncology}.
+2015;51(5):548-555.
+doi:\href{https://doi.org/10.1016/j.oraloncology.2015.01.016}{10.1016/j.oraloncology.2015.01.016}}
+
+\leavevmode\hypertarget{ref-wagenmakers2008}{}%
+\CSLLeftMargin{12. }
+\CSLRightInline{Wagenmakers E-J, Lee M, Lodewyckx T, Iverson GJ.
+Bayesian versus frequentist inference.
+In: \emph{Bayesian Evaluation of Informative Hypotheses}.
+Springer New York; 2008:181-207.
+doi:\href{https://doi.org/10.1007/978-0-387-09612-4_9}{10.1007/978-0-387-09612-4\_9}}
+
+\leavevmode\hypertarget{ref-gueorguieva2004}{}%
+\CSLLeftMargin{13. }
+\CSLRightInline{Gueorguieva R, Krystal JH. Move over {ANOVA}. \emph{Archives of General Psychiatry}. 2004;61(3):310. doi:\href{https://doi.org/10.1001/archpsyc.61.3.310}{10.1001/archpsyc.61.3.310}}
+
+\leavevmode\hypertarget{ref-schober2018}{}%
+\CSLLeftMargin{14. }
+\CSLRightInline{Schober P, Vetter TR.
+Repeated measures designs and analysis of longitudinal data.
+\emph{Anesthesia {\&} Analgesia}.
+2018;127(2):569-575.
+doi:\href{https://doi.org/10.1213/ane.0000000000003511}{10.1213/ane.0000000000003511}}
+
+\leavevmode\hypertarget{ref-pinheiro2006}{}%
+\CSLLeftMargin{15. }
+\CSLRightInline{Pinheiro J, Bates D. \emph{{Mixed-effects models in S and S-PLUS}}. Springer Science; Business Media; 2006. doi:\url{https://doi.org/10.1007/b98882}}
+
+\leavevmode\hypertarget{ref-vishwanath2009}{}%
+\CSLLeftMargin{16. }
+\CSLRightInline{Vishwanath K, Yuan H, Barry WT, Dewhirst MW, Ramanujam N.
+Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors.
+\emph{Neoplasia}.
+2009;11(9):889-900.
+doi:\href{https://doi.org/10.1593/neo.09580}{10.1593/neo.09580}}
+
+\leavevmode\hypertarget{ref-dennis2019}{}%
+\CSLLeftMargin{17. }
+\CSLRightInline{Dennis B, Ponciano JM, Taper ML, Lele SR.
+Errors in statistical inference under model misspecification: Evidence, hypothesis testing, and {AIC}.
+\emph{Frontiers in Ecology and Evolution}.
+2019;7.
+doi:\href{https://doi.org/10.3389/fevo.2019.00372}{10.3389/fevo.2019.00372}}
+
+\leavevmode\hypertarget{ref-wang2019}{}%
+\CSLLeftMargin{18. }
+\CSLRightInline{Wang B, Zhou Z, Wang H, Tu XM, Feng C.
+The p-value and model specification in statistics.
+\emph{General Psychiatry}.
+2019;32(3):e100081.
+doi:\href{https://doi.org/10.1136/gpsych-2019-100081}{10.1136/gpsych-2019-100081}}
+
+\leavevmode\hypertarget{ref-liu2010}{}%
+\CSLLeftMargin{19. }
+\CSLRightInline{Liu C, Cripe TP, Kim M-O.
+Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences.
+\emph{Molecular Therapy}.
+2010;18(9):1724-1730.
+doi:\href{https://doi.org/10.1038/mt.2010.127}{10.1038/mt.2010.127}}
+
+\leavevmode\hypertarget{ref-halsey2015}{}%
+\CSLLeftMargin{20. }
+\CSLRightInline{Halsey LG, Curran-Everett D, Vowler SL, Drummond GB.
+The fickle p value generates irreproducible results.
+\emph{Nature Methods}.
+2015;12(3):179-185.
+doi:\href{https://doi.org/10.1038/nmeth.3288}{10.1038/nmeth.3288}}
+
 \leavevmode\hypertarget{ref-abdi2010}{}%
-\textsc{Abdi, H.} (2010). {Holm's sequential Bonferroni procedure}. \emph{Encyclopedia of Research Design} \textbf{1}, 1--8.
+\CSLLeftMargin{21. }
+\CSLRightInline{Abdi H.
+{Holm's sequential Bonferroni procedure}.
+\emph{Encyclopedia of Research Design}.
+2010;1(8):1-8.
+doi:\href{https://doi.org/10.4135/9781412961288.n178}{10.4135/9781412961288.n178}}
+
+\leavevmode\hypertarget{ref-nakagawa2004}{}%
+\CSLLeftMargin{22. }
+\CSLRightInline{Nakagawa S.
+A farewell to bonferroni: The problems of low statistical power and publication bias.
+\emph{Behavioral Ecology}.
+2004;15(6):1044-1045.
+doi:\href{https://doi.org/10.1093/beheco/arh107}{10.1093/beheco/arh107}}
+
+\leavevmode\hypertarget{ref-gelman2012}{}%
+\CSLLeftMargin{23. }
+\CSLRightInline{Gelman A, Hill J, Yajima M.
+Why we (usually) don{{}}t have to worry about multiple comparisons.
+\emph{Journal of Research on Educational Effectiveness}.
+2012;5(2):189-211.
+doi:\href{https://doi.org/10.1080/19345747.2011.618213}{10.1080/19345747.2011.618213}}
 
 \leavevmode\hypertarget{ref-albers2019}{}%
-\textsc{Albers, C.} (2019).
+\CSLLeftMargin{24. }
+\CSLRightInline{Albers C.
 The problem with unadjusted multiple and sequential statistical testing.
-\emph{Nature Communications} \textbf{10}.
-doi:\href{https://doi.org/10.1038/s41467-019-09941-0}{10.1038/s41467-019-09941-0}.
+\emph{Nature Communications}.
+2019;10(1).
+doi:\href{https://doi.org/10.1038/s41467-019-09941-0}{10.1038/s41467-019-09941-0}}
+
+\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
+\CSLLeftMargin{25. }
+\CSLRightInline{Ugrinowitsch C, Fellingham GW, Ricard MD.
+Limitations of ordinary least squares models in analyzing repeated measures data.
+\emph{Medicine {\&} Science in Sports {\&} Exercise}.
+Published online December 2004:2144-2148.
+doi:\href{https://doi.org/10.1249/01.mss.0000147580.40591.75}{10.1249/01.mss.0000147580.40591.75}}
+
+\leavevmode\hypertarget{ref-huynh1976}{}%
+\CSLLeftMargin{26. }
+\CSLRightInline{Huynh H, Feldt LS.
+Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs.
+\emph{Journal of Educational Statistics}.
+1976;1(1):69-82.
+doi:\href{https://doi.org/10.3102/10769986001001069}{10.3102/10769986001001069}}
+
+\leavevmode\hypertarget{ref-greenhouse1959}{}%
+\CSLLeftMargin{27. }
+\CSLRightInline{Greenhouse SW, Geisser S. On methods in the analysis of profile data. \emph{Psychometrika}. 1959;24(2):95-112. doi:\href{https://doi.org/10.1007/bf02289823}{10.1007/bf02289823}}
+
+\leavevmode\hypertarget{ref-haverkamp2017}{}%
+\CSLLeftMargin{28. }
+\CSLRightInline{Haverkamp N, Beauducel A.
+Violation of the sphericity assumption and its effect on type-i error rates in repeated measures {ANOVA} and multi-level linear models ({MLM}).
+\emph{Frontiers in Psychology}.
+2017;8.
+doi:\href{https://doi.org/10.3389/fpsyg.2017.01841}{10.3389/fpsyg.2017.01841}}
+
+\leavevmode\hypertarget{ref-keselman2001}{}%
+\CSLLeftMargin{29. }
+\CSLRightInline{Keselman HJ, Algina J, Kowalchuk RK.
+The analysis of repeated measures designs: A review.
+\emph{British Journal of Mathematical and Statistical Psychology}.
+2001;54(1):1-20.
+doi:\href{https://doi.org/10.1348/000711001159357}{10.1348/000711001159357}}
+
+\leavevmode\hypertarget{ref-charan2013}{}%
+\CSLLeftMargin{30. }
+\CSLRightInline{Charan J, Kantharia N.
+How to calculate sample size in animal studies?
+\emph{Journal of Pharmacology and Pharmacotherapeutics}.
+2013;4(4):303.
+doi:\href{https://doi.org/10.4103/0976-500x.119726}{10.4103/0976-500x.119726}}
 
 \leavevmode\hypertarget{ref-barr2013}{}%
-\textsc{Barr, D.
-J., Levy, R., Scheepers, C.
-and Tily, H.
-J.} (2013).
+\CSLLeftMargin{31. }
+\CSLRightInline{Barr DJ, Levy R, Scheepers C, Tily HJ.
 Random effects structure for confirmatory hypothesis testing: Keep it maximal.
-\emph{Journal of Memory and Language} \textbf{68}, 255--278.
+\emph{Journal of Memory and Language}.
+2013;68(3):255-278.
+doi:\href{https://doi.org/10.1016/j.jml.2012.11.001}{10.1016/j.jml.2012.11.001}}
+
+\leavevmode\hypertarget{ref-rose2012}{}%
+\CSLLeftMargin{32. }
+\CSLRightInline{Rose NL, Yang H, Turner SD, Simpson GL.
+An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, {UK}.
+\emph{Geochimica et Cosmochimica Acta}.
+2012;82:113-135.
+doi:\href{https://doi.org/10.1016/j.gca.2010.12.026}{10.1016/j.gca.2010.12.026}}
+
+\leavevmode\hypertarget{ref-pedersen2019}{}%
+\CSLLeftMargin{33. }
+\CSLRightInline{Pedersen EJ, Miller DL, Simpson GL, Ross N.
+Hierarchical generalized additive models in ecology: An introduction with mgcv.
+\emph{{PeerJ}}.
+2019;7:e6876.
+doi:\href{https://doi.org/10.7717/peerj.6876}{10.7717/peerj.6876}}
+
+\leavevmode\hypertarget{ref-simpson2018}{}%
+\CSLLeftMargin{34. }
+\CSLRightInline{Simpson GL.
+Modelling palaeoecological time series using generalised additive models.
+\emph{Frontiers in Ecology and Evolution}.
+2018;6.
+doi:\href{https://doi.org/10.3389/fevo.2018.00149}{10.3389/fevo.2018.00149}}
+
+\leavevmode\hypertarget{ref-yang2012}{}%
+\CSLLeftMargin{35. }
+\CSLRightInline{Yang L, Qin G, Zhao N, Wang C, Song G.
+Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality.
+\emph{{BMC} Medical Research Methodology}.
+2012;12(1).
+doi:\href{https://doi.org/10.1186/1471-2288-12-165}{10.1186/1471-2288-12-165}}
 
 \leavevmode\hypertarget{ref-beck1998}{}%
-\textsc{Beck, N. and Jackman, S.} (1998). Beyond linearity by default: Generalized additive models. \emph{American Journal of Political Science} \textbf{42}, 596.
+\CSLLeftMargin{36. }
+\CSLRightInline{Beck N, Jackman S.
+Beyond linearity by default: Generalized additive models.
+\emph{American Journal of Political Science}.
+1998;42(2):596.
+doi:\href{https://doi.org/10.2307/2991772}{10.2307/2991772}}
+
+\leavevmode\hypertarget{ref-wood2017}{}%
+\CSLLeftMargin{37. }
+\CSLRightInline{Wood SN. \emph{Generalized Additive Models}. Chapman; Hall/{CRC}; 2017. doi:\href{https://doi.org/10.1201/9781315370279}{10.1201/9781315370279}}
+
+\leavevmode\hypertarget{ref-r}{}%
+\CSLLeftMargin{38. }
+\CSLRightInline{R Core Team. \emph{R: A Language and Environment for Statistical Computing}. R Foundation for Statistical Computing; 2020. \url{https://www.R-project.org/}}
+
+\leavevmode\hypertarget{ref-wood2016}{}%
+\CSLLeftMargin{39. }
+\CSLRightInline{Wood SN, Pya N, Säfken B.
+Smoothing parameter and model selection for general smooth models.
+\emph{Journal of the American Statistical Association}.
+2016;111(516):1548-1563.
+doi:\href{https://doi.org/10.1080/01621459.2016.1180986}{10.1080/01621459.2016.1180986}}
+
+\leavevmode\hypertarget{ref-west2014}{}%
+\CSLLeftMargin{40. }
+\CSLRightInline{West BT, Welch KB, Galecki AT.
+\emph{Linear Mixed Models: A Practical Guide Using Statistical Software, Second Edition}.
+Taylor \& Francis; 2014.
+\url{https://books.google.com/books?id=hjT6AwAAQBAJ}}
+
+\leavevmode\hypertarget{ref-wolfinger1996}{}%
+\CSLLeftMargin{41. }
+\CSLRightInline{Wolfinger RD.
+Heterogeneous variance: Covariance structures for repeated measures.
+\emph{Journal of Agricultural, Biological, and Environmental Statistics}.
+1996;1(2):205.
+doi:\href{https://doi.org/10.2307/1400366}{10.2307/1400366}}
+
+\leavevmode\hypertarget{ref-weiss2005}{}%
+\CSLLeftMargin{42. }
+\CSLRightInline{Weiss RE. \emph{Modeling Longitudinal Data}. Springer New York; 2005. doi:\href{https://doi.org/10.1007/0-387-28314-5}{10.1007/0-387-28314-5}}
+
+\leavevmode\hypertarget{ref-geisser1958}{}%
+\CSLLeftMargin{43. }
+\CSLRightInline{Geisser S, Greenhouse SW.
+An extension of box{{}}s results on the use of the {{}}f{{}} distribution in multivariate analysis.
+\emph{The Annals of Mathematical Statistics}.
+1958;29(3):885-891.
+doi:\href{https://doi.org/10.1214/aoms/1177706545}{10.1214/aoms/1177706545}}
+
+\leavevmode\hypertarget{ref-maxwell2017}{}%
+\CSLLeftMargin{44. }
+\CSLRightInline{Maxwell SE, Delaney HD, Kelley K. \emph{Designing Experiments and Analyzing Data}. Routledge; 2017. doi:\href{https://doi.org/10.4324/9781315642956}{10.4324/9781315642956}}
+
+\leavevmode\hypertarget{ref-molenberghs2004}{}%
+\CSLLeftMargin{45. }
+\CSLRightInline{Molenberghs G.
+Analyzing incomplete longitudinal clinical trial data.
+\emph{Biostatistics}.
+2004;5(3):445-464.
+doi:\href{https://doi.org/10.1093/biostatistics/kxh001}{10.1093/biostatistics/kxh001}}
+
+\leavevmode\hypertarget{ref-scheffer2002}{}%
+\CSLLeftMargin{46. }
+\CSLRightInline{Scheffer J. Dealing with missing data. \emph{Research Letters in the Information and Mathematical Sciences}. 2002;3:153-160.}
+
+\leavevmode\hypertarget{ref-potthoff2006}{}%
+\CSLLeftMargin{47. }
+\CSLRightInline{Potthoff RF, Tudor GE, Pieper KS, Hasselblad V.
+Can one assess whether missing data are missing at random in medical studies?
+\emph{Statistical Methods in Medical Research}.
+2006;15(3):213-234.
+doi:\href{https://doi.org/10.1191/0962280206sm448oa}{10.1191/0962280206sm448oa}}
+
+\leavevmode\hypertarget{ref-ma2012}{}%
+\CSLLeftMargin{48. }
+\CSLRightInline{Ma Y, Mazumdar M, Memtsoudis SG.
+Beyond repeated-measures analysis of variance.
+\emph{Regional Anesthesia and Pain Medicine}.
+2012;37(1):99-105.
+doi:\href{https://doi.org/10.1097/aap.0b013e31823ebc74}{10.1097/aap.0b013e31823ebc74}}
+
+\leavevmode\hypertarget{ref-nlme}{}%
+\CSLLeftMargin{49. }
+\CSLRightInline{Pinheiro J, Bates D, DebRoy S, Sarkar D, R Core Team. \emph{{nlme}: Linear and Nonlinear Mixed Effects Models}.; 2020. \url{https://CRAN.R-project.org/package=nlme}}
+
+\leavevmode\hypertarget{ref-nelder1972}{}%
+\CSLLeftMargin{50. }
+\CSLRightInline{Nelder JA, Wedderburn RWM.
+Generalized linear models.
+\emph{Journal of the Royal Statistical Society Series A (General)}.
+1972;135(3):370.
+doi:\href{https://doi.org/10.2307/2344614}{10.2307/2344614}}
+
+\leavevmode\hypertarget{ref-hastie1987}{}%
+\CSLLeftMargin{51. }
+\CSLRightInline{Hastie T, Tibshirani R.
+Generalized additive models: Some applications.
+\emph{Journal of the American Statistical Association}.
+1987;82(398):371-386.
+doi:\href{https://doi.org/10.1080/01621459.1987.10478440}{10.1080/01621459.1987.10478440}}
+
+\leavevmode\hypertarget{ref-hefley2017}{}%
+\CSLLeftMargin{52. }
+\CSLRightInline{Hefley TJ, Broms KM, Brost BM, et al.
+The basis function approach for modeling autocorrelation in ecological data.
+\emph{Ecology}.
+2017;98(3):632-646.
+doi:\href{https://doi.org/10.1002/ecy.1674}{10.1002/ecy.1674}}
+
+\leavevmode\hypertarget{ref-wegman1983}{}%
+\CSLLeftMargin{53. }
+\CSLRightInline{Wegman EJ, Wright IW.
+Splines in statistics.
+\emph{Journal of the American Statistical Association}.
+1983;78(382):351-365.
+doi:\href{https://doi.org/10.1080/01621459.1983.10477977}{10.1080/01621459.1983.10477977}}
+
+\leavevmode\hypertarget{ref-mcelreath2018}{}%
+\CSLLeftMargin{54. }
+\CSLRightInline{McElreath R. \emph{Statistical Rethinking}. Chapman; Hall/{CRC}; 2018. doi:\href{https://doi.org/10.1201/9781315372495}{10.1201/9781315372495}}
+
+\leavevmode\hypertarget{ref-miller2019}{}%
+\CSLLeftMargin{55. }
+\CSLRightInline{Miller DL. Bayesian views of generalized additive modelling. \emph{arXiv preprint arXiv:190201330}. Published online 2019.}
+
+\leavevmode\hypertarget{ref-laird1982}{}%
+\CSLLeftMargin{56. }
+\CSLRightInline{Laird NM, Ware JH. Random-effects models for longitudinal data. \emph{Biometrics}. 1982;38(4):963. doi:\href{https://doi.org/10.2307/2529876}{10.2307/2529876}}
+
+\leavevmode\hypertarget{ref-marra2012}{}%
+\CSLLeftMargin{57. }
+\CSLRightInline{Marra G, Wood SN.
+Coverage properties of confidence intervals for generalized additive model components.
+\emph{Scandinavian Journal of Statistics}.
+2012;39(1):53-74.
+doi:\href{https://doi.org/10.1111/j.1467-9469.2011.00760.x}{10.1111/j.1467-9469.2011.00760.x}}
+
+\leavevmode\hypertarget{ref-gratia}{}%
+\CSLLeftMargin{58. }
+\CSLRightInline{Simpson GL. \emph{Gratia: Graceful 'Ggplot'-Based Graphics and Other Functions for GAMs Fitted Using 'Mgcv'}.; 2020. \url{https://CRAN.R-project.org/package=gratia}}
+
+\leavevmode\hypertarget{ref-harezlak2018}{}%
+\CSLLeftMargin{59. }
+\CSLRightInline{Harezlak J, Ruppert D, Wand MP. \emph{Semiparametric Regression with r}. Springer New York; 2018. doi:\href{https://doi.org/10.1007/978-1-4939-8853-2}{10.1007/978-1-4939-8853-2}}
+
+\leavevmode\hypertarget{ref-lang2015}{}%
+\CSLLeftMargin{60. }
+\CSLRightInline{Lang TA, Altman DG.
+Basic statistical reporting for articles published in biomedical journals: The {``}statistical analyses and methods in the published literature{''} or the {SAMPL} guidelines.
+\emph{International Journal of Nursing Studies}.
+2015;52(1):5-9.
+doi:\href{https://doi.org/10.1016/j.ijnurstu.2014.09.006}{10.1016/j.ijnurstu.2014.09.006}}
+
+\leavevmode\hypertarget{ref-hastie1995}{}%
+\CSLLeftMargin{61. }
+\CSLRightInline{Hastie T, Tibshirani R.
+Generalized additive models for medical research.
+\emph{Statistical Methods in Medical Research}.
+1995;4(3):187-196.
+doi:\href{https://doi.org/10.1177/096228029500400302}{10.1177/096228029500400302}}
 
 \leavevmode\hypertarget{ref-begley2015}{}%
-\textsc{Begley, C. G. and Ioannidis, J. P. A.} (2015). Reproducibility in science. \emph{Circulation Research} \textbf{116}, 116--126.
-
-\leavevmode\hypertarget{ref-charan2013}{}%
-\textsc{Charan, J. and Kantharia, N.} (2013). How to calculate sample size in animal studies? \emph{Journal of Pharmacology and Pharmacotherapeutics} \textbf{4}, 303.
-
-\leavevmode\hypertarget{ref-demidov2018}{}%
-\textsc{Demidov, V., Maeda, A., Sugita, M., Madge, V., Sadanand, S., Flueraru, C.
-and Vitkin, I.
-A.} (2018).
-Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography.
-\emph{Scientific Reports} \textbf{8}.
-doi:\href{https://doi.org/10.1038/s41598-017-18635-w}{10.1038/s41598-017-18635-w}.
-
-\leavevmode\hypertarget{ref-dennis2019}{}%
-\textsc{Dennis, B., Ponciano, J.
-M., Taper, M.
-L.
-and Lele, S.
-R.} (2019).
-Errors in statistical inference under model misspecification: Evidence, hypothesis testing, and {AIC}.
-\emph{Frontiers in Ecology and Evolution} \textbf{7}.
-doi:\href{https://doi.org/10.3389/fevo.2019.00372}{10.3389/fevo.2019.00372}.
-
-\leavevmode\hypertarget{ref-geisser1958}{}%
-\textsc{Geisser, S.
-and Greenhouse, S.
-W.} (1958).
-An extension of box{{}}s results on the use of the {{}}f{{}} distribution in multivariate analysis.
-\emph{The Annals of Mathematical Statistics} \textbf{29}, 885--891.
-
-\leavevmode\hypertarget{ref-gelman2012}{}%
-\textsc{Gelman, A., Hill, J. and Yajima, M.} (2012). Why we (usually) don{{}}t have to worry about multiple comparisons. \emph{Journal of Research on Educational Effectiveness} \textbf{5}, 189--211.
-
-\leavevmode\hypertarget{ref-greenhouse1959}{}%
-\textsc{Greenhouse, S. W. and Geisser, S.} (1959). On methods in the analysis of profile data. \emph{Psychometrika} \textbf{24}, 95--112.
-
-\leavevmode\hypertarget{ref-greening2018}{}%
-\textsc{Greening, G.
-J., Miller, K.
-P., Spainhour, C.
-R., Cato, M.
-D.
-and Muldoon, T.
-J.} (2018).
-Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy.
-\emph{Biomedical Optics Express} \textbf{9}, 2871.
-
-\leavevmode\hypertarget{ref-gueorguieva2004}{}%
-\textsc{Gueorguieva, R. and Krystal, J. H.} (2004). Move over {ANOVA}. \emph{Archives of General Psychiatry} \textbf{61}, 310.
-
-\leavevmode\hypertarget{ref-halsey2015}{}%
-\textsc{Halsey, L. G., Curran-Everett, D., Vowler, S. L. and Drummond, G. B.} (2015). The fickle p value generates irreproducible results. \emph{Nature Methods} \textbf{12}, 179--185.
-
-\leavevmode\hypertarget{ref-harezlak2018}{}%
-\textsc{Harezlak, J., Ruppert, D. and Wand, M. P.} (2018). Semiparametric regression with r. Springer New York. doi:\href{https://doi.org/10.1007/978-1-4939-8853-2}{10.1007/978-1-4939-8853-2}.
-
-\leavevmode\hypertarget{ref-hastie1995}{}%
-\textsc{Hastie, T. and Tibshirani, R.} (1995). Generalized additive models for medical research. \emph{Statistical Methods in Medical Research} \textbf{4}, 187--196.
-
-\leavevmode\hypertarget{ref-hastie1987}{}%
-\textsc{Hastie, T. and Tibshirani, R.} (1987). Generalized additive models: Some applications. \emph{Journal of the American Statistical Association} \textbf{82}, 371--386.
-
-\leavevmode\hypertarget{ref-haverkamp2017}{}%
-\textsc{Haverkamp, N.
-and Beauducel, A.} (2017).
-Violation of the sphericity assumption and its effect on type-i error rates in repeated measures {ANOVA} and multi-level linear models ({MLM}).
-\emph{Frontiers in Psychology} \textbf{8}.
-doi:\href{https://doi.org/10.3389/fpsyg.2017.01841}{10.3389/fpsyg.2017.01841}.
-
-\leavevmode\hypertarget{ref-hefley2017}{}%
-\textsc{Hefley, T.
-J., Broms, K.
-M., Brost, B.
-M., Buderman, F.
-E., Kay, S.
-L., Scharf, H.
-R., Tipton, J.
-R., Williams, P.
-J.
-and Hooten, M.
-B.} (2017).
-The basis function approach for modeling autocorrelation in ecological data.
-\emph{Ecology} \textbf{98}, 632--646.
-
-\leavevmode\hypertarget{ref-huynh1976}{}%
-\textsc{Huynh, H.
-and Feldt, L.
-S.} (1976).
-Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs.
-\emph{Journal of Educational Statistics} \textbf{1}, 69--82.
-
-\leavevmode\hypertarget{ref-jones2018}{}%
-\textsc{Jones, J.
-D., Ramser, H.
-E., Woessner, A.
-E.
-and Quinn, K.
-P.} (2018).
-In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing.
-\emph{Communications Biology} \textbf{1}.
-doi:\href{https://doi.org/10.1038/s42003-018-0206-4}{10.1038/s42003-018-0206-4}.
-
-\leavevmode\hypertarget{ref-kamstra2015}{}%
-\textsc{Kamstra, J.
-I., Dijkstra, P.
-U., Leeuwen, M.
-van, Roodenburg, J.
-L.
-N.
-and Langendijk, J.
-A.} (2015).
-Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study.
-\emph{Oral Oncology} \textbf{51}, 548--555.
-
-\leavevmode\hypertarget{ref-keselman2001}{}%
-\textsc{Keselman, H.
-J., Algina, J.
-and Kowalchuk, R.
-K.} (2001).
-The analysis of repeated measures designs: A review.
-\emph{British Journal of Mathematical and Statistical Psychology} \textbf{54}, 1--20.
-
-\leavevmode\hypertarget{ref-laird1982}{}%
-\textsc{Laird, N. M. and Ware, J. H.} (1982). Random-effects models for longitudinal data. \emph{Biometrics} \textbf{38}, 963.
-
-\leavevmode\hypertarget{ref-lang2015}{}%
-\textsc{Lang, T.
-A.
-and Altman, D.
-G.} (2015).
-Basic statistical reporting for articles published in biomedical journals: The {``}statistical analyses and methods in the published literature{''} or the {SAMPL} guidelines.
-\emph{International Journal of Nursing Studies} \textbf{52}, 5--9.
-
-\leavevmode\hypertarget{ref-liu2010}{}%
-\textsc{Liu, C., Cripe, T.
-P.
-and Kim, M.-O.} (2010).
-Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences.
-\emph{Molecular Therapy} \textbf{18}, 1724--1730.
-
-\leavevmode\hypertarget{ref-ma2012}{}%
-\textsc{Ma, Y., Mazumdar, M. and Memtsoudis, S. G.} (2012). Beyond repeated-measures analysis of variance. \emph{Regional Anesthesia and Pain Medicine} \textbf{37}, 99--105.
-
-\leavevmode\hypertarget{ref-marra2012}{}%
-\textsc{Marra, G. and Wood, S. N.} (2012). Coverage properties of confidence intervals for generalized additive model components. \emph{Scandinavian Journal of Statistics} \textbf{39}, 53--74.
-
-\leavevmode\hypertarget{ref-maxwell2017}{}%
-\textsc{Maxwell, S. E., Delaney, H. D. and Kelley, K.} (2017). Designing experiments and analyzing data. Routledge. doi:\href{https://doi.org/10.4324/9781315642956}{10.4324/9781315642956}.
-
-\leavevmode\hypertarget{ref-mcelreath2018}{}%
-\textsc{McElreath, R.} (2018). Statistical rethinking. Chapman; Hall/{CRC}. doi:\href{https://doi.org/10.1201/9781315372495}{10.1201/9781315372495}.
-
-\leavevmode\hypertarget{ref-miller2019}{}%
-\textsc{Miller, D. L.} (2019). Bayesian views of generalized additive modelling. \emph{arXiv preprint arXiv:1902.01330}.
-
-\leavevmode\hypertarget{ref-molenberghs2004}{}%
-\textsc{Molenberghs, G.} (2004). Analyzing incomplete longitudinal clinical trial data. \emph{Biostatistics} \textbf{5}, 445--464.
-
-\leavevmode\hypertarget{ref-nakagawa2004}{}%
-\textsc{Nakagawa, S.} (2004). A farewell to bonferroni: The problems of low statistical power and publication bias. \emph{Behavioral Ecology} \textbf{15}, 1044--1045.
-
-\leavevmode\hypertarget{ref-nelder1972}{}%
-\textsc{Nelder, J. A. and Wedderburn, R. W. M.} (1972). Generalized linear models. \emph{Journal of the Royal Statistical Society. Series A (General)} \textbf{135}, 370.
-
-\leavevmode\hypertarget{ref-pavlov2018}{}%
-\textsc{Pavlov, M.
-V., Kalganova, T.
-I.
-and Lyubimtseva, Y.
-S.} (2018).
-Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy.
-\emph{Journal of Biomedical Optics} \textbf{23}, 1.
-
-\leavevmode\hypertarget{ref-pedersen2019}{}%
-\textsc{Pedersen, E. J., Miller, D. L., Simpson, G. L. and Ross, N.} (2019). Hierarchical generalized additive models in ecology: An introduction with mgcv. \emph{{PeerJ}} \textbf{7}, e6876.
-
-\leavevmode\hypertarget{ref-pinheiro2006}{}%
-\textsc{Pinheiro, J. and Bates, D.} (2006). {Mixed-effects models in S and S-PLUS}. Springer Science; Business Media. doi:\url{https://doi.org/10.1007/b98882}.
-
-\leavevmode\hypertarget{ref-nlme}{}%
-\textsc{Pinheiro, J., Bates, D., DebRoy, S., Sarkar, D. and R Core Team}. (2020). {nlme}: Linear and nonlinear mixed effects models. \url{https://CRAN.R-project.org/package=nlme}.
-
-\leavevmode\hypertarget{ref-potthoff2006}{}%
-\textsc{Potthoff, R.
-F., Tudor, G.
-E., Pieper, K.
-S.
-and Hasselblad, V.} (2006).
-Can one assess whether missing data are missing at random in medical studies?
-\emph{Statistical Methods in Medical Research} \textbf{15}, 213--234.
-
-\leavevmode\hypertarget{ref-r}{}%
-\textsc{R Core Team}. (2020). R: A language and environment for statistical computing. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.
-
-\leavevmode\hypertarget{ref-ritter2001}{}%
-\textsc{Ritter, G., Cohen, L., Williams, C., Richards, E., Old, L.
-and Welt, S.} (2001).
-{Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}.
-\emph{{Cancer Research}} \textbf{{61}}, 6851--6859.
-
-\leavevmode\hypertarget{ref-roblyer2011}{}%
-\textsc{Roblyer, D., Ueda, S., Cerussi, A., Tanamai, W., Durkin, A., Mehta, R., Hsiang, D., Butler, J.
-A., McLaren, C., Chen, W.-P., et al.} (2011).
-Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment.
-\emph{Proceedings of the National Academy of Sciences} \textbf{108}, 14626--14631.
-
-\leavevmode\hypertarget{ref-rose2012}{}%
-\textsc{Rose, N.
-L., Yang, H., Turner, S.
-D.
-and Simpson, G.
-L.} (2012).
-An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, {UK}.
-\emph{Geochimica et Cosmochimica Acta} \textbf{82}, 113--135.
-
-\leavevmode\hypertarget{ref-roth2017}{}%
-\textsc{Roth, E.
-M., Goldberg, A.
-C., Catapano, A.
-L., Torri, A., Yancopoulos, G.
-D., Stahl, N., Brunet, A., Lecorps, G.
-and Colhoun, H.
-M.} (2017).
-Antidrug antibodies in patients treated with alirocumab.
-\emph{New England Journal of Medicine} \textbf{376}, 1589--1590.
-
-\leavevmode\hypertarget{ref-scheffer2002}{}%
-\textsc{Scheffer, J.} (2002). Dealing with missing data. \emph{Research Letters in the Information and Mathematical Sciences} \textbf{3}, 153--160.
-
-\leavevmode\hypertarget{ref-schober2018}{}%
-\textsc{Schober, P. and Vetter, T. R.} (2018). Repeated measures designs and analysis of longitudinal data. \emph{Anesthesia {\&} Analgesia} \textbf{127}, 569--575.
-
-\leavevmode\hypertarget{ref-gratia}{}%
-\textsc{Simpson, G. L.} (2020). Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv'. \url{https://CRAN.R-project.org/package=gratia}.
-
-\leavevmode\hypertarget{ref-simpson2018}{}%
-\textsc{Simpson, G.
-L.} (2018).
-Modelling palaeoecological time series using generalised additive models.
-\emph{Frontiers in Ecology and Evolution} \textbf{6}.
-doi:\href{https://doi.org/10.3389/fevo.2018.00149}{10.3389/fevo.2018.00149}.
-
-\leavevmode\hypertarget{ref-sio2016}{}%
-\textsc{Sio, T.
-T., Atherton, P.
-J., Birckhead, B.
-J., Schwartz, D.
-J., Sloan, J.
-A., Seisler, D.
-K., Martenson, J.
-A., Loprinzi, C.
-L., Griffin, P.
-C., Morton, R.
-F., et al.} (2016).
-Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}).
-\emph{Supportive Care in Cancer} \textbf{24}, 3847--3855.
-
-\leavevmode\hypertarget{ref-skala2010}{}%
-\textsc{Skala, M.
-C., Fontanella, A., Lan, L., Izatt, J.
-A.
-and Dewhirst, M.
-W.} (2010).
-Longitudinal optical imaging of tumor metabolism and hemodynamics.
-\emph{Journal of Biomedical Optics} \textbf{15}, 011112.
-
-\leavevmode\hypertarget{ref-tank2020}{}%
-\textsc{Tank, A., Peterson, H.
-M., Pera, V., Tabassum, S., Leproux, A., O'Sullivan, T., Jones, E., Cabral, H., Ko, N., Mehta, R.
-S., et al.} (2020).
-Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens.
-\emph{Breast Cancer Research} \textbf{22}.
-doi:\href{https://doi.org/10.1186/s13058-020-01262-1}{10.1186/s13058-020-01262-1}.
-
-\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
-\textsc{Ugrinowitsch, C., Fellingham, G.
-W.
-and Ricard, M.
-D.} (2004).
-Limitations of ordinary least squares models in analyzing repeated measures data.
-\emph{Medicine {\&} Science in Sports {\&} Exercise}, 2144--2148.
-
-\leavevmode\hypertarget{ref-vishwanath2009}{}%
-\textsc{Vishwanath, K., Yuan, H., Barry, W.
-T., Dewhirst, M.
-W.
-and Ramanujam, N.} (2009).
-Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors.
-\emph{Neoplasia} \textbf{11}, 889--900.
-
-\leavevmode\hypertarget{ref-wagenmakers2008}{}%
-\textsc{Wagenmakers, E.-J., Lee, M., Lodewyckx, T. and Iverson, G. J.} (2008). Bayesian versus frequentist inference. In Bayesian evaluation of informative hypotheses. Springer New York. pp 181--207.
-
-\leavevmode\hypertarget{ref-wang2019}{}%
-\textsc{Wang, B., Zhou, Z., Wang, H., Tu, X. M. and Feng, C.} (2019). The p-value and model specification in statistics. \emph{General Psychiatry} \textbf{32}, e100081.
-
-\leavevmode\hypertarget{ref-wegman1983}{}%
-\textsc{Wegman, E. J. and Wright, I. W.} (1983). Splines in statistics. \emph{Journal of the American Statistical Association} \textbf{78}, 351--365.
-
-\leavevmode\hypertarget{ref-weiss2005}{}%
-\textsc{Weiss, R. E.} (2005). Modeling longitudinal data. Springer New York. doi:\href{https://doi.org/10.1007/0-387-28314-5}{10.1007/0-387-28314-5}.
+\CSLLeftMargin{62. }
+\CSLRightInline{Begley CG, Ioannidis JPA.
+Reproducibility in science.
+\emph{Circulation Research}.
+2015;116(1):116-126.
+doi:\href{https://doi.org/10.1161/circresaha.114.303819}{10.1161/circresaha.114.303819}}
 
 \leavevmode\hypertarget{ref-weissgerber2018}{}%
-\textsc{Weissgerber, T.
-L., Garcia-Valencia, O., Garovic, V.
-D., Milic, N.
-M.
-and Winham, S.
-J.} (2018).
+\CSLLeftMargin{63. }
+\CSLRightInline{Weissgerber TL, Garcia-Valencia O, Garovic VD, Milic NM, Winham SJ.
 Why we need to report more than {{}}data were analyzed by t-tests or {ANOVA}{{}}.
-\emph{{eLife}} \textbf{7}.
-doi:\href{https://doi.org/10.7554/elife.36163}{10.7554/elife.36163}.
-
-\leavevmode\hypertarget{ref-west2014}{}%
-\textsc{West, B.
-T., Welch, K.
-B.
-and Galecki, A.
-T.} (2014).
-Linear mixed models: A practical guide using statistical software, second edition.
-Taylor \& Francis.
-\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.
-
-\leavevmode\hypertarget{ref-wolfinger1996}{}%
-\textsc{Wolfinger, R. D.} (1996). Heterogeneous variance: Covariance structures for repeated measures. \emph{Journal of Agricultural, Biological, and Environmental Statistics} \textbf{1}, 205.
-
-\leavevmode\hypertarget{ref-wood2017}{}%
-\textsc{Wood, S. N.} (2017). Generalized additive models. Chapman; Hall/{CRC}. doi:\href{https://doi.org/10.1201/9781315370279}{10.1201/9781315370279}.
-
-\leavevmode\hypertarget{ref-wood2016}{}%
-\textsc{Wood, S. N., Pya, N. and Säfken, B.} (2016). Smoothing parameter and model selection for general smooth models. \emph{Journal of the American Statistical Association} \textbf{111}, 1548--1563.
-
-\leavevmode\hypertarget{ref-yang2012}{}%
-\textsc{Yang, L., Qin, G., Zhao, N., Wang, C.
-and Song, G.} (2012).
-Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality.
-\emph{{BMC} Medical Research Methodology} \textbf{12}.
-doi:\href{https://doi.org/10.1186/1471-2288-12-165}{10.1186/1471-2288-12-165}.
+\emph{{eLife}}.
+2018;7.
+doi:\href{https://doi.org/10.7554/elife.36163}{10.7554/elife.36163}}
 

### 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/00-Full_document_SIM_No_Appendix.tex
Created folder for Wiley submission

--- 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/00-Full_document_No_Appendix.tex
@@
 \hypersetup{
-  pdftitle={Using generalized additive models to analyze biomedical non-linear longitudinal data},
+  pdftitle={Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models},
   hidelinks,
@@
 \setcounter{secnumdepth}{5}
-\usepackage{helvet}
+\usepackage{mathptmx}
 \usepackage{lineno}
@@
 \usepackage{setspace}
-\usepackage[nomarkers,figuresonly]{endfloat}
+%\usepackage[nomarkers,figuresonly]{endfloat}
 %set a box to put the ORCID logo
@@
 
-\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
-\usepackage{etoolbox}
-\makeatletter
-\providecommand{\subtitle}[1]{% add subtitle to \maketitle
-  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
-}
-\makeatother
-\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
+\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}}
 \date{\vspace{-2.5em}}
@@
 
-In contrast, \emph{generalized additive models} (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
 Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented.
-We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing), allowing reliable inference from the data.
+We present the basic theory of GAMs, and using reported trends of oxygen saturation in tumors we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R.
+We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing).
 To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/l-q-response-1}
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_SIM_No_Appendix_files/figure-latex/l-q-response-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/basis-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_SIM_No_Appendix_files/figure-latex/basis-plot-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/sim-smooth-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_SIM_No_Appendix_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/plot-pairwise-comp-1}
+{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_SIM_No_Appendix_files/figure-latex/plot-pairwise-comp-1}
 

### 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/00-OUP_Appendix.Rmd
Created folder for Wiley submission

--- 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/00-Full_document_OUP.Rmd
@@
 ---
-title: '**Using generalized additive models to analyze biomedical non-linear longitudinal data**'
-subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
+title: |
+    | **Using generalized additive models to analyze biomedical non-linear longitudinal data**
+    | _Beyond repeated measures ANOVA and Linear Mixed Models_
+    | 
+    | SUPPLEMENTARY MATERIALS: APPENDIX
 header-includes:
@@
     includes:
-      in_header: preamble_OUP.sty #additional LaTeX formatting
+      in_header: preamble_OUP_Appendix.sty #additional LaTeX formatting
   bookdown::word_document2:
@@
 
+\newpage
 # Abstract
@@
 
-<!-- ```{r child='07-Appendix.Rmd'} -->
+```{r child='07-Appendix.Rmd'}
 
-<!-- ``` -->
+```

### 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/Full_document_SIM_rticles_No_Appendix.Rmd
Created folder for Wiley submission

--- 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/00-Full_document_OUP_No_Appendix.Rmd
@@
 ---
-title: '**Using generalized additive models to analyze biomedical non-linear longitudinal data**'
-subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
-header-includes:
-    \usepackage{placeins}
-output:  
-  bookdown::pdf_document2:
-    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
-    pandoc_args: --listings #calls the listings package to fit code within the page margins
-    keep_tex: yes #keep LaTeX file for submission
-    fig_caption: yes #allows captions in figures
-    toc: false #do not include table of contents
-    extra_dependencies:
-      subfig: null #allows for subfigures
-      breqn: null #line breaks for long equations
-      caption: ["font={small}"] #size of the figure captions
-      #float: null #allows for control of placement of figures
-    includes:
-      in_header: preamble_OUP.sty #additional LaTeX formatting
-  bookdown::word_document2:
-    fig_caption: yes #figure caption
-    keep_md: yes
-  bookdown::html_document2:
-    css: "style.css" #style for the HTML document
-csl: biostatistics.csl #style for references
-bibliography: refs.bib #references
-link-citations: yes #adds links to the citations
-'': default
----
-
-# Summary
-
-In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs).
+title: "Using generalized additive models to analyze biomedical non-linear longitudinal data: _Beyond Linear Mixed Models and Repeated Measures ANOVA_"
+author:
+- name: Ariel I. Mundo
+  num: a
+- name: Timothy J. Muldoon*
+  num: a
+- name: John R. Tipton
+  num: b
+address:
+- num: a
+  org: Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, US
+- num: b
+  org: Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, US
+corres: "*Timothy J. Muldoon, Engineering Research Center Suite 2540, Fayetteville, AR, 72701 \\email{tmuldoon@uark.edu}"
+presentaddress: This is sample for present address text this is sample for present address text
+authormark: Mundo \emph{et al}.
+articletype: Tutorial in Biostatistics
+received: 
+revised: 
+accepted: 
+abstract: "In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs).
 Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
-
 In contrast, _generalized additive models_ (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
@@
 We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing), allowing reliable inference from the data.
-To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research.
-
-# Keywords {-}
-
-cancer biology; tumor response; generalized additive models; simulation; R
+To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research."
+keywords: Longitudinal data; Biomedical data; Simulation; R;
+bibliography: refs.bib
+output:
+  bookdown::pdf_book:
+    pandoc_args: [ "--listings" ]
+    base_format: rticles::sim_article
+    keep_tex: yes
+    fig_caption: yes
+    includes:
+      in_header: preamble.sty #LaTeX formatting options
+---
 

### 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/Full_document_SIM_rticles_No_Appendix.tex
Created folder for Wiley submission

--- 2021-06-21 Manuscripts/Manuscript by chapters -for submission/00-Full_document.tex
@@
-\documentclass[,]{sagej}
-
-\usepackage{moreverb,url,natbib, multirow, tabularx}
-\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
+\documentclass[AMA,STIX1COL,]{WileyNJD-v2}
 
@@
 \usepackage{lineno}
-\usepackage{authblk}
+\usepackage[affil-it,blocks]{authblk}
 \usepackage{hyperref}
 \usepackage{graphicx}
-
+\usepackage{setspace}
+\usepackage[nomarkers,figuresonly]{endfloat}
 %set a box to put the ORCID logo
@@
 
+%to have page numbers with letters in the Appendix:
+% %solution from https://tex.stackexchange.com/questions/59572/custom-page-numbering-for-appendix
+\newcommand{\appendixpagenumbering}{
+  \break
+  \pagenumbering{arabic}
+  \renewcommand{\thepage}{\thesection-\arabic{page}}
+}
+
 %command for the package lineno
@@
 %}
-\usepackage{placeins}
-\usepackage{subfig}
-\usepackage{breqn}
-\usepackage[font={small}]{caption}
-\usepackage{float}
+
+%line spacing
+\doublespacing
+
+\articletype{Tutorial in Biostatistics}
+
+\raggedbottom
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 
-\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
-
-\runninghead{}
-
-\author{}
-
-\affiliation{}
-
-\maketitle
-
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
-
-In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs).
+\title{Using generalized additive models to analyze biomedical non-linear longitudinal data: \emph{Beyond Linear Mixed Models and Repeated Measures ANOVA}}
+
+\author[a]{Ariel I. Mundo}
+\author[a]{Timothy J. Muldoon*}
+\author[b]{John R. Tipton}
+
+\authormark{Mundo \emph{et al}.}
+
+\address[a]{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, US}
+\address[b]{Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, US}
+
+\corres{*Timothy J. Muldoon, Engineering Research Center Suite 2540, Fayetteville, AR, 72701 \email{tmuldoon@uark.edu}}
+
+\presentaddress{This is sample for present address text this is sample for present address text}
+
+\abstract{In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs).
 Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
-
-In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+In contrast, \emph{generalized additive models} (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
 Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
@@
 We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing), allowing reliable inference from the data.
-To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
+To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.}
+
+\keywords{Longitudinal data; Biomedical data; Simulation; R;}
+
+\maketitle
 
@@
 
-\FloatBarrier
-
 \hypertarget{challenges-presented-by-longitudinal-studies}{%
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/l-q-response-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SIM_rticles_No_Appendix_files/figure-latex/l-q-response-1}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SIM_rticles_No_Appendix_files/figure-latex/basis-plot-1}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SIM_rticles_No_Appendix_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-1}
+\begin{figure}
+
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SIM_rticles_No_Appendix_files/figure-latex/plot-pairwise-comp-1}
 
@@
 
-\FloatBarrier
+\hypertarget{supplementary-materials}{%
+\section*{Supplementary Materials}\label{supplementary-materials}}
+\addcontentsline{toc}{section}{Supplementary Materials}
+
+An Appendix which contains all the code used to create this manuscript, along with a basic workflow to implement GAMs in R is available as Supplementary Material in PDF.
+A GitHub repository containing all the code used for this paper along with detailed instructions for its use is available at \url{https://github.com/aimundo/GAMs-biomedical-research}.
 
@@
 
-\hypertarget{appendix}{%
-\section{(APPENDIX)}\label{appendix}}
-
-\hypertarget{code-for-manuscript-data}{%
-\section{Code for Manuscript data}\label{code-for-manuscript-data}}
-
-This section presents the code used to generate figures, models and simulated data from Sections 3 and 4 from the main manuscript.
-
-\hypertarget{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}{%
-\subsection{Compound symmetry and independent errors in linear and quadratic responses}\label{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}}
-
-This section simulated linear and quadratic data in the same manner as in Section \ref{simulation}.
-The linear simulations using Figure \ref{fig:linear-cases-Appendix} show in panels A and D the simulated mean responses and individual data points.
-Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-The following code produces a more comprehensive exploration of Figure \ref{fig:l-q-response} in the main manuscript.
-
-\begin{lstlisting}[language=R]
-##########Section for calculations###########
-
-## Example with linear response
-
-#This function simulates data using a linear or quadratic mean response and each with correlated
-#or uncorrelated errors. Each group has a different slope/concavity.
-example <- function(n_time = 6, #number of time points
-                    fun_type = "linear", #type of response
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  x <- seq(1,6, length.out = n_time)
-
-  #Create mean response matrix: linear or quadratic
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # quadratic response (non-linear)
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-
-  #create an array where individual observations per each time point for each group are to be stored. Currently using 10 observations per timepoint
-  y <- array(0, dim = c(length(x), 2, 10))
-
-  #Create array to store the "errors" for each group at each timepoint. The "errors" are the 
-  #between-group variability in the response.
-  errors <- array(0, dim = c(length(x), 2, 10))
-  #create an array where 10 observations per each time point for each group are to be stored
-
-  #The following cycles create independent or correlated responses.
-To each value of mu (mean response per group) a randomly generated error (correlated or uncorrelated) is added and thus the individual response is created.
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors: variance covariance matrix
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-#labeling y and errors  
-  dimnames(y) <- list(time = x, 
-                      treatment = 1:2, 
-                      subject = 1:10)
-
-  dimnames(errors) <- list(time = x, 
-                           treatment = 1:2, 
-                           subject = 1:10)
-
-  #labeling the mean response
-  dimnames(mu) <- list(time = x, 
-                       treatment = 1:2)
-
-  #convert y, mu and errors to  dataframes with time, treatment and subject columns
-  dat <- as.data.frame.table(y, 
-                             responseName = "y")
-  dat_errors <- as.data.frame.table(errors, 
-                                    responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, 
-                                responseName = "mu")
-
-  #join the dataframes to show mean response and errors per subject
-  dat <- left_join(dat, dat_errors, 
-                   by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, 
-                   by = c("time", "treatment"))
-  #add time
-  dat$time <- as.numeric(as.character(dat$time))
-  #label subjects per group
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, 
-                                  treatment, 
-                                  sep = "-")))
-
-  ## repeated measures ANOVA
-
-  fit_anova <- lm(y ~ time + treatment + time * treatment, data = dat)
-
-#LMEM: time and treatment interaction model, compound symmetry 
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  #create a prediction frame where the model can be used for plotting purposes
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  #add model predictions to the dataframe that has the simulated data
-  dat$pred_anova <- predict(fit_anova)
-  dat$pred_lmem <- predict(fit_lme)
-
-  #return everything in a list
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_anova=fit_anova,
-    fit_lme = fit_lme
-  ))
-}
-##################Section for plotting#################################
-#######################################################################
-#This function will create the plots for either a "linear" or "quadratic" response
-
-plot_example <- function(sim_dat) {
-  ## Plot the simulated data (scatterplot)
-
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = treatment, 
-               color = treatment)
-           ) +
-    geom_point(show.legend=FALSE) +
-    labs(y='response')+
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment),
-              show.legend=FALSE) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-  #plot the simulated data with trajectories per each subject
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = subject, 
-               color = treatment)
-           ) +
-    geom_line(aes(size = "Subjects"),
-              show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment,
-                  size = "Simulated Truth"), 
-              lty = 1,show.legend = FALSE) +
-    labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                face = "bold"),
-     text=element_text(size=30))+
-    thm
-
-  #plot the errors
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = errors,
-               group = subject, 
-               color = treatment)) +
-    geom_line(show.legend=FALSE) +
-     labs(y='errors')+
-     theme_classic()+
-     theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-   #plot the model predictions for rm-ANOVA
-  p4 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point(show.legend=FALSE)+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_anova), 
-                  group = subject, size = "Subjects"),show.legend = FALSE) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_anova, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population"),
-              show.legend=FALSE) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-   #plot the LMEM predictions
-  p5 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point()+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), 
-                  group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_lme, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population")) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 30, 
-                                  face = "bold"),
-        text=element_text(size=30))+
-    thm
-
-  return((p1+p3+p2+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-txt<-18
-
-#Store each plot in a separate object
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent")) 
-\end{lstlisting}
-
-\begin{figure}[H]
-\includegraphics{00-Full_document_files/figure-latex/linear-cases-Appendix-1} \caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are able to capture the trend of the data.}\label{fig:linear-cases-Appendix}
-\end{figure}
-
-For the quadratic response case, Figure \ref{fig:quadratic-cases-Appendix} shows the simulated responses using compound symmetry and independent errors.
-
-\begin{figure}[H]
-\includegraphics{00-Full_document_files/figure-latex/quadratic-cases-Appendix-1} \caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are not able to capture the changes in each group over time.}\label{fig:quadratic-cases-Appendix}
-\end{figure}
-
-\hypertarget{basis-functions-and-gams}{%
-\subsection{Basis functions and GAMs}\label{basis-functions-and-gams}}
-
-This code produces Figure \ref{fig:basis-plot} from the main manuscript.
-Briefly, a non-linear (quadratic) response is simulated a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
-The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
-
-\begin{lstlisting}[language=R]
-#generate the response: the same initial procedure from the previous section to simulate
-#the response
-set.seed(1)
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-
- #label each table
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
-
- #Convert to dataframes with subject, time and group columns
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
-
- #label subject per group
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
- #extract  "Group 1" to fit the GAM
-  dat<-subset(dat,treatment==1)
- #keep just the response and timepoint columns
-   dat<-dat[,c('y','time')]
-
-   #GAM model of time, 5 knots
-gm<-gam(y~s(time,k=5),data=dat)
-
-#model_matrix (also known as) 'design matrix'
-#will contain the smooths used to create  model 'gm'
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be 
-#multiplied by the coefficients
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final smooth by aggregating the basis functions
-
-smooth<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(smooth=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,
-            aes(x=time,
-                y=value_orig,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(size=sz,
-            show.legend=FALSE)+
-  geom_point(size=sz+1,
-             show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,
-             labeller = as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#penalized basis
-p12<-ggplot(basis_plot,
-            aes(x=time,
-                y=mod_val,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(show.legend = FALSE,
-            size=sz)+
-  geom_point(show.legend = FALSE,
-             size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,
-             labeller=as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#heatmap of the coefficients
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,
-            aes(x=Basis,
-                y=Basis))+
-  geom_tile(aes(fill = cof), 
-            colour = "black") +
-    scale_fill_gradient(low = "white",
-                        high = "#B50A2AFF")+ #color picked from KikiMedium
-  labs(x='Basis',
-       y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),
-            size=7,
-            show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,
-            aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  labs(y='Simulated \n response')+
-  geom_line(data=smooth,
-            aes(x=time,
-                y=smooth),
-            color="#6C581DFF",
-            size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/basis-plot-appendix-1}
-
-}
-
-\caption{Basis functions for a single smoother for time with five knots.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-B: Matrix of basis function weights.
-Each basis function is multiplied by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-C: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
-Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-D: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot-appendix}
-\end{figure}
-
-\hypertarget{tumor-data-simulation}{%
-\section{Longitudinal biomedical data simulation and GAMs}\label{tumor-data-simulation}}
-
-This section describes how to fit GAMs to longitudinal data using simulated data.
-First, data is simulated according to Section \ref{longitudinal-GAMs}, where reported data of oxygen saturation (\(\mbox{StO}_2\)) in tumors under either chemotherapy or saline control is used as a starting point to generate individual responses in each group.
-
-\begin{lstlisting}[language=R]
-set.seed(1)
-#Dataframe that contains the original reported trends
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
-
-## plot the mean response
-f1<-ggplot(dat, 
-           aes(x = Day, 
-               y = StO2, 
-               color = Group)) +
-    geom_line(size=1,
-              show.legend = FALSE)+
-    geom_point(show.legend = FALSE,
-               size=1.5,
-               alpha=0.5)+
-  labs(y=expression(paste(StO[2],
-                          ' (real)')))+
-  theme_classic()+
-  thm+
-    scale_x_continuous(breaks=c(0,5,10))+
-    scale_y_continuous(breaks=c(0,40))+
-  plot_layout(tag_level = 'new')+
-  theme(
-    plot.background = element_rect(fill = "transparent", 
-                                   color = NA),
-    axis.text=element_text(size=14)
-  )
-
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-n <- 10 #number of observations
-sd <- 10 #approximate sd from paper
-df <- 6
-dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, 
-           aes(x = Day, 
-               y = StO2_sim, 
-               color = Group)) +
-    geom_point(show.legend=FALSE,
-               size=1.5,
-               alpha=0.5)+
-    stat_summary(aes(y = StO2_sim,
-                     group=Group), 
-                 fun=mean, geom="line",
-                 size=1,
-                 show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],
-                         '(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-  thm+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
-\end{lstlisting}
-
-\hypertarget{workflow}{%
-\subsection{A basic Workflow for GAMs}\label{workflow}}
-
-This section shows a basic workflow to fit a series of increasingly complex GAMs to the simulated data from the previous section.
-Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
-
-\hypertarget{first-model}{%
-\subsubsection{First model}\label{first-model}}
-
-The first model fitted to the data is one that only accounts for different smooths by day.
-The model syntax specifies that \passthrough{\lstinline!gam\_00!} is the object that will contain all the model information, and that the model attempts to explain changes in \passthrough{\lstinline!StO2\_sim!} (simulated \(\mbox{StO}_2\)) using a smooth per \passthrough{\lstinline!Day!}.
-The model will use 5 knots (\passthrough{\lstinline!k=5!}) for the smooth.
-The smooth is constructed by default using thin plate regression splines.
-The smoothing parameter estimation method used is the restricted maximum likelihood (\passthrough{\lstinline!REML!}).
-
-\begin{lstlisting}[language=R]
-gam_00<-gam(StO2_sim ~ s(Day, k = 5),
-            method='REML',
-            data  = dat_sim)
-\end{lstlisting}
-
-To obtain model diagnostics, two methodologies are to be used: 1) graphical diagnostics, and 2) a model check.
-In the first case, the functions \passthrough{\lstinline!appraise!} and \passthrough{\lstinline!draw!} from the package \emph{gratia} can be used to obtain a single output with all the graphical diagnostics.
-For model check, the functions \passthrough{\lstinline!gam.check!} and \passthrough{\lstinline!summary!} from \emph{mgcv} provide detailed information about the model fit and its parameters.
-
-\hypertarget{graphical-diagnostics}{%
-\paragraph{Graphical diagnostics}\label{graphical-diagnostics}}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/first-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the first GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:first-GAM-diag}
-\end{figure}
-
-From the output of the function \passthrough{\lstinline!appraise!} in Figure \ref{fig:first-GAM-diag}, the major indicators of concern about the model are the QQ plot of residuals and the histogram of residuals.
-The QQ plot shows that the errors are not reasonably located along the 45\(^{\circ}\) line (which indicates normality), as there are multiple points that deviate from the trend, specially in the tails.
-The histogram also shows that the variation (residuals) is not following the assumption of a normal distribution.
-
-The \passthrough{\lstinline!draw!} function permits to plot the smooths as \passthrough{\lstinline!ggplot2!} objects, which eases subsequent manipulation, if desired.
-Because model \passthrough{\lstinline!gam\_00!} specifies only one smooth for the time covariate (Day), the plot only contains only one smooth.
-Note that the smooth shows an almost linear profile.
-
-\hypertarget{model-check}{%
-\paragraph{Model check}\label{model-check}}
-
-\begin{lstlisting}[language=R]
-#need to add figure number and caption
-gam.check(gam_00)
-\end{lstlisting}
-
-\includegraphics{00-Full_document_files/figure-latex/first-GAM-check-1}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 5 iterations.
-## Gradient range [-0.0003727881,-6.621452e-07]
-## (score 444.0118 & scale 450.6638).
-## Hessian positive definite, eigenvalue range [0.3881695,49.00676].
-## Model rank =  5 / 5 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##          k'  edf k-index p-value    
-## s(Day) 4.00 2.11    0.36  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-summary(gam_00)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ s(Day, k = 5)
-## 
-## Parametric coefficients:
-##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   22.967      2.123   10.82   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##          edf Ref.df     F  p-value    
-## s(Day) 2.114  2.565 7.633 0.000517 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.153   Deviance explained = 17.2%
-## -REML = 444.01  Scale est. = 450.66    n = 100
-\end{lstlisting}
-
-Special attention must be paid to the `k-index' from \passthrough{\lstinline!gam.check!}.
-This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
-If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless1).
-From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data.
-The \passthrough{\lstinline!edf!} (effective degrees of freedom) is an indicator of the complexity of the smooth.
-Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
-
-From the \passthrough{\lstinline!summary!} function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
-The link function is `identity' as the model does not make any transformation on the predictors.
-The `significance of smooth terms' \emph{p-value} indicates if each smooth is adding significance to the model.
-Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \(\approx\) 17\%.
-
-\hypertarget{second-model}{%
-\subsubsection{Second model}\label{second-model}}
-
-The major flaw of \passthrough{\lstinline!gam\_00!} is that this model is not taking into account the fact that the data is nested in groups.
-The next iteration is a model where a different smooth of time (Day) is assigned for each group using \passthrough{\lstinline!by=Group!} in the model syntax.
-
-\begin{lstlisting}[language=R]
-gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5),
-            method='REML',
-            data  = dat_sim)
-
-gam.check(gam_01)
-\end{lstlisting}
-
-\includegraphics{00-Full_document_files/figure-latex/second-GAM-1}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 7 iterations.
-## Gradient range [-5.51754e-05,2.671715e-06]
-## (score 423.3916 & scale 280.8777).
-## Hessian positive definite, eigenvalue range [0.3162258,48.5557].
-## Model rank =  9 / 9 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                         k'  edf k-index p-value    
-## s(Day):GroupControl   4.00 3.39    0.43  <2e-16 ***
-## s(Day):GroupTreatment 4.00 3.23    0.43  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-summary(gam_01)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ s(Day, by = Group, k = 5)
-## 
-## Parametric coefficients:
-##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   22.967      1.676    13.7   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                         edf Ref.df      F p-value    
-## s(Day):GroupControl   3.392  3.794  3.817  0.0304 *  
-## s(Day):GroupTreatment 3.229  3.682 21.174  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.472   Deviance explained = 50.8%
-## -REML = 423.39  Scale est. = 280.88    n = 100
-\end{lstlisting}
-
-Diagnostics for this model indicate that the k-index is still below 1 (0.43 from \passthrough{\lstinline!gam.check!}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
-Moreover, the smooths (plotted via the \passthrough{\lstinline!draw()!} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \(\approx\) 51\%.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/second-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the second GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:second-GAM-diag}
-\end{figure}
-
-\hypertarget{third-model}{%
-\subsubsection{Third model}\label{third-model}}
-
-Model \passthrough{\lstinline!gam\_00!} was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted, the model diagnostics, and, the low variance explained by the model.
-On the other hand, \passthrough{\lstinline!gam\_01!} takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \ref{longitudinal-GAMs}, in order to differentiate between each group a parametric term needs to be added to the model for the interaction of \emph{Day} and \emph{Group}.
-
-\begin{lstlisting}[language=R]
-#GAM for StO2
-
-m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
-            method='REML',
-            data  = dat_sim)
-
-gam.check(m1)
-\end{lstlisting}
-
-\includegraphics{00-Full_document_files/figure-latex/final-model-Appendix-1}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 10 iterations.
-## Gradient range [-8.164307e-08,1.500338e-08]
-## (score 355.8554 & scale 64.53344).
-## Hessian positive definite, eigenvalue range [1.174841,48.08834].
-## Model rank =  10 / 10 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                         k'  edf k-index p-value
-## s(Day):GroupControl   4.00 3.87    1.02    0.59
-## s(Day):GroupTreatment 4.00 3.88    1.02    0.54
-\end{lstlisting}
-
-\begin{lstlisting}[language=R]
-summary(m1)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ Group + s(Day, by = Group, k = 5)
-## 
-## Parametric coefficients:
-##                Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)       9.084      1.136   7.996 4.09e-12 ***
-## GroupTreatment   27.766      1.607  17.282  < 2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                         edf Ref.df     F p-value    
-## s(Day):GroupControl   3.873  3.990 17.57  <2e-16 ***
-## s(Day):GroupTreatment 3.879  3.991 89.33  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.879   Deviance explained = 88.9%
-## -REML = 355.86  Scale est. = 64.533    n = 100
-\end{lstlisting}
-
-The resulting model is \passthrough{\lstinline!m1!}, which is the model fitted in the main manuscript.
-By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\(\approx\) 1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 89\% of the variance in the data.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/final-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the final GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooths for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:final-GAM-diag}
-\end{figure}
-
-\hypertarget{comparing-models-via-aic}{%
-\subsubsection{Comparing models via AIC}\label{comparing-models-via-aic}}
-
-One final comparison that can be made for model selection involves the use of the Aikake Information Criterion (AIC).
-This metric is used to estimate information loss, which we want to minimize with an appropriate model.
-Therefore, when 2 or more models are compared, the model with lower AIC is preferred.
-In R, the comparison is done using the \passthrough{\lstinline!AIC!} function.
-
-\begin{lstlisting}[language=R]
-AIC(gam_00,gam_01,m1)
-\end{lstlisting}
-
-\begin{lstlisting}
-##               df      AIC
-## gam_00  4.564893 900.8257
-## gam_01  9.476137 858.6051
-## m1     10.980983 712.2067
-\end{lstlisting}
-
-The output in this case is expected: model \passthrough{\lstinline!m1!} has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
-The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
-
-\hypertarget{pairwise-comparisons-of-smooth-confidence-intervals}{%
-\paragraph{Pairwise comparisons of smooth confidence intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
-
-The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \ref{GAM-significance}.
-In this case, the ``design matrix'' is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
-Briefly, the ``design matrix'' (also known as the ``Xp matrix'') from the selected model (\passthrough{\lstinline!m1!}) is used to calculate a 95\% confidence interval of the difference between the smooth terms for each group.
-This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
-\textbf{All pairwise comparisons in this paper have been centered at the response scale to ease interpretation }.
-
-\begin{lstlisting}[language=R]
-##Pairwise comparisons
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-##matrix that contains the basis functions evaluated at the points in pdat
-    xp <- predict(m1, newdata = pdat, type = 'lpmatrix')
-
-#Find columns in xp where the name contains "Control"
-    c1 <- grepl('Control', colnames(xp))
-
-#Find columns in xp where the name contains 'Treatment'
-    c2 <- grepl('Treatment', colnames(xp))
-
-#Find rows in pdat that correspond to either 'Control' or 'Treatment'
-    r1 <- with(pdat, Group == 'Control')
-    r2 <- with(pdat, Group == 'Treatment')
-
-# In xp: find the rows that correspond to Control or Treatment, those that do not match will be
-    #set to zero. Then, substract the values from the rows corresponding to 'Control' from those that correspond
-    #to 'Treatment'
-    X <- xp[r1, ] - xp[r2, ]
-
-    ## remove columns that do not contain name 'Control' or 'Treatment'
-    X[, ! (c1 | c2)] <- 0
-    ## zero out the parametric cols, those that do not contain in the characters 's('
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
-
-    #Multiply matrix by model coefficients. X has (p,n) (rows, columns) and the coefficient matrix has
-    #dimensions (n,1). The resulting matrix has dimensions (p,1)
-    dif <- X %*% coef(m1)
-
-    #comp<-test %*% coef(gam1)[3:10]
-
-#Calculate standard error for the computed differences using the variance-covariance matrix
-    #of the model
-    se <- sqrt(rowSums((X %*% vcov(m1, unconditional = FALSE)) * X))
-    crit <- qt(0.05/2, df.residual(m1), lower.tail = FALSE)
-    #upper  limits
-    upr <- dif + (crit * se)
-    #lower limits
-    lwr <- dif - (crit * se)
-    #put all components in a dataframe for plotting
-    comp1<-data.frame(pair = paste('Control', 'Treatment', sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-
-#add time point sequence
-comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1))
-
-#use function from the pairwise comparison plot in the manuscript to get the shaded regions
-
-    my_list<-pairwise_limits(comp_StO2)
-  rib_col<-'#EDD03AFF' #color for the ribbon
-#plot the difference
-c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
-  #shaded region
-  annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill='#30123BFF',
-                alpha = 0.5,
-                ) +
-  annotate("text",
-             x=1.5,
-             y=-10,
-             label="Control",size=10
-           )+
-  #shaded region  
-  annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
-             alpha = 0.5
-           ) +
-  annotate("text",
-             x=6,
-             y=-10,
-             label="Treatment",
-             size=10
-           )+
-  #ribbon for difference confidence interval  
-  geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(color='black',size=1) +
-    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/pairwise-comp-workflow-fig-1}
-
-}
-
-\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!m1!} using a 95\% confidence interval for the difference between smooths.
-The comparison is centered at the response scale.
-Shaded regions indicate time intervals where each treatment group has a non-zero effect.}\label{fig:pairwise-comp-workflow-fig}
-\end{figure}
-
-Of notice, a convenient wrapper for the function described above exists in the package \passthrough{\lstinline!gratia!}.
-In this package, \passthrough{\lstinline!difference\_smooths!} is a function that makes the comparisons and produces Figure \ref{fig:pairwise-comp-workflow-fig} when is used on a fitted model.
-The function syntax and an example can be found at:
-
-\url{https://cran.r-project.org/web/packages/gratia/gratia.pdf}
-
-Keep in mind that this function \textbf{does not} center the pairwise comparison at the response scale, so it has to be shifted in order to be compared to the raw data.
-
-\hypertarget{gam-and-linear-model-plots-and-missing-data}{%
-\section{GAM and Linear model plots and Missing data}\label{gam-and-linear-model-plots-and-missing-data}}
-
-This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\passthrough{\lstinline!m1!}), linear model and GAM on data with missing observations are presented.
-Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inset are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
-
-\hypertarget{gam-and-linear-model-plots}{%
-\subsection{GAM and Linear model plots}\label{gam-and-linear-model-plots}}
-
-This code chunk creates panels B and D in Figure \ref{fig:sim-smooth-plot}.
-Note that this code uses the final GAM from the previous section (\passthrough{\lstinline!m1!}), so the simulated data and the model should be generated before running this section.
-
-\begin{lstlisting}[language=R]
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-#creates a dataframe using the length of the covariates for the GAM
-gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
-                         subject=factor(rep(1:10)))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#adds the predictions to the grid and creates a confidence interval for GAM
-gam_predict<-gam_predict%>%
-    mutate(fit = predict(m1,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(m1, gam_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=gam_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'complete')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-
-#posthoc comparisons for the linear model
-#library(multcomp)
-
-#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
-\end{lstlisting}
-
-\hypertarget{working-with-missing-data-in-gams}{%
-\subsection{Working with Missing data in GAMs}\label{working-with-missing-data-in-gams}}
-
-This code chunk first randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data.
-Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
-
-\begin{lstlisting}[language=R]
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
-
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'missing')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-      thm+
-  thm1
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/sim-smooth-plot-Appendix-1}
-
-}
-
-\caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
-\end{figure}
-
-\hypertarget{pairwise-comparisons-in-gams-full-and-missing-data-cases}{%
-\subsection{Pairwise comparisons in GAMs: full and missing data cases}\label{pairwise-comparisons-in-gams-full-and-missing-data-cases}}
-
-The next code chunk reproduces Figure \ref{fig:plot-pairwise-comp}. Here pairwise comparisons are made for the full and missing datasets.
-
-\begin{lstlisting}[language=R]
-##Pairwise comparisons
-
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-#this function takes the model, grid and groups to be compared using the lpmatrix
-#originally developed by G. Simpson:
-#https://fromthebottomoftheheap.net/2017/10/10/difference-splines-i/
-
-smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    #Find columns in xp where the name contains "Control" and "Treatment"
-    col1 <- grepl(g1, colnames(xp))
-    col2 <- grepl(g2, colnames(xp))
-    #Find rows in xp that correspond to each treatment
-    row1 <- with(newdata, Group == g1)
-    row2 <- with(newdata, Group == g2)
-    ## difference rows of xp for data from comparison
-    X <- xp[row1, ] - xp[row2, ]
-    ## zero out cols of X related to splines for other lochs
-    X[, ! (col1 | col2)] <- 0
-
-    ## zero out the parametric cols
-    #This line has been commented to keep the comparison at the response level,
-    #otherwise it gives the marginal change between smooths
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    #get standard error, critical value and boundaries
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(g1, g2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-}
-
-#use the function to calculate the difference in smooths
-comp1<-smooth_diff(m1,pdat,'Control','Treatment')
-
-#Create a dataframe with time, comparisons and labels for regions where difference exists
-comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
-
-pairwise_limits<-function(dataframe){
-    #extract values where the lower limit of the ribbon is greater than zero
-    #this is the region where the control group effect is greater
-    v1<-dataframe%>%
-        filter(lower>0)%>%
-        select(Day)
-    #get day  initial value
-    init1=v1$Day[[1]]
-    #get day final value
-    final1=v1$Day[[nrow(v1)]]
-
-    #extract values where the value of the upper limit of the ribbon is lower than zero
-    #this corresponds to the region where the treatment group effect is greater
-    v2<-comp_StO2_full%>%
-        filter(upper<0)%>%
-        select(Day)
-
-    init2=v2$Day[[1]]
-    final2=v2$Day[[nrow(v2)]]
-    #store values
-    my_list<-list(init1=init1,
-                  final1=final1,
-                  init2=init2,
-                  final2=final2)
-return(my_list)
-}
-
-my_list<-pairwise_limits(comp_StO2_full)
-rib_col<-'#EDD03AFF'
-
-c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
-    annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill='#30123BFF',
-                alpha = 0.5,
-                ) +
-  annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=8,
-           angle=90
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=8,
-           angle=90
-           )+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-###for missing data
-comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
-comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp2))
-
-missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    labs(x = 'Days', 
-         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
-                              )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme_classic()+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
-my_list<-pairwise_limits(comp_StO2_missing)
-
-c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    annotate("rect",
-             xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-             fill='#30123BFF',
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=8
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=8)+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-pair_comp<-c1+c2
-\end{lstlisting}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=0.75\linewidth,]{00-Full_document_files/figure-latex/plot-pairwise-comp-Appendix-1}
-
-}
-
-\caption{Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-B: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
-In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp-Appendix}
-\end{figure}
-
-\bibliographystyle{}
 \bibliography{refs.bib}

### 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/SIM_Appendix.Rmd
Created folder for Wiley submission

--- 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/00-OUP_Appendix.Rmd
@@
 title: |
-    | **Using generalized additive models to analyze biomedical non-linear longitudinal data**
-    | _Beyond repeated measures ANOVA and Linear Mixed Models_
+    | **Generalized additive models to analyze biomedical non-linear longitudinal data in R:**
+    | Beyond repeated measures ANOVA and Linear Mixed Models
     | 
@@
     includes:
-      in_header: preamble_OUP_Appendix.sty #additional LaTeX formatting
+      in_header: preamble_Appendix.sty #additional LaTeX formatting
   bookdown::word_document2:
@@
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented.
-We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing), allowing reliable inference from the data.
+We present the basic theory of GAMs, and using reported trends of oxygen saturation in tumors we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R.
+We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing).
 To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research.

### 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/SIM_Appendix.tex
Created folder for Wiley submission

--- 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/00-OUP_Appendix.tex
@@
 
-\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}\\
-\emph{Beyond repeated measures ANOVA and Linear Mixed Models}\\
+\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
+Beyond repeated measures ANOVA and Linear Mixed Models\\
 ~\\
@@
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented.
-We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing), allowing reliable inference from the data.
+We present the basic theory of GAMs, and using reported trends of oxygen saturation in tumors we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R.
+We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing).
 To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/l-q-response-1}
+{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/l-q-response-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/basis-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/basis-plot-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/sim-smooth-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/plot-pairwise-comp-1}
+{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/plot-pairwise-comp-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/linear-cases-Appendix-1}
+{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/linear-cases-Appendix-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/quadratic-cases-Appendix-1}
+{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/quadratic-cases-Appendix-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/basis-plot-appendix-1}
+{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/basis-plot-appendix-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/first-GAM-diag-1}
+{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/first-GAM-diag-1}
 
@@
 
-\begin{center}\includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/first-GAM-check-1} \end{center}
+\begin{center}\includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/first-GAM-check-1} \end{center}
 
@@
 
-\begin{center}\includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/second-GAM-1} \end{center}
+\begin{center}\includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/second-GAM-1} \end{center}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/second-GAM-diag-1}
+{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/second-GAM-diag-1}
 
@@
 
-\begin{center}\includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/final-model-Appendix-1} \end{center}
+\begin{center}\includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/final-model-Appendix-1} \end{center}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/final-GAM-diag-1}
+{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/final-GAM-diag-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/pairwise-comp-workflow-fig-1}
+{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/pairwise-comp-workflow-fig-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/sim-smooth-plot-Appendix-1}
+{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/sim-smooth-plot-Appendix-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-OUP_Appendix_files/figure-latex/plot-pairwise-comp-Appendix-1}
+{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/plot-pairwise-comp-Appendix-1}
 

### 2021-10-13 Manuscripts/Manuscript by chapters- SIM Revisions/00-Full_document_No_Appendix.Rmd
Started changes in Appendix, need to figure how to break up long code chunks and add text in between.

--- 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/00-Full_document_OUP_No_Appendix.Rmd
@@
     #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
+    fontsize: 12pt
+    fontfamily: helvetica
     pandoc_args: --listings #calls the listings package to fit code within the page margins
@@
     includes:
-      in_header: preamble_OUP.sty #additional LaTeX formatting
+      in_header: preamble.sty #additional LaTeX formatting
   bookdown::word_document2:
@@
     css: "style.css" #style for the HTML document
-csl: biostatistics.csl #style for references
+csl: american-medical-association.csl #style for references
 bibliography: refs.bib #references
@@
 
-# Summary
+# Abstract
 

### 2021-10-13 Manuscripts/Manuscript by chapters- SIM Revisions/00-Full_document_SIM_No_Appendix.Rmd
Started changes in Appendix, need to figure how to break up long code chunks and add text in between.

--- 2021-10-13 Manuscripts/Manuscript by chapters- SIM Revisions/00-Full_document_No_Appendix.Rmd
@@
 ---
-title: '**Using generalized additive models to analyze biomedical non-linear longitudinal data**'
-subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
+title: '**Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models**'
 header-includes:
@@
 
-In contrast, _generalized additive models_ (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
 Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented.
-We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing), allowing reliable inference from the data.
+We present the basic theory of GAMs, and using reported trends of oxygen saturation in tumors we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R.
+We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing).
 To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research.

### 2021-10-13 Manuscripts/Manuscript by chapters- SIM Revisions/04-Longitudinal analysis with GAMs.Rmd
Started changes in Appendix, need to figure how to break up long code chunks and add text in between.

--- 2021-06-10 Manuscripts/Manuscript by chapters -v2/04-Longitudinal analysis with GAMs.Rmd
@@
 The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (*Day*) and the other is a factor (*Group*).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
+Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the group means.
 Using $\textsf{R}$ and the package `mgcv` the model syntax is:
 
-`m1 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML, data = dat_sim)`
+```{r, eval = FALSE}
+gam_02 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
+```
 
-This syntax specifies that `m1` will store the model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using 5 knots.
+This syntax specifies that `gam_02` (named accordingly so it matches the model sequence from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using 5 knots.
 The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths [@simpson2018].
@@
 
-m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
+gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
             method='REML',
@@
 gam_predict<-gam_predict%>%
-    mutate(fit = predict(m1,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(m1, gam_predict,se.fit = TRUE,type='response')$se.fit)
+    mutate(fit = predict(gam_02,gam_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(gam_02, gam_predict,se.fit = TRUE,type='response')$se.fit)
 
@@
 
-#posthoc comparisons for the linear model
-#library(multcomp)
-
-#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
-
 ```
@@
 Consider the  simulated $\mbox{StO}_2$ values from Figure \@ref(fig:sim-smooth-plot)B.
-If 40% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \@ref(fig:sim-smooth-plot)D).
-Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.
+If 40% of the observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \@ref(fig:sim-smooth-plot)D).
+Although the confidence intervals have increased for both smooths, the model still shows different trends with as few as 4 observations per group at certain time points.
 
@@
 
-comp1<-smooth_diff(m1,pdat,'Control','Treatment')
+comp1<-smooth_diff(gam_02,pdat,'Control','Treatment')
 
@@
 
-With this expectation of different trends in each group, computing the difference between the trends will identify if the observed change is significant.
+With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant.
 The difference between groups with similar trends is likely to yield zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
@@
 
-On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the empirical Bayesian credible intervals of the smooths overlap between days 0 and 3.
-Consequently, the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that there is no evidence of a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.
+On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the empirical Bayesian credible intervals of the smooths shows the same trend of the full dataset.
+Consequently, the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that the Control Group has higher $\mbox{StO}_2$ before day 3, and is able to estimate the change on day 3 where the Treatment Group becomes significant as the full dataset smooth pairwise comparison.
 

### 2021-10-13 Manuscripts/Manuscript by chapters- SIM Revisions/07-Appendix.Rmd
Started changes in Appendix, need to figure how to break up long code chunks and add text in between.

--- 2021-06-22 Manuscripts/Manuscript by chapters -for submission/07-Appendix.Rmd
@@
 For model check, the functions `gam.check` and `summary` from  _mgcv_ provide detailed information about the model fit and its parameters.
-
+Keep in mind that `gam.check` is a function that also provides the graphical diagnostics obtained using _gratia_, if such graphical output is not desired the source code can be accessed typing `gam.check` in the Console, and the code without the graphical output can be used in a custom function (see \@ref(#gam-00-model-check) for an implementation of this).
 (ref:first-GAM-caption) Graphical diagnostics for the first GAM model.
@@
 
-#plot(layout2)
 ```
@@
 
-#### Model check
-
-Special attention must be paid to the 'k-index' from `gam.check`.
+#### Model check{#gam-00-model-check}
+
+Special attention must be paid to the parameter 'k-index' from `gam.check` (which calls the function `k.check` to perform the calculation).
 This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
-If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (<1).
+If the model is not adequately capturing the trends in the data, this is indicated by a low k-index value (<1).
+Because we plot the model diagnostics using `appraise` later, the graphical output from `gam.check` will be suppressed by creating a custom function to obtain just the model estimates, thus avoiding repetition of the diagnostic plots.
+This will be achieved by calling the source code of `gam.check` and using the appropriate code in a new function that will be called `gam.diagnostics`.
 
@@
 
-gam.check(gam_00)
+#Function that uses the source code of `gam.check` to obtain the estimates without the plots. The source can be checked by typing `gam.check` in the console.
+gam.diagnostics<-function (b, old.style = FALSE, type = c("deviance", "pearson", 
+    "response"), k.sample = 5000, k.rep = 200, rep = 0, 
+    level = 0.9, rl.col = 2, rep.col = "gray80", ...) 
+{
+    type <- match.arg(type)
+    resid <- residuals(b, type = type)
+    linpred <- if (is.matrix(b$linear.predictors) && !is.matrix(resid)) 
+        napredict(b$na.action, b$linear.predictors[, 1])
+    else napredict(b$na.action, b$linear.predictors)
+
+    fv <- if (inherits(b$family, "extended.family")) 
+        predict(b, type = "response")
+    else fitted(b)
+    if (is.matrix(fv) && !is.matrix(b$y)) 
+        fv <- fv[, 1]
+    gamm <- !(b$method %in% c("GCV", "GACV", "UBRE", 
+        "REML", "ML", "P-ML", "P-REML", 
+        "fREML"))
+    if (gamm) {
+        cat("\n'gamm' based fit - care required with interpretation.")
+        cat("\nChecks based on working residuals may be misleading.")
+    }
+    else {
+        cat("\nMethod:", b$method, "  Optimizer:", 
+            b$optimizer)
+        if (!is.null(b$outer.info)) {
+            if (b$optimizer[2] %in% c("newton", "bfgs")) {
+                boi <- b$outer.info
+                cat("\n", boi$conv, " after ", boi$iter, 
+                  " iteration", sep = "")
+                if (boi$iter == 1) 
+                  cat(".")
+                else cat("s.")
+                cat("\nGradient range [", min(boi$grad), 
+                  ",", max(boi$grad), "]", sep = "")
+                cat("\n(score ", b$gcv.ubre, " & scale ", 
+                  b$sig2, ").", sep = "")
+                ev <- eigen(boi$hess)$values
+                if (min(ev) > 0) 
+                  cat("\nHessian positive definite, ")
+                else cat("\n")
+                cat("eigenvalue range [", min(ev), ",", 
+                  max(ev), "].\n", sep = "")
+            }
+            else {
+                cat("\n")
+                print(b$outer.info)
+            }
+        }
+        else {
+            if (length(b$sp) == 0) 
+                cat("\nModel required no smoothing parameter selection")
+            else {
+                cat("\nSmoothing parameter selection converged after", 
+                  b$mgcv.conv$iter, "iteration")
+                if (b$mgcv.conv$iter > 1) 
+                  cat("s")
+                if (!b$mgcv.conv$fully.converged) 
+                  cat(" by steepest\ndescent step failure.\n")
+                else cat(".\n")
+                cat("The RMS", b$method, "score gradient at convergence was", 
+                  b$mgcv.conv$rms.grad, ".\n")
+                if (b$mgcv.conv$hess.pos.def) 
+                  cat("The Hessian was positive definite.\n")
+                else cat("The Hessian was not positive definite.\n")
+            }
+        }
+        if (!is.null(b$rank)) {
+            cat("Model rank = ", b$rank, "/", length(b$coefficients), 
+                "\n")
+        }
+    }
+    cat("\n")
+    kchck <- k.check(b, subsample = k.sample, n.rep = k.rep)
+    if (!is.null(kchck)) {
+        cat("Basis dimension (k) checking results. Low p-value (k-index<1) may\n")
+        cat("indicate that k is too low, especially if edf is close to k'.\n\n")
+        printCoefmat(kchck, digits = 3)
+    }
+}
+```
+
+```{r,gam-00-diagnostics, echo=TRUE, include=TRUE}
+gam.diagnostics(gam_00)
 
@@
 
-From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data.
+From the output, it can be seen that the `k-index` is 0.36, which indicates that the model is not capturing the variability in the data.
 The `edf` (effective degrees of freedom) is an indicator of the complexity of the smooth.
@@
 
-gam.check(gam_01)
+gam.diagnostics(gam_01)
 summary(gam_01)
@@
 
-gam.check(m1)
-summary(m1)
-```
-
-The resulting model is `m1`, which is the model fitted in the main manuscript.
+gam.diagnostics(gam_02)
+summary(gam_02)
+```
+
+The resulting model is `gam_02`, which is the model fitted in the main manuscript.
 By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
@@
 
-appr3<-appraise(m1)
-sm3<-draw(m1)
+appr3<-appraise(gam_02)
+sm3<-draw(gam_02)
 
@@
 
-AIC(gam_00,gam_01,m1)
-
-```
-
-The output in this case is expected: model `m1` has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
+AIC(gam_00,gam_01,gam_02)
+
+```
+
+The output in this case is expected: model `gam_02` has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
 The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
@@
 In this case, the "design matrix" is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
-Briefly, the "design matrix" (also known as the "Xp matrix") from the selected model (`m1`) is used to calculate a 95% confidence interval of the difference between the smooth terms for each group.
+Briefly, the "design matrix" (also known as the "Xp matrix") from the selected model (`gam_02`) is used to calculate a 95% confidence interval of the difference between the smooth terms for each group.
 This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
@@
 ##matrix that contains the basis functions evaluated at the points in pdat
-    xp <- predict(m1, newdata = pdat, type = 'lpmatrix')
+    xp <- predict(gam_02, newdata = pdat, type = 'lpmatrix')
 
@@
     #dimensions (n,1). The resulting matrix has dimensions (p,1)
-    dif <- X %*% coef(m1)
-
-    #comp<-test %*% coef(gam1)[3:10]
+    dif <- X %*% coef(gam_02)
 
@@
     #of the model
-    se <- sqrt(rowSums((X %*% vcov(m1, unconditional = FALSE)) * X))
-    crit <- qt(0.05/2, df.residual(m1), lower.tail = FALSE)
+    se <- sqrt(rowSums((X %*% vcov(gam_02, unconditional = FALSE)) * X))
+    crit <- qt(0.05/2, df.residual(gam_02), lower.tail = FALSE)
     #upper  limits
@@
 
-(ref:pairwise-comp-caption) Smooth pairwise comparisons for model `m1` using a 95% confidence interval for the difference between smooths.
+(ref:pairwise-comp-caption) Smooth pairwise comparisons for model `gam_02` using a 95% confidence interval for the difference between smooths.
 The comparison is centered at the response scale.
@@
 
-This section covers the code used to generate Figure \@ref(fig:sim-smooth-plot), where the simulated data, fit of the "final" GAM (`m1`), linear model  and GAM on data with missing observations are presented.
+This section covers the code used to generate Figure \@ref(fig:sim-smooth-plot), where the simulated data, fit of the "final" GAM (`gam_02`), linear model  and GAM on data with missing observations are presented.
 Note that panel A in Figure \@ref(fig:sim-smooth-plot) and the inset are generated in the code chunk where the data is simulated in Section \@ref(tumor-data-simulation), and are called later to build the figure.
@@
 This code chunk creates panels B and D in Figure \@ref(fig:sim-smooth-plot).
-Note that this code uses the final GAM from the previous section (`m1`), so the simulated data and the model should be generated before running this section.
+Note that this code uses the final GAM from the previous section (`gam_02`), so the simulated data and the model should be generated before running this section.
 
@@
 #use the function to calculate the difference in smooths
-comp1<-smooth_diff(m1,pdat,'Control','Treatment')
+comp1<-smooth_diff(gam_02,pdat,'Control','Treatment')
 

### 2021-10-13 Manuscripts/Manuscript by chapters- SIM Revisions/SIM_Appendix.tex
Started changes in Appendix, need to figure how to break up long code chunks and add text in between.

--- 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/SIM_Appendix.tex
@@
 ]{article}
+\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
+Beyond repeated measures ANOVA and Linear Mixed Models\\
+\strut \\
+SUPPLEMENTARY MATERIALS: APPENDIX}
+\date{\vspace{-2.5em}}
+
 \usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\usepackage{iftex}
+\ifPDFTeX
   \usepackage[T1]{fontenc}
@@
 \setcounter{secnumdepth}{5}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 %\usepackage{lineno}
@@
 \usepackage[font={small}]{caption}
-\ifluatex
+\ifLuaTeX
   \usepackage{selnolig}  % disable illegal ligatures
 \fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
-Beyond repeated measures ANOVA and Linear Mixed Models\\
-~\\
-SUPPLEMENTARY MATERIALS: APPENDIX}
-\date{\vspace{-2.5em}}
 
@@
 The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups.
+Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the group means.
 Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:
@@
 \begin{lstlisting}[language=R]
-m1 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
+gam_02 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
 \end{lstlisting}
 
-This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
+This syntax specifies that \passthrough{\lstinline!gam\_02!} (named accordingly so it matches the model sequence from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
 The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
@@
 Consider the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}B.
-If 40\% of the total observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \ref{fig:sim-smooth-plot}D).
-Although the confidence intervals have increased for both smooths, the model still shows different trends with as little as 4 observations per group at certain time points.
+If 40\% of the observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \ref{fig:sim-smooth-plot}D).
+Although the confidence intervals have increased for both smooths, the model still shows different trends with as few as 4 observations per group at certain time points.
 
@@
 
-On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the empirical Bayesian credible intervals of the smooths overlap between days 0 and 3.
-Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that there is no evidence of a significant difference between the groups during that period, but is still able to pick the change on day 3 as the full dataset smooth pairwise comparison.
+On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the empirical Bayesian credible intervals of the smooths shows the same trend of the full dataset.
+Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that the Control Group has higher \(\mbox{StO}_2\) before day 3, and is able to estimate the change on day 3 where the Treatment Group becomes significant as the full dataset smooth pairwise comparison.
 
@@
 \begin{CSLReferences}{0}{0}
-\leavevmode\hypertarget{ref-roblyer2011}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-roblyer2011}{}}%
 \CSLLeftMargin{{[}1{]} }
@@
 
-\leavevmode\hypertarget{ref-tank2020}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-tank2020}{}}%
 \CSLLeftMargin{{[}2{]} }
@@
 
-\leavevmode\hypertarget{ref-pavlov2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-pavlov2018}{}}%
 \CSLLeftMargin{{[}3{]} }
@@
 
-\leavevmode\hypertarget{ref-demidov2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-demidov2018}{}}%
 \CSLLeftMargin{{[}4{]} }
@@
 
-\leavevmode\hypertarget{ref-ritter2001}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-ritter2001}{}}%
 \CSLLeftMargin{{[}5{]} }
@@
 
-\leavevmode\hypertarget{ref-roth2017}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-roth2017}{}}%
 \CSLLeftMargin{{[}6{]} }
@@
 
-\leavevmode\hypertarget{ref-jones2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-jones2018}{}}%
 \CSLLeftMargin{{[}7{]} }
@@
 
-\leavevmode\hypertarget{ref-skala2010}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-skala2010}{}}%
 \CSLLeftMargin{{[}8{]} }
@@
 
-\leavevmode\hypertarget{ref-greening2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-greening2018}{}}%
 \CSLLeftMargin{{[}9{]} }
@@
 
-\leavevmode\hypertarget{ref-sio2016}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-sio2016}{}}%
 \CSLLeftMargin{{[}10{]} }
@@
 
-\leavevmode\hypertarget{ref-kamstra2015}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-kamstra2015}{}}%
 \CSLLeftMargin{{[}11{]} }
@@
 
-\leavevmode\hypertarget{ref-wagenmakers2008}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wagenmakers2008}{}}%
 \CSLLeftMargin{{[}12{]} }
@@
 
-\leavevmode\hypertarget{ref-gueorguieva2004}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-gueorguieva2004}{}}%
 \CSLLeftMargin{{[}13{]} }
@@
 
-\leavevmode\hypertarget{ref-schober2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-schober2018}{}}%
 \CSLLeftMargin{{[}14{]} }
@@
 
-\leavevmode\hypertarget{ref-pinheiro2006}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-pinheiro2006}{}}%
 \CSLLeftMargin{{[}15{]} }
@@
 
-\leavevmode\hypertarget{ref-vishwanath2009}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-vishwanath2009}{}}%
 \CSLLeftMargin{{[}16{]} }
@@
 
-\leavevmode\hypertarget{ref-dennis2019}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-dennis2019}{}}%
 \CSLLeftMargin{{[}17{]} }
@@
 
-\leavevmode\hypertarget{ref-wang2019}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wang2019}{}}%
 \CSLLeftMargin{{[}18{]} }
@@
 
-\leavevmode\hypertarget{ref-liu2010}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-liu2010}{}}%
 \CSLLeftMargin{{[}19{]} }
@@
 
-\leavevmode\hypertarget{ref-halsey2015}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-halsey2015}{}}%
 \CSLLeftMargin{{[}20{]} }
@@
 
-\leavevmode\hypertarget{ref-abdi2010}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-abdi2010}{}}%
 \CSLLeftMargin{{[}21{]} }
@@
 
-\leavevmode\hypertarget{ref-nakagawa2004}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-nakagawa2004}{}}%
 \CSLLeftMargin{{[}22{]} }
@@
 
-\leavevmode\hypertarget{ref-gelman2012}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-gelman2012}{}}%
 \CSLLeftMargin{{[}23{]} }
@@
 Hill, M.
-Yajima, Why we (usually) don{{}}t have to worry about multiple comparisons, Journal of Research on Educational Effectiveness.
+Yajima, Why we (usually) don{\textquotesingle}t have to worry about multiple comparisons, Journal of Research on Educational Effectiveness.
 5 (2012) 189--211.
@@
 
-\leavevmode\hypertarget{ref-albers2019}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-albers2019}{}}%
 \CSLLeftMargin{{[}24{]} }
@@
 
-\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-ugrinowitsch2004}{}}%
 \CSLLeftMargin{{[}25{]} }
@@
 
-\leavevmode\hypertarget{ref-huynh1976}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-huynh1976}{}}%
 \CSLLeftMargin{{[}26{]} }
@@
 
-\leavevmode\hypertarget{ref-greenhouse1959}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-greenhouse1959}{}}%
 \CSLLeftMargin{{[}27{]} }
@@
 
-\leavevmode\hypertarget{ref-haverkamp2017}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-haverkamp2017}{}}%
 \CSLLeftMargin{{[}28{]} }
@@
 
-\leavevmode\hypertarget{ref-keselman2001}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-keselman2001}{}}%
 \CSLLeftMargin{{[}29{]} }
@@
 
-\leavevmode\hypertarget{ref-charan2013}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-charan2013}{}}%
 \CSLLeftMargin{{[}30{]} }
@@
 
-\leavevmode\hypertarget{ref-barr2013}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-barr2013}{}}%
 \CSLLeftMargin{{[}31{]} }
@@
 
-\leavevmode\hypertarget{ref-rose2012}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-rose2012}{}}%
 \CSLLeftMargin{{[}32{]} }
@@
 
-\leavevmode\hypertarget{ref-pedersen2019}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-pedersen2019}{}}%
 \CSLLeftMargin{{[}33{]} }
@@
 
-\leavevmode\hypertarget{ref-simpson2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-simpson2018}{}}%
 \CSLLeftMargin{{[}34{]} }
@@
 
-\leavevmode\hypertarget{ref-yang2012}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-yang2012}{}}%
 \CSLLeftMargin{{[}35{]} }
@@
 
-\leavevmode\hypertarget{ref-beck1998}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-beck1998}{}}%
 \CSLLeftMargin{{[}36{]} }
@@
 
-\leavevmode\hypertarget{ref-wood2017}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wood2017}{}}%
 \CSLLeftMargin{{[}37{]} }
@@
 
-\leavevmode\hypertarget{ref-r}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-r}{}}%
 \CSLLeftMargin{{[}38{]} }
@@
 
-\leavevmode\hypertarget{ref-wood2016}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wood2016}{}}%
 \CSLLeftMargin{{[}39{]} }
@@
 
-\leavevmode\hypertarget{ref-west2014}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-west2014}{}}%
 \CSLLeftMargin{{[}40{]} }
@@
 
-\leavevmode\hypertarget{ref-wolfinger1996}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wolfinger1996}{}}%
 \CSLLeftMargin{{[}41{]} }
@@
 
-\leavevmode\hypertarget{ref-weiss2005}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-weiss2005}{}}%
 \CSLLeftMargin{{[}42{]} }
@@
 
-\leavevmode\hypertarget{ref-geisser1958}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-geisser1958}{}}%
 \CSLLeftMargin{{[}43{]} }
@@
 Geisser, S.W.
-Greenhouse, An extension of box{{}}s results on the use of the {{}}f{{}} distribution in multivariate analysis, The Annals of Mathematical Statistics.
+Greenhouse, An extension of box{\textquotesingle}s results on the use of the {\textdollar}f{\textdollar} distribution in multivariate analysis, The Annals of Mathematical Statistics.
 29 (1958) 885--891.
@@
 
-\leavevmode\hypertarget{ref-maxwell2017}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-maxwell2017}{}}%
 \CSLLeftMargin{{[}44{]} }
@@
 
-\leavevmode\hypertarget{ref-molenberghs2004}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-molenberghs2004}{}}%
 \CSLLeftMargin{{[}45{]} }
@@
 
-\leavevmode\hypertarget{ref-scheffer2002}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-scheffer2002}{}}%
 \CSLLeftMargin{{[}46{]} }
@@
 
-\leavevmode\hypertarget{ref-potthoff2006}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-potthoff2006}{}}%
 \CSLLeftMargin{{[}47{]} }
@@
 
-\leavevmode\hypertarget{ref-ma2012}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-ma2012}{}}%
 \CSLLeftMargin{{[}48{]} }
@@
 
-\leavevmode\hypertarget{ref-nlme}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-nlme}{}}%
 \CSLLeftMargin{{[}49{]} }
@@
 
-\leavevmode\hypertarget{ref-nelder1972}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-nelder1972}{}}%
 \CSLLeftMargin{{[}50{]} }
@@
 
-\leavevmode\hypertarget{ref-hastie1987}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-hastie1987}{}}%
 \CSLLeftMargin{{[}51{]} }
@@
 
-\leavevmode\hypertarget{ref-hefley2017}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-hefley2017}{}}%
 \CSLLeftMargin{{[}52{]} }
@@
 
-\leavevmode\hypertarget{ref-wegman1983}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wegman1983}{}}%
 \CSLLeftMargin{{[}53{]} }
@@
 
-\leavevmode\hypertarget{ref-mcelreath2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-mcelreath2018}{}}%
 \CSLLeftMargin{{[}54{]} }
@@
 
-\leavevmode\hypertarget{ref-miller2019}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-miller2019}{}}%
 \CSLLeftMargin{{[}55{]} }
@@
 
-\leavevmode\hypertarget{ref-laird1982}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-laird1982}{}}%
 \CSLLeftMargin{{[}56{]} }
@@
 
-\leavevmode\hypertarget{ref-marra2012}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
 \CSLLeftMargin{{[}57{]} }
@@
 
-\leavevmode\hypertarget{ref-gratia}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
 \CSLLeftMargin{{[}58{]} }
@@
 
-\leavevmode\hypertarget{ref-harezlak2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
 \CSLLeftMargin{{[}59{]} }
@@
 
-\leavevmode\hypertarget{ref-lang2015}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
 \CSLLeftMargin{{[}60{]} }
@@
 
-\leavevmode\hypertarget{ref-hastie1995}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-hastie1995}{}}%
 \CSLLeftMargin{{[}61{]} }
@@
 
-\leavevmode\hypertarget{ref-begley2015}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-begley2015}{}}%
 \CSLLeftMargin{{[}62{]} }
@@
 
-\leavevmode\hypertarget{ref-weissgerber2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-weissgerber2018}{}}%
 \CSLLeftMargin{{[}63{]} }
@@
 Milic, S.J.
-Winham, Why we need to report more than {{}}data were analyzed by t-tests or {ANOVA}{{}}, {eLife}.
+Winham, Why we need to report more than {\textquotesingle}data were analyzed by t-tests or {ANOVA}{\textquotesingle}, {eLife}.
 7 (2018).
@@
 For model check, the functions \passthrough{\lstinline!gam.check!} and \passthrough{\lstinline!summary!} from \emph{mgcv} provide detailed information about the model fit and its parameters.
+Keep in mind that \passthrough{\lstinline!gam.check!} is a function that also provides the graphical diagnostics obtained using \emph{gratia}, if such graphical output is not desired the source code can be accessed typing \passthrough{\lstinline!gam.check!} in the Console, and the code without the graphical output can be used in a custom function (see @ref(\#gam-00-model-check) for an implementation of this).
 
@@
 
-\hypertarget{model-check}{%
-\paragraph{Model check}\label{model-check}}
-
-Special attention must be paid to the `k-index' from \passthrough{\lstinline!gam.check!}.
+\hypertarget{gam-00-model-check}{%
+\paragraph{Model check}\label{gam-00-model-check}}
+
+Special attention must be paid to the parameter `k-index' from \passthrough{\lstinline!gam.check!} (which calls the function \passthrough{\lstinline!k.check!} to perform the calculation).
 This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
-If the model is not adequately capturing the trens in the data, this is indicated by a low k-index value (\textless1).
+If the model is not adequately capturing the trends in the data, this is indicated by a low k-index value (\textless1).
+Because we plot the model diagnostics using \passthrough{\lstinline!appraise!} later, the graphical output from \passthrough{\lstinline!gam.check!} will be suppressed by creating a custom function to obtain just the model estimates, thus avoiding repetition of the diagnostic plots.
+This will be achieved by calling the source code of \passthrough{\lstinline!gam.check!} and using the appropriate code in a new function that will be called \passthrough{\lstinline!gam.diagnostics!}.
 
 \begin{lstlisting}[language=R]
-gam.check(gam_00)
+#Function that uses the source code of `gam.check` to obtain the estimates without the plots. The source can be checked by typing `gam.check` in the console.
+gam.diagnostics<-function (b, old.style = FALSE, type = c("deviance", "pearson", 
+    "response"), k.sample = 5000, k.rep = 200, rep = 0, 
+    level = 0.9, rl.col = 2, rep.col = "gray80", ...) 
+{
+    type <- match.arg(type)
+    resid <- residuals(b, type = type)
+    linpred <- if (is.matrix(b$linear.predictors) && !is.matrix(resid)) 
+        napredict(b$na.action, b$linear.predictors[, 1])
+    else napredict(b$na.action, b$linear.predictors)
+
+    fv <- if (inherits(b$family, "extended.family")) 
+        predict(b, type = "response")
+    else fitted(b)
+    if (is.matrix(fv) && !is.matrix(b$y)) 
+        fv <- fv[, 1]
+    gamm <- !(b$method %in% c("GCV", "GACV", "UBRE", 
+        "REML", "ML", "P-ML", "P-REML", 
+        "fREML"))
+    if (gamm) {
+        cat("\n'gamm' based fit - care required with interpretation.")
+        cat("\nChecks based on working residuals may be misleading.")
+    }
+    else {
+        cat("\nMethod:", b$method, "  Optimizer:", 
+            b$optimizer)
+        if (!is.null(b$outer.info)) {
+            if (b$optimizer[2] %in% c("newton", "bfgs")) {
+                boi <- b$outer.info
+                cat("\n", boi$conv, " after ", boi$iter, 
+                  " iteration", sep = "")
+                if (boi$iter == 1) 
+                  cat(".")
+                else cat("s.")
+                cat("\nGradient range [", min(boi$grad), 
+                  ",", max(boi$grad), "]", sep = "")
+                cat("\n(score ", b$gcv.ubre, " & scale ", 
+                  b$sig2, ").", sep = "")
+                ev <- eigen(boi$hess)$values
+                if (min(ev) > 0) 
+                  cat("\nHessian positive definite, ")
+                else cat("\n")
+                cat("eigenvalue range [", min(ev), ",", 
+                  max(ev), "].\n", sep = "")
+            }
+            else {
+                cat("\n")
+                print(b$outer.info)
+            }
+        }
+        else {
+            if (length(b$sp) == 0) 
+                cat("\nModel required no smoothing parameter selection")
+            else {
+                cat("\nSmoothing parameter selection converged after", 
+                  b$mgcv.conv$iter, "iteration")
+                if (b$mgcv.conv$iter > 1) 
+                  cat("s")
+                if (!b$mgcv.conv$fully.converged) 
+                  cat(" by steepest\ndescent step failure.\n")
+                else cat(".\n")
+                cat("The RMS", b$method, "score gradient at convergence was", 
+                  b$mgcv.conv$rms.grad, ".\n")
+                if (b$mgcv.conv$hess.pos.def) 
+                  cat("The Hessian was positive definite.\n")
+                else cat("The Hessian was not positive definite.\n")
+            }
+        }
+        if (!is.null(b$rank)) {
+            cat("Model rank = ", b$rank, "/", length(b$coefficients), 
+                "\n")
+        }
+    }
+    cat("\n")
+    kchck <- k.check(b, subsample = k.sample, n.rep = k.rep)
+    if (!is.null(kchck)) {
+        cat("Basis dimension (k) checking results. Low p-value (k-index<1) may\n")
+        cat("indicate that k is too low, especially if edf is close to k'.\n\n")
+        printCoefmat(kchck, digits = 3)
+    }
+}
 \end{lstlisting}
 
-\begin{center}\includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/first-GAM-check-1} \end{center}
+\begin{lstlisting}[language=R]
+gam.diagnostics(gam_00)
+\end{lstlisting}
 
@@
 
-From the output, it can be seen that the k-index is 0.36, which indicates that the model is not capturing the variability in the data.
+From the output, it can be seen that the \passthrough{\lstinline!k-index!} is 0.36, which indicates that the model is not capturing the variability in the data.
 The \passthrough{\lstinline!edf!} (effective degrees of freedom) is an indicator of the complexity of the smooth.
@@
 
-gam.check(gam_01)
+gam.diagnostics(gam_01)
 \end{lstlisting}
-
-\begin{center}\includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/second-GAM-1} \end{center}
 
@@
 
-gam.check(m1)
+gam.diagnostics(gam_02)
 \end{lstlisting}
-
-\begin{center}\includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/final-model-Appendix-1} \end{center}
 
@@
 ##                         k'  edf k-index p-value
-## s(Day):GroupControl   4.00 3.87    1.02    0.59
-## s(Day):GroupTreatment 4.00 3.88    1.02    0.54
+## s(Day):GroupControl   4.00 3.87    1.02    0.58
+## s(Day):GroupTreatment 4.00 3.88    1.02    0.56
 \end{lstlisting}
@@
 \begin{lstlisting}[language=R]
-summary(m1)
+summary(gam_02)
 \end{lstlisting}
@@
 
-The resulting model is \passthrough{\lstinline!m1!}, which is the model fitted in the main manuscript.
+The resulting model is \passthrough{\lstinline!gam\_02!}, which is the model fitted in the main manuscript.
 By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
@@
 \begin{lstlisting}[language=R]
-AIC(gam_00,gam_01,m1)
+AIC(gam_00,gam_01,gam_02)
 \end{lstlisting}
@@
 ## gam_01  9.476137 858.6051
-## m1     10.980983 712.2067
+## gam_02 10.980983 712.2067
 \end{lstlisting}
 
-The output in this case is expected: model \passthrough{\lstinline!m1!} has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
+The output in this case is expected: model \passthrough{\lstinline!gam\_02!} has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
 The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
@@
 In this case, the ``design matrix'' is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
-Briefly, the ``design matrix'' (also known as the ``Xp matrix'') from the selected model (\passthrough{\lstinline!m1!}) is used to calculate a 95\% confidence interval of the difference between the smooth terms for each group.
+Briefly, the ``design matrix'' (also known as the ``Xp matrix'') from the selected model (\passthrough{\lstinline!gam\_02!}) is used to calculate a 95\% confidence interval of the difference between the smooth terms for each group.
 This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
@@
 
-\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!m1!} using a 95\% confidence interval for the difference between smooths.
+\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!gam\_02!} using a 95\% confidence interval for the difference between smooths.
 The comparison is centered at the response scale.
@@
 
-This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\passthrough{\lstinline!m1!}), linear model and GAM on data with missing observations are presented.
+This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\passthrough{\lstinline!gam\_02!}), linear model and GAM on data with missing observations are presented.
 Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inset are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
@@
 This code chunk creates panels B and D in Figure \ref{fig:sim-smooth-plot}.
-Note that this code uses the final GAM from the previous section (\passthrough{\lstinline!m1!}), so the simulated data and the model should be generated before running this section.
+Note that this code uses the final GAM from the previous section (\passthrough{\lstinline!gam\_02!}), so the simulated data and the model should be generated before running this section.
 
@@
   thm1
-
-#posthoc comparisons for the linear model
-#library(multcomp)
-
-#summary(glht(lm1, linfct = mcp(Group = 'Tukey')))
-#summary(glht(lm1, linfct=mcp(Group="Tukey", interaction_average=TRUE)))
 \end{lstlisting}

### 2021-10-15 Manuscripts/Manuscript by chapters- SIM Revisions/00-Full_document_No_Appendix.tex
Working on reviewers' comments

--- 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/00-Full_document_No_Appendix.tex
@@
 ]{article}
+\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
+\usepackage{etoolbox}
+\makeatletter
+\providecommand{\subtitle}[1]{% add subtitle to \maketitle
+  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
+}
+\makeatother
+\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
+\date{\vspace{-2.5em}}
+
 \usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\usepackage{iftex}
+\ifPDFTeX
   \usepackage[T1]{fontenc}
@@
 \setcounter{secnumdepth}{5}
-\usepackage{helvet}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\usepackage{mathptmx}
 \usepackage{lineno}
@@
 \usepackage[font={small}]{caption}
-\ifluatex
+\ifLuaTeX
   \usepackage{selnolig}  % disable illegal ligatures
 \fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
-\usepackage{etoolbox}
-\makeatletter
-\providecommand{\subtitle}[1]{% add subtitle to \maketitle
-  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
-}
-\makeatother
-\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
-\date{\vspace{-2.5em}}
 
@@
 
-Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{53}} Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{53}} Although different types of splines exist, cubic and thin plate regression splines will be briefly discussed next to give a general idea of the development within the GAM framework.
+
+Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
+The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{51}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{54}} In contrast, Wood\textsuperscript{\protect\hyperlink{ref-wood2003}{54}} provided a computationally efficient implementation of thin plate regression splines(TPRS), which do not require the knot placement, and require simply a choice on the number of basis to be used to create the smoother (for mathematical details see refs.
+\normalfont{[@wood2003]} and).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
+
+\textbf{To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-The timeline can be divided in equally spaced \emph{knots}, each knot being a region where a different set of basis functions will be used.
-Because there are six timepoints for this group, five knots can be used.
-The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is set using default values in the package \emph{mgcv} depending on the number of knots.
-In Figure \ref{fig:basis-plot}A, the four basis functions (and the intercept) are shown.
+A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
+Because the data has 6 unique time points, we can specify a maximum of six basis functions to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).}
+
+If five basis functions are used to fit a GAM for the data that appears in Figure \ref{fig:l-q-response}C, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A.
+The four basis functions (and the intercept) are shown.
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
@@
 Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1).
+Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is \textbf{negative} and less than 1).
 On the other hand, the third basis function has roughly doubled its value.
@@
 
-\caption{Basis functions for a single smoother for time with five knots.
+\caption{Basis functions for a single smoother for time.
 A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{54}} In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case).\textsuperscript{\protect\hyperlink{ref-miller2019}{55}} Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model.\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{56}} Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{55}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{57}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{55}} In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case).\textsuperscript{\protect\hyperlink{ref-miller2019}{56}} Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model.\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{57}} Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{56}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{58}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
 
@@
 \begin{lstlisting}[language=R]
-m1 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
+gam_02 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
 \end{lstlisting}
 
-This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
+This syntax specifies that \passthrough{\lstinline!gam\_02!} (named accordingly so it matches the model sequence from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
 The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{58}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{59}}
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{59}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{60}}
 
@@
 This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{60}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{61}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
 
 In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{61}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{62}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
 
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{62},\protect\hyperlink{ref-weissgerber2018}{63}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{63},\protect\hyperlink{ref-weissgerber2018}{64}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 \begin{CSLReferences}{0}{0}
-\leavevmode\hypertarget{ref-roblyer2011}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-roblyer2011}{}}%
 \CSLLeftMargin{1. }
@@
 
-\leavevmode\hypertarget{ref-tank2020}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-tank2020}{}}%
 \CSLLeftMargin{2. }
@@
 
-\leavevmode\hypertarget{ref-pavlov2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-pavlov2018}{}}%
 \CSLLeftMargin{3. }
@@
 
-\leavevmode\hypertarget{ref-demidov2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-demidov2018}{}}%
 \CSLLeftMargin{4. }
@@
 
-\leavevmode\hypertarget{ref-ritter2001}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-ritter2001}{}}%
 \CSLLeftMargin{5. }
@@
 
-\leavevmode\hypertarget{ref-roth2017}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-roth2017}{}}%
 \CSLLeftMargin{6. }
@@
 
-\leavevmode\hypertarget{ref-jones2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-jones2018}{}}%
 \CSLLeftMargin{7. }
@@
 
-\leavevmode\hypertarget{ref-skala2010}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-skala2010}{}}%
 \CSLLeftMargin{8. }
@@
 
-\leavevmode\hypertarget{ref-greening2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-greening2018}{}}%
 \CSLLeftMargin{9. }
@@
 
-\leavevmode\hypertarget{ref-sio2016}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-sio2016}{}}%
 \CSLLeftMargin{10. }
@@
 
-\leavevmode\hypertarget{ref-kamstra2015}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-kamstra2015}{}}%
 \CSLLeftMargin{11. }
@@
 
-\leavevmode\hypertarget{ref-wagenmakers2008}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wagenmakers2008}{}}%
 \CSLLeftMargin{12. }
@@
 
-\leavevmode\hypertarget{ref-gueorguieva2004}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-gueorguieva2004}{}}%
 \CSLLeftMargin{13. }
@@
 
-\leavevmode\hypertarget{ref-schober2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-schober2018}{}}%
 \CSLLeftMargin{14. }
@@
 
-\leavevmode\hypertarget{ref-pinheiro2006}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-pinheiro2006}{}}%
 \CSLLeftMargin{15. }
@@
 
-\leavevmode\hypertarget{ref-vishwanath2009}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-vishwanath2009}{}}%
 \CSLLeftMargin{16. }
@@
 
-\leavevmode\hypertarget{ref-dennis2019}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-dennis2019}{}}%
 \CSLLeftMargin{17. }
@@
 
-\leavevmode\hypertarget{ref-wang2019}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wang2019}{}}%
 \CSLLeftMargin{18. }
@@
 
-\leavevmode\hypertarget{ref-liu2010}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-liu2010}{}}%
 \CSLLeftMargin{19. }
@@
 
-\leavevmode\hypertarget{ref-halsey2015}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-halsey2015}{}}%
 \CSLLeftMargin{20. }
@@
 
-\leavevmode\hypertarget{ref-abdi2010}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-abdi2010}{}}%
 \CSLLeftMargin{21. }
@@
 
-\leavevmode\hypertarget{ref-nakagawa2004}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-nakagawa2004}{}}%
 \CSLLeftMargin{22. }
@@
 
-\leavevmode\hypertarget{ref-gelman2012}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-gelman2012}{}}%
 \CSLLeftMargin{23. }
 \CSLRightInline{Gelman A, Hill J, Yajima M.
-Why we (usually) don{{}}t have to worry about multiple comparisons.
+Why we (usually) don{\textquotesingle}t have to worry about multiple comparisons.
 \emph{Journal of Research on Educational Effectiveness}.
@@
 
-\leavevmode\hypertarget{ref-albers2019}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-albers2019}{}}%
 \CSLLeftMargin{24. }
@@
 
-\leavevmode\hypertarget{ref-ugrinowitsch2004}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-ugrinowitsch2004}{}}%
 \CSLLeftMargin{25. }
@@
 
-\leavevmode\hypertarget{ref-huynh1976}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-huynh1976}{}}%
 \CSLLeftMargin{26. }
@@
 
-\leavevmode\hypertarget{ref-greenhouse1959}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-greenhouse1959}{}}%
 \CSLLeftMargin{27. }
@@
 
-\leavevmode\hypertarget{ref-haverkamp2017}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-haverkamp2017}{}}%
 \CSLLeftMargin{28. }
@@
 
-\leavevmode\hypertarget{ref-keselman2001}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-keselman2001}{}}%
 \CSLLeftMargin{29. }
@@
 
-\leavevmode\hypertarget{ref-charan2013}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-charan2013}{}}%
 \CSLLeftMargin{30. }
@@
 
-\leavevmode\hypertarget{ref-barr2013}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-barr2013}{}}%
 \CSLLeftMargin{31. }
@@
 
-\leavevmode\hypertarget{ref-rose2012}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-rose2012}{}}%
 \CSLLeftMargin{32. }
@@
 
-\leavevmode\hypertarget{ref-pedersen2019}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-pedersen2019}{}}%
 \CSLLeftMargin{33. }
@@
 
-\leavevmode\hypertarget{ref-simpson2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-simpson2018}{}}%
 \CSLLeftMargin{34. }
@@
 
-\leavevmode\hypertarget{ref-yang2012}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-yang2012}{}}%
 \CSLLeftMargin{35. }
@@
 
-\leavevmode\hypertarget{ref-beck1998}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-beck1998}{}}%
 \CSLLeftMargin{36. }
@@
 
-\leavevmode\hypertarget{ref-wood2017}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wood2017}{}}%
 \CSLLeftMargin{37. }
@@
 
-\leavevmode\hypertarget{ref-r}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-r}{}}%
 \CSLLeftMargin{38. }
@@
 
-\leavevmode\hypertarget{ref-wood2016}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wood2016}{}}%
 \CSLLeftMargin{39. }
@@
 
-\leavevmode\hypertarget{ref-west2014}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-west2014}{}}%
 \CSLLeftMargin{40. }
@@
 
-\leavevmode\hypertarget{ref-wolfinger1996}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wolfinger1996}{}}%
 \CSLLeftMargin{41. }
@@
 
-\leavevmode\hypertarget{ref-weiss2005}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-weiss2005}{}}%
 \CSLLeftMargin{42. }
@@
 
-\leavevmode\hypertarget{ref-geisser1958}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-geisser1958}{}}%
 \CSLLeftMargin{43. }
 \CSLRightInline{Geisser S, Greenhouse SW.
-An extension of box{{}}s results on the use of the {{}}f{{}} distribution in multivariate analysis.
+An extension of box{\textquotesingle}s results on the use of the {\textdollar}f{\textdollar} distribution in multivariate analysis.
 \emph{The Annals of Mathematical Statistics}.
@@
 
-\leavevmode\hypertarget{ref-maxwell2017}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-maxwell2017}{}}%
 \CSLLeftMargin{44. }
@@
 
-\leavevmode\hypertarget{ref-molenberghs2004}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-molenberghs2004}{}}%
 \CSLLeftMargin{45. }
@@
 
-\leavevmode\hypertarget{ref-scheffer2002}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-scheffer2002}{}}%
 \CSLLeftMargin{46. }
@@
 
-\leavevmode\hypertarget{ref-potthoff2006}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-potthoff2006}{}}%
 \CSLLeftMargin{47. }
@@
 
-\leavevmode\hypertarget{ref-ma2012}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-ma2012}{}}%
 \CSLLeftMargin{48. }
@@
 
-\leavevmode\hypertarget{ref-nlme}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-nlme}{}}%
 \CSLLeftMargin{49. }
@@
 
-\leavevmode\hypertarget{ref-nelder1972}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-nelder1972}{}}%
 \CSLLeftMargin{50. }
@@
 
-\leavevmode\hypertarget{ref-hastie1987}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-hastie1987}{}}%
 \CSLLeftMargin{51. }
@@
 
-\leavevmode\hypertarget{ref-hefley2017}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-hefley2017}{}}%
 \CSLLeftMargin{52. }
@@
 
-\leavevmode\hypertarget{ref-wegman1983}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wegman1983}{}}%
 \CSLLeftMargin{53. }
@@
 
-\leavevmode\hypertarget{ref-mcelreath2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wood2003}{}}%
 \CSLLeftMargin{54. }
+\CSLRightInline{Wood SN. Thin plate regression splines. 2003;65(1):95-114. doi:\href{https://doi.org/10.1111/1467-9868.00374}{10.1111/1467-9868.00374}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-mcelreath2018}{}}%
+\CSLLeftMargin{55. }
 \CSLRightInline{McElreath R. \emph{Statistical Rethinking}. Chapman; Hall/{CRC}; 2018. doi:\href{https://doi.org/10.1201/9781315372495}{10.1201/9781315372495}}
 
-\leavevmode\hypertarget{ref-miller2019}{}%
-\CSLLeftMargin{55. }
+\leavevmode\vadjust pre{\hypertarget{ref-miller2019}{}}%
+\CSLLeftMargin{56. }
 \CSLRightInline{Miller DL. Bayesian views of generalized additive modelling. \emph{arXiv preprint arXiv:190201330}. Published online 2019.}
 
-\leavevmode\hypertarget{ref-laird1982}{}%
-\CSLLeftMargin{56. }
+\leavevmode\vadjust pre{\hypertarget{ref-laird1982}{}}%
+\CSLLeftMargin{57. }
 \CSLRightInline{Laird NM, Ware JH. Random-effects models for longitudinal data. \emph{Biometrics}. 1982;38(4):963. doi:\href{https://doi.org/10.2307/2529876}{10.2307/2529876}}
 
-\leavevmode\hypertarget{ref-marra2012}{}%
-\CSLLeftMargin{57. }
+\leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
+\CSLLeftMargin{58. }
 \CSLRightInline{Marra G, Wood SN.
@@
 
-\leavevmode\hypertarget{ref-gratia}{}%
-\CSLLeftMargin{58. }
+\leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
+\CSLLeftMargin{59. }
 \CSLRightInline{Simpson GL. \emph{Gratia: Graceful 'Ggplot'-Based Graphics and Other Functions for GAMs Fitted Using 'Mgcv'}.; 2020. \url{https://CRAN.R-project.org/package=gratia}}
 
-\leavevmode\hypertarget{ref-harezlak2018}{}%
-\CSLLeftMargin{59. }
+\leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
+\CSLLeftMargin{60. }
 \CSLRightInline{Harezlak J, Ruppert D, Wand MP. \emph{Semiparametric Regression with r}. Springer New York; 2018. doi:\href{https://doi.org/10.1007/978-1-4939-8853-2}{10.1007/978-1-4939-8853-2}}
 
-\leavevmode\hypertarget{ref-lang2015}{}%
-\CSLLeftMargin{60. }
+\leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
+\CSLLeftMargin{61. }
 \CSLRightInline{Lang TA, Altman DG.
@@
 
-\leavevmode\hypertarget{ref-hastie1995}{}%
-\CSLLeftMargin{61. }
+\leavevmode\vadjust pre{\hypertarget{ref-hastie1995}{}}%
+\CSLLeftMargin{62. }
 \CSLRightInline{Hastie T, Tibshirani R.
@@
 
-\leavevmode\hypertarget{ref-begley2015}{}%
-\CSLLeftMargin{62. }
+\leavevmode\vadjust pre{\hypertarget{ref-begley2015}{}}%
+\CSLLeftMargin{63. }
 \CSLRightInline{Begley CG, Ioannidis JPA.
@@
 
-\leavevmode\hypertarget{ref-weissgerber2018}{}%
-\CSLLeftMargin{63. }
+\leavevmode\vadjust pre{\hypertarget{ref-weissgerber2018}{}}%
+\CSLLeftMargin{64. }
 \CSLRightInline{Weissgerber TL, Garcia-Valencia O, Garovic VD, Milic NM, Winham SJ.
-Why we need to report more than {{}}data were analyzed by t-tests or {ANOVA}{{}}.
+Why we need to report more than {\textquotesingle}data were analyzed by t-tests or {ANOVA}{\textquotesingle}.
 \emph{{eLife}}.

### 2021-10-15 Manuscripts/Manuscript by chapters- SIM Revisions/03-GAM Theory.Rmd
Working on reviewers' comments

--- 2021-06-10 Manuscripts/Manuscript by chapters -v2/03-GAM Theory.Rmd
@@
 
-Splines (cubic, thin plate, etc.) are commonly used _basis functions_; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data [@wood2017;@simpson2018].
-Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as _basis functions_, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
+Although different types of splines exist, cubic and thin plate regression splines will be briefly discussed next to give a general idea of the development within the GAM framework.
 
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \@ref(fig:l-q-response)C.
+Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
+The use of CS as smoothers in GAMs was discussed within the original GAM framework [@hastie1987], but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as 'knots', the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit [@wood2003].
+In contrast, Wood [@wood2003] provided a computationally efficient implementation of thin plate regression splines(TPRS), which do not require the knot placement, and require simply a choice on the number of basis to be used to create the smoother (for mathematical details see refs.
+\normalfont{[@wood2003]} and [@wood2017]).
+
+**To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \@ref(fig:l-q-response)C.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-The timeline can be divided in equally spaced _knots_, each knot being a region where a different set of basis functions will be used.
-Because there are six timepoints for this group, five knots can be used.
-The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is set using default values in the package _mgcv_ depending on the number of knots.
-In Figure \@ref(fig:basis-plot)A, the four basis functions (and the intercept) are shown.
+A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
+Because the data has 6 unique time points, we can specify a maximum of six basis functions to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).**
+
+If five basis functions are used to fit a GAM for the data that appears in Figure \@ref(fig:l-q-response)C, the resulting fitting process is shown in Figure \@ref(fig:basis-plot)A.
+The four basis functions (and the intercept) are shown.
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
@@
 
-   #GAM model of time, 5 knots
+   #GAM model of time, 5 basis functions
 gm<-gam(y~s(time,k=5),data=dat)
@@
 
-#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
+#length of dataframe to be created: number of basis by number of timepoints (minus 1 for the intercept that we won't plot)
 ln<-6*(length(coef(gm)))
@@
 Figure \@ref(fig:basis-plot)C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1).
+Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is **negative** and less than 1).
 On the other hand, the third basis function has roughly doubled its value.
@@
 
-(ref:basis-plot-caption) Basis functions for a single smoother for time with five knots.
+(ref:basis-plot-caption) Basis functions for a single smoother for time.
 A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.

### 2021-10-17 Manuscripts/Manuscript by chapters- SIM Revisions/02-Challenges.Rmd
Fix comments appearance in code

--- 2021-06-22 Manuscripts/Manuscript by chapters -for submission/02-Challenges.Rmd
@@
 
-```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='(ref:l-q-response-caption)'}
+```{r, l-q-response, fig.width=12, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='(ref:l-q-response-caption)'}
 # linear response, correlated errors (subject effect)

### 2021-10-17 Manuscripts/Manuscript by chapters- SIM Revisions/07-Appendix.Rmd
Fix comments appearance in code

@@
 
-The following code produces a more comprehensive exploration of Figure \@ref(fig:l-q-response) in the main manuscript.
+The following code chunks produce a more comprehensive exploration of Figure \@ref(fig:l-q-response) in the main manuscript.
+
+First, a function is created to simulate data across six timepoints using a linear or quadratic mean response, with correlated or uncorrelated errors.
+Each group has a different slope/concavity.
+The main function is the same for both groups, but a change in the sign allows to invert the trend.
 
@@
 
-  #The following cycles create independent or correlated responses.
+  #The following loops create independent or correlated responses.
 To each value of mu (mean response per group) a randomly generated error (correlated or uncorrelated) is added and thus the individual response is created.
@@
 }
+
 ##################Section for plotting#################################

### 2021-10-17 Manuscripts/Manuscript by chapters- SIM Revisions/SIM_Appendix.tex
Fix comments appearance in code

@@
     sensitive=true,
+    commentstyle=\color{sred},
     % Border (above and below)
@@
 
-Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
-Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
-Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
+Although different types of splines exist, cubic and thin plate regression splines will be briefly discussed next to give a general idea of the development within the GAM framework.
+
+Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
+The use of CS as smoothers in GAMs was discussed within the original GAM framework {[}\protect\hyperlink{ref-hastie1987}{51}{]}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit {[}\protect\hyperlink{ref-wood2003}{54}{]}.
+In contrast, Wood {[}\protect\hyperlink{ref-wood2003}{54}{]} provided a computationally efficient implementation of thin plate regression splines(TPRS), which do not require the knot placement, and require simply a choice on the number of basis to be used to create the smoother (for mathematical details see refs.
+{[}\protect\hyperlink{ref-wood2003}{54}{]} and {[}\protect\hyperlink{ref-wood2017}{37}{]}).
+
+\textbf{To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-The timeline can be divided in equally spaced \emph{knots}, each knot being a region where a different set of basis functions will be used.
-Because there are six timepoints for this group, five knots can be used.
-The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept).
-The choice of basis functions is set using default values in the package \emph{mgcv} depending on the number of knots.
-In Figure \ref{fig:basis-plot}A, the four basis functions (and the intercept) are shown.
+A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
+Because the data has 6 unique time points, we can specify a maximum of six basis functions to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).}
+
+If five basis functions are used to fit a GAM for the data that appears in Figure \ref{fig:l-q-response}C, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A.
+The four basis functions (and the intercept) are shown.
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}\protect\hyperlink{ref-mcelreath2018}{54}{]}.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}\protect\hyperlink{ref-mcelreath2018}{55}{]}.
 In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) {[}\protect\hyperlink{ref-miller2019}{55}{]}.
-Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{56}{]}.
+GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) {[}\protect\hyperlink{ref-miller2019}{56}{]}.
+Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{57}{]}.
 Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.
 This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{55}{]} and {[}\protect\hyperlink{ref-marra2012}{57}{]}.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{56}{]} and {[}\protect\hyperlink{ref-marra2012}{58}{]}.
 With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
@@
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{58}{]}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{59}{]}.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{59}{]}.
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{60}{]}.
 
@@
 This ``model misspecification'' error, also is known as a ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{60}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{61}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
 
 In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{61}{]}.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{62}{]}.
 This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{62},\protect\hyperlink{ref-weissgerber2018}{63}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{63},\protect\hyperlink{ref-weissgerber2018}{64}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-wood2003}{}}%
+\CSLLeftMargin{{[}54{]} }
+\CSLRightInline{S.N. Wood, Thin plate regression splines, 65 (2003) 95--114. \url{https://doi.org/10.1111/1467-9868.00374}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-mcelreath2018}{}}%
-\CSLLeftMargin{{[}54{]} }
+\CSLLeftMargin{{[}55{]} }
 \CSLRightInline{R. McElreath, Statistical rethinking, Chapman; Hall/{CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-miller2019}{}}%
-\CSLLeftMargin{{[}55{]} }
+\CSLLeftMargin{{[}56{]} }
 \CSLRightInline{D.L. Miller, Bayesian views of generalized additive modelling, arXiv Preprint arXiv:1902.01330. (2019).}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-laird1982}{}}%
-\CSLLeftMargin{{[}56{]} }
+\CSLLeftMargin{{[}57{]} }
 \CSLRightInline{N.M. Laird, J.H. Ware, Random-effects models for longitudinal data, Biometrics. 38 (1982) 963. \url{https://doi.org/10.2307/2529876}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
-\CSLLeftMargin{{[}57{]} }
+\CSLLeftMargin{{[}58{]} }
 \CSLRightInline{G.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
-\CSLLeftMargin{{[}58{]} }
+\CSLLeftMargin{{[}59{]} }
 \CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
-\CSLLeftMargin{{[}59{]} }
+\CSLLeftMargin{{[}60{]} }
 \CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with r, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
-\CSLLeftMargin{{[}60{]} }
+\CSLLeftMargin{{[}61{]} }
 \CSLRightInline{T.A.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hastie1995}{}}%
-\CSLLeftMargin{{[}61{]} }
+\CSLLeftMargin{{[}62{]} }
 \CSLRightInline{T.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-begley2015}{}}%
-\CSLLeftMargin{{[}62{]} }
+\CSLLeftMargin{{[}63{]} }
 \CSLRightInline{C.G. Begley, J.P.A. Ioannidis, Reproducibility in science, Circulation Research. 116 (2015) 116--126. \url{https://doi.org/10.1161/circresaha.114.303819}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-weissgerber2018}{}}%
-\CSLLeftMargin{{[}63{]} }
+\CSLLeftMargin{{[}64{]} }
 \CSLRightInline{T.L.
@@
 
-The following code produces a more comprehensive exploration of Figure \ref{fig:l-q-response} in the main manuscript.
+The following code chunks produce a more comprehensive exploration of Figure \ref{fig:l-q-response} in the main manuscript.
+
+First, a function is created to simulate data across six timepoints using a linear or quadratic mean response, with correlated or uncorrelated errors.
+Each group has a different slope/concavity.
+The main function is the same for both groups, but a change in the sign allows to invert the trend.
 

### 2021-10-20 Manuscripts/Manuscript by chapters- SIM Revisions/00-Full_document_SIM_No_Appendix.tex
Need to update the GLMM part (add to LMEM part?)

--- 2021-10-15 Manuscripts/Manuscript by chapters- SIM Revisions/00-Full_document_No_Appendix.tex
@@
 ]{article}
-\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
-\usepackage{etoolbox}
-\makeatletter
-\providecommand{\subtitle}[1]{% add subtitle to \maketitle
-  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
-}
-\makeatother
-\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
+\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}}
 \date{\vspace{-2.5em}}
@@
 
-\hypertarget{missing-observations}{%
-\subsection{Missing observations}\label{missing-observations}}
-
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator.\textsuperscript{\protect\hyperlink{ref-molenberghs2004}{45}} Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR).\textsuperscript{\protect\hyperlink{ref-weiss2005}{42}} In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest.\textsuperscript{\protect\hyperlink{ref-scheffer2002}{46}} If the data are MCAR, this means that the missingness is completely unrelated to the collected information,\textsuperscript{\protect\hyperlink{ref-potthoff2006}{47}} and in the case of MNAR the missing values are dependent on their value.
-
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size.\textsuperscript{\protect\hyperlink{ref-ma2012}{48}} In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR.\textsuperscript{\protect\hyperlink{ref-west2014}{40}} For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
+\hypertarget{unbalanced-data}{%
+\subsection{Unbalanced data}\label{unbalanced-data}}
+
+In a longitudinal study, it is frequently the case that the number of observations is different across all subjects.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury in animals in preclinical research) leading to what is known as ``missing,'' ``incomplete'' or (more properly speaking) unbalanced data.\textsuperscript{\protect\hyperlink{ref-molenberghs2004}{45}} The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size.\textsuperscript{\protect\hyperlink{ref-ma2012}{46}}
+
+On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are \emph{missing at random} (MAR) or \emph{completely missing at random} (MCAR).\textsuperscript{\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}} In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest.\textsuperscript{\protect\hyperlink{ref-scheffer2002}{47}} If the data are MCAR, this means that the missingness is completely unrelated to the collected information,\textsuperscript{\protect\hyperlink{ref-potthoff2006}{48}} and in the case of MNAR the missing values are dependent on their value.
+For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
+
+However, it is worth reminding that essentially ``all models are wrong''\textsuperscript{\protect\hyperlink{ref-box1976}{49}} and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
+Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.
 
@@
 To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\)\textsuperscript{\protect\hyperlink{ref-r}{38}} and the package \emph{nlme}.\textsuperscript{\protect\hyperlink{ref-nlme}{49}}
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\)\textsuperscript{\protect\hyperlink{ref-r}{38}} and the package \emph{nlme}.\textsuperscript{\protect\hyperlink{ref-nlme}{50}}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/l-q-response-1}
+{\centering \includegraphics[width=1\linewidth]{00-Full_document_SIM_No_Appendix_files/figure-latex/l-q-response-1}
 
@@
 
-Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors.\textsuperscript{\protect\hyperlink{ref-nelder1972}{50}} In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{51}}.
+Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors.\textsuperscript{\protect\hyperlink{ref-nelder1972}{51}} In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{52}}.
 A GAM model can be written as:
@@
 A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{52}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{53}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
 In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{53}} Although different types of splines exist, cubic and thin plate regression splines will be briefly discussed next to give a general idea of the development within the GAM framework.
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{54}} Although different types of splines exist, cubic and thin plate regression splines will be briefly discussed next to give a general idea of the development within the GAM framework.
 
 Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{51}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{54}} In contrast, Wood\textsuperscript{\protect\hyperlink{ref-wood2003}{54}} provided a computationally efficient implementation of thin plate regression splines(TPRS), which do not require the knot placement, and require simply a choice on the number of basis to be used to create the smoother (for mathematical details see refs.
-\normalfont{[@wood2003]} and).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
+The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{52}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} In contrast, Wood\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} provided a computationally efficient implementation of thin plate regression splines(TPRS), which do not require the knot placement, and require simply a choice on the number of basis to be used to create the smoother (for mathematical details see refs.\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} and).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
 
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{55}} In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case).\textsuperscript{\protect\hyperlink{ref-miller2019}{56}} Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model.\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{57}} Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{56}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{58}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{56}} In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case).\textsuperscript{\protect\hyperlink{ref-miller2019}{57}} Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model.\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{58}} Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{57}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{59}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
 
@@
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in.\textsuperscript{\protect\hyperlink{ref-hefley2017}{52}}
+A detailed description on basis functions and correlations can be found in.\textsuperscript{\protect\hyperlink{ref-hefley2017}{53}}
 
@@
 The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{59}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{60}}
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{60}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{61}}
 
@@
 This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{61}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{62}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
 
 In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{62}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{53}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{63}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
 
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{63},\protect\hyperlink{ref-weissgerber2018}{64}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{64},\protect\hyperlink{ref-weissgerber2018}{65}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-ma2012}{}}%
+\CSLLeftMargin{46. }
+\CSLRightInline{Ma Y, Mazumdar M, Memtsoudis SG.
+Beyond repeated-measures analysis of variance.
+\emph{Regional Anesthesia and Pain Medicine}.
+2012;37(1):99-105.
+doi:\href{https://doi.org/10.1097/aap.0b013e31823ebc74}{10.1097/aap.0b013e31823ebc74}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-scheffer2002}{}}%
-\CSLLeftMargin{46. }
+\CSLLeftMargin{47. }
 \CSLRightInline{Scheffer J. Dealing with missing data. \emph{Research Letters in the Information and Mathematical Sciences}. 2002;3:153-160.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-potthoff2006}{}}%
-\CSLLeftMargin{47. }
+\CSLLeftMargin{48. }
 \CSLRightInline{Potthoff RF, Tudor GE, Pieper KS, Hasselblad V.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ma2012}{}}%
-\CSLLeftMargin{48. }
-\CSLRightInline{Ma Y, Mazumdar M, Memtsoudis SG.
-Beyond repeated-measures analysis of variance.
-\emph{Regional Anesthesia and Pain Medicine}.
-2012;37(1):99-105.
-doi:\href{https://doi.org/10.1097/aap.0b013e31823ebc74}{10.1097/aap.0b013e31823ebc74}}
+\leavevmode\vadjust pre{\hypertarget{ref-box1976}{}}%
+\CSLLeftMargin{49. }
+\CSLRightInline{Box GEP. Science and statistics. 1976;71(356):791-799. doi:\href{https://doi.org/10.1080/01621459.1976.10480949}{10.1080/01621459.1976.10480949}}
 
 \leavevmode\vadjust pre{\hypertarget{ref-nlme}{}}%
-\CSLLeftMargin{49. }
+\CSLLeftMargin{50. }
 \CSLRightInline{Pinheiro J, Bates D, DebRoy S, Sarkar D, R Core Team. \emph{{nlme}: Linear and Nonlinear Mixed Effects Models}.; 2020. \url{https://CRAN.R-project.org/package=nlme}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nelder1972}{}}%
-\CSLLeftMargin{50. }
+\CSLLeftMargin{51. }
 \CSLRightInline{Nelder JA, Wedderburn RWM.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hastie1987}{}}%
-\CSLLeftMargin{51. }
+\CSLLeftMargin{52. }
 \CSLRightInline{Hastie T, Tibshirani R.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hefley2017}{}}%
-\CSLLeftMargin{52. }
+\CSLLeftMargin{53. }
 \CSLRightInline{Hefley TJ, Broms KM, Brost BM, et al.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wegman1983}{}}%
-\CSLLeftMargin{53. }
+\CSLLeftMargin{54. }
 \CSLRightInline{Wegman EJ, Wright IW.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wood2003}{}}%
-\CSLLeftMargin{54. }
+\CSLLeftMargin{55. }
 \CSLRightInline{Wood SN. Thin plate regression splines. 2003;65(1):95-114. doi:\href{https://doi.org/10.1111/1467-9868.00374}{10.1111/1467-9868.00374}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-mcelreath2018}{}}%
-\CSLLeftMargin{55. }
+\CSLLeftMargin{56. }
 \CSLRightInline{McElreath R. \emph{Statistical Rethinking}. Chapman; Hall/{CRC}; 2018. doi:\href{https://doi.org/10.1201/9781315372495}{10.1201/9781315372495}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-miller2019}{}}%
-\CSLLeftMargin{56. }
+\CSLLeftMargin{57. }
 \CSLRightInline{Miller DL. Bayesian views of generalized additive modelling. \emph{arXiv preprint arXiv:190201330}. Published online 2019.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-laird1982}{}}%
-\CSLLeftMargin{57. }
+\CSLLeftMargin{58. }
 \CSLRightInline{Laird NM, Ware JH. Random-effects models for longitudinal data. \emph{Biometrics}. 1982;38(4):963. doi:\href{https://doi.org/10.2307/2529876}{10.2307/2529876}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
-\CSLLeftMargin{58. }
+\CSLLeftMargin{59. }
 \CSLRightInline{Marra G, Wood SN.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
-\CSLLeftMargin{59. }
+\CSLLeftMargin{60. }
 \CSLRightInline{Simpson GL. \emph{Gratia: Graceful 'Ggplot'-Based Graphics and Other Functions for GAMs Fitted Using 'Mgcv'}.; 2020. \url{https://CRAN.R-project.org/package=gratia}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
-\CSLLeftMargin{60. }
+\CSLLeftMargin{61. }
 \CSLRightInline{Harezlak J, Ruppert D, Wand MP. \emph{Semiparametric Regression with r}. Springer New York; 2018. doi:\href{https://doi.org/10.1007/978-1-4939-8853-2}{10.1007/978-1-4939-8853-2}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
-\CSLLeftMargin{61. }
+\CSLLeftMargin{62. }
 \CSLRightInline{Lang TA, Altman DG.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hastie1995}{}}%
-\CSLLeftMargin{62. }
+\CSLLeftMargin{63. }
 \CSLRightInline{Hastie T, Tibshirani R.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-begley2015}{}}%
-\CSLLeftMargin{63. }
+\CSLLeftMargin{64. }
 \CSLRightInline{Begley CG, Ioannidis JPA.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-weissgerber2018}{}}%
-\CSLLeftMargin{64. }
+\CSLLeftMargin{65. }
 \CSLRightInline{Weissgerber TL, Garcia-Valencia O, Garovic VD, Milic NM, Winham SJ.

### 2021-10-20 Manuscripts/Manuscript by chapters- SIM Revisions/02-Challenges.Rmd
Need to update the GLMM part (add to LMEM part?)

@@
 
-## Missing observations
+## Unbalanced data
 
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004].
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].
+**In a longitudinal study, it is frequently the case that the number of observations is different across all subjects.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury in animals in preclinical research) leading to what is known as "missing", "incomplete" or (more properly speaking) unbalanced data [@molenberghs2004].
+The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].**
+
+**On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are _missing at random_ (MAR) or _completely missing at random_ (MCAR) [@west2014; @weiss2005].
 In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002].
 If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
+For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.
+**
 
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014].
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.
+**However, it is worth reminding that essentially "all models are wrong" [@box1976] and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
+Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.**
 
@@
     theme_classic() +
-    theme(plot.title = element_text(size = 30, 
+    theme(plot.title = element_text(size = 20, 
                                   face = "bold"),
-        text=element_text(size=30))+
+        text=element_text(size=20))+
     thm
@@
 
-txt<-18
 A<-plot_example(example(fun_type = "linear", error_type = "correlated"))
@@
 
-```{r, l-q-response, fig.width=12, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='(ref:l-q-response-caption)'}
+```{r, l-q-response, fig.width=12, fig.height=7, out.width='100%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='(ref:l-q-response-caption)'}
 # linear response, correlated errors (subject effect)

### 2021-10-22 Manuscripts/Manuscript by chapters- SIM Revisions/00-Full_document_SIM_No_Appendix.tex
Made changes based on reviewers' comments.

@@
 
-In a longitudinal study, it is frequently the case that the number of observations is different across all subjects.
+\textbf{In a longitudinal study, it is frequently the case that the number of observations is different across all subjects.
 In biomedical research, this situation can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury in animals in preclinical research) leading to what is known as ``missing,'' ``incomplete'' or (more properly speaking) unbalanced data.\textsuperscript{\protect\hyperlink{ref-molenberghs2004}{45}} The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size.\textsuperscript{\protect\hyperlink{ref-ma2012}{46}}
-
-On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are \emph{missing at random} (MAR) or \emph{completely missing at random} (MCAR).\textsuperscript{\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}} In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest.\textsuperscript{\protect\hyperlink{ref-scheffer2002}{47}} If the data are MCAR, this means that the missingness is completely unrelated to the collected information,\textsuperscript{\protect\hyperlink{ref-potthoff2006}{48}} and in the case of MNAR the missing values are dependent on their value.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size.\textsuperscript{\protect\hyperlink{ref-ma2012}{46}}}
+
+\textbf{On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are \emph{missing at random} (MAR) or \emph{completely missing at random} (MCAR).\textsuperscript{\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}} In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest.\textsuperscript{\protect\hyperlink{ref-scheffer2002}{47}} If the data are MCAR, this means that the missingness is completely unrelated to the collected information,\textsuperscript{\protect\hyperlink{ref-potthoff2006}{48}} and in the case of MNAR the missing values are dependent on their value.
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
-
-However, it is worth reminding that essentially ``all models are wrong''\textsuperscript{\protect\hyperlink{ref-box1976}{49}} and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
-Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.
+}
+
+\textbf{However, it is worth reminding that essentially ``all models are wrong''\textsuperscript{\protect\hyperlink{ref-box1976}{49}} and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
+Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.}
 
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{56}} In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case).\textsuperscript{\protect\hyperlink{ref-miller2019}{57}} Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model.\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{58}} Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{57}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{59}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
+\textbf{Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{56}} In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than extremely ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using \{mgcv\} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case, which can be fitted using JAGS or Stan).\textsuperscript{\protect\hyperlink{ref-miller2019}{57}} The use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model.\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{58}} Therefore, the confidence intervals (CIs) calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (point-wise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}}
+
+\textbf{To understand this last part, it is worth reminding that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function described by the model exists, taking into account the randomness introduced by sampling.
+If random samples of the data are taken repeatedly, a model and CI could be obtained in each case.
+In a point-wise CI, the ``true'' function can exist outside of the CI in certain regions (because the model does not follow a fixed pattern like a linear function).
+This means that if repeated samples are taken and a model and point-wise CI is obtained in each case, the total number of CIs that contain the true function \emph{entirely} is much less than the nominal value of the CI (95\%, usually).
+}
+
+\textbf{Contrary to this, an \emph{across the function} CI (like those estimated for GAMs using \{mgcv\}) contains the true function through the entire range of the covariates (which would be time in the case of longitudinal data).
+Interestingly, for GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, \emph{on average} the percentage of CIs containing the true function would be close to the nominal value (if 100 random samples are obtained it would be expected that 95 of the fitted CIs contain the true function), thereby allowing more robust estimates from the model.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{57}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{59}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.}
 

### 2021-10-22 Manuscripts/Manuscript by chapters- SIM Revisions/03-GAM Theory.Rmd
Fix comments appearance in code
Made changes based on reviewers' comments.

@@
 In contrast, Wood [@wood2003] provided a computationally efficient implementation of thin plate regression splines(TPRS), which do not require the knot placement, and require simply a choice on the number of basis to be used to create the smoother (for mathematical details see refs.
-\normalfont{[@wood2003]} and [@wood2017]).
+[@wood2003] and [@wood2017]).
 
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief [@mcelreath2018].
-In the case of non-linear data, the belief that the _true_ trend of the data is likely to be smooth rather than "wiggly" introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs [@wood2017].
-GAMs are considered "empirical" Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the "Full Bayes" case) [@miller2019].
-Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model [@laird1982;@pedersen2019].
-Therefore, the confidence intervals calculated for the smooth terms using the package _mgcv_ are considered empirical Bayesian posterior credible intervals [@pedersen2019], which have good "frequentist" coverage (pointwise coverage or "single point" coverage), and _across the function_ coverage [@wood2017].
-This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain _on average_ the true function of the data 95% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
+**Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief [@mcelreath2018].
+In the case of non-linear data, the belief that the _true_ trend of the data is likely to be smooth rather than extremely "wiggly" introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs [@wood2017].
+Moreover, GAMs are considered "empirical" Bayesian models when fitted using {mgcv} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the "Full Bayes" case, which can be fitted using JAGS or Stan) [@miller2019].
+The use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model [@laird1982;@pedersen2019].
+Therefore, the confidence intervals (CIs) calculated for the smooth terms using the package _mgcv_ are considered empirical Bayesian posterior credible intervals [@pedersen2019], which have good "frequentist" coverage (point-wise coverage or "single point" coverage), and _across the function_ coverage [@wood2017].**
+
+**To understand this last part, it is worth reminding that a CI provides an estimate of the region where the "true" or "mean" value of a function described by the model exists, taking into account the randomness introduced by sampling.
+If random samples of the data are taken repeatedly, a model and CI could be obtained in each case.
+In a point-wise CI, the "true " function can exist outside of the CI in certain regions (because the model does not follow a fixed pattern like a linear function).
+This means that if repeated samples are taken and a model and point-wise CI is obtained in each case, the total number of CIs that contain the true function _entirely_ is much less than the nominal value of the CI (95%, usually).
+**
+
+**Contrary to this, an _across the function_ CI (like those estimated for GAMs using {mgcv}) contains the true function through the entire range of the covariates (which would be time in the case of longitudinal data).
+Interestingly, for GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, _on average_ the percentage of CIs containing the true function would be close to the nominal value (if 100 random samples are obtained it would be expected that 95 of the fitted CIs contain the true function), thereby allowing more robust estimates from the model.
 In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in [@miller2019;@wood2017;@simpson2018] and [@marra2012].
-With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as "empirical Bayesian" through the rest of this paper.
+With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as "empirical Bayesian" through the rest of this paper.**

### 2021-11-03 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.Rmd
-Working on reviewers' comments -Reorganized folders

---
title: |
    | **Generalized additive models to analyze biomedical non-linear longitudinal data in R:**
    | Beyond repeated measures ANOVA and Linear Mixed Models
    | 
    | Response to reviewers
header-includes:
    \usepackage{placeins}
output:  
  bookdown::pdf_document2:
    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
    pandoc_args: --listings #calls the listings package to fit code within the page margins
    keep_tex: yes #keep LaTeX file for submission
    fig_caption: yes #allows captions in figures
    toc: false #do not include table of contents
    extra_dependencies:
      subfig: null #allows for subfigures
      breqn: null #line breaks for long equations
      caption: ["font={small}"] #size of the figure captions
      #float: null #allows for control of placement of figures
    includes:
      in_header: latex_docs/preamble_Appendix.sty #additional LaTeX formatting
  bookdown::word_document2:
    fig_caption: yes #figure caption
    keep_md: yes
  bookdown::html_document2:
    css: "style.css" #style for the HTML document
csl: elsevier-with-titles.csl #style for references
bibliography: refs.bib #references
link-citations: yes #adds links to the citations
'': default
---

# General Comments

# Comments from reviewers
## Missing data

The title for Section 3.4 has been changed from "Missing observations" to "Unbalanced data". We have restructured this section in order to:

- Avoid confusion by using the term "missing data". We now refer to different number of observations as "unbalanced data" (L201-204).

- Make clear that LMEMs also can work with missing observations (L209)

- Emphasize that GAMs are not immune to missing data problems (L217-220)

## Unnecesary restriction of LMMs

## GLMMs

## GLMs, GAMs, and conditional distributions

## Thin plate regression splines

The following changes have been made:

- L297-300 no longer refer to "knots" when referring to the construction of the smooth, now indicating that the number of _basis functions_ is what is specified when working with thin plate regression splines (TPRS), in the updated manuscript this appears in L310-315.

## Penalised splines

## Bayesian

## Coverage of confidence intervals

## Differences in smooths

## Appendix

### 2021-11-03 Manuscripts/Manuscript_by_chapters-SIM_Revisions/SIM_Appendix.Rmd
-Working on reviewers' comments -Reorganized folders

--- 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/SIM_Appendix.Rmd
@@
     includes:
-      in_header: preamble_Appendix.sty #additional LaTeX formatting
+      in_header: latex_docs/preamble_Appendix.sty #additional LaTeX formatting
   bookdown::word_document2:
@@
     css: "style.css" #style for the HTML document
-csl: elsevier-with-titles.csl #style for references
-bibliography: refs.bib #references
+csl: bibliography/elsevier-with-titles.csl #style for references
+bibliography: bibliography/refs.bib #references
 link-citations: yes #adds links to the citations
@@
 
-```{r child='01-Background.Rmd'}
+```{r child='sections/01-Background.Rmd'}
 ```
@@
 
-```{r child='02-Challenges.Rmd'}
+```{r child='sections/02-Challenges.Rmd'}
 ```
@@
 
-```{r child='03-GAM Theory.Rmd'}
+```{r child='sections/03-GAM Theory.Rmd'}
 ```
@@
 
-```{r child='04-Longitudinal analysis with GAMs.Rmd'}
+```{r child='sections/04-Longitudinal analysis with GAMs.Rmd'}
 ```
@@
 
-```{r child='05-Discussion.Rmd'}
+```{r child='sections/05-Discussion.Rmd'}
 ```
@@
 
-```{r child='06-Conclusion.Rmd'}
+```{r child='sections/06-Conclusion.Rmd'}
 ```
@@
 
-```{r child='07-Appendix.Rmd'}
+```{r child='sections/07-Appendix.Rmd'}
 

### 2021-11-03 Manuscripts/Manuscript_by_chapters-SIM_Revisions/SIM_Appendix.tex
-Working on reviewers' comments -Reorganized folders

--- 2021-10-17 Manuscripts/Manuscript by chapters- SIM Revisions/SIM_Appendix.tex
@@
 \newbox{\myorcidaffilbox}
-\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}
+\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{latex_docs/orcid}}
 
@@
     sensitive=true,
-    commentstyle=\color{sred},
     % Border (above and below)
@@
 
-\hypertarget{missing-observations}{%
-\subsection{Missing observations}\label{missing-observations}}
-
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}\protect\hyperlink{ref-weiss2005}{42}{]}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{46}{]}.
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{47}{]}, and in the case of MNAR the missing values are dependent on their value.
-
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{48}{]}.
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}\protect\hyperlink{ref-west2014}{40}{]}.
+\hypertarget{unbalanced-data}{%
+\subsection{Unbalanced data}\label{unbalanced-data}}
+
+\textbf{In a longitudinal study, it is frequently the case that the number of observations is different across all subjects.
+In biomedical research, this situation can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury in animals in preclinical research) leading to what is known as ``missing,'' ``incomplete'' or (more properly speaking) unbalanced data {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
+The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{46}{]}.}
+
+\textbf{On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are \emph{missing at random} (MAR) or \emph{completely missing at random} (MCAR) {[}\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}{]}.
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{47}{]}.
+If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{48}{]}, and in the case of MNAR the missing values are dependent on their value.
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
+}
+
+\textbf{However, it is worth reminding that essentially ``all models are wrong'' {[}\protect\hyperlink{ref-box1976}{49}{]} and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
+Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.}
 
@@
 To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{49}{]}.
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{50}{]}.
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/l-q-response-1}
+{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/l-q-response-1}
 
@@
 
-Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors {[}\protect\hyperlink{ref-nelder1972}{50}{]}.
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{51}{]}.
+Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors {[}\protect\hyperlink{ref-nelder1972}{51}{]}.
+In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{52}{]}.
 A GAM model can be written as:
@@
 For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{52}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{53}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
 In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{54}{]}.
 Although different types of splines exist, cubic and thin plate regression splines will be briefly discussed next to give a general idea of the development within the GAM framework.
@@
 Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework {[}\protect\hyperlink{ref-hastie1987}{51}{]}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit {[}\protect\hyperlink{ref-wood2003}{54}{]}.
-In contrast, Wood {[}\protect\hyperlink{ref-wood2003}{54}{]} provided a computationally efficient implementation of thin plate regression splines(TPRS), which do not require the knot placement, and require simply a choice on the number of basis to be used to create the smoother (for mathematical details see refs.
-{[}\protect\hyperlink{ref-wood2003}{54}{]} and {[}\protect\hyperlink{ref-wood2017}{37}{]}).
+The use of CS as smoothers in GAMs was discussed within the original GAM framework {[}\protect\hyperlink{ref-hastie1987}{52}{]}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit {[}\protect\hyperlink{ref-wood2003}{55}{]}.
+In contrast, Wood {[}\protect\hyperlink{ref-wood2003}{55}{]} provided a computationally efficient implementation of thin plate regression splines(TPRS), which do not require the knot placement, and require simply a choice on the number of basis to be used to create the smoother (for mathematical details see refs.
+{[}\protect\hyperlink{ref-wood2003}{55}{]} and {[}\protect\hyperlink{ref-wood2017}{37}{]}).
 
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}\protect\hyperlink{ref-mcelreath2018}{55}{]}.
-In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) {[}\protect\hyperlink{ref-miller2019}{56}{]}.
-Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{57}{]}.
-Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{56}{]} and {[}\protect\hyperlink{ref-marra2012}{58}{]}.
-With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
+\textbf{Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}\protect\hyperlink{ref-mcelreath2018}{56}{]}.
+In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than extremely ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using \{mgcv\} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case, which can be fitted using JAGS or Stan) {[}\protect\hyperlink{ref-miller2019}{57}{]}.
+The use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{58}{]}.
+Therefore, the confidence intervals (CIs) calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (point-wise coverage or ``single point'' coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.}
+
+\textbf{To understand this last part, it is worth reminding that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function described by the model exists, taking into account the randomness introduced by sampling.
+If random samples of the data are taken repeatedly, a model and CI could be obtained in each case.
+In a point-wise CI, the ``true'' function can exist outside of the CI in certain regions (because the model does not follow a fixed pattern like a linear function).
+This means that if repeated samples are taken and a model and point-wise CI is obtained in each case, the total number of CIs that contain the true function \emph{entirely} is much less than the nominal value of the CI (95\%, usually).
+}
+
+\textbf{Contrary to this, an \emph{across the function} CI (like those estimated for GAMs using \{mgcv\}) contains the true function through the entire range of the covariates (which would be time in the case of longitudinal data).
+Interestingly, for GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, \emph{on average} the percentage of CIs containing the true function would be close to the nominal value (if 100 random samples are obtained it would be expected that 95 of the fitted CIs contain the true function), thereby allowing more robust estimates from the model.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{57}{]} and {[}\protect\hyperlink{ref-marra2012}{59}{]}.
+With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.}
 
@@
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in {[}\protect\hyperlink{ref-hefley2017}{52}{]}.
+A detailed description on basis functions and correlations can be found in {[}\protect\hyperlink{ref-hefley2017}{53}{]}.
 
@@
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{59}{]}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{60}{]}.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{60}{]}.
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{61}{]}.
 
@@
 This ``model misspecification'' error, also is known as a ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{61}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{62}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
 
 In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{62}{]}.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{53}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{63}{]}.
 This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{63},\protect\hyperlink{ref-weissgerber2018}{64}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{64},\protect\hyperlink{ref-weissgerber2018}{65}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-ma2012}{}}%
+\CSLLeftMargin{{[}46{]} }
+\CSLRightInline{Y.
+Ma, M.
+Mazumdar, S.G.
+Memtsoudis, Beyond repeated-measures analysis of variance, Regional Anesthesia and Pain Medicine.
+37 (2012) 99--105.
+\url{https://doi.org/10.1097/aap.0b013e31823ebc74}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-scheffer2002}{}}%
-\CSLLeftMargin{{[}46{]} }
+\CSLLeftMargin{{[}47{]} }
 \CSLRightInline{J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-potthoff2006}{}}%
-\CSLLeftMargin{{[}47{]} }
+\CSLLeftMargin{{[}48{]} }
 \CSLRightInline{R.F.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ma2012}{}}%
-\CSLLeftMargin{{[}48{]} }
-\CSLRightInline{Y.
-Ma, M.
-Mazumdar, S.G.
-Memtsoudis, Beyond repeated-measures analysis of variance, Regional Anesthesia and Pain Medicine.
-37 (2012) 99--105.
-\url{https://doi.org/10.1097/aap.0b013e31823ebc74}.}
+\leavevmode\vadjust pre{\hypertarget{ref-box1976}{}}%
+\CSLLeftMargin{{[}49{]} }
+\CSLRightInline{G.E.P. Box, Science and statistics, 71 (1976) 791--799. \url{https://doi.org/10.1080/01621459.1976.10480949}.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-nlme}{}}%
-\CSLLeftMargin{{[}49{]} }
+\CSLLeftMargin{{[}50{]} }
 \CSLRightInline{J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, {nlme}: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nelder1972}{}}%
-\CSLLeftMargin{{[}50{]} }
+\CSLLeftMargin{{[}51{]} }
 \CSLRightInline{J.A. Nelder, R.W.M. Wedderburn, Generalized linear models, Journal of the Royal Statistical Society. Series A (General). 135 (1972) 370. \url{https://doi.org/10.2307/2344614}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hastie1987}{}}%
-\CSLLeftMargin{{[}51{]} }
+\CSLLeftMargin{{[}52{]} }
 \CSLRightInline{T.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hefley2017}{}}%
-\CSLLeftMargin{{[}52{]} }
+\CSLLeftMargin{{[}53{]} }
 \CSLRightInline{T.J.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wegman1983}{}}%
-\CSLLeftMargin{{[}53{]} }
+\CSLLeftMargin{{[}54{]} }
 \CSLRightInline{E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wood2003}{}}%
-\CSLLeftMargin{{[}54{]} }
+\CSLLeftMargin{{[}55{]} }
 \CSLRightInline{S.N. Wood, Thin plate regression splines, 65 (2003) 95--114. \url{https://doi.org/10.1111/1467-9868.00374}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-mcelreath2018}{}}%
-\CSLLeftMargin{{[}55{]} }
+\CSLLeftMargin{{[}56{]} }
 \CSLRightInline{R. McElreath, Statistical rethinking, Chapman; Hall/{CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-miller2019}{}}%
-\CSLLeftMargin{{[}56{]} }
+\CSLLeftMargin{{[}57{]} }
 \CSLRightInline{D.L. Miller, Bayesian views of generalized additive modelling, arXiv Preprint arXiv:1902.01330. (2019).}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-laird1982}{}}%
-\CSLLeftMargin{{[}57{]} }
+\CSLLeftMargin{{[}58{]} }
 \CSLRightInline{N.M. Laird, J.H. Ware, Random-effects models for longitudinal data, Biometrics. 38 (1982) 963. \url{https://doi.org/10.2307/2529876}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
-\CSLLeftMargin{{[}58{]} }
+\CSLLeftMargin{{[}59{]} }
 \CSLRightInline{G.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
-\CSLLeftMargin{{[}59{]} }
+\CSLLeftMargin{{[}60{]} }
 \CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
-\CSLLeftMargin{{[}60{]} }
+\CSLLeftMargin{{[}61{]} }
 \CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with r, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
-\CSLLeftMargin{{[}61{]} }
+\CSLLeftMargin{{[}62{]} }
 \CSLRightInline{T.A.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hastie1995}{}}%
-\CSLLeftMargin{{[}62{]} }
+\CSLLeftMargin{{[}63{]} }
 \CSLRightInline{T.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-begley2015}{}}%
-\CSLLeftMargin{{[}63{]} }
+\CSLLeftMargin{{[}64{]} }
 \CSLRightInline{C.G. Begley, J.P.A. Ioannidis, Reproducibility in science, Circulation Research. 116 (2015) 116--126. \url{https://doi.org/10.1161/circresaha.114.303819}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-weissgerber2018}{}}%
-\CSLLeftMargin{{[}64{]} }
+\CSLLeftMargin{{[}65{]} }
 \CSLRightInline{T.L.

### 2021-11-03 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
-Working on reviewers' comments -Reorganized folders

--- 2021-10-20 Manuscripts/Manuscript by chapters- SIM Revisions/02-Challenges.Rmd
@@
 
-**In a longitudinal study, it is frequently the case that the number of observations is different across all subjects.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury in animals in preclinical research) leading to what is known as "missing", "incomplete" or (more properly speaking) unbalanced data [@molenberghs2004].
+In a longitudinal study, it is frequently the case that the number of observations is different across subjects.
+In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as "missing", "incomplete" or (more generally speaking) unbalanced data [@molenberghs2004].
 The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].**
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
 
-**On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are _missing at random_ (MAR) or _completely missing at random_ (MCAR) [@west2014; @weiss2005].
+On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are _missing at random_ (MAR) or _completely missing at random_ (MCAR) [@west2014; @weiss2005].
 In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002].
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.
+If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006].
+Missing observations can also be _missing not at random_ (MNAR) and in the case the missing observations are dependent on their value.
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
-**
 
-**However, it is worth reminding that essentially "all models are wrong" [@box1976] and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
-Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.**
+However, it is worth reminding that "all models are wrong" [@box1976] and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
+Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.
 

### 2021-11-03 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
-Working on reviewers' comments -Reorganized folders

--- 2021-10-22 Manuscripts/Manuscript by chapters- SIM Revisions/03-GAM Theory.Rmd
@@
 
-**To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \@ref(fig:l-q-response)C.
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \@ref(fig:l-q-response)C.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
 A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
-Because the data has 6 unique time points, we can specify a maximum of six basis functions to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).**
+Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).
 

### 2021-11-03 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/05-Discussion.Rmd
-Working on reviewers' comments -Reorganized folders

--- 2021-06-07 Manuscripts/Manuscript by chapters -v2/05-Discussion.Rmd
@@
 However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature [@vishwanath2009] using $\textsf{R}$ and the package _mgcv_[@wood2017] in Section \@ref(longitudinal-GAMs), while a basic workflow for model selection is in the Appendix.
-One of the features of GAMs is that they go beyond a mere _p-value_ to indicate differences between groups, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure  \@ref(fig:plot-pairwise-comp) indicate.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature [@vishwanath2009] using $\textsf{R}$ and the package _mgcv_[@wood2017] in Section \@ref(longitudinal-GAMs), while a basic workflow for model selection is in the [Appendix](#workflow).
+One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a _p-value_, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure  \@ref(fig:plot-pairwise-comp) indicate.
 The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( $\approx$ day 3 in Figure \@ref(fig:plot-pairwise-comp) A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.

### 2021-11-04 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.Rmd
Updated manuscript and response to reviewers

@@
     css: "style.css" #style for the HTML document
-csl: elsevier-with-titles.csl #style for references
-bibliography: refs.bib #references
+#csl: elsevier-with-titles.csl #style for references
+#bibliography: refs.bib #references
 link-citations: yes #adds links to the citations
@@
 
-- L297-300 no longer refer to "knots" when referring to the construction of the smooth, now indicating that the number of _basis functions_ is what is specified when working with thin plate regression splines (TPRS), in the updated manuscript this appears in L310-315.
+- The information on L289-294 pertaining splines has been changed.
+L296-307 now give a better description of cubic splines (CS), indicating their limitation due to knot placement.
+The lack of knot placement for thin plate regression splines (TPRS) is indicated as an advantage over CS.
+
+- L297-300 no longer refer to "knots" when referring to the construction of the smooth, now indicating that the number of _basis functions_ is what is specified when working with TPRS.
+In the updated manuscript this appears in L310-315.
 
@@
 
+- We have clarified the concept "empirical Bayesian", which appeared on L318-L319 in the original manuscript.
+In the revised manuscript, L335-336 now indicate that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
+
+- The sentence "Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model", which appears in L319 in the original manuscript has been removed from the text.
+Instead, the concept of REML has been moved to Section 6.2 L390-393, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in _mgcv_.
+
 ## Coverage of confidence intervals
+
+The content in L320-329 regarding confidence intervals (CIs) in the original submission has been changed in the following manner:
+
+- L339-L352 now contain a more detailed and accurate explanation on the differences and interpretation of "pointwise" CIs and "across-the-function" CIs.
+We also reference the work of Marra and Wood if the reader desires a more in-depth exploration.
 

### 2021-11-04 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.tex
Updated manuscript and response to reviewers

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
Beyond repeated measures ANOVA and Linear Mixed Models\\
\strut \\
Response to reviewers}
\author{}
\date{\vspace{-2.5em}}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
%\usepackage{lineno}
\usepackage[affil-it,blocks]{authblk}
\usepackage{hyperref}
\usepackage{graphicx}
%\usepackage[nomarkers,figuresonly]{endfloat}
%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{latex_docs/orcid}}

%add hyperlink to the box
\newcommand{\orcidaffila}[1]{%
  \href{https://orcid.org/0000-0002-6014-4538}{\usebox{\myorcidaffilbox}}}

\newcommand{\orcidaffilb}[1]{%
  \href{https://orcid.org/0000-0002-6135-8191}{\usebox{\myorcidaffilbox}}}

%to have page numbers with letters in the Appendix:
% %solution from https://tex.stackexchange.com/questions/59572/custom-page-numbering-for-appendix
\newcommand{\appendixpagenumbering}{
  \break
  \pagenumbering{arabic}
  \renewcommand{\thepage}{\thesection-\arabic{page}}
}

%command for the package lineno
%\linenumbers

%authors
\author{Ariel I. Mundo \orcidaffila{}}
\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
\author{John R. Tipton \orcidaffilb{}}
\affil{Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
\author{Timothy J. Muldoon*}
\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
\affil{tmuldoon@uark.edu}

%theme colors for the code chunks (originally from latex-solarized on GitHub)
%https://github.com/jez/latex-solarized
\usepackage{xcolor}
\definecolor{sbase03}{HTML}{002B36}
\definecolor{sbase02}{HTML}{073642}
\definecolor{sbase01}{HTML}{586E75}
\definecolor{sbase00}{HTML}{657B83}
\definecolor{sbase0}{HTML}{839496}
\definecolor{sbase1}{HTML}{93A1A1}
\definecolor{sbase2}{HTML}{EEE8D5}
\definecolor{sbase3}{HTML}{FDF6E3}
\definecolor{syellow}{HTML}{B58900}
\definecolor{sorange}{HTML}{CB4B16}
\definecolor{sred}{HTML}{DC322F}
\definecolor{smagenta}{HTML}{D33682}
\definecolor{sviolet}{HTML}{6C71C4}
\definecolor{sblue}{HTML}{268BD2}
\definecolor{scyan}{HTML}{2AA198}
\definecolor{sgreen}{HTML}{859900}
%command to set parameter(s) in package listings
\lstset{
    % How/what to match
    sensitive=true,
    % Border (above and below)
    frame=lines,
    % Extra margin on line (align with paragraph)
    xleftmargin=\parindent,
    % Put extra space under caption
    belowcaptionskip=1\baselineskip,
    % Colors
    backgroundcolor=\color{sbase3},
    basicstyle=\color{sbase00}\ttfamily,
    keywordstyle=\color{scyan},
    commentstyle=\color{sbase1},
    stringstyle=\color{sblue},
    numberstyle=\color{sviolet},
    identifierstyle=\color{sbase00},
    % Break long lines into multiple lines?
    breaklines=true,
    % Show a character for spaces?
    showstringspaces=false,
    tabsize=2
}

%\lstset{
%  breaklines=true,
%  stringstyle=\ttfamily,
%  backgroundcolor=\color{gray}
%}
\usepackage{placeins}
\usepackage{subfig}
\usepackage{breqn}
\usepackage[font={small}]{caption}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\begin{document}
\maketitle

\hypertarget{general-comments}{%
\section{General Comments}\label{general-comments}}

\hypertarget{comments-from-reviewers}{%
\section{Comments from reviewers}\label{comments-from-reviewers}}

\hypertarget{missing-data}{%
\subsection{Missing data}\label{missing-data}}

The title for Section 3.4 has been changed from ``Missing observations'' to ``Unbalanced data''. We have restructured this section in order to:

\begin{itemize}
\item
  Avoid confusion by using the term ``missing data''. We now refer to different number of observations as ``unbalanced data'' (L201-204).
\item
  Make clear that LMEMs also can work with missing observations (L209)
\item
  Emphasize that GAMs are not immune to missing data problems (L217-220)
\end{itemize}

\hypertarget{unnecesary-restriction-of-lmms}{%
\subsection{Unnecesary restriction of LMMs}\label{unnecesary-restriction-of-lmms}}

\hypertarget{glmms}{%
\subsection{GLMMs}\label{glmms}}

\hypertarget{glms-gams-and-conditional-distributions}{%
\subsection{GLMs, GAMs, and conditional distributions}\label{glms-gams-and-conditional-distributions}}

\hypertarget{thin-plate-regression-splines}{%
\subsection{Thin plate regression splines}\label{thin-plate-regression-splines}}

The following changes have been made:

\begin{itemize}
\item
  The information on L289-294 pertaining splines has been changed. L296-307 now give a better description of cubic splines (CS), indicating their limitation due to knot placement. The lack of knot placement for thin plate regression splines (TPRS) is indicated as an advantage over CS.
\item
  L297-300 no longer refer to ``knots'' when referring to the construction of the smooth, now indicating that the number of \emph{basis functions} is what is specified when working with TPRS. In the updated manuscript this appears in L310-315.
\end{itemize}

\hypertarget{penalised-splines}{%
\subsection{Penalised splines}\label{penalised-splines}}

\hypertarget{bayesian}{%
\subsection{Bayesian}\label{bayesian}}

\begin{itemize}
\item
  We have clarified the concept ``empirical Bayesian'', which appeared on L318-L319 in the original manuscript. In the revised manuscript, L335-336 now indicate that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
\item
  The sentence ``Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model'', which appears in L319 in the original manuscript has been removed from the text. Instead, the concept of REML has been moved to Section 6.2 L390-393, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in \emph{mgcv}.
\end{itemize}

\hypertarget{coverage-of-confidence-intervals}{%
\subsection{Coverage of confidence intervals}\label{coverage-of-confidence-intervals}}

The content in L320-329 regarding confidence intervals (CIs) in the original submission has been changed in the following manner:

\begin{itemize}
\tightlist
\item
  L339-L352 now contain a more detailed and accurate explanation on the differences and interpretation of ``pointwise'' CIs and ``across-the-function'' CIs. We also reference the work of Marra and Wood if the reader desires a more in-depth exploration.
\end{itemize}

\hypertarget{differences-in-smooths}{%
\subsection{Differences in smooths}\label{differences-in-smooths}}

\hypertarget{appendix}{%
\subsection{Appendix}\label{appendix}}

\end{document}

### 2021-11-04 Manuscripts/Manuscript_by_chapters-SIM_Revisions/00-Full_document_SIM_No_Appendix.tex
-Working on reviewers' comments -Reorganized folders
Updated manuscript and response to reviewers

--- 2021-10-20 Manuscripts/Manuscript by chapters- SIM Revisions/00-Full_document_SIM_No_Appendix.tex
@@
 
-In a longitudinal study, it is frequently the case that the number of observations is different across all subjects.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury in animals in preclinical research) leading to what is known as ``missing,'' ``incomplete'' or (more properly speaking) unbalanced data.\textsuperscript{\protect\hyperlink{ref-molenberghs2004}{45}} The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
+In a longitudinal study, it is frequently the case that the number of observations is different across subjects.
+In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing,'' ``incomplete'' or (more generally speaking) unbalanced data.\textsuperscript{\protect\hyperlink{ref-molenberghs2004}{45}} The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
 This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size.\textsuperscript{\protect\hyperlink{ref-ma2012}{46}}
 
-On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are \emph{missing at random} (MAR) or \emph{completely missing at random} (MCAR).\textsuperscript{\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}} In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest.\textsuperscript{\protect\hyperlink{ref-scheffer2002}{47}} If the data are MCAR, this means that the missingness is completely unrelated to the collected information,\textsuperscript{\protect\hyperlink{ref-potthoff2006}{48}} and in the case of MNAR the missing values are dependent on their value.
+On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are \emph{missing at random} (MAR) or \emph{completely missing at random} (MCAR).\textsuperscript{\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}} In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest.\textsuperscript{\protect\hyperlink{ref-scheffer2002}{47}} If the data are MCAR, this means that the missingness is completely unrelated to the collected information.\textsuperscript{\protect\hyperlink{ref-potthoff2006}{48}} Missing observations can also be \emph{missing not at random} (MNAR) and in the case the missing observations are dependent on their value.
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
 
-However, it is worth reminding that essentially ``all models are wrong''\textsuperscript{\protect\hyperlink{ref-box1976}{49}} and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
+However, it is worth reminding that ``all models are wrong''\textsuperscript{\protect\hyperlink{ref-box1976}{49}} and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
 Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.
@@
 
-\textbf{To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
 A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
-Because the data has 6 unique time points, we can specify a maximum of six basis functions to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).}
+Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).
 
@@
 Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is \textbf{negative} and less than 1).
-On the other hand, the third basis function has roughly doubled its value.
+Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and less than 1).
+On the other hand, the third basis function has roughly doubled its contribution to the smooth term.
 Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{56}} In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case).\textsuperscript{\protect\hyperlink{ref-miller2019}{57}} Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model.\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{58}} Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{57}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{59}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{56}} In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than extremely ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{57}} Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (point-wise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
+
+To understand this last part, it is worth reminding that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function exists, taking into account the randomness introduced by the sampling process.
+Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie.
+In a ``point-wise'' CI for non-linear data, the ``true'' function will lie outside of the CI in certain regions at a given frequency (because the data does not follow a linear trend).
+This means that if a point-wise CI is obtained for 100 random samples, the total number of CIs that contain the true function \emph{entirely} is much less than the nominal value of the CI (95\%, usually).
+
+Contrary to this, an \emph{across the function} CI (like those estimated for GAMs using \emph{mgcv}) contains the true function through the entire range of the covariates (which would be time in the case of longitudinal data).
+For GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, the percentage of CIs entirely containing the true function over the entire time period would be close to the nominal value (i.e., if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function), thereby allowing more robust estimates from the model.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{57}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{58}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
 
@@
 
-This syntax specifies that \passthrough{\lstinline!gam\_02!} (named accordingly so it matches the model sequence from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
-The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{60}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{61}}
+This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 basis functions.
+The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups.
+Although the default \passthrough{\lstinline!method!} used to estimate the smoothing parameters in \emph{mgcv} is generalized cross validation (GCV), Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} showed the restricted maximum likelihood (REML) to be more resistant to overfitting while also easing the quantification of uncertainty in the smooth parameters; therefore in this manuscript REML is always used for smooth parameter estimation.
+
+When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{59}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{60}}
 
@@
 This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{62}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{61}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
 
 In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{53}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{63}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{53}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{62}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
 
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{64},\protect\hyperlink{ref-weissgerber2018}{65}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{63},\protect\hyperlink{ref-weissgerber2018}{64}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-laird1982}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
 \CSLLeftMargin{58. }
-\CSLRightInline{Laird NM, Ware JH. Random-effects models for longitudinal data. \emph{Biometrics}. 1982;38(4):963. doi:\href{https://doi.org/10.2307/2529876}{10.2307/2529876}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
-\CSLLeftMargin{59. }
 \CSLRightInline{Marra G, Wood SN.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
+\CSLLeftMargin{59. }
+\CSLRightInline{Simpson GL. \emph{Gratia: Graceful 'Ggplot'-Based Graphics and Other Functions for GAMs Fitted Using 'Mgcv'}.; 2020. \url{https://CRAN.R-project.org/package=gratia}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
 \CSLLeftMargin{60. }
-\CSLRightInline{Simpson GL. \emph{Gratia: Graceful 'Ggplot'-Based Graphics and Other Functions for GAMs Fitted Using 'Mgcv'}.; 2020. \url{https://CRAN.R-project.org/package=gratia}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
+\CSLRightInline{Harezlak J, Ruppert D, Wand MP. \emph{Semiparametric Regression with r}. Springer New York; 2018. doi:\href{https://doi.org/10.1007/978-1-4939-8853-2}{10.1007/978-1-4939-8853-2}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
 \CSLLeftMargin{61. }
-\CSLRightInline{Harezlak J, Ruppert D, Wand MP. \emph{Semiparametric Regression with r}. Springer New York; 2018. doi:\href{https://doi.org/10.1007/978-1-4939-8853-2}{10.1007/978-1-4939-8853-2}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
-\CSLLeftMargin{62. }
 \CSLRightInline{Lang TA, Altman DG.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hastie1995}{}}%
-\CSLLeftMargin{63. }
+\CSLLeftMargin{62. }
 \CSLRightInline{Hastie T, Tibshirani R.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-begley2015}{}}%
-\CSLLeftMargin{64. }
+\CSLLeftMargin{63. }
 \CSLRightInline{Begley CG, Ioannidis JPA.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-weissgerber2018}{}}%
-\CSLLeftMargin{65. }
+\CSLLeftMargin{64. }
 \CSLRightInline{Weissgerber TL, Garcia-Valencia O, Garovic VD, Milic NM, Winham SJ.

### 2021-11-04 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated manuscript and response to reviewers

@@
 Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as _basis functions_, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
-Although different types of splines exist, cubic and thin plate regression splines will be briefly discussed next to give a general idea of the development within the GAM framework.
+Although different types of splines exist, cubic and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
 
@@
 The use of CS as smoothers in GAMs was discussed within the original GAM framework [@hastie1987], but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as 'knots', the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit [@wood2003].
-In contrast, Wood [@wood2003] provided a computationally efficient implementation of thin plate regression splines(TPRS), which do not require the knot placement, and require simply a choice on the number of basis to be used to create the smoother (for mathematical details see refs.
+In contrast, Wood [@wood2003] provided a computationally efficient implementation of thin plate regression splines (TPRS), which do not need knot placement.
+Instead, TPRS only require the number of basis to be used to create the smoother (for mathematical details see refs.
 [@wood2003] and [@wood2017]).
@@
 Figure \@ref(fig:basis-plot)C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is **negative** and less than 1).
-On the other hand, the third basis function has roughly doubled its value.
+Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and less than 1).
+On the other hand, the third basis function has roughly doubled its contribution to the smooth term.
 Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
@@
 
-**Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief [@mcelreath2018].
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief [@mcelreath2018].
 In the case of non-linear data, the belief that the _true_ trend of the data is likely to be smooth rather than extremely "wiggly" introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs [@wood2017].
-Moreover, GAMs are considered "empirical" Bayesian models when fitted using {mgcv} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the "Full Bayes" case, which can be fitted using JAGS or Stan) [@miller2019].
-The use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model [@laird1982;@pedersen2019].
-Therefore, the confidence intervals (CIs) calculated for the smooth terms using the package _mgcv_ are considered empirical Bayesian posterior credible intervals [@pedersen2019], which have good "frequentist" coverage (point-wise coverage or "single point" coverage), and _across the function_ coverage [@wood2017].**
+Moreover, GAMs are considered "empirical" Bayesian models when fitted using the package _mgcv_ because the smoothing parameters are estimated from the data (and not from a prior distribution as in the "fully Bayesian" case, which can be fitted using JAGS, Stan, or other probabilistic programming language) [@miller2019].
+Therefore, the confidence intervals (CIs) calculated for the smooth terms using _mgcv_ are considered empirical Bayesian posterior credible intervals [@pedersen2019], which have good "frequentist" coverage (point-wise coverage or "single point" coverage), and _across the function_ coverage [@wood2017].
 
-**To understand this last part, it is worth reminding that a CI provides an estimate of the region where the "true" or "mean" value of a function described by the model exists, taking into account the randomness introduced by sampling.
-If random samples of the data are taken repeatedly, a model and CI could be obtained in each case.
-In a point-wise CI, the "true " function can exist outside of the CI in certain regions (because the model does not follow a fixed pattern like a linear function).
-This means that if repeated samples are taken and a model and point-wise CI is obtained in each case, the total number of CIs that contain the true function _entirely_ is much less than the nominal value of the CI (95%, usually).
-**
+To understand this last part, it is worth reminding that a CI provides an estimate of the region where the “true” or “mean” value of a function exists, taking into account the randomness introduced by the sampling process.
+Because random samples from the population are used to calculate the "true" value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95%) where the function is expected to lie.
+In a "point-wise" CI for non-linear data, the "true" function will lie outside of the CI in certain regions at a given frequency (because the data does not follow a linear trend).
+This means that if a point-wise CI is obtained for 100 random samples, the total number of CIs that contain the true function _entirely_ is much less than the nominal value of the CI (95%, usually).
 
-**Contrary to this, an _across the function_ CI (like those estimated for GAMs using {mgcv}) contains the true function through the entire range of the covariates (which would be time in the case of longitudinal data).
-Interestingly, for GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, _on average_ the percentage of CIs containing the true function would be close to the nominal value (if 100 random samples are obtained it would be expected that 95 of the fitted CIs contain the true function), thereby allowing more robust estimates from the model.
+Contrary to this, an _across the function_ CI (like those estimated for GAMs using _mgcv_) contains the true function through the entire range of the covariates (which would be time in the case of longitudinal data).
+For GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, the percentage of CIs entirely containing the true function over the entire time period would be close to the nominal value (i.e., if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function), thereby allowing more robust estimates from the model.
 In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in [@miller2019;@wood2017;@simpson2018] and [@marra2012].
-With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as "empirical Bayesian" through the rest of this paper.**
+With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as "empirical Bayesian" through the rest of this paper.

### 2021-11-04 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated manuscript and response to reviewers

--- 2021-10-13 Manuscripts/Manuscript by chapters- SIM Revisions/04-Longitudinal analysis with GAMs.Rmd
@@
 
-This syntax specifies that `gam_02` (named accordingly so it matches the model sequence from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using 5 knots.
+This syntax specifies that `gam_02` (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using 5 basis functions.
 The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths [@simpson2018].
-The parametric term `Group` is added to quantify overall mean differences in the effect of treatment between groups, and the `method` chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) [@wood2017].
+The parametric term `Group` is added to quantify overall mean differences in the effect of treatment between groups.
+Although the default `method` used to estimate the smoothing parameters in _mgcv_ is generalized cross validation (GCV), Wood[@wood2017] showed the restricted maximum likelihood (REML) to be more resistant to overfitting while also easing the quantification of uncertainty in the smooth parameters; therefore in this manuscript REML is always used for smooth parameter estimation.
+
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot)B).

### 2021-11-05 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.Rmd
Updated
Updated response to reviewers

--- 2021-11-03 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.Rmd
@@
     | 
-    | Response to reviewers
+    | Journal of Submission: Statistics in Medicine
+    | Manuscript ID: SIM-21-0640
 header-includes:
@@
     toc: false #do not include table of contents
+    number_sections: false #do not number sections
     extra_dependencies:
@@
       in_header: latex_docs/preamble_Appendix.sty #additional LaTeX formatting
-  bookdown::word_document2:
-    fig_caption: yes #figure caption
-    keep_md: yes
-  bookdown::html_document2:
-    css: "style.css" #style for the HTML document
-csl: elsevier-with-titles.csl #style for references
-bibliography: refs.bib #references
+#csl: elsevier-with-titles.csl #style for references
+#bibliography: refs.bib #references
 link-citations: yes #adds links to the citations
@@
 
-# General Comments
+# General Comments to the Reviewers
+
+We would like to thank the reviewer for the careful and thorough analysis of this manuscript to _Statistics in Medicine_ and for the thoughtful comments and constructive suggestions, which helped improve the quality of this manuscript.
+We carefully considered the reviewer’s comments and in this document, explain how we revised the manuscript based on those comments and suggestions.
+
+# General Comments to the Editor
+
+Dr. Platt,
+
+The authors thank you for your determination that our manuscript may be suitable for resubmission in _Statistics in Medicine_ after addressing the reviewer's comments.
+To this end, we have addressed all critiques.
+We hope these revisions, submitted on December 5, 2021, improve the manuscript so it is deemed worthy of publication in _Statistics in Medicine_.
+Following are our detailed responses to reviewer comments.
+
+## Reviewer's Introduction
+Mundo and colleagues present a tutorial on the use of generalized additive models to analyze longitudinal data.
+A comparison with repeated measures ANOVA and linear mixed models is provided and a recurring example using simulated
+data is used to illustrate the differences among the methods and the advantages of GAM.
+While I like the general approach and aim of the manuscript, there are a number of inaccuracies and omissions, especially in the discussion of GAMs that make it difficult for me to support publication at this stage.
+I believe — given evidence elsewhere in the manuscript that the authors really do know the subject — that these inaccuracies and omissions stem from preparing a tutorial in a scientific paper format where word-length considerations come into play.
+Below I outline the main areas where I feel the GAM methodology is inaccurately presented or important topics omitted, and make suggestions to improve the manuscript.
+
+## \textcolor{reviewersblue} {Reply to Introduction}
+We appreciate Dr. Simpson's comments and critiques to our manuscript.
 
 # Comments from reviewers
-## Missing data
 
-The title for Section 3.4 has been changed from "Missing observations" to "Unbalanced data". We have restructured this section in order to:
+# Missing data
 
-- Avoid confusion by using the term "missing data". We now refer to different number of observations as "unbalanced data" (L201-204).
+## Reviewer's Comments
 
-- Make clear that LMEMs also can work with missing observations (L209)
+I found this use of “missing” data to be a little confusing.
+I understand what the authors are getting at, but it suggests that GAMs can handle this whereas rm-ANOVA and LMMs can’t.
+While I appreciate that rm-ANOVA might require
+balanced observations for the group errors, LMMs are just as able to handle “missingness” (in the sense implied by the authors) as GAMs are.
+Additionally, the “missingness” isn’t meant solely as missing in the statistical sense, but really it is due to irregular or incomplete sampling, more generally.
+It would be helpful to make the discussion of missingness about balanced data and to try to avoid terms like “missing” as that may be inferred to suggest GAMs and LMMs are immune to missing data problems (they aren’t), they
+just don’t require balance.
 
-- Emphasize that GAMs are not immune to missing data problems (L217-220)
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-## Unnecesary restriction of LMMs
+We appreciate the comments about clarity in the use of the term "missingness".
+Please note that the title for Section 3.4 has been changed from "Missing observations" to "Unbalanced data".
+We have restructured this section in the following manner:
 
-## GLMMs
+- The term "missing data" has been removed to avoid confusion. We now refer to different number of observations as "unbalanced data" (L201-204).
 
-## GLMs, GAMs, and conditional distributions
+- We have made clear that LMEMs can also work with missing observations (L209)
 
-## Thin plate regression splines
+- Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates (L217-220).
+
+# Unnecesary restriction of LMMs
+
+## Reviewer's Comments
+
+Why not use quadratic effects of time in the LMM?
+I appreciate this would make the model more complex, but trying to fit a quadratic effect with a linear model strikes me as futile and somewhat of a straw person argument to make.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We appreciate the reviewer's perspective regarding the somewhat contradictory argument of fitting a linear effects LMM to quadratic data.
+However, this is an intentional mistake because we follow the Statistical analysis logic followed by most Biomedical researchers: Fit a LMM, and if you get significant p-values that's good enough.
+We are trying to convey the point that although "significant" p-values can be obtained, visualizing the model fit is something that should be required, but that is seldomly done.
+**Incorporate in the section that a strenght of GAMs is that they can learn the function from the data, while LMMs can't.**
+
+# GLMMs
+
+# GLMs, GAMs, and conditional distributions
+
+# Thin plate regression splines
+
+## Reviewer's Comments
+
+The authors define the default basis in {mgcv}, the software being used here to illustrate fitting GAMs, as “…thin plate regression splines are an optimized version that work well with noisy data.” This is not a good description of what a thin plate regression spline (hereafter TPRS) is, and is not a good description of what the low-rank versions in {mgcv} are.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
@@
 
-- L297-300 no longer refer to "knots" when referring to the construction of the smooth, now indicating that the number of _basis functions_ is what is specified when working with thin plate regression splines (TPRS), in the updated manuscript this appears in L310-315.
+- The information on L289-294 pertaining splines has been changed.
+L301-309 now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
 
-## Penalised splines
+## Reviewer's Comments
 
-## Bayesian
+Much better, I believe, is to talk about this in terms of basis functions, of which there are five in this example.
+You can use the same terminology to refer to the CRS basis, and others.
+That in the CRS if you want _k_ basis functions you need _k_ knots (IIRC).
+CRS require you to specify the knot locations (or let the software spread them evenly through the data); TPRS don’t.
+Hence the paragraph starting on L195 is confusing and misleading.
+The basis functions are not piece-wise polynomial and there are not “region[s] where a different set of basis functions will be used”.
+Indeed, the functions you show
+operate throughout the range of the covariate.
 
-## Coverage of confidence intervals
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-## Differences in smooths
+- L297-300 no longer refer to "knots" when referring to the construction of the smooth, now indicating that the number of _basis functions_ is what is specified when constructing the smoother.
+In the updated manuscript, this change appears in L310-315, and is more clear to the reader due to the changes introduced in L301-L309.
 
-## Appendix
+# Penalised splines
+## Reviewer's Comments
+
+You talk a little about “wiggliness” and penalising the weights for each basis function towards 0 but you don’t really explain this most crucial concept of the model fitting problem and why modern GAMs are so much better than the
+GAMs developed by Hastie and Tibshirani when they first introduced GAMs to the world.
+You really need to define wiggliness beyond something that is used to avoid overfitting.
+Typically it is squared second derivative of the estimated function; so we limit the curvature of the fitted function by default.
+Technically the penalty is the penalty matrix - it is a matrix and it is fixed once we define the basis functions.
+What isn’t fixed is (are) the smoothness parameter(s) of the smooth; it is those that control how much penalty we subtract from the log-likelihood of the data given the model estimates.
+To speak of a “weak” or “strong” penalty was a little confusing for me.
+Hence fitting a GAM requires one to estimate parameters and one or more smoothness parameters for each smooth in the GAM, plus any parameters required for parametric terms.
+By glossing over these important concepts, the reader is left wondering what wiggliness is, how we measure overfit? etc.
+
+# Bayesian
+
+## Reviewer's Comments
+
+GAMs in {mgcv} are considered to be empirical Bayesian models, but in general GAMs can be fully Bayesian.
+You can fit fully Bayesian GAMs using INLA and JAGS using functions from {mgcv}, and the {brms} package allows full Bayesian GAMs to be estimated simply using Stan for example.
+This needs to be clarified.
+I’m not really sure what you mean by the sentence beginning “Moreover, the use of the restricted maximum…”.
+I think I get what you mean; by casting the wiggly parts of smooths as random effects we can estimate the fit using REML
+and standard linear mixed model software. However, you can fit the model using the full fat version of maximum likelihood and these models would still be empirical Bayes if fitted by {mgcv}.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+- We have clarified the concept "empirical Bayesian", which appeared on L318-L319 in the original manuscript.
+In the revised manuscript, L335-338 now indicate that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
+
+- The sentence "Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model", which appears in L319 in the original manuscript has been removed from the text.
+Instead, the concept of REML has been moved to Section 6.2 L392-395, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in _mgcv_.
+
+# Coverage of confidence intervals
+
+## Reviewer's Comments
+
+This entire section from L320 onward through to the end of Section 5 needs some work.
+When viewed from the Bayesian perspective, the intervals are Bayesian credible intervals.
+When viewed from a frequentist perspective, the same intervals are confidence intervals but instead of having the typical point-wise interpretation they have an across the function interpretation.
+The description of what this means is wrong — you are almost giving the incorrect definition of a confidence interval here.
+The interval either does or does not contain the true function.
+That is given.
+I’m sure this is just a slip then on L324-325.
+Also, your description implies a simultaneous interval, although I don’t think you intended this. One could interpret “95% of the time” as meaning 95% of the functions are contained in their entirety.
+
+What across-the-function means is simply that if we average the coverage of the interval over the entire function we get approximately the nominal coverage, 95% say.
+For this to occur then, some areas of the function must have more
+than nominal coverage and some areas less than the nominal coverage.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+The content in L320-329 regarding confidence intervals (CIs) in the original submission has been changed in the following manner:
+
+- L341-L355 now contain a more detailed and accurate explanation on the differences and interpretation of "poin-twise" CIs and "across-the-function" CIs.
+We removed the "95% of the time" phrase to avoid confusion, and instead we provide an explanation that focuses on random sampling: "if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function" (L352-354).
+We also reference the work of Marra and Wood if the reader desires a more in-depth exploration of across the function CIs.
+
+# Differences in smooths
+
+## Reviewer's Comments
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+# Appendix
+
+## Reviewer's Comments
+
+In the appendix I think you are needlessly restricting the size of the basis dimension to be k = 5, hence 4 basis functions per smooth when identifiability constraints are applied.
+Is there a reason I’m seeing here why you could leave this at the default k = 10 and really see the effect of the shrinkage as the EDF of the resulting smooths should be similar to the EDF you have with k = 5?
+This would also likely help with the k-index being low because there’s not a lot of shrinkage you can do when the maximum EDF possible is 4 (per smooth).
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+In L367 it is indicated that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
+The computational requisite in the smooth estimation of having the maximum number of basis equal to the number of unique values in the covariate makes it impossible to fit a GAM with a k>5, as the error "A term has fewer unique covariate combinations than specified maximum degrees of freedom" is thrown by _mgcv_.
+We consider that because the k-index for the model is 1.04 the basis dimension for the smooth is adequate (Wood says that "the further below 1 this is, th emore likely it is that there is missed pattern left in the residuals").
+
+## Reviewer's Comments
+
+Why change the model object name notation here?
+You had gam_00 previously and now you call the model m1?
+As you are already showing appraise() output for the diagnostic plots, you could use check.k() from {mgcv} to just get the basis dimension check.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+The name of the model has been changed to `gam_02`, and it appears with this notation in Section 6.2, L383 so it matches the workflow of model selection presented in the Appendix.

### 2021-11-05 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.tex
Updated response to reviewers

@@
 \strut \\
-Response to reviewers}
+Journal of Submission: Statistics in Medicine\\
+Manuscript ID: SIM-21-0640}
 \author{}
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 %\usepackage{lineno}
@@
 \usepackage{graphicx}
-%\usepackage[nomarkers,figuresonly]{endfloat}
+\usepackage[nomarkers,figuresonly]{endfloat}
 %set a box to put the ORCID logo
-\newbox{\myorcidaffilbox}
-\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{latex_docs/orcid}}
+%\newbox{\myorcidaffilbox}
+%\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{latex_docs/orcid}}
 
 %add hyperlink to the box
-\newcommand{\orcidaffila}[1]{%
-  \href{https://orcid.org/0000-0002-6014-4538}{\usebox{\myorcidaffilbox}}}
+%\newcommand{\orcidaffila}[1]{%
+%  \href{https://orcid.org/0000-0002-6014-4538}{\usebox{\myorcidaffilbox}}}
 
-\newcommand{\orcidaffilb}[1]{%
-  \href{https://orcid.org/0000-0002-6135-8191}{\usebox{\myorcidaffilbox}}}
+%\newcommand{\orcidaffilb}[1]{%
+%  \href{https://orcid.org/0000-0002-6135-8191}{\usebox{\myorcidaffilbox}}}
 
@@
 %authors
-\author{Ariel I. Mundo \orcidaffila{}}
-\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
-\author{John R. Tipton \orcidaffilb{}}
-\affil{Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
-\author{Timothy J. Muldoon*}
+\author{Corresponding author: Timothy J. Muldoon*}
 \affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
@@
 \usepackage{xcolor}
+\definecolor{reviewersblue}{HTML}{1A30AD}
 \definecolor{sbase03}{HTML}{002B36}
@@
 
-\hypertarget{general-comments}{%
-\section{General Comments}\label{general-comments}}
+\hypertarget{general-comments-to-the-reviewers}{%
+\section{General Comments to the Reviewers}\label{general-comments-to-the-reviewers}}
+
+We would like to thank the reviewer for the careful and thorough analysis of this manuscript to \emph{Statistics in Medicine} and for the thoughtful comments and constructive suggestions, which helped improve the quality of this manuscript.
+We carefully considered the reviewer's comments and in this document, explain how we revised the manuscript based on those comments and suggestions.
+
+\hypertarget{general-comments-to-the-editor}{%
+\section{General Comments to the Editor}\label{general-comments-to-the-editor}}
+
+Dr.~Platt,
+
+The authors thank you for your determination that our manuscript may be suitable for resubmission in \emph{Statistics in Medicine} after addressing the reviewer's comments.
+To this end, we have addressed all critiques.
+We hope these revisions, submitted on December 5, 2021, improve the manuscript so it is deemed worthy of publication in \emph{Statistics in Medicine}.
+Following are our detailed responses to reviewer comments.
+
+\hypertarget{reviewers-introduction}{%
+\subsection{Reviewer's Introduction}\label{reviewers-introduction}}
+
+Mundo and colleagues present a tutorial on the use of generalized additive models to analyze longitudinal data.
+A comparison with repeated measures ANOVA and linear mixed models is provided and a recurring example using simulated
+data is used to illustrate the differences among the methods and the advantages of GAM.
+While I like the general approach and aim of the manuscript, there are a number of inaccuracies and omissions, especially in the discussion of GAMs that make it difficult for me to support publication at this stage.
+I believe --- given evidence elsewhere in the manuscript that the authors really do know the subject --- that these inaccuracies and omissions stem from preparing a tutorial in a scientific paper format where word-length considerations come into play.
+Below I outline the main areas where I feel the GAM methodology is inaccurately presented or important topics omitted, and make suggestions to improve the manuscript.
+
+\hypertarget{section}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Introduction}}{}}\label{section}}
+
+We appreciate Dr.~Simpson's comments and critiques to our manuscript.
 
@@
 \hypertarget{missing-data}{%
-\subsection{Missing data}\label{missing-data}}
+\section{Missing data}\label{missing-data}}
 
-The title for Section 3.4 has been changed from ``Missing observations'' to ``Unbalanced data''. We have restructured this section in order to:
+\hypertarget{reviewers-comments}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments}}
+
+I found this use of ``missing'' data to be a little confusing.
+I understand what the authors are getting at, but it suggests that GAMs can handle this whereas rm-ANOVA and LMMs can't.
+While I appreciate that rm-ANOVA might require
+balanced observations for the group errors, LMMs are just as able to handle ``missingness'' (in the sense implied by the authors) as GAMs are.
+Additionally, the ``missingness'' isn't meant solely as missing in the statistical sense, but really it is due to irregular or incomplete sampling, more generally.
+It would be helpful to make the discussion of missingness about balanced data and to try to avoid terms like ``missing'' as that may be inferred to suggest GAMs and LMMs are immune to missing data problems (they aren't), they
+just don't require balance.
+
+\hypertarget{section-1}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-1}}
+
+We appreciate the comments about clarity in the use of the term ``missingness''.
+Please note that the title for Section 3.4 has been changed from ``Missing observations'' to ``Unbalanced data''.
+We have restructured this section in the following manner:
 
@@
 \item
-  Avoid confusion by using the term ``missing data''. We now refer to different number of observations as ``unbalanced data'' (L201-204).
+  The term ``missing data'' has been removed to avoid confusion. We now refer to different number of observations as ``unbalanced data'' (L201-204).
 \item
-  Make clear that LMEMs also can work with missing observations (L209)
+  We have made clear that LMEMs can also work with missing observations (L209)
 \item
-  Emphasize that GAMs are not immune to missing data problems (L217-220)
+  Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates (L217-220).
 \end{itemize}
@@
 \hypertarget{unnecesary-restriction-of-lmms}{%
-\subsection{Unnecesary restriction of LMMs}\label{unnecesary-restriction-of-lmms}}
+\section{Unnecesary restriction of LMMs}\label{unnecesary-restriction-of-lmms}}
+
+\hypertarget{reviewers-comments-1}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-1}}
+
+Why not use quadratic effects of time in the LMM?
+I appreciate this would make the model more complex, but trying to fit a quadratic effect with a linear model strikes me as futile and somewhat of a straw person argument to make.
+
+\hypertarget{section-2}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-2}}
+
+We appreciate the reviewer's perspective regarding the somewhat contradictory argument of fitting a linear effects LMM to quadratic data.
+However, this is an intentional mistake because we follow the Statistical analysis logic followed by most Biomedical researchers: Fit a LMM, and if you get significant p-values that's good enough.
+We are trying to convey the point that although ``significant'' p-values can be obtained, visualizing the model fit is something that should be required, but that is seldomly done.
+\textbf{Incorporate in the section that a strenght of GAMs is that they can learn the function from the data, while LMMs can't.}
 
 \hypertarget{glmms}{%
-\subsection{GLMMs}\label{glmms}}
+\section{GLMMs}\label{glmms}}
 
 \hypertarget{glms-gams-and-conditional-distributions}{%
-\subsection{GLMs, GAMs, and conditional distributions}\label{glms-gams-and-conditional-distributions}}
+\section{GLMs, GAMs, and conditional distributions}\label{glms-gams-and-conditional-distributions}}
 
 \hypertarget{thin-plate-regression-splines}{%
-\subsection{Thin plate regression splines}\label{thin-plate-regression-splines}}
+\section{Thin plate regression splines}\label{thin-plate-regression-splines}}
+
+\hypertarget{reviewers-comments-2}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-2}}
+
+The authors define the default basis in \{mgcv\}, the software being used here to illustrate fitting GAMs, as ``\ldots thin plate regression splines are an optimized version that work well with noisy data.'' This is not a good description of what a thin plate regression spline (hereafter TPRS) is, and is not a good description of what the low-rank versions in \{mgcv\} are.
+
+\hypertarget{section-3}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-3}}
 
@@
 \begin{itemize}
+\tightlist
 \item
   The information on L289-294 pertaining splines has been changed.
-L296-307 now give a better description of cubic splines (CS), indicating their limitation due to knot placement.
-The lack of knot placement for thin plate regression splines (TPRS) is indicated as an advantage over CS.
+L301-309 now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
+\end{itemize}
+
+\hypertarget{reviewers-comments-3}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-3}}
+
+Much better, I believe, is to talk about this in terms of basis functions, of which there are five in this example.
+You can use the same terminology to refer to the CRS basis, and others.
+That in the CRS if you want \emph{k} basis functions you need \emph{k} knots (IIRC).
+CRS require you to specify the knot locations (or let the software spread them evenly through the data); TPRS don't.
+Hence the paragraph starting on L195 is confusing and misleading.
+The basis functions are not piece-wise polynomial and there are not ``region{[}s{]} where a different set of basis functions will be used''.
+Indeed, the functions you show
+operate throughout the range of the covariate.
+
+\hypertarget{section-4}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-4}}
+
+\begin{itemize}
+\tightlist
 \item
-  L297-300 no longer refer to ``knots'' when referring to the construction of the smooth, now indicating that the number of \emph{basis functions} is what is specified when working with TPRS.
-In the updated manuscript this appears in L310-315.
+  L297-300 no longer refer to ``knots'' when referring to the construction of the smooth, now indicating that the number of \emph{basis functions} is what is specified when constructing the smoother.
+In the updated manuscript, this change appears in L310-315, and is more clear to the reader due to the changes introduced in L301-L309.
 \end{itemize}
@@
 \hypertarget{penalised-splines}{%
-\subsection{Penalised splines}\label{penalised-splines}}
+\section{Penalised splines}\label{penalised-splines}}
+
+\hypertarget{reviewers-comments-4}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-4}}
+
+You talk a little about ``wiggliness'' and penalising the weights for each basis function towards 0 but you don't really explain this most crucial concept of the model fitting problem and why modern GAMs are so much better than the
+GAMs developed by Hastie and Tibshirani when they first introduced GAMs to the world.
+You really need to define wiggliness beyond something that is used to avoid overfitting.
+Typically it is squared second derivative of the estimated function; so we limit the curvature of the fitted function by default.
+Technically the penalty is the penalty matrix - it is a matrix and it is fixed once we define the basis functions.
+What isn't fixed is (are) the smoothness parameter(s) of the smooth; it is those that control how much penalty we subtract from the log-likelihood of the data given the model estimates.
+To speak of a ``weak'' or ``strong'' penalty was a little confusing for me.
+Hence fitting a GAM requires one to estimate parameters and one or more smoothness parameters for each smooth in the GAM, plus any parameters required for parametric terms.
+By glossing over these important concepts, the reader is left wondering what wiggliness is, how we measure overfit? etc.
 
 \hypertarget{bayesian}{%
-\subsection{Bayesian}\label{bayesian}}
+\section{Bayesian}\label{bayesian}}
+
+\hypertarget{reviewers-comments-5}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-5}}
+
+GAMs in \{mgcv\} are considered to be empirical Bayesian models, but in general GAMs can be fully Bayesian.
+You can fit fully Bayesian GAMs using INLA and JAGS using functions from \{mgcv\}, and the \{brms\} package allows full Bayesian GAMs to be estimated simply using Stan for example.
+This needs to be clarified.
+I'm not really sure what you mean by the sentence beginning ``Moreover, the use of the restricted maximum\ldots{}''.
+I think I get what you mean; by casting the wiggly parts of smooths as random effects we can estimate the fit using REML
+and standard linear mixed model software. However, you can fit the model using the full fat version of maximum likelihood and these models would still be empirical Bayes if fitted by \{mgcv\}.
+
+\hypertarget{section-5}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-5}}
 
@@
   We have clarified the concept ``empirical Bayesian'', which appeared on L318-L319 in the original manuscript.
-In the revised manuscript, L335-336 now indicate that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
+In the revised manuscript, L335-338 now indicate that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
 \item
   The sentence ``Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model'', which appears in L319 in the original manuscript has been removed from the text.
-Instead, the concept of REML has been moved to Section 6.2 L390-393, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in \emph{mgcv}.
+Instead, the concept of REML has been moved to Section 6.2 L392-395, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in \emph{mgcv}.
 \end{itemize}
@@
 \hypertarget{coverage-of-confidence-intervals}{%
-\subsection{Coverage of confidence intervals}\label{coverage-of-confidence-intervals}}
+\section{Coverage of confidence intervals}\label{coverage-of-confidence-intervals}}
+
+\hypertarget{reviewers-comments-6}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-6}}
+
+This entire section from L320 onward through to the end of Section 5 needs some work.
+When viewed from the Bayesian perspective, the intervals are Bayesian credible intervals.
+When viewed from a frequentist perspective, the same intervals are confidence intervals but instead of having the typical point-wise interpretation they have an across the function interpretation.
+The description of what this means is wrong --- you are almost giving the incorrect definition of a confidence interval here.
+The interval either does or does not contain the true function.
+That is given.
+I'm sure this is just a slip then on L324-325.
+Also, your description implies a simultaneous interval, although I don't think you intended this.
+One could interpret ``95\% of the time'' as meaning 95\% of the functions are contained in their entirety.
+
+What across-the-function means is simply that if we average the coverage of the interval over the entire function we get approximately the nominal coverage, 95\% say.
+For this to occur then, some areas of the function must have more
+than nominal coverage and some areas less than the nominal coverage.
+
+\hypertarget{section-6}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-6}}
 
@@
 \item
-  L339-L352 now contain a more detailed and accurate explanation on the differences and interpretation of ``pointwise'' CIs and ``across-the-function'' CIs.
-We also reference the work of Marra and Wood if the reader desires a more in-depth exploration.
+  L341-L355 now contain a more detailed and accurate explanation on the differences and interpretation of ``poin-twise'' CIs and ``across-the-function'' CIs.
+We removed the ``95\% of the time'' phrase to avoid confusion, and instead we provide an explanation that focuses on random sampling: ``if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function'' (L352-354).
+We also reference the work of Marra and Wood if the reader desires a more in-depth exploration of across the function CIs.
 \end{itemize}
@@
 \hypertarget{differences-in-smooths}{%
-\subsection{Differences in smooths}\label{differences-in-smooths}}
+\section{Differences in smooths}\label{differences-in-smooths}}
+
+\hypertarget{reviewers-comments-7}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-7}}
+
+\hypertarget{section-7}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-7}}
 
 \hypertarget{appendix}{%
-\subsection{Appendix}\label{appendix}}
+\section{Appendix}\label{appendix}}
+
+\hypertarget{reviewers-comments-8}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-8}}
+
+In the appendix I think you are needlessly restricting the size of the basis dimension to be k = 5, hence 4 basis functions per smooth when identifiability constraints are applied.
+Is there a reason I'm seeing here why you could leave this at the default k = 10 and really see the effect of the shrinkage as the EDF of the resulting smooths should be similar to the EDF you have with k = 5?
+This would also likely help with the k-index being low because there's not a lot of shrinkage you can do when the maximum EDF possible is 4 (per smooth).
+
+\hypertarget{section-8}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-8}}
+
+In L367 it is indicated that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
+The computational requisite in the smooth estimation of having the maximum number of basis equal to the number of unique values in the covariate makes it impossible to fit a GAM with a k\textgreater5, as the error ``A term has fewer unique covariate combinations than specified maximum degrees of freedom'' is thrown by \emph{mgcv}.
+We consider that because the k-index for the model is 1.04 the basis dimension for the smooth is adequate (Wood says that ``the further below 1 this is, th emore likely it is that there is missed pattern left in the residuals'').
+
+\hypertarget{reviewers-comments-9}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-9}}
+
+Why change the model object name notation here?
+You had gam\_00 previously and now you call the model m1?
+As you are already showing appraise() output for the diagnostic plots, you could use check.k() from \{mgcv\} to just get the basis dimension check.
+
+\hypertarget{section-9}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-9}}
+
+The name of the model has been changed to \passthrough{\lstinline!gam\_02!}, and it appears with this notation in Section 6.2, L383 so it matches the workflow of model selection presented in the Appendix.
 

### 2021-11-05 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.Rmd
Updated response to reviewers

--- 2021-10-13 Manuscripts/Manuscript by chapters- SIM Revisions/00-Full_document_SIM_No_Appendix.Rmd
@@
     includes:
-      in_header: preamble.sty #additional LaTeX formatting
+      in_header: latex_docs/preamble.sty #additional LaTeX formatting
   bookdown::word_document2:
@@
   bookdown::html_document2:
-    css: "style.css" #style for the HTML document
-csl: american-medical-association.csl #style for references
-bibliography: refs.bib #references
+    css: "html_docs/style.css" #style for the HTML document
+csl: bibliography/american-medical-association.csl #style for references
+bibliography: bibliography/refs.bib #references
 link-citations: yes #adds links to the citations
 '': default
+# knit: |
+#   (function(inputFile, encoding) {
+#     outputFile<-gsub("Rmd$","pdf", basename(inputFile))
+#     rmarkdown::render(
+#       input=inputFile, 
+#       output_file= here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions/output", outputFile)
+#     )
+#   })
 ---
@@
 
-#knitr::opts_chunk$set(fig.pos = 'H', out.extra = "",error=TRUE)
-
 #global themes for plotting
@@
 thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
-
-#thm<-scale_color_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting
-#thm1<-scale_fill_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting
 

### 2021-11-05 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Updated response to reviewers

--- 2021-11-04 Manuscripts/Manuscript_by_chapters-SIM_Revisions/00-Full_document_SIM_No_Appendix.tex
@@
 
-{\centering \includegraphics[width=1\linewidth]{00-Full_document_SIM_No_Appendix_files/figure-latex/l-q-response-1}
+{\centering \includegraphics[width=1\linewidth]{Full_document_SIM_No_Appendix_files/figure-latex/l-q-response-1}
 
@@
 
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{54}} Although different types of splines exist, cubic and thin plate regression splines will be briefly discussed next to give a general idea of the development within the GAM framework.
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{54}} Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
 
 Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{52}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} In contrast, Wood\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} provided a computationally efficient implementation of thin plate regression splines(TPRS), which do not require the knot placement, and require simply a choice on the number of basis to be used to create the smoother (for mathematical details see refs.\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} and).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
+The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{52}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate.\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{55}}
+
+In contrast, thin regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner.\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see refs.\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} and).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_SIM_No_Appendix_files/figure-latex/basis-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SIM_No_Appendix_files/figure-latex/basis-plot-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_SIM_No_Appendix_files/figure-latex/sim-smooth-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SIM_No_Appendix_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_SIM_No_Appendix_files/figure-latex/plot-pairwise-comp-1}
+{\centering \includegraphics[width=0.75\linewidth]{Full_document_SIM_No_Appendix_files/figure-latex/plot-pairwise-comp-1}
 

### 2021-11-05 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated response to reviewers

@@
 Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as _basis functions_, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
-Although different types of splines exist, cubic and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
+Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
 
@@
 The use of CS as smoothers in GAMs was discussed within the original GAM framework [@hastie1987], but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as 'knots', the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit [@wood2003].
-In contrast, Wood [@wood2003] provided a computationally efficient implementation of thin plate regression splines (TPRS), which do not need knot placement.
-Instead, TPRS only require the number of basis to be used to create the smoother (for mathematical details see refs.
+A solution to the "knot" placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate [@wood2003; @wood2017].
+
+In contrast, thin regression splines (TPRS) provide a reasonable "low rank" (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner [@wood2003].
+Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see refs.
 [@wood2003] and [@wood2017]).

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.Rmd
Updated reviewer's comments

@@
 
-The name of the model has been changed to `gam_02`, and it appears with this notation in Section 6.2, L383 so it matches the workflow of model selection presented in the Appendix.
+- The name of the model has been changed to `gam_02`, and it appears with this notation in Section 6.2, L383 so it matches the workflow of model selection presented in the Appendix.
+- Additionally, we have used some of the source code of the function `gam.check` in order to use the graphical output of `gratia` and the numerical diagnostic information of `gam.check` without repeating the diagnostic plots.
+
+# Suggested Changes
+
+## Reviewer's Comments
+
+Essentially now that we have knot-free splines, we can focus on the concept of the user’s prior for the upper limit on the wiggliness of the each smooth; this is set using k.
+You also should explain that we need to increase k a little above this prior expectation because the basis of dimension k + K has a richer set of functions of complexity k than a basis of dimension k.
+From a practical point of view, the user really only needs to think about some anticipated amount of wiggliness and set k to be a little larger than this, then fit the model.
+Next the user should do the check.k() test on the size of the basis to see if it was large enough. If it isn’t, consider increasing k a bit and then refit and recheck k. Rinse and repeat.
+
+Beyond the usual model diagnostics and choosing the conditional distribution of the response, that’s all the user really needs to do to get fitting GAMs.
+And this basic concept is largely missing from the manuscript and the worked tutorial.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+## Reviewer's Comments
+
+For the Appendix, I would suggest that you break up the code a bit to include a little narrative code explaining what each sub-chunk is doing.
+At the moment you have these monolithic blocks of code that produce a figure and/or some
+printed console output and the user is largely left to figure out what the code is doing.
+I appreciate that there are comments, but for a tutorial, it would be better to have something more like a vignette rather than monolithic code
+blocks with comments.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+# Figures
+
+## Reviewer's Comments
+
+The figures are too tall — the aspect ratio used for the individual panels is wrong. You should be emphasising the time axis so you should have individual panels that are wider than they are tall.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We appreciate this comment regarding figure aesthetitcs, we have now made the figures wider to emphasize the time axis.
+
+## Reviewer's Comments
+
+Figure 3: on panel A, are the solid lines the truth in the main panel?
+If they are, you don’t need the inset plot, and if they aren’t, are they the mean of the simulated observations?
+If they are this latter, I think this is superfluous and it would be better to make the solid line the truth, about which you simulated noisy data.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+Although data simulation is a very common practice in Statistics/Ecology, it not so much in biomedical research.
+The main goal of Figure 3 A is to show that simulation is indeed a plausible tool to explore data behavior and to test statistical models in biomedical research.
+It might seem too basic or repetitive, but we are trying to convince the reader that simulation can indeed produce synthetic data that can be useful.
+
+# Minor Comments
+
+## Reviewer's Comments
+
+L57 _p_ values cannot be significant or otherwise, they just are.
+They might indicate statistically significant effects, where the weight of evidence against the null hypothesis is sufficient to meet some threshold of “significance”.
+I note the use of scare quotes so perhaps this sentence was intended in jest but even
+so it is confusing and unnecessary.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We appreciate the reviewer's comments here.
+However, the indicated sentence is not intended in jest.
+We are using in this sentence (and in the paper in general) a statistical approach that is understandable by biomedical researchers in general.
+To this day, the field values "significance" (defined by the arbitrarily set threshold of _p_<0.05).
+We are trying to convince the reader that low _p-values_ are possible with ill fit models using language that is understandable.
+
+## Reviewer's Comments
+
+Also here we have “non-linear data” which almost surely should be “non-linear effects” — data aren’t non-linear, but relationships between variables, “effects”, might be. See also L316
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We have changed the expression to "data that shows non-linear trends". The change has also been applied to L316 which is now L334 in the revised manuscript.
+
+## Reviewer's Comments
+
+L93 “distribution of the errors of the random effects” makes no sense.
+Also LMMs do not restrict random effects to be i.i.d.
+Gaussian — they could be i.i.d.
+gamma or t distributed for example.
+That typical R software (and mgcv from the GAM side) makes an assumption that the distribution of individual random effects terms is Gaussian, doesn’t mean the model class is so restricted.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+## Reviewer's Comments
+
+L185 here and throughout, check for situations where you have combined numerical citations to literature within narrative text; Here it should be “…see refs 40 & 42.)” for example, not superscript numerals.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+**leave this to the end as it can be fixed in the final .tex document**
+
+## Reviewer's Comments
+
+L244 “global” is perhaps more conventionally described as “population” in the mixed model literature.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We have changed the word from  "global" to "population".
+
+## Reviewer's Comments
+
+L283 I think it is better to say “…the mathematical space within which the true but unknown $f(x_t)$ is thought to exist.”
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We have updated this line to "the mathematical space within which the true but unknown $f(x_t\mid \beta_j)$  is thought to exist", in L290 in the revised manuscript.
+
+## Reviewer's Comments
+
+L293 You never really describe what “data-drive flexibility” is. Why are GAMs data driven and LMMs and GLMs not? This is a critical concept to convey I feel.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+This line has been removed from the revised manuscript.
+
+## Reviewer's Comments
+
+L351 Better to say “separate smooths” not “independent”.
+These factor-by smooths share a common basis, they just get an entirely separate set of model coeficients and smoothness parameters for each level of the factor.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+The sentence now reads "separate smooths for..."
+
+## Reviewer's Comments
+
+L354 “difference between group means” because the smooth encodes differences between the groups too, just not difference in the means. Be a little more specific here.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+The sentence now reads "any differences between the group means".
+
+## Reviewer's Comments
+
+L356 use straight quotes in code, not curly ones. And speficy the `family` here?
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+## Reviewer's Comments
+
+L359 “contain the fitted” instead of “store the”
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+The sentence now reads "contains the fitted model".
+
+## Reviewer's Comments
+
+L360 Not “knots” - there are k-1 basis functions, but as described above, because this is a TPRS there are knots at the unique data values and then we decompose the full basis and take the required eigenvectors as the basis to use.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+The updated sentence now reads "using four basis functions (plus the intercept)".
+
+## Reviewer's Comments
+
+L362 why highlight Gaussian process smooths?
+They’re not as useful as they seem as you need to optimise the length scale parameter yourself as otherwise, especially for longitudinal data they aren’t great - the length scale is the largest separate in time of any pair of observations.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+For the length of the covariates explored in this paper TPRS are sufficient.
+However, we want to make clear to the reader that other types of splines that might be appropriate in each case are available.
+To acknowledge this, L390 in the updated manuscript now reads "(a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console)".
+
+## Reviewer's Comments
+
+L366 See also Matteo Fasiolo’s {mgcViz} package
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We acknowledge the existence of Dr. Fasiolo's _mgcvViz_ package, but due to the introductory nature of this paper we prefer to focus only on _mgcv_ and _gratia_.
+
+## Reviewer's Comments
+
+L373 Do rm-ANOVA and LMMs require equally-spaced or complete observations?
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+Only rm-ANOVA requires equally-spaced and complete observations.
+To avoid confusion, the line now reads "Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does)".
+
+## Reviewer's Comments
+
+L377 delete “total”
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+The word "total" has been removed. The sentence now reads "If 40% of the observations are randomly deleted..."
+
+## Reviewer's Comments
+
+L381 “little” -> “few”
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+The word "few" has been removed. The sentence now reads "...with as few as 4 observations per group..."
+
+## Reviewer's Comments
+
+L393 “observed difference is significant” (not “change”)
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+The sentence has been rewritten. It now reads "...the observed difference is significant..."
+
+## Reviewer's Comments
+
+Also, it is (exceedingly) unlikely to yield 0. Better to say that the estimated difference is unlikely to be distinguishable from 0.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+## Reviewer's Comments
+
+L410 This needs to be reworded as what is written can be read as any difference between the control and treatment (and there is a difference as the controls are higher than the treated group before day 3) where you mean that the treatment group is only higher than the control group post day 3.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+This was an unintentional mistake based on some previous simulation results.
+We have updated the paragraph, and it now reads "shows that the Control Group has higher $\mbox{StO}_2$ before day 3, and is able to estimate the change on day 3 where the Treatment Group becomes significant as the full dataset smooth pairwise comparison." The change acknowledges that Control was higher before day 3, and that the Treatment Group becomes significant after day 3.
+
+## Reviewer's Comments
+
+L441 Delete “significant” here - we estimate the difference of the effects (what you call “change”) and make inference using the uncertainty estimates of that difference.
+Perhaps you mean “identify” or similar instead of “estimate” here, instead.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We have made the appropriate changes. The sentence now reads "The model is therefore able to identify changes between the groups at time points..."

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.tex
Updated reviewer's comments

@@
 
+\begin{itemize}
+\tightlist
+\item
   The name of the model has been changed to \passthrough{\lstinline!gam\_02!}, and it appears with this notation in Section 6.2, L383 so it matches the workflow of model selection presented in the Appendix.
+\item
+  Additionally, we have used some of the source code of the function \passthrough{\lstinline!gam.check!} in order to use the graphical output of \passthrough{\lstinline!gratia!} and the numerical diagnostic information of \passthrough{\lstinline!gam.check!} without repeating the diagnostic plots.
+\end{itemize}
+
+\hypertarget{suggested-changes}{%
+\section{Suggested Changes}\label{suggested-changes}}
+
+\hypertarget{reviewers-comments-10}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-10}}
+
+Essentially now that we have knot-free splines, we can focus on the concept of the user's prior for the upper limit on the wiggliness of the each smooth; this is set using k.
+You also should explain that we need to increase k a little above this prior expectation because the basis of dimension k + K has a richer set of functions of complexity k than a basis of dimension k.
+From a practical point of view, the user really only needs to think about some anticipated amount of wiggliness and set k to be a little larger than this, then fit the model.
+Next the user should do the check.k() test on the size of the basis to see if it was large enough. If it isn't, consider increasing k a bit and then refit and recheck k. Rinse and repeat.
+
+Beyond the usual model diagnostics and choosing the conditional distribution of the response, that's all the user really needs to do to get fitting GAMs.
+And this basic concept is largely missing from the manuscript and the worked tutorial.
+
+\hypertarget{section-10}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-10}}
+
+\hypertarget{reviewers-comments-11}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-11}}
+
+For the Appendix, I would suggest that you break up the code a bit to include a little narrative code explaining what each sub-chunk is doing.
+At the moment you have these monolithic blocks of code that produce a figure and/or some
+printed console output and the user is largely left to figure out what the code is doing.
+I appreciate that there are comments, but for a tutorial, it would be better to have something more like a vignette rather than monolithic code
+blocks with comments.
+
+\hypertarget{section-11}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-11}}
+
+\hypertarget{figures}{%
+\section{Figures}\label{figures}}
+
+\hypertarget{reviewers-comments-12}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-12}}
+
+The figures are too tall --- the aspect ratio used for the individual panels is wrong. You should be emphasising the time axis so you should have individual panels that are wider than they are tall.
+
+\hypertarget{section-12}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-12}}
+
+We appreciate this comment regarding figure aesthetitcs, we have now made the figures wider to emphasize the time axis.
+
+\hypertarget{reviewers-comments-13}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-13}}
+
+Figure 3: on panel A, are the solid lines the truth in the main panel?
+If they are, you don't need the inset plot, and if they aren't, are they the mean of the simulated observations?
+If they are this latter, I think this is superfluous and it would be better to make the solid line the truth, about which you simulated noisy data.
+
+\hypertarget{section-13}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-13}}
+
+Although data simulation is a very common practice in Statistics/Ecology, it not so much in biomedical research.
+The main goal of Figure 3 A is to show that simulation is indeed a plausible tool to explore data behavior and to test statistical models in biomedical research.
+It might seem too basic or repetitive, but we are trying to convince the reader that simulation can indeed produce synthetic data that can be useful.
+
+\hypertarget{minor-comments}{%
+\section{Minor Comments}\label{minor-comments}}
+
+\hypertarget{reviewers-comments-14}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-14}}
+
+L57 \emph{p} values cannot be significant or otherwise, they just are.
+They might indicate statistically significant effects, where the weight of evidence against the null hypothesis is sufficient to meet some threshold of ``significance''.
+I note the use of scare quotes so perhaps this sentence was intended in jest but even
+so it is confusing and unnecessary.
+
+\hypertarget{section-14}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-14}}
+
+We appreciate the reviewer's comments here.
+However, the indicated sentence is not intended in jest.
+We are using in this sentence (and in the paper in general) a statistical approach that is understandable by biomedical researchers in general.
+To this day, the field values ``significance'' (defined by the arbitrarily set threshold of \emph{p}\textless0.05).
+We are trying to convince the reader that low \emph{p-values} are possible with ill fit models using language that is understandable.
+
+\hypertarget{reviewers-comments-15}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-15}}
+
+Also here we have ``non-linear data'' which almost surely should be ``non-linear effects'' --- data aren't non-linear, but relationships between variables, ``effects'', might be. See also L316
+
+\hypertarget{section-15}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-15}}
+
+We have changed the expression to ``data that shows non-linear trends''. The change has also been applied to L316 which is now L334 in the revised manuscript.
+
+\hypertarget{reviewers-comments-16}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-16}}
+
+L93 ``distribution of the errors of the random effects'' makes no sense.
+Also LMMs do not restrict random effects to be i.i.d.
+Gaussian --- they could be i.i.d.
+gamma or t distributed for example.
+That typical R software (and mgcv from the GAM side) makes an assumption that the distribution of individual random effects terms is Gaussian, doesn't mean the model class is so restricted.
+
+\hypertarget{section-16}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-16}}
+
+\hypertarget{reviewers-comments-17}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-17}}
+
+L185 here and throughout, check for situations where you have combined numerical citations to literature within narrative text; Here it should be ``\ldots see refs 40 \& 42.)'' for example, not superscript numerals.
+
+\hypertarget{section-17}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-17}}
+
+\textbf{leave this to the end as it can be fixed in the final .tex document}
+
+\hypertarget{reviewers-comments-18}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-18}}
+
+L244 ``global'' is perhaps more conventionally described as ``population'' in the mixed model literature.
+
+\hypertarget{section-18}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-18}}
+
+We have changed the word from ``global'' to ``population''.
+
+\hypertarget{reviewers-comments-19}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-19}}
+
+L283 I think it is better to say ``\ldots the mathematical space within which the true but unknown \(f(x_t)\) is thought to exist.''
+
+\hypertarget{section-19}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-19}}
+
+We have updated this line to ``the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist'', in L290 in the revised manuscript.
+
+\hypertarget{reviewers-comments-20}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-20}}
+
+L293 You never really describe what ``data-drive flexibility'' is. Why are GAMs data driven and LMMs and GLMs not? This is a critical concept to convey I feel.
+
+\hypertarget{section-20}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-20}}
+
+This line has been removed from the revised manuscript.
+
+\hypertarget{reviewers-comments-21}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-21}}
+
+L351 Better to say ``separate smooths'' not ``independent''.
+These factor-by smooths share a common basis, they just get an entirely separate set of model coeficients and smoothness parameters for each level of the factor.
+
+\hypertarget{section-21}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-21}}
+
+The sentence now reads ``separate smooths for\ldots{}''
+
+\hypertarget{reviewers-comments-22}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-22}}
+
+L354 ``difference between group means'' because the smooth encodes differences between the groups too, just not difference in the means. Be a little more specific here.
+
+\hypertarget{section-22}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-22}}
+
+The sentence now reads ``any differences between the group means''.
+
+\hypertarget{reviewers-comments-23}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-23}}
+
+L356 use straight quotes in code, not curly ones. And speficy the \passthrough{\lstinline!family!} here?
+
+\hypertarget{section-23}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-23}}
+
+\hypertarget{reviewers-comments-24}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-24}}
+
+L359 ``contain the fitted'' instead of ``store the''
+
+\hypertarget{section-24}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-24}}
+
+The sentence now reads ``contains the fitted model''.
+
+\hypertarget{reviewers-comments-25}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-25}}
+
+L360 Not ``knots'' - there are k-1 basis functions, but as described above, because this is a TPRS there are knots at the unique data values and then we decompose the full basis and take the required eigenvectors as the basis to use.
+
+\hypertarget{section-25}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-25}}
+
+The updated sentence now reads ``using four basis functions (plus the intercept)''.
+
+\hypertarget{reviewers-comments-26}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-26}}
+
+L362 why highlight Gaussian process smooths?
+They're not as useful as they seem as you need to optimise the length scale parameter yourself as otherwise, especially for longitudinal data they aren't great - the length scale is the largest separate in time of any pair of observations.
+
+\hypertarget{section-26}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-26}}
+
+For the length of the covariates explored in this paper TPRS are sufficient.
+However, we want to make clear to the reader that other types of splines that might be appropriate in each case are available.
+To acknowledge this, L390 in the updated manuscript now reads ``(a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console)''.
+
+\hypertarget{reviewers-comments-27}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-27}}
+
+L366 See also Matteo Fasiolo's \{mgcViz\} package
+
+\hypertarget{section-27}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-27}}
+
+We acknowledge the existence of Dr.~Fasiolo's \emph{mgcvViz} package, but due to the introductory nature of this paper we prefer to focus only on \emph{mgcv} and \emph{gratia}.
+
+\hypertarget{reviewers-comments-28}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-28}}
+
+L373 Do rm-ANOVA and LMMs require equally-spaced or complete observations?
+
+\hypertarget{section-28}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-28}}
+
+Only rm-ANOVA requires equally-spaced and complete observations.
+To avoid confusion, the line now reads ``Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does)''.
+
+\hypertarget{reviewers-comments-29}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-29}}
+
+L377 delete ``total''
+
+\hypertarget{section-29}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-29}}
+
+The word ``total'' has been removed. The sentence now reads ``If 40\% of the observations are randomly deleted\ldots{}''
+
+\hypertarget{reviewers-comments-30}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-30}}
+
+L381 ``little'' -\textgreater{} ``few''
+
+\hypertarget{section-30}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-30}}
+
+The word ``few'' has been removed. The sentence now reads ``\ldots with as few as 4 observations per group\ldots{}''
+
+\hypertarget{reviewers-comments-31}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-31}}
+
+L393 ``observed difference is significant'' (not ``change'')
+
+\hypertarget{section-31}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-31}}
+
+The sentence has been rewritten. It now reads ``\ldots the observed difference is significant\ldots{}''
+
+\hypertarget{reviewers-comments-32}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-32}}
+
+Also, it is (exceedingly) unlikely to yield 0. Better to say that the estimated difference is unlikely to be distinguishable from 0.
+
+\hypertarget{section-32}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-32}}
+
+\hypertarget{reviewers-comments-33}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-33}}
+
+L410 This needs to be reworded as what is written can be read as any difference between the control and treatment (and there is a difference as the controls are higher than the treated group before day 3) where you mean that the treatment group is only higher than the control group post day 3.
+
+\hypertarget{section-33}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-33}}
+
+This was an unintentional mistake based on some previous simulation results.
+We have updated the paragraph, and it now reads ``shows that the Control Group has higher \(\mbox{StO}_2\) before day 3, and is able to estimate the change on day 3 where the Treatment Group becomes significant as the full dataset smooth pairwise comparison.'' The change acknowledges that Control was higher before day 3, and that the Treatment Group becomes significant after day 3.
+
+\hypertarget{reviewers-comments-34}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-34}}
+
+L441 Delete ``significant'' here - we estimate the difference of the effects (what you call ``change'') and make inference using the uncertainty estimates of that difference.
+Perhaps you mean ``identify'' or similar instead of ``estimate'' here, instead.
+
+\hypertarget{section-34}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-34}}
+
+We have made the appropriate changes. The sentence now reads ``The model is therefore able to identify changes between the groups at time points\ldots{}''
 

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Updated reviewer's comments

@@
 A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
-Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
+Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from data that shows non-linear trends, the validity of such metric is dependent on how adequate the model fits the data.
 In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}}.
@@
 When the linearity and compound symmetry assumptions are met, the rm-ANOVA model approximates well the global trend by group (Figure \ref{fig:l-q-response}B).
-Note that because the LMEM incorporates \emph{random effects}, is able to provide estimates for each subject and a ``global'' estimate (Figure \ref{fig:l-q-response}C).
+Note that because the LMEM incorporates \emph{random effects}, is able to provide estimates for each subject and a ``population'' estimate (Figure \ref{fig:l-q-response}C).
 
@@
 The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
+A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
 The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{53}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{56}} In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than extremely ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{57}} Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (point-wise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{56}} In the case of data that shows non-linear trends, the belief that the \emph{true} trend of the data is likely to be smooth rather than extremely ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{57}} Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (point-wise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
 
@@
 For the data in Figure \ref{fig:sim-smooth-plot}, A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
-To estimate this, the model incorporates independent smooths for \emph{Group} and \emph{Day}, respectively.
+To estimate this, the model incorporates separate smooths for \emph{Group} and \emph{Day}, respectively.
 The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}).
@@
 
-This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 basis functions.
-The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups.
+This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using four basis functions (plus the intercept).
+The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} (a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console).
+The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups.
 Although the default \passthrough{\lstinline!method!} used to estimate the smoothing parameters in \emph{mgcv} is generalized cross validation (GCV), Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} showed the restricted maximum likelihood (REML) to be more resistant to overfitting while also easing the quantification of uncertainty in the smooth parameters; therefore in this manuscript REML is always used for smooth parameter estimation.
@@
 
-Because GAMs do not require equally-spaced or complete observations for all subjects, they are advantageous to analyze longitudinal data where missingness exists.
+Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists.
 The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
@@
 
-On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the empirical Bayesian credible intervals of the smooths shows the same trend of the full dataset.
+On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the empirical Bayesian credible intervals of the smooths show the same trend of the full dataset.
 Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that the Control Group has higher \(\mbox{StO}_2\) before day 3, and is able to estimate the change on day 3 where the Treatment Group becomes significant as the full dataset smooth pairwise comparison.
@@
 
-Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
-Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}.
+Biomedical longitudinal data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
+Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}, where it is clear that these models do not capture the non-linear trend of the data, thereby causing a ``model misspecification error.''
+
 This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
+It could be argued that in Section \ref{simulation} a LMEM with quadratic effects could have been used, but because in reality the true function in the data is not known, using polynomial effects ends up causing more questions than answers in this case (e.g., what is the biological interpretation of a quadratic, cubic, or other effect of even higher order?).
 Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{61}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
 
-In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{53}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{62}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{53}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{62}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
 
@@
 One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
-The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
+The model is therefore able to identify changes between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
 

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/01-Background.Rmd
Updated reviewer's comments

--- 2021-06-28 Manuscripts/Manuscript by chapters- OUP submission/01-Background.Rmd
@@
 A _post hoc_ analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
-Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
+Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from data that shows non-linear trends, the validity of such metric is dependent on how adequate the model fits the data.
 In other words, _p-values_ are valid only if the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019].

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Updated reviewer's comments

@@
 When the linearity and compound symmetry assumptions are met, the rm-ANOVA model approximates well the global trend by group (Figure \@ref(fig:l-q-response)B).
-Note that because the LMEM incorporates _random effects_, is able to provide estimates for each subject and a "global" estimate (Figure \@ref(fig:l-q-response)C).
+Note that because the LMEM incorporates _random effects_, is able to provide estimates for each subject and a "population" estimate (Figure \@ref(fig:l-q-response)C).
 

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated reviewer's comments

@@
 The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using _basis function_ expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A _basis_ is a set of functions that spans the mathematical space where the smooths that approximate $f(x_t\mid \beta_j)$ exist [@simpson2018].
+A _basis_ is a set of functions that spans the mathematical space within which the true but unknown $f(x_t\mid \beta_j)$  is thought to exist [@simpson2018].
 For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
@@
 Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief [@mcelreath2018].
-In the case of non-linear data, the belief that the _true_ trend of the data is likely to be smooth rather than extremely "wiggly" introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs [@wood2017].
+In the case of data that shows non-linear trends, the belief that the _true_ trend of the data is likely to be smooth rather than extremely "wiggly" introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs [@wood2017].
 Moreover, GAMs are considered "empirical" Bayesian models when fitted using the package _mgcv_ because the smoothing parameters are estimated from the data (and not from a prior distribution as in the "fully Bayesian" case, which can be fitted using JAGS, Stan, or other probabilistic programming language) [@miller2019].

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated reviewer's comments

@@
 For the data in Figure \@ref(fig:sim-smooth-plot), A the main effect of interest is how $\mbox{StO}_2$ changes over time for each treatment.
-To estimate this, the model incorporates independent smooths for _Group_ and _Day_, respectively.
+To estimate this, the model incorporates separate smooths for _Group_ and _Day_, respectively.
 The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (*Day*) and the other is a factor (*Group*).
@@
 
-This syntax specifies that `gam_02` (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using 5 basis functions.
-The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths [@simpson2018].
+This syntax specifies that `gam_02` (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using four basis functions (plus the intercept).
+The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths [@simpson2018] (a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console).
 The parametric term `Group` is added to quantify overall mean differences in the effect of treatment between groups.
@@
 
-On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the empirical Bayesian credible intervals of the smooths shows the same trend of the full dataset.
+On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the empirical Bayesian credible intervals of the smooths show the same trend of the full dataset.
 Consequently, the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that the Control Group has higher $\mbox{StO}_2$ before day 3, and is able to estimate the change on day 3 where the Treatment Group becomes significant as the full dataset smooth pairwise comparison.

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/05-Discussion.Rmd
Updated reviewer's comments

@@
 
-Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
-Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \@ref(simulation).
+Biomedical longitudinal data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
+Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \@ref(simulation), where it is clear that these models do not capture the non-linear trend of the data, thereby causing a "model misspecification error".
+
 This "model misspecification" error, also is known as a "Type III" error [@dennis2019] is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
+It could be argued that in Section \@ref(simulation) a LMEM with quadratic effects could have been used, but because in reality the true function in the data is not known, using polynomial effects ends up causing more questions than answers in this case (e.g., what is the biological interpretation of a quadratic, cubic, or other effect of even higher order?).
 Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) [@lang2015] but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a $R^2$ value.
 
-In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
+In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.  
 It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) [@hefley2017; @pedersen2019] they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research [@hastie1995].
-This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \@ref(GAM-theory) is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \@ref(GAM-theory) is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
 
@@
 One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a _p-value_, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure  \@ref(fig:plot-pairwise-comp) indicate.
-The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( $\approx$ day 3 in Figure \@ref(fig:plot-pairwise-comp) A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
+The model is therefore able to identify changes between the groups at time points were data was not directly measured even with missing data exists ( $\approx$ day 3 in Figure \@ref(fig:plot-pairwise-comp) A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
 

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
updated discussion

@@
 Independent errors \(\varepsilon_{ijt}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
-In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy'').
+In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug,'' or ``saline'' vs.~``chemotherapy'').
 Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B).
@@
 
-When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)).
+When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEMs and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)).
 This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)).
@@
 
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data.
+For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data.
 As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
+One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
 
@@
 This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-It could be argued that in Section \ref{simulation} a LMEM with quadratic effects could have been used, but because in reality the true function in the data is not known, using polynomial effects ends up causing more questions than answers in this case (e.g., what is the biological interpretation of a quadratic, cubic, or other effect of even higher order?).
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{61}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
+We acknowledge that polynomial effects can be used in LMEMs, and that it could be argued that in Section \ref{simulation} a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects causes more questions than answers in this case (i.e., what is the appropriate polynomial effect to use?
+quadratic?, cubic?, what is the biological interpretation of a polynomial effect?).
+Moreover, polynomial effects have other associated issues that we have described in Section \{\#GAM-theory\}.
+
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{61}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.
 
 In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{53}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{62}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{53}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{62}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).
 

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
updated discussion

@@
 In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
-"novel drug" or "saline" vs.
+"novel drug", or "saline" vs.
 "chemotherapy").
@@
 
-When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
+When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is easily noticeable that LMEMs and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
 This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject (subject$_i$) within each group (group$_j$).
@@
 
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
+For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
 As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
updated discussion

@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with inference [@beck1998].
+One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials (which can be used in LMEMs), but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with inference [@beck1998].
 In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic and causes bias at the endpoints of the time period.

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/05-Discussion.Rmd
updated discussion

@@
 This "model misspecification" error, also is known as a "Type III" error [@dennis2019] is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-It could be argued that in Section \@ref(simulation) a LMEM with quadratic effects could have been used, but because in reality the true function in the data is not known, using polynomial effects ends up causing more questions than answers in this case (e.g., what is the biological interpretation of a quadratic, cubic, or other effect of even higher order?).
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) [@lang2015] but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a $R^2$ value.
+We acknowledge that polynomial effects can be used in LMEMs, and that it could be argued that in Section \@ref(simulation) a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects causes more questions than answers in this case (i.e., what is the appropriate polynomial effect to use?
+quadratic?, cubic?, what is the biological interpretation of a polynomial effect?).
+Moreover, polynomial effects have other associated issues that we have described in Section {#GAM-theory}.
+
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) [@lang2015] but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a $R^2$ value, which is not meaningful in the case of a Type III error.
 
@@
 It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) [@hefley2017; @pedersen2019] they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research [@hastie1995].
-This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \@ref(GAM-theory) is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \@ref(GAM-theory) is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship _a priori_ in order to provide it to the model (i.e., polynomial effects in LMEMs).
 

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/07-Appendix.Rmd
updated discussion

--- 2021-10-17 Manuscripts/Manuscript by chapters- SIM Revisions/07-Appendix.Rmd
@@
 
-```{r, quadratic-cases-Appendix, fig.width=14, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:quadratic-caption-appendix)'}
+```{r, quadratic-cases-Appendix, fig.width=12, fig.height=7, out.width='100%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:quadratic-caption-appendix)'}
 # linear response, correlated errors (subject effect)

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.Rmd
Updated LMEMs, need to incorporate in reviewer's comments.

@@
 We appreciate the reviewer's perspective regarding the somewhat contradictory argument of fitting a linear effects LMM to quadratic data.
-However, this is an intentional mistake because we follow the Statistical analysis logic followed by most Biomedical researchers: Fit a LMM, and if you get significant p-values that's good enough.
-We are trying to convey the point that although "significant" p-values can be obtained, visualizing the model fit is something that should be required, but that is seldomly done.
+However, this is an intentional approach because we follow the statistical analysis logic followed by most biomedical researchers: Fit a LMM or an rm-ANOVA, and if you get significant _p-values_ that's good enough.
+The purpose of the figure is to convey the point that visualizing the model fit is something necessary, but that is rarely done in the field.
 **Incorporate in the section that a strenght of GAMs is that they can learn the function from the data, while LMMs can't.**

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.tex
Updated LMEMs, need to incorporate in reviewer's comments.

@@
 We appreciate the reviewer's perspective regarding the somewhat contradictory argument of fitting a linear effects LMM to quadratic data.
-However, this is an intentional mistake because we follow the Statistical analysis logic followed by most Biomedical researchers: Fit a LMM, and if you get significant p-values that's good enough.
-We are trying to convey the point that although ``significant'' p-values can be obtained, visualizing the model fit is something that should be required, but that is seldomly done.
+However, this is an intentional approach because we follow the statistical analysis logic followed by most biomedical researchers: Fit a LMM or an rm-ANOVA, and if you get significant \emph{p-values} that's good enough.
+The purpose of the figure is to convey the point that visualizing the model fit is something necessary, but that is rarely done in the field.
 \textbf{Incorporate in the section that a strenght of GAMs is that they can learn the function from the data, while LMMs can't.}

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.Rmd
Updated LMEMs, need to incorporate in reviewer's comments.

@@
 In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs).
-Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
+Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies assume an expected trend in the measured response, which can results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
 
-In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
-Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
+In contrast, generalized additive models (GAMs) relax the assumptions made by LMEMs and rm-ANOVA, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+Therefore, GAMs present an excellent choice to analyze longitudinal data with non-linear trends in the context of biomedical research.
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Updated LMEMs, need to incorporate in reviewer's comments.

@@
 In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs).
-Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
-
-In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
-Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
+Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies assume an expected trend in the measured response, which can results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
+
+In contrast, generalized additive models (GAMs) relax the assumptions made by LMEMs and rm-ANOVA, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+Therefore, GAMs present an excellent choice to analyze longitudinal data with non-linear trends in the context of biomedical research.
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
@@
 The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times,\textsuperscript{\protect\hyperlink{ref-wagenmakers2008}{12}} and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}).\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}}
+The rm-ANOVA model makes three assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}).\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}}
 
@@
 Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)),\textsuperscript{\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}} consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni),\textsuperscript{\protect\hyperlink{ref-abdi2010}{21}} but they in turn reduce statistical power (1-\(\beta\))\textsuperscript{\protect\hyperlink{ref-nakagawa2004}{22}}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false).\textsuperscript{\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}} Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni),\textsuperscript{\protect\hyperlink{ref-abdi2010}{21}} but they in turn reduce statistical power (1-\(\beta\))\textsuperscript{\protect\hyperlink{ref-nakagawa2004}{22}}, and lead to increased Type II error (failing to reject the null hypothesis when it is false).\textsuperscript{\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}} Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
 
@@
 The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data.\textsuperscript{\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}} Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}} And even more importantly, LMEMs also assume a linear relationship between the response and time,\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} making them unsuitable to analyze non-linear data.
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}} And even more importantly, LMEMs also assume by default a linear relationship between the response and time,\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} making them unsuitable to analyze data non-linear with non-linear trends (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \{\#GAM-theory\}) .
+
+As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the trend of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
 Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data,\textsuperscript{\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}} health-environment interactions\textsuperscript{\protect\hyperlink{ref-yang2012}{35}} and the dynamics of government in political science.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and 3) can easily accommodate missing observations.
+Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data,\textsuperscript{\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}} health-environment interactions\textsuperscript{\protect\hyperlink{ref-yang2012}{35}} and the dynamics of government in political science.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements,\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and 3) can easily accommodate missing observations.
 Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
@@
 This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations are explained in detail.
+First, the limitations of LMEMs and rm-ANOVA regarding an expected trend of the response, constant correlation structures, and missing observations are explained in detail.
 Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
@@
 These statistical methodologies require certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
+From a practical view, the assumptions can be divided in three areas: 1) an assumed relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
 Each one of these assumptions is discussed below.
 
-\hypertarget{linear-relationship}{%
-\subsection{Linear relationship}\label{linear-relationship}}
+\hypertarget{assumed-relationship}{%
+\subsection{Assumed relationship}\label{assumed-relationship}}
 
@@
 
-\hypertarget{the-linear-mixed-model-case-lmem}{%
-\subsubsection{The Linear Mixed Model Case (LMEM)}\label{the-linear-mixed-model-case-lmem}}
+\hypertarget{the-linear-mixed-model-lmem-case}{%
+\subsubsection{The Linear Mixed Model (LMEM) Case}\label{the-linear-mixed-model-lmem-case}}
 
@@
 
+Of notice, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach, which is described in detail by Pinheiro and Bates\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}.
+However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates;\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} more importantly, polynomial effects lack biological or mechanistic interpretation\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} which limits their use in biomedical studies.
+
 \hypertarget{covariance-in-rm-anova-and-lmems}{%
@@
 A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs.
+We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear data with a linear model such as rm-ANOVA or a LMEM with ``default'' (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
 Of notice, in Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/01-Background.Rmd
Updated LMEMs, need to incorporate in reviewer's comments.

@@
 The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times [@wagenmakers2008], and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004; @schober2018].
+The rm-ANOVA model makes three assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004; @schober2018].
 
@@
 Additionally, the _p-value_ itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010; @halsey2015], consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012; @albers2019].
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when it is false) [@gelman2012; @albers2019].
 Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
@@
 However, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
-And even more importantly, LMEMs also assume a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze non-linear data.
+And even more importantly, LMEMs also assume by default a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze data non-linear with non-linear trends (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section {#GAM-theory}) .
 
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
+As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the trend of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
@@
 Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012; @pedersen2019; @simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998].
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017] and 3) can easily accommodate missing observations.
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017], and 3) can easily accommodate missing observations.
 Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.

### 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Updated LMEMs, need to incorporate in reviewer's comments.

@@
 These statistical methodologies require certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
+From a practical view, the assumptions can be divided in three areas: 1) an assumed relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
 Each one of these assumptions is discussed below.
 
-## Linear relationship
+## Assumed relationship
 
@@
 
-### The Linear Mixed Model Case (LMEM)
+### The Linear Mixed Model (LMEM) Case
 
@@
 However, the expected linear relationship of the covariates and the response in Equation \@ref(eq:linear-model) and in Equation \@ref(eq:LMEM) is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
+
+Of notice, LMEMs are capable of fitting non-linear trends using an "empirical" approach, which is described in detail by Pinheiro and Bates [@pinheiro2006] using polynomial fixed effects instead of linear effects such as in Equation \@ref(eq:LMEM).
+However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates [@beck1998]; more importantly, polynomial effects lack biological or mechanistic interpretation [@pinheiro2006] which limits their use in biomedical studies.
 
@@
 A more comprehensive exploration  of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs.
+We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
 Of notice, in Section \@ref(longitudinal-GAMs) we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.

### 2021-11-09 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.Rmd
Updated simulated data plot, need to fix panels

@@
 library(viridis)
+library(scico)
 set.seed(2021) #set seed for reproducibility
@@
 #global themes for plotting
-thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
-thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
+#thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
+#thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
+thm1<-scale_fill_scico_d(palette="oslo")
 

### 2021-11-09 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Updated simulated data plot, need to fix panels

@@
         text=element_text(size=20))+
-    thm
+    thm1
 

### 2021-11-09 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated simulated data plot, need to fix panels

@@
   theme_classic()+
-  thm
+  thm1
 

### 2021-11-09 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated simulated data plot, need to fix panels

@@
   theme_classic()+
-  thm+
+  thm1+
     scale_x_continuous(breaks=c(0,5,10))+
@@
 #plotting simulated data
-f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group)) +
-    geom_point(show.legend=FALSE,size=1.5,alpha=0.5)+
-    stat_summary(aes(y = StO2_sim,
-                     group=Group), 
-                 fun=mean, geom="line",size=1,show.legend = FALSE)+
+f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group, group=subject)) +
+    geom_point(show.legend=FALSE,size=1.5,alpha=0.6)+
+    geom_line(size=1.1, alpha=0.6)+
+    # stat_summary(aes(y = StO2_sim,
+    #                  group=Group), 
+    #              fun=mean, geom="line",size=1,show.legend = FALSE)+
   labs(y=expression(atop(StO[2],'(simulated)')))+
@@
   )+
-  thm+
+  thm1+
     scale_x_continuous(breaks=c(0,2,5,7,10))
@@
   )+
-      thm+
   thm1
@@
   )+
-  thm+
   thm1
@@
      )&
-  thm
+  thm1
 

### 2021-11-09 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Working on GLM and GLMM section

@@
+# The contemporary statistical landscape
+
+We consider that a proper understanding of GAMs would be incomplete without a brief survey of contemporary statistical models.
+Particularly because it seems to be a disconnect between biomedical research and the current state of Statistics.
+Today, the Statistical landscape has changed drastically and yet, biomedical research at large continues to be confined to a 't-test', 'ANOVA' mentality.
+The shortcomings that such view produces cannot be properly addressed without a brief mention of statistical methods.
+
+In contemporary Statistics,
+
+## Generalized linear models (GLM)
+
+Need the two first equations
+
+$$y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi)$$
+
+$$E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt})$$
+
+and this one
+
+$\eta_{ijt}=\beta_0+\beta_1x_{ijt}$ This is the GLM
+
+## Generalized linear mixed models (GLMMs)
+
+$\eta_{ijt}=\beta_0+\beta_1x_{ijt}+\mu_{ij}$ GLMM
+
 # Challenges presented by longitudinal studies

### 2021-11-10 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
updated

--- 2021-11-09 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
@@
+# The contemporary statistical landscape
+
+A proper understanding of GAMs would be incomplete without a brief mention of the current state in statistical models.
+The advances over the last two decades in computational power and the theoretical framework for statistical models have significantly altered what was once considered "standard" statistical knowledge.
+Today, the analysis of variance (ANOVA) is considered a model in the class of linear models (LM), where the response is described by fixed effects only.
+Linear mixed models (LMMs) are models where fixed and random effects are combined.
+Generalized linear models (GLMs) are those that have fixed effects but where the distribution of the response is not Gaussian.
+And Generalized Linear Mixed models (GLMMs) are a broader category of models where the distribution can be non-Gaussian, and can have fixed and random effects.
+
+In contemporary Statistics,
+
+## Generalized linear models (GLM)
+
+Need the two first equations
+
+$$y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi)$$
+
+$$E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt})$$
+
+and this one
+
+$\eta_{ijt}=\beta_0+\beta_1x_{ijt}$ This is the GLM
+
+## Generalized linear mixed models (GLMMs)
+
+$\eta_{ijt}=\beta_0+\beta_1x_{ijt}+\mu_{ij}$ GLMM
+
 # Challenges presented by longitudinal studies

### 2021-11-11 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/01-Background.Rmd
Updated

@@
 When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006].
-However, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013].
-And even more importantly, LMEMs also assume by default a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze data non-linear with non-linear trends (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section {#GAM-theory}) .
+However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent [@gueorguieva2004;@barr2013].
+And even more importantly, LMEMs also assume by default a linear relationship between the response and time [@pinheiro2006] (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section {#GAM-theory}) .
 

### 2021-11-11 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Updated

--- 2021-11-09 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
@@
+# The contemporary statistical landscape
+
+Although the limitations of rm-ANOVA and LMEMs to analyze non-linear trends in longitudinal data have been presented in a general way in the previous section, a proper understanding of the matter would be incomplete without a brief description of the current accepted knowledge regarding statistical models.
+The advances over the last two decades in computational power and the theoretical framework for statistical models have significantly changed the hierarchy of statistical models.
+Because the topic itself is far outside of the scope of this work
+
+## Linear Models (LMs) and Linear Mixed Models (LMEMs)
+
+Linear models (LMs) are those that assume a normal (Gaussian) distribution, such as ANOVA.
+These are by far the models most commonly used to analyze data within the biomedical research community, and only incorporate _fixed effects_.
+On the other hand, Linear Mixed Effect Models (LMEMs) are those that still assume normality but also incorporate _random effects_, as it has been described before.
+
+## Generalized Linear Models (GLMs)
+
+A major limitation of LMs is their assumption of normalit, and although it can be forced by the use of a transformation, this procedure can lead to poor model performance [@ohara2010], and can cause problems with the biological interpretation of the model estimates.
+McCullagh and Nelder [@nelder1972] defined General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
+To achieve this, consider the following equations:
+
+\begin{equation}
+y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi)
+(\#eq:GLM-y)
+\end{equation}
+
+Where $y_{ijt}$ is the observed response, which is assumed to follow some distribution $\mathcal{D}$ with some mean $\mu_{ijt}$ and a dispersion parameter $\phi$.
+The mean $\mu_{ijt}$ is also known as the _expected value_ ($E$) of the observed response.
+
+\begin{equation}
+E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt})
+(\#eq:GLM-Expectation)
+
+In contemporary Statistics,
+
+Need the two first equations
+
+and this one
+
+\begin{equation}
+$\eta_{ijt}=\beta_0+\beta_1x_{ijt}$ 
+(\#eq:GLM-eta)
+\end{equation}
+
+## Generalized linear mixed models (GLMMs)
+
+$\eta_{ijt}=\beta_0+\beta_1x_{ijt}+\mu_{ij}$ GLMM
+
 # Challenges presented by longitudinal studies

### 2021-11-12 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Updated

--- 2021-11-09 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
@@
-# The contemporary statistical landscape
-
-We consider that a proper understanding of GAMs would be incomplete without a brief survey of contemporary statistical models.
-Particularly because it seems to be a disconnect between biomedical research and the current state of Statistics.
-Today, the Statistical landscape has changed drastically and yet, biomedical research at large continues to be confined to a 't-test', 'ANOVA' mentality.
-The shortcomings that such view produces cannot be properly addressed without a brief mention of statistical methods.
-
-In contemporary Statistics,
-
-## Generalized linear models (GLM)
-
-Need the two first equations
-
-$$y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi)$$
-
-$$E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt})$$
-
-and this one
-
-$\eta_{ijt}=\beta_0+\beta_1x_{ijt}$ This is the GLM
-
-## Generalized linear mixed models (GLMMs)
-
-$\eta_{ijt}=\beta_0+\beta_1x_{ijt}+\mu_{ij}$ GLMM
-
 # Challenges presented by longitudinal studies
@@
 In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-Independent errors $\varepsilon_{ijt}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$ (independently and identically normally distributed with mean zero and variance $\sigma^2$).
+Independent errors $\varepsilon_{ijt}$ represent **random variation from the sampling process ** assumed to be $\sim N(0,\sigma^2)$ (independently and identically normally distributed with mean zero and variance $\sigma^2$).
 In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
@@
 
-### The Linear Mixed Model (LMEM) Case
+### The Linear Mixed Model (LMEM) Case{#LMEM-case}
 

### 2021-11-12 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated

@@
+# Linear Models, and beyond
+
+Although the limitations of rm-ANOVA and LMEMs to analyze non-linear trends in longitudinal data have been presented in the previous section, it is important to understand current state of statistical models to properly introduce GAMs.
+This is important because otherwise, the advances in statistical models over the last 30 years can be overlooked their limitations cannot be properly contextualized.
+
+## Linear Models (LMs) and Linear Mixed Models (LMEMs)
+
+Linear models (LMs) are those that assume a normal (Gaussian) distribution of the response, such as rm-ANOVA.
+These are by far the models most commonly used to analyze data within the biomedical research community, and only incorporate _fixed effects_.
+On the other hand, Linear Mixed Effect Models (LMEMs) are those that assume a normally-distributed response but also incorporate _random effects_, as it has been described before in Section \@ref(LMEM-case).
+
+In matrix notation, a LM has the general form:
+
+\begin{equation}
+$\textbf{y} = \textbf{X} \boldsymbol{\beta}$
+(\#eq:LM-matrix)
+\end{equation}
+
+Where $\textbf{y}$ a vector of observations, $\textbf{X}$ is a matrix that contains the measured values of the covariates, and $\boldsymbol{\beta}$ is a vector of the parameters to be estimated.
+The "model equation" form of Equation \@ref(eq:LM-matrix) for a rm-ANOVA model of two treatment groups corresponds to Equation \@ref(eq:linear-model).
+**do we need to take out the error term from anova?
+confounding lit about this.
+Dobson (p.89 has an error term in the normal linear model, but McCullagh, McCulloch and Stroup don't.
+Or better word that the errors come from the conditional distribution?**
+
+## Generalized Linear Models (GLMs)
+
+A major limitation of LMs is their assumption of normality, making data transformation necessary if it is non-normal.
+However, this procedure can lead to poor model performance [@ohara2010], and can cause problems with the biological interpretation of the model estimates.
+McCullagh and Nelder [@nelder1972] defined General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
+To achieve this, consider the following:
+
+\begin{equation}
+$y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi)$
+(\#eq:GLM-y)
+\end{equation}
+
+Where $y_{ijt}$ is the observation $i$ in group $j$ at time $t$, which is assumed to come randomly from some distribution of the exponential family $\mathcal{D}$, with some mean $\mu_{ijt}$ and a dispersion parameter $\phi$ (which in the Gaussian case is the variance $\sigma^{2}$).
+The mean $\mu_{ijt}$ is also known as the _expected value_ (also called _expectation_) ($E$) of the observed response.
+Then, the _linear predictor_ $\eta$, which defines the relationship between the mean and the covariates can be defined as
+
+\begin{equation}
+$\eta_{ijt}=\beta_0+\beta_1x_{ijt}$ 
+(\#eq:GLM-eta)
+\end{equation}
+
+Where $\eta_{ijt}$ is the linear predictor for each observation.
+Both $\beta_0$ (the intercept), and $\beta_1$ are the model parameters; and $x_{ijt}$ represents the covariates from each group at each time point.
+
+Finally,
+
+\begin{equation}
+$E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt})$
+(\#eq:GLM-Expectation)
+\end{equation}
+Where $E(y_{ijt})$ is the expectation, $g^-{1}$ is the inverse of the _link function_, which relates the parameters of the distribution to the linear combination of the  predictors in Equation \@ref(eq:GLM-eta).
+This shows that LMs (such as rm-ANOVA) are a special case of GLMs.
+
+## Generalized linear mixed models (GLMMs)
+
+Although GLMs relax the normality assumption, they only accomodate fixed effects.
+Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate _random effects_, which have an associated probability distribution.
+Therefore, in GLMMs the _linear predictor_ takes the form
+
+\begin{equation}
+$\eta_{ijt}=\beta_0+\beta_1x_{ijt}+\mu_{ij}$
+(\#eq:GLMM-eta)
+\end{equation}
+
+Where 
 # GAMs as a special case of Generalized Linear Models{#GAM-theory}

### 2021-11-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
updated GAM section

@@
 The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data.\textsuperscript{\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}} Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}} And even more importantly, LMEMs also assume by default a linear relationship between the response and time,\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} making them unsuitable to analyze data non-linear with non-linear trends (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \{\#GAM-theory\}) .
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}} And even more importantly, LMEMs also assume by default a linear relationship between the response and time\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .
 
@@
 In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
-Independent errors \(\varepsilon_{ijt}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
+Independent errors \(\varepsilon_{ijt}\) represent \textbf{random variation from the sampling process } assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
 In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug,'' or ``saline'' vs.~``chemotherapy'').
@@
 
-\hypertarget{the-linear-mixed-model-lmem-case}{%
-\subsubsection{The Linear Mixed Model (LMEM) Case}\label{the-linear-mixed-model-lmem-case}}
+\hypertarget{LMEM-case}{%
+\subsubsection{The Linear Mixed Model (LMEM) Case}\label{LMEM-case}}
 
@@
 Specifically, note that the fitted mean response of both models shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.
+
 The LMEM is only able to account for between-subject variation by providing estimates for each subject (Figure \ref{fig:l-q-response}F), but both models are unable to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare Figure \ref{fig:l-q-response}D with Figure \ref{fig:l-q-response}E, F).
@@
 Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
+It could be argued that a LMEM with quadratic effects should have been used to fit the data in Figure\ref{fig:l-q-response}, F.
+However, because in reality the true function is not known, choosing a polynomial degree causes more questions (e.g., is it quadratic?, cubic?, or a higher degree?).
+Additionally, polynomial effects have other limitations, which we cover in Section \ref{GAM-theory}.
 
 This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-The models fitted to the simulated data were an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
-In the following section, we present generalized additive models (GAMs) as a data-driven alternative method to analyze longitudinal non-linear data that overcomes the linearity assumption.
+The models fitted to the simulated data were an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response, which is only noticeable if the model fit and the response are visualized.
+In the following section, we provide a brief overview of linear models, general linear models and generalized linear mixed models before presenting the theory of GAMs.
+
+\hypertarget{linear-models-and-beyond}{%
+\section{Linear Models, and beyond}\label{linear-models-and-beyond}}
+
+In reality, rm-ANOVA and LMEMs are just \emph{special cases} of broader model classes (General Linear Models and Generalized Linear Mixed Models, respectively).
+Understanding this is critical to fully capture the constraints of such models and to understand how GAMs overcome those limitations.
+This section will briefly present the different Statistical model classes, indicate how rm-ANOVA and LMEMs fit within that framework and introduce the theory of GAMs.
+
+\hypertarget{linear-models-lms-and-linear-mixed-models-lmems}{%
+\subsection{Linear Models (LMs) and Linear Mixed Models (LMEMs)}\label{linear-models-lms-and-linear-mixed-models-lmems}}
+
+Linear models (LMs) are those that assume a normal (Gaussian) distribution of the response, such as rm-ANOVA.
+These are by far the models most commonly used to analyze data within the biomedical research community, and only incorporate \emph{fixed effects}.
+On the other hand, Linear Mixed Effect Models (LMEMs) are those that assume a normally-distributed response but also incorporate \emph{random effects}, as it has been described in Section \ref{LMEM-case}.
+
+In matrix notation, a LM has the general form:
+
+\begin{equation}
+\textbf{y} = \textbf{X} \boldsymbol{\beta}
+\label{eq:LM-matrix}
+\end{equation}
+
+Where \(\textbf{y}\) a vector of observations, \(\textbf{X}\) is a matrix that contains the measured values of the covariates, and \(\boldsymbol{\beta}\) is a vector of the parameters to be estimated.
+The ``model equation'' form of Equation \eqref{eq:LM-matrix} for a rm-ANOVA model of two treatment groups corresponds to Equation \eqref{eq:linear-model}.
+It is important to remember that the error term \(\varepsilon_{ijt}\) in Equation \eqref{eq:linear-model} is due to random sampling from the normal distribution.
+
+\hypertarget{generalized-linear-models-glms}{%
+\subsection{Generalized Linear Models (GLMs)}\label{generalized-linear-models-glms}}
+
+A major limitation of LMs is their assumption of normality.
+If the data is non-normal, a transformation is necessary in order to properly fit the model.
+However, transformation lead to poor model performance,\textsuperscript{\protect\hyperlink{ref-ohara2010}{51}} and can cause problems with the biological interpretation of the model estimates.
+McCullagh and Nelder\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} defined General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
+To achieve this, consider the following:
+
+\begin{equation}
+y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi)
+\label{eq:GLM-y}
+\end{equation}
+
+Where \(y_{ijt}\) is the observation \(i\) in group \(j\) at time \(t\), which is assumed to come randomly from some distribution of the exponential family \(\mathcal{D}\), with some mean \(\mu_{ijt}\) and a dispersion parameter \(\phi\) (which in the Gaussian case is the variance \(\sigma^{2}\)).
+The mean \(\mu_{ijt}\) is also known as the \emph{expected value} (also called \emph{expectation}) (\(E\)) of the observed response.
+Then, the \emph{linear predictor} \(\eta\), which defines the relationship between the mean and the covariates can be defined as
+
+\begin{equation}
+\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}
+\label{eq:GLM-eta}
+\end{equation}
+
+Where \(\eta_{ijt}\) is the linear predictor for each observation, \(\beta_0\) is the intercept, and \(\beta_{j}\) is the model parameter for each group; and \(x_{ijt}\) represents the covariates from each group at each time point.
+
+Finally,
+
+\begin{equation}
+E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt})
+\label{eq:GLM-Expectation}
+\end{equation}
+Where \(E(y_{ijt})\) is the expectation, \(g^-{1}\) is the inverse of the \emph{link function}, which relates the parameters of the distribution to the linear combination of the predictors in Equation \eqref{eq:GLM-eta}.
+Therefore, LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
+
+\hypertarget{generalized-linear-mixed-models-glmms}{%
+\subsection{Generalized linear mixed models (GLMMs)}\label{generalized-linear-mixed-models-glmms}}
+
+Although GLMs relax the normality assumption, they can only accommodate fixed effects.
+Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate \emph{random effects}, which have an associated probability distribution.\textsuperscript{\protect\hyperlink{ref-mcculloch2001}{53}} Therefore, in GLMMs the \emph{linear predictor} takes the form
+
+\begin{equation}
+\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\mu_{ij}
+\label{eq:GLMM-eta}
+\end{equation}
+
+Where \(\mu_{ij}\) corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of the previous equations.
+Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed.\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in refs\textsuperscript{\protect\hyperlink{ref-dobson2008}{54}} and.\textsuperscript{\protect\hyperlink{ref-stroup2013}{55}}
 
 \hypertarget{GAM-theory}{%
-\section{GAMs as a special case of Generalized Linear Models}\label{GAM-theory}}
+\subsubsection{GAMs as a special case of Generalized Linear Models}\label{GAM-theory}}
 
 \hypertarget{gams-and-basis-functions}{%
-\subsection{GAMs and Basis Functions}\label{gams-and-basis-functions}}
-
-Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors.\textsuperscript{\protect\hyperlink{ref-nelder1972}{51}} In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{52}}.
-A GAM model can be written as:
+\paragraph{GAMs and Basis Functions}\label{gams-and-basis-functions}}
+
+Generalized additive models (GAMs) are a family of regression-based methods for estimating smoothly varying trends and are an extension of the GLM family.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}} A GAM version of a linear model can be written as:
 
@@
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.
+Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and where the change of \(y_{ijt}\) over time instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta} is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.
 
@@
 A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{53}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
 In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{54}} Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{58}} Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
 
 Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{52}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate.\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{55}}
-
-In contrast, thin regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner.\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see refs.\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} and).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
+The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{56}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate.\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}
+
+In contrast, thin regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see refs.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} and).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
 
@@
 \hypertarget{a-bayesian-interpretation-of-gams}{%
-\section{A Bayesian interpretation of GAMs}\label{a-bayesian-interpretation-of-gams}}
-
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{56}} In the case of data that shows non-linear trends, the belief that the \emph{true} trend of the data is likely to be smooth rather than extremely ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{57}} Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (point-wise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
+\subsubsection{A Bayesian interpretation of GAMs}\label{a-bayesian-interpretation-of-gams}}
+
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}} In the case of data that shows non-linear trends, the belief that the \emph{true} trend of the data is likely to be smooth rather than extremely ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (point-wise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
 
@@
 For GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, the percentage of CIs entirely containing the true function over the entire time period would be close to the nominal value (i.e., if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function), thereby allowing more robust estimates from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{57}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{58}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{61}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{62}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
 
@@
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in.\textsuperscript{\protect\hyperlink{ref-hefley2017}{53}}
+A detailed description on basis functions and correlations can be found in.\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}}
 
@@
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{59}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{60}}
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{63}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{64}}
 
@@
 quadratic?, cubic?, what is the biological interpretation of a polynomial effect?).
-Moreover, polynomial effects have other associated issues that we have described in Section \{\#GAM-theory\}.
-
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{61}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.
+Moreover, polynomial effects have other associated issues that we have described in Section \ref{GAM-theory}.
+
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{65}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.
 
 In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{53}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{62}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{66}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).
 
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{63},\protect\hyperlink{ref-weissgerber2018}{64}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{67},\protect\hyperlink{ref-weissgerber2018}{68}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-ohara2010}{}}%
+\CSLLeftMargin{51. }
+\CSLRightInline{OHara R, Kotze J. Do not log-transform count data. Published online January 2010. doi:\href{https://doi.org/10.1038/npre.2010.4136.1}{10.1038/npre.2010.4136.1}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-nelder1972}{}}%
-\CSLLeftMargin{51. }
+\CSLLeftMargin{52. }
 \CSLRightInline{Nelder JA, Wedderburn RWM.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-mcculloch2001}{}}%
+\CSLLeftMargin{53. }
+\CSLRightInline{McCulloch C. \emph{Generalized, Linear, and Mixed Models}. John Wiley \& Sons; 2001.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-dobson2008}{}}%
+\CSLLeftMargin{54. }
+\CSLRightInline{Dobson A. \emph{An Introduction to Generalized Linear Models}. CRC Press; 2008.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-stroup2013}{}}%
+\CSLLeftMargin{55. }
+\CSLRightInline{Stroup W. \emph{Generalized Linear Mixed Models : Modern Concepts, Methods and Applications}. CRC Press, Taylor \& Francis Group; 2013.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-hastie1987}{}}%
-\CSLLeftMargin{52. }
+\CSLLeftMargin{56. }
 \CSLRightInline{Hastie T, Tibshirani R.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hefley2017}{}}%
-\CSLLeftMargin{53. }
+\CSLLeftMargin{57. }
 \CSLRightInline{Hefley TJ, Broms KM, Brost BM, et al.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wegman1983}{}}%
-\CSLLeftMargin{54. }
+\CSLLeftMargin{58. }
 \CSLRightInline{Wegman EJ, Wright IW.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wood2003}{}}%
-\CSLLeftMargin{55. }
+\CSLLeftMargin{59. }
 \CSLRightInline{Wood SN. Thin plate regression splines. 2003;65(1):95-114. doi:\href{https://doi.org/10.1111/1467-9868.00374}{10.1111/1467-9868.00374}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-mcelreath2018}{}}%
-\CSLLeftMargin{56. }
+\CSLLeftMargin{60. }
 \CSLRightInline{McElreath R. \emph{Statistical Rethinking}. Chapman; Hall/{CRC}; 2018. doi:\href{https://doi.org/10.1201/9781315372495}{10.1201/9781315372495}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-miller2019}{}}%
-\CSLLeftMargin{57. }
+\CSLLeftMargin{61. }
 \CSLRightInline{Miller DL. Bayesian views of generalized additive modelling. \emph{arXiv preprint arXiv:190201330}. Published online 2019.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
-\CSLLeftMargin{58. }
+\CSLLeftMargin{62. }
 \CSLRightInline{Marra G, Wood SN.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
-\CSLLeftMargin{59. }
+\CSLLeftMargin{63. }
 \CSLRightInline{Simpson GL. \emph{Gratia: Graceful 'Ggplot'-Based Graphics and Other Functions for GAMs Fitted Using 'Mgcv'}.; 2020. \url{https://CRAN.R-project.org/package=gratia}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
-\CSLLeftMargin{60. }
+\CSLLeftMargin{64. }
 \CSLRightInline{Harezlak J, Ruppert D, Wand MP. \emph{Semiparametric Regression with r}. Springer New York; 2018. doi:\href{https://doi.org/10.1007/978-1-4939-8853-2}{10.1007/978-1-4939-8853-2}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
-\CSLLeftMargin{61. }
+\CSLLeftMargin{65. }
 \CSLRightInline{Lang TA, Altman DG.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hastie1995}{}}%
-\CSLLeftMargin{62. }
+\CSLLeftMargin{66. }
 \CSLRightInline{Hastie T, Tibshirani R.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-begley2015}{}}%
-\CSLLeftMargin{63. }
+\CSLLeftMargin{67. }
 \CSLRightInline{Begley CG, Ioannidis JPA.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-weissgerber2018}{}}%
-\CSLLeftMargin{64. }
+\CSLLeftMargin{68. }
 \CSLRightInline{Weissgerber TL, Garcia-Valencia O, Garovic VD, Milic NM, Winham SJ.

### 2021-11-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/01-Background.Rmd
updated GAM section

@@
 However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent [@gueorguieva2004;@barr2013].
-And even more importantly, LMEMs also assume by default a linear relationship between the response and time [@pinheiro2006] (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section {#GAM-theory}) .
+And even more importantly, LMEMs also assume by default a linear relationship between the response and time [@pinheiro2006] (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \@ref(GAM-theory)) .
 

### 2021-11-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
updated GAM section

@@
 Specifically, note that the fitted mean response of both models shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.
+
 The LMEM is only able to account for between-subject variation by providing estimates for each subject (Figure \@ref(fig:l-q-response)F), but both models are unable to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare Figure \@ref(fig:l-q-response)D  with Figure \@ref(fig:l-q-response)E, F).
@@
 Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
+It could be argued that a LMEM with quadratic effects should have been used to fit the data in Figure\@ref(fig:l-q-response), F.
+However, because in reality the true function is not known, choosing a polynomial degree causes more questions (e.g., is it quadratic?, cubic?, or a higher degree?).
+Additionally, polynomial effects have other limitations, which we cover in Section \@ref(GAM-theory).
 
 This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-The models fitted to the simulated data were an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
-In the following section, we present generalized additive models (GAMs) as a data-driven  alternative method to analyze longitudinal non-linear data that overcomes the linearity assumption.
+The models fitted to the simulated data were an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response, which is only noticeable if the model fit and the response are visualized.
+In the following section, we provide a brief overview of linear models, general linear models and generalized linear mixed models before presenting the theory of GAMs.

### 2021-11-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
updated GAM section

--- 2021-11-09 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
@@
-# GAMs as a special case of Generalized Linear Models{#GAM-theory}
+# Linear Models, and beyond
 
-## GAMs and Basis Functions
+In reality, rm-ANOVA and LMEMs are just _special cases_ of broader model classes (General Linear Models and Generalized Linear Mixed Models, respectively).
+Understanding this is critical to fully capture the constraints of such models and to understand how GAMs overcome those limitations.
+This section will briefly present the different Statistical model classes, indicate how rm-ANOVA and LMEMs fit within that framework and introduce the theory of GAMs.
 
-Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors [@nelder1972].
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case[@simpson2018;@wood2017;@hastie1987].
-A GAM model can be written as:
+## Linear Models (LMs) and Linear Mixed Models (LMEMs)
+
+Linear models (LMs) are those that assume a normal (Gaussian) distribution of the response, such as rm-ANOVA.
+These are by far the models most commonly used to analyze data within the biomedical research community, and only incorporate _fixed effects_.
+On the other hand, Linear Mixed Effect Models (LMEMs) are those that assume a normally-distributed response but also incorporate _random effects_, as it has been described in Section \@ref(LMEM-case).
+
+In matrix notation, a LM has the general form:
+
+\begin{equation}
+\textbf{y} = \textbf{X} \boldsymbol{\beta}
+(\#eq:LM-matrix)
+\end{equation}
+
+Where $\textbf{y}$ a vector of observations, $\textbf{X}$ is a matrix that contains the measured values of the covariates, and $\boldsymbol{\beta}$ is a vector of the parameters to be estimated.
+The "model equation" form of Equation \@ref(eq:LM-matrix) for a rm-ANOVA model of two treatment groups corresponds to Equation \@ref(eq:linear-model).
+It is important to remember that the error term $\varepsilon_{ijt}$ in Equation \@ref(eq:linear-model) is due to random sampling from the normal distribution.
+
+## Generalized Linear Models (GLMs)
+
+A major limitation of LMs is their assumption of normality.
+If the data is non-normal, a transformation is necessary in order to properly fit the model.
+However, transformation lead to poor model performance [@ohara2010], and can cause problems with the biological interpretation of the model estimates.
+McCullagh and Nelder [@nelder1972] defined General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
+To achieve this, consider the following:
+
+\begin{equation}
+y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi)
+(\#eq:GLM-y)
+\end{equation}
+
+Where $y_{ijt}$ is the observation $i$ in group $j$ at time $t$, which is assumed to come randomly from some distribution of the exponential family $\mathcal{D}$, with some mean $\mu_{ijt}$ and a dispersion parameter $\phi$ (which in the Gaussian case is the variance $\sigma^{2}$).
+The mean $\mu_{ijt}$ is also known as the _expected value_ (also called _expectation_) ($E$) of the observed response.
+Then, the _linear predictor_ $\eta$, which defines the relationship between the mean and the covariates can be defined as
+
+\begin{equation}
+\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}
+(\#eq:GLM-eta)
+\end{equation}
+
+Where $\eta_{ijt}$ is the linear predictor for each observation, $\beta_0$ is the intercept, and $\beta_{j}$ is the model parameter for each group; and $x_{ijt}$ represents the covariates from each group at each time point.
+
+Finally,
+
+\begin{equation}
+E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt})
+(\#eq:GLM-Expectation)
+\end{equation}
+Where $E(y_{ijt})$ is the expectation, $g^-{1}$ is the inverse of the _link function_, which relates the parameters of the distribution to the linear combination of the  predictors in Equation \@ref(eq:GLM-eta).
+Therefore, LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
+
+## Generalized linear mixed models (GLMMs)
+
+Although GLMs relax the normality assumption, they can only accommodate fixed effects.
+Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate _random effects_, which have an associated probability distribution [@mcculloch2001].
+Therefore, in GLMMs the _linear predictor_ takes the form
+
+\begin{equation}
+\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\mu_{ij}
+(\#eq:GLMM-eta)
+\end{equation}
+
+Where $\mu_{ij}$ corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of the previous equations.
+Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed [@nelder1972].
+In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in refs [@dobson2008] and [@stroup2013].
+
+### GAMs as a special case of Generalized Linear Models{#GAM-theory}
+
+#### GAMs and Basis Functions
+
+Generalized additive models (GAMs) are a family of regression-based methods for estimating smoothly varying trends and are an extension of the GLM family [@simpson2018;@wood2017;@hastie1987].
+A GAM version of a linear model can be written as:
 
@@
 
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the _smooth function_ $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual error.
+Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, and where the change of $y_{ijt}$ over time instead of being defined by $\beta_{j}x_{ijt}$ as in Equation \@ref(eq:GLM-eta) is represented by the _smooth function_ $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual error.
 
@@
 
-# A Bayesian interpretation of GAMs
+### A Bayesian interpretation of GAMs
 

### 2021-11-16 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.Rmd
Updated sections

@@
 We appreciate the reviewer's perspective regarding the somewhat contradictory argument of fitting a linear effects LMM to quadratic data.
-However, this is an intentional approach because we follow the statistical analysis logic followed by most biomedical researchers: Fit a LMM or an rm-ANOVA, and if you get significant _p-values_ that's good enough.
-The purpose of the figure is to convey the point that visualizing the model fit is something necessary, but that is rarely done in the field.
-**Incorporate in the section that a strenght of GAMs is that they can learn the function from the data, while LMMs can't.**
+However, this is an intentional approach because we follow the statistical analysis logic followed by most biomedical researchers to analyze longitudinal data: Fit a LMM or an rm-ANOVA, and if you get significant _p-values_ that's good enough.
+The purpose of the figure is to convey that changing models (from rm-ANOVA to LMEM) without visualizing the model fit is in this case is something that will propagate bias in the results and does not provide any substantial improvement.
+However, we acknowledge that this point was not explicitly stated in the original manuscript and the following changes have been made:
+
+- L95: We mention that LMEMs assume a linear relationship "by default", but we indicate that polynomial effects can be used as well, though they have their own shortcomings.
+
+- L119-L120 now reads "the limitations of LMEMs and rm-ANOVA regarding an expected trend in the response", instead of "linearity of response" (L119 in the original manuscript).
+
+- L136 now reads "assumed relationship" instead of  "linear relationship" (L135 in the original manuscript).
+
+- Section 3.2 now is titled "Assumed relationship" instead of "Linear relationship"
+
+- L181-L184 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
+
+- We provide additional context on the purpose of using linear models to fit quadratic data in L250-L251 which now read "We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions."
+
+We believe that the aforementioned change makes a clear statement to the reader about our intentions in Fig.1.
+We want to emphasize that strictly speaking the term "default" might not be correct, but we are stretching the language to make the point that doing a simple switch from rm-ANOVA to a LMEMs (in R) without any additional changes results in the assumption of linear effects that still provide biased estimates.
+
+- L273-276 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
+1.
+Specifically, that in reality the true function is not known and that this complicates choosing the degree of the polynomial.
+We also reference Section 4.3.1, which provides more detail on the limitation of polynomial effects.
+
+- In Section 6 (Discussion)  we emphasize again the rationale for not using polynomial effects on the LMEM in Fig. 1 (L509-513).
 
@@
 
-# GLMs, GAMs, and conditional distributions
+## Reviewer's Comments
+
+You don’t even mention generalized linear mixed models; why?
+In the context of the tutorial they represent the logical bridge between GAM and LMM, especially when you consider that you are really fitting GAMMs here.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+- We appreciate Dr.
+Simpson's comment here and the equations provided, which we have now incorporated in the manuscript..
+We acknowledge that a brief overview of the different statistical model classes was missing in the original manuscript, and we have now extensively edited Section 4 to provide context on LMs, GLMs and GLMMs before introducing GAMs.
+The changes now appear between L282 (where Section 4 has been renamed to "Linear Models and Beyond") and L321.
+Because the different model classifications are not commonly known in the biomedical field, we contextualize the previously presented rm-ANOVA and LMEM into their respective classes to provide the reader with a clear understanding of where each model fits and also of where GAMs fit within this framework.
+We believe this properly addressed the reviewer comments.
+
+- We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the **linear case**.
+This change will avoid the propagation of confusion that Dr.
+Simpson correctly indicated in his comment.
+
+- We have also changed L328 to indicate that the term $\varepsilon_{ijt}$ is the residual, and not a separable error.
+Of notice, we are following the notation provided by Dr.
+Simpson in Eq.
+3 in the paper "Modelling Modelling Palaeoecological Time Series Using Generalised Additive Models".
+**do we need this here?
+I am adding it because we are following the same model specification that he has in this paper**
 
@@
 
-You talk a little about “wiggliness” and penalising the weights for each basis function towards 0 but you don’t really explain this most crucial concept of the model fitting problem and why modern GAMs are so much better than the
-GAMs developed by Hastie and Tibshirani when they first introduced GAMs to the world.
+You talk a little about “wiggliness” and penalising the weights for each basis function towards 0 but you don’t really explain this most crucial concept of the model fitting problem and why modern GAMs are so much better than the GAMs developed by Hastie and Tibshirani when they first introduced GAMs to the world.
 You really need to define wiggliness beyond something that is used to avoid overfitting.
@@
 By glossing over these important concepts, the reader is left wondering what wiggliness is, how we measure overfit? etc.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
@@
 ## Reviewer's Comments
+
+The essence of this is not “if the difference between [credible] intervals of the fitted smooths… is non-zero”.
+For a variety of reasons this is not how it works.
+Instead we compute the difference of the smooths and then compute a credible interval around this difference.
+Also, to interpret this correctly we really need to do a simultaneous interval as you are looking at all points over the range of the covariate.
+If we just use the usual interval we have to consider the under and over coverage of the Nychka across-the-function intervals.
+You should talk about this issue in this section if you don’t compute simultaneous intervals.
+This section could also do with a little more explanation; {gratia} (and hence I) tens to consider differences of smooths to exclude the groups means.
+Here, if I read the paper and the code correctly, you included the group means.
+This works OK for simple models like this, but what if you had two or more parametric factor terms in the model?
+Then you would not (at least by default) have any way to isolate the group means for the by factor independently of the levels of the other factor.
+You would have to create differences conditional upon a level of
+the other factor. Users, however, tend to expect differences to include the group means, as you have chosen to present things here. Hence a note exp
 
@@
 
-The figures are too tall — the aspect ratio used for the individual panels is wrong. You should be emphasising the time axis so you should have individual panels that are wider than they are tall.
+The figures are too tall — the aspect ratio used for the individual panels is wrong. You should be emphasizing the time axis so you should have individual panels that are wider than they are tall.
 
@@
 
-We appreciate this comment regarding figure aesthetitcs, we have now made the figures wider to emphasize the time axis.
+We appreciate this comment regarding figure aesthetics, we have now made the figures wider to emphasize the time axis.
 
@@
 It might seem too basic or repetitive, but we are trying to convince the reader that simulation can indeed produce synthetic data that can be useful.
+
+# Figures
+
+## Reviewer's Comments
+
+The figures are too tall — the aspect ratio used for the individual panels is wrong. You should be emphasising the time axis so you should have individual panels that are wider than they are tall.
+
+Figure 3: on panel A, are the solid lines the truth in the main panel?
+If they are, you don’t need the inset plot, and if they aren’t, are they the mean of the simulated observations?
+If they are this latter, I think this is superfluous and it would be better to make the solid line the truth, about which you simulated noisy data.
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We have changed the aspect of the figures to emphasize the time axis, making them wider.
+
+On Figure 3, we have added the original data to panel A, which now displays both the simulated data (thin lines) and the original data (truth) in order to simplify the figure.
+We have updated also places in the manuscript that reference this Figure as well.
 
@@
 To this day, the field values "significance" (defined by the arbitrarily set threshold of _p_<0.05).
-We are trying to convince the reader that low _p-values_ are possible with ill fit models using language that is understandable.
+We are trying to convince the reader that low _p-values_ are possible with ill-fitted models, and we use language that is understandable by the reader.
 
@@
 ## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+- We have changed the sentence to "the distribution of the random effects, which need to be independent", which also indicates that errors need not be normally-distributed.
 
@@
 This line has been removed from the revised manuscript.
+However, we acknowledge that we didn't clearly conveyed this point to the read.
+In Section 6 (Discussion, L525-526 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided.
 

### 2021-11-16 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.tex
Updated sections

@@
 We appreciate the reviewer's perspective regarding the somewhat contradictory argument of fitting a linear effects LMM to quadratic data.
-However, this is an intentional approach because we follow the statistical analysis logic followed by most biomedical researchers: Fit a LMM or an rm-ANOVA, and if you get significant \emph{p-values} that's good enough.
-The purpose of the figure is to convey the point that visualizing the model fit is something necessary, but that is rarely done in the field.
-\textbf{Incorporate in the section that a strenght of GAMs is that they can learn the function from the data, while LMMs can't.}
+However, this is an intentional approach because we follow the statistical analysis logic followed by most biomedical researchers to analyze longitudinal data: Fit a LMM or an rm-ANOVA, and if you get significant \emph{p-values} that's good enough.
+The purpose of the figure is to convey that changing models (from rm-ANOVA to LMEM) without visualizing the model fit is in this case is something that will propagate bias in the results and does not provide any substantial improvement.
+However, we acknowledge that this point was not explicitly stated in the original manuscript and the following changes have been made:
+
+\begin{itemize}
+\item
+  L95: We mention that LMEMs assume a linear relationship ``by default'', but we indicate that polynomial effects can be used as well, though they have their own shortcomings.
+\item
+  L119-L120 now reads ``the limitations of LMEMs and rm-ANOVA regarding an expected trend in the response'', instead of ``linearity of response'' (L119 in the original manuscript).
+\item
+  L136 now reads ``assumed relationship'' instead of ``linear relationship'' (L135 in the original manuscript).
+\item
+  Section 3.2 now is titled ``Assumed relationship'' instead of ``Linear relationship''
+\item
+  L181-L184 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
+\item
+  We provide additional context on the purpose of using linear models to fit quadratic data in L250-L251 which now read ``We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with''default'' (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.''
+\end{itemize}
+
+We believe that the aforementioned change makes a clear statement to the reader about our intentions in Fig.1.
+We want to emphasize that strictly speaking the term ``default'' might not be correct, but we are stretching the language to make the point that doing a simple switch from rm-ANOVA to a LMEMs (in R) without any additional changes results in the assumption of linear effects that still provide biased estimates.
+
+\begin{itemize}
+\item
+  L273-276 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
+1.
+Specifically, that in reality the true function is not known and that this complicates choosing the degree of the polynomial.
+We also reference Section 4.3.1, which provides more detail on the limitation of polynomial effects.
+\item
+  In Section 6 (Discussion) we emphasize again the rationale for not using polynomial effects on the LMEM in Fig. 1 (L509-513).
+\end{itemize}
 
@@
 
-\hypertarget{glms-gams-and-conditional-distributions}{%
-\section{GLMs, GAMs, and conditional distributions}\label{glms-gams-and-conditional-distributions}}
+\hypertarget{reviewers-comments-2}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-2}}
+
+You don't even mention generalized linear mixed models; why?
+In the context of the tutorial they represent the logical bridge between GAM and LMM, especially when you consider that you are really fitting GAMMs here.
+
+\hypertarget{section-3}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-3}}
+
+\begin{itemize}
+\item
+  We appreciate Dr.~Simpson's comment here and the equations provided, which we have now incorporated in the manuscript..
+We acknowledge that a brief overview of the different statistical model classes was missing in the original manuscript, and we have now extensively edited Section 4 to provide context on LMs, GLMs and GLMMs before introducing GAMs.
+The changes now appear between L282 (where Section 4 has been renamed to ``Linear Models and Beyond'') and L321.
+Because the different model classifications are not commonly known in the biomedical field, we contextualize the previously presented rm-ANOVA and LMEM into their respective classes to provide the reader with a clear understanding of where each model fits and also of where GAMs fit within this framework.
+We believe this properly addressed the reviewer comments.
+\item
+  We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the \textbf{linear case}.
+This change will avoid the propagation of confusion that Dr.~Simpson correctly indicated in his comment.
+\item
+  We have also changed L328 to indicate that the term \(\varepsilon_{ijt}\) is the residual, and not a separable error.
+Of notice, we are following the notation provided by Dr.~Simpson in Eq.
+3 in the paper ``Modelling Modelling Palaeoecological Time Series Using Generalised Additive Models''.
+\textbf{do we need this here?
+I am adding it because we are following the same model specification that he has in this paper}
+\end{itemize}
 
@@
 
-\hypertarget{reviewers-comments-2}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-2}}
+\hypertarget{reviewers-comments-3}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-3}}
 
@@
 
-\hypertarget{section-3}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-3}}
+\hypertarget{section-4}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-4}}
 
@@
 
-\hypertarget{reviewers-comments-3}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-3}}
+\hypertarget{reviewers-comments-4}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-4}}
 
@@
 
-\hypertarget{section-4}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-4}}
+\hypertarget{section-5}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-5}}
 
@@
 
-\hypertarget{reviewers-comments-4}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-4}}
+\hypertarget{reviewers-comments-5}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-5}}
 
@@
 
+\hypertarget{section-6}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-6}}
+
 \hypertarget{bayesian}{%
@@
 
-\hypertarget{reviewers-comments-5}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-5}}
+\hypertarget{reviewers-comments-6}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-6}}
 
@@
 
-\hypertarget{section-5}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-5}}
+\hypertarget{section-7}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-7}}
 
@@
 
-\hypertarget{reviewers-comments-6}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-6}}
+\hypertarget{reviewers-comments-7}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-7}}
 
@@
 
-\hypertarget{section-6}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-6}}
+\hypertarget{section-8}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-8}}
 
@@
 
-\hypertarget{reviewers-comments-7}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-7}}
+\hypertarget{reviewers-comments-8}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-8}}
 
-\hypertarget{section-7}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-7}}
+\hypertarget{section-9}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-9}}
 
@@
 
-\hypertarget{reviewers-comments-8}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-8}}
+\hypertarget{reviewers-comments-9}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-9}}
 
@@
 
-\hypertarget{section-8}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-8}}
+\hypertarget{section-10}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-10}}
 
@@
 
-\hypertarget{reviewers-comments-9}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-9}}
+\hypertarget{reviewers-comments-10}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-10}}
 
@@
 
-\hypertarget{section-9}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-9}}
+\hypertarget{section-11}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-11}}
 
@@
 
-\hypertarget{reviewers-comments-10}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-10}}
+\hypertarget{reviewers-comments-11}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-11}}
 
@@
 
-\hypertarget{section-10}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-10}}
+\hypertarget{section-12}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-12}}
 
-\hypertarget{reviewers-comments-11}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-11}}
+\hypertarget{reviewers-comments-12}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-12}}
 
@@
 
-\hypertarget{section-11}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-11}}
+\hypertarget{section-13}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-13}}
 
@@
 
-\hypertarget{reviewers-comments-12}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-12}}
-
-The figures are too tall --- the aspect ratio used for the individual panels is wrong. You should be emphasising the time axis so you should have individual panels that are wider than they are tall.
-
-\hypertarget{section-12}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-12}}
-
-We appreciate this comment regarding figure aesthetitcs, we have now made the figures wider to emphasize the time axis.
-
 \hypertarget{reviewers-comments-13}{%
 \subsection{Reviewer's Comments}\label{reviewers-comments-13}}
+
+The figures are too tall --- the aspect ratio used for the individual panels is wrong. You should be emphasizing the time axis so you should have individual panels that are wider than they are tall.
+
+\hypertarget{section-14}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-14}}
+
+We appreciate this comment regarding figure aesthetics, we have now made the figures wider to emphasize the time axis.
+
+\hypertarget{reviewers-comments-14}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-14}}
 
@@
 
-\hypertarget{section-13}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-13}}
+\hypertarget{section-15}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-15}}
 
@@
 
-\hypertarget{reviewers-comments-14}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-14}}
+\hypertarget{reviewers-comments-15}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-15}}
 
@@
 
-\hypertarget{section-14}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-14}}
+\hypertarget{section-16}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-16}}
 
@@
 To this day, the field values ``significance'' (defined by the arbitrarily set threshold of \emph{p}\textless0.05).
-We are trying to convince the reader that low \emph{p-values} are possible with ill fit models using language that is understandable.
+We are trying to convince the reader that low \emph{p-values} are possible with ill-fitted models, and we use language that is understandable by the reader.
 
-\hypertarget{reviewers-comments-15}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-15}}
+\hypertarget{reviewers-comments-16}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-16}}
 
@@
 
-\hypertarget{section-15}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-15}}
+\hypertarget{section-17}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-17}}
 
@@
 
-\hypertarget{reviewers-comments-16}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-16}}
+\hypertarget{reviewers-comments-17}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-17}}
 
@@
 
-\hypertarget{section-16}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-16}}
+\hypertarget{section-18}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-18}}
 
-\hypertarget{reviewers-comments-17}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-17}}
-
-L185 here and throughout, check for situations where you have combined numerical citations to literature within narrative text; Here it should be ``\ldots see refs 40 \& 42.)'' for example, not superscript numerals.
-
-\hypertarget{section-17}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-17}}
-
-\textbf{leave this to the end as it can be fixed in the final .tex document}
+\begin{itemize}
+\tightlist
+\item
+  We have changed the sentence to ``the distribution of the random effects, which need to be independent'', which also indicates that errors need not be normally-distributed.
+\end{itemize}
 
@@
 
-L244 ``global'' is perhaps more conventionally described as ``population'' in the mixed model literature.
+L185 here and throughout, check for situations where you have combined numerical citations to literature within narrative text; Here it should be ``\ldots see refs 40 \& 42.)'' for example, not superscript numerals.
 
-\hypertarget{section-18}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-18}}
+\hypertarget{section-19}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-19}}
 
-We have changed the word from ``global'' to ``population''.
+\textbf{leave this to the end as it can be fixed in the final .tex document}
 
@@
 
-L283 I think it is better to say ``\ldots the mathematical space within which the true but unknown \(f(x_t)\) is thought to exist.''
+L244 ``global'' is perhaps more conventionally described as ``population'' in the mixed model literature.
 
-\hypertarget{section-19}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-19}}
+\hypertarget{section-20}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-20}}
 
-We have updated this line to ``the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist'', in L290 in the revised manuscript.
+We have changed the word from ``global'' to ``population''.
 
@@
 
-L293 You never really describe what ``data-drive flexibility'' is. Why are GAMs data driven and LMMs and GLMs not? This is a critical concept to convey I feel.
+L283 I think it is better to say ``\ldots the mathematical space within which the true but unknown \(f(x_t)\) is thought to exist.''
 
-\hypertarget{section-20}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-20}}
+\hypertarget{section-21}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-21}}
 
-This line has been removed from the revised manuscript.
+We have updated this line to ``the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist'', in L290 in the revised manuscript.
 
@@
 
-L351 Better to say ``separate smooths'' not ``independent''.
-These factor-by smooths share a common basis, they just get an entirely separate set of model coeficients and smoothness parameters for each level of the factor.
+L293 You never really describe what ``data-drive flexibility'' is. Why are GAMs data driven and LMMs and GLMs not? This is a critical concept to convey I feel.
 
-\hypertarget{section-21}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-21}}
+\hypertarget{section-22}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-22}}
 
-The sentence now reads ``separate smooths for\ldots{}''
+This line has been removed from the revised manuscript.
+However, we acknowledge that we didn't clearly conveyed this point to the read.
+In Section 6 (Discussion, L525-526 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided.
 
@@
 
-L354 ``difference between group means'' because the smooth encodes differences between the groups too, just not difference in the means. Be a little more specific here.
+L351 Better to say ``separate smooths'' not ``independent''.
+These factor-by smooths share a common basis, they just get an entirely separate set of model coeficients and smoothness parameters for each level of the factor.
 
-\hypertarget{section-22}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-22}}
+\hypertarget{section-23}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-23}}
 
-The sentence now reads ``any differences between the group means''.
+The sentence now reads ``separate smooths for\ldots{}''
 
@@
 
-L356 use straight quotes in code, not curly ones. And speficy the \passthrough{\lstinline!family!} here?
+L354 ``difference between group means'' because the smooth encodes differences between the groups too, just not difference in the means. Be a little more specific here.
 
-\hypertarget{section-23}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-23}}
+\hypertarget{section-24}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-24}}
+
+The sentence now reads ``any differences between the group means''.
 
@@
 
-L359 ``contain the fitted'' instead of ``store the''
+L356 use straight quotes in code, not curly ones. And speficy the \passthrough{\lstinline!family!} here?
 
-\hypertarget{section-24}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-24}}
-
-The sentence now reads ``contains the fitted model''.
+\hypertarget{section-25}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-25}}
 
@@
 
-L360 Not ``knots'' - there are k-1 basis functions, but as described above, because this is a TPRS there are knots at the unique data values and then we decompose the full basis and take the required eigenvectors as the basis to use.
+L359 ``contain the fitted'' instead of ``store the''
 
-\hypertarget{section-25}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-25}}
+\hypertarget{section-26}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-26}}
 
-The updated sentence now reads ``using four basis functions (plus the intercept)''.
+The sentence now reads ``contains the fitted model''.
 
@@
 
+L360 Not ``knots'' - there are k-1 basis functions, but as described above, because this is a TPRS there are knots at the unique data values and then we decompose the full basis and take the required eigenvectors as the basis to use.
+
+\hypertarget{section-27}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-27}}
+
+The updated sentence now reads ``using four basis functions (plus the intercept)''.
+
+\hypertarget{reviewers-comments-27}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-27}}
+
 L362 why highlight Gaussian process smooths?
@@
 
-\hypertarget{section-26}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-26}}
+\hypertarget{section-28}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-28}}
 
@@
 
-\hypertarget{reviewers-comments-27}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-27}}
+\hypertarget{reviewers-comments-28}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-28}}
 
@@
 
-\hypertarget{section-27}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-27}}
+\hypertarget{section-29}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-29}}
 
@@
 
-\hypertarget{reviewers-comments-28}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-28}}
+\hypertarget{reviewers-comments-29}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-29}}
 
@@
 
-\hypertarget{section-28}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-28}}
+\hypertarget{section-30}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-30}}
 
@@
 
-\hypertarget{reviewers-comments-29}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-29}}
+\hypertarget{reviewers-comments-30}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-30}}
 
@@
 
-\hypertarget{section-29}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-29}}
+\hypertarget{section-31}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-31}}
 
 The word ``total'' has been removed. The sentence now reads ``If 40\% of the observations are randomly deleted\ldots{}''
-
-\hypertarget{reviewers-comments-30}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-30}}
-
-L381 ``little'' -\textgreater{} ``few''
-
-\hypertarget{section-30}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-30}}
-
-The word ``few'' has been removed. The sentence now reads ``\ldots with as few as 4 observations per group\ldots{}''
 
@@
 
-L393 ``observed difference is significant'' (not ``change'')
+L381 ``little'' -\textgreater{} ``few''
 
-\hypertarget{section-31}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-31}}
+\hypertarget{section-32}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-32}}
 
-The sentence has been rewritten. It now reads ``\ldots the observed difference is significant\ldots{}''
+The word ``few'' has been removed. The sentence now reads ``\ldots with as few as 4 observations per group\ldots{}''
 
@@
 
-Also, it is (exceedingly) unlikely to yield 0. Better to say that the estimated difference is unlikely to be distinguishable from 0.
+L393 ``observed difference is significant'' (not ``change'')
 
-\hypertarget{section-32}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-32}}
+\hypertarget{section-33}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-33}}
+
+The sentence has been rewritten. It now reads ``\ldots the observed difference is significant\ldots{}''
 
@@
 
+Also, it is (exceedingly) unlikely to yield 0. Better to say that the estimated difference is unlikely to be distinguishable from 0.
+
+\hypertarget{section-34}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-34}}
+
+\hypertarget{reviewers-comments-34}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-34}}
+
 L410 This needs to be reworded as what is written can be read as any difference between the control and treatment (and there is a difference as the controls are higher than the treated group before day 3) where you mean that the treatment group is only higher than the control group post day 3.
 
-\hypertarget{section-33}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-33}}
+\hypertarget{section-35}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-35}}
 
@@
 
-\hypertarget{reviewers-comments-34}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-34}}
+\hypertarget{reviewers-comments-35}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-35}}
 
@@
 
-\hypertarget{section-34}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-34}}
+\hypertarget{section-36}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-36}}
 

### 2021-11-16 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Updated sections

@@
 A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear data with a linear model such as rm-ANOVA or a LMEM with ``default'' (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
+We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with ``default'' (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
 Of notice, in Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
@@
 \end{equation}
-Where \(E(y_{ijt})\) is the expectation, \(g^-{1}\) is the inverse of the \emph{link function}, which relates the parameters of the distribution to the linear combination of the predictors in Equation \eqref{eq:GLM-eta}.
+Where \(E(y_{ijt})\) is the expectation, \(g^-{1}\) is the inverse of the \emph{link function} \(g\), which transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}).
 Therefore, LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
@@
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and where the change of \(y_{ijt}\) over time instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta} is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.
+Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and where the change of \(y_{ijt}\) over time instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta} is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual.
 
@@
 Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
-The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}A and the inset, respectively.
+The simulated and real data appear in Figure \ref{fig:sim-smooth-plot},A.
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SIM_No_Appendix_files/figure-latex/sim-smooth-plot-1}
+{\centering \includegraphics[width=1\linewidth]{Full_document_SIM_No_Appendix_files/figure-latex/sim-smooth-plot-1}
 
@@
 \caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+A: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
 Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
@@
 This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-We acknowledge that polynomial effects can be used in LMEMs, and that it could be argued that in Section \ref{simulation} a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects causes more questions than answers in this case (i.e., what is the appropriate polynomial effect to use?
+Polynomial effects can be used in LMEMs, and it could be argued that in Section \ref{simulation} a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects presents more questions than answers in this case (i.e., what is the appropriate polynomial effect to use?
 quadratic?, cubic?, what is the biological interpretation of a polynomial effect?).

### 2021-11-16 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Updated sections

@@
 A more comprehensive exploration  of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
+We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
 Of notice, in Section \@ref(longitudinal-GAMs) we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
@@
            ) +
-    geom_point(show.legend=FALSE) +
+    geom_point(alpha=0.5,
+               show.legend=FALSE) +
     labs(y='response')+
@@
                    color = treatment)) +
-    geom_point(show.legend=FALSE)+
+    geom_point(alpha=0.5,
+               show.legend=FALSE)+
     labs(y='response')+
     geom_line(aes(y = predict(sim_dat$fit_anova), 
-                  group = subject, size = "Subjects"),show.legend = FALSE) +
+                  group = subject, size = "Subjects"),
+              show.legend = FALSE) +
     geom_line(data = sim_dat$pred_dat, 
@@
                    color = treatment)) +
-    geom_point()+
+    geom_point(alpha=0.5)+
     labs(y='response')+

### 2021-11-16 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated sections

@@
 \end{equation}
-Where $E(y_{ijt})$ is the expectation, $g^-{1}$ is the inverse of the _link function_, which relates the parameters of the distribution to the linear combination of the  predictors in Equation \@ref(eq:GLM-eta).
+Where $E(y_{ijt})$ is the expectation, $g^-{1}$ is the inverse of the _link function_  $g$, which transforms the values from the response scale to the scale of the linear predictor $\eta$ (Equation \@ref(eq:GLM-eta)).
 Therefore, LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
@@
 
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, and where the change of $y_{ijt}$ over time instead of being defined by $\beta_{j}x_{ijt}$ as in Equation \@ref(eq:GLM-eta) is represented by the _smooth function_ $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual error.
+Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, and where the change of $y_{ijt}$ over time instead of being defined by $\beta_{j}x_{ijt}$ as in Equation \@ref(eq:GLM-eta) is represented by the _smooth function_ $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual.
 

### 2021-11-16 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated sections

@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors that appear in Figure 3C in [@vishwanath2009].
+The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors (Figure 3C in Vishwanath et.
+al.[@vishwanath2009]).
 In the paper, diffuse reflectance spectroscopy was used to quantify $\mbox{StO}_2$ changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
@@
 Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (matching the SD in the original paper).
-The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot)A and the inset, respectively.
+The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot),A.
 
@@
 #plotting simulated data
-f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group, group=subject)) +
-    geom_point(show.legend=FALSE,size=1.5,alpha=0.6)+
-    geom_line(size=1.1, alpha=0.6)+
-    # stat_summary(aes(y = StO2_sim,
-    #                  group=Group), 
-    #              fun=mean, geom="line",size=1,show.legend = FALSE)+
+f2<-ggplot(dat_sim, aes(x = Day, 
+                        y = StO2_sim, 
+                        color = Group, 
+                        group=subject)) +
+    geom_point(show.legend=FALSE,
+               size=1.5,
+               alpha=0.6)+
+    geom_line(size=0.6, alpha=0.4)+
+    geom_line(aes(x = Day, 
+               y = StO2,
+               color=Group),
+           size=1.5,
+           data=dat,
+           inherit.aes=FALSE,
+              show.legend = FALSE)+
   labs(y=expression(atop(StO[2],'(simulated)')))+
@@
                    ),
-              alpha=0.3,
+              alpha=0.5,
               data=gam_predict,
@@
     axis.text=element_text(size=22)
-  )+
-  thm1
+  )
 
@@
                    ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
+              alpha=0.5,
               data=lm_predict,
@@
                    ),
-              alpha=0.3,
+              alpha=0.5,
               data=m_predict,
@@
 (ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+A: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
 Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point.
@@
 
-```{r,sim-smooth-plot,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}
+```{r,sim-smooth-plot,fig.width=12, fig.height=7, out.width='100%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}
 
-mult_plot<-f2+inset_element(
-  f1, left = 0.01, 
-  bottom = 0.5, 
-  right = 0.5, 
-  top = 1.0)+
-  f3+f4+f6+
+mult_plot<-f2+f3+f4+f6+
    plot_annotation(tag_levels='A')&
@@
      text=element_text(size=18)
-     )&
-  thm1
+     )
 

### 2021-11-16 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/05-Discussion.Rmd
updated GAM section
Updated sections

@@
 This "model misspecification" error, also is known as a "Type III" error [@dennis2019] is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-We acknowledge that polynomial effects can be used in LMEMs, and that it could be argued that in Section \@ref(simulation) a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects causes more questions than answers in this case (i.e., what is the appropriate polynomial effect to use?
+Polynomial effects can be used in LMEMs, and it could be argued that in Section \@ref(simulation) a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects presents more questions than answers in this case (i.e., what is the appropriate polynomial effect to use?
 quadratic?, cubic?, what is the biological interpretation of a polynomial effect?).
-Moreover, polynomial effects have other associated issues that we have described in Section {#GAM-theory}.
+Moreover, polynomial effects have other associated issues that we have described in Section \@ref(GAM-theory).
 

### 2021-11-17 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.Rmd
Updated figures and theme using scico

@@
 #thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
-thm1<-scale_fill_scico_d(palette="oslo")
+thm1<-scale_fill_scico_d(palette="tokyo",begin=0.2, end=0.8, direction = -1, aesthetics = c("colour","fill"))
 

### 2021-11-17 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Updated figures and theme using scico

@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in.\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors (Figure 3C in Vishwanath et.
+al.\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}}).
+In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
 In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group.

### 2021-11-17 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated figures and theme using scico

@@
             aes(x=time,y=y))+
-  geom_point(size=sz+1)+
+  geom_point(size=sz+1,alpha=0.5)+
   thm1+

### 2021-11-17 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated figures and theme using scico

@@
             )
+#alpha for ribbon
+al<-0.8
 
@@
   theme_classic()+
-  thm1+
     scale_x_continuous(breaks=c(0,5,10))+
@@
     axis.text=element_text(size=14)
-  )
+  )+
+    thm1
 
@@
   )+
-  thm1+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    thm1
 
@@
                    ),
-              alpha=0.5,
+              alpha=al,
               data=gam_predict,
@@
     axis.text=element_text(size=22)
-  )
+  )+thm1
 
@@
                    ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.5,
+              alpha=al,
               data=lm_predict,
@@
                    ),
-              alpha=0.5,
+              alpha=al,
               data=m_predict,

### 2021-11-18 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.Rmd
Updated sections on wiggliness, GLMMs. Changed palette.

@@
 ## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+- We appreciate this comment regarding wiggliness.
+We have now added a new paragraph in Section 4.3.1 (L369-375) which briefly discuss wiggliness, and uses a physics analogy to convey the overall concept of penalization.
+Because the intention of our work is to present the practical advantages of GAMs we want to avoid including mathematical notation about penalization in the manuscript, which would be confusing and distracting to the readers from the main purpose of the paper.
+
+- We have substituted "weak" and "strong" for "low" and "high" (L373).
 

### 2021-11-18 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.tex
Updated sections on wiggliness, GLMMs. Changed palette.

@@
 
-You talk a little about ``wiggliness'' and penalising the weights for each basis function towards 0 but you don't really explain this most crucial concept of the model fitting problem and why modern GAMs are so much better than the
-GAMs developed by Hastie and Tibshirani when they first introduced GAMs to the world.
+You talk a little about ``wiggliness'' and penalising the weights for each basis function towards 0 but you don't really explain this most crucial concept of the model fitting problem and why modern GAMs are so much better than the GAMs developed by Hastie and Tibshirani when they first introduced GAMs to the world.
 You really need to define wiggliness beyond something that is used to avoid overfitting.
@@
 \subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-6}}
+
+\begin{itemize}
+\item
+  We appreciate this comment regarding wiggliness.
+We have now added a new paragraph in Section 4.3.1 (L369-375) which briefly discuss wiggliness, and uses a physics analogy to convey the overall concept of penalization.
+Because the intention of our work is to present the practical advantages of GAMs we want to avoid including mathematical notation about penalization in the manuscript, which would be confusing and distracting to the readers from the main purpose of the paper.
+\item
+  We have substituted ``weak'' and ``strong'' for ``low'' and ``high'' (L373).
+\end{itemize}
 
@@
 \subsection{Reviewer's Comments}\label{reviewers-comments-8}}
+
+The essence of this is not ``if the difference between {[}credible{]} intervals of the fitted smooths\ldots{} is non-zero''.
+For a variety of reasons this is not how it works.
+Instead we compute the difference of the smooths and then compute a credible interval around this difference.
+Also, to interpret this correctly we really need to do a simultaneous interval as you are looking at all points over the range of the covariate.
+If we just use the usual interval we have to consider the under and over coverage of the Nychka across-the-function intervals.
+You should talk about this issue in this section if you don't compute simultaneous intervals.
+This section could also do with a little more explanation; \{gratia\} (and hence I) tens to consider differences of smooths to exclude the groups means.
+Here, if I read the paper and the code correctly, you included the group means.
+This works OK for simple models like this, but what if you had two or more parametric factor terms in the model?
+Then you would not (at least by default) have any way to isolate the group means for the by factor independently of the levels of the other factor.
+You would have to create differences conditional upon a level of
+the other factor. Users, however, tend to expect differences to include the group means, as you have chosen to present things here. Hence a note exp
 
@@
 
+\hypertarget{figures-1}{%
+\section{Figures}\label{figures-1}}
+
+\hypertarget{reviewers-comments-15}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-15}}
+
+The figures are too tall --- the aspect ratio used for the individual panels is wrong. You should be emphasising the time axis so you should have individual panels that are wider than they are tall.
+
+Figure 3: on panel A, are the solid lines the truth in the main panel?
+If they are, you don't need the inset plot, and if they aren't, are they the mean of the simulated observations?
+If they are this latter, I think this is superfluous and it would be better to make the solid line the truth, about which you simulated noisy data.
+
+\hypertarget{section-16}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-16}}
+
+We have changed the aspect of the figures to emphasize the time axis, making them wider.
+
+On Figure 3, we have added the original data to panel A, which now displays both the simulated data (thin lines) and the original data (truth) in order to simplify the figure.
+We have updated also places in the manuscript that reference this Figure as well.
+
 \hypertarget{minor-comments}{%
@@
 
-\hypertarget{reviewers-comments-15}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-15}}
+\hypertarget{reviewers-comments-16}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-16}}
 
@@
 
-\hypertarget{section-16}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-16}}
+\hypertarget{section-17}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-17}}
 
@@
 
-\hypertarget{section-17}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-17}}
+\hypertarget{section-18}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-18}}
 
@@
 
-\hypertarget{reviewers-comments-17}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-17}}
+\hypertarget{reviewers-comments-18}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-18}}
 
@@
 
-\hypertarget{section-18}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-18}}
+\hypertarget{section-19}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-19}}
 
@@
 
-\hypertarget{reviewers-comments-18}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-18}}
-
-L185 here and throughout, check for situations where you have combined numerical citations to literature within narrative text; Here it should be ``\ldots see refs 40 \& 42.)'' for example, not superscript numerals.
-
-\hypertarget{section-19}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-19}}
-
-\textbf{leave this to the end as it can be fixed in the final .tex document}
-
 \hypertarget{reviewers-comments-19}{%
@@
 
-L244 ``global'' is perhaps more conventionally described as ``population'' in the mixed model literature.
+L185 here and throughout, check for situations where you have combined numerical citations to literature within narrative text; Here it should be ``\ldots see refs 40 \& 42.)'' for example, not superscript numerals.
 
@@
 
-We have changed the word from ``global'' to ``population''.
+\textbf{leave this to the end as it can be fixed in the final .tex document}
 
@@
 
-L283 I think it is better to say ``\ldots the mathematical space within which the true but unknown \(f(x_t)\) is thought to exist.''
+L244 ``global'' is perhaps more conventionally described as ``population'' in the mixed model literature.
 
@@
 
-We have updated this line to ``the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist'', in L290 in the revised manuscript.
+We have changed the word from ``global'' to ``population''.
 
@@
 
-L293 You never really describe what ``data-drive flexibility'' is. Why are GAMs data driven and LMMs and GLMs not? This is a critical concept to convey I feel.
+L283 I think it is better to say ``\ldots the mathematical space within which the true but unknown \(f(x_t)\) is thought to exist.''
 
@@
 \subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-22}}
+
+We have updated this line to ``the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist'', in L290 in the revised manuscript.
+
+\hypertarget{reviewers-comments-22}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-22}}
+
+L293 You never really describe what ``data-drive flexibility'' is. Why are GAMs data driven and LMMs and GLMs not? This is a critical concept to convey I feel.
+
+\hypertarget{section-23}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-23}}
 
@@
 
-\hypertarget{reviewers-comments-22}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-22}}
+\hypertarget{reviewers-comments-23}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-23}}
 
@@
 
-\hypertarget{section-23}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-23}}
-
-The sentence now reads ``separate smooths for\ldots{}''
-
-\hypertarget{reviewers-comments-23}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-23}}
-
-L354 ``difference between group means'' because the smooth encodes differences between the groups too, just not difference in the means. Be a little more specific here.
-
 \hypertarget{section-24}{%
@@
 
-The sentence now reads ``any differences between the group means''.
+The sentence now reads ``separate smooths for\ldots{}''
 
@@
 
-L356 use straight quotes in code, not curly ones. And speficy the \passthrough{\lstinline!family!} here?
+L354 ``difference between group means'' because the smooth encodes differences between the groups too, just not difference in the means. Be a little more specific here.
 
@@
 
+The sentence now reads ``any differences between the group means''.
+
 \hypertarget{reviewers-comments-25}{%
@@
 
-L359 ``contain the fitted'' instead of ``store the''
+L356 use straight quotes in code, not curly ones. And speficy the \passthrough{\lstinline!family!} here?
 
@@
 
-The sentence now reads ``contains the fitted model''.
-
 \hypertarget{reviewers-comments-26}{%
@@
 
-L360 Not ``knots'' - there are k-1 basis functions, but as described above, because this is a TPRS there are knots at the unique data values and then we decompose the full basis and take the required eigenvectors as the basis to use.
+L359 ``contain the fitted'' instead of ``store the''
 
@@
 
-The updated sentence now reads ``using four basis functions (plus the intercept)''.
+The sentence now reads ``contains the fitted model''.
 
@@
 
+L360 Not ``knots'' - there are k-1 basis functions, but as described above, because this is a TPRS there are knots at the unique data values and then we decompose the full basis and take the required eigenvectors as the basis to use.
+
+\hypertarget{section-28}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-28}}
+
+The updated sentence now reads ``using four basis functions (plus the intercept)''.
+
+\hypertarget{reviewers-comments-28}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-28}}
+
 L362 why highlight Gaussian process smooths?
@@
 
-\hypertarget{section-28}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-28}}
+\hypertarget{section-29}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-29}}
 
@@
 
-\hypertarget{reviewers-comments-28}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-28}}
-
-L366 See also Matteo Fasiolo's \{mgcViz\} package
-
-\hypertarget{section-29}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-29}}
-
-We acknowledge the existence of Dr.~Fasiolo's \emph{mgcvViz} package, but due to the introductory nature of this paper we prefer to focus only on \emph{mgcv} and \emph{gratia}.
-
 \hypertarget{reviewers-comments-29}{%
@@
 
-L373 Do rm-ANOVA and LMMs require equally-spaced or complete observations?
+L366 See also Matteo Fasiolo's \{mgcViz\} package
 
@@
 
-Only rm-ANOVA requires equally-spaced and complete observations.
-To avoid confusion, the line now reads ``Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does)''.
+We acknowledge the existence of Dr.~Fasiolo's \emph{mgcvViz} package, but due to the introductory nature of this paper we prefer to focus only on \emph{mgcv} and \emph{gratia}.
 
@@
 
-L377 delete ``total''
+L373 Do rm-ANOVA and LMMs require equally-spaced or complete observations?
 
@@
 
-The word ``total'' has been removed. The sentence now reads ``If 40\% of the observations are randomly deleted\ldots{}''
+Only rm-ANOVA requires equally-spaced and complete observations.
+To avoid confusion, the line now reads ``Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does)''.
 
@@
 
-L381 ``little'' -\textgreater{} ``few''
+L377 delete ``total''
 
@@
 
-The word ``few'' has been removed. The sentence now reads ``\ldots with as few as 4 observations per group\ldots{}''
+The word ``total'' has been removed. The sentence now reads ``If 40\% of the observations are randomly deleted\ldots{}''
 
@@
 
-L393 ``observed difference is significant'' (not ``change'')
+L381 ``little'' -\textgreater{} ``few''
 
@@
 
-The sentence has been rewritten. It now reads ``\ldots the observed difference is significant\ldots{}''
+The word ``few'' has been removed. The sentence now reads ``\ldots with as few as 4 observations per group\ldots{}''
 
@@
 
-Also, it is (exceedingly) unlikely to yield 0. Better to say that the estimated difference is unlikely to be distinguishable from 0.
+L393 ``observed difference is significant'' (not ``change'')
 
@@
 
+The sentence has been rewritten. It now reads ``\ldots the observed difference is significant\ldots{}''
+
 \hypertarget{reviewers-comments-34}{%
@@
 
-L410 This needs to be reworded as what is written can be read as any difference between the control and treatment (and there is a difference as the controls are higher than the treated group before day 3) where you mean that the treatment group is only higher than the control group post day 3.
+Also, it is (exceedingly) unlikely to yield 0. Better to say that the estimated difference is unlikely to be distinguishable from 0.
 
@@
 
+\hypertarget{reviewers-comments-35}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-35}}
+
+L410 This needs to be reworded as what is written can be read as any difference between the control and treatment (and there is a difference as the controls are higher than the treated group before day 3) where you mean that the treatment group is only higher than the control group post day 3.
+
+\hypertarget{section-36}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-36}}
+
 This was an unintentional mistake based on some previous simulation results.
@@
 
-\hypertarget{reviewers-comments-35}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-35}}
+\hypertarget{reviewers-comments-36}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-36}}
 
@@
 
-\hypertarget{section-36}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-36}}
+\hypertarget{section-37}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-37}}
 

### 2021-11-18 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.Rmd
Updated sections on wiggliness, GLMMs. Changed palette.

--- 2021-11-09 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.Rmd
@@
 #thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
-thm1<-scale_fill_scico_d(palette="oslo")
+thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
 

### 2021-11-18 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Updated sections on wiggliness, GLMMs. Changed palette.

@@
 
-The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, \(\textsf{R}\)\textsuperscript{\protect\hyperlink{ref-r}{38}} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models.\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}} Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
+The current advances in programming languages designed for statistical analysis (specifically R), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
+In particular, R\textsuperscript{\protect\hyperlink{ref-r}{38}} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+At the same time, R has many tools that simplify data simulation, an emerging strategy used to test statistical models.\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}} Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
 
@@
 Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies.\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in \(\textsf{R}\), in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
+Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in R, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
 
@@
 In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
-Independent errors \(\varepsilon_{ijt}\) represent \textbf{random variation from the sampling process } assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
+Independent errors \(\varepsilon_{ijt}\) represent \textbf{random variation from the sampling process } assumed to be \(\stackrel{i.i.d.}\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
 In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug,'' or ``saline'' vs.~``chemotherapy'').
@@
 Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response, this means that either presented as Equation \eqref{eq:linear-model}, Equation \eqref{eq:ANOVA-by-group} or Equation \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
+In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response.
+This means that either presented as Equation \eqref{eq:linear-model}, Equation \eqref{eq:ANOVA-by-group} or Equation \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data.
+If the data show non-linear trends, the rm-ANOVA model will approximate this behavior with non-parallel lines.
 
@@
 This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)).
-The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the ``global noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.
-
-For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data.
+The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the **``sampling** noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.
+
+For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data \textbf{that needs to be accounted for}.
 As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
@@
 In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study\textsuperscript{\protect\hyperlink{ref-wolfinger1996}{41}} (For an in-depth analysis of the covariance matrix see).\textsuperscript{\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}}
+The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study\textsuperscript{\protect\hyperlink{ref-wolfinger1996}{41}} (For an in-depth analysis of the covariance matrix see West\textsuperscript{\protect\hyperlink{ref-west2014}{40}} and Weiss\textsuperscript{\protect\hyperlink{ref-weiss2005}{42}}).
 
@@
 
-To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\)\textsuperscript{\protect\hyperlink{ref-r}{38}} and the package \emph{nlme}.\textsuperscript{\protect\hyperlink{ref-nlme}{50}}
+To visually demonstrate the limitations of rm-ANOVA an LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using R\textsuperscript{\protect\hyperlink{ref-r}{38}} and the package \emph{nlme}.\textsuperscript{\protect\hyperlink{ref-nlme}{50}}
 
@@
 A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with ``default'' (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
+We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with ``default'' (linear) effects and, 2) demonstrate that a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
 Of notice, in Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
@@
 Top row: linear response, bottom row: quadratic response.
-A: Simulated linear data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data.
-D: Simulated quadratic data with known mean response (thin lines) and individual responses (points) showing the dispersion of the data.
+A: Simulated linear data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
+D: Simulated quadratic data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
 B,E: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
@@
 The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation.
-C, F: Estimates from the LMEM in the linear and quadratic case.
+C, F: Estimates from the LMEM in the linear and quadratic case (subject: thin lines, population: thick lines) .
 The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data and grossly bias the initial estimates for each group in the quadratic case (bottom row).}\label{fig:l-q-response}
@@
 
-This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-The models fitted to the simulated data were an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response, which is only noticeable if the model fit and the response are visualized.
+This section has used simulation to better convey and visulize the limitations of linearity and correlation in the response in data with non-linear trends using an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
+Notice that the model misspecification is easily noticeable if the model fit and the response are visualized.
 In the following section, we provide a brief overview of linear models, general linear models and generalized linear mixed models before presenting the theory of GAMs.
@@
 In reality, rm-ANOVA and LMEMs are just \emph{special cases} of broader model classes (General Linear Models and Generalized Linear Mixed Models, respectively).
-Understanding this is critical to fully capture the constraints of such models and to understand how GAMs overcome those limitations.
-This section will briefly present the different Statistical model classes, indicate how rm-ANOVA and LMEMs fit within that framework and introduce the theory of GAMs.
+In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different Statistical model classes, indicate how rm-ANOVA and LMEMs fit within that framework and introduce the theory of GAMs.
 
@@
 
-Linear models (LMs) are those that assume a normal (Gaussian) distribution of the response, such as rm-ANOVA.
-These are by far the models most commonly used to analyze data within the biomedical research community, and only incorporate \emph{fixed effects}.
+Linear models (LMs) are those that assume a normal (Gaussian) distribution of the response, and only incorporate \emph{fixed effects} (such as rm-ANOVA).
+These are by far the models most commonly used to analyze data within the biomedical research community.
 On the other hand, Linear Mixed Effect Models (LMEMs) are those that assume a normally-distributed response but also incorporate \emph{random effects}, as it has been described in Section \ref{LMEM-case}.
 
-In matrix notation, a LM has the general form:
+In matrix notation, a LM has the general form
 
@@
 The ``model equation'' form of Equation \eqref{eq:LM-matrix} for a rm-ANOVA model of two treatment groups corresponds to Equation \eqref{eq:linear-model}.
-It is important to remember that the error term \(\varepsilon_{ijt}\) in Equation \eqref{eq:linear-model} is due to random sampling from the normal distribution.
+It is important to remember that the error term \(\varepsilon_{ijt}\) in Equation \eqref{eq:linear-model} represents the variability introduced by random sampling from the normal distribution.
 
@@
 McCullagh and Nelder\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} defined General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
-To achieve this, consider the following:
+To achieve this, consider the following
 
@@
 
-Where \(\eta_{ijt}\) is the linear predictor for each observation, \(\beta_0\) is the intercept, and \(\beta_{j}\) is the model parameter for each group; and \(x_{ijt}\) represents the covariates from each group at each time point.
+Where \(\eta_{ijt}\) is the linear predictor for each observation, \(\beta_0\) is the intercept, and \(\beta_{j}\) is the model parameter for each group; and \(x_{ijt}\) represents the covariates from each subject in each group at each time point.
 
@@
 \end{equation}
-Where \(E(y_{ijt})\) is the expectation, \(g^-{1}\) is the inverse of the \emph{link function} \(g\), which transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}).
+Where \(E(y_{ijt})\) is the expectation, \(g^{-1}\) is the inverse of the \emph{link function} (\(g\)).
+The link function transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}).
 Therefore, LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
@@
 
-Where \(\mu_{ij}\) corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of the previous equations.
-Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed.\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in refs\textsuperscript{\protect\hyperlink{ref-dobson2008}{54}} and.\textsuperscript{\protect\hyperlink{ref-stroup2013}{55}}
+Where \(\mu_{ij}\) corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of Equation \eqref{eq:GLM-eta}.
+Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed.\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in Dobson\textsuperscript{\protect\hyperlink{ref-dobson2008}{54}} and Stroup.\textsuperscript{\protect\hyperlink{ref-stroup2013}{55}}
 
@@
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and where the change of \(y_{ijt}\) over time instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta} is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual.
+Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and the change of \(y_{ijt}\) over time (instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta}) is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual due to \textbf{random sampling}.
 
@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
+One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
 
@@
 
-In contrast, thin regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see refs.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} and).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
+In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).
 
@@
 A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
-Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).
+Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth.
+It is important to note that is not necessary to specify a number of basis equal to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data.
+
+Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present.
+A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but it suffices to say that the wiggliness (or ``roughness'') of the fit is achieved by using a smoothness parameter \(\lambda\), which is applied to the second derivative of the spline.
+The second derivative of the spline is a measure of curvature, or the rate of change of the slope.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} A physics analogy to the penalization process would be that if we are driving a car on a hilly and curvy road, we want to minimize abrupt changes in speed in order to have a ``smooth'' ride, and we do this by penalizing the acceleration.
 
@@
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the ``wiggliness'' of the fit, each of the basis functions of Figure \ref{fig:basis-plot}A is weighted by multiplying it by a coefficient according to the matrix of Figure \ref{fig:basis-plot}B.
-The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the ``wiggliness'' of the smooth fit to prevent overfitting.
-A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
+To control the ``wiggliness'' of the fit, each of the basis functions of Figure \ref{fig:basis-plot}A is weighted by multiplying it by the smoothness parameter (\(\lambda\)) according to the matrix of Figure \ref{fig:basis-plot}B.
+The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
+A low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
 
@@
 For GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, the percentage of CIs entirely containing the true function over the entire time period would be close to the nominal value (i.e., if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function), thereby allowing more robust estimates from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{61}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{62}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in Miller,\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} and Marra.\textsuperscript{\protect\hyperlink{ref-marra2012}{62}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
 
@@
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in.\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}}
+A detailed description on basis functions and correlations can be found in Hefley.\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}}
 
@@
 \CSLLeftMargin{64. }
-\CSLRightInline{Harezlak J, Ruppert D, Wand MP. \emph{Semiparametric Regression with r}. Springer New York; 2018. doi:\href{https://doi.org/10.1007/978-1-4939-8853-2}{10.1007/978-1-4939-8853-2}}
+\CSLRightInline{Harezlak J, Ruppert D, Wand MP. \emph{Semiparametric Regression with {R}}. Springer New York; 2018. doi:\href{https://doi.org/10.1007/978-1-4939-8853-2}{10.1007/978-1-4939-8853-2}}
 

### 2021-11-18 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/01-Background.Rmd
Updated sections on wiggliness, GLMMs. Changed palette.

@@
 
-The current advances in programming languages designed for statistical analysis (specifically  $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, $\textsf{R}$ [@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016; @wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
+The current advances in programming languages designed for statistical analysis (specifically R), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
+In particular, R [@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016; @wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
 hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017].
+At the same time, R has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017].
 Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
@@
 The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in $\textsf{R}$, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
+Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in R, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
 

### 2021-11-18 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Updated sections on wiggliness, GLMMs. Changed palette.

@@
 In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-Independent errors $\varepsilon_{ijt}$ represent **random variation from the sampling process ** assumed to be $\sim N(0,\sigma^2)$ (independently and identically normally distributed with mean zero and variance $\sigma^2$).
+Independent errors $\varepsilon_{ijt}$ represent **random variation from the sampling process ** assumed to be $\stackrel{i.i.d.}\sim N(0,\sigma^2)$ (independently and identically normally distributed with mean zero and variance $\sigma^2$).
 In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
@@
 Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
+In other words, the rm-ANOVA model "expects" a linear relationship between the covariates and the response.
+This means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data.
+If the data show non-linear trends, the rm-ANOVA model will approximate this behavior with non-parallel lines.
 
@@
 This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject (subject$_i$) within each group (group$_j$).
-The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the "global noise" term $\varepsilon_{ijt}$ from Equation \@ref(eq:linear-model).
+The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the **"sampling** noise" term $\varepsilon_{ijt}$ from Equation \@ref(eq:linear-model).
 
-For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data.
+For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data **that needs to be accounted for**.
 As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
@@
 In a longitudinal study there is an expected _covariance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group.
-The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]).
+The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see West[@west2014] and Weiss[@weiss2005]).
 
@@
 
-To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \@ref(eq:linear-model)), and a LMEM  (Equation \@ref(eq:LMEM)) are fitted to each group, using $\textsf{R}$ [@r] and the package _nlme_ [@nlme].
+To visually demonstrate the limitations of rm-ANOVA an LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
+An rm-ANOVA model (Equation \@ref(eq:linear-model)), and a LMEM  (Equation \@ref(eq:LMEM)) are fitted to each group, using R [@r] and the package _nlme_ [@nlme].
 
@@
 A more comprehensive exploration  of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
+We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
 Of notice, in Section \@ref(longitudinal-GAMs) we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
@@
 The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation.
-C, F: Estimates from the LMEM in the linear and quadratic case.
+C, F: Estimates from the LMEM in the linear and quadratic case (subject: thin lines, population: thick lines) .
 The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data and grossly bias the initial estimates for each group in the quadratic case (bottom row).

### 2021-11-18 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated sections on wiggliness, GLMMs. Changed palette.

@@
 In reality, rm-ANOVA and LMEMs are just _special cases_ of broader model classes (General Linear Models and Generalized Linear Mixed Models, respectively).
-Understanding this is critical to fully capture the constraints of such models and to understand how GAMs overcome those limitations.
-This section will briefly present the different Statistical model classes, indicate how rm-ANOVA and LMEMs fit within that framework and introduce the theory of GAMs.
+In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different  Statistical model classes, indicate how rm-ANOVA and LMEMs fit within that framework and introduce the theory of GAMs.
 
@@
 
-Linear models (LMs) are those that assume a normal (Gaussian) distribution of the response, such as rm-ANOVA.
-These are by far the models most commonly used to analyze data within the biomedical research community, and only incorporate _fixed effects_.
+Linear models (LMs) are those that assume a normal (Gaussian) distribution of the response, and only incorporate _fixed effects_ (such as rm-ANOVA).
+These are by far the models most commonly used to analyze data within the biomedical research community.
 On the other hand, Linear Mixed Effect Models (LMEMs) are those that assume a normally-distributed response but also incorporate _random effects_, as it has been described in Section \@ref(LMEM-case).
 
-In matrix notation, a LM has the general form:
+In matrix notation, a LM has the general form
 
@@
 The "model equation" form of Equation \@ref(eq:LM-matrix) for a rm-ANOVA model of two treatment groups corresponds to Equation \@ref(eq:linear-model).
-It is important to remember that the error term $\varepsilon_{ijt}$ in Equation \@ref(eq:linear-model) is due to random sampling from the normal distribution.
+It is important to remember that the error term $\varepsilon_{ijt}$ in Equation \@ref(eq:linear-model) represents the variability introduced by random sampling from the normal distribution.
 
@@
 McCullagh and Nelder [@nelder1972] defined General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
-To achieve this, consider the following:
+To achieve this, consider the following
 
@@
 
-Where $\eta_{ijt}$ is the linear predictor for each observation, $\beta_0$ is the intercept, and $\beta_{j}$ is the model parameter for each group; and $x_{ijt}$ represents the covariates from each group at each time point.
+Where $\eta_{ijt}$ is the linear predictor for each observation, $\beta_0$ is the intercept, and $\beta_{j}$ is the model parameter for each group; and $x_{ijt}$ represents the covariates from each subject in each group at each time point.
 
@@
 \end{equation}
-Where $E(y_{ijt})$ is the expectation, $g^-{1}$ is the inverse of the _link function_  $g$, which transforms the values from the response scale to the scale of the linear predictor $\eta$ (Equation \@ref(eq:GLM-eta)).
+Where $E(y_{ijt})$ is the expectation, $g^{-1}$ is the inverse of the _link function_  ($g$).
+The link function transforms the values from the response scale to the scale of the linear predictor $\eta$ (Equation \@ref(eq:GLM-eta)).
 Therefore, LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
@@
 
-Where $\mu_{ij}$ corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of the previous equations.
+Where $\mu_{ij}$ corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of Equation \@ref(eq:GLM-eta).
 Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed [@nelder1972].
-In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in refs [@dobson2008] and [@stroup2013].
+In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in Dobson [@dobson2008] and Stroup [@stroup2013].
 
@@
 
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, and where the change of $y_{ijt}$ over time instead of being defined by $\beta_{j}x_{ijt}$ as in Equation \@ref(eq:GLM-eta) is represented by the _smooth function_ $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual.
+Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, and the change of $y_{ijt}$ over time (instead of being defined by $\beta_{j}x_{ijt}$ as in Equation \@ref(eq:GLM-eta)) is represented by the _smooth function_ $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual due to **random sampling**.
 
@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials (which can be used in LMEMs), but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with inference [@beck1998].
+One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with inference [@beck1998].
 In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic and causes bias at the endpoints of the time period.
@@
 
-In contrast, thin regression splines (TPRS) provide a reasonable "low rank" (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner [@wood2003].
-Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see refs.
-[@wood2003] and [@wood2017]).
+In contrast, thin plate regression splines (TPRS) provide a reasonable "low rank" (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner [@wood2003].
+Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood[@wood2003;@wood2017]).
 
@@
 A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
-Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).
+Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth.
+It is important to note that is not necessary to specify a number of basis equal to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data.
+
+Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too "wiggly" due to the noise present.
+A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but it suffices to say that the wiggliness (or "roughness") of the fit is achieved by using a smoothness parameter $\lambda$, which is applied to the second derivative of the spline.
+The second derivative of the spline is a measure of curvature, or the rate of change of the slope [@simpson2018; @wood2017].
+A physics analogy to the penalization process would be that if we are driving a car on a hilly and curvy road, we want to minimize abrupt changes in speed  in order to have a "smooth" ride, and we do this by penalizing the acceleration.
 
@@
 Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the "wiggliness" of the fit, each of the basis functions of Figure \@ref(fig:basis-plot)A is weighted by multiplying it by a coefficient according to the matrix of Figure \@ref(fig:basis-plot)B.
-The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the "wiggliness" of the smooth fit to prevent overfitting.
-A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.
+To control the "wiggliness" of the fit, each of the basis functions of Figure \@ref(fig:basis-plot)A is weighted by multiplying it by the smoothness parameter ($\lambda$) according to the matrix of Figure \@ref(fig:basis-plot)B.
+The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
+A low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
 
@@
 For GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, the percentage of CIs entirely containing the true function over the entire time period would be close to the nominal value (i.e., if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function), thereby allowing more robust estimates from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in [@miller2019;@wood2017;@simpson2018] and [@marra2012].
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in Miller [@miller2019], Wood[@wood2017], Simpson [@simpson2018] and Marra [@marra2012].
 With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as "empirical Bayesian" through the rest of this paper.

### 2021-11-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions/SIM_Appendix.Rmd
Updated code and figures

@@
 library(gratia)
+library(scico)
 library(viridis)
@@
 #global themes for plotting
-thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
-thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
-
-#thm<-scale_color_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting
-#thm1<-scale_fill_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting
+#thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
+#thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
+thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
 

### 2021-11-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions/SIM_Appendix.tex
Updated code and figures

--- 2021-10-17 Manuscripts/Manuscript by chapters- SIM Revisions/SIM_Appendix.tex
@@
 The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}\protect\hyperlink{ref-wagenmakers2008}{12}{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}{]}.
+The rm-ANOVA model makes three assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}{]}.
 
@@
 A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
-Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data.
+Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from data that shows non-linear trends, the validity of such metric is dependent on how adequate the model fits the data.
 In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}.
@@
 Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}{]}, consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}\protect\hyperlink{ref-abdi2010}{21}{]}), but they in turn reduce statistical power (1-\(\beta\)){[}\protect\hyperlink{ref-nakagawa2004}{22}{]}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) {[}\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}{]}.
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}\protect\hyperlink{ref-abdi2010}{21}{]}), but they in turn reduce statistical power (1-\(\beta\)){[}\protect\hyperlink{ref-nakagawa2004}{22}{]}, and lead to increased Type II error (failing to reject the null hypothesis when it is false) {[}\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}{]}.
 Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
@@
 When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}.
-And even more importantly, LMEMs also assume a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}, making them unsuitable to analyze non-linear data.
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
+However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}.
+And even more importantly, LMEMs also assume by default a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]} (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .
+
+As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the trend of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
@@
 Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]}.
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]} and 3) can easily accommodate missing observations.
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]}, and 3) can easily accommodate missing observations.
 Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
 
-The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
+The current advances in programming languages designed for statistical analysis (specifically R), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
+In particular, R {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
+At the same time, R has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
 Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
@@
 The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in \(\textsf{R}\), in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
+Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in R, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
 
@@
 In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
-Independent errors \(\varepsilon_{ijt}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
-In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy'').
+Independent errors \(\varepsilon_{ijt}\) represent \textbf{random variation from the sampling process } assumed to be \(\stackrel{i.i.d.}\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
+In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug,'' or ``saline'' vs.~``chemotherapy'').
 Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B).
@@
 Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response, this means that either presented as Equation \eqref{eq:linear-model}, Equation \eqref{eq:ANOVA-by-group} or Equation \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines.
-
-\hypertarget{the-linear-mixed-model-case-lmem}{%
-\subsubsection{The Linear Mixed Model Case (LMEM)}\label{the-linear-mixed-model-case-lmem}}
+In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response.
+This means that either presented as Equation \eqref{eq:linear-model}, Equation \eqref{eq:ANOVA-by-group} or Equation \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data.
+If the data show non-linear trends, the rm-ANOVA model will approximate this behavior with non-parallel lines.
+
+\hypertarget{LMEM-case}{%
+\subsubsection{The Linear Mixed Model (LMEM) Case}\label{LMEM-case}}
 
@@
 
-When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)).
+When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEMs and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)).
 This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)).
-The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the ``global noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.
-
-For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data.
+The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the **``sampling** noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.
+
+For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data \textbf{that needs to be accounted for}.
 As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
@@
 
+Of notice, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach, which is described in detail by Pinheiro and Bates {[}\protect\hyperlink{ref-pinheiro2006}{15}{]} using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}.
+However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates {[}\protect\hyperlink{ref-beck1998}{36}{]}; more importantly, polynomial effects lack biological or mechanistic interpretation {[}\protect\hyperlink{ref-pinheiro2006}{15}{]} which limits their use in biomedical studies.
+
 \hypertarget{covariance-in-rm-anova-and-lmems}{%
@@
 In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{41}{]} (For an in-depth analysis of the covariance matrix see {[}\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}{]}).
+The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{41}{]} (For an in-depth analysis of the covariance matrix see West{[}\protect\hyperlink{ref-west2014}{40}{]} and Weiss{[}\protect\hyperlink{ref-weiss2005}{42}{]}).
 
@@
 
-\hypertarget{missing-observations}{%
-\subsection{Missing observations}\label{missing-observations}}
-
-Missing observations are an issue that arises frequently in longitudinal studies.
-In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
-Dropout from patients and attrition or injury in animals are among the reasons for missing observations.
-Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}\protect\hyperlink{ref-weiss2005}{42}{]}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{46}{]}.
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{47}{]}, and in the case of MNAR the missing values are dependent on their value.
-
-An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{48}{]}.
-In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}\protect\hyperlink{ref-west2014}{40}{]}.
+\hypertarget{unbalanced-data}{%
+\subsection{Unbalanced data}\label{unbalanced-data}}
+
+In a longitudinal study, it is frequently the case that the number of observations is different across subjects.
+In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing,'' ``incomplete'' or (more generally speaking) unbalanced data {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
+The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
+This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{46}{]}.
+
+On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are \emph{missing at random} (MAR) or \emph{completely missing at random} (MCAR) {[}\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}{]}.
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{47}{]}.
+If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{48}{]}.
+Missing observations can also be \emph{missing not at random} (MNAR) and in the case the missing observations are dependent on their value.
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
+
+However, it is worth reminding that ``all models are wrong'' {[}\protect\hyperlink{ref-box1976}{49}{]} and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
+Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.
 
@@
 
-To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{49}{]}.
+To visually demonstrate the limitations of rm-ANOVA an LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using R {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{50}{]}.
 
@@
 A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
-We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs.
+We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with ``default'' (linear) effects and, 2) demonstrate that a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
 Of notice, in Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
@@
 Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
-
-This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data.
-The models fitted to the simulated data were an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
-In the following section, we present generalized additive models (GAMs) as a data-driven alternative method to analyze longitudinal non-linear data that overcomes the linearity assumption.
+It could be argued that a LMEM with quadratic effects should have been used to fit the data in Figure\ref{fig:l-q-response}, F.
+However, because in reality the true function is not known, choosing a polynomial degree causes more questions (e.g., is it quadratic?, cubic?, or a higher degree?).
+Additionally, polynomial effects have other limitations, which we cover in Section \ref{GAM-theory}.
+
+This section has used simulation to better convey and visulize the limitations of linearity and correlation in the response in data with non-linear trends using an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
+Notice that the model misspecification is easily noticeable if the model fit and the response are visualized.
+In the following section, we provide a brief overview of linear models, general linear models and generalized linear mixed models before presenting the theory of GAMs.
+
+\hypertarget{linear-models-and-beyond}{%
+\section{Linear Models, and beyond}\label{linear-models-and-beyond}}
+
+In reality, rm-ANOVA and LMEMs are just \emph{special cases} of broader model classes (General Linear Models and Generalized Linear Mixed Models, respectively).
+In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different Statistical model classes, indicate how rm-ANOVA and LMEMs fit within that framework and introduce the theory of GAMs.
+
+\hypertarget{linear-models-lms-and-linear-mixed-models-lmems}{%
+\subsection{Linear Models (LMs) and Linear Mixed Models (LMEMs)}\label{linear-models-lms-and-linear-mixed-models-lmems}}
+
+Linear models (LMs) are those that assume a normal (Gaussian) distribution of the response, and only incorporate \emph{fixed effects} (such as rm-ANOVA).
+These are by far the models most commonly used to analyze data within the biomedical research community.
+On the other hand, Linear Mixed Effect Models (LMEMs) are those that assume a normally-distributed response but also incorporate \emph{random effects}, as it has been described in Section \ref{LMEM-case}.
+
+In matrix notation, a LM has the general form
+
+\begin{equation}
+\textbf{y} = \textbf{X} \boldsymbol{\beta}
+\label{eq:LM-matrix}
+\end{equation}
+
+Where \(\textbf{y}\) a vector of observations, \(\textbf{X}\) is a matrix that contains the measured values of the covariates, and \(\boldsymbol{\beta}\) is a vector of the parameters to be estimated.
+The ``model equation'' form of Equation \eqref{eq:LM-matrix} for a rm-ANOVA model of two treatment groups corresponds to Equation \eqref{eq:linear-model}.
+It is important to remember that the error term \(\varepsilon_{ijt}\) in Equation \eqref{eq:linear-model} represents the variability introduced by random sampling from the normal distribution.
+
+\hypertarget{generalized-linear-models-glms}{%
+\subsection{Generalized Linear Models (GLMs)}\label{generalized-linear-models-glms}}
+
+A major limitation of LMs is their assumption of normality.
+If the data is non-normal, a transformation is necessary in order to properly fit the model.
+However, transformation lead to poor model performance {[}\protect\hyperlink{ref-ohara2010}{51}{]}, and can cause problems with the biological interpretation of the model estimates.
+McCullagh and Nelder {[}\protect\hyperlink{ref-nelder1972}{52}{]} defined General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
+To achieve this, consider the following
+
+\begin{equation}
+y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi)
+\label{eq:GLM-y}
+\end{equation}
+
+Where \(y_{ijt}\) is the observation \(i\) in group \(j\) at time \(t\), which is assumed to come randomly from some distribution of the exponential family \(\mathcal{D}\), with some mean \(\mu_{ijt}\) and a dispersion parameter \(\phi\) (which in the Gaussian case is the variance \(\sigma^{2}\)).
+The mean \(\mu_{ijt}\) is also known as the \emph{expected value} (also called \emph{expectation}) (\(E\)) of the observed response.
+Then, the \emph{linear predictor} \(\eta\), which defines the relationship between the mean and the covariates can be defined as
+
+\begin{equation}
+\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}
+\label{eq:GLM-eta}
+\end{equation}
+
+Where \(\eta_{ijt}\) is the linear predictor for each observation, \(\beta_0\) is the intercept, and \(\beta_{j}\) is the model parameter for each group; and \(x_{ijt}\) represents the covariates from each subject in each group at each time point.
+
+Finally,
+
+\begin{equation}
+E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt})
+\label{eq:GLM-Expectation}
+\end{equation}
+Where \(E(y_{ijt})\) is the expectation, \(g^{-1}\) is the inverse of the \emph{link function} (\(g\)).
+The link function transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}).
+Therefore, LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
+
+\hypertarget{generalized-linear-mixed-models-glmms}{%
+\subsection{Generalized linear mixed models (GLMMs)}\label{generalized-linear-mixed-models-glmms}}
+
+Although GLMs relax the normality assumption, they can only accommodate fixed effects.
+Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate \emph{random effects}, which have an associated probability distribution {[}\protect\hyperlink{ref-mcculloch2001}{53}{]}.
+Therefore, in GLMMs the \emph{linear predictor} takes the form
+
+\begin{equation}
+\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\mu_{ij}
+\label{eq:GLMM-eta}
+\end{equation}
+
+Where \(\mu_{ij}\) corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of Equation \eqref{eq:GLM-eta}.
+Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed {[}\protect\hyperlink{ref-nelder1972}{52}{]}.
+In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in Dobson {[}\protect\hyperlink{ref-dobson2008}{54}{]} and Stroup {[}\protect\hyperlink{ref-stroup2013}{55}{]}.
 
 \hypertarget{GAM-theory}{%
-\section{GAMs as a special case of Generalized Linear Models}\label{GAM-theory}}
+\subsubsection{GAMs as a special case of Generalized Linear Models}\label{GAM-theory}}
 
 \hypertarget{gams-and-basis-functions}{%
-\subsection{GAMs and Basis Functions}\label{gams-and-basis-functions}}
-
-Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors {[}\protect\hyperlink{ref-nelder1972}{50}{]}.
-In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case{[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{51}{]}.
-A GAM model can be written as:
+\paragraph{GAMs and Basis Functions}\label{gams-and-basis-functions}}
+
+Generalized additive models (GAMs) are a family of regression-based methods for estimating smoothly varying trends and are an extension of the GLM family {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}{]}.
+A GAM version of a linear model can be written as:
 
@@
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.
+Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and the change of \(y_{ijt}\) over time (instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta}) is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual due to \textbf{random sampling}.
 
@@
 This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
+One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
 In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
@@
 The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t\mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
+A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
 For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{52}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{57}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
 In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}.
-Although different types of splines exist, cubic and thin plate regression splines will be briefly discussed next to give a general idea of the development within the GAM framework.
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{58}{]}.
+Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
 
 Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework {[}\protect\hyperlink{ref-hastie1987}{51}{]}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit {[}\protect\hyperlink{ref-wood2003}{54}{]}.
-In contrast, Wood {[}\protect\hyperlink{ref-wood2003}{54}{]} provided a computationally efficient implementation of thin plate regression splines(TPRS), which do not require the knot placement, and require simply a choice on the number of basis to be used to create the smoother (for mathematical details see refs.
-{[}\protect\hyperlink{ref-wood2003}{54}{]} and {[}\protect\hyperlink{ref-wood2017}{37}{]}).
-
-\textbf{To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
+The use of CS as smoothers in GAMs was discussed within the original GAM framework {[}\protect\hyperlink{ref-hastie1987}{56}{]}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit {[}\protect\hyperlink{ref-wood2003}{59}{]}.
+A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}{]}.
+
+In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner {[}\protect\hyperlink{ref-wood2003}{59}{]}.
+Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood{[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}{]}).
+
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
 A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
-Because the data has 6 unique time points, we can specify a maximum of six basis functions to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).}
+Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth.
+It is important to note that is not necessary to specify a number of basis equal to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data.
+
+Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present.
+A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but it suffices to say that the wiggliness (or ``roughness'') of the fit is achieved by using a smoothness parameter \(\lambda\), which is applied to the second derivative of the spline.
+The second derivative of the spline is a measure of curvature, or the rate of change of the slope {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
+A physics analogy to the penalization process would be that if we are driving a car on a hilly and curvy road, we want to minimize abrupt changes in speed in order to have a ``smooth'' ride, and we do this by penalizing the acceleration.
 
@@
 \hypertarget{a-bayesian-interpretation-of-gams}{%
-\section{A Bayesian interpretation of GAMs}\label{a-bayesian-interpretation-of-gams}}
-
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}\protect\hyperlink{ref-mcelreath2018}{55}{]}.
-In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case) {[}\protect\hyperlink{ref-miller2019}{56}{]}.
-Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{57}{]}.
-Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{56}{]} and {[}\protect\hyperlink{ref-marra2012}{58}{]}.
+\subsubsection{A Bayesian interpretation of GAMs}\label{a-bayesian-interpretation-of-gams}}
+
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}\protect\hyperlink{ref-mcelreath2018}{60}{]}.
+In the case of data that shows non-linear trends, the belief that the \emph{true} trend of the data is likely to be smooth rather than extremely ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language) {[}\protect\hyperlink{ref-miller2019}{61}{]}.
+Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (point-wise coverage or ``single point'' coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+
+To understand this last part, it is worth reminding that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function exists, taking into account the randomness introduced by the sampling process.
+Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie.
+In a ``point-wise'' CI for non-linear data, the ``true'' function will lie outside of the CI in certain regions at a given frequency (because the data does not follow a linear trend).
+This means that if a point-wise CI is obtained for 100 random samples, the total number of CIs that contain the true function \emph{entirely} is much less than the nominal value of the CI (95\%, usually).
+
+Contrary to this, an \emph{across the function} CI (like those estimated for GAMs using \emph{mgcv}) contains the true function through the entire range of the covariates (which would be time in the case of longitudinal data).
+For GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, the percentage of CIs entirely containing the true function over the entire time period would be close to the nominal value (i.e., if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function), thereby allowing more robust estimates from the model.
+In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in Miller {[}\protect\hyperlink{ref-miller2019}{61}{]}, Wood{[}\protect\hyperlink{ref-wood2017}{37}{]}, Simpson {[}\protect\hyperlink{ref-simpson2018}{34}{]} and Marra {[}\protect\hyperlink{ref-marra2012}{62}{]}.
 With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
@@
 
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
+The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors (Figure 3C in Vishwanath et.
+al.{[}\protect\hyperlink{ref-vishwanath2009}{16}{]}).
 In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
@@
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in {[}\protect\hyperlink{ref-hefley2017}{52}{]}.
+A detailed description on basis functions and correlations can be found in Hefley {[}\protect\hyperlink{ref-hefley2017}{57}{]}.
 
 For the data in Figure \ref{fig:sim-smooth-plot}, A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
-To estimate this, the model incorporates independent smooths for \emph{Group} and \emph{Day}, respectively.
+To estimate this, the model incorporates separate smooths for \emph{Group} and \emph{Day}, respectively.
 The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}).
@@
 
-This syntax specifies that \passthrough{\lstinline!gam\_02!} (named accordingly so it matches the model sequence from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots.
-The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML) {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using four basis functions (plus the intercept).
+The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]} (a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console).
+The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups.
+Although the default \passthrough{\lstinline!method!} used to estimate the smoothing parameters in \emph{mgcv} is generalized cross validation (GCV), Wood{[}\protect\hyperlink{ref-wood2017}{37}{]} showed the restricted maximum likelihood (REML) to be more resistant to overfitting while also easing the quantification of uncertainty in the smooth parameters; therefore in this manuscript REML is always used for smooth parameter estimation.
+
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{59}{]}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{60}{]}.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{63}{]}.
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{64}{]}.
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/sim-smooth-plot-1}
+{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/sim-smooth-plot-1}
 
@@
 
-Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
-Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}.
+Biomedical longitudinal data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
+Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}, where it is clear that these models do not capture the non-linear trend of the data, thereby causing a ``model misspecification error.''
+
 This ``model misspecification'' error, also is known as a ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{61}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.
-
-In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{62}{]}.
-This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+Polynomial effects can be used in LMEMs, and it could be argued that in Section \ref{simulation} a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects presents more questions than answers in this case (i.e., what is the appropriate polynomial effect to use?
+quadratic?, cubic?, what is the biological interpretation of a polynomial effect?).
+Moreover, polynomial effects have other associated issues that we have described in Section \ref{GAM-theory}.
+
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{65}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.
+
+In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{66}{]}.
+This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).
 
@@
 One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
-The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
+The model is therefore able to identify changes between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
 
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{63},\protect\hyperlink{ref-weissgerber2018}{64}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{67},\protect\hyperlink{ref-weissgerber2018}{68}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ma2012}{}}%
-\CSLLeftMargin{{[}48{]} }
-\CSLRightInline{Y.
-Ma, M.
-Mazumdar, S.G.
-Memtsoudis, Beyond repeated-measures analysis of variance, Regional Anesthesia and Pain Medicine.
-37 (2012) 99--105.
-\url{https://doi.org/10.1097/aap.0b013e31823ebc74}.}
+\leavevmode\vadjust pre{\hypertarget{ref-box1976}{}}%
+\CSLLeftMargin{{[}49{]} }
+\CSLRightInline{G.E.P. Box, Science and statistics, 71 (1976) 791--799. \url{https://doi.org/10.1080/01621459.1976.10480949}.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-nlme}{}}%
-\CSLLeftMargin{{[}49{]} }
+\CSLLeftMargin{{[}50{]} }
 \CSLRightInline{J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, {nlme}: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.}
 
+\leavevmode\vadjust pre{\hypertarget{ref-ohara2010}{}}%
+\CSLLeftMargin{{[}51{]} }
+\CSLRightInline{R. OHara, J. Kotze, Do not log-transform count data, (2010). \url{https://doi.org/10.1038/npre.2010.4136.1}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-nelder1972}{}}%
-\CSLLeftMargin{{[}50{]} }
+\CSLLeftMargin{{[}52{]} }
 \CSLRightInline{J.A. Nelder, R.W.M. Wedderburn, Generalized linear models, Journal of the Royal Statistical Society. Series A (General). 135 (1972) 370. \url{https://doi.org/10.2307/2344614}.}
 
+\leavevmode\vadjust pre{\hypertarget{ref-mcculloch2001}{}}%
+\CSLLeftMargin{{[}53{]} }
+\CSLRightInline{C. McCulloch, Generalized, linear, and mixed models, John Wiley \& Sons, New York, 2001.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-dobson2008}{}}%
+\CSLLeftMargin{{[}54{]} }
+\CSLRightInline{A. Dobson, An introduction to generalized linear models, CRC Press, Boca Raton, 2008.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-stroup2013}{}}%
+\CSLLeftMargin{{[}55{]} }
+\CSLRightInline{W. Stroup, Generalized linear mixed models : Modern concepts, methods and applications, CRC Press, Taylor \& Francis Group, Boca Raton, 2013.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-hastie1987}{}}%
-\CSLLeftMargin{{[}51{]} }
+\CSLLeftMargin{{[}56{]} }
 \CSLRightInline{T.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hefley2017}{}}%
-\CSLLeftMargin{{[}52{]} }
+\CSLLeftMargin{{[}57{]} }
 \CSLRightInline{T.J.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wegman1983}{}}%
-\CSLLeftMargin{{[}53{]} }
+\CSLLeftMargin{{[}58{]} }
 \CSLRightInline{E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wood2003}{}}%
-\CSLLeftMargin{{[}54{]} }
+\CSLLeftMargin{{[}59{]} }
 \CSLRightInline{S.N. Wood, Thin plate regression splines, 65 (2003) 95--114. \url{https://doi.org/10.1111/1467-9868.00374}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-mcelreath2018}{}}%
-\CSLLeftMargin{{[}55{]} }
+\CSLLeftMargin{{[}60{]} }
 \CSLRightInline{R. McElreath, Statistical rethinking, Chapman; Hall/{CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-miller2019}{}}%
-\CSLLeftMargin{{[}56{]} }
+\CSLLeftMargin{{[}61{]} }
 \CSLRightInline{D.L. Miller, Bayesian views of generalized additive modelling, arXiv Preprint arXiv:1902.01330. (2019).}
 
-\leavevmode\vadjust pre{\hypertarget{ref-laird1982}{}}%
-\CSLLeftMargin{{[}57{]} }
-\CSLRightInline{N.M. Laird, J.H. Ware, Random-effects models for longitudinal data, Biometrics. 38 (1982) 963. \url{https://doi.org/10.2307/2529876}.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
-\CSLLeftMargin{{[}58{]} }
+\CSLLeftMargin{{[}62{]} }
 \CSLRightInline{G.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
-\CSLLeftMargin{{[}59{]} }
+\CSLLeftMargin{{[}63{]} }
 \CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
-\CSLLeftMargin{{[}60{]} }
-\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with r, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
+\CSLLeftMargin{{[}64{]} }
+\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with {R}, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
-\CSLLeftMargin{{[}61{]} }
+\CSLLeftMargin{{[}65{]} }
 \CSLRightInline{T.A.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hastie1995}{}}%
-\CSLLeftMargin{{[}62{]} }
+\CSLLeftMargin{{[}66{]} }
 \CSLRightInline{T.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-begley2015}{}}%
-\CSLLeftMargin{{[}63{]} }
+\CSLLeftMargin{{[}67{]} }
 \CSLRightInline{C.G. Begley, J.P.A. Ioannidis, Reproducibility in science, Circulation Research. 116 (2015) 116--126. \url{https://doi.org/10.1161/circresaha.114.303819}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-weissgerber2018}{}}%
-\CSLLeftMargin{{[}64{]} }
+\CSLLeftMargin{{[}68{]} }
 \CSLRightInline{T.L.
@@
   ## Plot the simulated data (scatterplot)
-
+  txt<-20
   p1 <- sim_dat$dat %>%
@@
     theme_classic() +
-    theme(plot.title = element_text(size = 30, 
+    theme(plot.title = element_text(size = txt, 
                                   face = "bold"),
-        text=element_text(size=30))+
-    thm
+        text=element_text(size=txt))+
+    thm1
 
@@
     theme_classic()+
-     theme(plot.title = element_text(size = 30, 
+     theme(plot.title = element_text(size = txt, 
                                 face = "bold"),
-     text=element_text(size=30))+
-    thm
+     text=element_text(size=txt))+
+    thm1
 
@@
      theme_classic()+
-     theme(plot.title = element_text(size = 30, 
+     theme(plot.title = element_text(size = txt, 
                                   face = "bold"),
-        text=element_text(size=30))+
-    thm
+        text=element_text(size=txt))+
+    thm1
 
@@
 }
-
-txt<-18
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/linear-cases-Appendix-1}
+{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/linear-cases-Appendix-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/quadratic-cases-Appendix-1}
+{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/quadratic-cases-Appendix-1}
 
@@
 This code produces Figure \ref{fig:basis-plot} from the main manuscript.
-Briefly, a non-linear (quadratic) response is simulated a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
+Briefly, a non-linear (quadratic) response is simulated, a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
 The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
@@
 
-   #GAM model of time, 5 knots
+   #GAM model of time, 5 basis
 gm<-gam(y~s(time,k=5),data=dat)
@@
 \begin{lstlisting}[language=R]
-set.seed(1)
-#Dataframe that contains the original reported trends
 dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
@@
             )
-
-## plot the mean response
-f1<-ggplot(dat, 
-           aes(x = Day, 
-               y = StO2, 
-               color = Group)) +
-    geom_line(size=1,
-              show.legend = FALSE)+
-    geom_point(show.legend = FALSE,
-               size=1.5,
-               alpha=0.5)+
-  labs(y=expression(paste(StO[2],
-                          ' (real)')))+
-  theme_classic()+
-  thm+
-    scale_x_continuous(breaks=c(0,5,10))+
-    scale_y_continuous(breaks=c(0,40))+
-  plot_layout(tag_level = 'new')+
-  theme(
-    plot.background = element_rect(fill = "transparent", 
-                                   color = NA),
-    axis.text=element_text(size=14)
-  )
+#alpha for ribbon
+al<-0.8
 
@@
 #subject = factor(paste(subject, treatment, sep = "-")))
+
 n <- 10 #number of observations
@@
 #plotting simulated data
-f2<-ggplot(dat_sim, 
-           aes(x = Day, 
+f2<-ggplot(dat_sim, aes(x = Day, 
                         y = StO2_sim, 
-               color = Group)) +
+                        color = Group, 
+                        group=subject)) +
     geom_point(show.legend=FALSE,
                size=1.5,
-               alpha=0.5)+
-    stat_summary(aes(y = StO2_sim,
-                     group=Group), 
-                 fun=mean, geom="line",
-                 size=1,
+               alpha=0.6)+
+    geom_line(size=0.6, alpha=0.4)+
+    geom_line(aes(x = Day, 
+               y = StO2,
+               color=Group),
+           size=1.5,
+           data=dat,
+           inherit.aes=FALSE,
               show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],
-                         '(simulated)')))+
+  labs(y=expression(atop(StO[2],'(simulated)')))+
   theme_classic()+
@@
   )+
-  thm+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    thm1
 \end{lstlisting}
@@
 The model syntax specifies that \passthrough{\lstinline!gam\_00!} is the object that will contain all the model information, and that the model attempts to explain changes in \passthrough{\lstinline!StO2\_sim!} (simulated \(\mbox{StO}_2\)) using a smooth per \passthrough{\lstinline!Day!}.
-The model will use 5 knots (\passthrough{\lstinline!k=5!}) for the smooth.
+The model will use 5 basis functions (\passthrough{\lstinline!k=5!}) for the smooth.
 The smooth is constructed by default using thin plate regression splines.
@@
 #use function from the pairwise comparison plot in the manuscript to get the shaded regions
-
 my_list<-pairwise_limits(comp_StO2)
-  rib_col<-'#EDD03AFF' #color for the ribbon
+
 #plot the difference
+rib_col<-'#8D7D82' #color for ribbon for confidence interval
+control_rib <- '#875F79' #color for ribbon for control region
+treat_rib <- '#A7D89E' #color for ribbon treatment region
+
 c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
@@
                 xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill='#30123BFF',
+                fill=control_rib,
                 alpha = 0.5,
@@
              xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
+             fill= treat_rib,
              alpha = 0.5
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/sim-smooth-plot-Appendix-1}
+{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/sim-smooth-plot-Appendix-1}
 
@@
 my_list<-pairwise_limits(comp_StO2_full)
-rib_col<-'#EDD03AFF'
+rib_col<-'#8D7D82' #color for ribbon for confidence interval
+control_rib <- '#875F79' #color for ribbon for control region
+treat_rib <- '#A7D89E' #color for ribbon treatment region
 
@@
              xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
+             fill=treat_rib,
              alpha = 0.5,
@@
              xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-             fill='#30123BFF',
-             alpha = 0.5,
+             fill=control_rib,
+             alpha = 0.5
              ) +
@@
              label="Control>Treatment",
-           size=8
+           size=8,
+           angle=90
            )+
@@
              xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill='#7A0403FF',
+             fill= treat_rib,
              alpha = 0.5,
@@
              label="Treatment>Control",
-             size=8)+
+             size=8,
+           angle=90
+           )+
     geom_ribbon(aes(ymin = lower, ymax = upper),

### 2021-11-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated sections on wiggliness, GLMMs. Changed palette.
Updated code and figures

@@
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in [@hefley2017].
+A detailed description on basis functions and correlations can be found in Hefley [@hefley2017].
 
@@
 al<-0.8
-
-## plot the mean response
-f1<-ggplot(dat, 
-           aes(x = Day, 
-               y = StO2, 
-               color = Group)) +
-    geom_line(size=1,
-              show.legend = FALSE)+
-    geom_point(show.legend = FALSE,
-               size=1.5,
-               alpha=0.5)+
-  labs(y=expression(paste(StO[2],
-                          ' (real)')))+
-  theme_classic()+
-    scale_x_continuous(breaks=c(0,5,10))+
-    scale_y_continuous(breaks=c(0,40))+
-  plot_layout(tag_level = 'new')+
-  theme(
-    plot.background = element_rect(fill = "transparent", 
-                                   color = NA),
-    axis.text=element_text(size=14)
-  )+
-    thm1
 
@@
 my_list<-pairwise_limits(comp_StO2_full)
-rib_col<-'#EDD03AFF' #color for ribbon
+rib_col<-'#8D7D82' #color for ribbon for confidence interval
+control_rib <- '#875F79' #color for ribbon for control region
+treat_rib <- '#A7D89E' #color for ribbon treatment region
 
@@
              xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-             fill='#30123BFF',
+             fill=control_rib,
              alpha = 0.5

### 2021-11-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/07-Appendix.Rmd
Updated code and figures

@@
 
-```{r, linear-cases-Appendix, fig.width=14, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:linear-cases-appendix)'}
+```{r, linear-cases-Appendix, fig.width=14, fig.height=12, out.width='100%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:linear-cases-appendix)'}
 # linear response, correlated errors (subject effect)
@@
 
-```{r, quadratic-cases-Appendix, fig.width=12, fig.height=7, out.width='100%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:quadratic-caption-appendix)'}
+```{r, quadratic-cases-Appendix, fig.width=14, fig.height=12, out.width='100%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:quadratic-caption-appendix)'}
 # linear response, correlated errors (subject effect)
@@
 This code produces Figure \@ref(fig:basis-plot) from the main manuscript.
-Briefly, a non-linear (quadratic) response is simulated a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
+Briefly, a non-linear (quadratic) response is simulated, a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
 The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
@@
 ```{r,sim-tumor-response-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE,eval=FALSE}
-set.seed(1)
-#Dataframe that contains the original reported trends
 dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
@@
   )+
-  thm+
-    scale_x_continuous(breaks=c(0,2,5,7,10))
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    thm1
 
@@
 The model syntax specifies that `gam_00` is the object that will contain all the model information, and that the model attempts to explain changes in `StO2_sim` (simulated $\mbox{StO}_2$) using a smooth per `Day`.
-The model will use 5 knots (`k=5`) for the smooth.
+The model will use 5 basis functions (`k=5`) for the smooth.
 The smooth is constructed by default using thin plate regression splines.

### 2021-11-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Started adding the changes on GAM theory section

@@
 y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
+\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
 \end{cases}
@@
 y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
+\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
 \end{cases}
@@
 
-A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}} A LMEM with interaction between time and treatment for a longitudinal study can be written as:
+A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to account for\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}}.
+A LMEM with interaction between time and treatment for a longitudinal study can be written as:
 
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
+y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij} +\varepsilon_{ijt}\\.
 \label{eq:LMEM}
@@
 
-When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEMs and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)).
-This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)).
-The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the **``sampling** noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.
-
-For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data \textbf{that needs to be accounted for}.
+When Equations \eqref{eq:linear-model} and \eqref{eq:LMEM} are compared, it is noticeable that LMEMs and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\alpha_{ij}\)).
+This term \(\alpha_{ij}\) corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)).
+The \emph{random} component can also be understood as modeling some ``noise'' in the response, but that does not arise from the sampling error term \(\varepsilon_{ijt}\) from Equations \eqref{eq:linear-model} through \eqref{eq:ANOVA-lines}.
+
+For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data that needs to be accounted for.
 As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
-The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
+The \emph{random effect} \(\alpha_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
+In essence, the \emph{random effect} in a LMEM enables fitting models with different intercepts at the subject-level\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
 However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
 
-Of notice, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach, which is described in detail by Pinheiro and Bates\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}.
-However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates;\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} more importantly, polynomial effects lack biological or mechanistic interpretation\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} which limits their use in biomedical studies.
+Of note, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach (using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}), which is described in detail by Pinheiro and Bates.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates;\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} more importantly, polynomial effects lack biological or mechanistic interpretation\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} which limits their use in biomedical studies.
 
@@
 
-To visually demonstrate the limitations of rm-ANOVA an LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
+To visually demonstrate the limitations of rm-ANOVA and LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
 An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using R\textsuperscript{\protect\hyperlink{ref-r}{38}} and the package \emph{nlme}.\textsuperscript{\protect\hyperlink{ref-nlme}{50}}
 
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}A).
+Briefly, two cases for the mean response for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}A).
 In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}C).
@@
 
-A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where a simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) is presented.
 We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with ``default'' (linear) effects and, 2) demonstrate that a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
@@
 
-In reality, rm-ANOVA and LMEMs are just \emph{special cases} of broader model classes (General Linear Models and Generalized Linear Mixed Models, respectively).
-In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different Statistical model classes, indicate how rm-ANOVA and LMEMs fit within that framework and introduce the theory of GAMs.
-
-\hypertarget{linear-models-lms-and-linear-mixed-models-lmems}{%
-\subsection{Linear Models (LMs) and Linear Mixed Models (LMEMs)}\label{linear-models-lms-and-linear-mixed-models-lmems}}
-
 Linear models (LMs) are those that assume a normal (Gaussian) distribution of the response, and only incorporate \emph{fixed effects} (such as rm-ANOVA).
@@
 
-In matrix notation, a LM has the general form
-
-\begin{equation}
-\textbf{y} = \textbf{X} \boldsymbol{\beta}
-\label{eq:LM-matrix}
-\end{equation}
-
-Where \(\textbf{y}\) a vector of observations, \(\textbf{X}\) is a matrix that contains the measured values of the covariates, and \(\boldsymbol{\beta}\) is a vector of the parameters to be estimated.
-The ``model equation'' form of Equation \eqref{eq:LM-matrix} for a rm-ANOVA model of two treatment groups corresponds to Equation \eqref{eq:linear-model}.
-It is important to remember that the error term \(\varepsilon_{ijt}\) in Equation \eqref{eq:linear-model} represents the variability introduced by random sampling from the normal distribution.
+In reality, rm-ANOVA and LMEMs are just \emph{special cases} of a broader class of models (General Linear Models and Generalized Linear Mixed Models, respectively).
+In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different classes of models, indicate how rm-ANOVA and LMEMs fit within this framework and introduce the theory of GAMs.
 
@@
 If the data is non-normal, a transformation is necessary in order to properly fit the model.
-However, transformation lead to poor model performance,\textsuperscript{\protect\hyperlink{ref-ohara2010}{51}} and can cause problems with the biological interpretation of the model estimates.
-McCullagh and Nelder\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} defined General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
-To achieve this, consider the following
+However, transformation can lead to poor model performance,\textsuperscript{\protect\hyperlink{ref-ohara2010}{51}} and can cause problems with the biological interpretation of the model estimates.
+McCullagh and Nelder\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} introduced General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
+To achieve this, consider the following model
 
 \begin{equation}
-y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi)
+y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi),
 \label{eq:GLM-y}
@@
 
-Where \(y_{ijt}\) is the observation \(i\) in group \(j\) at time \(t\), which is assumed to come randomly from some distribution of the exponential family \(\mathcal{D}\), with some mean \(\mu_{ijt}\) and a dispersion parameter \(\phi\) (which in the Gaussian case is the variance \(\sigma^{2}\)).
-The mean \(\mu_{ijt}\) is also known as the \emph{expected value} (also called \emph{expectation}) (\(E\)) of the observed response.
+where \(y_{ijt}\) is the observation \(i\) in group \(j\) at time \(t\), that is assumed to come from some distribution of the exponential family \(\mathcal{D}\), with some mean \(\mu_{ijt}\) and potentially, a dispersion parameter \(\phi\) (which in the Gaussian case is the variance \(\sigma^{2}\)).
+The mean \(\mu_{ijt}\) is also known as the \emph{expected value} (also called \emph{expectation}) (\(E(y_{ijt})\)) of the observed response \(y_{ijt}\).
 Then, the \emph{linear predictor} \(\eta\), which defines the relationship between the mean and the covariates can be defined as
@@
 \begin{equation}
-E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt})
+E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt}),
 \label{eq:GLM-Expectation}
 \end{equation}
-Where \(E(y_{ijt})\) is the expectation, \(g^{-1}\) is the inverse of the \emph{link function} (\(g\)).
-The link function transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}).
+
+where \(E(y_{ijt})\) is the expectation and \(g^{-1}\) is the inverse of a \emph{link function} (\(g\)).
+The link function transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}.
 Therefore, LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
@@
 
-Although GLMs relax the normality assumption, they can only accommodate fixed effects.
+Although GLMs relax the normality assumption, they only accommodate fixed effects.
 Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate \emph{random effects}, which have an associated probability distribution.\textsuperscript{\protect\hyperlink{ref-mcculloch2001}{53}} Therefore, in GLMMs the \emph{linear predictor} takes the form
@@
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\mu_{ij}
+\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\alpha_{ij}
 \label{eq:GLMM-eta}
@@
 
-Generalized additive models (GAMs) are a family of regression-based methods for estimating smoothly varying trends and are an extension of the GLM family.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}} A GAM version of a linear model can be written as:
+Notice that in the previous sections, the difference between GLMs and GLMMs resides on the linear predictors (Equations \eqref{eq:GLM-eta} and \eqref{eq:GLMM-eta}).
+A Generalized additive model (GAM) is an extension of the GLM where the linear predictor has the form
+
+\begin{equation}
+\eta_{ijt}=\beta_{0}+ \beta_{1} \times treatment_{j} +f(time_{t}|\beta_{j}).
+\label{eq:GAM-eta}
+\end{equation}
+
+Where \(\beta_{0}\) is the intercept, \(\beta_{1}\) is the coefficient for each treatment group, and \(f(time_{t}|\beta_{j})\) represents the \emph{smooth function} of time, with coefficient \(\beta_{j}\).
+In this sense, GAMs are a family of regression-based methods for estimating smoothly varying trends.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}} A GAM version of a linear model can be written as:
 
@@
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and the change of \(y_{ijt}\) over time (instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta}) is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual due to \textbf{random sampling}.
+Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and the change of \(y_{ijt}\) over time (instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta}) is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
 
@@
 The four basis functions (and the intercept) are shown.
-Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the ``wiggliness'' of the fit, each of the basis functions of Figure \ref{fig:basis-plot}A is weighted by multiplying it by the smoothness parameter (\(\lambda\)) according to the matrix of Figure \ref{fig:basis-plot}B.
-The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
+Each of the basis functions is evaulated at six different points (because there are six points on the timeline).
+To control the ``wiggliness'' of the fit, the coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\).
+The smoothness parameter is multiplied by the integrated square of the second derivative of the spline expansion.
+As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
 A low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
+The penalized estimates fitted for our example are shown in Figure \ref{fig:basis-plot}B.
 

### 2021-11-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Started adding the changes on GAM theory section

@@
 
-A LMEM is a class of statistical models that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to distinguish [@pinheiro2006;@west2014].
+A LMEM is a class of statistical models that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to account for[@pinheiro2006;@west2014].
 A LMEM with interaction between time and treatment for a longitudinal study  can be written as:
@@
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
+y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij} +\varepsilon_{ijt}\\.
 (\#eq:LMEM)
@@
 
-When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is easily noticeable that LMEMs and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$).
-This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject (subject$_i$) within each group (group$_j$).
-The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the **"sampling** noise" term $\varepsilon_{ijt}$ from Equation \@ref(eq:linear-model).
+When Equations \@ref(eq:linear-model) and \@ref(eq:LMEM) are compared, it is noticeable that LMEMs and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\alpha_{ij}$).
+This term $\alpha_{ij}$ corresponds to the  _random effect_, accounting for variability in each subject (subject$_i$) within each group (group$_j$).
+The _random_ component can also be understood as modeling some "noise" in the response, but that does not arise from the sampling error term $\varepsilon_{ijt}$ from Equations \@ref(eq:linear-model) through \@ref(eq:ANOVA-lines).
 
-For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data **that needs to be accounted for**.
+For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data that needs to be accounted for.
 As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
-The _random effect_ $\mu_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
-In essence, the _random effect_ in a LMEM enables to fit models with different slopes at the subject-level[@pinheiro2006].
+The _random effect_ $\alpha_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
+In essence, the _random effect_ in a LMEM enables fitting models with different intercepts at the subject-level[@pinheiro2006].
 However, the expected linear relationship of the covariates and the response in Equation \@ref(eq:linear-model) and in Equation \@ref(eq:LMEM) is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
 
-Of notice, LMEMs are capable of fitting non-linear trends using an "empirical" approach, which is described in detail by Pinheiro and Bates [@pinheiro2006] using polynomial fixed effects instead of linear effects such as in Equation \@ref(eq:LMEM).
+Of note, LMEMs are capable of fitting non-linear trends using an "empirical" approach (using polynomial fixed effects instead of linear effects such as in Equation \@ref(eq:LMEM)), which is described in detail by Pinheiro and Bates [@pinheiro2006].
 However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates [@beck1998]; more importantly, polynomial effects lack biological or mechanistic interpretation [@pinheiro2006] which limits their use in biomedical studies.
@@
 
-To visually demonstrate the limitations of rm-ANOVA an LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
+To visually demonstrate the limitations of rm-ANOVA and LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
 An rm-ANOVA model (Equation \@ref(eq:linear-model)), and a LMEM  (Equation \@ref(eq:LMEM)) are fitted to each group, using R [@r] and the package _nlme_ [@nlme].
 
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response)A).
+Briefly, two cases for the mean response for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response)A).
 In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response)C).
@@
 
-A more comprehensive exploration  of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+A more comprehensive exploration  of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where a simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) is presented.
 We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.

### 2021-11-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Started adding the changes on GAM theory section

@@
 # Linear Models, and beyond
-
-In reality, rm-ANOVA and LMEMs are just _special cases_ of broader model classes (General Linear Models and Generalized Linear Mixed Models, respectively).
-In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different  Statistical model classes, indicate how rm-ANOVA and LMEMs fit within that framework and introduce the theory of GAMs.
-
-## Linear Models (LMs) and Linear Mixed Models (LMEMs)
 
@@
 
-In matrix notation, a LM has the general form
-
-\begin{equation}
-\textbf{y} = \textbf{X} \boldsymbol{\beta}
-(\#eq:LM-matrix)
-\end{equation}
-
-Where $\textbf{y}$ a vector of observations, $\textbf{X}$ is a matrix that contains the measured values of the covariates, and $\boldsymbol{\beta}$ is a vector of the parameters to be estimated.
-The "model equation" form of Equation \@ref(eq:LM-matrix) for a rm-ANOVA model of two treatment groups corresponds to Equation \@ref(eq:linear-model).
-It is important to remember that the error term $\varepsilon_{ijt}$ in Equation \@ref(eq:linear-model) represents the variability introduced by random sampling from the normal distribution.
+In reality, rm-ANOVA and LMEMs are just _special cases_ of a broader class of models (General Linear Models and Generalized Linear Mixed Models, respectively).
+In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different classes of models, indicate how rm-ANOVA and LMEMs fit within this framework and introduce the theory of GAMs.
 
@@
 If the data is non-normal, a transformation is necessary in order to properly fit the model.
-However, transformation lead to poor model performance [@ohara2010], and can cause problems with the biological interpretation of the model estimates.
-McCullagh and Nelder [@nelder1972] defined General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
-To achieve this, consider the following
+However, transformation can lead to poor model performance [@ohara2010], and can cause problems with the biological interpretation of the model estimates.
+McCullagh and Nelder [@nelder1972] introduced General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
+To achieve this, consider the following model
 
 \begin{equation}
-y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi)
+y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi),
 (\#eq:GLM-y)
@@
 
-Where $y_{ijt}$ is the observation $i$ in group $j$ at time $t$, which is assumed to come randomly from some distribution of the exponential family $\mathcal{D}$, with some mean $\mu_{ijt}$ and a dispersion parameter $\phi$ (which in the Gaussian case is the variance $\sigma^{2}$).
-The mean $\mu_{ijt}$ is also known as the _expected value_ (also called _expectation_) ($E$) of the observed response.
+where $y_{ijt}$ is the observation $i$ in group $j$ at time $t$, that is assumed to come from some distribution of the exponential family $\mathcal{D}$, with some mean $\mu_{ijt}$ and potentially, a dispersion parameter $\phi$ (which in the Gaussian case is the variance $\sigma^{2}$).
+The mean $\mu_{ijt}$ is also known as the _expected value_ (also called _expectation_) ($E(y_{ijt})$) of the observed response $y_{ijt}$.
 Then, the _linear predictor_ $\eta$, which defines the relationship between the mean and the covariates can be defined as
@@
 \begin{equation}
-E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt})
+E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt}),
 (\#eq:GLM-Expectation)
 \end{equation}
-Where $E(y_{ijt})$ is the expectation, $g^{-1}$ is the inverse of the _link function_  ($g$).
-The link function transforms the values from the response scale to the scale of the linear predictor $\eta$ (Equation \@ref(eq:GLM-eta)).
+
+where $E(y_{ijt})$ is the expectation and $g^{-1}$ is the inverse of a _link function_  ($g$).
+The link function transforms the values from the response scale to the scale of the linear predictor $\eta$ (Equation \@ref(eq:GLM-eta).
 Therefore, LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
@@
 
-Although GLMs relax the normality assumption, they can only accommodate fixed effects.
+Although GLMs relax the normality assumption, they only accommodate fixed effects.
 Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate _random effects_, which have an associated probability distribution [@mcculloch2001].
@@
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\mu_{ij}
+\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\alpha_{ij}
 (\#eq:GLMM-eta)
@@
 
-Generalized additive models (GAMs) are a family of regression-based methods for estimating smoothly varying trends and are an extension of the GLM family [@simpson2018;@wood2017;@hastie1987].
+Notice that in the previous sections, the difference between GLMs and GLMMs resides on the linear predictors (Equations \@ref(eq:GLM-eta) and \@ref(eq:GLMM-eta)).
+A Generalized additive model (GAM) is an extension of the GLM where the linear predictor has the form
+
+\begin{equation}
+\eta_{ijt}=\beta_{0}+ \beta_{1} \times treatment_{j} +f(time_{t}|\beta_{j}).
+(\#eq:GAM-eta)
+\end{equation}
+
+Where $\beta_{0}$ is the intercept, $\beta_{1}$ is the coefficient for each treatment group, and $f(time_{t}|\beta_{j})$ represents the _smooth function_ of time, with coefficient $\beta_{j}$.
+In this sense, GAMs are a family of regression-based methods for estimating smoothly varying trends  [@simpson2018;@wood2017;@hastie1987].
 A GAM version of a linear model can be written as:
@@
 
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, and the change of $y_{ijt}$ over time (instead of being defined by $\beta_{j}x_{ijt}$ as in Equation \@ref(eq:GLM-eta)) is represented by the _smooth function_ $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual due to **random sampling**.
+Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, and the change of $y_{ijt}$ over time (instead of being defined by $\beta_{j}x_{ijt}$ as in Equation \@ref(eq:GLM-eta)) is represented by the _smooth function_ $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the deviation of each observation from the mean.
 
@@
 The four basis functions (and the intercept) are shown.
-Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the "wiggliness" of the fit, each of the basis functions of Figure \@ref(fig:basis-plot)A is weighted by multiplying it by the smoothness parameter ($\lambda$) according to the matrix of Figure \@ref(fig:basis-plot)B.
-The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
+Each of the basis functions is evaulated at six different points (because there are six points on the timeline).
+To control the "wiggliness" of the fit, the coefficients for each of the basis functions of Figure \@ref(fig:basis-plot)A are estimated using a penalized regression with smoothness parameter $\lambda$.
+The smoothness parameter is multiplied by the integrated square of the second derivative of the spline expansion.
+As $\lambda$ increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
 A low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
+The penalized estimates fitted for our example are shown in Figure \@ref(fig:basis-plot)B.
 

### 2021-11-29 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Updated physics analogy

@@
 \begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
+y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
 \label{eq:linear-model}
@@
 In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response.
-This means that either presented as Equation \eqref{eq:linear-model}, Equation \eqref{eq:ANOVA-by-group} or Equation \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data.
+This means that either presented as Equations \eqref{eq:linear-model}, \eqref{eq:ANOVA-by-group} or \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data.
 If the data show non-linear trends, the rm-ANOVA model will approximate this behavior with non-parallel lines.
@@
 As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
-The \emph{random effect} \(\alpha_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
+The random effect \(\alpha_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
 In essence, the \emph{random effect} in a LMEM enables fitting models with different intercepts at the subject-level\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
@@
 In a longitudinal study, it is frequently the case that the number of observations is different across subjects.
-In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing,'' ``incomplete'' or (more generally speaking) unbalanced data.\textsuperscript{\protect\hyperlink{ref-molenberghs2004}{45}} The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size.\textsuperscript{\protect\hyperlink{ref-ma2012}{46}}
+In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing,'' ``incomplete,'' or (more generally speaking) unbalanced data.\textsuperscript{\protect\hyperlink{ref-molenberghs2004}{45}} The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
+This is inconvenient because the remaining subjects might not accurately represent the population and statistical power is affected by this reduction in sample size.\textsuperscript{\protect\hyperlink{ref-ma2012}{46}}
 
@@
 where \(y_{ijt}\) is the observation \(i\) in group \(j\) at time \(t\), that is assumed to come from some distribution of the exponential family \(\mathcal{D}\), with some mean \(\mu_{ijt}\) and potentially, a dispersion parameter \(\phi\) (which in the Gaussian case is the variance \(\sigma^{2}\)).
-The mean \(\mu_{ijt}\) is also known as the \emph{expected value} (also called \emph{expectation}) (\(E(y_{ijt})\)) of the observed response \(y_{ijt}\).
+The mean (\(\mu_{ijt}\)) is also known as the \emph{expected value} (or \emph{expectation}), \(E(y_{ijt})\) of the observed response \(y_{ijt}\).
 Then, the \emph{linear predictor} \(\eta\), which defines the relationship between the mean and the covariates can be defined as
@@
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}
+\eta_{ijt}=\beta_0+\beta_{j}x_{ijt},
 \label{eq:GLM-eta}
@@
 
-Where \(\eta_{ijt}\) is the linear predictor for each observation, \(\beta_0\) is the intercept, and \(\beta_{j}\) is the model parameter for each group; and \(x_{ijt}\) represents the covariates from each subject in each group at each time point.
+where \(\eta_{ijt}\) is the linear predictor for each observation, \(\beta_0\) is the intercept, and \(\beta_{j}\) is the model parameter for each group; and \(x_{ijt}\) represents the covariates from each subject in each group at each time point.
 
@@
 
-where \(E(y_{ijt})\) is the expectation and \(g^{-1}\) is the inverse of a \emph{link function} (\(g\)).
+where \(E(y_{ijt})\) is the expectation, and \(g^{-1}\) is the inverse of a \emph{link function} (\(g\)).
 The link function transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}.
-Therefore, LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
+Therefore, it can be seen that LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
 
@@
 Although GLMs relax the normality assumption, they only accommodate fixed effects.
-Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate \emph{random effects}, which have an associated probability distribution.\textsuperscript{\protect\hyperlink{ref-mcculloch2001}{53}} Therefore, in GLMMs the \emph{linear predictor} takes the form
+Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate \emph{random effects}, which have an associated probability distribution.\textsuperscript{\protect\hyperlink{ref-mcculloch2001}{53}} Therefore, in GLMMs the linear predictor takes the form
 
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\alpha_{ij}
+\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\alpha_{ij},
 \label{eq:GLMM-eta}
@@
 
-Where \(\mu_{ij}\) corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of Equation \eqref{eq:GLM-eta}.
-Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed.\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in Dobson\textsuperscript{\protect\hyperlink{ref-dobson2008}{54}} and Stroup.\textsuperscript{\protect\hyperlink{ref-stroup2013}{55}}
+where \(\alpha_{ij}\) corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of Equation \eqref{eq:GLM-eta}.
+Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed,\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} and GLMs are a special case of GLMMs where there are no random effects.
+In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in Dobson\textsuperscript{\protect\hyperlink{ref-dobson2008}{54}} and Stroup.\textsuperscript{\protect\hyperlink{ref-stroup2013}{55}}
 
@@
 
-Notice that in the previous sections, the difference between GLMs and GLMMs resides on the linear predictors (Equations \eqref{eq:GLM-eta} and \eqref{eq:GLMM-eta}).
-A Generalized additive model (GAM) is an extension of the GLM where the linear predictor has the form
+Notice that in the previous sections, the difference between GLMs and GLMMs resides on their linear predictors (Equations \eqref{eq:GLM-eta} and \eqref{eq:GLMM-eta}).
+Generalized additive models (GAMs) are an extension of the GLM family that allow the estimation of smoothly varying trends where the relationship between the covariates and the response is modeled using \emph{smooth functions}\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}}.
+In a GAM, the linear predictor has the form
 
@@
 
-Where \(\beta_{0}\) is the intercept, \(\beta_{1}\) is the coefficient for each treatment group, and \(f(time_{t}|\beta_{j})\) represents the \emph{smooth function} of time, with coefficient \(\beta_{j}\).
-In this sense, GAMs are a family of regression-based methods for estimating smoothly varying trends.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}} A GAM version of a linear model can be written as:
+Where \(\beta_{0}\) is the intercept, \(\beta_{1}\) is the coefficient for each treatment group, and \(f(time_{t}|\beta_{j})\) represents the smooth function of time, with coefficient \(\beta_{j}\).
+A GAM version of a linear model can be written as:
 
@@
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and the change of \(y_{ijt}\) over time (instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta}) is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible \emph{smooth functions}.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
+Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and the change of \(y_{ijt}\) over time (instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta}) is represented by the smooth function \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
+
+In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is more advantageous as it allows more flexibility because it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
 One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
 
-The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using a \emph{basis function} expansion of the covariates and by estimating random coefficients associated with these basis functions.
+A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(x_t)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
+The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
 In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{58}} Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
-
-Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{56}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate.\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}
-
-In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{58}} Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
+
+Cubic splines (CS) are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
+The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{56}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the basis functions meet) to obtain the finite basis, which could affect the model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate.\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}
+
+In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} Like TPS, TPRS only require the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).
 
@@
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
+A computational requisite in \emph{mgcv} is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
 Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth.
-It is important to note that is not necessary to specify a number of basis equal to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data.
+It is important to note that is not necessary to specify a number of basis equal to the number of unique values in the independent variable; fewer basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data.
 
 Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present.
-A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but it suffices to say that the wiggliness (or ``roughness'') of the fit is achieved by using a smoothness parameter \(\lambda\), which is applied to the second derivative of the spline.
-The second derivative of the spline is a measure of curvature, or the rate of change of the slope.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} A physics analogy to the penalization process would be that if we are driving a car on a hilly and curvy road, we want to minimize abrupt changes in speed in order to have a ``smooth'' ride, and we do this by penalizing the acceleration.
-
-If five basis functions are used to fit a GAM for the data that appears in Figure \ref{fig:l-q-response}C, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A.
+A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the second derivative of the spline.
+The second derivative of the spline is a measure of curvature, or the rate of change of the slope.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} To better understand the smooth fitting process, we can use a physics analogy.
+
+Suppose that we are traveling by car through a mountainous country and that we want to map our route (using a GPS device) on a specific road that only has one lane and that goes through a mountain range within that country; moreover, we want to see how our route changes if we want to optimize fuel consumption.
+In this analogy, the exact middle of the road is ``the truth'' or the ``mean function'' which we want to approximate through the route we will create.
+Moreover, because we know the ``truth'' in advance, we know that the road has many turns and ascends and descends multiple times, but overall it does it on a ``smooth'' way.
+The first time we travel the road, we impose no restrictions on fuel consumption (similar to setting \(\lambda \rightarrow 0\)), and we are constantly driving on the road shoulder, making abrupt stops without apparent reason, and also accelerating the vehicle suddenly at irregular intervals.
+When the trip is completed and the route taken is displayed along the ``truth'' on the GPS device, we notice that our route was ``rough'' and with a lot of ``noise'' because the trajectory of the vehicle deviated from the ``truth'' many times.
+
+Besides being a certainly unpleasant driving experience, we know that such route does not accurately approximate the known ``truth.'' Because of this, we drive on the same road again, but this time to achieve our goal of a smooth drive, we want to optimize our fuel consumption by penalizing the acceleration of the vehicle (the second derivative of position with respect to time).
+Practically, this translates to maintaining constant speed as much as possible (which would also penalize sudden stops and acceleration bursts) while also staying as much as possible in the center of lane.
+We repeat the trip again and decide to set a mean value for our speed, which results in a certain fuel consumption value.
+If we repeat this process many times, we will eventually find a value of mean speed that produces the best overall fuel consumption (we find an optimal value for \(\lambda\)).
+This in a crude sense, is the process that we use to penalize the wiggliness of the smooth.
+
+With this in mind, if five basis functions are used to fit a GAM for the data that appears in Figure \ref{fig:l-q-response}C, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A.
 The four basis functions (and the intercept) are shown.
-Each of the basis functions is evaulated at six different points (because there are six points on the timeline).
-To control the ``wiggliness'' of the fit, the coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\).
-The smoothness parameter is multiplied by the integrated square of the second derivative of the spline expansion.
+Each of the basis functions is evaluated at six different points (because there are six points on the timeline).
+To control the ``wiggliness'' of the fit, the coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), which is multiplied by the integrated square of the second derivative of the spline expansion.
 As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.

### 2021-11-29 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Updated physics analogy

@@
 \begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
+y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
 (\#eq:linear-model)
@@
 In other words, the rm-ANOVA model "expects" a linear relationship between the covariates and the response.
-This means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data.
+This means that either presented as Equations \@ref(eq:linear-model), \@ref(eq:ANOVA-by-group) or \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data.
 If the data show non-linear trends, the rm-ANOVA model will approximate this behavior with non-parallel lines.
@@
 As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
-The _random effect_ $\alpha_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
+The random effect $\alpha_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
 In essence, the _random effect_ in a LMEM enables fitting models with different intercepts at the subject-level[@pinheiro2006].
@@
 In a longitudinal study, it is frequently the case that the number of observations is different across subjects.
-In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as "missing", "incomplete" or (more generally speaking) unbalanced data [@molenberghs2004].
+In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as "missing", "incomplete", or (more generally speaking) unbalanced data [@molenberghs2004].
 The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].
+This is inconvenient because the remaining subjects might not accurately represent the population and statistical power is affected by this reduction in sample size [@ma2012].
 

### 2021-11-29 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated physics analogy

@@
 where $y_{ijt}$ is the observation $i$ in group $j$ at time $t$, that is assumed to come from some distribution of the exponential family $\mathcal{D}$, with some mean $\mu_{ijt}$ and potentially, a dispersion parameter $\phi$ (which in the Gaussian case is the variance $\sigma^{2}$).
-The mean $\mu_{ijt}$ is also known as the _expected value_ (also called _expectation_) ($E(y_{ijt})$) of the observed response $y_{ijt}$.
+The mean ($\mu_{ijt}$) is also known as the _expected value_ (or _expectation_), $E(y_{ijt})$ of the observed response $y_{ijt}$.
 Then, the _linear predictor_ $\eta$, which defines the relationship between the mean and the covariates can be defined as
@@
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}
+\eta_{ijt}=\beta_0+\beta_{j}x_{ijt},
 (\#eq:GLM-eta)
@@
 
-Where $\eta_{ijt}$ is the linear predictor for each observation, $\beta_0$ is the intercept, and $\beta_{j}$ is the model parameter for each group; and $x_{ijt}$ represents the covariates from each subject in each group at each time point.
+where $\eta_{ijt}$ is the linear predictor for each observation, $\beta_0$ is the intercept, and $\beta_{j}$ is the model parameter for each group; and $x_{ijt}$ represents the covariates from each subject in each group at each time point.
 
@@
 
-where $E(y_{ijt})$ is the expectation and $g^{-1}$ is the inverse of a _link function_  ($g$).
+where $E(y_{ijt})$ is the expectation, and $g^{-1}$ is the inverse of a _link function_  ($g$).
 The link function transforms the values from the response scale to the scale of the linear predictor $\eta$ (Equation \@ref(eq:GLM-eta).
-Therefore, LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
+Therefore, it can be seen that LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
 
@@
 Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate _random effects_, which have an associated probability distribution [@mcculloch2001].
-Therefore, in GLMMs the _linear predictor_ takes the form
+Therefore, in GLMMs the linear predictor takes the form
 
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\alpha_{ij}
+\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\alpha_{ij},
 (\#eq:GLMM-eta)
@@
 
-Where $\mu_{ij}$ corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of Equation \@ref(eq:GLM-eta).
-Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed [@nelder1972].
+where $\alpha_{ij}$ corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of Equation \@ref(eq:GLM-eta).
+Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed [@nelder1972], and GLMs are a special case of GLMMs where there are no random effects.
 In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in Dobson [@dobson2008] and Stroup [@stroup2013].
@@
 
-Notice that in the previous sections, the difference between GLMs and GLMMs resides on the linear predictors (Equations \@ref(eq:GLM-eta) and \@ref(eq:GLMM-eta)).
-A Generalized additive model (GAM) is an extension of the GLM where the linear predictor has the form
+Notice that in the previous sections, the difference between GLMs and GLMMs resides on their linear predictors (Equations \@ref(eq:GLM-eta) and \@ref(eq:GLMM-eta)).
+Generalized additive models (GAMs) are an extension of the GLM family that allow the estimation of smoothly varying trends where the relationship between the covariates and the response is modeled using _smooth functions_[@simpson2018;@wood2017;@hastie1987].
+In a GAM, the linear predictor has the form
 
@@
 
-Where $\beta_{0}$ is the intercept, $\beta_{1}$ is the coefficient for each treatment group, and $f(time_{t}|\beta_{j})$ represents the _smooth function_ of time, with coefficient $\beta_{j}$.
-In this sense, GAMs are a family of regression-based methods for estimating smoothly varying trends  [@simpson2018;@wood2017;@hastie1987].
+Where $\beta_{0}$ is the intercept, $\beta_{1}$ is the coefficient for each treatment group, and $f(time_{t}|\beta_{j})$ represents the smooth function of time, with coefficient $\beta_{j}$.
 A GAM version of a linear model can be written as:
@@
 
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, and the change of $y_{ijt}$ over time (instead of being defined by $\beta_{j}x_{ijt}$ as in Equation \@ref(eq:GLM-eta)) is represented by the _smooth function_ $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the deviation of each observation from the mean.
+Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, and the change of $y_{ijt}$ over time (instead of being defined by $\beta_{j}x_{ijt}$ as in Equation \@ref(eq:GLM-eta)) is represented by the smooth function $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the deviation of each observation from the mean.
 
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
+In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is more advantageous as it allows more flexibility because it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
 One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with inference [@beck1998].
@@
 
-The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using _basis function_ expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A _basis_ is a set of functions that spans the mathematical space within which the true but unknown $f(x_t\mid \beta_j)$  is thought to exist [@simpson2018].
+The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using a _basis function_ expansion of the covariates and by estimating random coefficients associated with these basis functions.
+A _basis_ is a set of functions that spans the mathematical space within which the true but unknown $f(x_t)$  is thought to exist [@simpson2018].
 For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
+The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
 In the case of Equation \@ref(eq:GAM), the basis functions are contained in the expression $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
 
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as _basis functions_, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983].
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness [@wegman1983].
 Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
 
-Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework [@hastie1987], but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as 'knots', the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit [@wood2003].
+Cubic splines (CS) are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
+The use of CS as smoothers in GAMs was discussed within the original GAM framework [@hastie1987], but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as 'knots', the points where the basis functions meet) to obtain the finite basis, which could affect the model fit [@wood2003].
 A solution to the "knot" placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate [@wood2003; @wood2017].
@@
 In contrast, thin plate regression splines (TPRS) provide a reasonable "low rank" (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner [@wood2003].
-Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood[@wood2003;@wood2017]).
+Like TPS, TPRS only require the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood[@wood2003;@wood2017]).
 
@@
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
+A computational requisite in _mgcv_ is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
 Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth.
-It is important to note that is not necessary to specify a number of basis equal to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data.
+It is important to note that is not necessary to specify a number of basis equal to the number of unique values in the independent variable; fewer basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data.
 
 Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too "wiggly" due to the noise present.
-A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but it suffices to say that the wiggliness (or "roughness") of the fit is achieved by using a smoothness parameter $\lambda$, which is applied to the second derivative of the spline.
+A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence the wiggliness (or "roughness") of the fit is achieved by using a _smoothness parameter_ ($\lambda$), which is used to penalize the second derivative of the spline.
 The second derivative of the spline is a measure of curvature, or the rate of change of the slope [@simpson2018; @wood2017].
-A physics analogy to the penalization process would be that if we are driving a car on a hilly and curvy road, we want to minimize abrupt changes in speed  in order to have a "smooth" ride, and we do this by penalizing the acceleration.
+To better understand the smooth fitting process, we can use a physics analogy.
 
-If five basis functions are used to fit a GAM for the data that appears in Figure \@ref(fig:l-q-response)C, the resulting fitting process is shown in Figure \@ref(fig:basis-plot)A.
+Suppose that we are traveling by car through a mountainous country and that we want to map our route (using a GPS device) on a specific road that only has one lane and that goes through a mountain range within that country; moreover, we want to see how our route changes if we want to optimize fuel consumption.
+In this analogy, the exact middle of the road is "the truth" or the "mean function" which we want to approximate through the route we will create.
+Moreover, because we know the "truth" in advance, we know that the road has many turns and ascends and descends multiple times, but overall it does it on a "smooth" way.
+The first time we travel the road, we impose no restrictions on fuel consumption (similar to setting $\lambda \rightarrow 0$), and we are constantly driving on the road shoulder, making abrupt stops without apparent reason, and also accelerating the vehicle suddenly at irregular intervals.
+When the trip is completed and the route taken is displayed along the "truth" on the GPS device, we notice that our route was "rough" and with a lot of "noise" because the trajectory of the vehicle deviated from the "truth" many times.
+
+Besides being a certainly unpleasant driving experience, we know that such route does not accurately approximate the known "truth".
+Because of this, we drive on the same road again, but this time to achieve our goal of a smooth drive, we want to optimize our fuel consumption by penalizing the acceleration of the vehicle (the second derivative of position with respect to time).
+Practically, this translates to maintaining constant speed as much as possible (which would also penalize sudden stops and acceleration bursts) while also staying as much as possible in the center of lane.
+We repeat the trip again and decide to set a mean value for our speed, which results in a certain fuel consumption value.
+If we repeat this process many times, we will eventually find a value of mean speed that produces the best overall fuel consumption (we find an optimal value for $\lambda$).
+This in a crude sense, is the process that we use to penalize the wiggliness of the smooth.
+
+With this in mind, if five basis functions are used to fit a GAM for the data that appears in Figure \@ref(fig:l-q-response)C, the resulting fitting process is shown in Figure \@ref(fig:basis-plot)A.
 The four basis functions (and the intercept) are shown.
-Each of the basis functions is evaulated at six different points (because there are six points on the timeline).
-To control the "wiggliness" of the fit, the coefficients for each of the basis functions of Figure \@ref(fig:basis-plot)A are estimated using a penalized regression with smoothness parameter $\lambda$.
-The smoothness parameter is multiplied by the integrated square of the second derivative of the spline expansion.
+Each of the basis functions is evaluated at six different points (because there are six points on the timeline).
+To control the "wiggliness" of the fit, the coefficients for each of the basis functions of Figure \@ref(fig:basis-plot)A are estimated using a penalized regression with smoothness parameter $\lambda$, which is multiplied by the integrated square of the second derivative of the spline expansion.
 As $\lambda$ increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
@@
    #GAM model of time, 5 basis functions
-gm<-gam(y~s(time,k=5),data=dat)
+gm<-gam(y~s(time,k=5),data=dat, method= "REML")
 

### 2021-11-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Updated manuscript for GLM and GLMM notation

@@
 \begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
+y_{ijt} = \beta_0+\beta_1 \times treatment_{j} +\beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
 \label{eq:linear-model}
@@
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_{j}x_{ijt},
+\eta_{ijt}=\beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j},
 \label{eq:GLM-eta}
@@
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\alpha_{ij},
+\eta_{ijt}=\beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij},
 \label{eq:GLMM-eta}
@@
 
-Where \(\beta_{0}\) is the intercept, \(\beta_{1}\) is the coefficient for each treatment group, and \(f(time_{t}|\beta_{j})\) represents the smooth function of time, with coefficient \(\beta_{j}\).
+Where \(\beta_{0}\) is the intercept, \(\beta_{1}\) is the coefficient for each treatment group.
+Notice that we and \(f(time_{t}|\beta_{j})\) represents the smooth function of time, with coefficient \(\beta_{j}\).
 A GAM version of a linear model can be written as:
@@
 \begin{equation}
-  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
+  y_{ijt}=\beta_0+f(time_t\mid \beta_j)+\varepsilon_{ijt}
   \label{eq:GAM}
@@
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and the change of \(y_{ijt}\) over time (instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta}) is represented by the smooth function \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
+Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and the change of \(y_{ijt}\) over time (instead of being defined by \(\beta_{j}time_{ijt}\) as in Equation \eqref{eq:GLM-eta}) is represented by the smooth function \(f(time_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
 
 In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is more advantageous as it allows more flexibility because it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
+One possible set of functions for \(f(time_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(time_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
 
 The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using a \emph{basis function} expansion of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(x_t)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
+A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
 The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
-In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
+In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(time_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
@@
 Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present.
-A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the second derivative of the spline.
-The second derivative of the spline is a measure of curvature, or the rate of change of the slope.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} To better understand the smooth fitting process, we can use a physics analogy.
-
-Suppose that we are traveling by car through a mountainous country and that we want to map our route (using a GPS device) on a specific road that only has one lane and that goes through a mountain range within that country; moreover, we want to see how our route changes if we want to optimize fuel consumption.
-In this analogy, the exact middle of the road is ``the truth'' or the ``mean function'' which we want to approximate through the route we will create.
-Moreover, because we know the ``truth'' in advance, we know that the road has many turns and ascends and descends multiple times, but overall it does it on a ``smooth'' way.
-The first time we travel the road, we impose no restrictions on fuel consumption (similar to setting \(\lambda \rightarrow 0\)), and we are constantly driving on the road shoulder, making abrupt stops without apparent reason, and also accelerating the vehicle suddenly at irregular intervals.
-When the trip is completed and the route taken is displayed along the ``truth'' on the GPS device, we notice that our route was ``rough'' and with a lot of ``noise'' because the trajectory of the vehicle deviated from the ``truth'' many times.
-
-Besides being a certainly unpleasant driving experience, we know that such route does not accurately approximate the known ``truth.'' Because of this, we drive on the same road again, but this time to achieve our goal of a smooth drive, we want to optimize our fuel consumption by penalizing the acceleration of the vehicle (the second derivative of position with respect to time).
-Practically, this translates to maintaining constant speed as much as possible (which would also penalize sudden stops and acceleration bursts) while also staying as much as possible in the center of lane.
-We repeat the trip again and decide to set a mean value for our speed, which results in a certain fuel consumption value.
-If we repeat this process many times, we will eventually find a value of mean speed that produces the best overall fuel consumption (we find an optimal value for \(\lambda\)).
-This in a crude sense, is the process that we use to penalize the wiggliness of the smooth.
+A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood with a multiple \(\lambda\) times the integrated square of the second derivative of the spline.
+The second derivative of the spline is a measure of curvature, or the rate of change of the slope.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} Increasing the penalty by increasing \(\lambda\) results in models with less curvature.
+As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
+A low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
 
@@
 Each of the basis functions is evaluated at six different points (because there are six points on the timeline).
-To control the ``wiggliness'' of the fit, the coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), which is multiplied by the integrated square of the second derivative of the spline expansion.
-As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
-A low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
+The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\) estimated when fitting the model.
 The penalized estimates fitted for our example are shown in Figure \ref{fig:basis-plot}B.
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}} In the case of data that shows non-linear trends, the belief that the \emph{true} trend of the data is likely to be smooth rather than extremely ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (point-wise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}} In the case of data that shows non-linear trends, the prior that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good \emph{across the function} (``frequentist'') coverage\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
 
@@
 Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie.
-In a ``point-wise'' CI for non-linear data, the ``true'' function will lie outside of the CI in certain regions at a given frequency (because the data does not follow a linear trend).
-This means that if a point-wise CI is obtained for 100 random samples, the total number of CIs that contain the true function \emph{entirely} is much less than the nominal value of the CI (95\%, usually).
-
-Contrary to this, an \emph{across the function} CI (like those estimated for GAMs using \emph{mgcv}) contains the true function through the entire range of the covariates (which would be time in the case of longitudinal data).
-For GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, the percentage of CIs entirely containing the true function over the entire time period would be close to the nominal value (i.e., if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function), thereby allowing more robust estimates from the model.
+In an \emph{across the function} CI (like those estimated for GAMs using \emph{mgcv}) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%).
+In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI.
+For this to occur then, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
 In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in Miller,\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} and Marra.\textsuperscript{\protect\hyperlink{ref-marra2012}{62}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.

### 2021-11-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Updated manuscript for GLM and GLMM notation

@@
 \begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
+y_{ijt} = \beta_0+\beta_1 \times treatment_{j} +\beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
 (\#eq:linear-model)
@@
 
-In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
-Independent errors $\varepsilon_{ijt}$ represent **random variation from the sampling process ** assumed to be $\stackrel{i.i.d.}\sim N(0,\sigma^2)$ (independently and identically normally distributed with mean zero and variance $\sigma^2$).
+In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of treatment ($treatment_j$), time ($time_t$), and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
+Independent errors $\varepsilon_{ijt}$ represent random variation from the sampling process assumed to be $\stackrel{i.i.d.}\sim N(0,\sigma^2)$ (independently and identically normally distributed with mean zero and variance $\sigma^2$).
 In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs.
@@
 y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
-\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
+\beta_0 + \beta_2\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
+\beta_0 + \beta_1+\beta_2 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
 \end{cases}
@@
 
-To further simplify the expression, substitute $\widetilde{\beta_{0}}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B.
+To further simplify the expression, substitute $\widetilde{\beta_{0}}=\beta_0+\beta_{1}$ and $\widetilde{\beta_{1}}=\beta_{2}+\beta_{3}$ in the equation for Group B.
 This substitution allows for a different intercept and slope for Groups A and B.
@@
 y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
+\beta_0 + \beta_2\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
 \widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
@@
 A LMEM is a class of statistical models that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to account for[@pinheiro2006;@west2014].
-A LMEM with interaction between time and treatment for a longitudinal study  can be written as:
+A LMEM with interaction between time and treatment for a longitudinal study  can be written as
 
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij} +\varepsilon_{ijt}\\.
+y_{ijt} = \beta_0+ \beta_1 \times treatment_{j}+ \beta_2 \times time_{t} + \beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij} +\varepsilon_{ijt}\\.
 (\#eq:LMEM)

### 2021-11-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated manuscript for GLM and GLMM notation

@@
 
-Linear models (LMs) are those that assume a normal (Gaussian) distribution of the response, and only incorporate _fixed effects_ (such as rm-ANOVA).
+Linear models (LMs) are those that assume a normal (Gaussian) distribution of the errors, and only incorporate _fixed effects_ (such as rm-ANOVA).
 These are by far the models most commonly used to analyze data within the biomedical research community.
-On the other hand, Linear Mixed Effect Models (LMEMs) are those that assume a normally-distributed response but also incorporate _random effects_, as it has been described in Section \@ref(LMEM-case).
+On the other hand, Linear Mixed Effect Models (LMEMs) also incorporate _random effects_, as it has been described in Section \@ref(LMEM-case).
 
@@
 
-A major limitation of LMs is their assumption of normality.
-If the data is non-normal, a transformation is necessary in order to properly fit the model.
+A major limitation of LMs is their assumption of normality in the errors.
+If the residuals are non-normal, a transformation is necessary in order to properly fit the model.
 However, transformation can lead to poor model performance [@ohara2010], and can cause problems with the biological interpretation of the model estimates.
-McCullagh and Nelder [@nelder1972] introduced General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
+McCullagh and Nelder [@nelder1972] introduced General Linear Models (GLMs) as an extension of LMs, where the errors do not need to be normally distributed.
 To achieve this, consider the following model
@@
 The mean ($\mu_{ijt}$) is also known as the _expected value_ (or _expectation_), $E(y_{ijt})$ of the observed response $y_{ijt}$.
+
 Then, the _linear predictor_ $\eta$, which defines the relationship between the mean and the covariates can be defined as
@@
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_{j}x_{ijt},
+\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} +\beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j},
 (\#eq:GLM-eta)
@@
 
-where $\eta_{ijt}$ is the linear predictor for each observation, $\beta_0$ is the intercept, and $\beta_{j}$ is the model parameter for each group; and $x_{ijt}$ represents the covariates from each subject in each group at each time point.
+where $\eta_{ijt}$ is the linear predictor for each observation in each group at each timepoint, and $\beta_0$ (the intercept), $\beta_{1}, \beta_2,$ and $\beta_3$ are the model parameters for each group, which can be referred globally to as $\beta_j$.
+Finally, $time_{ijt}$ represents the covariates from each subject in each group at each time point.
 
@@
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\alpha_{ij},
+\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} + \beta_2 \times time_{t} + +\beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij},
 (\#eq:GLMM-eta)
@@
 
-Notice that in the previous sections, the difference between GLMs and GLMMs resides on their linear predictors (Equations \@ref(eq:GLM-eta) and \@ref(eq:GLMM-eta)).
-Generalized additive models (GAMs) are an extension of the GLM family that allow the estimation of smoothly varying trends where the relationship between the covariates and the response is modeled using _smooth functions_[@simpson2018;@wood2017;@hastie1987].
+Notice that in the previous sections, the difference between GLMs and GLMMs resides on their linear predictors (Equations \@ref(eq:GLM-eta), \@ref(eq:GLMM-eta)).
+Generalized additive models (GAMs) are an extension of the GLM family that allow the estimation of smoothly varying trends where the relationship between the covariates and the response is modeled using _smooth functions_ [@simpson2018;@wood2017;@hastie1987].
 In a GAM, the linear predictor has the form
@@
 
-Where $\beta_{0}$ is the intercept, $\beta_{1}$ is the coefficient for each treatment group, and $f(time_{t}|\beta_{j})$ represents the smooth function of time, with coefficient $\beta_{j}$.
+Where $\beta_{0}$ is the intercept, and $\beta_{1}$ is the coefficient for each treatment group.
+Notice that the construction of the predictor is similar to that of Equation \@ref(GLM-eta), but in this case the parametric terms involving the effect of time have been replaced by $f(time_{t}|\beta_{j})$, which represents the smooth function of time with with inputs as the covariates $time_t$ and parameters $\beta_{j}$.
 A GAM version of a linear model can be written as:
@@
 \begin{equation}
-  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
+  y_{ijt}=\beta_0+f(time_t\mid \beta_j)+\varepsilon_{ijt}
   (\#eq:GAM)
@@
 
-Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, and the change of $y_{ijt}$ over time (instead of being defined by $\beta_{j}x_{ijt}$ as in Equation \@ref(eq:GLM-eta)) is represented by the smooth function $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the deviation of each observation from the mean.
+Where $y_{ijt}$ is the response at time $t$ of subject $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the smooth function $f(time_t\mid \beta_j)$,  and $\varepsilon_{ijt}$ represents the deviation of each observation from the mean.
 
 In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is more advantageous as it allows more flexibility because it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with inference [@beck1998].
-In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic and causes bias at the endpoints of the time period.
+One possible set of functions for $f(time_t\mid \beta_j)$ that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with inference [@beck1998].
+In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(time_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic and causes bias at the endpoints of the time period.
 
 The smooth functional relationship between the covariates and the response in GAMs is specified  using a semi-parametric relationship that can be fit within the GLM framework, by using a _basis function_ expansion of the covariates and by estimating random coefficients associated with these basis functions.
-A _basis_ is a set of functions that spans the mathematical space within which the true but unknown $f(x_t)$  is thought to exist [@simpson2018].
-For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$.
+A _basis_ is a set of functions that spans the mathematical space within which the true but unknown $f(time_t)$  is thought to exist [@simpson2018].
+For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $treatment_j$, $time_t$, and $time_t \times treatment_j$.
 The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.
-In the case of Equation \@ref(eq:GAM), the basis functions are contained in the expression $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
+In the case of Equation \@ref(eq:GAM), the basis functions are contained in the expression $f(time_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
 
@@
 Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too "wiggly" due to the noise present.
-A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence the wiggliness (or "roughness") of the fit is achieved by using a _smoothness parameter_ ($\lambda$), which is used to penalize the second derivative of the spline.
-The second derivative of the spline is a measure of curvature, or the rate of change of the slope [@simpson2018; @wood2017].
-To better understand the smooth fitting process, we can use a physics analogy.
-
-Suppose that we are traveling by car through a mountainous country and that we want to map our route (using a GPS device) on a specific road that only has one lane and that goes through a mountain range within that country; moreover, we want to see how our route changes if we want to optimize fuel consumption.
-In this analogy, the exact middle of the road is "the truth" or the "mean function" which we want to approximate through the route we will create.
-Moreover, because we know the "truth" in advance, we know that the road has many turns and ascends and descends multiple times, but overall it does it on a "smooth" way.
-The first time we travel the road, we impose no restrictions on fuel consumption (similar to setting $\lambda \rightarrow 0$), and we are constantly driving on the road shoulder, making abrupt stops without apparent reason, and also accelerating the vehicle suddenly at irregular intervals.
-When the trip is completed and the route taken is displayed along the "truth" on the GPS device, we notice that our route was "rough" and with a lot of "noise" because the trajectory of the vehicle deviated from the "truth" many times.
-
-Besides being a certainly unpleasant driving experience, we know that such route does not accurately approximate the known "truth".
-Because of this, we drive on the same road again, but this time to achieve our goal of a smooth drive, we want to optimize our fuel consumption by penalizing the acceleration of the vehicle (the second derivative of position with respect to time).
-Practically, this translates to maintaining constant speed as much as possible (which would also penalize sudden stops and acceleration bursts) while also staying as much as possible in the center of lane.
-We repeat the trip again and decide to set a mean value for our speed, which results in a certain fuel consumption value.
-If we repeat this process many times, we will eventually find a value of mean speed that produces the best overall fuel consumption (we find an optimal value for $\lambda$).
-This in a crude sense, is the process that we use to penalize the wiggliness of the smooth.
+A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or "roughness") of the fit is achieved by using a _smoothness parameter_ ($\lambda$), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline.
+The second derivative of the spline is a measure of curvature, or the rate of change of the slope [@simpson2018; @wood2017], and increasing the penalty by increasing $\lambda$ results in models with less curvature.
+As $\lambda$ increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
+In other words,a low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
 
@@
 Each of the basis functions is evaluated at six different points (because there are six points on the timeline).
-To control the "wiggliness" of the fit, the coefficients for each of the basis functions of Figure \@ref(fig:basis-plot)A are estimated using a penalized regression with smoothness parameter $\lambda$, which is multiplied by the integrated square of the second derivative of the spline expansion.
-As $\lambda$ increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
-A low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
+The coefficients for each of the basis functions of Figure \@ref(fig:basis-plot)A are estimated using a penalized regression with smoothness parameter $\lambda$, that is estimated when fitting the model.
 The penalized estimates fitted for our example are shown in Figure \@ref(fig:basis-plot)B.
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief [@mcelreath2018].
-In the case of data that shows non-linear trends, the belief that the _true_ trend of the data is likely to be smooth rather than extremely "wiggly" introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs [@wood2017].
-Moreover, GAMs are considered "empirical" Bayesian models when fitted using the package _mgcv_ because the smoothing parameters are estimated from the data (and not from a prior distribution as in the "fully Bayesian" case, which can be fitted using JAGS, Stan, or other probabilistic programming language) [@miller2019].
-Therefore, the confidence intervals (CIs) calculated for the smooth terms using _mgcv_ are considered empirical Bayesian posterior credible intervals [@pedersen2019], which have good "frequentist" coverage (point-wise coverage or "single point" coverage), and _across the function_ coverage [@wood2017].
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge [@mcelreath2018].
+In the case of data that shows non-linear trends, the prior that the _true_ trend of the data is likely to be smooth rather than "wiggly" introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs [@wood2017].
+Moreover, GAMs are considered "empirical" Bayesian models when fitted using the package _mgcv_ because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the "fully Bayesian" case, which can be fitted using JAGS, Stan, or other probabilistic programming language) [@miller2019].
+Therefore, the confidence intervals (CIs) calculated for the smooth terms using _mgcv_ are considered empirical Bayesian credible intervals [@pedersen2019], which have good _across the function_ ("frequentist") coverage[@wood2017].
 
@@
 Because random samples from the population are used to calculate the "true" value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95%) where the function is expected to lie.
-In a "point-wise" CI for non-linear data, the "true" function will lie outside of the CI in certain regions at a given frequency (because the data does not follow a linear trend).
-This means that if a point-wise CI is obtained for 100 random samples, the total number of CIs that contain the true function _entirely_ is much less than the nominal value of the CI (95%, usually).
-
-Contrary to this, an _across the function_ CI (like those estimated for GAMs using _mgcv_) contains the true function through the entire range of the covariates (which would be time in the case of longitudinal data).
-For GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, the percentage of CIs entirely containing the true function over the entire time period would be close to the nominal value (i.e., if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function), thereby allowing more robust estimates from the model.
+In an _across the function_ CI (like those estimated for GAMs using _mgcv_) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95%).
+In other words, we expect that about 95% of the points that compose the true function will be covered by the across the function CI and for this to occur, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
 In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in Miller [@miller2019], Wood[@wood2017], Simpson [@simpson2018] and Marra [@marra2012].

### 2021-11-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated manuscript for GLM and GLMM notation

@@
                alpha=0.6)+
-    geom_line(size=0.6, alpha=0.4)+
+    geom_line(size=0.6, alpha=0.4,show.legend=FALSE)+
     geom_line(aes(x = Day, 
@@
     X <- xp[row1, ] - xp[row2, ]
-    ## zero out cols of X related to splines for other lochs
+   ## zero out cols of X not involved in the comparison
     X[, ! (col1 | col2)] <- 0
@@
 Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the _difference_ between the empirical Bayesian confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that time point(s).
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in `gam_02`) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
+**add sentence about the issue of the Nychka across the function intervals **
+
+around that  _difference_ between the empirical Bayesian confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that time point(s).
 The absence of a _p-value_ in this case might seem odd, but the empirical Bayesian confidence interval interpretation can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.

### 2021-11-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/07-Appendix.Rmd
Updated manuscript for GLM and GLMM notation

@@
     X <- xp[row1, ] - xp[row2, ]
-    ## zero out cols of X related to splines for other lochs
+    ## zero out cols of X not involved in the comparison
     X[, ! (col1 | col2)] <- 0

### 2021-12-01 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.Rmd
Updated reviewer's comments and Appendix

--- 2021-11-16 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.Rmd
@@
 
-- The term "missing data" has been removed to avoid confusion. We now refer to different number of observations as "unbalanced data" (L201-204).
+- L204-207: The term "missing data" has been removed to avoid confusion. We now refer to different number of observations as "unbalanced data".
 
-- We have made clear that LMEMs can also work with missing observations (L209)
+- L212: We have made clear that LMEMs can also work with missing observations.
 
-- Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates (L217-220).
+- L220-223: Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates.
 
@@
 
-- L181-L184 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
+- L180-L182 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
 
-- We provide additional context on the purpose of using linear models to fit quadratic data in L250-L251 which now read "We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions."
+- L246-L250: We provide additional context on the purpose of using linear models to fit quadratic data in  which now read "We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions."
 
 We believe that the aforementioned change makes a clear statement to the reader about our intentions in Fig.1.
-We want to emphasize that strictly speaking the term "default" might not be correct, but we are stretching the language to make the point that doing a simple switch from rm-ANOVA to a LMEMs (in R) without any additional changes results in the assumption of linear effects that still provide biased estimates.
+We want to emphasize that strictly speaking the term "default" (used for linear effects in a LMEM) might not be correct, but we are stretching the language to make the point that doing a simple switch from rm-ANOVA to a LMEMs (in R) without any additional changes results in the assumption of linear effects that still provide biased estimates.
 
-- L273-276 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
+- L271-274 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
 1.
 Specifically, that in reality the true function is not known and that this complicates choosing the degree of the polynomial.
-We also reference Section 4.3.1, which provides more detail on the limitation of polynomial effects.
+We also reference Section 4.3.1, which provides more detail on the limitations of polynomial effects.
 
-- In Section 6 (Discussion)  we emphasize again the rationale for not using polynomial effects on the LMEM in Fig. 1 (L509-513).
+- L505-509 In Section 6 (Discussion) we emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
 
@@
 
-- We appreciate Dr.
-Simpson's comment here and the equations provided, which we have now incorporated in the manuscript..
-We acknowledge that a brief overview of the different statistical model classes was missing in the original manuscript, and we have now extensively edited Section 4 to provide context on LMs, GLMs and GLMMs before introducing GAMs.
-The changes now appear between L282 (where Section 4 has been renamed to "Linear Models and Beyond") and L321.
-Because the different model classifications are not commonly known in the biomedical field, we contextualize the previously presented rm-ANOVA and LMEM into their respective classes to provide the reader with a clear understanding of where each model fits and also of where GAMs fit within this framework.
-We believe this properly addressed the reviewer comments.
+We appreciate Dr.
+Simpson's comment here and the equations provided, which we have now incorporated in the manuscript.
+We acknowledge that a brief overview of the different statistical model classes was missing in the original manuscript, and we have now extensively edited Section 4 to provide context on linear models (LMs), generalized linear models (GLMs) and generalized linear mixed models (GLMMs) before introducing GAMs.
 
-- We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the **linear case**.
+- L280- L318: Section 4 has been renamed to "Linear Models and Beyond" and the different Subsections cover LMs, GLMs, GLMMs and GAMs.
+Because the different model classes are not commonly known in the biomedical field, we contextualize the previously presented rm-ANOVA and LMEM into their respective classes to provide the reader with a clear understanding of where each model fits (Section 4.1).
+We also provide equations for the linear predictors in each case (Equations (6) and (8)) following the notation of Equations (1) through (4).
+We want to emphasize that because the target audience of this paper are biomedical researchers that are not familiar with Mathematical Statistics, we follow notation that is easy to read (e.g., using $time_t$ instead of $x_t$), although it might not be standard notation followed in Statistical publications.
+
+We believe that because Section 4 now provides context and follows a logical progression of LMs, GLMs, GLMMs and GAMs in a way that is easily understandable, these changes properly addressed the reviewer comments.
+
+- L323: We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the **linear case**.
 This change will avoid the propagation of confusion that Dr.
@@
 
-- We have also changed L328 to indicate that the term $\varepsilon_{ijt}$ is the residual, and not a separable error.
+- L325: We indicate that the term $\varepsilon_{ijt}$ is the residual, and not a separable error.
 Of notice, we are following the notation provided by Dr.
@@
 
-- The information on L289-294 pertaining splines has been changed.
-L301-309 now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
+- L343-357: This information that originally appeared in L289-294 pertaining splines has been changed, as we now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
+Here, we no longer refer to TPRS as an "optimized version" but we indicate that they are a truncated version of thin plate splines (TPS).
 
@@
 
-- L297-300 no longer refer to "knots" when referring to the construction of the smooth, now indicating that the number of _basis functions_ is what is specified when constructing the smoother.
-In the updated manuscript, this change appears in L310-315, and is more clear to the reader due to the changes introduced in L301-L309.
+- L358-365, which correspond to L297-300 on the original submission no longer refer to "knots" when referring to the construction of the smooth, now indicating that the number of _basis functions_ is what is specified when constructing the smoother.
+We believe that this will make sense to the reader  due to the information provided in L343-357.
 
@@
 
+- We appreciate this comment regarding wiggliness.
+For this particular topic, we have implemented changes that address the reviewer's concerns without adding additional notation (which will obscure the topic to the reader).
+We have used a minimalistic approach here in order to provide some facts about penalization without becoming too technical about it.
+
+L366-374 has been added, and is a paragraph which briefly discusses wiggliness, the smoothness parameter $\lambda$ and the penalization process using the integrated square of the seocnd derivative of the spline.
+Our goal here is for the reader to understand the effect that $\lambda$ in order to prevent overfitting.
+
+- L373: We have substituted "weak" and "strong" for "low" and "high".
+
 # Bayesian
@@
 
-- We have clarified the concept "empirical Bayesian", which appeared on L318-L319 in the original manuscript.
-In the revised manuscript, L335-338 now indicate that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
+- L392-393 now clarify this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
 
 - The sentence "Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model", which appears in L319 in the original manuscript has been removed from the text.
-Instead, the concept of REML has been moved to Section 6.2 L392-395, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in _mgcv_.
+Instead, the concept of REML has been moved to Section 5.2 L442-444, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in _mgcv_.
 
@@
 
-- L341-L355 now contain a more detailed and accurate explanation on the differences and interpretation of "poin-twise" CIs and "across-the-function" CIs.
-We removed the "95% of the time" phrase to avoid confusion, and instead we provide an explanation that focuses on random sampling: "if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function" (L352-354).
-We also reference the work of Marra and Wood if the reader desires a more in-depth exploration of across the function CIs.
+- L396-L403 now contain a more detailed and accurate explanation of"across-the-function" CIs.
+We have used the reviewer's comments to indicate that in GAM fitted using _mgcv_, the close to nominal across-the-function-coverage is true when averaged over the entire function.
 
@@
 
-In L367 it is indicated that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
+L416 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
 The computational requisite in the smooth estimation of having the maximum number of basis equal to the number of unique values in the covariate makes it impossible to fit a GAM with a k>5, as the error "A term has fewer unique covariate combinations than specified maximum degrees of freedom" is thrown by _mgcv_.
-We consider that because the k-index for the model is 1.04 the basis dimension for the smooth is adequate (Wood says that "the further below 1 this is, th emore likely it is that there is missed pattern left in the residuals").
+We consider that because the k-index for the model is 1.04 the basis dimension for the smooth is adequate (Wood says that "the further below 1 this is, the more likely it is that there is missed pattern left in the residuals").
+
+## Reviewer's Comments
+
+A.4.3 You say that gam_00 doesn’t account for the nesting of the data by Group.
+It does but only partially; you get a separate smooth per group and those smooth will be trying to also account for the different means of the response in the two groups.
+I think it is much clearer to say just that; it is better to include
+parametric terms for the group means so the smooths capture the time-course differences about these overall group means of $Y$ .
+
+## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We believe Dr.
+Simpson refers here to `gam_01` instead of `gam_00`, as the latter does not include `Group` as part of the model specification and because of this, we don't get a separate smooth per group, which can be seen in the resulting single smooth fitted by the model.
+However, we believe that the clarification he provided here is important, and we have added it to A.4.3 so it is more clear to the reader.
 
@@
 - The name of the model has been changed to `gam_02`, and it appears with this notation in Section 6.2, L383 so it matches the workflow of model selection presented in the Appendix.
-- Additionally, we have used some of the source code of the function `gam.check` in order to use the graphical output of `gratia` and the numerical diagnostic information of `gam.check` without repeating the diagnostic plots.
+- Additionally, we have used some of the source code of the function `gam.check` in order to use the graphical output of `gratia` and the numerical diagnostic information of `gam.check` without repeating the diagnostic plots, the code for this is now in the Appendix.
 
@@
 
+We provide the code for the figures and models from the manuscript in order to make the work reproducible, but we are aware that readers will not be interested specifically in such code, but rather in the code for the model selection workflow in the Appendix.
+For that part of the Appendix (Section A.4) we do follow the reviewer's comment and have updated the order of the code chunks in order to make it similar to a vignette.
+
 # Figures
@@
 It might seem too basic or repetitive, but we are trying to convince the reader that simulation can indeed produce synthetic data that can be useful.
+However, we do acknowledge that the inset was repetitive.
+
+We have updated Figure 3 so that panel A shows both the simulated and real data and therefore it is easier for the reader to see the simulation of the output.
 
@@
 
-We have changed the expression to "data that shows non-linear trends". The change has also been applied to L316 which is now L334 in the revised manuscript.
+We have changed the expression to "data that shows non-linear trends". The change has also been applied to L316 which is now L389 in the revised manuscript.
 
@@
 
-**leave this to the end as it can be fixed in the final .tex document**
+We have updated all the occurrences of citations within the narrative text by using the last name of the author followed by the citation, which fixes the issue.
 
@@
 
-We have updated this line to "the mathematical space within which the true but unknown $f(x_t\mid \beta_j)$  is thought to exist", in L290 in the revised manuscript.
+We have updated this line to "the mathematical space within which the true but unknown $f(x_t\mid \beta_j)$  is thought to exist", in L337 in the revised manuscript.
 
@@
 However, we acknowledge that we didn't clearly conveyed this point to the read.
-In Section 6 (Discussion, L525-526 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided.
+In Section 6 (Discussion, L525-527 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided.
 
@@
 ## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We use the code notation provided by the $LaTeX$ package `listings`, and it provides curly quotes for code.
+Regarding `family`, we believe that it is not necessary to include it here, but we indicate in L445-446 what the argument exists and that other distributions can be used.
 
@@
 For the length of the covariates explored in this paper TPRS are sufficient.
-However, we want to make clear to the reader that other types of splines that might be appropriate in each case are available.
-To acknowledge this, L390 in the updated manuscript now reads "(a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console)".
+However, we want to make clear to the reader that other types of splines that might be appropriate in certain situations are available.
+To acknowledge this, L440 in the updated manuscript now reads "(a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console)".
 

### 2021-12-01 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.tex
Updated reviewer's comments and Appendix

@@
 \item
-  The term ``missing data'' has been removed to avoid confusion. We now refer to different number of observations as ``unbalanced data'' (L201-204).
+  L204-207: The term ``missing data'' has been removed to avoid confusion. We now refer to different number of observations as ``unbalanced data''.
 \item
-  We have made clear that LMEMs can also work with missing observations (L209)
+  L212: We have made clear that LMEMs can also work with missing observations.
 \item
-  Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates (L217-220).
+  L220-223: Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates.
 \end{itemize}
@@
 \item
-  L181-L184 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
+  L180-L182 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
 \item
-  We provide additional context on the purpose of using linear models to fit quadratic data in L250-L251 which now read ``We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with''default'' (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.''
+  L246-L250: We provide additional context on the purpose of using linear models to fit quadratic data in which now read ``We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with''default'' (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.''
 \end{itemize}
@@
 We believe that the aforementioned change makes a clear statement to the reader about our intentions in Fig.1.
-We want to emphasize that strictly speaking the term ``default'' might not be correct, but we are stretching the language to make the point that doing a simple switch from rm-ANOVA to a LMEMs (in R) without any additional changes results in the assumption of linear effects that still provide biased estimates.
+We want to emphasize that strictly speaking the term ``default'' (used for linear effects in a LMEM) might not be correct, but we are stretching the language to make the point that doing a simple switch from rm-ANOVA to a LMEMs (in R) without any additional changes results in the assumption of linear effects that still provide biased estimates.
 
@@
 \item
-  L273-276 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
+  L271-274 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
 1.
 Specifically, that in reality the true function is not known and that this complicates choosing the degree of the polynomial.
-We also reference Section 4.3.1, which provides more detail on the limitation of polynomial effects.
+We also reference Section 4.3.1, which provides more detail on the limitations of polynomial effects.
 \item
-  In Section 6 (Discussion) we emphasize again the rationale for not using polynomial effects on the LMEM in Fig. 1 (L509-513).
+  L505-509 In Section 6 (Discussion) we emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
 \end{itemize}
@@
 
+We appreciate Dr.~Simpson's comment here and the equations provided, which we have now incorporated in the manuscript.
+We acknowledge that a brief overview of the different statistical model classes was missing in the original manuscript, and we have now extensively edited Section 4 to provide context on linear models (LMs), generalized linear models (GLMs) and generalized linear mixed models (GLMMs) before introducing GAMs.
+
+\begin{itemize}
+\tightlist
+\item
+  L280- L318: Section 4 has been renamed to ``Linear Models and Beyond'' and the different Subsections cover LMs, GLMs, GLMMs and GAMs.
+Because the different model classes are not commonly known in the biomedical field, we contextualize the previously presented rm-ANOVA and LMEM into their respective classes to provide the reader with a clear understanding of where each model fits (Section 4.1).
+We also provide equations for the linear predictors in each case (Equations (6) and (8)) following the notation of Equations (1) through (4).
+We want to emphasize that because the target audience of this paper are biomedical researchers that are not familiar with Mathematical Statistics, we follow notation that is easy to read (e.g., using \(time_t\) instead of \(x_t\)), although it might not be standard notation followed in Statistical publications.
+\end{itemize}
+
+We believe that because Section 4 now provides context and follows a logical progression of LMs, GLMs, GLMMs and GAMs in a way that is easily understandable, these changes properly addressed the reviewer comments.
+
 \begin{itemize}
 \item
-  We appreciate Dr.~Simpson's comment here and the equations provided, which we have now incorporated in the manuscript..
-We acknowledge that a brief overview of the different statistical model classes was missing in the original manuscript, and we have now extensively edited Section 4 to provide context on LMs, GLMs and GLMMs before introducing GAMs.
-The changes now appear between L282 (where Section 4 has been renamed to ``Linear Models and Beyond'') and L321.
-Because the different model classifications are not commonly known in the biomedical field, we contextualize the previously presented rm-ANOVA and LMEM into their respective classes to provide the reader with a clear understanding of where each model fits and also of where GAMs fit within this framework.
-We believe this properly addressed the reviewer comments.
-\item
-  We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the \textbf{linear case}.
+  L323: We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the \textbf{linear case}.
 This change will avoid the propagation of confusion that Dr.~Simpson correctly indicated in his comment.
 \item
-  We have also changed L328 to indicate that the term \(\varepsilon_{ijt}\) is the residual, and not a separable error.
+  L325: We indicate that the term \(\varepsilon_{ijt}\) is the residual, and not a separable error.
 Of notice, we are following the notation provided by Dr.~Simpson in Eq.
@@
 \item
-  The information on L289-294 pertaining splines has been changed.
-L301-309 now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
+  L343-357: This information that originally appeared in L289-294 pertaining splines has been changed, as we now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
+Here, we no longer refer to TPRS as an ``optimized version'' but we indicate that they are a truncated version of thin plate splines (TPS).
 \end{itemize}
@@
 \item
-  L297-300 no longer refer to ``knots'' when referring to the construction of the smooth, now indicating that the number of \emph{basis functions} is what is specified when constructing the smoother.
-In the updated manuscript, this change appears in L310-315, and is more clear to the reader due to the changes introduced in L301-L309.
+  L358-365, which correspond to L297-300 on the original submission no longer refer to ``knots'' when referring to the construction of the smooth, now indicating that the number of \emph{basis functions} is what is specified when constructing the smoother.
+We believe that this will make sense to the reader due to the information provided in L343-357.
 \end{itemize}
@@
 \begin{itemize}
+\tightlist
 \item
   We appreciate this comment regarding wiggliness.
-We have now added a new paragraph in Section 4.3.1 (L369-375) which briefly discuss wiggliness, and uses a physics analogy to convey the overall concept of penalization.
-Because the intention of our work is to present the practical advantages of GAMs we want to avoid including mathematical notation about penalization in the manuscript, which would be confusing and distracting to the readers from the main purpose of the paper.
+For this particular topic, we have implemented changes that address the reviewer's concerns without adding additional notation (which will obscure the topic to the reader).
+We have used a minimalistic approach here in order to provide some facts about penalization without becoming too technical about it.
+\end{itemize}
+
+L366-374 has been added, and is a paragraph which briefly discusses wiggliness, the smoothness parameter \(\lambda\) and the penalization process using the integrated square of the seocnd derivative of the spline.
+Our goal here is for the reader to understand the effect that \(\lambda\) in order to prevent overfitting.
+
+\begin{itemize}
+\tightlist
 \item
-  We have substituted ``weak'' and ``strong'' for ``low'' and ``high'' (L373).
+  L373: We have substituted ``weak'' and ``strong'' for ``low'' and ``high''.
 \end{itemize}
@@
 \item
-  We have clarified the concept ``empirical Bayesian'', which appeared on L318-L319 in the original manuscript.
-In the revised manuscript, L335-338 now indicate that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
+  L392-393 now clarify this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
 \item
   The sentence ``Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model'', which appears in L319 in the original manuscript has been removed from the text.
-Instead, the concept of REML has been moved to Section 6.2 L392-395, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in \emph{mgcv}.
+Instead, the concept of REML has been moved to Section 5.2 L442-444, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in \emph{mgcv}.
 \end{itemize}
@@
 \item
-  L341-L355 now contain a more detailed and accurate explanation on the differences and interpretation of ``poin-twise'' CIs and ``across-the-function'' CIs.
-We removed the ``95\% of the time'' phrase to avoid confusion, and instead we provide an explanation that focuses on random sampling: ``if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function'' (L352-354).
-We also reference the work of Marra and Wood if the reader desires a more in-depth exploration of across the function CIs.
+  L396-L403 now contain a more detailed and accurate explanation of''across-the-function'' CIs.
+We have used the reviewer's comments to indicate that in GAM fitted using \emph{mgcv}, the close to nominal across-the-function-coverage is true when averaged over the entire function.
 \end{itemize}
@@
 
-In L367 it is indicated that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
+L416 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
 The computational requisite in the smooth estimation of having the maximum number of basis equal to the number of unique values in the covariate makes it impossible to fit a GAM with a k\textgreater5, as the error ``A term has fewer unique covariate combinations than specified maximum degrees of freedom'' is thrown by \emph{mgcv}.
-We consider that because the k-index for the model is 1.04 the basis dimension for the smooth is adequate (Wood says that ``the further below 1 this is, th emore likely it is that there is missed pattern left in the residuals'').
+We consider that because the k-index for the model is 1.04 the basis dimension for the smooth is adequate (Wood says that ``the further below 1 this is, the more likely it is that there is missed pattern left in the residuals'').
 
@@
 \subsection{Reviewer's Comments}\label{reviewers-comments-10}}
+
+A.4.3 You say that gam\_00 doesn't account for the nesting of the data by Group.
+It does but only partially; you get a separate smooth per group and those smooth will be trying to also account for the different means of the response in the two groups.
+I think it is much clearer to say just that; it is better to include
+parametric terms for the group means so the smooths capture the time-course differences about these overall group means of \(Y\) .
+
+\hypertarget{section-11}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-11}}
+
+We believe Dr.~Simpson refers here to \passthrough{\lstinline!gam\_01!} instead of \passthrough{\lstinline!gam\_00!}, as the latter does not include \passthrough{\lstinline!Group!} as part of the model specification and because of this, we don't get a separate smooth per group, which can be seen in the resulting single smooth fitted by the model.
+However, we believe that the clarification he provided here is important, and we have added it to A.4.3 so it is more clear to the reader.
+
+\hypertarget{reviewers-comments-11}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-11}}
 
@@
 
-\hypertarget{section-11}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-11}}
+\hypertarget{section-12}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-12}}
 
@@
 \item
-  Additionally, we have used some of the source code of the function \passthrough{\lstinline!gam.check!} in order to use the graphical output of \passthrough{\lstinline!gratia!} and the numerical diagnostic information of \passthrough{\lstinline!gam.check!} without repeating the diagnostic plots.
+  Additionally, we have used some of the source code of the function \passthrough{\lstinline!gam.check!} in order to use the graphical output of \passthrough{\lstinline!gratia!} and the numerical diagnostic information of \passthrough{\lstinline!gam.check!} without repeating the diagnostic plots, the code for this is now in the Appendix.
 \end{itemize}
@@
 
-\hypertarget{reviewers-comments-11}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-11}}
+\hypertarget{reviewers-comments-12}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-12}}
 
@@
 
-\hypertarget{section-12}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-12}}
+\hypertarget{section-13}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-13}}
 
-\hypertarget{reviewers-comments-12}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-12}}
+\hypertarget{reviewers-comments-13}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-13}}
 
@@
 
-\hypertarget{section-13}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-13}}
+\hypertarget{section-14}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-14}}
+
+We provide the code for the figures and models from the manuscript in order to make the work reproducible, but we are aware that readers will not be interested specifically in such code, but rather in the code for the model selection workflow in the Appendix.
+For that part of the Appendix (Section A.4) we do follow the reviewer's comment and have updated the order of the code chunks in order to make it similar to a vignette.
 
@@
 
-\hypertarget{reviewers-comments-13}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-13}}
+\hypertarget{reviewers-comments-14}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-14}}
 
@@
 
-\hypertarget{section-14}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-14}}
+\hypertarget{section-15}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-15}}
 
@@
 
-\hypertarget{section-15}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-15}}
+\hypertarget{section-16}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-16}}
 
@@
 
-\hypertarget{section-18}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-18}}
+\hypertarget{section-19}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-19}}
 
-We have changed the expression to ``data that shows non-linear trends''. The change has also been applied to L316 which is now L334 in the revised manuscript.
+We have changed the expression to ``data that shows non-linear trends''. The change has also been applied to L316 which is now L389 in the revised manuscript.
 
-\hypertarget{reviewers-comments-18}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-18}}
+\hypertarget{reviewers-comments-19}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-19}}
 
@@
 
-\hypertarget{section-19}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-19}}
+\hypertarget{section-20}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-20}}
 
@@
 
-\hypertarget{reviewers-comments-19}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-19}}
-
-L185 here and throughout, check for situations where you have combined numerical citations to literature within narrative text; Here it should be ``\ldots see refs 40 \& 42.)'' for example, not superscript numerals.
-
-\hypertarget{section-20}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-20}}
-
-\textbf{leave this to the end as it can be fixed in the final .tex document}
-
 \hypertarget{reviewers-comments-20}{%
@@
 
-We have changed the word from ``global'' to ``population''.
+We have updated all the occurrences of citations within the narrative text by using the last name of the author followed by the citation, which fixes the issue.
 
@@
 
-This line has been removed from the revised manuscript.
-However, we acknowledge that we didn't clearly conveyed this point to the read.
-In Section 6 (Discussion, L525-526 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided.
+We have updated this line to ``the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist'', in L337 in the revised manuscript.
 
@@
 
-L351 Better to say ``separate smooths'' not ``independent''.
-These factor-by smooths share a common basis, they just get an entirely separate set of model coeficients and smoothness parameters for each level of the factor.
+L293 You never really describe what ``data-drive flexibility'' is. Why are GAMs data driven and LMMs and GLMs not? This is a critical concept to convey I feel.
 
@@
 
-The sentence now reads ``separate smooths for\ldots{}''
+This line has been removed from the revised manuscript.
+However, we acknowledge that we didn't clearly conveyed this point to the read.
+In Section 6 (Discussion, L525-527 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided.
 
@@
 
-L354 ``difference between group means'' because the smooth encodes differences between the groups too, just not difference in the means. Be a little more specific here.
+L351 Better to say ``separate smooths'' not ``independent''.
+These factor-by smooths share a common basis, they just get an entirely separate set of model coeficients and smoothness parameters for each level of the factor.
 
@@
 
+The sentence now reads ``any differences between the group means''.
+
 \hypertarget{reviewers-comments-26}{%
@@
 
-The sentence now reads ``contains the fitted model''.
+We use the code notation provided by the \(LaTeX\) package \passthrough{\lstinline!listings!}, and it provides curly quotes for code.
+Regarding \passthrough{\lstinline!family!}, we believe that it is not necessary to include it here, but we indicate in L445-446 what the argument exists and that other distributions can be used.
 
@@
 
-L362 why highlight Gaussian process smooths?
-They're not as useful as they seem as you need to optimise the length scale parameter yourself as otherwise, especially for longitudinal data they aren't great - the length scale is the largest separate in time of any pair of observations.
+L360 Not ``knots'' - there are k-1 basis functions, but as described above, because this is a TPRS there are knots at the unique data values and then we decompose the full basis and take the required eigenvectors as the basis to use.
 
@@
 
-For the length of the covariates explored in this paper TPRS are sufficient.
-However, we want to make clear to the reader that other types of splines that might be appropriate in each case are available.
-To acknowledge this, L390 in the updated manuscript now reads ``(a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console)''.
+The updated sentence now reads ``using four basis functions (plus the intercept)''.
 
@@
 
-L366 See also Matteo Fasiolo's \{mgcViz\} package
+L362 why highlight Gaussian process smooths?
+They're not as useful as they seem as you need to optimise the length scale parameter yourself as otherwise, especially for longitudinal data they aren't great - the length scale is the largest separate in time of any pair of observations.
 
@@
 
-We acknowledge the existence of Dr.~Fasiolo's \emph{mgcvViz} package, but due to the introductory nature of this paper we prefer to focus only on \emph{mgcv} and \emph{gratia}.
+For the length of the covariates explored in this paper TPRS are sufficient.
+However, we want to make clear to the reader that other types of splines that might be appropriate in certain situations are available.
+To acknowledge this, L440 in the updated manuscript now reads ``(a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console)''.
 
@@
 
+The sentence has been rewritten. It now reads ``\ldots the observed difference is significant\ldots{}''
+
 \hypertarget{reviewers-comments-35}{%
@@
 
+\hypertarget{reviewers-comments-36}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-36}}
+
+L410 This needs to be reworded as what is written can be read as any difference between the control and treatment (and there is a difference as the controls are higher than the treated group before day 3) where you mean that the treatment group is only higher than the control group post day 3.
+
+\hypertarget{section-37}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-37}}
+
 This was an unintentional mistake based on some previous simulation results.
@@
 
-\hypertarget{reviewers-comments-36}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-36}}
+\hypertarget{reviewers-comments-37}{%
+\subsection{Reviewer's Comments}\label{reviewers-comments-37}}
 
@@
 
-\hypertarget{section-37}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-37}}
+\hypertarget{section-38}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-38}}
 

### 2021-12-01 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Updated reviewer's comments and Appendix

@@
 
-In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
-Independent errors \(\varepsilon_{ijt}\) represent \textbf{random variation from the sampling process } assumed to be \(\stackrel{i.i.d.}\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
+In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of treatment (\(treatment_j\)), time (\(time_t\)), and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
+Independent errors \(\varepsilon_{ijt}\) represent random variation from the sampling process assumed to be \(\stackrel{i.i.d.}\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
 In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug,'' or ``saline'' vs.~``chemotherapy'').
@@
 
-To further simplify the expression, substitute \(\widetilde{\beta_{0}}=\beta_0+\beta_{2}\) and \(\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}\) in the equation for Group B.
+To further simplify the expression, substitute \(\widetilde{\beta_{0}}=\beta_0+\beta_{1}\) and \(\widetilde{\beta_{1}}=\beta_{2}+\beta_{3}\) in the equation for Group B.
 This substitution allows for a different intercept and slope for Groups A and B.
@@
 A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to account for\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}}.
-A LMEM with interaction between time and treatment for a longitudinal study can be written as:
+A LMEM with interaction between time and treatment for a longitudinal study can be written as
 
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij} +\varepsilon_{ijt}\\.
+y_{ijt} = \beta_0+ \beta_1 \times treatment_{j}+ \beta_2 \times time_{t} + \beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij} +\varepsilon_{ijt}\\.
 \label{eq:LMEM}
@@
 We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with ``default'' (linear) effects and, 2) demonstrate that a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
-Of notice, in Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
+In Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
 
@@
 Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
-It could be argued that a LMEM with quadratic effects should have been used to fit the data in Figure\ref{fig:l-q-response}, F.
+It could be argued that a LMEM with quadratic effects should have been used to fit the data in Figure\ref{fig:l-q-response}F.
 However, because in reality the true function is not known, choosing a polynomial degree causes more questions (e.g., is it quadratic?, cubic?, or a higher degree?).
@@
 
-Linear models (LMs) are those that assume a normal (Gaussian) distribution of the response, and only incorporate \emph{fixed effects} (such as rm-ANOVA).
+Linear models (LMs) are those that assume a normal (Gaussian) distribution of the errors, and only incorporate \emph{fixed effects} (such as rm-ANOVA).
 These are by far the models most commonly used to analyze data within the biomedical research community.
-On the other hand, Linear Mixed Effect Models (LMEMs) are those that assume a normally-distributed response but also incorporate \emph{random effects}, as it has been described in Section \ref{LMEM-case}.
+On the other hand, Linear Mixed Effect Models (LMEMs) also incorporate \emph{random effects}, as it has been described in Section \ref{LMEM-case}.
 
@@
 
-A major limitation of LMs is their assumption of normality.
-If the data is non-normal, a transformation is necessary in order to properly fit the model.
+A major limitation of LMs is their assumption of normality in the errors.
+If the residuals are non-normal, a transformation is necessary in order to properly fit the model.
 However, transformation can lead to poor model performance,\textsuperscript{\protect\hyperlink{ref-ohara2010}{51}} and can cause problems with the biological interpretation of the model estimates.
-McCullagh and Nelder\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} introduced General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
+McCullagh and Nelder\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} introduced General Linear Models (GLMs) as an extension of LMs, where the errors do not need to be normally distributed.
 To achieve this, consider the following model
@@
 The mean (\(\mu_{ijt}\)) is also known as the \emph{expected value} (or \emph{expectation}), \(E(y_{ijt})\) of the observed response \(y_{ijt}\).
+
 Then, the \emph{linear predictor} \(\eta\), which defines the relationship between the mean and the covariates can be defined as
@@
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j},
+\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} +\beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j},
 \label{eq:GLM-eta}
@@
 
-where \(\eta_{ijt}\) is the linear predictor for each observation, \(\beta_0\) is the intercept, and \(\beta_{j}\) is the model parameter for each group; and \(x_{ijt}\) represents the covariates from each subject in each group at each time point.
+where \(\eta_{ijt}\) is the linear predictor for each observation in each group at each timepoint, and \(\beta_0\) (the intercept), \(\beta_{1}, \beta_2,\) and \(\beta_3\) are the model parameters for each group, which can be referred globally to as \(\beta_j\).
+Finally, \(time_{ijt}\) represents the covariates from each subject in each group at each time point.
 
@@
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij},
+\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} + \beta_2 \times time_{t} + +\beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij},
 \label{eq:GLMM-eta}
@@
 
-Notice that in the previous sections, the difference between GLMs and GLMMs resides on their linear predictors (Equations \eqref{eq:GLM-eta} and \eqref{eq:GLMM-eta}).
-Generalized additive models (GAMs) are an extension of the GLM family that allow the estimation of smoothly varying trends where the relationship between the covariates and the response is modeled using \emph{smooth functions}\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}}.
-In a GAM, the linear predictor has the form
+Notice that in the previous sections, the difference between GLMs and GLMMs resides on their linear predictors (Equations \eqref{eq:GLM-eta}, \eqref{eq:GLMM-eta}).
+Generalized additive models (GAMs) are an extension of the GLM family that allow the estimation of smoothly varying trends where the relationship between the covariates and the response is modeled using \emph{smooth functions}.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}} In a GAM, the linear predictor has the form
 
@@
 
-Where \(\beta_{0}\) is the intercept, \(\beta_{1}\) is the coefficient for each treatment group.
-Notice that we and \(f(time_{t}|\beta_{j})\) represents the smooth function of time, with coefficient \(\beta_{j}\).
+Where \(\beta_{0}\) is the intercept, and \(\beta_{1}\) is the coefficient for each treatment group.
+Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time have been replaced by \(f(time_{t}|\beta_{j})\), which represents the smooth function of time with with inputs as the covariates \(time_t\) and parameters \(\beta_{j}\).
 A GAM version of a linear model can be written as:
@@
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and the change of \(y_{ijt}\) over time (instead of being defined by \(\beta_{j}time_{ijt}\) as in Equation \eqref{eq:GLM-eta}) is represented by the smooth function \(f(time_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
+Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the smooth function \(f(time_t\mid \beta_j)\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
 
@@
 The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using a \emph{basis function} expansion of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
+A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(treatment_j\), \(time_t\), and \(time_t \times treatment_j\).
 The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
@@
 Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present.
-A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood with a multiple \(\lambda\) times the integrated square of the second derivative of the spline.
-The second derivative of the spline is a measure of curvature, or the rate of change of the slope.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} Increasing the penalty by increasing \(\lambda\) results in models with less curvature.
+A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline.
+The second derivative of the spline is a measure of curvature, or the rate of change of the slope,\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} and increasing the penalty by increasing \(\lambda\) results in models with less curvature.
 As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
-A low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
+In other words,a low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
 
@@
 Each of the basis functions is evaluated at six different points (because there are six points on the timeline).
-The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\) estimated when fitting the model.
+The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), that is estimated when fitting the model.
 The penalized estimates fitted for our example are shown in Figure \ref{fig:basis-plot}B.
@@
 In an \emph{across the function} CI (like those estimated for GAMs using \emph{mgcv}) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%).
-In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI.
-For this to occur then, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
+In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI and for this to occur, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
 In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in Miller,\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} and Marra.\textsuperscript{\protect\hyperlink{ref-marra2012}{62}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
@@
 The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} (a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console).
+
 The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups.
 Although the default \passthrough{\lstinline!method!} used to estimate the smoothing parameters in \emph{mgcv} is generalized cross validation (GCV), Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} showed the restricted maximum likelihood (REML) to be more resistant to overfitting while also easing the quantification of uncertainty in the smooth parameters; therefore in this manuscript REML is always used for smooth parameter estimation.
+An additional argument (\passthrough{\lstinline!family!}) allows to specify the expected distribution of the response, but it is not used in this model because we expect a normally-distributed response (which is the default \passthrough{\lstinline!family!} in \emph{mgcv}).
 
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{63}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{64}}
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{63}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{64}}
 
@@
 Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that if the \emph{difference} between the empirical Bayesian confidence intervals of the fitted smooths for such groups is non-zero, then a significant difference exists at that time point(s).
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
+\textbf{add sentence about the issue of the Nychka across the function intervals }
+
+around that \emph{difference} between the empirical Bayesian confidence intervals of the fitted smooths for such groups is non-zero, then a significant difference exists at that time point(s).
 The absence of a \emph{p-value} in this case might seem odd, but the empirical Bayesian confidence interval interpretation can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.

### 2021-12-01 Manuscripts/Manuscript_by_chapters-SIM_Revisions/SIM_Appendix.tex
Updated reviewer's comments and Appendix

@@
 \begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
+y_{ijt} = \beta_0+\beta_1 \times treatment_{j} +\beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
 \label{eq:linear-model}
@@
 y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\beta_0 + \beta_2\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
+\beta_0 + \beta_1+\beta_2 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
 \end{cases}
@@
 y_{ijt} = \begin{cases}
-\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
-\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
+\beta_0 + \beta_2\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
+\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
 \end{cases}
@@
 
-A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to distinguish {[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}.
-A LMEM with interaction between time and treatment for a longitudinal study can be written as:
+A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to account for{[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}.
+A LMEM with interaction between time and treatment for a longitudinal study can be written as
 
 \begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
+y_{ijt} = \beta_0+ \beta_1 \times treatment_{j}+ \beta_2 \times time_{t} + \beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij} +\varepsilon_{ijt}\\.
 \label{eq:LMEM}
@@
 
-When Equation \eqref{eq:linear-model} and Equation \eqref{eq:LMEM} are compared, it is easily noticeable that LMEMs and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\mu_{ij}\)).
-This term \(\mu_{ij}\) is the one that corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)).
-The \emph{random} component can also be understood as used to model some ``noise'' in the response, but that is intended to be analyzed and disentangled from the **``sampling** noise'' term \(\varepsilon_{ijt}\) from Equation \eqref{eq:linear-model}.
-
-For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data \textbf{that needs to be accounted for}.
+When Equations \eqref{eq:linear-model} and \eqref{eq:LMEM} are compared, it is noticeable that LMEMs and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\alpha_{ij}\)).
+This term \(\alpha_{ij}\) corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)).
+The \emph{random} component can also be understood as modeling some ``noise'' in the response, but that does not arise from the sampling error term \(\varepsilon_{ijt}\) from Equations \eqref{eq:linear-model} through \eqref{eq:ANOVA-lines}.
+
+For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data that needs to be accounted for.
 As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
-The \emph{random effect} \(\mu_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence, the \emph{random effect} in a LMEM enables to fit models with different slopes at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
+The random effect \(\alpha_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
+In essence, the \emph{random effect} in a LMEM enables fitting models with different intercepts at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
 However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
 
-Of notice, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach, which is described in detail by Pinheiro and Bates {[}\protect\hyperlink{ref-pinheiro2006}{15}{]} using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}.
+Of note, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach (using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}), which is described in detail by Pinheiro and Bates {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
 However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates {[}\protect\hyperlink{ref-beck1998}{36}{]}; more importantly, polynomial effects lack biological or mechanistic interpretation {[}\protect\hyperlink{ref-pinheiro2006}{15}{]} which limits their use in biomedical studies.
@@
 In a longitudinal study, it is frequently the case that the number of observations is different across subjects.
-In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing,'' ``incomplete'' or (more generally speaking) unbalanced data {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
+In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing,'' ``incomplete,'' or (more generally speaking) unbalanced data {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
 The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{46}{]}.
+This is inconvenient because the remaining subjects might not accurately represent the population and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{46}{]}.
 
@@
 
-To visually demonstrate the limitations of rm-ANOVA an LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
+To visually demonstrate the limitations of rm-ANOVA and LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
 An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using R {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{50}{]}.
 
-Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}A).
+Briefly, two cases for the mean response for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}A).
 In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}C).
@@
 
-A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.
+A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where a simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) is presented.
 We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with ``default'' (linear) effects and, 2) demonstrate that a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
-Of notice, in Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
+In Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
 
@@
 
-In reality, rm-ANOVA and LMEMs are just \emph{special cases} of broader model classes (General Linear Models and Generalized Linear Mixed Models, respectively).
-In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different Statistical model classes, indicate how rm-ANOVA and LMEMs fit within that framework and introduce the theory of GAMs.
-
-\hypertarget{linear-models-lms-and-linear-mixed-models-lmems}{%
-\subsection{Linear Models (LMs) and Linear Mixed Models (LMEMs)}\label{linear-models-lms-and-linear-mixed-models-lmems}}
-
-Linear models (LMs) are those that assume a normal (Gaussian) distribution of the response, and only incorporate \emph{fixed effects} (such as rm-ANOVA).
+Linear models (LMs) are those that assume a normal (Gaussian) distribution of the errors, and only incorporate \emph{fixed effects} (such as rm-ANOVA).
 These are by far the models most commonly used to analyze data within the biomedical research community.
-On the other hand, Linear Mixed Effect Models (LMEMs) are those that assume a normally-distributed response but also incorporate \emph{random effects}, as it has been described in Section \ref{LMEM-case}.
-
-In matrix notation, a LM has the general form
-
-\begin{equation}
-\textbf{y} = \textbf{X} \boldsymbol{\beta}
-\label{eq:LM-matrix}
-\end{equation}
-
-Where \(\textbf{y}\) a vector of observations, \(\textbf{X}\) is a matrix that contains the measured values of the covariates, and \(\boldsymbol{\beta}\) is a vector of the parameters to be estimated.
-The ``model equation'' form of Equation \eqref{eq:LM-matrix} for a rm-ANOVA model of two treatment groups corresponds to Equation \eqref{eq:linear-model}.
-It is important to remember that the error term \(\varepsilon_{ijt}\) in Equation \eqref{eq:linear-model} represents the variability introduced by random sampling from the normal distribution.
+On the other hand, Linear Mixed Effect Models (LMEMs) also incorporate \emph{random effects}, as it has been described in Section \ref{LMEM-case}.
+
+In reality, rm-ANOVA and LMEMs are just \emph{special cases} of a broader class of models (General Linear Models and Generalized Linear Mixed Models, respectively).
+In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different classes of models, indicate how rm-ANOVA and LMEMs fit within this framework and introduce the theory of GAMs.
 
@@
 
-A major limitation of LMs is their assumption of normality.
-If the data is non-normal, a transformation is necessary in order to properly fit the model.
-However, transformation lead to poor model performance {[}\protect\hyperlink{ref-ohara2010}{51}{]}, and can cause problems with the biological interpretation of the model estimates.
-McCullagh and Nelder {[}\protect\hyperlink{ref-nelder1972}{52}{]} defined General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed.
-To achieve this, consider the following
+A major limitation of LMs is their assumption of normality in the errors.
+If the residuals are non-normal, a transformation is necessary in order to properly fit the model.
+However, transformation can lead to poor model performance {[}\protect\hyperlink{ref-ohara2010}{51}{]}, and can cause problems with the biological interpretation of the model estimates.
+McCullagh and Nelder {[}\protect\hyperlink{ref-nelder1972}{52}{]} introduced General Linear Models (GLMs) as an extension of LMs, where the errors do not need to be normally distributed.
+To achieve this, consider the following model
 
 \begin{equation}
-y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi)
+y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi),
 \label{eq:GLM-y}
@@
 
-Where \(y_{ijt}\) is the observation \(i\) in group \(j\) at time \(t\), which is assumed to come randomly from some distribution of the exponential family \(\mathcal{D}\), with some mean \(\mu_{ijt}\) and a dispersion parameter \(\phi\) (which in the Gaussian case is the variance \(\sigma^{2}\)).
-The mean \(\mu_{ijt}\) is also known as the \emph{expected value} (also called \emph{expectation}) (\(E\)) of the observed response.
+where \(y_{ijt}\) is the observation \(i\) in group \(j\) at time \(t\), that is assumed to come from some distribution of the exponential family \(\mathcal{D}\), with some mean \(\mu_{ijt}\) and potentially, a dispersion parameter \(\phi\) (which in the Gaussian case is the variance \(\sigma^{2}\)).
+The mean (\(\mu_{ijt}\)) is also known as the \emph{expected value} (or \emph{expectation}), \(E(y_{ijt})\) of the observed response \(y_{ijt}\).
+
 Then, the \emph{linear predictor} \(\eta\), which defines the relationship between the mean and the covariates can be defined as
@@
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}
+\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} +\beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j},
 \label{eq:GLM-eta}
@@
 
-Where \(\eta_{ijt}\) is the linear predictor for each observation, \(\beta_0\) is the intercept, and \(\beta_{j}\) is the model parameter for each group; and \(x_{ijt}\) represents the covariates from each subject in each group at each time point.
+where \(\eta_{ijt}\) is the linear predictor for each observation in each group at each timepoint, and \(\beta_0\) (the intercept), \(\beta_{1}, \beta_2,\) and \(\beta_3\) are the model parameters for each group, which can be referred globally to as \(\beta_j\).
+Finally, \(time_{ijt}\) represents the covariates from each subject in each group at each time point.
 
@@
 \begin{equation}
-E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt})
+E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt}),
 \label{eq:GLM-Expectation}
 \end{equation}
-Where \(E(y_{ijt})\) is the expectation, \(g^{-1}\) is the inverse of the \emph{link function} (\(g\)).
-The link function transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}).
-Therefore, LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
+
+where \(E(y_{ijt})\) is the expectation, and \(g^{-1}\) is the inverse of a \emph{link function} (\(g\)).
+The link function transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}.
+Therefore, it can be seen that LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
 
@@
 
-Although GLMs relax the normality assumption, they can only accommodate fixed effects.
+Although GLMs relax the normality assumption, they only accommodate fixed effects.
 Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate \emph{random effects}, which have an associated probability distribution {[}\protect\hyperlink{ref-mcculloch2001}{53}{]}.
-Therefore, in GLMMs the \emph{linear predictor} takes the form
+Therefore, in GLMMs the linear predictor takes the form
 
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\mu_{ij}
+\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} + \beta_2 \times time_{t} + +\beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij},
 \label{eq:GLMM-eta}
@@
 
-Where \(\mu_{ij}\) corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of Equation \eqref{eq:GLM-eta}.
-Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed {[}\protect\hyperlink{ref-nelder1972}{52}{]}.
+where \(\alpha_{ij}\) corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of Equation \eqref{eq:GLM-eta}.
+Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed {[}\protect\hyperlink{ref-nelder1972}{52}{]}, and GLMs are a special case of GLMMs where there are no random effects.
 In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in Dobson {[}\protect\hyperlink{ref-dobson2008}{54}{]} and Stroup {[}\protect\hyperlink{ref-stroup2013}{55}{]}.
@@
 
-Generalized additive models (GAMs) are a family of regression-based methods for estimating smoothly varying trends and are an extension of the GLM family {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}{]}.
+Notice that in the previous sections, the difference between GLMs and GLMMs resides on their linear predictors (Equations \eqref{eq:GLM-eta}, \eqref{eq:GLMM-eta}).
+Generalized additive models (GAMs) are an extension of the GLM family that allow the estimation of smoothly varying trends where the relationship between the covariates and the response is modeled using \emph{smooth functions} {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}{]}.
+In a GAM, the linear predictor has the form
+
+\begin{equation}
+\eta_{ijt}=\beta_{0}+ \beta_{1} \times treatment_{j} +f(time_{t}|\beta_{j}).
+\label{eq:GAM-eta}
+\end{equation}
+
+Where \(\beta_{0}\) is the intercept, and \(\beta_{1}\) is the coefficient for each treatment group.
+Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time have been replaced by \(f(time_{t}|\beta_{j})\), which represents the smooth function of time with with inputs as the covariates \(time_t\) and parameters \(\beta_{j}\).
 A GAM version of a linear model can be written as:
@@
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and the change of \(y_{ijt}\) over time (instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta}) is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual due to \textbf{random sampling}.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible \emph{smooth functions}.
-This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
-In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
-
-The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\).
-The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{57}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
-In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
-
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{58}{]}.
+Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the smooth function \(f(time_t\mid \beta_j)\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
+
+In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is more advantageous as it allows more flexibility because it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
+One possible set of functions for \(f(time_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
+In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(time_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
+
+The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using a \emph{basis function} expansion of the covariates and by estimating random coefficients associated with these basis functions.
+A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
+For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(treatment_j\), \(time_t\), and \(time_t \times treatment_j\).
+The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{57}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(time_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
+
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{58}{]}.
 Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
 
-Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework {[}\protect\hyperlink{ref-hastie1987}{56}{]}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit {[}\protect\hyperlink{ref-wood2003}{59}{]}.
+Cubic splines (CS) are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
+The use of CS as smoothers in GAMs was discussed within the original GAM framework {[}\protect\hyperlink{ref-hastie1987}{56}{]}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the basis functions meet) to obtain the finite basis, which could affect the model fit {[}\protect\hyperlink{ref-wood2003}{59}{]}.
 A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}{]}.
@@
 In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner {[}\protect\hyperlink{ref-wood2003}{59}{]}.
-Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood{[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}{]}).
+Like TPS, TPRS only require the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood{[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}{]}).
 
@@
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
+A computational requisite in \emph{mgcv} is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
 Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth.
-It is important to note that is not necessary to specify a number of basis equal to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data.
+It is important to note that is not necessary to specify a number of basis equal to the number of unique values in the independent variable; fewer basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data.
 
 Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present.
-A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but it suffices to say that the wiggliness (or ``roughness'') of the fit is achieved by using a smoothness parameter \(\lambda\), which is applied to the second derivative of the spline.
-The second derivative of the spline is a measure of curvature, or the rate of change of the slope {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}.
-A physics analogy to the penalization process would be that if we are driving a car on a hilly and curvy road, we want to minimize abrupt changes in speed in order to have a ``smooth'' ride, and we do this by penalizing the acceleration.
-
-If five basis functions are used to fit a GAM for the data that appears in Figure \ref{fig:l-q-response}C, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A.
+A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline.
+The second derivative of the spline is a measure of curvature, or the rate of change of the slope {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}, and increasing the penalty by increasing \(\lambda\) results in models with less curvature.
+As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
+In other words,a low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
+
+With this in mind, if five basis functions are used to fit a GAM for the data that appears in Figure \ref{fig:l-q-response}C, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A.
 The four basis functions (and the intercept) are shown.
-Each of the basis functions is composed of six different points (because there are six points on the timeline).
-To control the ``wiggliness'' of the fit, each of the basis functions of Figure \ref{fig:basis-plot}A is weighted by multiplying it by the smoothness parameter (\(\lambda\)) according to the matrix of Figure \ref{fig:basis-plot}B.
-The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
-A low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
+Each of the basis functions is evaluated at six different points (because there are six points on the timeline).
+The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), that is estimated when fitting the model.
+The penalized estimates fitted for our example are shown in Figure \ref{fig:basis-plot}B.
 
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief {[}\protect\hyperlink{ref-mcelreath2018}{60}{]}.
-In the case of data that shows non-linear trends, the belief that the \emph{true} trend of the data is likely to be smooth rather than extremely ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language) {[}\protect\hyperlink{ref-miller2019}{61}{]}.
-Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian posterior credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good ``frequentist'' coverage (point-wise coverage or ``single point'' coverage), and \emph{across the function} coverage {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge {[}\protect\hyperlink{ref-mcelreath2018}{60}{]}.
+In the case of data that shows non-linear trends, the prior that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language) {[}\protect\hyperlink{ref-miller2019}{61}{]}.
+Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good \emph{across the function} (``frequentist'') coverage{[}\protect\hyperlink{ref-wood2017}{37}{]}.
 
@@
 Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie.
-In a ``point-wise'' CI for non-linear data, the ``true'' function will lie outside of the CI in certain regions at a given frequency (because the data does not follow a linear trend).
-This means that if a point-wise CI is obtained for 100 random samples, the total number of CIs that contain the true function \emph{entirely} is much less than the nominal value of the CI (95\%, usually).
-
-Contrary to this, an \emph{across the function} CI (like those estimated for GAMs using \emph{mgcv}) contains the true function through the entire range of the covariates (which would be time in the case of longitudinal data).
-For GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, the percentage of CIs entirely containing the true function over the entire time period would be close to the nominal value (i.e., if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function), thereby allowing more robust estimates from the model.
+In an \emph{across the function} CI (like those estimated for GAMs using \emph{mgcv}) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%).
+In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI and for this to occur, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
 In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in Miller {[}\protect\hyperlink{ref-miller2019}{61}{]}, Wood{[}\protect\hyperlink{ref-wood2017}{37}{]}, Simpson {[}\protect\hyperlink{ref-simpson2018}{34}{]} and Marra {[}\protect\hyperlink{ref-marra2012}{62}{]}.
@@
 The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]} (a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console).
+
 The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups.
 Although the default \passthrough{\lstinline!method!} used to estimate the smoothing parameters in \emph{mgcv} is generalized cross validation (GCV), Wood{[}\protect\hyperlink{ref-wood2017}{37}{]} showed the restricted maximum likelihood (REML) to be more resistant to overfitting while also easing the quantification of uncertainty in the smooth parameters; therefore in this manuscript REML is always used for smooth parameter estimation.
+An additional argument (\passthrough{\lstinline!family!}) allows to specify the expected distribution of the response, but it is not used in this model because we expect a normally-distributed response (which is the default \passthrough{\lstinline!family!} in \emph{mgcv}).
 
@@
 Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{63}{]}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in {[}\protect\hyperlink{ref-wood2017}{37}{]} and {[}\protect\hyperlink{ref-harezlak2018}{64}{]}.
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood {[}\protect\hyperlink{ref-wood2017}{37}{]} and Harezlak {[}\protect\hyperlink{ref-harezlak2018}{64}{]}.
 
@@
 For model check, the functions \passthrough{\lstinline!gam.check!} and \passthrough{\lstinline!summary!} from \emph{mgcv} provide detailed information about the model fit and its parameters.
-Keep in mind that \passthrough{\lstinline!gam.check!} is a function that also provides the graphical diagnostics obtained using \emph{gratia}, if such graphical output is not desired the source code can be accessed typing \passthrough{\lstinline!gam.check!} in the Console, and the code without the graphical output can be used in a custom function (see @ref(\#gam-00-model-check) for an implementation of this).
+Keep in mind that \passthrough{\lstinline!gam.check!} is a function that also provides the graphical diagnostics obtained using \emph{gratia}, if such graphical output is not desired the source code can be accessed typing \passthrough{\lstinline!gam.check!} in the Console, and the code without the graphical output can be used in a custom function, which is the approach we follow later).
 
@@
 
-Special attention must be paid to the parameter `k-index' from \passthrough{\lstinline!gam.check!} (which calls the function \passthrough{\lstinline!k.check!} to perform the calculation).
+Special attention must be paid to the parameter `k-index' from \passthrough{\lstinline!gam.check!} (which calls \passthrough{\lstinline!k.check!} to perform the calculation).
 This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
@@
 This will be achieved by calling the source code of \passthrough{\lstinline!gam.check!} and using the appropriate code in a new function that will be called \passthrough{\lstinline!gam.diagnostics!}.
-
-\begin{lstlisting}[language=R]
-#Function that uses the source code of `gam.check` to obtain the estimates without the plots. The source can be checked by typing `gam.check` in the console.
-gam.diagnostics<-function (b, old.style = FALSE, type = c("deviance", "pearson", 
-    "response"), k.sample = 5000, k.rep = 200, rep = 0, 
-    level = 0.9, rl.col = 2, rep.col = "gray80", ...) 
-{
-    type <- match.arg(type)
-    resid <- residuals(b, type = type)
-    linpred <- if (is.matrix(b$linear.predictors) && !is.matrix(resid)) 
-        napredict(b$na.action, b$linear.predictors[, 1])
-    else napredict(b$na.action, b$linear.predictors)
-
-    fv <- if (inherits(b$family, "extended.family")) 
-        predict(b, type = "response")
-    else fitted(b)
-    if (is.matrix(fv) && !is.matrix(b$y)) 
-        fv <- fv[, 1]
-    gamm <- !(b$method %in% c("GCV", "GACV", "UBRE", 
-        "REML", "ML", "P-ML", "P-REML", 
-        "fREML"))
-    if (gamm) {
-        cat("\n'gamm' based fit - care required with interpretation.")
-        cat("\nChecks based on working residuals may be misleading.")
-    }
-    else {
-        cat("\nMethod:", b$method, "  Optimizer:", 
-            b$optimizer)
-        if (!is.null(b$outer.info)) {
-            if (b$optimizer[2] %in% c("newton", "bfgs")) {
-                boi <- b$outer.info
-                cat("\n", boi$conv, " after ", boi$iter, 
-                  " iteration", sep = "")
-                if (boi$iter == 1) 
-                  cat(".")
-                else cat("s.")
-                cat("\nGradient range [", min(boi$grad), 
-                  ",", max(boi$grad), "]", sep = "")
-                cat("\n(score ", b$gcv.ubre, " & scale ", 
-                  b$sig2, ").", sep = "")
-                ev <- eigen(boi$hess)$values
-                if (min(ev) > 0) 
-                  cat("\nHessian positive definite, ")
-                else cat("\n")
-                cat("eigenvalue range [", min(ev), ",", 
-                  max(ev), "].\n", sep = "")
-            }
-            else {
-                cat("\n")
-                print(b$outer.info)
-            }
-        }
-        else {
-            if (length(b$sp) == 0) 
-                cat("\nModel required no smoothing parameter selection")
-            else {
-                cat("\nSmoothing parameter selection converged after", 
-                  b$mgcv.conv$iter, "iteration")
-                if (b$mgcv.conv$iter > 1) 
-                  cat("s")
-                if (!b$mgcv.conv$fully.converged) 
-                  cat(" by steepest\ndescent step failure.\n")
-                else cat(".\n")
-                cat("The RMS", b$method, "score gradient at convergence was", 
-                  b$mgcv.conv$rms.grad, ".\n")
-                if (b$mgcv.conv$hess.pos.def) 
-                  cat("The Hessian was positive definite.\n")
-                else cat("The Hessian was not positive definite.\n")
-            }
-        }
-        if (!is.null(b$rank)) {
-            cat("Model rank = ", b$rank, "/", length(b$coefficients), 
-                "\n")
-        }
-    }
-    cat("\n")
-    kchck <- k.check(b, subsample = k.sample, n.rep = k.rep)
-    if (!is.null(kchck)) {
-        cat("Basis dimension (k) checking results. Low p-value (k-index<1) may\n")
-        cat("indicate that k is too low, especially if edf is close to k'.\n\n")
-        printCoefmat(kchck, digits = 3)
-    }
-}
-\end{lstlisting}
+We are not including in the Appendix the code for the function \passthrough{\lstinline!gam.diagnostics!} as it is rather long, but if desired it can be accessed by going to the \passthrough{\lstinline!Appendix.Rmd!} file in the GitHub repository and scrolling to this exact place (the code is not included in the final output, but is evaluated to create the function).
+
+We now call \passthrough{\lstinline!gam.diagnostics!} to provide the desired diagnostic output, as well as a summary of the fitted model, which is obtained by calling \passthrough{\lstinline!summary!}.
 
@@
 \end{lstlisting}
+
+From the output, it can be seen that the \passthrough{\lstinline!k-index!} is 0.36, which indicates that the model is not capturing the variability in the data.
+The \passthrough{\lstinline!edf!} (effective degrees of freedom) is an indicator of the complexity of the smooth.
+Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
 
@@
 
-From the output, it can be seen that the \passthrough{\lstinline!k-index!} is 0.36, which indicates that the model is not capturing the variability in the data.
-The \passthrough{\lstinline!edf!} (effective degrees of freedom) is an indicator of the complexity of the smooth.
-Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
-
 From the \passthrough{\lstinline!summary!} function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
@@
 The major flaw of \passthrough{\lstinline!gam\_00!} is that this model is not taking into account the fact that the data is nested in groups.
-The next iteration is a model where a different smooth of time (Day) is assigned for each group using \passthrough{\lstinline!by=Group!} in the model syntax.
+The next iteration is a model where a different smooth of time (Day) is assigned for each group using \passthrough{\lstinline!by = Group!} in the model syntax.
 
@@
 gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5),
-            method='REML',
+            method ='REML',
             data  = dat_sim)
@@
 \end{lstlisting}
+
+Diagnostics for this model indicate that the k-index is still below 1 (0.43 from \passthrough{\lstinline!gam.check!}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
+Moreover, the smooths (plotted via the \passthrough{\lstinline!draw()!} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
 
@@
 
-Diagnostics for this model indicate that the k-index is still below 1 (0.43 from \passthrough{\lstinline!gam.check!}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
-Moreover, the smooths (plotted via the \passthrough{\lstinline!draw()!} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
 From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \(\approx\) 51\%.
@@
 \end{lstlisting}
+
+The resulting model is \passthrough{\lstinline!gam\_02!}, which is the model fitted in the main manuscript.
+By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
+From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\(\approx\) 1.02).
 
@@
 
-The resulting model is \passthrough{\lstinline!gam\_02!}, which is the model fitted in the main manuscript.
-By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\(\approx\) 1.02), and \passthrough{\lstinline!summary!} now indicates that the model is able to capture 89\% of the variance in the data.
+From \passthrough{\lstinline!summary!}, the model is able to capture 89\% of the variance in the data, which is a substantial improvement over the variance explained by \passthrough{\lstinline!gam\_00!} and \passthrough{\lstinline!gam\_01!}.
 

### 2021-12-01 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Updated reviewer's comments and Appendix

@@
 We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
-Of notice, in Section \@ref(longitudinal-GAMs) we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
+In Section \@ref(longitudinal-GAMs) we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
 
@@
 Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
-It could be argued that a LMEM with quadratic effects should have been used to fit the data in Figure\@ref(fig:l-q-response), F.
+It could be argued that a LMEM with quadratic effects should have been used to fit the data in Figure\@ref(fig:l-q-response)F.
 However, because in reality the true function is not known, choosing a polynomial degree causes more questions (e.g., is it quadratic?, cubic?, or a higher degree?).

### 2021-12-01 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated reviewer's comments and Appendix

@@
 Where $\beta_{0}$ is the intercept, and $\beta_{1}$ is the coefficient for each treatment group.
-Notice that the construction of the predictor is similar to that of Equation \@ref(GLM-eta), but in this case the parametric terms involving the effect of time have been replaced by $f(time_{t}|\beta_{j})$, which represents the smooth function of time with with inputs as the covariates $time_t$ and parameters $\beta_{j}$.
+Notice that the construction of the predictor is similar to that of Equation \@ref(eq:GLM-eta), but in this case the parametric terms involving the effect of time have been replaced by $f(time_{t}|\beta_{j})$, which represents the smooth function of time with with inputs as the covariates $time_t$ and parameters $\beta_{j}$.
 A GAM version of a linear model can be written as:

### 2021-12-01 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated reviewer's comments and Appendix

@@
 The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths [@simpson2018] (a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console).
+
 The parametric term `Group` is added to quantify overall mean differences in the effect of treatment between groups.
 Although the default `method` used to estimate the smoothing parameters in _mgcv_ is generalized cross validation (GCV), Wood[@wood2017] showed the restricted maximum likelihood (REML) to be more resistant to overfitting while also easing the quantification of uncertainty in the smooth parameters; therefore in this manuscript REML is always used for smooth parameter estimation.
+An additional argument (`family`) allows to specify the expected distribution of the response, but it is not used in this model because we expect a normally-distributed response (which is the default `family` in _mgcv_).
 
@@
 Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_ [@gratia].
-A guide for model selection and diagnostics is in the [Appendix](#workflow), and an in-depth analysis can be found in [@wood2017] and [@harezlak2018].
+A guide for model selection and diagnostics is in the [Appendix](#workflow), and an in-depth analysis can be found in Wood [@wood2017] and Harezlak [@harezlak2018].
 

### 2021-12-01 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/07-Appendix.Rmd
Updated reviewer's comments and Appendix

@@
 For model check, the functions `gam.check` and `summary` from  _mgcv_ provide detailed information about the model fit and its parameters.
-Keep in mind that `gam.check` is a function that also provides the graphical diagnostics obtained using _gratia_, if such graphical output is not desired the source code can be accessed typing `gam.check` in the Console, and the code without the graphical output can be used in a custom function (see \@ref(#gam-00-model-check) for an implementation of this).
+Keep in mind that `gam.check` is a function that also provides the graphical diagnostics obtained using _gratia_, if such graphical output is not desired the source code can be accessed typing `gam.check` in the Console, and the code without the graphical output can be used in a custom function, which is the approach we follow later).
 (ref:first-GAM-caption) Graphical diagnostics for the first GAM model.
@@
 
-Special attention must be paid to the parameter 'k-index' from `gam.check` (which calls the function `k.check` to perform the calculation).
+Special attention must be paid to the parameter 'k-index' from `gam.check` (which calls `k.check` to perform the calculation).
 This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
@@
 This will be achieved by calling the source code of `gam.check` and using the appropriate code in a new function that will be called `gam.diagnostics`.
-
-```{r, first-GAM-check, fig.width=10, fig.height=7, out.width='75%', fig.align='center', echo=TRUE, include=TRUE, message=FALSE,warning=FALSE}
+We are not including in the Appendix the code for the function `gam.diagnostics` as it is rather long, but if desired it can be accessed by going to the `Appendix.Rmd` file in the GitHub repository and scrolling to this exact place (the code is not included in the final output, but is evaluated to create the function).
+
+```{r, first-GAM-check, fig.width=10, fig.height=7, out.width='75%', fig.align='center', echo=TRUE, include=FALSE, message=FALSE,warning=FALSE}
 
@@
 
+We now call `gam.diagnostics` to provide the desired diagnostic output, as well as a summary of the fitted model, which is obtained by calling `summary`.
+
 ```{r,gam-00-diagnostics, echo=TRUE, include=TRUE}
 gam.diagnostics(gam_00)
-
-summary(gam_00)
 
@@
 Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
+
+```{r, summary-gam_00}
+summary(gam_00)
+
+```
 
@@
 The major flaw of `gam_00` is that this model is not taking into account the fact that the data is nested in groups.
-The next iteration is a model where a different smooth of time (Day) is assigned for each group using `by=Group` in the model syntax.
+The next iteration is a model where a different smooth of time (Day) is assigned for each group using `by = Group` in the model syntax.
 
@@
 gam.diagnostics(gam_01)
-summary(gam_01)
 ```
@@
 Moreover, the smooths (plotted via the `draw()` function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
+
+```{r,summary-gam_01}
+
+summary(gam_01)
+```
+
 From `summary()`, the deviance explained by the model is $\approx$ 51%.
@@
 
-Model `gam_00` was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted, the model diagnostics, and, the low variance explained by the model.
+Model `gam_00` was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted (a single smooth), the model diagnostics, and, the low variance explained by the model.
 On the other hand, `gam_01` takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \@ref(longitudinal-GAMs), in order to differentiate between each group a parametric term needs to be added to the model for the interaction of _Day_ and _Group_.
+
+This is because in `gam_01`  separate smooths were fitted per group and those smooths also tried to account for the different means of the response in the two groups.
+Adding a parametric term for `Group` enables the smooths to capture the time course-differences of each group.
+The resulting model is `gam_02`, which is the model fitted in the main manuscript.
 
@@
 gam.diagnostics(gam_02)
+
+```
+
+By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
+From `gam.check`, the k-index is now at an acceptable value ($\approx$ 1.02).
+
+```{r,summary-gam_02}
 summary(gam_02)
@@
 
-The resulting model is `gam_02`, which is the model fitted in the main manuscript.
-By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From `gam.check`, the k-index is now at an acceptable value ($\approx$ 1.02), and `summary` now indicates that the model is able to capture 89% of the variance in the data.
+From `summary`, the model is able to capture 89% of the variance in the data, which is a substantial improvement over the variance explained by `gam_00` and `gam_01`.
 

### 2021-12-02 Manuscripts/Manuscript_by_chapters-SIM_Revisions/SIM_Appendix.tex
Started code for simultaneous interval

@@
 
-Mathematically speaking, a rm-ANOVA model with an interaction can be written as:
+Mathematically speaking, a rm-ANOVA model with an interaction can be written as
 
@@
 
-Model \passthrough{\lstinline!gam\_00!} was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted, the model diagnostics, and, the low variance explained by the model.
+Model \passthrough{\lstinline!gam\_00!} was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted (a single smooth), the model diagnostics, and, the low variance explained by the model.
 On the other hand, \passthrough{\lstinline!gam\_01!} takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \ref{longitudinal-GAMs}, in order to differentiate between each group a parametric term needs to be added to the model for the interaction of \emph{Day} and \emph{Group}.
+
+This is because in \passthrough{\lstinline!gam\_01!} separate smooths were fitted per group and those smooths also tried to account for the different means of the response in the two groups.
+Adding a parametric term for \passthrough{\lstinline!Group!} enables the smooths to capture the time course-differences of each group.
+The resulting model is \passthrough{\lstinline!gam\_02!}, which is the model fitted in the main manuscript.
 
@@
 
-The resulting model is \passthrough{\lstinline!gam\_02!}, which is the model fitted in the main manuscript.
 By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.

### 2021-12-02 Code and miscellaneous/Code_simultaenous_intervals.R

@@ -1,0 +1,133 @@
+set.seed(2021)
+
+thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
+
+#data
+dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
+            Day=rep(c(0,2,5,7,10),times=2),
+            Group=as.factor(rep(c("Control","Treatment"),each=5))
+)
+#alpha for ribbon
+al<-0.8
+
+simulate_data <- function(dat, n = 10, sd = 5) {
+    dat_sim <- dat %>%
+        slice(rep(1:n(), each = n)) %>%
+        group_by(Group, Day) %>%
+        mutate(
+            StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
+            subject=rep(1:10),
+            subject=factor(paste(subject, Group, sep = "-"))
+        ) %>%
+        ungroup()
+
+    return(dat_sim)
+}
+
+
+#subject = factor(paste(subject, treatment, sep = "-")))
+
+n <- 10 #number of observations
+sd <- 10 #approximate sd from paper
+df <- 6
+dat_sim <- simulate_data(dat, n, sd)
+
+#gam
+
+gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
+              method='REML',
+              data  = dat_sim)
+
+
+#creates a dataframe using the length of the covariates for the GAM
+gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
+                           Day = seq(0, 10, by = 0.1),
+                           subject=factor(rep(1:10)))
+
+#adds the predictions to the grid and creates a confidence interval for GAM
+gam_predict<-gam_predict%>%
+    mutate(fit = predict(gam_02,gam_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(gam_02, gam_predict,se.fit = TRUE,type='response')$se.fit)
+
+
+
+
+rmvn <- function(n, mu, sig) { ## MVN random deviates
+    L <- mroot(sig)
+    m <- ncol(L)
+    t(mu + L %*% matrix(rnorm(m*n), m, n))
+}
+
+
+
+Vb <- vcov(gam_02)
+newd <- expand_grid(Group = factor(c("Control", "Treatment")),
[…]

### 2021-12-05 Code and miscellaneous/Simultaneous-interval-differences.Rmd
Updated script for spline comparisons

---
title: "Untitled"
author: "Ariel Mundo"
date: "12/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,setup}
library(tidyverse)
library(mgcv)
library(gratia)
library(viridis)
library(scico)
set.seed(2021) #set seed for reproducibility
thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))

```

I follow the same approach from our paper here: simulated from the truth, 10 repeated measurements at each timepoint.

```{r,generate-simulated-data}
#data, "the truth"
dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
            Day=rep(c(0,2,5,7,10),times=2),
            Group=as.factor(rep(c("Control","Treatment"),each=5))
)

#simulate data
simulate_data <- function(dat, n = 10, sd = 5) {
    dat_sim <- dat %>%
        slice(rep(1:n(), each = n)) %>%
        group_by(Group, Day) %>%
        mutate(
            StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
            subject=rep(1:10),
            subject=factor(paste(subject, Group, sep = "-"))
        ) %>%
        ungroup()

    return(dat_sim)
}

#subject = factor(paste(subject, treatment, sep = "-")))

n <- 10 #number of observations
sd <- 10 #approximate sd from paper
df <- 6
dat_sim <- simulate_data(dat, n, sd)

```

Now, I fit the GAM.

```{r,fit-GAM}
#GAM, interaction of time and treatment

gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
              method='REML',
              data  = dat_sim)
```

Now I follow the exact code from G. Simpson's post to generate the simutaneous interval (link to the original post below):
https://fromthebottomoftheheap.net/2016/12/15/simultaneous-interval-revisited/

```{r,simulatenous-interval}
#Generate random values from multivariate normal
rmvn <- function(n, mu, sig) { ## MVN random deviates
    L <- mroot(sig)
    m <- ncol(L)
    t(mu + L %*% matrix(rnorm(m*n), m, n))
}

#Bayesian covariance matrix of the model coefficients
Vb <- vcov(gam_02)

#grid of values for confidence band
newd <- expand_grid(Group = factor(c("Control", "Treatment")),
                                   Day = seq(0, 10, by = 0.1),
                                   subject=factor(rep(1:10)))

#Generate predictions and standard errors
pred <- predict(gam_02, newd, se.fit = TRUE)
se.fit <- pred$se.fit

#number of draws
N <- 10000

#bias of the estimated model coefficients
BUdiff <- rmvn(N, mu = rep(0, nrow(Vb)), sig = Vb)

#evaluate the basis functions at the locations of x
Cg <- predict(gam_02, newd, type = "lpmatrix")

#Computes deviations between the fitted and true parameters
simDev <- Cg %*% t(BUdiff)

#compute absolute values of the standardized deviations from the true model
absDev <- abs(sweep(simDev, 1, se.fit, FUN = "/"))

#find maximum of the absolute standardized deviations at the grid of values of x for each simulation
masd <- apply(absDev, 2L, max)

#Find critical value used to scale the standard errors to yield the simultaneous interval

crit <- quantile(masd, prob = 0.95, type = 8)

pred <- transform(cbind(data.frame(pred), newd),
                  uprP = fit + (2 * se.fit),
                  lwrP = fit - (2 * se.fit),
                  uprS = fit + (crit * se.fit),
                  lwrS = fit - (crit * se.fit))

#plot the simultaneous and pointwise intervals
ggplot(pred, aes(x =Day,group=Group,color=Group,fill=Group)) +
    geom_path(aes(y=fit),lwd=2)+
    geom_ribbon(aes(ymin = lwrS, ymax = uprS), alpha = 0.2) +
    geom_ribbon(aes(ymin = lwrP, ymax = uprP), alpha = 0.2) +
    labs(y = "Tumor data",
         x = "Days")+
    thm1

```

Below is the code for the pairwise comparisons. I used G. Simpson's base code from:
https://fromthebottomoftheheap.net/2017/10/10/difference-splines-i/

But I have made some modifications (are they right?). First, in the original code the design matrix was calculated again (`xp`, L145 below). But from what I understand we have already calculated this matrix (`Cg` which we calculated for the simultaneous interval on L100 above), also I understand than instead of calculating `dif` using the model coefficients (L158 below), `dif` in our case is essentially `simDev`, which was already calculated above (L103)

```{simultaneous-diff}
##calculate simultaneous diff

    xp <- predict(gam_02, newdata = newdata, type = 'lpmatrix')
    #Find columns in xp where the name contains "Control" and "Treatment"
    col1 <- grepl("Control", colnames(Cg))
    col2 <- grepl("Treatment", colnames(Cg))
    #Find rows in xp that correspond to each treatment
    row1 <- with(newd, Group == "Control")
    row2 <- with(newd, Group == "Treatment")
    ## difference rows of xp for data from comparison
    X <- Cg[row1, ] - Cg[row2, ]
    ## zero out cols of X not involved in the comparison
    X[, ! (col1 | col2)] <- 0
    ## zero out the parametric cols
    #X[, !grepl('^s\\(', colnames(Cg))] <- 0
    dif <- X %*% coef(gam_02)
    se <- sqrt(rowSums((X %*% vcov(gam_02, unconditional = TRUE)) * X))
    crit <- qt(0.05/2, df.residual(gam_02), lower.tail = FALSE)
    upr <- dif + (crit * se)
    lwr <- dif - (crit * se)
   comp1<- data.frame(pair = paste("Control", "Treatment", sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)

comp_StO2_full <- cbind(Day = seq(0, 10, length = 1010),
                        rbind(comp1)) %>%
    mutate(interval=case_when(
        upper>0 & lower<0~"no-diff",
        upper<0~"less",
        lower>0~"greater"
    ))

c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper),
                alpha = 0.5,
                fill="gray") +
    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
    geom_line(color='black',size=1) +

    facet_wrap(~ pair) +
    theme_classic()+
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
        text=element_text(size=18),
        legend.title=element_blank()
    )
c1

### 2021-12-05 Code and miscellaneous/spline-smooths.R

@@ -1,0 +1,346 @@
+library(gratia)
+library(mvnfast)
+library(tidyverse)
+library(patchwork)
+load_mgcv()
+set.seed(2021)
+##
+## re-write two gratia functions to provide pointwise and simultaneous smooths ----
+##
+
+calc_difference_R <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
+  ## make sure f1 and f2 are characters
+  f1 <-  as.character(f1)
+  f2 <-  as.character(f2)
+  cnames <- colnames(Xp)
+  ## columns of Xp associated with pair of smooths
+  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
+  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
+  ## rows of Xp associated with pair of smooths
+  r1 <- data[[by_var]] == f1
+  r2 <- data[[by_var]] == f2
+
+  ## difference rows of Xp for pair of smooths
+  X <- Xp[r1, ] - Xp[r2, ]
+
+  ## zero the cols related to other splines
+  X[, ! (c1 | c2)] <- 0
+
+  ## zero out the parametric cols
+  X[, !grepl('^s\\(', cnames)] <- 0
+
+  ## compute difference
+  sm_diff <- drop(X %*% coefs)
+  se <- sqrt(rowSums((X %*% V) * X))
+  nr <- NROW(X)
+
+  ## Calculate posterior simulation for smooths
+  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
+  rownames(coefs_sim) <- rownames(V)
+  simDev <- X %*% coefs_sim
+  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
+  masd <- apply(absDev, 2, max)
+  crit_s <- quantile(masd, prob = 0.95, type = 8)
+
+
+  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
+              level_1 = rep(f1, nr),
+              level_2 = rep(f2, nr),
+              diff = sm_diff, se = se,
+              lower_s = sm_diff - crit_s * se,
+              upper_s = sm_diff + crit_s*se)
+
+  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
[…]

### 2021-12-10 Code and miscellaneous/Simultaneous-interval-differences.Rmd
Uncommented the "zero parametric cols" line in Section: Longitudinal data analysis with GAMs

@@
     ## zero out the parametric cols
-    #X[, !grepl('^s\\(', colnames(Cg))] <- 0
+    X[, !grepl('^s\\(', colnames(Cg))] <- 0
     dif <- X %*% coef(gam_02)

### 2021-12-10 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Uncommented the "zero parametric cols" line in Section: Longitudinal data analysis with GAMs

@@
     ## zero out the parametric cols
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
+    X[, !grepl('^s\\(', colnames(xp))] <- 0
     dif <- X %*% coef(model)

### 2021-12-10 Code and miscellaneous/spline-smooths.R

@@ -206,7 +206,7 @@ p1 <- ggplot(ci, aes(x = x, y = est, group = smooth)) +
                 mapping = aes(ymin = lower, ymax = upper, x = x, group = smooth),
                 fill = "grey80", inherit.aes = FALSE, alpha = 0.5)
 
-p1
+p1 #this plot here produces confidence intervals that appear to be shifted, why?
 
 
 

### 2021-12-14 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated pairwise comparisons, simultaneous CI for smooths

@@
 
-#creates a dataframe using the length of the covariates for the GAM
-gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
-                         subject=factor(rep(1:10)))
+## point-wise interval
+ci <- confint(gam_02, parm = "s(Day)", partial_match = TRUE, type = "confidence")
+## simultaneous interval
+si <- confint(gam_02, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
+
+# mean shift for Treatment group
+const <- coef(gam_02)[2]
+
+#pointwise confidence interval
+ci <- ci %>%
+mutate(est = case_when(Group == "Treatment" ~ est + const,
+TRUE ~ est),
+lower = case_when(Group == "Treatment" ~ lower + const,
+TRUE ~ lower),
+upper = case_when(Group == "Treatment" ~ upper + const,
+TRUE ~ upper))
+
+#simultaneous interval
+si <- si %>%
+mutate(est = case_when(Group == "Treatment" ~ est + const,
+TRUE ~ est),
+lower = case_when(Group == "Treatment" ~ lower + const,
+TRUE ~ lower),
+upper = case_when(Group == "Treatment" ~ upper + const,
+TRUE ~ upper))
 
@@
 
-#adds the predictions to the grid and creates a confidence interval for GAM
-gam_predict<-gam_predict%>%
-    mutate(fit = predict(gam_02,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(gam_02, gam_predict,se.fit = TRUE,type='response')$se.fit)
-
 #using lm
@@
 
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=al,
-              data=gam_predict,
-            show.legend=FALSE,
-            inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
+#plot smooths and pointwise and simulatenous confidence intervals for GAM
+f3<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
+geom_line(lwd = 1) +
+geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
+inherit.aes = FALSE, alpha = 0.7,
+show.legend=FALSE) +
+geom_ribbon(data = si,
+mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
+inherit.aes = FALSE, alpha = 0.4,
+show.legend=FALSE)+
+    geom_point(data=dat_sim, aes(x = Day, 
+                        y = StO2_sim, 
+                        color = Group), 
+                        size=1.5,
+                        alpha=0.6, 
+               inherit.aes = FALSE,
                show.legend = FALSE)+
-  #facet_wrap(~Group)+
     labs(y=expression(atop(StO[2],'complete')))+
@@
 
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
+ci <- confint(mod_m1, parm = "s(Day)", partial_match = TRUE, type = "confidence")
+## simultaneous interval
+si <- confint(mod_m1, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
 
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
+# mean shift for group 2
+const <- coef(mod_m1)[2]
 
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=al,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
+ci <- ci %>%
+mutate(est = case_when(Group == "Treatment" ~ est + const,
+TRUE ~ est),
+lower = case_when(Group == "Treatment" ~ lower + const,
+TRUE ~ lower),
+upper = case_when(Group == "Treatment" ~ upper + const,
+TRUE ~ upper))
+
+si <- si %>%
+mutate(est = case_when(Group == "Treatment" ~ est + const,
+TRUE ~ est),
+lower = case_when(Group == "Treatment" ~ lower + const,
+TRUE ~ lower),
+upper = case_when(Group == "Treatment" ~ upper + const,
+TRUE ~ upper))
+
+f6<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
+geom_line(lwd = 1) +
+geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
+inherit.aes = FALSE, alpha = 0.7,
+show.legend=FALSE) +
+geom_ribbon(data = si,
+mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
+inherit.aes = FALSE, alpha = 0.4,
+show.legend=FALSE)+
+    geom_point(data=dat_missing, aes(x = Day, 
+                        y = StO2_sim, 
+                        color = Group), 
+                            size=1.5,
+                        alpha=0.6, 
+               inherit.aes = FALSE)+
     labs(y=expression(atop(StO[2],'missing')))+
@@
   theme(
-    axis.text=element_text(size=22)
-  )+thm1
+    axis.text=element_text(size=22))+
+        thm1
 
@@
    plot_annotation(tag_levels='A')&
-   ylim(c(-7,75)) &
+   ylim(c(-20,75)) &
   theme(
@@
 ```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
+
 ##Pairwise comparisons
 
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
+calc_difference_R <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
+  ## make sure f1 and f2 are characters
+  f1 <-  as.character(f1)
+  f2 <-  as.character(f2)
+  cnames <- colnames(Xp)
+  ## columns of Xp associated with pair of smooths
+  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
+  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
+  ## rows of Xp associated with pair of smooths
+  r1 <- data[[by_var]] == f1
+  r2 <- data[[by_var]] == f2
 
-#this function takes the model, grid and groups to be compared using the lpmatrix
+  ## difference rows of Xp for pair of smooths
+  X <- Xp[r1, ] - Xp[r2, ]
 
-smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    #Find columns in xp where the name contains "Control" and "Treatment"
-    col1 <- grepl(g1, colnames(xp))
-    col2 <- grepl(g2, colnames(xp))
-    #Find rows in xp that correspond to each treatment
-    row1 <- with(newdata, Group == g1)
-    row2 <- with(newdata, Group == g2)
-    ## difference rows of xp for data from comparison
-    X <- xp[row1, ] - xp[row2, ]
-   ## zero out cols of X not involved in the comparison
-    X[, ! (col1 | col2)] <- 0
+  ## zero the cols related to other splines
+ # X[, ! (c1 | c2)] <- 0
+
   ## zero out the parametric cols
-    X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(g1, g2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
+  #X[, !grepl('^s\\(', cnames)] <- 0
+
+  ## compute difference
+  sm_diff <- drop(X %*% coefs)
+  se <- sqrt(rowSums((X %*% V) * X))
+  nr <- NROW(X)
+
+  ## Calculate posterior simulation for smooths
+  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
+  rownames(coefs_sim) <- rownames(V)
+  simDev <- X %*% coefs_sim
+  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
+  masd <- apply(absDev, 2, max)
+  crit_s <- quantile(masd, prob = 0.95, type = 8)
+
+  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
+              level_1 = rep(f1, nr),
+              level_2 = rep(f2, nr),
+              diff = sm_diff, se = se,
+              lower_s = sm_diff - crit_s * se,
+              upper_s = sm_diff + crit_s*se)
+
+  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
+  ## Only need rows associated with one of the levels
+  out <- bind_cols(out, data[r1, smooth_var])
+
+  out
 }
 
-comp1<-smooth_diff(gam_02,pdat,'Control','Treatment')
+#does both ci and si
+difference_smooths_R <- function(model,
+                                 smooth,
+                                 n = 100,
+                                 ci_level = 0.95,
+                                 newdata = NULL,
+                                 partial_match = TRUE,
+                                 unconditional = FALSE,
+                                 frequentist = FALSE,
+                                 nrep = 10000,
+                                 include_means = TRUE,
+                                 ...) {
+  if (missing(smooth)) {
+    stop("Must specify a smooth to difference via 'smooth'.")
+  }
 
-comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
+  # smooths in model
+  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
+  # select smooths
+  select <-
+    gratia:::check_user_select_smooths(smooths = S, select = smooth,
+                                       partial_match = partial_match)#,
+  # model_name = expr_label(substitute(object)))
+  sm_ids <- which(select)
+  smooths <- gratia::get_smooths_by_id(model, sm_ids)
+  sm_data <- map(sm_ids, gratia:::smooth_data,
+                 model = model, n = n, include_all = TRUE)
+  sm_data <- bind_rows(sm_data)
+  by_var <- by_variable(smooths[[1L]])
+  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
+  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
+                                   stringsAsFactor = FALSE))
+  names(pairs) <- paste0("f", 1:2)
+
+  Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
+  V <- gratia:::get_vcov(model, unconditional = unconditional,
+                         frequentist = frequentist)
+  coefs <- coef(model)
+
+  out <- pmap(pairs, calc_difference_R, smooth = smooth, by_var = by_var,
+              smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
+              coefs = coefs, nrep = nrep)
+  out <- bind_rows(out)
+  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
+
+  out <- add_column(out,
+                    lower = out$diff - (crit * out$se),
+                    upper = out$diff + (crit * out$se),
+                    .after = 6L)
+  out
+}
+
+#compute difference between smooths and calculate confidence interval: complete data
+diff_complete <- difference_smooths_R(gam_02, smooth = "s(Day)", newdata = newdat,
+                     unconditional = TRUE, frequentist = FALSE,
+                     n=100, partial_match = TRUE, nrep=10000)
 
@@
     #this is the region where the control group effect is greater
+
     v1<-dataframe%>%
-        filter(lower>0)%>%
+        filter(lower_s>0)%>%
         select(Day)
@@
     final1=v1$Day[[nrow(v1)]]
-
     #extract values where the value of the upper limit of the ribbon is lower than zero
     #this corresponds to the region where the treatment group effect is greater
-    v2<-comp_StO2_full%>%
-        filter(upper<0)%>%
+    v2<-dataframe%>%
+        filter(upper_s<0)%>%
         select(Day)
@@
     return(my_list)
+
 }
 
-my_list<-pairwise_limits(comp_StO2_full)
+my_list<-pairwise_limits(diff_complete)
 rib_col<-'#8D7D82' #color for ribbon for confidence interval
@@
 
-c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
+c1 <- ggplot() +
+  geom_line(data = diff_complete, aes(x = Day, y = diff),size=1, alpha=0.5) +
     annotate("rect",
@@
                 alpha = 0.5,
-                ) +
+                )+
     annotate("text",
@@
              fill=treat_rib,
+             alpha = 0.5
+             ) +
+  annotate("text",
+             x=6,
+             y=-18,
+             label="Treatment>Control",
+             size=8,
+           angle=90
+           )+
+  geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s, ymax = upper_s),
+              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
+  geom_hline(yintercept = 0, lty = 2, color = "red")+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    theme(
+    axis.text=element_text(size=22))+
+    theme_classic()
+
+###compute difference in smooths for missing data
+diff_missing <- difference_smooths_R(mod_m1, smooth = "s(Day)", newdata = newdat,
+                     unconditional = TRUE, frequentist = FALSE,
+                     n=100, partial_match = TRUE, nrep=10000)
+
+v2<- diff_missing %>%
+        filter(upper_s<0)%>%
+        select(Day)
+init2=v2$Day[[1]]
+final2=v2$Day[[nrow(v2)]]
+
+c2 <- ggplot() +
+  geom_line(data = diff_missing, aes(x = Day, y = diff),size=1, alpha=0.5) +
+    annotate("rect",
+             xmin =init2, xmax = final2,ymin=-Inf,ymax=Inf,
+             fill=treat_rib,
              alpha = 0.5,
@@
            )+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+  geom_ribbon(data = diff_missing, aes(x = Day, ymin = lower_s, ymax = upper_s),
+              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
+  geom_hline(yintercept = 0, lty = 2, color = "red")+
     scale_x_continuous(breaks=c(0,2,5,7,10))+
     theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-###for missing data
-comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
-comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp2))
-
-missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    labs(x = 'Days', 
-         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
-                              )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme_classic()+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
-my_list<-pairwise_limits(comp_StO2_missing)
-
-c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    annotate("rect",
-             xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-             fill=control_rib,
-             alpha = 0.5
-             ) +
-  annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=8,
-           angle=90
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill= treat_rib,
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=8,
-           angle=90
-           )+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
+    axis.text=element_text(size=22))+
+    theme_classic()
 
@@
      )
-
 ```

### 2021-12-14 Code and miscellaneous/spline-smooths.R

@@ -57,7 +57,7 @@ calc_difference_R <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, c
   out
 }
 
-
+#does both ci and si
 difference_smooths_R <- function(model,
                                  smooth,
                                  n = 100,
@@ -118,7 +118,7 @@ difference_smooths_R <- function(model,
 ## Simulate some data like we have in our example ----
 ##
 
-# set.seed(2)
+#  set.seed(2)
 # n <- 50
 # dat <- data.frame(x=seq(1:n), group = factor(rep(c(1, 2), each = n)))
 # # create the grouped smooths

### 2021-12-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Uncommented the "zero parametric cols" line in Section: Longitudinal data analysis with GAMs
Updated figures with simultaneous and pointwise CI

@@
 \caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+\textbf{A}: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
 Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot}
+\textbf{B}: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
+Lines represent trends for each group, shaded regions are 95\% across the function (inner region) and simultaneous (outer region) confidence intervals.
+\textbf{C}: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
+\textbf{D}: Smooths for the GAM model for the simulated data with missing observations (40\%).
+Lines represent trends for each group, shaded regions are 95\% across the function (inner region) and simultaneous (outer region) confidence intervals.}\label{fig:sim-smooth-plot}
 \end{figure}
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{Full_document_SIM_No_Appendix_files/figure-latex/plot-pairwise-comp-1}
+{\centering \includegraphics[width=1\linewidth]{Full_document_SIM_No_Appendix_files/figure-latex/plot-pairwise-comp-1}
 

### 2021-12-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated figures with simultaneous and pointwise CI

@@
 Moreover, GAMs are considered "empirical" Bayesian models when fitted using the package _mgcv_ because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the "fully Bayesian" case, which can be fitted using JAGS, Stan, or other probabilistic programming language) [@miller2019].
-Therefore, the confidence intervals (CIs) calculated for the smooth terms using _mgcv_ are considered empirical Bayesian credible intervals [@pedersen2019], which have good _across the function_ ("frequentist") coverage[@wood2017].
+Therefore, the confidence intervals (CIs) calculated by default for the smooth terms using _mgcv_ are considered empirical Bayesian credible intervals [@pedersen2019], which have good _across the function_ ("frequentist") coverage[@wood2017].
 
@@
 In other words, we expect that about 95% of the points that compose the true function will be covered by the across the function CI and for this to occur, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
+
+**add about simultaneous interval**
 In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in Miller [@miller2019], Wood[@wood2017], Simpson [@simpson2018] and Marra [@marra2012].

### 2021-12-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated figures with simultaneous and pointwise CI

@@
                show.legend = FALSE)+
-    labs(y=expression(atop(StO[2],'complete')))+
+    labs(y=expression(atop(StO[2],'(complete observations)')))+
     scale_x_continuous(breaks=c(0,2,5,7,10))+
@@
 inherit.aes = FALSE, alpha = 0.4,
-show.legend=FALSE)+
+show.legend=TRUE)+
     geom_point(data=dat_missing, aes(x = Day, 
@@
                inherit.aes = FALSE)+
-    labs(y=expression(atop(StO[2],'missing')))+
+    labs(y=expression(atop(StO[2],'(missing observations)')))+
     scale_x_continuous(breaks=c(0,2,5,7,10))+
@@
 (ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+**A**: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
 Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95% confidence intervals.
-C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40% of its observations missing.
-Lines represent trends for each group, shaded regions are 95% empirical Bayesian confidence intervals.
+**B**: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
+Lines represent trends for each group, shaded regions are 95% across the function (inner region) and simultaneous (outer region) confidence intervals.
+**C**: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
+**D**: Smooths for the GAM model for the simulated data with missing observations (40%).
+Lines represent trends for each group, shaded regions are 95%  across the function (inner region) and simultaneous (outer region) confidence intervals.
 
@@
 
-calc_difference_R <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
+difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
   ## make sure f1 and f2 are characters
@@
 #does both ci and si
-difference_smooths_R <- function(model,
+difference_smooths <- function(model,
                                  smooth,
@@
 
-  out <- pmap(pairs, calc_difference_R, smooth = smooth, by_var = by_var,
+  out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
               smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
@@
 #compute difference between smooths and calculate confidence interval: complete data
-diff_complete <- difference_smooths_R(gam_02, smooth = "s(Day)", newdata = newdat,
+diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
                      unconditional = TRUE, frequentist = FALSE,
@@
              label="Control>Treatment",
-           size=8,
+           size=6,
            angle=90
@@
              label="Treatment>Control",
-             size=8,
+             size=6,
            angle=90
@@
     scale_x_continuous(breaks=c(0,2,5,7,10))+
+    labs(y="Difference\n(Complete observations)")+
+    theme_classic()+
     theme(
-    axis.text=element_text(size=22))+
-    theme_classic()
+    axis.text=element_text(size=22))
 
 ###compute difference in smooths for missing data
-diff_missing <- difference_smooths_R(mod_m1, smooth = "s(Day)", newdata = newdat,
+diff_missing <- difference_smooths(mod_m1, smooth = "s(Day)", newdata = newdat,
                      unconditional = TRUE, frequentist = FALSE,
@@
     scale_x_continuous(breaks=c(0,2,5,7,10))+
+    labs(y="Difference\n(Missing observations)")+
+    theme_classic()+
     theme(
-    axis.text=element_text(size=22))+
-    theme_classic()
+    axis.text=element_text(size=22))
 
@@
 
-```{r,plot-pairwise-comp,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-comp)'}
+```{r,plot-pairwise-comp,fig.width=12, fig.height=7, out.width='100%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-comp)'}
 

### 2021-12-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.Rmd
Updated reviewer's comments, Appendix, Main manuscript

@@
 
-- L325: We indicate that the term $\varepsilon_{ijt}$ is the residual, and not a separable error.
-Of notice, we are following the notation provided by Dr.
-Simpson in Eq.
-3 in the paper "Modelling Modelling Palaeoecological Time Series Using Generalised Additive Models".
-**do we need this here?
-I am adding it because we are following the same model specification that he has in this paper**
+- L325: We indicate that the term $\varepsilon_{ijt}$ is the deivation of each observation from the mean.
 
@@
 
-- L396-L403 now contain a more detailed and accurate explanation of"across-the-function" CIs.
+- L396-L404 now contain a more detailed and accurate explanation of"across-the-function" CIs.
 We have used the reviewer's comments to indicate that in GAM fitted using _mgcv_, the close to nominal across-the-function-coverage is true when averaged over the entire function.
+
+- L405-411 now provide a brief explanation of simultaneous CIs.
+With this, we want to provide the reader with an intuitive understanding on why simultaneous CIs are more robust than across the function CIs.
+
+- We have modified Figure 3 to now include across the function and simultaneous CIs.
+Figs.
+3B,D now show the fitted smooths and both intervals, and we explain their inclusion in L473-478.
+In this way, the reader has a visual understanding of the difference between across the function and simultaneous CIs.
 
@@
 You would have to create differences conditional upon a level of
-the other factor. Users, however, tend to expect differences to include the group means, as you have chosen to present things here. Hence a note exp
+the other factor.
+Users, however, tend to expect differences to include the group means, as you have chosen to present things here.
+Hence a note explaining the implications of this choice would be useful to forewarn the user.
 
 ## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We have made major changes to this section based on the reviewer's comments.
+
+- We have removed the sentence "if the difference between the credible intervals of the fitted smooths...
+is non-zero", and now L484 reads: "is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference."
+
+- L490 now reads "unlikely to be distinguishable from zero".
+
+- L493-502 now indicate that we are computing the difference using a simultaneous CI, and that we _included_ the group means in our pairwise comparison.
+We also indicate to the reader the fact that if multiple parametric terms exist in the model the inclusion of the means can become problematic.
+
+However, we believe that including the means makes the comparison more intuitive, as in that way the researcher can directly compare both the fitted smooths and the pairwise comparisons to infer any significant changes and their associated magnitude.
+In the Appendix we also have made the corrections to the calculation of the pairwise comparisons in Section A.4.4.1 (the GAM workflow section) so the reader is aware again that in our code we keep the group means.
 

### 2021-12-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Updated reviewer's comments, Appendix, Main manuscript

@@
 Top row: linear response, bottom row: quadratic response.
-A: Simulated linear data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
-D: Simulated quadratic data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
-B,E: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
+\textbf{A}: Simulated linear data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
+\textbf{D}: Simulated quadratic data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
+\textbf{B,E}: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
 Points represent the original raw data.
 The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation.
-C, F: Estimates from the LMEM in the linear and quadratic case (subject: thin lines, population: thick lines) .
+\textbf{C, F}: Estimates from the LMEM in the linear and quadratic case (subject: thin lines, population: thick lines) .
 The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data and grossly bias the initial estimates for each group in the quadratic case (bottom row).}\label{fig:l-q-response}
@@
 \caption{Basis functions for a single smoother for time.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-B: Matrix of basis function weights.
+\textbf{A}: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
+\textbf{B}: Matrix of basis function weights.
 Each basis function is multiplied by a coefficient which can be positive or negative.
 The coefficient determines the overall effect of each basis in the final smoother.
-C: Weighted basis functions.
+\textbf{C}: Weighted basis functions.
 Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
 Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-D: Smoother for time and original data points.
+\textbf{D}: Smoother for time and original data points.
 The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}} In the case of data that shows non-linear trends, the prior that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good \emph{across the function} (``frequentist'') coverage\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}} In the case of data that shows non-linear trends, the prior that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Therefore, the confidence intervals (CIs) calculated by default for the smooth terms using \emph{mgcv} are considered empirical Bayesian credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good \emph{across the function} (``frequentist'') coverage\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
 
@@
 Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie.
-In an \emph{across the function} CI (like those estimated for GAMs using \emph{mgcv}) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%).
+In an \emph{across the function} CI (like those estimated by default for GAMs using \emph{mgcv}) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%).
 In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI and for this to occur, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in Miller,\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} and Marra.\textsuperscript{\protect\hyperlink{ref-marra2012}{62}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
+
+However, besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Suppose we chose a nominal value (say, 95\%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
+Briefly, to obtain a simultaneous CI we use simulation to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI.\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}} Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups, as we discuss in the following section.
+
+In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller,\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson,\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} Marra,\textsuperscript{\protect\hyperlink{ref-marra2012}{63}} and Ruppert.\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}} With this brief introduction to the Bayesian interpretation of GAMs, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends.
 
@@
 Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
-The simulated and real data appear in Figure \ref{fig:sim-smooth-plot},A.
+The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}A.
 
@@
 
-For the data in Figure \ref{fig:sim-smooth-plot}, A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
+For the data in Figure \ref{fig:sim-smooth-plot}A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
 To estimate this, the model incorporates separate smooths for \emph{Group} and \emph{Day}, respectively.
@@
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{63}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{64}}
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{64}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{65}}
 
@@
 Consider the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}B.
-If 40\% of the observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \ref{fig:sim-smooth-plot}D).
-Although the confidence intervals have increased for both smooths, the model still shows different trends with as few as 4 observations per group at certain time points.
+If 40\% of the observations are randomly deleted and the same interaction GAM (\passthrough{\lstinline!gam\_02!}) is fitted, the resulting smooths appear in (Figure \ref{fig:sim-smooth-plot}D).
+Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.
+
+Additionally, note that in Figures \ref{fig:sim-smooth-plot}B,D we show two CIs for each of the fitted smooths (shaded regions).
+The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, we expect them to be wider than the across the function CI, resulting in the wide shaded regions.
+For the dataset with missing observations, the smooths CIs overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset.
+However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.
 
@@
 \textbf{B}: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% across the function (inner region) and simultaneous (outer region) confidence intervals.
+Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.
 \textbf{C}: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
 \textbf{D}: Smooths for the GAM model for the simulated data with missing observations (40\%).
-Lines represent trends for each group, shaded regions are 95\% across the function (inner region) and simultaneous (outer region) confidence intervals.}\label{fig:sim-smooth-plot}
+Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.}\label{fig:sim-smooth-plot}
 \end{figure}
@@
 \caption{Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-B: Pairwise comparisons for the dataset with missing observations.
+\textbf{A}: Pairwise comparisons for the full dataset.
+\textbf{B}: Pairwise comparisons for the dataset with missing observations.
 Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
-In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp}
+In both cases the effect of treatment is significant after day 3.
+In both cases, a simultaneous CI has been computed around the smooth difference.}\label{fig:plot-pairwise-comp}
 \end{figure}
@@
 In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
-\textbf{add sentence about the issue of the Nychka across the function intervals }
-
-around that \emph{difference} between the empirical Bayesian confidence intervals of the fitted smooths for such groups is non-zero, then a significant difference exists at that time point(s).
-The absence of a \emph{p-value} in this case might seem odd, but the empirical Bayesian confidence interval interpretation can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
+
+The absence of a \emph{p-value} in this context might seem odd, but its interpretation can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment.
 This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
@@
 With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant.
-The difference between groups with similar trends is likely to yield zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
+A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
 
 Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}B and Figure \ref{fig:sim-smooth-plot}D.
-Figure \ref{fig:plot-pairwise-comp} shows the comparison between each treatment group for the full and missing datasets.
+Figure \ref{fig:plot-pairwise-comp} shows the comparison between each treatment group for the full and missing datasets with a simultaneous CI computed around the difference.
 Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared.
-Of notice, the pairwise comparison has been set on the response scale (see Appendix for code details), because otherwise the comparison appears shifted and is not intuitively easy to relate to the original data.
-
-With this correction in mind, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
+Notice that because we have included the means of each group, there is correspondence between the scale of the original data the scale of the pairwise comparisons.
+This can be seen as in Figure \ref{fig:sim-smooth-plot}B at day 5 there is essentially a 50\% difference between \(\mbox{StO}_2\) in both groups, which corresponds to the -50 difference in \ref{fig:plot-pairwise-comp}A.
+However, if there are multiple parametric terms in the model (more factors that need to be specified in the model) such inclusion of the means can become problematic.
+However, we believe that the model we have presented here suffices in a wide range of situations where adding the group means is a relatively easy implementation that can help better visualize the estimation from the model from a biological perspective.
+
+In Figure \ref{fig:plot-pairwise-comp}A the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
 Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by \(\approx\) 3 day it is the ``Treatment'' group the one that on average, has greater \(\mbox{StO}_2\).
@@
 
-On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the empirical Bayesian credible intervals of the smooths show the same trend of the full dataset.
-Consequently, the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that the Control Group has higher \(\mbox{StO}_2\) before day 3, and is able to estimate the change on day 3 where the Treatment Group becomes significant as the full dataset smooth pairwise comparison.
+On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that because the confidence intervals overlap for the first two days there is essentially no difference between the groups.
+However, because the model was still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes significant as in the full dataset smooth pairwise comparison.
 
@@
 
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{65}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{66}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.
 
 In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{66}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{67}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).
 
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{67},\protect\hyperlink{ref-weissgerber2018}{68}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{68},\protect\hyperlink{ref-weissgerber2018}{69}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-ruppert2003}{}}%
+\CSLLeftMargin{62. }
+\CSLRightInline{Ruppert D. \emph{Semiparametric Regression}. Cambridge University Press; 2003.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
-\CSLLeftMargin{62. }
+\CSLLeftMargin{63. }
 \CSLRightInline{Marra G, Wood SN.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
-\CSLLeftMargin{63. }
+\CSLLeftMargin{64. }
 \CSLRightInline{Simpson GL. \emph{Gratia: Graceful 'Ggplot'-Based Graphics and Other Functions for GAMs Fitted Using 'Mgcv'}.; 2020. \url{https://CRAN.R-project.org/package=gratia}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
-\CSLLeftMargin{64. }
+\CSLLeftMargin{65. }
 \CSLRightInline{Harezlak J, Ruppert D, Wand MP. \emph{Semiparametric Regression with {R}}. Springer New York; 2018. doi:\href{https://doi.org/10.1007/978-1-4939-8853-2}{10.1007/978-1-4939-8853-2}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
-\CSLLeftMargin{65. }
+\CSLLeftMargin{66. }
 \CSLRightInline{Lang TA, Altman DG.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hastie1995}{}}%
-\CSLLeftMargin{66. }
+\CSLLeftMargin{67. }
 \CSLRightInline{Hastie T, Tibshirani R.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-begley2015}{}}%
-\CSLLeftMargin{67. }
+\CSLLeftMargin{68. }
 \CSLRightInline{Begley CG, Ioannidis JPA.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-weissgerber2018}{}}%
-\CSLLeftMargin{68. }
+\CSLLeftMargin{69. }
 \CSLRightInline{Weissgerber TL, Garcia-Valencia O, Garovic VD, Milic NM, Winham SJ.

### 2021-12-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/SIM_Appendix.tex
Updated reviewer's comments, Appendix, Main manuscript

@@
 Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language) {[}\protect\hyperlink{ref-miller2019}{61}{]}.
-Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good \emph{across the function} (``frequentist'') coverage{[}\protect\hyperlink{ref-wood2017}{37}{]}.
+Therefore, the confidence intervals (CIs) calculated by default for the smooth terms using \emph{mgcv} are considered empirical Bayesian credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good \emph{across the function} (``frequentist'') coverage{[}\protect\hyperlink{ref-wood2017}{37}{]}.
 
@@
 Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie.
-In an \emph{across the function} CI (like those estimated for GAMs using \emph{mgcv}) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%).
+In an \emph{across the function} CI (like those estimated by default for GAMs using \emph{mgcv}) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%).
 In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI and for this to occur, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in Miller {[}\protect\hyperlink{ref-miller2019}{61}{]}, Wood{[}\protect\hyperlink{ref-wood2017}{37}{]}, Simpson {[}\protect\hyperlink{ref-simpson2018}{34}{]} and Marra {[}\protect\hyperlink{ref-marra2012}{62}{]}.
-With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.
+
+However, besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+Suppose we chose a nominal value (say, 95\%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
+Briefly, to obtain a simultaneous CI we use simulation to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI {[}\protect\hyperlink{ref-ruppert2003}{62}{]}.
+Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups, as we discuss in the following section.
+
+In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller {[}\protect\hyperlink{ref-miller2019}{61}{]}, Wood{[}\protect\hyperlink{ref-wood2017}{37}{]}, Simpson {[}\protect\hyperlink{ref-simpson2018}{34}{]}, Marra {[}\protect\hyperlink{ref-marra2012}{63}{]}, and Ruppert {[}\protect\hyperlink{ref-ruppert2003}{62}{]}.
+With this brief introduction to the Bayesian interpretation of GAMs, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends.
 
@@
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{63}{]}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood {[}\protect\hyperlink{ref-wood2017}{37}{]} and Harezlak {[}\protect\hyperlink{ref-harezlak2018}{64}{]}.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{64}{]}.
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood {[}\protect\hyperlink{ref-wood2017}{37}{]} and Harezlak {[}\protect\hyperlink{ref-harezlak2018}{65}{]}.
 
@@
 \caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+\textbf{A}: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
 Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot}
+\textbf{B}: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
+Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.
+\textbf{C}: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
+\textbf{D}: Smooths for the GAM model for the simulated data with missing observations (40\%).
+Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.}\label{fig:sim-smooth-plot}
 \end{figure}
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/plot-pairwise-comp-1}
+{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/plot-pairwise-comp-1}
 
@@
 
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{65}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{66}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.
 
 In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{66}{]}.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{67}{]}.
 This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{67},\protect\hyperlink{ref-weissgerber2018}{68}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{68},\protect\hyperlink{ref-weissgerber2018}{69}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-ruppert2003}{}}%
+\CSLLeftMargin{{[}62{]} }
+\CSLRightInline{D. Ruppert, Semiparametric regression, Cambridge University Press, Cambridge New York, 2003.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
-\CSLLeftMargin{{[}62{]} }
+\CSLLeftMargin{{[}63{]} }
 \CSLRightInline{G.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
-\CSLLeftMargin{{[}63{]} }
+\CSLLeftMargin{{[}64{]} }
 \CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
-\CSLLeftMargin{{[}64{]} }
+\CSLLeftMargin{{[}65{]} }
 \CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with {R}, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
-\CSLLeftMargin{{[}65{]} }
+\CSLLeftMargin{{[}66{]} }
 \CSLRightInline{T.A.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hastie1995}{}}%
-\CSLLeftMargin{{[}66{]} }
+\CSLLeftMargin{{[}67{]} }
 \CSLRightInline{T.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-begley2015}{}}%
-\CSLLeftMargin{{[}67{]} }
+\CSLLeftMargin{{[}68{]} }
 \CSLRightInline{C.G. Begley, J.P.A. Ioannidis, Reproducibility in science, Circulation Research. 116 (2015) 116--126. \url{https://doi.org/10.1161/circresaha.114.303819}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-weissgerber2018}{}}%
-\CSLLeftMargin{{[}68{]} }
+\CSLLeftMargin{{[}69{]} }
 \CSLRightInline{T.L.
@@
 \caption{Basis functions for a single smoother for time.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-B: Matrix of basis function weights.
+\textbf{A}: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
+\textbf{B}: Matrix of basis function weights.
 Each basis function is multiplied by a coefficient which can be positive or negative.
 The coefficient determines the overall effect of each basis in the final smoother.
-C: Weighted basis functions.
+\textbf{C}: Weighted basis functions.
 Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
 Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-D: Smoother for time and original data points.
+\textbf{D}: Smoother for time and original data points.
 The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot-appendix}
@@
 Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
+For simplicity, the confidence intervals (CIs) shown in this section for the models are the across the function CIs created by \passthrough{\lstinline!mgcv!} by default.
+However, for the pairwise comparisons of the third model we use \emph{simultaneous intervals} as in the main manuscript, and the code for creating simultaneous CIs for the smooths can be found in Section \ref{GAM-linear-plot}.
 
@@
 ##                         k'  edf k-index p-value
-## s(Day):GroupControl   4.00 3.87    1.02    0.58
-## s(Day):GroupTreatment 4.00 3.88    1.02    0.56
+## s(Day):GroupControl   4.00 3.87    1.02    0.55
+## s(Day):GroupTreatment 4.00 3.88    1.02    0.58
 \end{lstlisting}
@@
 The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \ref{GAM-significance}.
+
 In this case, the ``design matrix'' is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
@@
 This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
-\textbf{All pairwise comparisons in this paper have been centered at the response scale to ease interpretation }.
+
+We want to emphasize that for the model used here we have included the group means in order to keep the pairwise comparisons on the scale of the response (a change from the code which appears in the chunk with a comment that says \passthrough{\lstinline!\#\#\#IMPORTANT!}).
+This approach works well for models like \passthrough{\lstinline!gam\_02!} but implementing this on models with more parametric terms can be challenging.
+However, we do believe that the model presented in the paper covers a wide range of situations and our approach here for the pairwise comparisons will be useful for most biomedical researchers.
 
@@
 ##Pairwise comparisons
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-##matrix that contains the basis functions evaluated at the points in pdat
-    xp <- predict(gam_02, newdata = pdat, type = 'lpmatrix')
-
-#Find columns in xp where the name contains "Control"
-    c1 <- grepl('Control', colnames(xp))
-
-#Find columns in xp where the name contains 'Treatment'
-    c2 <- grepl('Treatment', colnames(xp))
-
-#Find rows in pdat that correspond to either 'Control' or 'Treatment'
-    r1 <- with(pdat, Group == 'Control')
-    r2 <- with(pdat, Group == 'Treatment')
-
-# In xp: find the rows that correspond to Control or Treatment, those that do not match will be
-    #set to zero. Then, substract the values from the rows corresponding to 'Control' from those that correspond
-    #to 'Treatment'
-    X <- xp[r1, ] - xp[r2, ]
-
-    ## remove columns that do not contain name 'Control' or 'Treatment'
-    X[, ! (c1 | c2)] <- 0
-    ## zero out the parametric cols, those that do not contain in the characters 's('
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
-
-    #Multiply matrix by model coefficients. X has (p,n) (rows, columns) and the coefficient matrix has
-    #dimensions (n,1). The resulting matrix has dimensions (p,1)
-    dif <- X %*% coef(gam_02)
-
-#Calculate standard error for the computed differences using the variance-covariance matrix
-    #of the model
-    se <- sqrt(rowSums((X %*% vcov(gam_02, unconditional = FALSE)) * X))
-    crit <- qt(0.05/2, df.residual(gam_02), lower.tail = FALSE)
-    #upper  limits
-    upr <- dif + (crit * se)
-    #lower limits
-    lwr <- dif - (crit * se)
-    #put all components in a dataframe for plotting
-    comp1<-data.frame(pair = paste('Control', 'Treatment', sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-
-#add time point sequence
-comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1))
-
-#use function from the pairwise comparison plot in the manuscript to get the shaded regions
-my_list<-pairwise_limits(comp_StO2)
-
-#plot the difference
+difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
+  ## make sure f1 and f2 are characters
+  f1 <-  as.character(f1)
+  f2 <-  as.character(f2)
+  cnames <- colnames(Xp)
+  ## columns of Xp associated with pair of smooths
+  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
+  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
+  ## rows of Xp associated with pair of smooths
+  r1 <- data[[by_var]] == f1
+  r2 <- data[[by_var]] == f2
+
+  ## difference rows of Xp for pair of smooths
+  X <- Xp[r1, ] - Xp[r2, ]
+
+  ### IMPORTANT: Keep group means####
+  ## zero the cols related to other splines
+ # X[, ! (c1 | c2)] <- 0
+
+  ## zero out the parametric cols
+  #X[, !grepl('^s\\(', cnames)] <- 0
+
+  ## compute difference
+  sm_diff <- drop(X %*% coefs)
+  se <- sqrt(rowSums((X %*% V) * X))
+  nr <- NROW(X)
+
+  ## Calculate posterior simulation for smooths
+  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
+  rownames(coefs_sim) <- rownames(V)
+  simDev <- X %*% coefs_sim
+  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
+  masd <- apply(absDev, 2, max)
+  crit_s <- quantile(masd, prob = 0.95, type = 8)
+
+  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
+              level_1 = rep(f1, nr),
+              level_2 = rep(f2, nr),
+              diff = sm_diff, se = se,
+              lower_s = sm_diff - crit_s * se,
+              upper_s = sm_diff + crit_s*se)
+
+  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
+  ## Only need rows associated with one of the levels
+  out <- bind_cols(out, data[r1, smooth_var])
+
+  out
+}
+
+#does both ci and si
+difference_smooths <- function(model,
+                                 smooth,
+                                 n = 100,
+                                 ci_level = 0.95,
+                                 newdata = NULL,
+                                 partial_match = TRUE,
+                                 unconditional = FALSE,
+                                 frequentist = FALSE,
+                                 nrep = 10000,
+                                 include_means = TRUE,
+                                 ...) {
+  if (missing(smooth)) {
+    stop("Must specify a smooth to difference via 'smooth'.")
+  }
+
+  # smooths in model
+  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
+  # select smooths
+  select <-
+    gratia:::check_user_select_smooths(smooths = S, select = smooth,
+                                       partial_match = partial_match)#,
+  # model_name = expr_label(substitute(object)))
+  sm_ids <- which(select)
+  smooths <- gratia::get_smooths_by_id(model, sm_ids)
+  sm_data <- map(sm_ids, gratia:::smooth_data,
+                 model = model, n = n, include_all = TRUE)
+  sm_data <- bind_rows(sm_data)
+  by_var <- by_variable(smooths[[1L]])
+  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
+  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
+                                   stringsAsFactor = FALSE))
+  names(pairs) <- paste0("f", 1:2)
+
+  Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
+  V <- gratia:::get_vcov(model, unconditional = unconditional,
+                         frequentist = frequentist)
+  coefs <- coef(model)
+
+  out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
+              smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
+              coefs = coefs, nrep = nrep)
+  out <- bind_rows(out)
+  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
+
+  out <- add_column(out,
+                    lower = out$diff - (crit * out$se),
+                    upper = out$diff + (crit * out$se),
+                    .after = 6L)
+  out
+}
+
+#compute difference between smooths and calculate confidence interval: complete data
+diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
+                     unconditional = TRUE, frequentist = FALSE,
+                     n=100, partial_match = TRUE, nrep=10000)
+
+#function to obtain values for the shading regions
+pairwise_limits<-function(dataframe){
+    #extract values where the lower limit of the ribbon is greater than zero
+    #this is the region where the control group effect is greater
+
+    v1<-dataframe%>%
+        filter(lower_s>0)%>%
+        select(Day)
+    #get day  initial value
+    init1=v1$Day[[1]]
+    #get day final value
+    final1=v1$Day[[nrow(v1)]]
+    #extract values where the value of the upper limit of the ribbon is lower than zero
+    #this corresponds to the region where the treatment group effect is greater
+    v2<-dataframe%>%
+        filter(upper_s<0)%>%
+        select(Day)
+
+    init2=v2$Day[[1]]
+    final2=v2$Day[[nrow(v2)]]
+    #store values
+    my_list<-list(init1=init1,
+                  final1=final1,
+                  init2=init2,
+                  final2=final2)
+    return(my_list)
+
+}
+
+my_list<-pairwise_limits(diff_complete)
 rib_col<-'#8D7D82' #color for ribbon for confidence interval
@@
 
-c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
-  #shaded region
+c1 <- ggplot() +
+  geom_line(data = diff_complete, aes(x = Day, y = diff),size=1, alpha=0.5) +
     annotate("rect",
@@
                 alpha = 0.5,
-                ) +
+                )+
     annotate("text",
              x=1.5,
-             y=-10,
-             label="Control",size=10
+             y=-18,
+             label="Control>Treatment",
+           size=6,
+           angle=90
            )+
-  #shaded region  
     annotate("rect",
              xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill= treat_rib,
+             fill=treat_rib,
              alpha = 0.5
@@
              x=6,
-             y=-10,
-             label="Treatment",
-             size=10
+             y=-18,
+             label="Treatment>Control",
+             size=6,
+           angle=90
            )+
-  #ribbon for difference confidence interval  
-  geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(color='black',size=1) +
-    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
-    facet_wrap(~ pair) +
+  geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s, ymax = upper_s),
+              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
+  geom_hline(yintercept = 0, lty = 2, color = "red")+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    labs(y="Difference\n(Complete observations)")+
     theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
     theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
+    axis.text=element_text(size=22))
 \end{lstlisting}
@@
 
-\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!gam\_02!} using a 95\% confidence interval for the difference between smooths.
-The comparison is centered at the response scale.
+\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!gam\_02!} using a 95\% simultaneous CI for the difference between smooths.
+The comparison includes the group means and therefore can be directly correlated with the magnitude of the response.
 Shaded regions indicate time intervals where each treatment group has a non-zero effect.}\label{fig:pairwise-comp-workflow-fig}
@@
 
-Of notice, a convenient wrapper for the function described above exists in the package \passthrough{\lstinline!gratia!}.
+Of notice, the package \passthrough{\lstinline!gratia!} contains a function that allows to do pairwise comparisons.
 In this package, \passthrough{\lstinline!difference\_smooths!} is a function that makes the comparisons and produces Figure \ref{fig:pairwise-comp-workflow-fig} when is used on a fitted model.
@@
 
-Keep in mind that this function \textbf{does not} center the pairwise comparison at the response scale, so it has to be shifted in order to be compared to the raw data.
+Keep in mind that this function removes the group means for the pairwise comparison so the result will not be directly interpretable with the scale of the measured response.
 
@@
 
-\hypertarget{gam-and-linear-model-plots}{%
-\subsubsection{GAM and Linear model plots}\label{gam-and-linear-model-plots}}
-
-This code chunk creates panels B and D in Figure \ref{fig:sim-smooth-plot}.
+\hypertarget{GAM-linear-plot}{%
+\subsubsection{GAM and Linear model plots}\label{GAM-linear-plot}}
+
+This code chunk creates panels B and C in Figure \ref{fig:sim-smooth-plot}.
 Note that this code uses the final GAM from the previous section (\passthrough{\lstinline!gam\_02!}), so the simulated data and the model should be generated before running this section.
@@
 
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=gam_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
+#plot smooths and pointwise and simulatenous confidence intervals for GAM
+f3<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
+geom_line(lwd = 1) +
+geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
+inherit.aes = FALSE, alpha = 0.7,
+show.legend=FALSE) +
+geom_ribbon(data = si,
+mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
+inherit.aes = FALSE, alpha = 0.4,
+show.legend=FALSE)+
+    geom_point(data=dat_sim, aes(x = Day, 
+                        y = StO2_sim, 
+                        color = Group), 
+                        size=1.5,
+                        alpha=0.6, 
+               inherit.aes = FALSE,
                show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'complete')))+
+    labs(y=expression(atop(StO[2],'(complete observations)')))+
     scale_x_continuous(breaks=c(0,2,5,7,10))+
@@
     axis.text=element_text(size=22)
-  )+
-  thm1
+  )+thm1
 
@@
                    ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
+              alpha=al,
               data=lm_predict,
@@
     axis.text=element_text(size=22)
-  )+
-  thm1
+  )+thm1
 \end{lstlisting}
@@
 #correspond to the row numbers to be randomly erased from the original dataset
-
 missing <- sample(1:100, 40)
@@
 
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
-
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'missing')))+
+ci <- confint(mod_m1, parm = "s(Day)", partial_match = TRUE, type = "confidence")
+## simultaneous interval
+si <- confint(mod_m1, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
+
+# mean shift for group 2
+const <- coef(mod_m1)[2]
+
+ci <- ci %>%
+mutate(est = case_when(Group == "Treatment" ~ est + const,
+TRUE ~ est),
+lower = case_when(Group == "Treatment" ~ lower + const,
+TRUE ~ lower),
+upper = case_when(Group == "Treatment" ~ upper + const,
+TRUE ~ upper))
+
+si <- si %>%
+mutate(est = case_when(Group == "Treatment" ~ est + const,
+TRUE ~ est),
+lower = case_when(Group == "Treatment" ~ lower + const,
+TRUE ~ lower),
+upper = case_when(Group == "Treatment" ~ upper + const,
+TRUE ~ upper))
+
+#plot model and original data
+f6<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
+geom_line(lwd = 1) +
+geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
+inherit.aes = FALSE, alpha = 0.7,
+show.legend=FALSE) +
+geom_ribbon(data = si,
+mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
+inherit.aes = FALSE, alpha = 0.4,
+show.legend=TRUE)+
+    geom_point(data=dat_missing, aes(x = Day, 
+                        y = StO2_sim, 
+                        color = Group), 
+                            size=1.5,
+                        alpha=0.6, 
+               inherit.aes = FALSE)+
+    labs(y=expression(atop(StO[2],'(missing observations)')))+
     scale_x_continuous(breaks=c(0,2,5,7,10))+
@@
   theme(
-    axis.text=element_text(size=22)
-  )+
+    axis.text=element_text(size=22))+
         thm1
@@
 \caption{Simulated data and smooths for oxygen saturation in tumors.
-A: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
+\textbf{A}: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
 Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% confidence intervals.
-C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-D: Smooths for the GAM model for the simulated data with 40\% of its observations missing.
-Lines represent trends for each group, shaded regions are 95\% empirical Bayesian confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
+\textbf{B}: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
+Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.
+\textbf{C}: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
+\textbf{D}: Smooths for the GAM model for the simulated data with missing observations (40\%).
+Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
 \end{figure}
@@
 
-The next code chunk reproduces Figure \ref{fig:plot-pairwise-comp}. Here pairwise comparisons are made for the full and missing datasets.
+The next code chunk reproduces Figure \ref{fig:plot-pairwise-comp}.
+Here pairwise comparisons are made for the full and missing datasets by computing the difference between the smooths and a 95\% simultaneous CI.
 
@@
 
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-#this function takes the model, grid and groups to be compared using the lpmatrix
-#originally developed by G. Simpson:
-#https://fromthebottomoftheheap.net/2017/10/10/difference-splines-i/
-
-smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    #Find columns in xp where the name contains "Control" and "Treatment"
-    col1 <- grepl(g1, colnames(xp))
-    col2 <- grepl(g2, colnames(xp))
-    #Find rows in xp that correspond to each treatment
-    row1 <- with(newdata, Group == g1)
-    row2 <- with(newdata, Group == g2)
-    ## difference rows of xp for data from comparison
-    X <- xp[row1, ] - xp[row2, ]
-    ## zero out cols of X not involved in the comparison
-    X[, ! (col1 | col2)] <- 0
+difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
+  ## make sure f1 and f2 are characters
+  f1 <-  as.character(f1)
+  f2 <-  as.character(f2)
+  cnames <- colnames(Xp)
+  ## columns of Xp associated with pair of smooths
+  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
+  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
+  ## rows of Xp associated with pair of smooths
+  r1 <- data[[by_var]] == f1
+  r2 <- data[[by_var]] == f2
+
+  ## difference rows of Xp for pair of smooths
+  X <- Xp[r1, ] - Xp[r2, ]
+
+  ## zero the cols related to other splines
+ # X[, ! (c1 | c2)] <- 0
 
   ## zero out the parametric cols
-    #This line has been commented to keep the comparison at the response level,
-    #otherwise it gives the marginal change between smooths
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    #get standard error, critical value and boundaries
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(g1, g2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-}
-
-#use the function to calculate the difference in smooths
-comp1<-smooth_diff(gam_02,pdat,'Control','Treatment')
-
-#Create a dataframe with time, comparisons and labels for regions where difference exists
-comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
-
+  #X[, !grepl('^s\\(', cnames)] <- 0
+
+  ## compute difference
+  sm_diff <- drop(X %*% coefs)
+  se <- sqrt(rowSums((X %*% V) * X))
+  nr <- NROW(X)
+
+  ## Calculate posterior simulation for smooths
+  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
+  rownames(coefs_sim) <- rownames(V)
+  simDev <- X %*% coefs_sim
+  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
+  masd <- apply(absDev, 2, max)
+  crit_s <- quantile(masd, prob = 0.95, type = 8)
+
+  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
+              level_1 = rep(f1, nr),
+              level_2 = rep(f2, nr),
+              diff = sm_diff, se = se,
+              lower_s = sm_diff - crit_s * se,
+              upper_s = sm_diff + crit_s*se)
+
+  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
+  ## Only need rows associated with one of the levels
+  out <- bind_cols(out, data[r1, smooth_var])
+
+  out
+}
+
+#does both ci and si
+difference_smooths <- function(model,
+                                 smooth,
+                                 n = 100,
+                                 ci_level = 0.95,
+                                 newdata = NULL,
+                                 partial_match = TRUE,
+                                 unconditional = FALSE,
+                                 frequentist = FALSE,
+                                 nrep = 10000,
+                                 include_means = TRUE,
+                                 ...) {
+  if (missing(smooth)) {
+    stop("Must specify a smooth to difference via 'smooth'.")
+  }
+
+  # smooths in model
+  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
+  # select smooths
+  select <-
+    gratia:::check_user_select_smooths(smooths = S, select = smooth,
+                                       partial_match = partial_match)#,
+  # model_name = expr_label(substitute(object)))
+  sm_ids <- which(select)
+  smooths <- gratia::get_smooths_by_id(model, sm_ids)
+  sm_data <- map(sm_ids, gratia:::smooth_data,
+                 model = model, n = n, include_all = TRUE)
+  sm_data <- bind_rows(sm_data)
+  by_var <- by_variable(smooths[[1L]])
+  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
+  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
+                                   stringsAsFactor = FALSE))
+  names(pairs) <- paste0("f", 1:2)
+
+  Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
+  V <- gratia:::get_vcov(model, unconditional = unconditional,
+                         frequentist = frequentist)
+  coefs <- coef(model)
+
+  out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
+              smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
+              coefs = coefs, nrep = nrep)
+  out <- bind_rows(out)
+  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
+
+  out <- add_column(out,
+                    lower = out$diff - (crit * out$se),
+                    upper = out$diff + (crit * out$se),
+                    .after = 6L)
+  out
+}
+
+#compute difference between smooths and calculate confidence interval: complete data
+diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
+                     unconditional = TRUE, frequentist = FALSE,
+                     n=100, partial_match = TRUE, nrep=10000)
+
+#function to obtain values for the shading regions
 pairwise_limits<-function(dataframe){
@@
     return(my_list)
-}
-
-my_list<-pairwise_limits(comp_StO2_full)
+
+}
+
+my_list<-pairwise_limits(diff_complete)
 rib_col<-'#8D7D82' #color for ribbon for confidence interval
@@
              xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
+             fill=treat_rib,
+             alpha = 0.5
+             ) +
+  annotate("text",
+             x=6,
+             y=-18,
+             label="Treatment>Control",
+             size=6,
+           angle=90
+           )+
+  geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s, ymax = upper_s),
+              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
+  geom_hline(yintercept = 0, lty = 2, color = "red")+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    labs(y="Difference\n(Complete observations)")+
+    theme_classic()+
+    theme(
+    axis.text=element_text(size=22))
+
+###compute difference in smooths for missing data
+diff_missing <- difference_smooths(mod_m1, smooth = "s(Day)", newdata = newdat,
+                     unconditional = TRUE, frequentist = FALSE,
+                     n=100, partial_match = TRUE, nrep=10000)
+
+v2<- diff_missing %>%
+        filter(upper_s<0)%>%
+        select(Day)
+init2=v2$Day[[1]]
+final2=v2$Day[[nrow(v2)]]
+
+c2 <- ggplot() +
+  geom_line(data = diff_missing, aes(x = Day, y = diff),size=1, alpha=0.5) +
+    annotate("rect",
+             xmin =init2, xmax = final2,ymin=-Inf,ymax=Inf,
              fill=treat_rib,
@@
              label="Treatment>Control",
-             size=8,
+             size=6,
            angle=90
            )+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
+  geom_ribbon(data = diff_missing, aes(x = Day, ymin = lower_s, ymax = upper_s),
+              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
+  geom_hline(yintercept = 0, lty = 2, color = "red")+
     scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-###for missing data
-comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
-comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp2))
-
-missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    labs(x = 'Days', 
-         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
-                              )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
+    labs(y="Difference\n(Missing observations)")+
     theme_classic()+
     theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
-my_list<-pairwise_limits(comp_StO2_missing)
-
-c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    annotate("rect",
-             xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-             fill=control_rib,
-             alpha = 0.5
-             ) +
-  annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=8,
-           angle=90
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill= treat_rib,
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=8,
-           angle=90
-           )+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
+    axis.text=element_text(size=22))
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/plot-pairwise-comp-Appendix-1}
+{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/plot-pairwise-comp-Appendix-1}
 
@@
 \caption{Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-B: Pairwise comparisons for the dataset with missing observations.
+\textbf{A}: Pairwise comparisons for the full dataset.
+\textbf{B}: Pairwise comparisons for the dataset with missing observations.
 Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
-In both cases the effect of treatment is significant after day 3.}\label{fig:plot-pairwise-comp-Appendix}
+In both cases the effect of treatment is significant after day 3.
+In both cases, a simultaneous CI has been computed around the smooth difference.}\label{fig:plot-pairwise-comp-Appendix}
 \end{figure}

### 2021-12-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Started code for simultaneous interval
Updated reviewer's comments, Appendix, Main manuscript

@@
 Top row: linear response, bottom row: quadratic response.
-A: Simulated linear data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
-D: Simulated quadratic data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
-B,E: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
+**A**: Simulated linear data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
+**D**: Simulated quadratic data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
+**B,E**: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
 Points represent the original raw data.
 The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation.
-C, F: Estimates from the LMEM in the linear and quadratic case (subject: thin lines, population: thick lines) .
+**C, F**: Estimates from the LMEM in the linear and quadratic case (subject: thin lines, population: thick lines) .
 The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data and grossly bias the initial estimates for each group in the quadratic case (bottom row).

### 2021-12-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated reviewer's comments, Appendix, Main manuscript

@@
 (ref:basis-plot-caption) Basis functions for a single smoother for time.
-A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-B: Matrix of basis function weights.
+**A**: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
+**B**: Matrix of basis function weights.
 Each basis function is multiplied by a coefficient which can be positive or negative.
 The coefficient determines the overall effect of each basis in the final smoother.
-C: Weighted basis functions.
+**C**: Weighted basis functions.
 Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
 Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-D: Smoother for time and original data points.
+**D**: Smoother for time and original data points.
 The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.
@@
 Because random samples from the population are used to calculate the "true" value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95%) where the function is expected to lie.
-In an _across the function_ CI (like those estimated for GAMs using _mgcv_) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95%).
+In an _across the function_ CI (like those estimated by default for GAMs using _mgcv_) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95%).
 In other words, we expect that about 95% of the points that compose the true function will be covered by the across the function CI and for this to occur, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
 
-**add about simultaneous interval**
-In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in Miller [@miller2019], Wood[@wood2017], Simpson [@simpson2018] and Marra [@marra2012].
-With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as "empirical Bayesian" through the rest of this paper.
+However, besides the across the function CI, "simultaneous" or "whole function" CIs can also be computed, which contain the _whole function_ with a specified probability [@wood2017].
+Suppose we chose a nominal value (say, 95%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
+Briefly, to obtain a simultaneous CI we use simulation to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI [@ruppert2003].
+Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups, as we discuss in the following section.
+
+In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller [@miller2019], Wood[@wood2017], Simpson [@simpson2018], Marra [@marra2012], and Ruppert [@ruppert2003].
+With this brief introduction to the Bayesian interpretation of GAMs, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends.

### 2021-12-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated reviewer's comments, Appendix, Main manuscript

@@
 Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (matching the SD in the original paper).
-The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot),A.
+The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot)A.
 
@@
 
-For the data in Figure \@ref(fig:sim-smooth-plot), A the main effect of interest is how $\mbox{StO}_2$ changes over time for each treatment.
+For the data in Figure \@ref(fig:sim-smooth-plot)A the main effect of interest is how $\mbox{StO}_2$ changes over time for each treatment.
 To estimate this, the model incorporates separate smooths for _Group_ and _Day_, respectively.
@@
 Consider the  simulated $\mbox{StO}_2$ values from Figure \@ref(fig:sim-smooth-plot)B.
-If 40% of the observations are randomly deleted and the same interaction GAM fitted for the complete dataset is used, the resulting smooths are still able to show a different trend for each group, but because the empirical Bayesian credible intervals for the smooths overlap during the first 3 days with fewer data points, the trend is less pronounced than in the full dataset (Figure \@ref(fig:sim-smooth-plot)D).
-Although the confidence intervals have increased for both smooths, the model still shows different trends with as few as 4 observations per group at certain time points.
+If 40% of the observations are randomly deleted and the same interaction GAM (`gam_02`) is fitted, the resulting smooths appear in (Figure \@ref(fig:sim-smooth-plot)D).
+Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.
+
+Additionally, note that in Figures \@ref(fig:sim-smooth-plot)B,D we show two CIs for each of the fitted smooths (shaded regions).
+The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, we expect them to be wider than the across the function CI, resulting in the wide shaded regions.
+For the dataset with missing observations, the smooths CIs overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset.
+However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.
 
@@
 **B**: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95% across the function (inner region) and simultaneous (outer region) confidence intervals.
+Lines represent trends for each group, shaded regions are 95% across the function (narrow region) and simultaneous (wide region) confidence intervals.
 **C**: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
 **D**: Smooths for the GAM model for the simulated data with missing observations (40%).
-Lines represent trends for each group, shaded regions are 95%  across the function (inner region) and simultaneous (outer region) confidence intervals.
+Lines represent trends for each group, shaded regions are 95%  across the function (narrow region) and simultaneous (wide region) confidence intervals.
 
@@
 (ref:pairwise-comp) Pairwise comparisons for smooth terms.
-A: Pairwise comparisons for the full dataset.
-B: Pairwise comparisons for the dataset with missing observations.
+**A**: Pairwise comparisons for the full dataset.
+**B**: Pairwise comparisons for the dataset with missing observations.
 Significant differences exist where the 95% empirical Bayesian credible interval does not cover 0.
 In both cases the effect of treatment is significant after day 3.
+In both cases, a simultaneous CI has been computed around the smooth difference.
 
@@
 In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in `gam_02`) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
-**add sentence about the issue of the Nychka across the function intervals **
 
-around that  _difference_ between the empirical Bayesian confidence intervals of the fitted smooths for such groups is  non-zero, then a significant difference exists at that time point(s).
-The absence of a _p-value_ in this case might seem odd, but the empirical Bayesian confidence interval interpretation can be conceptualized in the following manner: Different trends in each group are an indication of an effect by the treatment.
+The absence of a _p-value_ in this context might seem odd, but its interpretation can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment.
 This is what happens for the simulated data in Figure \@ref(fig:sim-smooth-plot)A, where the chemotherapy causes $\mbox{StO}_2$ to increase over time.
@@
 With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant.
-The difference between groups with similar trends is likely to yield zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
+A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
 
 Consider the calculation of pairwise differences for the smooths in Figure \@ref(fig:sim-smooth-plot)B and Figure \@ref(fig:sim-smooth-plot)D.
-Figure \@ref(fig:plot-pairwise-comp) shows the comparison between each treatment group for the full and missing datasets.
+Figure \@ref(fig:plot-pairwise-comp) shows the comparison between each treatment group for the full and missing datasets with a simultaneous CI computed around the difference.
 Here, the  "Control" group is used as the reference to which "Treatment" group is being compared.
-Of notice, the pairwise comparison has been set on the response scale (see Appendix for code details), because otherwise the comparison appears shifted and is not intuitively easy to relate to the original data.
+Notice that because we have included the means of each group, there is correspondence between the scale of the original data the scale of the pairwise comparisons.
+This can be seen as in Figure \@ref(fig:sim-smooth-plot)B at day 5 there is essentially a 50% difference between $\mbox{StO}_2$ in both groups, which corresponds to the -50 difference in \@ref(fig:plot-pairwise-comp)A.
+However, if there are multiple parametric terms in the model (more factors that need to be specified in the model) such inclusion of the means can become problematic.
+However, we believe that the model we have presented here suffices in a wide range of situations where adding the group means is a relatively easy implementation that can help better visualize the estimation from the model from a biological perspective.
 
-With this correction in mind, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
+In Figure \@ref(fig:plot-pairwise-comp)A the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
 Notice that the shaded region between days 0 and $\approx$ 2 for the full dataset indicates that through that time, the "Control" group has higher mean $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by $\approx$ 3 day  it is the "Treatment" group the one that on average, has greater $\mbox{StO}_2$.
@@
 
-On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the empirical Bayesian credible intervals of the smooths show the same trend of the full dataset.
-Consequently, the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that the Control Group has higher $\mbox{StO}_2$ before day 3, and is able to estimate the change on day 3 where the Treatment Group becomes significant as the full dataset smooth pairwise comparison.
+On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that because the confidence intervals overlap for the first two days there is essentially no difference between the groups.
+However, because the model was still able to pick the overall trend in $\mbox{StO}_2$, the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes significant as in the full dataset smooth pairwise comparison.
 

### 2021-12-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/07-Appendix.Rmd
Updated reviewer's comments, Appendix, Main manuscript

@@
 
+<<testing>>
 ## Example with linear response
@@
 Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
+For simplicity, the confidence intervals (CIs) shown in this section for the models are the across the function CIs created by `mgcv` by default.
+However, for the pairwise comparisons of the third model we use _simultaneous intervals_ as in the main manuscript, and the code for creating simultaneous CIs for the smooths can be found in Section \@ref(GAM-linear-plot).
 
@@
 The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \@ref(GAM-significance).
+
 In this case, the "design matrix" is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
@@
 This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
-**All pairwise comparisons in this paper have been centered at the response scale to ease interpretation **.
+
+We want to emphasize that for the model used here we have included the group means in order to keep the pairwise comparisons on the scale of the response (a change from the code which appears in the chunk with a comment that says `###IMPORTANT`).
+This approach works well for models like `gam_02` but implementing this on models with more parametric terms can be challenging.
+However, we do believe that the model presented in the paper covers a wide range of situations and our approach here for the pairwise comparisons will be useful for most biomedical researchers.
 
@@
              x=6,
-             y=-10,
-             label="Treatment",
-             size=10
+             y=-18,
+             label="Treatment>Control",
+             size=6,
+           angle=90
            )+
-  #ribbon for difference confidence interval  
-  geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(color='black',size=1) +
-    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
-    facet_wrap(~ pair) +
+  geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s, ymax = upper_s),
+              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
+  geom_hline(yintercept = 0, lty = 2, color = "red")+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    labs(y="Difference\n(Complete observations)")+
     theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
     theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-```
-
-(ref:pairwise-comp-caption) Smooth pairwise comparisons for model `gam_02` using a 95% confidence interval for the difference between smooths.
-The comparison is centered at the response scale.
+    axis.text=element_text(size=22))
+
+```
+
+(ref:pairwise-comp-caption) Smooth pairwise comparisons for model `gam_02` using a 95% simultaneous CI for the difference between smooths.
+The comparison includes the group means and therefore can be directly correlated with the magnitude of the response.
 Shaded regions indicate time intervals where each treatment group has a non-zero effect.
@@
 
-Of notice, a convenient wrapper for the function described above exists in the package `gratia`.
+Of notice, the package `gratia` contains a function that allows to do pairwise comparisons.
 In this package, `difference_smooths` is a function that makes the comparisons and produces Figure \@ref(fig:pairwise-comp-workflow-fig) when is used on a fitted model.
@@
 
-Keep in mind that this function **does not** center the pairwise comparison at the response scale, so it has to be shifted in order to be compared to the raw data.
+Keep in mind that this function removes the group means for the pairwise comparison so the result will not be directly interpretable with the scale of the measured response.
 
@@
 
-### GAM and Linear model plots
-
-This code chunk creates panels B and D in Figure \@ref(fig:sim-smooth-plot).
+### GAM and Linear model plots{#GAM-linear-plot}
+
+This code chunk creates panels B and C in Figure \@ref(fig:sim-smooth-plot).
 Note that this code uses the final GAM from the previous section (`gam_02`), so the simulated data and the model should be generated before running this section.
@@
 
-The next code chunk reproduces Figure \@ref(fig:plot-pairwise-comp). Here pairwise comparisons are made for the full and missing datasets.
-
-```{r,pairwise-comp-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE, eval=FALSE}
+The next code chunk reproduces Figure \@ref(fig:plot-pairwise-comp).
+Here pairwise comparisons are made for the full and missing datasets by computing the difference between the smooths and a 95% simultaneous CI.
+
+```{r,pairwise-comp-Appendix,include=TRUE,message=FALSE,warning=FALSE, eval=FALSE}
 ##Pairwise comparisons
 
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-#this function takes the model, grid and groups to be compared using the lpmatrix
-#originally developed by G. Simpson:
-#https://fromthebottomoftheheap.net/2017/10/10/difference-splines-i/
-
-smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    #Find columns in xp where the name contains "Control" and "Treatment"
-    col1 <- grepl(g1, colnames(xp))
-    col2 <- grepl(g2, colnames(xp))
-    #Find rows in xp that correspond to each treatment
-    row1 <- with(newdata, Group == g1)
-    row2 <- with(newdata, Group == g2)
-    ## difference rows of xp for data from comparison
-    X <- xp[row1, ] - xp[row2, ]
-    ## zero out cols of X not involved in the comparison
-    X[, ! (col1 | col2)] <- 0
+difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
+  ## make sure f1 and f2 are characters
+  f1 <-  as.character(f1)
+  f2 <-  as.character(f2)
+  cnames <- colnames(Xp)
+  ## columns of Xp associated with pair of smooths
+  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
+  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
+  ## rows of Xp associated with pair of smooths
+  r1 <- data[[by_var]] == f1
+  r2 <- data[[by_var]] == f2
+
+  ## difference rows of Xp for pair of smooths
+  X <- Xp[r1, ] - Xp[r2, ]
+
+  ## zero the cols related to other splines
+ # X[, ! (c1 | c2)] <- 0
 
   ## zero out the parametric cols
-    #This line has been commented to keep the comparison at the response level,
-    #otherwise it gives the marginal change between smooths
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    #get standard error, critical value and boundaries
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(g1, g2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-}
-
-#use the function to calculate the difference in smooths
-comp1<-smooth_diff(gam_02,pdat,'Control','Treatment')
-
-#Create a dataframe with time, comparisons and labels for regions where difference exists
-comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
-
+  #X[, !grepl('^s\\(', cnames)] <- 0
+
+  ## compute difference
+  sm_diff <- drop(X %*% coefs)
+  se <- sqrt(rowSums((X %*% V) * X))
+  nr <- NROW(X)
+
+  ## Calculate posterior simulation for smooths
+  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
+  rownames(coefs_sim) <- rownames(V)
+  simDev <- X %*% coefs_sim
+  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
+  masd <- apply(absDev, 2, max)
+  crit_s <- quantile(masd, prob = 0.95, type = 8)
+
+  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
+              level_1 = rep(f1, nr),
+              level_2 = rep(f2, nr),
+              diff = sm_diff, se = se,
+              lower_s = sm_diff - crit_s * se,
+              upper_s = sm_diff + crit_s*se)
+
+  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
+  ## Only need rows associated with one of the levels
+  out <- bind_cols(out, data[r1, smooth_var])
+
+  out
+}
+
+#does both ci and si
+difference_smooths <- function(model,
+                                 smooth,
+                                 n = 100,
+                                 ci_level = 0.95,
+                                 newdata = NULL,
+                                 partial_match = TRUE,
+                                 unconditional = FALSE,
+                                 frequentist = FALSE,
+                                 nrep = 10000,
+                                 include_means = TRUE,
+                                 ...) {
+  if (missing(smooth)) {
+    stop("Must specify a smooth to difference via 'smooth'.")
+  }
+
+  # smooths in model
+  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
+  # select smooths
+  select <-
+    gratia:::check_user_select_smooths(smooths = S, select = smooth,
+                                       partial_match = partial_match)#,
+  # model_name = expr_label(substitute(object)))
+  sm_ids <- which(select)
+  smooths <- gratia::get_smooths_by_id(model, sm_ids)
+  sm_data <- map(sm_ids, gratia:::smooth_data,
+                 model = model, n = n, include_all = TRUE)
+  sm_data <- bind_rows(sm_data)
+  by_var <- by_variable(smooths[[1L]])
+  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
+  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
+                                   stringsAsFactor = FALSE))
+  names(pairs) <- paste0("f", 1:2)
+
+  Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
+  V <- gratia:::get_vcov(model, unconditional = unconditional,
+                         frequentist = frequentist)
+  coefs <- coef(model)
+
+  out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
+              smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
+              coefs = coefs, nrep = nrep)
+  out <- bind_rows(out)
+  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
+
+  out <- add_column(out,
+                    lower = out$diff - (crit * out$se),
+                    upper = out$diff + (crit * out$se),
+                    .after = 6L)
+  out
+}
+
+#compute difference between smooths and calculate confidence interval: complete data
+diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
+                     unconditional = TRUE, frequentist = FALSE,
+                     n=100, partial_match = TRUE, nrep=10000)
+
+#function to obtain values for the shading regions
 pairwise_limits<-function(dataframe){

### 2021-12-16 Manuscripts/Manuscript_by_chapters-SIM_Revisions/SIM_Appendix.tex
Updated ribbon limits on GAM smooth models for better visualization

@@
 mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
-inherit.aes = FALSE, alpha = 0.4,
+inherit.aes = FALSE, alpha = 0.3,
 show.legend=FALSE)+
@@
                show.legend = FALSE)+
+    geom_line(data=si,aes(Day,upper,color=Group), size=0.8, alpha=0.7)+
+    geom_line(data=si,aes(Day,lower,color=Group), size=0.8, alpha=0.7)+
     labs(y=expression(atop(StO[2],'(complete observations)')))+
@@
 show.legend=TRUE)+
+    geom_line(data=si,aes(Day,upper,color=Group), size=0.8, alpha=0.7)+
+    geom_line(data=si,aes(Day,lower,color=Group), size=0.8, alpha=0.7)+
     geom_point(data=dat_missing, aes(x = Day, 

### 2021-12-21 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Checked Bibtex entries

@@
 ]{article}
-\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}}
-\date{\vspace{-2.5em}}
-
 \usepackage{amsmath,amssymb}
@@
 
+\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}}
+\date{\vspace{-2.5em}}
+
 \begin{document}
@@
 Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response,\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}} antibody expression,\textsuperscript{\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}} and cell metabolism\textsuperscript{\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals\textsuperscript{\protect\hyperlink{ref-greening2018}{9}}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients,\textsuperscript{\protect\hyperlink{ref-sio2016}{10}} to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients.\textsuperscript{\protect\hyperlink{ref-kamstra2015}{11}}
+Tumor response\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}}, antibody expression\textsuperscript{\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}}, and cell metabolism\textsuperscript{\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
+Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals\textsuperscript{\protect\hyperlink{ref-greening2018}{9}}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients\textsuperscript{\protect\hyperlink{ref-sio2016}{10}}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients\textsuperscript{\protect\hyperlink{ref-kamstra2015}{11}}.
 
 Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times,\textsuperscript{\protect\hyperlink{ref-wagenmakers2008}{12}} and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}).\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}}
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
+The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times\textsuperscript{\protect\hyperlink{ref-wagenmakers2008}{12}}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
+The rm-ANOVA model makes three assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations})\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}}.
+
+The expected linear behavior of the response through time is a key requisite in rm-ANOVA\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
+This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
 In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings.\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}} These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}}.
+These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
 Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
@@
 In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}}.
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid.\textsuperscript{\protect\hyperlink{ref-wang2019}{18}}
-
-Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)),\textsuperscript{\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}} consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni),\textsuperscript{\protect\hyperlink{ref-abdi2010}{21}} but they in turn reduce statistical power (1-\(\beta\))\textsuperscript{\protect\hyperlink{ref-nakagawa2004}{22}}, and lead to increased Type II error (failing to reject the null hypothesis when it is false).\textsuperscript{\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}} Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases.\textsuperscript{\protect\hyperlink{ref-ugrinowitsch2004}{25}} Corrections can be made in rm-ANOVA in the absence of compound symmetry,\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}} but the effectiveness of the correction is limited by the size of the sample, the number of measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and group sizes.\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}} In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons\textsuperscript{\protect\hyperlink{ref-charan2013}{30}} which might cause the corrections for lack of compound symmetry to be ineffective.
+For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid\textsuperscript{\protect\hyperlink{ref-wang2019}{18}}.
+
+Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\))\textsuperscript{\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}}, consequently biasing the conclusions of the study.
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni\textsuperscript{\protect\hyperlink{ref-abdi2010}{21}}), but they in turn reduce statistical power (1-\(\beta\))\textsuperscript{\protect\hyperlink{ref-nakagawa2004}{22}}, and lead to increased Type II error (failing to reject the null hypothesis when it is false)\textsuperscript{\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}}.
+Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
+
+On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases\textsuperscript{\protect\hyperlink{ref-ugrinowitsch2004}{25}}.
+Corrections can be made in rm-ANOVA in the absence of compound symmetry\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and group sizes\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}}.
+In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons\textsuperscript{\protect\hyperlink{ref-charan2013}{30}} which might cause the corrections for lack of compound symmetry to be ineffective.
 
@@
 In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13}} This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
+When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13}}.
+This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
 At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
@@
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data.\textsuperscript{\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}} Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}} And even more importantly, LMEMs also assume by default a linear relationship between the response and time\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .
+The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data\textsuperscript{\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}}.
+Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
+However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}}.
+And even more importantly, LMEMs also assume by default a linear relationship between the response and time\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .
 
@@
 Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data,\textsuperscript{\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}} health-environment interactions\textsuperscript{\protect\hyperlink{ref-yang2012}{35}} and the dynamics of government in political science.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements,\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and 3) can easily accommodate missing observations.
+Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data\textsuperscript{\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}}, health-environment interactions\textsuperscript{\protect\hyperlink{ref-yang2012}{35}} and the dynamics of government in political science\textsuperscript{\protect\hyperlink{ref-beck1998}{36}}.
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, and 3) can easily accommodate missing observations.
 Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
@@
 In particular, R\textsuperscript{\protect\hyperlink{ref-r}{38}} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, R has many tools that simplify data simulation, an emerging strategy used to test statistical models.\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}} Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
+At the same time, R has many tools that simplify data simulation, an emerging strategy used to test statistical models\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}.
+Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
 
@@
 Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies.\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
+Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}}.
+The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
 Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in R, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
@@
 Independent errors \(\varepsilon_{ijt}\) represent random variation from the sampling process assumed to be \(\stackrel{i.i.d.}\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
-In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug,'' or ``saline'' vs.~``chemotherapy'').
+In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'', or ``saline'' vs.~``chemotherapy'').
 Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B).
@@
 
-Of note, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach (using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}), which is described in detail by Pinheiro and Bates.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates;\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} more importantly, polynomial effects lack biological or mechanistic interpretation\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} which limits their use in biomedical studies.
+Of note, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach (using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}), which is described in detail by Pinheiro and Bates\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
+However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates\textsuperscript{\protect\hyperlink{ref-beck1998}{36}}; more importantly, polynomial effects lack biological or mechanistic interpretation\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} which limits their use in biomedical studies.
 
@@
 In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time.\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}} However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time;\textsuperscript{\protect\hyperlink{ref-maxwell2017}{44}} and it is higher between consecutive measurements.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}} Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and they are not robust if the group sizes are unbalanced.\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}} Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
+Under this assumption, the between-subject variance and within-subject correlation are constant across time\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}}.
+However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time\textsuperscript{\protect\hyperlink{ref-maxwell2017}{44}}; and it is higher between consecutive measurements\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}}.
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and they are not robust if the group sizes are unbalanced\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}}.
+Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
+
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
+Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
 Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
@@
 In a longitudinal study, it is frequently the case that the number of observations is different across subjects.
-In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing,'' ``incomplete,'' or (more generally speaking) unbalanced data.\textsuperscript{\protect\hyperlink{ref-molenberghs2004}{45}} The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population and statistical power is affected by this reduction in sample size.\textsuperscript{\protect\hyperlink{ref-ma2012}{46}}
-
-On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are \emph{missing at random} (MAR) or \emph{completely missing at random} (MCAR).\textsuperscript{\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}} In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest.\textsuperscript{\protect\hyperlink{ref-scheffer2002}{47}} If the data are MCAR, this means that the missingness is completely unrelated to the collected information.\textsuperscript{\protect\hyperlink{ref-potthoff2006}{48}} Missing observations can also be \emph{missing not at random} (MNAR) and in the case the missing observations are dependent on their value.
+In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing'', ``incomplete'', or (more generally speaking) unbalanced data\textsuperscript{\protect\hyperlink{ref-molenberghs2004}{45}}.
+The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
+This is inconvenient because the remaining subjects might not accurately represent the population and statistical power is affected by this reduction in sample size\textsuperscript{\protect\hyperlink{ref-ma2012}{46}}.
+
+On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are \emph{missing at random} (MAR) or \emph{completely missing at random} (MCAR)\textsuperscript{\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}}.
+In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest\textsuperscript{\protect\hyperlink{ref-scheffer2002}{47}}.
+If the data are MCAR, this means that the missingness is completely unrelated to the collected information\textsuperscript{\protect\hyperlink{ref-potthoff2006}{48}}.
+Missing observations can also be \emph{missing not at random} (MNAR) and in the case the missing observations are dependent on their value.
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
@@
 To visually demonstrate the limitations of rm-ANOVA and LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using R\textsuperscript{\protect\hyperlink{ref-r}{38}} and the package \emph{nlme}.\textsuperscript{\protect\hyperlink{ref-nlme}{50}}
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using R\textsuperscript{\protect\hyperlink{ref-r}{38}} and the package \emph{nlme}\textsuperscript{\protect\hyperlink{ref-nlme}{50}}.
 
@@
 If the residuals are non-normal, a transformation is necessary in order to properly fit the model.
-However, transformation can lead to poor model performance,\textsuperscript{\protect\hyperlink{ref-ohara2010}{51}} and can cause problems with the biological interpretation of the model estimates.
+However, transformation can lead to poor model performance\textsuperscript{\protect\hyperlink{ref-ohara2010}{51}}, and can cause problems with the biological interpretation of the model estimates.
 McCullagh and Nelder\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} introduced General Linear Models (GLMs) as an extension of LMs, where the errors do not need to be normally distributed.
@@
 Although GLMs relax the normality assumption, they only accommodate fixed effects.
-Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate \emph{random effects}, which have an associated probability distribution.\textsuperscript{\protect\hyperlink{ref-mcculloch2001}{53}} Therefore, in GLMMs the linear predictor takes the form
+Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate \emph{random effects}, which have an associated probability distribution\textsuperscript{\protect\hyperlink{ref-mcculloch2001}{53}}.
+Therefore, in GLMMs the linear predictor takes the form
 
@@
 where \(\alpha_{ij}\) corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of Equation \eqref{eq:GLM-eta}.
-Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed,\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} and GLMs are a special case of GLMMs where there are no random effects.
-In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in Dobson\textsuperscript{\protect\hyperlink{ref-dobson2008}{54}} and Stroup.\textsuperscript{\protect\hyperlink{ref-stroup2013}{55}}
+Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}}, and GLMs are a special case of GLMMs where there are no random effects.
+In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in Dobson\textsuperscript{\protect\hyperlink{ref-dobson2008}{54}} and Stroup\textsuperscript{\protect\hyperlink{ref-stroup2013}{55}}.
 
@@
 Notice that in the previous sections, the difference between GLMs and GLMMs resides on their linear predictors (Equations \eqref{eq:GLM-eta}, \eqref{eq:GLMM-eta}).
-Generalized additive models (GAMs) are an extension of the GLM family that allow the estimation of smoothly varying trends where the relationship between the covariates and the response is modeled using \emph{smooth functions}.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}} In a GAM, the linear predictor has the form
+Generalized additive models (GAMs) are an extension of the GLM family that allow the estimation of smoothly varying trends where the relationship between the covariates and the response is modeled using \emph{smooth functions}\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}}.
+In a GAM, the linear predictor has the form
 
@@
 In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is more advantageous as it allows more flexibility because it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(time_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(time_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
+One possible set of functions for \(f(time_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference\textsuperscript{\protect\hyperlink{ref-beck1998}{36}}.
+In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(time_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
 
 The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using a \emph{basis function} expansion of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(treatment_j\), \(time_t\), and \(time_t \times treatment_j\).
-The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}}.
+For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(treatment_j\), \(time_t\), and \(time_t \times treatment_j\).
+The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
 In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(time_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{58}} Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness\textsuperscript{\protect\hyperlink{ref-wegman1983}{58}}.
+Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
 
 Cubic splines (CS) are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{56}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the basis functions meet) to obtain the finite basis, which could affect the model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate.\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}
-
-In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} Like TPS, TPRS only require the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).
+The use of CS as smoothers in GAMs was discussed within the original GAM framework\textsuperscript{\protect\hyperlink{ref-hastie1987}{56}}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots', the points where the basis functions meet) to obtain the finite basis, which could affect the model fit\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}.
+A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}.
+
+In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}.
+Like TPS, TPRS only require the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).
 
@@
 A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline.
-The second derivative of the spline is a measure of curvature, or the rate of change of the slope,\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} and increasing the penalty by increasing \(\lambda\) results in models with less curvature.
+The second derivative of the spline is a measure of curvature, or the rate of change of the slope\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}}, and increasing the penalty by increasing \(\lambda\) results in models with less curvature.
 As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}} In the case of data that shows non-linear trends, the prior that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Therefore, the confidence intervals (CIs) calculated by default for the smooth terms using \emph{mgcv} are considered empirical Bayesian credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good \emph{across the function} (``frequentist'') coverage\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}}.
+In the case of data that shows non-linear trends, the prior that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
+Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language)\textsuperscript{\protect\hyperlink{ref-miller2019}{61}}.
+Therefore, the confidence intervals (CIs) calculated by default for the smooth terms using \emph{mgcv} are considered empirical Bayesian credible intervals\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}}, which have good \emph{across the function} (``frequentist'') coverage\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
 
@@
 
-However, besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Suppose we chose a nominal value (say, 95\%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
-Briefly, to obtain a simultaneous CI we use simulation to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI.\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}} Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups, as we discuss in the following section.
-
-In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller,\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson,\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} Marra,\textsuperscript{\protect\hyperlink{ref-marra2012}{63}} and Ruppert.\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}} With this brief introduction to the Bayesian interpretation of GAMs, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends.
+However, besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
+Suppose we chose a nominal value (say, 95\%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
+Briefly, to obtain a simultaneous CI we use simulation to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}.
+Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups, as we discuss in the following section.
+
+In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller\textsuperscript{\protect\hyperlink{ref-miller2019}{61}}, Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}}, Marra\textsuperscript{\protect\hyperlink{ref-marra2012}{63}}, and Ruppert\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}.
+With this brief introduction to the Bayesian interpretation of GAMs, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends.
 
@@
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in Hefley.\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}}
+A detailed description on basis functions and correlations can be found in Hefley\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}}.
 
@@
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{64}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{65}}
-
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}\textsuperscript{\protect\hyperlink{ref-gratia}{64}}.
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak\textsuperscript{\protect\hyperlink{ref-harezlak2018}{65}}.
+
+One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?''.
+The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
 This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
@@
 Biomedical longitudinal data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
-Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}, where it is clear that these models do not capture the non-linear trend of the data, thereby causing a ``model misspecification error.''
+Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}, where it is clear that these models do not capture the non-linear trend of the data, thereby causing a ``model misspecification error''.
 
@@
 In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{67}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research\textsuperscript{\protect\hyperlink{ref-hastie1995}{67}}.
+This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).
 
@@
 We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{68},\protect\hyperlink{ref-weissgerber2018}{69}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research\textsuperscript{\protect\hyperlink{ref-begley2015}{68},\protect\hyperlink{ref-weissgerber2018}{69}}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 \CSLRightInline{Geisser S, Greenhouse SW.
-An extension of box{\textquotesingle}s results on the use of the {\textdollar}f{\textdollar} distribution in multivariate analysis.
+An extension of {B}ox{\textquotesingle}s results on the use of the {\emph{F}} distribution in multivariate analysis.
 \emph{The Annals of Mathematical Statistics}.

### 2021-12-22 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.Rmd
Made changes of revisions, need to re-structure Appendix

@@
 In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs).
-Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies assume an expected trend in the measured response, which can results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
+Although LMEMs are less restrictive than rm-ANOVA in terms of assumptions about imbalanced observations and correlation between observations, both methodologies assume a linear trend in the measured response.
+It is common in biomedical research that the true trend response is nonlinear and in these cases the linearity assumption of rm-ANOVA and LMEMs can lead to biased estimates and unreliable inference.
 
@@
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We present the basic theory of GAMs, and using reported trends of oxygen saturation in tumors we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R.
+We present the basic theory of GAMs and using reported trends of oxygen saturation in tumors, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R.
 We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing).

### 2021-12-22 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Made changes of revisions, need to re-structure Appendix

@@
 This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We present the basic theory of GAMs, and using reported trends of oxygen saturation in tumors we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R.
+We present the basic theory of GAMs and using reported trends of oxygen saturation in tumors, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R.
 We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing).
@@
 Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}}, antibody expression\textsuperscript{\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}}, and cell metabolism\textsuperscript{\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
+Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points, with subjects of analysis ranging from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
+Tumor response\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}}, antibody expression\textsuperscript{\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}}, and cell metabolism\textsuperscript{\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}} are examples of different situations where researchers have used longitudinal designs to study some physiological response.
 Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals\textsuperscript{\protect\hyperlink{ref-greening2018}{9}}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients\textsuperscript{\protect\hyperlink{ref-sio2016}{10}}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients\textsuperscript{\protect\hyperlink{ref-kamstra2015}{11}}.
@@
 This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
+In longitudinal biomedical research, non-linear trends are the norm rather than the exception.
 A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}}.
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
+Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
 
 A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
-Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from data that shows non-linear trends, the validity of such metric is dependent on how adequate the model fits the data.
+Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from data with non-linear trends, the validity of such a metric is dependent on how adequate the model fits the data.
 In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}}.
@@
 Corrections can be made in rm-ANOVA in the absence of compound symmetry\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and group sizes\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}}.
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons\textsuperscript{\protect\hyperlink{ref-charan2013}{30}} which might cause the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
+In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons,\textsuperscript{\protect\hyperlink{ref-charan2013}{30}} which might cause the corrections for lack of compound symmetry to be ineffective.
+
+Due to a variety of causes, the number of observations during a study can be different between all subjects.
 For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible.
@@
 However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}}.
-And even more importantly, LMEMs also assume by default a linear relationship between the response and time\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the trend of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
+And even more importantly, LMEMs also assume by default a linear relationship between the response and time\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} (polynomial effects can be used, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .
+
+As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the trend of the response, there is a need for biomedical researchers to explore the use of statistical tools that allow the data (and not a model assumption) to determine the trend of the fitted model and to enable appropriate inference.
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
@@
 Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data\textsuperscript{\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}}, health-environment interactions\textsuperscript{\protect\hyperlink{ref-yang2012}{35}} and the dynamics of government in political science\textsuperscript{\protect\hyperlink{ref-beck1998}{36}}.
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, and 3) can easily accommodate incomplete observations.
+Therefore, GAMs provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
 
 The current advances in programming languages designed for statistical analysis (specifically R), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, R\textsuperscript{\protect\hyperlink{ref-r}{38}} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, R has many tools that simplify data simulation, an emerging strategy used to test statistical models\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}.
-Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
+In particular, R\textsuperscript{\protect\hyperlink{ref-r}{38}} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}} that speed up the initial stages of the analysis and enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills.
+At the same time, R has many tools that simplify data simulation.
+
+Data simulation methods are an emerging technique that allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation.
+In addition, simulation can be also used for power calculations and study design questions\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}.
 
@@
 
-In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of treatment (\(treatment_j\)), time (\(time_t\)), and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
+In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of treatment (\(treatment_j\)), time (\(time_t\)), and their interaction \(time_t\times treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
 Independent errors \(\varepsilon_{ijt}\) represent random variation from the sampling process assumed to be \(\stackrel{i.i.d.}\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
@@
 As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
-The random effect \(\alpha_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
+The random effect \(\alpha_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\alpha_{ij} \sim N(0,\sigma^2_\alpha)\).
 In essence, the \emph{random effect} in a LMEM enables fitting models with different intercepts at the subject-level\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
@@
 Of note, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach (using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}), which is described in detail by Pinheiro and Bates\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
-However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates\textsuperscript{\protect\hyperlink{ref-beck1998}{36}}; more importantly, polynomial effects lack biological or mechanistic interpretation\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} which limits their use in biomedical studies.
+However, polynomial fits have limited predictive power, cause bias on the boundaries of the covariates\textsuperscript{\protect\hyperlink{ref-beck1998}{36}}, and more importantly, their lack of biological or mechanistic interpretation limits their use in biomedical studies\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
 
@@
 Under this assumption, the between-subject variance and within-subject correlation are constant across time\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}}.
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time\textsuperscript{\protect\hyperlink{ref-maxwell2017}{44}}; and it is higher between consecutive measurements\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}}.
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and they are not robust if the group sizes are unbalanced\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}}.
+However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time\textsuperscript{\protect\hyperlink{ref-maxwell2017}{44}}; and is higher between consecutive measurements\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}}.
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}} their effectiveness is dependent on sample size and number of repeated measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and it has been shown that corrections are not robust if the group sizes are unbalanced\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}}.
 Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
@@
 \hypertarget{simulation}{%
-\subsection{What do an rm-ANOVA and LMEM fit look like? A visual representation using simulated data}\label{simulation}}
+\subsection{What does the fit of an rm-ANOVA and LMEM look like? A visual representation using simulated data}\label{simulation}}
 
 To visually demonstrate the limitations of rm-ANOVA and LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using R\textsuperscript{\protect\hyperlink{ref-r}{38}} and the package \emph{nlme}\textsuperscript{\protect\hyperlink{ref-nlme}{50}}.
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group using R\textsuperscript{\protect\hyperlink{ref-r}{38}} and the package \emph{nlme}\textsuperscript{\protect\hyperlink{ref-nlme}{50}}.
 
 Briefly, two cases for the mean response for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}C).
-In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments where a strong non-linear trend is present.
+In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}D).
+In both the linear and quadratic simulated data, the groups start with the same mean value in order to simulate the expected temporal evolution of some physiological quantity, starting at a common initial condition.
 
@@
 From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D).
+Thus, the response per subject at each timepoint in both the linear and quadratic simulation corresponds to the mean response per group plus the error (represented by the points in Figure \ref{fig:l-q-response} A, D).
 
@@
 Points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation.
+The rm-ANOVA model not only fails to pick the trend of the quadratic data (E) but also assigns a global estimate that does not take into account the between-subject variation.
 \textbf{C, F}: Estimates from the LMEM in the linear and quadratic case (subject: thin lines, population: thick lines) .
@@
 
-The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response}A).
-When the linearity and compound symmetry assumptions are met, the rm-ANOVA model approximates well the global trend by group (Figure \ref{fig:l-q-response}B).
+The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data (\ref{fig:l-q-response}B), as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response}A).
 Note that because the LMEM incorporates \emph{random effects}, is able to provide estimates for each subject and a ``population'' estimate (Figure \ref{fig:l-q-response}C).
@@
 Notice that the model misspecification is easily noticeable if the model fit and the response are visualized.
-In the following section, we provide a brief overview of linear models, general linear models and generalized linear mixed models before presenting the theory of GAMs.
+In the following section, we provide a brief overview of linear models, general linear models and generalized linear mixed models before presenting the theory of GAMs, a class of semi-parametric models that can fit non-linear trends in data and that overcome the limitations of rm-ANOVA and LMEMs in the analysis of biomedical longitudinal data.
 
@@
 In reality, rm-ANOVA and LMEMs are just \emph{special cases} of a broader class of models (General Linear Models and Generalized Linear Mixed Models, respectively).
-In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different classes of models, indicate how rm-ANOVA and LMEMs fit within this framework and introduce the theory of GAMs.
+In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different classes of models and indicate how rm-ANOVA, LMEMs, and GAMs fit within this framework.
 
@@
 
-where \(y_{ijt}\) is the observation \(i\) in group \(j\) at time \(t\), that is assumed to come from some distribution of the exponential family \(\mathcal{D}\), with some mean \(\mu_{ijt}\) and potentially, a dispersion parameter \(\phi\) (which in the Gaussian case is the variance \(\sigma^{2}\)).
-The mean (\(\mu_{ijt}\)) is also known as the \emph{expected value} (or \emph{expectation}), \(E(y_{ijt})\) of the observed response \(y_{ijt}\).
+where \(y_{ijt}\) is the observation \(i\) in group \(j\) at time \(t\), that is assumed to come from some distribution of the exponential family \(\mathcal{D}\), with some mean \(\mu_{ijt}\), and potentially, a dispersion parameter \(\phi\) (which in the Gaussian case is the variance \(\sigma^{2}\)).
+The mean (\(\mu_{ijt}\)) is also known as the \emph{expected value} (or \emph{expectation}) \(E(y_{ijt})\) of the observed response \(y_{ijt}\).
 
@@
 
-where \(\eta_{ijt}\) is the linear predictor for each observation in each group at each timepoint, and \(\beta_0\) (the intercept), \(\beta_{1}, \beta_2,\) and \(\beta_3\) are the model parameters for each group, which can be referred globally to as \(\beta_j\).
-Finally, \(time_{ijt}\) represents the covariates from each subject in each group at each time point.
+where \(\eta_{ijt}\) is the linear predictor for each observation \(i\) in each group \(j\), at each timepoint \(t\), and \(\beta_0\) (the intercept), \(\beta_{1}, \beta_2,\) and \(\beta_3\) are the model parameters for each group for the linear mean from Equation (\eqref{eq:linear-model}).
+Finally, \(time_{t}\) represents the covariates from each subject in each group at each time point.
 
@@
 \begin{equation}
-\eta_{ijt}=\beta_{0}+ \beta_{1} \times treatment_{j} +f(time_{t}|\beta_{j}).
+\eta_{ijt}=\beta_{0}+ \beta_{1} \times treatment_{j} +f(time_{t}|\beta_{j}),
 \label{eq:GAM-eta}
@@
 
-Where \(\beta_{0}\) is the intercept, and \(\beta_{1}\) is the coefficient for each treatment group.
-Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time have been replaced by \(f(time_{t}|\beta_{j})\), which represents the smooth function of time with with inputs as the covariates \(time_t\) and parameters \(\beta_{j}\).
+where \(\beta_{0}\) is the intercept, and \(\beta_{1}\) is the coefficient for each treatment group.
+Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time, and the interaction between time and treatment have been replaced by \(f(time_{t}|\beta_{j})\), which represents the smooth function of time with inputs as the covariates \(time_t\) and parameters \(\beta_{j}\).
 A GAM version of a linear model can be written as:
@@
 \begin{equation}
-  y_{ijt}=\beta_0+f(time_t\mid \beta_j)+\varepsilon_{ijt}
+  y_{ijt}=\beta_0+f(time_t\mid \beta_j)+\varepsilon_{ijt},
   \label{eq:GAM}
@@
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the smooth function \(f(time_t\mid \beta_j)\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is more advantageous as it allows more flexibility because it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
+where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the smooth function \(f(time_t\mid \beta_j)\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
+
+In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
 One possible set of functions for \(f(time_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference\textsuperscript{\protect\hyperlink{ref-beck1998}{36}}.
@@
 
-The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using a \emph{basis function} expansion of the covariates and by estimating random coefficients associated with these basis functions.
+The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, using a \emph{basis function} expansion of the covariates and estimating random coefficients associated with these basis functions.
 A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}}.
@@
 
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness\textsuperscript{\protect\hyperlink{ref-wegman1983}{58}}.
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions that have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness\textsuperscript{\protect\hyperlink{ref-wegman1983}{58}}.
 Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
@@
 Cubic splines (CS) are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework\textsuperscript{\protect\hyperlink{ref-hastie1987}{56}}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots', the points where the basis functions meet) to obtain the finite basis, which could affect the model fit\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}.
+The use of CS as smoothers in GAMs was discussed within the original GAM framework\textsuperscript{\protect\hyperlink{ref-hastie1987}{56}}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots', the points where the basis functions meet) to obtain the finite basis, which affects model fit\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}.
 A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}.
@@
 In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}.
-Like TPS, TPRS only require the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
+Like TPS, TPRS only requires specifying the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).
+
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}D.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
@@
 
-With this in mind, if five basis functions are used to fit a GAM for the data that appears in Figure \ref{fig:l-q-response}C, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A.
-The four basis functions (and the intercept) are shown.
-Each of the basis functions is evaluated at six different points (because there are six points on the timeline).
+With this in mind, if five basis functions are used to fit a GAM for the data of Group 1 (concave down function) that appears in Figure \ref{fig:l-q-response}D, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A, where the four basis functions (and the intercept) are shown.
+Each of the five basis functions is evaluated at six different points (because there are six points on the timeline).
 The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), that is estimated when fitting the model.
-The penalized estimates fitted for our example are shown in Figure \ref{fig:basis-plot}B.
-
-To get the weighted basis functions, each basis (from Figure Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
+The penalized coefficient estimates fitted for our example are shown in Figure \ref{fig:basis-plot}B.
+
+To get the weighted basis functions, each basis (from Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
 Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and less than 1).
+Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
 On the other hand, the third basis function has roughly doubled its contribution to the smooth term.
 Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (orange line), along the simulated values per group, which appear as points.
+The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (brown line), along the simulated values per group, which appear as points.
 
@@
 \textbf{C}: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
+Each of the five basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
 Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
 \textbf{D}: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
+The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for Group 1 shown as points.}\label{fig:basis-plot}
 \end{figure}
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}}.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge and observed data\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}}.
 In the case of data that shows non-linear trends, the prior that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
@@
 
-To understand this last part, it is worth reminding that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function exists, taking into account the randomness introduced by the sampling process.
+To understand across the function coverage, recall that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function exists, taking into account the randomness introduced by the sampling process.
 Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie.
-In an \emph{across the function} CI (like those estimated by default for GAMs using \emph{mgcv}) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%).
-In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI and for this to occur, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
-
-However, besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
+In an \emph{across the function} CI (like those estimated by default for GAMs using \emph{mgcv}), if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%).
+In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI.
+As a consequence, some areas of CI for the function must have more than nominal coverage and some areas less than the nominal coverage.
+
+Besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
 Suppose we chose a nominal value (say, 95\%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
-Briefly, to obtain a simultaneous CI we use simulation to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}.
-Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups, as we discuss in the following section.
+Briefly, to obtain a simultaneous CI we use simulation from the empirical Bayesian posterior distribution to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}.
+Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
 
@@
 The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors (Figure 3C in Vishwanath et.
-al.\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}}).
-In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
+al.\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}}), where diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
 In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group.
 Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
-The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}A.
+The simulation based on the real data appears in Figure \ref{fig:sim-smooth-plot}A.
 
@@
 
-An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination.
-In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time.
+An interaction effect is typically the main interest in longitudinal biomedical data, as the interaction takes into account treatment, time, and their combination.
+In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect for each treatment to account for the repeated measures over time.
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in Hefley\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}}.
-
-For the data in Figure \ref{fig:sim-smooth-plot}A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
-To estimate this, the model incorporates separate smooths for \emph{Group} and \emph{Day}, respectively.
+A detailed description on the close relationship between basis functions and correlation functions can be found in Hefley et.
+al.\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}}.
+
+For the data in Figure \ref{fig:sim-smooth-plot}A, the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
+To estimate this, the model incorporates separate smooths for each \emph{Group} as a function of \emph{Day}.
 The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}).
 Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the group means.
-Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:
+Using R and the package \passthrough{\lstinline!mgcv!} the model syntax is:
 
 \begin{lstlisting}[language=R]
-gam_02 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
+gam_02 <- gam(StO2_sim ~ Group + s(Day, by = Group, k = 5), method ='REML', data = dat_sim)
 \end{lstlisting}
 
-This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using four basis functions (plus the intercept).
+This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using five basis functions.
 The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} (a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console).
-
-The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups.
+Finally, the parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, as we have indicated above.
+
 Although the default \passthrough{\lstinline!method!} used to estimate the smoothing parameters in \emph{mgcv} is generalized cross validation (GCV), Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} showed the restricted maximum likelihood (REML) to be more resistant to overfitting while also easing the quantification of uncertainty in the smooth parameters; therefore in this manuscript REML is always used for smooth parameter estimation.
 An additional argument (\passthrough{\lstinline!family!}) allows to specify the expected distribution of the response, but it is not used in this model because we expect a normally-distributed response (which is the default \passthrough{\lstinline!family!} in \emph{mgcv}).
-
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}\textsuperscript{\protect\hyperlink{ref-gratia}{64}}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak\textsuperscript{\protect\hyperlink{ref-harezlak2018}{65}}.
-
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?''.
-The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
-This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
-
-Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists.
-The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
-However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
-Consider the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}B.
-If 40\% of the observations are randomly deleted and the same interaction GAM (\passthrough{\lstinline!gam\_02!}) is fitted, the resulting smooths appear in (Figure \ref{fig:sim-smooth-plot}D).
-Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.
-
-Additionally, note that in Figures \ref{fig:sim-smooth-plot}B,D we show two CIs for each of the fitted smooths (shaded regions).
-The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, we expect them to be wider than the across the function CI, resulting in the wide shaded regions.
-For the dataset with missing observations, the smooths CIs overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset.
-However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.
 
@@
 
+When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}\textsuperscript{\protect\hyperlink{ref-gratia}{64}}.
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak\textsuperscript{\protect\hyperlink{ref-harezlak2018}{65}}.
+
+One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?''.
+The fit of an rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model}, is presented in Figure \ref{fig:sim-smooth-plot}C.
+This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
+
+Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists.
+The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
+However, this usually causes the resulting smooths to have wider confidence intervals and less ability to discern differences in trends.
+To exemplify this, consider the random deletion of 40\% of the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}A.
+If the same interaction GAM (\passthrough{\lstinline!gam\_02!}) is fitted to this data with missing observations, the resulting smooths appear in (Figure \ref{fig:sim-smooth-plot}D).
+Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.
+
+Additionally, note that in Figure \ref{fig:sim-smooth-plot}B,D we show two CIs for each of the fitted smooths (shaded regions).
+The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, they are wider than the across the function CI, resulting in the wide shaded regions.
+For the dataset with missing observations, the CIs for the smooths overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset.
+However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.
+
+\hypertarget{GAM-significance}{%
+\subsection{Determination of significance in GAMs for longitudinal data}\label{GAM-significance}}
+
+At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
+Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
+
 \begin{figure}
@@
 \end{figure}
-
-\hypertarget{GAM-significance}{%
-\subsection{Determination of significance in GAMs for longitudinal data}\label{GAM-significance}}
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
 

### 2021-12-22 Manuscripts/Manuscript_by_chapters-SIM_Revisions/SIM_Appendix.tex
Made changes of revisions, need to re-structure Appendix

@@
 ]{article}
-\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
-Beyond repeated measures ANOVA and Linear Mixed Models\\
-\strut \\
-SUPPLEMENTARY MATERIALS: APPENDIX}
-\date{\vspace{-2.5em}}
-
 \usepackage{amsmath,amssymb}
@@
 
+\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
+Beyond repeated measures ANOVA and Linear Mixed Models\\
+\strut \\
+SUPPLEMENTARY MATERIALS: APPENDIX}
+\date{\vspace{-2.5em}}
+
 \begin{document}
@@
 Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response {[}\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}{]}, antibody expression {[}\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}{]}, and cell metabolism {[}\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
+Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points, with subjects of analysis ranging from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
+Tumor response {[}\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}{]}, antibody expression {[}\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}{]}, and cell metabolism {[}\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}{]} are examples of different situations where researchers have used longitudinal designs to study some physiological response.
 Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}\protect\hyperlink{ref-greening2018}{9}{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}\protect\hyperlink{ref-sio2016}{10}{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}\protect\hyperlink{ref-kamstra2015}{11}{]}.
@@
 This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
+In longitudinal biomedical research, non-linear trends are the norm rather than the exception.
 A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
+Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
 
 A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
-Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from data that shows non-linear trends, the validity of such metric is dependent on how adequate the model fits the data.
+Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from data with non-linear trends, the validity of such a metric is dependent on how adequate the model fits the data.
 In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}.
@@
 Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and group sizes {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}\protect\hyperlink{ref-charan2013}{30}{]} which might cause the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
+In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons, {[}\protect\hyperlink{ref-charan2013}{30}{]} which might cause the corrections for lack of compound symmetry to be ineffective.
+
+Due to a variety of causes, the number of observations during a study can be different between all subjects.
 For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible.
@@
 However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}.
-And even more importantly, LMEMs also assume by default a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]} (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the trend of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
+And even more importantly, LMEMs also assume by default a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]} (polynomial effects can be used, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .
+
+As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the trend of the response, there is a need for biomedical researchers to explore the use of statistical tools that allow the data (and not a model assumption) to determine the trend of the fitted model and to enable appropriate inference.
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
@@
 Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]}.
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]}, and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]}, and 3) can easily accommodate incomplete observations.
+Therefore, GAMs provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
 
 The current advances in programming languages designed for statistical analysis (specifically R), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, R {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, R has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
-Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
+In particular, R {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that speed up the initial stages of the analysis and enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills.
+At the same time, R has many tools that simplify data simulation.
+
+Data simulation methods are an emerging technique that allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation.
+In addition, simulation can be also used for power calculations and study design questions {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
 
@@
 
-In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of treatment (\(treatment_j\)), time (\(time_t\)), and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
+In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of treatment (\(treatment_j\)), time (\(time_t\)), and their interaction \(time_t\times treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
 Independent errors \(\varepsilon_{ijt}\) represent random variation from the sampling process assumed to be \(\stackrel{i.i.d.}\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
-In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug,'' or ``saline'' vs.~``chemotherapy'').
+In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'', or ``saline'' vs.~``chemotherapy'').
 Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B).
@@
 As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
-The random effect \(\alpha_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
+The random effect \(\alpha_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\alpha_{ij} \sim N(0,\sigma^2_\alpha)\).
 In essence, the \emph{random effect} in a LMEM enables fitting models with different intercepts at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
@@
 Of note, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach (using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}), which is described in detail by Pinheiro and Bates {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates {[}\protect\hyperlink{ref-beck1998}{36}{]}; more importantly, polynomial effects lack biological or mechanistic interpretation {[}\protect\hyperlink{ref-pinheiro2006}{15}{]} which limits their use in biomedical studies.
+However, polynomial fits have limited predictive power, cause bias on the boundaries of the covariates {[}\protect\hyperlink{ref-beck1998}{36}{]}, and more importantly, their lack of biological or mechanistic interpretation limits their use in biomedical studies {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
 
@@
 Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}{]}.
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}\protect\hyperlink{ref-maxwell2017}{44}{]}; and it is higher between consecutive measurements {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and they are not robust if the group sizes are unbalanced {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
+However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}\protect\hyperlink{ref-maxwell2017}{44}{]}; and is higher between consecutive measurements {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]} their effectiveness is dependent on sample size and number of repeated measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and it has been shown that corrections are not robust if the group sizes are unbalanced {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
 Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
@@
 In a longitudinal study, it is frequently the case that the number of observations is different across subjects.
-In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing,'' ``incomplete,'' or (more generally speaking) unbalanced data {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
+In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing'', ``incomplete'', or (more generally speaking) unbalanced data {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
 The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
@@
 \hypertarget{simulation}{%
-\subsection{What do an rm-ANOVA and LMEM fit look like? A visual representation using simulated data}\label{simulation}}
+\subsection{What does the fit of an rm-ANOVA and LMEM look like? A visual representation using simulated data}\label{simulation}}
 
 To visually demonstrate the limitations of rm-ANOVA and LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using R {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{50}{]}.
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group using R {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{50}{]}.
 
 Briefly, two cases for the mean response for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}C).
-In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments where a strong non-linear trend is present.
+In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}D).
+In both the linear and quadratic simulated data, the groups start with the same mean value in order to simulate the expected temporal evolution of some physiological quantity, starting at a common initial condition.
 
@@
 From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D).
+Thus, the response per subject at each timepoint in both the linear and quadratic simulation corresponds to the mean response per group plus the error (represented by the points in Figure \ref{fig:l-q-response} A, D).
+\textbf{Move simulated to the front: the simulated response per subject}
 
@@
 
-This section has used simulation to better convey and visulize the limitations of linearity and correlation in the response in data with non-linear trends using an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
+This section has used simulation to better convey and visualize the limitations of linearity and correlation in the response in data with non-linear trends using an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
 Notice that the model misspecification is easily noticeable if the model fit and the response are visualized.
-In the following section, we provide a brief overview of linear models, general linear models and generalized linear mixed models before presenting the theory of GAMs.
+In the following section, we provide a brief overview of linear models, general linear models and generalized linear mixed models before presenting the theory of GAMs, a class of semi-parametric models that can fit non-linear trends in data and that overcome the limitations of rm-ANOVA and LMEMs in the analysis of biomedical longitudinal data.
 
@@
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} + \beta_2 \times time_{t} + +\beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij},
+\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} + \beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij},
 \label{eq:GLMM-eta}
@@
 
-Where \(\beta_{0}\) is the intercept, and \(\beta_{1}\) is the coefficient for each treatment group.
-Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time have been replaced by \(f(time_{t}|\beta_{j})\), which represents the smooth function of time with with inputs as the covariates \(time_t\) and parameters \(\beta_{j}\).
-A GAM version of a linear model can be written as:
+where \(\beta_{0}\) is the intercept, and \(\beta_{1}\) is the coefficient for each treatment group.
+Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time, and the interaction between time and treatment have been replaced by \(f(time_{t}|\beta_{j})\).
+The smooth term \(f(time_{t}|\beta_{j})\) gives a different smooth response for each treatment.
+If the smooth term represented a linear relationship, then \(f(time_{t}|\beta_{j})= \beta_2 \times time_t+\beta_3 \times time_t \times treatment_j\); however, in general, the smooth term is a more flexible function than a linear relationship, with parameter vectors \(\beta_{j}\) for each treatment.
+A GAM version of a linear model can be written as
 
 \begin{equation}
-  y_{ijt}=\beta_0+f(time_t\mid \beta_j)+\varepsilon_{ijt}
+  y_{ijt}=\beta_0+ \beta_1 \times treatment_j + f(time_t\mid \beta_j)+\varepsilon_{ijt},
   \label{eq:GAM}
@@
 
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the smooth function \(f(time_t\mid \beta_j)\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is more advantageous as it allows more flexibility because it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
+where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
+
+In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
 One possible set of functions for \(f(time_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
@@
 
-The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using a \emph{basis function} expansion of the covariates and by estimating random coefficients associated with these basis functions.
+The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, using a \emph{basis function} expansion of the covariates and estimating random coefficients associated with these basis functions.
 A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
@@
 
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{58}{]}.
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions that have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{58}{]}.
 Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
@@
 Cubic splines (CS) are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework {[}\protect\hyperlink{ref-hastie1987}{56}{]}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the basis functions meet) to obtain the finite basis, which could affect the model fit {[}\protect\hyperlink{ref-wood2003}{59}{]}.
+The use of CS as smoothers in GAMs was discussed within the original GAM framework {[}\protect\hyperlink{ref-hastie1987}{56}{]}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots', the points where the basis functions meet) to obtain the finite basis, which affects model fit {[}\protect\hyperlink{ref-wood2003}{59}{]}.
 A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}{]}.
@@
 In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner {[}\protect\hyperlink{ref-wood2003}{59}{]}.
-Like TPS, TPRS only require the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood{[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}{]}).
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
+Like TPS, TPRS only requires specifying the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood{[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}{]}).
+
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}D.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge {[}\protect\hyperlink{ref-mcelreath2018}{60}{]}.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge and observed data {[}\protect\hyperlink{ref-mcelreath2018}{60}{]}.
 In the case of data that shows non-linear trends, the prior that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
@@
 
-To understand this last part, it is worth reminding that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function exists, taking into account the randomness introduced by the sampling process.
+To understand across the function coverage, recall that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function exists, taking into account the randomness introduced by the sampling process.
 Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie.
-In an \emph{across the function} CI (like those estimated by default for GAMs using \emph{mgcv}) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%).
-In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI and for this to occur, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
-
-However, besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability {[}\protect\hyperlink{ref-wood2017}{37}{]}.
+In an \emph{across the function} CI (like those estimated by default for GAMs using \emph{mgcv}), if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%).
+In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI.
+As a consequence, some areas of CI for the function must have more than nominal coverage and some areas less than the nominal coverage.
+
+Besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability {[}\protect\hyperlink{ref-wood2017}{37}{]}.
 Suppose we chose a nominal value (say, 95\%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
-Briefly, to obtain a simultaneous CI we use simulation to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI {[}\protect\hyperlink{ref-ruppert2003}{62}{]}.
-Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups, as we discuss in the following section.
+Briefly, to obtain a simultaneous CI we use simulation from the empirical Bayesian posterior distribution to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI {[}\protect\hyperlink{ref-ruppert2003}{62}{]}.
+Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
 
@@
 The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors (Figure 3C in Vishwanath et.
-al.{[}\protect\hyperlink{ref-vishwanath2009}{16}{]}).
-In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
+al.{[}\protect\hyperlink{ref-vishwanath2009}{16}{]}), where diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
 In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group.
 Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
-The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}A.
+The simulation based on the real data appears in Figure \ref{fig:sim-smooth-plot}A.
 
@@
 
-An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination.
-In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time.
+An interaction effect is typically the main interest in longitudinal biomedical data, as the interaction takes into account treatment, time, and their combination.
+In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect for each treatment to account for the repeated measures over time.
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in Hefley {[}\protect\hyperlink{ref-hefley2017}{57}{]}.
-
-For the data in Figure \ref{fig:sim-smooth-plot}A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
-To estimate this, the model incorporates separate smooths for \emph{Group} and \emph{Day}, respectively.
+A detailed description on the close relationship between basis functions and correlation functions can be found in Hefley et.
+al.
+{[}\protect\hyperlink{ref-hefley2017}{57}{]}.
+
+For the data in Figure \ref{fig:sim-smooth-plot}A, the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
+To estimate this, the model incorporates separate smooths for each \emph{Group} as a function of \emph{Day}.
 The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}).
 Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the group means.
-Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:
+Using R and the package \passthrough{\lstinline!mgcv!} the model syntax is:
 
 \begin{lstlisting}[language=R]
-gam_02 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
+gam_02 <- gam(StO2_sim ~ Group + s(Day, by = Group, k = 5), method ='REML', data = dat_sim)
 \end{lstlisting}
 
-This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using four basis functions (plus the intercept).
+This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using five basis functions.
 The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]} (a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console).
-
-The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups.
+Finally, the parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, as we have indicated above.
+
 Although the default \passthrough{\lstinline!method!} used to estimate the smoothing parameters in \emph{mgcv} is generalized cross validation (GCV), Wood{[}\protect\hyperlink{ref-wood2017}{37}{]} showed the restricted maximum likelihood (REML) to be more resistant to overfitting while also easing the quantification of uncertainty in the smooth parameters; therefore in this manuscript REML is always used for smooth parameter estimation.
 An additional argument (\passthrough{\lstinline!family!}) allows to specify the expected distribution of the response, but it is not used in this model because we expect a normally-distributed response (which is the default \passthrough{\lstinline!family!} in \emph{mgcv}).
-
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{64}{]}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood {[}\protect\hyperlink{ref-wood2017}{37}{]} and Harezlak {[}\protect\hyperlink{ref-harezlak2018}{65}{]}.
-
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
-This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
-
-Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists.
-The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
-However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
-Consider the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}B.
-If 40\% of the observations are randomly deleted and the same interaction GAM (\passthrough{\lstinline!gam\_02!}) is fitted, the resulting smooths appear in (Figure \ref{fig:sim-smooth-plot}D).
-Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.
-
-Additionally, note that in Figures \ref{fig:sim-smooth-plot}B,D we show two CIs for each of the fitted smooths (shaded regions).
-The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, we expect them to be wider than the across the function CI, resulting in the wide shaded regions.
-For the dataset with missing observations, the smooths CIs overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset.
-However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.
 
@@
 
+When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{64}{]}.
+A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood {[}\protect\hyperlink{ref-wood2017}{37}{]} and Harezlak {[}\protect\hyperlink{ref-harezlak2018}{65}{]}.
+
+One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?''.
+The fit of an rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model}, is presented in Figure \ref{fig:sim-smooth-plot}C.
+This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
+
+Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists.
+The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
+However, this usually causes the resulting smooths to have wider confidence intervals and less ability to discern differences in trends.
+To exemplify this, consider the random deletion of 40\% of the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}A.
+If the same interaction GAM (\passthrough{\lstinline!gam\_02!}) is fitted to this data with missing observations, the resulting smooths appear in (Figure \ref{fig:sim-smooth-plot}D).
+Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.
+
+Additionally, note that in Figure \ref{fig:sim-smooth-plot}B,D we show two CIs for each of the fitted smooths (shaded regions).
+The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, they are wider than the across the function CI, resulting in the wide shaded regions.
+For the dataset with missing observations, the CIs for the smooths overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset.
+However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.
+
+\hypertarget{GAM-significance}{%
+\subsection{Determination of significance in GAMs for longitudinal data}\label{GAM-significance}}
+
+At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
+Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
+
 \begin{figure}
@@
 
-\hypertarget{GAM-significance}{%
-\subsection{Determination of significance in GAMs for longitudinal data}\label{GAM-significance}}
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
-
 The absence of a \emph{p-value} in this context might seem odd, but its interpretation can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment.
@@
 
-Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}B and Figure \ref{fig:sim-smooth-plot}D.
+Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}B, D.
 Figure \ref{fig:plot-pairwise-comp} shows the comparison between each treatment group for the full and missing datasets with a simultaneous CI computed around the difference.
@@
 
-Biomedical longitudinal data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
-Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}, where it is clear that these models do not capture the non-linear trend of the data, thereby causing a ``model misspecification error.''
-
+Biomedical longitudinal data is particularly challenging to analyze due to the frequency of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
+Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}.
+When these models do not capture the non-linear trend of the data, this results in a ``model misspecification error''.
 This ``model misspecification'' error, also is known as a ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Polynomial effects can be used in LMEMs, and it could be argued that in Section \ref{simulation} a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects presents more questions than answers in this case (i.e., what is the appropriate polynomial effect to use?
-quadratic?, cubic?, what is the biological interpretation of a polynomial effect?).
-Moreover, polynomial effects have other associated issues that we have described in Section \ref{GAM-theory}.
-
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{66}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.
+
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{66}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) or \emph{p-value} value, whose interpretation is not clear in the case of a Type III error.
 
 In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{67}{]}.
-This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (e.g., which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{67}{]}.
+This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs.
+T
+
+The purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model, such as in the case of polynomial effects in LMEMs.
+In contrast, because GAMs let the data speak for themselves they can provide an accurate representation of the effect of time in a biological process.
 
 However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} using \(\textsf{R}\) and the package \emph{mgcv}{[}\protect\hyperlink{ref-wood2017}{37}{]} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
-One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
-The model is therefore able to identify changes between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
-
-We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} using R and the package \emph{mgcv}{[}\protect\hyperlink{ref-wood2017}{37}{]} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
+
+One of the features of GAMs is that their Bayesian interpretation allows for inference about differences between groups without the need of a \emph{p-value}, and in addition can provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
+The model is therefore able to identify changes between the groups at time points were data was not directly measured even in the case of incomplete observations ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
+
+We have used R as the software of choice for this paper because R not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
 In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{68},\protect\hyperlink{ref-weissgerber2018}{69}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
@@
 108 (2011) 14626--14631.
-\url{https://doi.org/10.1073/pnas.1013103108}.}
+https://doi.org/\href{https://doi.org/10.1073/pnas.1013103108}{10.1073/pnas.1013103108}.}
 
@@
 22 (2020).
-\url{https://doi.org/10.1186/s13058-020-01262-1}.}
+https://doi.org/\href{https://doi.org/10.1186/s13058-020-01262-1}{10.1186/s13058-020-01262-1}.}
 
@@
 23 (2018) 1.
-\url{https://doi.org/10.1117/1.jbo.23.9.091410}.}
+https://doi.org/\href{https://doi.org/10.1117/1.jbo.23.9.091410}{10.1117/1.jbo.23.9.091410}.}
 
@@
 8 (2018).
-\url{https://doi.org/10.1038/s41598-017-18635-w}.}
+https://doi.org/\href{https://doi.org/10.1038/s41598-017-18635-w}{10.1038/s41598-017-18635-w}.}
 
@@
 376 (2017) 1589--1590.
-\url{https://doi.org/10.1056/nejmc1616623}.}
+https://doi.org/\href{https://doi.org/10.1056/nejmc1616623}{10.1056/nejmc1616623}.}
 
@@
 1 (2018).
-\url{https://doi.org/10.1038/s42003-018-0206-4}.}
+https://doi.org/\href{https://doi.org/10.1038/s42003-018-0206-4}{10.1038/s42003-018-0206-4}.}
 
@@
 15 (2010) 011112.
-\url{https://doi.org/10.1117/1.3285584}.}
+https://doi.org/\href{https://doi.org/10.1117/1.3285584}{10.1117/1.3285584}.}
 
@@
 9 (2018) 2871.
-\url{https://doi.org/10.1364/boe.9.002871}.}
+https://doi.org/\href{https://doi.org/10.1364/boe.9.002871}{10.1364/boe.9.002871}.}
 
@@
 24 (2016) 3847--3855.
-\url{https://doi.org/10.1007/s00520-016-3213-3}.}
+https://doi.org/\href{https://doi.org/10.1007/s00520-016-3213-3}{10.1007/s00520-016-3213-3}.}
 
@@
 51 (2015) 548--555.
-\url{https://doi.org/10.1016/j.oraloncology.2015.01.016}.}
+https://doi.org/\href{https://doi.org/10.1016/j.oraloncology.2015.01.016}{10.1016/j.oraloncology.2015.01.016}.}
 
@@
 181--207.
-\url{https://doi.org/10.1007/978-0-387-09612-4_9}.}
+https://doi.org/\href{https://doi.org/10.1007/978-0-387-09612-4_9}{10.1007/978-0-387-09612-4\_9}.}
 
@@
 \CSLLeftMargin{{[}13{]} }
-\CSLRightInline{R. Gueorguieva, J.H. Krystal, Move over {ANOVA}, Archives of General Psychiatry. 61 (2004) 310. \url{https://doi.org/10.1001/archpsyc.61.3.310}.}
+\CSLRightInline{R.
+Gueorguieva, J.H.
+Krystal, Move over {ANOVA}, Archives of General Psychiatry.
+61 (2004) 310.
+https://doi.org/\href{https://doi.org/10.1001/archpsyc.61.3.310}{10.1001/archpsyc.61.3.310}.}
 
@@
 127 (2018) 569--575.
-\url{https://doi.org/10.1213/ane.0000000000003511}.}
+https://doi.org/\href{https://doi.org/10.1213/ane.0000000000003511}{10.1213/ane.0000000000003511}.}
 
@@
 11 (2009) 889--900.
-\url{https://doi.org/10.1593/neo.09580}.}
+https://doi.org/\href{https://doi.org/10.1593/neo.09580}{10.1593/neo.09580}.}
 
@@
 7 (2019).
-\url{https://doi.org/10.3389/fevo.2019.00372}.}
+https://doi.org/\href{https://doi.org/10.3389/fevo.2019.00372}{10.3389/fevo.2019.00372}.}
 
@@
 32 (2019) e100081.
-\url{https://doi.org/10.1136/gpsych-2019-100081}.}
+https://doi.org/\href{https://doi.org/10.1136/gpsych-2019-100081}{10.1136/gpsych-2019-100081}.}
 
@@
 18 (2010) 1724--1730.
-\url{https://doi.org/10.1038/mt.2010.127}.}
+https://doi.org/\href{https://doi.org/10.1038/mt.2010.127}{10.1038/mt.2010.127}.}
 
@@
 12 (2015) 179--185.
-\url{https://doi.org/10.1038/nmeth.3288}.}
+https://doi.org/\href{https://doi.org/10.1038/nmeth.3288}{10.1038/nmeth.3288}.}
 
@@
 \CSLLeftMargin{{[}21{]} }
-\CSLRightInline{H. Abdi, {Holm's sequential Bonferroni procedure}, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.}
+\CSLRightInline{H.
+Abdi, {Holm's sequential Bonferroni procedure}, Encyclopedia of Research Design.
+1 (2010) 1--8.
+https://doi.org/\href{https://doi.org/10.4135/9781412961288.n178}{10.4135/9781412961288.n178}.}
 
@@
 15 (2004) 1044--1045.
-\url{https://doi.org/10.1093/beheco/arh107}.}
+https://doi.org/\href{https://doi.org/10.1093/beheco/arh107}{10.1093/beheco/arh107}.}
 
@@
 5 (2012) 189--211.
-\url{https://doi.org/10.1080/19345747.2011.618213}.}
+https://doi.org/\href{https://doi.org/10.1080/19345747.2011.618213}{10.1080/19345747.2011.618213}.}
 
@@
 \CSLLeftMargin{{[}24{]} }
-\CSLRightInline{C. Albers, The problem with unadjusted multiple and sequential statistical testing, Nature Communications. 10 (2019). \url{https://doi.org/10.1038/s41467-019-09941-0}.}
+\CSLRightInline{C.
+Albers, The problem with unadjusted multiple and sequential statistical testing, Nature Communications.
+10 (2019).
+https://doi.org/\href{https://doi.org/10.1038/s41467-019-09941-0}{10.1038/s41467-019-09941-0}.}
 
@@
 (2004) 2144--2148.
-\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.}
+https://doi.org/\href{https://doi.org/10.1249/01.mss.0000147580.40591.75}{10.1249/01.mss.0000147580.40591.75}.}
 
@@
 1 (1976) 69--82.
-\url{https://doi.org/10.3102/10769986001001069}.}
+https://doi.org/\href{https://doi.org/10.3102/10769986001001069}{10.3102/10769986001001069}.}
 
@@
 \CSLLeftMargin{{[}27{]} }
-\CSLRightInline{S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.}
+\CSLRightInline{S.W.
+Greenhouse, S.
+Geisser, On methods in the analysis of profile data, Psychometrika.
+24 (1959) 95--112.
+https://doi.org/\href{https://doi.org/10.1007/bf02289823}{10.1007/bf02289823}.}
 
@@
 8 (2017).
-\url{https://doi.org/10.3389/fpsyg.2017.01841}.}
+https://doi.org/\href{https://doi.org/10.3389/fpsyg.2017.01841}{10.3389/fpsyg.2017.01841}.}
 
@@
 54 (2001) 1--20.
-\url{https://doi.org/10.1348/000711001159357}.}
+https://doi.org/\href{https://doi.org/10.1348/000711001159357}{10.1348/000711001159357}.}
 
@@
 4 (2013) 303.
-\url{https://doi.org/10.4103/0976-500x.119726}.}
+https://doi.org/\href{https://doi.org/10.4103/0976-500x.119726}{10.4103/0976-500x.119726}.}
 
@@
 68 (2013) 255--278.
-\url{https://doi.org/10.1016/j.jml.2012.11.001}.}
+https://doi.org/\href{https://doi.org/10.1016/j.jml.2012.11.001}{10.1016/j.jml.2012.11.001}.}
 
@@
 82 (2012) 113--135.
-\url{https://doi.org/10.1016/j.gca.2010.12.026}.}
+https://doi.org/\href{https://doi.org/10.1016/j.gca.2010.12.026}{10.1016/j.gca.2010.12.026}.}
 
@@
 7 (2019) e6876.
-\url{https://doi.org/10.7717/peerj.6876}.}
+https://doi.org/\href{https://doi.org/10.7717/peerj.6876}{10.7717/peerj.6876}.}
 
@@
 \CSLLeftMargin{{[}34{]} }
-\CSLRightInline{G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.}
+\CSLRightInline{G.L.
+Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution.
+6 (2018).
+https://doi.org/\href{https://doi.org/10.3389/fevo.2018.00149}{10.3389/fevo.2018.00149}.}
 
@@
 12 (2012).
-\url{https://doi.org/10.1186/1471-2288-12-165}.}
+https://doi.org/\href{https://doi.org/10.1186/1471-2288-12-165}{10.1186/1471-2288-12-165}.}
 
@@
 \CSLLeftMargin{{[}36{]} }
-\CSLRightInline{N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. 42 (1998) 596. \url{https://doi.org/10.2307/2991772}.}
+\CSLRightInline{N.
+Beck, S.
+Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science.
+42 (1998) 596.
+https://doi.org/\href{https://doi.org/10.2307/2991772}{10.2307/2991772}.}
 
@@
 \CSLLeftMargin{{[}37{]} }
-\CSLRightInline{S.N. Wood, Generalized additive models, Chapman; Hall/{CRC}, 2017. \url{https://doi.org/10.1201/9781315370279}.}
+\CSLRightInline{S.N. Wood, Generalized additive models, Chapman; Hall/{CRC}, 2017. https://doi.org/\href{https://doi.org/10.1201/9781315370279}{10.1201/9781315370279}.}
 
@@
 111 (2016) 1548--1563.
-\url{https://doi.org/10.1080/01621459.2016.1180986}.}
+https://doi.org/\href{https://doi.org/10.1080/01621459.2016.1180986}{10.1080/01621459.2016.1180986}.}
 
@@
 1 (1996) 205.
-\url{https://doi.org/10.2307/1400366}.}
+https://doi.org/\href{https://doi.org/10.2307/1400366}{10.2307/1400366}.}
 
@@
 \CSLLeftMargin{{[}42{]} }
-\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://doi.org/10.1007/0-387-28314-5}.}
+\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. https://doi.org/\href{https://doi.org/10.1007/0-387-28314-5}{10.1007/0-387-28314-5}.}
 
@@
 Geisser, S.W.
-Greenhouse, An extension of box{\textquotesingle}s results on the use of the {\textdollar}f{\textdollar} distribution in multivariate analysis, The Annals of Mathematical Statistics.
+Greenhouse, An extension of {B}ox{\textquotesingle}s results on the use of the {\emph{F}} distribution in multivariate analysis, The Annals of Mathematical Statistics.
 29 (1958) 885--891.
-\url{https://doi.org/10.1214/aoms/1177706545}.}
+https://doi.org/\href{https://doi.org/10.1214/aoms/1177706545}{10.1214/aoms/1177706545}.}
 
@@
 \CSLLeftMargin{{[}44{]} }
-\CSLRightInline{S.E. Maxwell, H.D. Delaney, K. Kelley, Designing experiments and analyzing data, Routledge, 2017. \url{https://doi.org/10.4324/9781315642956}.}
+\CSLRightInline{S.E. Maxwell, H.D. Delaney, K. Kelley, Designing experiments and analyzing data, Routledge, 2017. https://doi.org/\href{https://doi.org/10.4324/9781315642956}{10.4324/9781315642956}.}
 
@@
 \CSLLeftMargin{{[}45{]} }
-\CSLRightInline{G. Molenberghs, Analyzing incomplete longitudinal clinical trial data, Biostatistics. 5 (2004) 445--464. \url{https://doi.org/10.1093/biostatistics/kxh001}.}
+\CSLRightInline{G.
+Molenberghs, Analyzing incomplete longitudinal clinical trial data, Biostatistics.
+5 (2004) 445--464.
+https://doi.org/\href{https://doi.org/10.1093/biostatistics/kxh001}{10.1093/biostatistics/kxh001}.}
 
@@
 37 (2012) 99--105.
-\url{https://doi.org/10.1097/aap.0b013e31823ebc74}.}
+https://doi.org/\href{https://doi.org/10.1097/aap.0b013e31823ebc74}{10.1097/aap.0b013e31823ebc74}.}
 
@@
 15 (2006) 213--234.
-\url{https://doi.org/10.1191/0962280206sm448oa}.}
+https://doi.org/\href{https://doi.org/10.1191/0962280206sm448oa}{10.1191/0962280206sm448oa}.}
 
@@
 \CSLLeftMargin{{[}49{]} }
-\CSLRightInline{G.E.P. Box, Science and statistics, 71 (1976) 791--799. \url{https://doi.org/10.1080/01621459.1976.10480949}.}
+\CSLRightInline{G.E.P. Box, Science and statistics, 71 (1976) 791--799. https://doi.org/\href{https://doi.org/10.1080/01621459.1976.10480949}{10.1080/01621459.1976.10480949}.}
 
@@
 \CSLLeftMargin{{[}51{]} }
-\CSLRightInline{R. OHara, J. Kotze, Do not log-transform count data, (2010). \url{https://doi.org/10.1038/npre.2010.4136.1}.}
+\CSLRightInline{R. OHara, J. Kotze, Do not log-transform count data, (2010). https://doi.org/\href{https://doi.org/10.1038/npre.2010.4136.1}{10.1038/npre.2010.4136.1}.}
 
@@
 \CSLLeftMargin{{[}52{]} }
-\CSLRightInline{J.A. Nelder, R.W.M. Wedderburn, Generalized linear models, Journal of the Royal Statistical Society. Series A (General). 135 (1972) 370. \url{https://doi.org/10.2307/2344614}.}
+\CSLRightInline{J.A.
+Nelder, R.W.M.
+Wedderburn, Generalized linear models, Journal of the Royal Statistical Society.
+Series A (General).
+135 (1972) 370.
+https://doi.org/\href{https://doi.org/10.2307/2344614}{10.2307/2344614}.}
 
@@
 82 (1987) 371--386.
-\url{https://doi.org/10.1080/01621459.1987.10478440}.}
+https://doi.org/\href{https://doi.org/10.1080/01621459.1987.10478440}{10.1080/01621459.1987.10478440}.}
 
@@
 98 (2017) 632--646.
-\url{https://doi.org/10.1002/ecy.1674}.}
+https://doi.org/\href{https://doi.org/10.1002/ecy.1674}{10.1002/ecy.1674}.}
 
@@
 \CSLLeftMargin{{[}58{]} }
-\CSLRightInline{E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.}
+\CSLRightInline{E.J.
+Wegman, I.W.
+Wright, Splines in statistics, Journal of the American Statistical Association.
+78 (1983) 351--365.
+https://doi.org/\href{https://doi.org/10.1080/01621459.1983.10477977}{10.1080/01621459.1983.10477977}.}
 
@@
 \CSLLeftMargin{{[}59{]} }
-\CSLRightInline{S.N. Wood, Thin plate regression splines, 65 (2003) 95--114. \url{https://doi.org/10.1111/1467-9868.00374}.}
+\CSLRightInline{S.N. Wood, Thin plate regression splines, 65 (2003) 95--114. https://doi.org/\href{https://doi.org/10.1111/1467-9868.00374}{10.1111/1467-9868.00374}.}
 
@@
 \CSLLeftMargin{{[}60{]} }
-\CSLRightInline{R. McElreath, Statistical rethinking, Chapman; Hall/{CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
+\CSLRightInline{R. McElreath, Statistical rethinking, Chapman; Hall/{CRC}, 2018. https://doi.org/\href{https://doi.org/10.1201/9781315372495}{10.1201/9781315372495}.}
 
@@
 39 (2012) 53--74.
-\url{https://doi.org/10.1111/j.1467-9469.2011.00760.x}.}
+https://doi.org/\href{https://doi.org/10.1111/j.1467-9469.2011.00760.x}{10.1111/j.1467-9469.2011.00760.x}.}
 
@@
 \CSLLeftMargin{{[}65{]} }
-\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with {R}, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
+\CSLRightInline{J.
+Harezlak, D.
+Ruppert, M.P.
+Wand, Semiparametric regression with {R}, Springer New York, 2018.
+https://doi.org/\href{https://doi.org/10.1007/978-1-4939-8853-2}{10.1007/978-1-4939-8853-2}.}
 
@@
 52 (2015) 5--9.
-\url{https://doi.org/10.1016/j.ijnurstu.2014.09.006}.}
+https://doi.org/\href{https://doi.org/10.1016/j.ijnurstu.2014.09.006}{10.1016/j.ijnurstu.2014.09.006}.}
 
@@
 4 (1995) 187--196.
-\url{https://doi.org/10.1177/096228029500400302}.}
+https://doi.org/\href{https://doi.org/10.1177/096228029500400302}{10.1177/096228029500400302}.}
 
@@
 \CSLLeftMargin{{[}68{]} }
-\CSLRightInline{C.G. Begley, J.P.A. Ioannidis, Reproducibility in science, Circulation Research. 116 (2015) 116--126. \url{https://doi.org/10.1161/circresaha.114.303819}.}
+\CSLRightInline{C.G.
+Begley, J.P.A.
+Ioannidis, Reproducibility in science, Circulation Research.
+116 (2015) 116--126.
+https://doi.org/\href{https://doi.org/10.1161/circresaha.114.303819}{10.1161/circresaha.114.303819}.}
 
@@
 7 (2018).
-\url{https://doi.org/10.7554/elife.36163}.}
+https://doi.org/\href{https://doi.org/10.7554/elife.36163}{10.7554/elife.36163}.}
 
@@
 
-\hypertarget{appendix}{%
-\section{APPENDIX}\label{appendix}}
+\section{Appendix A: Code and Functions for Plotting}
 
@@
 
+\section {Appendix B: Basis Functions}
+
 \hypertarget{basis-functions-and-gams}{%
-\subsection{Basis functions and GAMs}\label{basis-functions-and-gams}}
+\section{Basis functions and GAMs}\label{basis-functions-and-gams}}
 
@@
 \textbf{C}: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
+Each of the five basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
 Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
 \textbf{D}: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot-appendix}
+The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for Group 1 shown as points.}\label{fig:basis-plot-appendix}
 \end{figure}
@@
 
-\hypertarget{workflow}{%
-\subsection{A basic Workflow for GAMs}\label{workflow}}
+\section {Appendix C: A basic Workflow for GAMs}
 
@@
 \hypertarget{first-model}{%
-\subsubsection{First model}\label{first-model}}
+\subsection{First model}\label{first-model}}
 
@@
 \hypertarget{graphical-diagnostics}{%
-\paragraph{Graphical diagnostics}\label{graphical-diagnostics}}
+\subsubsection{Graphical diagnostics}\label{graphical-diagnostics}}
 
@@
 \hypertarget{gam-00-model-check}{%
-\paragraph{Model check}\label{gam-00-model-check}}
+\subsubsection{Model check}\label{gam-00-model-check}}
 
@@
 \hypertarget{second-model}{%
-\subsubsection{Second model}\label{second-model}}
+\subsection{Second model}\label{second-model}}
 
@@
 \hypertarget{third-model}{%
-\subsubsection{Third model}\label{third-model}}
+\subsection{Third model}\label{third-model}}
 
@@
 \hypertarget{comparing-models-via-aic}{%
-\subsubsection{Comparing models via AIC}\label{comparing-models-via-aic}}
+\subsection{Comparing models via AIC}\label{comparing-models-via-aic}}
 
@@
 \hypertarget{pairwise-comparisons-of-smooth-confidence-intervals}{%
-\paragraph{Pairwise comparisons of smooth confidence intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
+\subsubsection{Pairwise comparisons of smooth confidence intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
 

### 2021-12-22 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/01-Background.Rmd
Made changes of revisions, need to re-structure Appendix

@@
 Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001; @roth2017], and cell metabolism [@jones2018; @skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response.
+Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points, with subjects of analysis ranging from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
+Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001; @roth2017], and cell metabolism [@jones2018; @skala2010] are examples of different situations where researchers have used longitudinal designs to study some  physiological response.
 Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients [@kamstra2015].
@@
 This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
+In longitudinal biomedical research, non-linear trends are the norm rather than the exception.
 A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011; @skala2010; @vishwanath2009].
 These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
+Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
 
 A _post hoc_ analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
-Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from data that shows non-linear trends, the validity of such metric is dependent on how adequate the model fits the data.
+Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from data with non-linear trends, the validity of such a metric is dependent on how adequate the model fits the data.
 In other words, _p-values_ are valid only if the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019].
@@
 Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976; @greenhouse1959], but the effectiveness of the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001].
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not "large" due to ethical and budgetary reasons [@charan2013] which might cause the corrections for lack of compound symmetry to be ineffective.
+In the case of biomedical research, where living subjects are frequently used, sample sizes are often not "large" due to ethical and budgetary reasons, [@charan2013] which might cause the corrections for lack of compound symmetry to be ineffective.
 
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
+Due to a variety of causes, the number of observations during a study can be different between all subjects.
 For example, in a clinical trial patients may voluntarily withdraw, whereas attrition  due to injury or weight loss in preclinical animal studies is possible.
@@
 However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent [@gueorguieva2004;@barr2013].
-And even more importantly, LMEMs also assume by default a linear relationship between the response and time [@pinheiro2006] (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \@ref(GAM-theory)) .
+And even more importantly, LMEMs also assume by default a linear relationship between the response and time [@pinheiro2006] (polynomial effects can be used, but this approach has its own shortcomings as we discuss in Section \@ref(GAM-theory)) .
 
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the trend of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
+As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the trend of the response, there is a need for biomedical researchers to explore the use of statistical tools that allow the data (and not a model assumption) to determine the trend of the fitted model and to enable appropriate inference.
 In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
@@
 Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012; @pedersen2019; @simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998].
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017], and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
+There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017], and 3) can easily accommodate incomplete observations.
+Therefore, GAMs provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
 
 The current advances in programming languages designed for statistical analysis (specifically R), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, R [@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016; @wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.
-hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, R has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017].
-Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
+In particular, R [@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016; @wood2017] that speed up the initial stages of the analysis and enable the use of advanced modeling structures (e.g.
+hierarchical models, confidence interval comparisons) without requiring advanced programming skills.
+At the same time, R has many tools that simplify data simulation.
+
+Data simulation methods are an emerging technique that allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation.
+In addition, simulation can be also used for power calculations and study design questions [@haverkamp2017].
 

### 2021-12-22 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Made changes of revisions, need to re-structure Appendix

@@
 
-In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of treatment ($treatment_j$), time ($time_t$), and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
+In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of treatment ($treatment_j$), time ($time_t$), and their interaction $time_t\times treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively.
 Independent errors $\varepsilon_{ijt}$ represent random variation from the sampling process assumed to be $\stackrel{i.i.d.}\sim N(0,\sigma^2)$ (independently and identically normally distributed with mean zero and variance $\sigma^2$).
@@
 As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.
-The random effect $\alpha_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be $\mu_{ij} \sim N(0,\sigma^2_\mu)$.
+The random effect $\alpha_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be $\alpha_{ij} \sim N(0,\sigma^2_\alpha)$.
 In essence, the _random effect_ in a LMEM enables fitting models with different intercepts at the subject-level[@pinheiro2006].
@@
 Of note, LMEMs are capable of fitting non-linear trends using an "empirical" approach (using polynomial fixed effects instead of linear effects such as in Equation \@ref(eq:LMEM)), which is described in detail by Pinheiro and Bates [@pinheiro2006].
-However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates [@beck1998]; more importantly, polynomial effects lack biological or mechanistic interpretation [@pinheiro2006] which limits their use in biomedical studies.
+However, polynomial fits have limited predictive power, cause bias on the boundaries of the covariates [@beck1998], and more importantly, their lack of biological or mechanistic interpretation limits their use in biomedical studies [@pinheiro2006].
 
@@
 Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976].
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)[@huynh1976;@greenhouse1959] the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001].
+However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time [@maxwell2017]; and is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004].
+Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)[@huynh1976;@greenhouse1959] their effectiveness is dependent on sample size and number of repeated measurements[@haverkamp2017], and it has been shown that corrections are not robust if the group sizes are unbalanced [@keselman2001].
 Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
@@
 
-## What do an rm-ANOVA and LMEM fit look like? A visual representation using simulated data{#simulation}
+## What does the fit of an rm-ANOVA and LMEM look like? A visual representation using simulated data{#simulation}
 
 To visually demonstrate the limitations of rm-ANOVA and LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \@ref(eq:linear-model)), and a LMEM  (Equation \@ref(eq:LMEM)) are fitted to each group, using R [@r] and the package _nlme_ [@nlme].
+An rm-ANOVA model (Equation \@ref(eq:linear-model)), and a LMEM  (Equation \@ref(eq:LMEM)) are fitted to each group using R [@r] and the package _nlme_ [@nlme].
 
 Briefly, two cases for the mean response for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response)A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response)C).
-In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments where a strong non-linear trend is present.
+In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response)D).
+In both the linear and quadratic simulated data, the groups start with the same mean value in order to simulate the expected temporal evolution of some physiological quantity, starting at a common initial condition.
 
@@
 From the mean response per group (linear or quadratic), the variability or "error" of individual responses  within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \@ref(fig:l-q-response) B,D).
+Thus, the response per subject at each timepoint in both the linear and quadratic simulation corresponds to the mean response per group plus the error (represented by the points in Figure \@ref(fig:l-q-response) A, D).
+**Move simulated to the front: the simulated response per subject**
 
@@
 Points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation.
+The rm-ANOVA model not only fails to pick the trend of the quadratic data (E) but also assigns a global estimate that does not take into account the between-subject variation.
 **C, F**: Estimates from the LMEM in the linear and quadratic case (subject: thin lines, population: thick lines) .
@@
 
-The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response)A).
-When the linearity and compound symmetry assumptions are met, the rm-ANOVA model approximates well the global trend by group (Figure \@ref(fig:l-q-response)B).
+The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data (\@ref(fig:l-q-response)B), as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response)A).
 Note that because the LMEM incorporates _random effects_, is able to provide estimates for each subject and a "population" estimate (Figure \@ref(fig:l-q-response)C).
@@
 
-This section has used simulation to better convey and visulize the limitations of linearity and correlation in the response in data with non-linear trends using an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
+This section has used simulation to better convey and visualize the limitations of linearity and correlation in the response in data with non-linear trends using an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
 Notice that the model misspecification is easily noticeable if the model fit and the response are visualized.
-In the following section, we provide a brief overview of linear models, general linear models and generalized linear mixed models before presenting the theory of GAMs.
+In the following section, we provide a brief overview of linear models, general linear models and generalized linear mixed models before presenting the theory of GAMs, a class of semi-parametric models that can fit non-linear trends in data and that overcome the limitations of rm-ANOVA and LMEMs in the analysis of biomedical longitudinal data.

### 2021-12-22 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Made changes of revisions, need to re-structure Appendix

@@
 In reality, rm-ANOVA and LMEMs are just _special cases_ of a broader class of models (General Linear Models and Generalized Linear Mixed Models, respectively).
-In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different classes of models, indicate how rm-ANOVA and LMEMs fit within this framework and introduce the theory of GAMs.
+In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different classes of models and indicate how rm-ANOVA, LMEMs, and GAMs fit within this framework.
 
@@
 
-where $y_{ijt}$ is the observation $i$ in group $j$ at time $t$, that is assumed to come from some distribution of the exponential family $\mathcal{D}$, with some mean $\mu_{ijt}$ and potentially, a dispersion parameter $\phi$ (which in the Gaussian case is the variance $\sigma^{2}$).
-The mean ($\mu_{ijt}$) is also known as the _expected value_ (or _expectation_), $E(y_{ijt})$ of the observed response $y_{ijt}$.
+where $y_{ijt}$ is the observation $i$ in group $j$ at time $t$, that is assumed to come from some distribution of the exponential family $\mathcal{D}$, with some mean $\mu_{ijt}$, and potentially, a dispersion parameter $\phi$ (which in the Gaussian case is the variance $\sigma^{2}$).
+The mean ($\mu_{ijt}$) is also known as the _expected value_ (or _expectation_) $E(y_{ijt})$ of the observed response $y_{ijt}$.
 
@@
 
-where $\eta_{ijt}$ is the linear predictor for each observation in each group at each timepoint, and $\beta_0$ (the intercept), $\beta_{1}, \beta_2,$ and $\beta_3$ are the model parameters for each group, which can be referred globally to as $\beta_j$.
-Finally, $time_{ijt}$ represents the covariates from each subject in each group at each time point.
+where $\eta_{ijt}$ is the linear predictor for each observation $i$ in each group $j$, at each timepoint $t$, and $\beta_0$ (the intercept), $\beta_{1}, \beta_2,$ and $\beta_3$ are the model parameters for each group for the linear mean from Equation (\@ref(eq:linear-model)).
+Finally, $time_{t}$ represents the covariates from each subject in each group at each time point.
 
@@
 \begin{equation}
-\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} + \beta_2 \times time_{t} + +\beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij},
+\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} + \beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij},
 (\#eq:GLMM-eta)
@@
 \begin{equation}
-\eta_{ijt}=\beta_{0}+ \beta_{1} \times treatment_{j} +f(time_{t}|\beta_{j}).
+\eta_{ijt}=\beta_{0}+ \beta_{1} \times treatment_{j} +f(time_{t}|\beta_{j}),
 (\#eq:GAM-eta)
@@
 
-Where $\beta_{0}$ is the intercept, and $\beta_{1}$ is the coefficient for each treatment group.
-Notice that the construction of the predictor is similar to that of Equation \@ref(eq:GLM-eta), but in this case the parametric terms involving the effect of time have been replaced by $f(time_{t}|\beta_{j})$, which represents the smooth function of time with with inputs as the covariates $time_t$ and parameters $\beta_{j}$.
-A GAM version of a linear model can be written as:
+where $\beta_{0}$ is the intercept, and $\beta_{1}$ is the coefficient for each treatment group.
+Notice that the construction of the predictor is similar to that of Equation \@ref(eq:GLM-eta), but in this case the parametric terms involving the effect of time, and the interaction between time and treatment have been replaced by $f(time_{t}|\beta_{j})$.
+The smooth term $f(time_{t}|\beta_{j})$ gives a different smooth response for each treatment.
+If the smooth term represented a linear relationship, then $f(time_{t}|\beta_{j})= \beta_2 \times time_t+\beta_3 \times time_t \times treatment_j$; however, in general, the smooth term is a more flexible function than a linear relationship, with parameter vectors $\beta_{j}$ for each treatment.
+A GAM version of a linear model can be written as
 
 \begin{equation}
-  y_{ijt}=\beta_0+f(time_t\mid \beta_j)+\varepsilon_{ijt}
+  y_{ijt}=\beta_0+ \beta_1 \times treatment_j + f(time_t\mid \beta_j)+\varepsilon_{ijt},
   (\#eq:GAM)
@@
 
-Where $y_{ijt}$ is the response at time $t$ of subject $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the smooth function $f(time_t\mid \beta_j)$,  and $\varepsilon_{ijt}$ represents the deviation of each observation from the mean.
+where $y_{ijt}$ is the response at time $t$ of subject $i$ in group $j$, and $\varepsilon_{ijt}$ represents the deviation of each observation from the mean.
 
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is more advantageous as it allows more flexibility because it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
+In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
 One possible set of functions for $f(time_t\mid \beta_j)$ that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with inference [@beck1998].
@@
 
-The smooth functional relationship between the covariates and the response in GAMs is specified  using a semi-parametric relationship that can be fit within the GLM framework, by using a _basis function_ expansion of the covariates and by estimating random coefficients associated with these basis functions.
+The smooth functional relationship between the covariates and the response in GAMs is specified  using a semi-parametric relationship that can be fit within the GLM framework, using a _basis function_ expansion of the covariates and estimating random coefficients associated with these basis functions.
 A _basis_ is a set of functions that spans the mathematical space within which the true but unknown $f(time_t)$  is thought to exist [@simpson2018].
@@
 
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness [@wegman1983].
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions that have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness [@wegman1983].
 Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
@@
 Cubic splines (CS) are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework [@hastie1987], but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as 'knots', the points where the basis functions meet) to obtain the finite basis, which could affect the model fit [@wood2003].
+The use of CS as smoothers in GAMs was discussed within the original GAM framework [@hastie1987], but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as 'knots', the points where the basis functions meet) to obtain the finite basis, which affects model fit [@wood2003].
 A solution to the "knot" placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate [@wood2003; @wood2017].
@@
 In contrast, thin plate regression splines (TPRS) provide a reasonable "low rank" (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner [@wood2003].
-Like TPS, TPRS only require the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood[@wood2003;@wood2017]).
+Like TPS, TPRS only requires specifying the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood[@wood2003;@wood2017]).
 
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \@ref(fig:l-q-response)C.
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \@ref(fig:l-q-response)D.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
@@
 
-With this in mind, if five basis functions are used to fit a GAM for the data that appears in Figure \@ref(fig:l-q-response)C, the resulting fitting process is shown in Figure \@ref(fig:basis-plot)A.
-The four basis functions (and the intercept) are shown.
-Each of the basis functions is evaluated at six different points (because there are six points on the timeline).
+With this in mind, if five basis functions are used to fit a GAM for the data of Group 1 (concave down function) that appears in Figure \@ref(fig:l-q-response)D, the resulting fitting process is shown in Figure \@ref(fig:basis-plot)A, where the four basis functions (and the intercept) are shown.
+Each of the five basis functions is evaluated at six different points (because there are six points on the timeline).
 The coefficients for each of the basis functions of Figure \@ref(fig:basis-plot)A are estimated using a penalized regression with smoothness parameter $\lambda$, that is estimated when fitting the model.
-The penalized estimates fitted for our example are shown in Figure \@ref(fig:basis-plot)B.
+The penalized coefficient estimates fitted for our example are shown in Figure \@ref(fig:basis-plot)B.
 
@@
 
-To get the weighted basis functions, each basis (from Figure Figure \@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients in Figure \@ref(fig:basis-plot)B, thereby increasing or decreasing the original basis functions.
+To get the weighted basis functions, each basis (from Figure \@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients in Figure \@ref(fig:basis-plot)B, thereby increasing or decreasing the original basis functions.
 Figure \@ref(fig:basis-plot)C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and less than 1).
+Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
 On the other hand, the third basis function has roughly doubled its contribution to the smooth term.
 Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of _time_ is shown in Figure \@ref(fig:basis-plot)D (orange line), along the simulated values per group, which appear as points.
+The resulting smooth term for the effect of _time_ is shown in Figure \@ref(fig:basis-plot)D (brown line), along the simulated values per group, which appear as points.
 
@@
 **C**: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
+Each of the five basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
 Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
 **D**: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.
+The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for Group 1 shown as points.
 
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge [@mcelreath2018].
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge and observed data [@mcelreath2018].
 In the case of data that shows non-linear trends, the prior that the _true_ trend of the data is likely to be smooth rather than "wiggly" introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs [@wood2017].
@@
 
-To understand this last part, it is worth reminding that a CI provides an estimate of the region where the “true” or “mean” value of a function exists, taking into account the randomness introduced by the sampling process.
+To understand across the function coverage, recall that a CI provides an estimate of the region where the “true” or “mean” value of a function exists, taking into account the randomness introduced by the sampling process.
 Because random samples from the population are used to calculate the "true" value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95%) where the function is expected to lie.
-In an _across the function_ CI (like those estimated by default for GAMs using _mgcv_) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95%).
-In other words, we expect that about 95% of the points that compose the true function will be covered by the across the function CI and for this to occur, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
+In an _across the function_ CI (like those estimated by default for GAMs using _mgcv_), if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95%).
+In other words, we expect that about 95% of the points that compose the true function will be covered by the across the function CI.
+As a consequence, some areas of CI for the function must have more than nominal coverage and some areas less than the nominal coverage.
 
-However, besides the across the function CI, "simultaneous" or "whole function" CIs can also be computed, which contain the _whole function_ with a specified probability [@wood2017].
+Besides the across the function CI, "simultaneous" or "whole function" CIs can also be computed, which contain the _whole function_ with a specified probability [@wood2017].
 Suppose we chose a nominal value (say, 95%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
-Briefly, to obtain a simultaneous CI we use simulation to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI [@ruppert2003].
-Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups, as we discuss in the following section.
+Briefly, to obtain a simultaneous CI we use simulation from the empirical Bayesian posterior distribution to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI [@ruppert2003].
+Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
 

### 2021-12-22 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Made changes of revisions, need to re-structure Appendix

@@
 The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors (Figure 3C in Vishwanath et.
-al.[@vishwanath2009]).
-In the paper, diffuse reflectance spectroscopy was used to quantify $\mbox{StO}_2$ changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
+al.[@vishwanath2009]), where diffuse reflectance spectroscopy was used to quantify $\mbox{StO}_2$ changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
 In the "Treatment" group (chemotherapy) an increase in $\mbox{StO}_2$ is observed through time, while a decrease is seen in the "Control" (saline) group.
 Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (matching the SD in the original paper).
-The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot)A.
+The simulation based on the real data appears in Figure \@ref(fig:sim-smooth-plot)A.
 
@@
 
-An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination.
-In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the _time_ effect to account for the repeated measures over time.
+An interaction effect is typically the main interest in longitudinal biomedical data, as the interaction takes into account treatment, time, and their combination.
+In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the _time_ effect for each treatment to account for the repeated measures over time.
 Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in Hefley [@hefley2017].
+A detailed description on the close relationship between basis functions and correlation functions can be found in Hefley et.
+al.
+[@hefley2017].
 
-For the data in Figure \@ref(fig:sim-smooth-plot)A the main effect of interest is how $\mbox{StO}_2$ changes over time for each treatment.
-To estimate this, the model incorporates separate smooths for _Group_ and _Day_, respectively.
+For the data in Figure \@ref(fig:sim-smooth-plot)A, the main effect of interest is how $\mbox{StO}_2$ changes over time for each treatment.
+To estimate this, the model incorporates separate smooths for each _Group_ as a function of _Day_.
 The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (*Day*) and the other is a factor (*Group*).
 Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the group means.
-Using $\textsf{R}$ and the package `mgcv` the model syntax is:
+Using R and the package `mgcv` the model syntax is:
 
 ```{r, eval = FALSE}
-gam_02 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
+gam_02 <- gam(StO2_sim ~ Group + s(Day, by = Group, k = 5), method ='REML', data = dat_sim)
 ```
 
-This syntax specifies that `gam_02` (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using four basis functions (plus the intercept).
+This syntax specifies that `gam_02` (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using five basis functions.
 The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths [@simpson2018] (a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console).
+Finally, the parametric term `Group` is added to quantify overall mean differences in the effect of treatment between groups, as we have indicated above.
 
-The parametric term `Group` is added to quantify overall mean differences in the effect of treatment between groups.
 Although the default `method` used to estimate the smoothing parameters in _mgcv_ is generalized cross validation (GCV), Wood[@wood2017] showed the restricted maximum likelihood (REML) to be more resistant to overfitting while also easing the quantification of uncertainty in the smooth parameters; therefore in this manuscript REML is always used for smooth parameter estimation.
 An additional argument (`family`) allows to specify the expected distribution of the response, but it is not used in this model because we expect a normally-distributed response (which is the default `family` in _mgcv_).
-
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot)B).
-Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_ [@gratia].
-A guide for model selection and diagnostics is in the [Appendix](#workflow), and an in-depth analysis can be found in Wood [@wood2017] and Harezlak [@harezlak2018].
-
-One question that might arise at this point is "what is the fit that an rm-ANOVA model produces for the simulated data?".
-The rm-ANOVA model, which corresponds to Equation \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot)C.
-This is a typical case of model misspecification: The slopes of each group are different, which would lead to a _p-value_ indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \@ref(fig:sim-smooth-plot)B) .
 
@@
 
-Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists.
-The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
-However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
-Consider the  simulated $\mbox{StO}_2$ values from Figure \@ref(fig:sim-smooth-plot)B.
-If 40% of the observations are randomly deleted and the same interaction GAM (`gam_02`) is fitted, the resulting smooths appear in (Figure \@ref(fig:sim-smooth-plot)D).
-Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.
-
-Additionally, note that in Figures \@ref(fig:sim-smooth-plot)B,D we show two CIs for each of the fitted smooths (shaded regions).
-The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, we expect them to be wider than the across the function CI, resulting in the wide shaded regions.
-For the dataset with missing observations, the smooths CIs overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset.
-However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.
-
 ```{r,missing-data,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
@@
                inherit.aes = FALSE)+
-    labs(y=expression(atop(StO[2],'(missing observations)')))+
+    labs(y=expression(atop(StO[2],'(incomplete observations)')))+
     scale_x_continuous(breaks=c(0,2,5,7,10))+
@@
 ```
+
+When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot)B).
+Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_ [@gratia].
+A guide for model selection and diagnostics is in the [Appendix](#workflow), and an in-depth analysis can be found in Wood [@wood2017] and Harezlak [@harezlak2018].
+
+One question that might arise at this point is "what is the fit that an rm-ANOVA model produces for the simulated data?".
+The fit of an rm-ANOVA model, which corresponds to Equation \@ref(eq:linear-model), is presented in Figure \@ref(fig:sim-smooth-plot)C.
+This is a typical case of model misspecification: The slopes of each group are different, which would lead to a _p-value_ indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \@ref(fig:sim-smooth-plot)B) .
+
+Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists.
+The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
+However, this usually causes the resulting smooths to have wider confidence intervals and less ability to discern differences in trends.
+To exemplify this, consider the random deletion of 40% of the simulated $\mbox{StO}_2$ values from Figure \@ref(fig:sim-smooth-plot)A.
+If the same interaction GAM (`gam_02`) is fitted to this data with missing observations, the resulting smooths appear in (Figure \@ref(fig:sim-smooth-plot)D).
+Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.
+
+Additionally, note that in Figure \@ref(fig:sim-smooth-plot)B,D we show two CIs for each of the fitted smooths (shaded regions).
+The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, they are wider than the across the function CI, resulting in the wide shaded regions.
+For the dataset with missing observations, the CIs for the smooths overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset.
+However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.
 
@@
 
+## Determination of significance in GAMs for longitudinal data{#GAM-significance}
+
+At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
+Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs can use a different approach to estimate significance.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in `gam_02`) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
+
 (ref:pairwise-comp) Pairwise comparisons for smooth terms.
@@
 
-## Determination of significance in GAMs for longitudinal data{#GAM-significance}
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in `gam_02`) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
-
 The absence of a _p-value_ in this context might seem odd, but its interpretation can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment.
@@
 
-Consider the calculation of pairwise differences for the smooths in Figure \@ref(fig:sim-smooth-plot)B and Figure \@ref(fig:sim-smooth-plot)D.
+Consider the calculation of pairwise differences for the smooths in Figure \@ref(fig:sim-smooth-plot)B, D.
 Figure \@ref(fig:plot-pairwise-comp) shows the comparison between each treatment group for the full and missing datasets with a simultaneous CI computed around the difference.

### 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.tex
New directory, knit separate appendices, separate directory for scripts

--- 2021-12-01 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.tex
@@
 \item
-  L325: We indicate that the term \(\varepsilon_{ijt}\) is the residual, and not a separable error.
-Of notice, we are following the notation provided by Dr.~Simpson in Eq.
-3 in the paper ``Modelling Modelling Palaeoecological Time Series Using Generalised Additive Models''.
-\textbf{do we need this here?
-I am adding it because we are following the same model specification that he has in this paper}
+  L325: We indicate that the term \(\varepsilon_{ijt}\) is the deivation of each observation from the mean.
 \end{itemize}
@@
 \begin{itemize}
-\tightlist
 \item
-  L396-L403 now contain a more detailed and accurate explanation of''across-the-function'' CIs.
+  L396-L404 now contain a more detailed and accurate explanation of''across-the-function'' CIs.
 We have used the reviewer's comments to indicate that in GAM fitted using \emph{mgcv}, the close to nominal across-the-function-coverage is true when averaged over the entire function.
+\item
+  L405-411 now provide a brief explanation of simultaneous CIs.
+With this, we want to provide the reader with an intuitive understanding on why simultaneous CIs are more robust than across the function CIs.
+\item
+  We have modified Figure 3 to now include across the function and simultaneous CIs.
+Figs.
+3B,D now show the fitted smooths and both intervals, and we explain their inclusion in L473-478.
+In this way, the reader has a visual understanding of the difference between across the function and simultaneous CIs.
 \end{itemize}
@@
 You would have to create differences conditional upon a level of
-the other factor. Users, however, tend to expect differences to include the group means, as you have chosen to present things here. Hence a note exp
+the other factor.
+Users, however, tend to expect differences to include the group means, as you have chosen to present things here.
+Hence a note explaining the implications of this choice would be useful to forewarn the user.
 
@@
 \subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-9}}
+
+We have made major changes to this section based on the reviewer's comments.
+
+\begin{itemize}
+\item
+  We have removed the sentence ``if the difference between the credible intervals of the fitted smooths\ldots{} is non-zero'', and now L484 reads: ``is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.''
+\item
+  L490 now reads ``unlikely to be distinguishable from zero''.
+\item
+  L493-502 now indicate that we are computing the difference using a simultaneous CI, and that we \emph{included} the group means in our pairwise comparison.
+We also indicate to the reader the fact that if multiple parametric terms exist in the model the inclusion of the means can become problematic.
+\end{itemize}
+
+However, we believe that including the means makes the comparison more intuitive, as in that way the researcher can directly compare both the fitted smooths and the pairwise comparisons to infer any significant changes and their associated magnitude.
+In the Appendix we also have made the corrections to the calculation of the pairwise comparisons in Section A.4.4.1 (the GAM workflow section) so the reader is aware again that in our code we keep the group means.
 

### 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/Full_document_SIM_No_Appendix.Rmd
New directory, knit separate appendices, separate directory for scripts

--- 2021-12-22 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.Rmd
@@
 
-```{r libraries, message = FALSE,include=FALSE}
+```{r setup, message = FALSE,include=FALSE}
 #the packages needed to run the code
@@
 library(gratia)
-library(viridis)
 library(scico)
@@
 options(bookdown.post.latex = remove_author)
+
 ```

### 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/Full_document_SIM_No_Appendix.tex
New directory, knit separate appendices, separate directory for scripts

--- 2021-12-22 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
@@
 In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs).
-Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies assume an expected trend in the measured response, which can results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
+Although LMEMs are less restrictive than rm-ANOVA in terms of assumptions about imbalanced observations and correlation between observations, both methodologies assume a linear trend in the measured response.
+It is common in biomedical research that the true trend response is nonlinear and in these cases the linearity assumption of rm-ANOVA and LMEMs can lead to biased estimates and unreliable inference.
 
@@
 Thus, the response per subject at each timepoint in both the linear and quadratic simulation corresponds to the mean response per group plus the error (represented by the points in Figure \ref{fig:l-q-response} A, D).
+\textbf{Move simulated to the front: the simulated response per subject}
 
@@
 
-This section has used simulation to better convey and visulize the limitations of linearity and correlation in the response in data with non-linear trends using an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
+This section has used simulation to better convey and visualize the limitations of linearity and correlation in the response in data with non-linear trends using an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
 Notice that the model misspecification is easily noticeable if the model fit and the response are visualized.
@@
 where \(\beta_{0}\) is the intercept, and \(\beta_{1}\) is the coefficient for each treatment group.
-Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time, and the interaction between time and treatment have been replaced by \(f(time_{t}|\beta_{j})\), which represents the smooth function of time with inputs as the covariates \(time_t\) and parameters \(\beta_{j}\).
-A GAM version of a linear model can be written as:
+Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time, and the interaction between time and treatment have been replaced by \(f(time_{t}|\beta_{j})\).
+The smooth term \(f(time_{t}|\beta_{j})\) gives a different smooth response for each treatment.
+If the smooth term represented a linear relationship, then \(f(time_{t}|\beta_{j})= \beta_2 \times time_t+\beta_3 \times time_t \times treatment_j\); however, in general, the smooth term is a more flexible function than a linear relationship, with parameter vectors \(\beta_{j}\) for each treatment.
+A GAM version of a linear model can be written as
 
 \begin{equation}
-  y_{ijt}=\beta_0+f(time_t\mid \beta_j)+\varepsilon_{ijt},
+  y_{ijt}=\beta_0+ \beta_1 \times treatment_j + f(time_t\mid \beta_j)+\varepsilon_{ijt},
   \label{eq:GAM}
@@
 
-where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the smooth function \(f(time_t\mid \beta_j)\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
+where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
 
@@
 
-Biomedical longitudinal data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
-Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}, where it is clear that these models do not capture the non-linear trend of the data, thereby causing a ``model misspecification error''.
-
+Biomedical longitudinal data is particularly challenging to analyze due to the frequency of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
+Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}.
+When these models do not capture the non-linear trend of the data, this results in a ``model misspecification error''.
 This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Polynomial effects can be used in LMEMs, and it could be argued that in Section \ref{simulation} a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects presents more questions than answers in this case (i.e., what is the appropriate polynomial effect to use?
-quadratic?, cubic?, what is the biological interpretation of a polynomial effect?).
-Moreover, polynomial effects have other associated issues that we have described in Section \ref{GAM-theory}.
-
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{66}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.
+
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{66}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) or \emph{p-value} value, whose interpretation is not clear in the case of a Type III error.
 
 In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research\textsuperscript{\protect\hyperlink{ref-hastie1995}{67}}.
-This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (e.g., which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research\textsuperscript{\protect\hyperlink{ref-hastie1995}{67}}.
+This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs.
+T
+
+The purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model, such as in the case of polynomial effects in LMEMs.
+In contrast, because GAMs let the data speak for themselves they can provide an accurate representation of the effect of time in a biological process.
 
 However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} using \(\textsf{R}\) and the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
-One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
-The model is therefore able to identify changes between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
-
-We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} using R and the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
+
+One of the features of GAMs is that their Bayesian interpretation allows for inference about differences between groups without the need of a \emph{p-value}, and in addition can provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
+The model is therefore able to identify changes between the groups at time points were data was not directly measured even in the case of incomplete observations ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
+
+We have used R as the software of choice for this paper because R not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
 In this regard, reproducibility is still an issue in biomedical research\textsuperscript{\protect\hyperlink{ref-begley2015}{68},\protect\hyperlink{ref-weissgerber2018}{69}}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.

### 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_A.Rmd
New directory, knit separate appendices, separate directory for scripts

--- 2021-12-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/07-Appendix.Rmd
@@
-# (APPENDIX) Appendix {-}
+---
+title: | 
+    | SUPPLEMENTARY MATERIALS for
+    |
+    | **Generalized additive models to analyze biomedical non-linear longitudinal data in R:**
+    | Beyond repeated measures ANOVA and Linear Mixed Models
+    | 
+    | APPENDIX A: WORKFLOW FOR GAMs FOR BIOMEDICAL LONGITUDINAL DATA 
+header-includes:
+    \usepackage{placeins}
+output:  
+  bookdown::pdf_document2:
+    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
+    pandoc_args: --listings #calls the listings package to fit code within the page margins
+    keep_tex: yes #keep LaTeX file for submission
+    fig_caption: yes #allows captions in figures
+    toc: false #do not include table of contents
+    extra_dependencies:
+      subfig: null #allows for subfigures
+      breqn: null #line breaks for long equations
+      caption: ["font={small}"] #size of the figure captions
+      float: null #allows for control of placement of figures
+    includes:
+      in_header: latex_docs/preamble_Appendix_A.sty #additional LaTeX formatting
+  bookdown::word_document2:
+    fig_caption: yes #figure caption
+    keep_md: yes
+  bookdown::html_document2:
+    css: "style.css" #style for the HTML document
+link-citations: yes #adds links to the citations
+'': default
+urlcolor: blue
+---
 
-<!-- allows for page numbering using letters for the Appendix -->
-\appendixpagenumbering
+\newpage
 
@@
 \counterwithin{figure}{section}
-# APPENDIX
 
-This section presents the code used to generate figures, models and simulated data from the main manuscript.
+This appendix shows a basic workflow to fit a series of increasingly complex GAMs to simulated data that follows the trends presented in Section 5.1 in the main manuscript.
+Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
+For simplicity, the confidence intervals (CIs) shown in this section for the models are the across the function CIs created by `mgcv` by default.
+However, for the pairwise comparisons of the third model we use _simultaneous intervals_ as in the main manuscript, and the code for creating simultaneous CIs for the smooths can be found in **Appendix XXX**.
 
-## Compound symmetry and independent errors in linear and quadratic responses
+## Setup and data simulation
 
-This section simulates linear and quadratic data in the same manner as in Section \@ref(simulation) in the main manuscript.
-The linear simulations using Figure \@ref(fig:linear-cases-Appendix) show in panels A and D the simulated mean responses and individual data points.
-Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
+First, we load the libraries needed for all analyses and figures. We also set seed for reproducibility.
 
-The following code chunks produce a more comprehensive exploration of Figure \@ref(fig:l-q-response) in the main manuscript.
+```{r,setup, warning=FALSE, message=FALSE}
+library(patchwork)
+library(tidyverse)
+library(mvnfast)
+library(nlme)
+library(mgcv)
+library(gratia)
+library(here)
+library(scico)
+set.seed(2021) #set seed for reproducibility
 
-First, a function is created to simulate data across six timepoints using a linear or quadratic mean response, with correlated or uncorrelated errors.
-Each group has a different slope/concavity.
-The main function is the same for both groups, but a change in the sign allows to invert the trend.
-
-```{r, responses-Appendix,include=TRUE,message=FALSE,echo=TRUE}
-##########Section for calculations###########
-
-<<testing>>
-## Example with linear response
-
-#This function simulates data using a linear or quadratic mean response and each with correlated
-#or uncorrelated errors. Each group has a different slope/concavity.
-example <- function(n_time = 6, #number of time points
-                    fun_type = "linear", #type of response
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  x <- seq(1,6, length.out = n_time)
-
-  #Create mean response matrix: linear or quadratic
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # quadratic response (non-linear)
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-
-  #create an array where individual observations per each time point for each group are to be stored. Currently using 10 observations per timepoint
-  y <- array(0, dim = c(length(x), 2, 10))
-
-  #Create array to store the "errors" for each group at each timepoint. The "errors" are the 
-  #between-group variability in the response.
-  errors <- array(0, dim = c(length(x), 2, 10))
-  #create an array where 10 observations per each time point for each group are to be stored
-
-  #The following loops create independent or correlated responses.
-To each value of mu (mean response per group) a randomly generated error (correlated or uncorrelated) is added and thus the individual response is created.
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors: variance covariance matrix
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-#labeling y and errors  
-  dimnames(y) <- list(time = x, 
-                      treatment = 1:2, 
-                      subject = 1:10)
-
-  dimnames(errors) <- list(time = x, 
-                           treatment = 1:2, 
-                           subject = 1:10)
-
-  #labeling the mean response
-  dimnames(mu) <- list(time = x, 
-                       treatment = 1:2)
-
-  #convert y, mu and errors to  dataframes with time, treatment and subject columns
-  dat <- as.data.frame.table(y, 
-                             responseName = "y")
-  dat_errors <- as.data.frame.table(errors, 
-                                    responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, 
-                                responseName = "mu")
-
-  #join the dataframes to show mean response and errors per subject
-  dat <- left_join(dat, dat_errors, 
-                   by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, 
-                   by = c("time", "treatment"))
-  #add time
-  dat$time <- as.numeric(as.character(dat$time))
-  #label subjects per group
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, 
-                                  treatment, 
-                                  sep = "-")))
-
-  ## repeated measures ANOVA
-
-  fit_anova <- lm(y ~ time + treatment + time * treatment, data = dat)
-
-#LMEM: time and treatment interaction model, compound symmetry 
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  #create a prediction frame where the model can be used for plotting purposes
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  #add model predictions to the dataframe that has the simulated data
-  dat$pred_anova <- predict(fit_anova)
-  dat$pred_lmem <- predict(fit_lme)
-
-  #return everything in a list
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_anova=fit_anova,
-    fit_lme = fit_lme
-  ))
-}
-
-##################Section for plotting#################################
-
-#This function will create the plots for either a "linear" or "quadratic" response
-
-plot_example <- function(sim_dat) {
-  ## Plot the simulated data (scatterplot)
-  txt<-20
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = treatment, 
-               color = treatment)
-           ) +
-    geom_point(show.legend=FALSE) +
-    labs(y='response')+
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment),
-              show.legend=FALSE) +
-    theme_classic() +
-    theme(plot.title = element_text(size = txt, 
-                                  face = "bold"),
-        text=element_text(size=txt))+
-    thm1
-
-  #plot the simulated data with trajectories per each subject
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = subject, 
-               color = treatment)
-           ) +
-    geom_line(aes(size = "Subjects"),
-              show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment,
-                  size = "Simulated Truth"), 
-              lty = 1,show.legend = FALSE) +
-    labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    theme_classic()+
-     theme(plot.title = element_text(size = txt, 
-                                face = "bold"),
-     text=element_text(size=txt))+
-    thm1
-
-  #plot the errors
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = errors,
-               group = subject, 
-               color = treatment)) +
-    geom_line(show.legend=FALSE) +
-     labs(y='errors')+
-     theme_classic()+
-     theme(plot.title = element_text(size = txt, 
-                                  face = "bold"),
-        text=element_text(size=txt))+
-    thm1
-
-   #plot the model predictions for rm-ANOVA
-  p4 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point(show.legend=FALSE)+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_anova), 
-                  group = subject, size = "Subjects"),show.legend = FALSE) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_anova, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population"),
-              show.legend=FALSE) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = txt, 
-                                  face = "bold"),
-        text=element_text(size=txt))+
-    thm1
-
-   #plot the LMEM predictions
-  p5 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point()+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), 
-                  group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_lme, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population")) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = txt, 
-                                  face = "bold"),
-        text=element_text(size=txt))+
-    thm1
-
-  return((p1+p3+p2+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-#Store each plot in a separate object
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
+#alpha for ribbon in the smooth plots
+al<-0.8
 
@@
 
-(ref:linear-cases-appendix) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are able to capture the trend of the data.
+Next, we create a data frame with the same trends from the main manuscript:
 
-```{r, linear-cases-Appendix, fig.width=14, fig.height=12, out.width='100%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:linear-cases-appendix)'}
-# linear response, correlated errors (subject effect)
-A1/B1+plot_annotation(tag_levels = 'A')
-```
+- Two treatment groups (Control and Treatment)
+- Five time points (days 0, 2, 5, 7,10)
+- Trends in $\mbox{StO}_2$ for both groups
 
-For the quadratic response case, Figure \@ref(fig:quadratic-cases-Appendix) shows the simulated responses using compound symmetry and independent errors.
-
-(ref:quadratic-caption-appendix) Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are not able to capture the changes in each group over time.
-
-```{r, quadratic-cases-Appendix, fig.width=14, fig.height=12, out.width='100%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:quadratic-caption-appendix)'}
-# linear response, correlated errors (subject effect)
-C1/D1+plot_annotation(tag_levels = 'A')
-```
-
-## Basis functions and GAMs
-
-This code produces Figure \@ref(fig:basis-plot) from the main manuscript.
-Briefly, a non-linear (quadratic) response is simulated, a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
-The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
-
-```{r,basis-Appendix, echo=TRUE,include=TRUE,message=FALSE,warning=FALSE,eval=FALSE}
-#generate the response: the same initial procedure from the previous section to simulate
-#the response
-set.seed(1)
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-
- #label each table
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
-
- #Convert to dataframes with subject, time and group columns
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
-
- #label subject per group
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
- #extract  "Group 1" to fit the GAM
-  dat<-subset(dat,treatment==1)
- #keep just the response and timepoint columns
-   dat<-dat[,c('y','time')]
-
-   #GAM model of time, 5 basis
-gm<-gam(y~s(time,k=5),data=dat)
-
-#model_matrix (also known as) 'design matrix'
-#will contain the smooths used to create  model 'gm'
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of basis by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be 
-#multiplied by the coefficients
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final smooth by aggregating the basis functions
-
-smooth<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(smooth=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,
-            aes(x=time,
-                y=value_orig,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(size=sz,
-            show.legend=FALSE)+
-  geom_point(size=sz+1,
-             show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,
-             labeller = as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#penalized basis
-p12<-ggplot(basis_plot,
-            aes(x=time,
-                y=mod_val,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(show.legend = FALSE,
-            size=sz)+
-  geom_point(show.legend = FALSE,
-             size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,
-             labeller=as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#heatmap of the coefficients
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,
-            aes(x=Basis,
-                y=Basis))+
-  geom_tile(aes(fill = cof), 
-            colour = "black") +
-    scale_fill_gradient(low = "white",
-                        high = "#B50A2AFF")+ #color picked from KikiMedium
-  labs(x='Basis',
-       y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),
-            size=7,
-            show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,
-            aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  labs(y='Simulated \n response')+
-  geom_line(data=smooth,
-            aes(x=time,
-                y=smooth),
-            color="#6C581DFF",
-            size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-
-```
-
-```{r,basis-plot-appendix, ref.label=I('basis-plot')}
-
-```
-
-## Longitudinal biomedical data simulation and GAMs{#tumor-data-simulation}
-
-This section describes how to fit GAMs to longitudinal data using simulated data.
-First, data is simulated according to Section \@ref(longitudinal-GAMs), where reported data of oxygen saturation ($\mbox{StO}_2$) in tumors under either chemotherapy or saline control is used as a starting point to generate individual responses in each group.
-
-```{r,sim-tumor-response-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE,eval=FALSE}
+```{r,data, echo= TRUE}
 dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
@@
             )
-#alpha for ribbon
-al<-0.8
+```
 
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
+We do not include a plot of the simulated data here because it follows the same trend from Figure 3A in the main manuscript (however, it can be easily plotted if desired).
+Finally, we call the function `simulate_data` in the script `simulate_data.R` to create 10 replicates at each time point from a normal distribution with a standard deviation of 10% in $\mbox{StO2}$.
 
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-
+```{r,simulation}
 n <- 10 #number of observations
 sd <- 10 #approximate sd from paper
-df <- 6
+source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","simulate_data.R"))
 dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, aes(x = Day, 
-                        y = StO2_sim, 
-                        color = Group, 
-                        group=subject)) +
-    geom_point(show.legend=FALSE,
-               size=1.5,
-               alpha=0.6)+
-    geom_line(size=0.6, alpha=0.4,show.legend=FALSE)+
-    geom_line(aes(x = Day, 
-               y = StO2,
-               color=Group),
-           size=1.5,
-           data=dat,
-           inherit.aes=FALSE,
-              show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],'(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    thm1
-
 ```
 
-## A basic Workflow for GAMs{#workflow}
-
-This section shows a basic workflow to fit a series of increasingly complex GAMs to the simulated data from the previous section.
-Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
-For simplicity, the confidence intervals (CIs) shown in this section for the models are the across the function CIs created by `mgcv` by default.
-However, for the pairwise comparisons of the third model we use _simultaneous intervals_ as in the main manuscript, and the code for creating simultaneous CIs for the smooths can be found in Section \@ref(GAM-linear-plot).
-
-### First model
+## First model
 
@@
 For model check, the functions `gam.check` and `summary` from  _mgcv_ provide detailed information about the model fit and its parameters.
-Keep in mind that `gam.check` is a function that also provides the graphical diagnostics obtained using _gratia_, if such graphical output is not desired the source code can be accessed typing `gam.check` in the Console, and the code without the graphical output can be used in a custom function, which is the approach we follow later).
-(ref:first-GAM-caption) Graphical diagnostics for the first GAM model.
-Left: Graphical diagnostics provided by the function `appraise` from the package _gratia_.
-Right: Fitted smooth for the model, provided by the function `draw`.
+Keep in mind that `gam.check` is a function that also provides the graphical diagnostics obtained using _gratia_, if such graphical output is not desired the source code can be accessed typing `gam.check` in the Console, and the code without the graphical output can be used in a custom function (which is the approach we follow later on this Appendix).
 
@@
 
-#### Graphical diagnostics
+### Graphical diagnostics
 
@@
 
+(ref:first-GAM-caption) Graphical diagnostics for the first GAM model.
+Left: Graphical diagnostics provided by the function `appraise` from the package _gratia_.
+Right: Fitted smooth for the model, provided by the function `draw`.
+
 From the output of the function `appraise` in Figure \@ref(fig:first-GAM-diag), the major indicators of concern about the model are the QQ plot of residuals and the histogram of residuals.
@@
 
-#### Model check{#gam-00-model-check}
+### Model check{#gam-00-model-check}
 
@@
 Because we plot the model diagnostics using `appraise` later, the graphical output from `gam.check` will be suppressed by creating a custom function to obtain just the model estimates, thus avoiding repetition of the diagnostic plots.
-This will be achieved by calling the source code of `gam.check` and using the appropriate code in a new function that will be called `gam.diagnostics`.
-We are not including in the Appendix the code for the function `gam.diagnostics` as it is rather long, but if desired it can be accessed by going to the `Appendix.Rmd` file in the GitHub repository and scrolling to this exact place (the code is not included in the final output, but is evaluated to create the function).
+This will be achieved by calling the source code of `gam.check` and using the appropriate code in a new function that will be called `gam_diagnostics`, which can be found in the script `gam_diagnostics.R`.
 
-```{r, first-GAM-check, fig.width=10, fig.height=7, out.width='75%', fig.align='center', echo=TRUE, include=FALSE, message=FALSE,warning=FALSE}
-
-#Function that uses the source code of `gam.check` to obtain the estimates without the plots. The source can be checked by typing `gam.check` in the console.
-gam.diagnostics<-function (b, old.style = FALSE, type = c("deviance", "pearson", 
-    "response"), k.sample = 5000, k.rep = 200, rep = 0, 
-    level = 0.9, rl.col = 2, rep.col = "gray80", ...) 
-{
-    type <- match.arg(type)
-    resid <- residuals(b, type = type)
-    linpred <- if (is.matrix(b$linear.predictors) && !is.matrix(resid)) 
-        napredict(b$na.action, b$linear.predictors[, 1])
-    else napredict(b$na.action, b$linear.predictors)
-
-    fv <- if (inherits(b$family, "extended.family")) 
-        predict(b, type = "response")
-    else fitted(b)
-    if (is.matrix(fv) && !is.matrix(b$y)) 
-        fv <- fv[, 1]
-    gamm <- !(b$method %in% c("GCV", "GACV", "UBRE", 
-        "REML", "ML", "P-ML", "P-REML", 
-        "fREML"))
-    if (gamm) {
-        cat("\n'gamm' based fit - care required with interpretation.")
-        cat("\nChecks based on working residuals may be misleading.")
-    }
-    else {
-        cat("\nMethod:", b$method, "  Optimizer:", 
-            b$optimizer)
-        if (!is.null(b$outer.info)) {
-            if (b$optimizer[2] %in% c("newton", "bfgs")) {
-                boi <- b$outer.info
-                cat("\n", boi$conv, " after ", boi$iter, 
-                  " iteration", sep = "")
-                if (boi$iter == 1) 
-                  cat(".")
-                else cat("s.")
-                cat("\nGradient range [", min(boi$grad), 
-                  ",", max(boi$grad), "]", sep = "")
-                cat("\n(score ", b$gcv.ubre, " & scale ", 
-                  b$sig2, ").", sep = "")
-                ev <- eigen(boi$hess)$values
-                if (min(ev) > 0) 
-                  cat("\nHessian positive definite, ")
-                else cat("\n")
-                cat("eigenvalue range [", min(ev), ",", 
-                  max(ev), "].\n", sep = "")
-            }
-            else {
-                cat("\n")
-                print(b$outer.info)
-            }
-        }
-        else {
-            if (length(b$sp) == 0) 
-                cat("\nModel required no smoothing parameter selection")
-            else {
-                cat("\nSmoothing parameter selection converged after", 
-                  b$mgcv.conv$iter, "iteration")
-                if (b$mgcv.conv$iter > 1) 
-                  cat("s")
-                if (!b$mgcv.conv$fully.converged) 
-                  cat(" by steepest\ndescent step failure.\n")
-                else cat(".\n")
-                cat("The RMS", b$method, "score gradient at convergence was", 
-                  b$mgcv.conv$rms.grad, ".\n")
-                if (b$mgcv.conv$hess.pos.def) 
-                  cat("The Hessian was positive definite.\n")
-                else cat("The Hessian was not positive definite.\n")
-            }
-        }
-        if (!is.null(b$rank)) {
-            cat("Model rank = ", b$rank, "/", length(b$coefficients), 
-                "\n")
-        }
-    }
-    cat("\n")
-    kchck <- k.check(b, subsample = k.sample, n.rep = k.rep)
-    if (!is.null(kchck)) {
-        cat("Basis dimension (k) checking results. Low p-value (k-index<1) may\n")
-        cat("indicate that k is too low, especially if edf is close to k'.\n\n")
-        printCoefmat(kchck, digits = 3)
-    }
-}
-```
-
-We now call `gam.diagnostics` to provide the desired diagnostic output, as well as a summary of the fitted model, which is obtained by calling `summary`.
+In the next code chunk, we call `gam.diagnostics` to provide the desired diagnostic output:
 
 ```{r,gam-00-diagnostics, echo=TRUE, include=TRUE}
-gam.diagnostics(gam_00)
+source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","gam_diagnostics.R"))
+gam_diagnostics(gam_00)
 
@@
 
-From the output, it can be seen that the `k-index` is 0.36, which indicates that the model is not capturing the variability in the data.
-The `edf` (effective degrees of freedom) is an indicator of the complexity of the smooth.
-Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
+And now obtain a summary of the fitted model, which is obtained by calling `summary`.
 
@@
 
+From the output of `gam_diagnostics`, it can be seen that the `k-index` is 0.35, which indicates that the model is not capturing the variability in the data.
+The `edf` (effective degrees of freedom) is an indicator of the complexity of the smooth.
+Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
+
 From the `summary` function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
@@
 Here, the _p-value_ is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to $\approx$ 17%.
+Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to $\approx$ 12%.
 
-### Second model
+## Second model
 
@@
 The next iteration is a model where a different smooth of time (Day) is assigned for each group using `by = Group` in the model syntax.
+Below we fit such model, which is saved as `gam_01`, and use `gam_diagnostiscs` to obtain information from the model fit.
 
@@
 
-gam.diagnostics(gam_01)
+source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","gam_diagnostics.R"))
+gam_diagnostics(gam_01)
 ```
 
-Diagnostics for this model indicate that the k-index is still below 1 (0.43 from `gam.check`), and that the residuals are still not following a normal distribution (Figure  \@ref(fig:second-GAM-diag)).
+Diagnostics for this model indicate that the k-index is still below 1 (0.58 from `gam_diagnostics` output), and that the residuals are still not following a normal distribution (Figure  \@ref(fig:second-GAM-diag)).
 Moreover, the smooths (plotted via the `draw()` function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
@@
 
-From `summary()`, the deviance explained by the model is $\approx$ 51%.
+From `summary()`, the deviance explained by the model is $\approx$ 65%.
 
@@
 
-### Third model
+## Third model
 
 Model `gam_00` was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted (a single smooth), the model diagnostics, and, the low variance explained by the model.
-On the other hand, `gam_01` takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \@ref(longitudinal-GAMs), in order to differentiate between each group a parametric term needs to be added to the model for the interaction of _Day_ and _Group_.
+On the other hand, `gam_01` takes into account the nesting within each group and provides better variance explanation, but as indicated in Section 5.2 in the main manuscript, in order to differentiate between each group a parametric term needs to be added to the model for the interaction of _Day_ and _Group_.
 
@@
 
-gam.diagnostics(gam_02)
+source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","gam_diagnostics.R"))
+
+gam_diagnostics(gam_02)
 
@@
 By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From `gam.check`, the k-index is now at an acceptable value ($\approx$ 1.02).
+From `gam_diagnostics`, the k-index is now at an acceptable value ($\approx$ **1.23**).
 
@@
 
-From `summary`, the model is able to capture 89% of the variance in the data, which is a substantial improvement over the variance explained by `gam_00` and `gam_01`.
+From `summary`, the model is able to capture 86.7% of the variance in the data, which is a substantial improvement over the variance explained by `gam_00` and `gam_01`.
 
@@
 
-The output in this case is expected: model `gam_02` has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
+The output in this case is expected: model `gam_02` has a lower AIC **(710.54)** whereas the initial two models have higher AICs (**884.43** and **805.30**).
 The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
 
-#### Pairwise comparisons of smooth confidence intervals
+## Pairwise comparisons of smooth confidence intervals
 
-The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \@ref(GAM-significance).
+The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in Section 5.3 in the main manuscript.
 
@@
 
-We want to emphasize that for the model used here we have included the group means in order to keep the pairwise comparisons on the scale of the response (a change from the code which appears in the chunk with a comment that says `###IMPORTANT`).
-This approach works well for models like `gam_02` but implementing this on models with more parametric terms can be challenging.
+We want to emphasize that for through the manuscript and for the comparisons on `gam_02`, we included the group means in order to keep the pairwise comparisons on the scale of the response.
+By default, pairwise comparisons available in other software packages (such as in G.
+Simpson's _gratia_) do not include the group means.
+However, we decided to include them because for our approach, it is easier to see the magnitude in the change of the difference between two treatment groups when the means are included.
+
+The change we allude to (mean group inclusion) can be found in the script `pointwise_comparisons.R`, at the line where we have added a comment that reads:
+
+```{r, eval=FALSE}
+######IMPORTANT: uncommenting the following two lines
+    #removes the group means from the comparison######
+```
+
+The inclusion of group means works well for models like `gam_02`, but implementing this on models with more parametric terms can be challenging and we want the reader to be aware of this.
+
 However, we do believe that the model presented in the paper covers a wide range of situations and our approach here for the pairwise comparisons will be useful for most biomedical researchers.
+We first call the script `pointwise_comparisons.R` to estimate the pointwise difference, then call the script `difference_smooths.R` which calculates the simultaneous confidence interval for the difference, and finally `pairwise_limits.R`, which estimates the regions to be shaded for the intervals with significance in the final figure.
+We use `difference_smooths` on model `gam_02` and save the estimates in `diff_complete`.
 
-```{r,pairwise-comp-workflow,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
-##Pairwise comparisons
-difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
-  ## make sure f1 and f2 are characters
-  f1 <-  as.character(f1)
-  f2 <-  as.character(f2)
-  cnames <- colnames(Xp)
-  ## columns of Xp associated with pair of smooths
-  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
-  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
-  ## rows of Xp associated with pair of smooths
-  r1 <- data[[by_var]] == f1
-  r2 <- data[[by_var]] == f2
+```{r,pairwise-comp-workflow-A,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
 
-  ## difference rows of Xp for pair of smooths
-  X <- Xp[r1, ] - Xp[r2, ]
-
-  ### IMPORTANT: Keep group means####
-  ## zero the cols related to other splines
- # X[, ! (c1 | c2)] <- 0
-
-  ## zero out the parametric cols
-  #X[, !grepl('^s\\(', cnames)] <- 0
-
-  ## compute difference
-  sm_diff <- drop(X %*% coefs)
-  se <- sqrt(rowSums((X %*% V) * X))
-  nr <- NROW(X)
-
-  ## Calculate posterior simulation for smooths
-  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
-  rownames(coefs_sim) <- rownames(V)
-  simDev <- X %*% coefs_sim
-  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
-  masd <- apply(absDev, 2, max)
-  crit_s <- quantile(masd, prob = 0.95, type = 8)
-
-  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
-              level_1 = rep(f1, nr),
-              level_2 = rep(f2, nr),
-              diff = sm_diff, se = se,
-              lower_s = sm_diff - crit_s * se,
-              upper_s = sm_diff + crit_s*se)
-
-  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
-  ## Only need rows associated with one of the levels
-  out <- bind_cols(out, data[r1, smooth_var])
-
-  out
-}
-
-#does both ci and si
-difference_smooths <- function(model,
-                                 smooth,
-                                 n = 100,
-                                 ci_level = 0.95,
-                                 newdata = NULL,
-                                 partial_match = TRUE,
-                                 unconditional = FALSE,
-                                 frequentist = FALSE,
-                                 nrep = 10000,
-                                 include_means = TRUE,
-                                 ...) {
-  if (missing(smooth)) {
-    stop("Must specify a smooth to difference via 'smooth'.")
-  }
-
-  # smooths in model
-  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
-  # select smooths
-  select <-
-    gratia:::check_user_select_smooths(smooths = S, select = smooth,
-                                       partial_match = partial_match)#,
-  # model_name = expr_label(substitute(object)))
-  sm_ids <- which(select)
-  smooths <- gratia::get_smooths_by_id(model, sm_ids)
-  sm_data <- map(sm_ids, gratia:::smooth_data,
-                 model = model, n = n, include_all = TRUE)
-  sm_data <- bind_rows(sm_data)
-  by_var <- by_variable(smooths[[1L]])
-  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
-  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
-                                   stringsAsFactor = FALSE))
-  names(pairs) <- paste0("f", 1:2)
-
-  Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
-  V <- gratia:::get_vcov(model, unconditional = unconditional,
-                         frequentist = frequentist)
-  coefs <- coef(model)
-
-  out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
-              smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
-              coefs = coefs, nrep = nrep)
-  out <- bind_rows(out)
-  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
-
-  out <- add_column(out,
-                    lower = out$diff - (crit * out$se),
-                    upper = out$diff + (crit * out$se),
-                    .after = 6L)
-  out
-}
-
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","pointwise_comparisons.R"))
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","difference_smooths.R"))
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","pairwise_limits.R"))
 #compute difference between smooths and calculate confidence interval: complete data
@@
                      n=100, partial_match = TRUE, nrep=10000)
+my_list<-pairwise_limits(diff_complete)
+```
 
-#function to obtain values for the shading regions
-pairwise_limits<-function(dataframe){
-    #extract values where the lower limit of the ribbon is greater than zero
-    #this is the region where the control group effect is greater
+Next, we plot the comparisons from `diff_complete` and create a `ggplot2` object (c1) so we can also shade the regions of significant differences easily.
 
-    v1<-dataframe%>%
-        filter(lower_s>0)%>%
-        select(Day)
-    #get day  initial value
-    init1=v1$Day[[1]]
-    #get day final value
-    final1=v1$Day[[nrow(v1)]]
-    #extract values where the value of the upper limit of the ribbon is lower than zero
-    #this corresponds to the region where the treatment group effect is greater
-    v2<-dataframe%>%
-        filter(upper_s<0)%>%
-        select(Day)
-
-    init2=v2$Day[[1]]
-    final2=v2$Day[[nrow(v2)]]
-    #store values
-    my_list<-list(init1=init1,
-                  final1=final1,
-                  init2=init2,
-                  final2=final2)
-    return(my_list)
-
-}
-
-my_list<-pairwise_limits(diff_complete)
+```{r,pairwise-comp-workflow-B,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
 rib_col<-'#8D7D82' #color for ribbon for confidence interval
@@
 
+The resulting plot (Figure \@ref(fig:pairwise-comp-workflow-fig)) shows the estimated difference in the smooths with a 95% simultaneous CI.
+As a reminder, because we have kept the group means, we can directly see that the Treatment starts to have a significant effect around day 3.
+As therapy progresses, the effect continues, and the magnitude of the difference ($\approx$ 40%) at day five corresponds directly with the magnitude in the increase in $\mbox{StO}_2$ in the group at the same time point in the data.
+
 (ref:pairwise-comp-caption) Smooth pairwise comparisons for model `gam_02` using a 95% simultaneous CI for the difference between smooths.
@@
 ```{r, pairwise-comp-workflow-fig, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-comp-caption)'}
+knitr::opts_chunk$set(fig.pos = "!H", out.extra = "")
 c1
 ```
+ \FloatBarrier
 
-Of notice, the package `gratia` contains a function that allows to do pairwise comparisons.
-In this package, `difference_smooths` is a function that makes the comparisons and produces Figure \@ref(fig:pairwise-comp-workflow-fig) when is used on a fitted model.
-The function syntax and an example can be found at:
+## Additional resources
 
-https://cran.r-project.org/web/packages/gratia/gratia.pdf
+Multiple and excellent resources are available for biomedical researchers that want to gain more insight on the theory and computation of GAMs.
+Here we provide a brief list of resources that cover additional topics as well as available packages that are worth considering:
 
-Keep in mind that this function removes the group means for the pairwise comparison so the result will not be directly interpretable with the scale of the measured response.
+- Gavin Simpson's package _gratia_, which provides convenient wrapper functions for plotting and pairwise comparisons. 
+CRAN page:
+[https://cran.r-project.org/web/packages/gratia/gratia.pdf](https://cran.r-project.org/web/packages/gratia/gratia.pdf)
 
-## GAM and Linear model plots and Missing data
+Vignette:
 
-This section covers the code used to generate Figure \@ref(fig:sim-smooth-plot), where the simulated data, fit of the "final" GAM (`gam_02`), linear model  and GAM on data with missing observations are presented.
-Note that panel A in Figure \@ref(fig:sim-smooth-plot) and the inset are generated in the code chunk where the data is simulated in Section \@ref(tumor-data-simulation), and are called later to build the figure.
+[https://gavinsimpson.github.io/gratia/](https://gavinsimpson.github.io/gratia/)
 
-### GAM and Linear model plots{#GAM-linear-plot}
+- Gavin Simpson's blog _From the bottom of the heap_, which covers a wide range of topics in GAM modeling, news about updates on _gratia_ and provides tutorials on GAM fitting.
 
-This code chunk creates panels B and C in Figure \@ref(fig:sim-smooth-plot).
-Note that this code uses the final GAM from the previous section (`gam_02`), so the simulated data and the model should be generated before running this section.
+Link:
 
-```{r, plotting-appendix, echo=TRUE, include=TRUE, message=FALSE, warning=FALSE,eval=FALSE}
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
+[https://fromthebottomoftheheap.net/](https://fromthebottomoftheheap.net/)
 
-## point-wise interval
-ci <- confint(gam_02, parm = "s(Day)", partial_match = TRUE, type = "confidence")
-## simultaneous interval
-si <- confint(gam_02, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
+- Matteo Fasiolo's package _mgcViz_, ane extension of the _mgcv_ package. Provides visual tools for Generalized Additive Models that exploit the additive structure of such models.
 
-# mean shift for Treatment group
-const <- coef(gam_02)[2]
+CRAN Page:
 
-#pointwise confidence interval
-ci <- ci %>%
-mutate(est = case_when(Group == "Treatment" ~ est + const,
-TRUE ~ est),
-lower = case_when(Group == "Treatment" ~ lower + const,
-TRUE ~ lower),
-upper = case_when(Group == "Treatment" ~ upper + const,
-TRUE ~ upper))
+[https://cran.r-project.org/web/packages/mgcViz/index.html](https://cran.r-project.org/web/packages/mgcViz/index.html)
 
-#simultaneous interval
-si <- si %>%
-mutate(est = case_when(Group == "Treatment" ~ est + const,
-TRUE ~ est),
-lower = case_when(Group == "Treatment" ~ lower + const,
-TRUE ~ lower),
-upper = case_when(Group == "Treatment" ~ upper + const,
-TRUE ~ upper))
+Vignette:
+[https://mfasiolo.github.io/mgcViz/articles/mgcviz.html](https://mfasiolo.github.io/mgcViz/articles/mgcviz.html)
 
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#plot smooths and pointwise and simulatenous confidence intervals for GAM
-f3<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
-geom_line(lwd = 1) +
-geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
-inherit.aes = FALSE, alpha = 0.7,
-show.legend=FALSE) +
-geom_ribbon(data = si,
-mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
-inherit.aes = FALSE, alpha = 0.4,
-show.legend=FALSE)+
-    geom_point(data=dat_sim, aes(x = Day, 
-                        y = StO2_sim, 
-                        color = Group), 
-                        size=1.5,
-                        alpha=0.6, 
-               inherit.aes = FALSE,
-               show.legend = FALSE)+
-    labs(y=expression(atop(StO[2],'(complete observations)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+thm1
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=al,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+thm1
-
-```
-
-## Working with Missing data in GAMs
-
-This code chunk first randomly deletes 40% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data.
-Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
-
-```{r,missing-data-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE,eval=FALSE}
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-ci <- confint(mod_m1, parm = "s(Day)", partial_match = TRUE, type = "confidence")
-## simultaneous interval
-si <- confint(mod_m1, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
-
-# mean shift for group 2
-const <- coef(mod_m1)[2]
-
-ci <- ci %>%
-mutate(est = case_when(Group == "Treatment" ~ est + const,
-TRUE ~ est),
-lower = case_when(Group == "Treatment" ~ lower + const,
-TRUE ~ lower),
-upper = case_when(Group == "Treatment" ~ upper + const,
-TRUE ~ upper))
-
-si <- si %>%
-mutate(est = case_when(Group == "Treatment" ~ est + const,
-TRUE ~ est),
-lower = case_when(Group == "Treatment" ~ lower + const,
-TRUE ~ lower),
-upper = case_when(Group == "Treatment" ~ upper + const,
-TRUE ~ upper))
-
-#plot model and original data
-f6<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
-geom_line(lwd = 1) +
-geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
-inherit.aes = FALSE, alpha = 0.7,
-show.legend=FALSE) +
-geom_ribbon(data = si,
-mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
-inherit.aes = FALSE, alpha = 0.4,
-show.legend=TRUE)+
-    geom_point(data=dat_missing, aes(x = Day, 
-                        y = StO2_sim, 
-                        color = Group), 
-                            size=1.5,
-                        alpha=0.6, 
-               inherit.aes = FALSE)+
-    labs(y=expression(atop(StO[2],'(missing observations)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22))+
-        thm1
-
-```
-
-```{r,sim-smooth-plot-Appendix, ref.label=I('sim-smooth-plot')}
-
-```
-
-## Pairwise comparisons in GAMs: full and missing data cases
-
-The next code chunk reproduces Figure \@ref(fig:plot-pairwise-comp).
-Here pairwise comparisons are made for the full and missing datasets by computing the difference between the smooths and a 95% simultaneous CI.
-
-```{r,pairwise-comp-Appendix,include=TRUE,message=FALSE,warning=FALSE, eval=FALSE}
-##Pairwise comparisons
-
-difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
-  ## make sure f1 and f2 are characters
-  f1 <-  as.character(f1)
-  f2 <-  as.character(f2)
-  cnames <- colnames(Xp)
-  ## columns of Xp associated with pair of smooths
-  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
-  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
-  ## rows of Xp associated with pair of smooths
-  r1 <- data[[by_var]] == f1
-  r2 <- data[[by_var]] == f2
-
-  ## difference rows of Xp for pair of smooths
-  X <- Xp[r1, ] - Xp[r2, ]
-
-  ## zero the cols related to other splines
- # X[, ! (c1 | c2)] <- 0
-
-  ## zero out the parametric cols
-  #X[, !grepl('^s\\(', cnames)] <- 0
-
-  ## compute difference
-  sm_diff <- drop(X %*% coefs)
-  se <- sqrt(rowSums((X %*% V) * X))
-  nr <- NROW(X)
-
-  ## Calculate posterior simulation for smooths
-  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
-  rownames(coefs_sim) <- rownames(V)
-  simDev <- X %*% coefs_sim
-  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
-  masd <- apply(absDev, 2, max)
-  crit_s <- quantile(masd, prob = 0.95, type = 8)
-
-  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
-              level_1 = rep(f1, nr),
-              level_2 = rep(f2, nr),
-              diff = sm_diff, se = se,
-              lower_s = sm_diff - crit_s * se,
-              upper_s = sm_diff + crit_s*se)
-
-  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
-  ## Only need rows associated with one of the levels
-  out <- bind_cols(out, data[r1, smooth_var])
-
-  out
-}
-
-#does both ci and si
-difference_smooths <- function(model,
-                                 smooth,
-                                 n = 100,
-                                 ci_level = 0.95,
-                                 newdata = NULL,
-                                 partial_match = TRUE,
-                                 unconditional = FALSE,
-                                 frequentist = FALSE,
-                                 nrep = 10000,
-                                 include_means = TRUE,
-                                 ...) {
-  if (missing(smooth)) {
-    stop("Must specify a smooth to difference via 'smooth'.")
-  }
-
-  # smooths in model
-  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
-  # select smooths
-  select <-
-    gratia:::check_user_select_smooths(smooths = S, select = smooth,
-                                       partial_match = partial_match)#,
-  # model_name = expr_label(substitute(object)))
-  sm_ids <- which(select)
-  smooths <- gratia::get_smooths_by_id(model, sm_ids)
-  sm_data <- map(sm_ids, gratia:::smooth_data,
-                 model = model, n = n, include_all = TRUE)
-  sm_data <- bind_rows(sm_data)
-  by_var <- by_variable(smooths[[1L]])
-  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
-  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
-                                   stringsAsFactor = FALSE))
-  names(pairs) <- paste0("f", 1:2)
-
-  Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
-  V <- gratia:::get_vcov(model, unconditional = unconditional,
-                         frequentist = frequentist)
-  coefs <- coef(model)
-
-  out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
-              smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
-              coefs = coefs, nrep = nrep)
-  out <- bind_rows(out)
-  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
-
-  out <- add_column(out,
-                    lower = out$diff - (crit * out$se),
-                    upper = out$diff + (crit * out$se),
-                    .after = 6L)
-  out
-}
-
-#compute difference between smooths and calculate confidence interval: complete data
-diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
-                     unconditional = TRUE, frequentist = FALSE,
-                     n=100, partial_match = TRUE, nrep=10000)
-
-#function to obtain values for the shading regions
-pairwise_limits<-function(dataframe){
-    #extract values where the lower limit of the ribbon is greater than zero
-    #this is the region where the control group effect is greater
-
-    v1<-dataframe%>%
-        filter(lower_s>0)%>%
-        select(Day)
-    #get day  initial value
-    init1=v1$Day[[1]]
-    #get day final value
-    final1=v1$Day[[nrow(v1)]]
-    #extract values where the value of the upper limit of the ribbon is lower than zero
-    #this corresponds to the region where the treatment group effect is greater
-    v2<-dataframe%>%
-        filter(upper_s<0)%>%
-        select(Day)
-
-    init2=v2$Day[[1]]
-    final2=v2$Day[[nrow(v2)]]
-    #store values
-    my_list<-list(init1=init1,
-                  final1=final1,
-                  init2=init2,
-                  final2=final2)
-    return(my_list)
-
-}
-
-my_list<-pairwise_limits(diff_complete)
-rib_col<-'#8D7D82' #color for ribbon for confidence interval
-control_rib <- '#875F79' #color for ribbon for control region
-treat_rib <- '#A7D89E' #color for ribbon treatment region
-
-c1 <- ggplot() +
-  geom_line(data = diff_complete, aes(x = Day, y = diff),size=1, alpha=0.5) +
-    annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill=control_rib,
-                alpha = 0.5,
-                )+
-    annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=6,
-           angle=90
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill=treat_rib,
-             alpha = 0.5
-             ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=6,
-           angle=90
-           )+
-  geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s, ymax = upper_s),
-              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
-  geom_hline(yintercept = 0, lty = 2, color = "red")+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    labs(y="Difference\n(Complete observations)")+
-    theme_classic()+
-    theme(
-    axis.text=element_text(size=22))
-
-###compute difference in smooths for missing data
-diff_missing <- difference_smooths(mod_m1, smooth = "s(Day)", newdata = newdat,
-                     unconditional = TRUE, frequentist = FALSE,
-                     n=100, partial_match = TRUE, nrep=10000)
-
-v2<- diff_missing %>%
-        filter(upper_s<0)%>%
-        select(Day)
-init2=v2$Day[[1]]
-final2=v2$Day[[nrow(v2)]]
-
-c2 <- ggplot() +
-  geom_line(data = diff_missing, aes(x = Day, y = diff),size=1, alpha=0.5) +
-    annotate("rect",
-             xmin =init2, xmax = final2,ymin=-Inf,ymax=Inf,
-             fill=treat_rib,
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=6,
-           angle=90
-           )+
-  geom_ribbon(data = diff_missing, aes(x = Day, ymin = lower_s, ymax = upper_s),
-              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
-  geom_hline(yintercept = 0, lty = 2, color = "red")+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    labs(y="Difference\n(Missing observations)")+
-    theme_classic()+
-    theme(
-    axis.text=element_text(size=22))
-
-pair_comp<-c1+c2
-```
-
-```{r,plot-pairwise-comp-Appendix,ref.label=I('plot-pairwise-comp')}
-
-```
+- The book "Generalized Additive Models: An Introduction with R" (the 2nd.
+Edition can be found [here](https://www.routledge.com/Generalized-Additive-Models-An-Introduction-with-R-Second-Edition/Wood/p/book/9781498728331)) by Simon Wood is an excellent resource for in-depth material on GAM theory, mathematical derivation, examples, and detailed descriptions on the computational aspect of _mgcv_.

### 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_A.tex
New directory, knit separate appendices, separate directory for scripts

--- 2021-12-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/SIM_Appendix.tex
@@
 \PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
 %
@@
 \hypersetup{
-  hidelinks,
+  colorlinks=true,
+  linkcolor={Maroon},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={blue},
   pdfcreator={LaTeX via pandoc}}
@@
 \setcounter{secnumdepth}{5}
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 %\usepackage{lineno}
@@
 
-%to have page numbers with letters in the Appendix:
-% %solution from https://tex.stackexchange.com/questions/59572/custom-page-numbering-for-appendix
-\newcommand{\appendixpagenumbering}{
-  \break
-  \pagenumbering{arabic}
-  \renewcommand{\thepage}{\thesection-\arabic{page}}
-}
+%start sections and page numbers with A
+\setcounter{page}{1}
+\renewcommand\thesection{A}
+\renewcommand\thesubsection{\thesection.\arabic{subsection}}
+\renewcommand{\thepage}{A-\arabic{page}}
 
@@
 \author{Ariel I. Mundo \orcidaffila{}}
-\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
+%\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
 \author{John R. Tipton \orcidaffilb{}}
-\affil{Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
+%\affil{Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
 \author{Timothy J. Muldoon*}
-\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
+%\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
 \affil{tmuldoon@uark.edu}
@@
 \usepackage[font={small}]{caption}
+\usepackage{float}
 \ifLuaTeX
@@
 
+\title{SUPPLEMENTARY MATERIALS for\\
+\strut \\
+\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
+Beyond repeated measures ANOVA and Linear Mixed Models\\
+\strut \\
+APPENDIX A: WORKFLOW FOR GAMs FOR BIOMEDICAL LONGITUDINAL DATA}
+\author{}
+\date{\vspace{-2.5em}}
+
 \begin{document}
@@
 
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
-
-In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs).
-Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
-
-In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
-Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
-This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We present the basic theory of GAMs, and using reported trends of oxygen saturation in tumors we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R.
-We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing).
-To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
-
-\hypertarget{keywords}{%
-\section*{Keywords}\label{keywords}}
-\addcontentsline{toc}{section}{Keywords}
-
-cancer biology; tumor response; generalized additive models; simulation; R
-
-\hypertarget{background}{%
-\section{Background}\label{background}}
-
-Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response {[}\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}{]}, antibody expression {[}\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}{]}, and cell metabolism {[}\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}\protect\hyperlink{ref-greening2018}{9}{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}\protect\hyperlink{ref-sio2016}{10}{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}\protect\hyperlink{ref-kamstra2015}{11}{]}.
-
-Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}\protect\hyperlink{ref-wagenmakers2008}{12}{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}{]}.
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
-
-A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
-Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from data that shows non-linear trends, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}.
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}\protect\hyperlink{ref-wang2019}{18}{]}.
-
-Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}{]}, consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}\protect\hyperlink{ref-abdi2010}{21}{]}), but they in turn reduce statistical power (1-\(\beta\)){[}\protect\hyperlink{ref-nakagawa2004}{22}{]}, and lead to increased Type II error (failing to reject the null hypothesis when it is false) {[}\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}{]}.
-Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and group sizes {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}\protect\hyperlink{ref-charan2013}{30}{]} which might cause the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible.
-It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}\protect\hyperlink{ref-gueorguieva2004}{13}{]}.
-This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
-At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
-
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}.
-And even more importantly, LMEMs also assume by default a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]} (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the trend of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
-In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]}.
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]}, and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
-
-The current advances in programming languages designed for statistical analysis (specifically R), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, R {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, R has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
-Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
-
-This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding an expected trend of the response, constant correlation structures, and missing observations are explained in detail.
-Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in R, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
-
-In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models to improve the standards for reproducibility in biomedical research.
-
-\hypertarget{challenges-presented-by-longitudinal-studies}{%
-\section{Challenges presented by longitudinal studies}\label{challenges-presented-by-longitudinal-studies}}
-
-\hypertarget{the-repeated-measures-anova-and-linear-mixed-model}{%
-\subsection{The repeated measures ANOVA and Linear Mixed Model}\label{the-repeated-measures-anova-and-linear-mixed-model}}
-
-The \emph{repeated measures analysis of variance} (rm-ANOVA) and the \emph{linear mixed model} (LMEM) are the most commonly used statistical analysis for longitudinal data in biomedical research.
-These statistical methodologies require certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1) an assumed relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
-Each one of these assumptions is discussed below.
-
-\hypertarget{assumed-relationship}{%
-\subsection{Assumed relationship}\label{assumed-relationship}}
-
-\hypertarget{the-repeated-measures-anova-case}{%
-\subsubsection{The repeated measures ANOVA case}\label{the-repeated-measures-anova-case}}
-
-In a longitudinal biomedical study, two or more groups of subjects (e.g., human subject, mice, samples) are subject to different treatments (e.g., a ``treatment'' group receives a novel drug or intervention vs.~a ``control'' group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response is modeled with \emph{fixed} components.
-The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug/intervention in the ``treatment'' group.
-
-Mathematically speaking, a rm-ANOVA model with an interaction can be written as
-
-\begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times treatment_{j} +\beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
-\label{eq:linear-model}
-\end{equation}
-
-In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of treatment (\(treatment_j\)), time (\(time_t\)), and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
-Independent errors \(\varepsilon_{ijt}\) represent random variation from the sampling process assumed to be \(\stackrel{i.i.d.}\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
-In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug,'' or ``saline'' vs.~``chemotherapy'').
-Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B).
-With this notation, the linear model then can be expressed as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_2\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
-\beta_0 + \beta_1+\beta_2 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
-\end{cases}
-\label{eq:ANOVA-by-group}
-\end{equation}
-
-To further simplify the expression, substitute \(\widetilde{\beta_{0}}=\beta_0+\beta_{1}\) and \(\widetilde{\beta_{1}}=\beta_{2}+\beta_{3}\) in the equation for Group B.
-This substitution allows for a different intercept and slope for Groups A and B.
-The model is then written as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_2\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
-\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
-\end{cases}
-\label{eq:ANOVA-lines}
-\end{equation}
-
-Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response.
-This means that either presented as Equations \eqref{eq:linear-model}, \eqref{eq:ANOVA-by-group} or \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear trends, the rm-ANOVA model will approximate this behavior with non-parallel lines.
-
-\hypertarget{LMEM-case}{%
-\subsubsection{The Linear Mixed Model (LMEM) Case}\label{LMEM-case}}
-
-A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to account for{[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}.
-A LMEM with interaction between time and treatment for a longitudinal study can be written as
-
-\begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times treatment_{j}+ \beta_2 \times time_{t} + \beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij} +\varepsilon_{ijt}\\.
-\label{eq:LMEM}
-\end{equation}
-
-When Equations \eqref{eq:linear-model} and \eqref{eq:LMEM} are compared, it is noticeable that LMEMs and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\alpha_{ij}\)).
-This term \(\alpha_{ij}\) corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)).
-The \emph{random} component can also be understood as modeling some ``noise'' in the response, but that does not arise from the sampling error term \(\varepsilon_{ijt}\) from Equations \eqref{eq:linear-model} through \eqref{eq:ANOVA-lines}.
-
-For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data that needs to be accounted for.
-As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
-The random effect \(\alpha_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence, the \emph{random effect} in a LMEM enables fitting models with different intercepts at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
-
-Of note, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach (using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}), which is described in detail by Pinheiro and Bates {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates {[}\protect\hyperlink{ref-beck1998}{36}{]}; more importantly, polynomial effects lack biological or mechanistic interpretation {[}\protect\hyperlink{ref-pinheiro2006}{15}{]} which limits their use in biomedical studies.
-
-\hypertarget{covariance-in-rm-anova-and-lmems}{%
-\subsection{Covariance in rm-ANOVA and LMEMs}\label{covariance-in-rm-anova-and-lmems}}
-
-In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{41}{]} (For an in-depth analysis of the covariance matrix see West{[}\protect\hyperlink{ref-west2014}{40}{]} and Weiss{[}\protect\hyperlink{ref-weiss2005}{42}{]}).
-
-In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}{]}.
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}\protect\hyperlink{ref-maxwell2017}{44}{]}; and it is higher between consecutive measurements {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and they are not robust if the group sizes are unbalanced {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
-Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
-
-\hypertarget{unbalanced-data}{%
-\subsection{Unbalanced data}\label{unbalanced-data}}
-
-In a longitudinal study, it is frequently the case that the number of observations is different across subjects.
-In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing,'' ``incomplete,'' or (more generally speaking) unbalanced data {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
-The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{46}{]}.
-
-On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are \emph{missing at random} (MAR) or \emph{completely missing at random} (MCAR) {[}\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}{]}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{47}{]}.
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{48}{]}.
-Missing observations can also be \emph{missing not at random} (MNAR) and in the case the missing observations are dependent on their value.
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
-
-However, it is worth reminding that ``all models are wrong'' {[}\protect\hyperlink{ref-box1976}{49}{]} and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
-Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.
-
-\hypertarget{simulation}{%
-\subsection{What do an rm-ANOVA and LMEM fit look like? A visual representation using simulated data}\label{simulation}}
-
-To visually demonstrate the limitations of rm-ANOVA and LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using R {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{50}{]}.
-
-Briefly, two cases for the mean response for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}C).
-In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments where a strong non-linear trend is present.
-
-Specifically, the rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect.
-In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group.
-From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D).
-
-A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where a simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) is presented.
-We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with ``default'' (linear) effects and, 2) demonstrate that a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
-In Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/l-q-response-1}
-
-}
-
-\caption{Simulated responses from two groups with correlated errors using a LMEM and a rm-ANOVA model.
-Top row: linear response, bottom row: quadratic response.
-\textbf{A}: Simulated linear data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
-\textbf{D}: Simulated quadratic data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
-\textbf{B,E}: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
-Points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation.
-\textbf{C, F}: Estimates from the LMEM in the linear and quadratic case (subject: thin lines, population: thick lines) .
-The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data and grossly bias the initial estimates for each group in the quadratic case (bottom row).}\label{fig:l-q-response}
-\end{figure}
-
-The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response}A).
-When the linearity and compound symmetry assumptions are met, the rm-ANOVA model approximates well the global trend by group (Figure \ref{fig:l-q-response}B).
-Note that because the LMEM incorporates \emph{random effects}, is able to provide estimates for each subject and a ``population'' estimate (Figure \ref{fig:l-q-response}C).
-
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}D.
-Here, the mean response per group was simulated using a quadratic function, and errors and individual responses were produced as in Figure \ref{fig:l-q-response}A.
-The mean response in the simulated data with quadratic behavior changes in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model (Equation \eqref{eq:linear-model}) or a LMEM (Equation \eqref{eq:LMEM}) to this data produces the fit that appears in Figure \ref{fig:l-q-response}E, F.
-
-Comparing the fitted responses of the LMEM and the rm-ANOVA models used in the simulated quadratic data (Figure \ref{fig:l-q-response}E, F) indicates that the models are not capturing the changes within each group.
-Specifically, note that the fitted mean response of both models shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.
-
-The LMEM is only able to account for between-subject variation by providing estimates for each subject (Figure \ref{fig:l-q-response}F), but both models are unable to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare Figure \ref{fig:l-q-response}D with Figure \ref{fig:l-q-response}E, F).
-If such a change has important physiological implications, both rm-ANOVA and LMEMs omit it from the fitted mean response.
-Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
-It could be argued that a LMEM with quadratic effects should have been used to fit the data in Figure\ref{fig:l-q-response}F.
-However, because in reality the true function is not known, choosing a polynomial degree causes more questions (e.g., is it quadratic?, cubic?, or a higher degree?).
-Additionally, polynomial effects have other limitations, which we cover in Section \ref{GAM-theory}.
-
-This section has used simulation to better convey and visulize the limitations of linearity and correlation in the response in data with non-linear trends using an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
-Notice that the model misspecification is easily noticeable if the model fit and the response are visualized.
-In the following section, we provide a brief overview of linear models, general linear models and generalized linear mixed models before presenting the theory of GAMs.
-
-\hypertarget{linear-models-and-beyond}{%
-\section{Linear Models, and beyond}\label{linear-models-and-beyond}}
-
-Linear models (LMs) are those that assume a normal (Gaussian) distribution of the errors, and only incorporate \emph{fixed effects} (such as rm-ANOVA).
-These are by far the models most commonly used to analyze data within the biomedical research community.
-On the other hand, Linear Mixed Effect Models (LMEMs) also incorporate \emph{random effects}, as it has been described in Section \ref{LMEM-case}.
-
-In reality, rm-ANOVA and LMEMs are just \emph{special cases} of a broader class of models (General Linear Models and Generalized Linear Mixed Models, respectively).
-In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different classes of models, indicate how rm-ANOVA and LMEMs fit within this framework and introduce the theory of GAMs.
-
-\hypertarget{generalized-linear-models-glms}{%
-\subsection{Generalized Linear Models (GLMs)}\label{generalized-linear-models-glms}}
-
-A major limitation of LMs is their assumption of normality in the errors.
-If the residuals are non-normal, a transformation is necessary in order to properly fit the model.
-However, transformation can lead to poor model performance {[}\protect\hyperlink{ref-ohara2010}{51}{]}, and can cause problems with the biological interpretation of the model estimates.
-McCullagh and Nelder {[}\protect\hyperlink{ref-nelder1972}{52}{]} introduced General Linear Models (GLMs) as an extension of LMs, where the errors do not need to be normally distributed.
-To achieve this, consider the following model
-
-\begin{equation}
-y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi),
-\label{eq:GLM-y}
-\end{equation}
-
-where \(y_{ijt}\) is the observation \(i\) in group \(j\) at time \(t\), that is assumed to come from some distribution of the exponential family \(\mathcal{D}\), with some mean \(\mu_{ijt}\) and potentially, a dispersion parameter \(\phi\) (which in the Gaussian case is the variance \(\sigma^{2}\)).
-The mean (\(\mu_{ijt}\)) is also known as the \emph{expected value} (or \emph{expectation}), \(E(y_{ijt})\) of the observed response \(y_{ijt}\).
-
-Then, the \emph{linear predictor} \(\eta\), which defines the relationship between the mean and the covariates can be defined as
-
-\begin{equation}
-\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} +\beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j},
-\label{eq:GLM-eta}
-\end{equation}
-
-where \(\eta_{ijt}\) is the linear predictor for each observation in each group at each timepoint, and \(\beta_0\) (the intercept), \(\beta_{1}, \beta_2,\) and \(\beta_3\) are the model parameters for each group, which can be referred globally to as \(\beta_j\).
-Finally, \(time_{ijt}\) represents the covariates from each subject in each group at each time point.
-
-Finally,
-
-\begin{equation}
-E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt}),
-\label{eq:GLM-Expectation}
-\end{equation}
-
-where \(E(y_{ijt})\) is the expectation, and \(g^{-1}\) is the inverse of a \emph{link function} (\(g\)).
-The link function transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}.
-Therefore, it can be seen that LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
-
-\hypertarget{generalized-linear-mixed-models-glmms}{%
-\subsection{Generalized linear mixed models (GLMMs)}\label{generalized-linear-mixed-models-glmms}}
-
-Although GLMs relax the normality assumption, they only accommodate fixed effects.
-Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate \emph{random effects}, which have an associated probability distribution {[}\protect\hyperlink{ref-mcculloch2001}{53}{]}.
-Therefore, in GLMMs the linear predictor takes the form
-
-\begin{equation}
-\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} + \beta_2 \times time_{t} + +\beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij},
-\label{eq:GLMM-eta}
-\end{equation}
-
-where \(\alpha_{ij}\) corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of Equation \eqref{eq:GLM-eta}.
-Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed {[}\protect\hyperlink{ref-nelder1972}{52}{]}, and GLMs are a special case of GLMMs where there are no random effects.
-In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in Dobson {[}\protect\hyperlink{ref-dobson2008}{54}{]} and Stroup {[}\protect\hyperlink{ref-stroup2013}{55}{]}.
-
-\hypertarget{GAM-theory}{%
-\subsubsection{GAMs as a special case of Generalized Linear Models}\label{GAM-theory}}
-
-\hypertarget{gams-and-basis-functions}{%
-\paragraph{GAMs and Basis Functions}\label{gams-and-basis-functions}}
-
-Notice that in the previous sections, the difference between GLMs and GLMMs resides on their linear predictors (Equations \eqref{eq:GLM-eta}, \eqref{eq:GLMM-eta}).
-Generalized additive models (GAMs) are an extension of the GLM family that allow the estimation of smoothly varying trends where the relationship between the covariates and the response is modeled using \emph{smooth functions} {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}{]}.
-In a GAM, the linear predictor has the form
-
-\begin{equation}
-\eta_{ijt}=\beta_{0}+ \beta_{1} \times treatment_{j} +f(time_{t}|\beta_{j}).
-\label{eq:GAM-eta}
-\end{equation}
-
-Where \(\beta_{0}\) is the intercept, and \(\beta_{1}\) is the coefficient for each treatment group.
-Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time have been replaced by \(f(time_{t}|\beta_{j})\), which represents the smooth function of time with with inputs as the covariates \(time_t\) and parameters \(\beta_{j}\).
-A GAM version of a linear model can be written as:
-
-\begin{equation}
-  y_{ijt}=\beta_0+f(time_t\mid \beta_j)+\varepsilon_{ijt}
-  \label{eq:GAM}
-\end{equation}
-
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the smooth function \(f(time_t\mid \beta_j)\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is more advantageous as it allows more flexibility because it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(time_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
-In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(time_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
-
-The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using a \emph{basis function} expansion of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(treatment_j\), \(time_t\), and \(time_t \times treatment_j\).
-The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{57}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
-In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(time_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
-
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{58}{]}.
-Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
-
-Cubic splines (CS) are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework {[}\protect\hyperlink{ref-hastie1987}{56}{]}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the basis functions meet) to obtain the finite basis, which could affect the model fit {[}\protect\hyperlink{ref-wood2003}{59}{]}.
-A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}{]}.
-
-In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner {[}\protect\hyperlink{ref-wood2003}{59}{]}.
-Like TPS, TPRS only require the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood{[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}{]}).
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-A computational requisite in \emph{mgcv} is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
-Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth.
-It is important to note that is not necessary to specify a number of basis equal to the number of unique values in the independent variable; fewer basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data.
-
-Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present.
-A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline.
-The second derivative of the spline is a measure of curvature, or the rate of change of the slope {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}, and increasing the penalty by increasing \(\lambda\) results in models with less curvature.
-As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
-In other words,a low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
-
-With this in mind, if five basis functions are used to fit a GAM for the data that appears in Figure \ref{fig:l-q-response}C, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A.
-The four basis functions (and the intercept) are shown.
-Each of the basis functions is evaluated at six different points (because there are six points on the timeline).
-The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), that is estimated when fitting the model.
-The penalized estimates fitted for our example are shown in Figure \ref{fig:basis-plot}B.
-
-To get the weighted basis functions, each basis (from Figure Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
-Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and less than 1).
-On the other hand, the third basis function has roughly doubled its contribution to the smooth term.
-Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (orange line), along the simulated values per group, which appear as points.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/basis-plot-1}
-
-}
-
-\caption{Basis functions for a single smoother for time.
-\textbf{A}: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-\textbf{B}: Matrix of basis function weights.
-Each basis function is multiplied by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-\textbf{C}: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
-Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-\textbf{D}: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
-\end{figure}
-
-\hypertarget{a-bayesian-interpretation-of-gams}{%
-\subsubsection{A Bayesian interpretation of GAMs}\label{a-bayesian-interpretation-of-gams}}
-
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge {[}\protect\hyperlink{ref-mcelreath2018}{60}{]}.
-In the case of data that shows non-linear trends, the prior that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language) {[}\protect\hyperlink{ref-miller2019}{61}{]}.
-Therefore, the confidence intervals (CIs) calculated by default for the smooth terms using \emph{mgcv} are considered empirical Bayesian credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good \emph{across the function} (``frequentist'') coverage{[}\protect\hyperlink{ref-wood2017}{37}{]}.
-
-To understand this last part, it is worth reminding that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function exists, taking into account the randomness introduced by the sampling process.
-Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie.
-In an \emph{across the function} CI (like those estimated by default for GAMs using \emph{mgcv}) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%).
-In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI and for this to occur, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
-
-However, besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-Suppose we chose a nominal value (say, 95\%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
-Briefly, to obtain a simultaneous CI we use simulation to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI {[}\protect\hyperlink{ref-ruppert2003}{62}{]}.
-Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups, as we discuss in the following section.
-
-In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller {[}\protect\hyperlink{ref-miller2019}{61}{]}, Wood{[}\protect\hyperlink{ref-wood2017}{37}{]}, Simpson {[}\protect\hyperlink{ref-simpson2018}{34}{]}, Marra {[}\protect\hyperlink{ref-marra2012}{63}{]}, and Ruppert {[}\protect\hyperlink{ref-ruppert2003}{62}{]}.
-With this brief introduction to the Bayesian interpretation of GAMs, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends.
-
-\hypertarget{longitudinal-GAMs}{%
-\section{The analyisis of longitudinal biomedical data using GAMs}\label{longitudinal-GAMs}}
-
-The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs.
-This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}.
-The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
-
-\hypertarget{simulated-data}{%
-\subsection{Simulated data}\label{simulated-data}}
-
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors (Figure 3C in Vishwanath et.
-al.{[}\protect\hyperlink{ref-vishwanath2009}{16}{]}).
-In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
-In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group.
-Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
-The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}A.
-
-\hypertarget{an-interaction-gam-for-longitudinal-data}{%
-\subsection{An interaction GAM for longitudinal data}\label{an-interaction-gam-for-longitudinal-data}}
-
-An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination.
-In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time.
-Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in Hefley {[}\protect\hyperlink{ref-hefley2017}{57}{]}.
-
-For the data in Figure \ref{fig:sim-smooth-plot}A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
-To estimate this, the model incorporates separate smooths for \emph{Group} and \emph{Day}, respectively.
-The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the group means.
-Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:
-
-\begin{lstlisting}[language=R]
-gam_02 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
-\end{lstlisting}
-
-This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using four basis functions (plus the intercept).
-The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]} (a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console).
-
-The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups.
-Although the default \passthrough{\lstinline!method!} used to estimate the smoothing parameters in \emph{mgcv} is generalized cross validation (GCV), Wood{[}\protect\hyperlink{ref-wood2017}{37}{]} showed the restricted maximum likelihood (REML) to be more resistant to overfitting while also easing the quantification of uncertainty in the smooth parameters; therefore in this manuscript REML is always used for smooth parameter estimation.
-An additional argument (\passthrough{\lstinline!family!}) allows to specify the expected distribution of the response, but it is not used in this model because we expect a normally-distributed response (which is the default \passthrough{\lstinline!family!} in \emph{mgcv}).
-
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{64}{]}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood {[}\protect\hyperlink{ref-wood2017}{37}{]} and Harezlak {[}\protect\hyperlink{ref-harezlak2018}{65}{]}.
-
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
-This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
-
-Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists.
-The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
-However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
-Consider the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}B.
-If 40\% of the observations are randomly deleted and the same interaction GAM (\passthrough{\lstinline!gam\_02!}) is fitted, the resulting smooths appear in (Figure \ref{fig:sim-smooth-plot}D).
-Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.
-
-Additionally, note that in Figures \ref{fig:sim-smooth-plot}B,D we show two CIs for each of the fitted smooths (shaded regions).
-The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, we expect them to be wider than the across the function CI, resulting in the wide shaded regions.
-For the dataset with missing observations, the smooths CIs overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset.
-However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/sim-smooth-plot-1}
-
-}
-
-\caption{Simulated data and smooths for oxygen saturation in tumors.
-\textbf{A}: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-\textbf{B}: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.
-\textbf{C}: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-\textbf{D}: Smooths for the GAM model for the simulated data with missing observations (40\%).
-Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.}\label{fig:sim-smooth-plot}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/plot-pairwise-comp-1}
-
-}
-
-\caption{Pairwise comparisons for smooth terms.
-\textbf{A}: Pairwise comparisons for the full dataset.
-\textbf{B}: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
-In both cases the effect of treatment is significant after day 3.
-In both cases, a simultaneous CI has been computed around the smooth difference.}\label{fig:plot-pairwise-comp}
-\end{figure}
-
-\hypertarget{GAM-significance}{%
-\subsection{Determination of significance in GAMs for longitudinal data}\label{GAM-significance}}
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
-
-The absence of a \emph{p-value} in this context might seem odd, but its interpretation can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment.
-This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
-
-With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant.
-A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
-
-Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}B and Figure \ref{fig:sim-smooth-plot}D.
-Figure \ref{fig:plot-pairwise-comp} shows the comparison between each treatment group for the full and missing datasets with a simultaneous CI computed around the difference.
-Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared.
-Notice that because we have included the means of each group, there is correspondence between the scale of the original data the scale of the pairwise comparisons.
-This can be seen as in Figure \ref{fig:sim-smooth-plot}B at day 5 there is essentially a 50\% difference between \(\mbox{StO}_2\) in both groups, which corresponds to the -50 difference in \ref{fig:plot-pairwise-comp}A.
-However, if there are multiple parametric terms in the model (more factors that need to be specified in the model) such inclusion of the means can become problematic.
-However, we believe that the model we have presented here suffices in a wide range of situations where adding the group means is a relatively easy implementation that can help better visualize the estimation from the model from a biological perspective.
-
-In Figure \ref{fig:plot-pairwise-comp}A the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by \(\approx\) 3 day it is the ``Treatment'' group the one that on average, has greater \(\mbox{StO}_2\).
-This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the given model.
-Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in mean \(\mbox{StO}_2\) occurs.
-
-On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that because the confidence intervals overlap for the first two days there is essentially no difference between the groups.
-However, because the model was still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes significant as in the full dataset smooth pairwise comparison.
-
-In a sense, the pairwise smooth comparison is more informative than a \emph{post-hoc} \emph{p-value}.
-For biomedical studies, the smooth comparison is able to provide an estimate of \emph{when} and by \emph{how much} a biological process becomes significant.
-This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-Biomedical longitudinal data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
-Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}, where it is clear that these models do not capture the non-linear trend of the data, thereby causing a ``model misspecification error.''
-
-This ``model misspecification'' error, also is known as a ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Polynomial effects can be used in LMEMs, and it could be argued that in Section \ref{simulation} a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects presents more questions than answers in this case (i.e., what is the appropriate polynomial effect to use?
-quadratic?, cubic?, what is the biological interpretation of a polynomial effect?).
-Moreover, polynomial effects have other associated issues that we have described in Section \ref{GAM-theory}.
-
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{66}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.
-
-In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{67}{]}.
-This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).
-
-However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} using \(\textsf{R}\) and the package \emph{mgcv}{[}\protect\hyperlink{ref-wood2017}{37}{]} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
-One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
-The model is therefore able to identify changes between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
-
-We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{68},\protect\hyperlink{ref-weissgerber2018}{69}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
-Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
-We have made all the data and code used in this paper accessible, and we hope that this will encourage other researchers to do the same with future projects.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-We have presented GAMs as a method to analyze longitudinal biomedical data.
-Future directions of this work will include simulation-based estimations of statistical power using GAMs, as well as demonstrating the prediction capabilities of these models using large datasets.
-By making the data and code used in this paper accessible, we hope to address the need of creating and sharing reproducible work in biomedical research.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-This work was supported by the National Science Foundation Career Award (CBET 1751554, TJM) and the Arkansas Biosciences Institute.
-
-\hypertarget{declaration-of-conflicting-interests}{%
-\section{Declaration of Conflicting Interests}\label{declaration-of-conflicting-interests}}
-
-The Authors declare that there is no conflict of interest.
-
-\hypertarget{supplementary-materials}{%
-\section*{Supplementary Materials}\label{supplementary-materials}}
-\addcontentsline{toc}{section}{Supplementary Materials}
-
-An Appendix which contains all the code used to create this manuscript, along with a basic workflow to implement GAMs in R is available as Supplementary Material in PDF.
-A GitHub repository containing all the code used for this paper along with detailed instructions for its use is available at \url{https://github.com/aimundo/GAMs-biomedical-research}.
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-roblyer2011}{}}%
-\CSLLeftMargin{{[}1{]} }
-\CSLRightInline{D.
-Roblyer, S.
-Ueda, A.
-Cerussi, W.
-Tanamai, A.
-Durkin, R.
-Mehta, D.
-Hsiang, J.A.
-Butler, C.
-McLaren, W.-P.
-Chen, B.
-Tromberg, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences.
-108 (2011) 14626--14631.
-\url{https://doi.org/10.1073/pnas.1013103108}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tank2020}{}}%
-\CSLLeftMargin{{[}2{]} }
-\CSLRightInline{A.
-Tank, H.M.
-Peterson, V.
-Pera, S.
-Tabassum, A.
-Leproux, T.
-O'Sullivan, E.
-Jones, H.
-Cabral, N.
-Ko, R.S.
-Mehta, B.J.
-Tromberg, D.
-Roblyer, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research.
-22 (2020).
-\url{https://doi.org/10.1186/s13058-020-01262-1}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-pavlov2018}{}}%
-\CSLLeftMargin{{[}3{]} }
-\CSLRightInline{M.V.
-Pavlov, T.I.
-Kalganova, Y.S.
-Lyubimtseva, Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy, Journal of Biomedical Optics.
-23 (2018) 1.
-\url{https://doi.org/10.1117/1.jbo.23.9.091410}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-demidov2018}{}}%
-\CSLLeftMargin{{[}4{]} }
-\CSLRightInline{V.
-Demidov, A.
-Maeda, M.
-Sugita, V.
-Madge, S.
-Sadanand, C.
-Flueraru, I.A.
-Vitkin, Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography, Scientific Reports.
-8 (2018).
-\url{https://doi.org/10.1038/s41598-017-18635-w}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ritter2001}{}}%
-\CSLLeftMargin{{[}5{]} }
-\CSLRightInline{G.
-Ritter, L.
-Cohen, C.
-Williams, E.
-Richards, L.
-Old, S.
-Welt, {Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}, {Cancer Research}.
-{61} (2001) 6851--6859.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-roth2017}{}}%
-\CSLLeftMargin{{[}6{]} }
-\CSLRightInline{E.M.
-Roth, A.C.
-Goldberg, A.L.
-Catapano, A.
-Torri, G.D.
-Yancopoulos, N.
-Stahl, A.
-Brunet, G.
-Lecorps, H.M.
-Colhoun, Antidrug antibodies in patients treated with alirocumab, New England Journal of Medicine.
-376 (2017) 1589--1590.
-\url{https://doi.org/10.1056/nejmc1616623}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jones2018}{}}%
-\CSLLeftMargin{{[}7{]} }
-\CSLRightInline{J.D.
-Jones, H.E.
-Ramser, A.E.
-Woessner, K.P.
-Quinn, In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing, Communications Biology.
-1 (2018).
-\url{https://doi.org/10.1038/s42003-018-0206-4}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-skala2010}{}}%
-\CSLLeftMargin{{[}8{]} }
-\CSLRightInline{M.C.
-Skala, A.
-Fontanella, L.
-Lan, J.A.
-Izatt, M.W.
-Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics.
-15 (2010) 011112.
-\url{https://doi.org/10.1117/1.3285584}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-greening2018}{}}%
-\CSLLeftMargin{{[}9{]} }
-\CSLRightInline{G.J.
-Greening, K.P.
-Miller, C.R.
-Spainhour, M.D.
-Cato, T.J.
-Muldoon, Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy, Biomedical Optics Express.
-9 (2018) 2871.
-\url{https://doi.org/10.1364/boe.9.002871}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-sio2016}{}}%
-\CSLLeftMargin{{[}10{]} }
-\CSLRightInline{T.T.
-Sio, P.J.
-Atherton, B.J.
-Birckhead, D.J.
-Schwartz, J.A.
-Sloan, D.K.
-Seisler, J.A.
-Martenson, C.L.
-Loprinzi, P.C.
-Griffin, R.F.
-Morton, J.C.
-Anders, T.J.
-Stoffel, R.E.
-Haselow, R.B.
-Mowat, M.A.N.
-Wittich, J.D.
-Bearden, R.C.
-Miller, Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}), Supportive Care in Cancer.
-24 (2016) 3847--3855.
-\url{https://doi.org/10.1007/s00520-016-3213-3}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-kamstra2015}{}}%
-\CSLLeftMargin{{[}11{]} }
-\CSLRightInline{J.I.
-Kamstra, P.U.
-Dijkstra, M.
-van Leeuwen, J.L.N.
-Roodenburg, J.A.
-Langendijk, Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study, Oral Oncology.
-51 (2015) 548--555.
-\url{https://doi.org/10.1016/j.oraloncology.2015.01.016}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wagenmakers2008}{}}%
-\CSLLeftMargin{{[}12{]} }
-\CSLRightInline{E.-J.
-Wagenmakers, M.
-Lee, T.
-Lodewyckx, G.J.
-Iverson, Bayesian versus frequentist inference, in: Bayesian Evaluation of Informative Hypotheses, Springer New York, 2008: pp.
-181--207.
-\url{https://doi.org/10.1007/978-0-387-09612-4_9}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gueorguieva2004}{}}%
-\CSLLeftMargin{{[}13{]} }
-\CSLRightInline{R. Gueorguieva, J.H. Krystal, Move over {ANOVA}, Archives of General Psychiatry. 61 (2004) 310. \url{https://doi.org/10.1001/archpsyc.61.3.310}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-schober2018}{}}%
-\CSLLeftMargin{{[}14{]} }
-\CSLRightInline{P.
-Schober, T.R.
-Vetter, Repeated measures designs and analysis of longitudinal data, Anesthesia {\&} Analgesia.
-127 (2018) 569--575.
-\url{https://doi.org/10.1213/ane.0000000000003511}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-pinheiro2006}{}}%
-\CSLLeftMargin{{[}15{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science; Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-vishwanath2009}{}}%
-\CSLLeftMargin{{[}16{]} }
-\CSLRightInline{K.
-Vishwanath, H.
-Yuan, W.T.
-Barry, M.W.
-Dewhirst, N.
-Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia.
-11 (2009) 889--900.
-\url{https://doi.org/10.1593/neo.09580}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-dennis2019}{}}%
-\CSLLeftMargin{{[}17{]} }
-\CSLRightInline{B.
-Dennis, J.M.
-Ponciano, M.L.
-Taper, S.R.
-Lele, Errors in statistical inference under model misspecification: Evidence, hypothesis testing, and {AIC}, Frontiers in Ecology and Evolution.
-7 (2019).
-\url{https://doi.org/10.3389/fevo.2019.00372}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wang2019}{}}%
-\CSLLeftMargin{{[}18{]} }
-\CSLRightInline{B.
-Wang, Z.
-Zhou, H.
-Wang, X.M.
-Tu, C.
-Feng, The p-value and model specification in statistics, General Psychiatry.
-32 (2019) e100081.
-\url{https://doi.org/10.1136/gpsych-2019-100081}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-liu2010}{}}%
-\CSLLeftMargin{{[}19{]} }
-\CSLRightInline{C.
-Liu, T.P.
-Cripe, M.-O.
-Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy.
-18 (2010) 1724--1730.
-\url{https://doi.org/10.1038/mt.2010.127}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-halsey2015}{}}%
-\CSLLeftMargin{{[}20{]} }
-\CSLRightInline{L.G.
-Halsey, D.
-Curran-Everett, S.L.
-Vowler, G.B.
-Drummond, The fickle p value generates irreproducible results, Nature Methods.
-12 (2015) 179--185.
-\url{https://doi.org/10.1038/nmeth.3288}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-abdi2010}{}}%
-\CSLLeftMargin{{[}21{]} }
-\CSLRightInline{H. Abdi, {Holm's sequential Bonferroni procedure}, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nakagawa2004}{}}%
-\CSLLeftMargin{{[}22{]} }
-\CSLRightInline{S.
-Nakagawa, A farewell to bonferroni: The problems of low statistical power and publication bias, Behavioral Ecology.
-15 (2004) 1044--1045.
-\url{https://doi.org/10.1093/beheco/arh107}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gelman2012}{}}%
-\CSLLeftMargin{{[}23{]} }
-\CSLRightInline{A.
-Gelman, J.
-Hill, M.
-Yajima, Why we (usually) don{\textquotesingle}t have to worry about multiple comparisons, Journal of Research on Educational Effectiveness.
-5 (2012) 189--211.
-\url{https://doi.org/10.1080/19345747.2011.618213}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-albers2019}{}}%
-\CSLLeftMargin{{[}24{]} }
-\CSLRightInline{C. Albers, The problem with unadjusted multiple and sequential statistical testing, Nature Communications. 10 (2019). \url{https://doi.org/10.1038/s41467-019-09941-0}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ugrinowitsch2004}{}}%
-\CSLLeftMargin{{[}25{]} }
-\CSLRightInline{C.
-Ugrinowitsch, G.W.
-Fellingham, M.D.
-Ricard, Limitations of ordinary least squares models in analyzing repeated measures data, Medicine {\&} Science in Sports {\&} Exercise.
-(2004) 2144--2148.
-\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-huynh1976}{}}%
-\CSLLeftMargin{{[}26{]} }
-\CSLRightInline{H.
-Huynh, L.S.
-Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics.
-1 (1976) 69--82.
-\url{https://doi.org/10.3102/10769986001001069}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-greenhouse1959}{}}%
-\CSLLeftMargin{{[}27{]} }
-\CSLRightInline{S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-haverkamp2017}{}}%
-\CSLLeftMargin{{[}28{]} }
-\CSLRightInline{N.
-Haverkamp, A.
-Beauducel, Violation of the sphericity assumption and its effect on type-i error rates in repeated measures {ANOVA} and multi-level linear models ({MLM}), Frontiers in Psychology.
-8 (2017).
-\url{https://doi.org/10.3389/fpsyg.2017.01841}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-keselman2001}{}}%
-\CSLLeftMargin{{[}29{]} }
-\CSLRightInline{H.J.
-Keselman, J.
-Algina, R.K.
-Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematical and Statistical Psychology.
-54 (2001) 1--20.
-\url{https://doi.org/10.1348/000711001159357}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-charan2013}{}}%
-\CSLLeftMargin{{[}30{]} }
-\CSLRightInline{J.
-Charan, N.
-Kantharia, How to calculate sample size in animal studies?, Journal of Pharmacology and Pharmacotherapeutics.
-4 (2013) 303.
-\url{https://doi.org/10.4103/0976-500x.119726}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-barr2013}{}}%
-\CSLLeftMargin{{[}31{]} }
-\CSLRightInline{D.J.
-Barr, R.
-Levy, C.
-Scheepers, H.J.
-Tily, Random effects structure for confirmatory hypothesis testing: Keep it maximal, Journal of Memory and Language.
-68 (2013) 255--278.
-\url{https://doi.org/10.1016/j.jml.2012.11.001}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rose2012}{}}%
-\CSLLeftMargin{{[}32{]} }
-\CSLRightInline{N.L.
-Rose, H.
-Yang, S.D.
-Turner, G.L.
-Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, {UK}, Geochimica Et Cosmochimica Acta.
-82 (2012) 113--135.
-\url{https://doi.org/10.1016/j.gca.2010.12.026}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-pedersen2019}{}}%
-\CSLLeftMargin{{[}33{]} }
-\CSLRightInline{E.J.
-Pedersen, D.L.
-Miller, G.L.
-Simpson, N.
-Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, {PeerJ}.
-7 (2019) e6876.
-\url{https://doi.org/10.7717/peerj.6876}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-simpson2018}{}}%
-\CSLLeftMargin{{[}34{]} }
-\CSLRightInline{G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-yang2012}{}}%
-\CSLLeftMargin{{[}35{]} }
-\CSLRightInline{L.
-Yang, G.
-Qin, N.
-Zhao, C.
-Wang, G.
-Song, Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality, {BMC} Medical Research Methodology.
-12 (2012).
-\url{https://doi.org/10.1186/1471-2288-12-165}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-beck1998}{}}%
-\CSLLeftMargin{{[}36{]} }
-\CSLRightInline{N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. 42 (1998) 596. \url{https://doi.org/10.2307/2991772}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wood2017}{}}%
-\CSLLeftMargin{{[}37{]} }
-\CSLRightInline{S.N. Wood, Generalized additive models, Chapman; Hall/{CRC}, 2017. \url{https://doi.org/10.1201/9781315370279}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-r}{}}%
-\CSLLeftMargin{{[}38{]} }
-\CSLRightInline{R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wood2016}{}}%
-\CSLLeftMargin{{[}39{]} }
-\CSLRightInline{S.N.
-Wood, N.
-Pya, B.
-Säfken, Smoothing parameter and model selection for general smooth models, Journal of the American Statistical Association.
-111 (2016) 1548--1563.
-\url{https://doi.org/10.1080/01621459.2016.1180986}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-west2014}{}}%
-\CSLLeftMargin{{[}40{]} }
-\CSLRightInline{B.T.
-West, K.B.
-Welch, A.T.
-Galecki, Linear mixed models: A practical guide using statistical software, second edition, Taylor \& Francis, 2014.
-\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wolfinger1996}{}}%
-\CSLLeftMargin{{[}41{]} }
-\CSLRightInline{R.D.
-Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics.
-1 (1996) 205.
-\url{https://doi.org/10.2307/1400366}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-weiss2005}{}}%
-\CSLLeftMargin{{[}42{]} }
-\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://doi.org/10.1007/0-387-28314-5}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-geisser1958}{}}%
-\CSLLeftMargin{{[}43{]} }
-\CSLRightInline{S.
-Geisser, S.W.
-Greenhouse, An extension of box{\textquotesingle}s results on the use of the {\textdollar}f{\textdollar} distribution in multivariate analysis, The Annals of Mathematical Statistics.
-29 (1958) 885--891.
-\url{https://doi.org/10.1214/aoms/1177706545}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-maxwell2017}{}}%
-\CSLLeftMargin{{[}44{]} }
-\CSLRightInline{S.E. Maxwell, H.D. Delaney, K. Kelley, Designing experiments and analyzing data, Routledge, 2017. \url{https://doi.org/10.4324/9781315642956}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-molenberghs2004}{}}%
-\CSLLeftMargin{{[}45{]} }
-\CSLRightInline{G. Molenberghs, Analyzing incomplete longitudinal clinical trial data, Biostatistics. 5 (2004) 445--464. \url{https://doi.org/10.1093/biostatistics/kxh001}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ma2012}{}}%
-\CSLLeftMargin{{[}46{]} }
-\CSLRightInline{Y.
-Ma, M.
-Mazumdar, S.G.
-Memtsoudis, Beyond repeated-measures analysis of variance, Regional Anesthesia and Pain Medicine.
-37 (2012) 99--105.
-\url{https://doi.org/10.1097/aap.0b013e31823ebc74}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-scheffer2002}{}}%
-\CSLLeftMargin{{[}47{]} }
-\CSLRightInline{J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-potthoff2006}{}}%
-\CSLLeftMargin{{[}48{]} }
-\CSLRightInline{R.F.
-Potthoff, G.E.
-Tudor, K.S.
-Pieper, V.
-Hasselblad, Can one assess whether missing data are missing at random in medical studies?, Statistical Methods in Medical Research.
-15 (2006) 213--234.
-\url{https://doi.org/10.1191/0962280206sm448oa}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-box1976}{}}%
-\CSLLeftMargin{{[}49{]} }
-\CSLRightInline{G.E.P. Box, Science and statistics, 71 (1976) 791--799. \url{https://doi.org/10.1080/01621459.1976.10480949}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlme}{}}%
-\CSLLeftMargin{{[}50{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, {nlme}: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ohara2010}{}}%
-\CSLLeftMargin{{[}51{]} }
-\CSLRightInline{R. OHara, J. Kotze, Do not log-transform count data, (2010). \url{https://doi.org/10.1038/npre.2010.4136.1}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nelder1972}{}}%
-\CSLLeftMargin{{[}52{]} }
-\CSLRightInline{J.A. Nelder, R.W.M. Wedderburn, Generalized linear models, Journal of the Royal Statistical Society. Series A (General). 135 (1972) 370. \url{https://doi.org/10.2307/2344614}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-mcculloch2001}{}}%
-\CSLLeftMargin{{[}53{]} }
-\CSLRightInline{C. McCulloch, Generalized, linear, and mixed models, John Wiley \& Sons, New York, 2001.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-dobson2008}{}}%
-\CSLLeftMargin{{[}54{]} }
-\CSLRightInline{A. Dobson, An introduction to generalized linear models, CRC Press, Boca Raton, 2008.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-stroup2013}{}}%
-\CSLLeftMargin{{[}55{]} }
-\CSLRightInline{W. Stroup, Generalized linear mixed models : Modern concepts, methods and applications, CRC Press, Taylor \& Francis Group, Boca Raton, 2013.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hastie1987}{}}%
-\CSLLeftMargin{{[}56{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association.
-82 (1987) 371--386.
-\url{https://doi.org/10.1080/01621459.1987.10478440}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hefley2017}{}}%
-\CSLLeftMargin{{[}57{]} }
-\CSLRightInline{T.J.
-Hefley, K.M.
-Broms, B.M.
-Brost, F.E.
-Buderman, S.L.
-Kay, H.R.
-Scharf, J.R.
-Tipton, P.J.
-Williams, M.B.
-Hooten, The basis function approach for modeling autocorrelation in ecological data, Ecology.
-98 (2017) 632--646.
-\url{https://doi.org/10.1002/ecy.1674}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wegman1983}{}}%
-\CSLLeftMargin{{[}58{]} }
-\CSLRightInline{E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wood2003}{}}%
-\CSLLeftMargin{{[}59{]} }
-\CSLRightInline{S.N. Wood, Thin plate regression splines, 65 (2003) 95--114. \url{https://doi.org/10.1111/1467-9868.00374}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-mcelreath2018}{}}%
-\CSLLeftMargin{{[}60{]} }
-\CSLRightInline{R. McElreath, Statistical rethinking, Chapman; Hall/{CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-miller2019}{}}%
-\CSLLeftMargin{{[}61{]} }
-\CSLRightInline{D.L. Miller, Bayesian views of generalized additive modelling, arXiv Preprint arXiv:1902.01330. (2019).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ruppert2003}{}}%
-\CSLLeftMargin{{[}62{]} }
-\CSLRightInline{D. Ruppert, Semiparametric regression, Cambridge University Press, Cambridge New York, 2003.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
-\CSLLeftMargin{{[}63{]} }
-\CSLRightInline{G.
-Marra, S.N.
-Wood, Coverage properties of confidence intervals for generalized additive model components, Scandinavian Journal of Statistics.
-39 (2012) 53--74.
-\url{https://doi.org/10.1111/j.1467-9469.2011.00760.x}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
-\CSLLeftMargin{{[}64{]} }
-\CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
-\CSLLeftMargin{{[}65{]} }
-\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with {R}, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
-\CSLLeftMargin{{[}66{]} }
-\CSLRightInline{T.A.
-Lang, D.G.
-Altman, Basic statistical reporting for articles published in biomedical journals: The {``}statistical analyses and methods in the published literature{''} or the {SAMPL} guidelines, International Journal of Nursing Studies.
-52 (2015) 5--9.
-\url{https://doi.org/10.1016/j.ijnurstu.2014.09.006}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hastie1995}{}}%
-\CSLLeftMargin{{[}67{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models for medical research, Statistical Methods in Medical Research.
-4 (1995) 187--196.
-\url{https://doi.org/10.1177/096228029500400302}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-begley2015}{}}%
-\CSLLeftMargin{{[}68{]} }
-\CSLRightInline{C.G. Begley, J.P.A. Ioannidis, Reproducibility in science, Circulation Research. 116 (2015) 116--126. \url{https://doi.org/10.1161/circresaha.114.303819}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-weissgerber2018}{}}%
-\CSLLeftMargin{{[}69{]} }
-\CSLRightInline{T.L.
-Weissgerber, O.
-Garcia-Valencia, V.D.
-Garovic, N.M.
-Milic, S.J.
-Winham, Why we need to report more than {\textquotesingle}data were analyzed by t-tests or {ANOVA}{\textquotesingle}, {eLife}.
-7 (2018).
-\url{https://doi.org/10.7554/elife.36163}.}
-
-\end{CSLReferences}
-
-\hypertarget{appendix-appendix}{%
-\appendix}
-
-\appendixpagenumbering
-
 \counterwithin{figure}{section}
 
-\hypertarget{appendix}{%
-\section{APPENDIX}\label{appendix}}
-
-This section presents the code used to generate figures, models and simulated data from the main manuscript.
-
-\hypertarget{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}{%
-\subsection{Compound symmetry and independent errors in linear and quadratic responses}\label{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}}
-
-This section simulates linear and quadratic data in the same manner as in Section \ref{simulation} in the main manuscript.
-The linear simulations using Figure \ref{fig:linear-cases-Appendix} show in panels A and D the simulated mean responses and individual data points.
-Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-The following code chunks produce a more comprehensive exploration of Figure \ref{fig:l-q-response} in the main manuscript.
-
-First, a function is created to simulate data across six timepoints using a linear or quadratic mean response, with correlated or uncorrelated errors.
-Each group has a different slope/concavity.
-The main function is the same for both groups, but a change in the sign allows to invert the trend.
-
-\begin{lstlisting}[language=R]
-##########Section for calculations###########
-
-## Example with linear response
-
-#This function simulates data using a linear or quadratic mean response and each with correlated
-#or uncorrelated errors. Each group has a different slope/concavity.
-example <- function(n_time = 6, #number of time points
-                    fun_type = "linear", #type of response
-                    error_type = "correlated") {
-
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  x <- seq(1,6, length.out = n_time)
-
-  #Create mean response matrix: linear or quadratic
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # quadratic response (non-linear)
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-
-  #create an array where individual observations per each time point for each group are to be stored. Currently using 10 observations per timepoint
-  y <- array(0, dim = c(length(x), 2, 10))
-
-  #Create array to store the "errors" for each group at each timepoint. The "errors" are the 
-  #between-group variability in the response.
-  errors <- array(0, dim = c(length(x), 2, 10))
-  #create an array where 10 observations per each time point for each group are to be stored
-
-  #The following loops create independent or correlated responses.
-To each value of mu (mean response per group) a randomly generated error (correlated or uncorrelated) is added and thus the individual response is created.
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors: variance covariance matrix
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-#labeling y and errors  
-  dimnames(y) <- list(time = x, 
-                      treatment = 1:2, 
-                      subject = 1:10)
-
-  dimnames(errors) <- list(time = x, 
-                           treatment = 1:2, 
-                           subject = 1:10)
-
-  #labeling the mean response
-  dimnames(mu) <- list(time = x, 
-                       treatment = 1:2)
-
-  #convert y, mu and errors to  dataframes with time, treatment and subject columns
-  dat <- as.data.frame.table(y, 
-                             responseName = "y")
-  dat_errors <- as.data.frame.table(errors, 
-                                    responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, 
-                                responseName = "mu")
-
-  #join the dataframes to show mean response and errors per subject
-  dat <- left_join(dat, dat_errors, 
-                   by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, 
-                   by = c("time", "treatment"))
-  #add time
-  dat$time <- as.numeric(as.character(dat$time))
-  #label subjects per group
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, 
-                                  treatment, 
-                                  sep = "-")))
-
-  ## repeated measures ANOVA
-
-  fit_anova <- lm(y ~ time + treatment + time * treatment, data = dat)
-
-#LMEM: time and treatment interaction model, compound symmetry 
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  #create a prediction frame where the model can be used for plotting purposes
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  #add model predictions to the dataframe that has the simulated data
-  dat$pred_anova <- predict(fit_anova)
-  dat$pred_lmem <- predict(fit_lme)
-
-  #return everything in a list
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_anova=fit_anova,
-    fit_lme = fit_lme
-  ))
-}
-
-##################Section for plotting#################################
-
-#This function will create the plots for either a "linear" or "quadratic" response
-
-plot_example <- function(sim_dat) {
-  ## Plot the simulated data (scatterplot)
-  txt<-20
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = treatment, 
-               color = treatment)
-           ) +
-    geom_point(show.legend=FALSE) +
-    labs(y='response')+
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment),
-              show.legend=FALSE) +
-    theme_classic() +
-    theme(plot.title = element_text(size = txt, 
-                                  face = "bold"),
-        text=element_text(size=txt))+
-    thm1
-
-  #plot the simulated data with trajectories per each subject
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = subject, 
-               color = treatment)
-           ) +
-    geom_line(aes(size = "Subjects"),
-              show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment,
-                  size = "Simulated Truth"), 
-              lty = 1,show.legend = FALSE) +
-    labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    theme_classic()+
-     theme(plot.title = element_text(size = txt, 
-                                face = "bold"),
-     text=element_text(size=txt))+
-    thm1
-
-  #plot the errors
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = errors,
-               group = subject, 
-               color = treatment)) +
-    geom_line(show.legend=FALSE) +
-     labs(y='errors')+
-     theme_classic()+
-     theme(plot.title = element_text(size = txt, 
-                                  face = "bold"),
-        text=element_text(size=txt))+
-    thm1
-
-   #plot the model predictions for rm-ANOVA
-  p4 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point(show.legend=FALSE)+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_anova), 
-                  group = subject, size = "Subjects"),show.legend = FALSE) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_anova, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population"),
-              show.legend=FALSE) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = txt, 
-                                  face = "bold"),
-        text=element_text(size=txt))+
-    thm1
-
-   #plot the LMEM predictions
-  p5 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point()+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), 
-                  group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_lme, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population")) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = txt, 
-                                  face = "bold"),
-        text=element_text(size=txt))+
-    thm1
-
-  return((p1+p3+p2+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
-
-#Store each plot in a separate object
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent")) 
-\end{lstlisting}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/linear-cases-Appendix-1}
-
-}
-
-\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are able to capture the trend of the data.}\label{fig:linear-cases-Appendix}
-\end{figure}
-
-For the quadratic response case, Figure \ref{fig:quadratic-cases-Appendix} shows the simulated responses using compound symmetry and independent errors.
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/quadratic-cases-Appendix-1}
-
-}
-
-\caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
-In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
-Both rm-ANOVA and the LMEM are not able to capture the changes in each group over time.}\label{fig:quadratic-cases-Appendix}
-\end{figure}
-
-\hypertarget{basis-functions-and-gams}{%
-\subsection{Basis functions and GAMs}\label{basis-functions-and-gams}}
-
-This code produces Figure \ref{fig:basis-plot} from the main manuscript.
-Briefly, a non-linear (quadratic) response is simulated, a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
-The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
-
-\begin{lstlisting}[language=R]
-#generate the response: the same initial procedure from the previous section to simulate
-#the response
-set.seed(1)
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-
- #label each table
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
-
- #Convert to dataframes with subject, time and group columns
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
-
- #label subject per group
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
- #extract  "Group 1" to fit the GAM
-  dat<-subset(dat,treatment==1)
- #keep just the response and timepoint columns
-   dat<-dat[,c('y','time')]
-
-   #GAM model of time, 5 basis
-gm<-gam(y~s(time,k=5),data=dat)
-
-#model_matrix (also known as) 'design matrix'
-#will contain the smooths used to create  model 'gm'
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of basis by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be 
-#multiplied by the coefficients
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final smooth by aggregating the basis functions
-
-smooth<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(smooth=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,
-            aes(x=time,
-                y=value_orig,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(size=sz,
-            show.legend=FALSE)+
-  geom_point(size=sz+1,
-             show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,
-             labeller = as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#penalized basis
-p12<-ggplot(basis_plot,
-            aes(x=time,
-                y=mod_val,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(show.legend = FALSE,
-            size=sz)+
-  geom_point(show.legend = FALSE,
-             size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,
-             labeller=as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#heatmap of the coefficients
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,
-            aes(x=Basis,
-                y=Basis))+
-  geom_tile(aes(fill = cof), 
-            colour = "black") +
-    scale_fill_gradient(low = "white",
-                        high = "#B50A2AFF")+ #color picked from KikiMedium
-  labs(x='Basis',
-       y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),
-            size=7,
-            show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,
-            aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  labs(y='Simulated \n response')+
-  geom_line(data=smooth,
-            aes(x=time,
-                y=smooth),
-            color="#6C581DFF",
-            size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-\end{lstlisting}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/basis-plot-appendix-1}
-
-}
-
-\caption{Basis functions for a single smoother for time.
-\textbf{A}: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-\textbf{B}: Matrix of basis function weights.
-Each basis function is multiplied by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-\textbf{C}: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
-Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-\textbf{D}: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot-appendix}
-\end{figure}
-
-\hypertarget{tumor-data-simulation}{%
-\subsection{Longitudinal biomedical data simulation and GAMs}\label{tumor-data-simulation}}
-
-This section describes how to fit GAMs to longitudinal data using simulated data.
-First, data is simulated according to Section \ref{longitudinal-GAMs}, where reported data of oxygen saturation (\(\mbox{StO}_2\)) in tumors under either chemotherapy or saline control is used as a starting point to generate individual responses in each group.
+This appendix shows a basic workflow to fit a series of increasingly complex GAMs to simulated data that follows the trends presented in Section 5.1 in the main manuscript.
+Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
+For simplicity, the confidence intervals (CIs) shown in this section for the models are the across the function CIs created by \passthrough{\lstinline!mgcv!} by default.
+However, for the pairwise comparisons of the third model we use \emph{simultaneous intervals} as in the main manuscript, and the code for creating simultaneous CIs for the smooths can be found in \textbf{Appendix XXX}.
+
+\hypertarget{setup-and-data-simulation}{%
+\subsection{Setup and data simulation}\label{setup-and-data-simulation}}
+
+First, we load the libraries needed for all analyses and figures. We also set seed for reproducibility.
+
+\begin{lstlisting}[language=R]
+library(patchwork)
+library(tidyverse)
+library(mvnfast)
+library(nlme)
+library(mgcv)
+library(gratia)
+library(here)
+library(scico)
+set.seed(2021) #set seed for reproducibility
+
+#alpha for ribbon in the smooth plots
+al<-0.8
+\end{lstlisting}
+
+Next, we create a data frame with the same trends from the main manuscript:
+
+\begin{itemize}
+\tightlist
+\item
+  Two treatment groups (Control and Treatment)
+\item
+  Five time points (days 0, 2, 5, 7,10)
+\item
+  Trends in \(\mbox{StO}_2\) for both groups
+\end{itemize}
 
@@
             )
-#alpha for ribbon
-al<-0.8
-
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-
+\end{lstlisting}
+
+We do not include a plot of the simulated data here because it follows the same trend from Figure 3A in the main manuscript (however, it can be easily plotted if desired).
+Finally, we call the function \passthrough{\lstinline!simulate\_data!} in the script \passthrough{\lstinline!simulate\_data.R!} to create 10 replicates at each time point from a normal distribution with a standard deviation of 10\% in \(\mbox{StO2}\).
+
+\begin{lstlisting}[language=R]
 n <- 10 #number of observations
 sd <- 10 #approximate sd from paper
-df <- 6
+source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","simulate_data.R"))
 dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, aes(x = Day, 
-                        y = StO2_sim, 
-                        color = Group, 
-                        group=subject)) +
-    geom_point(show.legend=FALSE,
-               size=1.5,
-               alpha=0.6)+
-    geom_line(size=0.6, alpha=0.4,show.legend=FALSE)+
-    geom_line(aes(x = Day, 
-               y = StO2,
-               color=Group),
-           size=1.5,
-           data=dat,
-           inherit.aes=FALSE,
-              show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],'(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    thm1
-\end{lstlisting}
-
-\hypertarget{workflow}{%
-\subsection{A basic Workflow for GAMs}\label{workflow}}
-
-This section shows a basic workflow to fit a series of increasingly complex GAMs to the simulated data from the previous section.
-Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
-For simplicity, the confidence intervals (CIs) shown in this section for the models are the across the function CIs created by \passthrough{\lstinline!mgcv!} by default.
-However, for the pairwise comparisons of the third model we use \emph{simultaneous intervals} as in the main manuscript, and the code for creating simultaneous CIs for the smooths can be found in Section \ref{GAM-linear-plot}.
+\end{lstlisting}
 
 \hypertarget{first-model}{%
-\subsubsection{First model}\label{first-model}}
+\subsection{First model}\label{first-model}}
 
@@
 For model check, the functions \passthrough{\lstinline!gam.check!} and \passthrough{\lstinline!summary!} from \emph{mgcv} provide detailed information about the model fit and its parameters.
-Keep in mind that \passthrough{\lstinline!gam.check!} is a function that also provides the graphical diagnostics obtained using \emph{gratia}, if such graphical output is not desired the source code can be accessed typing \passthrough{\lstinline!gam.check!} in the Console, and the code without the graphical output can be used in a custom function, which is the approach we follow later).
+Keep in mind that \passthrough{\lstinline!gam.check!} is a function that also provides the graphical diagnostics obtained using \emph{gratia}, if such graphical output is not desired the source code can be accessed typing \passthrough{\lstinline!gam.check!} in the Console, and the code without the graphical output can be used in a custom function (which is the approach we follow later on this Appendix).
 
 \hypertarget{graphical-diagnostics}{%
-\paragraph{Graphical diagnostics}\label{graphical-diagnostics}}
+\subsubsection{Graphical diagnostics}\label{graphical-diagnostics}}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/first-GAM-diag-1}
+{\centering \includegraphics[width=0.75\linewidth]{Appendix_A_files/figure-latex/first-GAM-diag-1}
 
@@
 Because we plot the model diagnostics using \passthrough{\lstinline!appraise!} later, the graphical output from \passthrough{\lstinline!gam.check!} will be suppressed by creating a custom function to obtain just the model estimates, thus avoiding repetition of the diagnostic plots.
-This will be achieved by calling the source code of \passthrough{\lstinline!gam.check!} and using the appropriate code in a new function that will be called \passthrough{\lstinline!gam.diagnostics!}.
-We are not including in the Appendix the code for the function \passthrough{\lstinline!gam.diagnostics!} as it is rather long, but if desired it can be accessed by going to the \passthrough{\lstinline!Appendix.Rmd!} file in the GitHub repository and scrolling to this exact place (the code is not included in the final output, but is evaluated to create the function).
-
-We now call \passthrough{\lstinline!gam.diagnostics!} to provide the desired diagnostic output, as well as a summary of the fitted model, which is obtained by calling \passthrough{\lstinline!summary!}.
-
-\begin{lstlisting}[language=R]
-gam.diagnostics(gam_00)
+This will be achieved by calling the source code of \passthrough{\lstinline!gam.check!} and using the appropriate code in a new function that will be called \passthrough{\lstinline!gam\_diagnostics!}, which can be found in the script \passthrough{\lstinline!gam\_diagnostics.R!}.
+
+In the next code chunk, we call \passthrough{\lstinline!gam.diagnostics!} to provide the desired diagnostic output:
+
+\begin{lstlisting}[language=R]
+source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","gam_diagnostics.R"))
+gam_diagnostics(gam_00)
 \end{lstlisting}
@@
 ## full convergence after 5 iterations.
-## Gradient range [-0.0003727881,-6.621452e-07]
-## (score 444.0118 & scale 450.6638).
-## Hessian positive definite, eigenvalue range [0.3881695,49.00676].
+## Gradient range [-2.830761e-07,-3.314641e-08]
+## (score 435.8783 & scale 388.2273).
+## Hessian positive definite, eigenvalue range [0.009374965,49.00017].
 ## Model rank =  5 / 5 
@@
 ##          k'  edf k-index p-value    
-## s(Day) 4.00 2.11    0.36  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-From the output, it can be seen that the \passthrough{\lstinline!k-index!} is 0.36, which indicates that the model is not capturing the variability in the data.
+## s(Day) 4.00 1.18    0.35  <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{lstlisting}
+
+And now obtain a summary of the fitted model, which is obtained by calling \passthrough{\lstinline!summary!}.
+
+\begin{lstlisting}[language=R]
+summary(gam_00)
+\end{lstlisting}
+
+\begin{lstlisting}
+## 
+## Family: gaussian 
+## Link function: identity 
+## 
+## Formula:
+## StO2_sim ~ s(Day, k = 5)
+## 
+## Parametric coefficients:
+##             Estimate Std. Error t value Pr(>|t|)    
+## (Intercept)    20.17       1.97   10.24   <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Approximate significance of smooth terms:
+##          edf Ref.df     F p-value   
+## s(Day) 1.184  1.342 8.861 0.00136 **
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.108   Deviance explained = 11.9%
+## -REML = 435.88  Scale est. = 388.23    n = 100
+\end{lstlisting}
+
+From the output of \passthrough{\lstinline!gam\_diagnostics!}, it can be seen that the \passthrough{\lstinline!k-index!} is 0.35, which indicates that the model is not capturing the variability in the data.
 The \passthrough{\lstinline!edf!} (effective degrees of freedom) is an indicator of the complexity of the smooth.
 Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
-
-\begin{lstlisting}[language=R]
-summary(gam_00)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ s(Day, k = 5)
-## 
-## Parametric coefficients:
-##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   22.967      2.123   10.82   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##          edf Ref.df     F  p-value    
-## s(Day) 2.114  2.565 7.633 0.000517 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.153   Deviance explained = 17.2%
-## -REML = 444.01  Scale est. = 450.66    n = 100
-\end{lstlisting}
 
@@
 Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \(\approx\) 17\%.
+Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \(\approx\) 12\%.
 
 \hypertarget{second-model}{%
-\subsubsection{Second model}\label{second-model}}
+\subsection{Second model}\label{second-model}}
 
@@
 The next iteration is a model where a different smooth of time (Day) is assigned for each group using \passthrough{\lstinline!by = Group!} in the model syntax.
+Below we fit such model, which is saved as \passthrough{\lstinline!gam\_01!}, and use \passthrough{\lstinline!gam\_diagnostiscs!} to obtain information from the model fit.
 
@@
 
-gam.diagnostics(gam_01)
+source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","gam_diagnostics.R"))
+gam_diagnostics(gam_01)
 \end{lstlisting}
@@
 ## Method: REML   Optimizer: outer newton
-## full convergence after 7 iterations.
-## Gradient range [-5.51754e-05,2.671715e-06]
-## (score 423.3916 & scale 280.8777).
-## Hessian positive definite, eigenvalue range [0.3162258,48.5557].
+## full convergence after 9 iterations.
+## Gradient range [-5.875578e-07,1.45076e-08]
+## (score 400.4803 & scale 164.8662).
+## Hessian positive definite, eigenvalue range [0.892525,48.57694].
 ## Model rank =  9 / 9 
@@
 ##                         k'  edf k-index p-value    
-## s(Day):GroupControl   4.00 3.39    0.43  <2e-16 ***
-## s(Day):GroupTreatment 4.00 3.23    0.43  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-Diagnostics for this model indicate that the k-index is still below 1 (0.43 from \passthrough{\lstinline!gam.check!}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
+## s(Day):GroupControl   4.00 3.69    0.58  <2e-16 ***
+## s(Day):GroupTreatment 4.00 3.72    0.58  <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{lstlisting}
+
+Diagnostics for this model indicate that the k-index is still below 1 (0.58 from \passthrough{\lstinline!gam\_diagnostics!} output), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
 Moreover, the smooths (plotted via the \passthrough{\lstinline!draw()!} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
@@
 ##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   22.967      1.676    13.7   <2e-16 ***
+## (Intercept)   20.169      1.284   15.71   <2e-16 ***
 ## ---
@@
 ##                         edf Ref.df      F  p-value    
-## s(Day):GroupControl   3.392  3.794  3.817  0.0304 *  
-## s(Day):GroupTreatment 3.229  3.682 21.174  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.472   Deviance explained = 50.8%
-## -REML = 423.39  Scale est. = 280.88    n = 100
-\end{lstlisting}
-
-From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \(\approx\) 51\%.
+## s(Day):GroupControl   3.688  3.942  9.084 2.33e-05 ***
+## s(Day):GroupTreatment 3.723  3.954 33.465  < 2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.621   Deviance explained =   65%
+## -REML = 400.48  Scale est. = 164.87    n = 100
+\end{lstlisting}
+
+From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \(\approx\) 65\%.
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/second-GAM-diag-1}
+{\centering \includegraphics[width=0.75\linewidth]{Appendix_A_files/figure-latex/second-GAM-diag-1}
 
@@
 \hypertarget{third-model}{%
-\subsubsection{Third model}\label{third-model}}
+\subsection{Third model}\label{third-model}}
 
 Model \passthrough{\lstinline!gam\_00!} was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted (a single smooth), the model diagnostics, and, the low variance explained by the model.
-On the other hand, \passthrough{\lstinline!gam\_01!} takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \ref{longitudinal-GAMs}, in order to differentiate between each group a parametric term needs to be added to the model for the interaction of \emph{Day} and \emph{Group}.
+On the other hand, \passthrough{\lstinline!gam\_01!} takes into account the nesting within each group and provides better variance explanation, but as indicated in Section 5.2 in the main manuscript, in order to differentiate between each group a parametric term needs to be added to the model for the interaction of \emph{Day} and \emph{Group}.
 
@@
 
-gam.diagnostics(gam_02)
+source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","gam_diagnostics.R"))
+
+gam_diagnostics(gam_02)
 \end{lstlisting}
@@
 ## full convergence after 10 iterations.
-## Gradient range [-8.164307e-08,1.500338e-08]
-## (score 355.8554 & scale 64.53344).
-## Hessian positive definite, eigenvalue range [1.174841,48.08834].
+## Gradient range [-2.934421e-07,2.346107e-07]
+## (score 355.5265 & scale 63.48881).
+## Hessian positive definite, eigenvalue range [1.21617,48.08942].
 ## Model rank =  10 / 10 
@@
 ##                         k'  edf k-index p-value
-## s(Day):GroupControl   4.00 3.87    1.02    0.55
-## s(Day):GroupTreatment 4.00 3.88    1.02    0.58
+## s(Day):GroupControl   4.00 3.88    1.23    0.98
+## s(Day):GroupTreatment 4.00 3.90    1.23    0.97
 \end{lstlisting}
@@
 By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\(\approx\) 1.02).
+From \passthrough{\lstinline!gam\_diagnostics!}, the k-index is now at an acceptable value (\(\approx\) \textbf{1.23}).
 
@@
 ##                Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)       9.084      1.136   7.996 4.09e-12 ***
-## GroupTreatment   27.766      1.607  17.282  < 2e-16 ***
+## (Intercept)      10.574      1.127   9.384 5.36e-15 ***
+## GroupTreatment   19.191      1.594  12.043  < 2e-16 ***
 ## ---
@@
 ##                         edf Ref.df     F p-value    
-## s(Day):GroupControl   3.873  3.990 17.57  <2e-16 ***
-## s(Day):GroupTreatment 3.879  3.991 89.33  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.879   Deviance explained = 88.9%
-## -REML = 355.86  Scale est. = 64.533    n = 100
-\end{lstlisting}
-
-From \passthrough{\lstinline!summary!}, the model is able to capture 89\% of the variance in the data, which is a substantial improvement over the variance explained by \passthrough{\lstinline!gam\_00!} and \passthrough{\lstinline!gam\_01!}.
+## s(Day):GroupControl   3.881  3.991 23.54  <2e-16 ***
+## s(Day):GroupTreatment 3.904  3.994 87.29  <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## R-sq.(adj) =  0.854   Deviance explained = 86.7%
+## -REML = 355.53  Scale est. = 63.489    n = 100
+\end{lstlisting}
+
+From \passthrough{\lstinline!summary!}, the model is able to capture 86.7\% of the variance in the data, which is a substantial improvement over the variance explained by \passthrough{\lstinline!gam\_00!} and \passthrough{\lstinline!gam\_01!}.
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/final-GAM-diag-1}
+{\centering \includegraphics[width=0.75\linewidth]{Appendix_A_files/figure-latex/final-GAM-diag-1}
 
@@
 ##               df      AIC
-## gam_00  4.564893 900.8257
-## gam_01  9.476137 858.6051
-## gam_02 10.980983 712.2067
-\end{lstlisting}
-
-The output in this case is expected: model \passthrough{\lstinline!gam\_02!} has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
+## gam_00  3.342420 884.4234
+## gam_01  9.895821 805.3071
+## gam_02 10.985458 710.5476
+\end{lstlisting}
+
+The output in this case is expected: model \passthrough{\lstinline!gam\_02!} has a lower AIC \textbf{(710.54)} whereas the initial two models have higher AICs (\textbf{884.43} and \textbf{805.30}).
 The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
@@
 \hypertarget{pairwise-comparisons-of-smooth-confidence-intervals}{%
-\paragraph{Pairwise comparisons of smooth confidence intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
-
-The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \ref{GAM-significance}.
+\subsection{Pairwise comparisons of smooth confidence intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
+
+The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in Section 5.3 in the main manuscript.
 
@@
 
-We want to emphasize that for the model used here we have included the group means in order to keep the pairwise comparisons on the scale of the response (a change from the code which appears in the chunk with a comment that says \passthrough{\lstinline!\#\#\#IMPORTANT!}).
-This approach works well for models like \passthrough{\lstinline!gam\_02!} but implementing this on models with more parametric terms can be challenging.
+We want to emphasize that for through the manuscript and for the comparisons on \passthrough{\lstinline!gam\_02!}, we included the group means in order to keep the pairwise comparisons on the scale of the response.
+By default, pairwise comparisons available in other software packages (such as in G.
+Simpson's \emph{gratia}) do not include the group means.
+However, we decided to include them because for our approach, it is easier to see the magnitude in the change of the difference between two treatment groups when the means are included.
+
+The change we allude to (mean group inclusion) can be found in the script \passthrough{\lstinline!pointwise\_comparisons.R!}, at the line where we have added a comment that reads:
+
+\begin{lstlisting}[language=R]
+######IMPORTANT: uncommenting the following two lines
+    #removes the group means from the comparison######
+\end{lstlisting}
+
+The inclusion of group means works well for models like \passthrough{\lstinline!gam\_02!}, but implementing this on models with more parametric terms can be challenging and we want the reader to be aware of this.
+
 However, we do believe that the model presented in the paper covers a wide range of situations and our approach here for the pairwise comparisons will be useful for most biomedical researchers.
-
-\begin{lstlisting}[language=R]
-##Pairwise comparisons
-difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
-  ## make sure f1 and f2 are characters
-  f1 <-  as.character(f1)
-  f2 <-  as.character(f2)
-  cnames <- colnames(Xp)
-  ## columns of Xp associated with pair of smooths
-  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
-  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
-  ## rows of Xp associated with pair of smooths
-  r1 <- data[[by_var]] == f1
-  r2 <- data[[by_var]] == f2
-
-  ## difference rows of Xp for pair of smooths
-  X <- Xp[r1, ] - Xp[r2, ]
-
-  ### IMPORTANT: Keep group means####
-  ## zero the cols related to other splines
- # X[, ! (c1 | c2)] <- 0
-
-  ## zero out the parametric cols
-  #X[, !grepl('^s\\(', cnames)] <- 0
-
-  ## compute difference
-  sm_diff <- drop(X %*% coefs)
-  se <- sqrt(rowSums((X %*% V) * X))
-  nr <- NROW(X)
-
-  ## Calculate posterior simulation for smooths
-  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
-  rownames(coefs_sim) <- rownames(V)
-  simDev <- X %*% coefs_sim
-  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
-  masd <- apply(absDev, 2, max)
-  crit_s <- quantile(masd, prob = 0.95, type = 8)
-
-  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
-              level_1 = rep(f1, nr),
-              level_2 = rep(f2, nr),
-              diff = sm_diff, se = se,
-              lower_s = sm_diff - crit_s * se,
-              upper_s = sm_diff + crit_s*se)
-
-  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
-  ## Only need rows associated with one of the levels
-  out <- bind_cols(out, data[r1, smooth_var])
-
-  out
-}
-
-#does both ci and si
-difference_smooths <- function(model,
-                                 smooth,
-                                 n = 100,
-                                 ci_level = 0.95,
-                                 newdata = NULL,
-                                 partial_match = TRUE,
-                                 unconditional = FALSE,
-                                 frequentist = FALSE,
-                                 nrep = 10000,
-                                 include_means = TRUE,
-                                 ...) {
-  if (missing(smooth)) {
-    stop("Must specify a smooth to difference via 'smooth'.")
-  }
-
-  # smooths in model
-  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
-  # select smooths
-  select <-
-    gratia:::check_user_select_smooths(smooths = S, select = smooth,
-                                       partial_match = partial_match)#,
-  # model_name = expr_label(substitute(object)))
-  sm_ids <- which(select)
-  smooths <- gratia::get_smooths_by_id(model, sm_ids)
-  sm_data <- map(sm_ids, gratia:::smooth_data,
-                 model = model, n = n, include_all = TRUE)
-  sm_data <- bind_rows(sm_data)
-  by_var <- by_variable(smooths[[1L]])
-  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
-  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
-                                   stringsAsFactor = FALSE))
-  names(pairs) <- paste0("f", 1:2)
-
-  Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
-  V <- gratia:::get_vcov(model, unconditional = unconditional,
-                         frequentist = frequentist)
-  coefs <- coef(model)
-
-  out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
-              smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
-              coefs = coefs, nrep = nrep)
-  out <- bind_rows(out)
-  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
-
-  out <- add_column(out,
-                    lower = out$diff - (crit * out$se),
-                    upper = out$diff + (crit * out$se),
-                    .after = 6L)
-  out
-}
-
+We first call the script \passthrough{\lstinline!pointwise\_comparisons.R!} to estimate the pointwise difference, then call the script \passthrough{\lstinline!difference\_smooths.R!} which calculates the simultaneous confidence interval for the difference, and finally \passthrough{\lstinline!pairwise\_limits.R!}, which estimates the regions to be shaded for the intervals with significance in the final figure.
+We use \passthrough{\lstinline!difference\_smooths!} on model \passthrough{\lstinline!gam\_02!} and save the estimates in \passthrough{\lstinline!diff\_complete!}.
+
+\begin{lstlisting}[language=R]
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","pointwise_comparisons.R"))
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","difference_smooths.R"))
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","pairwise_limits.R"))
 #compute difference between smooths and calculate confidence interval: complete data
@@
                      n=100, partial_match = TRUE, nrep=10000)
-
-#function to obtain values for the shading regions
-pairwise_limits<-function(dataframe){
-    #extract values where the lower limit of the ribbon is greater than zero
-    #this is the region where the control group effect is greater
-
-    v1<-dataframe%>%
-        filter(lower_s>0)%>%
-        select(Day)
-    #get day  initial value
-    init1=v1$Day[[1]]
-    #get day final value
-    final1=v1$Day[[nrow(v1)]]
-    #extract values where the value of the upper limit of the ribbon is lower than zero
-    #this corresponds to the region where the treatment group effect is greater
-    v2<-dataframe%>%
-        filter(upper_s<0)%>%
-        select(Day)
-
-    init2=v2$Day[[1]]
-    final2=v2$Day[[nrow(v2)]]
-    #store values
-    my_list<-list(init1=init1,
-                  final1=final1,
-                  init2=init2,
-                  final2=final2)
-    return(my_list)
-
-}
-
 my_list<-pairwise_limits(diff_complete)
+\end{lstlisting}
+
+Next, we plot the comparisons from \passthrough{\lstinline!diff\_complete!} and create a \passthrough{\lstinline!ggplot2!} object (c1) so we can also shade the regions of significant differences easily.
+
+\begin{lstlisting}[language=R]
 rib_col<-'#8D7D82' #color for ribbon for confidence interval
@@
 
+The resulting plot (Figure \ref{fig:pairwise-comp-workflow-fig}) shows the estimated difference in the smooths with a 95\% simultaneous CI.
+As a reminder, because we have kept the group means, we can directly see that the Treatment starts to have a significant effect around day 3.
+As therapy progresses, the effect continues, and the magnitude of the difference (\(\approx\) 40\%) at day five corresponds directly with the magnitude in the increase in \(\mbox{StO}_2\) in the group at the same time point in the data.
+
 \begin{figure}
 
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/pairwise-comp-workflow-fig-1}
+{\centering \includegraphics[width=0.75\linewidth]{Appendix_A_files/figure-latex/pairwise-comp-workflow-fig-1}
 
@@
 \end{figure}
-
-Of notice, the package \passthrough{\lstinline!gratia!} contains a function that allows to do pairwise comparisons.
-In this package, \passthrough{\lstinline!difference\_smooths!} is a function that makes the comparisons and produces Figure \ref{fig:pairwise-comp-workflow-fig} when is used on a fitted model.
-The function syntax and an example can be found at:
-
+ \FloatBarrier
+
+\hypertarget{additional-resources}{%
+\subsection{Additional resources}\label{additional-resources}}
+
+Multiple and excellent resources are available for biomedical researchers that want to gain more insight on the theory and computation of GAMs.
+Here we provide a brief list of resources that cover additional topics as well as available packages that are worth considering:
+
+\begin{itemize}
+\tightlist
+\item
+  Gavin Simpson's package \emph{gratia}, which provides convenient wrapper functions for plotting and pairwise comparisons.
+  CRAN page:
   \url{https://cran.r-project.org/web/packages/gratia/gratia.pdf}
-
-Keep in mind that this function removes the group means for the pairwise comparison so the result will not be directly interpretable with the scale of the measured response.
-
-\hypertarget{gam-and-linear-model-plots-and-missing-data}{%
-\subsection{GAM and Linear model plots and Missing data}\label{gam-and-linear-model-plots-and-missing-data}}
-
-This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\passthrough{\lstinline!gam\_02!}), linear model and GAM on data with missing observations are presented.
-Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inset are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
-
-\hypertarget{GAM-linear-plot}{%
-\subsubsection{GAM and Linear model plots}\label{GAM-linear-plot}}
-
-This code chunk creates panels B and C in Figure \ref{fig:sim-smooth-plot}.
-Note that this code uses the final GAM from the previous section (\passthrough{\lstinline!gam\_02!}), so the simulated data and the model should be generated before running this section.
-
-\begin{lstlisting}[language=R]
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-## point-wise interval
-ci <- confint(gam_02, parm = "s(Day)", partial_match = TRUE, type = "confidence")
-## simultaneous interval
-si <- confint(gam_02, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
-
-# mean shift for Treatment group
-const <- coef(gam_02)[2]
-
-#pointwise confidence interval
-ci <- ci %>%
-mutate(est = case_when(Group == "Treatment" ~ est + const,
-TRUE ~ est),
-lower = case_when(Group == "Treatment" ~ lower + const,
-TRUE ~ lower),
-upper = case_when(Group == "Treatment" ~ upper + const,
-TRUE ~ upper))
-
-#simultaneous interval
-si <- si %>%
-mutate(est = case_when(Group == "Treatment" ~ est + const,
-TRUE ~ est),
-lower = case_when(Group == "Treatment" ~ lower + const,
-TRUE ~ lower),
-upper = case_when(Group == "Treatment" ~ upper + const,
-TRUE ~ upper))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#plot smooths and pointwise and simulatenous confidence intervals for GAM
-f3<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
-geom_line(lwd = 1) +
-geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
-inherit.aes = FALSE, alpha = 0.7,
-show.legend=FALSE) +
-geom_ribbon(data = si,
-mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
-inherit.aes = FALSE, alpha = 0.4,
-show.legend=FALSE)+
-    geom_point(data=dat_sim, aes(x = Day, 
-                        y = StO2_sim, 
-                        color = Group), 
-                        size=1.5,
-                        alpha=0.6, 
-               inherit.aes = FALSE,
-               show.legend = FALSE)+
-    labs(y=expression(atop(StO[2],'(complete observations)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+thm1
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=al,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+thm1
-\end{lstlisting}
-
-\hypertarget{working-with-missing-data-in-gams}{%
-\subsection{Working with Missing data in GAMs}\label{working-with-missing-data-in-gams}}
-
-This code chunk first randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data.
-Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
-
-\begin{lstlisting}[language=R]
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-ci <- confint(mod_m1, parm = "s(Day)", partial_match = TRUE, type = "confidence")
-## simultaneous interval
-si <- confint(mod_m1, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
-
-# mean shift for group 2
-const <- coef(mod_m1)[2]
-
-ci <- ci %>%
-mutate(est = case_when(Group == "Treatment" ~ est + const,
-TRUE ~ est),
-lower = case_when(Group == "Treatment" ~ lower + const,
-TRUE ~ lower),
-upper = case_when(Group == "Treatment" ~ upper + const,
-TRUE ~ upper))
-
-si <- si %>%
-mutate(est = case_when(Group == "Treatment" ~ est + const,
-TRUE ~ est),
-lower = case_when(Group == "Treatment" ~ lower + const,
-TRUE ~ lower),
-upper = case_when(Group == "Treatment" ~ upper + const,
-TRUE ~ upper))
-
-#plot model and original data
-f6<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
-geom_line(lwd = 1) +
-geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
-inherit.aes = FALSE, alpha = 0.7,
-show.legend=FALSE) +
-geom_ribbon(data = si,
-mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
-inherit.aes = FALSE, alpha = 0.4,
-show.legend=TRUE)+
-    geom_point(data=dat_missing, aes(x = Day, 
-                        y = StO2_sim, 
-                        color = Group), 
-                            size=1.5,
-                        alpha=0.6, 
-               inherit.aes = FALSE)+
-    labs(y=expression(atop(StO[2],'(missing observations)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22))+
-        thm1
-\end{lstlisting}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/sim-smooth-plot-Appendix-1}
-
-}
-
-\caption{Simulated data and smooths for oxygen saturation in tumors.
-\textbf{A}: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-\textbf{B}: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.
-\textbf{C}: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-\textbf{D}: Smooths for the GAM model for the simulated data with missing observations (40\%).
-Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
-\end{figure}
-
-\hypertarget{pairwise-comparisons-in-gams-full-and-missing-data-cases}{%
-\subsection{Pairwise comparisons in GAMs: full and missing data cases}\label{pairwise-comparisons-in-gams-full-and-missing-data-cases}}
-
-The next code chunk reproduces Figure \ref{fig:plot-pairwise-comp}.
-Here pairwise comparisons are made for the full and missing datasets by computing the difference between the smooths and a 95\% simultaneous CI.
-
-\begin{lstlisting}[language=R]
-##Pairwise comparisons
-
-difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
-  ## make sure f1 and f2 are characters
-  f1 <-  as.character(f1)
-  f2 <-  as.character(f2)
-  cnames <- colnames(Xp)
-  ## columns of Xp associated with pair of smooths
-  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
-  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
-  ## rows of Xp associated with pair of smooths
-  r1 <- data[[by_var]] == f1
-  r2 <- data[[by_var]] == f2
-
-  ## difference rows of Xp for pair of smooths
-  X <- Xp[r1, ] - Xp[r2, ]
-
-  ## zero the cols related to other splines
- # X[, ! (c1 | c2)] <- 0
-
-  ## zero out the parametric cols
-  #X[, !grepl('^s\\(', cnames)] <- 0
-
-  ## compute difference
-  sm_diff <- drop(X %*% coefs)
-  se <- sqrt(rowSums((X %*% V) * X))
-  nr <- NROW(X)
-
-  ## Calculate posterior simulation for smooths
-  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
-  rownames(coefs_sim) <- rownames(V)
-  simDev <- X %*% coefs_sim
-  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
-  masd <- apply(absDev, 2, max)
-  crit_s <- quantile(masd, prob = 0.95, type = 8)
-
-  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
-              level_1 = rep(f1, nr),
-              level_2 = rep(f2, nr),
-              diff = sm_diff, se = se,
-              lower_s = sm_diff - crit_s * se,
-              upper_s = sm_diff + crit_s*se)
-
-  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
-  ## Only need rows associated with one of the levels
-  out <- bind_cols(out, data[r1, smooth_var])
-
-  out
-}
-
-#does both ci and si
-difference_smooths <- function(model,
-                                 smooth,
-                                 n = 100,
-                                 ci_level = 0.95,
-                                 newdata = NULL,
-                                 partial_match = TRUE,
-                                 unconditional = FALSE,
-                                 frequentist = FALSE,
-                                 nrep = 10000,
-                                 include_means = TRUE,
-                                 ...) {
-  if (missing(smooth)) {
-    stop("Must specify a smooth to difference via 'smooth'.")
-  }
-
-  # smooths in model
-  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
-  # select smooths
-  select <-
-    gratia:::check_user_select_smooths(smooths = S, select = smooth,
-                                       partial_match = partial_match)#,
-  # model_name = expr_label(substitute(object)))
-  sm_ids <- which(select)
-  smooths <- gratia::get_smooths_by_id(model, sm_ids)
-  sm_data <- map(sm_ids, gratia:::smooth_data,
-                 model = model, n = n, include_all = TRUE)
-  sm_data <- bind_rows(sm_data)
-  by_var <- by_variable(smooths[[1L]])
-  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
-  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
-                                   stringsAsFactor = FALSE))
-  names(pairs) <- paste0("f", 1:2)
-
-  Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
-  V <- gratia:::get_vcov(model, unconditional = unconditional,
-                         frequentist = frequentist)
-  coefs <- coef(model)
-
-  out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
-              smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
-              coefs = coefs, nrep = nrep)
-  out <- bind_rows(out)
-  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
-
-  out <- add_column(out,
-                    lower = out$diff - (crit * out$se),
-                    upper = out$diff + (crit * out$se),
-                    .after = 6L)
-  out
-}
-
-#compute difference between smooths and calculate confidence interval: complete data
-diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
-                     unconditional = TRUE, frequentist = FALSE,
-                     n=100, partial_match = TRUE, nrep=10000)
-
-#function to obtain values for the shading regions
-pairwise_limits<-function(dataframe){
-    #extract values where the lower limit of the ribbon is greater than zero
-    #this is the region where the control group effect is greater
-
-    v1<-dataframe%>%
-        filter(lower_s>0)%>%
-        select(Day)
-    #get day  initial value
-    init1=v1$Day[[1]]
-    #get day final value
-    final1=v1$Day[[nrow(v1)]]
-    #extract values where the value of the upper limit of the ribbon is lower than zero
-    #this corresponds to the region where the treatment group effect is greater
-    v2<-dataframe%>%
-        filter(upper_s<0)%>%
-        select(Day)
-
-    init2=v2$Day[[1]]
-    final2=v2$Day[[nrow(v2)]]
-    #store values
-    my_list<-list(init1=init1,
-                  final1=final1,
-                  init2=init2,
-                  final2=final2)
-    return(my_list)
-
-}
-
-my_list<-pairwise_limits(diff_complete)
-rib_col<-'#8D7D82' #color for ribbon for confidence interval
-control_rib <- '#875F79' #color for ribbon for control region
-treat_rib <- '#A7D89E' #color for ribbon treatment region
-
-c1 <- ggplot() +
-  geom_line(data = diff_complete, aes(x = Day, y = diff),size=1, alpha=0.5) +
-    annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill=control_rib,
-                alpha = 0.5,
-                )+
-    annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=6,
-           angle=90
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill=treat_rib,
-             alpha = 0.5
-             ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=6,
-           angle=90
-           )+
-  geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s, ymax = upper_s),
-              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
-  geom_hline(yintercept = 0, lty = 2, color = "red")+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    labs(y="Difference\n(Complete observations)")+
-    theme_classic()+
-    theme(
-    axis.text=element_text(size=22))
-
-###compute difference in smooths for missing data
-diff_missing <- difference_smooths(mod_m1, smooth = "s(Day)", newdata = newdat,
-                     unconditional = TRUE, frequentist = FALSE,
-                     n=100, partial_match = TRUE, nrep=10000)
-
-v2<- diff_missing %>%
-        filter(upper_s<0)%>%
-        select(Day)
-init2=v2$Day[[1]]
-final2=v2$Day[[nrow(v2)]]
-
-c2 <- ggplot() +
-  geom_line(data = diff_missing, aes(x = Day, y = diff),size=1, alpha=0.5) +
-    annotate("rect",
-             xmin =init2, xmax = final2,ymin=-Inf,ymax=Inf,
-             fill=treat_rib,
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=6,
-           angle=90
-           )+
-  geom_ribbon(data = diff_missing, aes(x = Day, ymin = lower_s, ymax = upper_s),
-              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
-  geom_hline(yintercept = 0, lty = 2, color = "red")+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    labs(y="Difference\n(Missing observations)")+
-    theme_classic()+
-    theme(
-    axis.text=element_text(size=22))
-
-pair_comp<-c1+c2
-\end{lstlisting}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/plot-pairwise-comp-Appendix-1}
-
-}
-
-\caption{Pairwise comparisons for smooth terms.
-\textbf{A}: Pairwise comparisons for the full dataset.
-\textbf{B}: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
-In both cases the effect of treatment is significant after day 3.
-In both cases, a simultaneous CI has been computed around the smooth difference.}\label{fig:plot-pairwise-comp-Appendix}
-\end{figure}
+\end{itemize}
+
+Vignette:
+
+\url{https://gavinsimpson.github.io/gratia/}
+
+\begin{itemize}
+\tightlist
+\item
+  Gavin Simpson's blog \emph{From the bottom of the heap}, which covers a wide range of topics in GAM modeling, news about updates on \emph{gratia} and provides tutorials on GAM fitting.
+\end{itemize}
+
+Link:
+
+\url{https://fromthebottomoftheheap.net/}
+
+\begin{itemize}
+\tightlist
+\item
+  Matteo Fasiolo's package \emph{mgcViz}, ane extension of the \emph{mgcv} package. Provides visual tools for Generalized Additive Models that exploit the additive structure of such models.
+\end{itemize}
+
+CRAN Page:
+
+\url{https://cran.r-project.org/web/packages/mgcViz/index.html}
+
+Vignette:
+\url{https://mfasiolo.github.io/mgcViz/articles/mgcviz.html}
+
+\begin{itemize}
+\tightlist
+\item
+  The book ``Generalized Additive Models: An Introduction with R'' (the 2nd.
+Edition can be found \href{https://www.routledge.com/Generalized-Additive-Models-An-Introduction-with-R-Second-Edition/Wood/p/book/9781498728331}{here}) by Simon Wood is an excellent resource for in-depth material on GAM theory, mathematical derivation, examples, and detailed descriptions on the computational aspect of \emph{mgcv}.
+\end{itemize}
 

### 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/SIM_Appendix.Rmd
New directory, knit separate appendices, separate directory for scripts

--- 2021-11-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions/SIM_Appendix.Rmd
@@
     includes:
-      in_header: latex_docs/preamble_Appendix.sty #additional LaTeX formatting
+      in_header: latex_docs/preamble_Appendix_A.sty #additional LaTeX formatting
   bookdown::word_document2:

### 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/02-Challenges.Rmd
New directory, knit separate appendices, separate directory for scripts

--- 2021-12-22 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
@@
 ```{r,include=FALSE,message=FALSE,echo=FALSE}
-## Example with linear response. This function generates either linear or quadratic mean responses with correlated or uncorrelated errors and fits a linear model to the data.
-example <- function(n_time = 6, #number of time points
-                    fun_type = "linear", #type of reponse
-                    error_type = "correlated") {
 
-  if (!(fun_type %in% c("linear", "quadratic")))
-    stop('fun_type must be either "linear", or "quadratic"')
-  if (!(error_type %in% c("correlated", "independent")))
-    stop('fun_type must be either "correlated", or "independent"')
-
-  x <- seq(1,6, length.out = n_time)
-
-  #Create mean response matrix: linear or quadratic
-  mu <- matrix(0, length(x), 2)
-  # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
-    mu[, 2] <- 0.25*x+2
-  } else {
-    # quadratic response (non-linear)
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
-  }
-
-  #create an array where individual observations per each time point for each group are to be stored. Currently using 10 observations per timepoint
-  y <- array(0, dim = c(length(x), 2, 10))
-
-  #Create array to store the "errors" for each group at each timepoint. The "errors" are the 
-  #between-group variability in the response.
-  errors <- array(0, dim = c(length(x), 2, 10))
-  #create an array where 10 observations per each time point for each group are to be stored
-
-  #The following cycles create independent or correlated responses.
-To each value of mu (mean response per group) a randomly generated error (correlated or uncorrelated) is added and thus the individual response is created.
-  if (error_type == "independent") {
-    ## independent errors
-    for (i in 1:2) {
-      for (j in 1:10) {
-        errors[, i, j] <- rnorm(6, 0, 0.25)
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  } else {
-    for (i in 1:2) {     # number of treatments
-      for (j in 1:10) {  # number of subjects
-        # compound symmetry errors: variance covariance matrix
-        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-        y[, i, j] <- mu[, i] + errors[, i, j]
-      }
-    }
-  }
-
-  ## subject random effects
-
-  ## visualizing the difference between independent errors and compound symmetry
-  ## why do we need to account for this -- overly confident inference
-
-#labelling y and errors  
-  dimnames(y) <- list(time = x, 
-                      treatment = 1:2, 
-                      subject = 1:10)
-
-  dimnames(errors) <- list(time = x, 
-                           treatment = 1:2, 
-                           subject = 1:10)
-
-  #labeling the mean response
-  dimnames(mu) <- list(time = x, 
-                       treatment = 1:2)
-
-  #convert y, mu and errors to  dataframes with time, treatment and subject columns
-  dat <- as.data.frame.table(y, 
-                             responseName = "y")
-  dat_errors <- as.data.frame.table(errors, 
-                                    responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, 
-                                responseName = "mu")
-
-  #join the dataframes to show mean response and errors per subject
-  dat <- left_join(dat, dat_errors, 
-                   by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, 
-                   by = c("time", "treatment"))
-  #add time
-  dat$time <- as.numeric(as.character(dat$time))
-  #label subjects per group
-  dat <- dat %>%
-    mutate(subject = factor(paste(subject, 
-                                  treatment, 
-                                  sep = "-")))
-
-   ## repeated measures ANOVA in R
-#time and treatment interaction model
-  fit_anova <- lm(y ~ time + treatment + time * treatment, data = dat)
-
-  #LMEM with compound symmetry
-
-  fit_lme <- lme(y ~ treatment + time + treatment:time,
-                 data = dat,
-                 random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
-
-  #create a prediction frame where the model can be used for plotting purposes
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
-
-  #add model predictions to the dataframe that has the simulated data
-  dat$pred_anova <- predict(fit_anova)
-  dat$pred_lmem <- predict(fit_lme)
-
-  #return everything in a list
-  return(list(
-    dat = dat,
-    pred_dat = pred_dat,
-    fit_lme = fit_lme,
-    fit_anova=fit_anova
-
-  ))
-}
-##################Section for plotting#################################
-#######################################################################
-#This function will create the plots for either a "linear" or "quadratic" response
-
-plot_example <- function(sim_dat) {
-  ## Plot the simulated data (scatterplot)
-  p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = treatment, 
-               color = treatment)
-           ) +
-    geom_point(alpha=0.5,
-               show.legend=FALSE) +
-    labs(y='response')+
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment),
-              size=3,
-              show.legend=FALSE) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 20, 
-                                  face = "bold"),
-        text=element_text(size=20))+
-    thm1
-
-   #plot the model predictions for rm-ANOVA
-  p2 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point(alpha=0.5,
-               show.legend=FALSE)+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_anova), 
-                  group = subject, size = "Subjects"),
-              show.legend = FALSE) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_anova, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population"),
-              show.legend=FALSE) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 20, 
-                                  face = "bold"),
-        text=element_text(size=20))+
-    thm1
-
-   #plot the model predictions for LMEM
-  p4 <- ggplot(sim_dat$dat, 
-               aes(x = time, 
-                   y = y, 
-                   color = treatment)) +
-    geom_point(alpha=0.5)+
-    labs(y='response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme), 
-                  group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat, 
-              aes(y = predict(sim_dat$fit_lme, 
-                              level = 0, 
-                              newdata = sim_dat$pred_dat), 
-                  size = "Population")) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions", 
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = 20, 
-                                  face = "bold"),
-        text=element_text(size=20))+
-    thm1
-
-  return((p1+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-
-}
+#calls the scripts example.R and plot_example.R
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","example.R"))
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","plot_example.R"))
 
@@
 ```{r, l-q-response, fig.width=12, fig.height=7, out.width='100%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='(ref:l-q-response-caption)'}
-# linear response, correlated errors (subject effect)
-#par(mar = c(0.5, 0.5, 0.5, 0.5))
 A/C+plot_annotation(tag_levels = 'A')

### 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/03-GAM Theory.Rmd
New directory, knit separate appendices, separate directory for scripts

--- 2021-12-22 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
@@
 
-```{r,basis-functions-plot, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-
- #label each table
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
-
- #Convert to dataframes with subject, time and group columns
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
-
- #label subject per group
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
- #extract  "Group 1" to fit the GAM
-  dat<-subset(dat,treatment==1)
- #keep just the response and timepoint columns
-   dat<-dat[,c('y','time')]
-
-   #GAM model of time, 5 basis functions
-gm<-gam(y~s(time,k=5),data=dat, method= "REML")
-
-#model_matrix (also known as) 'design matrix'
-#will contain the smooths used to create  model 'gm'
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of basis by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be 
-#multiplied by the coefficients
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final smooth by aggregating the basis functions
-
-smooth<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(smooth=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,
-            aes(x=time,
-                y=value_orig,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(size=sz,
-            show.legend=FALSE)+
-  geom_point(size=sz+1,
-             show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,
-             labeller = as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm1
-
-#penalized basis
-p12<-ggplot(basis_plot,
-            aes(x=time,
-                y=mod_val,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(show.legend = FALSE,
-            size=sz)+
-  geom_point(show.legend = FALSE,
-             size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,
-             labeller=as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm1
-
-#heatmap of the  coefficients
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,
-            aes(x=Basis,
-                y=Basis))+
-  geom_tile(aes(fill = cof), 
-            colour = "black") +
-    scale_fill_gradient(low = "white",
-                        high = "#B50A2AFF")+ 
-  labs(x='Basis',
-       y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),
-            size=7,
-            show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,
-            aes(x=time,y=y))+
-  geom_point(size=sz+1,alpha=0.5)+
-  thm1+
-  labs(y='Simulated \n response')+
-  geom_line(data=smooth,
-            aes(x=time,
-                y=smooth),
-            color="#6C581DFF",
-            size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-
-```
-
 To get the weighted basis functions, each basis (from Figure \@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients in Figure \@ref(fig:basis-plot)B, thereby increasing or decreasing the original basis functions.
@@
 par(mar = c(2, 2, 2, 2))
-
+#calls the object b_plot from script basis_functions.R which contains the composite figure
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","basis_functions.R"))
 b_plot

### 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/04-Longitudinal analysis with GAMs.Rmd
New directory, knit separate appendices, separate directory for scripts

--- 2021-12-22 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
@@
 ```{r,sim-tumor-response,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
+
+#data from Vishwanath et. al.
 dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
@@
 
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
 #subject = factor(paste(subject, treatment, sep = "-")))
@@
 df <- 6
+
+#calls simulate_data.R to generate simulated responses below
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","simulate_data.R"))
+
 dat_sim <- simulate_data(dat, n, sd)
@@
 
-##Pairwise comparisons
+##Pairwise comparisons: calls pointwise_comparisons.R and difference_smooths.R to create the 
+#confidence intervals (pointwise and simultaneous) for a pairwise comparison between the smooths.
 
-difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
-  ## make sure f1 and f2 are characters
-  f1 <-  as.character(f1)
-  f2 <-  as.character(f2)
-  cnames <- colnames(Xp)
-  ## columns of Xp associated with pair of smooths
-  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
-  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
-  ## rows of Xp associated with pair of smooths
-  r1 <- data[[by_var]] == f1
-  r2 <- data[[by_var]] == f2
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","pointwise_comparisons.R"))
 
-  ## difference rows of Xp for pair of smooths
-  X <- Xp[r1, ] - Xp[r2, ]
-
-  ## zero the cols related to other splines
- # X[, ! (c1 | c2)] <- 0
-
-  ## zero out the parametric cols
-  #X[, !grepl('^s\\(', cnames)] <- 0
-
-  ## compute difference
-  sm_diff <- drop(X %*% coefs)
-  se <- sqrt(rowSums((X %*% V) * X))
-  nr <- NROW(X)
-
-  ## Calculate posterior simulation for smooths
-  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
-  rownames(coefs_sim) <- rownames(V)
-  simDev <- X %*% coefs_sim
-  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
-  masd <- apply(absDev, 2, max)
-  crit_s <- quantile(masd, prob = 0.95, type = 8)
-
-  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
-              level_1 = rep(f1, nr),
-              level_2 = rep(f2, nr),
-              diff = sm_diff, se = se,
-              lower_s = sm_diff - crit_s * se,
-              upper_s = sm_diff + crit_s*se)
-
-  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
-  ## Only need rows associated with one of the levels
-  out <- bind_cols(out, data[r1, smooth_var])
-
-  out
-}
-
-#does both ci and si
-difference_smooths <- function(model,
-                                 smooth,
-                                 n = 100,
-                                 ci_level = 0.95,
-                                 newdata = NULL,
-                                 partial_match = TRUE,
-                                 unconditional = FALSE,
-                                 frequentist = FALSE,
-                                 nrep = 10000,
-                                 include_means = TRUE,
-                                 ...) {
-  if (missing(smooth)) {
-    stop("Must specify a smooth to difference via 'smooth'.")
-  }
-
-  # smooths in model
-  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
-  # select smooths
-  select <-
-    gratia:::check_user_select_smooths(smooths = S, select = smooth,
-                                       partial_match = partial_match)#,
-  # model_name = expr_label(substitute(object)))
-  sm_ids <- which(select)
-  smooths <- gratia::get_smooths_by_id(model, sm_ids)
-  sm_data <- map(sm_ids, gratia:::smooth_data,
-                 model = model, n = n, include_all = TRUE)
-  sm_data <- bind_rows(sm_data)
-  by_var <- by_variable(smooths[[1L]])
-  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
-  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
-                                   stringsAsFactor = FALSE))
-  names(pairs) <- paste0("f", 1:2)
-
-  Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
-  V <- gratia:::get_vcov(model, unconditional = unconditional,
-                         frequentist = frequentist)
-  coefs <- coef(model)
-
-  out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
-              smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
-              coefs = coefs, nrep = nrep)
-  out <- bind_rows(out)
-  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
-
-  out <- add_column(out,
-                    lower = out$diff - (crit * out$se),
-                    upper = out$diff + (crit * out$se),
-                    .after = 6L)
-  out
-}
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","difference_smooths.R"))
 
@@
 
-#function to obtain values for the shading regions
-pairwise_limits<-function(dataframe){
-    #extract values where the lower limit of the ribbon is greater than zero
-    #this is the region where the control group effect is greater
-
-    v1<-dataframe%>%
-        filter(lower_s>0)%>%
-        select(Day)
-    #get day  initial value
-    init1=v1$Day[[1]]
-    #get day final value
-    final1=v1$Day[[nrow(v1)]]
-    #extract values where the value of the upper limit of the ribbon is lower than zero
-    #this corresponds to the region where the treatment group effect is greater
-    v2<-dataframe%>%
-        filter(upper_s<0)%>%
-        select(Day)
-
-    init2=v2$Day[[1]]
-    final2=v2$Day[[nrow(v2)]]
-    #store values
-    my_list<-list(init1=init1,
-                  final1=final1,
-                  init2=init2,
-                  final2=final2)
-    return(my_list)
-
-}
-
-my_list<-pairwise_limits(diff_complete)
+####These are the aesthetics for the components of the plot
 rib_col<-'#8D7D82' #color for ribbon for confidence interval
@@
 
+#calls the function in the script pairwise_limits.R 
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","pairwise_limits.R"))
+
+my_list<-pairwise_limits(diff_complete)
+
+#plot the pairwise comparisons
 c1 <- ggplot() +

### 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/05-Discussion.Rmd
New directory, knit separate appendices, separate directory for scripts

# Discussion

Biomedical longitudinal data is particularly challenging to analyze due to the frequency of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA. Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \@ref(simulation). When these models do not capture the non-linear trend of the data, this results in a "model misspecification error". This "model misspecification" error, also is known as a "Type III" error [@dennis2019] is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.

Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) [@lang2015] but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a $R^2$ or _p-value_ value, whose interpretation is not clear in the case of a Type III error.

In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.  
It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (e.g., which are palaeoecology, geochemistry, and ecology) [@hefley2017; @pedersen2019] they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research [@hastie1995]. This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs. T

The purpose of Section \@ref(GAM-theory) is to demonstrate that at its core the principle  underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship _a priori_ in order to provide it to the model, such as in the case of polynomial effects in LMEMs. In contrast, because GAMs let the data speak for themselves they can provide an accurate representation of the effect of time in a biological process.

However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented. We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature [@vishwanath2009] using R and the package _mgcv_[@wood2017] in Section \@ref(longitudinal-GAMs), while a basic workflow for model selection is in the [Appendix](#workflow).

One of the features of GAMs is that their Bayesian interpretation allows for inference about differences between groups without the need of a _p-value_, and in addition can provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure  \@ref(fig:plot-pairwise-comp) indicate. The model is therefore able to identify changes between the groups at time points were data was not directly measured even in the case of incomplete observations ( $\approx$ day 3 in Figure \@ref(fig:plot-pairwise-comp) A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.

We have used R as the software of choice for this paper because R not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible. In this regard, reproducibility is still an issue in biomedical research [@begley2015;@weissgerber2018], but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field. Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research. We have made all the data and code used in this paper accessible, and we hope that this will encourage other researchers to do the same with future projects.

### 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts/example.R

@@ -1,0 +1,130 @@
+## Example with linear response. This function generates either linear or quadratic mean
+#responses with correlated or uncorrelated errors and fits a linear model to the data.
+
+example <- function(n_time = 6, #number of time points
+                    fun_type = "linear", #type of reponse
+                    error_type = "correlated") {
+
+    if (!(fun_type %in% c("linear", "quadratic")))
+        stop('fun_type must be either "linear", or "quadratic"')
+    if (!(error_type %in% c("correlated", "independent")))
+        stop('fun_type must be either "correlated", or "independent"')
+
+
+    x <- seq(1,6, length.out = n_time)
+
+    #Create mean response matrix: linear or quadratic
+    mu <- matrix(0, length(x), 2)
+    # linear response
+    if (fun_type == "linear") {
+        mu[, 1] <- - (0.25*x)+2
+        mu[, 2] <- 0.25*x+2
+    } else {
+        # quadratic response (non-linear)
+
+        mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
+        mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
+    }
+
+    #create an array where individual observations per each time point for each group are to be stored. Currently using 10 observations per timepoint
+    y <- array(0, dim = c(length(x), 2, 10))
+
+    #Create array to store the "errors" for each group at each timepoint. The "errors" are the
+    #between-group variability in the response.
+    errors <- array(0, dim = c(length(x), 2, 10))
+    #create an array where 10 observations per each time point for each group are to be stored
+
+    #The following cycles create independent or correlated responses. To each value of mu (mean response per group) a randomly generated error (correlated or uncorrelated) is added and thus the individual response is created.
+    if (error_type == "independent") {
+        ## independent errors
+        for (i in 1:2) {
+            for (j in 1:10) {
+                errors[, i, j] <- rnorm(6, 0, 0.25)
+                y[, i, j] <- mu[, i] + errors[, i, j]
+            }
+        }
+    } else {
+        for (i in 1:2) {     # number of treatments
+            for (j in 1:10) {  # number of subjects
+                # compound symmetry errors: variance covariance matrix
+                errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
+                y[, i, j] <- mu[, i] + errors[, i, j]
+            }
+        }
+    }
+
+
+    ## subject random effects
+
[…]

### 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts/plot_example.R

@@ -1,0 +1,81 @@
+
+## This function plots the rm-ANOVA and LMEM for the data simulated in example.R
+plot_example <- function(sim_dat) {
+    # Plot the simulated data (scatterplot)
+    p1 <- sim_dat$dat %>%
+        ggplot(aes(x = time,
+                   y = y,
+                   group = treatment,
+                   color = treatment)
+        ) +
+        geom_point(alpha=0.5,
+                   show.legend=FALSE) +
+        labs(y='response')+
+        geom_line(aes(x = time,
+                      y = mu,
+                      color = treatment),
+                  size=3,
+                  show.legend=FALSE) +
+        theme_classic() +
+        theme(plot.title = element_text(size = 20,
+                                        face = "bold"),
+              text=element_text(size=20))+
+        thm1
+
+
+    #plot the model predictions for rm-ANOVA
+    p2 <- ggplot(sim_dat$dat,
+                 aes(x = time,
+                     y = y,
+                     color = treatment)) +
+        geom_point(alpha=0.5,
+                   show.legend=FALSE)+
+        labs(y='response')+
+        geom_line(aes(y = predict(sim_dat$fit_anova),
+                      group = subject, size = "Subjects"),
+                  show.legend = FALSE) +
+        geom_line(data = sim_dat$pred_dat,
+                  aes(y = predict(sim_dat$fit_anova,
+                                  level = 0,
+                                  newdata = sim_dat$pred_dat),
+                      size = "Population"),
+                  show.legend=FALSE) +
+        guides(color = guide_legend(override.aes = list(size = 2)))+
+        scale_size_manual(name = "Predictions",
+                          values=c("Subjects" = 0.5, "Population" = 3)) +
+        theme_classic() +
+        theme(plot.title = element_text(size = 20,
+                                        face = "bold"),
+              text=element_text(size=20))+
+        thm1
+
+
+    #plot the model predictions for LMEM
+    p4 <- ggplot(sim_dat$dat,
+                 aes(x = time,
+                     y = y,
+                     color = treatment)) +
+        geom_point(alpha=0.5)+
+        labs(y='response')+
+        geom_line(aes(y = predict(sim_dat$fit_lme),
+                      group = subject, size = "Subjects")) +
+        geom_line(data = sim_dat$pred_dat,
+                  aes(y = predict(sim_dat$fit_lme,
+                                  level = 0,
[…]

### 2021-12-24 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/Full_document_SIM_No_Appendix.tex
Working on Appendix B

@@
 ]{article}
+\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}}
+\date{\vspace{-2.5em}}
+
 \usepackage{amsmath,amssymb}
@@
 
-\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}}
-\date{\vspace{-2.5em}}
-
 \begin{document}
@@
 Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points, with subjects of analysis ranging from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}}, antibody expression\textsuperscript{\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}}, and cell metabolism\textsuperscript{\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}} are examples of different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals\textsuperscript{\protect\hyperlink{ref-greening2018}{9}}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients\textsuperscript{\protect\hyperlink{ref-sio2016}{10}}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients\textsuperscript{\protect\hyperlink{ref-kamstra2015}{11}}.
+Tumor response,\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}} antibody expression,\textsuperscript{\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}} and cell metabolism\textsuperscript{\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}} are examples of different situations where researchers have used longitudinal designs to study some physiological response.
+Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals\textsuperscript{\protect\hyperlink{ref-greening2018}{9}}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients,\textsuperscript{\protect\hyperlink{ref-sio2016}{10}} to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients.\textsuperscript{\protect\hyperlink{ref-kamstra2015}{11}}
 
 Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times\textsuperscript{\protect\hyperlink{ref-wagenmakers2008}{12}}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations})\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}}.
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
-This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
+The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times,\textsuperscript{\protect\hyperlink{ref-wagenmakers2008}{12}} and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
+The rm-ANOVA model makes three assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}).\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}}
+
+The expected linear behavior of the response through time is a key requisite in rm-ANOVA.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
 In longitudinal biomedical research, non-linear trends are the norm rather than the exception.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}}.
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
+A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings.\textsuperscript{\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}} These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
 Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
@@
 In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}}.
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid\textsuperscript{\protect\hyperlink{ref-wang2019}{18}}.
-
-Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\))\textsuperscript{\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}}, consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni\textsuperscript{\protect\hyperlink{ref-abdi2010}{21}}), but they in turn reduce statistical power (1-\(\beta\))\textsuperscript{\protect\hyperlink{ref-nakagawa2004}{22}}, and lead to increased Type II error (failing to reject the null hypothesis when it is false)\textsuperscript{\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}}.
-Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases\textsuperscript{\protect\hyperlink{ref-ugrinowitsch2004}{25}}.
-Corrections can be made in rm-ANOVA in the absence of compound symmetry\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and group sizes\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}}.
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons,\textsuperscript{\protect\hyperlink{ref-charan2013}{30}} which might cause the corrections for lack of compound symmetry to be ineffective.
+For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid.\textsuperscript{\protect\hyperlink{ref-wang2019}{18}}
+
+Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)),\textsuperscript{\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}} consequently biasing the conclusions of the study.
+Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni),\textsuperscript{\protect\hyperlink{ref-abdi2010}{21}} but they in turn reduce statistical power (1-\(\beta\))\textsuperscript{\protect\hyperlink{ref-nakagawa2004}{22}}, and lead to increased Type II error (failing to reject the null hypothesis when it is false).\textsuperscript{\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}} Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
+
+On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases.\textsuperscript{\protect\hyperlink{ref-ugrinowitsch2004}{25}} Corrections can be made in rm-ANOVA in the absence of compound symmetry,\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}} but the effectiveness of the correction is limited by the size of the sample, the number of measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and group sizes.\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}} In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons,\textsuperscript{\protect\hyperlink{ref-charan2013}{30}} which might cause the corrections for lack of compound symmetry to be ineffective.
 
@@
 In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13}}.
-This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
+When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13}} This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
 At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
@@
 During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data\textsuperscript{\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}}.
-Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
-However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}}.
-And even more importantly, LMEMs also assume by default a linear relationship between the response and time\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} (polynomial effects can be used, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .
+The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data.\textsuperscript{\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}} Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}} And even more importantly, LMEMs also assume by default a linear relationship between the response and time\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} (polynomial effects can be used, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .
 
@@
 Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data\textsuperscript{\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}}, health-environment interactions\textsuperscript{\protect\hyperlink{ref-yang2012}{35}} and the dynamics of government in political science\textsuperscript{\protect\hyperlink{ref-beck1998}{36}}.
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, and 3) can easily accommodate incomplete observations.
+Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data,\textsuperscript{\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}} health-environment interactions\textsuperscript{\protect\hyperlink{ref-yang2012}{35}} and the dynamics of government in political science.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements,\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and 3) can easily accommodate incomplete observations.
 Therefore, GAMs provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
@@
 Data simulation methods are an emerging technique that allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation.
-In addition, simulation can be also used for power calculations and study design questions\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}.
+In addition, simulation can be also used for power calculations and study design questions.\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}
 
@@
 Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}}.
-The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
+Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies.\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
 Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in R, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
@@
 Independent errors \(\varepsilon_{ijt}\) represent random variation from the sampling process assumed to be \(\stackrel{i.i.d.}\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
-In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'', or ``saline'' vs.~``chemotherapy'').
+In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug,'' or ``saline'' vs.~``chemotherapy'').
 Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B).
@@
 
-Of note, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach (using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}), which is described in detail by Pinheiro and Bates\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
-However, polynomial fits have limited predictive power, cause bias on the boundaries of the covariates\textsuperscript{\protect\hyperlink{ref-beck1998}{36}}, and more importantly, their lack of biological or mechanistic interpretation limits their use in biomedical studies\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
+Of note, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach (using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}), which is described in detail by Pinheiro and Bates.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} However, polynomial fits have limited predictive power, cause bias on the boundaries of the covariates,\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} and more importantly, their lack of biological or mechanistic interpretation limits their use in biomedical studies.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}
 
@@
 In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}}.
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time\textsuperscript{\protect\hyperlink{ref-maxwell2017}{44}}; and is higher between consecutive measurements\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}}.
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}} their effectiveness is dependent on sample size and number of repeated measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and it has been shown that corrections are not robust if the group sizes are unbalanced\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}}.
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
+Under this assumption, the between-subject variance and within-subject correlation are constant across time.\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}} However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time;\textsuperscript{\protect\hyperlink{ref-maxwell2017}{44}} and is higher between consecutive measurements.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}} Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)\textsuperscript{\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}} their effectiveness is dependent on sample size and number of repeated measurements\textsuperscript{\protect\hyperlink{ref-haverkamp2017}{28}}, and it has been shown that corrections are not robust if the group sizes are unbalanced.\textsuperscript{\protect\hyperlink{ref-keselman2001}{29}} Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
+
+In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
 Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
@@
 In a longitudinal study, it is frequently the case that the number of observations is different across subjects.
-In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing'', ``incomplete'', or (more generally speaking) unbalanced data\textsuperscript{\protect\hyperlink{ref-molenberghs2004}{45}}.
-The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population and statistical power is affected by this reduction in sample size\textsuperscript{\protect\hyperlink{ref-ma2012}{46}}.
-
-On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are \emph{missing at random} (MAR) or \emph{completely missing at random} (MCAR)\textsuperscript{\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest\textsuperscript{\protect\hyperlink{ref-scheffer2002}{47}}.
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information\textsuperscript{\protect\hyperlink{ref-potthoff2006}{48}}.
-Missing observations can also be \emph{missing not at random} (MNAR) and in the case the missing observations are dependent on their value.
+In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing,'' ``incomplete,'' or (more generally speaking) unbalanced data.\textsuperscript{\protect\hyperlink{ref-molenberghs2004}{45}} The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
+This is inconvenient because the remaining subjects might not accurately represent the population and statistical power is affected by this reduction in sample size.\textsuperscript{\protect\hyperlink{ref-ma2012}{46}}
+
+On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are \emph{missing at random} (MAR) or \emph{completely missing at random} (MCAR).\textsuperscript{\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}} In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest.\textsuperscript{\protect\hyperlink{ref-scheffer2002}{47}} If the data are MCAR, this means that the missingness is completely unrelated to the collected information.\textsuperscript{\protect\hyperlink{ref-potthoff2006}{48}} Missing observations can also be \emph{missing not at random} (MNAR) and in the case the missing observations are dependent on their value.
 For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
@@
 To visually demonstrate the limitations of rm-ANOVA and LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group using R\textsuperscript{\protect\hyperlink{ref-r}{38}} and the package \emph{nlme}\textsuperscript{\protect\hyperlink{ref-nlme}{50}}.
+An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group using R\textsuperscript{\protect\hyperlink{ref-r}{38}} and the package \emph{nlme}.\textsuperscript{\protect\hyperlink{ref-nlme}{50}}
 
@@
 If the residuals are non-normal, a transformation is necessary in order to properly fit the model.
-However, transformation can lead to poor model performance\textsuperscript{\protect\hyperlink{ref-ohara2010}{51}}, and can cause problems with the biological interpretation of the model estimates.
+However, transformation can lead to poor model performance,\textsuperscript{\protect\hyperlink{ref-ohara2010}{51}} and can cause problems with the biological interpretation of the model estimates.
 McCullagh and Nelder\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} introduced General Linear Models (GLMs) as an extension of LMs, where the errors do not need to be normally distributed.
@@
 Although GLMs relax the normality assumption, they only accommodate fixed effects.
-Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate \emph{random effects}, which have an associated probability distribution\textsuperscript{\protect\hyperlink{ref-mcculloch2001}{53}}.
-Therefore, in GLMMs the linear predictor takes the form
+Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate \emph{random effects}, which have an associated probability distribution.\textsuperscript{\protect\hyperlink{ref-mcculloch2001}{53}} Therefore, in GLMMs the linear predictor takes the form
 
@@
 where \(\alpha_{ij}\) corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of Equation \eqref{eq:GLM-eta}.
-Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}}, and GLMs are a special case of GLMMs where there are no random effects.
-In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in Dobson\textsuperscript{\protect\hyperlink{ref-dobson2008}{54}} and Stroup\textsuperscript{\protect\hyperlink{ref-stroup2013}{55}}.
+Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed,\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} and GLMs are a special case of GLMMs where there are no random effects.
+In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in Dobson\textsuperscript{\protect\hyperlink{ref-dobson2008}{54}} and Stroup.\textsuperscript{\protect\hyperlink{ref-stroup2013}{55}}
 
@@
 Notice that in the previous sections, the difference between GLMs and GLMMs resides on their linear predictors (Equations \eqref{eq:GLM-eta}, \eqref{eq:GLMM-eta}).
-Generalized additive models (GAMs) are an extension of the GLM family that allow the estimation of smoothly varying trends where the relationship between the covariates and the response is modeled using \emph{smooth functions}\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}}.
-In a GAM, the linear predictor has the form
+Generalized additive models (GAMs) are an extension of the GLM family that allow the estimation of smoothly varying trends where the relationship between the covariates and the response is modeled using \emph{smooth functions}.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}} In a GAM, the linear predictor has the form
 
@@
 where \(\beta_{0}\) is the intercept, and \(\beta_{1}\) is the coefficient for each treatment group.
-Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time, and the interaction between time and treatment have been replaced by \(f(time_{t}|\beta_{j})\).
-The smooth term \(f(time_{t}|\beta_{j})\) gives a different smooth response for each treatment.
-If the smooth term represented a linear relationship, then \(f(time_{t}|\beta_{j})= \beta_2 \times time_t+\beta_3 \times time_t \times treatment_j\); however, in general, the smooth term is a more flexible function than a linear relationship, with parameter vectors \(\beta_{j}\) for each treatment.
-A GAM version of a linear model can be written as
+Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time, and the interaction between time and treatment have been replaced by the smooth term \(f(time_{t}|\beta_{j})\).
+The smooth term \(f(time_{t}|\beta_{j})\) gives a different smooth response for each treatment..\footnote{If the smooth term represented a linear relationship, then \(f(time_{t}|\beta_{j})= \beta_2 \times time_t+\beta_3 \times time_t \times treatment_j\); however, in general, the smooth term is a more flexible function than a linear relationship, with parameter vectors \(\beta_{j}\) for each treatment} A GAM version of a linear model can be written as
 
@@
 In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is advantageous as it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(time_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference\textsuperscript{\protect\hyperlink{ref-beck1998}{36}}.
-In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(time_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
+One possible set of functions for \(f(time_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(time_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
 
 The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, using a \emph{basis function} expansion of the covariates and estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}}.
-For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(treatment_j\), \(time_t\), and \(time_t \times treatment_j\).
-The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
+A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(treatment_j\), \(time_t\), and \(time_t \times treatment_j\).
+The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
 In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(time_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
 
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions that have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness\textsuperscript{\protect\hyperlink{ref-wegman1983}{58}}.
-Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
+Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions that have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{58}} Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
 
 Cubic splines (CS) are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework\textsuperscript{\protect\hyperlink{ref-hastie1987}{56}}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots', the points where the basis functions meet) to obtain the finite basis, which affects model fit\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}.
-A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}.
-
-In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}.
-Like TPS, TPRS only requires specifying the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}D.
+The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{56}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the basis functions meet) to obtain the finite basis, which affects model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate.\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}
+
+In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an computationally efficient.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} Like TPS, TPRS only requires specifying the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).
+
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 that appears in Figure \ref{fig:l-q-response}D.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
@@
 A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline.
-The second derivative of the spline is a measure of curvature, or the rate of change of the slope\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}}, and increasing the penalty by increasing \(\lambda\) results in models with less curvature.
+The second derivative of the spline is a measure of curvature, or the rate of change of the slope,\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} and increasing the penalty by increasing \(\lambda\) results in models with less curvature.
 As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
@@
 
-With this in mind, if five basis functions are used to fit a GAM for the data of Group 1 (concave down function) that appears in Figure \ref{fig:l-q-response}D, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A, where the four basis functions (and the intercept) are shown.
+With this in mind, if four basis functions (plus the intercept) are used to fit a GAM for the data of Group 1 (concave down function) that appears in Figure \ref{fig:l-q-response}D, the resulting fitting process is shown in Figure \ref{fig:basis-plot}.
+In Figure \ref{fig:basis-plot}A the four basis functions (and the intercept) are shown.
 Each of the five basis functions is evaluated at six different points (because there are six points on the timeline).
 The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), that is estimated when fitting the model.
-The penalized coefficient estimates fitted for our example are shown in Figure \ref{fig:basis-plot}B.
-
-To get the weighted basis functions, each basis (from Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
-Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
+The penalized coefficient estimates fitted are shown in Figure \ref{fig:basis-plot}B.
+
+To get the weighed basis functions, each basis (from Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
+Figure \ref{fig:basis-plot}C shows the resulting weighed basis functions.
+Note that the magnitude of the weighing for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
 On the other hand, the third basis function has roughly doubled its contribution to the smooth term.
 Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (brown line), along the simulated values per group, which appear as points.
+The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (brown line), along the simulated values which appear as points.
 
@@
 The coefficient determines the overall effect of each basis in the final smoother.
-\textbf{C}: Weighted basis functions.
-Each of the five basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
+\textbf{C}: Weighed basis functions.
+Each of the four basis functions (and the intercept) of panel A has been weighed by the corresponding coefficient shown in Panel B.
 Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
@@
 
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge and observed data\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}}.
-In the case of data that shows non-linear trends, the prior that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
-Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language)\textsuperscript{\protect\hyperlink{ref-miller2019}{61}}.
-Therefore, the confidence intervals (CIs) calculated by default for the smooth terms using \emph{mgcv} are considered empirical Bayesian credible intervals\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}}, which have good \emph{across the function} (``frequentist'') coverage\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
+Bayes' theorem states that the probability of an event can be calculated using prior knowledge and observed data.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}} In the case of data that shows non-linear trends, the prior that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Therefore, the confidence intervals (CIs) calculated by default for the smooth terms using \emph{mgcv} are considered empirical Bayesian credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good \emph{across the function} (``frequentist'') coverage\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
 
@@
 
-Besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.
-Suppose we chose a nominal value (say, 95\%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
-Briefly, to obtain a simultaneous CI we use simulation from the empirical Bayesian posterior distribution to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}.
-Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
-
-In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller\textsuperscript{\protect\hyperlink{ref-miller2019}{61}}, Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}}, Marra\textsuperscript{\protect\hyperlink{ref-marra2012}{63}}, and Ruppert\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}.
-With this brief introduction to the Bayesian interpretation of GAMs, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends.
+Besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Suppose we chose a nominal value (say, 95\%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
+Briefly, to obtain a simultaneous CI we use simulation from the empirical Bayesian posterior distribution to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI.\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}
+
+In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller,\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson,\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} Marra,\textsuperscript{\protect\hyperlink{ref-marra2012}{63}} and Ruppert.\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}} What we want to convey is that a Bayesian interpretation of GAMs allows for robust estimation using simultaneous empirical Bayesian CIs, as their stimates can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
+
+With this in mind, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends, and use simultaneous empirical Bayesian CIs to assess significance between treatment groups.
 
@@
 This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}.
-The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
+A brief guide for model selection and diagnostics appears in Appendix A, and the code for the simulated data and figures can be found in Appendix B.
 
@@
 A detailed description on the close relationship between basis functions and correlation functions can be found in Hefley et.
-al.\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}}.
+al..\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}}
 
@@
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}\textsuperscript{\protect\hyperlink{ref-gratia}{64}}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak\textsuperscript{\protect\hyperlink{ref-harezlak2018}{65}}.
-
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?''.
-The fit of an rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model}, is presented in Figure \ref{fig:sim-smooth-plot}C.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}\textsuperscript{\protect\hyperlink{ref-gratia}{64}} as we show in Appendix A.
+Additional discussions on model selection can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{65}}
+
+One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The fit of an rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model}, is presented in Figure \ref{fig:sim-smooth-plot}C.
 This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
@@
 Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.
 
@@
 
-The absence of a \emph{p-value} in this context might seem odd, but its interpretation can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment.
-This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
-
-With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant.
+The absence of a \emph{p-value} in this context might seem odd, but its interpretation of the comparison can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
+
+With this expectation of different trends in each group, computing the difference between the trends will identify if and \emph{when} the observed difference is significant.
 A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
+Therefore, the estimation of the difference between the smooths and the correspondent simultaneous empirical Bayesian CI of this difference provides an estimation of significance that is more robust that a mere \emph{p-value} because the estimated difference is able to indicates the significant time interval and we know that on average, the correction provided by the simultaneous empirical Bayesian CI ensures that it will contain \emph{the whole function} at a nominal value (say, 95\%).
 
 Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}B, D.
-Figure \ref{fig:plot-pairwise-comp} shows the comparison between each treatment group for the full and missing datasets with a simultaneous CI computed around the difference.
+Figure \ref{fig:plot-pairwise-comp} shows the difference between each treatment group trend for the full and missing datasets with a simultaneous CI computed around the difference.
 Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared.
-Notice that because we have included the means of each group, there is correspondence between the scale of the original data the scale of the pairwise comparisons.
-This can be seen as in Figure \ref{fig:sim-smooth-plot}B at day 5 there is essentially a 50\% difference between \(\mbox{StO}_2\) in both groups, which corresponds to the -50 difference in \ref{fig:plot-pairwise-comp}A.
-However, if there are multiple parametric terms in the model (more factors that need to be specified in the model) such inclusion of the means can become problematic.
+Notice that because we have included the means of each group, there is correspondence between the scale of the original data and the scale of the pairwise comparisons.
+This can be seen in Figure \ref{fig:sim-smooth-plot}B, where at day 5 there is essentially a difference of 50\% between \(\mbox{StO}_2\) in both groups, which corresponds to the -50\% difference in \ref{fig:plot-pairwise-comp}A.
+However, if there are multiple parametric terms in the model (more factors that need to be specified), such inclusion of the means can become problematic.
 However, we believe that the model we have presented here suffices in a wide range of situations where adding the group means is a relatively easy implementation that can help better visualize the estimation from the model from a biological perspective.
 
-In Figure \ref{fig:plot-pairwise-comp}A the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by \(\approx\) 3 day it is the ``Treatment'' group the one that on average, has greater \(\mbox{StO}_2\).
+In Figure \ref{fig:plot-pairwise-comp}A, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a different effect than the other.
+Notice that the shaded region between days 1 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by day \(\approx\) 3 it is the ``Treatment'' group that statistically on average has greater \(\mbox{StO}_2\).
 This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the given model.
@@
 
-On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that because the confidence intervals overlap for the first two days there is essentially no difference between the groups.
-However, because the model was still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes significant as in the full dataset smooth pairwise comparison.
-
-In a sense, the pairwise smooth comparison is more informative than a \emph{post-hoc} \emph{p-value}.
-For biomedical studies, the smooth comparison is able to provide an estimate of \emph{when} and by \emph{how much} a biological process becomes significant.
-This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.
+On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that because the confidence intervals overlap zero for the first two days there is statistically significant no difference between the groups.
+However, because the model is still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes statistically significant as in the full dataset smooth pairwise comparison.
+
+For biomedical studies, the ability of smooth comparisons to provide an estimate of \emph{when} and by \emph{how much} a biological process becomes significant is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.
 
@@
 Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}.
-When these models do not capture the non-linear trend of the data, this results in a ``model misspecification error''.
-This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
+When these models do not capture the non-linear trend of the data, this results in a ``model misspecification error.'' This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
 
@@
 
-In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (e.g., which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research\textsuperscript{\protect\hyperlink{ref-hastie1995}{67}}.
-This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs.
-T
-
-The purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model, such as in the case of polynomial effects in LMEMs.
+In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (e.g., which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{67}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs.
+
+However, in Section \ref{GAM-theory} we demonstrated that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model, such as in the case of polynomial effects in LMEMs where additionally there is no biological interpretation for such effect.
 In contrast, because GAMs let the data speak for themselves they can provide an accurate representation of the effect of time in a biological process.
@@
 However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} using R and the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
-
-One of the features of GAMs is that their Bayesian interpretation allows for inference about differences between groups without the need of a \emph{p-value}, and in addition can provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} using R and the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in Appendix A.
+
+One of the features of GAMs is that their Bayesian interpretation allows for inference about differences between groups without the need of a \emph{p-value}, thereby providing a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
 The model is therefore able to identify changes between the groups at time points were data was not directly measured even in the case of incomplete observations ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
@@
 We have used R as the software of choice for this paper because R not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research\textsuperscript{\protect\hyperlink{ref-begley2015}{68},\protect\hyperlink{ref-weissgerber2018}{69}}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{68},\protect\hyperlink{ref-weissgerber2018}{69}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.

### 2021-12-24 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_B.Rmd
Working on Appendix B

--- 2021-12-01 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/07-Appendix.Rmd
@@
-# (APPENDIX) Appendix {-}
+---
+title: | 
+    | SUPPLEMENTARY MATERIALS for
+    |
+    | **Generalized additive models to analyze biomedical non-linear longitudinal data in R:**
+    | Beyond repeated measures ANOVA and Linear Mixed Models
+    | 
+    | APPENDIX B: CODE AND FUNCTIONS 
+header-includes:
+    \usepackage{placeins}
+output:  
+  bookdown::pdf_document2:
+    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
+    pandoc_args: --listings #calls the listings package to fit code within the page margins
+    keep_tex: yes #keep LaTeX file for submission
+    fig_caption: yes #allows captions in figures
+    toc: false #do not include table of contents
+    extra_dependencies:
+      subfig: null #allows for subfigures
+      breqn: null #line breaks for long equations
+      caption: ["font={small}"] #size of the figure captions
+      float: null #allows for control of placement of figures
+    includes:
+      in_header: latex_docs/preamble_Appendix_B.sty #additional LaTeX formatting
+  bookdown::word_document2:
+    fig_caption: yes #figure caption
+    keep_md: yes
+  bookdown::html_document2:
+    css: "style.css" #style for the HTML document
+link-citations: yes #adds links to the citations
+'': default
+urlcolor: blue
+---
 
-<!-- allows for page numbering using letters for the Appendix -->
-\appendixpagenumbering
+\newpage
 
@@
 \counterwithin{figure}{section}
-# APPENDIX
 
-This section presents the code used to generate figures, models and simulated data from the main manuscript.
+This appendix shows the code for the functions used through the main manuscript, which can be found in the _scripts_ folder in the GitHub repository.
+We provide a brief explanation of the purpose of each function.
 
-## Compound symmetry and independent errors in linear and quadratic responses
+## Setup
 
-This section simulates linear and quadratic data in the same manner as in Section \@ref(simulation) in the main manuscript.
-The linear simulations using Figure \@ref(fig:linear-cases-Appendix) show in panels A and D the simulated mean responses and individual data points.
-Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
+First, we load all required libraries and set seed.
 
-The following code chunks produce a more comprehensive exploration of Figure \@ref(fig:l-q-response) in the main manuscript.
+```{r,setup, warning=FALSE, message=FALSE}
+library(patchwork)
+library(tidyverse)
+library(mvnfast)
+library(nlme)
+library(mgcv)
+library(gratia)
+library(here)
+library(scico)
+set.seed(2021) #set seed for reproducibility
 
-First, a function is created to simulate data across six timepoints using a linear or quadratic mean response, with correlated or uncorrelated errors.
-Each group has a different slope/concavity.
-The main function is the same for both groups, but a change in the sign allows to invert the trend.
+#alpha for ribbon in the smooth plots
+al<-0.8
 
-```{r, responses-Appendix,include=TRUE,message=FALSE,echo=TRUE}
+thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
+
+```
+
+## Linear and quadratic longitudinal trends
+
+### Function for linear and quadratic trends, rm-ANOVA and LMEM fits
+
+The first function is `example.R`, which allows to simulate linear and quadratic data in the same manner as in Section 3.5 in the main manuscript and estimates rm-ANOVA and LMEM with interaction to the data.
+The error for each simulated trend can be correlated or uncorrelated
+
+```{r, example-Appendix,include=TRUE,message=FALSE, eval=FALSE}
 ##########Section for calculations###########
@@
 }
+```
 
-##################Section for plotting#################################
+### A composite plot for the trends
 
-#This function will create the plots for either a "linear" or "quadratic" response
+Function `plot_example.R` uses the output of `example.R` to show the fit of a rm-ANOVA and a LMEM.
+It can be used to show an expanded version of Figure 1 in the main manuscript, presenting simulated data with correlated and uncorrelated errors and the correspoding rm-ANOVA and LMEM fits, which we do in the next subsection.
 
+```{r,plot_example-Appendix, include=TRUE, eval=FALSE}
+
+## This function plots the rm-ANOVA and LMEM for the data simulated in example.R
 plot_example <- function(sim_dat) {
-  ## Plot the simulated data (scatterplot)
-  txt<-20
+    # Plot the simulated data (scatterplot)
     p1 <- sim_dat$dat %>%
@@
         ) +
-    geom_point(show.legend=FALSE) +
+        geom_point(alpha=0.5,
+                   show.legend=FALSE) +
         labs(y='response')+
@@
                       color = treatment),
+                  size=3,
                   show.legend=FALSE) +
         theme_classic() +
-    theme(plot.title = element_text(size = txt, 
+        theme(plot.title = element_text(size = 20,
                                         face = "bold"),
-        text=element_text(size=txt))+
-    thm1
-
-  #plot the simulated data with trajectories per each subject
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = subject, 
-               color = treatment)
-           ) +
-    geom_line(aes(size = "Subjects"),
-              show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment,
-                  size = "Simulated Truth"), 
-              lty = 1,show.legend = FALSE) +
-    labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    theme_classic()+
-     theme(plot.title = element_text(size = txt, 
-                                face = "bold"),
-     text=element_text(size=txt))+
-    thm1
-
-  #plot the errors
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = errors,
-               group = subject, 
-               color = treatment)) +
-    geom_line(show.legend=FALSE) +
-     labs(y='errors')+
-     theme_classic()+
-     theme(plot.title = element_text(size = txt, 
-                                  face = "bold"),
-        text=element_text(size=txt))+
+              text=element_text(size=20))+
         thm1
@@
     #plot the model predictions for rm-ANOVA
-  p4 <- ggplot(sim_dat$dat, 
+    p2 <- ggplot(sim_dat$dat,
                  aes(x = time,
@@
                      color = treatment)) +
-    geom_point(show.legend=FALSE)+
+        geom_point(alpha=0.5,
+                   show.legend=FALSE)+
         labs(y='response')+
         geom_line(aes(y = predict(sim_dat$fit_anova),
-                  group = subject, size = "Subjects"),show.legend = FALSE) +
+                      group = subject, size = "Subjects"),
+                  show.legend = FALSE) +
         geom_line(data = sim_dat$pred_dat,
@@
         theme_classic() +
-    theme(plot.title = element_text(size = txt, 
+        theme(plot.title = element_text(size = 20,
                                         face = "bold"),
-        text=element_text(size=txt))+
+              text=element_text(size=20))+
         thm1
 
-   #plot the LMEM predictions
-  p5 <- ggplot(sim_dat$dat, 
+    #plot the model predictions for LMEM
+    p4 <- ggplot(sim_dat$dat,
                  aes(x = time,
@@
                      color = treatment)) +
-    geom_point()+
+        geom_point(alpha=0.5)+
         labs(y='response')+
@@
         theme_classic() +
-    theme(plot.title = element_text(size = txt, 
+        theme(plot.title = element_text(size = 20,
                                         face = "bold"),
-        text=element_text(size=txt))+
+              text=element_text(size=20))+
         thm1
@@
 
-#Store each plot in a separate object
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
+```
 
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
+### Plotting rm-ANOVA and LMEM fits for linear and quadratic trends in data
 
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
+In this subsection, we use `example.R` and a modified version `plot_example.R` (`plot_example_Appendix.R`) to create an expanded version of Figure 1 on the main manuscript.
+The only difference between `plot_example.R` and `plot_example_Appendix.R` is the inclusion of a `ggplot2` object (`p3`) that allows to plot the simulated errors.
 
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"))
+Figure \@ref(fig:linear-cases-Appendix) show in panels A and D the simulated mean responses and individual data points.
+Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
+In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
+Panels D and H show the predictions from the rm-ANOVA model.
+
+#### Fits for linear trends
+
+The chunk below calls both `example.R` and `plot_example_Appendix.R` to simulate data and create the composite plots.
+
+```{r,eval=TRUE,echo=TRUE, warning=FALSE}
+
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","example.R"))
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","plot_example_Appendix.R"))
+
+A1<-plot_example_Appendix(example(fun_type = "linear", error_type = "correlated"))
+
+B1<-plot_example_Appendix(example(fun_type = "linear", error_type = "independent"))
+
+C1<-plot_example_Appendix(example(fun_type = "quadratic", error_type = "correlated"))
+
+D1<-plot_example_Appendix(example(fun_type = "quadratic", error_type = "independent"))
 
@@
 
+#### Fits for quadratic trends
 For the quadratic response case, Figure \@ref(fig:quadratic-cases-Appendix) shows the simulated responses using compound symmetry and independent errors.
@@
 ```
-
-## Basis functions and GAMs
-
-This code produces Figure \@ref(fig:basis-plot) from the main manuscript.
-Briefly, a non-linear (quadratic) response is simulated, a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
-The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
-
-```{r,basis-Appendix, echo=TRUE,include=TRUE,message=FALSE,warning=FALSE,eval=FALSE}
-#generate the response: the same initial procedure from the previous section to simulate
-#the response
-set.seed(1)
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-
- #label each table
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
-
- #Convert to dataframes with subject, time and group columns
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
-
- #label subject per group
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
- #extract  "Group 1" to fit the GAM
-  dat<-subset(dat,treatment==1)
- #keep just the response and timepoint columns
-   dat<-dat[,c('y','time')]
-
-   #GAM model of time, 5 basis
-gm<-gam(y~s(time,k=5),data=dat)
-
-#model_matrix (also known as) 'design matrix'
-#will contain the smooths used to create  model 'gm'
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of basis by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be 
-#multiplied by the coefficients
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final smooth by aggregating the basis functions
-
-smooth<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(smooth=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,
-            aes(x=time,
-                y=value_orig,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(size=sz,
-            show.legend=FALSE)+
-  geom_point(size=sz+1,
-             show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,
-             labeller = as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#penalized basis
-p12<-ggplot(basis_plot,
-            aes(x=time,
-                y=mod_val,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(show.legend = FALSE,
-            size=sz)+
-  geom_point(show.legend = FALSE,
-             size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,
-             labeller=as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#heatmap of the coefficients
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,
-            aes(x=Basis,
-                y=Basis))+
-  geom_tile(aes(fill = cof), 
-            colour = "black") +
-    scale_fill_gradient(low = "white",
-                        high = "#B50A2AFF")+ #color picked from KikiMedium
-  labs(x='Basis',
-       y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),
-            size=7,
-            show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,
-            aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  labs(y='Simulated \n response')+
-  geom_line(data=smooth,
-            aes(x=time,
-                y=smooth),
-            color="#6C581DFF",
-            size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-
-```
-
-```{r,basis-plot-appendix, ref.label=I('basis-plot')}
-
-```
-
-## Longitudinal biomedical data simulation and GAMs{#tumor-data-simulation}
-
-This section describes how to fit GAMs to longitudinal data using simulated data.
-First, data is simulated according to Section \@ref(longitudinal-GAMs), where reported data of oxygen saturation ($\mbox{StO}_2$) in tumors under either chemotherapy or saline control is used as a starting point to generate individual responses in each group.
-
-```{r,sim-tumor-response-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE,eval=FALSE}
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
-#alpha for ribbon
-al<-0.8
-
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-
-n <- 10 #number of observations
-sd <- 10 #approximate sd from paper
-df <- 6
-dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, aes(x = Day, 
-                        y = StO2_sim, 
-                        color = Group, 
-                        group=subject)) +
-    geom_point(show.legend=FALSE,
-               size=1.5,
-               alpha=0.6)+
-    geom_line(size=0.6, alpha=0.4,show.legend=FALSE)+
-    geom_line(aes(x = Day, 
-               y = StO2,
-               color=Group),
-           size=1.5,
-           data=dat,
-           inherit.aes=FALSE,
-              show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],'(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    thm1
-
-```
-
-## A basic Workflow for GAMs{#workflow}
-
-This section shows a basic workflow to fit a series of increasingly complex GAMs to the simulated data from the previous section.
-Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
-
-### First model
-
-The first model fitted to the data is one that only accounts for different smooths by day.
-The model syntax specifies that `gam_00` is the object that will contain all the model information, and that the model attempts to explain changes in `StO2_sim` (simulated $\mbox{StO}_2$) using a smooth per `Day`.
-The model will use 5 basis functions (`k=5`) for the smooth.
-The smooth is constructed by default using thin plate regression splines.
-The smoothing parameter estimation method used is the restricted maximum likelihood (`REML`).
-
-```{r, first-GAM, echo=TRUE, include=TRUE, message=FALSE,warning=FALSE}
-
-gam_00<-gam(StO2_sim ~ s(Day, k = 5),
-            method='REML',
-            data  = dat_sim)
-
-```
-
-To obtain model diagnostics, two methodologies are to be used: 1) graphical diagnostics, and 2) a model check.
-In the first case, the functions `appraise` and `draw` from the package _gratia_ can be used to obtain a single output with all the graphical diagnostics.
-For model check, the functions `gam.check` and `summary` from  _mgcv_ provide detailed information about the model fit and its parameters.
-Keep in mind that `gam.check` is a function that also provides the graphical diagnostics obtained using _gratia_, if such graphical output is not desired the source code can be accessed typing `gam.check` in the Console, and the code without the graphical output can be used in a custom function, which is the approach we follow later).
-(ref:first-GAM-caption) Graphical diagnostics for the first GAM model.
-Left: Graphical diagnostics provided by the function `appraise` from the package _gratia_.
-Right: Fitted smooth for the model, provided by the function `draw`.
-
-```{r, plotting layouts,echo=TRUE, include=FALSE}
-
-#see https://patchwork.data-imaginist.com/reference/area.html
-layout1 <- c(
-  area(1, 1),
-  area( 1, 2),
-  area(2, 1),
-  area(2, 2),
-  area(1, 3, 2)
-)
-
-layout2 <- c(
-  area(1, 1),
-  area( 1, 2),
-  area(2, 1),
-  area(2, 2),
-  area(1,3,2,5)
-)
-
-```
-
-#### Graphical diagnostics
-
-```{r, first-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:first-GAM-caption)'}
-
-appr1<-appraise(gam_00)
-sm1<-draw(gam_00)
-visual_check<-appr1+sm1
-
-visual_check+plot_layout(design=layout1)
-```
-
-From the output of the function `appraise` in Figure \@ref(fig:first-GAM-diag), the major indicators of concern about the model are the QQ plot of residuals and the histogram of residuals.
-The QQ plot shows that the errors are not reasonably located along the 45$^{\circ}$ line (which indicates normality), as there are multiple points that deviate from the trend, specially in the tails.
-The histogram also shows that the variation (residuals) is not following the assumption of a normal distribution.
-
-The `draw` function permits to plot the smooths as `ggplot2` objects, which eases subsequent manipulation, if desired.
-Because model `gam_00` specifies only one smooth for the time covariate (Day), the plot only contains only one smooth.
-Note that the smooth shows an almost linear profile.
-
-#### Model check{#gam-00-model-check}
-
-Special attention must be paid to the parameter 'k-index' from `gam.check` (which calls `k.check` to perform the calculation).
-This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
-If the model is not adequately capturing the trends in the data, this is indicated by a low k-index value (<1).
-Because we plot the model diagnostics using `appraise` later, the graphical output from `gam.check` will be suppressed by creating a custom function to obtain just the model estimates, thus avoiding repetition of the diagnostic plots.
-This will be achieved by calling the source code of `gam.check` and using the appropriate code in a new function that will be called `gam.diagnostics`.
-We are not including in the Appendix the code for the function `gam.diagnostics` as it is rather long, but if desired it can be accessed by going to the `Appendix.Rmd` file in the GitHub repository and scrolling to this exact place (the code is not included in the final output, but is evaluated to create the function).
-
-```{r, first-GAM-check, fig.width=10, fig.height=7, out.width='75%', fig.align='center', echo=TRUE, include=FALSE, message=FALSE,warning=FALSE}
-
-#Function that uses the source code of `gam.check` to obtain the estimates without the plots. The source can be checked by typing `gam.check` in the console.
-gam.diagnostics<-function (b, old.style = FALSE, type = c("deviance", "pearson", 
-    "response"), k.sample = 5000, k.rep = 200, rep = 0, 
-    level = 0.9, rl.col = 2, rep.col = "gray80", ...) 
-{
-    type <- match.arg(type)
-    resid <- residuals(b, type = type)
-    linpred <- if (is.matrix(b$linear.predictors) && !is.matrix(resid)) 
-        napredict(b$na.action, b$linear.predictors[, 1])
-    else napredict(b$na.action, b$linear.predictors)
-
-    fv <- if (inherits(b$family, "extended.family")) 
-        predict(b, type = "response")
-    else fitted(b)
-    if (is.matrix(fv) && !is.matrix(b$y)) 
-        fv <- fv[, 1]
-    gamm <- !(b$method %in% c("GCV", "GACV", "UBRE", 
-        "REML", "ML", "P-ML", "P-REML", 
-        "fREML"))
-    if (gamm) {
-        cat("\n'gamm' based fit - care required with interpretation.")
-        cat("\nChecks based on working residuals may be misleading.")
-    }
-    else {
-        cat("\nMethod:", b$method, "  Optimizer:", 
-            b$optimizer)
-        if (!is.null(b$outer.info)) {
-            if (b$optimizer[2] %in% c("newton", "bfgs")) {
-                boi <- b$outer.info
-                cat("\n", boi$conv, " after ", boi$iter, 
-                  " iteration", sep = "")
-                if (boi$iter == 1) 
-                  cat(".")
-                else cat("s.")
-                cat("\nGradient range [", min(boi$grad), 
-                  ",", max(boi$grad), "]", sep = "")
-                cat("\n(score ", b$gcv.ubre, " & scale ", 
-                  b$sig2, ").", sep = "")
-                ev <- eigen(boi$hess)$values
-                if (min(ev) > 0) 
-                  cat("\nHessian positive definite, ")
-                else cat("\n")
-                cat("eigenvalue range [", min(ev), ",", 
-                  max(ev), "].\n", sep = "")
-            }
-            else {
-                cat("\n")
-                print(b$outer.info)
-            }
-        }
-        else {
-            if (length(b$sp) == 0) 
-                cat("\nModel required no smoothing parameter selection")
-            else {
-                cat("\nSmoothing parameter selection converged after", 
-                  b$mgcv.conv$iter, "iteration")
-                if (b$mgcv.conv$iter > 1) 
-                  cat("s")
-                if (!b$mgcv.conv$fully.converged) 
-                  cat(" by steepest\ndescent step failure.\n")
-                else cat(".\n")
-                cat("The RMS", b$method, "score gradient at convergence was", 
-                  b$mgcv.conv$rms.grad, ".\n")
-                if (b$mgcv.conv$hess.pos.def) 
-                  cat("The Hessian was positive definite.\n")
-                else cat("The Hessian was not positive definite.\n")
-            }
-        }
-        if (!is.null(b$rank)) {
-            cat("Model rank = ", b$rank, "/", length(b$coefficients), 
-                "\n")
-        }
-    }
-    cat("\n")
-    kchck <- k.check(b, subsample = k.sample, n.rep = k.rep)
-    if (!is.null(kchck)) {
-        cat("Basis dimension (k) checking results. Low p-value (k-index<1) may\n")
-        cat("indicate that k is too low, especially if edf is close to k'.\n\n")
-        printCoefmat(kchck, digits = 3)
-    }
-}
-```
-
-We now call `gam.diagnostics` to provide the desired diagnostic output, as well as a summary of the fitted model, which is obtained by calling `summary`.
-
-```{r,gam-00-diagnostics, echo=TRUE, include=TRUE}
-gam.diagnostics(gam_00)
-
-```
-
-From the output, it can be seen that the `k-index` is 0.36, which indicates that the model is not capturing the variability in the data.
-The `edf` (effective degrees of freedom) is an indicator of the complexity of the smooth.
-Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
-
-```{r, summary-gam_00}
-summary(gam_00)
-
-```
-
-From the `summary` function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
-The link function is 'identity' as the model does not make any transformation on the predictors.
-The 'significance of smooth terms' _p-value_ indicates if each smooth is adding significance to the model.
-Here, the _p-value_ is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to $\approx$ 17%.
-
-### Second model
-
-The major flaw of `gam_00` is that this model is not taking into account the fact that the data is nested in groups.
-The next iteration is a model where a different smooth of time (Day) is assigned for each group using `by = Group` in the model syntax.
-
-```{r, second-GAM, echo=TRUE, fig.width=10, fig.height=7, out.width='75%', fig.align='center', include=TRUE, message=FALSE,warning=FALSE}
-
-gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5),
-            method ='REML',
-            data  = dat_sim)
-
-gam.diagnostics(gam_01)
-```
-
-Diagnostics for this model indicate that the k-index is still below 1 (0.43 from `gam.check`), and that the residuals are still not following a normal distribution (Figure  \@ref(fig:second-GAM-diag)).
-Moreover, the smooths (plotted via the `draw()` function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-
-```{r,summary-gam_01}
-
-summary(gam_01)
-```
-
-From `summary()`, the deviance explained by the model is $\approx$ 51%.
-
-(ref:second-GAM-caption) Graphical diagnostics for the second GAM model.
-Left: Graphical diagnostics provided by the function `appraise` from the package _gratia_.
-Right: Fitted smooth for the model, provided by the function `draw`.
-
-```{r, second-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:second-GAM-caption)'}
-#need to add figure number and caption
-appr2<-appraise(gam_01)
-sm2<-draw(gam_01)
-
-visual_check2<-appr2+sm2
-
-visual_check2+plot_layout(design=layout1)
-```
-
-### Third model
-
-Model `gam_00` was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted (a single smooth), the model diagnostics, and, the low variance explained by the model.
-On the other hand, `gam_01` takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \@ref(longitudinal-GAMs), in order to differentiate between each group a parametric term needs to be added to the model for the interaction of _Day_ and _Group_.
-
-This is because in `gam_01`  separate smooths were fitted per group and those smooths also tried to account for the different means of the response in the two groups.
-Adding a parametric term for `Group` enables the smooths to capture the time course-differences of each group.
-The resulting model is `gam_02`, which is the model fitted in the main manuscript.
-
-```{r,final-model-Appendix,fig.width=10, fig.height=7, out.width='75%', fig.align='center', echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
-
-#GAM for StO2
-
-gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
-            method='REML',
-            data  = dat_sim)
-
-gam.diagnostics(gam_02)
-
-```
-
-By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From `gam.check`, the k-index is now at an acceptable value ($\approx$ 1.02).
-
-```{r,summary-gam_02}
-summary(gam_02)
-```
-
-From `summary`, the model is able to capture 89% of the variance in the data, which is a substantial improvement over the variance explained by `gam_00` and `gam_01`.
-
-(ref:final-GAM-caption) Graphical diagnostics for the final GAM model.
-Left: Graphical diagnostics provided by the function `appraise` from the package _gratia_.
-Right: Fitted smooths for the model, provided by the function `draw`.
-
-```{r, final-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:final-GAM-caption)'}
-
-appr3<-appraise(gam_02)
-sm3<-draw(gam_02)
-
-visual_check3<-appr3+sm3
-
-visual_check3+plot_layout(design=layout2)
-```
-
-### Comparing models via AIC
-
-One final  comparison that can be made for model selection involves the use of the Aikake Information Criterion (AIC).
-This metric is used to estimate information loss, which we want to minimize with an appropriate model.
-Therefore, when 2 or more models are compared, the model with lower AIC is preferred.
-In R, the comparison is done using the `AIC` function.
-
-```{r,model-selection-AIC}
-
-AIC(gam_00,gam_01,gam_02)
-
-```
-
-The output in this case is expected: model `gam_02` has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
-The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
-
-#### Pairwise comparisons of smooth confidence intervals
-
-The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \@ref(GAM-significance).
-In this case, the "design matrix" is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
-Briefly, the "design matrix" (also known as the "Xp matrix") from the selected model (`gam_02`) is used to calculate a 95% confidence interval of the difference between the smooth terms for each group.
-This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
-**All pairwise comparisons in this paper have been centered at the response scale to ease interpretation **.
-
-```{r,pairwise-comp-workflow,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
-##Pairwise comparisons
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-##matrix that contains the basis functions evaluated at the points in pdat
-    xp <- predict(gam_02, newdata = pdat, type = 'lpmatrix')
-
-#Find columns in xp where the name contains "Control"
-    c1 <- grepl('Control', colnames(xp))
-
-#Find columns in xp where the name contains 'Treatment'
-    c2 <- grepl('Treatment', colnames(xp))
-
-#Find rows in pdat that correspond to either 'Control' or 'Treatment'
-    r1 <- with(pdat, Group == 'Control')
-    r2 <- with(pdat, Group == 'Treatment')
-
-# In xp: find the rows that correspond to Control or Treatment, those that do not match will be
-    #set to zero. Then, substract the values from the rows corresponding to 'Control' from those that correspond
-    #to 'Treatment'
-    X <- xp[r1, ] - xp[r2, ]
-
-    ## remove columns that do not contain name 'Control' or 'Treatment'
-    X[, ! (c1 | c2)] <- 0
-    ## zero out the parametric cols, those that do not contain in the characters 's('
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
-
-    #Multiply matrix by model coefficients. X has (p,n) (rows, columns) and the coefficient matrix has
-    #dimensions (n,1). The resulting matrix has dimensions (p,1)
-    dif <- X %*% coef(gam_02)
-
-#Calculate standard error for the computed differences using the variance-covariance matrix
-    #of the model
-    se <- sqrt(rowSums((X %*% vcov(gam_02, unconditional = FALSE)) * X))
-    crit <- qt(0.05/2, df.residual(gam_02), lower.tail = FALSE)
-    #upper  limits
-    upr <- dif + (crit * se)
-    #lower limits
-    lwr <- dif - (crit * se)
-    #put all components in a dataframe for plotting
-    comp1<-data.frame(pair = paste('Control', 'Treatment', sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-
-#add time point sequence
-comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1))
-
-#use function from the pairwise comparison plot in the manuscript to get the shaded regions
-my_list<-pairwise_limits(comp_StO2)
-
-#plot the difference
-rib_col<-'#8D7D82' #color for ribbon for confidence interval
-control_rib <- '#875F79' #color for ribbon for control region
-treat_rib <- '#A7D89E' #color for ribbon treatment region
-
-c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
-  #shaded region
-  annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill=control_rib,
-                alpha = 0.5,
-                ) +
-  annotate("text",
-             x=1.5,
-             y=-10,
-             label="Control",size=10
-           )+
-  #shaded region  
-  annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill= treat_rib,
-             alpha = 0.5
-           ) +
-  annotate("text",
-             x=6,
-             y=-10,
-             label="Treatment",
-             size=10
-           )+
-  #ribbon for difference confidence interval  
-  geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(color='black',size=1) +
-    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-```
-
-(ref:pairwise-comp-caption) Smooth pairwise comparisons for model `gam_02` using a 95% confidence interval for the difference between smooths.
-The comparison is centered at the response scale.
-Shaded regions indicate time intervals where each treatment group has a non-zero effect.
-
-```{r, pairwise-comp-workflow-fig, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-comp-caption)'}
-c1
-```
-
-Of notice, a convenient wrapper for the function described above exists in the package `gratia`.
-In this package, `difference_smooths` is a function that makes the comparisons and produces Figure \@ref(fig:pairwise-comp-workflow-fig) when is used on a fitted model.
-The function syntax and an example can be found at:
-
-https://cran.r-project.org/web/packages/gratia/gratia.pdf
-
-Keep in mind that this function **does not** center the pairwise comparison at the response scale, so it has to be shifted in order to be compared to the raw data.
-
-## GAM and Linear model plots and Missing data
-
-This section covers the code used to generate Figure \@ref(fig:sim-smooth-plot), where the simulated data, fit of the "final" GAM (`gam_02`), linear model  and GAM on data with missing observations are presented.
-Note that panel A in Figure \@ref(fig:sim-smooth-plot) and the inset are generated in the code chunk where the data is simulated in Section \@ref(tumor-data-simulation), and are called later to build the figure.
-
-### GAM and Linear model plots
-
-This code chunk creates panels B and D in Figure \@ref(fig:sim-smooth-plot).
-Note that this code uses the final GAM from the previous section (`gam_02`), so the simulated data and the model should be generated before running this section.
-
-```{r, plotting-appendix, echo=TRUE, include=TRUE, message=FALSE, warning=FALSE,eval=FALSE}
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-#creates a dataframe using the length of the covariates for the GAM
-gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1),
-                         subject=factor(rep(1:10)))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#adds the predictions to the grid and creates a confidence interval for GAM
-gam_predict<-gam_predict%>%
-    mutate(fit = predict(gam_02,gam_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(gam_02, gam_predict,se.fit = TRUE,type='response')$se.fit)
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#plot smooths and confidence interval for GAM
-f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=gam_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=gam_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'complete')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-  thm1
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=0.3,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-  thm1
-
-```
-
-## Working with Missing data in GAMs
-
-This code chunk first randomly deletes 40% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data.
-Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
-
-```{r,missing-data-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE,eval=FALSE}
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-m_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = seq(0, 10, by = 0.1))
-
-#adds the predictions to the grid and creates a confidence interval
-m_predict<-m_predict%>%
-    mutate(fit = predict(mod_m1,m_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(mod_m1, m_predict,se.fit = TRUE,type='response')$se.fit)
-
-f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),
-                   fill=Group
-                   ),
-              alpha=0.3,
-              data=m_predict,
-            show.legend=FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=m_predict,
-              show.legend = TRUE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(atop(StO[2],'missing')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-  thm1
-
-```
-
-```{r,sim-smooth-plot-Appendix, ref.label=I('sim-smooth-plot')}
-
-```
-
-## Pairwise comparisons in GAMs: full and missing data cases
-
-The next code chunk reproduces Figure \@ref(fig:plot-pairwise-comp). Here pairwise comparisons are made for the full and missing datasets.
-
-```{r,pairwise-comp-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE, eval=FALSE}
-##Pairwise comparisons
-
-pdat <- expand.grid(Day = seq(0, 10, length = 400),
-                    Group = c('Control', 'Treatment'))
-
-#this function takes the model, grid and groups to be compared using the lpmatrix
-#originally developed by G. Simpson:
-#https://fromthebottomoftheheap.net/2017/10/10/difference-splines-i/
-
-smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
-                        unconditional = FALSE) {
-    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
-    #Find columns in xp where the name contains "Control" and "Treatment"
-    col1 <- grepl(g1, colnames(xp))
-    col2 <- grepl(g2, colnames(xp))
-    #Find rows in xp that correspond to each treatment
-    row1 <- with(newdata, Group == g1)
-    row2 <- with(newdata, Group == g2)
-    ## difference rows of xp for data from comparison
-    X <- xp[row1, ] - xp[row2, ]
-    ## zero out cols of X not involved in the comparison
-    X[, ! (col1 | col2)] <- 0
-
-    ## zero out the parametric cols
-    #This line has been commented to keep the comparison at the response level,
-    #otherwise it gives the marginal change between smooths
-    #X[, !grepl('^s\\(', colnames(xp))] <- 0
-    dif <- X %*% coef(model)
-    #get standard error, critical value and boundaries
-    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
-    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
-    upr <- dif + (crit * se)
-    lwr <- dif - (crit * se)
-    data.frame(pair = paste(g1, g2, sep = '-'),
-               diff = dif,
-               se = se,
-               upper = upr,
-               lower = lwr)
-}
-
-#use the function to calculate the difference in smooths
-comp1<-smooth_diff(gam_02,pdat,'Control','Treatment')
-
-#Create a dataframe with time, comparisons and labels for regions where difference exists
-comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp1)) %>%
-  mutate(interval=case_when(
-    upper>0 & lower<0~"no-diff",
-    upper<0~"less",
-    lower>0~"greater"
-  ))
-
-pairwise_limits<-function(dataframe){
-    #extract values where the lower limit of the ribbon is greater than zero
-    #this is the region where the control group effect is greater
-    v1<-dataframe%>%
-        filter(lower>0)%>%
-        select(Day)
-    #get day  initial value
-    init1=v1$Day[[1]]
-    #get day final value
-    final1=v1$Day[[nrow(v1)]]
-
-    #extract values where the value of the upper limit of the ribbon is lower than zero
-    #this corresponds to the region where the treatment group effect is greater
-    v2<-comp_StO2_full%>%
-        filter(upper<0)%>%
-        select(Day)
-
-    init2=v2$Day[[1]]
-    final2=v2$Day[[nrow(v2)]]
-    #store values
-    my_list<-list(init1=init1,
-                  final1=final1,
-                  init2=init2,
-                  final2=final2)
-return(my_list)
-}
-
-my_list<-pairwise_limits(comp_StO2_full)
-rib_col<-'#8D7D82' #color for ribbon for confidence interval
-control_rib <- '#875F79' #color for ribbon for control region
-treat_rib <- '#A7D89E' #color for ribbon treatment region
-
-c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
-    annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill=control_rib,
-                alpha = 0.5,
-                ) +
-  annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=8,
-           angle=90
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill=treat_rib,
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=8,
-           angle=90
-           )+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-###for missing data
-comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
-comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
-                   rbind(comp2))
-
-missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    labs(x = 'Days', 
-         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
-                              )))+
-  scale_x_continuous(breaks=c(0,2,5,7,10))+
-  theme_classic()+
-  theme(
-     text=element_text(size=18),
-     legend.title=element_blank()
-     )
-
-my_list<-pairwise_limits(comp_StO2_missing)
-
-c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
-    annotate("rect",
-             xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-             fill=control_rib,
-             alpha = 0.5
-             ) +
-  annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=8,
-           angle=90
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill= treat_rib,
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=8,
-           angle=90
-           )+
-    geom_ribbon(aes(ymin = lower, ymax = upper),
-                alpha = 0.5,
-                fill=rib_col) +
-    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
-    geom_line(color='black',size=1) +
-    facet_wrap(~ pair) +
-    theme_classic()+
-    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    theme(
-        text=element_text(size=18),
-        legend.title=element_blank()
-    )
-
-pair_comp<-c1+c2
-```
-
-```{r,plot-pairwise-comp-Appendix,ref.label=I('plot-pairwise-comp')}
-
-```

### 2021-12-24 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_B.tex
Working on Appendix B

--- 2021-12-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/SIM_Appendix.tex
@@
 ]{article}
-\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
+\title{SUPPLEMENTARY MATERIALS for\\
+\strut \\
+\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
 Beyond repeated measures ANOVA and Linear Mixed Models\\
 \strut \\
-SUPPLEMENTARY MATERIALS: APPENDIX}
+APPENDIX B: CODE AND FUNCTIONS}
+\author{}
 \date{\vspace{-2.5em}}
@@
 
-%to have page numbers with letters in the Appendix:
-% %solution from https://tex.stackexchange.com/questions/59572/custom-page-numbering-for-appendix
-\newcommand{\appendixpagenumbering}{
-  \break
-  \pagenumbering{arabic}
-  \renewcommand{\thepage}{\thesection-\arabic{page}}
-}
+%start sections and page numbers with A
+\setcounter{page}{1}
+\renewcommand\thesection{B}
+\renewcommand\thesubsection{\thesection.\arabic{subsection}}
+\renewcommand{\thepage}{B-\arabic{page}}
 
@@
 
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
-
-In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs).
-Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.
-
-In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
-Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research.
-This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
-We present the basic theory of GAMs, and using reported trends of oxygen saturation in tumors we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R.
-We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing).
-To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
-
-\hypertarget{keywords}{%
-\section*{Keywords}\label{keywords}}
-\addcontentsline{toc}{section}{Keywords}
-
-cancer biology; tumor response; generalized additive models; simulation; R
-
-\hypertarget{background}{%
-\section{Background}\label{background}}
-
-Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study).
-Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others.
-Tumor response {[}\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}{]}, antibody expression {[}\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}{]}, and cell metabolism {[}\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response.
-Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}\protect\hyperlink{ref-greening2018}{9}{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}\protect\hyperlink{ref-sio2016}{10}{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}\protect\hyperlink{ref-kamstra2015}{11}{]}.
-
-Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study.
-The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}\protect\hyperlink{ref-wagenmakers2008}{12}{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA).
-The rm-ANOVA model makes three assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}{]}.
-
-The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference.
-In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies.
-A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation.
-Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.
-
-A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
-Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from data that shows non-linear trends, the validity of such metric is dependent on how adequate the model fits the data.
-In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}.
-For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}\protect\hyperlink{ref-wang2019}{18}{]}.
-
-Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}{]}, consequently biasing the conclusions of the study.
-Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}\protect\hyperlink{ref-abdi2010}{21}{]}), but they in turn reduce statistical power (1-\(\beta\)){[}\protect\hyperlink{ref-nakagawa2004}{22}{]}, and lead to increased Type II error (failing to reject the null hypothesis when it is false) {[}\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}{]}.
-Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.
-
-On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and group sizes {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}\protect\hyperlink{ref-charan2013}{30}{]} which might cause the corrections for lack of compound symmetry to be ineffective.
-
-Due to a variety of causes, the number of observations during a study can vary between all subjects.
-For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible.
-It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}\protect\hyperlink{ref-gueorguieva2004}{13}{]}.
-This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
-At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.
-
-During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data.
-The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}.
-And even more importantly, LMEMs also assume by default a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]} (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .
-
-As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the trend of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference.
-In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data.
-Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.
-Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]}.
-There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]}, and 3) can easily accommodate missing observations.
-Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.
-
-The current advances in programming languages designed for statistical analysis (specifically R), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs.
-In particular, R {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user.
-At the same time, R has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.
-Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.
-
-This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding an expected trend of the response, constant correlation structures, and missing observations are explained in detail.
-Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
-Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}.
-The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies.
-Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in R, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.
-
-In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models to improve the standards for reproducibility in biomedical research.
-
-\hypertarget{challenges-presented-by-longitudinal-studies}{%
-\section{Challenges presented by longitudinal studies}\label{challenges-presented-by-longitudinal-studies}}
-
-\hypertarget{the-repeated-measures-anova-and-linear-mixed-model}{%
-\subsection{The repeated measures ANOVA and Linear Mixed Model}\label{the-repeated-measures-anova-and-linear-mixed-model}}
-
-The \emph{repeated measures analysis of variance} (rm-ANOVA) and the \emph{linear mixed model} (LMEM) are the most commonly used statistical analysis for longitudinal data in biomedical research.
-These statistical methodologies require certain assumptions for the model to be valid.
-From a practical view, the assumptions can be divided in three areas: 1) an assumed relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects.
-Each one of these assumptions is discussed below.
-
-\hypertarget{assumed-relationship}{%
-\subsection{Assumed relationship}\label{assumed-relationship}}
-
-\hypertarget{the-repeated-measures-anova-case}{%
-\subsubsection{The repeated measures ANOVA case}\label{the-repeated-measures-anova-case}}
-
-In a longitudinal biomedical study, two or more groups of subjects (e.g., human subject, mice, samples) are subject to different treatments (e.g., a ``treatment'' group receives a novel drug or intervention vs.~a ``control'' group that receives a placebo), and measurements from each subject within each group are collected at specific time points.
-The collected response is modeled with \emph{fixed} components.
-The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug/intervention in the ``treatment'' group.
-
-Mathematically speaking, a rm-ANOVA model with an interaction can be written as
-
-\begin{equation}
-y_{ijt} = \beta_0+\beta_1 \times treatment_{j} +\beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
-\label{eq:linear-model}
-\end{equation}
-
-In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of treatment (\(treatment_j\)), time (\(time_t\)), and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively.
-Independent errors \(\varepsilon_{ijt}\) represent random variation from the sampling process assumed to be \(\stackrel{i.i.d.}\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
-In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug,'' or ``saline'' vs.~``chemotherapy'').
-Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B).
-With this notation, the linear model then can be expressed as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_2\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
-\beta_0 + \beta_1+\beta_2 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
-\end{cases}
-\label{eq:ANOVA-by-group}
-\end{equation}
-
-To further simplify the expression, substitute \(\widetilde{\beta_{0}}=\beta_0+\beta_{1}\) and \(\widetilde{\beta_{1}}=\beta_{2}+\beta_{3}\) in the equation for Group B.
-This substitution allows for a different intercept and slope for Groups A and B.
-The model is then written as
-
-\begin{equation}
-y_{ijt} = \begin{cases}
-\beta_0 + \beta_2\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
-\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
-\end{cases}
-\label{eq:ANOVA-lines}
-\end{equation}
-
-Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group).
-In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response.
-This means that either presented as Equations \eqref{eq:linear-model}, \eqref{eq:ANOVA-by-group} or \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data.
-If the data show non-linear trends, the rm-ANOVA model will approximate this behavior with non-parallel lines.
-
-\hypertarget{LMEM-case}{%
-\subsubsection{The Linear Mixed Model (LMEM) Case}\label{LMEM-case}}
-
-A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to account for{[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}.
-A LMEM with interaction between time and treatment for a longitudinal study can be written as
-
-\begin{equation}
-y_{ijt} = \beta_0+ \beta_1 \times treatment_{j}+ \beta_2 \times time_{t} + \beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij} +\varepsilon_{ijt}\\.
-\label{eq:LMEM}
-\end{equation}
-
-When Equations \eqref{eq:linear-model} and \eqref{eq:LMEM} are compared, it is noticeable that LMEMs and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\alpha_{ij}\)).
-This term \(\alpha_{ij}\) corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)).
-The \emph{random} component can also be understood as modeling some ``noise'' in the response, but that does not arise from the sampling error term \(\varepsilon_{ijt}\) from Equations \eqref{eq:linear-model} through \eqref{eq:ANOVA-lines}.
-
-For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data that needs to be accounted for.
-As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value.
-The random effect \(\alpha_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\mu_{ij} \sim N(0,\sigma^2_\mu)\).
-In essence, the \emph{random effect} in a LMEM enables fitting models with different intercepts at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.
-
-Of note, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach (using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}), which is described in detail by Pinheiro and Bates {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-However, polynomial fits have limited predictive power and cause bias on the boundaries of the covariates {[}\protect\hyperlink{ref-beck1998}{36}{]}; more importantly, polynomial effects lack biological or mechanistic interpretation {[}\protect\hyperlink{ref-pinheiro2006}{15}{]} which limits their use in biomedical studies.
-
-\hypertarget{covariance-in-rm-anova-and-lmems}{%
-\subsection{Covariance in rm-ANOVA and LMEMs}\label{covariance-in-rm-anova-and-lmems}}
-
-In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group.
-The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{41}{]} (For an in-depth analysis of the covariance matrix see West{[}\protect\hyperlink{ref-west2014}{40}{]} and Weiss{[}\protect\hyperlink{ref-weiss2005}{42}{]}).
-
-In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity'').
-Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}{]}.
-However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}\protect\hyperlink{ref-maxwell2017}{44}{]}; and it is higher between consecutive measurements {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}.
-Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]} the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and they are not robust if the group sizes are unbalanced {[}\protect\hyperlink{ref-keselman2001}{29}{]}.
-Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.
-
-In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.
-Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself.
-Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.
-
-\hypertarget{unbalanced-data}{%
-\subsection{Unbalanced data}\label{unbalanced-data}}
-
-In a longitudinal study, it is frequently the case that the number of observations is different across subjects.
-In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing,'' ``incomplete,'' or (more generally speaking) unbalanced data {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}.
-The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis.
-This is inconvenient because the remaining subjects might not accurately represent the population and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{46}{]}.
-
-On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are \emph{missing at random} (MAR) or \emph{completely missing at random} (MCAR) {[}\protect\hyperlink{ref-west2014}{40},\protect\hyperlink{ref-weiss2005}{42}{]}.
-In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{47}{]}.
-If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{48}{]}.
-Missing observations can also be \emph{missing not at random} (MNAR) and in the case the missing observations are dependent on their value.
-For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.
-
-However, it is worth reminding that ``all models are wrong'' {[}\protect\hyperlink{ref-box1976}{49}{]} and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
-Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.
-
-\hypertarget{simulation}{%
-\subsection{What do an rm-ANOVA and LMEM fit look like? A visual representation using simulated data}\label{simulation}}
-
-To visually demonstrate the limitations of rm-ANOVA and LMEMs for longitudinal data with non-linear trends, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each.
-An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using R {[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme} {[}\protect\hyperlink{ref-nlme}{50}{]}.
-
-Briefly, two cases for the mean response for each group are considered: in the first case, the mean response in each group is a linear function over time with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}A).
-In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}C).
-In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point.
-This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments where a strong non-linear trend is present.
-
-Specifically, the rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect.
-In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group.
-From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time).
-Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D).
-
-A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where a simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) is presented.
-We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with ``default'' (linear) effects and, 2) demonstrate that a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
-In Section \ref{longitudinal-GAMs} we use simulated data that does follow reported trends in the biomedical literature to implement GAMs.
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/l-q-response-1}
-
-}
-
-\caption{Simulated responses from two groups with correlated errors using a LMEM and a rm-ANOVA model.
-Top row: linear response, bottom row: quadratic response.
-\textbf{A}: Simulated linear data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
-\textbf{D}: Simulated quadratic data with known mean response (thick lines) and individual responses (points) showing the dispersion of the data.
-\textbf{B,E}: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic).
-Points represent the original raw data.
-The rm-ANOVA model not only fails to pick the trend of the quadratic data (D) but also assigns a global estimate that does not take between-subject variation.
-\textbf{C, F}: Estimates from the LMEM in the linear and quadratic case (subject: thin lines, population: thick lines) .
-The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data and grossly bias the initial estimates for each group in the quadratic case (bottom row).}\label{fig:l-q-response}
-\end{figure}
-
-The simulation shows that the fits produced by the LMEM and the rm-ANOVA model are good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response}A).
-When the linearity and compound symmetry assumptions are met, the rm-ANOVA model approximates well the global trend by group (Figure \ref{fig:l-q-response}B).
-Note that because the LMEM incorporates \emph{random effects}, is able to provide estimates for each subject and a ``population'' estimate (Figure \ref{fig:l-q-response}C).
-
-However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}D.
-Here, the mean response per group was simulated using a quadratic function, and errors and individual responses were produced as in Figure \ref{fig:l-q-response}A.
-The mean response in the simulated data with quadratic behavior changes in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group.
-Fitting an rm-ANOVA model (Equation \eqref{eq:linear-model}) or a LMEM (Equation \eqref{eq:LMEM}) to this data produces the fit that appears in Figure \ref{fig:l-q-response}E, F.
-
-Comparing the fitted responses of the LMEM and the rm-ANOVA models used in the simulated quadratic data (Figure \ref{fig:l-q-response}E, F) indicates that the models are not capturing the changes within each group.
-Specifically, note that the fitted mean response of both models shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.
-
-The LMEM is only able to account for between-subject variation by providing estimates for each subject (Figure \ref{fig:l-q-response}F), but both models are unable to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare Figure \ref{fig:l-q-response}D with Figure \ref{fig:l-q-response}E, F).
-If such a change has important physiological implications, both rm-ANOVA and LMEMs omit it from the fitted mean response.
-Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.
-It could be argued that a LMEM with quadratic effects should have been used to fit the data in Figure\ref{fig:l-q-response}F.
-However, because in reality the true function is not known, choosing a polynomial degree causes more questions (e.g., is it quadratic?, cubic?, or a higher degree?).
-Additionally, polynomial effects have other limitations, which we cover in Section \ref{GAM-theory}.
-
-This section has used simulation to better convey and visulize the limitations of linearity and correlation in the response in data with non-linear trends using an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response.
-Notice that the model misspecification is easily noticeable if the model fit and the response are visualized.
-In the following section, we provide a brief overview of linear models, general linear models and generalized linear mixed models before presenting the theory of GAMs.
-
-\hypertarget{linear-models-and-beyond}{%
-\section{Linear Models, and beyond}\label{linear-models-and-beyond}}
-
-Linear models (LMs) are those that assume a normal (Gaussian) distribution of the errors, and only incorporate \emph{fixed effects} (such as rm-ANOVA).
-These are by far the models most commonly used to analyze data within the biomedical research community.
-On the other hand, Linear Mixed Effect Models (LMEMs) also incorporate \emph{random effects}, as it has been described in Section \ref{LMEM-case}.
-
-In reality, rm-ANOVA and LMEMs are just \emph{special cases} of a broader class of models (General Linear Models and Generalized Linear Mixed Models, respectively).
-In order to fully capture the constraints of such models and to understand how GAMs overcome those limitations this section will briefly provide an overview of the different classes of models, indicate how rm-ANOVA and LMEMs fit within this framework and introduce the theory of GAMs.
-
-\hypertarget{generalized-linear-models-glms}{%
-\subsection{Generalized Linear Models (GLMs)}\label{generalized-linear-models-glms}}
-
-A major limitation of LMs is their assumption of normality in the errors.
-If the residuals are non-normal, a transformation is necessary in order to properly fit the model.
-However, transformation can lead to poor model performance {[}\protect\hyperlink{ref-ohara2010}{51}{]}, and can cause problems with the biological interpretation of the model estimates.
-McCullagh and Nelder {[}\protect\hyperlink{ref-nelder1972}{52}{]} introduced General Linear Models (GLMs) as an extension of LMs, where the errors do not need to be normally distributed.
-To achieve this, consider the following model
-
-\begin{equation}
-y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi),
-\label{eq:GLM-y}
-\end{equation}
-
-where \(y_{ijt}\) is the observation \(i\) in group \(j\) at time \(t\), that is assumed to come from some distribution of the exponential family \(\mathcal{D}\), with some mean \(\mu_{ijt}\) and potentially, a dispersion parameter \(\phi\) (which in the Gaussian case is the variance \(\sigma^{2}\)).
-The mean (\(\mu_{ijt}\)) is also known as the \emph{expected value} (or \emph{expectation}), \(E(y_{ijt})\) of the observed response \(y_{ijt}\).
-
-Then, the \emph{linear predictor} \(\eta\), which defines the relationship between the mean and the covariates can be defined as
-
-\begin{equation}
-\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} +\beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j},
-\label{eq:GLM-eta}
-\end{equation}
-
-where \(\eta_{ijt}\) is the linear predictor for each observation in each group at each timepoint, and \(\beta_0\) (the intercept), \(\beta_{1}, \beta_2,\) and \(\beta_3\) are the model parameters for each group, which can be referred globally to as \(\beta_j\).
-Finally, \(time_{ijt}\) represents the covariates from each subject in each group at each time point.
-
-Finally,
-
-\begin{equation}
-E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt}),
-\label{eq:GLM-Expectation}
-\end{equation}
-
-where \(E(y_{ijt})\) is the expectation, and \(g^{-1}\) is the inverse of a \emph{link function} (\(g\)).
-The link function transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}.
-Therefore, it can be seen that LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
-
-\hypertarget{generalized-linear-mixed-models-glmms}{%
-\subsection{Generalized linear mixed models (GLMMs)}\label{generalized-linear-mixed-models-glmms}}
-
-Although GLMs relax the normality assumption, they only accommodate fixed effects.
-Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate \emph{random effects}, which have an associated probability distribution {[}\protect\hyperlink{ref-mcculloch2001}{53}{]}.
-Therefore, in GLMMs the linear predictor takes the form
-
-\begin{equation}
-\eta_{ijt}=\beta_0+\beta_1 \times treatment_{j} + \beta_2 \times time_{t} + +\beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij},
-\label{eq:GLMM-eta}
-\end{equation}
-
-where \(\alpha_{ij}\) corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of Equation \eqref{eq:GLM-eta}.
-Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed {[}\protect\hyperlink{ref-nelder1972}{52}{]}, and GLMs are a special case of GLMMs where there are no random effects.
-In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in Dobson {[}\protect\hyperlink{ref-dobson2008}{54}{]} and Stroup {[}\protect\hyperlink{ref-stroup2013}{55}{]}.
-
-\hypertarget{GAM-theory}{%
-\subsubsection{GAMs as a special case of Generalized Linear Models}\label{GAM-theory}}
-
-\hypertarget{gams-and-basis-functions}{%
-\paragraph{GAMs and Basis Functions}\label{gams-and-basis-functions}}
-
-Notice that in the previous sections, the difference between GLMs and GLMMs resides on their linear predictors (Equations \eqref{eq:GLM-eta}, \eqref{eq:GLMM-eta}).
-Generalized additive models (GAMs) are an extension of the GLM family that allow the estimation of smoothly varying trends where the relationship between the covariates and the response is modeled using \emph{smooth functions} {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}{]}.
-In a GAM, the linear predictor has the form
-
-\begin{equation}
-\eta_{ijt}=\beta_{0}+ \beta_{1} \times treatment_{j} +f(time_{t}|\beta_{j}).
-\label{eq:GAM-eta}
-\end{equation}
-
-Where \(\beta_{0}\) is the intercept, and \(\beta_{1}\) is the coefficient for each treatment group.
-Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time have been replaced by \(f(time_{t}|\beta_{j})\), which represents the smooth function of time with with inputs as the covariates \(time_t\) and parameters \(\beta_{j}\).
-A GAM version of a linear model can be written as:
-
-\begin{equation}
-  y_{ijt}=\beta_0+f(time_t\mid \beta_j)+\varepsilon_{ijt}
-  \label{eq:GAM}
-\end{equation}
-
-Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the smooth function \(f(time_t\mid \beta_j)\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.
-
-In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is more advantageous as it allows more flexibility because it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response.
-One possible set of functions for \(f(time_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}.
-In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(time_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.
-
-The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using a \emph{basis function} expansion of the covariates and by estimating random coefficients associated with these basis functions.
-A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}.
-For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(treatment_j\), \(time_t\), and \(time_t \times treatment_j\).
-The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{57}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions.
-In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(time_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.
-
-Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as basis functions as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible, and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{58}{]}.
-Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.
-
-Cubic splines (CS) are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness.
-The use of CS as smoothers in GAMs was discussed within the original GAM framework {[}\protect\hyperlink{ref-hastie1987}{56}{]}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the basis functions meet) to obtain the finite basis, which could affect the model fit {[}\protect\hyperlink{ref-wood2003}{59}{]}.
-A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}{]}.
-
-In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner {[}\protect\hyperlink{ref-wood2003}{59}{]}.
-Like TPS, TPRS only require the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood{[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}{]}).
-
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C.
-The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
-A computational requisite in \emph{mgcv} is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable.
-Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth.
-It is important to note that is not necessary to specify a number of basis equal to the number of unique values in the independent variable; fewer basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data.
-
-Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present.
-A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline.
-The second derivative of the spline is a measure of curvature, or the rate of change of the slope {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}, and increasing the penalty by increasing \(\lambda\) results in models with less curvature.
-As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
-In other words,a low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
-
-With this in mind, if five basis functions are used to fit a GAM for the data that appears in Figure \ref{fig:l-q-response}C, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A.
-The four basis functions (and the intercept) are shown.
-Each of the basis functions is evaluated at six different points (because there are six points on the timeline).
-The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), that is estimated when fitting the model.
-The penalized estimates fitted for our example are shown in Figure \ref{fig:basis-plot}B.
-
-To get the weighted basis functions, each basis (from Figure Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
-Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and less than 1).
-On the other hand, the third basis function has roughly doubled its contribution to the smooth term.
-Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (orange line), along the simulated values per group, which appear as points.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/basis-plot-1}
-
-}
-
-\caption{Basis functions for a single smoother for time.
-\textbf{A}: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-\textbf{B}: Matrix of basis function weights.
-Each basis function is multiplied by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-\textbf{C}: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
-Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-\textbf{D}: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
-\end{figure}
-
-\hypertarget{a-bayesian-interpretation-of-gams}{%
-\subsubsection{A Bayesian interpretation of GAMs}\label{a-bayesian-interpretation-of-gams}}
-
-Bayes' theorem states that the probability of an event can be calculated using prior knowledge {[}\protect\hyperlink{ref-mcelreath2018}{60}{]}.
-In the case of data that shows non-linear trends, the prior that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language) {[}\protect\hyperlink{ref-miller2019}{61}{]}.
-Therefore, the confidence intervals (CIs) calculated by default for the smooth terms using \emph{mgcv} are considered empirical Bayesian credible intervals {[}\protect\hyperlink{ref-pedersen2019}{33}{]}, which have good \emph{across the function} (``frequentist'') coverage{[}\protect\hyperlink{ref-wood2017}{37}{]}.
-
-To understand this last part, it is worth reminding that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function exists, taking into account the randomness introduced by the sampling process.
-Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie.
-In an \emph{across the function} CI (like those estimated by default for GAMs using \emph{mgcv}) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%).
-In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI and for this to occur, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage.
-
-However, besides the across the function CI, ``simultaneous'' or ``whole function'' CIs can also be computed, which contain the \emph{whole function} with a specified probability {[}\protect\hyperlink{ref-wood2017}{37}{]}.
-Suppose we chose a nominal value (say, 95\%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
-Briefly, to obtain a simultaneous CI we use simulation to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI {[}\protect\hyperlink{ref-ruppert2003}{62}{]}.
-Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups, as we discuss in the following section.
-
-In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller {[}\protect\hyperlink{ref-miller2019}{61}{]}, Wood{[}\protect\hyperlink{ref-wood2017}{37}{]}, Simpson {[}\protect\hyperlink{ref-simpson2018}{34}{]}, Marra {[}\protect\hyperlink{ref-marra2012}{63}{]}, and Ruppert {[}\protect\hyperlink{ref-ruppert2003}{62}{]}.
-With this brief introduction to the Bayesian interpretation of GAMs, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends.
-
-\hypertarget{longitudinal-GAMs}{%
-\section{The analyisis of longitudinal biomedical data using GAMs}\label{longitudinal-GAMs}}
-
-The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs.
-This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}.
-The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
-
-\hypertarget{simulated-data}{%
-\subsection{Simulated data}\label{simulated-data}}
-
-The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors (Figure 3C in Vishwanath et.
-al.{[}\protect\hyperlink{ref-vishwanath2009}{16}{]}).
-In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10).
-In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group.
-Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
-The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}A.
-
-\hypertarget{an-interaction-gam-for-longitudinal-data}{%
-\subsection{An interaction GAM for longitudinal data}\label{an-interaction-gam-for-longitudinal-data}}
-
-An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination.
-In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time.
-Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry.
-A detailed description on basis functions and correlations can be found in Hefley {[}\protect\hyperlink{ref-hefley2017}{57}{]}.
-
-For the data in Figure \ref{fig:sim-smooth-plot}A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment.
-To estimate this, the model incorporates separate smooths for \emph{Group} and \emph{Day}, respectively.
-The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}).
-Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the group means.
-Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:
+\counterwithin{figure}{section}
+
+This appendix shows the code for the functions used through the main manuscript, which can be found in the \emph{scripts} folder in the GitHub repository.
+We provide a brief explanation of the purpose of each function.
+
+\hypertarget{setup}{%
+\subsection{Setup}\label{setup}}
+
+First, we load all required libraries and set seed.
 
 \begin{lstlisting}[language=R]
-gam_02 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
+library(patchwork)
+library(tidyverse)
+library(mvnfast)
+library(nlme)
+library(mgcv)
+library(gratia)
+library(here)
+library(scico)
+set.seed(2021) #set seed for reproducibility
+
+#alpha for ribbon in the smooth plots
+al<-0.8
+
+thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
 \end{lstlisting}
 
-This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using four basis functions (plus the intercept).
-The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths {[}\protect\hyperlink{ref-simpson2018}{34}{]} (a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console).
-
-The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups.
-Although the default \passthrough{\lstinline!method!} used to estimate the smoothing parameters in \emph{mgcv} is generalized cross validation (GCV), Wood{[}\protect\hyperlink{ref-wood2017}{37}{]} showed the restricted maximum likelihood (REML) to be more resistant to overfitting while also easing the quantification of uncertainty in the smooth parameters; therefore in this manuscript REML is always used for smooth parameter estimation.
-An additional argument (\passthrough{\lstinline!family!}) allows to specify the expected distribution of the response, but it is not used in this model because we expect a normally-distributed response (which is the default \passthrough{\lstinline!family!} in \emph{mgcv}).
-
-When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia} {[}\protect\hyperlink{ref-gratia}{64}{]}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood {[}\protect\hyperlink{ref-wood2017}{37}{]} and Harezlak {[}\protect\hyperlink{ref-harezlak2018}{65}{]}.
-
-One question that might arise at this point is ``what is the fit that an rm-ANOVA model produces for the simulated data?'' The rm-ANOVA model, which corresponds to Equation \eqref{eq:linear-model} is presented in Figure \ref{fig:sim-smooth-plot}C.
-This is a typical case of model misspecification: The slopes of each group are different, which would lead to a \emph{p-value} indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \ref{fig:sim-smooth-plot}B) .
-
-Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists.
-The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
-However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends.
-Consider the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}B.
-If 40\% of the observations are randomly deleted and the same interaction GAM (\passthrough{\lstinline!gam\_02!}) is fitted, the resulting smooths appear in (Figure \ref{fig:sim-smooth-plot}D).
-Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.
-
-Additionally, note that in Figures \ref{fig:sim-smooth-plot}B,D we show two CIs for each of the fitted smooths (shaded regions).
-The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, we expect them to be wider than the across the function CI, resulting in the wide shaded regions.
-For the dataset with missing observations, the smooths CIs overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset.
-However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/sim-smooth-plot-1}
-
-}
-
-\caption{Simulated data and smooths for oxygen saturation in tumors.
-\textbf{A}: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-\textbf{B}: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.
-\textbf{C}: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-\textbf{D}: Smooths for the GAM model for the simulated data with missing observations (40\%).
-Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.}\label{fig:sim-smooth-plot}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/plot-pairwise-comp-1}
-
-}
-
-\caption{Pairwise comparisons for smooth terms.
-\textbf{A}: Pairwise comparisons for the full dataset.
-\textbf{B}: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
-In both cases the effect of treatment is significant after day 3.
-In both cases, a simultaneous CI has been computed around the smooth difference.}\label{fig:plot-pairwise-comp}
-\end{figure}
-
-\hypertarget{GAM-significance}{%
-\subsection{Determination of significance in GAMs for longitudinal data}\label{GAM-significance}}
-
-At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
-
-The absence of a \emph{p-value} in this context might seem odd, but its interpretation can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment.
-This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
-
-With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant.
-A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
-
-Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}B and Figure \ref{fig:sim-smooth-plot}D.
-Figure \ref{fig:plot-pairwise-comp} shows the comparison between each treatment group for the full and missing datasets with a simultaneous CI computed around the difference.
-Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared.
-Notice that because we have included the means of each group, there is correspondence between the scale of the original data the scale of the pairwise comparisons.
-This can be seen as in Figure \ref{fig:sim-smooth-plot}B at day 5 there is essentially a 50\% difference between \(\mbox{StO}_2\) in both groups, which corresponds to the -50 difference in \ref{fig:plot-pairwise-comp}A.
-However, if there are multiple parametric terms in the model (more factors that need to be specified in the model) such inclusion of the means can become problematic.
-However, we believe that the model we have presented here suffices in a wide range of situations where adding the group means is a relatively easy implementation that can help better visualize the estimation from the model from a biological perspective.
-
-In Figure \ref{fig:plot-pairwise-comp}A the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by \(\approx\) 3 day it is the ``Treatment'' group the one that on average, has greater \(\mbox{StO}_2\).
-This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the given model.
-Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in mean \(\mbox{StO}_2\) occurs.
-
-On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that because the confidence intervals overlap for the first two days there is essentially no difference between the groups.
-However, because the model was still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes significant as in the full dataset smooth pairwise comparison.
-
-In a sense, the pairwise smooth comparison is more informative than a \emph{post-hoc} \emph{p-value}.
-For biomedical studies, the smooth comparison is able to provide an estimate of \emph{when} and by \emph{how much} a biological process becomes significant.
-This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-Biomedical longitudinal data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
-Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}, where it is clear that these models do not capture the non-linear trend of the data, thereby causing a ``model misspecification error.''
-
-This ``model misspecification'' error, also is known as a ``Type III'' error {[}\protect\hyperlink{ref-dennis2019}{17}{]} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-Polynomial effects can be used in LMEMs, and it could be argued that in Section \ref{simulation} a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects presents more questions than answers in this case (i.e., what is the appropriate polynomial effect to use?
-quadratic?, cubic?, what is the biological interpretation of a polynomial effect?).
-Moreover, polynomial effects have other associated issues that we have described in Section \ref{GAM-theory}.
-
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) {[}\protect\hyperlink{ref-lang2015}{66}{]} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.
-
-In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) {[}\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}{]} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research {[}\protect\hyperlink{ref-hastie1995}{67}{]}.
-This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).
-
-However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature {[}\protect\hyperlink{ref-vishwanath2009}{16}{]} using \(\textsf{R}\) and the package \emph{mgcv}{[}\protect\hyperlink{ref-wood2017}{37}{]} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in the \protect\hyperlink{workflow}{Appendix}.
-One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a \emph{p-value}, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
-The model is therefore able to identify changes between the groups at time points were data was not directly measured even with missing data exists ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
-
-We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research {[}\protect\hyperlink{ref-begley2015}{68},\protect\hyperlink{ref-weissgerber2018}{69}{]}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
-Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
-We have made all the data and code used in this paper accessible, and we hope that this will encourage other researchers to do the same with future projects.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-We have presented GAMs as a method to analyze longitudinal biomedical data.
-Future directions of this work will include simulation-based estimations of statistical power using GAMs, as well as demonstrating the prediction capabilities of these models using large datasets.
-By making the data and code used in this paper accessible, we hope to address the need of creating and sharing reproducible work in biomedical research.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-This work was supported by the National Science Foundation Career Award (CBET 1751554, TJM) and the Arkansas Biosciences Institute.
-
-\hypertarget{declaration-of-conflicting-interests}{%
-\section{Declaration of Conflicting Interests}\label{declaration-of-conflicting-interests}}
-
-The Authors declare that there is no conflict of interest.
-
-\hypertarget{supplementary-materials}{%
-\section*{Supplementary Materials}\label{supplementary-materials}}
-\addcontentsline{toc}{section}{Supplementary Materials}
-
-An Appendix which contains all the code used to create this manuscript, along with a basic workflow to implement GAMs in R is available as Supplementary Material in PDF.
-A GitHub repository containing all the code used for this paper along with detailed instructions for its use is available at \url{https://github.com/aimundo/GAMs-biomedical-research}.
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-roblyer2011}{}}%
-\CSLLeftMargin{{[}1{]} }
-\CSLRightInline{D.
-Roblyer, S.
-Ueda, A.
-Cerussi, W.
-Tanamai, A.
-Durkin, R.
-Mehta, D.
-Hsiang, J.A.
-Butler, C.
-McLaren, W.-P.
-Chen, B.
-Tromberg, Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment, Proceedings of the National Academy of Sciences.
-108 (2011) 14626--14631.
-\url{https://doi.org/10.1073/pnas.1013103108}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tank2020}{}}%
-\CSLLeftMargin{{[}2{]} }
-\CSLRightInline{A.
-Tank, H.M.
-Peterson, V.
-Pera, S.
-Tabassum, A.
-Leproux, T.
-O'Sullivan, E.
-Jones, H.
-Cabral, N.
-Ko, R.S.
-Mehta, B.J.
-Tromberg, D.
-Roblyer, Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens, Breast Cancer Research.
-22 (2020).
-\url{https://doi.org/10.1186/s13058-020-01262-1}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-pavlov2018}{}}%
-\CSLLeftMargin{{[}3{]} }
-\CSLRightInline{M.V.
-Pavlov, T.I.
-Kalganova, Y.S.
-Lyubimtseva, Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy, Journal of Biomedical Optics.
-23 (2018) 1.
-\url{https://doi.org/10.1117/1.jbo.23.9.091410}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-demidov2018}{}}%
-\CSLLeftMargin{{[}4{]} }
-\CSLRightInline{V.
-Demidov, A.
-Maeda, M.
-Sugita, V.
-Madge, S.
-Sadanand, C.
-Flueraru, I.A.
-Vitkin, Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography, Scientific Reports.
-8 (2018).
-\url{https://doi.org/10.1038/s41598-017-18635-w}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ritter2001}{}}%
-\CSLLeftMargin{{[}5{]} }
-\CSLRightInline{G.
-Ritter, L.
-Cohen, C.
-Williams, E.
-Richards, L.
-Old, S.
-Welt, {Serological analysis of human anti-human antibody responses in colon cancer patients treated with repeated doses of humanized monoclonal antibody A33}, {Cancer Research}.
-{61} (2001) 6851--6859.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-roth2017}{}}%
-\CSLLeftMargin{{[}6{]} }
-\CSLRightInline{E.M.
-Roth, A.C.
-Goldberg, A.L.
-Catapano, A.
-Torri, G.D.
-Yancopoulos, N.
-Stahl, A.
-Brunet, G.
-Lecorps, H.M.
-Colhoun, Antidrug antibodies in patients treated with alirocumab, New England Journal of Medicine.
-376 (2017) 1589--1590.
-\url{https://doi.org/10.1056/nejmc1616623}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jones2018}{}}%
-\CSLLeftMargin{{[}7{]} }
-\CSLRightInline{J.D.
-Jones, H.E.
-Ramser, A.E.
-Woessner, K.P.
-Quinn, In vivo multiphoton microscopy detects longitudinal metabolic changes associated with delayed skin wound healing, Communications Biology.
-1 (2018).
-\url{https://doi.org/10.1038/s42003-018-0206-4}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-skala2010}{}}%
-\CSLLeftMargin{{[}8{]} }
-\CSLRightInline{M.C.
-Skala, A.
-Fontanella, L.
-Lan, J.A.
-Izatt, M.W.
-Dewhirst, Longitudinal optical imaging of tumor metabolism and hemodynamics, Journal of Biomedical Optics.
-15 (2010) 011112.
-\url{https://doi.org/10.1117/1.3285584}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-greening2018}{}}%
-\CSLLeftMargin{{[}9{]} }
-\CSLRightInline{G.J.
-Greening, K.P.
-Miller, C.R.
-Spainhour, M.D.
-Cato, T.J.
-Muldoon, Effects of isoflurane anesthesia on physiological parameters in murine subcutaneous tumor allografts measured via diffuse reflectance spectroscopy, Biomedical Optics Express.
-9 (2018) 2871.
-\url{https://doi.org/10.1364/boe.9.002871}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-sio2016}{}}%
-\CSLLeftMargin{{[}10{]} }
-\CSLRightInline{T.T.
-Sio, P.J.
-Atherton, B.J.
-Birckhead, D.J.
-Schwartz, J.A.
-Sloan, D.K.
-Seisler, J.A.
-Martenson, C.L.
-Loprinzi, P.C.
-Griffin, R.F.
-Morton, J.C.
-Anders, T.J.
-Stoffel, R.E.
-Haselow, R.B.
-Mowat, M.A.N.
-Wittich, J.D.
-Bearden, R.C.
-Miller, Repeated measures analyses of dermatitis symptom evolution in breast cancer patients receiving radiotherapy in a phase 3 randomized trial of mometasone furoate vs placebo (N06C4 {[}alliance{]}), Supportive Care in Cancer.
-24 (2016) 3847--3855.
-\url{https://doi.org/10.1007/s00520-016-3213-3}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-kamstra2015}{}}%
-\CSLLeftMargin{{[}11{]} }
-\CSLRightInline{J.I.
-Kamstra, P.U.
-Dijkstra, M.
-van Leeuwen, J.L.N.
-Roodenburg, J.A.
-Langendijk, Mouth opening in patients irradiated for head and neck cancer: A prospective repeated measures study, Oral Oncology.
-51 (2015) 548--555.
-\url{https://doi.org/10.1016/j.oraloncology.2015.01.016}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wagenmakers2008}{}}%
-\CSLLeftMargin{{[}12{]} }
-\CSLRightInline{E.-J.
-Wagenmakers, M.
-Lee, T.
-Lodewyckx, G.J.
-Iverson, Bayesian versus frequentist inference, in: Bayesian Evaluation of Informative Hypotheses, Springer New York, 2008: pp.
-181--207.
-\url{https://doi.org/10.1007/978-0-387-09612-4_9}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gueorguieva2004}{}}%
-\CSLLeftMargin{{[}13{]} }
-\CSLRightInline{R. Gueorguieva, J.H. Krystal, Move over {ANOVA}, Archives of General Psychiatry. 61 (2004) 310. \url{https://doi.org/10.1001/archpsyc.61.3.310}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-schober2018}{}}%
-\CSLLeftMargin{{[}14{]} }
-\CSLRightInline{P.
-Schober, T.R.
-Vetter, Repeated measures designs and analysis of longitudinal data, Anesthesia {\&} Analgesia.
-127 (2018) 569--575.
-\url{https://doi.org/10.1213/ane.0000000000003511}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-pinheiro2006}{}}%
-\CSLLeftMargin{{[}15{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, {Mixed-effects models in S and S-PLUS}, Springer Science; Business Media, 2006. https://doi.org/\url{https://doi.org/10.1007/b98882}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-vishwanath2009}{}}%
-\CSLLeftMargin{{[}16{]} }
-\CSLRightInline{K.
-Vishwanath, H.
-Yuan, W.T.
-Barry, M.W.
-Dewhirst, N.
-Ramanujam, Using optical spectroscopy to longitudinally monitor physiological changes within solid tumors, Neoplasia.
-11 (2009) 889--900.
-\url{https://doi.org/10.1593/neo.09580}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-dennis2019}{}}%
-\CSLLeftMargin{{[}17{]} }
-\CSLRightInline{B.
-Dennis, J.M.
-Ponciano, M.L.
-Taper, S.R.
-Lele, Errors in statistical inference under model misspecification: Evidence, hypothesis testing, and {AIC}, Frontiers in Ecology and Evolution.
-7 (2019).
-\url{https://doi.org/10.3389/fevo.2019.00372}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wang2019}{}}%
-\CSLLeftMargin{{[}18{]} }
-\CSLRightInline{B.
-Wang, Z.
-Zhou, H.
-Wang, X.M.
-Tu, C.
-Feng, The p-value and model specification in statistics, General Psychiatry.
-32 (2019) e100081.
-\url{https://doi.org/10.1136/gpsych-2019-100081}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-liu2010}{}}%
-\CSLLeftMargin{{[}19{]} }
-\CSLRightInline{C.
-Liu, T.P.
-Cripe, M.-O.
-Kim, Statistical issues in longitudinal data analysis for treatment efficacy studies in the biomedical sciences, Molecular Therapy.
-18 (2010) 1724--1730.
-\url{https://doi.org/10.1038/mt.2010.127}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-halsey2015}{}}%
-\CSLLeftMargin{{[}20{]} }
-\CSLRightInline{L.G.
-Halsey, D.
-Curran-Everett, S.L.
-Vowler, G.B.
-Drummond, The fickle p value generates irreproducible results, Nature Methods.
-12 (2015) 179--185.
-\url{https://doi.org/10.1038/nmeth.3288}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-abdi2010}{}}%
-\CSLLeftMargin{{[}21{]} }
-\CSLRightInline{H. Abdi, {Holm's sequential Bonferroni procedure}, Encyclopedia of Research Design. 1 (2010) 1--8. \url{https://doi.org/10.4135/9781412961288.n178}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nakagawa2004}{}}%
-\CSLLeftMargin{{[}22{]} }
-\CSLRightInline{S.
-Nakagawa, A farewell to bonferroni: The problems of low statistical power and publication bias, Behavioral Ecology.
-15 (2004) 1044--1045.
-\url{https://doi.org/10.1093/beheco/arh107}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gelman2012}{}}%
-\CSLLeftMargin{{[}23{]} }
-\CSLRightInline{A.
-Gelman, J.
-Hill, M.
-Yajima, Why we (usually) don{\textquotesingle}t have to worry about multiple comparisons, Journal of Research on Educational Effectiveness.
-5 (2012) 189--211.
-\url{https://doi.org/10.1080/19345747.2011.618213}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-albers2019}{}}%
-\CSLLeftMargin{{[}24{]} }
-\CSLRightInline{C. Albers, The problem with unadjusted multiple and sequential statistical testing, Nature Communications. 10 (2019). \url{https://doi.org/10.1038/s41467-019-09941-0}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ugrinowitsch2004}{}}%
-\CSLLeftMargin{{[}25{]} }
-\CSLRightInline{C.
-Ugrinowitsch, G.W.
-Fellingham, M.D.
-Ricard, Limitations of ordinary least squares models in analyzing repeated measures data, Medicine {\&} Science in Sports {\&} Exercise.
-(2004) 2144--2148.
-\url{https://doi.org/10.1249/01.mss.0000147580.40591.75}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-huynh1976}{}}%
-\CSLLeftMargin{{[}26{]} }
-\CSLRightInline{H.
-Huynh, L.S.
-Feldt, Estimation of the box correction for degrees of freedom from sample data in randomized block and split-plot designs, Journal of Educational Statistics.
-1 (1976) 69--82.
-\url{https://doi.org/10.3102/10769986001001069}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-greenhouse1959}{}}%
-\CSLLeftMargin{{[}27{]} }
-\CSLRightInline{S.W. Greenhouse, S. Geisser, On methods in the analysis of profile data, Psychometrika. 24 (1959) 95--112. \url{https://doi.org/10.1007/bf02289823}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-haverkamp2017}{}}%
-\CSLLeftMargin{{[}28{]} }
-\CSLRightInline{N.
-Haverkamp, A.
-Beauducel, Violation of the sphericity assumption and its effect on type-i error rates in repeated measures {ANOVA} and multi-level linear models ({MLM}), Frontiers in Psychology.
-8 (2017).
-\url{https://doi.org/10.3389/fpsyg.2017.01841}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-keselman2001}{}}%
-\CSLLeftMargin{{[}29{]} }
-\CSLRightInline{H.J.
-Keselman, J.
-Algina, R.K.
-Kowalchuk, The analysis of repeated measures designs: A review, British Journal of Mathematical and Statistical Psychology.
-54 (2001) 1--20.
-\url{https://doi.org/10.1348/000711001159357}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-charan2013}{}}%
-\CSLLeftMargin{{[}30{]} }
-\CSLRightInline{J.
-Charan, N.
-Kantharia, How to calculate sample size in animal studies?, Journal of Pharmacology and Pharmacotherapeutics.
-4 (2013) 303.
-\url{https://doi.org/10.4103/0976-500x.119726}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-barr2013}{}}%
-\CSLLeftMargin{{[}31{]} }
-\CSLRightInline{D.J.
-Barr, R.
-Levy, C.
-Scheepers, H.J.
-Tily, Random effects structure for confirmatory hypothesis testing: Keep it maximal, Journal of Memory and Language.
-68 (2013) 255--278.
-\url{https://doi.org/10.1016/j.jml.2012.11.001}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rose2012}{}}%
-\CSLLeftMargin{{[}32{]} }
-\CSLRightInline{N.L.
-Rose, H.
-Yang, S.D.
-Turner, G.L.
-Simpson, An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, {UK}, Geochimica Et Cosmochimica Acta.
-82 (2012) 113--135.
-\url{https://doi.org/10.1016/j.gca.2010.12.026}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-pedersen2019}{}}%
-\CSLLeftMargin{{[}33{]} }
-\CSLRightInline{E.J.
-Pedersen, D.L.
-Miller, G.L.
-Simpson, N.
-Ross, Hierarchical generalized additive models in ecology: An introduction with mgcv, {PeerJ}.
-7 (2019) e6876.
-\url{https://doi.org/10.7717/peerj.6876}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-simpson2018}{}}%
-\CSLLeftMargin{{[}34{]} }
-\CSLRightInline{G.L. Simpson, Modelling palaeoecological time series using generalised additive models, Frontiers in Ecology and Evolution. 6 (2018). \url{https://doi.org/10.3389/fevo.2018.00149}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-yang2012}{}}%
-\CSLLeftMargin{{[}35{]} }
-\CSLRightInline{L.
-Yang, G.
-Qin, N.
-Zhao, C.
-Wang, G.
-Song, Using a generalized additive model with autoregressive terms to study the effects of daily temperature on mortality, {BMC} Medical Research Methodology.
-12 (2012).
-\url{https://doi.org/10.1186/1471-2288-12-165}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-beck1998}{}}%
-\CSLLeftMargin{{[}36{]} }
-\CSLRightInline{N. Beck, S. Jackman, Beyond linearity by default: Generalized additive models, American Journal of Political Science. 42 (1998) 596. \url{https://doi.org/10.2307/2991772}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wood2017}{}}%
-\CSLLeftMargin{{[}37{]} }
-\CSLRightInline{S.N. Wood, Generalized additive models, Chapman; Hall/{CRC}, 2017. \url{https://doi.org/10.1201/9781315370279}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-r}{}}%
-\CSLLeftMargin{{[}38{]} }
-\CSLRightInline{R Core Team, R: A language and environment for statistical computing, R Foundation for Statistical Computing, Vienna, Austria, 2020. \url{https://www.R-project.org/}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wood2016}{}}%
-\CSLLeftMargin{{[}39{]} }
-\CSLRightInline{S.N.
-Wood, N.
-Pya, B.
-Säfken, Smoothing parameter and model selection for general smooth models, Journal of the American Statistical Association.
-111 (2016) 1548--1563.
-\url{https://doi.org/10.1080/01621459.2016.1180986}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-west2014}{}}%
-\CSLLeftMargin{{[}40{]} }
-\CSLRightInline{B.T.
-West, K.B.
-Welch, A.T.
-Galecki, Linear mixed models: A practical guide using statistical software, second edition, Taylor \& Francis, 2014.
-\url{https://books.google.com/books?id=hjT6AwAAQBAJ}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wolfinger1996}{}}%
-\CSLLeftMargin{{[}41{]} }
-\CSLRightInline{R.D.
-Wolfinger, Heterogeneous variance: Covariance structures for repeated measures, Journal of Agricultural, Biological, and Environmental Statistics.
-1 (1996) 205.
-\url{https://doi.org/10.2307/1400366}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-weiss2005}{}}%
-\CSLLeftMargin{{[}42{]} }
-\CSLRightInline{R.E. Weiss, Modeling longitudinal data, Springer New York, 2005. \url{https://doi.org/10.1007/0-387-28314-5}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-geisser1958}{}}%
-\CSLLeftMargin{{[}43{]} }
-\CSLRightInline{S.
-Geisser, S.W.
-Greenhouse, An extension of box{\textquotesingle}s results on the use of the {\textdollar}f{\textdollar} distribution in multivariate analysis, The Annals of Mathematical Statistics.
-29 (1958) 885--891.
-\url{https://doi.org/10.1214/aoms/1177706545}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-maxwell2017}{}}%
-\CSLLeftMargin{{[}44{]} }
-\CSLRightInline{S.E. Maxwell, H.D. Delaney, K. Kelley, Designing experiments and analyzing data, Routledge, 2017. \url{https://doi.org/10.4324/9781315642956}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-molenberghs2004}{}}%
-\CSLLeftMargin{{[}45{]} }
-\CSLRightInline{G. Molenberghs, Analyzing incomplete longitudinal clinical trial data, Biostatistics. 5 (2004) 445--464. \url{https://doi.org/10.1093/biostatistics/kxh001}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ma2012}{}}%
-\CSLLeftMargin{{[}46{]} }
-\CSLRightInline{Y.
-Ma, M.
-Mazumdar, S.G.
-Memtsoudis, Beyond repeated-measures analysis of variance, Regional Anesthesia and Pain Medicine.
-37 (2012) 99--105.
-\url{https://doi.org/10.1097/aap.0b013e31823ebc74}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-scheffer2002}{}}%
-\CSLLeftMargin{{[}47{]} }
-\CSLRightInline{J. Scheffer, Dealing with missing data, Research Letters in the Information and Mathematical Sciences. 3 (2002) 153--160.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-potthoff2006}{}}%
-\CSLLeftMargin{{[}48{]} }
-\CSLRightInline{R.F.
-Potthoff, G.E.
-Tudor, K.S.
-Pieper, V.
-Hasselblad, Can one assess whether missing data are missing at random in medical studies?, Statistical Methods in Medical Research.
-15 (2006) 213--234.
-\url{https://doi.org/10.1191/0962280206sm448oa}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-box1976}{}}%
-\CSLLeftMargin{{[}49{]} }
-\CSLRightInline{G.E.P. Box, Science and statistics, 71 (1976) 791--799. \url{https://doi.org/10.1080/01621459.1976.10480949}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlme}{}}%
-\CSLLeftMargin{{[}50{]} }
-\CSLRightInline{J. Pinheiro, D. Bates, S. DebRoy, D. Sarkar, R Core Team, {nlme}: Linear and nonlinear mixed effects models, 2020. \url{https://CRAN.R-project.org/package=nlme}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ohara2010}{}}%
-\CSLLeftMargin{{[}51{]} }
-\CSLRightInline{R. OHara, J. Kotze, Do not log-transform count data, (2010). \url{https://doi.org/10.1038/npre.2010.4136.1}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nelder1972}{}}%
-\CSLLeftMargin{{[}52{]} }
-\CSLRightInline{J.A. Nelder, R.W.M. Wedderburn, Generalized linear models, Journal of the Royal Statistical Society. Series A (General). 135 (1972) 370. \url{https://doi.org/10.2307/2344614}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-mcculloch2001}{}}%
-\CSLLeftMargin{{[}53{]} }
-\CSLRightInline{C. McCulloch, Generalized, linear, and mixed models, John Wiley \& Sons, New York, 2001.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-dobson2008}{}}%
-\CSLLeftMargin{{[}54{]} }
-\CSLRightInline{A. Dobson, An introduction to generalized linear models, CRC Press, Boca Raton, 2008.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-stroup2013}{}}%
-\CSLLeftMargin{{[}55{]} }
-\CSLRightInline{W. Stroup, Generalized linear mixed models : Modern concepts, methods and applications, CRC Press, Taylor \& Francis Group, Boca Raton, 2013.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hastie1987}{}}%
-\CSLLeftMargin{{[}56{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models: Some applications, Journal of the American Statistical Association.
-82 (1987) 371--386.
-\url{https://doi.org/10.1080/01621459.1987.10478440}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hefley2017}{}}%
-\CSLLeftMargin{{[}57{]} }
-\CSLRightInline{T.J.
-Hefley, K.M.
-Broms, B.M.
-Brost, F.E.
-Buderman, S.L.
-Kay, H.R.
-Scharf, J.R.
-Tipton, P.J.
-Williams, M.B.
-Hooten, The basis function approach for modeling autocorrelation in ecological data, Ecology.
-98 (2017) 632--646.
-\url{https://doi.org/10.1002/ecy.1674}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wegman1983}{}}%
-\CSLLeftMargin{{[}58{]} }
-\CSLRightInline{E.J. Wegman, I.W. Wright, Splines in statistics, Journal of the American Statistical Association. 78 (1983) 351--365. \url{https://doi.org/10.1080/01621459.1983.10477977}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wood2003}{}}%
-\CSLLeftMargin{{[}59{]} }
-\CSLRightInline{S.N. Wood, Thin plate regression splines, 65 (2003) 95--114. \url{https://doi.org/10.1111/1467-9868.00374}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-mcelreath2018}{}}%
-\CSLLeftMargin{{[}60{]} }
-\CSLRightInline{R. McElreath, Statistical rethinking, Chapman; Hall/{CRC}, 2018. \url{https://doi.org/10.1201/9781315372495}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-miller2019}{}}%
-\CSLLeftMargin{{[}61{]} }
-\CSLRightInline{D.L. Miller, Bayesian views of generalized additive modelling, arXiv Preprint arXiv:1902.01330. (2019).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ruppert2003}{}}%
-\CSLLeftMargin{{[}62{]} }
-\CSLRightInline{D. Ruppert, Semiparametric regression, Cambridge University Press, Cambridge New York, 2003.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
-\CSLLeftMargin{{[}63{]} }
-\CSLRightInline{G.
-Marra, S.N.
-Wood, Coverage properties of confidence intervals for generalized additive model components, Scandinavian Journal of Statistics.
-39 (2012) 53--74.
-\url{https://doi.org/10.1111/j.1467-9469.2011.00760.x}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
-\CSLLeftMargin{{[}64{]} }
-\CSLRightInline{G.L. Simpson, Gratia: Graceful 'ggplot'-based graphics and other functions for GAMs fitted using 'mgcv', 2020. \url{https://CRAN.R-project.org/package=gratia}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
-\CSLLeftMargin{{[}65{]} }
-\CSLRightInline{J. Harezlak, D. Ruppert, M.P. Wand, Semiparametric regression with {R}, Springer New York, 2018. \url{https://doi.org/10.1007/978-1-4939-8853-2}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
-\CSLLeftMargin{{[}66{]} }
-\CSLRightInline{T.A.
-Lang, D.G.
-Altman, Basic statistical reporting for articles published in biomedical journals: The {``}statistical analyses and methods in the published literature{''} or the {SAMPL} guidelines, International Journal of Nursing Studies.
-52 (2015) 5--9.
-\url{https://doi.org/10.1016/j.ijnurstu.2014.09.006}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hastie1995}{}}%
-\CSLLeftMargin{{[}67{]} }
-\CSLRightInline{T.
-Hastie, R.
-Tibshirani, Generalized additive models for medical research, Statistical Methods in Medical Research.
-4 (1995) 187--196.
-\url{https://doi.org/10.1177/096228029500400302}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-begley2015}{}}%
-\CSLLeftMargin{{[}68{]} }
-\CSLRightInline{C.G. Begley, J.P.A. Ioannidis, Reproducibility in science, Circulation Research. 116 (2015) 116--126. \url{https://doi.org/10.1161/circresaha.114.303819}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-weissgerber2018}{}}%
-\CSLLeftMargin{{[}69{]} }
-\CSLRightInline{T.L.
-Weissgerber, O.
-Garcia-Valencia, V.D.
-Garovic, N.M.
-Milic, S.J.
-Winham, Why we need to report more than {\textquotesingle}data were analyzed by t-tests or {ANOVA}{\textquotesingle}, {eLife}.
-7 (2018).
-\url{https://doi.org/10.7554/elife.36163}.}
-
-\end{CSLReferences}
-
-\hypertarget{appendix-appendix}{%
-\appendix}
-
-\appendixpagenumbering
-
-\counterwithin{figure}{section}
-
-\hypertarget{appendix}{%
-\section{APPENDIX}\label{appendix}}
-
-This section presents the code used to generate figures, models and simulated data from the main manuscript.
-
-\hypertarget{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}{%
-\subsection{Compound symmetry and independent errors in linear and quadratic responses}\label{compound-symmetry-and-independent-errors-in-linear-and-quadratic-responses}}
-
-This section simulates linear and quadratic data in the same manner as in Section \ref{simulation} in the main manuscript.
-The linear simulations using Figure \ref{fig:linear-cases-Appendix} show in panels A and D the simulated mean responses and individual data points.
-Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
-In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
-Panels D and H show the predictions from the rm-ANOVA model.
-
-The following code chunks produce a more comprehensive exploration of Figure \ref{fig:l-q-response} in the main manuscript.
-
-First, a function is created to simulate data across six timepoints using a linear or quadratic mean response, with correlated or uncorrelated errors.
-Each group has a different slope/concavity.
-The main function is the same for both groups, but a change in the sign allows to invert the trend.
+\hypertarget{linear-and-quadratic-longitudinal-trends}{%
+\subsection{Linear and quadratic longitudinal trends}\label{linear-and-quadratic-longitudinal-trends}}
+
+\hypertarget{function-for-linear-and-quadratic-trends-rm-anova-and-lmem-fits}{%
+\subsubsection{Function for linear and quadratic trends, rm-ANOVA and LMEM fits}\label{function-for-linear-and-quadratic-trends-rm-anova-and-lmem-fits}}
+
+The first function is \passthrough{\lstinline!example.R!}, which allows to simulate linear and quadratic data in the same manner as in Section 3.5 in the main manuscript and estimates rm-ANOVA and LMEM with interaction to the data.
+The error for each simulated trend can be correlated or uncorrelated
 
@@
 }
-
-##################Section for plotting#################################
-
-#This function will create the plots for either a "linear" or "quadratic" response
-
+\end{lstlisting}
+
+\hypertarget{a-composite-plot-for-the-trends}{%
+\subsubsection{A composite plot for the trends}\label{a-composite-plot-for-the-trends}}
+
+Function \passthrough{\lstinline!plot\_example.R!} uses the output of \passthrough{\lstinline!example.R!} to show the fit of a rm-ANOVA and a LMEM.
+It can be used to show an expanded version of Figure 1 in the main manuscript, presenting simulated data with correlated and uncorrelated errors and the correspoding rm-ANOVA and LMEM fits, which we do in the next subsection.
+
+\begin{lstlisting}[language=R]
+## This function plots the rm-ANOVA and LMEM for the data simulated in example.R
 plot_example <- function(sim_dat) {
-  ## Plot the simulated data (scatterplot)
-  txt<-20
+    # Plot the simulated data (scatterplot)
     p1 <- sim_dat$dat %>%
@@
                       color = treatment),
+                  size=3,
                   show.legend=FALSE) +
         theme_classic() +
-    theme(plot.title = element_text(size = txt, 
+        theme(plot.title = element_text(size = 20,
                                         face = "bold"),
-        text=element_text(size=txt))+
+              text=element_text(size=20))+
         thm1
 
-  #plot the simulated data with trajectories per each subject
-  p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = y, 
-               group = subject, 
-               color = treatment)
-           ) +
-    geom_line(aes(size = "Subjects"),
-              show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, 
-                  y = mu, 
-                  color = treatment,
-                  size = "Simulated Truth"), 
-              lty = 1,show.legend = FALSE) +
-    labs(y='response')+
-    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    theme_classic()+
-     theme(plot.title = element_text(size = txt, 
-                                face = "bold"),
-     text=element_text(size=txt))+
-    thm1
-
-  #plot the errors
-   p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, 
-               y = errors,
-               group = subject, 
-               color = treatment)) +
-    geom_line(show.legend=FALSE) +
-     labs(y='errors')+
-     theme_classic()+
-     theme(plot.title = element_text(size = txt, 
-                                  face = "bold"),
-        text=element_text(size=txt))+
-    thm1
-
     #plot the model predictions for rm-ANOVA
-  p4 <- ggplot(sim_dat$dat, 
+    p2 <- ggplot(sim_dat$dat,
                  aes(x = time,
@@
 }
-
-#Store each plot in a separate object
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"))
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"))
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent")) 
 \end{lstlisting}
 
+\hypertarget{plotting-rm-anova-and-lmem-fits-for-linear-and-quadratic-trends-in-data}{%
+\subsubsection{Plotting rm-ANOVA and LMEM fits for linear and quadratic trends in data}\label{plotting-rm-anova-and-lmem-fits-for-linear-and-quadratic-trends-in-data}}
+
+In this subsection, we use \passthrough{\lstinline!example.R!} and a modified version \passthrough{\lstinline!plot\_example.R!} (\passthrough{\lstinline!plot\_example\_Appendix.R!}) to create an expanded version of Figure 1 on the main manuscript.
+The only difference between \passthrough{\lstinline!plot\_example.R!} and \passthrough{\lstinline!plot\_example\_Appendix.R!} is the inclusion of a \passthrough{\lstinline!ggplot2!} object (\passthrough{\lstinline!p3!}) that allows to plot the simulated errors.
+
+Figure \ref{fig:linear-cases-Appendix} show in panels A and D the simulated mean responses and individual data points.
+Panels C and G show a visual interpretation of ``correlation'' in the responses: In panel C, subjects that have a value of the random error \(\varepsilon\) either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response.
+In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.
+Panels D and H show the predictions from the rm-ANOVA model.
+
+\hypertarget{fits-for-linear-trends}{%
+\paragraph{Fits for linear trends}\label{fits-for-linear-trends}}
+
+The chunk below calls both \passthrough{\lstinline!example.R!} and \passthrough{\lstinline!plot\_example\_Appendix.R!} to simulate data and create the composite plots.
+
+\begin{lstlisting}[language=R]
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","example.R"))
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","plot_example_Appendix.R"))
+
+A1<-plot_example_Appendix(example(fun_type = "linear", error_type = "correlated"))
+
+B1<-plot_example_Appendix(example(fun_type = "linear", error_type = "independent"))
+
+C1<-plot_example_Appendix(example(fun_type = "quadratic", error_type = "correlated"))
+
+D1<-plot_example_Appendix(example(fun_type = "quadratic", error_type = "independent")) 
+\end{lstlisting}
+
 \begin{figure}
 
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/linear-cases-Appendix-1}
+{\centering \includegraphics[width=1\linewidth]{Appendix_B_files/figure-latex/linear-cases-Appendix-1}
 
@@
 
+\hypertarget{fits-for-quadratic-trends}{%
+\paragraph{Fits for quadratic trends}\label{fits-for-quadratic-trends}}
+
 For the quadratic response case, Figure \ref{fig:quadratic-cases-Appendix} shows the simulated responses using compound symmetry and independent errors.
@@
 
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/quadratic-cases-Appendix-1}
+{\centering \includegraphics[width=1\linewidth]{Appendix_B_files/figure-latex/quadratic-cases-Appendix-1}
 
@@
 
-\hypertarget{basis-functions-and-gams}{%
-\subsection{Basis functions and GAMs}\label{basis-functions-and-gams}}
-
-This code produces Figure \ref{fig:basis-plot} from the main manuscript.
-Briefly, a non-linear (quadratic) response is simulated, a gam model is fitted and the basis are extracted in order to explain how the smooth is constructed.
-The code for data simulation is used again here for the sake of keeping the same structure, although the data can be simulated in a more simple fashion.
-
-\begin{lstlisting}[language=R]
-#generate the response: the same initial procedure from the previous section to simulate
-#the response
-set.seed(1)
-n_time = 6
- x <- seq(1,6, length.out = n_time)
- mu <- matrix(0, length(x), 2)
- mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
- mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
- y <- array(0, dim = c(length(x), 2, 10))
- errors <- array(0, dim = c(length(x), 2, 10))
- for (i in 1:2) {     # number of treatments
-     for (j in 1:10) {  # number of subjects
-         # compound symmetry errors
-         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
-         y[, i, j] <- mu[, i] + errors[, i, j]
-     }
- }
-
- #label each table
-  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
- dimnames(mu) <- list(time = x, treatment = 1:2)
-
- #Convert to dataframes with subject, time and group columns
- dat <- as.data.frame.table(y, responseName = "y")
- dat_errors <- as.data.frame.table(errors, responseName = "errors")
- dat_mu <- as.data.frame.table(mu, responseName = "mu")
- dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
- dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
- dat$time <- as.numeric(as.character(dat$time))
-
- #label subject per group
- dat <- dat %>%
-     mutate(subject = factor(paste(subject, treatment, sep = "-")))
-
- #extract  "Group 1" to fit the GAM
-  dat<-subset(dat,treatment==1)
- #keep just the response and timepoint columns
-   dat<-dat[,c('y','time')]
-
-   #GAM model of time, 5 basis
-gm<-gam(y~s(time,k=5),data=dat)
-
-#model_matrix (also known as) 'design matrix'
-#will contain the smooths used to create  model 'gm'
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
-#basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-  pivot_longer(
-    cols=starts_with("s")
-  )%>%
-  arrange(name) #ordering
-
-#length of dataframe to be created: number of basis by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-  mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be 
-#multiplied by the coefficients
-
-#creating labeller to change the labels in the basis plots
-
-basis_names<-c(
-  `1`="Intercept",
-  `2`="1",
-  `3`="2",
-  `4`="3",
-  `5`="4"
-)
-
-#calculating the final smooth by aggregating the basis functions
-
-smooth<-basis_plot%>% 
-  group_by(time)%>%
-  summarize(smooth=sum(mod_val))
-
-#original basis
-sz<-1
-p11<-ggplot(basis_plot,
-            aes(x=time,
-                y=value_orig,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(size=sz,
-            show.legend=FALSE)+
-  geom_point(size=sz+1,
-             show.legend = FALSE)+
-  labs(y='Basis functions')+
-  facet_wrap(~Basis,
-             labeller = as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#penalized basis
-p12<-ggplot(basis_plot,
-            aes(x=time,
-                y=mod_val,
-                colour=as.factor(Basis)
-                )
-            )+
-  geom_line(show.legend = FALSE,
-            size=sz)+
-  geom_point(show.legend = FALSE,
-             size=sz+1)+
-  labs(y='Penalized \n basis functions')+
-  scale_y_continuous(breaks=seq(-1,1,1))+
-  facet_wrap(~Basis,
-             labeller=as_labeller(basis_names)
-             )+
-  theme_classic()+
-  thm
-
-#heatmap of the coefficients
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,
-            aes(x=Basis,
-                y=Basis))+
-  geom_tile(aes(fill = cof), 
-            colour = "black") +
-    scale_fill_gradient(low = "white",
-                        high = "#B50A2AFF")+ #color picked from KikiMedium
-  labs(x='Basis',
-       y='Basis')+
-  scale_x_discrete(labels=x_labels)+
-  geom_text(aes(label=round(cof,2)),
-            size=7,
-            show.legend = FALSE)+
-  theme_classic()+
-  theme(legend.title = element_blank())
-
-#plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,
-            aes(x=time,y=y))+
-  geom_point(size=sz+1)+
-  labs(y='Simulated \n response')+
-  geom_line(data=smooth,
-            aes(x=time,
-                y=smooth),
-            color="#6C581DFF",
-            size=sz+1)+
-  theme_classic()
-
-#Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-  theme(
-     text=element_text(size=18)
-     )
-\end{lstlisting}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/basis-plot-appendix-1}
-
-}
-
-\caption{Basis functions for a single smoother for time.
-\textbf{A}: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2.
-\textbf{B}: Matrix of basis function weights.
-Each basis function is multiplied by a coefficient which can be positive or negative.
-The coefficient determines the overall effect of each basis in the final smoother.
-\textbf{C}: Weighted basis functions.
-Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
-Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
-\textbf{D}: Smoother for time and original data points.
-The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot-appendix}
-\end{figure}
-
-\hypertarget{tumor-data-simulation}{%
-\subsection{Longitudinal biomedical data simulation and GAMs}\label{tumor-data-simulation}}
-
-This section describes how to fit GAMs to longitudinal data using simulated data.
-First, data is simulated according to Section \ref{longitudinal-GAMs}, where reported data of oxygen saturation (\(\mbox{StO}_2\)) in tumors under either chemotherapy or saline control is used as a starting point to generate individual responses in each group.
-
-\begin{lstlisting}[language=R]
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
-#alpha for ribbon
-al<-0.8
-
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")
-
-simulate_data <- function(dat, n = 10, sd = 5) {
-    dat_sim <- dat %>%
-        slice(rep(1:n(), each = n)) %>%
-        group_by(Group, Day) %>%
-        mutate(
-               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-               subject=rep(1:10),
-               subject=factor(paste(subject, Group, sep = "-"))
-               ) %>%
-        ungroup()
-
-    return(dat_sim)
-}
-
-#subject = factor(paste(subject, treatment, sep = "-")))
-
-n <- 10 #number of observations
-sd <- 10 #approximate sd from paper
-df <- 6
-dat_sim <- simulate_data(dat, n, sd)
-
-#plotting simulated data
-f2<-ggplot(dat_sim, aes(x = Day, 
-                        y = StO2_sim, 
-                        color = Group, 
-                        group=subject)) +
-    geom_point(show.legend=FALSE,
-               size=1.5,
-               alpha=0.6)+
-    geom_line(size=0.6, alpha=0.4,show.legend=FALSE)+
-    geom_line(aes(x = Day, 
-               y = StO2,
-               color=Group),
-           size=1.5,
-           data=dat,
-           inherit.aes=FALSE,
-              show.legend = FALSE)+
-  labs(y=expression(atop(StO[2],'(simulated)')))+
-  theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    thm1
-\end{lstlisting}
-
-\hypertarget{workflow}{%
-\subsection{A basic Workflow for GAMs}\label{workflow}}
-
-This section shows a basic workflow to fit a series of increasingly complex GAMs to the simulated data from the previous section.
-Graphical and parameter diagnostics for goodness of fit are discussed, as well as model comparison via AIC (Aikake Information Criterion).
-For simplicity, the confidence intervals (CIs) shown in this section for the models are the across the function CIs created by \passthrough{\lstinline!mgcv!} by default.
-However, for the pairwise comparisons of the third model we use \emph{simultaneous intervals} as in the main manuscript, and the code for creating simultaneous CIs for the smooths can be found in Section \ref{GAM-linear-plot}.
-
-\hypertarget{first-model}{%
-\subsubsection{First model}\label{first-model}}
-
-The first model fitted to the data is one that only accounts for different smooths by day.
-The model syntax specifies that \passthrough{\lstinline!gam\_00!} is the object that will contain all the model information, and that the model attempts to explain changes in \passthrough{\lstinline!StO2\_sim!} (simulated \(\mbox{StO}_2\)) using a smooth per \passthrough{\lstinline!Day!}.
-The model will use 5 basis functions (\passthrough{\lstinline!k=5!}) for the smooth.
-The smooth is constructed by default using thin plate regression splines.
-The smoothing parameter estimation method used is the restricted maximum likelihood (\passthrough{\lstinline!REML!}).
-
-\begin{lstlisting}[language=R]
-gam_00<-gam(StO2_sim ~ s(Day, k = 5),
-            method='REML',
-            data  = dat_sim)
-\end{lstlisting}
-
-To obtain model diagnostics, two methodologies are to be used: 1) graphical diagnostics, and 2) a model check.
-In the first case, the functions \passthrough{\lstinline!appraise!} and \passthrough{\lstinline!draw!} from the package \emph{gratia} can be used to obtain a single output with all the graphical diagnostics.
-For model check, the functions \passthrough{\lstinline!gam.check!} and \passthrough{\lstinline!summary!} from \emph{mgcv} provide detailed information about the model fit and its parameters.
-Keep in mind that \passthrough{\lstinline!gam.check!} is a function that also provides the graphical diagnostics obtained using \emph{gratia}, if such graphical output is not desired the source code can be accessed typing \passthrough{\lstinline!gam.check!} in the Console, and the code without the graphical output can be used in a custom function, which is the approach we follow later).
-
-\hypertarget{graphical-diagnostics}{%
-\paragraph{Graphical diagnostics}\label{graphical-diagnostics}}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/first-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the first GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:first-GAM-diag}
-\end{figure}
-
-From the output of the function \passthrough{\lstinline!appraise!} in Figure \ref{fig:first-GAM-diag}, the major indicators of concern about the model are the QQ plot of residuals and the histogram of residuals.
-The QQ plot shows that the errors are not reasonably located along the 45\(^{\circ}\) line (which indicates normality), as there are multiple points that deviate from the trend, specially in the tails.
-The histogram also shows that the variation (residuals) is not following the assumption of a normal distribution.
-
-The \passthrough{\lstinline!draw!} function permits to plot the smooths as \passthrough{\lstinline!ggplot2!} objects, which eases subsequent manipulation, if desired.
-Because model \passthrough{\lstinline!gam\_00!} specifies only one smooth for the time covariate (Day), the plot only contains only one smooth.
-Note that the smooth shows an almost linear profile.
-
-\hypertarget{gam-00-model-check}{%
-\paragraph{Model check}\label{gam-00-model-check}}
-
-Special attention must be paid to the parameter `k-index' from \passthrough{\lstinline!gam.check!} (which calls \passthrough{\lstinline!k.check!} to perform the calculation).
-This parameter indicates if the basis dimension of the smooth is adequate, i.e., it checks that the basis used to create the smooth are adequate to capture the trends in the data.
-If the model is not adequately capturing the trends in the data, this is indicated by a low k-index value (\textless1).
-Because we plot the model diagnostics using \passthrough{\lstinline!appraise!} later, the graphical output from \passthrough{\lstinline!gam.check!} will be suppressed by creating a custom function to obtain just the model estimates, thus avoiding repetition of the diagnostic plots.
-This will be achieved by calling the source code of \passthrough{\lstinline!gam.check!} and using the appropriate code in a new function that will be called \passthrough{\lstinline!gam.diagnostics!}.
-We are not including in the Appendix the code for the function \passthrough{\lstinline!gam.diagnostics!} as it is rather long, but if desired it can be accessed by going to the \passthrough{\lstinline!Appendix.Rmd!} file in the GitHub repository and scrolling to this exact place (the code is not included in the final output, but is evaluated to create the function).
-
-We now call \passthrough{\lstinline!gam.diagnostics!} to provide the desired diagnostic output, as well as a summary of the fitted model, which is obtained by calling \passthrough{\lstinline!summary!}.
-
-\begin{lstlisting}[language=R]
-gam.diagnostics(gam_00)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 5 iterations.
-## Gradient range [-0.0003727881,-6.621452e-07]
-## (score 444.0118 & scale 450.6638).
-## Hessian positive definite, eigenvalue range [0.3881695,49.00676].
-## Model rank =  5 / 5 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##          k'  edf k-index p-value    
-## s(Day) 4.00 2.11    0.36  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-From the output, it can be seen that the \passthrough{\lstinline!k-index!} is 0.36, which indicates that the model is not capturing the variability in the data.
-The \passthrough{\lstinline!edf!} (effective degrees of freedom) is an indicator of the complexity of the smooth.
-Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
-
-\begin{lstlisting}[language=R]
-summary(gam_00)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ s(Day, k = 5)
-## 
-## Parametric coefficients:
-##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   22.967      2.123   10.82   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##          edf Ref.df     F  p-value    
-## s(Day) 2.114  2.565 7.633 0.000517 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.153   Deviance explained = 17.2%
-## -REML = 444.01  Scale est. = 450.66    n = 100
-\end{lstlisting}
-
-From the \passthrough{\lstinline!summary!} function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
-The link function is `identity' as the model does not make any transformation on the predictors.
-The `significance of smooth terms' \emph{p-value} indicates if each smooth is adding significance to the model.
-Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \(\approx\) 17\%.
-
-\hypertarget{second-model}{%
-\subsubsection{Second model}\label{second-model}}
-
-The major flaw of \passthrough{\lstinline!gam\_00!} is that this model is not taking into account the fact that the data is nested in groups.
-The next iteration is a model where a different smooth of time (Day) is assigned for each group using \passthrough{\lstinline!by = Group!} in the model syntax.
-
-\begin{lstlisting}[language=R]
-gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5),
-            method ='REML',
-            data  = dat_sim)
-
-gam.diagnostics(gam_01)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 7 iterations.
-## Gradient range [-5.51754e-05,2.671715e-06]
-## (score 423.3916 & scale 280.8777).
-## Hessian positive definite, eigenvalue range [0.3162258,48.5557].
-## Model rank =  9 / 9 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                         k'  edf k-index p-value    
-## s(Day):GroupControl   4.00 3.39    0.43  <2e-16 ***
-## s(Day):GroupTreatment 4.00 3.23    0.43  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{lstlisting}
-
-Diagnostics for this model indicate that the k-index is still below 1 (0.43 from \passthrough{\lstinline!gam.check!}), and that the residuals are still not following a normal distribution (Figure \ref{fig:second-GAM-diag}).
-Moreover, the smooths (plotted via the \passthrough{\lstinline!draw()!} function) appear with a fairly linear profile, which indicates they are still not capturing the trends observed in the data.
-
-\begin{lstlisting}[language=R]
-summary(gam_01)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ s(Day, by = Group, k = 5)
-## 
-## Parametric coefficients:
-##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)   22.967      1.676    13.7   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                         edf Ref.df      F p-value    
-## s(Day):GroupControl   3.392  3.794  3.817  0.0304 *  
-## s(Day):GroupTreatment 3.229  3.682 21.174  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.472   Deviance explained = 50.8%
-## -REML = 423.39  Scale est. = 280.88    n = 100
-\end{lstlisting}
-
-From \passthrough{\lstinline!summary()!}, the deviance explained by the model is \(\approx\) 51\%.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/second-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the second GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooth for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:second-GAM-diag}
-\end{figure}
-
-\hypertarget{third-model}{%
-\subsubsection{Third model}\label{third-model}}
-
-Model \passthrough{\lstinline!gam\_00!} was built for didactic purposes to cover the simplest case, but it does not account for the nesting of the data by Group, which is apparent from the type of smooth fitted (a single smooth), the model diagnostics, and, the low variance explained by the model.
-On the other hand, \passthrough{\lstinline!gam\_01!} takes into account the nesting within each group and provides better variance explanation, but as indicated in Section \ref{longitudinal-GAMs}, in order to differentiate between each group a parametric term needs to be added to the model for the interaction of \emph{Day} and \emph{Group}.
-
-This is because in \passthrough{\lstinline!gam\_01!} separate smooths were fitted per group and those smooths also tried to account for the different means of the response in the two groups.
-Adding a parametric term for \passthrough{\lstinline!Group!} enables the smooths to capture the time course-differences of each group.
-The resulting model is \passthrough{\lstinline!gam\_02!}, which is the model fitted in the main manuscript.
-
-\begin{lstlisting}[language=R]
-#GAM for StO2
-
-gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
-            method='REML',
-            data  = dat_sim)
-
-gam.diagnostics(gam_02)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Method: REML   Optimizer: outer newton
-## full convergence after 10 iterations.
-## Gradient range [-8.164307e-08,1.500338e-08]
-## (score 355.8554 & scale 64.53344).
-## Hessian positive definite, eigenvalue range [1.174841,48.08834].
-## Model rank =  10 / 10 
-## 
-## Basis dimension (k) checking results. Low p-value (k-index<1) may
-## indicate that k is too low, especially if edf is close to k'.
-## 
-##                         k'  edf k-index p-value
-## s(Day):GroupControl   4.00 3.87    1.02    0.55
-## s(Day):GroupTreatment 4.00 3.88    1.02    0.58
-\end{lstlisting}
-
-By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From \passthrough{\lstinline!gam.check!}, the k-index is now at an acceptable value (\(\approx\) 1.02).
-
-\begin{lstlisting}[language=R]
-summary(gam_02)
-\end{lstlisting}
-
-\begin{lstlisting}
-## 
-## Family: gaussian 
-## Link function: identity 
-## 
-## Formula:
-## StO2_sim ~ Group + s(Day, by = Group, k = 5)
-## 
-## Parametric coefficients:
-##                Estimate Std. Error t value Pr(>|t|)    
-## (Intercept)       9.084      1.136   7.996 4.09e-12 ***
-## GroupTreatment   27.766      1.607  17.282  < 2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Approximate significance of smooth terms:
-##                         edf Ref.df     F p-value    
-## s(Day):GroupControl   3.873  3.990 17.57  <2e-16 ***
-## s(Day):GroupTreatment 3.879  3.991 89.33  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## R-sq.(adj) =  0.879   Deviance explained = 88.9%
-## -REML = 355.86  Scale est. = 64.533    n = 100
-\end{lstlisting}
-
-From \passthrough{\lstinline!summary!}, the model is able to capture 89\% of the variance in the data, which is a substantial improvement over the variance explained by \passthrough{\lstinline!gam\_00!} and \passthrough{\lstinline!gam\_01!}.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/final-GAM-diag-1}
-
-}
-
-\caption{Graphical diagnostics for the final GAM model.
-Left: Graphical diagnostics provided by the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.
-Right: Fitted smooths for the model, provided by the function \passthrough{\lstinline!draw!}.}\label{fig:final-GAM-diag}
-\end{figure}
-
-\hypertarget{comparing-models-via-aic}{%
-\subsubsection{Comparing models via AIC}\label{comparing-models-via-aic}}
-
-One final comparison that can be made for model selection involves the use of the Aikake Information Criterion (AIC).
-This metric is used to estimate information loss, which we want to minimize with an appropriate model.
-Therefore, when 2 or more models are compared, the model with lower AIC is preferred.
-In R, the comparison is done using the \passthrough{\lstinline!AIC!} function.
-
-\begin{lstlisting}[language=R]
-AIC(gam_00,gam_01,gam_02)
-\end{lstlisting}
-
-\begin{lstlisting}
-##               df      AIC
-## gam_00  4.564893 900.8257
-## gam_01  9.476137 858.6051
-## gam_02 10.980983 712.2067
-\end{lstlisting}
-
-The output in this case is expected: model \passthrough{\lstinline!gam\_02!} has a lower AIC (712.46) whereas the initial two models have higher AICs (900 and 858).
-The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
-
-\hypertarget{pairwise-comparisons-of-smooth-confidence-intervals}{%
-\paragraph{Pairwise comparisons of smooth confidence intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}
-
-The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \ref{GAM-significance}.
-
-In this case, the ``design matrix'' is used to estimate the pairwise comparisons (see main manuscript for details and associated references).
-Briefly, the ``design matrix'' (also known as the ``Xp matrix'') from the selected model (\passthrough{\lstinline!gam\_02!}) is used to calculate a 95\% confidence interval of the difference between the smooth terms for each group.
-This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).
-
-We want to emphasize that for the model used here we have included the group means in order to keep the pairwise comparisons on the scale of the response (a change from the code which appears in the chunk with a comment that says \passthrough{\lstinline!\#\#\#IMPORTANT!}).
-This approach works well for models like \passthrough{\lstinline!gam\_02!} but implementing this on models with more parametric terms can be challenging.
-However, we do believe that the model presented in the paper covers a wide range of situations and our approach here for the pairwise comparisons will be useful for most biomedical researchers.
-
-\begin{lstlisting}[language=R]
-##Pairwise comparisons
-difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
-  ## make sure f1 and f2 are characters
-  f1 <-  as.character(f1)
-  f2 <-  as.character(f2)
-  cnames <- colnames(Xp)
-  ## columns of Xp associated with pair of smooths
-  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
-  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
-  ## rows of Xp associated with pair of smooths
-  r1 <- data[[by_var]] == f1
-  r2 <- data[[by_var]] == f2
-
-  ## difference rows of Xp for pair of smooths
-  X <- Xp[r1, ] - Xp[r2, ]
-
-  ### IMPORTANT: Keep group means####
-  ## zero the cols related to other splines
- # X[, ! (c1 | c2)] <- 0
-
-  ## zero out the parametric cols
-  #X[, !grepl('^s\\(', cnames)] <- 0
-
-  ## compute difference
-  sm_diff <- drop(X %*% coefs)
-  se <- sqrt(rowSums((X %*% V) * X))
-  nr <- NROW(X)
-
-  ## Calculate posterior simulation for smooths
-  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
-  rownames(coefs_sim) <- rownames(V)
-  simDev <- X %*% coefs_sim
-  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
-  masd <- apply(absDev, 2, max)
-  crit_s <- quantile(masd, prob = 0.95, type = 8)
-
-  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
-              level_1 = rep(f1, nr),
-              level_2 = rep(f2, nr),
-              diff = sm_diff, se = se,
-              lower_s = sm_diff - crit_s * se,
-              upper_s = sm_diff + crit_s*se)
-
-  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
-  ## Only need rows associated with one of the levels
-  out <- bind_cols(out, data[r1, smooth_var])
-
-  out
-}
-
-#does both ci and si
-difference_smooths <- function(model,
-                                 smooth,
-                                 n = 100,
-                                 ci_level = 0.95,
-                                 newdata = NULL,
-                                 partial_match = TRUE,
-                                 unconditional = FALSE,
-                                 frequentist = FALSE,
-                                 nrep = 10000,
-                                 include_means = TRUE,
-                                 ...) {
-  if (missing(smooth)) {
-    stop("Must specify a smooth to difference via 'smooth'.")
-  }
-
-  # smooths in model
-  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
-  # select smooths
-  select <-
-    gratia:::check_user_select_smooths(smooths = S, select = smooth,
-                                       partial_match = partial_match)#,
-  # model_name = expr_label(substitute(object)))
-  sm_ids <- which(select)
-  smooths <- gratia::get_smooths_by_id(model, sm_ids)
-  sm_data <- map(sm_ids, gratia:::smooth_data,
-                 model = model, n = n, include_all = TRUE)
-  sm_data <- bind_rows(sm_data)
-  by_var <- by_variable(smooths[[1L]])
-  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
-  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
-                                   stringsAsFactor = FALSE))
-  names(pairs) <- paste0("f", 1:2)
-
-  Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
-  V <- gratia:::get_vcov(model, unconditional = unconditional,
-                         frequentist = frequentist)
-  coefs <- coef(model)
-
-  out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
-              smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
-              coefs = coefs, nrep = nrep)
-  out <- bind_rows(out)
-  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
-
-  out <- add_column(out,
-                    lower = out$diff - (crit * out$se),
-                    upper = out$diff + (crit * out$se),
-                    .after = 6L)
-  out
-}
-
-#compute difference between smooths and calculate confidence interval: complete data
-diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
-                     unconditional = TRUE, frequentist = FALSE,
-                     n=100, partial_match = TRUE, nrep=10000)
-
-#function to obtain values for the shading regions
-pairwise_limits<-function(dataframe){
-    #extract values where the lower limit of the ribbon is greater than zero
-    #this is the region where the control group effect is greater
-
-    v1<-dataframe%>%
-        filter(lower_s>0)%>%
-        select(Day)
-    #get day  initial value
-    init1=v1$Day[[1]]
-    #get day final value
-    final1=v1$Day[[nrow(v1)]]
-    #extract values where the value of the upper limit of the ribbon is lower than zero
-    #this corresponds to the region where the treatment group effect is greater
-    v2<-dataframe%>%
-        filter(upper_s<0)%>%
-        select(Day)
-
-    init2=v2$Day[[1]]
-    final2=v2$Day[[nrow(v2)]]
-    #store values
-    my_list<-list(init1=init1,
-                  final1=final1,
-                  init2=init2,
-                  final2=final2)
-    return(my_list)
-
-}
-
-my_list<-pairwise_limits(diff_complete)
-rib_col<-'#8D7D82' #color for ribbon for confidence interval
-control_rib <- '#875F79' #color for ribbon for control region
-treat_rib <- '#A7D89E' #color for ribbon treatment region
-
-c1 <- ggplot() +
-  geom_line(data = diff_complete, aes(x = Day, y = diff),size=1, alpha=0.5) +
-    annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill=control_rib,
-                alpha = 0.5,
-                )+
-    annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=6,
-           angle=90
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill=treat_rib,
-             alpha = 0.5
-             ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=6,
-           angle=90
-           )+
-  geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s, ymax = upper_s),
-              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
-  geom_hline(yintercept = 0, lty = 2, color = "red")+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    labs(y="Difference\n(Complete observations)")+
-    theme_classic()+
-    theme(
-    axis.text=element_text(size=22))
-\end{lstlisting}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{SIM_Appendix_files/figure-latex/pairwise-comp-workflow-fig-1}
-
-}
-
-\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!gam\_02!} using a 95\% simultaneous CI for the difference between smooths.
-The comparison includes the group means and therefore can be directly correlated with the magnitude of the response.
-Shaded regions indicate time intervals where each treatment group has a non-zero effect.}\label{fig:pairwise-comp-workflow-fig}
-\end{figure}
-
-Of notice, the package \passthrough{\lstinline!gratia!} contains a function that allows to do pairwise comparisons.
-In this package, \passthrough{\lstinline!difference\_smooths!} is a function that makes the comparisons and produces Figure \ref{fig:pairwise-comp-workflow-fig} when is used on a fitted model.
-The function syntax and an example can be found at:
-
-\url{https://cran.r-project.org/web/packages/gratia/gratia.pdf}
-
-Keep in mind that this function removes the group means for the pairwise comparison so the result will not be directly interpretable with the scale of the measured response.
-
-\hypertarget{gam-and-linear-model-plots-and-missing-data}{%
-\subsection{GAM and Linear model plots and Missing data}\label{gam-and-linear-model-plots-and-missing-data}}
-
-This section covers the code used to generate Figure \ref{fig:sim-smooth-plot}, where the simulated data, fit of the ``final'' GAM (\passthrough{\lstinline!gam\_02!}), linear model and GAM on data with missing observations are presented.
-Note that panel A in Figure \ref{fig:sim-smooth-plot} and the inset are generated in the code chunk where the data is simulated in Section \ref{tumor-data-simulation}, and are called later to build the figure.
-
-\hypertarget{GAM-linear-plot}{%
-\subsubsection{GAM and Linear model plots}\label{GAM-linear-plot}}
-
-This code chunk creates panels B and C in Figure \ref{fig:sim-smooth-plot}.
-Note that this code uses the final GAM from the previous section (\passthrough{\lstinline!gam\_02!}), so the simulated data and the model should be generated before running this section.
-
-\begin{lstlisting}[language=R]
-#linear model
-lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)
-
-## point-wise interval
-ci <- confint(gam_02, parm = "s(Day)", partial_match = TRUE, type = "confidence")
-## simultaneous interval
-si <- confint(gam_02, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
-
-# mean shift for Treatment group
-const <- coef(gam_02)[2]
-
-#pointwise confidence interval
-ci <- ci %>%
-mutate(est = case_when(Group == "Treatment" ~ est + const,
-TRUE ~ est),
-lower = case_when(Group == "Treatment" ~ lower + const,
-TRUE ~ lower),
-upper = case_when(Group == "Treatment" ~ upper + const,
-TRUE ~ upper))
-
-#simultaneous interval
-si <- si %>%
-mutate(est = case_when(Group == "Treatment" ~ est + const,
-TRUE ~ est),
-lower = case_when(Group == "Treatment" ~ lower + const,
-TRUE ~ lower),
-upper = case_when(Group == "Treatment" ~ upper + const,
-TRUE ~ upper))
-
-#creates a dataframe using the length of the covariates for rm-ANOVA
-lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
-                         Day = c(0:10),
-                        subject=factor(rep(1:10)),
-                          )
-lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))
-
-#using lm
-lm_predict<-lm_predict%>%
-    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
-           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)
-
-#plot smooths and pointwise and simulatenous confidence intervals for GAM
-f3<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
-geom_line(lwd = 1) +
-geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
-inherit.aes = FALSE, alpha = 0.7,
-show.legend=FALSE) +
-geom_ribbon(data = si,
-mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
-inherit.aes = FALSE, alpha = 0.4,
-show.legend=FALSE)+
-    geom_point(data=dat_sim, aes(x = Day, 
-                        y = StO2_sim, 
-                        color = Group), 
-                        size=1.5,
-                        alpha=0.6, 
-               inherit.aes = FALSE,
-               show.legend = FALSE)+
-    labs(y=expression(atop(StO[2],'(complete observations)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+thm1
-
-#plot linear fit for rm-ANOVA
-f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
-    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
-  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
-                   ymax=(fit + 2*se.fit),fill=Group),
-              alpha=al,
-              data=lm_predict,
-              show.legend = FALSE,
-                inherit.aes=FALSE) +
-  geom_line(aes(y=fit,
-                color=Group),
-              size=1,data=lm_predict,
-              show.legend = FALSE)+
-  #facet_wrap(~Group)+
-  labs(y=expression(paste('StO'[2],' (simulated)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22)
-  )+thm1
-\end{lstlisting}
-
-\hypertarget{working-with-missing-data-in-gams}{%
-\subsection{Working with Missing data in GAMs}\label{working-with-missing-data-in-gams}}
-
-This code chunk first randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data.
-Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.
-
-\begin{lstlisting}[language=R]
-#missing data
-#create a sequence of 40 random numbers between 1 and 100, these numbers will
-#correspond to the row numbers to be randomly erased from the original dataset
-missing <- sample(1:100, 40)
-
-#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA
-
-ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))
-
-#create a new dataframe, remove the StO2 column
-dat_missing <- dat_sim[,-1]
-
-#add NAs at the ind positions
-dat_missing$StO2_sim[ind]<-NA
-
-#Count the number of remaining observations per day (original dataset had 10 per group per day)
-dat_missing %>%
-    group_by(Day,Group) %>%
-    filter(!is.na(StO2_sim))%>%
-  count(Day)
-
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
-#appraise the model
-appraise(mod_m1)
-
-ci <- confint(mod_m1, parm = "s(Day)", partial_match = TRUE, type = "confidence")
-## simultaneous interval
-si <- confint(mod_m1, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
-
-# mean shift for group 2
-const <- coef(mod_m1)[2]
-
-ci <- ci %>%
-mutate(est = case_when(Group == "Treatment" ~ est + const,
-TRUE ~ est),
-lower = case_when(Group == "Treatment" ~ lower + const,
-TRUE ~ lower),
-upper = case_when(Group == "Treatment" ~ upper + const,
-TRUE ~ upper))
-
-si <- si %>%
-mutate(est = case_when(Group == "Treatment" ~ est + const,
-TRUE ~ est),
-lower = case_when(Group == "Treatment" ~ lower + const,
-TRUE ~ lower),
-upper = case_when(Group == "Treatment" ~ upper + const,
-TRUE ~ upper))
-
-#plot model and original data
-f6<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
-geom_line(lwd = 1) +
-geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
-inherit.aes = FALSE, alpha = 0.7,
-show.legend=FALSE) +
-geom_ribbon(data = si,
-mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
-inherit.aes = FALSE, alpha = 0.4,
-show.legend=TRUE)+
-    geom_point(data=dat_missing, aes(x = Day, 
-                        y = StO2_sim, 
-                        color = Group), 
-                            size=1.5,
-                        alpha=0.6, 
-               inherit.aes = FALSE)+
-    labs(y=expression(atop(StO[2],'(missing observations)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22))+
-        thm1
-\end{lstlisting}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/sim-smooth-plot-Appendix-1}
-
-}
-
-\caption{Simulated data and smooths for oxygen saturation in tumors.
-\textbf{A}: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment.
-Simulated data is from a normal distribution with standard deviation of 10\% with 10 observations per time point.
-\textbf{B}: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment.
-Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.
-\textbf{C}: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time.
-\textbf{D}: Smooths for the GAM model for the simulated data with missing observations (40\%).
-Lines represent trends for each group, shaded regions are 95\% across the function (narrow region) and simultaneous (wide region) confidence intervals.}\label{fig:sim-smooth-plot-Appendix}
-\end{figure}
-
-\hypertarget{pairwise-comparisons-in-gams-full-and-missing-data-cases}{%
-\subsection{Pairwise comparisons in GAMs: full and missing data cases}\label{pairwise-comparisons-in-gams-full-and-missing-data-cases}}
-
-The next code chunk reproduces Figure \ref{fig:plot-pairwise-comp}.
-Here pairwise comparisons are made for the full and missing datasets by computing the difference between the smooths and a 95\% simultaneous CI.
-
-\begin{lstlisting}[language=R]
-##Pairwise comparisons
-
-difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
-  ## make sure f1 and f2 are characters
-  f1 <-  as.character(f1)
-  f2 <-  as.character(f2)
-  cnames <- colnames(Xp)
-  ## columns of Xp associated with pair of smooths
-  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
-  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
-  ## rows of Xp associated with pair of smooths
-  r1 <- data[[by_var]] == f1
-  r2 <- data[[by_var]] == f2
-
-  ## difference rows of Xp for pair of smooths
-  X <- Xp[r1, ] - Xp[r2, ]
-
-  ## zero the cols related to other splines
- # X[, ! (c1 | c2)] <- 0
-
-  ## zero out the parametric cols
-  #X[, !grepl('^s\\(', cnames)] <- 0
-
-  ## compute difference
-  sm_diff <- drop(X %*% coefs)
-  se <- sqrt(rowSums((X %*% V) * X))
-  nr <- NROW(X)
-
-  ## Calculate posterior simulation for smooths
-  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
-  rownames(coefs_sim) <- rownames(V)
-  simDev <- X %*% coefs_sim
-  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
-  masd <- apply(absDev, 2, max)
-  crit_s <- quantile(masd, prob = 0.95, type = 8)
-
-  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
-              level_1 = rep(f1, nr),
-              level_2 = rep(f2, nr),
-              diff = sm_diff, se = se,
-              lower_s = sm_diff - crit_s * se,
-              upper_s = sm_diff + crit_s*se)
-
-  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
-  ## Only need rows associated with one of the levels
-  out <- bind_cols(out, data[r1, smooth_var])
-
-  out
-}
-
-#does both ci and si
-difference_smooths <- function(model,
-                                 smooth,
-                                 n = 100,
-                                 ci_level = 0.95,
-                                 newdata = NULL,
-                                 partial_match = TRUE,
-                                 unconditional = FALSE,
-                                 frequentist = FALSE,
-                                 nrep = 10000,
-                                 include_means = TRUE,
-                                 ...) {
-  if (missing(smooth)) {
-    stop("Must specify a smooth to difference via 'smooth'.")
-  }
-
-  # smooths in model
-  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
-  # select smooths
-  select <-
-    gratia:::check_user_select_smooths(smooths = S, select = smooth,
-                                       partial_match = partial_match)#,
-  # model_name = expr_label(substitute(object)))
-  sm_ids <- which(select)
-  smooths <- gratia::get_smooths_by_id(model, sm_ids)
-  sm_data <- map(sm_ids, gratia:::smooth_data,
-                 model = model, n = n, include_all = TRUE)
-  sm_data <- bind_rows(sm_data)
-  by_var <- by_variable(smooths[[1L]])
-  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
-  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
-                                   stringsAsFactor = FALSE))
-  names(pairs) <- paste0("f", 1:2)
-
-  Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
-  V <- gratia:::get_vcov(model, unconditional = unconditional,
-                         frequentist = frequentist)
-  coefs <- coef(model)
-
-  out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
-              smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
-              coefs = coefs, nrep = nrep)
-  out <- bind_rows(out)
-  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
-
-  out <- add_column(out,
-                    lower = out$diff - (crit * out$se),
-                    upper = out$diff + (crit * out$se),
-                    .after = 6L)
-  out
-}
-
-#compute difference between smooths and calculate confidence interval: complete data
-diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
-                     unconditional = TRUE, frequentist = FALSE,
-                     n=100, partial_match = TRUE, nrep=10000)
-
-#function to obtain values for the shading regions
-pairwise_limits<-function(dataframe){
-    #extract values where the lower limit of the ribbon is greater than zero
-    #this is the region where the control group effect is greater
-
-    v1<-dataframe%>%
-        filter(lower_s>0)%>%
-        select(Day)
-    #get day  initial value
-    init1=v1$Day[[1]]
-    #get day final value
-    final1=v1$Day[[nrow(v1)]]
-    #extract values where the value of the upper limit of the ribbon is lower than zero
-    #this corresponds to the region where the treatment group effect is greater
-    v2<-dataframe%>%
-        filter(upper_s<0)%>%
-        select(Day)
-
-    init2=v2$Day[[1]]
-    final2=v2$Day[[nrow(v2)]]
-    #store values
-    my_list<-list(init1=init1,
-                  final1=final1,
-                  init2=init2,
-                  final2=final2)
-    return(my_list)
-
-}
-
-my_list<-pairwise_limits(diff_complete)
-rib_col<-'#8D7D82' #color for ribbon for confidence interval
-control_rib <- '#875F79' #color for ribbon for control region
-treat_rib <- '#A7D89E' #color for ribbon treatment region
-
-c1 <- ggplot() +
-  geom_line(data = diff_complete, aes(x = Day, y = diff),size=1, alpha=0.5) +
-    annotate("rect",
-                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-                fill=control_rib,
-                alpha = 0.5,
-                )+
-    annotate("text",
-             x=1.5,
-             y=-18,
-             label="Control>Treatment",
-           size=6,
-           angle=90
-           )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill=treat_rib,
-             alpha = 0.5
-             ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=6,
-           angle=90
-           )+
-  geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s, ymax = upper_s),
-              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
-  geom_hline(yintercept = 0, lty = 2, color = "red")+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    labs(y="Difference\n(Complete observations)")+
-    theme_classic()+
-    theme(
-    axis.text=element_text(size=22))
-
-###compute difference in smooths for missing data
-diff_missing <- difference_smooths(mod_m1, smooth = "s(Day)", newdata = newdat,
-                     unconditional = TRUE, frequentist = FALSE,
-                     n=100, partial_match = TRUE, nrep=10000)
-
-v2<- diff_missing %>%
-        filter(upper_s<0)%>%
-        select(Day)
-init2=v2$Day[[1]]
-final2=v2$Day[[nrow(v2)]]
-
-c2 <- ggplot() +
-  geom_line(data = diff_missing, aes(x = Day, y = diff),size=1, alpha=0.5) +
-    annotate("rect",
-             xmin =init2, xmax = final2,ymin=-Inf,ymax=Inf,
-             fill=treat_rib,
-             alpha = 0.5,
-    ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=6,
-           angle=90
-           )+
-  geom_ribbon(data = diff_missing, aes(x = Day, ymin = lower_s, ymax = upper_s),
-              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
-  geom_hline(yintercept = 0, lty = 2, color = "red")+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    labs(y="Difference\n(Missing observations)")+
-    theme_classic()+
-    theme(
-    axis.text=element_text(size=22))
-
-pair_comp<-c1+c2
-\end{lstlisting}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{SIM_Appendix_files/figure-latex/plot-pairwise-comp-Appendix-1}
-
-}
-
-\caption{Pairwise comparisons for smooth terms.
-\textbf{A}: Pairwise comparisons for the full dataset.
-\textbf{B}: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
-In both cases the effect of treatment is significant after day 3.
-In both cases, a simultaneous CI has been computed around the smooth difference.}\label{fig:plot-pairwise-comp-Appendix}
-\end{figure}
-
 \end{document}

### 2021-12-24 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/02-Challenges.Rmd
Working on Appendix B

@@
 Thus, the response per subject at each timepoint in both the linear and quadratic simulation corresponds to the mean response per group plus the error (represented by the points in Figure \@ref(fig:l-q-response) A, D).
-**Move simulated to the front: the simulated response per subject**
 
-A more comprehensive exploration  of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where a simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) is presented.
+A more comprehensive exploration  of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data can be obtained from the code that appears in Appendix B, (Figure A.1 and Figure A.2), where a simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) is presented.
 We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.

### 2021-12-24 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/03-GAM Theory.Rmd
Working on Appendix B

@@
 where $\beta_{0}$ is the intercept, and $\beta_{1}$ is the coefficient for each treatment group.
-Notice that the construction of the predictor is similar to that of Equation \@ref(eq:GLM-eta), but in this case the parametric terms involving the effect of time, and the interaction between time and treatment have been replaced by $f(time_{t}|\beta_{j})$.
+Notice that the construction of the predictor is similar to that of Equation \@ref(eq:GLM-eta), but in this case the parametric terms involving the effect of time, and the interaction between time and treatment have been replaced by the smooth term $f(time_{t}|\beta_{j})$.
 The smooth term $f(time_{t}|\beta_{j})$ gives a different smooth response for each treatment.
-If the smooth term represented a linear relationship, then $f(time_{t}|\beta_{j})= \beta_2 \times time_t+\beta_3 \times time_t \times treatment_j$; however, in general, the smooth term is a more flexible function than a linear relationship, with parameter vectors $\beta_{j}$ for each treatment.
+^[If the smooth term represented a linear relationship, then $f(time_{t}|\beta_{j})= \beta_2 \times time_t+\beta_3 \times time_t \times treatment_j$; however, in general, the smooth term is a more flexible function than a linear relationship, with parameter vectors $\beta_{j}$ for each treatment].
 A GAM version of a linear model can be written as
@@
 
-In contrast, thin plate regression splines (TPRS) provide a reasonable "low rank" (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner [@wood2003].
+In contrast, thin plate regression splines (TPRS) provide a reasonable "low rank" (truncated) approximation to the optimal TPS estimation, which can be implemented in an computationally efficient [@wood2003].
 Like TPS, TPRS only requires specifying the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood[@wood2003;@wood2017]).
 
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \@ref(fig:l-q-response)D.
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 that appears in Figure \@ref(fig:l-q-response)D.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
@@
 
-With this in mind, if five basis functions are used to fit a GAM for the data of Group 1 (concave down function) that appears in Figure \@ref(fig:l-q-response)D, the resulting fitting process is shown in Figure \@ref(fig:basis-plot)A, where the four basis functions (and the intercept) are shown.
+With this in mind, if four basis functions (plus the intercept) are used to fit a GAM for the data of Group 1 (concave down function) that appears in Figure \@ref(fig:l-q-response)D, the resulting fitting process is shown in Figure \@ref(fig:basis-plot).
+In Figure \@ref(fig:basis-plot)A the four basis functions (and the intercept) are shown.
 Each of the five basis functions is evaluated at six different points (because there are six points on the timeline).
 The coefficients for each of the basis functions of Figure \@ref(fig:basis-plot)A are estimated using a penalized regression with smoothness parameter $\lambda$, that is estimated when fitting the model.
-The penalized coefficient estimates fitted for our example are shown in Figure \@ref(fig:basis-plot)B.
+The penalized coefficient estimates fitted are shown in Figure \@ref(fig:basis-plot)B.
 
-To get the weighted basis functions, each basis (from Figure \@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients in Figure \@ref(fig:basis-plot)B, thereby increasing or decreasing the original basis functions.
-Figure \@ref(fig:basis-plot)C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
+To get the weighed basis functions, each basis (from Figure \@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients in Figure \@ref(fig:basis-plot)B, thereby increasing or decreasing the original basis functions.
+Figure \@ref(fig:basis-plot)C shows the resulting weighed basis functions.
+Note that the magnitude of the weighing for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
 On the other hand, the third basis function has roughly doubled its contribution to the smooth term.
 Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of _time_ is shown in Figure \@ref(fig:basis-plot)D (brown line), along the simulated values per group, which appear as points.
+The resulting smooth term for the effect of _time_ is shown in Figure \@ref(fig:basis-plot)D (brown line), along the simulated values which appear as points.
 
@@
 The coefficient determines the overall effect of each basis in the final smoother.
-**C**: Weighted basis functions.
-Each of the five basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B.
+**C**: Weighed basis functions.
+Each of the four basis functions (and the intercept) of panel A has been weighed by the corresponding coefficient shown in Panel B.
 Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
@@
 Briefly, to obtain a simultaneous CI we use simulation from the empirical Bayesian posterior distribution to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI [@ruppert2003].
-Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
 
 In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller [@miller2019], Wood[@wood2017], Simpson [@simpson2018], Marra [@marra2012], and Ruppert [@ruppert2003].
-With this brief introduction to the Bayesian interpretation of GAMs, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends.
+What we want to convey is that a Bayesian interpretation of GAMs allows for robust estimation using simultaneous empirical Bayesian CIs, as their stimates can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
+
+With this in mind, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends, and use simultaneous empirical Bayesian CIs to assess significance between treatment groups.

### 2021-12-24 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/04-Longitudinal analysis with GAMs.Rmd
Working on Appendix B

@@
 This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using $\textsf{R}$ and the package `mgcv`.
-The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.
+A brief guide for model selection and diagnostics appears in Appendix A, and the code for the simulated data and figures can be found in Appendix B.
 
@@
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot)B).
-Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_ [@gratia].
-A guide for model selection and diagnostics is in the [Appendix](#workflow), and an in-depth analysis can be found in Wood [@wood2017] and Harezlak [@harezlak2018].
+Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_ [@gratia] as we show in Appendix A.
+Additional discussions on model selection can be found in Wood [@wood2017] and Harezlak [@harezlak2018].
 
@@
 Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in `gam_02`) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in `gam_02`) is that the difference between them can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.
 
@@
 
-The absence of a _p-value_ in this context might seem odd, but its interpretation can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment.
-This is what happens for the simulated data in Figure \@ref(fig:sim-smooth-plot)A, where the chemotherapy causes $\mbox{StO}_2$ to increase over time.
+The absence of a _p-value_ in this context might seem odd, but its interpretation of the comparison can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \@ref(fig:sim-smooth-plot)A, where the chemotherapy causes $\mbox{StO}_2$ to increase over time.
 
-With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant.
+With this expectation of different trends in each group, computing the difference between the trends will identify if and _when_ the observed difference is significant.
 A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
+Therefore, the estimation of the difference between the smooths and the correspondent simultaneous empirical Bayesian CI of this difference provides an estimation of significance that is more robust that a mere _p-value_ because the estimated difference is able to indicates the significant time interval and we know that on average, the correction provided by the simultaneous empirical Bayesian CI ensures that it will contain _the whole function_ at a nominal value (say, 95%).
 
 Consider the calculation of pairwise differences for the smooths in Figure \@ref(fig:sim-smooth-plot)B, D.
-Figure \@ref(fig:plot-pairwise-comp) shows the comparison between each treatment group for the full and missing datasets with a simultaneous CI computed around the difference.
+Figure \@ref(fig:plot-pairwise-comp) shows the difference between each treatment group trend for the full and missing datasets with a simultaneous CI computed around the difference.
 Here, the  "Control" group is used as the reference to which "Treatment" group is being compared.
-Notice that because we have included the means of each group, there is correspondence between the scale of the original data the scale of the pairwise comparisons.
-This can be seen as in Figure \@ref(fig:sim-smooth-plot)B at day 5 there is essentially a 50% difference between $\mbox{StO}_2$ in both groups, which corresponds to the -50 difference in \@ref(fig:plot-pairwise-comp)A.
-However, if there are multiple parametric terms in the model (more factors that need to be specified in the model) such inclusion of the means can become problematic.
+Notice that because we have included the means of each group, there is correspondence between the scale of the original data and the scale of the pairwise comparisons.
+This can be seen in Figure \@ref(fig:sim-smooth-plot)B, where at day 5 there is essentially a difference of 50% between $\mbox{StO}_2$ in both groups, which corresponds to the -50% difference in \@ref(fig:plot-pairwise-comp)A.
+However, if there are multiple parametric terms in the model (more factors that need to be specified), such inclusion of the means can become problematic.
 However, we believe that the model we have presented here suffices in a wide range of situations where adding the group means is a relatively easy implementation that can help better visualize the estimation from the model from a biological perspective.
 
-In Figure \@ref(fig:plot-pairwise-comp)A the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and $\approx$ 2 for the full dataset indicates that through that time, the "Control" group has higher mean $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by $\approx$ 3 day  it is the "Treatment" group the one that on average, has greater $\mbox{StO}_2$.
+In Figure \@ref(fig:plot-pairwise-comp)A, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a different effect than the other.
+Notice that the shaded region between days 1 and $\approx$ 2 for the full dataset indicates that through that time, the "Control" group has higher mean $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by day $\approx$ 3  it is the "Treatment" group that statistically on average has greater $\mbox{StO}_2$.
 This would suggest that the effect of chemotherapy in the "Treatment" group becomes significant after day 3 for the given model.
@@
 
-On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that because the confidence intervals overlap for the first two days there is essentially no difference between the groups.
-However, because the model was still able to pick the overall trend in $\mbox{StO}_2$, the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes significant as in the full dataset smooth pairwise comparison.
+On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that because the confidence intervals overlap zero for the first two days there is statistically significant no difference between the groups.
+However, because the model is still able to pick the overall trend in $\mbox{StO}_2$, the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes statistically significant as in the full dataset smooth pairwise comparison.
 
-In a sense, the pairwise smooth comparison is more informative than a _post-hoc_ _p-value_.
-For biomedical studies, the smooth comparison is able to provide an estimate of _when_ and by _how much_ a biological process becomes significant.
-This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.
+For biomedical studies, the ability of smooth comparisons to provide an estimate of _when_ and by _how much_ a biological process becomes significant is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.

### 2021-12-24 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/05-Discussion.Rmd
Working on Appendix B

@@
 This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs.
-T
 
-The purpose of Section \@ref(GAM-theory) is to demonstrate that at its core the principle  underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship _a priori_ in order to provide it to the model, such as in the case of polynomial effects in LMEMs.
+However, in Section \@ref(GAM-theory) we demonstrated that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship _a priori_ in order to provide it to the model, such as in the case of polynomial effects in LMEMs where additionally there is no biological interpretation for such effect.
 In contrast, because GAMs let the data speak for themselves they can provide an accurate representation of the effect of time in a biological process.
@@
 However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature [@vishwanath2009] using R and the package _mgcv_[@wood2017] in Section \@ref(longitudinal-GAMs), while a basic workflow for model selection is in the [Appendix](#workflow).
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature [@vishwanath2009] using R and the package _mgcv_[@wood2017] in Section \@ref(longitudinal-GAMs), while a basic workflow for model selection is in Appendix A.
 
-One of the features of GAMs is that their Bayesian interpretation allows for inference about differences between groups without the need of a _p-value_, and in addition can provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure  \@ref(fig:plot-pairwise-comp) indicate.
+One of the features of GAMs is that their Bayesian interpretation allows for inference about differences between groups without the need of a _p-value_, thereby providing a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure  \@ref(fig:plot-pairwise-comp) indicate.
 The model is therefore able to identify changes between the groups at time points were data was not directly measured even in the case of incomplete observations ( $\approx$ day 3 in Figure \@ref(fig:plot-pairwise-comp) A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.

### 2021-12-24 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts/plot_example_Appendix.R

@@ -1,39 +1,73 @@
 
 ## This function plots the rm-ANOVA and LMEM for the data simulated in example.R
-plot_example <- function(sim_dat) {
+
+plot_example_Appendix <- function(sim_dat) {
     # Plot the simulated data (scatterplot)
+    txt<-20
     p1 <- sim_dat$dat %>%
         ggplot(aes(x = time,
                    y = y,
                    group = treatment,
                    color = treatment)
         ) +
-        geom_point(alpha=0.5,
-                   show.legend=FALSE) +
+        geom_point(show.legend=FALSE) +
         labs(y='response')+
         geom_line(aes(x = time,
                       y = mu,
                       color = treatment),
-                  size=3,
                   show.legend=FALSE) +
         theme_classic() +
-        theme(plot.title = element_text(size = 20,
+        theme(plot.title = element_text(size = txt,
                                         face = "bold"),
-              text=element_text(size=20))+
+              text=element_text(size=txt))+
         thm1
 
+    #plot the simulated data with trajectories per each subject
+    p2 <- sim_dat$dat %>%
+        ggplot(aes(x = time,
+                   y = y,
+                   group = subject,
+                   color = treatment)
+        ) +
+        geom_line(aes(size = "Subjects"),
+                  show.legend = FALSE) +
+        # facet_wrap(~ treatment) +
+        geom_line(aes(x = time,
+                      y = mu,
+                      color = treatment,
+                      size = "Simulated Truth"),
+                  lty = 1,show.legend = FALSE) +
+        labs(y='response')+
+        scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
+        theme_classic()+
+        theme(plot.title = element_text(size = txt,
+                                        face = "bold"),
+              text=element_text(size=txt))+
+        thm1
+
+    #plot the errors
+    p3 <- sim_dat$dat %>%
+        ggplot(aes(x = time,
+                   y = errors,
+                   group = subject,
+                   color = treatment)) +
+        geom_line(show.legend=FALSE) +
+        labs(y='errors')+
+        theme_classic()+
+        theme(plot.title = element_text(size = txt,
+                                        face = "bold"),
+              text=element_text(size=txt))+
+        thm1
 
     #plot the model predictions for rm-ANOVA
-    p2 <- ggplot(sim_dat$dat,
+    p4 <- ggplot(sim_dat$dat,
[…]

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/Full_document_SIM_No_Appendix.tex
Appendix B, functions and code

@@
 Thus, the response per subject at each timepoint in both the linear and quadratic simulation corresponds to the mean response per group plus the error (represented by the points in Figure \ref{fig:l-q-response} A, D).
-\textbf{Move simulated to the front: the simulated response per subject}
-
-A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data appears in the Appendix (Figure A.1 and Figure A.2), where a simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) is presented.
+
+A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data can be obtained from the code that appears in Appendix B, (Figures B.1 and B.2), where a simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) is presented.
 We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with ``default'' (linear) effects and, 2) demonstrate that a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_B.Rmd
Appendix B, functions and code

@@
     | 
-    | APPENDIX B: CODE AND FUNCTIONS 
+    | APPENDIX B: CODE 
 header-includes:
@@
 
-The first function is `example.R`, which allows to simulate linear and quadratic data in the same manner as in Section 3.5 in the main manuscript and estimates rm-ANOVA and LMEM with interaction to the data.
-The error for each simulated trend can be correlated or uncorrelated
+The first function is `example.R`, which allows to simulate linear and quadratic data in the same manner as in Section 3.5 in the main manuscript.
+Both rm-ANOVA and LMEM with interaction are fitted to the data.
+The error for each simulated trend can be correlated or uncorrelated as well.
 
@@
 Function `plot_example.R` uses the output of `example.R` to show the fit of a rm-ANOVA and a LMEM.
-It can be used to show an expanded version of Figure 1 in the main manuscript, presenting simulated data with correlated and uncorrelated errors and the correspoding rm-ANOVA and LMEM fits, which we do in the next subsection.
+This function can be used to show an expanded version of Figure 1 in the main manuscript, presenting simulated data with correlated and uncorrelated errors and how the individual trends vary in each case.
+The corresponding rm-ANOVA and LMEM fits are also presented, and we show the complete output in the next subsection.
 
@@
 plot_example <- function(sim_dat) {
-    # Plot the simulated data (scatterplot)
+    txt<-20
     p1 <- sim_dat$dat %>%
@@
         ) +
-        geom_point(alpha=0.5,
-                   show.legend=FALSE) +
+        geom_point(show.legend=FALSE) +
         labs(y='response')+
@@
                       color = treatment),
-                  size=3,
                   show.legend=FALSE) +
         theme_classic() +
-        theme(plot.title = element_text(size = 20,
+        theme(plot.title = element_text(size = txt,
                                         face = "bold"),
-              text=element_text(size=20))+
+              text=element_text(size=txt))+
+        thm1
+
+    #plot the simulated data with trajectories per each subject
+    p2 <- sim_dat$dat %>%
+        ggplot(aes(x = time,
+                   y = y,
+                   group = subject,
+                   color = treatment)
+        ) +
+        geom_line(aes(size = "Subjects"),
+                  show.legend = FALSE) +
+        # facet_wrap(~ treatment) +
+        geom_line(aes(x = time,
+                      y = mu,
+                      color = treatment,
+                      size = "Simulated Truth"),
+                  lty = 1,show.legend = FALSE) +
+        labs(y='response')+
+        scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
+        theme_classic()+
+        theme(plot.title = element_text(size = txt,
+                                        face = "bold"),
+              text=element_text(size=txt))+
+        thm1
+
+    #plot the errors
+    p3 <- sim_dat$dat %>%
+        ggplot(aes(x = time,
+                   y = errors,
+                   group = subject,
+                   color = treatment)) +
+        geom_line(show.legend=FALSE) +
+        labs(y='errors')+
+        theme_classic()+
+        theme(plot.title = element_text(size = txt,
+                                        face = "bold"),
+              text=element_text(size=txt))+
         thm1
@@
     #plot the model predictions for rm-ANOVA
-    p2 <- ggplot(sim_dat$dat,
+    p4 <- ggplot(sim_dat$dat,
                  aes(x = time,
@@
                      color = treatment)) +
-        geom_point(alpha=0.5,
-                   show.legend=FALSE)+
+        geom_point(show.legend=FALSE)+
         labs(y='response')+
         geom_line(aes(y = predict(sim_dat$fit_anova),
-                      group = subject, size = "Subjects"),
-                  show.legend = FALSE) +
+                      group = subject, size = "Subjects"),show.legend = FALSE) +
         geom_line(data = sim_dat$pred_dat,
@@
         theme_classic() +
-        theme(plot.title = element_text(size = 20,
+        theme(plot.title = element_text(size = txt,
                                         face = "bold"),
-              text=element_text(size=20))+
+              text=element_text(size=txt))+
         thm1
 
-    #plot the model predictions for LMEM
-    p4 <- ggplot(sim_dat$dat,
+    #plot the LMEM predictions
+    p5 <- ggplot(sim_dat$dat,
                  aes(x = time,
@@
                      color = treatment)) +
-        geom_point(alpha=0.5)+
+        geom_point()+
         labs(y='response')+
@@
         theme_classic() +
-        theme(plot.title = element_text(size = 20,
+        theme(plot.title = element_text(size = txt,
                                         face = "bold"),
-              text=element_text(size=20))+
+              text=element_text(size=txt))+
         thm1
 
+    if(option=='simple'){
+        return((p1+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
+    }
+    else {
     return((p1+p3+p2+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
+    }
 
@@
 
-In this subsection, we use `example.R` and a modified version `plot_example.R` (`plot_example_Appendix.R`) to create an expanded version of Figure 1 on the main manuscript.
-The only difference between `plot_example.R` and `plot_example_Appendix.R` is the inclusion of a `ggplot2` object (`p3`) that allows to plot the simulated errors.
+In this subsection, we use `example.R` and `plot_example.R` to create an expanded version of Figure 1 on the main manuscript.
+The difference here is that in the main manuscript `plot_example.R` uses `option='simple'` to create the plot, and therefore only shows the simulated data, and the rm-ANOVA and LMEM fit.
+Here, we use the option `composite` to generate the additional panels for the uncorrelated errors and their respective rm-ANOVA and LMEM fits.
 
@@
 source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","example.R"))
-source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","plot_example_Appendix.R"))
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","plot_example.R"))
 
-A1<-plot_example_Appendix(example(fun_type = "linear", error_type = "correlated"))
+A1<-plot_example(example(fun_type = "linear", error_type = "correlated"),option='composite')
 
-B1<-plot_example_Appendix(example(fun_type = "linear", error_type = "independent"))
+B1<-plot_example(example(fun_type = "linear", error_type = "independent"),option='composite')
 
-C1<-plot_example_Appendix(example(fun_type = "quadratic", error_type = "correlated"))
+C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"),option='composite')
 
-D1<-plot_example_Appendix(example(fun_type = "quadratic", error_type = "independent"))
+D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"),option='composite')
 
@@
 (ref:linear-cases-appendix) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
+**A, F**:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
+**B, G**: Generated errors showing the difference in the behavior of correlated and independent errors.
+**C, H**: Simulated data with thin lines representing individual trajectories.
+**D, I**: Estimations from the rm-ANOVA model for the mean group response.
+**E, J**: Estimations from the LMEM for the mean group response and individual responses (thin lines).
 In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
@@
 (ref:quadratic-caption-appendix) Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
+**A, F**:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
+**B, G**: Generated errors showing the difference in the behavior of correlated and independent errors.
+**C, H**: Simulated data with thin lines representing individual trajectories.
+**D, I**: Estimations from the rm-ANOVA model for the mean group response.
+**E, J**: Estimations from the LMEM for the mean group response and individual responses (thin lines).
 In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
@@
 ```
+
+## Basis Functions
+
+The next chunk presents the code used to create Figure 2 in the main manuscript.
+Here, we simulate data from the same concave down quadratic trend from Figure \@ref(fig:quadratic-cases-Appendix)A and use the package `mgcv` to fit a GAM to fit the trend of the data over time.
+The GAM in this case can be represented as
+
+ $$Response= \beta_0 + f(time_t|\beta_1)$$
+ Because we only use the data from one group and we do not account for interaction. The model in `mgcv` is written as:
+
+`gm<-gam( y~ s(time,k = 5),data = dat, method = "REML")`
+
+From the model, we extract the original and weighed four basis functions (and the intercept), and add these last weighed basis functions to construct the smoother.
+
+```{r,basis-function-Appendix, eval=FALSE}
+
+#basis functions: this script creates Fig 2. by calculating a GAM for the Group 1 data of Figure 1,
+# extracting the basis functions
+#and creates objects p11,p12,p13,p14 for plotting, which are combined in b_plot to create the
+#final composite figure
+n_time = 6
+x <- seq(1,6, length.out = n_time)
+mu <- matrix(0, length(x), 2)
+mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
+mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
+y <- array(0, dim = c(length(x), 2, 10))
+errors <- array(0, dim = c(length(x), 2, 10))
+for (i in 1:2) {     # number of treatments
+    for (j in 1:10) {  # number of subjects
+        # compound symmetry errors
+        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
+        y[, i, j] <- mu[, i] + errors[, i, j]
+    }
+}
+
+#label each table
+dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
+dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
+dimnames(mu) <- list(time = x, treatment = 1:2)
+
+#Convert to dataframes with subject, time and group columns
+dat <- as.data.frame.table(y, responseName = "y")
+dat_errors <- as.data.frame.table(errors, responseName = "errors")
+dat_mu <- as.data.frame.table(mu, responseName = "mu")
+dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
+dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
+dat$time <- as.numeric(as.character(dat$time))
+
+#label subject per group
+dat <- dat %>%
+    mutate(subject = factor(paste(subject, treatment, sep = "-")))
+
+#extract  "Group 1" to fit the GAM
+dat<-subset(dat,treatment==1)
+#keep just the response and timepoint columns
+dat<-dat[,c('y','time')]
+
+#GAM model of time, 5 basis functions
+gm<-gam(y~s(time,k=5),data=dat, method= "REML")
+
+#model_matrix (also known as) 'design matrix'
+#will contain the smooths used to create  model 'gm'
+model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
+
+time<-c(1:6)
+
+basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
+#basis<-model_matrix[1:6,-1] #extracting basis
+colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
+basis<-basis %>% #pivoting to long format
+    pivot_longer(
+        cols=starts_with("s")
+    )%>%
+    arrange(name) #ordering
+
+#length of dataframe to be created: number of basis by number of timepoints (minus 1 for the intercept that we won't plot)
+ln<-6*(length(coef(gm)))
+
+basis_plot<-data.frame(Basis=integer(ln),
+                       value_orig=double(ln),
+                       time=integer(ln),
+                       cof=double(ln)
+)
+
+basis_plot$time<-rep(time) #pasting timepoints
+basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
+basis_plot$value_orig<-basis$value #pasting basis values
+basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
+basis_plot<-basis_plot%>%
+    mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be
+#multiplied by the coefficients
+
+#creating labeller to change the labels in the basis plots
+
+basis_names<-c(
+    `1`="Intercept",
+    `2`="1",
+    `3`="2",
+    `4`="3",
+    `5`="4"
+)
+
+#calculating the final smooth by aggregating the basis functions
+
+smooth<-basis_plot%>%
+    group_by(time)%>%
+    summarize(smooth=sum(mod_val))
+
+#original basis
+sz<-1
+p11<-ggplot(basis_plot,
+            aes(x=time,
+                y=value_orig,
+                colour=as.factor(Basis)
+            )
+)+
+    geom_line(size=sz,
+              show.legend=FALSE)+
+    geom_point(size=sz+1,
+               show.legend = FALSE)+
+    labs(y='Basis functions')+
+    facet_wrap(~Basis,
+               labeller = as_labeller(basis_names)
+    )+
+    theme_classic()+
+    thm1
+
+#penalized basis
+p12<-ggplot(basis_plot,
+            aes(x=time,
+                y=mod_val,
+                colour=as.factor(Basis)
+            )
+)+
+    geom_line(show.legend = FALSE,
+              size=sz)+
+    geom_point(show.legend = FALSE,
+               size=sz+1)+
+    labs(y='Penalized \n basis functions')+
+    scale_y_continuous(breaks=seq(-1,1,1))+
+    facet_wrap(~Basis,
+               labeller=as_labeller(basis_names)
+    )+
+    theme_classic()+
+    thm1
+
+#heatmap of the  coefficients
+x_labels<-c("Intercept","1","2","3","4")
+p13<-ggplot(basis_plot,
+            aes(x=Basis,
+                y=Basis))+
+    geom_tile(aes(fill = cof),
+              colour = "black") +
+    scale_fill_gradient(low = "white",
+                        high = "#B50A2AFF")+
+    labs(x='Basis',
+         y='Basis')+
+    scale_x_discrete(labels=x_labels)+
+    geom_text(aes(label=round(cof,2)),
+              size=7,
+              show.legend = FALSE)+
+    theme_classic()+
+    theme(legend.title = element_blank())
+
+#plotting simulated datapoints and smooth term
+p14<-ggplot(data=dat,
+            aes(x=time,y=y))+
+    geom_point(size=sz+1,alpha=0.5)+
+    thm1+
+    labs(y='Simulated \n response')+
+    geom_line(data=smooth,
+              aes(x=time,
+                  y=smooth),
+              color="#6C581DFF",
+              size=sz+1)+
+    theme_classic()
+
+#Combining all
+b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
+    theme(
+        text=element_text(size=18)
+    )
+
+```
+
+## Function for data simulation
+
+The next chunk presents the code of the function `simulate_data.R`, which we use in the main manuscript to simulate data that follows the reported trends in $\mbox{StO}_2$ in the literature.
+Although the default standard deviation (SD) to generate the normally-distributed data in the function is 10%, we use a standard deviation of 10% in the main manuscript.
+We use the default number of 10 observations at each time point in the simulated data.
+Note that the input for the function is `dat`, needs to be a dataframe that contains the original values for the trends in each group over time from the literature.
+In our manuscript, we use the dataframe
+
+```{r,vishwanath, eval=FALSE}
+
+#data from Vishwanath et. al.
+dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
+            Day=rep(c(0,2,5,7,10),times=2),
+            Group=as.factor(rep(c("Control","Treatment"),each=5))
+            )
+```
+
+As the input for the function. Here, dataframe contains the trends in $\mbox{StO}_2$ for each group through the five time points and the labels for each treatment group (`Control` or `Treatment`).
+
+```{r,simulate-data-Appendix,eval=FALSE}
+#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
+#Because physiologically StO2 cannot go below 0%, data is
+#generated with a cutoff value of 0.0001 (the "StO2_sim")
+
+simulate_data <- function(dat, n = 10, sd = 5) {
+    dat_sim <- dat %>%
+        slice(rep(1:n(), each = n)) %>%
+        group_by(Group, Day) %>%
+        mutate(
+            StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
+            subject=rep(1:10),
+            subject=factor(paste(subject, Group, sep = "-"))
+        ) %>%
+        ungroup()
+
+    return(dat_sim)
+}
+
+```
+
+## Pointwise and simultaneous confidence intervals for smooths
+
+In this subsection we present the code used to create the pointwise and simultaneous empirical Bayesian confidence intervals (CIs) that are shown along the fitted smooths for the data in Figure 3 in the main manuscript.
+The computation is the same in the case of the complete and incomplete simulated data, and therefore we present here only the code for the incomplete data case.
+
+We do not use a custom function in this case but instead extract the pointwise and simultaneous CIs after fitting the GAM by using the function `confint` from base R.
+The boundaries for the pairwise and simultaneous CIs are obtained by using the type `"confindence"` or `"simultaneous"` in `confint` and are stored in `ci` and `si` respectively.
+
+Because we want to include the group means for plotting purposes, we extract the intercept (`const`) and add it to the Treatment group estimates in both `ci` and `si` order to shift the interval so it matches the scale of the response.
+Finally, we create a `ggplot2` object for plotting.
+
+```{r,pointwise-plot, eval=FALSE}
+#the same model used for the full dataset
+mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing)
+#appraise the model
+appraise(mod_m1)
+
+ci <- confint(mod_m1, parm = "s(Day)", partial_match = TRUE, type = "confidence")
+## simultaneous interval
+si <- confint(mod_m1, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
+
+# mean shift for group 2
+const <- coef(mod_m1)[2]
+
+ci <- ci %>%
+mutate(est = case_when(Group == "Treatment" ~ est + const,
+TRUE ~ est),
+lower = case_when(Group == "Treatment" ~ lower + const,
+TRUE ~ lower),
+upper = case_when(Group == "Treatment" ~ upper + const,
+TRUE ~ upper))
+
+si <- si %>%
+mutate(est = case_when(Group == "Treatment" ~ est + const,
+TRUE ~ est),
+lower = case_when(Group == "Treatment" ~ lower + const,
+TRUE ~ lower),
+upper = case_when(Group == "Treatment" ~ upper + const,
+TRUE ~ upper))
+
+f6<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
+geom_line(lwd = 1) +
+geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
+inherit.aes = FALSE, alpha = 0.7,
+show.legend=FALSE) +
+geom_ribbon(data = si,
+mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
+inherit.aes = FALSE, alpha = 0.4,
+show.legend=TRUE)+
+    geom_line(data=si,aes(Day,upper,color=Group), size=0.8, alpha=0.7)+
+    geom_line(data=si,aes(Day,lower,color=Group), size=0.8, alpha=0.7)+
+    geom_point(data=dat_missing, aes(x = Day, 
+                        y = StO2_sim, 
+                        color = Group), 
+                            size=1.5,
+                        alpha=0.6, 
+               inherit.aes = FALSE)+
+    labs(y=expression(atop(StO[2],'(incomplete observations)')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+  theme(
+    axis.text=element_text(size=22))+
+        thm1
+
+```
+
+## Pairwise comparisons
+
+This subsection presents the code used in functions `pointwise_comparisons.R` and `difference_smooths.R`, which generate the pointwise and simultaenous CI for the estimated difference between the smooths.
+In essence, `pointwise_comparisons.R` computes the difference between the smooths and the pointwise CI.
+Then, `difference_comparisons.R` uses the output of `pointwise_comparisons.R` to estimate the simultaneous CI.
+
+The chunk below presents the code for `pointwise_comparisons.R`.
+Note that as we indicate in Appendix A in this case we include the group means to estimate the confidence intervals so they can be directly compared with the original data.
+This works for the model we used in the paper, but if there are more parametric terms in the model the computation is not straightforward and will require a careful implementation because with an increase in parametric terms the number of pairwise comparisons also increases greatly.
+
+```{pointwise-appendix,eval=FALSE}
+
+##this function determines the pointwise confidence interval of a difference between two smooths
+
+difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
+    ## make sure f1 and f2 are characters
+    f1 <-  as.character(f1)
+    f2 <-  as.character(f2)
+    cnames <- colnames(Xp)
+    ## columns of Xp associated with pair of smooths
+    c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
+    c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
+    ## rows of Xp associated with pair of smooths
+    r1 <- data[[by_var]] == f1
+    r2 <- data[[by_var]] == f2
+
+    ## difference rows of Xp for pair of smooths
+    X <- Xp[r1, ] - Xp[r2, ]
+
+    ######IMPORTANT: uncommenting the following two lines
+    #removes the group means from the comparison######
+
+    ## zero the cols related to other splines
+    # X[, ! (c1 | c2)] <- 0
+
+    ## zero out the parametric cols
+    #X[, !grepl('^s\\(', cnames)] <- 0
+
+    ## compute difference
+    sm_diff <- drop(X %*% coefs)
+    se <- sqrt(rowSums((X %*% V) * X))
+    nr <- NROW(X)
+
+    ## Calculate posterior simulation for smooths
+    coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
+    rownames(coefs_sim) <- rownames(V)
+    simDev <- X %*% coefs_sim
+    absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
+    masd <- apply(absDev, 2, max)
+    crit_s <- quantile(masd, prob = 0.95, type = 8)
+
+    out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
+                level_1 = rep(f1, nr),
+                level_2 = rep(f2, nr),
+                diff = sm_diff, se = se,
+                lower_s = sm_diff - crit_s * se,
+                upper_s = sm_diff + crit_s*se)
+
+    out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
+    ## Only need rows associated with one of the levels
+    out <- bind_cols(out, data[r1, smooth_var])
+
+    out
+}
+```
+
+Then, `difference_smooths.R` takes the output from the previous function and uses it to compute the simultaneous CI.
+We present the code below.
+
+```{r, difference-smooths-Appendix, eval=FALSE}
+#this function calculates the pointwise (by calling pointwise_comparisons.R) CI and the simultaneous CI
+# for a pairwise comparison between two smooths
+difference_smooths <- function(model,
+                               smooth,
+                               n = 100,
+                               ci_level = 0.95,
+                               newdata = NULL,
+                               partial_match = TRUE,
+                               unconditional = FALSE,
+                               frequentist = FALSE,
+                               nrep = 10000,
+                               include_means = TRUE,
+                               ...) {
+    if (missing(smooth)) {
+        stop("Must specify a smooth to difference via 'smooth'.")
+    }
+
+    # smooths in model
+    S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
+    # select smooths
+    select <-
+        gratia:::check_user_select_smooths(smooths = S, select = smooth,
+                                           partial_match = partial_match)#,
+    # model_name = expr_label(substitute(object)))
+    sm_ids <- which(select)
+    smooths <- gratia::get_smooths_by_id(model, sm_ids)
+    sm_data <- map(sm_ids, gratia:::smooth_data,
+                   model = model, n = n, include_all = TRUE)
+    sm_data <- bind_rows(sm_data)
+    by_var <- by_variable(smooths[[1L]])
+    smooth_var <- gratia:::smooth_variable(smooths[[1L]])
+    pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
+                                     stringsAsFactor = FALSE))
+    names(pairs) <- paste0("f", 1:2)
+
+    Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
+    V <- gratia:::get_vcov(model, unconditional = unconditional,
+                           frequentist = frequentist)
+    coefs <- coef(model)
+
+    out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
+                smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
+                coefs = coefs, nrep = nrep)
+    out <- bind_rows(out)
+    crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
+
+    out <- add_column(out,
+                      lower = out$diff - (crit * out$se),
+                      upper = out$diff + (crit * out$se),
+                      .after = 6L)
+    out
+}
+
+```
+
+## Function for plotting limits of the pairwise comparisons
+
+The next function has the purpose of extracting the time intervals where the simultaneous CI does not cover zero from the object where the output of `difference_smooths.R` is stored in order to overlay two rectangles that help visualize the regions where each group is statistically significant.
+
+```{r,pairwise-limits-Appendix,eval=FALSE}
+
+#function to obtain values for the shading regions of the pairwise comparison between the smooths
+
+pairwise_limits<-function(dataframe){
+    #extract values where the lower limit of the ribbon is greater than zero
+    #this is the region where the control group effect is greater
+
+    v1<-dataframe%>%
+        filter(lower_s>0)%>%
+        select(Day)
+    #get day  initial value
+    init1=v1$Day[[1]]
+    #get day final value
+    final1=v1$Day[[nrow(v1)]]
+    #extract values where the value of the upper limit of the ribbon is lower than zero
+    #this corresponds to the region where the treatment group effect is greater
+    v2<-dataframe%>%
+        filter(upper_s<0)%>%
+        select(Day)
+
+    init2=v2$Day[[1]]
+    final2=v2$Day[[nrow(v2)]]
+    #store values
+    my_list<-list(init1=init1,
+                  final1=final1,
+                  init2=init2,
+                  final2=final2)
+    return(my_list)
+
+}
+
+```
+
+## GAM diagnostics function
+
+In Appendix A we discuss the use of quantitative and graphical diagnostics to assess the goodness of fit of a GAM.
+The package `mgcv` has the function `gam.check` to provide such information, but its graphical diagnostics are made using base R graphics and therefore it is not straightforward to use them in conjunction with a `ggplot2` object.
+Therefore we create the function `gam_diagnostics` that uses the same code from `gam.check` but without the graphical output.
+In this way, we can use `appraise` from the package `gratia` to create a graphical output in ggplot2 format and the quantitative information from `gam.check`.
+
+```{r,gam-diagnostics-Appendix, eval=FALSE}
+#Function that uses the source code of `gam.check` to obtain the estimates without the plots. The source can be checked by typing `gam.check` in the console.
+gam_diagnostics<-function (b, old.style = FALSE, type = c("deviance", "pearson",
+                                                          "response"), k.sample = 5000, k.rep = 200, rep = 0,
+                           level = 0.9, rl.col = 2, rep.col = "gray80", ...)
+{
+    type <- match.arg(type)
+    resid <- residuals(b, type = type)
+    linpred <- if (is.matrix(b$linear.predictors) && !is.matrix(resid))
+        napredict(b$na.action, b$linear.predictors[, 1])
+    else napredict(b$na.action, b$linear.predictors)
+
+    fv <- if (inherits(b$family, "extended.family"))
+        predict(b, type = "response")
+    else fitted(b)
+    if (is.matrix(fv) && !is.matrix(b$y))
+        fv <- fv[, 1]
+    gamm <- !(b$method %in% c("GCV", "GACV", "UBRE",
+                              "REML", "ML", "P-ML", "P-REML",
+                              "fREML"))
+    if (gamm) {
+        cat("\n'gamm' based fit - care required with interpretation.")
+        cat("\nChecks based on working residuals may be misleading.")
+    }
+    else {
+        cat("\nMethod:", b$method, "  Optimizer:",
+            b$optimizer)
+        if (!is.null(b$outer.info)) {
+            if (b$optimizer[2] %in% c("newton", "bfgs")) {
+                boi <- b$outer.info
+                cat("\n", boi$conv, " after ", boi$iter,
+                    " iteration", sep = "")
+                if (boi$iter == 1)
+                    cat(".")
+                else cat("s.")
+                cat("\nGradient range [", min(boi$grad),
+                    ",", max(boi$grad), "]", sep = "")
+                cat("\n(score ", b$gcv.ubre, " & scale ",
+                    b$sig2, ").", sep = "")
+                ev <- eigen(boi$hess)$values
+                if (min(ev) > 0)
+                    cat("\nHessian positive definite, ")
+                else cat("\n")
+                cat("eigenvalue range [", min(ev), ",",
+                    max(ev), "].\n", sep = "")
+            }
+            else {
+                cat("\n")
+                print(b$outer.info)
+            }
+        }
+        else {
+            if (length(b$sp) == 0)
+                cat("\nModel required no smoothing parameter selection")
+            else {
+                cat("\nSmoothing parameter selection converged after",
+                    b$mgcv.conv$iter, "iteration")
+                if (b$mgcv.conv$iter > 1)
+                    cat("s")
+                if (!b$mgcv.conv$fully.converged)
+                    cat(" by steepest\ndescent step failure.\n")
+                else cat(".\n")
+                cat("The RMS", b$method, "score gradient at convergence was",
+                    b$mgcv.conv$rms.grad, ".\n")
+                if (b$mgcv.conv$hess.pos.def)
+                    cat("The Hessian was positive definite.\n")
+                else cat("The Hessian was not positive definite.\n")
+            }
+        }
+        if (!is.null(b$rank)) {
+            cat("Model rank = ", b$rank, "/", length(b$coefficients),
+                "\n")
+        }
+    }
+    cat("\n")
+    kchck <- k.check(b, subsample = k.sample, n.rep = k.rep)
+    if (!is.null(kchck)) {
+        cat("Basis dimension (k) checking results. Low p-value (k-index<1) may\n")
+        cat("indicate that k is too low, especially if edf is close to k'.\n\n")
+        printCoefmat(kchck, digits = 3)
+    }
+}
+
+```

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_B.tex
Appendix B, functions and code

@@
 \strut \\
-APPENDIX B: CODE AND FUNCTIONS}
+APPENDIX B: CODE}
 \author{}
@@
 
-The first function is \passthrough{\lstinline!example.R!}, which allows to simulate linear and quadratic data in the same manner as in Section 3.5 in the main manuscript and estimates rm-ANOVA and LMEM with interaction to the data.
-The error for each simulated trend can be correlated or uncorrelated
+The first function is \passthrough{\lstinline!example.R!}, which allows to simulate linear and quadratic data in the same manner as in Section 3.5 in the main manuscript.
+Both rm-ANOVA and LMEM with interaction are fitted to the data.
+The error for each simulated trend can be correlated or uncorrelated as well.
 
@@
 Function \passthrough{\lstinline!plot\_example.R!} uses the output of \passthrough{\lstinline!example.R!} to show the fit of a rm-ANOVA and a LMEM.
-It can be used to show an expanded version of Figure 1 in the main manuscript, presenting simulated data with correlated and uncorrelated errors and the correspoding rm-ANOVA and LMEM fits, which we do in the next subsection.
+This function can be used to show an expanded version of Figure 1 in the main manuscript, presenting simulated data with correlated and uncorrelated errors and how the individual trends vary in each case.
+The corresponding rm-ANOVA and LMEM fits are also presented, and we show the complete output in the next subsection.
 
@@
 
-In this subsection, we use \passthrough{\lstinline!example.R!} and a modified version \passthrough{\lstinline!plot\_example.R!} (\passthrough{\lstinline!plot\_example\_Appendix.R!}) to create an expanded version of Figure 1 on the main manuscript.
-The only difference between \passthrough{\lstinline!plot\_example.R!} and \passthrough{\lstinline!plot\_example\_Appendix.R!} is the inclusion of a \passthrough{\lstinline!ggplot2!} object (\passthrough{\lstinline!p3!}) that allows to plot the simulated errors.
+In this subsection, we use \passthrough{\lstinline!example.R!} and \passthrough{\lstinline!plot\_example.R!} to create an expanded version of Figure 1 on the main manuscript.
+The difference here is that in the main manuscript \passthrough{\lstinline!plot\_example.R!} uses \passthrough{\lstinline!option='simple'!} to create the plot, and therefore only shows the simulated data, and the rm-ANOVA and LMEM fit.
+Here, we use the option \passthrough{\lstinline!composite!} to generate the additional panels for the uncorrelated errors and their respective rm-ANOVA and LMEM fits.
 
@@
 source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","example.R"))
-source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","plot_example_Appendix.R"))
+source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","plot_example.R"))
 
-A1<-plot_example_Appendix(example(fun_type = "linear", error_type = "correlated"))
+A1<-plot_example(example(fun_type = "linear", error_type = "correlated"),option='composite')
 
-B1<-plot_example_Appendix(example(fun_type = "linear", error_type = "independent"))
+B1<-plot_example(example(fun_type = "linear", error_type = "independent"),option='composite')
 
-C1<-plot_example_Appendix(example(fun_type = "quadratic", error_type = "correlated"))
+C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"),option='composite')
 
-D1<-plot_example_Appendix(example(fun_type = "quadratic", error_type = "independent")) 
+D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"),option='composite') 
 \end{lstlisting}
@@
 \caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
+\textbf{A, F}:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
+\textbf{B, G}: Generated errors showing the difference in the behavior of correlated and independent errors.
+\textbf{C, H}: Simulated data with thin lines representing individual trajectories.
+\textbf{D, I}: Estimations from the rm-ANOVA model for the mean group response.
+\textbf{E, J}: Estimations from the LMEM for the mean group response and individual responses (thin lines).
 In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
@@
 \caption{Simulated quadratic responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model and a LMEM.
-A, F:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
-B,G: Generated errors showing the difference in the behavior of correlated and independent errors.
-C,H: Simulated data with thin lines representing individual trajectories.
-D,I: Estimations from the rm-ANOVA model for the mean group response.
-E, J: Estimations from the LMEM for the mean group response and individual responses (thin lines).
+\textbf{A, F}:Simulated data with known mean response and individual responses (points) showing the dispersion of the data.
+\textbf{B, G}: Generated errors showing the difference in the behavior of correlated and independent errors.
+\textbf{C, H}: Simulated data with thin lines representing individual trajectories.
+\textbf{D, I}: Estimations from the rm-ANOVA model for the mean group response.
+\textbf{E, J}: Estimations from the LMEM for the mean group response and individual responses (thin lines).
 In all panels, thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.
@@
 
+\hypertarget{basis-functions}{%
+\subsection{Basis Functions}\label{basis-functions}}
+
+The next chunk presents the code used to create Figure 2 in the main manuscript.
+Here, we simulate data from the same concave down quadratic trend from Figure \ref{fig:quadratic-cases-Appendix}A and use the package \passthrough{\lstinline!mgcv!} to fit a GAM to fit the trend of the data over time.
+The GAM in this case can be represented as
+
+\[Response= \beta_0 + f(time_t|\beta_1)\]
+Because we only use the data from one group and we do not account for interaction. The model in \passthrough{\lstinline!mgcv!} is written as:
+
+\passthrough{\lstinline!gm<-gam( y\~ s(time,k = 5),data = dat, method = "REML")!}
+
+From the model, we extract the original and weighed four basis functions (and the intercept), and add these last weighed basis functions to construct the smoother.
+
+\begin{lstlisting}[language=R]
+#basis functions: this script creates Fig 2. by calculating a GAM for the Group 1 data of Figure 1,
+# extracting the basis functions
+#and creates objects p11,p12,p13,p14 for plotting, which are combined in b_plot to create the
+#final composite figure
+n_time = 6
+x <- seq(1,6, length.out = n_time)
+mu <- matrix(0, length(x), 2)
+mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
+mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
+y <- array(0, dim = c(length(x), 2, 10))
+errors <- array(0, dim = c(length(x), 2, 10))
+for (i in 1:2) {     # number of treatments
+    for (j in 1:10) {  # number of subjects
+        # compound symmetry errors
+        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
+        y[, i, j] <- mu[, i] + errors[, i, j]
+    }
+}
+
+#label each table
+dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
+dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
+dimnames(mu) <- list(time = x, treatment = 1:2)
+
+#Convert to dataframes with subject, time and group columns
+dat <- as.data.frame.table(y, responseName = "y")
+dat_errors <- as.data.frame.table(errors, responseName = "errors")
+dat_mu <- as.data.frame.table(mu, responseName = "mu")
+dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
+dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
+dat$time <- as.numeric(as.character(dat$time))
+
+#label subject per group
+dat <- dat %>%
+    mutate(subject = factor(paste(subject, treatment, sep = "-")))
+
+#extract  "Group 1" to fit the GAM
+dat<-subset(dat,treatment==1)
+#keep just the response and timepoint columns
+dat<-dat[,c('y','time')]
+
+#GAM model of time, 5 basis functions
+gm<-gam(y~s(time,k=5),data=dat, method= "REML")
+
+#model_matrix (also known as) 'design matrix'
+#will contain the smooths used to create  model 'gm'
+model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
+
+time<-c(1:6)
+
+basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
+#basis<-model_matrix[1:6,-1] #extracting basis
+colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
+basis<-basis %>% #pivoting to long format
+    pivot_longer(
+        cols=starts_with("s")
+    )%>%
+    arrange(name) #ordering
+
+#length of dataframe to be created: number of basis by number of timepoints (minus 1 for the intercept that we won't plot)
+ln<-6*(length(coef(gm)))
+
+basis_plot<-data.frame(Basis=integer(ln),
+                       value_orig=double(ln),
+                       time=integer(ln),
+                       cof=double(ln)
+)
+
+basis_plot$time<-rep(time) #pasting timepoints
+basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
+basis_plot$value_orig<-basis$value #pasting basis values
+basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
+basis_plot<-basis_plot%>%
+    mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be
+#multiplied by the coefficients
+
+#creating labeller to change the labels in the basis plots
+
+basis_names<-c(
+    `1`="Intercept",
+    `2`="1",
+    `3`="2",
+    `4`="3",
+    `5`="4"
+)
+
+#calculating the final smooth by aggregating the basis functions
+
+smooth<-basis_plot%>%
+    group_by(time)%>%
+    summarize(smooth=sum(mod_val))
+
+#original basis
+sz<-1
+p11<-ggplot(basis_plot,
+            aes(x=time,
+                y=value_orig,
+                colour=as.factor(Basis)
+            )
+)+
+    geom_line(size=sz,
+              show.legend=FALSE)+
+    geom_point(size=sz+1,
+               show.legend = FALSE)+
+    labs(y='Basis functions')+
+    facet_wrap(~Basis,
+               labeller = as_labeller(basis_names)
+    )+
+    theme_classic()+
+    thm1
+
+#penalized basis
+p12<-ggplot(basis_plot,
+            aes(x=time,
+                y=mod_val,
+                colour=as.factor(Basis)
+            )
+)+
+    geom_line(show.legend = FALSE,
+              size=sz)+
+    geom_point(show.legend = FALSE,
+               size=sz+1)+
+    labs(y='Penalized \n basis functions')+
+    scale_y_continuous(breaks=seq(-1,1,1))+
+    facet_wrap(~Basis,
+               labeller=as_labeller(basis_names)
+    )+
+    theme_classic()+
+    thm1
+
+#heatmap of the  coefficients
+x_labels<-c("Intercept","1","2","3","4")
+p13<-ggplot(basis_plot,
+            aes(x=Basis,
+                y=Basis))+
+    geom_tile(aes(fill = cof),
+              colour = "black") +
+    scale_fill_gradient(low = "white",
+                        high = "#B50A2AFF")+
+    labs(x='Basis',
+         y='Basis')+
+    scale_x_discrete(labels=x_labels)+
+    geom_text(aes(label=round(cof,2)),
+              size=7,
+              show.legend = FALSE)+
+    theme_classic()+
+    theme(legend.title = element_blank())
+
+#plotting simulated datapoints and smooth term
+p14<-ggplot(data=dat,
+            aes(x=time,y=y))+
+    geom_point(size=sz+1,alpha=0.5)+
+    thm1+
+    labs(y='Simulated \n response')+
+    geom_line(data=smooth,
+              aes(x=time,
+                  y=smooth),
+              color="#6C581DFF",
+              size=sz+1)+
+    theme_classic()
+
+#Combining all
+b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
+    theme(
+        text=element_text(size=18)
+    )
+\end{lstlisting}
+
+\hypertarget{function-for-data-simulation}{%
+\subsection{Function for data simulation}\label{function-for-data-simulation}}
+
+The next chunk presents the code of the function \passthrough{\lstinline!simulate\_data.R!}, which we use in the main manuscript to simulate data that follows the reported trends in \(\mbox{StO}_2\) in the literature.
+Although the default standard deviation (SD) to generate the normally-distributed data in the function is 10\%, we use a standard deviation of 10\% in the main manuscript.
+We use the default number of 10 observations at each time point in the simulated data.
+Note that the input for the function is \passthrough{\lstinline!dat!}, needs to be a dataframe that contains the original values for the trends in each group over time from the literature.
+In our manuscript, we use the dataframe
+
+\begin{lstlisting}[language=R]
+#data from Vishwanath et. al.
+dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
+            Day=rep(c(0,2,5,7,10),times=2),
+            Group=as.factor(rep(c("Control","Treatment"),each=5))
+            )
+\end{lstlisting}
+
+As the input for the function.
+Here, dataframe contains the trends in \(\mbox{StO}_2\) for each group through the five time points and the labels for each treatment group (\passthrough{\lstinline!Control!} or \passthrough{\lstinline!Treatment!}).
+
+\begin{lstlisting}[language=R]
+#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
+#Because physiologically StO2 cannot go below 0%, data is
+#generated with a cutoff value of 0.0001 (the "StO2_sim")
+
+simulate_data <- function(dat, n = 10, sd = 5) {
+    dat_sim <- dat %>%
+        slice(rep(1:n(), each = n)) %>%
+        group_by(Group, Day) %>%
+        mutate(
+            StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
+            subject=rep(1:10),
+            subject=factor(paste(subject, Group, sep = "-"))
+        ) %>%
+        ungroup()
+
+    return(dat_sim)
+}
+\end{lstlisting}
+
+\hypertarget{pointwise-and-simultaneous-confidence-intervals-for-smooths}{%
+\subsection{Pointwise and simultaneous confidence intervals for smooths}\label{pointwise-and-simultaneous-confidence-intervals-for-smooths}}
+
+In this subsection we present the code used to create the pointwise and simultaneous empirical Bayesian confidence intervals (CIs) that are shown along the fitted smooths for the data in Figure 3 in the main manuscript.
+The computation is the same in the case of the complete and incomplete simulated data, and therefore we present here only the code for the incomplete data case.
+
+We do not use a custom function in this case but instead extract the pointwise and simultaneous CIs after fitting the GAM by using the function \passthrough{\lstinline!confint!} from base R.
+The boundaries for the pairwise and simultaneous CIs are obtained by using the type \passthrough{\lstinline!"confindence"!} or \passthrough{\lstinline!"simultaneous"!} in \passthrough{\lstinline!confint!} and are stored in \passthrough{\lstinline!ci!} and \passthrough{\lstinline!si!} respectively.
+
+Because we want to include the group means for plotting purposes, we extract the intercept (\passthrough{\lstinline!const!}) and add it to the Treatment group estimates in both \passthrough{\lstinline!ci!} and \passthrough{\lstinline!si!} order to shift the interval so it matches the scale of the response.
+Finally, we create a \passthrough{\lstinline!ggplot2!} object for plotting.
+
+\begin{lstlisting}[language=R]
+#the same model used for the full dataset
+mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing)
+#appraise the model
+appraise(mod_m1)
+
+ci <- confint(mod_m1, parm = "s(Day)", partial_match = TRUE, type = "confidence")
+## simultaneous interval
+si <- confint(mod_m1, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
+
+# mean shift for group 2
+const <- coef(mod_m1)[2]
+
+ci <- ci %>%
+mutate(est = case_when(Group == "Treatment" ~ est + const,
+TRUE ~ est),
+lower = case_when(Group == "Treatment" ~ lower + const,
+TRUE ~ lower),
+upper = case_when(Group == "Treatment" ~ upper + const,
+TRUE ~ upper))
+
+si <- si %>%
+mutate(est = case_when(Group == "Treatment" ~ est + const,
+TRUE ~ est),
+lower = case_when(Group == "Treatment" ~ lower + const,
+TRUE ~ lower),
+upper = case_when(Group == "Treatment" ~ upper + const,
+TRUE ~ upper))
+
+f6<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
+geom_line(lwd = 1) +
+geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
+inherit.aes = FALSE, alpha = 0.7,
+show.legend=FALSE) +
+geom_ribbon(data = si,
+mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
+inherit.aes = FALSE, alpha = 0.4,
+show.legend=TRUE)+
+    geom_line(data=si,aes(Day,upper,color=Group), size=0.8, alpha=0.7)+
+    geom_line(data=si,aes(Day,lower,color=Group), size=0.8, alpha=0.7)+
+    geom_point(data=dat_missing, aes(x = Day, 
+                        y = StO2_sim, 
+                        color = Group), 
+                            size=1.5,
+                        alpha=0.6, 
+               inherit.aes = FALSE)+
+    labs(y=expression(atop(StO[2],'(incomplete observations)')))+
+    scale_x_continuous(breaks=c(0,2,5,7,10))+
+      theme_classic()+
+  theme(
+    axis.text=element_text(size=22))+
+        thm1
+\end{lstlisting}
+
+\hypertarget{pairwise-comparisons}{%
+\subsection{Pairwise comparisons}\label{pairwise-comparisons}}
+
+This subsection presents the code used in functions \passthrough{\lstinline!pointwise\_comparisons.R!} and \passthrough{\lstinline!difference\_smooths.R!}, which generate the pointwise and simultaenous CI for the estimated difference between the smooths.
+In essence, \passthrough{\lstinline!pointwise\_comparisons.R!} computes the difference between the smooths and the pointwise CI.
+Then, \passthrough{\lstinline!difference\_comparisons.R!} uses the output of \passthrough{\lstinline!pointwise\_comparisons.R!} to estimate the simultaneous CI.
+
+The chunk below presents the code for \passthrough{\lstinline!pointwise\_comparisons.R!}.
+Note that as we indicate in Appendix A in this case we include the group means to estimate the confidence intervals so they can be directly compared with the original data.
+This works for the model we used in the paper, but if there are more parametric terms in the model the computation is not straightforward and will require a careful implementation because with an increase in parametric terms the number of pairwise comparisons also increases greatly.
+
+\begin{lstlisting}
+##this function determines the pointwise confidence interval of a difference between two smooths
+
+difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
+    ## make sure f1 and f2 are characters
+    f1 <-  as.character(f1)
+    f2 <-  as.character(f2)
+    cnames <- colnames(Xp)
+    ## columns of Xp associated with pair of smooths
+    c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
+    c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
+    ## rows of Xp associated with pair of smooths
+    r1 <- data[[by_var]] == f1
+    r2 <- data[[by_var]] == f2
+
+    ## difference rows of Xp for pair of smooths
+    X <- Xp[r1, ] - Xp[r2, ]
+
+    ######IMPORTANT: uncommenting the following two lines
+    #removes the group means from the comparison######
+
+    ## zero the cols related to other splines
+    # X[, ! (c1 | c2)] <- 0
+
+    ## zero out the parametric cols
+    #X[, !grepl('^s\\(', cnames)] <- 0
+
+    ## compute difference
+    sm_diff <- drop(X %*% coefs)
+    se <- sqrt(rowSums((X %*% V) * X))
+    nr <- NROW(X)
+
+    ## Calculate posterior simulation for smooths
+    coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
+    rownames(coefs_sim) <- rownames(V)
+    simDev <- X %*% coefs_sim
+    absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
+    masd <- apply(absDev, 2, max)
+    crit_s <- quantile(masd, prob = 0.95, type = 8)
+
+    out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
+                level_1 = rep(f1, nr),
+                level_2 = rep(f2, nr),
+                diff = sm_diff, se = se,
+                lower_s = sm_diff - crit_s * se,
+                upper_s = sm_diff + crit_s*se)
+
+    out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
+    ## Only need rows associated with one of the levels
+    out <- bind_cols(out, data[r1, smooth_var])
+
+    out
+}
+\end{lstlisting}
+
+Then, \passthrough{\lstinline!difference\_smooths.R!} takes the output from the previous function and uses it to compute the simultaneous CI.
+We present the code below.
+
+\begin{lstlisting}[language=R]
+#this function calculates the pointwise (by calling pointwise_comparisons.R) CI and the simultaneous CI
+# for a pairwise comparison between two smooths
+difference_smooths <- function(model,
+                               smooth,
+                               n = 100,
+                               ci_level = 0.95,
+                               newdata = NULL,
+                               partial_match = TRUE,
+                               unconditional = FALSE,
+                               frequentist = FALSE,
+                               nrep = 10000,
+                               include_means = TRUE,
+                               ...) {
+    if (missing(smooth)) {
+        stop("Must specify a smooth to difference via 'smooth'.")
+    }
+
+    # smooths in model
+    S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
+    # select smooths
+    select <-
+        gratia:::check_user_select_smooths(smooths = S, select = smooth,
+                                           partial_match = partial_match)#,
+    # model_name = expr_label(substitute(object)))
+    sm_ids <- which(select)
+    smooths <- gratia::get_smooths_by_id(model, sm_ids)
+    sm_data <- map(sm_ids, gratia:::smooth_data,
+                   model = model, n = n, include_all = TRUE)
+    sm_data <- bind_rows(sm_data)
+    by_var <- by_variable(smooths[[1L]])
+    smooth_var <- gratia:::smooth_variable(smooths[[1L]])
+    pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
+                                     stringsAsFactor = FALSE))
+    names(pairs) <- paste0("f", 1:2)
+
+    Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
+    V <- gratia:::get_vcov(model, unconditional = unconditional,
+                           frequentist = frequentist)
+    coefs <- coef(model)
+
+    out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
+                smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
+                coefs = coefs, nrep = nrep)
+    out <- bind_rows(out)
+    crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
+
+    out <- add_column(out,
+                      lower = out$diff - (crit * out$se),
+                      upper = out$diff + (crit * out$se),
+                      .after = 6L)
+    out
+}
+\end{lstlisting}
+
+\hypertarget{function-for-plotting-limits-of-the-pairwise-comparisons}{%
+\subsection{Function for plotting limits of the pairwise comparisons}\label{function-for-plotting-limits-of-the-pairwise-comparisons}}
+
+The next function has the purpose of extracting the time intervals where the simultaneous CI does not cover zero from the object where the output of \passthrough{\lstinline!difference\_smooths.R!} is stored in order to overlay two rectangles that help visualize the regions where each group is statistically significant.
+
+\begin{lstlisting}[language=R]
+#function to obtain values for the shading regions of the pairwise comparison between the smooths
+
+pairwise_limits<-function(dataframe){
+    #extract values where the lower limit of the ribbon is greater than zero
+    #this is the region where the control group effect is greater
+
+    v1<-dataframe%>%
+        filter(lower_s>0)%>%
+        select(Day)
+    #get day  initial value
+    init1=v1$Day[[1]]
+    #get day final value
+    final1=v1$Day[[nrow(v1)]]
+    #extract values where the value of the upper limit of the ribbon is lower than zero
+    #this corresponds to the region where the treatment group effect is greater
+    v2<-dataframe%>%
+        filter(upper_s<0)%>%
+        select(Day)
+
+    init2=v2$Day[[1]]
+    final2=v2$Day[[nrow(v2)]]
+    #store values
+    my_list<-list(init1=init1,
+                  final1=final1,
+                  init2=init2,
+                  final2=final2)
+    return(my_list)
+
+}
+\end{lstlisting}
+
+\hypertarget{gam-diagnostics-function}{%
+\subsection{GAM diagnostics function}\label{gam-diagnostics-function}}
+
+In Appendix A we discuss the use of quantitative and graphical diagnostics to assess the goodness of fit of a GAM.
+The package \passthrough{\lstinline!mgcv!} has the function \passthrough{\lstinline!gam.check!} to provide such information, but its graphical diagnostics are made using base R graphics and therefore it is not straightforward to use them in conjunction with a \passthrough{\lstinline!ggplot2!} object.
+Therefore we create the function \passthrough{\lstinline!gam\_diagnostics!} that uses the same code from \passthrough{\lstinline!gam.check!} but without the graphical output.
+In this way, we can use \passthrough{\lstinline!appraise!} from the package \passthrough{\lstinline!gratia!} to create a graphical output in ggplot2 format and the quantitative information from \passthrough{\lstinline!gam.check!}.
+
+\begin{lstlisting}[language=R]
+#Function that uses the source code of `gam.check` to obtain the estimates without the plots. The source can be checked by typing `gam.check` in the console.
+gam_diagnostics<-function (b, old.style = FALSE, type = c("deviance", "pearson",
+                                                          "response"), k.sample = 5000, k.rep = 200, rep = 0,
+                           level = 0.9, rl.col = 2, rep.col = "gray80", ...)
+{
+    type <- match.arg(type)
+    resid <- residuals(b, type = type)
+    linpred <- if (is.matrix(b$linear.predictors) && !is.matrix(resid))
+        napredict(b$na.action, b$linear.predictors[, 1])
+    else napredict(b$na.action, b$linear.predictors)
+
+    fv <- if (inherits(b$family, "extended.family"))
+        predict(b, type = "response")
+    else fitted(b)
+    if (is.matrix(fv) && !is.matrix(b$y))
+        fv <- fv[, 1]
+    gamm <- !(b$method %in% c("GCV", "GACV", "UBRE",
+                              "REML", "ML", "P-ML", "P-REML",
+                              "fREML"))
+    if (gamm) {
+        cat("\n'gamm' based fit - care required with interpretation.")
+        cat("\nChecks based on working residuals may be misleading.")
+    }
+    else {
+        cat("\nMethod:", b$method, "  Optimizer:",
+            b$optimizer)
+        if (!is.null(b$outer.info)) {
+            if (b$optimizer[2] %in% c("newton", "bfgs")) {
+                boi <- b$outer.info
+                cat("\n", boi$conv, " after ", boi$iter,
+                    " iteration", sep = "")
+                if (boi$iter == 1)
+                    cat(".")
+                else cat("s.")
+                cat("\nGradient range [", min(boi$grad),
+                    ",", max(boi$grad), "]", sep = "")
+                cat("\n(score ", b$gcv.ubre, " & scale ",
+                    b$sig2, ").", sep = "")
+                ev <- eigen(boi$hess)$values
+                if (min(ev) > 0)
+                    cat("\nHessian positive definite, ")
+                else cat("\n")
+                cat("eigenvalue range [", min(ev), ",",
+                    max(ev), "].\n", sep = "")
+            }
+            else {
+                cat("\n")
+                print(b$outer.info)
+            }
+        }
+        else {
+            if (length(b$sp) == 0)
+                cat("\nModel required no smoothing parameter selection")
+            else {
+                cat("\nSmoothing parameter selection converged after",
+                    b$mgcv.conv$iter, "iteration")
+                if (b$mgcv.conv$iter > 1)
+                    cat("s")
+                if (!b$mgcv.conv$fully.converged)
+                    cat(" by steepest\ndescent step failure.\n")
+                else cat(".\n")
+                cat("The RMS", b$method, "score gradient at convergence was",
+                    b$mgcv.conv$rms.grad, ".\n")
+                if (b$mgcv.conv$hess.pos.def)
+                    cat("The Hessian was positive definite.\n")
+                else cat("The Hessian was not positive definite.\n")
+            }
+        }
+        if (!is.null(b$rank)) {
+            cat("Model rank = ", b$rank, "/", length(b$coefficients),
+                "\n")
+        }
+    }
+    cat("\n")
+    kchck <- k.check(b, subsample = k.sample, n.rep = k.rep)
+    if (!is.null(kchck)) {
+        cat("Basis dimension (k) checking results. Low p-value (k-index<1) may\n")
+        cat("indicate that k is too low, especially if edf is close to k'.\n\n")
+        printCoefmat(kchck, digits = 3)
+    }
+}
+\end{lstlisting}
+
 \end{document}

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/02-Challenges.Rmd
Appendix B, functions and code

@@
 
-A more comprehensive exploration  of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data can be obtained from the code that appears in Appendix B, (Figure A.1 and Figure A.2), where a simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) is presented.
+A more comprehensive exploration  of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal data can be obtained from the code that appears in Appendix B, (Figures B.1 and B.2), where a simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) is presented.
 We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.
@@
 
-A<-plot_example(example(fun_type = "linear", error_type = "correlated"))
+A<-plot_example(example(fun_type = "linear", error_type = "correlated"),option='simple')
 
-C<-plot_example(example(fun_type = "quadratic", error_type = "correlated"))
+C<-plot_example(example(fun_type = "quadratic", error_type = "correlated"), option='simple')
 

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/04-Longitudinal analysis with GAMs.Rmd
Appendix B, functions and code

@@
 #the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
+mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing)
 #appraise the model

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts/example.R

@@ -2,7 +2,7 @@
 #responses with correlated or uncorrelated errors and fits a linear model to the data.
 
 example <- function(n_time = 6, #number of time points
-                    fun_type = "linear", #type of reponse
+                    fun_type = "linear", #type of response
                     error_type = "correlated") {
 
     if (!(fun_type %in% c("linear", "quadratic")))

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts/plot_example.R

@@ -1,39 +1,74 @@
 
 ## This function plots the rm-ANOVA and LMEM for the data simulated in example.R
-plot_example <- function(sim_dat) {
-    # Plot the simulated data (scatterplot)
+plot_example <- function(sim_dat,
+                         option="simple") {
+    txt<-20
+
+    #plot simulated data (scatterplot)
     p1 <- sim_dat$dat %>%
         ggplot(aes(x = time,
                    y = y,
                    group = treatment,
                    color = treatment)
         ) +
-        geom_point(alpha=0.5,
-                   show.legend=FALSE) +
+        geom_point(show.legend=FALSE) +
         labs(y='response')+
         geom_line(aes(x = time,
                       y = mu,
                       color = treatment),
-                  size=3,
                   show.legend=FALSE) +
         theme_classic() +
-        theme(plot.title = element_text(size = 20,
+        theme(plot.title = element_text(size = txt,
                                         face = "bold"),
-              text=element_text(size=20))+
+              text=element_text(size=txt))+
         thm1
 
+    #plot the simulated data with trajectories per each subject
+    p2 <- sim_dat$dat %>%
+        ggplot(aes(x = time,
+                   y = y,
+                   group = subject,
+                   color = treatment)
+        ) +
+        geom_line(aes(size = "Subjects"),
+                  show.legend = FALSE) +
+        # facet_wrap(~ treatment) +
+        geom_line(aes(x = time,
+                      y = mu,
+                      color = treatment,
+                      size = "Simulated Truth"),
+                  lty = 1,show.legend = FALSE) +
+        labs(y='response')+
+        scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
+        theme_classic()+
+        theme(plot.title = element_text(size = txt,
+                                        face = "bold"),
+              text=element_text(size=txt))+
+        thm1
+
+    #plot the errors
+    p3 <- sim_dat$dat %>%
+        ggplot(aes(x = time,
+                   y = errors,
+                   group = subject,
+                   color = treatment)) +
+        geom_line(show.legend=FALSE) +
+        labs(y='errors')+
+        theme_classic()+
+        theme(plot.title = element_text(size = txt,
+                                        face = "bold"),
+              text=element_text(size=txt))+
+        thm1
 
[…]

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts/plot_example_Appendix.R

@@ -1,114 +1,0 @@
-
-## This function plots the rm-ANOVA and LMEM for the data simulated in example.R
-
-plot_example_Appendix <- function(sim_dat) {
-    # Plot the simulated data (scatterplot)
-    txt<-20
-    p1 <- sim_dat$dat %>%
-        ggplot(aes(x = time,
-                   y = y,
-                   group = treatment,
-                   color = treatment)
-        ) +
-        geom_point(show.legend=FALSE) +
-        labs(y='response')+
-        geom_line(aes(x = time,
-                      y = mu,
-                      color = treatment),
-                  show.legend=FALSE) +
-        theme_classic() +
-        theme(plot.title = element_text(size = txt,
-                                        face = "bold"),
-              text=element_text(size=txt))+
-        thm1
-
-    #plot the simulated data with trajectories per each subject
-    p2 <- sim_dat$dat %>%
-        ggplot(aes(x = time,
-                   y = y,
-                   group = subject,
-                   color = treatment)
-        ) +
-        geom_line(aes(size = "Subjects"),
-                  show.legend = FALSE) +
-        # facet_wrap(~ treatment) +
-        geom_line(aes(x = time,
-                      y = mu,
-                      color = treatment,
-                      size = "Simulated Truth"),
-                  lty = 1,show.legend = FALSE) +
-        labs(y='response')+
-        scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-        theme_classic()+
-        theme(plot.title = element_text(size = txt,
-                                        face = "bold"),
-              text=element_text(size=txt))+
-        thm1
-
-    #plot the errors
-    p3 <- sim_dat$dat %>%
-        ggplot(aes(x = time,
-                   y = errors,
-                   group = subject,
-                   color = treatment)) +
-        geom_line(show.legend=FALSE) +
-        labs(y='errors')+
-        theme_classic()+
-        theme(plot.title = element_text(size = txt,
-                                        face = "bold"),
-              text=element_text(size=txt))+
-        thm1
-
-    #plot the model predictions for rm-ANOVA
-    p4 <- ggplot(sim_dat$dat,
-                 aes(x = time,
-                     y = y,
-                     color = treatment)) +
-        geom_point(show.legend=FALSE)+
-        labs(y='response')+
-        geom_line(aes(y = predict(sim_dat$fit_anova),
-                      group = subject, size = "Subjects"),show.legend = FALSE) +
[…]

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.Rmd
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

--- 2021-12-01 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.Rmd
@@
     | 
+    | Response to Reviewer's Comments
+    |
     | Journal of Submission: Statistics in Medicine
+    |
     | Manuscript ID: SIM-21-0640
@@
 ## \textcolor{reviewersblue} {Reply to Introduction}
-We appreciate Dr. Simpson's comments and critiques to our manuscript.
+We greatly appreciate Dr. Simpson's comments and critiques to our manuscript.
 
@@
 
-- L204-207: The term "missing data" has been removed to avoid confusion. We now refer to different number of observations as "unbalanced data".
+- L206-L209: The term "missing data" has been removed to avoid confusion. We now refer to different number of observations as "unbalanced data".
 
-- L212: We have made clear that LMEMs can also work with missing observations.
+- L214: We have made clear that LMEMs can also work with missing observations.
 
-- L220-223: Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates.
+- L222-L225: Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates.
 
@@
 We appreciate the reviewer's perspective regarding the somewhat contradictory argument of fitting a linear effects LMM to quadratic data.
-However, this is an intentional approach because we follow the statistical analysis logic followed by most biomedical researchers to analyze longitudinal data: Fit a LMM or an rm-ANOVA, and if you get significant _p-values_ that's good enough.
-The purpose of the figure is to convey that changing models (from rm-ANOVA to LMEM) without visualizing the model fit is in this case is something that will propagate bias in the results and does not provide any substantial improvement.
+However, this is an intentional approach because we follow the statistical analysis logic followed by most biomedical researchers to analyze longitudinal data: Fit a LMM or an rm-ANOVA, and if a _post-hoc_ analysis gives _p-values_ below a certain threshold (usually 0.05) that's good enough.
+The purpose of the figure is to convey that changing models (from rm-ANOVA to LMEM) without visualizing the model fit is in this case is something that will propagate bias in the results and does not provide any substantial improvement to the analysis.
 However, we acknowledge that this point was not explicitly stated in the original manuscript and the following changes have been made:
 
-- L95: We mention that LMEMs assume a linear relationship "by default", but we indicate that polynomial effects can be used as well, though they have their own shortcomings.
+- L97: We mention that LMEMs assume a linear relationship "by default", but we indicate that polynomial effects can be used as well, though they have their own shortcomings.
 
-- L119-L120 now reads "the limitations of LMEMs and rm-ANOVA regarding an expected trend in the response", instead of "linearity of response" (L119 in the original manuscript).
+- L121-L122 now reads "the limitations of LMEMs and rm-ANOVA regarding an expected trend in the response", instead of "linearity of response" (L119 in the original manuscript).
 
-- L136 now reads "assumed relationship" instead of  "linear relationship" (L135 in the original manuscript).
+- L138 now reads "assumed relationship" instead of  "linear relationship" (L135 in the original manuscript).
 
@@
 
-- L180-L182 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
+- L181-L184 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
 
-- L246-L250: We provide additional context on the purpose of using linear models to fit quadratic data in  which now read "We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions."
+- L248-L251: We provide additional context on the purpose of using linear models to fit quadratic data in  which now read "We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions."
 
 We believe that the aforementioned change makes a clear statement to the reader about our intentions in Fig.1.
-We want to emphasize that strictly speaking the term "default" (used for linear effects in a LMEM) might not be correct, but we are stretching the language to make the point that doing a simple switch from rm-ANOVA to a LMEMs (in R) without any additional changes results in the assumption of linear effects that still provide biased estimates.
+We want to emphasize that strictly speaking the term "default" (used for linear effects in a LMEM) might not be correct, but we are stretching the language to make the point that doing a simple switch from rm-ANOVA to a LMEMs (in R) without any additional considerations on the data results in the assumption of linear effects that still provide biased estimates.
 
-- L271-274 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
+- L272-L274 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
 1.
@@
 
-- L505-509 In Section 6 (Discussion) we emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
+- L552 In Section 6 (Discussion) we emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
 
@@
 
-- L280- L318: Section 4 has been renamed to "Linear Models and Beyond" and the different Subsections cover LMs, GLMs, GLMMs and GAMs.
+- L283- L328: Section 4 has been renamed to "Linear Models and Beyond" and the different Subsections cover LMs, GLMs, GLMMs and GAMs.
 Because the different model classes are not commonly known in the biomedical field, we contextualize the previously presented rm-ANOVA and LMEM into their respective classes to provide the reader with a clear understanding of where each model fits (Section 4.1).
 We also provide equations for the linear predictors in each case (Equations (6) and (8)) following the notation of Equations (1) through (4).
-We want to emphasize that because the target audience of this paper are biomedical researchers that are not familiar with Mathematical Statistics, we follow notation that is easy to read (e.g., using $time_t$ instead of $x_t$), although it might not be standard notation followed in Statistical publications.
+We want to emphasize that because the target audience of this paper are biomedical researchers that are not familiar with Mathematical Statistics, we follow notation that is easy to read (e.g., using $time_t$ instead of $x_t$), although it might not be notation followed in other Statistical publications.
 
@@
 
-- L323: We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the **linear case**.
+- L326: We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the **linear case**.
 This change will avoid the propagation of confusion that Dr.
@@
 
-- L325: We indicate that the term $\varepsilon_{ijt}$ is the residual, and not a separable error.
-Of notice, we are following the notation provided by Dr.
-Simpson in Eq.
-3 in the paper "Modelling Modelling Palaeoecological Time Series Using Generalised Additive Models".
-**do we need this here?
-I am adding it because we are following the same model specification that he has in this paper**
+- L327-L328: We indicate that the term $\varepsilon_{ijt}$ is the deviation of each observation from the mean.
 
@@
 
-- L343-357: This information that originally appeared in L289-294 pertaining splines has been changed, as we now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
+- L351-L356: The information that originally appeared in L289-294 pertaining splines has been changed, as we now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
 Here, we no longer refer to TPRS as an "optimized version" but we indicate that they are a truncated version of thin plate splines (TPS).
@@
 
-- L358-365, which correspond to L297-300 on the original submission no longer refer to "knots" when referring to the construction of the smooth, now indicating that the number of _basis functions_ is what is specified when constructing the smoother.
-We believe that this will make sense to the reader  due to the information provided in L343-357.
+- L358-L360: These lines correspond to L297-L300 on the original submission no longer refer to "knots" when referring to the construction of the smooth, now indicating that the number of _basis functions_ is what is specified when constructing the smoother.
+We believe that this will make sense to the reader  due to the information provided in L351-354.
 
@@
 - We appreciate this comment regarding wiggliness.
-For this particular topic, we have implemented changes that address the reviewer's concerns without adding additional notation (which will obscure the topic to the reader).
-We have used a minimalistic approach here in order to provide some facts about penalization without becoming too technical about it.
+For this particular topic, we have implemented changes that address the reviewer's concerns.
+We have used a minimalist approach here in order to provide some facts about penalization to give a general idea of the logic behind it but without adding additional notation, which we believe will obscure the topic to the reader.
 
-L366-374 has been added, and is a paragraph which briefly discusses wiggliness, the smoothness parameter $\lambda$ and the penalization process using the integrated square of the seocnd derivative of the spline.
+L369-377 has been added, and is a paragraph which briefly discusses wiggliness, the smoothness parameter $\lambda$ and the penalization process using the integrated square of the second derivative of the spline.
 Our goal here is for the reader to understand the effect that $\lambda$ in order to prevent overfitting.
 
-- L373: We have substituted "weak" and "strong" for "low" and "high".
+- L376: We have substituted "weak" and "strong" for "low" and "high".
 
@@
 
-- L392-393 now clarify this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
+- L395-397 now clarify this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
 
 - The sentence "Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model", which appears in L319 in the original manuscript has been removed from the text.
-Instead, the concept of REML has been moved to Section 5.2 L442-444, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in _mgcv_.
+Instead, the concept of REML has been moved to Section 5.2 L458-463, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in _mgcv_.
 
@@
 
-- L396-L403 now contain a more detailed and accurate explanation of"across-the-function" CIs.
+- L400-L408 now contain a more detailed and accurate explanation of"across-the-function" CIs.
 We have used the reviewer's comments to indicate that in GAM fitted using _mgcv_, the close to nominal across-the-function-coverage is true when averaged over the entire function.
+
+- L409-L414 now provide a brief explanation of simultaneous CIs.
+With this explanation we want to provide the reader with an intuitive understanding on the advantages of simultaneous CIs over across the function CIs when computing comparisons between different groups (which we discuss in L417-L419).
+
+- We have modified Figure 3 to now include across the function and simultaneous CIs.
+Fig.
+3B,D now show the fitted smooths and both intervals (across the function and simultaneous), and we explain the rationale for their inclusion in L480-L485.
+In this way, the reader has a visual understanding of the difference between across the function and simultaneous CIs.
 
@@
 You would have to create differences conditional upon a level of
-the other factor. Users, however, tend to expect differences to include the group means, as you have chosen to present things here. Hence a note exp
+the other factor.
+Users, however, tend to expect differences to include the group means, as you have chosen to present things here.
+Hence a note explaining the implications of this choice would be useful to forewarn the user.
 
 ## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We have made major changes to this section based on the reviewer's comments.
+
+- We have removed the sentence "if the difference between the credible intervals of the fitted smooths...
+is non-zero", and now L491-L492 now read: "is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference."
+
+- L497 now reads "unlikely to be distinguishable from zero".
+
+- L493-502 now indicate that we are computing a simultaneous CI around the difference, and that we _included_ the group means in our pairwise comparison.
+We also indicate to the reader the fact that if multiple parametric terms exist in the model the inclusion of the means can become problematic.
+
+However, we want to point that our belief is that including the means makes the comparison more intuitive, as in that way the researcher can directly compare both the fitted smooths and the pairwise comparisons to infer any significant changes and their associated magnitude.
+In  Appendix A we also have made the corrections to the calculation of the pairwise comparisons in Section A.5 (the GAM workflow section) so the reader is aware again that in our code we keep the group means.
 
@@
 
-L416 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
+L431 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
 The computational requisite in the smooth estimation of having the maximum number of basis equal to the number of unique values in the covariate makes it impossible to fit a GAM with a k>5, as the error "A term has fewer unique covariate combinations than specified maximum degrees of freedom" is thrown by _mgcv_.
@@
 Simpson refers here to `gam_01` instead of `gam_00`, as the latter does not include `Group` as part of the model specification and because of this, we don't get a separate smooth per group, which can be seen in the resulting single smooth fitted by the model.
-However, we believe that the clarification he provided here is important, and we have added it to A.4.3 so it is more clear to the reader.
+However, we believe that the clarification he provided here is important, and we have added it to the first paragraph of Section A.4 (Third model) so it is more clear to the reader.
 
@@
 - The name of the model has been changed to `gam_02`, and it appears with this notation in Section 6.2, L383 so it matches the workflow of model selection presented in the Appendix.
-- Additionally, we have used some of the source code of the function `gam.check` in order to use the graphical output of `gratia` and the numerical diagnostic information of `gam.check` without repeating the diagnostic plots, the code for this is now in the Appendix.
+
+- Additionally, we have used some of the source code of the function `gam.check` in order to use the graphical output of `gratia` and the numerical diagnostic information of `gam.check` without repeating the diagnostic plots, the code for this is now in the Appendix B.
 
@@
 
+Our goal in the main manuscript is to familiarize biomedical researchers with the theory of GAMs and their advantages to analyze non-linear trends in longitudinal data.
+We have incorporated Dr.
+Simpson's comments about the concept of GAM fitting in the Section "Final Considerations" in Appendix A.
+Our rationale for adding the information Dr.
+Simpson mentions here is that we want the reader to understand that fitting GAMs needs has to be more than a mechanic exercise in order to be effective.
+Too often, researchers fit a statistical model in a mechanic fashion, and our goal in this work is to move away from this.
+
+After presenting the theory in the main manuscript and the workflow for model selection in Appendix A, we consider that the reader will better appreciate that using `k.check` is the last step of the chain of thought required to fit a GAM.
+
 ## Reviewer's Comments
@@
 
-We provide the code for the figures and models from the manuscript in order to make the work reproducible, but we are aware that readers will not be interested specifically in such code, but rather in the code for the model selection workflow in the Appendix.
-For that part of the Appendix (Section A.4) we do follow the reviewer's comment and have updated the order of the code chunks in order to make it similar to a vignette.
+We have taken the following approach to address this comment:
+
+- We now have two Appendices: Appendix A, which is a tutorial for GAM selection, and Appendix B, which has the code used through the manuscript.
+
+Additionally, the R functions used through the manuscript are now stored in a directory called _scripts_ so they can be easily accessed if desired.
+
+Because we consider that the majority of the readers will be interested in Appendix A, we have used a vignette-style approach by breaking up the code chunks to provide narrative in order to make it more understandable.
+
+Readers interested in the code itself can then revise Appendix B, and in this way we avoid to show long sections of code upfront to the reader by using a single Appendix.
 
@@
 
-We have changed the expression to "data that shows non-linear trends". The change has also been applied to L316 which is now L389 in the revised manuscript.
+We have changed the expression to "data that shows non-linear trends". The change has also been applied to L316 which is now L393 in the revised manuscript.
 
@@
 However, we acknowledge that we didn't clearly conveyed this point to the read.
-In Section 6 (Discussion, L525-527 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided.
+In Section 6 (Discussion, L548-550 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided.
 
@@
 We use the code notation provided by the $LaTeX$ package `listings`, and it provides curly quotes for code.
-Regarding `family`, we believe that it is not necessary to include it here, but we indicate in L445-446 what the argument exists and that other distributions can be used.
+Regarding `family`, we believe that it is not necessary to include it here, but we indicate in L461-463 what the argument exists and that other distributions can be used.
 
@@
 However, we want to make clear to the reader that other types of splines that might be appropriate in certain situations are available.
-To acknowledge this, L440 in the updated manuscript now reads "(a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console)".
+To acknowledge this, L455 in the updated manuscript now reads "(a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console)".
 
@@
 
-We acknowledge the existence of Dr. Fasiolo's _mgcvViz_ package, but due to the introductory nature of this paper we prefer to focus only on _mgcv_ and _gratia_.
+We acknowledge the existence of Dr. Fasiolo's _mgcvViz_ package in Appendix A in the Section "Additional Resources".
 
@@
 
+The sentence in L497-L498 now reads "is unlikely to be distinguishable from zero".
+
 ## Reviewer's Comments
@@
 
-We have made the appropriate changes. The sentence now reads "The model is therefore able to identify changes between the groups at time points..."
+We have made the appropriate changes. The sentence now reads "The model is therefore able to identify changes between the groups at time points..." (L562)

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.tex
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

--- 2021-12-01 Manuscripts/Manuscript_by_chapters-SIM_Revisions/response_to_reviewers/Response_to_reviewers.tex
@@
 ]{article}
-\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
-Beyond repeated measures ANOVA and Linear Mixed Models\\
-\strut \\
-Journal of Submission: Statistics in Medicine\\
-Manuscript ID: SIM-21-0640}
-\author{}
-\date{\vspace{-2.5em}}
-
 \usepackage{amsmath,amssymb}
@@
 
+\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
+Beyond repeated measures ANOVA and Linear Mixed Models\\
+\strut \\
+Response to Reviewer's Comments\\
+\strut \\
+Journal of Submission: Statistics in Medicine\\
+\strut \\
+Manuscript ID: SIM-21-0640}
+\author{}
+\date{\vspace{-2.5em}}
+
 \begin{document}
@@
 
-We appreciate Dr.~Simpson's comments and critiques to our manuscript.
+We greatly appreciate Dr.~Simpson's comments and critiques to our manuscript.
 
@@
 \item
-  L204-207: The term ``missing data'' has been removed to avoid confusion. We now refer to different number of observations as ``unbalanced data''.
+  L206-L209: The term ``missing data'' has been removed to avoid confusion. We now refer to different number of observations as ``unbalanced data''.
 \item
-  L212: We have made clear that LMEMs can also work with missing observations.
+  L214: We have made clear that LMEMs can also work with missing observations.
 \item
-  L220-223: Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates.
+  L222-L225: Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates.
 \end{itemize}
@@
 We appreciate the reviewer's perspective regarding the somewhat contradictory argument of fitting a linear effects LMM to quadratic data.
-However, this is an intentional approach because we follow the statistical analysis logic followed by most biomedical researchers to analyze longitudinal data: Fit a LMM or an rm-ANOVA, and if you get significant \emph{p-values} that's good enough.
-The purpose of the figure is to convey that changing models (from rm-ANOVA to LMEM) without visualizing the model fit is in this case is something that will propagate bias in the results and does not provide any substantial improvement.
+However, this is an intentional approach because we follow the statistical analysis logic followed by most biomedical researchers to analyze longitudinal data: Fit a LMM or an rm-ANOVA, and if a \emph{post-hoc} analysis gives \emph{p-values} below a certain threshold (usually 0.05) that's good enough.
+The purpose of the figure is to convey that changing models (from rm-ANOVA to LMEM) without visualizing the model fit is in this case is something that will propagate bias in the results and does not provide any substantial improvement to the analysis.
 However, we acknowledge that this point was not explicitly stated in the original manuscript and the following changes have been made:
@@
 \item
-  L95: We mention that LMEMs assume a linear relationship ``by default'', but we indicate that polynomial effects can be used as well, though they have their own shortcomings.
+  L97: We mention that LMEMs assume a linear relationship ``by default'', but we indicate that polynomial effects can be used as well, though they have their own shortcomings.
 \item
-  L119-L120 now reads ``the limitations of LMEMs and rm-ANOVA regarding an expected trend in the response'', instead of ``linearity of response'' (L119 in the original manuscript).
+  L121-L122 now reads ``the limitations of LMEMs and rm-ANOVA regarding an expected trend in the response'', instead of ``linearity of response'' (L119 in the original manuscript).
 \item
-  L136 now reads ``assumed relationship'' instead of ``linear relationship'' (L135 in the original manuscript).
+  L138 now reads ``assumed relationship'' instead of ``linear relationship'' (L135 in the original manuscript).
 \item
@@
 \item
-  L180-L182 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
+  L181-L184 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
 \item
-  L246-L250: We provide additional context on the purpose of using linear models to fit quadratic data in which now read ``We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with''default'' (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.''
+  L248-L251: We provide additional context on the purpose of using linear models to fit quadratic data in which now read ``We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with''default'' (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.''
 \end{itemize}
@@
 We believe that the aforementioned change makes a clear statement to the reader about our intentions in Fig.1.
-We want to emphasize that strictly speaking the term ``default'' (used for linear effects in a LMEM) might not be correct, but we are stretching the language to make the point that doing a simple switch from rm-ANOVA to a LMEMs (in R) without any additional changes results in the assumption of linear effects that still provide biased estimates.
+We want to emphasize that strictly speaking the term ``default'' (used for linear effects in a LMEM) might not be correct, but we are stretching the language to make the point that doing a simple switch from rm-ANOVA to a LMEMs (in R) without any additional considerations on the data results in the assumption of linear effects that still provide biased estimates.
 
@@
 \item
-  L271-274 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
+  L272-L274 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
 1.
@@
 \item
-  L505-509 In Section 6 (Discussion) we emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
+  L552 In Section 6 (Discussion) we emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
 \end{itemize}
@@
 \item
-  L280- L318: Section 4 has been renamed to ``Linear Models and Beyond'' and the different Subsections cover LMs, GLMs, GLMMs and GAMs.
+  L283- L328: Section 4 has been renamed to ``Linear Models and Beyond'' and the different Subsections cover LMs, GLMs, GLMMs and GAMs.
 Because the different model classes are not commonly known in the biomedical field, we contextualize the previously presented rm-ANOVA and LMEM into their respective classes to provide the reader with a clear understanding of where each model fits (Section 4.1).
 We also provide equations for the linear predictors in each case (Equations (6) and (8)) following the notation of Equations (1) through (4).
-We want to emphasize that because the target audience of this paper are biomedical researchers that are not familiar with Mathematical Statistics, we follow notation that is easy to read (e.g., using \(time_t\) instead of \(x_t\)), although it might not be standard notation followed in Statistical publications.
+We want to emphasize that because the target audience of this paper are biomedical researchers that are not familiar with Mathematical Statistics, we follow notation that is easy to read (e.g., using \(time_t\) instead of \(x_t\)), although it might not be notation followed in other Statistical publications.
 \end{itemize}
@@
 \item
-  L323: We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the \textbf{linear case}.
+  L326: We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the \textbf{linear case}.
 This change will avoid the propagation of confusion that Dr.~Simpson correctly indicated in his comment.
 \item
-  L325: We indicate that the term \(\varepsilon_{ijt}\) is the residual, and not a separable error.
-Of notice, we are following the notation provided by Dr.~Simpson in Eq.
-3 in the paper ``Modelling Modelling Palaeoecological Time Series Using Generalised Additive Models''.
-\textbf{do we need this here?
-I am adding it because we are following the same model specification that he has in this paper}
+  L327-L328: We indicate that the term \(\varepsilon_{ijt}\) is the deviation of each observation from the mean.
 \end{itemize}
@@
 \item
-  L343-357: This information that originally appeared in L289-294 pertaining splines has been changed, as we now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
+  L351-L356: The information that originally appeared in L289-294 pertaining splines has been changed, as we now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
 Here, we no longer refer to TPRS as an ``optimized version'' but we indicate that they are a truncated version of thin plate splines (TPS).
@@
 \item
-  L358-365, which correspond to L297-300 on the original submission no longer refer to ``knots'' when referring to the construction of the smooth, now indicating that the number of \emph{basis functions} is what is specified when constructing the smoother.
-We believe that this will make sense to the reader due to the information provided in L343-357.
+  L358-L360: These lines correspond to L297-L300 on the original submission no longer refer to ``knots'' when referring to the construction of the smooth, now indicating that the number of \emph{basis functions} is what is specified when constructing the smoother.
+We believe that this will make sense to the reader due to the information provided in L351-354.
 \end{itemize}
@@
   We appreciate this comment regarding wiggliness.
-For this particular topic, we have implemented changes that address the reviewer's concerns without adding additional notation (which will obscure the topic to the reader).
-We have used a minimalistic approach here in order to provide some facts about penalization without becoming too technical about it.
+For this particular topic, we have implemented changes that address the reviewer's concerns.
+We have used a minimalist approach here in order to provide some facts about penalization to give a general idea of the logic behind it but without adding additional notation, which we believe will obscure the topic to the reader.
 \end{itemize}
 
-L366-374 has been added, and is a paragraph which briefly discusses wiggliness, the smoothness parameter \(\lambda\) and the penalization process using the integrated square of the seocnd derivative of the spline.
+L369-377 has been added, and is a paragraph which briefly discusses wiggliness, the smoothness parameter \(\lambda\) and the penalization process using the integrated square of the second derivative of the spline.
 Our goal here is for the reader to understand the effect that \(\lambda\) in order to prevent overfitting.
@@
 \item
-  L373: We have substituted ``weak'' and ``strong'' for ``low'' and ``high''.
+  L376: We have substituted ``weak'' and ``strong'' for ``low'' and ``high''.
 \end{itemize}
@@
 \item
-  L392-393 now clarify this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
+  L395-397 now clarify this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
 \item
   The sentence ``Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model'', which appears in L319 in the original manuscript has been removed from the text.
-Instead, the concept of REML has been moved to Section 5.2 L442-444, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in \emph{mgcv}.
+Instead, the concept of REML has been moved to Section 5.2 L458-463, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in \emph{mgcv}.
 \end{itemize}
@@
 \begin{itemize}
-\tightlist
 \item
-  L396-L403 now contain a more detailed and accurate explanation of''across-the-function'' CIs.
+  L400-L408 now contain a more detailed and accurate explanation of''across-the-function'' CIs.
 We have used the reviewer's comments to indicate that in GAM fitted using \emph{mgcv}, the close to nominal across-the-function-coverage is true when averaged over the entire function.
+\item
+  L409-L414 now provide a brief explanation of simultaneous CIs.
+With this explanation we want to provide the reader with an intuitive understanding on the advantages of simultaneous CIs over across the function CIs when computing comparisons between different groups (which we discuss in L417-L419).
+\item
+  We have modified Figure 3 to now include across the function and simultaneous CIs.
+Fig.
+3B,D now show the fitted smooths and both intervals (across the function and simultaneous), and we explain the rationale for their inclusion in L480-L485.
+In this way, the reader has a visual understanding of the difference between across the function and simultaneous CIs.
 \end{itemize}
@@
 \subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-9}}
+
+We have made major changes to this section based on the reviewer's comments.
+
+\begin{itemize}
+\item
+  We have removed the sentence ``if the difference between the credible intervals of the fitted smooths\ldots{} is non-zero'', and now L491-L492 now read: ``is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.''
+\item
+  L497 now reads ``unlikely to be distinguishable from zero''.
+\item
+  L493-502 now indicate that we are computing a simultaneous CI around the difference, and that we \emph{included} the group means in our pairwise comparison.
+We also indicate to the reader the fact that if multiple parametric terms exist in the model the inclusion of the means can become problematic.
+\end{itemize}
+
+However, we want to point that our belief is that including the means makes the comparison more intuitive, as in that way the researcher can directly compare both the fitted smooths and the pairwise comparisons to infer any significant changes and their associated magnitude.
+In Appendix A we also have made the corrections to the calculation of the pairwise comparisons in Section A.5 (the GAM workflow section) so the reader is aware again that in our code we keep the group means.
 
@@
 
-L416 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
+L431 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
 The computational requisite in the smooth estimation of having the maximum number of basis equal to the number of unique values in the covariate makes it impossible to fit a GAM with a k\textgreater5, as the error ``A term has fewer unique covariate combinations than specified maximum degrees of freedom'' is thrown by \emph{mgcv}.
@@
 We believe Dr.~Simpson refers here to \passthrough{\lstinline!gam\_01!} instead of \passthrough{\lstinline!gam\_00!}, as the latter does not include \passthrough{\lstinline!Group!} as part of the model specification and because of this, we don't get a separate smooth per group, which can be seen in the resulting single smooth fitted by the model.
-However, we believe that the clarification he provided here is important, and we have added it to A.4.3 so it is more clear to the reader.
+However, we believe that the clarification he provided here is important, and we have added it to the first paragraph of Section A.4 (Third model) so it is more clear to the reader.
 
@@
 \begin{itemize}
-\tightlist
 \item
@@
 \item
-  Additionally, we have used some of the source code of the function \passthrough{\lstinline!gam.check!} in order to use the graphical output of \passthrough{\lstinline!gratia!} and the numerical diagnostic information of \passthrough{\lstinline!gam.check!} without repeating the diagnostic plots, the code for this is now in the Appendix.
+  Additionally, we have used some of the source code of the function \passthrough{\lstinline!gam.check!} in order to use the graphical output of \passthrough{\lstinline!gratia!} and the numerical diagnostic information of \passthrough{\lstinline!gam.check!} without repeating the diagnostic plots, the code for this is now in the Appendix B.
 \end{itemize}
@@
 
+Our goal in the main manuscript is to familiarize biomedical researchers with the theory of GAMs and their advantages to analyze non-linear trends in longitudinal data.
+We have incorporated Dr.~Simpson's comments about the concept of GAM fitting in the Section ``Final Considerations'' in Appendix A.
+Our rationale for adding the information Dr.~Simpson mentions here is that we want the reader to understand that fitting GAMs needs has to be more than a mechanic exercise in order to be effective.
+Too often, researchers fit a statistical model in a mechanic fashion, and our goal in this work is to move away from this.
+
+After presenting the theory in the main manuscript and the workflow for model selection in Appendix A, we consider that the reader will better appreciate that using \passthrough{\lstinline!k.check!} is the last step of the chain of thought required to fit a GAM.
+
 \hypertarget{reviewers-comments-13}{%
@@
 
-We provide the code for the figures and models from the manuscript in order to make the work reproducible, but we are aware that readers will not be interested specifically in such code, but rather in the code for the model selection workflow in the Appendix.
-For that part of the Appendix (Section A.4) we do follow the reviewer's comment and have updated the order of the code chunks in order to make it similar to a vignette.
+We have taken the following approach to address this comment:
+
+\begin{itemize}
+\tightlist
+\item
+  We now have two Appendices: Appendix A, which is a tutorial for GAM selection, and Appendix B, which has the code used through the manuscript.
+\end{itemize}
+
+Additionally, the R functions used through the manuscript are now stored in a directory called \emph{scripts} so they can be easily accessed if desired.
+
+Because we consider that the majority of the readers will be interested in Appendix A, we have used a vignette-style approach by breaking up the code chunks to provide narrative in order to make it more understandable.
+
+Readers interested in the code itself can then revise Appendix B, and in this way we avoid to show long sections of code upfront to the reader by using a single Appendix.
 
@@
 
-We have changed the expression to ``data that shows non-linear trends''. The change has also been applied to L316 which is now L389 in the revised manuscript.
+We have changed the expression to ``data that shows non-linear trends''. The change has also been applied to L316 which is now L393 in the revised manuscript.
 
@@
 We use the code notation provided by the \(LaTeX\) package \passthrough{\lstinline!listings!}, and it provides curly quotes for code.
-Regarding \passthrough{\lstinline!family!}, we believe that it is not necessary to include it here, but we indicate in L445-446 what the argument exists and that other distributions can be used.
+Regarding \passthrough{\lstinline!family!}, we believe that it is not necessary to include it here, but we indicate in L461-463 what the argument exists and that other distributions can be used.
 
@@
 However, we want to make clear to the reader that other types of splines that might be appropriate in certain situations are available.
-To acknowledge this, L440 in the updated manuscript now reads ``(a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console)''.
+To acknowledge this, L455 in the updated manuscript now reads ``(a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console)''.
 
@@
 
-We acknowledge the existence of Dr.~Fasiolo's \emph{mgcvViz} package, but due to the introductory nature of this paper we prefer to focus only on \emph{mgcv} and \emph{gratia}.
+We acknowledge the existence of Dr.~Fasiolo's \emph{mgcvViz} package in Appendix A in the Section ``Additional Resources''.
 
@@
 
+The sentence in L497-L498 now reads ``is unlikely to be distinguishable from zero''.
+
 \hypertarget{reviewers-comments-36}{%
@@
 
-We have made the appropriate changes. The sentence now reads ``The model is therefore able to identify changes between the groups at time points\ldots{}''
+We have made the appropriate changes. The sentence now reads ``The model is therefore able to identify changes between the groups at time points\ldots{}'' (L562)
 

### 2021-12-26 Code and miscellaneous/Interval_comparisons.Rmd
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

---
title: Testing the simultaneous CI functions
---

```{r setup, message = FALSE,include=FALSE}
#the packages needed to run the code

library(patchwork)
library(tidyverse)
library(mvnfast)
library(nlme)
library(mgcv)
library(gratia)
library(scico)
set.seed(2021) #set seed for reproducibility
thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
```

I want to test if there's a difference because when knitting I get the message:

`In gratia:::get_vcov(model, unconditional = unconditional, frequentist = frequentist) :Covariance corrected for smoothness uncertainty not available.`

```{r,generate-simulated-data}
#data, "the truth"
dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
            Day=rep(c(0,2,5,7,10),times=2),
            Group=as.factor(rep(c("Control","Treatment"),each=5))
)

#simulate data
simulate_data <- function(dat, n = 10, sd = 5) {
    dat_sim <- dat %>%
        slice(rep(1:n(), each = n)) %>%
        group_by(Group, Day) %>%
        mutate(
            StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
            subject=rep(1:10),
            subject=factor(paste(subject, Group, sep = "-"))
        ) %>%
        ungroup()

    return(dat_sim)
}

#subject = factor(paste(subject, treatment, sep = "-")))

n <- 10 #number of observations
sd <- 10 #approximate sd from paper
df <- 6
dat_sim <- simulate_data(dat, n, sd)

#plot data

c<-ggplot(dat_sim, aes(x = Day, 
                        y = StO2_sim, 
                        color = Group,
                        group=subject)) +

    geom_line(size=0.6, alpha=0.4,show.legend=FALSE)+
    geom_line(aes(x = Day, 
               y = StO2,
               color=Group),
           size=1.5,
           data=dat,
           inherit.aes=FALSE,
              show.legend = FALSE)+
  labs(y=expression(atop(StO[2],'(simulated)')))+
  theme_classic()+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    thm1

```

Now, I fit the GAM.

```{r,fit-GAM}
#GAM, interaction of time and treatment

gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
              method='REML',
              data  = dat_sim)
```

Now I follow the exact code from G. Simpson's post to generate the simultaneous interval (link to the original post below):
https://fromthebottomoftheheap.net/2016/12/15/simultaneous-interval-revisited/

```{r,simulatenous-interval}
#Generate random values from multivariate normal
rmvn <- function(n, mu, sig) { ## MVN random deviates
    L <- mroot(sig)
    m <- ncol(L)
    t(mu + L %*% matrix(rnorm(m*n), m, n))
}

#Bayesian covariance matrix of the model coefficients
Vb <- vcov(gam_02)

#grid of values for confidence band
newd <- expand_grid(Group = factor(c("Control", "Treatment")),
                                   Day = seq(0, 10, by = 0.1),
                                   subject=factor(rep(1:10)))

#Generate predictions and standard errors
pred <- predict(gam_02, newd, se.fit = TRUE)
se.fit <- pred$se.fit

#number of draws
N <- 10000

#bias of the estimated model coefficients
BUdiff <- rmvn(N, mu = rep(0, nrow(Vb)), sig = Vb)

#evaluate the basis functions at the locations of x
Cg <- predict(gam_02, newd, type = "lpmatrix")

#Computes deviations between the fitted and true parameters
simDev <- Cg %*% t(BUdiff)

#compute absolute values of the standardized deviations from the true model
absDev <- abs(sweep(simDev, 1, se.fit, FUN = "/"))

#find maximum of the absolute standardized deviations at the grid of values of x for each simulation
masd <- apply(absDev, 2L, max)

#Find critical value used to scale the standard errors to yield the simultaneous interval

crit <- quantile(masd, prob = 0.95, type = 8)

pred <- transform(cbind(data.frame(pred), newd),
                  uprP = fit + (2 * se.fit),
                  lwrP = fit - (2 * se.fit),
                  uprS = fit + (crit * se.fit),
                  lwrS = fit - (crit * se.fit))

#plot the simultaneous and pointwise intervals
a<-ggplot(pred, aes(x =Day,group=Group,color=Group,fill=Group)) +
    geom_path(aes(y=fit),lwd=2,show.legend = FALSE)+
    geom_ribbon(aes(ymin = lwrS, ymax = uprS), alpha = 0.2,show.legend = FALSE) +
    geom_ribbon(aes(ymin = lwrP, ymax = uprP), alpha = 0.2,show.legend = FALSE) +
    labs(y = "G. Simpson's code")+
   geom_hline(yintercept = 60,color="red")+
    geom_hline(yintercept = 20,color="red")+
    geom_hline(yintercept = 0,color="red")+
    theme_classic()+
    thm1

```

Now I will use the code from our paper to get the same intervals

```{r,code-from-paper}

ci <- confint(gam_02, parm = "s(Day)", partial_match = TRUE, type = "confidence")
## simultaneous interval
si <- confint(gam_02, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)

# mean shift for group 2

Intercept <-coef(gam_02)[1]
const <- coef(gam_02)[2]+Intercept

ci <- ci %>%
mutate(est = case_when(Group == "Treatment" ~ est + const,
TRUE ~ est+Intercept),
lower = case_when(Group == "Treatment" ~ lower + const,
TRUE ~ lower+Intercept),
upper = case_when(Group == "Treatment" ~ upper + const,
TRUE ~ upper+Intercept))

si <- si %>%
mutate(est = case_when(Group == "Treatment" ~ est + const,
TRUE ~ est+Intercept),
lower = case_when(Group == "Treatment" ~ lower + const,
TRUE ~ lower+Intercept),
upper = case_when(Group == "Treatment" ~ upper + const,
TRUE ~ upper+Intercept))

b<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
geom_line(lwd = 1,show.legend = FALSE) +
geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
inherit.aes = FALSE, alpha = 0.7,
show.legend=FALSE) +
geom_ribbon(data = si,
mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
inherit.aes = FALSE, alpha = 0.4,
show.legend=FALSE)+
    geom_line(data=si,aes(Day,upper,color=Group),show.legend = FALSE)+
    geom_line(data=si,aes(Day,lower,color=Group),show.legend = FALSE)+
    geom_hline(yintercept = 60,color="red")+
    geom_hline(yintercept = 20,color="red")+
    geom_hline(yintercept = 0,color="red")+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
     labs(y = "Our code")+
      theme_classic()+
        thm1

```

Now I plot both results and is strange, but the intervals that we get from G. Simpson's code are wider! why?

```{r,plotting-data-and-intervals}
#plot both
c+a+b+scale_y_continuous(breaks=c(-20,0,20,40,60))

```

```{r,poitwise-diff}
##this function determines the pointwise confidence interval of a difference between two smooths

difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
    ## make sure f1 and f2 are characters
    f1 <-  as.character(f1)
    f2 <-  as.character(f2)
    cnames <- colnames(Xp)
    ## columns of Xp associated with pair of smooths
    c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
    c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
    ## rows of Xp associated with pair of smooths
    r1 <- data[[by_var]] == f1
    r2 <- data[[by_var]] == f2

    ## difference rows of Xp for pair of smooths
    X <- Xp[r1, ] - Xp[r2, ]

    ######IMPORTANT: uncommenting the following two lines
    #removes the group means from the comparison######

    ## zero the cols related to other splines
    # X[, ! (c1 | c2)] <- 0

    ## zero out the parametric cols
    #X[, !grepl('^s\\(', cnames)] <- 0

    ## compute difference
    sm_diff <- drop(X %*% coefs)
    se <- sqrt(rowSums((X %*% V) * X))
    nr <- NROW(X)

    ## Calculate posterior simulation for smooths
    coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
    rownames(coefs_sim) <- rownames(V)
    simDev <- X %*% coefs_sim
    absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
    masd <- apply(absDev, 2, max)
    crit_s <- quantile(masd, prob = 0.95, type = 8)

    out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
                level_1 = rep(f1, nr),
                level_2 = rep(f2, nr),
                diff = sm_diff, se = se,
                lower_s = sm_diff - crit_s * se,
                upper_s = sm_diff + crit_s*se)

    out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
    ## Only need rows associated with one of the levels
    out <- bind_cols(out, data[r1, smooth_var])

    out
}

```

And this:

```{r,difference-smooths}
#this function calculates the pointwise (by calling pointwise_comparisons.R) CI and the simultaneous CI
# for a pairwise comparison between two smooths
difference_smooths <- function(model,
                               smooth,
                               n = 100,
                               ci_level = 0.95,
                               newdata = NULL,
                               partial_match = TRUE,
                               unconditional = FALSE,
                               frequentist = FALSE,
                               nrep = 10000,
                               include_means = TRUE,
                               ...) {
    if (missing(smooth)) {
        stop("Must specify a smooth to difference via 'smooth'.")
    }

    # smooths in model
    S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
    # select smooths
    select <-
        gratia:::check_user_select_smooths(smooths = S, select = smooth,
                                           partial_match = partial_match)#,
    # model_name = expr_label(substitute(object)))
    sm_ids <- which(select)
    smooths <- gratia::get_smooths_by_id(model, sm_ids)
    sm_data <- map(sm_ids, gratia:::smooth_data,
                   model = model, n = n, include_all = TRUE)
    sm_data <- bind_rows(sm_data)
    by_var <- by_variable(smooths[[1L]])
    smooth_var <- gratia:::smooth_variable(smooths[[1L]])
    pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
                                     stringsAsFactor = FALSE))
    names(pairs) <- paste0("f", 1:2)

    Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
    V <- gratia:::get_vcov(model, unconditional = unconditional,
                           frequentist = frequentist)
    coefs <- coef(model)

    out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
                smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
                coefs = coefs, nrep = nrep)
    out <- bind_rows(out)
    crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)

    out <- add_column(out,
                      lower = out$diff - (crit * out$se),
                      upper = out$diff + (crit * out$se),
                      .after = 6L)
    out
}
```

Let's get the fit then:

```{r,testing}
diff_testing <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
                     unconditional = TRUE, frequentist = FALSE,
                     n=100, partial_match = TRUE, nrep=10000)

```

### 2021-12-26 Code and miscellaneous/Simultaneous-interval-differences.Rmd
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

@@
 
-Now I follow the exact code from G. Simpson's post to generate the simutaneous interval (link to the original post below):
+Now I follow the exact code from G. Simpson's post to generate the simultaneous interval (link to the original post below):
 https://fromthebottomoftheheap.net/2016/12/15/simultaneous-interval-revisited/

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/Full_document_SIM_No_Appendix.Rmd
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

@@
 set.seed(2021) #set seed for reproducibility
-
-#global themes for plotting
-#thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
-#thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
 thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
@@
 
-\newpage
 # References

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/Full_document_SIM_No_Appendix.tex
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

--- 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/Full_document_SIM_No_Appendix.tex
@@
 where \(\beta_{0}\) is the intercept, and \(\beta_{1}\) is the coefficient for each treatment group.
-Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time, and the interaction between time and treatment have been replaced by \(f(time_{t}|\beta_{j})\).
+Notice that the construction of the predictor is similar to that of Equation \eqref{eq:GLM-eta}, but in this case the parametric terms involving the effect of time, and the interaction between time and treatment have been replaced by the smooth term \(f(time_{t}|\beta_{j})\).
 The smooth term \(f(time_{t}|\beta_{j})\) gives a different smooth response for each treatment.
-If the smooth term represented a linear relationship, then \(f(time_{t}|\beta_{j})= \beta_2 \times time_t+\beta_3 \times time_t \times treatment_j\); however, in general, the smooth term is a more flexible function than a linear relationship, with parameter vectors \(\beta_{j}\) for each treatment.
-A GAM version of a linear model can be written as
+\footnote{If the smooth term represented a linear relationship, then \(f(time_{t}|\beta_{j})= \beta_2 \times time_t+\beta_3 \times time_t \times treatment_j\); however, in general, the smooth term is a more flexible function than a linear relationship, with parameter vectors \(\beta_{j}\) for each treatment} A GAM version of a linear model can be written as
 
@@
 
-In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}.
+In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an computationally efficient\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}.
 Like TPS, TPRS only requires specifying the number of basis functions to be used to create the smoother (for mathematical details on both TPS and TPRS see Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}).
 
-To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}D.
+To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 that appears in Figure \ref{fig:l-q-response}D.
 The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time.
@@
 As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
-In other words,a low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
-
-With this in mind, if five basis functions are used to fit a GAM for the data of Group 1 (concave down function) that appears in Figure \ref{fig:l-q-response}D, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A, where the four basis functions (and the intercept) are shown.
+In other words, a low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
+
+With this in mind, if four basis functions (plus the intercept) are used to fit a GAM for the data of Group 1 (concave down function) that appears in Figure \ref{fig:l-q-response}D, the resulting fitting process is shown in Figure \ref{fig:basis-plot}.
+In Figure \ref{fig:basis-plot}A the four basis functions (and the intercept) are shown.
 Each of the five basis functions is evaluated at six different points (because there are six points on the timeline).
 The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), that is estimated when fitting the model.
-The penalized coefficient estimates fitted for our example are shown in Figure \ref{fig:basis-plot}B.
-
-To get the weighted basis functions, each basis (from Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
-Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions.
-Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
+The penalized coefficient estimates fitted are shown in Figure \ref{fig:basis-plot}B.
+
+To get the weighed basis functions, each basis (from Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
+Figure \ref{fig:basis-plot}C shows the resulting weighed basis functions.
+Note that the magnitude of the weighing for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
 On the other hand, the third basis function has roughly doubled its contribution to the smooth term.
 Finally, the weighted basis functions are added at each timepoint to produce the smooth term.
-The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (brown line), along the simulated values per group, which appear as points.
+The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (brown line), along the simulated values which appear as points.
 
@@
 Briefly, to obtain a simultaneous CI we use simulation from the empirical Bayesian posterior distribution to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}.
-Overall, a simultaneous CI provides more robust estimates that can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
 
 In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller\textsuperscript{\protect\hyperlink{ref-miller2019}{61}}, Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}}, Marra\textsuperscript{\protect\hyperlink{ref-marra2012}{63}}, and Ruppert\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}.
-With this brief introduction to the Bayesian interpretation of GAMs, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends.
+What we want to convey is that a Bayesian interpretation of GAMs allows for robust estimation using simultaneous empirical Bayesian CIs, as their stimates can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
+
+With this in mind, in the next section we consider the use of GAMs to analyze longitudinal biomedical data with non-linear trends, and use simultaneous empirical Bayesian CIs to assess significance between treatment groups.
 
@@
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}\textsuperscript{\protect\hyperlink{ref-gratia}{64}}.
-A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak\textsuperscript{\protect\hyperlink{ref-harezlak2018}{65}}.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}\textsuperscript{\protect\hyperlink{ref-gratia}{64}} as we show in Appendix A.
+Additional discussions on model selection can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak\textsuperscript{\protect\hyperlink{ref-harezlak2018}{65}}.
 
@@
 \textbf{A}: Pairwise comparisons for the full dataset.
-\textbf{B}: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the 95\% empirical Bayesian credible interval does not cover 0.
+\textbf{B}: Pairwise comparisons for the dataset with incomplete observations.
+Significant differences exist where the 95\% empirical Bayesian simultaneous CI does not cover 0.
 In both cases the effect of treatment is significant after day 3.
-In both cases, a simultaneous CI has been computed around the smooth difference.}\label{fig:plot-pairwise-comp}
+For the difference, we have included the means so the the value of the difference has direct correspondence with the scale of the response (Figure \ref{fig:sim-smooth-plot}A).}\label{fig:plot-pairwise-comp}
 \end{figure}
 
-The absence of a \emph{p-value} in this context might seem odd, but its interpretation can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment.
-This is what happens for the simulated data in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
-
-With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant.
+The absence of a \emph{p-value} in this context might seem odd, but its interpretation of the comparison can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
+
+With this expectation of different trends in each group, computing the difference between the trends will identify if and \emph{when} the observed difference is significant.
 A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
-
-Consider the calculation of pairwise differences for the smooths in Figure \ref{fig:sim-smooth-plot}B, D.
-Figure \ref{fig:plot-pairwise-comp} shows the comparison between each treatment group for the full and missing datasets with a simultaneous CI computed around the difference.
+Therefore, the estimation of the difference between the smooths and the correspondent simultaneous empirical Bayesian CI of this difference provides an estimation of significance that is more robust that a mere \emph{p-value} because the estimated difference is able to indicates the significant time interval and we know that on average, the correction provided by the simultaneous empirical Bayesian CI ensures that it will contain \emph{the whole function} at a nominal value (say, 95\%).
+
+Consider the calculation of pairwise differences for the fitted smooths that appear in Figure \ref{fig:sim-smooth-plot}B, D.
+Figure \ref{fig:plot-pairwise-comp} shows the difference between each treatment group trend for the full and missing datasets with a simultaneous CI computed around the difference.
 Here, the ``Control'' group is used as the reference to which ``Treatment'' group is being compared.
-Notice that because we have included the means of each group, there is correspondence between the scale of the original data the scale of the pairwise comparisons.
-This can be seen as in Figure \ref{fig:sim-smooth-plot}B at day 5 there is essentially a 50\% difference between \(\mbox{StO}_2\) in both groups, which corresponds to the -50 difference in \ref{fig:plot-pairwise-comp}A.
-However, if there are multiple parametric terms in the model (more factors that need to be specified in the model) such inclusion of the means can become problematic.
+Notice that because we have included the means of each group, there is correspondence between the scale of the original data and the scale of the pairwise comparisons.
+This can be seen in Figure \ref{fig:sim-smooth-plot}B, where at day 5 there is essentially a difference of 50\% between \(\mbox{StO}_2\) in both groups, which corresponds to the -50\% difference in \ref{fig:plot-pairwise-comp}A.
+However, if there are multiple parametric terms in the model (more factors that need to be specified), such inclusion of the means can become problematic.
 However, we believe that the model we have presented here suffices in a wide range of situations where adding the group means is a relatively easy implementation that can help better visualize the estimation from the model from a biological perspective.
 
-In Figure \ref{fig:plot-pairwise-comp}A the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
-Notice that the shaded region between days 0 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by \(\approx\) 3 day it is the ``Treatment'' group the one that on average, has greater \(\mbox{StO}_2\).
+In Figure \ref{fig:plot-pairwise-comp}A, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a different effect than the other.
+Notice that the shaded region between days 1 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by day \(\approx\) 3 it is the ``Treatment'' group that statistically on average has greater \(\mbox{StO}_2\).
 This would suggest that the effect of chemotherapy in the ``Treatment'' group becomes significant after day 3 for the given model.
@@
 
-In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
+In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.
 It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (e.g., which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research\textsuperscript{\protect\hyperlink{ref-hastie1995}{67}}.
 This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs.
-T
-
-The purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model, such as in the case of polynomial effects in LMEMs.
+
+However, in Section \ref{GAM-theory} we demonstrated that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model, such as in the case of polynomial effects in LMEMs where additionally there is no biological interpretation for such effect.
 In contrast, because GAMs let the data speak for themselves they can provide an accurate representation of the effect of time in a biological process.
@@
 \begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\newpage
 
@@
 \CSLRightInline{Rose NL, Yang H, Turner SD, Simpson GL.
-An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to scotland, {UK}.
+An assessment of the mechanisms for the transfer of lead and mercury from atmospherically contaminated organic soils to lake sediments with particular reference to {S}cotland, {UK}.
 \emph{Geochimica et Cosmochimica Acta}.

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_A.Rmd
Working on Appendix B
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

@@
 al<-0.8
+
+thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
 
@@
 
+## Final considerations
+
+In this Appendix we have shown a basic model selection workflow for GAMs.
+Our goal here is to familiarize the reader with the logic behind the construction of each model and the kind of diagnostic information that needs to be checked to ensure the model is adequate.
+One final consideration must be given to the conditional distribution of the response.
+In the main manuscript and through this workflow, we have assumed a normal distribution for the response, which we believe is appropriate in many biomedical scenarios.
+However, we must remind the reader that the normal distribution is not intended to be used as a "one size fits all" distribution.
+Depending on the type of data (counts, binary outcomes) the user can choose different conditional distributions in _mgcv_.
+
+We also want to indicate that one of the major advantages of fitting GAMs in R using _mgcv_ is that once the appropriate conditional distribution has been chosen and model diagnostics are assessed, the user only needs to choose a basis dimension (the number of basis functions to use), and check if the basis dimension is adequately capturing the trend of the data using `k.check`.
+If that is  not the case, then the user needs to increase `k` a bit and check `k.check` again.
+The process might seem mechanic, but we hope that with the theory presented in the main manuscript and the workflow of this Appendix the user has a good understanding of what the model is doing and the rationale for choosing a GAM as a tool for statistical analysis.
+
 ## Additional resources
@@
 Edition can be found [here](https://www.routledge.com/Generalized-Additive-Models-An-Introduction-with-R-Second-Edition/Wood/p/book/9781498728331)) by Simon Wood is an excellent resource for in-depth material on GAM theory, mathematical derivation, examples, and detailed descriptions on the computational aspect of _mgcv_.
+
+Finally, Noam Ross has assembled a list of online tools and resources for learning about and using GAMs in R, which cover a range of very useful videos, slides, and courses from different authors.
+The list can be found in Noam Ross's GitHub profile here:
+
+[Resources for Learning About and Using GAMs in R](https://github.com/noamross/gam-resources)

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_A.tex
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

@@
 al<-0.8
+
+thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
 \end{lstlisting}
@@
 
+\hypertarget{final-considerations}{%
+\subsection{Final considerations}\label{final-considerations}}
+
+In this Appendix we have shown a basic model selection workflow for GAMs.
+Our goal here is to familiarize the reader with the logic behind the construction of each model and the kind of diagnostic information that needs to be checked to ensure the model is adequate.
+One final consideration must be given to the conditional distribution of the response.
+In the main manuscript and through this workflow, we have assumed a normal distribution for the response, which we believe is appropriate in many biomedical scenarios.
+However, we must remind the reader that the normal distribution is not intended to be used as a ``one size fits all'' distribution.
+Depending on the type of data (counts, binary outcomes) the user can choose different conditional distributions in \emph{mgcv}.
+
+We also want to indicate that one of the major advantages of fitting GAMs in R using \emph{mgcv} is that once the appropriate conditional distribution has been chosen and model diagnostics are assessed, the user only needs to choose a basis dimension (the number of basis functions to use), and check if the basis dimension is adequately capturing the trend of the data using \passthrough{\lstinline!k.check!}.
+If that is not the case, then the user needs to increase \passthrough{\lstinline!k!} a bit and check \passthrough{\lstinline!k.check!} again.
+The process might seem mechanic, but we hope that with the theory presented in the main manuscript and the workflow of this Appendix the user has a good understanding of what the model is doing and the rationale for choosing a GAM as a tool for statistical analysis.
+
 \hypertarget{additional-resources}{%
@@
 
+Finally, Noam Ross has assembled a list of online tools and resources for learning about and using GAMs in R, which cover a range of very useful videos, slides, and courses from different authors.
+The list can be found in Noam Ross's GitHub profile here:
+
+\href{https://github.com/noamross/gam-resources}{Resources for Learning About and Using GAMs in R}
+
 \end{document}

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_B.Rmd
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

@@
 The next chunk presents the code used to create Figure 2 in the main manuscript.
-Here, we simulate data from the same concave down quadratic trend from Figure \@ref(fig:quadratic-cases-Appendix)A and use the package `mgcv` to fit a GAM to fit the trend of the data over time.
+Here, we simulate data from the same concave down quadratic trend from Figure \@ref(fig:quadratic-cases-Appendix)A and use the package _mgcv_ to fit a GAM to fit the trend of the data over time.
 The GAM in this case can be represented as
@@
  $$Response= \beta_0 + f(time_t|\beta_1)$$
- Because we only use the data from one group and we do not account for interaction. The model in `mgcv` is written as:
+ Because we only use the data from one group and we do not account for interaction. The model in _mgcv_ is written as:
 
@@
 Because we want to include the group means for plotting purposes, we extract the intercept (`const`) and add it to the Treatment group estimates in both `ci` and `si` order to shift the interval so it matches the scale of the response.
-Finally, we create a `ggplot2` object for plotting.
+Finally, we create a _ggplot2_ object for plotting.
 
@@
 In Appendix A we discuss the use of quantitative and graphical diagnostics to assess the goodness of fit of a GAM.
-The package `mgcv` has the function `gam.check` to provide such information, but its graphical diagnostics are made using base R graphics and therefore it is not straightforward to use them in conjunction with a `ggplot2` object.
+The package _mgcv_ has the function `gam.check` to provide such information, but its graphical diagnostics are made using base R graphics and therefore it is not straightforward to use them in conjunction with a _ggplot2_ object.
 Therefore we create the function `gam_diagnostics` that uses the same code from `gam.check` but without the graphical output.
-In this way, we can use `appraise` from the package `gratia` to create a graphical output in ggplot2 format and the quantitative information from `gam.check`.
+In this way, we can use `appraise` from the package _gratia_ to create a graphical output in _ggplot2_ format and the quantitative information from `gam.check`.
 

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_B.tex
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

@@
 ]{article}
-\title{SUPPLEMENTARY MATERIALS for\\
-\strut \\
-\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
-Beyond repeated measures ANOVA and Linear Mixed Models\\
-\strut \\
-APPENDIX B: CODE}
-\author{}
-\date{\vspace{-2.5em}}
-
 \usepackage{amsmath,amssymb}
@@
 
+\title{SUPPLEMENTARY MATERIALS for\\
+\strut \\
+\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
+Beyond repeated measures ANOVA and Linear Mixed Models\\
+\strut \\
+APPENDIX B: CODE}
+\author{}
+\date{\vspace{-2.5em}}
+
 \begin{document}
@@
 The next chunk presents the code used to create Figure 2 in the main manuscript.
-Here, we simulate data from the same concave down quadratic trend from Figure \ref{fig:quadratic-cases-Appendix}A and use the package \passthrough{\lstinline!mgcv!} to fit a GAM to fit the trend of the data over time.
+Here, we simulate data from the same concave down quadratic trend from Figure \ref{fig:quadratic-cases-Appendix}A and use the package \emph{mgcv} to fit a GAM to fit the trend of the data over time.
 The GAM in this case can be represented as
@@
 \[Response= \beta_0 + f(time_t|\beta_1)\]
-Because we only use the data from one group and we do not account for interaction. The model in \passthrough{\lstinline!mgcv!} is written as:
+Because we only use the data from one group and we do not account for interaction. The model in \emph{mgcv} is written as:
 
@@
 Because we want to include the group means for plotting purposes, we extract the intercept (\passthrough{\lstinline!const!}) and add it to the Treatment group estimates in both \passthrough{\lstinline!ci!} and \passthrough{\lstinline!si!} order to shift the interval so it matches the scale of the response.
-Finally, we create a \passthrough{\lstinline!ggplot2!} object for plotting.
+Finally, we create a \emph{ggplot2} object for plotting.
 
@@
 In Appendix A we discuss the use of quantitative and graphical diagnostics to assess the goodness of fit of a GAM.
-The package \passthrough{\lstinline!mgcv!} has the function \passthrough{\lstinline!gam.check!} to provide such information, but its graphical diagnostics are made using base R graphics and therefore it is not straightforward to use them in conjunction with a \passthrough{\lstinline!ggplot2!} object.
+The package \emph{mgcv} has the function \passthrough{\lstinline!gam.check!} to provide such information, but its graphical diagnostics are made using base R graphics and therefore it is not straightforward to use them in conjunction with a \emph{ggplot2} object.
 Therefore we create the function \passthrough{\lstinline!gam\_diagnostics!} that uses the same code from \passthrough{\lstinline!gam.check!} but without the graphical output.
-In this way, we can use \passthrough{\lstinline!appraise!} from the package \passthrough{\lstinline!gratia!} to create a graphical output in ggplot2 format and the quantitative information from \passthrough{\lstinline!gam.check!}.
+In this way, we can use \passthrough{\lstinline!appraise!} from the package \emph{gratia} to create a graphical output in \emph{ggplot2} format and the quantitative information from \passthrough{\lstinline!gam.check!}.
 

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/03-GAM Theory.Rmd
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

@@
 The smooth term $f(time_{t}|\beta_{j})$ gives a different smooth response for each treatment.
-^[If the smooth term represented a linear relationship, then $f(time_{t}|\beta_{j})= \beta_2 \times time_t+\beta_3 \times time_t \times treatment_j$; however, in general, the smooth term is a more flexible function than a linear relationship, with parameter vectors $\beta_{j}$ for each treatment].
-A GAM version of a linear model can be written as
+^[If the smooth term represented a linear relationship, then $f(time_{t}|\beta_{j})= \beta_2 \times time_t+\beta_3 \times time_t \times treatment_j$; however, in general, the smooth term is a more flexible function than a linear relationship, with parameter vectors $\beta_{j}$ for each treatment] A GAM version of a linear model can be written as
 
@@
 As $\lambda$ increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
-In other words,a low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
+In other words, a low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.
 

### 2021-12-26 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/04-Longitudinal analysis with GAMs.Rmd
Updated scripts and made a file to compare CIs from our code and G. Simpson's approach

@@
 # mean shift for Treatment group
-const <- coef(gam_02)[2]
+Intercept<-coef(gam_02)[1]
+const <- coef(gam_02)[2]+Intercept
 
@@
 mutate(est = case_when(Group == "Treatment" ~ est + const,
-TRUE ~ est),
+TRUE ~ est+Intercept),
 lower = case_when(Group == "Treatment" ~ lower + const,
-TRUE ~ lower),
+TRUE ~ lower+Intercept),
 upper = case_when(Group == "Treatment" ~ upper + const,
-TRUE ~ upper))
+TRUE ~ upper+Intercept))
 
@@
 # mean shift for group 2
-const <- coef(mod_m1)[2]
+Intercept<-coef(mod_m1)[1]
+const <- coef(mod_m1)[2]+Intercept
 
@@
     scale_x_continuous(breaks=c(0,2,5,7,10))+
-    labs(y="Difference\n(Missing observations)")+
+    labs(y="Difference\n(Incomplete observations)")+
     theme_classic()+
@@
 **A**: Pairwise comparisons for the full dataset.
-**B**: Pairwise comparisons for the dataset with missing observations.
-Significant differences exist where the 95% empirical Bayesian credible interval does not cover 0.
+**B**: Pairwise comparisons for the dataset with incomplete observations.
+Significant differences exist where the 95% empirical Bayesian simultaneous CI does not cover 0.
 In both cases the effect of treatment is significant after day 3.
-In both cases, a simultaneous CI has been computed around the smooth difference.
+For the difference, we have included the means so the the value of the difference has direct correspondence with the scale of the response (Figure \@ref(fig:sim-smooth-plot)A).
 
@@
 
-Consider the calculation of pairwise differences for the smooths in Figure \@ref(fig:sim-smooth-plot)B, D.
+Consider the calculation of pairwise differences for the fitted smooths that appear in Figure \@ref(fig:sim-smooth-plot)B, D.
 Figure \@ref(fig:plot-pairwise-comp) shows the difference between each treatment group trend for the full and missing datasets with a simultaneous CI computed around the difference.

### 2021-12-27 Code and miscellaneous/Interval_comparisons.Rmd
Checking code as there is something going on with the CI computation

@@
     geom_path(aes(y=fit),lwd=2,show.legend = FALSE)+
-    geom_ribbon(aes(ymin = lwrS, ymax = uprS), alpha = 0.2,show.legend = FALSE) +
-    geom_ribbon(aes(ymin = lwrP, ymax = uprP), alpha = 0.2,show.legend = FALSE) +
+    geom_ribbon(aes(ymin = lwrS, ymax = uprS), alpha = 0.4,show.legend = FALSE) +
+    geom_ribbon(aes(ymin = lwrP, ymax = uprP), alpha = 0.7,show.legend = FALSE) +
     labs(y = "G. Simpson's code")+
    geom_hline(yintercept = 60,color="red")+
+    geom_hline(yintercept = 35.77,color="red")+
     geom_hline(yintercept = 20,color="red")+
     geom_hline(yintercept = 0,color="red")+
+    scale_x_continuous(breaks=seq(1,10,1))+
     theme_classic()+
@@
 
-ci <- confint(gam_02, parm = "s(Day)", partial_match = TRUE, type = "confidence")
+ci <- confint(gam_02, parm = "s(Day)", partial_match = TRUE, type ="confidence")
 ## simultaneous interval
-si <- confint(gam_02, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
+si <- confint(gam_02, parm = "s(Day)", type = "simultaneous", partial_match = TRUE,nsim=10000)
 
@@
     geom_hline(yintercept = 0,color="red")+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
+    geom_hline(yintercept = 35.77,color="red")+
+    scale_x_continuous(breaks=seq(1,10,1))+
      labs(y = "Our code")+
+      theme_classic()+
+        thm1
+
+```
+
+Additional test: use the approach we used initially to draw the confidence intervals, see if there's a difference:
+
+```{r}
+
+#creates a dataframe using the length of the covariates for the GAM
+gam_predict <- expand_grid(Group = factor(c("Control", "Treatment")),
+                         Day = seq(0, 10, by = 0.1),
+                         subject=factor(rep(1:10)))
+
+#adds the predictions to the grid and creates a confidence interval for GAM
+gam_predict<-gam_predict%>%
+    mutate(fit = predict(gam_02,gam_predict,se.fit = TRUE,type='response')$fit,
+           se.fit = predict(gam_02, gam_predict,se.fit = TRUE,type='response')$se.fit)
+
+#plot smooths and confidence interval for GAM
+f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
+  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),
+                   fill=Group
+                   ),
+              alpha=0.3,
+              data=gam_predict,
+            show.legend=FALSE,
+                inherit.aes=FALSE) +
+  geom_line(aes(y=fit,
+                color=Group),
+              size=1,data=gam_predict,
+              show.legend = FALSE)+
+    geom_hline(yintercept = 60,color="red")+
+    geom_hline(yintercept = 20,color="red")+
+    geom_hline(yintercept = 0,color="red")+
+    geom_hline(yintercept = 35.77,color="red")+
+  labs(y='previous method')+
+    scale_x_continuous(breaks=seq(1,10,1))+
       theme_classic()+
@@
 #plot both
-c+a+b+scale_y_continuous(breaks=c(-20,0,20,40,60))
+c+a+b+f3+scale_y_continuous(breaks=c(-20,0,20,40,60))&theme_bw()
+
+#plot the pointiwse intervals only
+
+ggplot()+
+geom_ribbon(aes(ymin = lwrP, ymax = uprP,x=Day,group=Group,fill='red'),
+            alpha = 0.1,
+            show.legend = FALSE,
+            data=pred) +
+    geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
+                   ymax=(fit + 2*se.fit),
+                   group=Group
+                   ),
+              alpha=0.1,
+              fill='green',
+              data=gam_predict,
+            show.legend=FALSE,
+            inherit.aes = FALSE)+
+    geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = 'yellow'),
+inherit.aes = FALSE, alpha = 0.8,
+show.legend=FALSE)
+
+ggplot()+
+    geom_ribbon(aes(ymin = lwrS, ymax = uprS,x=Day,group=Group),
+                fill='yellow',
+                alpha = 0.4,
+                show.legend = FALSE,data=pred) +
+geom_ribbon(data = si,
+mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth),
+fill ='red',
+inherit.aes = FALSE, alpha = 0.4,
+show.legend=FALSE)
 
 ```
+
+From the plots, this is apparent:
+
+The limits of the CI interval are the same in G.Simpson's code and in our original approach to the matter, they overlap perfectly. From our code however, the pointwise CI is slightly wider.
+
+For the simultaneous CIs, the CI computed by G. Simpson's code is wider than the one we obtain using the `confint` function. Why is this?
 

### 2021-12-28 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/04-Longitudinal analysis with GAMs.Rmd
Fixed error on gratia::vcov

@@
 #the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing)
+mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5),
+              method='REML',
+              data  = dat_missing)
 #appraise the model

### 2021-12-29 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_A.Rmd
revision to appendix

@@
 For simplicity, the confidence intervals (CIs) shown in this section for the models are the across the function CIs created by `mgcv` by default.
-However, for the pairwise comparisons of the third model we use _simultaneous intervals_ as in the main manuscript, and the code for creating simultaneous CIs for the smooths can be found in **Appendix XXX**.
+However, for the pairwise comparisons of the third model we use _simultaneous intervals_ as in the main manuscript, and the code for creating simultaneous CIs for the smooths can be found in Appendix B.
 
 ## Setup and data simulation
+
+```{r, include = FALSE}
+knitr::opts_chunk$set(tidy = TRUE)
+```
 
@@
 #alpha for ribbon in the smooth plots
-al<-0.8
+al <- 0.8
 
-thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
+thm1 <- scale_fill_scico_d(palette = "tokyo", begin = 0.3, end = 0.8,
+                         direction = -1, aesthetics = c("colour", "fill"))
 
@@
 ```{r,data, echo= TRUE}
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
+dat<-tibble(StO2  = c(4, 27, 3, 2, 0.5, 7,4, 50, 45, 56),
+            Day   = rep(c(0, 2, 5, 7, 10), times = 2),
+            Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
 ```
@@
 sd <- 10 #approximate sd from paper
-source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","simulate_data.R"))
+source(here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final",
+            "scripts", "simulate_data.R"))
 dat_sim <- simulate_data(dat, n, sd)
@@
 
-The first model fitted to the data is one that only accounts for different smooths by day.
-The model syntax specifies that `gam_00` is the object that will contain all the model information, and that the model attempts to explain changes in `StO2_sim` (simulated $\mbox{StO}_2$) using a smooth per `Day`.
-The model will use 5 basis functions (`k=5`) for the smooth.
+The first model fitted to the data is one that only assumes a single smooth over time shared by both the treatment and control groups The model syntax specifies that `gam_00` is the object that will contain all the model information, and that the model attempts to explain changes in `StO2_sim` (simulated $\mbox{StO}_2$) using a smooth per `Day`.
+The model will use four basis functions (`k = 5`) for the smooth.
 The smooth is constructed by default using thin plate regression splines.
@@
 gam_00<-gam(StO2_sim ~ s(Day, k = 5),
-            method='REML',
+            method   ='REML',
             data     = dat_sim)
@@
 
-```{r, first-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:first-GAM-caption)'}
+The following code chunk generates graphical checks for the fitted GAM model with a single shared smooth shared between the treatment and control groups.
 
-appr1<-appraise(gam_00)
-sm1<-draw(gam_00)
-visual_check<-appr1+sm1
+```{r, first-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center', message=FALSE, warning=FALSE,fig.cap='(ref:first-GAM-caption)'}
 
-visual_check+plot_layout(design=layout1)
+appr1 <- appraise(gam_00)
+sm1 <- draw(gam_00)
+visual_check <- appr1 + sm1
+
+visual_check + plot_layout(design = layout1)
 ```
@@
 ```{r,gam-00-diagnostics, echo=TRUE, include=TRUE}
-source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","gam_diagnostics.R"))
+source(here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final",
+            "scripts", "gam_diagnostics.R"))
 gam_diagnostics(gam_00)
@@
 
+From the output of `gam_diagnostics`, it can be seen that the `k-index` is 0.35, which indicates that the model is not capturing the variability in the data.
+The `edf` (effective degrees of freedom) is an indicator of the complexity of the smooth.
+Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
 And now obtain a summary of the fitted model, which is obtained by calling `summary`.
@@
 
-From the output of `gam_diagnostics`, it can be seen that the `k-index` is 0.35, which indicates that the model is not capturing the variability in the data.
-The `edf` (effective degrees of freedom) is an indicator of the complexity of the smooth.
-Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
-
 From the `summary` function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
 The link function is 'identity' as the model does not make any transformation on the predictors.
-The 'significance of smooth terms' _p-value_ indicates if each smooth is adding significance to the model.
+The 'significance of smooth terms' _p-value_ indicates if each smooth is significantly different from a constant mean 0 under the model.
 Here, the _p-value_ is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the 'deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to $\approx$ 12%.
+Finally, the 'deviance explained' indicates how much of the variation in the data the model is able to capture, which in this case corresponds to $\approx$ 12%.
 
@@
 
-gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5),
+gam_01<-gam(StO2_sim ~ s(Day, by=Group, k = 5),
             method ='REML',
@@
 
-source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","gam_diagnostics.R"))
 gam_diagnostics(gam_01)
@@
 #need to add figure number and caption
-appr2<-appraise(gam_01)
-sm2<-draw(gam_01)
+appr2 <- appraise(gam_01)
+sm2 <- draw(gam_01)
 
-visual_check2<-appr2+sm2
+visual_check2 <- appr2 + sm2
 
-visual_check2+plot_layout(design=layout1)
+visual_check2 + plot_layout(design = layout1)
 ```
@@
 gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
-            method='REML',
+            method ='REML',
             data   = dat_sim)
-
-source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","gam_diagnostics.R"))
 
@@
 
-By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag)) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From `gam_diagnostics`, the k-index is now at an acceptable value ($\approx$ **1.23**).
+By using `appraise()` and `draw` on this model (Figure  \@ref(fig:final-GAM-diag) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably close to an normal distribution, and the smooths are capturing the trend of the data within each group.
+From `gam_diagnostics`, the k-index is now at an acceptable value ($\approx$ 1.23).
 
@@
 
-appr3<-appraise(gam_02)
-sm3<-draw(gam_02)
+appr3 <- appraise(gam_02)
+sm3 <- draw(gam_02)
 
-visual_check3<-appr3+sm3
+visual_check3 <- appr3 + sm3
 
-visual_check3+plot_layout(design=layout2)
+visual_check3 + plot_layout(design = layout2)
 ```
@@
 
-AIC(gam_00,gam_01,gam_02)
+AIC(gam_00, gam_01, gam_02)
 
@@
 
-The output in this case is expected: model `gam_02` has a lower AIC **(710.54)** whereas the initial two models have higher AICs (**884.43** and **805.30**).
+The output in this case is expected: model `gam_02` has a lower AIC (710.54) whereas the initial two models have higher AICs (884.43 and 805.30).
 The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
@@
 
-We want to emphasize that for through the manuscript and for the comparisons on `gam_02`, we included the group means in order to keep the pairwise comparisons on the scale of the response.
+We want to emphasize that throught the manuscript and for the comparisons on `gam_02`, we included the group means in order to keep the pairwise comparisons on the scale of the response.
 By default, pairwise comparisons available in other software packages (such as in G.
@@
 
-source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","pointwise_comparisons.R"))
-source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","difference_smooths.R"))
-source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","pairwise_limits.R"))
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
+                  "scripts","pointwise_comparisons.R"))
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final",
+                  "scripts", "difference_smooths.R"))
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
+                  "scripts", "pairwise_limits.R"))
 #compute difference between smooths and calculate confidence interval: complete data
-diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
-                     unconditional = TRUE, frequentist = FALSE,
-                     n=100, partial_match = TRUE, nrep=10000)
-my_list<-pairwise_limits(diff_complete)
+diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", 
+                                    newdata = newdat, 
+                                    unconditional = TRUE,
+                                    frequentist = FALSE,
+                                    n = 100, partial_match = TRUE, 
+                                    nrep=10000)
+my_list <- pairwise_limits(diff_complete)
 ```
@@
 ```{r,pairwise-comp-workflow-B,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
-rib_col<-'#8D7D82' #color for ribbon for confidence interval
+rib_col <- '#8D7D82'     #color for ribbon for confidence interval
 control_rib <- '#875F79' #color for ribbon for control region
@@
 c1 <- ggplot() +
-  geom_line(data = diff_complete, aes(x = Day, y = diff),size=1, alpha=0.5) +
-  annotate("rect",
-              xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-              fill=control_rib,
-              alpha = 0.5,
-              )+
-  annotate("text",
-           x=1.5,
-           y=-18,
-           label="Control>Treatment",
-         size=6,
-         angle=90
-         )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill=treat_rib,
-             alpha = 0.5
-             ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=6,
-           angle=90
-           )+
-  geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s, ymax = upper_s),
+  geom_line(data = diff_complete, aes(x = Day, y = diff), 
+            size = 1, alpha = 0.5) +
+  annotate("rect", xmin = my_list$init1, xmax = my_list$final1,
+           ymin = -Inf, ymax = Inf, fill = control_rib,
+           alpha = 0.5) + 
+  annotate("text", x = 1.5, y = -18, label = "Control>Treatment",
+           size = 6, angle = 90) +
+  annotate("rect", xmin = my_list$init2, xmax = my_list$final2,
+           ymin = -Inf, ymax = Inf, fill = treat_rib,
+           alpha = 0.5) +
+  annotate("text", x = 6, y = -18, label = "Treatment>Control",
+           size = 6, angle = 90) +
+  geom_ribbon(data = diff_complete, 
+              aes(x = Day, ymin = lower_s, ymax = upper_s),
               alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
-  geom_hline(yintercept = 0, lty = 2, color = "red")+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    labs(y="Difference\n(Complete observations)")+
+  geom_hline(yintercept = 0, lty = 2, color = "red") +
+  scale_x_continuous(breaks = c(0, 2, 5, 7, 10)) +
+  labs(y = "Difference\n(Complete observations)") +
   theme_classic()+
-    theme(
-    axis.text=element_text(size=22))
+  theme(axis.text = element_text(size = 22))
 
@@
 If that is  not the case, then the user needs to increase `k` a bit and check `k.check` again.
-The process might seem mechanic, but we hope that with the theory presented in the main manuscript and the workflow of this Appendix the user has a good understanding of what the model is doing and the rationale for choosing a GAM as a tool for statistical analysis.
+The process might seem mechanistic, but we hope that with the theory presented in the main manuscript and the workflow of this Appendix the user has a good understanding of what the model is doing and the rationale for choosing a GAM as a tool for statistical analysis.
 

### 2021-12-29 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_A.tex
revision to appendix

@@
 \PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
 %
@@
 \usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
+\usepackage{ifxetex,ifluatex}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
   \usepackage[T1]{fontenc}
@@
   colorlinks=true,
-  linkcolor={Maroon},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={blue},
+  linkcolor=Maroon,
+  filecolor=Maroon,
+  citecolor=Blue,
+  urlcolor=blue,
   pdfcreator={LaTeX via pandoc}}
@@
 \usepackage{float}
-\ifLuaTeX
+\ifluatex
   \usepackage{selnolig}  % disable illegal ligatures
@@
 \title{SUPPLEMENTARY MATERIALS for\\
-\strut \\
+~\\
 \textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
 Beyond repeated measures ANOVA and Linear Mixed Models\\
-\strut \\
+~\\
 APPENDIX A: WORKFLOW FOR GAMs FOR BIOMEDICAL LONGITUDINAL DATA}
@@
 For simplicity, the confidence intervals (CIs) shown in this section for the models are the across the function CIs created by \passthrough{\lstinline!mgcv!} by default.
-However, for the pairwise comparisons of the third model we use \emph{simultaneous intervals} as in the main manuscript, and the code for creating simultaneous CIs for the smooths can be found in \textbf{Appendix XXX}.
+However, for the pairwise comparisons of the third model we use \emph{simultaneous intervals} as in the main manuscript, and the code for creating simultaneous CIs for the smooths can be found in Appendix B.
 
@@
 #alpha for ribbon in the smooth plots
-al<-0.8
+al <- 0.8
 
-thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
+thm1 <- scale_fill_scico_d(palette = "tokyo", begin = 0.3, end = 0.8,
+                         direction = -1, aesthetics = c("colour", "fill"))
 \end{lstlisting}
@@
 \begin{lstlisting}[language=R]
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
+dat<-tibble(StO2  = c(4, 27, 3, 2, 0.5, 7,4, 50, 45, 56),
+            Day   = rep(c(0, 2, 5, 7, 10), times = 2),
+            Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
 \end{lstlisting}
@@
 
-The first model fitted to the data is one that only accounts for different smooths by day.
-The model syntax specifies that \passthrough{\lstinline!gam\_00!} is the object that will contain all the model information, and that the model attempts to explain changes in \passthrough{\lstinline!StO2\_sim!} (simulated \(\mbox{StO}_2\)) using a smooth per \passthrough{\lstinline!Day!}.
-The model will use 5 basis functions (\passthrough{\lstinline!k=5!}) for the smooth.
+The first model fitted to the data is one that only assumes a single smooth over time shared by both the treatment and control groups The model syntax specifies that \passthrough{\lstinline!gam\_00!} is the object that will contain all the model information, and that the model attempts to explain changes in \passthrough{\lstinline!StO2\_sim!} (simulated \(\mbox{StO}_2\)) using a smooth per \passthrough{\lstinline!Day!}.
+The model will use four basis functions (\passthrough{\lstinline!k = 5!}) for the smooth.
 The smooth is constructed by default using thin plate regression splines.
@@
 \subsubsection{Graphical diagnostics}\label{graphical-diagnostics}}
+
+The following code chunk generates graphical checks for the fitted GAM model with a single shared smooth shared between the treatment and control groups.
+
+\begin{lstlisting}[language=R]
+appr1 <- appraise(gam_00)
+sm1 <- draw(gam_00)
+visual_check <- appr1 + sm1
+
+visual_check + plot_layout(design = layout1)
+\end{lstlisting}
 
@@
 \begin{lstlisting}[language=R]
-source(here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","gam_diagnostics.R"))
+source(here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final",
+            "scripts", "gam_diagnostics.R"))
 gam_diagnostics(gam_00)
@@
 
+From the output of \passthrough{\lstinline!gam\_diagnostics!}, it can be seen that the \passthrough{\lstinline!k-index!} is 0.35, which indicates that the model is not capturing the variability in the data.
+The \passthrough{\lstinline!edf!} (effective degrees of freedom) is an indicator of the complexity of the smooth.
+Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
 And now obtain a summary of the fitted model, which is obtained by calling \passthrough{\lstinline!summary!}.
@@
 
-From the output of \passthrough{\lstinline!gam\_diagnostics!}, it can be seen that the \passthrough{\lstinline!k-index!} is 0.35, which indicates that the model is not capturing the variability in the data.
-The \passthrough{\lstinline!edf!} (effective degrees of freedom) is an indicator of the complexity of the smooth.
-Here the complexity of the smooth is comparable to that of a 4th degree polynomial.
-
 From the \passthrough{\lstinline!summary!} function, information about the assumed distribution of the errors (Gaussian in this case) and the link function can be obtained.
 The link function is `identity' as the model does not make any transformation on the predictors.
-The `significance of smooth terms' \emph{p-value} indicates if each smooth is adding significance to the model.
+The `significance of smooth terms' \emph{p-value} indicates if each smooth is significantly different from a constant mean 0 under the model.
 Here, the \emph{p-value} is low but we have seen that there are issues with the model from the previous outputs.
-Finally, the `deviance explained' indicates how much of the data the model is able to capture, which in this case corresponds to \(\approx\) 12\%.
+Finally, the `deviance explained' indicates how much of the variation in the data the model is able to capture, which in this case corresponds to \(\approx\) 12\%.
 
@@
 \begin{lstlisting}[language=R]
-gam_01<-gam(StO2_sim ~ s(Day, by=Group,k = 5),
+gam_01<-gam(StO2_sim ~ s(Day, by=Group, k = 5),
             method ='REML',
@@
 ## full convergence after 9 iterations.
-## Gradient range [-5.875578e-07,1.45076e-08]
+## Gradient range [-5.875578e-07,1.450759e-08]
 ## (score 400.4803 & scale 164.8662).
@@
 
-By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag}) we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably distributed, and the smooths are capturing the trend of the data within each group.
-From \passthrough{\lstinline!gam\_diagnostics!}, the k-index is now at an acceptable value (\(\approx\) \textbf{1.23}).
+By using \passthrough{\lstinline!appraise()!} and \passthrough{\lstinline!draw!} on this model (Figure \ref{fig:final-GAM-diag} we see that the trend on the QQ plot has improved, the histogram of the residuals appears to be reasonably close to an normal distribution, and the smooths are capturing the trend of the data within each group.
+From \passthrough{\lstinline!gam\_diagnostics!}, the k-index is now at an acceptable value (\(\approx\) 1.23).
 
@@
 \begin{lstlisting}[language=R]
-AIC(gam_00,gam_01,gam_02)
+AIC(gam_00, gam_01, gam_02)
 \end{lstlisting}
@@
 
-The output in this case is expected: model \passthrough{\lstinline!gam\_02!} has a lower AIC \textbf{(710.54)} whereas the initial two models have higher AICs (\textbf{884.43} and \textbf{805.30}).
+The output in this case is expected: model \passthrough{\lstinline!gam\_02!} has a lower AIC (710.54) whereas the initial two models have higher AICs (884.43 and 805.30).
 The AIC should not be considered as the only estimator of model quality, instead to be used as complimentary information to the graphical diagnostics and model checks described above.
@@
 
-We want to emphasize that for through the manuscript and for the comparisons on \passthrough{\lstinline!gam\_02!}, we included the group means in order to keep the pairwise comparisons on the scale of the response.
+We want to emphasize that throught the manuscript and for the comparisons on \passthrough{\lstinline!gam\_02!}, we included the group means in order to keep the pairwise comparisons on the scale of the response.
 By default, pairwise comparisons available in other software packages (such as in G.
@@
 \begin{lstlisting}[language=R]
-source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","pointwise_comparisons.R"))
-source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","difference_smooths.R"))
-source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","pairwise_limits.R"))
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
+                  "scripts","pointwise_comparisons.R"))
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final",
+                  "scripts", "difference_smooths.R"))
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
+                  "scripts", "pairwise_limits.R"))
 #compute difference between smooths and calculate confidence interval: complete data
-diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
-                     unconditional = TRUE, frequentist = FALSE,
-                     n=100, partial_match = TRUE, nrep=10000)
-my_list<-pairwise_limits(diff_complete)
+diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", 
+                                    newdata = newdat, 
+                                    unconditional = TRUE,
+                                    frequentist = FALSE,
+                                    n = 100, partial_match = TRUE, 
+                                    nrep=10000)
+my_list <- pairwise_limits(diff_complete)
 \end{lstlisting}
@@
 \begin{lstlisting}[language=R]
-rib_col<-'#8D7D82' #color for ribbon for confidence interval
+rib_col <- '#8D7D82'     #color for ribbon for confidence interval
 control_rib <- '#875F79' #color for ribbon for control region
@@
 c1 <- ggplot() +
-  geom_line(data = diff_complete, aes(x = Day, y = diff),size=1, alpha=0.5) +
-  annotate("rect",
-              xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
-              fill=control_rib,
-              alpha = 0.5,
-              )+
-  annotate("text",
-           x=1.5,
-           y=-18,
-           label="Control>Treatment",
-         size=6,
-         angle=90
-         )+
-    annotate("rect",
-             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
-             fill=treat_rib,
-             alpha = 0.5
-             ) +
-  annotate("text",
-             x=6,
-             y=-18,
-             label="Treatment>Control",
-             size=6,
-           angle=90
-           )+
-  geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s, ymax = upper_s),
+  geom_line(data = diff_complete, aes(x = Day, y = diff), 
+            size = 1, alpha = 0.5) +
+  annotate("rect", xmin = my_list$init1, xmax = my_list$final1,
+           ymin = -Inf, ymax = Inf, fill = control_rib,
+           alpha = 0.5) + 
+  annotate("text", x = 1.5, y = -18, label = "Control>Treatment",
+           size = 6, angle = 90) +
+  annotate("rect", xmin = my_list$init2, xmax = my_list$final2,
+           ymin = -Inf, ymax = Inf, fill = treat_rib,
+           alpha = 0.5) +
+  annotate("text", x = 6, y = -18, label = "Treatment>Control",
+           size = 6, angle = 90) +
+  geom_ribbon(data = diff_complete, 
+              aes(x = Day, ymin = lower_s, ymax = upper_s),
               alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
-  geom_hline(yintercept = 0, lty = 2, color = "red")+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-    labs(y="Difference\n(Complete observations)")+
+  geom_hline(yintercept = 0, lty = 2, color = "red") +
+  scale_x_continuous(breaks = c(0, 2, 5, 7, 10)) +
+  labs(y = "Difference\n(Complete observations)") +
   theme_classic()+
-    theme(
-    axis.text=element_text(size=22))
+  theme(axis.text = element_text(size = 22))
 \end{lstlisting}
@@
 If that is not the case, then the user needs to increase \passthrough{\lstinline!k!} a bit and check \passthrough{\lstinline!k.check!} again.
-The process might seem mechanic, but we hope that with the theory presented in the main manuscript and the workflow of this Appendix the user has a good understanding of what the model is doing and the rationale for choosing a GAM as a tool for statistical analysis.
+The process might seem mechanistic, but we hope that with the theory presented in the main manuscript and the workflow of this Appendix the user has a good understanding of what the model is doing and the rationale for choosing a GAM as a tool for statistical analysis.
 

### 2021-12-29 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_B.Rmd
revision to appendix

@@
 
+```{r, echo = FALSE}
+knitr::opts_chunk$set(tidy = TRUE)
+```
+
 \newpage
@@
 #alpha for ribbon in the smooth plots
-al<-0.8
+al<-     0.8
 
-thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
+thm1 <- scale_fill_scico_d(palette = "tokyo", begin = 0.3, end = 0.8,
+                           direction = -1, aesthetics = c("colour", "fill"))
 
@@
 
-The first function is `example.R`, which allows to simulate linear and quadratic data in the same manner as in Section 3.5 in the main manuscript.
+The first function is `example()` which is in the file named `example.R` in the `scripts/` folder, which allows to simulate linear and quadratic data in the same manner as in Section 3.5 in the main manuscript.
 Both rm-ANOVA and LMEM with interaction are fitted to the data.
@@
   mu <- matrix(0, length(x), 2)
+
+  if (fun_type == "linear") {
     # linear response
-  if (fun_type == "linear") {
-    mu[, 1] <- - (0.25*x)+2  
+    mu[, 1] <- -(0.25*x)+2  
     mu[, 2] <- 0.25*x+2
@@
     # quadratic response (non-linear)
-
-    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
-    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
+    mu[, 1] <- -(0.25 * x^2) + 1.5 * x - 1.25
+    mu[, 2] <- (0.25 * x^2) - 1.5 * x + 1.25
   }
@@
 
-  #Create array to store the "errors" for each group at each timepoint. The "errors" are the 
-  #between-group variability in the response.
+  #Create array to store the "errors" for each group at each timepoint. The "errors" are the between-group variability in the response.
   errors <- array(0, dim = c(length(x), 2, 10))
@@
   #labeling y and errors  
-  dimnames(y) <- list(time = x, 
-                      treatment = 1:2, 
-                      subject = 1:10)
+  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
 
-  dimnames(errors) <- list(time = x, 
-                           treatment = 1:2, 
-                           subject = 1:10)
+  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
 
   #labeling the mean response
-  dimnames(mu) <- list(time = x, 
-                       treatment = 1:2)
+  dimnames(mu) <- list(time = x, treatment = 1:2)
 
   #convert y, mu and errors to  dataframes with time, treatment and subject columns
-  dat <- as.data.frame.table(y, 
-                             responseName = "y")
-  dat_errors <- as.data.frame.table(errors, 
-                                    responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, 
-                                responseName = "mu")
+  dat <- as.data.frame.table(y, responseName = "y")
+  dat_errors <- as.data.frame.table(errors, responseName = "errors")
+  dat_mu <- as.data.frame.table(mu, responseName = "mu")
 
   #join the dataframes to show mean response and errors per subject
-  dat <- left_join(dat, dat_errors, 
-                   by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, 
-                   by = c("time", "treatment"))
+  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
+  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
   #add time
@@
                  random = ~ 1 | subject,
-                 correlation = corCompSymm(form = ~ 1 | subject)
-  )
+                 correlation = corCompSymm(form = ~ 1 | subject))
 
   #create a prediction frame where the model can be used for plotting purposes
-  pred_dat <- expand.grid(
-    treatment = factor(1:2), 
-    time = unique(dat$time)
-  )
+  pred_dat <- expand.grid(treatment = factor(1:2), 
+                          time = unique(dat$time))
 
@@
     pred_dat  = pred_dat,
-    fit_anova=fit_anova,
+    fit_anova = fit_anova,
     fit_lme   = fit_lme
@@
 
-Function `plot_example.R` uses the output of `example.R` to show the fit of a rm-ANOVA and a LMEM.
+The function `plot_example()` from the file `plot_example.R` in the folder `scripts/` uses the output of `example.R` to show the fit of a rm-ANOVA and a LMEM.
 This function can be used to show an expanded version of Figure 1 in the main manuscript, presenting simulated data with correlated and uncorrelated errors and how the individual trends vary in each case.
@@
 plot_example <- function(sim_dat) {
-    txt<-20
+  txt <- 20
   p1 <- sim_dat$dat %>%
-        ggplot(aes(x = time,
-                   y = y,
-                   group = treatment,
-                   color = treatment)
-        ) +
-        geom_point(show.legend=FALSE) +
-        labs(y='response')+
-        geom_line(aes(x = time,
-                      y = mu,
-                      color = treatment),
-                  show.legend=FALSE) +
+    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
+    geom_point(show.legend = FALSE) +
+    labs(y = 'response')+
+    geom_line(aes(x = time, y = mu, color = treatment), 
+              show.legend = FALSE) +
     theme_classic() +
-        theme(plot.title = element_text(size = txt,
-                                        face = "bold"),
-              text=element_text(size=txt))+
+    theme(plot.title = element_text(size = txt, face = "bold"),
+          text=element_text(size = txt)) +
     thm1
@@
   p2 <- sim_dat$dat %>%
-        ggplot(aes(x = time,
-                   y = y,
-                   group = subject,
-                   color = treatment)
-        ) +
-        geom_line(aes(size = "Subjects"),
-                  show.legend = FALSE) +
+    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
+    geom_line(aes(size = "Subjects"), show.legend = FALSE) +
     # facet_wrap(~ treatment) +
-        geom_line(aes(x = time,
-                      y = mu,
-                      color = treatment,
+    geom_line(aes(x = time, y = mu, color = treatment, 
                   size = "Simulated Truth"),
-                  lty = 1,show.legend = FALSE) +
-        labs(y='response')+
-        scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-        theme_classic()+
-        theme(plot.title = element_text(size = txt,
-                                        face = "bold"),
-              text=element_text(size=txt))+
+              lty = 1, show.legend = FALSE) +
+    labs(y = 'response') +
+    scale_size_manual(name = "Type", 
+                      values = c("Subjects" = 0.5, "Simulated Truth" = 3)) +
+    theme_classic() +
+    theme(plot.title = element_text(size = txt, face = "bold"),
+          text = element_text(size = txt)) +
     thm1
@@
   p3 <- sim_dat$dat %>%
-        ggplot(aes(x = time,
-                   y = errors,
-                   group = subject,
-                   color = treatment)) +
-        geom_line(show.legend=FALSE) +
-        labs(y='errors')+
-        theme_classic()+
-        theme(plot.title = element_text(size = txt,
-                                        face = "bold"),
-              text=element_text(size=txt))+
+    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
+    geom_line(show.legend = FALSE) +
+    labs(y = 'errors') +
+    theme_classic() +
+    theme(plot.title = element_text(size = txt, face = "bold"),
+          text = element_text(size = txt))+
     thm1
@@
   #plot the model predictions for rm-ANOVA
-    p4 <- ggplot(sim_dat$dat,
-                 aes(x = time,
-                     y = y,
-                     color = treatment)) +
-        geom_point(show.legend=FALSE)+
-        labs(y='response')+
+  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
+    geom_point(show.legend = FALSE) +
+    labs(y = 'response')+
     geom_line(aes(y = predict(sim_dat$fit_anova), 
-                      group = subject, size = "Subjects"),show.legend = FALSE) +
+                  group = subject, 
+                  size = "Subjects"),
+              show.legend = FALSE) +
     geom_line(data = sim_dat$pred_dat,
@@
               show.legend=FALSE) +
-        guides(color = guide_legend(override.aes = list(size = 2)))+
+    guides(color = guide_legend(override.aes = list(size = 2))) +
     scale_size_manual(name = "Predictions",
@@
     theme_classic() +
-        theme(plot.title = element_text(size = txt,
-                                        face = "bold"),
-              text=element_text(size=txt))+
+    theme(plot.title = element_text(size = txt, face = "bold"),
+          text = element_text(size = txt)) +
     thm1
@@
   #plot the LMEM predictions
-    p5 <- ggplot(sim_dat$dat,
-                 aes(x = time,
-                     y = y,
-                     color = treatment)) +
+  p5 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
     geom_point()+
-        labs(y='response')+
+    labs(y = 'response')+
     geom_line(aes(y = predict(sim_dat$fit_lme),
@@
     theme_classic() +
-        theme(plot.title = element_text(size = txt,
-                                        face = "bold"),
-              text=element_text(size=txt))+
+    theme(plot.title = element_text(size = txt, face = "bold"),
+          text = element_text(size=txt))+
     thm1
@@
   if(option=='simple'){
-        return((p1+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
+    return((p1 + p4 + p5) + plot_layout(nrow = 1) + 
+             plot_annotation(tag_levels = 'A'))
   }
   else {
-    return((p1+p3+p2+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
+    return((p1 + p3 + p2 + p4 + p5) + plot_layout(nrow = 1) + 
+             plot_annotation(tag_levels = 'A'))
   }
-
 }
@@
 
-In this subsection, we use `example.R` and `plot_example.R` to create an expanded version of Figure 1 on the main manuscript.
-The difference here is that in the main manuscript `plot_example.R` uses `option='simple'` to create the plot, and therefore only shows the simulated data, and the rm-ANOVA and LMEM fit.
+In this subsection, we use the `example()` and `plot_example()` functions to create an expanded version of Figure 1 in the main manuscript.
+The difference here is that in the main manuscript `plot_example()` uses `option='simple'` to create the plot, and therefore only shows the simulated data, and the rm-ANOVA and LMEM fit to the simulated data with correlated errors.
 Here, we use the option `composite` to generate the additional panels for the uncorrelated errors and their respective rm-ANOVA and LMEM fits.
@@
 
-The chunk below calls both `example.R` and `plot_example_Appendix.R` to simulate data and create the composite plots.
+The chunk below sources both `example.R` and `plot_example_Appendix.R` files to simulate data and create the composite plots.
 
@@
 
-source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","example.R"))
-source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","plot_example.R"))
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
+                  "scripts", "example.R"))
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
+                  "scripts", "plot_example.R"))
 
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"),option='composite')
+A1 <- plot_example(example(fun_type = "linear", error_type = "correlated"),
+                 option = 'composite')
 
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"),option='composite')
+B1 <- plot_example(example(fun_type = "linear", error_type = "independent"),
+                 option = 'composite')
 
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"),option='composite')
+C1 <- plot_example(example(fun_type = "quadratic", error_type = "correlated"), 
+                 option = 'composite')
 
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"),option='composite')
+D1 <- plot_example(example(fun_type = "quadratic", error_type = "independent"), 
+                 option = 'composite')
 
@@
 # linear response, correlated errors (subject effect)
-A1/B1+plot_annotation(tag_levels = 'A')
+A1 / B1 + plot_annotation(tag_levels = 'A')
 ```
@@
 # linear response, correlated errors (subject effect)
-C1/D1+plot_annotation(tag_levels = 'A')
+C1 / D1 + plot_annotation(tag_levels = 'A')
 ```
@@
 
-From the model, we extract the original and weighed four basis functions (and the intercept), and add these last weighed basis functions to construct the smoother.
+From the model, we extract the original and weighted four basis functions (and the intercept), and add these last weighted basis functions to construct the smoother.
 
@@
 mu <- matrix(0, length(x), 2)
-mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25 #mean response
-mu[, 2] <- (0.25 * x^2) -1.5*x+1.25 #mean response
+mu[, 1] <- -(0.25 * x^2) + 1.5 * x - 1.25 #mean response
+mu[, 2] <- (0.25 * x^2) - 1.5 * x + 1.25 #mean response
 y <- array(0, dim = c(length(x), 2, 10))
@@
 #extract  "Group 1" to fit the GAM
-dat<-subset(dat,treatment==1)
+dat <- subset(dat, treatment==1)
 #keep just the response and timepoint columns
-dat<-dat[,c('y','time')]
+dat <- dat[,c('y', 'time')]
 
 #GAM model of time, 5 basis functions
-gm<-gam(y~s(time,k=5),data=dat, method= "REML")
+gm <- gam(y ~ s(time, k = 5), data = dat, method = "REML")
 
@@
 #will contain the smooths used to create  model 'gm'
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
+model_matrix <- as.data.frame(predict(gm, type = 'lpmatrix'))
 
-time<-c(1:6)
+time <- c(1:6)
 
-basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
+basis <- model_matrix[1:6, ] #extracting basis (because the values are repeated after every 6 rows)
 #basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-    pivot_longer(
-        cols=starts_with("s")
-    )%>%
+colnames(basis)[colnames(basis) == "(Intercept)"] <- "s(time).0"
+basis <- basis %>% #pivoting to long format
+  pivot_longer(cols = starts_with("s")) %>%
   arrange(name) #ordering
@@
 #length of dataframe to be created: number of basis by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
+ln <- 6 * (length(coef(gm)))
 
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
+basis_plot <- data.frame(Basis = integer(ln),
+                         value_orig = double(ln),
+                         time = integer(ln),
+                         coef = double(ln))
 
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-    mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be
+basis_plot$time <- rep(time) #pasting timepoints
+basis_plot$Basis <- factor(rep(c(1:5), each = 6)) #pasting basis number values
+basis_plot$value_orig <- basis$value #pasting basis values
+basis_plot$coef <- rep(coef(gm)[1:5], each = 6) #pasting coefficients
+basis_plot <- basis_plot %>%
+  mutate(mod_val = value_orig * coef) #the create the predicted values the bases need to be
 #multiplied by the coefficients
@@
 
-basis_names<-c(
-    `1`="Intercept",
-    `2`="1",
-    `3`="2",
-    `4`="3",
-    `5`="4"
-)
+basis_names <- c(`1`="Intercept", `2`="1", `3`="2", `4`="3", `5`="4")
 
@@
 
-smooth<-basis_plot%>%
-    group_by(time)%>%
-    summarize(smooth=sum(mod_val))
+smooth <- basis_plot %>%
+  group_by(time) %>%
+  summarize(smooth = sum(mod_val))
 
 #original basis
-sz<-1
-p11<-ggplot(basis_plot,
-            aes(x=time,
-                y=value_orig,
-                colour=as.factor(Basis)
-            )
-)+
-    geom_line(size=sz,
-              show.legend=FALSE)+
-    geom_point(size=sz+1,
-               show.legend = FALSE)+
-    labs(y='Basis functions')+
-    facet_wrap(~Basis,
-               labeller = as_labeller(basis_names)
-    )+
-    theme_classic()+
+sz <- 1
+p11 <- ggplot(basis_plot,
+              aes(x = time, y = value_orig, colour = as.factor(Basis))) +
+  geom_line(size=sz, show.legend = FALSE) +
+  geom_point(size = sz + 1, show.legend = FALSE) +
+  labs(y = 'Basis functions') +
+  facet_wrap(~ Basis, labeller = as_labeller(basis_names)) +
+  theme_classic() +
   thm1
@@
 #penalized basis
-p12<-ggplot(basis_plot,
-            aes(x=time,
-                y=mod_val,
-                colour=as.factor(Basis)
-            )
-)+
-    geom_line(show.legend = FALSE,
-              size=sz)+
-    geom_point(show.legend = FALSE,
-               size=sz+1)+
-    labs(y='Penalized \n basis functions')+
-    scale_y_continuous(breaks=seq(-1,1,1))+
-    facet_wrap(~Basis,
-               labeller=as_labeller(basis_names)
-    )+
-    theme_classic()+
+p12 <- ggplot(basis_plot,
+              aes(x = time, y = mod_val, colour = as.factor(Basis))) +
+  geom_line(show.legend = FALSE, size = sz) +
+  geom_point(show.legend = FALSE, size = sz + 1) +
+  labs(y = 'Penalized \n basis functions') +
+  scale_y_continuous(breaks = seq(-1, 1, 1)) +
+  facet_wrap(~ Basis, labeller = as_labeller(basis_names)) +
+  theme_classic() +
   thm1
@@
 #heatmap of the  coefficients
-x_labels<-c("Intercept","1","2","3","4")
-p13<-ggplot(basis_plot,
-            aes(x=Basis,
-                y=Basis))+
-    geom_tile(aes(fill = cof),
-              colour = "black") +
-    scale_fill_gradient(low = "white",
-                        high = "#B50A2AFF")+
-    labs(x='Basis',
-         y='Basis')+
-    scale_x_discrete(labels=x_labels)+
-    geom_text(aes(label=round(cof,2)),
-              size=7,
-              show.legend = FALSE)+
+x_labels <- c("Intercept", "1", "2", "3", "4")
+p13 <- ggplot(basis_plot,
+              aes(x = Basis, y = Basis)) +
+  geom_tile(aes(fill = coef), colour = "black") +
+  scale_fill_gradient(low = "white", high = "#B50A2AFF") +
+  labs(x = 'Basis', y = 'Basis') +
+  scale_x_discrete(labels = x_labels) +
+  geom_text(aes(label = round(coef, 2)),
+            size = 7, show.legend = FALSE) +
   theme_classic()+
@@
 #plotting simulated datapoints and smooth term
-p14<-ggplot(data=dat,
-            aes(x=time,y=y))+
-    geom_point(size=sz+1,alpha=0.5)+
-    thm1+
-    labs(y='Simulated \n response')+
-    geom_line(data=smooth,
-              aes(x=time,
-                  y=smooth),
-              color="#6C581DFF",
-              size=sz+1)+
+p14<-ggplot(data = dat,
+            aes(x = time, y = y)) +
+  geom_point(size = sz + 1, alpha = 0.5) +
+  thm1 +
+  labs(y = 'Simulated \n response') +
+  geom_line(data = smooth, aes(x = time, y = smooth),
+            color = "#6C581DFF", size = sz + 1)+
   theme_classic()
@@
 #Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-    theme(
-        text=element_text(size=18)
-    )
+b_plot <- p11 + p13 + p12 + p14 + plot_annotation(tag_levels = 'A') &
+  theme(text = element_text(size=18))
 
@@
 
-The next chunk presents the code of the function `simulate_data.R`, which we use in the main manuscript to simulate data that follows the reported trends in $\mbox{StO}_2$ in the literature.
+The next chunk presents the code of the function `simulate_data()` from the file `simulate_data.R` in the `scripts/` folder, which we use in the main manuscript to simulate data that follows the reported trends in $\mbox{StO}_2$ in the literature.
 Although the default standard deviation (SD) to generate the normally-distributed data in the function is 10%, we use a standard deviation of 10% in the main manuscript.
@@
 #data from Vishwanath et. al.
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
+dat <- tibble(StO2 = c(4, 27, 3, 2, 0.5, 7, 4, 50, 45, 56),
+              Day = rep(c(0, 2, 5, 7, 10), times = 2),
+              Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
 ```
@@
       StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-            subject=rep(1:10),
-            subject=factor(paste(subject, Group, sep = "-"))
+      subject = rep(1:10),
+      subject = factor(paste(subject, Group, sep = "-"))
     ) %>%
@@
 
-We do not use a custom function in this case but instead extract the pointwise and simultaneous CIs after fitting the GAM by using the function `confint` from base R.
+We do not use a custom function in this case but instead extract the pointwise and simultaneous CIs after fitting the GAM by using the function `confint` methods from the _gratia_ package.
 The boundaries for the pairwise and simultaneous CIs are obtained by using the type `"confindence"` or `"simultaneous"` in `confint` and are stored in `ci` and `si` respectively.
@@
 #the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing)
+mod_m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5), 
+              data  = dat_missing)
 #appraise the model
@@
 
-ci <- confint(mod_m1, parm = "s(Day)", partial_match = TRUE, type = "confidence")
+ci <- confint(mod_m1, parm = "s(Day)", partial_match = TRUE,
+              type = "confidence")
 ## simultaneous interval
-si <- confint(mod_m1, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
+si <- confint(mod_m1, parm = "s(Day)", type = "simultaneous",
+              partial_match = TRUE)
 
@@
 
-f6<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
+f6 <- ggplot(ci, aes(x = Day, y = est, group = smooth)) +
   geom_line(lwd = 1) +
-geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
-inherit.aes = FALSE, alpha = 0.7,
-show.legend=FALSE) +
+  geom_ribbon(data = ci, 
+              aes(ymin = lower, ymax = upper, x = Day, 
+                  group = smooth, fill = Group),
+              inherit.aes = FALSE, alpha = 0.7, show.legend = FALSE) +
   geom_ribbon(data = si,
-mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
-inherit.aes = FALSE, alpha = 0.4,
-show.legend=TRUE)+
-    geom_line(data=si,aes(Day,upper,color=Group), size=0.8, alpha=0.7)+
-    geom_line(data=si,aes(Day,lower,color=Group), size=0.8, alpha=0.7)+
-    geom_point(data=dat_missing, aes(x = Day, 
-                        y = StO2_sim, 
-                        color = Group), 
-                            size=1.5,
-                        alpha=0.6, 
-               inherit.aes = FALSE)+
-    labs(y=expression(atop(StO[2],'(incomplete observations)')))+
-    scale_x_continuous(breaks=c(0,2,5,7,10))+
-      theme_classic()+
-  theme(
-    axis.text=element_text(size=22))+
+              aes(ymin = lower, ymax = upper, x = Day, 
+                  group = smooth, fill =Group),
+              inherit.aes = FALSE, alpha = 0.4, show.legend = TRUE) +
+  geom_line(data = si, aes(Day, upper, color = Group), 
+            size = 0.8, alpha = 0.7) +
+  geom_point(data = dat_missing, aes(x = Day, y = StO2_sim, color=Group), 
+             size = 1.5, alpha = 0.6, inherit.aes = FALSE) +
+  labs(y = expression(atop(StO[2], '(incomplete observations)'))) +
+  scale_x_continuous(breaks = c(0, 2, 5, 7, 10)) +
+  theme_classic() +
+  theme(axis.text=element_text(size = 22)) +
   thm1
@@
 
-This subsection presents the code used in functions `pointwise_comparisons.R` and `difference_smooths.R`, which generate the pointwise and simultaenous CI for the estimated difference between the smooths.
-In essence, `pointwise_comparisons.R` computes the difference between the smooths and the pointwise CI.
-Then, `difference_comparisons.R` uses the output of `pointwise_comparisons.R` to estimate the simultaneous CI.
+This subsection presents the code used in functions `pointwise_comparisons()` and `difference_smooths()` from the files `pointwise_comparisons.R` and `difference_smooths.R` in the `scripts/` folder, which generate the across the function (pointwise) and simultaenous CI for the estimated difference between the smooths.
+In essence, `pointwise_comparisons()` computes the difference between the smooths and the pointwise CI.
+Then, `difference_comparisons()` uses the output of `pointwise_comparisons.R` to estimate the simultaneous CI.
 
@@
 
-difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
+difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, 
+                                 data, Xp, V, coefs, nrep = 1000) {
   ## make sure f1 and f2 are characters
@@
   ## columns of Xp associated with pair of smooths
-    c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
-    c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
+  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1),
+              cnames, fixed = TRUE)
+  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), 
+              cnames, fixed = TRUE)
   ## rows of Xp associated with pair of smooths
@@
 
-Then, `difference_smooths.R` takes the output from the previous function and uses it to compute the simultaneous CI.
+Then, `difference_smooths()` function from the file `difference_smooths.R` in the `scripts/` folder takes the output from the previous function and uses it to compute the simultaneous CI.
 We present the code below.
@@
 # for a pairwise comparison between two smooths
-difference_smooths <- function(model,
-                               smooth,
-                               n = 100,
-                               ci_level = 0.95,
-                               newdata = NULL,
-                               partial_match = TRUE,
-                               unconditional = FALSE,
-                               frequentist = FALSE,
-                               nrep = 10000,
-                               include_means = TRUE,
-                               ...) {
+difference_smooths <- function(model, smooth, n = 100, ci_level = 0.95, 
+                               newdata = NULL, partial_match = TRUE,
+                               unconditional = FALSE, frequentist = FALSE,
+                               nrep = 10000, include_means = TRUE, ...) {
   if (missing(smooth)) {
@@
     gratia:::check_user_select_smooths(smooths = S, select = smooth,
-                                           partial_match = partial_match)#,
-    # model_name = expr_label(substitute(object)))
+                                       partial_match = partial_match)
+
   sm_ids <- which(select)
@@
 
-    out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
-                smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
-                coefs = coefs, nrep = nrep)
+  out <- pmap(pairs, difference_pointwise, smooth = smooth, 
+              by_var = by_var, smooth_var = smooth_var, data = sm_data,
+              Xp = Xp, V = V, coefs = coefs, nrep = nrep)
   out <- bind_rows(out)
@@
 
-pairwise_limits<-function(dataframe){
+pairwise_limits <- function(dataframe) {
   #extract values where the lower limit of the ribbon is greater than zero
@@
 
-    v1<-dataframe%>%
-        filter(lower_s>0)%>%
+  v1 <- dataframe %>%
+    filter(lower_s > 0) %>%
     select(Day)
   #get day  initial value
-    init1=v1$Day[[1]]
+  init1 = v1$Day[[1]]
   #get day final value
-    final1=v1$Day[[nrow(v1)]]
+  final1 = v1$Day[[nrow(v1)]]
   #extract values where the value of the upper limit of the ribbon is lower than zero
   #this corresponds to the region where the treatment group effect is greater
-    v2<-dataframe%>%
-        filter(upper_s<0)%>%
+  v2 <- dataframe %>%
+    filter(upper_s < 0) %>%
     select(Day)
 
-    init2=v2$Day[[1]]
-    final2=v2$Day[[nrow(v2)]]
+  init2 = v2$Day[[1]]
+  final2 = v2$Day[[nrow(v2)]]
   #store values
-    my_list<-list(init1=init1,
-                  final1=final1,
-                  init2=init2,
-                  final2=final2)
+  my_list <- list(init1 = init1,
+                  final1 = final1,
+                  init2 = init2,
+                  final2 = final2)
   return(my_list)
@@
 #Function that uses the source code of `gam.check` to obtain the estimates without the plots. The source can be checked by typing `gam.check` in the console.
-gam_diagnostics<-function (b, old.style = FALSE, type = c("deviance", "pearson",
-                                                          "response"), k.sample = 5000, k.rep = 200, rep = 0,
-                           level = 0.9, rl.col = 2, rep.col = "gray80", ...)
-{
+gam_diagnostics <- function (b, old.style = FALSE, 
+                           type = c("deviance", "pearson", "response"), 
+                           k.sample = 5000, k.rep = 200, rep = 0,
+                           level = 0.9, rl.col = 2, rep.col = "gray80", 
+                           ...) {
   type <- match.arg(type)
@@
   if (gamm) {
-        cat("\n'gamm' based fit - care required with interpretation.")
-        cat("\nChecks based on working residuals may be misleading.")
+    message("\n'gamm' based fit - care required with interpretation.")
+    message("\nChecks based on working residuals may be misleading.")
   }
   else {
-        cat("\nMethod:", b$method, "  Optimizer:",
-            b$optimizer)
+    message("\nMethod:", b$method, "  Optimizer:", b$optimizer)
     if (!is.null(b$outer.info)) {
@@
         boi <- b$outer.info
-                cat("\n", boi$conv, " after ", boi$iter,
-                    " iteration", sep = "")
+        message("\n", boi$conv, " after ", boi$iter, " iteration", sep = "")
         if (boi$iter == 1)
-                    cat(".")
-                else cat("s.")
-                cat("\nGradient range [", min(boi$grad),
-                    ",", max(boi$grad), "]", sep = "")
-                cat("\n(score ", b$gcv.ubre, " & scale ",
-                    b$sig2, ").", sep = "")
+          message(".")
+        else message("s.")
+        message("\nGradient range [", min(boi$grad), ",", max(boi$grad), "]", sep = "")
+        message("\n(score ", b$gcv.ubre, " & scale ", b$sig2, ").", sep = "")
         ev <- eigen(boi$hess)$values
         if (min(ev) > 0)
-                    cat("\nHessian positive definite, ")
-                else cat("\n")
-                cat("eigenvalue range [", min(ev), ",",
-                    max(ev), "].\n", sep = "")
+          message("\nHessian positive definite, ")
+        else message("\n")
+        message("eigenvalue range [", min(ev), ",", max(ev), "].\n", sep = "")
       }
       else {
-                cat("\n")
+        message("\n")
         print(b$outer.info)
@@
       if (length(b$sp) == 0)
-                cat("\nModel required no smoothing parameter selection")
+        message("\nModel required no smoothing parameter selection")
       else {
-                cat("\nSmoothing parameter selection converged after",
+        message("\nSmoothing parameter selection converged after",
             b$mgcv.conv$iter, "iteration")
         if (b$mgcv.conv$iter > 1)
-                    cat("s")
+          message("s")
         if (!b$mgcv.conv$fully.converged)
-                    cat(" by steepest\ndescent step failure.\n")
-                else cat(".\n")
-                cat("The RMS", b$method, "score gradient at convergence was",
+          message(" by steepest\ndescent step failure.\n")
+        else message(".\n")
+        message("The RMS", b$method, "score gradient at convergence was",
             b$mgcv.conv$rms.grad, ".\n")
         if (b$mgcv.conv$hess.pos.def)
-                    cat("The Hessian was positive definite.\n")
-                else cat("The Hessian was not positive definite.\n")
+          message("The Hessian was positive definite.\n")
+        else message("The Hessian was not positive definite.\n")
       }
@@
     if (!is.null(b$rank)) {
-            cat("Model rank = ", b$rank, "/", length(b$coefficients),
-                "\n")
+      message("Model rank = ", b$rank, "/", length(b$coefficients), "\n")
     }
   }
-    cat("\n")
+  message("\n")
   kchck <- k.check(b, subsample = k.sample, n.rep = k.rep)
   if (!is.null(kchck)) {
-        cat("Basis dimension (k) checking results. Low p-value (k-index<1) may\n")
-        cat("indicate that k is too low, especially if edf is close to k'.\n\n")
+    message("Basis dimension (k) checking results. Low p-value (k-index<1) may\n")
+    message("indicate that k is too low, especially if edf is close to k'.\n\n")
     printCoefmat(kchck, digits = 3)

### 2021-12-29 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_B.tex
revision to appendix

@@
 \title{SUPPLEMENTARY MATERIALS for\\
-\strut \\
+~\\
 \textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
 Beyond repeated measures ANOVA and Linear Mixed Models\\
-\strut \\
+~\\
 APPENDIX B: CODE}
@@
 #alpha for ribbon in the smooth plots
-al<-0.8
-
-thm1<-scale_fill_scico_d(palette="tokyo",begin=0.3, end=0.8, direction = -1, aesthetics = c("colour","fill"))
+al<-     0.8
+
+thm1 <- scale_fill_scico_d(palette = "tokyo", begin = 0.3, end = 0.8,
+                           direction = -1, aesthetics = c("colour", "fill"))
 \end{lstlisting}
@@
 
-The first function is \passthrough{\lstinline!example.R!}, which allows to simulate linear and quadratic data in the same manner as in Section 3.5 in the main manuscript.
+The first function is \passthrough{\lstinline!example()!} which is in the file named \passthrough{\lstinline!example.R!} in the \passthrough{\lstinline!scripts/!} folder, which allows to simulate linear and quadratic data in the same manner as in Section 3.5 in the main manuscript.
 Both rm-ANOVA and LMEM with interaction are fitted to the data.
@@
   #labeling y and errors  
-  dimnames(y) <- list(time = x, 
-                      treatment = 1:2, 
-                      subject = 1:10)
-
-  dimnames(errors) <- list(time = x, 
-                           treatment = 1:2, 
-                           subject = 1:10)
+  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
+
+  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
 
   #labeling the mean response
-  dimnames(mu) <- list(time = x, 
-                       treatment = 1:2)
+  dimnames(mu) <- list(time = x, treatment = 1:2)
 
   #convert y, mu and errors to  dataframes with time, treatment and subject columns
-  dat <- as.data.frame.table(y, 
-                             responseName = "y")
-  dat_errors <- as.data.frame.table(errors, 
-                                    responseName = "errors")
-  dat_mu <- as.data.frame.table(mu, 
-                                responseName = "mu")
+  dat <- as.data.frame.table(y, responseName = "y")
+  dat_errors <- as.data.frame.table(errors, responseName = "errors")
+  dat_mu <- as.data.frame.table(mu, responseName = "mu")
 
   #join the dataframes to show mean response and errors per subject
-  dat <- left_join(dat, dat_errors, 
-                   by = c("time", "treatment", "subject"))
-  dat <- left_join(dat, dat_mu, 
-                   by = c("time", "treatment"))
+  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
+  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
   #add time
@@
 
-Function \passthrough{\lstinline!plot\_example.R!} uses the output of \passthrough{\lstinline!example.R!} to show the fit of a rm-ANOVA and a LMEM.
+Function \passthrough{\lstinline!plot\_example()!} from the file \passthrough{\lstinline!plot\_example.R!} in the folder \passthrough{\lstinline!scripts/!} uses the output of \passthrough{\lstinline!example.R!} to show the fit of a rm-ANOVA and a LMEM.
 This function can be used to show an expanded version of Figure 1 in the main manuscript, presenting simulated data with correlated and uncorrelated errors and how the individual trends vary in each case.
@@
   p3 <- sim_dat$dat %>%
-        ggplot(aes(x = time,
-                   y = errors,
+    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
+    geom_line(show.legend = FALSE) +
+    labs(y = 'errors') +
+    theme_classic() +
+    theme(plot.title = element_text(size = txt, face = "bold"),
+          text = element_text(size = txt))+
+    thm1
+
+  #plot the model predictions for rm-ANOVA
+  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
+    geom_point(show.legend = FALSE) +
+    labs(y = 'response')+
+    geom_line(aes(y = predict(sim_dat$fit_anova), 
                   group = subject, 
-                   color = treatment)) +
-        geom_line(show.legend=FALSE) +
-        labs(y='errors')+
-        theme_classic()+
-        theme(plot.title = element_text(size = txt,
-                                        face = "bold"),
-              text=element_text(size=txt))+
-        thm1
-
-    #plot the model predictions for rm-ANOVA
-    p4 <- ggplot(sim_dat$dat,
-                 aes(x = time,
-                     y = y,
-                     color = treatment)) +
-        geom_point(show.legend=FALSE)+
-        labs(y='response')+
-        geom_line(aes(y = predict(sim_dat$fit_anova),
-                      group = subject, size = "Subjects"),show.legend = FALSE) +
+                  size = "Subjects"),
+              show.legend = FALSE) +
     geom_line(data = sim_dat$pred_dat,
@@
   if(option=='simple'){
-        return((p1+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
+    return((p1 + p4 + p5) + plot_layout(nrow = 1) + 
+             plot_annotation(tag_levels = 'A'))
   }
   else {
-    return((p1+p3+p2+p4+p5)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
-    }
-
+    return((p1 + p3 + p2 + p4 + p5) + plot_layout(nrow = 1) + 
+             plot_annotation(tag_levels = 'A'))
+  }
 }
@@
 
-The chunk below calls both \passthrough{\lstinline!example.R!} and \passthrough{\lstinline!plot\_example\_Appendix.R!} to simulate data and create the composite plots.
+The chunk below sources both \passthrough{\lstinline!example.R!} and \passthrough{\lstinline!plot\_example\_Appendix.R!} to simulate data and create the composite plots.
 
 \begin{lstlisting}[language=R]
-source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","example.R"))
-source(here::here("Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts","plot_example.R"))
-
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"),option='composite')
-
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"),option='composite')
-
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"),option='composite')
-
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"),option='composite') 
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
+                  "scripts", "example.R"))
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
+                  "scripts", "plot_example.R"))
+
+A1<-plot_example(example(fun_type = "linear", error_type = "correlated"),
+                 option = 'composite')
+
+B1<-plot_example(example(fun_type = "linear", error_type = "independent"),
+                 option = 'composite')
+
+C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"), 
+                 option = 'composite')
+
+D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"), 
+                 option = 'composite') 
 \end{lstlisting}
@@
 #will contain the smooths used to create  model 'gm'
-model_matrix<-as.data.frame(predict(gm,type='lpmatrix'))
-
-time<-c(1:6)
-
-basis<-model_matrix[1:6,] #extracting basis (because the values are repeated after every 6 rows)
+model_matrix <- as.data.frame(predict(gm, type = 'lpmatrix'))
+
+time <- c(1:6)
+
+basis <- model_matrix[1:6, ] #extracting basis (because the values are repeated after every 6 rows)
 #basis<-model_matrix[1:6,-1] #extracting basis
-colnames(basis)[colnames(basis)=="(Intercept)"]<-"s(time).0"
-basis<-basis %>% #pivoting to long format
-    pivot_longer(
-        cols=starts_with("s")
-    )%>%
+colnames(basis)[colnames(basis) == "(Intercept)"] <- "s(time).0"
+basis <- basis %>% #pivoting to long format
+  pivot_longer(cols = starts_with("s")) %>%
   arrange(name) #ordering
@@
 #length of dataframe to be created: number of basis by number of timepoints (minus 1 for the intercept that we won't plot)
-ln<-6*(length(coef(gm)))
-
-basis_plot<-data.frame(Basis=integer(ln),
-                       value_orig=double(ln),
-                       time=integer(ln),
-                       cof=double(ln)
-)
-
-basis_plot$time<-rep(time) #pasting timepoints
-basis_plot$Basis<-factor(rep(c(1:5),each=6)) #pasting basis number values
-basis_plot$value_orig<-basis$value #pasting basis values
-basis_plot$cof<-rep(coef(gm)[1:5],each=6) #pasting coefficients
-basis_plot<-basis_plot%>%
-    mutate(mod_val=value_orig*cof) #the create the predicted values the bases need to be
+ln <- 6 * (length(coef(gm)))
+
+basis_plot <- data.frame(Basis = integer(ln),
+                         value_orig = double(ln),
+                         time = integer(ln),
+                         coef = double(ln))
+
+basis_plot$time <- rep(time) #pasting timepoints
+basis_plot$Basis <- factor(rep(c(1:5), each = 6)) #pasting basis number values
+basis_plot$value_orig <- basis$value #pasting basis values
+basis_plot$coef <- rep(coef(gm)[1:5], each = 6) #pasting coefficients
+basis_plot <- basis_plot %>%
+  mutate(mod_val = value_orig * coef) #the create the predicted values the bases need to be
 #multiplied by the coefficients
@@
 #Combining all
-b_plot<-p11+p13+p12+p14+plot_annotation(tag_levels='A')&
-    theme(
-        text=element_text(size=18)
-    )
+b_plot <- p11 + p13 + p12 + p14 + plot_annotation(tag_levels = 'A') &
+  theme(text = element_text(size=18))
 \end{lstlisting}
@@
 
-The next chunk presents the code of the function \passthrough{\lstinline!simulate\_data.R!}, which we use in the main manuscript to simulate data that follows the reported trends in \(\mbox{StO}_2\) in the literature.
+The next chunk presents the code of the function \passthrough{\lstinline!simulate\_data()!} from the file \passthrough{\lstinline!simulate\_data.R!} in the \passthrough{\lstinline!scripts/!} folder, which we use in the main manuscript to simulate data that follows the reported trends in \(\mbox{StO}_2\) in the literature.
 Although the default standard deviation (SD) to generate the normally-distributed data in the function is 10\%, we use a standard deviation of 10\% in the main manuscript.
@@
 #data from Vishwanath et. al.
-dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
-            Day=rep(c(0,2,5,7,10),times=2),
-            Group=as.factor(rep(c("Control","Treatment"),each=5))
-            )
+dat <- tibble(StO2 = c(4, 27, 3, 2, 0.5, 7, 4, 50, 45, 56),
+              Day = rep(c(0, 2, 5, 7, 10), times = 2),
+              Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
 \end{lstlisting}
@@
 
-Then, \passthrough{\lstinline!difference\_smooths.R!} takes the output from the previous function and uses it to compute the simultaneous CI.
+Then, \passthrough{\lstinline!difference\_smooths()!} function from the file \passthrough{\lstinline!difference\_smooths.R!} in the \passthrough{\lstinline!scripts/!} folder takes the output from the previous function and uses it to compute the simultaneous CI.
 We present the code below.
@@
     if (!is.null(b$rank)) {
-            cat("Model rank = ", b$rank, "/", length(b$coefficients),
-                "\n")
-        }
-    }
-    cat("\n")
+      message("Model rank = ", b$rank, "/", length(b$coefficients), "\n")
+    }
+  }
+  message("\n")
   kchck <- k.check(b, subsample = k.sample, n.rep = k.rep)
   if (!is.null(kchck)) {
-        cat("Basis dimension (k) checking results. Low p-value (k-index<1) may\n")
-        cat("indicate that k is too low, especially if edf is close to k'.\n\n")
+    message("Basis dimension (k) checking results. Low p-value (k-index<1) may\n")
+    message("indicate that k is too low, especially if edf is close to k'.\n\n")
     printCoefmat(kchck, digits = 3)

### 2021-12-29 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_A.tex
Updated

@@
 \PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
 %
@@
 \usepackage{lmodern}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\usepackage{iftex}
+\ifPDFTeX
   \usepackage[T1]{fontenc}
@@
   colorlinks=true,
-  linkcolor=Maroon,
-  filecolor=Maroon,
-  citecolor=Blue,
-  urlcolor=blue,
+  linkcolor={Maroon},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={blue},
   pdfcreator={LaTeX via pandoc}}
@@
 \usepackage{float}
-\ifluatex
+\ifLuaTeX
   \usepackage{selnolig}  % disable illegal ligatures
@@
 \title{SUPPLEMENTARY MATERIALS for\\
-~\\
+\strut \\
 \textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
 Beyond repeated measures ANOVA and Linear Mixed Models\\
-~\\
+\strut \\
 APPENDIX A: WORKFLOW FOR GAMs FOR BIOMEDICAL LONGITUDINAL DATA}
@@
 ## full convergence after 9 iterations.
-## Gradient range [-5.875578e-07,1.450759e-08]
+## Gradient range [-5.875578e-07,1.45076e-08]
 ## (score 400.4803 & scale 164.8662).

### 2021-12-29 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_B.tex
Updated

@@
 \title{SUPPLEMENTARY MATERIALS for\\
-~\\
+\strut \\
 \textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
 Beyond repeated measures ANOVA and Linear Mixed Models\\
-~\\
+\strut \\
 APPENDIX B: CODE}
@@
 
-Function \passthrough{\lstinline!plot\_example()!} from the file \passthrough{\lstinline!plot\_example.R!} in the folder \passthrough{\lstinline!scripts/!} uses the output of \passthrough{\lstinline!example.R!} to show the fit of a rm-ANOVA and a LMEM.
+The function \passthrough{\lstinline!plot\_example()!} from the file \passthrough{\lstinline!plot\_example.R!} in the folder \passthrough{\lstinline!scripts/!} uses the output of \passthrough{\lstinline!example.R!} to show the fit of a rm-ANOVA and a LMEM.
 This function can be used to show an expanded version of Figure 1 in the main manuscript, presenting simulated data with correlated and uncorrelated errors and how the individual trends vary in each case.
@@
 
-In this subsection, we use \passthrough{\lstinline!example.R!} and \passthrough{\lstinline!plot\_example.R!} to create an expanded version of Figure 1 on the main manuscript.
-The difference here is that in the main manuscript \passthrough{\lstinline!plot\_example.R!} uses \passthrough{\lstinline!option='simple'!} to create the plot, and therefore only shows the simulated data, and the rm-ANOVA and LMEM fit.
+In this subsection, we use the \passthrough{\lstinline!example()!} and \passthrough{\lstinline!plot\_example()!} functions to create an expanded version of Figure 1 in the main manuscript.
+The difference here is that in the main manuscript \passthrough{\lstinline!plot\_example()!} uses \passthrough{\lstinline!option='simple'!} to create the plot, and therefore only shows the simulated data, and the rm-ANOVA and LMEM fit to the simulated data with correlated errors.
 Here, we use the option \passthrough{\lstinline!composite!} to generate the additional panels for the uncorrelated errors and their respective rm-ANOVA and LMEM fits.
@@
 
-The chunk below sources both \passthrough{\lstinline!example.R!} and \passthrough{\lstinline!plot\_example\_Appendix.R!} to simulate data and create the composite plots.
+The chunk below sources both \passthrough{\lstinline!example.R!} and \passthrough{\lstinline!plot\_example\_Appendix.R!} files to simulate data and create the composite plots.
 
@@
 
-A1<-plot_example(example(fun_type = "linear", error_type = "correlated"),
+A1 <- plot_example(example(fun_type = "linear", error_type = "correlated"),
                  option = 'composite')
 
-B1<-plot_example(example(fun_type = "linear", error_type = "independent"),
+B1 <- plot_example(example(fun_type = "linear", error_type = "independent"),
                  option = 'composite')
 
-C1<-plot_example(example(fun_type = "quadratic", error_type = "correlated"), 
+C1 <- plot_example(example(fun_type = "quadratic", error_type = "correlated"), 
                  option = 'composite')
 
-D1<-plot_example(example(fun_type = "quadratic", error_type = "independent"), 
+D1 <- plot_example(example(fun_type = "quadratic", error_type = "independent"), 
                  option = 'composite') 
@@
 
-We do not use a custom function in this case but instead extract the pointwise and simultaneous CIs after fitting the GAM by using the function \passthrough{\lstinline!confint!} from base R.
+We do not use a custom function in this case but instead extract the pointwise and simultaneous CIs after fitting the GAM by using the function \passthrough{\lstinline!confint!} methods from the \emph{gratia} package.
 The boundaries for the pairwise and simultaneous CIs are obtained by using the type \passthrough{\lstinline!"confindence"!} or \passthrough{\lstinline!"simultaneous"!} in \passthrough{\lstinline!confint!} and are stored in \passthrough{\lstinline!ci!} and \passthrough{\lstinline!si!} respectively.
@@
 
-This subsection presents the code used in functions \passthrough{\lstinline!pointwise\_comparisons.R!} and \passthrough{\lstinline!difference\_smooths.R!}, which generate the pointwise and simultaenous CI for the estimated difference between the smooths.
-In essence, \passthrough{\lstinline!pointwise\_comparisons.R!} computes the difference between the smooths and the pointwise CI.
-Then, \passthrough{\lstinline!difference\_comparisons.R!} uses the output of \passthrough{\lstinline!pointwise\_comparisons.R!} to estimate the simultaneous CI.
+This subsection presents the code used in functions \passthrough{\lstinline!pointwise\_comparisons()!} and \passthrough{\lstinline!difference\_smooths()!} from the files \passthrough{\lstinline!pointwise\_comparisons.R!} and \passthrough{\lstinline!difference\_smooths.R!} in the \passthrough{\lstinline!scripts/!} folder, which generate the across the function (pointwise) and simultaenous CI for the estimated difference between the smooths.
+In essence, \passthrough{\lstinline!pointwise\_comparisons()!} computes the difference between the smooths and the pointwise CI.
+Then, \passthrough{\lstinline!difference\_comparisons()!} uses the output of \passthrough{\lstinline!pointwise\_comparisons.R!} to estimate the simultaneous CI.
 

### 2021-12-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.Rmd
Updated manuscript, checked lines in response to reviewer's document

@@
     | 
-    | Response to Reviewer's Comments
+    | **Response to Reviewer's Comments**
     |
@@
 
-- L214: We have made clear that LMEMs can also work with missing observations.
+- L213: We have made clear that LMEMs can also work with missing observations.
 
-- L222-L225: Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates.
+- L221-L224: Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates.
 
@@
 
-- L97: We mention that LMEMs assume a linear relationship "by default", but we indicate that polynomial effects can be used as well, though they have their own shortcomings.
+- L95-L97: We mention that LMEMs assume a linear relationship "by default", but we indicate that polynomial effects can be used as well, though they have their own shortcomings.
 
-- L121-L122 now reads "the limitations of LMEMs and rm-ANOVA regarding an expected trend in the response", instead of "linearity of response" (L119 in the original manuscript).
+- L120-L121 now reads "the limitations of LMEMs and rm-ANOVA regarding an expected trend in the response", instead of "linearity of response" (L119 in the original manuscript).
 
-- L138 now reads "assumed relationship" instead of  "linear relationship" (L135 in the original manuscript).
+- L137 now reads "assumed relationship" instead of  "linear relationship" (L135 in the original manuscript).
 
@@
 
-- L181-L184 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
+- L180-L183 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
 
-- L248-L251: We provide additional context on the purpose of using linear models to fit quadratic data in  which now read "We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions."
+- L247-L250: We provide additional context on the purpose of using linear models to fit quadratic data in  which now read "We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with "default" (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions."
 
@@
 
-- L272-L274 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
+- L270-L273 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
 1.
@@
 
-- L552 In Section 6 (Discussion) we emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
+- L547-L549 In Section 6 (Discussion) emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
 
@@
 
-- L283- L328: Section 4 has been renamed to "Linear Models and Beyond" and the different Subsections cover LMs, GLMs, GLMMs and GAMs.
+- L282- L327: Section 4 has been renamed to "Linear Models and Beyond" and the different Subsections cover LMs, GLMs, GLMMs and GAMs.
 Because the different model classes are not commonly known in the biomedical field, we contextualize the previously presented rm-ANOVA and LMEM into their respective classes to provide the reader with a clear understanding of where each model fits (Section 4.1).
@@
 
-- L326: We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the **linear case**.
+- L325: We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the **linear case**.
 This change will avoid the propagation of confusion that Dr.
@@
 
-- L327-L328: We indicate that the term $\varepsilon_{ijt}$ is the deviation of each observation from the mean.
+- L326-L327: We indicate that the term $\varepsilon_{ijt}$ is the deviation of each observation from the mean.
 
@@
 
-- L351-L356: The information that originally appeared in L289-294 pertaining splines has been changed, as we now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
+- L350-L358: The information that originally appeared in L289-294 pertaining splines has been changed, as we now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
 Here, we no longer refer to TPRS as an "optimized version" but we indicate that they are a truncated version of thin plate splines (TPS).
@@
 The basis functions are not piece-wise polynomial and there are not “region[s] where a different set of basis functions will be used”.
-Indeed, the functions you show
-operate throughout the range of the covariate.
+Indeed, the functions you show operate throughout the range of the covariate.
 
@@
 
-- L358-L360: These lines correspond to L297-L300 on the original submission no longer refer to "knots" when referring to the construction of the smooth, now indicating that the number of _basis functions_ is what is specified when constructing the smoother.
-We believe that this will make sense to the reader  due to the information provided in L351-354.
+- L363-L366: These lines correspond to L297-L300 on the original submission no longer refer to "knots" when referring to the construction of the smooth, now indicating that the number of _basis functions_ is what is specified when constructing the smoother.
+We believe that this will make sense to the reader  due to the information provided in L350-353.
 
@@
 
-L369-377 has been added, and is a paragraph which briefly discusses wiggliness, the smoothness parameter $\lambda$ and the penalization process using the integrated square of the second derivative of the spline.
+L367-375 has been added, and is a paragraph which briefly discusses wiggliness, the smoothness parameter $\lambda$ and the penalization process using the integrated square of the second derivative of the spline.
 Our goal here is for the reader to understand the effect that $\lambda$ in order to prevent overfitting.
 
-- L376: We have substituted "weak" and "strong" for "low" and "high".
+- L374-L375: We have substituted "weak" and "strong" for "low" and "high".
 
@@
 
-- L395-397 now clarify this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
+- L395 now clarifies this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
 
 - The sentence "Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model", which appears in L319 in the original manuscript has been removed from the text.
-Instead, the concept of REML has been moved to Section 5.2 L458-463, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in _mgcv_.
+Instead, the concept of REML has been moved to Section 5.2 L456-459, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in _mgcv_.
 
@@
 
-- L400-L408 now contain a more detailed and accurate explanation of"across-the-function" CIs.
-We have used the reviewer's comments to indicate that in GAM fitted using _mgcv_, the close to nominal across-the-function-coverage is true when averaged over the entire function.
+- L398-L405 now contain a more detailed and accurate explanation of"across-the-function" CIs.
+We have used the reviewer's comments to indicate that in GAM fitted using _mgcv_, the close to nominal across-the-function coverage is true when averaged over the entire function.
 
-- L409-L414 now provide a brief explanation of simultaneous CIs.
+- L406-L412 now provide a brief explanation of simultaneous CIs.
 With this explanation we want to provide the reader with an intuitive understanding on the advantages of simultaneous CIs over across the function CIs when computing comparisons between different groups (which we discuss in L417-L419).
@@
 Fig.
-3B,D now show the fitted smooths and both intervals (across the function and simultaneous), and we explain the rationale for their inclusion in L480-L485.
+3B,D now show the fitted smooths and both intervals (across the function and simultaneous), and we explain the rationale for their inclusion in L477-L482.
 In this way, the reader has a visual understanding of the difference between across the function and simultaneous CIs.
@@
 Then you would not (at least by default) have any way to isolate the group means for the by factor independently of the levels of the other factor.
-You would have to create differences conditional upon a level of
-the other factor.
+You would have to create differences conditional upon a level of the other factor.
 Users, however, tend to expect differences to include the group means, as you have chosen to present things here.
@@
 - We have removed the sentence "if the difference between the credible intervals of the fitted smooths...
-is non-zero", and now L491-L492 now read: "is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference."
+is non-zero", and now L487-L489 now read: "is that the difference between them can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference."
 
-- L497 now reads "unlikely to be distinguishable from zero".
+- L494 now reads "unlikely to be distinguishable from zero".
 
-- L493-502 now indicate that we are computing a simultaneous CI around the difference, and that we _included_ the group means in our pairwise comparison.
+- L504-510 now indicate that we are computing a simultaneous CI around the difference, and that we _included_ the group means in our pairwise comparison.
 We also indicate to the reader the fact that if multiple parametric terms exist in the model the inclusion of the means can become problematic.
@@
 
-L431 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
+L429 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
 The computational requisite in the smooth estimation of having the maximum number of basis equal to the number of unique values in the covariate makes it impossible to fit a GAM with a k>5, as the error "A term has fewer unique covariate combinations than specified maximum degrees of freedom" is thrown by _mgcv_.
@@
 
-We have changed the expression to "data that shows non-linear trends". The change has also been applied to L316 which is now L393 in the revised manuscript.
+We have changed the expression to "data that shows non-linear trends". The change has also been applied to L316 which is now L391 in the revised manuscript.
 
@@
 
-- We have changed the sentence to "the distribution of the random effects, which need to be independent", which also indicates that errors need not be normally-distributed.
+- We have changed the sentence to "the distribution of the random effects, which need to be independent", which also indicates that errors need not be normally-distributed (L95 in the revised manuscript).
 
@@
 
-We have updated this line to "the mathematical space within which the true but unknown $f(x_t\mid \beta_j)$  is thought to exist", in L337 in the revised manuscript.
+We have updated this line to "the mathematical space within which the true but unknown $f(x_t\mid \beta_j)$  is thought to exist", in L339 in the revised manuscript.
 
@@
 However, we acknowledge that we didn't clearly conveyed this point to the read.
-In Section 6 (Discussion, L548-550 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided.
+In Section 6 (Discussion), L549-550 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided.
 
@@
 L351 Better to say “separate smooths” not “independent”.
-These factor-by smooths share a common basis, they just get an entirely separate set of model coeficients and smoothness parameters for each level of the factor.
+These factor-by smooths share a common basis, they just get an entirely separate set of model coefficients and smoothness parameters for each level of the factor.
 
@@
 
-The sentence now reads "separate smooths for..."
+The sentence now reads "separate smooths for..." (L441 in the revised manuscript)
 
@@
 
-The sentence now reads "any differences between the group means".
+The sentence now reads "any differences between the group means" (L444).
 
@@
 We use the code notation provided by the $LaTeX$ package `listings`, and it provides curly quotes for code.
-Regarding `family`, we believe that it is not necessary to include it here, but we indicate in L461-463 what the argument exists and that other distributions can be used.
+Regarding `family`, we believe that it is not necessary to include it here, but we indicate in L459-460 what the argument exists and that other distributions can be used.
 
@@
 
-The sentence now reads "contains the fitted model".
+The sentence now reads "contains the fitted model" (L449-L450).
 
@@
 
-The updated sentence now reads "using four basis functions (plus the intercept)".
+The updated sentence now reads "using four basis functions (plus the intercept)" (L451).
 
@@
 However, we want to make clear to the reader that other types of splines that might be appropriate in certain situations are available.
-To acknowledge this, L455 in the updated manuscript now reads "(a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console)".
+To acknowledge this, L453 in the updated manuscript now reads "(a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console)".
 
@@
 
-The word "total" has been removed. The sentence now reads "If 40% of the observations are randomly deleted..."
+The word "total" has been removed. The sentence now reads "consider the random deletion of 40% of the..." (L474)
 
@@
 
-The word "few" has been removed. The sentence now reads "...with as few as 4 observations per group..."
+The word "few" has been removed. The sentence now reads "...with as few as 4 observations per group..." (L482)
 
@@
 
-The sentence has been rewritten. It now reads "...the observed difference is significant..."
+The sentence has been rewritten. It now reads "...the observed difference is significant..." (L494).
 
@@
 
-The sentence in L497-L498 now reads "is unlikely to be distinguishable from zero".
+The sentence in L494-L495 now reads "is unlikely to be distinguishable from zero".
 
@@
 This was an unintentional mistake based on some previous simulation results.
-We have updated the paragraph, and it now reads "shows that the Control Group has higher $\mbox{StO}_2$ before day 3, and is able to estimate the change on day 3 where the Treatment Group becomes significant as the full dataset smooth pairwise comparison." The change acknowledges that Control was higher before day 3, and that the Treatment Group becomes significant after day 3.
+We have updated the paragraph, and it now reads "However, because the model is still able to pick the overall trend in $\mbox{StO}_2$, the pairwise comparison is able to estimate the change on day 3 where the mean difference between groups becomes statistically significant as in the full dataset smooth pairwise comparison" (L520-L522).
+The change acknowledges that Control was higher before day 3, and that the Treatment Group becomes significant after day 3 in relation to L513-L517.
 
@@
 
-We have made the appropriate changes. The sentence now reads "The model is therefore able to identify changes between the groups at time points..." (L562)
+We have made the appropriate changes. The sentence now reads "The GAM model is therefore able to identify changes between the groups at time points..." (L557)

### 2021-12-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.tex
Updated manuscript, checked lines in response to reviewer's document

@@
 \strut \\
-Response to Reviewer's Comments\\
+\textbf{Response to Reviewer's Comments}\\
 \strut \\
@@
 \item
-  L214: We have made clear that LMEMs can also work with missing observations.
+  L213: We have made clear that LMEMs can also work with missing observations.
 \item
-  L222-L225: Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates.
+  L221-L224: Emphasis has been provided to the fact that GAMs are not immune to missing data problems, and that researchers need to minimize missing observation rates.
 \end{itemize}
@@
 \item
-  L97: We mention that LMEMs assume a linear relationship ``by default'', but we indicate that polynomial effects can be used as well, though they have their own shortcomings.
+  L95-L97: We mention that LMEMs assume a linear relationship ``by default'', but we indicate that polynomial effects can be used as well, though they have their own shortcomings.
 \item
-  L121-L122 now reads ``the limitations of LMEMs and rm-ANOVA regarding an expected trend in the response'', instead of ``linearity of response'' (L119 in the original manuscript).
+  L120-L121 now reads ``the limitations of LMEMs and rm-ANOVA regarding an expected trend in the response'', instead of ``linearity of response'' (L119 in the original manuscript).
 \item
-  L138 now reads ``assumed relationship'' instead of ``linear relationship'' (L135 in the original manuscript).
+  L137 now reads ``assumed relationship'' instead of ``linear relationship'' (L135 in the original manuscript).
 \item
@@
 \item
-  L181-L184 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
+  L180-L183 now indicate that polynomial effects can be used in a LMEM, but that they have little predictive power and lack biological or mechanistic interpretation.
 \item
-  L248-L251: We provide additional context on the purpose of using linear models to fit quadratic data in which now read ``We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with''default'' (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.''
+  L247-L250: We provide additional context on the purpose of using linear models to fit quadratic data in which now read ``We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used to 1) present the consequences of modeling non-linear trends in data with a linear model such as rm-ANOVA or a LMEM with''default'' (linear) effects and, 2) demonstrate that, a visual assessment of model fit is an important tool that helps determine the validity of any statistical assumptions.''
 \end{itemize}
@@
 \item
-  L272-L274 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
+  L270-L273 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
 1.
@@
 \item
-  L552 In Section 6 (Discussion) we emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
+  L547-L549 In Section 6 (Discussion) emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
 \end{itemize}
@@
 \item
-  L283- L328: Section 4 has been renamed to ``Linear Models and Beyond'' and the different Subsections cover LMs, GLMs, GLMMs and GAMs.
+  L282- L327: Section 4 has been renamed to ``Linear Models and Beyond'' and the different Subsections cover LMs, GLMs, GLMMs and GAMs.
 Because the different model classes are not commonly known in the biomedical field, we contextualize the previously presented rm-ANOVA and LMEM into their respective classes to provide the reader with a clear understanding of where each model fits (Section 4.1).
@@
 \item
-  L326: We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the \textbf{linear case}.
+  L325: We also acknowledge that the GAM model (Equation 5 in the original manuscript and Equation 10 in the revised document) corresponds to the \textbf{linear case}.
 This change will avoid the propagation of confusion that Dr.~Simpson correctly indicated in his comment.
 \item
-  L327-L328: We indicate that the term \(\varepsilon_{ijt}\) is the deviation of each observation from the mean.
+  L326-L327: We indicate that the term \(\varepsilon_{ijt}\) is the deviation of each observation from the mean.
 \end{itemize}
@@
 \item
-  L351-L356: The information that originally appeared in L289-294 pertaining splines has been changed, as we now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
+  L350-L358: The information that originally appeared in L289-294 pertaining splines has been changed, as we now describe cubic splines (CS), thin plate splines (TPS) and thin plate regression splines (TPRS) in a manner that conveys their general properties and advantages (or disadvantages) to a non-Statistical reader.
 Here, we no longer refer to TPRS as an ``optimized version'' but we indicate that they are a truncated version of thin plate splines (TPS).
@@
 The basis functions are not piece-wise polynomial and there are not ``region{[}s{]} where a different set of basis functions will be used''.
-Indeed, the functions you show
-operate throughout the range of the covariate.
+Indeed, the functions you show operate throughout the range of the covariate.
 
@@
 \item
-  L358-L360: These lines correspond to L297-L300 on the original submission no longer refer to ``knots'' when referring to the construction of the smooth, now indicating that the number of \emph{basis functions} is what is specified when constructing the smoother.
-We believe that this will make sense to the reader due to the information provided in L351-354.
+  L363-L366: These lines correspond to L297-L300 on the original submission no longer refer to ``knots'' when referring to the construction of the smooth, now indicating that the number of \emph{basis functions} is what is specified when constructing the smoother.
+We believe that this will make sense to the reader due to the information provided in L350-353.
 \end{itemize}
@@
 
-L369-377 has been added, and is a paragraph which briefly discusses wiggliness, the smoothness parameter \(\lambda\) and the penalization process using the integrated square of the second derivative of the spline.
+L367-375 has been added, and is a paragraph which briefly discusses wiggliness, the smoothness parameter \(\lambda\) and the penalization process using the integrated square of the second derivative of the spline.
 Our goal here is for the reader to understand the effect that \(\lambda\) in order to prevent overfitting.
@@
 \item
-  L376: We have substituted ``weak'' and ``strong'' for ``low'' and ``high''.
+  L374-L375: We have substituted ``weak'' and ``strong'' for ``low'' and ``high''.
 \end{itemize}
@@
 \item
-  L395-397 now clarify this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
+  L395 now clarifies this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
 \item
   The sentence ``Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model'', which appears in L319 in the original manuscript has been removed from the text.
-Instead, the concept of REML has been moved to Section 5.2 L458-463, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in \emph{mgcv}.
+Instead, the concept of REML has been moved to Section 5.2 L456-459, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in \emph{mgcv}.
 \end{itemize}
@@
 \item
-  L400-L408 now contain a more detailed and accurate explanation of''across-the-function'' CIs.
-We have used the reviewer's comments to indicate that in GAM fitted using \emph{mgcv}, the close to nominal across-the-function-coverage is true when averaged over the entire function.
+  L398-L405 now contain a more detailed and accurate explanation of''across-the-function'' CIs.
+We have used the reviewer's comments to indicate that in GAM fitted using \emph{mgcv}, the close to nominal across-the-function coverage is true when averaged over the entire function.
 \item
-  L409-L414 now provide a brief explanation of simultaneous CIs.
+  L406-L412 now provide a brief explanation of simultaneous CIs.
 With this explanation we want to provide the reader with an intuitive understanding on the advantages of simultaneous CIs over across the function CIs when computing comparisons between different groups (which we discuss in L417-L419).
@@
 \item
-  We have removed the sentence ``if the difference between the credible intervals of the fitted smooths\ldots{} is non-zero'', and now L491-L492 now read: ``is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.''
+  We have removed the sentence ``if the difference between the credible intervals of the fitted smooths\ldots{} is non-zero'', and now L487-L489 now read: ``is that the difference between them can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.''
 \item
-  L497 now reads ``unlikely to be distinguishable from zero''.
+  L494 now reads ``unlikely to be distinguishable from zero''.
 \item
-  L493-502 now indicate that we are computing a simultaneous CI around the difference, and that we \emph{included} the group means in our pairwise comparison.
+  L504-510 now indicate that we are computing a simultaneous CI around the difference, and that we \emph{included} the group means in our pairwise comparison.
 We also indicate to the reader the fact that if multiple parametric terms exist in the model the inclusion of the means can become problematic.
@@
 
-L431 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
+L429 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
 The computational requisite in the smooth estimation of having the maximum number of basis equal to the number of unique values in the covariate makes it impossible to fit a GAM with a k\textgreater5, as the error ``A term has fewer unique covariate combinations than specified maximum degrees of freedom'' is thrown by \emph{mgcv}.
@@
 
-We have changed the expression to ``data that shows non-linear trends''. The change has also been applied to L316 which is now L393 in the revised manuscript.
+We have changed the expression to ``data that shows non-linear trends''. The change has also been applied to L316 which is now L391 in the revised manuscript.
 
@@
 \item
-  We have changed the sentence to ``the distribution of the random effects, which need to be independent'', which also indicates that errors need not be normally-distributed.
+  We have changed the sentence to ``the distribution of the random effects, which need to be independent'', which also indicates that errors need not be normally-distributed (L95 in the revised manuscript).
 \end{itemize}
@@
 
-We have updated this line to ``the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist'', in L337 in the revised manuscript.
+We have updated this line to ``the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist'', in L339 in the revised manuscript.
 
@@
 L351 Better to say ``separate smooths'' not ``independent''.
-These factor-by smooths share a common basis, they just get an entirely separate set of model coeficients and smoothness parameters for each level of the factor.
+These factor-by smooths share a common basis, they just get an entirely separate set of model coefficients and smoothness parameters for each level of the factor.
 
@@
 
-The sentence now reads ``separate smooths for\ldots{}''
+The sentence now reads ``separate smooths for\ldots{}'' (L441 in the revised manuscript)
 
@@
 
-The sentence now reads ``any differences between the group means''.
+The sentence now reads ``any differences between the group means'' (L444).
 
@@
 We use the code notation provided by the \(LaTeX\) package \passthrough{\lstinline!listings!}, and it provides curly quotes for code.
-Regarding \passthrough{\lstinline!family!}, we believe that it is not necessary to include it here, but we indicate in L461-463 what the argument exists and that other distributions can be used.
+Regarding \passthrough{\lstinline!family!}, we believe that it is not necessary to include it here, but we indicate in L459-460 what the argument exists and that other distributions can be used.
 
@@
 
-The sentence now reads ``contains the fitted model''.
+The sentence now reads ``contains the fitted model'' (L449-L450).
 
@@
 
-The updated sentence now reads ``using four basis functions (plus the intercept)''.
+The updated sentence now reads ``using four basis functions (plus the intercept)'' (L451).
 
@@
 However, we want to make clear to the reader that other types of splines that might be appropriate in certain situations are available.
-To acknowledge this, L455 in the updated manuscript now reads ``(a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console)''.
+To acknowledge this, L453 in the updated manuscript now reads ``(a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console)''.
 
@@
 
-The word ``total'' has been removed. The sentence now reads ``If 40\% of the observations are randomly deleted\ldots{}''
+The word ``total'' has been removed. The sentence now reads ``consider the random deletion of 40\% of the\ldots{}'' (L474)
 
@@
 
-The word ``few'' has been removed. The sentence now reads ``\ldots with as few as 4 observations per group\ldots{}''
+The word ``few'' has been removed. The sentence now reads ``\ldots with as few as 4 observations per group\ldots{}'' (L482)
 
@@
 
-The sentence has been rewritten. It now reads ``\ldots the observed difference is significant\ldots{}''
+The sentence has been rewritten. It now reads ``\ldots the observed difference is significant\ldots{}'' (L494).
 
@@
 
-The sentence in L497-L498 now reads ``is unlikely to be distinguishable from zero''.
+The sentence in L494-L495 now reads ``is unlikely to be distinguishable from zero''.
 
@@
 This was an unintentional mistake based on some previous simulation results.
-We have updated the paragraph, and it now reads ``shows that the Control Group has higher \(\mbox{StO}_2\) before day 3, and is able to estimate the change on day 3 where the Treatment Group becomes significant as the full dataset smooth pairwise comparison.'' The change acknowledges that Control was higher before day 3, and that the Treatment Group becomes significant after day 3.
+We have updated the paragraph, and it now reads ``However, because the model is still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the mean difference between groups becomes statistically significant as in the full dataset smooth pairwise comparison'' (L520-L522).
+The change acknowledges that Control was higher before day 3, and that the Treatment Group becomes significant after day 3 in relation to L513-L517.
 
@@
 
-We have made the appropriate changes. The sentence now reads ``The model is therefore able to identify changes between the groups at time points\ldots{}'' (L562)
+We have made the appropriate changes. The sentence now reads ``The GAM model is therefore able to identify changes between the groups at time points\ldots{}'' (L557)
 

### 2021-12-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated manuscript, checked lines in response to reviewer's document

@@
 
-The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs.
+The previous sections provided the basic understanding of the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs.
 This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using $\textsf{R}$ and the package `mgcv`.
@@
 
-This syntax specifies that `gam_02` (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using five basis functions.
+This syntax specifies that `gam_02` (named this way so it matches the model workflow from Appendix A) contains the fitted model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using four basis functions (plus intercept).
 The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths [@simpson2018] (a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console).
@@
 
-Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists.
-The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
+Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where unbalanced data exists.
+The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are incomplete.
 However, this usually causes the resulting smooths to have wider confidence intervals and less ability to discern differences in trends.
 To exemplify this, consider the random deletion of 40% of the simulated $\mbox{StO}_2$ values from Figure \@ref(fig:sim-smooth-plot)A.
-If the same interaction GAM (`gam_02`) is fitted to this data with missing observations, the resulting smooths appear in (Figure \@ref(fig:sim-smooth-plot)D).
+If the same interaction GAM (`gam_02`) is fitted to this unbalanced data, the resulting smooths appear in (Figure \@ref(fig:sim-smooth-plot)D).
 Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.
@@
 The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, they are wider than the across the function CI, resulting in the wide shaded regions.
-For the dataset with missing observations, the CIs for the smooths overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset.
+For the dataset with incomplete observations, the CIs for the smooths overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset.
 However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.
@@
 Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in `gam_02`) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that the difference between the separate smoothers per group (such as in `gam_02`) can be computed, followed by the estimation of a confidence interval around this difference.
 
@@
 
-The absence of a _p-value_ in this context might seem odd, but its interpretation can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment.
+The absence of a _p-value_ in this context might seem odd, but can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment.
 This is what happens for the simulated data in Figure \@ref(fig:sim-smooth-plot)A, where the chemotherapy causes $\mbox{StO}_2$ to increase over time.
@@
 A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
+Therefore, the computation of the difference between the smooths and the corresponding simultaneous empirical Bayesian CI around this difference is able to provide an estimation of when and by how much there is a significant difference between the different groups.
+Additionally, the correction provided by the simultaneous empirical Bayesian CI makes the estimation robust as we know that on average, the simultaneous CI will contain the _whole function_ at a nominal value (say, 95%).
 
-Consider the calculation of pairwise differences for the smooths in Figure \@ref(fig:sim-smooth-plot)B, D.
-Figure \@ref(fig:plot-pairwise-comp) shows the comparison between each treatment group for the full and missing datasets with a simultaneous CI computed around the difference.
+To visualize this, consider the calculation of pairwise differences for the smooths in Figure \@ref(fig:sim-smooth-plot)B, D.
+Figure \@ref(fig:plot-pairwise-comp) shows the comparison between each treatment group for the full and incomplete datasets with a simultaneous CI computed around the difference.
 Here, the  "Control" group is used as the reference to which "Treatment" group is being compared.
@@
 
-In Figure \@ref(fig:plot-pairwise-comp)A the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other.
+In Figure \@ref(fig:plot-pairwise-comp)A the shaded regions over the confidence interval (where the CI does not cover 0) indicate the time interval where each group has a higher effect than the other.
 Notice that the shaded region between days 0 and $\approx$ 2 for the full dataset indicates that through that time, the "Control" group has higher mean $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by $\approx$ 3 day  it is the "Treatment" group the one that on average, has greater $\mbox{StO}_2$.
@@
 
-On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that because the confidence intervals overlap for the first two days there is essentially no difference between the groups.
-However, because the model was still able to pick the overall trend in $\mbox{StO}_2$, the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes significant as in the full dataset smooth pairwise comparison.
+On the data with incomplete observations (Figure \@ref(fig:sim-smooth-plot)D), the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that because the confidence intervals overlap for the first two days there is no statistically significant difference between the groups.
+However, because the model was still able to pick the overall trend in $\mbox{StO}_2$, the pairwise comparison is able to estimate the change on day 3 where the mean difference between groups becomes significant as in the full dataset smooth pairwise comparison.
 

### 2021-12-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/05-Discussion.Rmd
Updated manuscript, checked lines in response to reviewer's document

--- 2021-11-08 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/05-Discussion.Rmd
@@
 
-Biomedical longitudinal data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
-Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \@ref(simulation), where it is clear that these models do not capture the non-linear trend of the data, thereby causing a "model misspecification error".
+Biomedical longitudinal data is particularly challenging to analyze due to the frequency of incomplete observations and different correlation structures in the data, which limit the use of rm-ANOVA.
+Although LMEMs can handle unbalanced observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \@ref(simulation).
+When these models do not capture the non-linear trend of the data, this results in a "model misspecification error".
+This "model misspecification" error, also is known as a "Type III" error [@dennis2019] is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
 
-This "model misspecification" error, also is known as a "Type III" error [@dennis2019] is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model.
-It could be argued that in Section \@ref(simulation) a LMEM with quadratic effects could have been used, but because in reality the true function in the data is not known, using polynomial effects ends up causing more questions than answers in this case (e.g., what is the biological interpretation of a quadratic, cubic, or other effect of even higher order?).
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) [@lang2015] but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a $R^2$ value.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines) [@lang2015] but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a $R^2$ or _p-value_ value, whose interpretation is not clear in the case of a Type III error.
 
 In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.  
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology) [@hefley2017; @pedersen2019] they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research [@hastie1995].
-This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \@ref(GAM-theory) is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (e.g., which are palaeoecology, geochemistry, and ecology) [@hefley2017; @pedersen2019] they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research [@hastie1995].
+This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs.
+T
+
+The purpose of Section \@ref(GAM-theory) is to demonstrate that at its core the principle  underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data.
+The use of basis functions is a major advantage over models where the user has to know the non-linear relationship _a priori_, such as in the case of polynomial effects in LMEMs.
+In contrast, because GAMs let the data speak for themselves they can provide an accurate representation of the effect of time in a biological process.
 
 However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
-We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature [@vishwanath2009] using $\textsf{R}$ and the package _mgcv_[@wood2017] in Section \@ref(longitudinal-GAMs), while a basic workflow for model selection is in the [Appendix](#workflow).
-One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a _p-value_, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure  \@ref(fig:plot-pairwise-comp) indicate.
-The model is therefore able to identify changes between the groups at time points were data was not directly measured even with missing data exists ( $\approx$ day 3 in Figure \@ref(fig:plot-pairwise-comp) A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
+We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature [@vishwanath2009] using R and the package _mgcv_[@wood2017] in Section \@ref(longitudinal-GAMs), while a basic workflow for model selection is in the [Appendix](#workflow).
 
-We have used $\textsf{R}$ as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research [@begley2015;@weissgerber2018], but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+One of the features of GAMs is that their Bayesian interpretation allows for inference about differences between groups without the need of a _p-value_, and in addition can provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure  \@ref(fig:plot-pairwise-comp) indicate.
+The GAM is therefore able to identify changes between the groups at time points where data was not directly measured even in the case of incomplete observations ( $\approx$ day 3 in Figure \@ref(fig:plot-pairwise-comp) A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
+
+We have used R as the software of choice for this paper because R provides a fully developed environment to fit GAMs, enables simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations), and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work more reproducible.
+In this regard, reproducibility is still an issue in biomedical research [@begley2015;@weissgerber2018], and is becoming apparent that what other disciplines have experienced in this aspect is likely to impact this field sooner rather than later.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.

### 2021-12-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/Full_document_SIM_No_Appendix.tex
Updated manuscript, checked lines in response to reviewer's document

@@
 
-In contrast, generalized additive models (GAMs) relax the assumptions made by LMEMs and rm-ANOVA, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures.
+In contrast, generalized additive models (GAMs) relax the assumptions made by LMEMs and rm-ANOVA, and allow the data to determine the fit of the model while permitting incomplete observations and different correlation structures.
 Therefore, GAMs present an excellent choice to analyze longitudinal data with non-linear trends in the context of biomedical research.
@@
 We present the basic theory of GAMs and using reported trends of oxygen saturation in tumors, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R.
-We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing).
+We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when incomplete observations exist (with 40\% of the simulated observations missing).
 To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
@@
 It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points.
-In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
-When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13}}.
+In each of these scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated.
+When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with incomplete observations from the analysis\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13}}.
 This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects.
@@
 Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age).
-When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
+When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate incomplete observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}}.
 However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}}.
@@
 This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas.
-First, the limitations of LMEMs and rm-ANOVA regarding an expected trend of the response, constant correlation structures, and missing observations are explained in detail.
+First, the limitations of LMEMs and rm-ANOVA regarding an expected trend of the response, constant correlation structures, and complete observations are explained in detail.
 Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations.
@@
 
-However, it is worth reminding that ``all models are wrong''\textsuperscript{\protect\hyperlink{ref-box1976}{49}} and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
+However, it is worth reminding that ``all models are wrong''\textsuperscript{\protect\hyperlink{ref-box1976}{49}} and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of incomplete data, such as sampling bias or a drastic reduction in statistical power.
 Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.
@@
 
-where \(\eta_{ijt}\) is the linear predictor for each observation \(i\) in each group \(j\), at each timepoint \(t\), and \(\beta_0\) (the intercept), \(\beta_{1}, \beta_2,\) and \(\beta_3\) are the model parameters for each group for the linear mean from Equation (\eqref{eq:linear-model}).
-Finally, \(time_{t}\) represents the covariates from each subject in each group at each time point.
+where \(\eta_{ijt}\) is the linear predictor for each observation \(i\) in each group \(j\), at each timepoint \(t\).
+Following the notation from Equation \eqref{eq:linear-model} the model parameters for each group are \(\beta_0\) (the intercept), \(\beta_{1}\), \(\beta_2\), and \(\beta_3\); \(time_{t}\) represents the covariates from each subject in each group at each time point, \(treatment_j\) represents the different treatment levels, and \(time_t \times treatment_j\) represents their interaction.
 
@@
 where \(E(y_{ijt})\) is the expectation, and \(g^{-1}\) is the inverse of a \emph{link function} (\(g\)).
-The link function transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}.
+The link function transforms the values from the response scale to the scale of the linear predictor \(\eta\) (Equation \eqref{eq:GLM-eta}).
 Therefore, it can be seen that LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
@@
 The use of CS as smoothers in GAMs was discussed within the original GAM framework\textsuperscript{\protect\hyperlink{ref-hastie1987}{56}}, but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots', the points where the basis functions meet) to obtain the finite basis, which affects model fit\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}.
-A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}.
-
+A solution to the knot placement limitation of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}.
 In contrast, thin plate regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an computationally efficient\textsuperscript{\protect\hyperlink{ref-wood2003}{59}}.
@@
 Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present.
-A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline.
-The second derivative of the spline is a measure of curvature, or the rate of change of the slope\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}}, and increasing the penalty by increasing \(\lambda\) results in models with less curvature.
+A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline smooth.
+The second derivative of the spline smooth is a measure of curvature, or the rate of change of the slope\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}}, and increasing the penalty by increasing \(\lambda\) results in models with less curvature.
 As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
@@
 
-With this in mind, if four basis functions (plus the intercept) are used to fit a GAM for the data of Group 1 (concave down function) that appears in Figure \ref{fig:l-q-response}D, the resulting fitting process is shown in Figure \ref{fig:basis-plot}.
+With this in mind, if four basis functions (plus the intercept) are used to fit a GAM for the data of Group 1 (concave down trend) that appears in Figure \ref{fig:l-q-response}D, the resulting fitting process is shown in Figure \ref{fig:basis-plot}.
 In Figure \ref{fig:basis-plot}A the four basis functions (and the intercept) are shown.
@@
 Figure \ref{fig:basis-plot}C shows the resulting weighed basis functions.
-Note that the magnitude of the weighing for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
+Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
 On the other hand, the third basis function has roughly doubled its contribution to the smooth term.
@@
 In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI.
-As a consequence, some areas of CI for the function must have more than nominal coverage and some areas less than the nominal coverage.
+As a consequence, some areas of the CI for the function have more than nominal coverage and some areas less than the nominal coverage.
 
@@
 In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller\textsuperscript{\protect\hyperlink{ref-miller2019}{61}}, Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}}, Marra\textsuperscript{\protect\hyperlink{ref-marra2012}{63}}, and Ruppert\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}.
-What we want to convey is that a Bayesian interpretation of GAMs allows for robust estimation using simultaneous empirical Bayesian CIs, as their stimates can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
+What we want to convey is that a Bayesian interpretation of GAMs allows for robust estimation using simultaneous empirical Bayesian CIs, as their estimates can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
 
@@
 
-The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs.
+The previous sections provided the basic understanding of the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs.
 This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}.
@@
 
-This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using five basis functions.
+This syntax specifies that \passthrough{\lstinline!gam\_02!} (named this way so it matches the model workflow from Appendix A) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using four basis functions (plus intercept).
 The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} (a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console).
@@
 
-Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists.
-The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
+Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where unbalanced data exists.
+The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are incomplete.
 However, this usually causes the resulting smooths to have wider confidence intervals and less ability to discern differences in trends.
 To exemplify this, consider the random deletion of 40\% of the simulated \(\mbox{StO}_2\) values from Figure \ref{fig:sim-smooth-plot}A.
-If the same interaction GAM (\passthrough{\lstinline!gam\_02!}) is fitted to this data with missing observations, the resulting smooths appear in (Figure \ref{fig:sim-smooth-plot}D).
+If the same interaction GAM (\passthrough{\lstinline!gam\_02!}) is fitted to this data with unbalanced observations, the resulting smooths appear in (Figure \ref{fig:sim-smooth-plot}D).
 Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.
@@
 Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in \passthrough{\lstinline!gam\_02!}) is that the difference between them can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that the difference between the separate smoothers per group (such as in \passthrough{\lstinline!gam\_02!}) can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.
 
@@
 
-The absence of a \emph{p-value} in this context might seem odd, but its interpretation of the comparison can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
-
-With this expectation of different trends in each group, computing the difference between the trends will identify if and \emph{when} the observed difference is significant.
+The absence of a \emph{p-value} in this context might seem odd, but the pairwise difference in smooths can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
+
+With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant.
 A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
-Therefore, the estimation of the difference between the smooths and the correspondent simultaneous empirical Bayesian CI of this difference provides an estimation of significance that is more robust that a mere \emph{p-value} because the estimated difference is able to indicates the significant time interval and we know that on average, the correction provided by the simultaneous empirical Bayesian CI ensures that it will contain \emph{the whole function} at a nominal value (say, 95\%).
-
-Consider the calculation of pairwise differences for the fitted smooths that appear in Figure \ref{fig:sim-smooth-plot}B, D.
+Therefore, the computation of both the difference between the smooths and the corresponding simultaneous empirical Bayesian CI around this difference is able to provide an estimation of when and by how much there is a significant difference between the different groups.
+Additionally, the correction provided by the simultaneous empirical Bayesian CI makes the estimation robust as we know that on average, the simultaneous CI will contain the \emph{whole function} at a nominal value (say, 95\%).
+
+To visualize this, consider the calculation of pairwise differences for the fitted smooths that appear in Figure \ref{fig:sim-smooth-plot}B, D.
 Figure \ref{fig:plot-pairwise-comp} shows the difference between each treatment group trend for the full and missing datasets with a simultaneous CI computed around the difference.
@@
 
-In Figure \ref{fig:plot-pairwise-comp}A, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a different effect than the other.
+In Figure \ref{fig:plot-pairwise-comp}A, the shaded regions over the confidence interval (where the CI does not cover 0) indicate the time interval where each group has a different mean effect than the other.
 Notice that the shaded region between days 1 and \(\approx\) 2 for the full dataset indicates that through that time, the ``Control'' group has higher mean \(\mbox{StO}_2\), but as therapy progresses the effect is reversed and by day \(\approx\) 3 it is the ``Treatment'' group that statistically on average has greater \(\mbox{StO}_2\).
@@
 
-On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that because the confidence intervals overlap zero for the first two days there is statistically significant no difference between the groups.
-However, because the model is still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes statistically significant as in the full dataset smooth pairwise comparison.
+On the data with missing observations (Figure \ref{fig:sim-smooth-plot}D), the smooth pairwise comparison (Figure \ref{fig:plot-pairwise-comp}B) shows that because the confidence intervals overlap zero for the first two days there is no statistically significant difference between the groups.
+However, because the model is still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the mean difference between groups becomes statistically significant as in the full dataset smooth pairwise comparison.
 
@@
 
-Biomedical longitudinal data is particularly challenging to analyze due to the frequency of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA.
-Although LMEMs can handle missing observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}.
+Biomedical longitudinal data is particularly challenging to analyze due to the frequency of incomplete observations and different correlation structures in the data, which limit the use of rm-ANOVA.
+Although LMEMs can handle unbalanced observations and different correlation structures, both LMEMs and rm-ANOVA yield biased estimates when they are used to fit data with non-linear trends as we have visually demonstrated in Section \ref{simulation}.
 When these models do not capture the non-linear trend of the data, this results in a ``model misspecification error''.
@@
 
-However, in Section \ref{GAM-theory} we demonstrated that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model, such as in the case of polynomial effects in LMEMs where additionally there is no biological interpretation for such effect.
+However, in Section \ref{GAM-theory} we demonstrated that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data.
+The use of basis functions is a major advantage over models where the user has to know the non-linear relationship \emph{a priori}, such as in the case of polynomial effects in LMEMs.
+In addition, there is no biological interpretation for such polynomial assumptions.
 In contrast, because GAMs let the data speak for themselves they can provide an accurate representation of the effect of time in a biological process.
@@
 One of the features of GAMs is that their Bayesian interpretation allows for inference about differences between groups without the need of a \emph{p-value}, thereby providing a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure \ref{fig:plot-pairwise-comp} indicate.
-The model is therefore able to identify changes between the groups at time points were data was not directly measured even in the case of incomplete observations ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
-
-We have used R as the software of choice for this paper because R not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible.
-In this regard, reproducibility is still an issue in biomedical research\textsuperscript{\protect\hyperlink{ref-begley2015}{68},\protect\hyperlink{ref-weissgerber2018}{69}}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field.
+The GAM is therefore able to identify changes between the groups at time points where data was not directly measured and in the case of incomplete observations ( \(\approx\) day 3 in Figure \ref{fig:plot-pairwise-comp} A, B ).
+This more nuanced inference can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
+
+We have used R as the software of choice for this paper because it provides a fully developed environment to fit GAMs, enables simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations), and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work more reproducible.
+In this regard, reproducibility is still an issue in biomedical research\textsuperscript{\protect\hyperlink{ref-begley2015}{68},\protect\hyperlink{ref-weissgerber2018}{69}}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact this field sooner rather than later.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 
-An Appendix which contains all the code used to create this manuscript, along with a basic workflow to implement GAMs in R is available as Supplementary Material in PDF.
+Two Appendices which contain a basic workflow to implement GAMs in R and all the code used to create this manuscript are available as PDFs in the Supplementary Material.
 A GitHub repository containing all the code used for this paper along with detailed instructions for its use is available at \url{https://github.com/aimundo/GAMs-biomedical-research}.

### 2021-12-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/02-Challenges.Rmd
Updated manuscript, checked lines in response to reviewer's document

@@
 
-However, it is worth reminding that "all models are wrong" [@box1976] and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power.
+However, it is worth reminding that "all models are wrong" [@box1976] and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of incomplete data, such as sampling bias or a drastic reduction in statistical power.
 Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.

### 2021-12-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/03-GAM Theory.Rmd
Updated manuscript, checked lines in response to reviewer's document

@@
 
-where $\eta_{ijt}$ is the linear predictor for each observation $i$ in each group $j$, at each timepoint $t$, and $\beta_0$ (the intercept), $\beta_{1}, \beta_2,$ and $\beta_3$ are the model parameters for each group for the linear mean from Equation (\@ref(eq:linear-model)).
-Finally, $time_{t}$ represents the covariates from each subject in each group at each time point.
+where $\eta_{ijt}$ is the linear predictor for each observation $i$ in each group $j$, at each timepoint $t$.
+Following the notation from Equation \@ref(eq:linear-model) the model parameters for each group are $\beta_0$ (the intercept), $\beta_{1}$, $\beta_2$, and $\beta_3$; $time_{t}$ represents the covariates from each subject in each group at each time point, $treatment_j$ represents the different treatment levels, and $time_t \times treatment_j$ represents their interaction.
 
@@
 where $E(y_{ijt})$ is the expectation, and $g^{-1}$ is the inverse of a _link function_  ($g$).
-The link function transforms the values from the response scale to the scale of the linear predictor $\eta$ (Equation \@ref(eq:GLM-eta).
+The link function transforms the values from the response scale to the scale of the linear predictor $\eta$ (Equation \@ref(eq:GLM-eta)).
 Therefore, it can be seen that LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.
@@
 The use of CS as smoothers in GAMs was discussed within the original GAM framework [@hastie1987], but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as 'knots', the points where the basis functions meet) to obtain the finite basis, which affects model fit [@wood2003].
-A solution to the "knot" placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate [@wood2003; @wood2017].
-
+A solution to the knot placement limitation of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate [@wood2003; @wood2017].
 In contrast, thin plate regression splines (TPRS) provide a reasonable "low rank" (truncated) approximation to the optimal TPS estimation, which can be implemented in an computationally efficient [@wood2003].
@@
 Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too "wiggly" due to the noise present.
-A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or "roughness") of the fit is achieved by using a _smoothness parameter_ ($\lambda$), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline.
-The second derivative of the spline is a measure of curvature, or the rate of change of the slope [@simpson2018; @wood2017], and increasing the penalty by increasing $\lambda$ results in models with less curvature.
+A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence controlling the wiggliness (or "roughness") of the fit is achieved by using a _smoothness parameter_ ($\lambda$), which is used to penalize the likelihood by multiplying it with the integrated square of the second derivative of the spline smooth.
+The second derivative of the spline smooth is a measure of curvature, or the rate of change of the slope [@simpson2018; @wood2017], and increasing the penalty by increasing $\lambda$ results in models with less curvature.
 As $\lambda$ increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting.
@@
 
-With this in mind, if four basis functions (plus the intercept) are used to fit a GAM for the data of Group 1 (concave down function) that appears in Figure \@ref(fig:l-q-response)D, the resulting fitting process is shown in Figure \@ref(fig:basis-plot).
+With this in mind, if four basis functions (plus the intercept) are used to fit a GAM for the data of Group 1 (concave down trend) that appears in Figure \@ref(fig:l-q-response)D, the resulting fitting process is shown in Figure \@ref(fig:basis-plot).
 In Figure \@ref(fig:basis-plot)A the four basis functions (and the intercept) are shown.
@@
 Figure \@ref(fig:basis-plot)C shows the resulting weighed basis functions.
-Note that the magnitude of the weighing for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
+Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
 On the other hand, the third basis function has roughly doubled its contribution to the smooth term.
@@
 In other words, we expect that about 95% of the points that compose the true function will be covered by the across the function CI.
-As a consequence, some areas of CI for the function must have more than nominal coverage and some areas less than the nominal coverage.
+As a consequence, some areas of the CI for the function have more than nominal coverage and some areas less than the nominal coverage.
 
@@
 In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller [@miller2019], Wood[@wood2017], Simpson [@simpson2018], Marra [@marra2012], and Ruppert [@ruppert2003].
-What we want to convey is that a Bayesian interpretation of GAMs allows for robust estimation using simultaneous empirical Bayesian CIs, as their stimates can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
+What we want to convey is that a Bayesian interpretation of GAMs allows for robust estimation using simultaneous empirical Bayesian CIs, as their estimates can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
 

### 2021-12-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/04-Longitudinal analysis with GAMs.Rmd
Updated manuscript, checked lines in response to reviewer's document

@@
 
-Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists.
-The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing.
+Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where unbalanced data exists.
+The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are incomplete.
 However, this usually causes the resulting smooths to have wider confidence intervals and less ability to discern differences in trends.
 To exemplify this, consider the random deletion of 40% of the simulated $\mbox{StO}_2$ values from Figure \@ref(fig:sim-smooth-plot)A.
-If the same interaction GAM (`gam_02`) is fitted to this data with missing observations, the resulting smooths appear in (Figure \@ref(fig:sim-smooth-plot)D).
+If the same interaction GAM (`gam_02`) is fitted to this data with unbalanced observations, the resulting smooths appear in (Figure \@ref(fig:sim-smooth-plot)D).
 Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas.
@@
 Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in `gam_02`) is that the difference between them can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that the difference between the separate smoothers per group (such as in `gam_02`) can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.
 
@@
 
-The absence of a _p-value_ in this context might seem odd, but its interpretation of the comparison can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \@ref(fig:sim-smooth-plot)A, where the chemotherapy causes $\mbox{StO}_2$ to increase over time.
+The absence of a _p-value_ in this context might seem odd, but the pairwise difference in smooths can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \@ref(fig:sim-smooth-plot)A, where the chemotherapy causes $\mbox{StO}_2$ to increase over time.
 
-With this expectation of different trends in each group, computing the difference between the trends will identify if and _when_ the observed difference is significant.
+With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant.
 A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).
-Therefore, the estimation of the difference between the smooths and the correspondent simultaneous empirical Bayesian CI of this difference provides an estimation of significance that is more robust that a mere _p-value_ because the estimated difference is able to indicates the significant time interval and we know that on average, the correction provided by the simultaneous empirical Bayesian CI ensures that it will contain _the whole function_ at a nominal value (say, 95%).
+Therefore, the computation of both the difference between the smooths and the corresponding simultaneous empirical Bayesian CI around this difference is able to provide an estimation of when and by how much there is a significant difference between the different groups.
+Additionally, the correction provided by the simultaneous empirical Bayesian CI makes the estimation robust as we know that on average, the simultaneous CI will contain the _whole function_ at a nominal value (say, 95%).
 
-Consider the calculation of pairwise differences for the fitted smooths that appear in Figure \@ref(fig:sim-smooth-plot)B, D.
+To visualize this, consider the calculation of pairwise differences for the fitted smooths that appear in Figure \@ref(fig:sim-smooth-plot)B, D.
 Figure \@ref(fig:plot-pairwise-comp) shows the difference between each treatment group trend for the full and missing datasets with a simultaneous CI computed around the difference.
@@
 
-In Figure \@ref(fig:plot-pairwise-comp)A, the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a different effect than the other.
+In Figure \@ref(fig:plot-pairwise-comp)A, the shaded regions over the confidence interval (where the CI does not cover 0) indicate the time interval where each group has a different mean effect than the other.
 Notice that the shaded region between days 1 and $\approx$ 2 for the full dataset indicates that through that time, the "Control" group has higher mean $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by day $\approx$ 3  it is the "Treatment" group that statistically on average has greater $\mbox{StO}_2$.
@@
 
-On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that because the confidence intervals overlap zero for the first two days there is statistically significant no difference between the groups.
-However, because the model is still able to pick the overall trend in $\mbox{StO}_2$, the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes statistically significant as in the full dataset smooth pairwise comparison.
+On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that because the confidence intervals overlap zero for the first two days there is no statistically significant difference between the groups.
+However, because the model is still able to pick the overall trend in $\mbox{StO}_2$, the pairwise comparison is able to estimate the change on day 3 where the mean difference between groups becomes statistically significant as in the full dataset smooth pairwise comparison.
 

### 2021-12-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts/plot_example.R

@@ -11,7 +11,8 @@ plot_example <- function(sim_dat,
                    group = treatment,
                    color = treatment)
         ) +
-        geom_point(show.legend=FALSE) +
+        geom_point(show.legend=FALSE,
+                   alpha = 0.5) +
         labs(y='response')+
         geom_line(aes(x = time,
                       y = mu,
@@ -65,7 +66,8 @@ plot_example <- function(sim_dat,
                  aes(x = time,
                      y = y,
                      color = treatment)) +
-        geom_point(show.legend=FALSE)+
+        geom_point(show.legend = FALSE,
+                   alpha=0.5)+
         labs(y='response')+
         geom_line(aes(y = predict(sim_dat$fit_anova),
                       group = subject, size = "Subjects"),show.legend = FALSE) +
@@ -91,7 +93,7 @@ plot_example <- function(sim_dat,
                  aes(x = time,
                      y = y,
                      color = treatment)) +
-        geom_point()+
+        geom_point(alpha = 0.5)+
         labs(y='response')+
         geom_line(aes(y = predict(sim_dat$fit_lme),
                       group = subject, size = "Subjects")) +

### 2022-01-04 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.Rmd
Updated

@@
 To this end, we have addressed all critiques.
-We hope these revisions, submitted on December 5, 2021, improve the manuscript so it is deemed worthy of publication in _Statistics in Medicine_.
+We hope these revisions, submitted on January 4th, 2022, improve the manuscript so it is deemed worthy of publication in _Statistics in Medicine_.
 Following are our detailed responses to reviewer comments.
@@
 
+The updated sentence in L59 now reads "significant effects (_p-value_<0.05)".
+
 ## Reviewer's Comments
@@
 However, we acknowledge that we didn't clearly conveyed this point to the read.
-In Section 6 (Discussion), L549-550 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided.
+In Section 6 (Discussion), L549-550 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided _a priori_ by the user.
 
@@
 
-The word "few" has been removed. The sentence now reads "...with as few as 4 observations per group..." (L482)
+The sentence now reads "...with as few as 4 observations per group..." (L482)
 

### 2022-01-04 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.tex
Updated

@@
 To this end, we have addressed all critiques.
-We hope these revisions, submitted on December 5, 2021, improve the manuscript so it is deemed worthy of publication in \emph{Statistics in Medicine}.
+We hope these revisions, submitted on January 4th, 2022, improve the manuscript so it is deemed worthy of publication in \emph{Statistics in Medicine}.
 Following are our detailed responses to reviewer comments.
@@
 
+The updated sentence in L59 now reads ``significant effects (\emph{p-value}\textless0.05)''.
+
 \hypertarget{reviewers-comments-18}{%
@@
 However, we acknowledge that we didn't clearly conveyed this point to the read.
-In Section 6 (Discussion), L549-550 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided.
+In Section 6 (Discussion), L549-550 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided \emph{a priori} by the user.
 
@@
 
-The word ``few'' has been removed. The sentence now reads ``\ldots with as few as 4 observations per group\ldots{}'' (L482)
+The sentence now reads ``\ldots with as few as 4 observations per group\ldots{}'' (L482)
 

### 2022-01-04 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/Full_document_SIM_No_Appendix.tex
Updated

@@
 \hypersetup{
-  pdftitle={Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models},
+  pdftitle={Generalized additive models to analyze non-linear trends in biomedical longitudinal data using R: Beyond repeated measures ANOVA and Linear Mixed Models},
   hidelinks,
@@
 
-\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}}
+\title{\textbf{Generalized additive models to analyze non-linear trends in biomedical longitudinal data using R: Beyond repeated measures ANOVA and Linear Mixed Models}}
 \date{\vspace{-2.5em}}
@@
 Therefore, GAMs present an excellent choice to analyze longitudinal data with non-linear trends in the context of biomedical research.
-This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data.
+This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on data with non-linear trends.
 We present the basic theory of GAMs and using reported trends of oxygen saturation in tumors, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R.
-We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when incomplete observations exist (with 40\% of the simulated observations missing).
+We also show that GAMs are able to produce estimates with non-linear trends even when incomplete observations exist (with 40\% of the simulated observations missing).
 To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
@@
 A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
-Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from data with non-linear trends, the validity of such a metric is dependent on how adequate the model fits the data.
+Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant effects ( \emph{p-value}\textless0.05)'' from data with non-linear trends, the validity of such a metric is dependent on how adequate the model fits the data.
 In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}}.

### 2022-01-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.Rmd
Updated, made changes

@@
 title: |
-    | **Generalized additive models to analyze biomedical non-linear longitudinal data in R:**
+    | **Generalized additive models to analyze non-linear trends in biomedical longitudinal data using R:**
     | Beyond repeated measures ANOVA and Linear Mixed Models
@@
     | **Response to Reviewer's Comments**
-    |
     | Journal of Submission: Statistics in Medicine
-    |
-    | Manuscript ID: SIM-21-0640
+subtitle: 'Manuscript ID: SIM-21-0640.R1'
 header-includes:
@@
     fig_caption: yes #allows captions in figures
-    toc: false #do not include table of contents
+    toc: true #include table of contents
+    toc_depth: 2
     number_sections: false #do not number sections
@@
 
-We would like to thank the reviewer for the careful and thorough analysis of this manuscript to _Statistics in Medicine_ and for the thoughtful comments and constructive suggestions, which helped improve the quality of this manuscript.
-We carefully considered the reviewer’s comments and in this document, explain how we revised the manuscript based on those comments and suggestions.
+We would like to thank the reviewers for the careful and thorough analysis of this manuscript to _Statistics in Medicine_ and for the thoughtful comments and constructive suggestions, which helped improve the quality of this manuscript.
+We carefully considered the reviewers comments and in this document, explain how we revised the manuscript based on those comments and suggestions.
 
@@
 Following are our detailed responses to reviewer comments.
+
+# Reviewer #1
 
@@
 
-# Comments from reviewers
+## Comments from Reviewer
 
-# Missing data
+## Missing data
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
@@
 
-# Unnecesary restriction of LMMs
+## Unnecesary restriction of LMMs
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-- L270-L273 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
-1.
+- L270-L273 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Figure 1.
 Specifically, that in reality the true function is not known and that this complicates choosing the degree of the polynomial.
@@
 
-- L547-L549 In Section 6 (Discussion) emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
+- L550-L552 In Section 6 (Discussion) emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
 
-# GLMMs
+## GLMMs
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-# Thin plate regression splines
+## Thin plate regression splines
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-# Penalised splines
-## Reviewer's Comments
+## Penalised splines
+### Reviewer's Comments
 
@@
 
-# Bayesian
+## Bayesian
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-- L395 now clarifies this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
+- L394-L395 now clarifies this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
 
 - The sentence "Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model", which appears in L319 in the original manuscript has been removed from the text.
-Instead, the concept of REML has been moved to Section 5.2 L456-459, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in _mgcv_.
+Instead, the concept of REML has been moved to Section 5.2 L457-460, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in _mgcv_.
 
-# Coverage of confidence intervals
+## Coverage of confidence intervals
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-- L406-L412 now provide a brief explanation of simultaneous CIs.
-With this explanation we want to provide the reader with an intuitive understanding on the advantages of simultaneous CIs over across the function CIs when computing comparisons between different groups (which we discuss in L417-L419).
+- L406-L413 now provide a brief explanation of simultaneous CIs and how they correct for the across the function CI.
+With this explanation we want to provide the reader with an intuitive understanding on the advantages of simultaneous CIs over across the function CIs when computing comparisons between different groups (which we discuss in L485-L503).
 
 - We have modified Figure 3 to now include across the function and simultaneous CIs.
-Fig.
-3B,D now show the fitted smooths and both intervals (across the function and simultaneous), and we explain the rationale for their inclusion in L477-L482.
+Figure 3B, D now show the fitted smooths and both intervals (across the function and simultaneous), and we explain the rationale for their inclusion in L478-L483.
 In this way, the reader has a visual understanding of the difference between across the function and simultaneous CIs.
 
-# Differences in smooths
+## Differences in smooths
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-- L494 now reads "unlikely to be distinguishable from zero".
+- L497-L498 The sentence now reads "unlikely to be distinguishable from zero".
 
-- L504-510 now indicate that we are computing a simultaneous CI around the difference, and that we _included_ the group means in our pairwise comparison.
+- L507-513 now indicate that we are computing a simultaneous CI around the difference, and that we _included_ the group means in our pairwise comparison.
 We also indicate to the reader the fact that if multiple parametric terms exist in the model the inclusion of the means can become problematic.
@@
 
-# Appendix
+## Appendix
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-L429 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
+L430 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
 The computational requisite in the smooth estimation of having the maximum number of basis equal to the number of unique values in the covariate makes it impossible to fit a GAM with a k>5, as the error "A term has fewer unique covariate combinations than specified maximum degrees of freedom" is thrown by _mgcv_.
@@
 
-# Suggested Changes
+## Suggested Changes
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 Simpson mentions here is that we want the reader to understand that fitting GAMs needs has to be more than a mechanic exercise in order to be effective.
-Too often, researchers fit a statistical model in a mechanic fashion, and our goal in this work is to move away from this.
+Too often, researchers fit a statistical model in a mechanic fashion, and our goal in this work is to convince the read to move away from such approach.
 
@@
 
-# Figures
+## Figures
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-# Minor Comments
+## Minor Comments
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
 This line has been removed from the revised manuscript.
-However, we acknowledge that we didn't clearly conveyed this point to the read.
+However, we acknowledge that we didn't clearly conveyed this point to the reader.
 In Section 6 (Discussion), L549-550 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided _a priori_ by the user.
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-The sentence now reads "separate smooths for..." (L441 in the revised manuscript)
+The sentence now reads "separate smooths for..." (L442 in the revised manuscript)
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-The sentence now reads "any differences between the group means" (L444).
+The sentence now reads "any differences between the group means" (L445).
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-We use the code notation provided by the $LaTeX$ package `listings`, and it provides curly quotes for code.
+We use the code notation provided by the $LaTeX$ package `listings`, and have updated the quotes to be straight in all code chunks.
 Regarding `family`, we believe that it is not necessary to include it here, but we indicate in L459-460 what the argument exists and that other distributions can be used.
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-The sentence now reads "contains the fitted model" (L449-L450).
+The sentence now reads "contains the fitted model" (L450-L451).
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-The updated sentence now reads "using four basis functions (plus the intercept)" (L451).
+The updated sentence now reads "using four basis functions (plus the intercept)" (L452).
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 However, we want to make clear to the reader that other types of splines that might be appropriate in certain situations are available.
-To acknowledge this, L453 in the updated manuscript now reads "(a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console)".
+To acknowledge this, L454 in the updated manuscript now reads "(a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console)".
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
 Only rm-ANOVA requires equally-spaced and complete observations.
-To avoid confusion, the line now reads "Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does)".
+To avoid confusion, the line now reads "Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does)" (L471).
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-The word "total" has been removed. The sentence now reads "consider the random deletion of 40% of the..." (L474)
+The word "total" has been removed. The sentence now reads "consider the random deletion of 40% of the..." (L475)
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-The sentence now reads "...with as few as 4 observations per group..." (L482)
+The sentence now reads "...with as few as 4 observations per group..." (L483)
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-The sentence has been rewritten. It now reads "...the observed difference is significant..." (L494).
+The sentence has been rewritten. It now reads "...the observed difference is significant..." (L496-L497).
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-The sentence in L494-L495 now reads "is unlikely to be distinguishable from zero".
+The sentence in L497-L498 now reads "is unlikely to be distinguishable from zero".
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
 This was an unintentional mistake based on some previous simulation results.
-We have updated the paragraph, and it now reads "However, because the model is still able to pick the overall trend in $\mbox{StO}_2$, the pairwise comparison is able to estimate the change on day 3 where the mean difference between groups becomes statistically significant as in the full dataset smooth pairwise comparison" (L520-L522).
-The change acknowledges that Control was higher before day 3, and that the Treatment Group becomes significant after day 3 in relation to L513-L517.
+We have updated the paragraph, and it now reads "However, because the model is still able to pick the overall trend in $\mbox{StO}_2$, the pairwise comparison is able to estimate the change on day 3 where the mean difference between groups becomes statistically significant as in the full dataset smooth pairwise comparison" (L523-L525).
+The change acknowledges that Control was higher before day 3, and that the Treatment Group becomes significant after day 3 in relation to L515-L519.
 
-## Reviewer's Comments
+### Reviewer's Comments
 
@@
 
-## \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
 
-We have made the appropriate changes. The sentence now reads "The GAM model is therefore able to identify changes between the groups at time points..." (L557)
+We have made the appropriate changes. The sentence now reads "The GAM is therefore able to identify changes between the groups at time points..." (L566)
+
+# Reviewer # 2
+
+## Reviewer's Introduction
+
+This is a well-written paper that gives a useful overview of methods that are underused in the field.
+It is written with great clarity all is well-targeted towards this journal's audience.
+I recommend accepting with only minor revisions.
+
+## \textcolor{reviewersblue} {Reply to Introduction}
+
+The authors thank the reviewer for deeming our within the scope of the Journal.
+
+## Comments from Reviewer
+
+There is one are where I feel expanded discussion would improve the paper: interpretability.
+A common challenge in presenting GAMs is the lack of interpretability of coefficients.
+While, in general, linear model coefficients simultaneously can be used to present effect sizes and significance, GAM models do not present estimands with such clear meaning.
+This is an area that practitioners will run into in taking up this new method, and it would serve them best to address it directly.
+A common approach to this is thinking carefully about the quantity of interest (such as the difference in outcomes at two specific values of a predictor, rather than a slope), and calculating this value, with uncertainties, using posterior samples.
+This is a natural extension of section 6.3., where one can discuss how calculations from the posterior can be used to determine the value of these quantities of interest in addition to significance.
+It would be useful to discuss posterior probabilities and the notion of a q-value as a Bayesian measure of significance that can be quantitative in addition to visual.
+The {gratia} package provides some tools for sampling from GAM posteriors that could be demonstrated.
+
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+Although we acknowledge that these concepts are important, in our manuscript we want to focus on the visualization part as we believe is more intuitive and informative in a biomedical context.
+We have addressed the comment from the reviewer by providing context about interpretability and posterior simulation without doing an in-depth discussion which would be beyond the scope of our work.
+To this end, the following changes have been made:
+
+L409-L413: We describe the use of simulation from the empirical Bayesian posterior distribution to correct for the coverage of the across the function CI, in a similar fashion of the correction that _q-values_ provide.
+
+L485-493: We indicate here that _p-values_ for GAMs do not have the same interpretation of effect size as in unlike linear models, but that we can use the model to estimate the instantaneous effect size by doing pairwise comparisons and constructing a simultaneous CI around the difference.
+Additionally, in Appendix B, Section B.5 we indicate that we use {gratia} to estimate the simulatenous and pointwise CIs.
+
+## Reviewer's Small Notes
+
+L14: GAMs aren't really what permit missing data or other correlative structures, all of which can be incorporated into LMEMs.
+I see this is addressed down in section 3.4, but it is somewhat misleading here.
+
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We have corrected the line above and now L11-L12 make clear that LMEMs can work with unbalanced data and non-constant correlation as well.
+Later, in L16-L17 we indicate that GAMs can also permit incomplete observations and different correlation structures.
+We believe this will clear any confusion about the capabilities of LMEMs.
+
+### Reviewer's Comments
+
+L99: LMEMs do not need normally distributed or independent data, though non-normal hierarchical effects tend to require more advanced tooling.
+
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+L95: we have removed the part about normality, and we now only indicate that random effects need to be independent.
+
+### Reviewer's Comments
+
+L179: I believe equation 4 need a term of random effect times time
+
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+Our random effect $\alpha_{ij}$ is just a random intercept for each individual, and therefore it does not vary over time, which is the general case of a LMEM we cover in the paper.
+
+### Reviewer's Comments
+
+L303: I believe it is important to note here that GAMs do not extrapolate well either.
+While they don't suffer from steep slopes right at the end of the data, beyond the range of the data they only reflect the assumptions built into the basis functions, be that flat values or linear extrapolation of the slope.
+
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We have incorporated the reviewer suggestion in Section 6 (Discussion), as we believe it is the appropriate place to briefly present the limitations of GAMs regarding extrapolation.
+In L552-L557 we have followed the basic structure of the reviewer's comment in order to indicate the extrapolation limitations from GAMs.
+The section of the paragraph now reads:
+
+"This does not mean, however, that as any other statistical model GAMs do not have certain limitations.
+In particular, beyond the range of the data GAMs only reflect the assumptions built into the basis functions, be that flat values or linear extrapolation of the slope.
+Therefore, researchers need to be careful when using GAMs for extrapolating purposes.
+In addition, both polynomial and GAMs show higher variance in estimates near the boundary of the data, but additive models generally have less variance than polynomials."
+
+We believe this change acknowledges the limitations of GAMs regarding extrapolation.
+
+### Reviewer's Comments
+
+L342, Figure 2: I believe the Y axis of (C), "Penalized Basis Functions", should be ""Weighted Basis Functions".
+
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+The label on the y axis now reads 'Weighted Basis Functions'.
+
+### Reviewer's Comments
+
+L360: The word "framework" is used twice here
+
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We believe the reviewer refers to L331 in the original manuscript. In the revised manuscript the sentence now reads:
+
+"The previous sections provided the basic understanding of the GAM framework and how these models..." (L421).
+
+### Reviewer's Comments
+
+Finally, I was pleased to find I could reproduce all your results and re-render the results from the GitHub repository. I commend the authors on this good work.
+
+### \textcolor{reviewersblue} {Reply to Reviewer's Comments}
+
+We appreciate the dedication of the reviewer to ensure the reproducibility of our results.

### 2022-01-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.tex
Updated, made changes

@@
 \usepackage[nomarkers,figuresonly]{endfloat}
+\usepackage{setspace}
+\setlength{\affilsep}{0.5em}
+
 %set a box to put the ORCID logo
@@
 %authors
+\singlespacing
 \author{Corresponding author: Timothy J. Muldoon*}
@@
 
-\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
+\title{\textbf{Generalized additive models to analyze non-linear trends in biomedical longitudinal data using R:}\\
 Beyond repeated measures ANOVA and Linear Mixed Models\\
@@
 \textbf{Response to Reviewer's Comments}\\
-\strut \\
-Journal of Submission: Statistics in Medicine\\
-\strut \\
-Manuscript ID: SIM-21-0640}
+Journal of Submission: Statistics in Medicine}
+\usepackage{etoolbox}
+\makeatletter
+\providecommand{\subtitle}[1]{% add subtitle to \maketitle
+  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
+}
+\makeatother
+\subtitle{Manuscript ID: SIM-21-0640.R1}
 \author{}
@@
 
+{
+\setcounter{tocdepth}{2}
+\tableofcontents
+}
 \hypertarget{general-comments-to-the-reviewers}{%
@@
 
-We would like to thank the reviewer for the careful and thorough analysis of this manuscript to \emph{Statistics in Medicine} and for the thoughtful comments and constructive suggestions, which helped improve the quality of this manuscript.
-We carefully considered the reviewer's comments and in this document, explain how we revised the manuscript based on those comments and suggestions.
+We would like to thank the reviewers for the careful and thorough analysis of this manuscript to \emph{Statistics in Medicine} and for the thoughtful comments and constructive suggestions, which helped improve the quality of this manuscript.
+We carefully considered the reviewers comments and in this document, explain how we revised the manuscript based on those comments and suggestions.
 
@@
 Following are our detailed responses to reviewer comments.
+
+\hypertarget{reviewer-1}{%
+\section{Reviewer \#1}\label{reviewer-1}}
 
@@
 
-\hypertarget{comments-from-reviewers}{%
-\section{Comments from reviewers}\label{comments-from-reviewers}}
+\hypertarget{comments-from-reviewer}{%
+\subsection{Comments from Reviewer}\label{comments-from-reviewer}}
 
 \hypertarget{missing-data}{%
-\section{Missing data}\label{missing-data}}
+\subsection{Missing data}\label{missing-data}}
 
 \hypertarget{reviewers-comments}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments}}
 
@@
 \hypertarget{section-1}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-1}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-1}}
 
@@
 \hypertarget{unnecesary-restriction-of-lmms}{%
-\section{Unnecesary restriction of LMMs}\label{unnecesary-restriction-of-lmms}}
+\subsection{Unnecesary restriction of LMMs}\label{unnecesary-restriction-of-lmms}}
 
 \hypertarget{reviewers-comments-1}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-1}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-1}}
 
@@
 \hypertarget{section-2}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-2}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-2}}
 
@@
 \item
-  L270-L273 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Fig.
-1.
+  L270-L273 now include some of the reasons that complicate the use of polynomial effects in a LMEM for the data in Figure 1.
 Specifically, that in reality the true function is not known and that this complicates choosing the degree of the polynomial.
@@
 \item
-  L547-L549 In Section 6 (Discussion) emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
+  L550-L552 In Section 6 (Discussion) emphasize again the limitations of polynomial effects when used on biological data, such as in the case of the simulated data in Fig. 1.
 \end{itemize}
@@
 \hypertarget{glmms}{%
-\section{GLMMs}\label{glmms}}
+\subsection{GLMMs}\label{glmms}}
 
 \hypertarget{reviewers-comments-2}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-2}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-2}}
 
@@
 \hypertarget{section-3}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-3}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-3}}
 
@@
 \hypertarget{thin-plate-regression-splines}{%
-\section{Thin plate regression splines}\label{thin-plate-regression-splines}}
+\subsection{Thin plate regression splines}\label{thin-plate-regression-splines}}
 
 \hypertarget{reviewers-comments-3}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-3}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-3}}
 
@@
 \hypertarget{section-4}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-4}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-4}}
 
@@
 \hypertarget{reviewers-comments-4}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-4}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-4}}
 
@@
 \hypertarget{section-5}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-5}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-5}}
 
@@
 \hypertarget{penalised-splines}{%
-\section{Penalised splines}\label{penalised-splines}}
+\subsection{Penalised splines}\label{penalised-splines}}
 
 \hypertarget{reviewers-comments-5}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-5}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-5}}
 
@@
 \hypertarget{section-6}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-6}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-6}}
 
@@
 \hypertarget{bayesian}{%
-\section{Bayesian}\label{bayesian}}
+\subsection{Bayesian}\label{bayesian}}
 
 \hypertarget{reviewers-comments-6}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-6}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-6}}
 
@@
 \hypertarget{section-7}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-7}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-7}}
 
@@
 \item
-  L395 now clarifies this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
+  L394-L395 now clarifies this by indicating that Stan, JAGS or other probabilistic programming language can be used to estimate GAMs using a full Bayesian approach.
 \item
   The sentence ``Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model'', which appears in L319 in the original manuscript has been removed from the text.
-Instead, the concept of REML has been moved to Section 5.2 L456-459, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in \emph{mgcv}.
+Instead, the concept of REML has been moved to Section 5.2 L457-460, where we state some of the reasons indicated by Wood when choosing restricted maximum likelihood (REML) over the default general cross validation (GCV) method for smooth parameter estimation in \emph{mgcv}.
 \end{itemize}
@@
 \hypertarget{coverage-of-confidence-intervals}{%
-\section{Coverage of confidence intervals}\label{coverage-of-confidence-intervals}}
+\subsection{Coverage of confidence intervals}\label{coverage-of-confidence-intervals}}
 
 \hypertarget{reviewers-comments-7}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-7}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-7}}
 
@@
 \hypertarget{section-8}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-8}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-8}}
 
@@
 \item
-  L406-L412 now provide a brief explanation of simultaneous CIs.
-With this explanation we want to provide the reader with an intuitive understanding on the advantages of simultaneous CIs over across the function CIs when computing comparisons between different groups (which we discuss in L417-L419).
+  L406-L413 now provide a brief explanation of simultaneous CIs and how they correct for the across the function CI.
+With this explanation we want to provide the reader with an intuitive understanding on the advantages of simultaneous CIs over across the function CIs when computing comparisons between different groups (which we discuss in L485-L503).
 \item
   We have modified Figure 3 to now include across the function and simultaneous CIs.
-Fig.
-3B,D now show the fitted smooths and both intervals (across the function and simultaneous), and we explain the rationale for their inclusion in L477-L482.
+Figure 3B, D now show the fitted smooths and both intervals (across the function and simultaneous), and we explain the rationale for their inclusion in L478-L483.
 In this way, the reader has a visual understanding of the difference between across the function and simultaneous CIs.
@@
 \hypertarget{differences-in-smooths}{%
-\section{Differences in smooths}\label{differences-in-smooths}}
+\subsection{Differences in smooths}\label{differences-in-smooths}}
 
 \hypertarget{reviewers-comments-8}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-8}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-8}}
 
@@
 \hypertarget{section-9}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-9}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-9}}
 
@@
 \item
-  L494 now reads ``unlikely to be distinguishable from zero''.
+  L497-L498 The sentence now reads ``unlikely to be distinguishable from zero''.
 \item
-  L504-510 now indicate that we are computing a simultaneous CI around the difference, and that we \emph{included} the group means in our pairwise comparison.
+  L507-513 now indicate that we are computing a simultaneous CI around the difference, and that we \emph{included} the group means in our pairwise comparison.
 We also indicate to the reader the fact that if multiple parametric terms exist in the model the inclusion of the means can become problematic.
@@
 \hypertarget{appendix}{%
-\section{Appendix}\label{appendix}}
+\subsection{Appendix}\label{appendix}}
 
 \hypertarget{reviewers-comments-9}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-9}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-9}}
 
@@
 \hypertarget{section-10}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-10}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-10}}
 
-L429 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
+L430 (in the revised manuscript) indicates that the simulated data used to fit the GAM that the reviewer alludes to has only 5 unique covariates (days 0, 2, 5, 7 and 10).
 The computational requisite in the smooth estimation of having the maximum number of basis equal to the number of unique values in the covariate makes it impossible to fit a GAM with a k\textgreater5, as the error ``A term has fewer unique covariate combinations than specified maximum degrees of freedom'' is thrown by \emph{mgcv}.
@@
 \hypertarget{reviewers-comments-10}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-10}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-10}}
 
@@
 \hypertarget{section-11}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-11}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-11}}
 
@@
 \hypertarget{reviewers-comments-11}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-11}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-11}}
 
@@
 \hypertarget{section-12}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-12}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-12}}
 
@@
 \hypertarget{suggested-changes}{%
-\section{Suggested Changes}\label{suggested-changes}}
+\subsection{Suggested Changes}\label{suggested-changes}}
 
 \hypertarget{reviewers-comments-12}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-12}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-12}}
 
@@
 \hypertarget{section-13}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-13}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-13}}
 
@@
 Our rationale for adding the information Dr.~Simpson mentions here is that we want the reader to understand that fitting GAMs needs has to be more than a mechanic exercise in order to be effective.
-Too often, researchers fit a statistical model in a mechanic fashion, and our goal in this work is to move away from this.
+Too often, researchers fit a statistical model in a mechanic fashion, and our goal in this work is to convince the read to move away from such approach.
 
@@
 \hypertarget{reviewers-comments-13}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-13}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-13}}
 
@@
 \hypertarget{section-14}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-14}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-14}}
 
@@
 \hypertarget{figures}{%
-\section{Figures}\label{figures}}
+\subsection{Figures}\label{figures}}
 
 \hypertarget{reviewers-comments-14}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-14}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-14}}
 
@@
 \hypertarget{section-15}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-15}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-15}}
 
@@
 \hypertarget{reviewers-comments-15}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-15}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-15}}
 
@@
 \hypertarget{section-16}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-16}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-16}}
 
@@
 \hypertarget{figures-1}{%
-\section{Figures}\label{figures-1}}
+\subsection{Figures}\label{figures-1}}
 
 \hypertarget{reviewers-comments-16}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-16}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-16}}
 
@@
 \hypertarget{section-17}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-17}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-17}}
 
@@
 \hypertarget{minor-comments}{%
-\section{Minor Comments}\label{minor-comments}}
+\subsection{Minor Comments}\label{minor-comments}}
 
 \hypertarget{reviewers-comments-17}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-17}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-17}}
 
@@
 \hypertarget{section-18}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-18}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-18}}
 
@@
 \hypertarget{reviewers-comments-18}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-18}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-18}}
 
@@
 \hypertarget{section-19}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-19}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-19}}
 
@@
 \hypertarget{reviewers-comments-19}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-19}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-19}}
 
@@
 \hypertarget{section-20}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-20}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-20}}
 
@@
 \hypertarget{reviewers-comments-20}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-20}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-20}}
 
@@
 \hypertarget{section-21}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-21}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-21}}
 
@@
 \hypertarget{reviewers-comments-21}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-21}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-21}}
 
@@
 \hypertarget{section-22}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-22}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-22}}
 
@@
 \hypertarget{reviewers-comments-22}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-22}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-22}}
 
@@
 \hypertarget{section-23}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-23}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-23}}
 
@@
 \hypertarget{reviewers-comments-23}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-23}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-23}}
 
@@
 \hypertarget{section-24}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-24}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-24}}
 
 This line has been removed from the revised manuscript.
-However, we acknowledge that we didn't clearly conveyed this point to the read.
+However, we acknowledge that we didn't clearly conveyed this point to the reader.
 In Section 6 (Discussion), L549-550 we now clearly state that a major advantage of GAMs is that they can learn non-linear trends from the data, thus eliminating the requirement of other models such as LMEMs where the relationship needs to be provided \emph{a priori} by the user.
@@
 \hypertarget{reviewers-comments-24}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-24}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-24}}
 
@@
 \hypertarget{section-25}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-25}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-25}}
 
-The sentence now reads ``separate smooths for\ldots{}'' (L441 in the revised manuscript)
+The sentence now reads ``separate smooths for\ldots{}'' (L442 in the revised manuscript)
 
 \hypertarget{reviewers-comments-25}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-25}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-25}}
 
@@
 \hypertarget{section-26}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-26}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-26}}
 
-The sentence now reads ``any differences between the group means'' (L444).
+The sentence now reads ``any differences between the group means'' (L445).
 
 \hypertarget{reviewers-comments-26}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-26}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-26}}
 
@@
 \hypertarget{section-27}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-27}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-27}}
 
-We use the code notation provided by the \(LaTeX\) package \passthrough{\lstinline!listings!}, and it provides curly quotes for code.
+We use the code notation provided by the \(LaTeX\) package \passthrough{\lstinline!listings!}, and have updated the quotes to be straight in all code chunks.
 Regarding \passthrough{\lstinline!family!}, we believe that it is not necessary to include it here, but we indicate in L459-460 what the argument exists and that other distributions can be used.
@@
 \hypertarget{reviewers-comments-27}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-27}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-27}}
 
@@
 \hypertarget{section-28}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-28}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-28}}
 
-The sentence now reads ``contains the fitted model'' (L449-L450).
+The sentence now reads ``contains the fitted model'' (L450-L451).
 
 \hypertarget{reviewers-comments-28}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-28}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-28}}
 
@@
 \hypertarget{section-29}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-29}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-29}}
 
-The updated sentence now reads ``using four basis functions (plus the intercept)'' (L451).
+The updated sentence now reads ``using four basis functions (plus the intercept)'' (L452).
 
 \hypertarget{reviewers-comments-29}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-29}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-29}}
 
@@
 \hypertarget{section-30}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-30}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-30}}
 
@@
 However, we want to make clear to the reader that other types of splines that might be appropriate in certain situations are available.
-To acknowledge this, L453 in the updated manuscript now reads ``(a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console)''.
+To acknowledge this, L454 in the updated manuscript now reads ``(a description of all the available smooths can be found by typing \passthrough{\lstinline!?mgcv::smooth.terms!} in the Console)''.
 
 \hypertarget{reviewers-comments-30}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-30}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-30}}
 
@@
 \hypertarget{section-31}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-31}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-31}}
 
@@
 \hypertarget{reviewers-comments-31}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-31}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-31}}
 
@@
 \hypertarget{section-32}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-32}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-32}}
 
 Only rm-ANOVA requires equally-spaced and complete observations.
-To avoid confusion, the line now reads ``Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does)''.
+To avoid confusion, the line now reads ``Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does)'' (L471).
 
 \hypertarget{reviewers-comments-32}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-32}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-32}}
 
@@
 \hypertarget{section-33}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-33}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-33}}
 
-The word ``total'' has been removed. The sentence now reads ``consider the random deletion of 40\% of the\ldots{}'' (L474)
+The word ``total'' has been removed. The sentence now reads ``consider the random deletion of 40\% of the\ldots{}'' (L475)
 
 \hypertarget{reviewers-comments-33}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-33}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-33}}
 
@@
 \hypertarget{section-34}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-34}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-34}}
 
-The sentence now reads ``\ldots with as few as 4 observations per group\ldots{}'' (L482)
+The sentence now reads ``\ldots with as few as 4 observations per group\ldots{}'' (L483)
 
 \hypertarget{reviewers-comments-34}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-34}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-34}}
 
@@
 \hypertarget{section-35}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-35}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-35}}
 
-The sentence has been rewritten. It now reads ``\ldots the observed difference is significant\ldots{}'' (L494).
+The sentence has been rewritten. It now reads ``\ldots the observed difference is significant\ldots{}'' (L496-L497).
 
 \hypertarget{reviewers-comments-35}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-35}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-35}}
 
@@
 \hypertarget{section-36}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-36}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-36}}
 
-The sentence in L494-L495 now reads ``is unlikely to be distinguishable from zero''.
+The sentence in L497-L498 now reads ``is unlikely to be distinguishable from zero''.
 
 \hypertarget{reviewers-comments-36}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-36}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-36}}
 
@@
 \hypertarget{section-37}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-37}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-37}}
 
 This was an unintentional mistake based on some previous simulation results.
-We have updated the paragraph, and it now reads ``However, because the model is still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the mean difference between groups becomes statistically significant as in the full dataset smooth pairwise comparison'' (L520-L522).
-The change acknowledges that Control was higher before day 3, and that the Treatment Group becomes significant after day 3 in relation to L513-L517.
+We have updated the paragraph, and it now reads ``However, because the model is still able to pick the overall trend in \(\mbox{StO}_2\), the pairwise comparison is able to estimate the change on day 3 where the mean difference between groups becomes statistically significant as in the full dataset smooth pairwise comparison'' (L523-L525).
+The change acknowledges that Control was higher before day 3, and that the Treatment Group becomes significant after day 3 in relation to L515-L519.
 
 \hypertarget{reviewers-comments-37}{%
-\subsection{Reviewer's Comments}\label{reviewers-comments-37}}
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-37}}
 
@@
 \hypertarget{section-38}{%
-\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-38}}
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-38}}
 
-We have made the appropriate changes. The sentence now reads ``The GAM model is therefore able to identify changes between the groups at time points\ldots{}'' (L557)
+We have made the appropriate changes. The sentence now reads ``The GAM is therefore able to identify changes between the groups at time points\ldots{}'' (L566)
+
+\hypertarget{reviewer-2}{%
+\section{Reviewer \# 2}\label{reviewer-2}}
+
+\hypertarget{reviewers-introduction-1}{%
+\subsection{Reviewer's Introduction}\label{reviewers-introduction-1}}
+
+This is a well-written paper that gives a useful overview of methods that are underused in the field.
+It is written with great clarity all is well-targeted towards this journal's audience.
+I recommend accepting with only minor revisions.
+
+\hypertarget{section-39}{%
+\subsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Introduction}}{}}\label{section-39}}
+
+The authors thank the reviewer for deeming our within the scope of the Journal.
+
+\hypertarget{comments-from-reviewer-1}{%
+\subsection{Comments from Reviewer}\label{comments-from-reviewer-1}}
+
+There is one are where I feel expanded discussion would improve the paper: interpretability.
+A common challenge in presenting GAMs is the lack of interpretability of coefficients.
+While, in general, linear model coefficients simultaneously can be used to present effect sizes and significance, GAM models do not present estimands with such clear meaning.
+This is an area that practitioners will run into in taking up this new method, and it would serve them best to address it directly.
+A common approach to this is thinking carefully about the quantity of interest (such as the difference in outcomes at two specific values of a predictor, rather than a slope), and calculating this value, with uncertainties, using posterior samples.
+This is a natural extension of section 6.3., where one can discuss how calculations from the posterior can be used to determine the value of these quantities of interest in addition to significance.
+It would be useful to discuss posterior probabilities and the notion of a q-value as a Bayesian measure of significance that can be quantitative in addition to visual.
+The \{gratia\} package provides some tools for sampling from GAM posteriors that could be demonstrated.
+
+\hypertarget{section-40}{%
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-40}}
+
+Although we acknowledge that these concepts are important, in our manuscript we want to focus on the visualization part as we believe is more intuitive and informative in a biomedical context.
+We have addressed the comment from the reviewer by providing context about interpretability and posterior simulation without doing an in-depth discussion which would be beyond the scope of our work.
+To this end, the following changes have been made:
+
+L409-L413: We describe the use of simulation from the empirical Bayesian posterior distribution to correct for the coverage of the across the function CI, in a similar fashion of the correction that \emph{q-values} provide.
+
+L485-493: We indicate here that \emph{p-values} for GAMs do not have the same interpretation of effect size as in unlike linear models, but that we can use the model to estimate the instantaneous effect size by doing pairwise comparisons and constructing a simultaneous CI around the difference.
+Additionally, in Appendix B, Section B.5 we indicate that we use \{gratia\} to estimate the simulatenous and pointwise CIs.
+
+\hypertarget{reviewers-small-notes}{%
+\subsection{Reviewer's Small Notes}\label{reviewers-small-notes}}
+
+L14: GAMs aren't really what permit missing data or other correlative structures, all of which can be incorporated into LMEMs.
+I see this is addressed down in section 3.4, but it is somewhat misleading here.
+
+\hypertarget{section-41}{%
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-41}}
+
+We have corrected the line above and now L11-L12 make clear that LMEMs can work with unbalanced data and non-constant correlation as well.
+Later, in L16-L17 we indicate that GAMs can also permit incomplete observations and different correlation structures.
+We believe this will clear any confusion about the capabilities of LMEMs.
+
+\hypertarget{reviewers-comments-38}{%
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-38}}
+
+L99: LMEMs do not need normally distributed or independent data, though non-normal hierarchical effects tend to require more advanced tooling.
+
+\hypertarget{section-42}{%
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-42}}
+
+L95: we have removed the part about normality, and we now only indicate that random effects need to be independent.
+
+\hypertarget{reviewers-comments-39}{%
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-39}}
+
+L179: I believe equation 4 need a term of random effect times time
+
+\hypertarget{section-43}{%
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-43}}
+
+Our random effect \(\alpha_{ij}\) is just a random intercept for each individual, and therefore it does not vary over time, which is the general case of a LMEM we cover in the paper.
+
+\hypertarget{reviewers-comments-40}{%
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-40}}
+
+L303: I believe it is important to note here that GAMs do not extrapolate well either.
+While they don't suffer from steep slopes right at the end of the data, beyond the range of the data they only reflect the assumptions built into the basis functions, be that flat values or linear extrapolation of the slope.
+
+\hypertarget{section-44}{%
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-44}}
+
+We have incorporated the reviewer suggestion in Section 6 (Discussion), as we believe it is the appropriate place to briefly present the limitations of GAMs regarding extrapolation.
+In L552-L557 we have followed the basic structure of the reviewer's comment in order to indicate the extrapolation limitations from GAMs.
+The section of the paragraph now reads:
+
+``This does not mean, however, that as any other statistical model GAMs do not have certain limitations.
+In particular, beyond the range of the data GAMs only reflect the assumptions built into the basis functions, be that flat values or linear extrapolation of the slope.
+Therefore, researchers need to be careful when using GAMs for extrapolating purposes.
+In addition, both polynomial and GAMs show higher variance in estimates near the boundary of the data, but additive models generally have less variance than polynomials.''
+
+We believe this change acknowledges the limitations of GAMs regarding extrapolation.
+
+\hypertarget{reviewers-comments-41}{%
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-41}}
+
+L342, Figure 2: I believe the Y axis of (C), ``Penalized Basis Functions'', should be ````Weighted Basis Functions''.
+
+\hypertarget{section-45}{%
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-45}}
+
+The label on the y axis now reads `Weighted Basis Functions'.
+
+\hypertarget{reviewers-comments-42}{%
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-42}}
+
+L360: The word ``framework'' is used twice here
+
+\hypertarget{section-46}{%
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-46}}
+
+We believe the reviewer refers to L331 in the original manuscript. In the revised manuscript the sentence now reads:
+
+``The previous sections provided the basic understanding of the GAM framework and how these models\ldots{}'' (L421).
+
+\hypertarget{reviewers-comments-43}{%
+\subsubsection{Reviewer's Comments}\label{reviewers-comments-43}}
+
+Finally, I was pleased to find I could reproduce all your results and re-render the results from the GitHub repository. I commend the authors on this good work.
+
+\hypertarget{section-47}{%
+\subsubsection{\texorpdfstring{\textcolor{reviewersblue} {Reply to Reviewer's Comments}}{}}\label{section-47}}
+
+We appreciate the dedication of the reviewer to ensure the reproducibility of our results.
 

### 2022-01-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/Full_document_SIM_No_Appendix.tex
Updated, made changes

@@
     % How/what to match
+    language=R,
+    upquote=true,
     sensitive=true,
@@
 In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs).
-Although LMEMs are less restrictive than rm-ANOVA in terms of assumptions about imbalanced observations and correlation between observations, both methodologies assume a linear trend in the measured response.
+Although LMEMs are less restrictive than rm-ANOVA as they can work with unbalanced data and non-constant correlation between observations, both methodologies assume a linear trend in the measured response.
 It is common in biomedical research that the true trend response is nonlinear and in these cases the linearity assumption of rm-ANOVA and LMEMs can lead to biased estimates and unreliable inference.
 
-In contrast, generalized additive models (GAMs) relax the assumptions made by LMEMs and rm-ANOVA, and allow the data to determine the fit of the model while permitting incomplete observations and different correlation structures.
+In contrast, GAMs relax the linearity assumption of rm-ANOVA and LMEMs and allow the data to determine the fit of the model while also permitting incomplete observations and different correlation structures.
 Therefore, GAMs present an excellent choice to analyze longitudinal data with non-linear trends in the context of biomedical research.
@@
 
-To get the weighed basis functions, each basis (from Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
-Figure \ref{fig:basis-plot}C shows the resulting weighed basis functions.
+To get the weighted basis functions, each basis (from Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions.
+Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions.
 Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
@@
 The coefficient determines the overall effect of each basis in the final smoother.
-\textbf{C}: Weighed basis functions.
-Each of the four basis functions (and the intercept) of panel A has been weighed by the corresponding coefficient shown in Panel B.
+\textbf{C}: Weighted basis functions.
+Each of the four basis functions (and the intercept) of panel A has been weighted by the corresponding coefficient shown in Panel B.
 Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
@@
 Suppose we chose a nominal value (say, 95\%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
-Briefly, to obtain a simultaneous CI we use simulation from the empirical Bayesian posterior distribution to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}.
-
-In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller\textsuperscript{\protect\hyperlink{ref-miller2019}{61}}, Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}}, Marra\textsuperscript{\protect\hyperlink{ref-marra2012}{63}}, and Ruppert\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}.
+Briefly, to obtain a simultaneous CI we simulate 10,000 draws from the empirical Bayesian posterior distribution of the fitted smooths.
+Then, we obtain the maximum absolute standardized deviation of the differences in smooth estimates which is used to correct the coverage of the across the function CI\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}} in a similar fashion to how \emph{q-values} correct \emph{p-values} to control false positive discovery rates\textsuperscript{\protect\hyperlink{ref-storey2003}{63}}.
+
+In-depth theory of the Bayesian interpretation of GAMs and details on the computation of simultaneous and across the function CIs are beyond the scope of this paper, but can be found in Miller\textsuperscript{\protect\hyperlink{ref-miller2019}{61}}, Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}}, Marra\textsuperscript{\protect\hyperlink{ref-marra2012}{64}}, and Ruppert\textsuperscript{\protect\hyperlink{ref-ruppert2003}{62}}.
 What we want to convey is that a Bayesian interpretation of GAMs allows for robust estimation using simultaneous empirical Bayesian CIs, as their estimates can be used to make comparisons between different groups in a similar way that multiple comparisons adjustments make inference from ANOVA models more reliable.
@@
 When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B).
-Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}\textsuperscript{\protect\hyperlink{ref-gratia}{64}} as we show in Appendix A.
-Additional discussions on model selection can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak\textsuperscript{\protect\hyperlink{ref-harezlak2018}{65}}.
+Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}\textsuperscript{\protect\hyperlink{ref-gratia}{65}} as we show in Appendix A.
+Additional discussions on model selection can be found in Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and Harezlak\textsuperscript{\protect\hyperlink{ref-harezlak2018}{66}}.
 
@@
 At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a \emph{post-hoc} analysis is required to answer such question by calculating some \emph{p-values} after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that the difference between the separate smoothers per group (such as in \passthrough{\lstinline!gam\_02!}) can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.
+In linear models (such as rm-ANOVA), if there is a significant \emph{p-value} after a \emph{post-hoc} analysis we then can make inference about the effect size using the slope or the intercept from the model.
+In GAMs however, there is no single \emph{p-value} to determine the significance of an effect as in linear models.
+Therefore, the coefficients of GAMs do not provide a simple interpretation as in the linear model case, but the changes in slope at specific timepoints can be used to determine the instantaneous effect size.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that the difference between the separate smoothers per group (such as in \passthrough{\lstinline!gam\_02!}) can be computed pairwise, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.
 
@@
 
-The absence of a \emph{p-value} in this context might seem odd, but the pairwise difference in smooths can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
-
+The pairwise difference in smooths can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \ref{fig:sim-smooth-plot}A, where the chemotherapy causes \(\mbox{StO}_2\) to increase over time.
 With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant.
@@
 
-Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{66}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) or \emph{p-value} value, whose interpretation is not clear in the case of a Type III error.
+Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{67}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) or \emph{p-value} value, whose interpretation is not clear in the case of a Type III error.
 
 In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.
-It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (e.g., which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research\textsuperscript{\protect\hyperlink{ref-hastie1995}{67}}.
+It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (e.g., which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research\textsuperscript{\protect\hyperlink{ref-hastie1995}{68}}.
 This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs.
@@
 However, in Section \ref{GAM-theory} we demonstrated that at its core the principle underlying GAMs is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data.
-The use of basis functions is a major advantage over models where the user has to know the non-linear relationship \emph{a priori}, such as in the case of polynomial effects in LMEMs.
-In addition, there is no biological interpretation for such polynomial assumptions.
-In contrast, because GAMs let the data speak for themselves they can provide an accurate representation of the effect of time in a biological process.
-
-However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
+The use of basis functions is a major advantage over models where the user has to know the non-linear relationship \emph{a priori}, such as in the case of polynomial effects in LMEMs where in addition, there is no biological interpretation for such polynomial assumptions.
+This does not mean, however, that as any other statistical model GAMs do not have certain limitations.
+In particular, beyond the range of the data GAMs only reflect the assumptions built into the basis functions, be that flat values or linear extrapolation of the slope.
+Therefore, researchers need to be careful when using GAMs for extrapolating purposes.
+In addition, both polynomial and GAMs show higher variance in estimates near the boundary of the data, but additive models generally have less variance than polynomials.
+However, because GAMs let the data speak for themselves, they provide estimates that are consistent with non-linear trends and therefore can be used to obtain an accurate representation of the effect of time in a biological process.
+
+Beyond the theory, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented.
 We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} using R and the package \emph{mgcv}\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} in Section \ref{longitudinal-GAMs}, while a basic workflow for model selection is in Appendix A.
@@
 We have used R as the software of choice for this paper because it provides a fully developed environment to fit GAMs, enables simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations), and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work more reproducible.
-In this regard, reproducibility is still an issue in biomedical research\textsuperscript{\protect\hyperlink{ref-begley2015}{68},\protect\hyperlink{ref-weissgerber2018}{69}}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact this field sooner rather than later.
+In this regard, reproducibility is still an issue in biomedical research\textsuperscript{\protect\hyperlink{ref-begley2015}{69},\protect\hyperlink{ref-weissgerber2018}{70}}, but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact this field sooner rather than later.
 Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-storey2003}{}}%
+\CSLLeftMargin{63. }
+\CSLRightInline{Storey JD.
+The positive false discovery rate: A bayesian interpretation and the q-value.
+\emph{The Annals of Statistics}.
+2003;31(6).
+doi:\href{https://doi.org/10.1214/aos/1074290335}{10.1214/aos/1074290335}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-marra2012}{}}%
-\CSLLeftMargin{63. }
+\CSLLeftMargin{64. }
 \CSLRightInline{Marra G, Wood SN.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gratia}{}}%
-\CSLLeftMargin{64. }
+\CSLLeftMargin{65. }
 \CSLRightInline{Simpson GL. \emph{Gratia: Graceful 'Ggplot'-Based Graphics and Other Functions for GAMs Fitted Using 'Mgcv'}.; 2020. \url{https://CRAN.R-project.org/package=gratia}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-harezlak2018}{}}%
-\CSLLeftMargin{65. }
+\CSLLeftMargin{66. }
 \CSLRightInline{Harezlak J, Ruppert D, Wand MP. \emph{Semiparametric Regression with {R}}. Springer New York; 2018. doi:\href{https://doi.org/10.1007/978-1-4939-8853-2}{10.1007/978-1-4939-8853-2}}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lang2015}{}}%
-\CSLLeftMargin{66. }
+\CSLLeftMargin{67. }
 \CSLRightInline{Lang TA, Altman DG.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hastie1995}{}}%
-\CSLLeftMargin{67. }
+\CSLLeftMargin{68. }
 \CSLRightInline{Hastie T, Tibshirani R.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-begley2015}{}}%
-\CSLLeftMargin{68. }
+\CSLLeftMargin{69. }
 \CSLRightInline{Begley CG, Ioannidis JPA.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-weissgerber2018}{}}%
-\CSLLeftMargin{69. }
+\CSLLeftMargin{70. }
 \CSLRightInline{Weissgerber TL, Garcia-Valencia O, Garovic VD, Milic NM, Winham SJ.

### 2022-01-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_A.tex
Updated, made changes

@@
 \lstset{
+    language=R,
+    upquote=true,
     % How/what to match
@@
 
-#alpha for ribbon in the smooth plots
+# alpha for ribbon in the smooth plots
 al <- 0.8
 
-thm1 <- scale_fill_scico_d(palette = "tokyo", begin = 0.3, end = 0.8,
-                         direction = -1, aesthetics = c("colour", "fill"))
+thm1 <- scale_fill_scico_d(palette = "tokyo", begin = 0.3, end = 0.8, direction = -1,
+    aesthetics = c("colour", "fill"))
 \end{lstlisting}
@@
 \begin{lstlisting}[language=R]
-dat<-tibble(StO2  = c(4, 27, 3, 2, 0.5, 7,4, 50, 45, 56),
-            Day   = rep(c(0, 2, 5, 7, 10), times = 2),
-            Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
+dat <- tibble(StO2 = c(4, 27, 3, 2, 0.5, 7, 4, 50, 45, 56), Day = rep(c(0, 2, 5,
+    7, 10), times = 2), Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
 \end{lstlisting}
@@
 sd <- 10  #approximate sd from paper
-source(here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final",
-            "scripts", "simulate_data.R"))
+source(here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", "scripts",
+    "simulate_data.R"))
 dat_sim <- simulate_data(dat, n, sd)
@@
 \begin{lstlisting}[language=R]
-gam_00<-gam(StO2_sim ~ s(Day, k = 5),
-            method   ='REML',
-            data     = dat_sim)
+gam_00 <- gam(StO2_sim ~ s(Day, k = 5), method = "REML", data = dat_sim)
 \end{lstlisting}
@@
 \begin{lstlisting}[language=R]
-source(here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final",
-            "scripts", "gam_diagnostics.R"))
+source(here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", "scripts",
+    "gam_diagnostics.R"))
 gam_diagnostics(gam_00)
@@
 \begin{lstlisting}[language=R]
-gam_01<-gam(StO2_sim ~ s(Day, by=Group, k = 5),
-            method ='REML',
-            data   = dat_sim)
+gam_01 <- gam(StO2_sim ~ s(Day, by = Group, k = 5), method = "REML", data = dat_sim)
 
@@
 \begin{lstlisting}[language=R]
-#GAM for StO2
+# GAM for StO2
 
-gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
-            method ='REML',
-            data   = dat_sim)
+gam_02 <- gam(StO2_sim ~ Group + s(Day, by = Group, k = 5), method = "REML", data = dat_sim)
 
@@
 \begin{lstlisting}[language=R]
-######IMPORTANT: uncommenting the following two lines
-    #removes the group means from the comparison######
+###### IMPORTANT: uncommenting the following two lines removes the group means
+###### from the comparison######
 \end{lstlisting}
@@
 \begin{lstlisting}[language=R]
-source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
-                  "scripts","pointwise_comparisons.R"))
-source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final",
-                  "scripts", "difference_smooths.R"))
-source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
-                  "scripts", "pairwise_limits.R"))
-#compute difference between smooths and calculate confidence interval: complete data
-diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", 
-                                    newdata = newdat, 
-                                    unconditional = TRUE,
-                                    frequentist = FALSE,
-                                    n = 100, partial_match = TRUE, 
-                                    nrep=10000)
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", "scripts",
+    "pointwise_comparisons.R"))
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", "scripts",
+    "difference_smooths.R"))
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", "scripts",
+    "pairwise_limits.R"))
+# compute difference between smooths and calculate confidence interval:
+# complete data
+diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
+    unconditional = TRUE, frequentist = FALSE, n = 100, partial_match = TRUE, nrep = 10000)
 my_list <- pairwise_limits(diff_complete)
@@
 \begin{lstlisting}[language=R]
-rib_col <- '#8D7D82'     #color for ribbon for confidence interval
-control_rib <- '#875F79' #color for ribbon for control region
-treat_rib <- '#A7D89E'   #color for ribbon treatment region
+rib_col <- "#8D7D82"  #color for ribbon for confidence interval
+control_rib <- "#875F79"  #color for ribbon for control region
+treat_rib <- "#A7D89E"  #color for ribbon treatment region
 
-c1 <- ggplot() +
-  geom_line(data = diff_complete, aes(x = Day, y = diff), 
-            size = 1, alpha = 0.5) +
-  annotate("rect", xmin = my_list$init1, xmax = my_list$final1,
-           ymin = -Inf, ymax = Inf, fill = control_rib,
-           alpha = 0.5) + 
-  annotate("text", x = 1.5, y = -18, label = "Control>Treatment",
-           size = 6, angle = 90) +
-  annotate("rect", xmin = my_list$init2, xmax = my_list$final2,
-           ymin = -Inf, ymax = Inf, fill = treat_rib,
-           alpha = 0.5) +
-  annotate("text", x = 6, y = -18, label = "Treatment>Control",
-           size = 6, angle = 90) +
-  geom_ribbon(data = diff_complete, 
-              aes(x = Day, ymin = lower_s, ymax = upper_s),
-              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
-  geom_hline(yintercept = 0, lty = 2, color = "red") +
-  scale_x_continuous(breaks = c(0, 2, 5, 7, 10)) +
-  labs(y = "Difference\n(Complete observations)") +
-  theme_classic()+
-  theme(axis.text = element_text(size = 22))
+c1 <- ggplot() + geom_line(data = diff_complete, aes(x = Day, y = diff), size = 1,
+    alpha = 0.5) + annotate("rect", xmin = my_list$init1, xmax = my_list$final1,
+    ymin = -Inf, ymax = Inf, fill = control_rib, alpha = 0.5) + annotate("text",
+    x = 1.5, y = -18, label = "Control>Treatment", size = 6, angle = 90) + annotate("rect",
+    xmin = my_list$init2, xmax = my_list$final2, ymin = -Inf, ymax = Inf, fill = treat_rib,
+    alpha = 0.5) + annotate("text", x = 6, y = -18, label = "Treatment>Control",
+    size = 6, angle = 90) + geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s,
+    ymax = upper_s), alpha = 0.5, fill = rib_col, inherit.aes = FALSE) + geom_hline(yintercept = 0,
+    lty = 2, color = "red") + scale_x_continuous(breaks = c(0, 2, 5, 7, 10)) + labs(y = "Difference\n(Complete observations)") +
+    theme_classic() + theme(axis.text = element_text(size = 22))
 \end{lstlisting}

### 2022-01-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_B.tex
Updated, made changes

@@
 
-#alpha for ribbon in the smooth plots
-al<-     0.8
+# alpha for ribbon in the smooth plots
+al <- 0.8
 
-thm1 <- scale_fill_scico_d(palette = "tokyo", begin = 0.3, end = 0.8,
-                           direction = -1, aesthetics = c("colour", "fill"))
+thm1 <- scale_fill_scico_d(palette = "tokyo", begin = 0.3, end = 0.8, direction = -1,
+    aesthetics = c("colour", "fill"))
 \end{lstlisting}
@@
 \begin{lstlisting}[language=R]
-## This function plots the rm-ANOVA and LMEM for the data simulated in example.R
+## This function plots the rm-ANOVA and LMEM for the data simulated in
+## example.R
 plot_example <- function(sim_dat) {
@@
     p1 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
-    geom_point(show.legend = FALSE) +
-    labs(y = 'response')+
-    geom_line(aes(x = time, y = mu, color = treatment), 
-              show.legend = FALSE) +
-    theme_classic() +
-    theme(plot.title = element_text(size = txt, face = "bold"),
-          text=element_text(size = txt)) +
-    thm1
+        ggplot(aes(x = time, y = y, group = treatment, color = treatment)) + geom_point(show.legend = FALSE) +
+        labs(y = "response") + geom_line(aes(x = time, y = mu, color = treatment),
+        show.legend = FALSE) + theme_classic() + theme(plot.title = element_text(size = txt,
+        face = "bold"), text = element_text(size = txt)) + thm1
 
-  #plot the simulated data with trajectories per each subject
+    # plot the simulated data with trajectories per each subject
     p2 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
-    geom_line(aes(size = "Subjects"), show.legend = FALSE) +
-    # facet_wrap(~ treatment) +
-    geom_line(aes(x = time, y = mu, color = treatment, 
-                  size = "Simulated Truth"),
-              lty = 1, show.legend = FALSE) +
-    labs(y = 'response') +
-    scale_size_manual(name = "Type", 
-                      values = c("Subjects" = 0.5, "Simulated Truth" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = txt, face = "bold"),
-          text = element_text(size = txt)) +
-    thm1
+        ggplot(aes(x = time, y = y, group = subject, color = treatment)) + geom_line(aes(size = "Subjects"),
+        show.legend = FALSE) + # facet_wrap(~ treatment) + show.legend = FALSE)
+        show.legend = FALSE) + # facet_wrap(~ treatment) + + # facet_wrap(~
+        show.legend = FALSE) + # facet_wrap(~ treatment) + treatment) +
+    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"),
+        lty = 1, show.legend = FALSE) + labs(y = "response") + scale_size_manual(name = "Type",
+        values = c(Subjects = 0.5, `Simulated Truth` = 3)) + theme_classic() + theme(plot.title = element_text(size = txt,
+        face = "bold"), text = element_text(size = txt)) + thm1
 
-  #plot the errors
+    # plot the errors
     p3 <- sim_dat$dat %>%
-    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
-    geom_line(show.legend = FALSE) +
-    labs(y = 'errors') +
-    theme_classic() +
-    theme(plot.title = element_text(size = txt, face = "bold"),
-          text = element_text(size = txt))+
-    thm1
+        ggplot(aes(x = time, y = errors, group = subject, color = treatment)) + geom_line(show.legend = FALSE) +
+        labs(y = "errors") + theme_classic() + theme(plot.title = element_text(size = txt,
+        face = "bold"), text = element_text(size = txt)) + thm1
 
-  #plot the model predictions for rm-ANOVA
-  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point(show.legend = FALSE) +
-    labs(y = 'response')+
-    geom_line(aes(y = predict(sim_dat$fit_anova), 
-                  group = subject, 
-                  size = "Subjects"),
-              show.legend = FALSE) +
-    geom_line(data = sim_dat$pred_dat,
-              aes(y = predict(sim_dat$fit_anova,
-                              level = 0,
-                              newdata = sim_dat$pred_dat),
-                  size = "Population"),
-              show.legend=FALSE) +
-    guides(color = guide_legend(override.aes = list(size = 2))) +
-    scale_size_manual(name = "Predictions",
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = txt, face = "bold"),
-          text = element_text(size = txt)) +
-    thm1
+    # plot the model predictions for rm-ANOVA
+    p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) + geom_point(show.legend = FALSE) +
+        labs(y = "response") + geom_line(aes(y = predict(sim_dat$fit_anova), group = subject,
+        size = "Subjects"), show.legend = FALSE) + geom_line(data = sim_dat$pred_dat,
+        aes(y = predict(sim_dat$fit_anova, level = 0, newdata = sim_dat$pred_dat),
+            size = "Population"), show.legend = FALSE) + guides(color = guide_legend(override.aes = list(size = 2))) +
+        scale_size_manual(name = "Predictions", values = c(Subjects = 0.5, Population = 3)) +
+        theme_classic() + theme(plot.title = element_text(size = txt, face = "bold"),
+        text = element_text(size = txt)) + thm1
 
-  #plot the LMEM predictions
-  p5 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
-    geom_point()+
-    labs(y = 'response')+
-    geom_line(aes(y = predict(sim_dat$fit_lme),
-                  group = subject, size = "Subjects")) +
-    geom_line(data = sim_dat$pred_dat,
-              aes(y = predict(sim_dat$fit_lme, 
-                              level = 0,
-                              newdata = sim_dat$pred_dat),
-                  size = "Population")) +
-    guides(color = guide_legend(override.aes = list(size = 2)))+
-    scale_size_manual(name = "Predictions",
-                      values=c("Subjects" = 0.5, "Population" = 3)) +
-    theme_classic() +
-    theme(plot.title = element_text(size = txt, face = "bold"),
-          text = element_text(size=txt))+
-    thm1
+    # plot the LMEM predictions
+    p5 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) + geom_point() +
+        labs(y = "response") + geom_line(aes(y = predict(sim_dat$fit_lme), group = subject,
+        size = "Subjects")) + geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme,
+        level = 0, newdata = sim_dat$pred_dat), size = "Population")) + guides(color = guide_legend(override.aes = list(size = 2))) +
+        scale_size_manual(name = "Predictions", values = c(Subjects = 0.5, Population = 3)) +
+        theme_classic() + theme(plot.title = element_text(size = txt, face = "bold"),
+        text = element_text(size = txt)) + thm1
 
-  if(option=='simple'){
-    return((p1 + p4 + p5) + plot_layout(nrow = 1) + 
-             plot_annotation(tag_levels = 'A'))
-  }
-  else {
-    return((p1 + p3 + p2 + p4 + p5) + plot_layout(nrow = 1) + 
-             plot_annotation(tag_levels = 'A'))
+    if (option == "simple") {
+        return((p1 + p4 + p5) + plot_layout(nrow = 1) + plot_annotation(tag_levels = "A"))
+    } else {
+        return((p1 + p3 + p2 + p4 + p5) + plot_layout(nrow = 1) + plot_annotation(tag_levels = "A"))
     }
@@
 \begin{lstlisting}[language=R]
-source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
-                  "scripts", "example.R"))
-source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", 
-                  "scripts", "plot_example.R"))
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", "scripts",
+    "example.R"))
+source(here::here("Manuscripts", "Manuscript_by_chapters-SIM_Revisions_final", "scripts",
+    "plot_example.R"))
 
-A1 <- plot_example(example(fun_type = "linear", error_type = "correlated"),
-                 option = 'composite')
+A1 <- plot_example(example(fun_type = "linear", error_type = "correlated"), option = "composite")
 
-B1 <- plot_example(example(fun_type = "linear", error_type = "independent"),
-                 option = 'composite')
+B1 <- plot_example(example(fun_type = "linear", error_type = "independent"), option = "composite")
 
-C1 <- plot_example(example(fun_type = "quadratic", error_type = "correlated"), 
-                 option = 'composite')
+C1 <- plot_example(example(fun_type = "quadratic", error_type = "correlated"), option = "composite")
 
-D1 <- plot_example(example(fun_type = "quadratic", error_type = "independent"), 
-                 option = 'composite') 
+D1 <- plot_example(example(fun_type = "quadratic", error_type = "independent"), option = "composite")
 \end{lstlisting}
@@
 \begin{lstlisting}[language=R]
-#data from Vishwanath et. al.
-dat <- tibble(StO2 = c(4, 27, 3, 2, 0.5, 7, 4, 50, 45, 56),
-              Day = rep(c(0, 2, 5, 7, 10), times = 2),
-              Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
+# data from Vishwanath et. al.
+dat <- tibble(StO2 = c(4, 27, 3, 2, 0.5, 7, 4, 50, 45, 56), Day = rep(c(0, 2, 5,
+    7, 10), times = 2), Group = as.factor(rep(c("Control", "Treatment"), each = 5)))
 \end{lstlisting}
@@
 \begin{lstlisting}[language=R]
-#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
-#Because physiologically StO2 cannot go below 0%, data is
-#generated with a cutoff value of 0.0001 (the "StO2_sim")
+# This function simulates data for the tumor data using default parameters of
+# 10 observations per time point,and Standard deviation (sd) of 5%. Because
+# physiologically StO2 cannot go below 0%, data is generated with a cutoff
+# value of 0.0001 (the 'StO2_sim')
 
@@
         group_by(Group, Day) %>%
-    mutate(
-      StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
-      subject = rep(1:10),
-      subject = factor(paste(subject, Group, sep = "-"))
-    ) %>%
+        mutate(StO2_sim = pmax(rnorm(n, StO2, sd), 1e-04), subject = rep(1:10), subject = factor(paste(subject,
+            Group, sep = "-"))) %>%
         ungroup()
@@
 \begin{lstlisting}[language=R]
-#the same model used for the full dataset
-mod_m1 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5), 
-              data  = dat_missing)
-#appraise the model
+# the same model used for the full dataset
+mod_m1 <- gam(StO2_sim ~ Group + s(Day, by = Group, k = 5), data = dat_missing)
+# appraise the model
 appraise(mod_m1)
 
-ci <- confint(mod_m1, parm = "s(Day)", partial_match = TRUE,
-              type = "confidence")
+ci <- confint(mod_m1, parm = "s(Day)", partial_match = TRUE, type = "confidence")
 ## simultaneous interval
-si <- confint(mod_m1, parm = "s(Day)", type = "simultaneous",
-              partial_match = TRUE)
+si <- confint(mod_m1, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)
 
@@
 ci <- ci %>%
-  mutate(est = case_when(Group == "Treatment" ~ est + const,
-                         TRUE ~ est),
-         lower = case_when(Group == "Treatment" ~ lower + const,
-                           TRUE ~ lower),
-         upper = case_when(Group == "Treatment" ~ upper + const,
-                           TRUE ~ upper))
+    mutate(est = case_when(Group == "Treatment" ~ est + const, TRUE ~ est), lower = case_when(Group ==
+        "Treatment" ~ lower + const, TRUE ~ lower), upper = case_when(Group == "Treatment" ~
+        upper + const, TRUE ~ upper))
 
 si <- si %>%
-  mutate(est = case_when(Group == "Treatment" ~ est + const,
-                         TRUE ~ est),
-         lower = case_when(Group == "Treatment" ~ lower + const,
-                           TRUE ~ lower),
-         upper = case_when(Group == "Treatment" ~ upper + const,
-                           TRUE ~ upper))
+    mutate(est = case_when(Group == "Treatment" ~ est + const, TRUE ~ est), lower = case_when(Group ==
+        "Treatment" ~ lower + const, TRUE ~ lower), upper = case_when(Group == "Treatment" ~
+        upper + const, TRUE ~ upper))
 
-f6 <- ggplot(ci, aes(x = Day, y = est, group = smooth)) +
-  geom_line(lwd = 1) +
-  geom_ribbon(data = ci, 
-              aes(ymin = lower, ymax = upper, x = Day, 
-                  group = smooth, fill = Group),
-              inherit.aes = FALSE, alpha = 0.7, show.legend = FALSE) +
-  geom_ribbon(data = si,
-              aes(ymin = lower, ymax = upper, x = Day, 
-                  group = smooth, fill =Group),
-              inherit.aes = FALSE, alpha = 0.4, show.legend = TRUE) +
-  geom_line(data = si, aes(Day, upper, color = Group), 
-            size = 0.8, alpha = 0.7) +
-  geom_point(data = dat_missing, aes(x = Day, y = StO2_sim, color=Group), 
-             size = 1.5, alpha = 0.6, inherit.aes = FALSE) +
-  labs(y = expression(atop(StO[2], '(incomplete observations)'))) +
-  scale_x_continuous(breaks = c(0, 2, 5, 7, 10)) +
-  theme_classic() +
-  theme(axis.text=element_text(size = 22)) +
-  thm1
+f6 <- ggplot(ci, aes(x = Day, y = est, group = smooth)) + geom_line(lwd = 1) + geom_ribbon(data = ci,
+    aes(ymin = lower, ymax = upper, x = Day, group = smooth, fill = Group), inherit.aes = FALSE,
+    alpha = 0.7, show.legend = FALSE) + geom_ribbon(data = si, aes(ymin = lower,
+    ymax = upper, x = Day, group = smooth, fill = Group), inherit.aes = FALSE, alpha = 0.4,
+    show.legend = TRUE) + geom_line(data = si, aes(Day, upper, color = Group), size = 0.8,
+    alpha = 0.7) + geom_point(data = dat_missing, aes(x = Day, y = StO2_sim, color = Group),
+    size = 1.5, alpha = 0.6, inherit.aes = FALSE) + labs(y = expression(atop(StO[2],
+    "(incomplete observations)"))) + scale_x_continuous(breaks = c(0, 2, 5, 7, 10)) +
+    theme_classic() + theme(axis.text = element_text(size = 22)) + thm1
 \end{lstlisting}
@@
 \begin{lstlisting}[language=R]
-#this function calculates the pointwise (by calling pointwise_comparisons.R) CI and the simultaneous CI
-# for a pairwise comparison between two smooths
-difference_smooths <- function(model, smooth, n = 100, ci_level = 0.95, 
-                               newdata = NULL, partial_match = TRUE,
-                               unconditional = FALSE, frequentist = FALSE,
-                               nrep = 10000, include_means = TRUE, ...) {
+# this function calculates the pointwise (by calling pointwise_comparisons.R)
+# CI and the simultaneous CI for a pairwise comparison between two smooths
+difference_smooths <- function(model, smooth, n = 100, ci_level = 0.95, newdata = NULL,
+    partial_match = TRUE, unconditional = FALSE, frequentist = FALSE, nrep = 10000,
+    include_means = TRUE, ...) {
     if (missing(smooth)) {
@@
     # smooths in model
-  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
+    S <- gratia::smooths(model)  # vector of smooth labels - 's(x)'
     # select smooths
-  select <-
-    gratia:::check_user_select_smooths(smooths = S, select = smooth,
-                                       partial_match = partial_match)
+    select <- gratia:::check_user_select_smooths(smooths = S, select = smooth, partial_match = partial_match)
 
@@
     smooths <- gratia::get_smooths_by_id(model, sm_ids)
-  sm_data <- map(sm_ids, gratia:::smooth_data,
-                 model = model, n = n, include_all = TRUE)
+    sm_data <- map(sm_ids, gratia:::smooth_data, model = model, n = n, include_all = TRUE)
     sm_data <- bind_rows(sm_data)
@@
     smooth_var <- gratia:::smooth_variable(smooths[[1L]])
-  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
-                                   stringsAsFactor = FALSE))
+    pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)), stringsAsFactor = FALSE))
     names(pairs) <- paste0("f", 1:2)
@@
     Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
-  V <- gratia:::get_vcov(model, unconditional = unconditional,
-                         frequentist = frequentist)
+    V <- gratia:::get_vcov(model, unconditional = unconditional, frequentist = frequentist)
     coefs <- coef(model)
 
-  out <- pmap(pairs, difference_pointwise, smooth = smooth, 
-              by_var = by_var, smooth_var = smooth_var, data = sm_data,
-              Xp = Xp, V = V, coefs = coefs, nrep = nrep)
+    out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var, smooth_var = smooth_var,
+        data = sm_data, Xp = Xp, V = V, coefs = coefs, nrep = nrep)
     out <- bind_rows(out)
-  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)
+    crit <- qnorm((1 - ci_level)/2, lower.tail = FALSE)
 
-  out <- add_column(out,
-                    lower = out$diff - (crit * out$se),
-                    upper = out$diff + (crit * out$se),
-                    .after = 6L)
+    out <- add_column(out, lower = out$diff - (crit * out$se), upper = out$diff +
+        (crit * out$se), .after = 6L)
     out
@@
 \begin{lstlisting}[language=R]
-#function to obtain values for the shading regions of the pairwise comparison between the smooths
+# function to obtain values for the shading regions of the pairwise comparison
+# between the smooths
 
 pairwise_limits <- function(dataframe) {
-  #extract values where the lower limit of the ribbon is greater than zero
-  #this is the region where the control group effect is greater
+    # extract values where the lower limit of the ribbon is greater than zero
+    # this is the region where the control group effect is greater
 
@@
         select(Day)
-  #get day  initial value
+    # get day initial value
     init1 = v1$Day[[1]]
-  #get day final value
+    # get day final value
     final1 = v1$Day[[nrow(v1)]]
-  #extract values where the value of the upper limit of the ribbon is lower than zero
-  #this corresponds to the region where the treatment group effect is greater
+    # extract values where the value of the upper limit of the ribbon is lower
+    # than zero this corresponds to the region where the treatment group effect
+    # is greater
     v2 <- dataframe %>%
@@
     final2 = v2$Day[[nrow(v2)]]
-  #store values
-  my_list <- list(init1 = init1,
-                  final1 = final1,
-                  init2 = init2,
-                  final2 = final2)
+    # store values
+    my_list <- list(init1 = init1, final1 = final1, init2 = init2, final2 = final2)
     return(my_list)
@@
 \begin{lstlisting}[language=R]
-#Function that uses the source code of `gam.check` to obtain the estimates without the plots. The source can be checked by typing `gam.check` in the console.
-gam_diagnostics <- function (b, old.style = FALSE, 
-                           type = c("deviance", "pearson", "response"), 
-                           k.sample = 5000, k.rep = 200, rep = 0,
-                           level = 0.9, rl.col = 2, rep.col = "gray80", 
-                           ...) {
+# Function that uses the source code of `gam.check` to obtain the estimates
+# without the plots. The source can be checked by typing `gam.check` in the
+# console.
+gam_diagnostics <- function(b, old.style = FALSE, type = c("deviance", "pearson",
+    "response"), k.sample = 5000, k.rep = 200, rep = 0, level = 0.9, rl.col = 2,
+    rep.col = "gray80", ...) {
     type <- match.arg(type)
@@
     linpred <- if (is.matrix(b$linear.predictors) && !is.matrix(resid))
-    napredict(b$na.action, b$linear.predictors[, 1])
-  else napredict(b$na.action, b$linear.predictors)
+        napredict(b$na.action, b$linear.predictors[, 1]) else napredict(b$na.action, b$linear.predictors)
 
     fv <- if (inherits(b$family, "extended.family"))
-    predict(b, type = "response")
-  else fitted(b)
+        predict(b, type = "response") else fitted(b)
     if (is.matrix(fv) && !is.matrix(b$y))
         fv <- fv[, 1]
-  gamm <- !(b$method %in% c("GCV", "GACV", "UBRE",
-                            "REML", "ML", "P-ML", "P-REML",
+    gamm <- !(b$method %in% c("GCV", "GACV", "UBRE", "REML", "ML", "P-ML", "P-REML",
         "fREML"))
@@
         message("\nChecks based on working residuals may be misleading.")
-  }
-  else {
+    } else {
         message("\nMethod:", b$method, "  Optimizer:", b$optimizer)
@@
                 if (boi$iter == 1)
-          message(".")
-        else message("s.")
-        message("\nGradient range [", min(boi$grad), ",", max(boi$grad), "]", sep = "")
+                  message(".") else message("s.")
+                message("\nGradient range [", min(boi$grad), ",", max(boi$grad),
+                  "]", sep = "")
                 message("\n(score ", b$gcv.ubre, " & scale ", b$sig2, ").", sep = "")
@@
                 if (min(ev) > 0)
-          message("\nHessian positive definite, ")
-        else message("\n")
+                  message("\nHessian positive definite, ") else message("\n")
                 message("eigenvalue range [", min(ev), ",", max(ev), "].\n", sep = "")
-      }
-      else {
+            } else {
                 message("\n")
@@
             }
-    }
-    else {
+        } else {
             if (length(b$sp) == 0)
-        message("\nModel required no smoothing parameter selection")
-      else {
-        message("\nSmoothing parameter selection converged after",
-            b$mgcv.conv$iter, "iteration")
+                message("\nModel required no smoothing parameter selection") else {
+                message("\nSmoothing parameter selection converged after", b$mgcv.conv$iter,
+                  "iteration")
                 if (b$mgcv.conv$iter > 1)
@@
                 if (!b$mgcv.conv$fully.converged)
-          message(" by steepest\ndescent step failure.\n")
-        else message(".\n")
+                  message(" by steepest\ndescent step failure.\n") else message(".\n")
                 message("The RMS", b$method, "score gradient at convergence was",
@@
                 if (b$mgcv.conv$hess.pos.def)
-          message("The Hessian was positive definite.\n")
-        else message("The Hessian was not positive definite.\n")
+                  message("The Hessian was positive definite.\n") else message("The Hessian was not positive definite.\n")
             }

### 2022-01-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/03-GAM Theory.Rmd
Updated, made changes

@@
 
-To get the weighed basis functions, each basis (from Figure \@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients in Figure \@ref(fig:basis-plot)B, thereby increasing or decreasing the original basis functions.
-Figure \@ref(fig:basis-plot)C shows the resulting weighed basis functions.
+To get the weighted basis functions, each basis (from Figure \@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients in Figure \@ref(fig:basis-plot)B, thereby increasing or decreasing the original basis functions.
+Figure \@ref(fig:basis-plot)C shows the resulting weighted basis functions.
 Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall contribution to the smoother term (because the coefficient for that basis function is negative and its magnitude is less than one).
@@
 The coefficient determines the overall effect of each basis in the final smoother.
-**C**: Weighed basis functions.
-Each of the four basis functions (and the intercept) of panel A has been weighed by the corresponding coefficient shown in Panel B.
+**C**: Weighted basis functions.
+Each of the four basis functions (and the intercept) of panel A has been weighted by the corresponding coefficient shown in Panel B.
 Note the corresponding increase (or decrease) in magnitude of each weighted basis function.
@@
 Suppose we chose a nominal value (say, 95%) and compute a simultaneous CI; if we obtain 100 repeated samples and compute a simultaneous CI in each case, we would expect that the true function lies completely within the computed simultaneous CI in 95 of those repeated samples.
-Briefly, to obtain a simultaneous CI we use simulation from the empirical Bayesian posterior distribution to obtain the maximum absolute standardized deviation of the model estimates, which is used to correct the coverage of the across the function CI [@ruppert2003].
+Briefly, to obtain a simultaneous CI we simulate 10,000 draws from the empirical Bayesian posterior distribution of the fitted smooths.
+Then, we obtain the maximum absolute standardized deviation of the differences in smooth estimates which is used to correct the coverage of the across the function CI [@ruppert2003] in a similar fashion to how _q-values_ correct _p-values_ to control false positive discovery rates [@storey2003].
 

### 2022-01-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/sections/04-Longitudinal analysis with GAMs.Rmd
Updated, made changes

@@
 At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups.
-Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs can use a different approach to estimate significance.
-In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that the difference between the separate smoothers per group (such as in `gam_02`) can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.
+In linear models (such as rm-ANOVA), if there is a significant _p-value_ after a _post-hoc_ analysis we then can make inference about the effect size using the slope or the intercept from the model.
+In GAMs however, there is no single _p-value_ to determine the significance of an effect as in linear models.
+Therefore, the coefficients of GAMs do not provide a simple interpretation as in the linear model case, but the changes in slope at specific timepoints can be used to determine the instantaneous effect size.
+In essence, the idea behind the estimation of significance in GAMs across different treatment groups is that the difference between the separate smoothers per group (such as in `gam_02`) can be computed pairwise, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.
 
@@
 
-The absence of a _p-value_ in this context might seem odd, but the pairwise difference in smooths can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \@ref(fig:sim-smooth-plot)A, where the chemotherapy causes $\mbox{StO}_2$ to increase over time.
-
+The pairwise difference in smooths can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment as in Figure \@ref(fig:sim-smooth-plot)A, where the chemotherapy causes $\mbox{StO}_2$ to increase over time.
 With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant.

### 2022-01-19 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/scripts/basis_functions.R

@@ -123,7 +123,7 @@ p12<-ggplot(basis_plot,
               size=sz)+
     geom_point(show.legend = FALSE,
                size=sz+1)+
-    labs(y='Penalized \n basis functions')+
+    labs(y='Weighted \n basis functions')+
     scale_y_continuous(breaks=seq(-1,1,1))+
     facet_wrap(~Basis,
                labeller=as_labeller(basis_names)

### 2022-01-20 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.Rmd
Made final fixes to RtR document, ready for resubmission.

@@
 - We have removed the sentence "if the difference between the credible intervals of the fitted smooths...
-is non-zero", and now L487-L489 now read: "is that the difference between them can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference."
+is non-zero", and now L491-L493 now read: "is that the difference between them can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference."
 
@@
 
-L283 I think it is better to say “…the mathematical space within which the true but unknown $f(x_t)$ is thought to exist.”
+L283 I think it is better to say “…the mathematical space within which the true but unknown $f(time_t)$ is thought to exist.”
 
@@
 We use the code notation provided by the $LaTeX$ package `listings`, and have updated the quotes to be straight in all code chunks.
-Regarding `family`, we believe that it is not necessary to include it here, but we indicate in L459-460 what the argument exists and that other distributions can be used.
+Regarding `family`, we believe that it is not necessary to include it here, but we indicate in L460-L461 what the argument exists and that other distributions can be used.
 

### 2022-01-20 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.tex
Made final fixes to RtR document, ready for resubmission.

@@
 ]{article}
+\title{\textbf{Generalized additive models to analyze non-linear trends in biomedical longitudinal data using R:}\\
+Beyond repeated measures ANOVA and Linear Mixed Models\\
+\strut \\
+\textbf{Response to Reviewer's Comments}\\
+Journal of Submission: Statistics in Medicine}
+\usepackage{etoolbox}
+\makeatletter
+\providecommand{\subtitle}[1]{% add subtitle to \maketitle
+  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
+}
+\makeatother
+\subtitle{Manuscript ID: SIM-21-0640.R1}
+\author{}
+\date{\vspace{-2.5em}}
+
 \usepackage{amsmath,amssymb}
@@
 
-\title{\textbf{Generalized additive models to analyze non-linear trends in biomedical longitudinal data using R:}\\
-Beyond repeated measures ANOVA and Linear Mixed Models\\
-\strut \\
-\textbf{Response to Reviewer's Comments}\\
-Journal of Submission: Statistics in Medicine}
-\usepackage{etoolbox}
-\makeatletter
-\providecommand{\subtitle}[1]{% add subtitle to \maketitle
-  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
-}
-\makeatother
-\subtitle{Manuscript ID: SIM-21-0640.R1}
-\author{}
-\date{\vspace{-2.5em}}
-
 \begin{document}
@@
 \item
-  We have removed the sentence ``if the difference between the credible intervals of the fitted smooths\ldots{} is non-zero'', and now L487-L489 now read: ``is that the difference between them can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.''
+  We have removed the sentence ``if the difference between the credible intervals of the fitted smooths\ldots{} is non-zero'', and now L491-L493 now read: ``is that the difference between them can be computed, followed by the estimation of an empirical Bayesian simultaneous CI around this difference.''
 \item
@@
 
-L283 I think it is better to say ``\ldots the mathematical space within which the true but unknown \(f(x_t)\) is thought to exist.''
+L283 I think it is better to say ``\ldots the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist.''
 
@@
 We use the code notation provided by the \(LaTeX\) package \passthrough{\lstinline!listings!}, and have updated the quotes to be straight in all code chunks.
-Regarding \passthrough{\lstinline!family!}, we believe that it is not necessary to include it here, but we indicate in L459-460 what the argument exists and that other distributions can be used.
+Regarding \passthrough{\lstinline!family!}, we believe that it is not necessary to include it here, but we indicate in L460-L461 what the argument exists and that other distributions can be used.
 

### 2022-01-20 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/response_to_reviewers/Response_to_reviewers.tex
Updated date

@@
 To this end, we have addressed all critiques.
-We hope these revisions, submitted on January 4th, 2022, improve the manuscript so it is deemed worthy of publication in \emph{Statistics in Medicine}.
+We hope these revisions, submitted on January 20th, 2022, improve the manuscript so it is deemed worthy of publication in \emph{Statistics in Medicine}.
 Following are our detailed responses to reviewer comments.
